### 2021-10-30 manuscript/manuscript.tex
restructuring

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  english,
  man]{apa6}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={The title},
  pdfauthor={First Author1 \& Ernst-August Doelle1,2},
  pdflang={en-EN},
  pdfkeywords={keywords},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
% Manuscript styling
\usepackage{upgreek}
\captionsetup{font=singlespacing,justification=justified}

% Table formatting
\usepackage{longtable}
\usepackage{lscape}
% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
\usepackage{multirow}		% Table styling
\usepackage{tabularx}		% Control Column width
\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
\usepackage{threeparttablex}            % Lets threeparttable work with longtable

% Create new environments so endfloat can handle them
% \newenvironment{ltable}
%   {\begin{landscape}\centering\begin{threeparttable}}
%   {\end{threeparttable}\end{landscape}}
\newenvironment{lltable}{\begin{landscape}\centering\begin{ThreePartTable}}{\end{ThreePartTable}\end{landscape}}

% Enables adjusting longtable caption width to table width
% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
\makeatletter
\newcommand\LastLTentrywidth{1em}
\newlength\longtablewidth
\setlength{\longtablewidth}{1in}
\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}

% \setlength{\parindent}{0.5in}
% \setlength{\parskip}{0pt plus 0pt minus 0pt}

% \usepackage{etoolbox}
\makeatletter
\patchcmd{\HyOrg@maketitle}
  {\section{\normalfont\normalsize\abstractname}}
  {\section*{\normalfont\normalsize\abstractname}}
  {}{\typeout{Failed to patch abstract.}}
\patchcmd{\HyOrg@maketitle}
  {\section{\protect\normalfont{\@title}}}
  {\section*{\protect\normalfont{\@title}}}
  {}{\typeout{Failed to patch title.}}
\makeatother
\shorttitle{Title}
\keywords{keywords\newline\indent Word count: X}
\DeclareDelayedFloatFlavor{ThreePartTable}{table}
\DeclareDelayedFloatFlavor{lltable}{table}
\DeclareDelayedFloatFlavor*{longtable}{table}
\makeatletter
\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother
\usepackage{lineno}

\linenumbers
\usepackage{csquotes}
\ifxetex
  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
  \usepackage{polyglossia}
  \setmainlanguage[]{english}
\else
  \usepackage[shorthands=off,main=english]{babel}
\fi

\title{The title}
\author{First Author\textsuperscript{1} \& Ernst-August Doelle\textsuperscript{1,2}}
\date{}

\authornote{

Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.

Enter author note here.

The authors made the following contributions. First Author: Conceptualization, Writing - Original Draft Preparation, Writing - Review \& Editing; Ernst-August Doelle: Writing - Review \& Editing.

Correspondence concerning this article should be addressed to First Author, Postal address. E-mail: \href{mailto:my@email.com}{\nolinkurl{my@email.com}}

}

\affiliation{\vspace{0.5cm}\textsuperscript{1} Wilhelm-Wundt-University\\\textsuperscript{2} Konstanz Business School}

\abstract{
One or two sentences providing a \textbf{basic introduction} to the field, comprehensible to a scientist in any discipline.

Two to three sentences of \textbf{more detailed background}, comprehensible to scientists in related disciplines.

One sentence clearly stating the \textbf{general problem} being addressed by this particular study.

One sentence summarizing the main result (with the words ``\textbf{here we show}'' or their equivalent).

Two or three sentences explaining what the \textbf{main result} reveals in direct comparison to what was thought to be the case previously, or how the main result adds to previous knowledge.

One or two sentences to put the results into a more \textbf{general context}.

Two or three sentences to provide a \textbf{broader perspective}, readily comprehensible to a scientist in any discipline.
}

\begin{document}
\maketitle

\hypertarget{methods}{%
\section{Methods}\label{methods}}

We report how we determined our sample size, all data exclusions (if any), all manipulations, and all measures in the study.

\hypertarget{participants}{%
\subsection{Participants}\label{participants}}

\hypertarget{material}{%
\subsection{Material}\label{material}}

\hypertarget{procedure}{%
\subsection{Procedure}\label{procedure}}

\hypertarget{data-analysis}{%
\subsection{Data analysis}\label{data-analysis}}

We used R (Version 4.1.0; R Core Team, 2021) and the R-package \emph{papaja} (Version 0.1.0.9997; Aust \& Barth, 2020) for all our analyses.

\hypertarget{results}{%
\section{Results}\label{results}}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

\newpage

\hypertarget{references}{%
\section{References}\label{references}}

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-R-papaja}{}%
Aust, F., \& Barth, M. (2020). \emph{papaja: Create APA manuscripts with R Markdown}. Retrieved from \url{https://github.com/crsh/papaja}

\leavevmode\hypertarget{ref-R-base}{}%
R Core Team. (2021). \emph{R: A language and environment for statistical computing}. Vienna, Austria: R Foundation for Statistical Computing. Retrieved from \url{https://www.R-project.org/}

\endgroup

\end{document}

### 2021-10-30 R/fringilidae.R

@@ -1,0 +1,31 @@
+library(datelife)
+opentree_chronograms = utils::data("opentree_chronograms", package = "datelife"),
+dq = datelife::make_datelife_query(input = "fringillidae", get_spp_from_taxon = TRUE),
+dr = datelife::get_datelife_result(input = dq, cache = opentree_chronograms),
+summ = datelife::get_taxon_summary(datelife_query = dq, datelife_result = dr),
+phyloall = summarize_datelife_result(datelife_query = dq,
+  datelife_result = dr, summary_format = "phylo_all",
+  taxon_summary = "none"),
+cb_cols1 = c("#0072B2", "#D55E00", "#440154FF", "#CC79A7", "#440154FF", "#009E73",
+  "#9ad0f3", "#95D840FF", "#F0E442"),
+cb_cols2 = c("#440154FF", "#009E73", "#e79f00", "#9ad0f3", "#0072B2", "#D55E00",
+  "#CC79A7", "#F0E442", "#95D840FF"),
+ltt_plot = plot_ltt_phyloall(taxon = "Fringillidae", phy = phyloall, ltt_colors = cb_cols1, tax_datedotol = NULL,
+    file_name = "ltt_plot_poster.pdf", file_dir = getwd(), height = 3, width = 7.9, add_legend = FALSE, add_title = FALSE),
+datedotol = get_dated_otol_induced_subtree(input = dq, ott_ids = dq$ott_ids),
+bestgrove = get_best_grove(datelife_result = dr),
+sdm_matrix = get_sdm_matrix(datelife_result = bestgrove$best_grove),
+median_matrix = datelife_result_median_matrix(datelife_result = bestgrove$best_grove),
+otol = get_otol_synthetic_tree(input = dq),
+phy_sdm = summary_matrix_to_phylo_all(summ_matrix = sdm_matrix, target_tree = otol),
+phy_median = summary_matrix_to_phylo_all(summ_matrix = median_matrix, target_tree = otol),
+ltt_summ = plot_ltt_summary(taxon = "Fringillidae", phy = phyloall, phy_sdm, phy_median,
+      file_name = "ltt_summ_poster.pdf", file_dir = "~//Desktop//datelife_paper1//", height = 3, width = 7.9,
+      add_legend = TRUE),
+each_calibrations = get_all_calibrations(phyloall, each = TRUE)
+# , # inactivating the following because having trouble with getting otol for some reason
[…]

### 2021-10-30 data-raw/fringilidae.R

@@ -1,0 +1,82 @@
+## code to prepare paper datasets goes here
+
+utils::data("opentree_chronograms", package = "datelife")
+
+dq = datelife::make_datelife_query(input = "fringillidae", get_spp_from_taxon = TRUE)
+usethis::use_data(dq, overwrite = TRUE)
+
+dr = datelife::get_datelife_result(input = dq, cache = opentree_chronograms)
+usethis::use_data(dr, overwrite = TRUE)
+
+summ = datelife::get_taxon_summary(datelife_query = dq, datelife_result = dr)
+usethis::use_data(summ, overwrite = TRUE)
+
+phyloall = datelife::summarize_datelife_result(datelife_query = dq,
+                                    datelife_result = dr, 
+                                    summary_format = "phylo_all",
+                                    taxon_summary = "none")
+usethis::use_data(phyloall, overwrite = TRUE)
+
+
+# LTT plot
+cb_cols1 = c("#0072B2", 
+             "#D55E00", 
+             "#440154FF", 
+             "#CC79A7", 
+             "#440154FF", 
+             "#009E73",
+             "#9ad0f3",
+             "#95D840FF",
+             "#F0E442")
+cb_cols2 = c("#440154FF", 
+             "#009E73", 
+             "#e79f00", 
+             "#9ad0f3", 
+             "#0072B2", 
+             "#D55E00",
+             "#CC79A7", 
+             "#F0E442", 
+             "#95D840FF")
+ltt_plot = datelifeplot::plot_ltt_phyloall(taxon = "Fringillidae", 
+                             phy = phyloall, 
+                             ltt_colors = cb_cols1, 
+                             tax_datedotol = NULL,
+                             file_name = "ltt_plot_poster.pdf", 
+                             file_dir = getwd(), 
+                             height = 3, 
+                             width = 7.9, 
+                             add_legend = FALSE, 
+                             add_title = FALSE)
+usethis::use_data(ltt_plot, overwrite = TRUE)
+
+# datedotol = datelife::get_dated_otol_induced_subtree(input = dq, 
+#                                           ott_ids = dq$ott_ids)
+
+bestgrove = datelife::get_best_grove(datelife_result = dr)
+usethis::use_data(bestgrove, overwrite = TRUE)
+
[…]

### 2021-10-30 manuscript/manuscript.Rmd
abstract

---
title             : "DateLife: Leveraging databases and analytical tools to reveal the dated Tree of Life"
shorttitle        : "Title"

author:
  - name          : "Luna L. Sánchez Reyes"
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : ""
    email         : "sanchez.reyes.luna@gmail.com"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Data curation
      - Investigation
      - Software
      - Visualization
      - Validation
      - Writing - Original Draft Preparation
      - Writing - Review & Editing
  - name          : "Emily Jane McTavish"
    affiliation   : "1"
    role:
      - Resources
      - Software
      - Writing - Review & Editing
  - name          : "Brian O'Meara"
    affiliation   : "2"
    role:
      - Conceptualization
      - Funding acquisition
      - Methodology
      - Resources
      - Software
      - Supervision
      - Writing - Review & Editing

affiliation:
  - id            : "1"
    institution   : "University of California, Merced"
  - id            : "2"
    institution   : "University of Tennessee, Knoxville"

authornote: |
  School of Natural Sciences, University of California, Merced, Science and Engineering Building 1.

  Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 425 Hesler Biology Building, Knoxville, TN 37996, USA.

abstract: |
  Time of evolutionary origin is fundamental for understanding biological processes.

  The combination of new analytical techniques, availability of more fossil and molecular
  data, and efforts to improve data sharing practices in biology has resulted in a steady accumulation
  of time of lineage divergence in public and open databases such as TreeBASE, Dryad, and Open Tree
  of Life for a large quantity and diversity of organisms in the last few decades.

  However, getting a tree with branch lengths proportional to time remains difficult for many biologists and the non-academic community, despite its importance for biological research, medicine, education, and science communication.

  Here we present `datelife`, a service implemented via an R package and a web site
  (<http://www.datelife.org/>) for efficient reuse, summary and reanalysis of expert, peer-reviewed, public data on time of lineage divergence.

  Main results:
    1. blah
    2. blah
    3. blah

  Results in a general context:
  All source and summary chronograms can be saved in formats that permit easy reuse and reanalysis. Summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research in biology.
  How well this trees work for this purpose still needs to be tested.

  `datelife` will be useful to increase awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.

  <!-- https://tinyurl.com/ybremelq -->

keywords          : "keywords"
wordcount         : "X"

bibliography      : ["paper_references.bib"]

floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
---

```{r setup, include = FALSE}
library("papaja")
r_refs("paper_references.bib")
```

```{r analysis-preferences}
# Seed for random number generation
set.seed(50)
knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
```

# Methods
We report how we determined our sample size, all data exclusions (if any), all manipulations, and all measures in the study. <!-- 21-word solution (Simmons, Nelson & Simonsohn, 2012; retrieved from http://ssrn.com/abstract=2160588) -->

## Participants

## Material

## Procedure

## Data analysis
We used `r cite_r("r-references.bib")` for all our analyses.

# Results

# Discussion

\newpage

# References

\begingroup
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}

<div id="refs" custom-style="Bibliography"></div>
\endgroup

### 2021-10-30 manuscript/manuscript.tex
abstract

@@
 \hypersetup{
-  pdftitle={The title},
-  pdfauthor={First Author1 \& Ernst-August Doelle1,2},
+  pdftitle={DateLife: Leveraging databases and analytical tools to reveal the dated Tree of Life},
+  pdfauthor={Luna L. Sánchez Reyes1,2, Emily Jane McTavish1, \& Brian O'Meara2},
   pdflang={en-EN},
@@
 
-\title{The title}
-\author{First Author\textsuperscript{1} \& Ernst-August Doelle\textsuperscript{1,2}}
+\title{DateLife: Leveraging databases and analytical tools to reveal the dated Tree of Life}
+\author{Luna L. Sánchez Reyes\textsuperscript{1,2}, Emily Jane McTavish\textsuperscript{1}, \& Brian O'Meara\textsuperscript{2}}
 \date{}
@@
 
-Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.
+School of Natural Sciences, University of California, Merced, Science and Engineering Building 1.
 
-Enter author note here.
+Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 425 Hesler Biology Building, Knoxville, TN 37996, USA.
 
-The authors made the following contributions. First Author: Conceptualization, Writing - Original Draft Preparation, Writing - Review \& Editing; Ernst-August Doelle: Writing - Review \& Editing.
+The authors made the following contributions.
+Luna L.
+Sánchez Reyes: Data curation, Investigation, Software, Visualization, Validation, Writing - Original Draft Preparation, Writing - Review \& Editing; Emily Jane McTavish: Resources, Software, Writing - Review \& Editing; Brian O'Meara: Conceptualization, Funding acquisition, Methodology, Resources, Software, Supervision, Writing - Review \& Editing.
 
-Correspondence concerning this article should be addressed to First Author, Postal address. E-mail: \href{mailto:my@email.com}{\nolinkurl{my@email.com}}
+Correspondence concerning this article should be addressed to Luna L. Sánchez Reyes, . E-mail: \href{mailto:sanchez.reyes.luna@gmail.com}{\nolinkurl{sanchez.reyes.luna@gmail.com}}
 
@@
 
-\affiliation{\vspace{0.5cm}\textsuperscript{1} Wilhelm-Wundt-University\\\textsuperscript{2} Konstanz Business School}
+\affiliation{\vspace{0.5cm}\textsuperscript{1} University of California, Merced\\\textsuperscript{2} University of Tennessee, Knoxville}
 
@@
 Two or three sentences to provide a \textbf{broader perspective}, readily comprehensible to a scientist in any discipline.
+
+The combination of new analytical techniques, availability of more fossil and molecular
+data, and better practices in data sharing in biology has resulted in a steady accumulation
+of chronograms in public and open databases such as TreeBASE, Dryad, and Open Tree
+of Life for a large quantity and diversity of organisms in the last few decades.
+However, getting a tree with branch lengths proportional to time remains difficult for many biologists and the non-academic community, despite its importance in many areas of research, education, and science communication.
+\texttt{datelife} is a service implemented via an R package and a web site
+(\url{http://www.datelife.org/}) for efficient reuse, summary and reanalysis of published data on lineage divergence times.
+The main workflow starts with at least two taxon names as input, either as tip labels on a tree, or as a simple comma separated character string.
+A name search is then performed across the chronogram database and positively identified source trees are pruned to maintain queried taxa only and stored as a named list of patristic distance matrices.
+Source chronogram data can be summarised using branch length summary statistics or variance minimizing approaches to generate a single summary chronogram.
+Source chronogram data can also be used as calibration points to date a tree containing some or all names from the initial query.
+If there is no information available for any queried taxa, data can be simulated.
+All source and summary chronograms can be saved in formats that permit easy reuse and reanalysis.
+Summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research in biology.
+How well this trees work for this purpose still needs to be tested.
+\texttt{datelife} will be useful to increase awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
 }

### 2021-10-30 R/fringilidae.R

@@ -1,31 +1,12 @@
-library(datelife)
-opentree_chronograms = utils::data("opentree_chronograms", package = "datelife"),
-dq = datelife::make_datelife_query(input = "fringillidae", get_spp_from_taxon = TRUE),
-dr = datelife::get_datelife_result(input = dq, cache = opentree_chronograms),
-summ = datelife::get_taxon_summary(datelife_query = dq, datelife_result = dr),
-phyloall = summarize_datelife_result(datelife_query = dq,
-  datelife_result = dr, summary_format = "phylo_all",
-  taxon_summary = "none"),
-cb_cols1 = c("#0072B2", "#D55E00", "#440154FF", "#CC79A7", "#440154FF", "#009E73",
-  "#9ad0f3", "#95D840FF", "#F0E442"),
-cb_cols2 = c("#440154FF", "#009E73", "#e79f00", "#9ad0f3", "#0072B2", "#D55E00",
-  "#CC79A7", "#F0E442", "#95D840FF"),
-ltt_plot = plot_ltt_phyloall(taxon = "Fringillidae", phy = phyloall, ltt_colors = cb_cols1, tax_datedotol = NULL,
-    file_name = "ltt_plot_poster.pdf", file_dir = getwd(), height = 3, width = 7.9, add_legend = FALSE, add_title = FALSE),
-datedotol = get_dated_otol_induced_subtree(input = dq, ott_ids = dq$ott_ids),
-bestgrove = get_best_grove(datelife_result = dr),
-sdm_matrix = get_sdm_matrix(datelife_result = bestgrove$best_grove),
-median_matrix = datelife_result_median_matrix(datelife_result = bestgrove$best_grove),
-otol = get_otol_synthetic_tree(input = dq),
-phy_sdm = summary_matrix_to_phylo_all(summ_matrix = sdm_matrix, target_tree = otol),
-phy_median = summary_matrix_to_phylo_all(summ_matrix = median_matrix, target_tree = otol),
-ltt_summ = plot_ltt_summary(taxon = "Fringillidae", phy = phyloall, phy_sdm, phy_median,
-      file_name = "ltt_summ_poster.pdf", file_dir = "~//Desktop//datelife_paper1//", height = 3, width = 7.9,
-      add_legend = TRUE),
-each_calibrations = get_all_calibrations(phyloall, each = TRUE)
-# , # inactivating the following because having trouble with getting otol for some reason
[…]

### 2021-10-30 data-raw/fringilidae.R

@@ -2,7 +2,7 @@
 
 utils::data("opentree_chronograms", package = "datelife")
 
-dq = datelife::make_datelife_query(input = "fringillidae", get_spp_from_taxon = TRUE)
+dq = datelife::make_datelife_query(input = "fringilidae", get_spp_from_taxon = TRUE)
 usethis::use_data(dq, overwrite = TRUE)
 
 dr = datelife::get_datelife_result(input = dq, cache = opentree_chronograms)
@@ -12,44 +12,44 @@ summ = datelife::get_taxon_summary(datelife_query = dq, datelife_result = dr)
 usethis::use_data(summ, overwrite = TRUE)
 
 phyloall = datelife::summarize_datelife_result(datelife_query = dq,
-                                    datelife_result = dr, 
+                                    datelife_result = dr,
                                     summary_format = "phylo_all",
                                     taxon_summary = "none")
 usethis::use_data(phyloall, overwrite = TRUE)
 
 
 # LTT plot
-cb_cols1 = c("#0072B2", 
-             "#D55E00", 
-             "#440154FF", 
-             "#CC79A7", 
-             "#440154FF", 
+cb_cols1 = c("#0072B2",
+             "#D55E00",
+             "#440154FF",
+             "#CC79A7",
+             "#440154FF",
              "#009E73",
              "#9ad0f3",
              "#95D840FF",
              "#F0E442")
-cb_cols2 = c("#440154FF", 
-             "#009E73", 
-             "#e79f00", 
-             "#9ad0f3", 
-             "#0072B2", 
+cb_cols2 = c("#440154FF",
+             "#009E73",
+             "#e79f00",
+             "#9ad0f3",
+             "#0072B2",
              "#D55E00",
-             "#CC79A7", 
-             "#F0E442", 
+             "#CC79A7",
+             "#F0E442",
              "#95D840FF")
-ltt_plot = datelifeplot::plot_ltt_phyloall(taxon = "Fringillidae", 
-                             phy = phyloall, 
-                             ltt_colors = cb_cols1, 
+ltt_plot = datelifeplot::plot_ltt_phyloall(taxon = "Fringillidae",
+                             phy = phyloall,
+                             ltt_colors = cb_cols1,
                              tax_datedotol = NULL,
-                             file_name = "ltt_plot_poster.pdf", 
-                             file_dir = getwd(), 
-                             height = 3, 
[…]

### 2021-10-30 manuscript/manuscript.Rmd
sections

@@
 ---
-title             : "DateLife: Leveraging databases and analytical tools to reveal the dated Tree of Life"
-shorttitle        : "Title"
+title             : "DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life"
+shorttitle        : "DateLife: revealing the dated Tree of Life"
 
@@
 
-keywords          : "keywords"
-wordcount         : "X"
+keywords          : "Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations"
+wordcount         : "55"
 
@@
 output            : papaja::apa6_pdf
+
+nocite: |
+  @barker2012going, @barker2015new, @burns2014phylogenetics, @claramunt2015new, @gibb2015new, @Hedges2015, @hooper2017chromosomal, @Jetz2012, @price2014niche
 ---
@@
 knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
+data(opentree_chronograms)
+rpackage_available <- FALSE
+availability <- ifelse(rpackage_available, "is available", "will be available")
 ```
 
-# Methods
+# Introduction
+
+# Description
 We report how we determined our sample size, all data exclusions (if any), all manipulations, and all measures in the study.
@@
 
-## Participants
+# Benchmark
 
-## Material
-
-## Procedure
-
-## Data analysis
+# Example
 We used `r cite_r("r-references.bib")` for all our analyses.
 
-# Results
+# Discussion
 
-# Discussion
+# Conclusions
+
+# Availability
+
+`datelife` is free and open source and it can be used through its current website
+<http://www.datelife.org/query/>, through its R package, and through Phylotastic's project web portal <http://phylo.cs.nmsu.edu:3000/>.
+`datelife`'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
+`datelife`'s R package stable version `r availability`
+for installation from the CRAN repository (<https://cran.r-project.org/package=datelife>)
+using the command `install.packages(pkgs = "datelife")` from within R. Development versions
+are available from the GitHub repository (<https://github.com/phylotastic/datelife>)
+and can be installed using the command `devtools::install_github("phylotastic/datelife")`.
+
+# Supplementary Material
+
+Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities are available at [datelifeMS1](https://github.com/LunaSare/datelifeMS1), [datelife_examples](https://github.com/LunaSare/datelife_examples), and [datelife_benchmark](https://github.com/LunaSare/datelife_benchmark) repositories in LLSR's GitHub account.
+
+# Funding
+
+Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603  to Datelife project and DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), and the Phylotastic project Grant ABI-1458572.
+
+# Acknowledgements
+
+We thank colleagues from the O'Meara Lab at the University
+of Tennesse Knoxville for suggestions, discussions and software testing.
+The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
+that led to initial work on this project.
+The team that assembled `datelife`'s first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
+The Open Tree of Life project that provides the open, metadata rich repository of
+trees used for `datelife`.
+The many scientists who publish their chronograms in an open, reusable form, and
+the scientists who curate them for deposition in the Open Tree of Life repository.
+The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.
 

### 2021-10-30 manuscript/manuscript.tex
sections

@@
 \hypersetup{
-  pdftitle={DateLife: Leveraging databases and analytical tools to reveal the dated Tree of Life},
+  pdftitle={DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life},
   pdfauthor={Luna L. Sánchez Reyes1,2, Emily Jane McTavish1, \& Brian O'Meara2},
   pdflang={en-EN},
-  pdfkeywords={keywords},
+  pdfkeywords={Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations},
   hidelinks,
@@
 \makeatother
-\shorttitle{Title}
-\keywords{keywords\newline\indent Word count: X}
+\shorttitle{DateLife: revealing the dated Tree of Life}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations\newline\indent Word count: 55}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 
-\title{DateLife: Leveraging databases and analytical tools to reveal the dated Tree of Life}
+\title{DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life}
 \author{Luna L. Sánchez Reyes\textsuperscript{1,2}, Emily Jane McTavish\textsuperscript{1}, \& Brian O'Meara\textsuperscript{2}}
@@
 \abstract{
-One or two sentences providing a \textbf{basic introduction} to the field, comprehensible to a scientist in any discipline.
-
-Two to three sentences of \textbf{more detailed background}, comprehensible to scientists in related disciplines.
-
-One sentence clearly stating the \textbf{general problem} being addressed by this particular study.
-
-One sentence summarizing the main result (with the words ``\textbf{here we show}'' or their equivalent).
-
-Two or three sentences explaining what the \textbf{main result} reveals in direct comparison to what was thought to be the case previously, or how the main result adds to previous knowledge.
-
-One or two sentences to put the results into a more \textbf{general context}.
-
-Two or three sentences to provide a \textbf{broader perspective}, readily comprehensible to a scientist in any discipline.
+Time of evolutionary origin is fundamental for understanding biological processes.
 
 The combination of new analytical techniques, availability of more fossil and molecular
-data, and better practices in data sharing in biology has resulted in a steady accumulation
-of chronograms in public and open databases such as TreeBASE, Dryad, and Open Tree
+data, and efforts to improve data sharing practices in biology has resulted in a steady accumulation
+of time of lineage divergence in public and open databases such as TreeBASE, Dryad, and Open Tree
 of Life for a large quantity and diversity of organisms in the last few decades.
-However, getting a tree with branch lengths proportional to time remains difficult for many biologists and the non-academic community, despite its importance in many areas of research, education, and science communication.
-\texttt{datelife} is a service implemented via an R package and a web site
-(\url{http://www.datelife.org/}) for efficient reuse, summary and reanalysis of published data on lineage divergence times.
-The main workflow starts with at least two taxon names as input, either as tip labels on a tree, or as a simple comma separated character string.
-A name search is then performed across the chronogram database and positively identified source trees are pruned to maintain queried taxa only and stored as a named list of patristic distance matrices.
-Source chronogram data can be summarised using branch length summary statistics or variance minimizing approaches to generate a single summary chronogram.
-Source chronogram data can also be used as calibration points to date a tree containing some or all names from the initial query.
-If there is no information available for any queried taxa, data can be simulated.
+
+However, getting a tree with branch lengths proportional to time remains difficult for many biologists and the non-academic community, despite its importance for biological research, medicine, education, and science communication.
+
+Here we present \texttt{datelife}, a service implemented via an R package and a web site
+(\url{http://www.datelife.org/}) for efficient reuse, summary and reanalysis of expert, peer-reviewed, public data on time of lineage divergence.
+
+Main results:
+1. blah
+2. blah
+3. blah
+
+Results in a general context:
 All source and summary chronograms can be saved in formats that permit easy reuse and reanalysis.
@@
 How well this trees work for this purpose still needs to be tested.
+
 \texttt{datelife} will be useful to increase awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
@@
 
-\hypertarget{methods}{%
-\section{Methods}\label{methods}}
+\begin{verbatim}
+## Warning in data(opentree_chronograms): data set 'opentree_chronograms' not found
+\end{verbatim}
+
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
+
+\hypertarget{description}{%
+\section{Description}\label{description}}
 
@@
 
-\hypertarget{participants}{%
-\subsection{Participants}\label{participants}}
+\hypertarget{benchmark}{%
+\section{Benchmark}\label{benchmark}}
 
-\hypertarget{material}{%
-\subsection{Material}\label{material}}
-
-\hypertarget{procedure}{%
-\subsection{Procedure}\label{procedure}}
-
-\hypertarget{data-analysis}{%
-\subsection{Data analysis}\label{data-analysis}}
+\hypertarget{example}{%
+\section{Example}\label{example}}
 
@@
 
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
 \hypertarget{discussion}{%
 \section{Discussion}\label{discussion}}
+
+\hypertarget{conclusions}{%
+\section{Conclusions}\label{conclusions}}
+
+\hypertarget{availability}{%
+\section{Availability}\label{availability}}
+
+\texttt{datelife} is free and open source and it can be used through its current website
+\url{http://www.datelife.org/query/}, through its R package, and through Phylotastic's project web portal \url{http://phylo.cs.nmsu.edu:3000/}.
+\texttt{datelife}'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
+\texttt{datelife}'s R package stable version will be available
+for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
+using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R. Development versions
+are available from the GitHub repository (\url{https://github.com/phylotastic/datelife})
+and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.
+
+\hypertarget{supplementary-material}{%
+\section{Supplementary Material}\label{supplementary-material}}
+
+Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities are available at \href{https://github.com/LunaSare/datelifeMS1}{datelifeMS1}, \href{https://github.com/LunaSare/datelife_examples}{datelife\_examples}, and \href{https://github.com/LunaSare/datelife_benchmark}{datelife\_benchmark} repositories in LLSR's GitHub account.
+
+\hypertarget{funding}{%
+\section{Funding}\label{funding}}
+
+Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to Datelife project and DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), and the Phylotastic project Grant ABI-1458572.
+
+\hypertarget{acknowledgements}{%
+\section{Acknowledgements}\label{acknowledgements}}
+
+We thank colleagues from the O'Meara Lab at the University
+of Tennesse Knoxville for suggestions, discussions and software testing.
+The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
+that led to initial work on this project.
+The team that assembled \texttt{datelife}'s first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
+The Open Tree of Life project that provides the open, metadata rich repository of
+trees used for \texttt{datelife}.
+The many scientists who publish their chronograms in an open, reusable form, and
+the scientists who curate them for deposition in the Open Tree of Life repository.
+The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.
 
@@
 
+\leavevmode\hypertarget{ref-barker2012going}{}%
+Barker, F.
+K., Burns, K.
+J., Klicka, J., Lanyon, S.
+M., \& Lovette, I.
+J.
+(2012).
+Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds.
+\emph{Systematic Biology}, \emph{62}(2), 298--320.
+
+\leavevmode\hypertarget{ref-barker2015new}{}%
+Barker, F.
+K., Burns, K.
+J., Klicka, J., Lanyon, S.
+M., \& Lovette, I.
+J.
+(2015).
+New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
+\emph{The Auk: Ornithological Advances}, \emph{132}(2), 333--348.
+
+\leavevmode\hypertarget{ref-burns2014phylogenetics}{}%
+Burns, K.
+J., Shultz, A.
+J., Title, P.
+O., Mason, N.
+A., Barker, F.
+K., Klicka, J., \ldots{} Lovette, I.
+J.
+(2014).
+Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds.
+\emph{Molecular Phylogenetics and Evolution}, \emph{75}, 41--77.
+
+\leavevmode\hypertarget{ref-claramunt2015new}{}%
+Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{Science Advances}, \emph{1}(11), e1501005.
+
+\leavevmode\hypertarget{ref-gibb2015new}{}%
+Gibb, G.
+C., England, R., Hartig, G., McLenachan, P.
+A., Taylor Smith, B.
+L., McComish, B.
+J., \ldots{} Penny, D.
+(2015).
+New zealand passerines help clarify the diversification of major songbird lineages during the oligocene.
+\emph{Genome Biology and Evolution}, \emph{7}(11), 2983--2995.
+
+\leavevmode\hypertarget{ref-Hedges2015}{}%
+Hedges, S.
+B., Marin, J., Suleski, M., Paymer, M., \& Kumar, S.
+(2015).
+Tree of life reveals clock-like speciation and diversification.
+\emph{Molecular Biology and Evolution}, \emph{32}(4), 835--845.
+\url{https://doi.org/10.1093/molbev/msv037}
+
+\leavevmode\hypertarget{ref-hooper2017chromosomal}{}%
+Hooper, D. M., \& Price, T. D. (2017). Chromosomal inversion differences correlate with range overlap in passerine birds. \emph{Nature Ecology \& Evolution}, \emph{1}(10), 1526.
+
+\leavevmode\hypertarget{ref-Jetz2012}{}%
+Jetz, W., Thomas, G., Joy, J.
+J.
+B., Hartmann, K., \& Mooers, A.
+(2012).
+The global diversity of birds in space and time.
+\emph{Nature}, \emph{491}(7424), 444--448.
+\url{https://doi.org/10.1038/nature11631}
+
+\leavevmode\hypertarget{ref-price2014niche}{}%
+Price, T.
+D., Hooper, D.
+M., Buchanan, C.
+D., Johansson, U.
+S., Tietze, D.
+T., Alström, P., \ldots{} others.
+(2014).
+Niche filling slows the diversification of himalayan songbirds.
+\emph{Nature}, \emph{509}(7499), 222.
+
 \leavevmode\hypertarget{ref-R-base}{}%

### 2021-10-30 manuscript/paper_3-main.Rmd
sections

---
output:
  pdf_document: default
  html_document: default
---

Clade ages represent a fundamental piece of information for evolutionary
understanding in many areas of research, from developmental to conservation biology [@Felsenstein1985a; @Webb2000], from historical biogeography to species diversification studies [@posadas2006historical; @Morlon2014]. The primary information needed for these time estimates comes from the fossil record. Coupled with phylogenies with branch lengths based on molecular and/or morphological data, the time of divergence of extant and extinct lineages can be reconstructed with molecular dating methods.
The number of studies publishing phylogenies with branch lengths proportional to geological time (hereafter chronograms) have constantly increased in number for the last two decades [@Kumar2017].
Still, generating a chronogram is not an easy task unless you have specialized training: it requires inferring a tree, understanding what fossil data are available and their limits, and where fossils go on the tree. That is why there has been an urge for promoting and facilitating reuse of the vast amount of phylogenetic and time of lineage divergence data that has been generated and made available in publications, for the advantage of research relying on this information [@webb2005phylomatic; @Stoltzfus2013].

Wide interest from the scientific community to make information from phylogenies in general and chronograms in particular available for consultation and reuse has spurred the creation of public platforms with various goals and characteristics. TreeBASE [@piel2009treebase; @vos2012nexml], the Dryad repository (<http://datadryad.org/>), and the Open Tree of Life (OToL) repository [@Hinchliff2015; @mctavish2015phylesystem] are platforms that store and make available published phylogenies and chronograms for easy scientific reuse. Trees in all of them can be queried using automatised web procedures, which permit personalized, large scale searches that are also very fast.
OToL stores trees with branch length information from a wide range of living organisms, implementing a metadata structure that stores the branch length units (i.e., time or relative susbtitution rates). Treebase and Dryad repositories also contain trees from all groups of life, but the former did not store branch length information until recently (and lacks consistent metadata on what any branch lengths stored mean) and Dryad stores many other types of biological data using metadata that does not allow automatic distinction of types of trees and branch length units, impairing the automatised access to time of lineage divergence information.

Besides keeping a repository to easily store and share expert phylogenetic and chronogram knowledge, OToL also has the primary goal of synthesising all trees in their repository to expose to the community a single tree of all life depicting the phylogenetic relationships among known lineages.
All or parts of this synthetic tree can be reused for any purpose. However, it currently only focus on synthesizing tree topology, meaning that it does not expose branch length data of any type.
<!-- let alone time of divergence data from their stored chronograms. -->
The Timetree of Life project focuses on the synthesis of a single chronogram of life [@Hedges2006] and presents a very accessible, attractive interface. However, the thousands of chronograms this NSF-funded project have compiled for synthesis are only publicly available for visual examination in their website or for download as images, but large scale download remains prohibited by their site. The latest version of their synthetic chronogram [@Kumar2017] can be queried only through their website in a non-automatised fashion, and only subsets of it can be reused for analyses with the permission of the authors.
Other platforms such as SuperSmart [@antonelli2017supersmart] and phylogenerator [@pearse2013phylogenerator] are focused in automatised _de novo_ chronogram inference, by reusing DNA sequence data to reconstruct phylogenetic trees. However, expert fossil information necessary for subsequent molecular dating analyses still needs to be compiled and curated by the user, rendering them a challenging tool to obtain data on time of lineage divergence for the non-specialist. Additionally, neither Supersmart nor phylogenerator are designed to provide information from already created expert chronograms, since it might not be their objective.

<!-- discussion:
The Angiosperm Phylogeny Website [APW; @stevens2001apw] also specializes in constructing a single synthetic tree, but it focuses in showing phylogenetic relationships between seed plant families only, and it does not store or make available the source trees.

Platforms dedicated to store data on fossils and fossil calibrations: PBDB and The fossil calibration database
can provide data on time of lineage divergence, but it is often not in a phylogenetic context.

To our knowledge, there is no system allowing the comparison of available source information (apw does this a little bit, but it is a verbal comparison more than a graphical or visual comparison). Other types of technical comparisons: which trees are better (which trees represent reality better), but this is of course a very difficult issue.
 -->
A tool for efficient reuse of expert, published data on time of lineage divergence should have an open and fully public chronogram database storing data in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram information as needed by the user.
A prototype service aiming to meet this characteristics was developed over a series of hackathons at the National Evolutionary Synthesis Center [@Stoltzfus2013].
Here we present the formal description and implementation of the `datelife` service, constituted by an R package and a web site (<http://www.datelife.org/>).
<!-- There is still much room for improvement, and flaws and limitations are addressed below.  -->
The current implementation of `datelife` performs the basic tasks described above. It features a system for maintenance of an open database of chronograms pulled from public repositories, methods to summarize and compare source chronograms, and new functions to visualize and graphically compare source and summary chronograms.
<!-- , for the benefit of the broader community. -->
<!-- as well as tools for comparison of source and summary trees.  NOT YET u_u-->
<!-- R packages for benchmarking of functionalities and demonstrating services were also developed and made available.>

<!--(BOLD and OToL species names, are they standardized with each other? I think yes-->

\begin{center}
\textsc{Description}
\end{center}

The basic `datelife` workflow is shown in figure \ref{fig:workflow} and consists of:

\begin{enumerate}
\item A user providing at least two taxon names as input. Taxon names can be given at the species level or above. The user can provide them as a simple comma separated character string, or as tip labels on a tree. The tree can be provided in newick format (as a character string) or as a "phylo" R object, and can be with or without branch lengths.
\item A name search is then performed across the chronogram database. Source trees with at least two matching input names are selected, and taxa that do not match the original query are dropped from these source trees.
<!-- The resulting pruned source chronograms are hereafter referred as source chronograms.  --> <!--# This was generating some confussion, so I'm calling them pruned source chronograms now-->
Finally, each pruned source chronogram is transformed to a patristic distance matrix. This format facilitates and greatly speeds up all further analyses and summarizing algorithms. For the sake of keeping track of authorship later, the matrices are associated to the citation of the original study.
\item  At this point, the user can already obtain different summary information from their query, allowing the user to take informed decisions for the next steps of the workflow. Types of summary information provided are: a) all pruned source chronograms, b) age of the mrca (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of succesful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most taxa matched.
\item  Then, source chronogram data can be used as calibration points to date a tree with or without branch lengths containing some or all names from the initial query. %<!--, a taxonomic tree-->
\item  If there is no information available for any queried taxa, users can also create both age and phylogenetic data for this missing taxa with a variety of algorithms described below.
\item  Finally, users can easily save all source and summary chronograms in formats that permit easy reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
\end{enumerate}

<!--To perform each task:-->
To gather, process, and present information, `datelife` builds up from functions
available in several R packages including rotl [@Michonneau2016], ape [@Paradis2004],
geiger [@Harmon2008], paleotree [@Bapst2012a], bold [@Chamberlain2018], phytools [@Revell2012], taxize [@Chamberlain2013; @Chamberlain2018], phyloch [@Heibl2008], phylocomr [@Ooms2018] and rphylotastic [@Omeara2019].

A `datelife` search currently accepts scientific names only. It can be any named clade or binomial specific. <!--common names (implemented in phylotastic)-->
The search is performed at the species level, so when input names correspond
to named clades, `datelife` pulls all accepted species names within the
clade from OToL's reference taxonomy [@rees2017automated] and uses all of those.
Searches at the infraspecies level are not currently allowed, so input names belonging to subspecies or any other infraspecific category are collapsed to the species level.
`datelife` processes input names with the taxon name resolution service [TNRS; @Boyle2013],
<!--which "corrects misspelled names and authorities, standardizes variant spellings,
and converts nomenclatural synonyms to accepted names" (Boyle et al. 2013)-->
which corrects potentially misspelled names and typos, and standardizes spelling
variations and synonyms <!--to a taxonomic reference of users choosing, default
to otol taxonomy -->, increasing the probability of correctly finding the
queried taxa in `datelife`'s chronogram database.

Then, a name matching procedure is performed across `datelife`'s chronogram database which is assembled from OToL's tree repository [known as the phylesystem; @mctavish2015phylesystem]. Compared to other existing open tree repositories OToL's metadata rich tree store is the only one that supports search, identification, and handling of chronograms in an automatised fashion. Also, the tip names in their stored trees correspond to scientific names at or below the species level. Finally, all their chronograms come from peer-reviewed published studies generated by specialists in the targeted lineages, arguably representing expert knowledge on time of lineage divergence.
<!-- The TreeBASE repository stores trees without branch lengths, so chronograms stored there do not provide relevant information on time of lineage divergence. The Dryad repository stores all types of biological data besides chronograms, and their metadata does not allow differentiating between branch lengths units.  -->
<!-- branch length units that might represent other types of time of divergence that are not proportional to absolute (geologic) time.  -->

Information from pruned source chronograms can then be summarised with simple summary statistics of tree branch length, such as the median or the mean. A  much slower, Super Distance Matrix (SDM) approach for supertree reconstruction with branch lengths [@Criscuolo2006] is also implemented via the ape package [@Paradis2004].<!-- and the hierarchical clustering method of ttol??-->
<!-- describe what SDM does and why clustering does not work here -->
The resulting summary patristic distance matrix could be clustered with classic algorithms to return a tree. However, the resulting trees are often non-ultrameric and do not reflect the source chronogram data (see datelife_examples package). Instead, we obtained a distribution of age data from the summary matrix available for nodes on a consensus tree. The Branch Length Adjuster (BLADJ) algorithm [@Webb2008] was then used to fix known node ages and then distribute nodes with unknown ages evenly over the consensus tree, minimizing age variance in the resulting chronogram.

For tree dating, the congruification algorithm described by @Eastman2013
is implemented to find shared nodes between trees (congruent nodes). The ages of these nodes are then used as calibrations to date any given tree. Currently implemented methods for tree dating are BLADJ, MrBayes [@Huelsenbeck2001; @Ronquist2003] and PATHd8 [@Britton2007], a non-clock, rate-smoothing dating method.

<!-- and simulations are performed with various algorithms: -->
<!-- MrBayes can be used when adding taxa at -->
<!-- random, following a reference taxonomy or a topological constraint. It draws ages -->
<!-- from a pure birth model, as implemented by Jetz and collaborators [-@Jetz2012]. -->

<!-- Details on each step are further developed in `datelife`'s R package documentation -->
<!-- `datelife workflow` vignette at (<https://LINK>). -->

\begin{center}
\textsc{Benchmark}
\end{center}
<!--**Benchmark: Testing DateLife computing performance**-->
<!-- Good to test 3 features on a package: performance, speed and scalability-->

`datelife`'s code speed was tested on an Apple iMac
with one 3.4 GHz Intel Core i5 processor. <!--No multiprocessor tests were performed-->
We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
Query processing time increases roughly linearly with number of input taxon names, and
increases considerably if taxonomic name resolution with TNRS [@Boyle2013] is activated.
<!-- Defining tnrs here again, bc we only talk about it two times, so it easy to forget it -->
Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig. \ref{fig:runtime1}).
<!--Speed of other datelife workflows? NO, bc it refkects speed from other software mainly. But I will put a sentence about it...
Summarizing `datelife` results processing times: shows sdm speed, clustering methods speed;
generating new trees: shows bladj speed;
Adding dates processing time: reflects more mrbayes speed than anything;
get_bold_otol_tree running time: reflects bold downloading data speed.-->
<!--Speed with different number of chronograms in database? NO-->
<!--Speed of web interface and of r package in computers with different capacities? NOOOO-->
`datelife`'s code performance was evaluated with a set of unit tests designed and
implemented with the R package testthat [@RCoreTeam2018] that were run both locally
with the devtools package [@RCoreTeam2018], and on a public server --via
GitHub, using the continuous integration tool Travis CI (<https://travis-ci.org>). At
present, unit tests cover more than 50% of `datelife`'s code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->

\begin{center}
\textsc{Example}
\end{center}
<!--**Biological example: Testing DateLife accuracy**-->
<!--Bird (or reptile) chronograms, too long time...
finches is good
nothofagus
Look for all chronograms containing any birds
Or, look for chronograms containing small and old lineages
Determine which clade of birds has the more chronograms (have been dated
more times) and use that as biological example. Fringillidae is a good candidate
Find a clade with at least one chronogram containing all clade's species.
(Penguins look good, but they are giving weird results in SDM)

Remove this chronogram from datelife Results.

Make sdm and median trees and Compare

add taxa with different methods and Compare

Use ltts to compare for now.

think of a test to compare trees, topology- and date-wise

test improvement in synthetic otol tree after using bold sequences. Is it really better resolved? Probably not.
Do not include this analysis in paper, maybe in examples, but i think it is not worthy.
-->
In this section we demonstrate the types of outputs that can be obtained with `datelife`, using as an example the bird family Fringillidae of true finches. We performed a higher-taxon search to obtain data on lineage divergence available in `datelife`'s database for all recognised species within the Fringillidae (`r length(dq$cleaned_names)` species according to the Open Tree of Life taxonomy). There are `r length(dr)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dr)))` different studies (Fig. \ref{fig:schronograms1}).
<!-- In case we want to cite the studies in here, but they are already cited in the figure caption: [Fig. \ref{fig:schronograms1}; @barker2012going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche]. -->
Data from these source chronograms was used to generate two types of summary chronograms, median and SDM. As explained in the "Description" section, data from source chronograms was first summarised into a single distance matrix (using the median and the SDM method respectively) and then the available node ages were used as fixed ages over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig. \ref{fig:summaries}). Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM. With both methods, ages are generally consistent with source ages, but there are some biological examples in which this is not true (see Flaws, Limitations and Prospects section).
<!-- (Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
<!-- and hierarchical clustering method which appears to push ages back -->

Data from source chronograms was also used to date tree topologies with no branch length information and trees with branch lengths in relative substitution rates (Figs. \ref{fig:cvbladj} and \ref{fig:cvbold}). As a form of cross validation, we used tree topologies from each study and calibrated them using information from all other source chronograms. In the absence of branch length data, the ages of internal nodes were approximately recovered in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvbladj}).
Maximum tree ages were only approximately recovered in one case (study 2; Fig. \ref{fig:cvbladj}).
<!-- We need to think abput the root! is it comparable across source chronograms?? not all of them have the same sampling... so probably not the same root and thus not comparable. Address this -->
To exemplify dating a tree with branch lengths provided by the user, we obtained DNA sequence data using the Barcode of Life Data System [BOLD; ratnasingham2007bold] and branch lengths were successfully generated for all source chronograms. However, dating with PATHd8 [@Britton2007] using congruified calibrations, was only successful in
<!-- two studies when expanding calibrations to make them agree (studies 3 and 5), and from  -->
three cases (studies 3, 5, and 9, shown in Fig. \ref{fig:cvbold}). From these, two trees have a different sampling than the original source chronogram, mainly because DNA data for some species is absent from the BOLD. Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
More examples and code used to generate these trees were developes on an open repository that is available for consultation and reuse at <https://github.com/LunaSare/datelife_examples>.

<!--Show code here, in the paper? Or show it only in the examples? Or only in the vignette? Write down what about datelife will be shown in each of these.
Paper should show outputs from web site and package.
Vignette will show the R code used to generate the outputs displayed here and in the website.
datelife examples will explain more on what is done in every step.
-->

\begin{center}
\textsc{Flaws, Limitations and Prospects}
\end{center}

The main goal of `datelife` is to make expert information on time of lineage divergence easily accesible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter. It is a very fast tool that fulfills the quality of openness and does not require any expert biological knowledge from users --besides the names of the organisms they want to work with-- for any of its functionalities. However, it has many flaws. Some of them can be overcome, some of them might represent limitations.

Up to the time this manuscript was written, `datelife`'s chronogram database had `r length(opentree_chronograms$trees)` chronograms, pulled entirely from OToL's tree repository, the only public tree repository from where `datelife` can currently get chronograms to construct its database. This represents `r round(length(opentree_chronograms$trees)*100/3987, digits = 2)`% of the largest existing chronogram database, TimeTree, which has a collection of 3,998 chronograms as of `r format(sys.Date(), format="%B %d %Y")`. Unforutnately, TimeTree's database is not open for scientific reuse nor automatised data mining [@Kumar2017].
In 2015, a synthetic chronogram was constructed from 2,274 chronograms available at the time on the TimeTree database [@Hedges2015]. This is the only synthetic TimeTree chronogram that has been made publicly available and deposited on the OToL repository, and is part of datelife's database now. Hence, the amount of lineages represented in datelife's database is at least as substantial as TimeTree's, ensuring that some information will be available for any given taxon or lineage. Regrettably, this does not ensure that the full state of knowledge of time of divergence of the taxon/lineage will be available.
<!--
Even though datelifes chronograms cover this many species from the tree of life compared to this many covered by TimeTree (database or synthetic chronogram?)-->
Incorporation of more published chronograms into `datelife`'s database is crucial to improve its services. One option to increase our database is the Dryad data repository.  Methods to automatically mine chronograms from Dryad could be designed and implemented. However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined on a second step, by a curator.
<!-- Can we mine journal databases? -->
<!-- At the same time, the prospect is to increase awereness and keep inviting and doing work encouraging scientists to make their data sets available in OToL's database. -->
Consequently, we would like to emphasize on the importance of sharing chronogram data for the benefit of the scientific community as a whole, into repositories that require expert input and manual curation, such as OToL's tree repository [@mctavish2015phylesystem].

Another potential concern comes from summary chronograms. We currently summarize by default all source chronograms that overlap with at least two taxa. Users can subset source data if they have reasons to choose some source chronograms over others. Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely. To our knowledge, there is no objective way to determine if an expert chronogram is better than another. Some criteria that have been put forward are the level of lineage sampling and the number of calibrations used. Scientists usually also favor chronograms constructed using primary calibrations (ages obtained from the fossil or geological record) to ones constructed with secondary calibrations (ages coming from other chronograms). It has been observed with simulations that divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with bayesian inference methods when priors are implemented in similar ways in both analyses [@schenk2016sec]. Yet, there are different ways to use secondary calibrations
<!-- give some examples or alternative ways of using calibrations-->
and that same bias might not be encountered with dating methods that do not require setting priors, i.e., Maximum Likelihood methods such as r8s [@sanderson2003r8s]. Certainly, further studies are required to fully understand the effect of using secondary calibrations on time estimates and downstream anlyses.

Furthermore, even chronograms obtained with primary fossil data can show substantial variation in time estimates between clades, as observed from the comparison of source chronograms in the Fringillidae example.
This observation is often encountered in the literature
<!-- Different source chronograms often show substantial variation in clade ages  -->
(see, for example, the ongoing debate about crown group age of angiosperms [@barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time]. For some studies, especially ones based on branch lengths (e.g., studies of species diversification, timing of evolutionary events, phenotypic trait evolution), using a different chronogram may return different results [@title2016macrophylogenies]. Stitching together these chronograms can create a larger tree that uses information from multiple studies, but the effect of uncertainties and errors here on downstream analyses is still largely unknown.

Summarizing chronograms might also imply summarizing fundamentally distinct evolutionary hypotheses. For example, two different researchers working on the same clade both carefully select and argument their choices of fossil calibrations. Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses teh same one to calibrate outside the clade, the resulting age estimates will probably differ substantially (the placement of calibrations is proved to deeply affect estimated times of lineage divergence). Trying to summarize the resulting chronograms into a single one using simple summary statistics might erase all types of relevant information from the source chronograms. Accordingly, the prevailing view in our research community is that we should favor time of lineage divergence estimates obtained from a single analysis, using fossil data as primary sources of calibrations, and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history [@antonelli2017supersmart]. <!-- Will this be implemented in future `datelife` versions? Probably not-->
However, the exercise of summarizing different chronograms ha sthe potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypothesis. Choosing the elements of the chronograms that we are going tp keep and the ones that we are going to discard is key, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram. <!-- Example -->

<!-- still requires more research. -->
<!-- Often, there is a large discrepancy in time of lineage divergence estimates across studies from expert knowledge, that is well known and documented for different groups of organisms [e.g., angiosperms; @magallon2015metacalibrated].  -->
<!-- Comparison of available age data for a wide range of organisms shown here suggest that this is a widespread phenomenon that requires further attention. -->
<!-- In the case of the example presented here, most chronograms are new, meaning that we expect that data was generated with high quality and the software used should be very advanced. However, this is just an assumption.  -->

Alternatively, one could try to choose the "best" chronogram from a set of possible evolutionary hypotheses. Several characteristics of the data used for dating analyses as well as from the output chronogram itself, could be used to score quality of source chronograms. Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals. To facilitate subsetting of source chronograms following different criteria by the users, this information should be included as metadata manually entered by curators in the future.

<!-- Still, even if all source chronograms have been generated by excellent standards and using similar methods, the evolutionary history they depict might be very different.  -->

In other areas of biological research, such as ecology and conservation biology, it has been shown that at least some data on lineage divergence represents a relevant improvement for testing alternative hypothesis using phylogenetic distance [@Webb2008]. Hence, we integrated into datelife's workflow different ways of creating branch lengths in the absence of starting branch length information for taxa lacking this information (BLADJ option). Making up branch lengths in this or other ways is accepted in scientific publications: @Jetz2012, created a time-calibrated tree of all 9,993 bird species, where 67% had molecular data adn the rest was simulated; @rabosky2018inverse created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a tree of 353,185 seed plants where only 23% had molecular data. Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch imputation methods. There has yet to be a thorough analysis of what can go wrong when one goes beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make sure results are substantially similar between trees fully reconstructed with molecular or other data, and trees that are reconstructed using taxonomy by resolving polytomies at random following a statistical model.

<!-- There is evidence that different chronograms will result in very different macroevolutionary histories inferred from them [@title2016macrophylogenies]. However, it is unknown how using very different age data could affect hypothesis testing in other areas of research and more effort is needed in that directions -->

\begin{center}
\textsc{Conclusions}
\end{center}

Divergence time information is key to many areas of evolutionary studies: trait evolution,
diversification, biogeography, macroecology and more. It is also crucial for science communication and education, but generating chronograms _de novo_ is difficult,
especially for those who want to use phylogenies but who are not systematists, or
do not have the time to acquire and develop the necessary knowledge and data curation skills. Moreover, years of primarily public funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.

<!--Mention again uses shown here and outline potential uses not shown here-->
`datelife` allows easy and fast summarization of publicly available information
on time of lineage divergence. This provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identification of regions that require more research or that have conflicting information.
<!-- It can also summarize public data, or use it to generate new chronograms based on trees provided by the user. -->
Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.  `datelife` helps with awareness of the existing variation in expert time of divergence data, and will foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.

<!-- `datelife` is also part of the reproducibility of science movement. It strives to facilitate usage of public data sets as well as encouraging the making of information publicly available. Acknowledging source information and describing workflows and analyses. -->

<!-- it allows exploration of alternative taxonomic and phylogenetic schemes; -->
<!-- it allows rapid ; -->

\begin{center}
\textsc{Availability}
\end{center}

\begin{center}
\textsc{Supplementary Material}
\end{center}

<!-- Vignettes?
Dryad also?? since it is all in github already, maybe not necessary-->

\begin{center}
\textsc{Funding}
\end{center}

<!-- Any separate funding from the University of Tennessee, Knoxville?? -->

\begin{center}
\textsc{Acknowledgements}
\end{center}

### 2021-10-30 manuscript/manuscript.Rmd
abstract

@@
 abstract: |
-  Time of evolutionary origin is fundamental for understanding biological processes.
+  Time of evolutionary origin is fundamental for research in the natural sciences, as well as for education, science communication and policy.
+  Despite an increased availability of fossil and molecular data, and time-efficient analytical techniques, achieving a high-quality reconstruction of time of evolutionary origin as a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is still a difficult and time-consuming task for a majority of interested parties.
+  Yet, the amount of published chronograms has increased significantly in the past two decades, and a non-negligeable proportion of these data have been steadily accumulating in public, open databases such as TreeBASE and Open Tree
+  of Life, exposing a wealth of expertly-curated and peer-reviewed data on time of evolutionary origin in a programatic and reusable way, for a large quantity and diversity of organisms.
+  This trend results from intensive and localized efforts for improving data sharing practices, as well as incentivizing open science in biology.
+Despite these trends, accessibility for time data is not that good.
+  R has become a widely used element of the biological data analyst toolkit.
+Hence,to improve accessibility of time data we developed an R package that access these data and eases interaction, reanalysis and reuse of it, incorporation of these data into the evolutionary workflow.
 
-  The combination of new analytical techniques, availability of more fossil and molecular
-  data, and efforts to improve data sharing practices in biology has resulted in a steady accumulation
-  of time of lineage divergence in public and open databases such as TreeBASE, Dryad, and Open Tree
-  of Life for a large quantity and diversity of organisms in the last few decades.
-
-  However, getting a tree with branch lengths proportional to time remains difficult for many biologists and the non-academic community, despite its importance for biological research, medicine, education, and science communication.
-
-  Here we present `datelife`, a service implemented via an R package and a web site
-  (<http://www.datelife.org/>) for efficient reuse, summary and reanalysis of expert, peer-reviewed, public data on time of lineage divergence.
+  Here we present `datelife`, a service implemented as an R package and a web site
+  (www.datelife.org) for increased accessibility, efficient reuse, summary and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
 

### 2021-10-30 manuscript/manuscript.tex
abstract

@@
 \abstract{
-Time of evolutionary origin is fundamental for understanding biological processes.
+Time of evolutionary origin is fundamental for research in the natural sciences, as well as for education, science communication and policy.
+Despite an increased availability of fossil and molecular data, and time-efficient analytical techniques, achieving a high-quality reconstruction of time of evolutionary origin as a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is still a difficult and time-consuming task for a majority of interested parties.
+Yet, the amount of published chronograms has increased significantly in the past two decades, and a non-negligeable proportion of these data have been steadily accumulating in public, open databases such as TreeBASE and Open Tree
+of Life, exposing a wealth of expertly-curated and peer-reviewed data on time of evolutionary origin in a programatic and reusable way, for a large quantity and diversity of organisms.
+This trend results from intensive and localized efforts for improving data sharing practices, as well as incentivizing open science in biology.
+Despite these trends, accessibility for time data is not that good.
+R has become a widely used element of the biological data analyst toolkit.
+Hence,to improve accessibility of time data we developed an R package that access these data and eases interaction, reanalysis and reuse of it, incorporation of these data into the evolutionary workflow.
 
-The combination of new analytical techniques, availability of more fossil and molecular
-data, and efforts to improve data sharing practices in biology has resulted in a steady accumulation
-of time of lineage divergence in public and open databases such as TreeBASE, Dryad, and Open Tree
-of Life for a large quantity and diversity of organisms in the last few decades.
-
-However, getting a tree with branch lengths proportional to time remains difficult for many biologists and the non-academic community, despite its importance for biological research, medicine, education, and science communication.
-
-Here we present \texttt{datelife}, a service implemented via an R package and a web site
-(\url{http://www.datelife.org/}) for efficient reuse, summary and reanalysis of expert, peer-reviewed, public data on time of lineage divergence.
+Here we present \texttt{datelife}, a service implemented as an R package and a web site
+(www.datelife.org) for increased accessibility, efficient reuse, summary and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
 

### 2021-10-30 manuscript/manuscript.Rmd
description and benchmark

@@
 keywords          : "Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations"
-wordcount         : "55"
+wordcount         : "571"
 
@@
 # Description
-We report how we determined our sample size, all data exclusions (if any), all manipulations, and all measures in the study.
-<!-- 21-word solution (Simmons, Nelson & Simonsohn, 2012; retrieved from http://ssrn.com/abstract=2160588) -->
+
+The `datelife` workflow builds off from functions in several R packages (rotl [@Michonneau2016], ape [@Paradis2004],
+geiger [@Harmon2008], paleotree [@Bapst2012a], bold [@Chamberlain2018], phytools [@Revell2012], taxize [@Chamberlain2013; @Chamberlain2018], phyloch [@Heibl2008], and phylocomr [@Ooms2018]).
+
+The basic `datelife` workflow is shown in figure \ref{fig:workflow} and consists of:
+
+\begin{enumerate}
+\item A user providing at least two taxon names as input.
+Taxon names can be given at the species level or above.
+The user can provide them as a simple comma separated character string, or as tip labels on a tree.
+The tree can be provided in newick format (as a character string) or as a "phylo" R object, and can be with or without branch lengths.
+\item A name search is then performed across the chronogram database.
+Source trees with at least two matching input names are selected, and taxa that do not match the original query are dropped from these source trees.
+<!-- The resulting pruned source chronograms are hereafter referred as source chronograms.  --> <!--# This was generating some confussion, so I'm calling them pruned source chronograms now-->
+Finally, each pruned source chronogram is transformed to a patristic distance matrix.
+This format facilitates and greatly speeds up all further analyses and summarizing algorithms.
+For the sake of keeping track of authorship later, the matrices are associated to the citation of the original study.
+\item  At this point, the user can already obtain different summary information from their query, allowing the user to take informed decisions for the next steps of the workflow.
+Types of summary information provided are: a) all pruned source chronograms, b) age of the mrca (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of succesful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most taxa matched.
+\item  Then, source chronogram data can be used as secondary calibration points to date a tree with or without branch lengths containing some or all taxon names from the initial query.
+%<!--, a taxonomic tree-->
+\item  If there is no information available for any queried taxa, users can also create both age and phylogenetic data for this missing taxa with a variety of algorithms described below.
+\item  Finally, users can easily save all source and summary chronograms in formats that permit easy reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
+\end{enumerate}
 
 # Benchmark
+
+`datelife`'s code speed was tested on an Apple iMac
+with one 3.4 GHz Intel Core i5 processor. <!--No multiprocessor tests were performed-->
+We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
+Query processing time increases roughly linearly with number of input taxon names, and
+increases considerably if the taxonomic name resolution service [TNRS; @Boyle2013] is activated.
+<!-- Defining tnrs here again, bc we only talk about it two times, so it easy to forget it -->
+Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
+Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig.
+\ref{fig:runtime1}).
+<!--Speed of other datelife workflows? NO, bc it reflects speed from other software mainly. But I will put a sentence about it...
+Summarizing `datelife` results processing times: shows sdm speed, clustering methods speed;
+generating new trees: shows bladj speed;
+Adding dates processing time: reflects more mrbayes speed than anything;
+get_bold_otol_tree running time: reflects bold downloading data speed.-->
+<!--Speed with different number of chronograms in database? NO-->
+<!--Speed of web interface and of r package in computers with different capacities? NOOOO-->
+`datelife`'s code performance was evaluated with a set of unit tests designed and
+implemented with the R package testthat [@RCoreTeam2018] that were run both locally
+with the devtools package [@RCoreTeam2018], and on a public server --via
+GitHub, using the continuous integration tool Travis CI (<https://travis-ci.org>). At
+present, unit tests cover more than 30% of `datelife`'s code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->
 

### 2021-10-30 manuscript/manuscript.tex
description and benchmark

@@
 
-We report how we determined our sample size, all data exclusions (if any), all manipulations, and all measures in the study.
+The \texttt{datelife} workflow builds off from functions in several R packages (rotl (Michonneau, Brown, \& Winter, 2016), ape (Paradis, Claude, \& Strimmer, 2004),
+geiger (Harmon, Weir, Brock, Glor, \& Challenger, 2008), paleotree (Bapst, 2012), bold (Chamberlain, 2018), phytools (Revell, 2012), taxize (Chamberlain, 2018; Chamberlain \& Szöcs, 2013), phyloch (Heibl, 2008), and phylocomr (Ooms \& Chamberlain, 2018)).
+
+The basic \texttt{datelife} workflow is shown in figure \ref{fig:workflow} and consists of:
+
+\begin{enumerate}
+\item A user providing at least two taxon names as input.
+Taxon names can be given at the species level or above.
+The user can provide them as a simple comma separated character string, or as tip labels on a tree.
+The tree can be provided in newick format (as a character string) or as a "phylo" R object, and can be with or without branch lengths.
+\item A name search is then performed across the chronogram database.
+Source trees with at least two matching input names are selected, and taxa that do not match the original query are dropped from these source trees.
+<!-- The resulting pruned source chronograms are hereafter referred as source chronograms.  --> <!--# This was generating some confussion, so I'm calling them pruned source chronograms now-->
+Finally, each pruned source chronogram is transformed to a patristic distance matrix.
+This format facilitates and greatly speeds up all further analyses and summarizing algorithms.
+For the sake of keeping track of authorship later, the matrices are associated to the citation of the original study.
+\item  At this point, the user can already obtain different summary information from their query, allowing the user to take informed decisions for the next steps of the workflow.
+Types of summary information provided are: a) all pruned source chronograms, b) age of the mrca (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of succesful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most taxa matched.
+\item  Then, source chronogram data can be used as secondary calibration points to date a tree with or without branch lengths containing some or all taxon names from the initial query.
+%<!--, a taxonomic tree-->
+\item  If there is no information available for any queried taxa, users can also create both age and phylogenetic data for this missing taxa with a variety of algorithms described below.
+\item  Finally, users can easily save all source and summary chronograms in formats that permit easy reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
+\end{enumerate}
 
@@
 \section{Benchmark}\label{benchmark}}
+
+\texttt{datelife}'s code speed was tested on an Apple iMac
+with one 3.4 GHz Intel Core i5 processor.
+We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
+Query processing time increases roughly linearly with number of input taxon names, and
+increases considerably if the taxonomic name resolution service (TNRS; Boyle et al., 2013) is activated.
+Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
+Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig.
+\ref{fig:runtime1}).
+\texttt{datelife}'s code performance was evaluated with a set of unit tests designed and
+implemented with the R package testthat (R Core Team, 2018) that were run both locally
+with the devtools package (R Core Team, 2018), and on a public server --via
+GitHub, using the continuous integration tool Travis CI (\url{https://travis-ci.org}). At
+present, unit tests cover more than 30\% of \texttt{datelife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).
 
@@
 
+\leavevmode\hypertarget{ref-Bapst2012a}{}%
+Bapst, D.
+W.
+(2012).
+Paleotree: An R package for paleontological and phylogenetic analyses of evolution.
+\emph{Methods in Ecology and Evolution}, \emph{3}(5), 803--807.
+\url{https://doi.org/10.1111/j.2041-210X.2012.00223.x}
+
 \leavevmode\hypertarget{ref-barker2012going}{}%
@@
 
+\leavevmode\hypertarget{ref-Boyle2013}{}%
+Boyle, B., Hopkins, N., Lu, Z., Raygoza Garay, J.
+A., Mozzherin, D., Rees, T., \ldots{} Enquist, B.
+J.
+(2013).
+The taxonomic name resolution service: An online tool for automated standardization of plant names.
+\emph{BMC Bioinformatics}, \emph{14}(1).
+\url{https://doi.org/10.1186/1471-2105-14-16}
+
 \leavevmode\hypertarget{ref-burns2014phylogenetics}{}%
@@
 \emph{Molecular Phylogenetics and Evolution}, \emph{75}, 41--77.
+
+\leavevmode\hypertarget{ref-Chamberlain2018}{}%
+Chamberlain, S. (2018). \emph{bold: Interface to Bold Systems API}. Retrieved from \url{https://CRAN.R-project.org/package=bold}
+
+\leavevmode\hypertarget{ref-Chamberlain2013}{}%
+Chamberlain, S.
+A., \& Szöcs, E.
+(2013).
+taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}.
+\emph{F1000Research}, \emph{2}(191), 1--29.
+\url{https://doi.org/10.12688/f1000research.2-191.v2}
 
@@
 
+\leavevmode\hypertarget{ref-Harmon2008}{}%
+Harmon, L., Weir, J., Brock, C., Glor, R., \& Challenger, W. (2008). GEIGER: investigating evolutionary radiations. \emph{Bioinformatics}, \emph{24}, 129--131.
+
 \leavevmode\hypertarget{ref-Hedges2015}{}%
@@
 \url{https://doi.org/10.1093/molbev/msv037}
+
+\leavevmode\hypertarget{ref-Heibl2008}{}%
+Heibl, C. (2008). \emph{PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages.} Retrieved from \url{http://www.christophheibl.de/Rpackages.html}
 
@@
 
+\leavevmode\hypertarget{ref-Michonneau2016}{}%
+Michonneau, F., Brown, J.
+W., \& Winter, D.
+J.
+(2016).
+rotl: an R package to interact with the Open Tree of Life data.
+\emph{Methods in Ecology and Evolution}, \emph{7}(12), 1476--1481.
+\url{https://doi.org/10.1111/2041-210X.12593}
+
+\leavevmode\hypertarget{ref-Ooms2018}{}%
+Ooms, J., \& Chamberlain, S. (2018). \emph{Phylocomr: Interface to 'phylocom'}. Retrieved from \url{https://CRAN.R-project.org/package=phylocomr}
+
+\leavevmode\hypertarget{ref-Paradis2004}{}%
+Paradis, E., Claude, J., \& Strimmer, K. (2004). APE: analyses of phylogenetics and evolution in R language. \emph{Bioinformatics}, \emph{20}, 289--290.
+
 \leavevmode\hypertarget{ref-price2014niche}{}%
@@
 
+\leavevmode\hypertarget{ref-RCoreTeam2018}{}%
+R Core Team. (2018). \emph{R: a language and environment for statistical computing}. Vienna, Austria: R Foundation for Statistical Computing.
+
 \leavevmode\hypertarget{ref-R-base}{}%
 R Core Team. (2021). \emph{R: A language and environment for statistical computing}. Vienna, Austria: R Foundation for Statistical Computing. Retrieved from \url{https://www.R-project.org/}
+
+\leavevmode\hypertarget{ref-Revell2012}{}%
+Revell, L. J. (2012). Phytools: An r package for phylogenetic comparative biology (and other things). \emph{Methods in Ecology and Evolution}, \emph{3}, 217--223.
 

### 2021-10-30 manuscript/paper_3-main.Rmd
description and benchmark

@@
 
-The basic `datelife` workflow is shown in figure \ref{fig:workflow} and consists of:
-
-\begin{enumerate}
-\item A user providing at least two taxon names as input.
-Taxon names can be given at the species level or above.
-The user can provide them as a simple comma separated character string, or as tip labels on a tree.
-The tree can be provided in newick format (as a character string) or as a "phylo" R object, and can be with or without branch lengths.
-\item A name search is then performed across the chronogram database.
-Source trees with at least two matching input names are selected, and taxa that do not match the original query are dropped from these source trees.
-<!-- The resulting pruned source chronograms are hereafter referred as source chronograms.  --> <!--# This was generating some confussion, so I'm calling them pruned source chronograms now-->
-Finally, each pruned source chronogram is transformed to a patristic distance matrix.
-This format facilitates and greatly speeds up all further analyses and summarizing algorithms.
-For the sake of keeping track of authorship later, the matrices are associated to the citation of the original study.
-\item  At this point, the user can already obtain different summary information from their query, allowing the user to take informed decisions for the next steps of the workflow.
-Types of summary information provided are: a) all pruned source chronograms, b) age of the mrca (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of succesful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most taxa matched.
-\item  Then, source chronogram data can be used as calibration points to date a tree with or without branch lengths containing some or all names from the initial query. %<!--, a taxonomic tree-->
-\item  If there is no information available for any queried taxa, users can also create both age and phylogenetic data for this missing taxa with a variety of algorithms described below.
-\item  Finally, users can easily save all source and summary chronograms in formats that permit easy reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
-\end{enumerate}
-
-<!--To perform each task:-->
-To gather, process, and present information, `datelife` builds up from functions
-available in several R packages including rotl [@Michonneau2016], ape [@Paradis2004],
-geiger [@Harmon2008], paleotree [@Bapst2012a], bold [@Chamberlain2018], phytools [@Revell2012], taxize [@Chamberlain2013; @Chamberlain2018], phyloch [@Heibl2008], phylocomr [@Ooms2018] and rphylotastic [@Omeara2019].
-
 A `datelife` search currently accepts scientific names only. It can be any named clade or binomial specific. <!--common names (implemented in phylotastic)-->
@@
 <!-- Good to test 3 features on a package: performance, speed and scalability-->
-
-`datelife`'s code speed was tested on an Apple iMac
-with one 3.4 GHz Intel Core i5 processor. <!--No multiprocessor tests were performed-->
-We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
-Query processing time increases roughly linearly with number of input taxon names, and
-increases considerably if taxonomic name resolution with TNRS [@Boyle2013] is activated.
-<!-- Defining tnrs here again, bc we only talk about it two times, so it easy to forget it -->
-Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
-Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig.
-\ref{fig:runtime1}).
-<!--Speed of other datelife workflows? NO, bc it refkects speed from other software mainly. But I will put a sentence about it...
-Summarizing `datelife` results processing times: shows sdm speed, clustering methods speed;
-generating new trees: shows bladj speed;
-Adding dates processing time: reflects more mrbayes speed than anything;
-get_bold_otol_tree running time: reflects bold downloading data speed.-->
-<!--Speed with different number of chronograms in database? NO-->
-<!--Speed of web interface and of r package in computers with different capacities? NOOOO-->
-`datelife`'s code performance was evaluated with a set of unit tests designed and
-implemented with the R package testthat [@RCoreTeam2018] that were run both locally
-with the devtools package [@RCoreTeam2018], and on a public server --via
-GitHub, using the continuous integration tool Travis CI (<https://travis-ci.org>). At
-present, unit tests cover more than 50% of `datelife`'s code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->
 

### 2021-10-30 manuscript/manuscript.Rmd
results

@@
 availability <- ifelse(rpackage_available, "is available", "will be available")
+dq <- datelife::make_datelife_query(input = "fringillidae", get_spp_from_taxon = TRUE)
 ```
@@
 
-# Description
+# Implementation/Description/Workflow
 
-The `datelife` workflow builds off from functions in several R packages (rotl [@Michonneau2016], ape [@Paradis2004],
+The `datelife` workflow builds off of functions from several R packages (rotl [@Michonneau2016], ape [@Paradis2004],
 geiger [@Harmon2008], paleotree [@Bapst2012a], bold [@Chamberlain2018], phytools [@Revell2012], taxize [@Chamberlain2013; @Chamberlain2018], phyloch [@Heibl2008], and phylocomr [@Ooms2018]).
@@
 
-# Example
-We used `r cite_r("r-references.bib")` for all our analyses.
+# Results
+
+<!--We used `r cite_r("r-references.bib")` for all our analyses.-->
+
+## Case study
+
+We illustrate the `datelife` workflow using the family of true finches, Fringillidae as an example.
+To contextualize, a college educator wishes to know the state-of-the-art on time of evolutionary origin of species belonging to the true finches.
+One option is to go to the website at www.datelife.org and perform an interactive run. However, the educator wants the students to practice their R skills. 
+The first step is to run a higher-taxon-name 'datelife` query.
+This will get taxon names for all recognised species within any higher taxon.
+The Fringillidae has `r length(dq$cleaned_names)` species, according to the Open Tree of Life taxonomy.
+Once with a curated set of query taxon names, the next step is to run a `datelife` search.
+This will find all chronograms that contain at least two queried taxon names, and will save the information on time of lineage divergence as (an R "data frame") table.
+ There are `r length(dr)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dr)))` different studies (Fig. \ref{fig:schronograms1}).
+<!-- In case we want to cite the studies in here, but they are already cited in the figure caption: [Fig.
+\ref{fig:schronograms1}; @barker2012going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
+-->
+The final step is to summarize the available information using the two alternative types of summary chronograms, median and SDM.
+As explained in the "Description" section, data from source chronograms is first summarised into a single distance matrix (using the median and the SDM method respectively) and then the available node ages are used as fixed ages over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig.
+\ref{fig:summaries}).
+Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
+With both methods, ages are generally consistent with source ages, but there are some biological examples in which this is not true (see Discussion).
+<!-- (Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
+<!-- and hierarchical clustering method which appears to push ages back -->
+
+## Cross-validation test
+Data from source chronograms can be also used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates (Figs.
+\ref{fig:cvbladj} and \ref{fig:cvbold}).
+As a form of cross validation, we took tree topologies from each study and calibrated them using time of lineage divergence data from all other source chronograms.
+In the absence of branch lengths, the ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig.
+\ref{fig:cvbladj}).
+Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cvbladj}).
+<!-- We need to think about the root! is it comparable across source chronograms?? not all of them have the same sampling... so probably not the same root and thus not comparable. Address this -->
+We also demonstrate the usage of PATHd8 [@Britton2007] as an alternative method to BLADJ.
+For this, we run a `datelife` branch length reconstruction that searches for DNA sequence data from the Barcode of Life Data System [BOLD; ratnasingham2007bold] to generate branch lengths.
+We were able to successfully generate a tree with BOLD branch lengths for all of the Fringillidae source chronograms.
+However, dating with PATHd8 using congruified calibrations, was only successful in
+<!-- two studies when expanding calibrations to make them agree (studies 3 and 5), and from  -->
+three cases (studies 3, 5, and 9, shown in Fig.
+\ref{fig:cvbold}).
+From these, two trees have a different sampling than the original source chronogram, mainly because DNA BOLD  data for some species is absent from the database.
+Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
+More examples and code used to generate these trees were developed on an open repository that is available for consultation and reuse at <https://github.com/LunaSare/datelife_examples>.
 

### 2021-10-30 manuscript/paper_3-main.Rmd
results

@@
 -->
-In this section we demonstrate the types of outputs that can be obtained with `datelife`, using as an example the bird family Fringillidae of true finches.
-We performed a higher-taxon search to obtain data on lineage divergence available in `datelife`'s database for all recognised species within the Fringillidae (`r length(dq$cleaned_names)` species according to the Open Tree of Life taxonomy).
-There are `r length(dr)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dr)))` different studies (Fig.
-\ref{fig:schronograms1}).
-<!-- In case we want to cite the studies in here, but they are already cited in the figure caption: [Fig.
-\ref{fig:schronograms1}; @barker2012going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
--->
-Data from these source chronograms was used to generate two types of summary chronograms, median and SDM.
-As explained in the "Description" section, data from source chronograms was first summarised into a single distance matrix (using the median and the SDM method respectively) and then the available node ages were used as fixed ages over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig.
-\ref{fig:summaries}).
-Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
-With both methods, ages are generally consistent with source ages, but there are some biological examples in which this is not true (see Flaws, Limitations and Prospects section).
-<!-- (Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
-<!-- and hierarchical clustering method which appears to push ages back -->
-
-Data from source chronograms was also used to date tree topologies with no branch length information and trees with branch lengths in relative substitution rates (Figs.
-\ref{fig:cvbladj} and \ref{fig:cvbold}).
-As a form of cross validation, we used tree topologies from each study and calibrated them using information from all other source chronograms.
-In the absence of branch length data, the ages of internal nodes were approximately recovered in almost all cases (except for studies 3, and 5; Fig.
-\ref{fig:cvbladj}).
-Maximum tree ages were only approximately recovered in one case (study 2; Fig. \ref{fig:cvbladj}).
-<!-- We need to think abput the root! is it comparable across source chronograms?? not all of them have the same sampling... so probably not the same root and thus not comparable. Address this -->
-To exemplify dating a tree with branch lengths provided by the user, we obtained DNA sequence data using the Barcode of Life Data System [BOLD; ratnasingham2007bold] and branch lengths were successfully generated for all source chronograms.
-However, dating with PATHd8 [@Britton2007] using congruified calibrations, was only successful in
-<!-- two studies when expanding calibrations to make them agree (studies 3 and 5), and from  -->
-three cases (studies 3, 5, and 9, shown in Fig.
-\ref{fig:cvbold}).
-From these, two trees have a different sampling than the original source chronogram, mainly because DNA data for some species is absent from the BOLD.
-Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
-More examples and code used to generate these trees were developes on an open repository that is available for consultation and reuse at <https://github.com/LunaSare/datelife_examples>.
 

### 2021-11-02 manuscript/manuscript.Rmd
implementation

@@
 keywords          : "Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations"
-wordcount         : "571"
+wordcount         : "2628"
 
@@
 knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
-data(opentree_chronograms)
+data(opentree_chronograms, package = "datelife")
 rpackage_available <- FALSE
 availability <- ifelse(rpackage_available, "is available", "will be available")
-dq <- datelife::make_datelife_query(input = "fringillidae", get_spp_from_taxon = TRUE)
+load("../data/dq.rda")
+load("../data/dr.rda")
 ```
@@
 
+The main goal of `datelife` is to generate a chronogram for any given combination of taxon names, based on expert scientific information.
+
 The `datelife` workflow builds off of functions from several R packages (rotl [@Michonneau2016], ape [@Paradis2004],
@@
 
-The basic `datelife` workflow is shown in figure \ref{fig:workflow} and consists of:
+The basic `datelife` workflow is shown in figure \ref{fig:workflow}, largely:
 
 \begin{enumerate}
-\item A user providing at least two taxon names as input.
-Taxon names can be given at the species level or above.
-The user can provide them as a simple comma separated character string, or as tip labels on a tree.
-The tree can be provided in newick format (as a character string) or as a "phylo" R object, and can be with or without branch lengths.
-\item A name search is then performed across the chronogram database.
-Source trees with at least two matching input names are selected, and taxa that do not match the original query are dropped from these source trees.
-<!-- The resulting pruned source chronograms are hereafter referred as source chronograms.  --> <!--# This was generating some confussion, so I'm calling them pruned source chronograms now-->
-Finally, each pruned source chronogram is transformed to a patristic distance matrix.
-This format facilitates and greatly speeds up all further analyses and summarizing algorithms.
-For the sake of keeping track of authorship later, the matrices are associated to the citation of the original study.
-\item  At this point, the user can already obtain different summary information from their query, allowing the user to take informed decisions for the next steps of the workflow.
-Types of summary information provided are: a) all pruned source chronograms, b) age of the mrca (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of succesful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most taxa matched.
-\item  Then, source chronogram data can be used as secondary calibration points to date a tree with or without branch lengths containing some or all taxon names from the initial query.
+\item It starts with an input consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree.
+The tree can be provided in newick format, also as a character string, or as a "phylo" R object, and can have any type of branch lengths or none.
+\item The input taxon names are cleaned with TNRS and saved as a `datelifeQuery` object.
+If taxon names are taxonomic groups above the species level, `datelife` has two alternative behaviors.
+If the "get species from taxon" flag is active, `datelife` will retrieve all species within a higher taxon name and add the species names to the input.
+If the flag is inactive, `datelife` will drop the higher taxon names from the input.
+The cleaned input taxon names are searched across the source chronogram database.
+Source chronograms with at least two matching input taxon names are singled out and pruned down to preserve only input taxon names in the tips of the chronograms.
+<!-- The resulting pruned source chronograms are hereafter referred as source chronograms.  --> <!-- This was generating some confusion, so I'm calling them pruned source chronograms now-->
+Then, each pruned source chronogram is transformed to a patristic distance matrix.
+This format facilitates and greatly speeds up all downstream analyses and summaries.
+The matrices are associated to the citation of the original study and stored as a `datelifeResult` object.
+\item  At this point, various summary data can be obtained to inform decisions for the next steps of the analysis workflow.
+Types of summary information provided are: a) all pruned source chronograms, b) age of the MRCA (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of successful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most matching input taxon names.
+\item  Finally, time of lineage divcergence obtained from the pruned source chronograms can be used as secondary calibration points to date a tree with or without branch lengths containing some or all input taxon names.
 %<!--, a taxonomic tree-->
-\item  If there is no information available for any queried taxa, users can also create both age and phylogenetic data for this missing taxa with a variety of algorithms described below.
-\item  Finally, users can easily save all source and summary chronograms in formats that permit easy reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
+\item  If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms described below.
+\item  Users can easily save all source and summary chronograms in formats that permit easy reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
 \end{enumerate}
@@
 We illustrate the `datelife` workflow using the family of true finches, Fringillidae as an example.
-To contextualize, a college educator wishes to know the state-of-the-art on time of evolutionary origin of species belonging to the true finches.
+To contextualize, a college educator wishes to know the state-of-the-art on time of evolutionary origin of species belonging to the true finches using `datelife`.
 One option is to go to the website at www.datelife.org and perform an interactive run. However, the educator wants the students to practice their R skills. 
-The first step is to run a higher-taxon-name 'datelife` query.
+The first step is to run a higher-taxon-name `datelife` query.
 This will get taxon names for all recognised species within any higher taxon.
@@
 
+The main goal of `datelife` is to make expert information on time of lineage divergence easily accesible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
+It is a very fast tool that fulfills the quality of openness and does not require any expert biological knowledge from users --besides the names of the organisms they want to work with-- for any of its functionalities.
+However, it has many flaws.
+Some of them can be overcome, some of them might represent limitations.
+
+Up to the time this manuscript was written, `datelife`'s chronogram database had `r length(opentree_chronograms$trees)` chronograms, pulled entirely from OpenTree's tree repository, the only public tree repository from where `datelife` can currently get chronograms to construct its database.
+This represents `r round(length(opentree_chronograms$trees)*100/3987, digits = 2)`% of the largest existing chronogram database, TimeTree, which has a collection of 3,998 chronograms as of `r format(Sys.Date(), format="%B %d %Y")`.
+Unfortunately, TimeTree's database is not open for scientific reuse nor automatised data mining [@Kumar2017].
+In 2015, a synthetic chronogram was constructed from 2,274 chronograms available at the time on the TimeTree database [@Hedges2015].
+This is the only synthetic TimeTree chronogram that has been made publicly available and deposited on the OpenTree repository, and is part of datelife's database now.
+Hence, the amount of lineages represented in datelife's database is at least as substantial as TimeTree's, ensuring that some information will be available for any given taxon or lineage.
+Regrettably, this does not ensure that the full state of knowledge of time of divergence of the taxon/lineage will be available.
+<!--
+Even though datelifes chronograms cover this many species from the tree of life compared to this many covered by TimeTree (database or synthetic chronogram?)-->
+Incorporation of more published chronograms into `datelife`'s database is crucial to improve its services.
+One option to increase our database is the Dryad data repository.
+Methods to automatically mine chronograms from Dryad could be designed and implemented.
+However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined on a second step, by a curator.
+<!-- Can we mine journal databases? -->
+<!-- At the same time, the prospect is to increase awereness and keep inviting and doing work encouraging scientists to make their data sets available in OpenTree's database. -->
+Consequently, we would like to emphasize on the importance of sharing chronogram data for the benefit of the scientific community as a whole, into repositories that require expert input and manual curation, such as OpenTree's tree repository [@mctavish2015phylesystem].
+
+Another potential concern comes from summary chronograms.
+We currently summarize by default all source chronograms that overlap with at least two taxa.
+Users can subset source data if they have reasons to choose some source chronograms over others.
+Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely.
+To our knowledge, there is no objective way to determine if an expert chronogram is better than another.
+Some criteria that have been put forward are the level of lineage sampling and the number of calibrations used.
+Scientists usually also favor chronograms constructed using primary calibrations (ages obtained from the fossil or geological record) to ones constructed with secondary calibrations (ages coming from other chronograms).
+It has been observed with simulations that divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with bayesian inference methods when priors are implemented in similar ways in both analyses [@schenk2016sec].
+Yet, there are different ways to use secondary calibrations
+<!-- give some examples or alternative ways of using calibrations-->
+and that same bias might not be encountered with dating methods that do not require setting priors, i.e., Maximum Likelihood methods such as r8s [@sanderson2003r8s].
+Certainly, further studies are required to fully understand the effect of using secondary calibrations on time estimates and downstream anlyses.
+
+Furthermore, even chronograms obtained with primary fossil data can show substantial variation in time estimates between clades, as observed from the comparison of source chronograms in the Fringillidae example.
+This observation is often encountered in the literature
+<!-- Different source chronograms often show substantial variation in clade ages  -->
+(see, for example, the ongoing debate about crown group age of angiosperms [@barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time].
+For some studies, especially ones based on branch lengths (e.g., studies of species diversification, timing of evolutionary events, phenotypic trait evolution), using a different chronogram may return different results [@title2016macrophylogenies].
+Stitching together these chronograms can create a larger tree that uses information from multiple studies, but the effect of uncertainties and errors here on downstream analyses is still largely unknown.
+
+Summarizing chronograms might also imply summarizing fundamentally distinct evolutionary hypotheses.
+For example, two different researchers working on the same clade both carefully select and argument their choices of fossil calibrations.
+Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses teh same one to calibrate outside the clade, the resulting age estimates will probably differ substantially (the placement of calibrations is proved to deeply affect estimated times of lineage divergence).
+Trying to summarize the resulting chronograms into a single one using simple summary statistics might erase all types of relevant information from the source chronograms.
+Accordingly, the prevailing view in our research community is that we should favor time of lineage divergence estimates obtained from a single analysis, using fossil data as primary sources of calibrations, and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history [@antonelli2017supersmart].
+<!-- Will this be implemented in future `datelife` versions?
+Probably not-->
+However, the exercise of summarizing different chronograms ha sthe potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypothesis.
+Choosing the elements of the chronograms that we are going tp keep and the ones that we are going to discard is key, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram.
+<!-- Example -->
+
+<!-- still requires more research. -->
+<!-- Often, there is a large discrepancy in time of lineage divergence estimates across studies from expert knowledge, that is well known and documented for different groups of organisms [e.g., angiosperms; @magallon2015metacalibrated].
+-->
+<!-- Comparison of available age data for a wide range of organisms shown here suggest that this is a widespread phenomenon that requires further attention. -->
+<!-- In the case of the example presented here, most chronograms are new, meaning that we expect that data was generated with high quality and the software used should be very advanced.
+However, this is just an assumption.
+-->
+
+Alternatively, one could try to choose the "best" chronogram from a set of possible evolutionary hypotheses.
+Several characteristics of the data used for dating analyses as well as from the output chronogram itself, could be used to score quality of source chronograms.
+Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
+To facilitate subsetting of source chronograms following different criteria by the users, this information should be included as metadata manually entered by curators in the future.
+
+<!-- Still, even if all source chronograms have been generated by excellent standards and using similar methods, the evolutionary history they depict might be very different.  -->
+
+In other areas of biological research, such as ecology and conservation biology, it has been shown that at least some data on lineage divergence represents a relevant improvement for testing alternative hypothesis using phylogenetic distance [@Webb2008].
+Hence, we integrated into datelife's workflow different ways of creating branch lengths in the absence of starting branch length information for taxa lacking this information (BLADJ option).
+Making up branch lengths in this or other ways is accepted in scientific publications: @Jetz2012, created a time-calibrated tree of all 9,993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a tree of 353,185 seed plants where only 23% had molecular data.
+Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch imputation methods.
+There has yet to be a thorough analysis of what can go wrong when one goes beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make sure results are substantially similar between trees fully reconstructed with molecular or other data, and trees that are reconstructed using taxonomy by resolving polytomies at random following a statistical model.
+
+<!-- There is evidence that different chronograms will result in very different macroevolutionary histories inferred from them [@title2016macrophylogenies].
+However, it is unknown how using very different age data could affect hypothesis testing in other areas of research and more effort is needed in that directions -->
+
 # Conclusions
+
+Divergence time information is key to many areas of evolutionary studies: trait evolution,
+diversification, biogeography, macroecology and more. It is also crucial for science communication and education, but generating chronograms _de novo_ is difficult,
+especially for those who want to use phylogenies but who are not systematists, or
+do not have the time to acquire and develop the necessary knowledge and data curation skills.
+Moreover, years of primarily public funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
+
+<!--Mention again uses shown here and outline potential uses not shown here-->
+`datelife` allows easy and fast summarization of publicly available information
+on time of lineage divergence.
+This provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identification of regions that require more research or that have conflicting information.
+<!-- It can also summarize public data, or use it to generate new chronograms based on trees provided by the user. -->
+Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
+`datelife` helps with awareness of the existing variation in expert time of divergence data, and will foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
 

### 2021-11-02 manuscript/manuscript.tex
implementation

@@
 \shorttitle{DateLife: revealing the dated Tree of Life}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations\newline\indent Word count: 55}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations\newline\indent Word count: 2628}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 
-\begin{verbatim}
-## Warning in data(opentree_chronograms): data set 'opentree_chronograms' not found
-\end{verbatim}
-
 \hypertarget{introduction}{%
@@
 
-\hypertarget{description}{%
-\section{Description}\label{description}}
+\hypertarget{implementationdescriptionworkflow}{%
+\section{Implementation/Description/Workflow}\label{implementationdescriptionworkflow}}
 
-The \texttt{datelife} workflow builds off from functions in several R packages (rotl (Michonneau, Brown, \& Winter, 2016), ape (Paradis, Claude, \& Strimmer, 2004),
+The main goal of \texttt{datelife} is to generate a chronogram for any given combination of taxon names, based on expert scientific information.
+
+The \texttt{datelife} workflow builds off of functions from several R packages (rotl (Michonneau, Brown, \& Winter, 2016), ape (Paradis, Claude, \& Strimmer, 2004),
 geiger (Harmon, Weir, Brock, Glor, \& Challenger, 2008), paleotree (Bapst, 2012), bold (Chamberlain, 2018), phytools (Revell, 2012), taxize (Chamberlain, 2018; Chamberlain \& Szöcs, 2013), phyloch (Heibl, 2008), and phylocomr (Ooms \& Chamberlain, 2018)).
 
-The basic \texttt{datelife} workflow is shown in figure \ref{fig:workflow} and consists of:
+The basic \texttt{datelife} workflow is shown in figure \ref{fig:workflow}, largely:
 
 \begin{enumerate}
-\item A user providing at least two taxon names as input.
-Taxon names can be given at the species level or above.
-The user can provide them as a simple comma separated character string, or as tip labels on a tree.
-The tree can be provided in newick format (as a character string) or as a "phylo" R object, and can be with or without branch lengths.
-\item A name search is then performed across the chronogram database.
-Source trees with at least two matching input names are selected, and taxa that do not match the original query are dropped from these source trees.
-<!-- The resulting pruned source chronograms are hereafter referred as source chronograms.  --> <!--# This was generating some confussion, so I'm calling them pruned source chronograms now-->
-Finally, each pruned source chronogram is transformed to a patristic distance matrix.
-This format facilitates and greatly speeds up all further analyses and summarizing algorithms.
-For the sake of keeping track of authorship later, the matrices are associated to the citation of the original study.
-\item  At this point, the user can already obtain different summary information from their query, allowing the user to take informed decisions for the next steps of the workflow.
-Types of summary information provided are: a) all pruned source chronograms, b) age of the mrca (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of succesful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most taxa matched.
-\item  Then, source chronogram data can be used as secondary calibration points to date a tree with or without branch lengths containing some or all taxon names from the initial query.
+\item It starts with an input consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree.
+The tree can be provided in newick format, also as a character string, or as a "phylo" R object, and can have any type of branch lengths or none.
+\item The input taxon names are cleaned with TNRS and saved as a `datelifeQuery` object.
+If taxon names are taxonomic groups above the species level, `datelife` has two alternative behaviors.
+If the "get species from taxon" flag is active, `datelife` will retrieve all species within a higher taxon name and add the species names to the input.
+If the flag is inactive, `datelife` will drop the higher taxon names from the input.
+The cleaned input taxon names are searched across the source chronogram database.
+Source chronograms with at least two matching input taxon names are singled out and pruned down to preserve only input taxon names in the tips of the chronograms.
+<!-- The resulting pruned source chronograms are hereafter referred as source chronograms.  --> <!-- This was generating some confusion, so I'm calling them pruned source chronograms now-->
+Then, each pruned source chronogram is transformed to a patristic distance matrix.
+This format facilitates and greatly speeds up all downstream analyses and summaries.
+The matrices are associated to the citation of the original study and stored as a `datelifeResult` object.
+\item  At this point, various summary data can be obtained to inform decisions for the next steps of the analysis workflow.
+Types of summary information provided are: a) all pruned source chronograms, b) age of the MRCA (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of successful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most matching input taxon names.
+\item  Finally, time of lineage divcergence obtained from the pruned source chronograms can be used as secondary calibration points to date a tree with or without branch lengths containing some or all input taxon names.
 %<!--, a taxonomic tree-->
-\item  If there is no information available for any queried taxa, users can also create both age and phylogenetic data for this missing taxa with a variety of algorithms described below.
-\item  Finally, users can easily save all source and summary chronograms in formats that permit easy reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
+\item  If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms described below.
+\item  Users can easily save all source and summary chronograms in formats that permit easy reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
 \end{enumerate}
@@
 
-\hypertarget{example}{%
-\section{Example}\label{example}}
+\hypertarget{results}{%
+\section{Results}\label{results}}
 
-We used R (Version 4.1.0; R Core Team, 2021) and the R-package \emph{papaja} (Version 0.1.0.9997; Aust \& Barth, 2020) for all our analyses.
+\hypertarget{case-study}{%
+\subsection{Case study}\label{case-study}}
+
+We illustrate the \texttt{datelife} workflow using the family of true finches, Fringillidae as an example.
+To contextualize, a college educator wishes to know the state-of-the-art on time of evolutionary origin of species belonging to the true finches using \texttt{datelife}.
+One option is to go to the website at www.datelife.org and perform an interactive run. However, the educator wants the students to practice their R skills.
+The first step is to run a higher-taxon-name \texttt{datelife} query.
+This will get taxon names for all recognised species within any higher taxon.
+The Fringillidae has 289 species, according to the Open Tree of Life taxonomy.
+Once with a curated set of query taxon names, the next step is to run a \texttt{datelife} search.
+This will find all chronograms that contain at least two queried taxon names, and will save the information on time of lineage divergence as (an R \enquote{data frame}) table.
+There are 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Fig. \ref{fig:schronograms1}).
+The final step is to summarize the available information using the two alternative types of summary chronograms, median and SDM.
+As explained in the \enquote{Description} section, data from source chronograms is first summarised into a single distance matrix (using the median and the SDM method respectively) and then the available node ages are used as fixed ages over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig.
+\ref{fig:summaries}).
+Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
+With both methods, ages are generally consistent with source ages, but there are some biological examples in which this is not true (see Discussion).
+
+\hypertarget{cross-validation-test}{%
+\subsection{Cross-validation test}\label{cross-validation-test}}
+
+Data from source chronograms can be also used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates (Figs.
+\ref{fig:cvbladj} and \ref{fig:cvbold}).
+As a form of cross validation, we took tree topologies from each study and calibrated them using time of lineage divergence data from all other source chronograms.
+In the absence of branch lengths, the ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig.
+\ref{fig:cvbladj}).
+Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cvbladj}).
+We also demonstrate the usage of PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007) as an alternative method to BLADJ.
+For this, we run a \texttt{datelife} branch length reconstruction that searches for DNA sequence data from the Barcode of Life Data System {[}BOLD; ratnasingham2007bold{]} to generate branch lengths.
+We were able to successfully generate a tree with BOLD branch lengths for all of the Fringillidae source chronograms.
+However, dating with PATHd8 using congruified calibrations, was only successful in
+three cases (studies 3, 5, and 9, shown in Fig.
+\ref{fig:cvbold}).
+From these, two trees have a different sampling than the original source chronogram, mainly because DNA BOLD data for some species is absent from the database.
+Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
+More examples and code used to generate these trees were developed on an open repository that is available for consultation and reuse at \url{https://github.com/LunaSare/datelife_examples}.
 
@@
 
+The main goal of \texttt{datelife} is to make expert information on time of lineage divergence easily accesible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
+It is a very fast tool that fulfills the quality of openness and does not require any expert biological knowledge from users --besides the names of the organisms they want to work with-- for any of its functionalities.
+However, it has many flaws.
+Some of them can be overcome, some of them might represent limitations.
+
+Up to the time this manuscript was written, \texttt{datelife}'s chronogram database had 231 chronograms, pulled entirely from OpenTree's tree repository, the only public tree repository from where \texttt{datelife} can currently get chronograms to construct its database.
+This represents 5.79\% of the largest existing chronogram database, TimeTree, which has a collection of 3,998 chronograms as of November 01 2021.
+Unfortunately, TimeTree's database is not open for scientific reuse nor automatised data mining (Kumar, Stecher, Suleski, \& Hedges, 2017).
+In 2015, a synthetic chronogram was constructed from 2,274 chronograms available at the time on the TimeTree database (Hedges, Marin, Suleski, Paymer, \& Kumar, 2015).
+This is the only synthetic TimeTree chronogram that has been made publicly available and deposited on the OpenTree repository, and is part of datelife's database now.
+Hence, the amount of lineages represented in datelife's database is at least as substantial as TimeTree's, ensuring that some information will be available for any given taxon or lineage.
+Regrettably, this does not ensure that the full state of knowledge of time of divergence of the taxon/lineage will be available.
+Incorporation of more published chronograms into \texttt{datelife}'s database is crucial to improve its services.
+One option to increase our database is the Dryad data repository.
+Methods to automatically mine chronograms from Dryad could be designed and implemented.
+However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined on a second step, by a curator.
+Consequently, we would like to emphasize on the importance of sharing chronogram data for the benefit of the scientific community as a whole, into repositories that require expert input and manual curation, such as OpenTree's tree repository (McTavish et al., 2015).
+
+Another potential concern comes from summary chronograms.
+We currently summarize by default all source chronograms that overlap with at least two taxa.
+Users can subset source data if they have reasons to choose some source chronograms over others.
+Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely.
+To our knowledge, there is no objective way to determine if an expert chronogram is better than another.
+Some criteria that have been put forward are the level of lineage sampling and the number of calibrations used.
+Scientists usually also favor chronograms constructed using primary calibrations (ages obtained from the fossil or geological record) to ones constructed with secondary calibrations (ages coming from other chronograms).
+It has been observed with simulations that divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with bayesian inference methods when priors are implemented in similar ways in both analyses (Schenk, 2016).
+Yet, there are different ways to use secondary calibrations
+and that same bias might not be encountered with dating methods that do not require setting priors, i.e., Maximum Likelihood methods such as r8s (Sanderson, 2003).
+Certainly, further studies are required to fully understand the effect of using secondary calibrations on time estimates and downstream anlyses.
+
+Furthermore, even chronograms obtained with primary fossil data can show substantial variation in time estimates between clades, as observed from the comparison of source chronograms in the Fringillidae example.
+This observation is often encountered in the literature
+(see, for example, the ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández, 2015; Ramshaw et al., 1972; Sanderson \& Doyle, 2001).
+For some studies, especially ones based on branch lengths (e.g., studies of species diversification, timing of evolutionary events, phenotypic trait evolution), using a different chronogram may return different results (Title \& Rabosky, 2016).
+Stitching together these chronograms can create a larger tree that uses information from multiple studies, but the effect of uncertainties and errors here on downstream analyses is still largely unknown.
+
+Summarizing chronograms might also imply summarizing fundamentally distinct evolutionary hypotheses.
+For example, two different researchers working on the same clade both carefully select and argument their choices of fossil calibrations.
+Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses teh same one to calibrate outside the clade, the resulting age estimates will probably differ substantially (the placement of calibrations is proved to deeply affect estimated times of lineage divergence).
+Trying to summarize the resulting chronograms into a single one using simple summary statistics might erase all types of relevant information from the source chronograms.
+Accordingly, the prevailing view in our research community is that we should favor time of lineage divergence estimates obtained from a single analysis, using fossil data as primary sources of calibrations, and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history (Antonelli et al., 2017).
+However, the exercise of summarizing different chronograms ha sthe potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypothesis.
+Choosing the elements of the chronograms that we are going tp keep and the ones that we are going to discard is key, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram.
+
+Alternatively, one could try to choose the \enquote{best} chronogram from a set of possible evolutionary hypotheses.
+Several characteristics of the data used for dating analyses as well as from the output chronogram itself, could be used to score quality of source chronograms.
+Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
+To facilitate subsetting of source chronograms following different criteria by the users, this information should be included as metadata manually entered by curators in the future.
+
+In other areas of biological research, such as ecology and conservation biology, it has been shown that at least some data on lineage divergence represents a relevant improvement for testing alternative hypothesis using phylogenetic distance (Webb, Ackerly, \& Kembel, 2008).
+Hence, we integrated into datelife's workflow different ways of creating branch lengths in the absence of starting branch length information for taxa lacking this information (BLADJ option).
+Making up branch lengths in this or other ways is accepted in scientific publications: Jetz, Thomas, Joy, Hartmann, and Mooers (2012), created a time-calibrated tree of all 9,993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
+(2018) created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37\% had molecular data; Smith and Brown (2018) constructed a tree of 353,185 seed plants where only 23\% had molecular data.
+Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch imputation methods.
+There has yet to be a thorough analysis of what can go wrong when one goes beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make sure results are substantially similar between trees fully reconstructed with molecular or other data, and trees that are reconstructed using taxonomy by resolving polytomies at random following a statistical model.
+
 \hypertarget{conclusions}{%
 \section{Conclusions}\label{conclusions}}
+
+Divergence time information is key to many areas of evolutionary studies: trait evolution,
+diversification, biogeography, macroecology and more. It is also crucial for science communication and education, but generating chronograms \emph{de novo} is difficult,
+especially for those who want to use phylogenies but who are not systematists, or
+do not have the time to acquire and develop the necessary knowledge and data curation skills.
+Moreover, years of primarily public funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
+
+\texttt{datelife} allows easy and fast summarization of publicly available information
+on time of lineage divergence.
+This provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identification of regions that require more research or that have conflicting information.
+Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
+\texttt{datelife} helps with awareness of the existing variation in expert time of divergence data, and will foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
 
@@
 \hypertarget{refs}{}
-\leavevmode\hypertarget{ref-R-papaja}{}%
-Aust, F., \& Barth, M. (2020). \emph{papaja: Create APA manuscripts with R Markdown}. Retrieved from \url{https://github.com/crsh/papaja}
+\leavevmode\hypertarget{ref-antonelli2017supersmart}{}%
+Antonelli, A., Hettling, H., Condamine, F.
+L., Vos, K., Nilsson, R.
+H., Sanderson, M.
+J., \ldots{} Vos, R.
+A.
+(2017).
+Toward a self-updating platform for estimating rates of speciation and migration, ages, and relationships of Taxa.
+\emph{Systematic Biology}, \emph{66}(2), 153--166.
+\url{https://doi.org/10.1093/sysbio/syw066}
 
@@
 \url{https://doi.org/10.1111/j.2041-210X.2012.00223.x}
+
+\leavevmode\hypertarget{ref-barba2018constraining}{}%
+Barba-Montoya, J., Reis, M.
+dos, Schneider, H., Donoghue, P.
+C., \& Yang, Z.
+(2018).
+Constraining uncertainty in the timescale of angiosperm evolution and the veracity of a cretaceous terrestrial revolution.
+\emph{New Phytologist}, \emph{218}(2), 819--834.
 
@@
 \url{https://doi.org/10.1186/1471-2105-14-16}
+
+\leavevmode\hypertarget{ref-Britton2007}{}%
+Britton, T., Anderson, C.
+L., Jacquet, D., Lundqvist, S., \& Bremer, K.
+(2007).
+Estimating Divergence Times in Large Phylogenetic Trees.
+\emph{Systematic Biology}, \emph{56}(788777878), 741--752.
+\url{https://doi.org/10.1080/10635150701613783}
 
@@
 
+\leavevmode\hypertarget{ref-Kumar2017}{}%
+Kumar, S., Stecher, G., Suleski, M., \& Hedges, S.
+B.
+(2017).
+TimeTree: A Resource for Timelines, Timetrees, and Divergence Times.
+\emph{Molecular Biology and Evolution}, \emph{34}(7), 1812--1819.
+\url{https://doi.org/10.1093/molbev/msx116}
+
+\leavevmode\hypertarget{ref-magallon2015metacalibrated}{}%
+Magallón, S., Gómez-Acevedo, S., Sánchez-Reyes, L.
+L., \& Hernández-Hernández, T.
+(2015).
+A metacalibrated time-tree documents the early rise of flowering plant phylogenetic diversity.
+\emph{New Phytologist}, \emph{207}(2), 437--453.
+
+\leavevmode\hypertarget{ref-mctavish2015phylesystem}{}%
+McTavish, E.
+J., Hinchliff, C.
+E., Allman, J.
+F., Brown, J.
+W., Cranston, K.
+A., Holder, M.
+T., \ldots{} Smith, S.
+A.
+(2015).
+Phylesystem: A git-based data store for community-curated phylogenetic estimates.
+\emph{Bioinformatics}, \emph{31}(17), 2794--2800.
+
 \leavevmode\hypertarget{ref-Michonneau2016}{}%
@@
 
+\leavevmode\hypertarget{ref-rabosky2018inverse}{}%
+Rabosky, D.
+L., Chang, J., Title, P.
+O., Cowman, P.
+F., Sallan, L., Friedman, M., \ldots{} others.
+(2018).
+An inverse latitudinal gradient in speciation rate for marine fishes.
+\emph{Nature}, \emph{559}(7714), 392.
+
+\leavevmode\hypertarget{ref-ramshaw1972time}{}%
+Ramshaw, J., Richardson, D., Meatyard, B., Brown, R., Richardson, M., Thompson, E., \& Boulter, D.
+(1972).
+The time of origin of the flowering plants determined by using amino acid sequence data of cytochrome c.
+\emph{New Phytologist}, \emph{71}(5), 773--779.
+
 \leavevmode\hypertarget{ref-RCoreTeam2018}{}%
 R Core Team. (2018). \emph{R: a language and environment for statistical computing}. Vienna, Austria: R Foundation for Statistical Computing.
-
-\leavevmode\hypertarget{ref-R-base}{}%
-R Core Team. (2021). \emph{R: A language and environment for statistical computing}. Vienna, Austria: R Foundation for Statistical Computing. Retrieved from \url{https://www.R-project.org/}
 
@@
 
+\leavevmode\hypertarget{ref-sanderson2003r8s}{}%
+Sanderson, M. J. (2003). R8s: Inferring absolute rates of molecular evolution and divergence times in the absence of a molecular clock. \emph{Bioinformatics}, \emph{19}(2), 301--302.
+
+\leavevmode\hypertarget{ref-sanderson2001sources}{}%
+Sanderson, M.
+J., \& Doyle, J.
+A.
+(2001).
+Sources of error and confidence intervals in estimating the age of angiosperms from rbcL and 18S rDNA data.
+\emph{American Journal of Botany}, \emph{88}(8), 1499--1516.
+
+\leavevmode\hypertarget{ref-schenk2016sec}{}%
+Schenk, J. J. (2016). Consequences of secondary calibrations on divergence time estimates. \emph{PLoS ONE}, \emph{11}(1). \url{https://doi.org/10.1371/journal.pone.0148228}
+
+\leavevmode\hypertarget{ref-smith2018constructing}{}%
+Smith, S. A., \& Brown, J. W. (2018). Constructing a broadly inclusive seed plant phylogeny. \emph{American Journal of Botany}, \emph{105}(3), 302--314.
+
+\leavevmode\hypertarget{ref-title2016macrophylogenies}{}%
+Title, P.
+O., \& Rabosky, D.
+L.
+(2016).
+Do Macrophylogenies Yield Stable Macroevolutionary Inferences?
+An Example from Squamate Reptiles.
+\emph{Systematic Biology}, syw102.
+\url{https://doi.org/10.1093/sysbio/syw102}
+
+\leavevmode\hypertarget{ref-Webb2008}{}%
+Webb, C.
+O., Ackerly, D.
+D., \& Kembel, S.
+W.
+(2008).
+Phylocom: Software for the analysis of phylogenetic community structure and trait evolution.
+\emph{Bioinformatics}, \emph{24}(18), 2098--2100.
+\url{https://doi.org/10.1093/bioinformatics/btn358}
+
 \endgroup

### 2021-11-02 manuscript/paper_3-main.Rmd
implementation

@@
 
-The main goal of `datelife` is to make expert information on time of lineage divergence easily accesible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
-It is a very fast tool that fulfills the quality of openness and does not require any expert biological knowledge from users --besides the names of the organisms they want to work with-- for any of its functionalities.
-However, it has many flaws.
-Some of them can be overcome, some of them might represent limitations.
-
-Up to the time this manuscript was written, `datelife`'s chronogram database had `r length(opentree_chronograms$trees)` chronograms, pulled entirely from OToL's tree repository, the only public tree repository from where `datelife` can currently get chronograms to construct its database.
-This represents `r round(length(opentree_chronograms$trees)*100/3987, digits = 2)`% of the largest existing chronogram database, TimeTree, which has a collection of 3,998 chronograms as of `r format(sys.Date(), format="%B %d %Y")`.
-Unforutnately, TimeTree's database is not open for scientific reuse nor automatised data mining [@Kumar2017].
-In 2015, a synthetic chronogram was constructed from 2,274 chronograms available at the time on the TimeTree database [@Hedges2015].
-This is the only synthetic TimeTree chronogram that has been made publicly available and deposited on the OToL repository, and is part of datelife's database now.
-Hence, the amount of lineages represented in datelife's database is at least as substantial as TimeTree's, ensuring that some information will be available for any given taxon or lineage.
-Regrettably, this does not ensure that the full state of knowledge of time of divergence of the taxon/lineage will be available.
-<!--
-Even though datelifes chronograms cover this many species from the tree of life compared to this many covered by TimeTree (database or synthetic chronogram?)-->
-Incorporation of more published chronograms into `datelife`'s database is crucial to improve its services.
-One option to increase our database is the Dryad data repository.
-Methods to automatically mine chronograms from Dryad could be designed and implemented.
-However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined on a second step, by a curator.
-<!-- Can we mine journal databases? -->
-<!-- At the same time, the prospect is to increase awereness and keep inviting and doing work encouraging scientists to make their data sets available in OToL's database. -->
-Consequently, we would like to emphasize on the importance of sharing chronogram data for the benefit of the scientific community as a whole, into repositories that require expert input and manual curation, such as OToL's tree repository [@mctavish2015phylesystem].
-
-Another potential concern comes from summary chronograms.
-We currently summarize by default all source chronograms that overlap with at least two taxa.
-Users can subset source data if they have reasons to choose some source chronograms over others.
-Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely.
-To our knowledge, there is no objective way to determine if an expert chronogram is better than another.
-Some criteria that have been put forward are the level of lineage sampling and the number of calibrations used.
-Scientists usually also favor chronograms constructed using primary calibrations (ages obtained from the fossil or geological record) to ones constructed with secondary calibrations (ages coming from other chronograms).
-It has been observed with simulations that divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with bayesian inference methods when priors are implemented in similar ways in both analyses [@schenk2016sec].
-Yet, there are different ways to use secondary calibrations
-<!-- give some examples or alternative ways of using calibrations-->
-and that same bias might not be encountered with dating methods that do not require setting priors, i.e., Maximum Likelihood methods such as r8s [@sanderson2003r8s].
-Certainly, further studies are required to fully understand the effect of using secondary calibrations on time estimates and downstream anlyses.
-
-Furthermore, even chronograms obtained with primary fossil data can show substantial variation in time estimates between clades, as observed from the comparison of source chronograms in the Fringillidae example.
-This observation is often encountered in the literature
-<!-- Different source chronograms often show substantial variation in clade ages  -->
-(see, for example, the ongoing debate about crown group age of angiosperms [@barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time].
-For some studies, especially ones based on branch lengths (e.g., studies of species diversification, timing of evolutionary events, phenotypic trait evolution), using a different chronogram may return different results [@title2016macrophylogenies].
-Stitching together these chronograms can create a larger tree that uses information from multiple studies, but the effect of uncertainties and errors here on downstream analyses is still largely unknown.
-
-Summarizing chronograms might also imply summarizing fundamentally distinct evolutionary hypotheses.
-For example, two different researchers working on the same clade both carefully select and argument their choices of fossil calibrations.
-Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses teh same one to calibrate outside the clade, the resulting age estimates will probably differ substantially (the placement of calibrations is proved to deeply affect estimated times of lineage divergence).
-Trying to summarize the resulting chronograms into a single one using simple summary statistics might erase all types of relevant information from the source chronograms.
-Accordingly, the prevailing view in our research community is that we should favor time of lineage divergence estimates obtained from a single analysis, using fossil data as primary sources of calibrations, and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history [@antonelli2017supersmart].
-<!-- Will this be implemented in future `datelife` versions?
-Probably not-->
-However, the exercise of summarizing different chronograms ha sthe potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypothesis.
-Choosing the elements of the chronograms that we are going tp keep and the ones that we are going to discard is key, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram.
-<!-- Example -->
-
-<!-- still requires more research. -->
-<!-- Often, there is a large discrepancy in time of lineage divergence estimates across studies from expert knowledge, that is well known and documented for different groups of organisms [e.g., angiosperms; @magallon2015metacalibrated].
--->
-<!-- Comparison of available age data for a wide range of organisms shown here suggest that this is a widespread phenomenon that requires further attention. -->
-<!-- In the case of the example presented here, most chronograms are new, meaning that we expect that data was generated with high quality and the software used should be very advanced.
-However, this is just an assumption.
--->
-
-Alternatively, one could try to choose the "best" chronogram from a set of possible evolutionary hypotheses.
-Several characteristics of the data used for dating analyses as well as from the output chronogram itself, could be used to score quality of source chronograms.
-Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
-To facilitate subsetting of source chronograms following different criteria by the users, this information should be included as metadata manually entered by curators in the future.
-
-<!-- Still, even if all source chronograms have been generated by excellent standards and using similar methods, the evolutionary history they depict might be very different.  -->
-
-In other areas of biological research, such as ecology and conservation biology, it has been shown that at least some data on lineage divergence represents a relevant improvement for testing alternative hypothesis using phylogenetic distance [@Webb2008].
-Hence, we integrated into datelife's workflow different ways of creating branch lengths in the absence of starting branch length information for taxa lacking this information (BLADJ option).
-Making up branch lengths in this or other ways is accepted in scientific publications: @Jetz2012, created a time-calibrated tree of all 9,993 bird species, where 67% had molecular data adn the rest was simulated; @rabosky2018inverse created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a tree of 353,185 seed plants where only 23% had molecular data.
-Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch imputation methods.
-There has yet to be a thorough analysis of what can go wrong when one goes beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make sure results are substantially similar between trees fully reconstructed with molecular or other data, and trees that are reconstructed using taxonomy by resolving polytomies at random following a statistical model.
-
-<!-- There is evidence that different chronograms will result in very different macroevolutionary histories inferred from them [@title2016macrophylogenies].
-However, it is unknown how using very different age data could affect hypothesis testing in other areas of research and more effort is needed in that directions -->
-
 \begin{center}
@@
 \end{center}
-
-Divergence time information is key to many areas of evolutionary studies: trait evolution,
-diversification, biogeography, macroecology and more. It is also crucial for science communication and education, but generating chronograms _de novo_ is difficult,
-especially for those who want to use phylogenies but who are not systematists, or
-do not have the time to acquire and develop the necessary knowledge and data curation skills.
-Moreover, years of primarily public funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
-
-<!--Mention again uses shown here and outline potential uses not shown here-->
-`datelife` allows easy and fast summarization of publicly available information
-on time of lineage divergence.
-This provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identification of regions that require more research or that have conflicting information.
-<!-- It can also summarize public data, or use it to generate new chronograms based on trees provided by the user. -->
-Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
-`datelife` helps with awareness of the existing variation in expert time of divergence data, and will foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
 

### 2021-11-02 manuscript/manuscript.Rmd
figures

@@
 keywords          : "Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations"
-wordcount         : "2628"
+wordcount         : "2650"
 
@@
 classoption       : "man"
-output            : papaja::apa6_pdf
+output            : 
+  papaja::apa6_pdf:
+    includes      :
+      # in_header: preamble.tex
+      after_body  : [figures.tex]
 
@@
 Source chronograms with at least two matching input taxon names are singled out and pruned down to preserve only input taxon names in the tips of the chronograms.
-<!-- The resulting pruned source chronograms are hereafter referred as source chronograms.  --> <!-- This was generating some confusion, so I'm calling them pruned source chronograms now-->
 Then, each pruned source chronogram is transformed to a patristic distance matrix.

### 2021-11-02 manuscript/manuscript.tex
figures

@@
 \shorttitle{DateLife: revealing the dated Tree of Life}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations\newline\indent Word count: 2628}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations\newline\indent Word count: 2650}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 
+% in this file I leave the figure captions outside\ caption{} because I want them
+to be formatted in the same way as the general text (double spaced and linenumbered)
+\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}
+
+\begin{figure}[!h]
+\includegraphics{../figures/Fig1.pdf}
+\caption{Stylized DateLife workflow.
+This shows the general workflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{www.datelife.org/query/}.
+Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.}
+\label{fig:workflow}
+\end{figure}
+% \begin{center}
+% \textsc{Figure \ref{fig:workflow}}
+% \end{center}
+% Stylized DateLife workflow.
+This shows the general workflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{http://www.datelife.org/}.
+Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
+\newpage
+\begin{figure}[!h]
+\includegraphics[width=1\linewidth]{../figures/fig_runtime_main.pdf}
+\caption{Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names.
+We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray).
+We also performed a search using the already processed query for comparison (light blue).}
+\label{fig:runtime1}
+\end{figure}
+% \begin{center}
+% \textsc{Figure \ref{fig:runtime1}}
+% \end{center}
+%Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names.
+We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray).
+We also performed a search using the already processed query for comparison (light blue).
+\newpage
+\begin{figure}[!h]
+\includegraphics[width=1\linewidth]{../figures/fig_schronograms1.pdf}
+\caption{Lineage through time (LTT) plots of source chronograms containing all or a subset of species from the bird family Fringillidae of true finches.
+Arrows indicate maximum age of each chronogram.
+Numbers reference to chronograms' original publications 1: Barker et al.
+(\protect\hyperlink{ref-barker2012going}{2012}), 2: Barker et al.
+(\protect\hyperlink{ref-barker2015new}{2015}), 3: Burns et al.
+(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), 4: Claramunt and Cracraft (\protect\hyperlink{ref-claramunt2015new}{2015}), 5: Gibb et al.
+(\protect\hyperlink{ref-gibb2015new}{2015}), 6: Hedges et al.
+(\protect\hyperlink{ref-Hedges2015}{2015}), 7: Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), 8: Jetz et al.
+(\protect\hyperlink{ref-Jetz2012}{2012}), 9: Price et al.
+(\protect\hyperlink{ref-price2014niche}{2014}).
+}
+\label{fig:schronograms1}
+\end{figure}
+% \begin{center}
+% \textsc{Figure \ref{fig:schronograms1}}
+% \end{center}
+%Lineage through time (LTT) plots of source chronograms containing all or a subset of species from the bird family Fringillidae of true finches.
+Arrows indicate maximum age of each chronogram.
+Numbers reference to chronograms' original publications 1: Barker et al.
+(\protect\hyperlink{ref-barker2012going}{2012}), 2: Barker et al.
+(\protect\hyperlink{ref-barker2015new}{2015}), 3: Burns et al.
+(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), 4: Claramunt and Cracraft (\protect\hyperlink{ref-claramunt2015new}{2015}), 5: Gibb et al.
+(\protect\hyperlink{ref-gibb2015new}{2015}), 6: Hedges et al.
+(\protect\hyperlink{ref-Hedges2015}{2015}), 7: Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), 8: Jetz et al.
+(\protect\hyperlink{ref-Jetz2012}{2012}), 9: Price et al.
+(\protect\hyperlink{ref-price2014niche}{2014}).
+\newpage
+\begin{figure}[!h]
+\includegraphics{../figures/fig_summaries.pdf}
+\caption{LTT plots of median (top) and Supermatrix Distance Method (SDM; bottom) chronograms summarising information from source chronograms found for the Fringillidae.
+Arrows indicate tree maximum age.}
+\label{fig:summaries}
+\end{figure}
+% \begin{center}
+% \textsc{Figure \ref{fig:summaries}}
+% \end{center}
+% LTT plots of median (top) and Supermatrix Distance Method (SDM; bottom) chronograms summarising information from source chronograms found for the Fringillidae. Arrows indicate tree maximum age.
+\newpage
+\begin{figure}[!h]
+\includegraphics{../figures/fig_crossval_bladj.pdf}
+\caption{LTT plots showing results from the cross-validation analyses of trees without branch lengths dated using BLADJ.
+The dating analysis can only be performed in trees with more than 2 tips, thus excluding chronogram from study 4; its data was still used as calibration for the other source chronograms.}
+\label{fig:cvbladj}
+\end{figure}
+% \begin{center}
+% \textsc{Figure \ref{fig:cvbladj}}
+% \end{center}
+% LTT plots showing results from the cross-validation analyses of trees without branch lengths dated using BLADJ.
+The dating analysis can only be performed in trees with more than 2 tips, thus excluding chronogram from study 4; its data was still used as calibration for the other source chronograms.
+\newpage
+\begin{figure}[!h]
+\includegraphics{../figures/fig_crossval_boldsumm.pdf}
+\caption{LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8.
+We could construct a tree with branch lengths for all source chronograms.
+However, dating with PATHd8 was only successful in three source chronograms shown here.}
+\label{fig:cvbold}
+\end{figure}
+% \begin{center}
+% \textsc{Figure \ref{fig:cvbold}}
+% \end{center}
+FIGURES
+%LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8.
+We could construct a tree with branch lengths for all source chronograms.
+However, dating with PATHd8 was only successful in three source chronograms shown here.
+
+%\end{linenumbers}
+
 \end{document}

### 2021-11-02 manuscript/figures-alt.tex
figures

--- 2021-10-30 manuscript/manuscript.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-  english,
-  man]{apa6}
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life},
-  pdfauthor={Luna L. Sánchez Reyes1,2, Emily Jane McTavish1, \& Brian O'Meara2},
-  pdflang={en-EN},
-  pdfkeywords={Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage{graphicx,grffile}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-% Manuscript styling
-\usepackage{upgreek}
-\captionsetup{font=singlespacing,justification=justified}
-
-% Table formatting
-\usepackage{longtable}
-\usepackage{lscape}
-% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
-\usepackage{multirow}		% Table styling
-\usepackage{tabularx}		% Control Column width
-\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
-\usepackage{threeparttablex}            % Lets threeparttable work with longtable
-
-% Create new environments so endfloat can handle them
-% \newenvironment{ltable}
-%   {\begin{landscape}\centering\begin{threeparttable}}
-%   {\end{threeparttable}\end{landscape}}
-\newenvironment{lltable}{\begin{landscape}\centering\begin{ThreePartTable}}{\end{ThreePartTable}\end{landscape}}
-
-% Enables adjusting longtable caption width to table width
-% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
-\makeatletter
-\newcommand\LastLTentrywidth{1em}
-\newlength\longtablewidth
-\setlength{\longtablewidth}{1in}
-\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}
-
-% \setlength{\parindent}{0.5in}
-% \setlength{\parskip}{0pt plus 0pt minus 0pt}
-
-% \usepackage{etoolbox}
-\makeatletter
-\patchcmd{\HyOrg@maketitle}
-  {\section{\normalfont\normalsize\abstractname}}
-  {\section*{\normalfont\normalsize\abstractname}}
-  {}{\typeout{Failed to patch abstract.}}
-\patchcmd{\HyOrg@maketitle}
-  {\section{\protect\normalfont{\@title}}}
-  {\section*{\protect\normalfont{\@title}}}
-  {}{\typeout{Failed to patch title.}}
-\makeatother
-\shorttitle{DateLife: revealing the dated Tree of Life}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations\newline\indent Word count: 55}
-\DeclareDelayedFloatFlavor{ThreePartTable}{table}
-\DeclareDelayedFloatFlavor{lltable}{table}
-\DeclareDelayedFloatFlavor*{longtable}{table}
-\makeatletter
-\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
-\makeatother
-\usepackage{lineno}
-
-\linenumbers
-\usepackage{csquotes}
-\ifxetex
-  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
-  \usepackage{polyglossia}
-  \setmainlanguage[]{english}
-\else
-  \usepackage[shorthands=off,main=english]{babel}
-\fi
-
-\title{DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life}
-\author{Luna L. Sánchez Reyes\textsuperscript{1,2}, Emily Jane McTavish\textsuperscript{1}, \& Brian O'Meara\textsuperscript{2}}
-\date{}
-
-\authornote{
-
-School of Natural Sciences, University of California, Merced, Science and Engineering Building 1.
-
-Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 425 Hesler Biology Building, Knoxville, TN 37996, USA.
-
-The authors made the following contributions.
-Luna L.
-Sánchez Reyes: Data curation, Investigation, Software, Visualization, Validation, Writing - Original Draft Preparation, Writing - Review \& Editing; Emily Jane McTavish: Resources, Software, Writing - Review \& Editing; Brian O'Meara: Conceptualization, Funding acquisition, Methodology, Resources, Software, Supervision, Writing - Review \& Editing.
-
-Correspondence concerning this article should be addressed to Luna L. Sánchez Reyes, . E-mail: \href{mailto:sanchez.reyes.luna@gmail.com}{\nolinkurl{sanchez.reyes.luna@gmail.com}}
-
-}
-
-\affiliation{\vspace{0.5cm}\textsuperscript{1} University of California, Merced\\\textsuperscript{2} University of Tennessee, Knoxville}
-
-\abstract{
-Time of evolutionary origin is fundamental for research in the natural sciences, as well as for education, science communication and policy.
-Despite an increased availability of fossil and molecular data, and time-efficient analytical techniques, achieving a high-quality reconstruction of time of evolutionary origin as a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is still a difficult and time-consuming task for a majority of interested parties.
-Yet, the amount of published chronograms has increased significantly in the past two decades, and a non-negligeable proportion of these data have been steadily accumulating in public, open databases such as TreeBASE and Open Tree
-of Life, exposing a wealth of expertly-curated and peer-reviewed data on time of evolutionary origin in a programatic and reusable way, for a large quantity and diversity of organisms.
-This trend results from intensive and localized efforts for improving data sharing practices, as well as incentivizing open science in biology.
-Despite these trends, accessibility for time data is not that good.
-R has become a widely used element of the biological data analyst toolkit.
-Hence,to improve accessibility of time data we developed an R package that access these data and eases interaction, reanalysis and reuse of it, incorporation of these data into the evolutionary workflow.
-
-Here we present \texttt{datelife}, a service implemented as an R package and a web site
-(www.datelife.org) for increased accessibility, efficient reuse, summary and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
-
-Main results:
-1. blah
-2. blah
-3. blah
-
-Results in a general context:
-All source and summary chronograms can be saved in formats that permit easy reuse and reanalysis.
-Summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research in biology.
-How well this trees work for this purpose still needs to be tested.
-
-\texttt{datelife} will be useful to increase awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
-}
-
-\begin{document}
-\maketitle
-
-\begin{verbatim}
-## Warning in data(opentree_chronograms): data set 'opentree_chronograms' not found
-\end{verbatim}
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-\hypertarget{description}{%
-\section{Description}\label{description}}
-
-The \texttt{datelife} workflow builds off from functions in several R packages (rotl (Michonneau, Brown, \& Winter, 2016), ape (Paradis, Claude, \& Strimmer, 2004),
-geiger (Harmon, Weir, Brock, Glor, \& Challenger, 2008), paleotree (Bapst, 2012), bold (Chamberlain, 2018), phytools (Revell, 2012), taxize (Chamberlain, 2018; Chamberlain \& Szöcs, 2013), phyloch (Heibl, 2008), and phylocomr (Ooms \& Chamberlain, 2018)).
-
-The basic \texttt{datelife} workflow is shown in figure \ref{fig:workflow} and consists of:
-
-\begin{enumerate}
-\item A user providing at least two taxon names as input.
-Taxon names can be given at the species level or above.
-The user can provide them as a simple comma separated character string, or as tip labels on a tree.
-The tree can be provided in newick format (as a character string) or as a "phylo" R object, and can be with or without branch lengths.
-\item A name search is then performed across the chronogram database.
-Source trees with at least two matching input names are selected, and taxa that do not match the original query are dropped from these source trees.
-<!-- The resulting pruned source chronograms are hereafter referred as source chronograms.  --> <!--# This was generating some confussion, so I'm calling them pruned source chronograms now-->
-Finally, each pruned source chronogram is transformed to a patristic distance matrix.
-This format facilitates and greatly speeds up all further analyses and summarizing algorithms.
-For the sake of keeping track of authorship later, the matrices are associated to the citation of the original study.
-\item  At this point, the user can already obtain different summary information from their query, allowing the user to take informed decisions for the next steps of the workflow.
-Types of summary information provided are: a) all pruned source chronograms, b) age of the mrca (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of succesful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most taxa matched.
-\item  Then, source chronogram data can be used as secondary calibration points to date a tree with or without branch lengths containing some or all taxon names from the initial query.
-%<!--, a taxonomic tree-->
-\item  If there is no information available for any queried taxa, users can also create both age and phylogenetic data for this missing taxa with a variety of algorithms described below.
-\item  Finally, users can easily save all source and summary chronograms in formats that permit easy reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
-\end{enumerate}
-
-\hypertarget{benchmark}{%
-\section{Benchmark}\label{benchmark}}
-
-\texttt{datelife}'s code speed was tested on an Apple iMac
-with one 3.4 GHz Intel Core i5 processor.
-We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
-Query processing time increases roughly linearly with number of input taxon names, and
-increases considerably if the taxonomic name resolution service (TNRS; Boyle et al., 2013) is activated.
-Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
-Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig.
-\ref{fig:runtime1}).
-\texttt{datelife}'s code performance was evaluated with a set of unit tests designed and
-implemented with the R package testthat (R Core Team, 2018) that were run both locally
-with the devtools package (R Core Team, 2018), and on a public server --via
-GitHub, using the continuous integration tool Travis CI (\url{https://travis-ci.org}). At
-present, unit tests cover more than 30\% of \texttt{datelife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).
-
-\hypertarget{example}{%
-\section{Example}\label{example}}
-
-We used R (Version 4.1.0; R Core Team, 2021) and the R-package \emph{papaja} (Version 0.1.0.9997; Aust \& Barth, 2020) for all our analyses.
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-\hypertarget{conclusions}{%
-\section{Conclusions}\label{conclusions}}
-
-\hypertarget{availability}{%
-\section{Availability}\label{availability}}
-
-\texttt{datelife} is free and open source and it can be used through its current website
-\url{http://www.datelife.org/query/}, through its R package, and through Phylotastic's project web portal \url{http://phylo.cs.nmsu.edu:3000/}.
-\texttt{datelife}'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
-\texttt{datelife}'s R package stable version will be available
-for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
-using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R. Development versions
-are available from the GitHub repository (\url{https://github.com/phylotastic/datelife})
-and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.
-
-\hypertarget{supplementary-material}{%
-\section{Supplementary Material}\label{supplementary-material}}
-
-Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities are available at \href{https://github.com/LunaSare/datelifeMS1}{datelifeMS1}, \href{https://github.com/LunaSare/datelife_examples}{datelife\_examples}, and \href{https://github.com/LunaSare/datelife_benchmark}{datelife\_benchmark} repositories in LLSR's GitHub account.
-
-\hypertarget{funding}{%
-\section{Funding}\label{funding}}
-
-Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to Datelife project and DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), and the Phylotastic project Grant ABI-1458572.
-
-\hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
-
-We thank colleagues from the O'Meara Lab at the University
-of Tennesse Knoxville for suggestions, discussions and software testing.
-The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
-that led to initial work on this project.
-The team that assembled \texttt{datelife}'s first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
-The Open Tree of Life project that provides the open, metadata rich repository of
-trees used for \texttt{datelife}.
-The many scientists who publish their chronograms in an open, reusable form, and
-the scientists who curate them for deposition in the Open Tree of Life repository.
-The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.
+% in this file I leave the figure captions outside\ caption{} because I want them
+to be formatted in the same way as the general text (double spaced and linenumbered)
+\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}
 
 \newpage
+\begin{figure}[!h]
+\includegraphics{../figures/Fig1.pdf}
+\caption{}
+\label{fig:workflow}
+\end{figure}
+% \begin{center}
+% \textsc{Figure \ref{fig:workflow}}
+% \end{center}
+Stylized DateLife workflow.
+This shows the general worflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{http://www.datelife.org/}.
+Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
+\newpage
+\begin{figure}[!h]
+\includegraphics[width=1\linewidth]{../figures/fig_runtime_main.pdf}
+\caption{}
+\label{fig:runtime1}
+\end{figure}
+% \begin{center}
+% \textsc{Figure \ref{fig:runtime1}}
+% \end{center}
+Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names.
+We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray).
+We also performed a search using the already processed query for comparison (light blue).
+\newpage
+\begin{figure}[!h]
+\includegraphics[width=1\linewidth]{../figures/fig_schronograms1.pdf}
+\caption{}
+\label{fig:schronograms1}
+\end{figure}
+% \begin{center}
+% \textsc{Figure \ref{fig:schronograms1}}
+% \end{center}
+Lineage through time (LTT) plots of source chronograms containing all or a subset of species from the bird family Fringillidae of true finches.
+Arrows indicate maximum age of each chronogram.
+Numbers reference to chronograms' original publications 1: Barker et al.
+(\protect\hyperlink{ref-barker2012going}{2012}), 2: Barker et al.
+(\protect\hyperlink{ref-barker2015new}{2015}), 3: Burns et al.
+(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), 4: Claramunt and Cracraft (\protect\hyperlink{ref-claramunt2015new}{2015}), 5: Gibb et al.
+(\protect\hyperlink{ref-gibb2015new}{2015}), 6: Hedges et al.
+(\protect\hyperlink{ref-Hedges2015}{2015}), 7: Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), 8: Jetz et al.
+(\protect\hyperlink{ref-Jetz2012}{2012}), 9: Price et al.
+(\protect\hyperlink{ref-price2014niche}{2014}).
+\newpage
+\begin{figure}[!h]
+\includegraphics{../figures/fig_summaries.pdf}
+\caption{}
+\label{fig:summaries}
+\end{figure}
+% \begin{center}
+% \textsc{Figure \ref{fig:summaries}}
+% \end{center}
+LTT plots of median (top) and Supermatrix Distance Method (SDM; bottom) chronograms summarising information from source chronograms found for the Fringillidae. Arrows indicate tree maximum age.
+\newpage
+\begin{figure}[!h]
+\includegraphics{../figures/fig_crossval_bladj.pdf}
+\caption{}
+\label{fig:cvbladj}
+\end{figure}
+% \begin{center}
+% \textsc{Figure \ref{fig:cvbladj}}
+% \end{center}
+LTT plots showing results from the cross-validation analyses of trees without branch lengths dated using BLADJ.
+The dating analysis can only be performed in trees with more than 2 tips, thus excluding chronogram from study 4; its data was still used as calibration for the other source chronograms.
+\newpage
+\begin{figure}[!h]
+\includegraphics{../figures/fig_crossval_boldsumm.pdf}
+\caption{}
+\label{fig:cvbold}
+\end{figure}
+% \begin{center}
+% \textsc{Figure \ref{fig:cvbold}}
+% \end{center}
+LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8.
+We could construct a tree with branch lengths for all source chronograms.
+However, dating with PATHd8 was only successful in three source chronograms shown here.
 
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-\hypertarget{refs}{}
-\leavevmode\hypertarget{ref-R-papaja}{}%
-Aust, F., \& Barth, M. (2020). \emph{papaja: Create APA manuscripts with R Markdown}. Retrieved from \url{https://github.com/crsh/papaja}
-
-\leavevmode\hypertarget{ref-Bapst2012a}{}%
-Bapst, D.
-W.
-(2012).
-Paleotree: An R package for paleontological and phylogenetic analyses of evolution.
-\emph{Methods in Ecology and Evolution}, \emph{3}(5), 803--807.
-\url{https://doi.org/10.1111/j.2041-210X.2012.00223.x}
-
-\leavevmode\hypertarget{ref-barker2012going}{}%
-Barker, F.
-K., Burns, K.
-J., Klicka, J., Lanyon, S.
-M., \& Lovette, I.
-J.
-(2012).
-Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds.
-\emph{Systematic Biology}, \emph{62}(2), 298--320.
-
-\leavevmode\hypertarget{ref-barker2015new}{}%
-Barker, F.
-K., Burns, K.
-J., Klicka, J., Lanyon, S.
-M., \& Lovette, I.
-J.
-(2015).
-New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
-\emph{The Auk: Ornithological Advances}, \emph{132}(2), 333--348.
-
-\leavevmode\hypertarget{ref-Boyle2013}{}%
-Boyle, B., Hopkins, N., Lu, Z., Raygoza Garay, J.
-A., Mozzherin, D., Rees, T., \ldots{} Enquist, B.
-J.
-(2013).
-The taxonomic name resolution service: An online tool for automated standardization of plant names.
-\emph{BMC Bioinformatics}, \emph{14}(1).
-\url{https://doi.org/10.1186/1471-2105-14-16}
-
-\leavevmode\hypertarget{ref-burns2014phylogenetics}{}%
-Burns, K.
-J., Shultz, A.
-J., Title, P.
-O., Mason, N.
-A., Barker, F.
-K., Klicka, J., \ldots{} Lovette, I.
-J.
-(2014).
-Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds.
-\emph{Molecular Phylogenetics and Evolution}, \emph{75}, 41--77.
-
-\leavevmode\hypertarget{ref-Chamberlain2018}{}%
-Chamberlain, S. (2018). \emph{bold: Interface to Bold Systems API}. Retrieved from \url{https://CRAN.R-project.org/package=bold}
-
-\leavevmode\hypertarget{ref-Chamberlain2013}{}%
-Chamberlain, S.
-A., \& Szöcs, E.
-(2013).
-taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}.
-\emph{F1000Research}, \emph{2}(191), 1--29.
-\url{https://doi.org/10.12688/f1000research.2-191.v2}
-
-\leavevmode\hypertarget{ref-claramunt2015new}{}%
-Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{Science Advances}, \emph{1}(11), e1501005.
-
-\leavevmode\hypertarget{ref-gibb2015new}{}%
-Gibb, G.
-C., England, R., Hartig, G., McLenachan, P.
-A., Taylor Smith, B.
-L., McComish, B.
-J., \ldots{} Penny, D.
-(2015).
-New zealand passerines help clarify the diversification of major songbird lineages during the oligocene.
-\emph{Genome Biology and Evolution}, \emph{7}(11), 2983--2995.
-
-\leavevmode\hypertarget{ref-Harmon2008}{}%
-Harmon, L., Weir, J., Brock, C., Glor, R., \& Challenger, W. (2008). GEIGER: investigating evolutionary radiations. \emph{Bioinformatics}, \emph{24}, 129--131.
-
-\leavevmode\hypertarget{ref-Hedges2015}{}%
-Hedges, S.
-B., Marin, J., Suleski, M., Paymer, M., \& Kumar, S.
-(2015).
-Tree of life reveals clock-like speciation and diversification.
-\emph{Molecular Biology and Evolution}, \emph{32}(4), 835--845.
-\url{https://doi.org/10.1093/molbev/msv037}
-
-\leavevmode\hypertarget{ref-Heibl2008}{}%
-Heibl, C. (2008). \emph{PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages.} Retrieved from \url{http://www.christophheibl.de/Rpackages.html}
-
-\leavevmode\hypertarget{ref-hooper2017chromosomal}{}%
-Hooper, D. M., \& Price, T. D. (2017). Chromosomal inversion differences correlate with range overlap in passerine birds. \emph{Nature Ecology \& Evolution}, \emph{1}(10), 1526.
-
-\leavevmode\hypertarget{ref-Jetz2012}{}%
-Jetz, W., Thomas, G., Joy, J.
-J.
-B., Hartmann, K., \& Mooers, A.
-(2012).
-The global diversity of birds in space and time.
-\emph{Nature}, \emph{491}(7424), 444--448.
-\url{https://doi.org/10.1038/nature11631}
-
-\leavevmode\hypertarget{ref-Michonneau2016}{}%
-Michonneau, F., Brown, J.
-W., \& Winter, D.
-J.
-(2016).
-rotl: an R package to interact with the Open Tree of Life data.
-\emph{Methods in Ecology and Evolution}, \emph{7}(12), 1476--1481.
-\url{https://doi.org/10.1111/2041-210X.12593}
-
-\leavevmode\hypertarget{ref-Ooms2018}{}%
-Ooms, J., \& Chamberlain, S. (2018). \emph{Phylocomr: Interface to 'phylocom'}. Retrieved from \url{https://CRAN.R-project.org/package=phylocomr}
-
-\leavevmode\hypertarget{ref-Paradis2004}{}%
-Paradis, E., Claude, J., \& Strimmer, K. (2004). APE: analyses of phylogenetics and evolution in R language. \emph{Bioinformatics}, \emph{20}, 289--290.
-
-\leavevmode\hypertarget{ref-price2014niche}{}%
-Price, T.
-D., Hooper, D.
-M., Buchanan, C.
-D., Johansson, U.
-S., Tietze, D.
-T., Alström, P., \ldots{} others.
-(2014).
-Niche filling slows the diversification of himalayan songbirds.
-\emph{Nature}, \emph{509}(7499), 222.
-
-\leavevmode\hypertarget{ref-RCoreTeam2018}{}%
-R Core Team. (2018). \emph{R: a language and environment for statistical computing}. Vienna, Austria: R Foundation for Statistical Computing.
-
-\leavevmode\hypertarget{ref-R-base}{}%
-R Core Team. (2021). \emph{R: A language and environment for statistical computing}. Vienna, Austria: R Foundation for Statistical Computing. Retrieved from \url{https://www.R-project.org/}
-
-\leavevmode\hypertarget{ref-Revell2012}{}%
-Revell, L. J. (2012). Phytools: An r package for phylogenetic comparative biology (and other things). \emph{Methods in Ecology and Evolution}, \emph{3}, 217--223.
-
-\endgroup
-
-\end{document}
+% \end{linenumbers}

### 2021-11-02 manuscript/figures.tex
figures

--- 2021-10-30 manuscript/manuscript.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-  english,
-  man]{apa6}
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life},
-  pdfauthor={Luna L. Sánchez Reyes1,2, Emily Jane McTavish1, \& Brian O'Meara2},
-  pdflang={en-EN},
-  pdfkeywords={Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage{graphicx,grffile}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-% Manuscript styling
-\usepackage{upgreek}
-\captionsetup{font=singlespacing,justification=justified}
+% in this file I leave the figure captions outside\ caption{} because I want them
+to be formatted in the same way as the general text (double spaced and linenumbered)
+\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}
 
-% Table formatting
-\usepackage{longtable}
-\usepackage{lscape}
-% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
-\usepackage{multirow}		% Table styling
-\usepackage{tabularx}		% Control Column width
-\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
-\usepackage{threeparttablex}            % Lets threeparttable work with longtable
+\begin{figure}[!h]
+\includegraphics{../figures/Fig1.pdf}
+\caption{Stylized DateLife workflow.
+This shows the general workflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{www.datelife.org/query/}.
+Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.}
+\label{fig:workflow}
+\end{figure}
+% \begin{center}
+% \textsc{Figure \ref{fig:workflow}}
+% \end{center}
+% Stylized DateLife workflow.
+This shows the general workflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{http://www.datelife.org/}.
+Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
+\newpage
+\begin{figure}[!h]
+\includegraphics[width=1\linewidth]{../figures/fig_runtime_main.pdf}
+\caption{Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names.
+We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray).
+We also performed a search using the already processed query for comparison (light blue).}
+\label{fig:runtime1}
+\end{figure}
+% \begin{center}
+% \textsc{Figure \ref{fig:runtime1}}
+% \end{center}
+%Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names.
+We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray).
+We also performed a search using the already processed query for comparison (light blue).
+\newpage
+\begin{figure}[!h]
+\includegraphics[width=1\linewidth]{../figures/fig_schronograms1.pdf}
+\caption{Lineage through time (LTT) plots of source chronograms containing all or a subset of species from the bird family Fringillidae of true finches.
+Arrows indicate maximum age of each chronogram.
+Numbers reference to chronograms' original publications 1: Barker et al.
+(\protect\hyperlink{ref-barker2012going}{2012}), 2: Barker et al.
+(\protect\hyperlink{ref-barker2015new}{2015}), 3: Burns et al.
+(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), 4: Claramunt and Cracraft (\protect\hyperlink{ref-claramunt2015new}{2015}), 5: Gibb et al.
+(\protect\hyperlink{ref-gibb2015new}{2015}), 6: Hedges et al.
+(\protect\hyperlink{ref-Hedges2015}{2015}), 7: Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), 8: Jetz et al.
+(\protect\hyperlink{ref-Jetz2012}{2012}), 9: Price et al.
+(\protect\hyperlink{ref-price2014niche}{2014}).
+}
+\label{fig:schronograms1}
+\end{figure}
+% \begin{center}
+% \textsc{Figure \ref{fig:schronograms1}}
+% \end{center}
+%Lineage through time (LTT) plots of source chronograms containing all or a subset of species from the bird family Fringillidae of true finches.
+Arrows indicate maximum age of each chronogram.
+Numbers reference to chronograms' original publications 1: Barker et al.
+(\protect\hyperlink{ref-barker2012going}{2012}), 2: Barker et al.
+(\protect\hyperlink{ref-barker2015new}{2015}), 3: Burns et al.
+(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), 4: Claramunt and Cracraft (\protect\hyperlink{ref-claramunt2015new}{2015}), 5: Gibb et al.
+(\protect\hyperlink{ref-gibb2015new}{2015}), 6: Hedges et al.
+(\protect\hyperlink{ref-Hedges2015}{2015}), 7: Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), 8: Jetz et al.
+(\protect\hyperlink{ref-Jetz2012}{2012}), 9: Price et al.
+(\protect\hyperlink{ref-price2014niche}{2014}).
+\newpage
+\begin{figure}[!h]
+\includegraphics{../figures/fig_summaries.pdf}
+\caption{LTT plots of median (top) and Supermatrix Distance Method (SDM; bottom) chronograms summarising information from source chronograms found for the Fringillidae.
+Arrows indicate tree maximum age.}
+\label{fig:summaries}
+\end{figure}
+% \begin{center}
+% \textsc{Figure \ref{fig:summaries}}
+% \end{center}
+% LTT plots of median (top) and Supermatrix Distance Method (SDM; bottom) chronograms summarising information from source chronograms found for the Fringillidae. Arrows indicate tree maximum age.
+\newpage
+\begin{figure}[!h]
+\includegraphics{../figures/fig_crossval_bladj.pdf}
+\caption{LTT plots showing results from the cross-validation analyses of trees without branch lengths dated using BLADJ.
+The dating analysis can only be performed in trees with more than 2 tips, thus excluding chronogram from study 4; its data was still used as calibration for the other source chronograms.}
+\label{fig:cvbladj}
+\end{figure}
+% \begin{center}
+% \textsc{Figure \ref{fig:cvbladj}}
+% \end{center}
+% LTT plots showing results from the cross-validation analyses of trees without branch lengths dated using BLADJ.
+The dating analysis can only be performed in trees with more than 2 tips, thus excluding chronogram from study 4; its data was still used as calibration for the other source chronograms.
+\newpage
+\begin{figure}[!h]
+\includegraphics{../figures/fig_crossval_boldsumm.pdf}
+\caption{LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8.
+We could construct a tree with branch lengths for all source chronograms.
+However, dating with PATHd8 was only successful in three source chronograms shown here.}
+\label{fig:cvbold}
+\end{figure}
+% \begin{center}
+% \textsc{Figure \ref{fig:cvbold}}
+% \end{center}
+FIGURES
+%LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8.
+We could construct a tree with branch lengths for all source chronograms.
+However, dating with PATHd8 was only successful in three source chronograms shown here.
 
-% Create new environments so endfloat can handle them
-% \newenvironment{ltable}
-%   {\begin{landscape}\centering\begin{threeparttable}}
-%   {\end{threeparttable}\end{landscape}}
-\newenvironment{lltable}{\begin{landscape}\centering\begin{ThreePartTable}}{\end{ThreePartTable}\end{landscape}}
-
-% Enables adjusting longtable caption width to table width
-% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
-\makeatletter
-\newcommand\LastLTentrywidth{1em}
-\newlength\longtablewidth
-\setlength{\longtablewidth}{1in}
-\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}
-
-% \setlength{\parindent}{0.5in}
-% \setlength{\parskip}{0pt plus 0pt minus 0pt}
-
-% \usepackage{etoolbox}
-\makeatletter
-\patchcmd{\HyOrg@maketitle}
-  {\section{\normalfont\normalsize\abstractname}}
-  {\section*{\normalfont\normalsize\abstractname}}
-  {}{\typeout{Failed to patch abstract.}}
-\patchcmd{\HyOrg@maketitle}
-  {\section{\protect\normalfont{\@title}}}
-  {\section*{\protect\normalfont{\@title}}}
-  {}{\typeout{Failed to patch title.}}
-\makeatother
-\shorttitle{DateLife: revealing the dated Tree of Life}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations\newline\indent Word count: 55}
-\DeclareDelayedFloatFlavor{ThreePartTable}{table}
-\DeclareDelayedFloatFlavor{lltable}{table}
-\DeclareDelayedFloatFlavor*{longtable}{table}
-\makeatletter
-\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
-\makeatother
-\usepackage{lineno}
-
-\linenumbers
-\usepackage{csquotes}
-\ifxetex
-  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
-  \usepackage{polyglossia}
-  \setmainlanguage[]{english}
-\else
-  \usepackage[shorthands=off,main=english]{babel}
-\fi
-
-\title{DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life}
-\author{Luna L. Sánchez Reyes\textsuperscript{1,2}, Emily Jane McTavish\textsuperscript{1}, \& Brian O'Meara\textsuperscript{2}}
-\date{}
-
-\authornote{
-
-School of Natural Sciences, University of California, Merced, Science and Engineering Building 1.
-
-Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 425 Hesler Biology Building, Knoxville, TN 37996, USA.
-
-The authors made the following contributions.
-Luna L.
-Sánchez Reyes: Data curation, Investigation, Software, Visualization, Validation, Writing - Original Draft Preparation, Writing - Review \& Editing; Emily Jane McTavish: Resources, Software, Writing - Review \& Editing; Brian O'Meara: Conceptualization, Funding acquisition, Methodology, Resources, Software, Supervision, Writing - Review \& Editing.
-
-Correspondence concerning this article should be addressed to Luna L. Sánchez Reyes, . E-mail: \href{mailto:sanchez.reyes.luna@gmail.com}{\nolinkurl{sanchez.reyes.luna@gmail.com}}
-
-}
-
-\affiliation{\vspace{0.5cm}\textsuperscript{1} University of California, Merced\\\textsuperscript{2} University of Tennessee, Knoxville}
-
-\abstract{
-Time of evolutionary origin is fundamental for research in the natural sciences, as well as for education, science communication and policy.
-Despite an increased availability of fossil and molecular data, and time-efficient analytical techniques, achieving a high-quality reconstruction of time of evolutionary origin as a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is still a difficult and time-consuming task for a majority of interested parties.
-Yet, the amount of published chronograms has increased significantly in the past two decades, and a non-negligeable proportion of these data have been steadily accumulating in public, open databases such as TreeBASE and Open Tree
-of Life, exposing a wealth of expertly-curated and peer-reviewed data on time of evolutionary origin in a programatic and reusable way, for a large quantity and diversity of organisms.
-This trend results from intensive and localized efforts for improving data sharing practices, as well as incentivizing open science in biology.
-Despite these trends, accessibility for time data is not that good.
-R has become a widely used element of the biological data analyst toolkit.
-Hence,to improve accessibility of time data we developed an R package that access these data and eases interaction, reanalysis and reuse of it, incorporation of these data into the evolutionary workflow.
-
-Here we present \texttt{datelife}, a service implemented as an R package and a web site
-(www.datelife.org) for increased accessibility, efficient reuse, summary and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
-
-Main results:
-1. blah
-2. blah
-3. blah
-
-Results in a general context:
-All source and summary chronograms can be saved in formats that permit easy reuse and reanalysis.
-Summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research in biology.
-How well this trees work for this purpose still needs to be tested.
-
-\texttt{datelife} will be useful to increase awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
-}
-
-\begin{document}
-\maketitle
-
-\begin{verbatim}
-## Warning in data(opentree_chronograms): data set 'opentree_chronograms' not found
-\end{verbatim}
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-\hypertarget{description}{%
-\section{Description}\label{description}}
-
-The \texttt{datelife} workflow builds off from functions in several R packages (rotl (Michonneau, Brown, \& Winter, 2016), ape (Paradis, Claude, \& Strimmer, 2004),
-geiger (Harmon, Weir, Brock, Glor, \& Challenger, 2008), paleotree (Bapst, 2012), bold (Chamberlain, 2018), phytools (Revell, 2012), taxize (Chamberlain, 2018; Chamberlain \& Szöcs, 2013), phyloch (Heibl, 2008), and phylocomr (Ooms \& Chamberlain, 2018)).
-
-The basic \texttt{datelife} workflow is shown in figure \ref{fig:workflow} and consists of:
-
-\begin{enumerate}
-\item A user providing at least two taxon names as input.
-Taxon names can be given at the species level or above.
-The user can provide them as a simple comma separated character string, or as tip labels on a tree.
-The tree can be provided in newick format (as a character string) or as a "phylo" R object, and can be with or without branch lengths.
-\item A name search is then performed across the chronogram database.
-Source trees with at least two matching input names are selected, and taxa that do not match the original query are dropped from these source trees.
-<!-- The resulting pruned source chronograms are hereafter referred as source chronograms.  --> <!--# This was generating some confussion, so I'm calling them pruned source chronograms now-->
-Finally, each pruned source chronogram is transformed to a patristic distance matrix.
-This format facilitates and greatly speeds up all further analyses and summarizing algorithms.
-For the sake of keeping track of authorship later, the matrices are associated to the citation of the original study.
-\item  At this point, the user can already obtain different summary information from their query, allowing the user to take informed decisions for the next steps of the workflow.
-Types of summary information provided are: a) all pruned source chronograms, b) age of the mrca (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of succesful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most taxa matched.
-\item  Then, source chronogram data can be used as secondary calibration points to date a tree with or without branch lengths containing some or all taxon names from the initial query.
-%<!--, a taxonomic tree-->
-\item  If there is no information available for any queried taxa, users can also create both age and phylogenetic data for this missing taxa with a variety of algorithms described below.
-\item  Finally, users can easily save all source and summary chronograms in formats that permit easy reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
-\end{enumerate}
-
-\hypertarget{benchmark}{%
-\section{Benchmark}\label{benchmark}}
-
-\texttt{datelife}'s code speed was tested on an Apple iMac
-with one 3.4 GHz Intel Core i5 processor.
-We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
-Query processing time increases roughly linearly with number of input taxon names, and
-increases considerably if the taxonomic name resolution service (TNRS; Boyle et al., 2013) is activated.
-Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
-Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig.
-\ref{fig:runtime1}).
-\texttt{datelife}'s code performance was evaluated with a set of unit tests designed and
-implemented with the R package testthat (R Core Team, 2018) that were run both locally
-with the devtools package (R Core Team, 2018), and on a public server --via
-GitHub, using the continuous integration tool Travis CI (\url{https://travis-ci.org}). At
-present, unit tests cover more than 30\% of \texttt{datelife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).
-
-\hypertarget{example}{%
-\section{Example}\label{example}}
-
-We used R (Version 4.1.0; R Core Team, 2021) and the R-package \emph{papaja} (Version 0.1.0.9997; Aust \& Barth, 2020) for all our analyses.
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-\hypertarget{conclusions}{%
-\section{Conclusions}\label{conclusions}}
-
-\hypertarget{availability}{%
-\section{Availability}\label{availability}}
-
-\texttt{datelife} is free and open source and it can be used through its current website
-\url{http://www.datelife.org/query/}, through its R package, and through Phylotastic's project web portal \url{http://phylo.cs.nmsu.edu:3000/}.
-\texttt{datelife}'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
-\texttt{datelife}'s R package stable version will be available
-for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
-using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R. Development versions
-are available from the GitHub repository (\url{https://github.com/phylotastic/datelife})
-and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.
-
-\hypertarget{supplementary-material}{%
-\section{Supplementary Material}\label{supplementary-material}}
-
-Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities are available at \href{https://github.com/LunaSare/datelifeMS1}{datelifeMS1}, \href{https://github.com/LunaSare/datelife_examples}{datelife\_examples}, and \href{https://github.com/LunaSare/datelife_benchmark}{datelife\_benchmark} repositories in LLSR's GitHub account.
-
-\hypertarget{funding}{%
-\section{Funding}\label{funding}}
-
-Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to Datelife project and DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), and the Phylotastic project Grant ABI-1458572.
-
-\hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
-
-We thank colleagues from the O'Meara Lab at the University
-of Tennesse Knoxville for suggestions, discussions and software testing.
-The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
-that led to initial work on this project.
-The team that assembled \texttt{datelife}'s first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
-The Open Tree of Life project that provides the open, metadata rich repository of
-trees used for \texttt{datelife}.
-The many scientists who publish their chronograms in an open, reusable form, and
-the scientists who curate them for deposition in the Open Tree of Life repository.
-The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.
-
-\newpage
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-\hypertarget{refs}{}
-\leavevmode\hypertarget{ref-R-papaja}{}%
-Aust, F., \& Barth, M. (2020). \emph{papaja: Create APA manuscripts with R Markdown}. Retrieved from \url{https://github.com/crsh/papaja}
-
-\leavevmode\hypertarget{ref-Bapst2012a}{}%
-Bapst, D.
-W.
-(2012).
-Paleotree: An R package for paleontological and phylogenetic analyses of evolution.
-\emph{Methods in Ecology and Evolution}, \emph{3}(5), 803--807.
-\url{https://doi.org/10.1111/j.2041-210X.2012.00223.x}
-
-\leavevmode\hypertarget{ref-barker2012going}{}%
-Barker, F.
-K., Burns, K.
-J., Klicka, J., Lanyon, S.
-M., \& Lovette, I.
-J.
-(2012).
-Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds.
-\emph{Systematic Biology}, \emph{62}(2), 298--320.
-
-\leavevmode\hypertarget{ref-barker2015new}{}%
-Barker, F.
-K., Burns, K.
-J., Klicka, J., Lanyon, S.
-M., \& Lovette, I.
-J.
-(2015).
-New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
-\emph{The Auk: Ornithological Advances}, \emph{132}(2), 333--348.
-
-\leavevmode\hypertarget{ref-Boyle2013}{}%
-Boyle, B., Hopkins, N., Lu, Z., Raygoza Garay, J.
-A., Mozzherin, D., Rees, T., \ldots{} Enquist, B.
-J.
-(2013).
-The taxonomic name resolution service: An online tool for automated standardization of plant names.
-\emph{BMC Bioinformatics}, \emph{14}(1).
-\url{https://doi.org/10.1186/1471-2105-14-16}
-
-\leavevmode\hypertarget{ref-burns2014phylogenetics}{}%
-Burns, K.
-J., Shultz, A.
-J., Title, P.
-O., Mason, N.
-A., Barker, F.
-K., Klicka, J., \ldots{} Lovette, I.
-J.
-(2014).
-Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds.
-\emph{Molecular Phylogenetics and Evolution}, \emph{75}, 41--77.
-
-\leavevmode\hypertarget{ref-Chamberlain2018}{}%
-Chamberlain, S. (2018). \emph{bold: Interface to Bold Systems API}. Retrieved from \url{https://CRAN.R-project.org/package=bold}
-
-\leavevmode\hypertarget{ref-Chamberlain2013}{}%
-Chamberlain, S.
-A., \& Szöcs, E.
-(2013).
-taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}.
-\emph{F1000Research}, \emph{2}(191), 1--29.
-\url{https://doi.org/10.12688/f1000research.2-191.v2}
-
-\leavevmode\hypertarget{ref-claramunt2015new}{}%
-Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{Science Advances}, \emph{1}(11), e1501005.
-
-\leavevmode\hypertarget{ref-gibb2015new}{}%
-Gibb, G.
-C., England, R., Hartig, G., McLenachan, P.
-A., Taylor Smith, B.
-L., McComish, B.
-J., \ldots{} Penny, D.
-(2015).
-New zealand passerines help clarify the diversification of major songbird lineages during the oligocene.
-\emph{Genome Biology and Evolution}, \emph{7}(11), 2983--2995.
-
-\leavevmode\hypertarget{ref-Harmon2008}{}%
-Harmon, L., Weir, J., Brock, C., Glor, R., \& Challenger, W. (2008). GEIGER: investigating evolutionary radiations. \emph{Bioinformatics}, \emph{24}, 129--131.
-
-\leavevmode\hypertarget{ref-Hedges2015}{}%
-Hedges, S.
-B., Marin, J., Suleski, M., Paymer, M., \& Kumar, S.
-(2015).
-Tree of life reveals clock-like speciation and diversification.
-\emph{Molecular Biology and Evolution}, \emph{32}(4), 835--845.
-\url{https://doi.org/10.1093/molbev/msv037}
-
-\leavevmode\hypertarget{ref-Heibl2008}{}%
-Heibl, C. (2008). \emph{PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages.} Retrieved from \url{http://www.christophheibl.de/Rpackages.html}
-
-\leavevmode\hypertarget{ref-hooper2017chromosomal}{}%
-Hooper, D. M., \& Price, T. D. (2017). Chromosomal inversion differences correlate with range overlap in passerine birds. \emph{Nature Ecology \& Evolution}, \emph{1}(10), 1526.
-
-\leavevmode\hypertarget{ref-Jetz2012}{}%
-Jetz, W., Thomas, G., Joy, J.
-J.
-B., Hartmann, K., \& Mooers, A.
-(2012).
-The global diversity of birds in space and time.
-\emph{Nature}, \emph{491}(7424), 444--448.
-\url{https://doi.org/10.1038/nature11631}
-
-\leavevmode\hypertarget{ref-Michonneau2016}{}%
-Michonneau, F., Brown, J.
-W., \& Winter, D.
-J.
-(2016).
-rotl: an R package to interact with the Open Tree of Life data.
-\emph{Methods in Ecology and Evolution}, \emph{7}(12), 1476--1481.
-\url{https://doi.org/10.1111/2041-210X.12593}
-
-\leavevmode\hypertarget{ref-Ooms2018}{}%
-Ooms, J., \& Chamberlain, S. (2018). \emph{Phylocomr: Interface to 'phylocom'}. Retrieved from \url{https://CRAN.R-project.org/package=phylocomr}
-
-\leavevmode\hypertarget{ref-Paradis2004}{}%
-Paradis, E., Claude, J., \& Strimmer, K. (2004). APE: analyses of phylogenetics and evolution in R language. \emph{Bioinformatics}, \emph{20}, 289--290.
-
-\leavevmode\hypertarget{ref-price2014niche}{}%
-Price, T.
-D., Hooper, D.
-M., Buchanan, C.
-D., Johansson, U.
-S., Tietze, D.
-T., Alström, P., \ldots{} others.
-(2014).
-Niche filling slows the diversification of himalayan songbirds.
-\emph{Nature}, \emph{509}(7499), 222.
-
-\leavevmode\hypertarget{ref-RCoreTeam2018}{}%
-R Core Team. (2018). \emph{R: a language and environment for statistical computing}. Vienna, Austria: R Foundation for Statistical Computing.
-
-\leavevmode\hypertarget{ref-R-base}{}%
-R Core Team. (2021). \emph{R: A language and environment for statistical computing}. Vienna, Austria: R Foundation for Statistical Computing. Retrieved from \url{https://www.R-project.org/}
-
-\leavevmode\hypertarget{ref-Revell2012}{}%
-Revell, L. J. (2012). Phytools: An r package for phylogenetic comparative biology (and other things). \emph{Methods in Ecology and Evolution}, \emph{3}, 217--223.
-
-\endgroup
-
-\end{document}
+%\end{linenumbers}

### 2021-11-02 manuscript/manuscript.Rmd
abstract

@@
   This trend results from intensive and localized efforts for improving data sharing practices, as well as incentivizing open science in biology.
-Despite these trends, accessibility for time data is not that good.
-  R has become a widely used element of the biological data analyst toolkit.
-Hence,to improve accessibility of time data we developed an R package that access these data and eases interaction, reanalysis and reuse of it, incorporation of these data into the evolutionary workflow.
+Despite these trends, accessibility to state-of-the-art knowledge on time of evolutionary origin is still reduced.
 
-  Here we present `datelife`, a service implemented as an R package and a web site
-  (www.datelife.org) for increased accessibility, efficient reuse, summary and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
+  Here we present `datelife`, a service implemented as an R package and an Rshiny website application
+  available at www.datelife.org/query/, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
 
-  Main results:
-    1. blah
-    2. blah
-    3. blah
+  The main workflow of `datelife` is to construct a chronogram for any given combination of taxon names, by searching a local chronogram database constructed and curated from the Open Tree of Life (OpenTree), which incorporates phylogenetic data from the TreeBASE database as well.
+  We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms.
+Additionally, time data extracted from source chronograms can be usedas secondary calibration points to add branch lengths proportional to absolute time to a tree topology using alternative dating methods.
 
-  Results in a general context:
-  All source and summary chronograms can be saved in formats that permit easy reuse and reanalysis.
   Summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research in biology.
-  How well this trees work for this purpose still needs to be tested.
+How well this chronograms work for this purpose still needs to be tested.
 
-  `datelife` will be useful to increase awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
+  `datelife` will be useful to increase awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.
 
-  <!-- https://tinyurl.com/ybremelq -->
+  <!--  R has become mainstream in the biological-data analyst toolkit.
+Hence, to improve accessibility of time of evolutionary origin data we developed an R package that facilitates interaction, reanalysis and reuse of it, as well as incorporation of these data into the evolutionary analysis workflow.
+-->
 
-keywords          : "Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations"
+keywords          : "Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations"
 wordcount         : "2650"
@@
     includes      :
-      # in_header: preamble.tex
       after_body  : [figures.tex]
@@
 # Implementation/Description/Workflow
-
-The main goal of `datelife` is to generate a chronogram for any given combination of taxon names, based on expert scientific information.
 

### 2021-11-02 manuscript/manuscript.Rmd
intro

@@
 keywords          : "Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations"
-wordcount         : "2650"
+wordcount         : "2949"
 
@@
 # Introduction
+
+Time of evolutionary origin represents a fundamental piece of information for
+understanding biological processes in many areas of research, from developmental to conservation biology [@Felsenstein1985a; @Webb2000], from historical biogeography to species diversification studies [@posadas2006historical; @Morlon2014].
+The number of studies publishing phylogenies with branch lengths proportional to absolute time (hereafter chronograms) have constantly increased in number for the last two decades [@Kumar2017].
+Still, generating a chronogram is not an easy task unless you have specialized training, and a wealth of time and resources: it requires inferring a phylogenetic tree using genetic markers, obtaining independent time data from the fossil record, and understanding the placement of fossils on the tree as well as their limits for analysis.
+That is why there has been an urge for promoting and facilitating the reuse of the vast amount of phylogenetic and evolutionary time data that has been made available in publications, for the advantage of research relying on this information [@webb2005phylomatic; @Stoltzfus2013].
+
+A tool for efficient reuse of expert, published data on time of evolutionary origin should have an open and fully public chronogram database storing data in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram information as needed by the user.
+A prototype service aiming to meet this criteria was developed over a series of hackathons at the National Evolutionary Synthesis Center [@Stoltzfus2013].
+Here we present the full implementation of the `datelife` service, constituted by an R package and a web site available at www.datelife.org/query/.
+The current implementation of `datelife` performs the tasks described above.
+It features an algorithm for automatic curation and maintenance of an open database of chronograms pulled from the OpenTree public repository, methods to summarize and compare source chronograms, and new functions to visualize and graphically compare source and summary chronograms.
 

### 2021-11-02 manuscript/manuscript.tex
intro

@@
   pdflang={en-EN},
-  pdfkeywords={Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations},
+  pdfkeywords={Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations},
   hidelinks,
@@
 \shorttitle{DateLife: revealing the dated Tree of Life}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations\newline\indent Word count: 2650}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 2949}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 This trend results from intensive and localized efforts for improving data sharing practices, as well as incentivizing open science in biology.
-Despite these trends, accessibility for time data is not that good.
-R has become a widely used element of the biological data analyst toolkit.
-Hence,to improve accessibility of time data we developed an R package that access these data and eases interaction, reanalysis and reuse of it, incorporation of these data into the evolutionary workflow.
+Despite these trends, accessibility to state-of-the-art knowledge on time of evolutionary origin is still reduced.
 
-Here we present \texttt{datelife}, a service implemented as an R package and a web site
-(www.datelife.org) for increased accessibility, efficient reuse, summary and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
+Here we present \texttt{datelife}, a service implemented as an R package and an Rshiny website application
+available at www.datelife.org/query/, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
 
-Main results:
-1. blah
-2. blah
-3. blah
+The main workflow of \texttt{datelife} is to construct a chronogram for any given combination of taxon names, by searching a local chronogram database constructed and curated from the Open Tree of Life (OpenTree), which incorporates phylogenetic data from the TreeBASE database as well.
+We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms.
+Additionally, time data extracted from source chronograms can be usedas secondary calibration points to add branch lengths proportional to absolute time to a tree topology using alternative dating methods.
 
-Results in a general context:
-All source and summary chronograms can be saved in formats that permit easy reuse and reanalysis.
 Summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research in biology.
-How well this trees work for this purpose still needs to be tested.
+How well this chronograms work for this purpose still needs to be tested.
 
-\texttt{datelife} will be useful to increase awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
+\texttt{datelife} will be useful to increase awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.
 }
@@
 
+Time of evolutionary origin represents a fundamental piece of information for
+understanding biological processes in many areas of research, from developmental to conservation biology (Felsenstein, 1985; Webb, 2000), from historical biogeography to species diversification studies (Morlon, 2014; Posadas, Crisci, \& Katinas, 2006).
+The number of studies publishing phylogenies with branch lengths proportional to absolute time (hereafter chronograms) have constantly increased in number for the last two decades (Kumar, Stecher, Suleski, \& Hedges, 2017).
+Still, generating a chronogram is not an easy task unless you have specialized training, and a wealth of time and resources: it requires inferring a phylogenetic tree using genetic markers, obtaining independent time data from the fossil record, and understanding the placement of fossils on the tree as well as their limits for analysis.
+That is why there has been an urge for promoting and facilitating the reuse of the vast amount of phylogenetic and evolutionary time data that has been made available in publications, for the advantage of research relying on this information (Stoltzfus et al., 2013; Webb \& Donoghue, 2005).
+
+A tool for efficient reuse of expert, published data on time of evolutionary origin should have an open and fully public chronogram database storing data in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram information as needed by the user.
+A prototype service aiming to meet this criteria was developed over a series of hackathons at the National Evolutionary Synthesis Center (Stoltzfus et al., 2013).
+Here we present the full implementation of the \texttt{datelife} service, constituted by an R package and a web site available at www.datelife.org/query/.
+The current implementation of \texttt{datelife} performs the tasks described above.
+It features an algorithm for automatic curation and maintenance of an open database of chronograms pulled from the OpenTree public repository, methods to summarize and compare source chronograms, and new functions to visualize and graphically compare source and summary chronograms.
+
 \hypertarget{implementationdescriptionworkflow}{%
 \section{Implementation/Description/Workflow}\label{implementationdescriptionworkflow}}
-
-The main goal of \texttt{datelife} is to generate a chronogram for any given combination of taxon names, based on expert scientific information.
 
@@
 Up to the time this manuscript was written, \texttt{datelife}'s chronogram database had 231 chronograms, pulled entirely from OpenTree's tree repository, the only public tree repository from where \texttt{datelife} can currently get chronograms to construct its database.
-This represents 5.79\% of the largest existing chronogram database, TimeTree, which has a collection of 3,998 chronograms as of November 01 2021.
-Unfortunately, TimeTree's database is not open for scientific reuse nor automatised data mining (Kumar, Stecher, Suleski, \& Hedges, 2017).
+This represents 5.79\% of the largest existing chronogram database, TimeTree, which has a collection of 3,998 chronograms as of November 02 2021.
+Unfortunately, TimeTree's database is not open for scientific reuse nor automatised data mining (Kumar et al., 2017).
 In 2015, a synthetic chronogram was constructed from 2,274 chronograms available at the time on the TimeTree database (Hedges, Marin, Suleski, Paymer, \& Kumar, 2015).
@@
 
+\leavevmode\hypertarget{ref-Felsenstein1985a}{}%
+Felsenstein, J. (1985). Phylogenies and the Comparative Method. \emph{The American Naturalist}, \emph{125}(1), 1--15. Retrieved from \url{http://www.jstor.org/stable/2461605}
+
 \leavevmode\hypertarget{ref-gibb2015new}{}%
@@
 
+\leavevmode\hypertarget{ref-Morlon2014}{}%
+Morlon, H. (2014). Phylogenetic approaches for studying diversification. \emph{Ecology Letters}, \emph{17}(4), 508--525. \url{https://doi.org/10.1111/ele.12251}
+
 \leavevmode\hypertarget{ref-Ooms2018}{}%
@@
 Paradis, E., Claude, J., \& Strimmer, K. (2004). APE: analyses of phylogenetics and evolution in R language. \emph{Bioinformatics}, \emph{20}, 289--290.
+
+\leavevmode\hypertarget{ref-posadas2006historical}{}%
+Posadas, P., Crisci, J. V., \& Katinas, L. (2006). Historical biogeography: A review of its basic concepts and critical issues. \emph{Journal of Arid Environments}, \emph{66}(3), 389--403.
 
@@
 
+\leavevmode\hypertarget{ref-Stoltzfus2013}{}%
+Stoltzfus, A., Lapp, H., Matasci, N., Deus, H., Sidlauskas, B., Zmasek, C.
+M., \ldots{} Jordan, G.
+(2013).
+Phylotastic!
+Making tree-of-life knowledge accessible, reusable and convenient.
+\emph{BMC Bioinformatics}, \emph{14}.
+\url{https://doi.org/10.1186/1471-2105-14-158}
+
 \leavevmode\hypertarget{ref-title2016macrophylogenies}{}%
@@
 
+\leavevmode\hypertarget{ref-Webb2000}{}%
+Webb, C. O. (2000). Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees. \emph{The American Naturalist}, \emph{156}(2), 145--155.
+
 \leavevmode\hypertarget{ref-Webb2008}{}%
@@
 \url{https://doi.org/10.1093/bioinformatics/btn358}
+
+\leavevmode\hypertarget{ref-webb2005phylomatic}{}%
+Webb, C. O., \& Donoghue, M. J. (2005). Phylomatic: Tree assembly for applied phylogenetics. \emph{Molecular Ecology Notes}, \emph{5}(1), 181--183.
 

### 2021-11-02 manuscript/paper_3-main.Rmd
intro

@@
 
-Clade ages represent a fundamental piece of information for evolutionary
-understanding in many areas of research, from developmental to conservation biology [@Felsenstein1985a; @Webb2000], from historical biogeography to species diversification studies [@posadas2006historical; @Morlon2014].
  The primary information needed for these time estimates comes from the fossil record.
 Coupled with phylogenies with branch lengths based on molecular and/or morphological data, the time of divergence of extant and extinct lineages can be reconstructed with molecular dating methods.
-The number of studies publishing phylogenies with branch lengths proportional to geological time (hereafter chronograms) have constantly increased in number for the last two decades [@Kumar2017].
-Still, generating a chronogram is not an easy task unless you have specialized training: it requires inferring a tree, understanding what fossil data are available and their limits, and where fossils go on the tree.
-That is why there has been an urge for promoting and facilitating reuse of the vast amount of phylogenetic and time of lineage divergence data that has been generated and made available in publications, for the advantage of research relying on this information [@webb2005phylomatic; @Stoltzfus2013].
 
@@
  -->
-A tool for efficient reuse of expert, published data on time of lineage divergence should have an open and fully public chronogram database storing data in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram information as needed by the user.
-A prototype service aiming to meet this characteristics was developed over a series of hackathons at the National Evolutionary Synthesis Center [@Stoltzfus2013].
-Here we present the formal description and implementation of the `datelife` service, constituted by an R package and a web site (<http://www.datelife.org/>).
-<!-- There is still much room for improvement, and flaws and limitations are addressed below.  -->
-The current implementation of `datelife` performs the basic tasks described above.
-It features a system for maintenance of an open database of chronograms pulled from public repositories, methods to summarize and compare source chronograms, and new functions to visualize and graphically compare source and summary chronograms.
+
 <!-- , for the benefit of the broader community. -->

### 2021-11-03 manuscript/figures.tex
lab meeting edits

@@
 \includegraphics[width=1\linewidth]{../figures/fig_runtime_main.pdf}
-\caption{Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names.
-We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray).
+\caption{Input taxon name processing and chronogram database search computation time increases with number of input taxon names.
+We sampled N bird species names for each input size class, 100 times, and then performed a \texttt{datelife} search using the Taxon Names Resoultion Service (TNRS; dark gray), and without using TNRS (light gray).
 We also performed a search using the already processed query for comparison (light blue).}
@@
 % \end{center}
-%Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names.
-We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray).
-We also performed a search using the already processed query for comparison (light blue).
+% 
 \newpage

### 2021-11-03 manuscript/manuscript.Rmd
lab meeting edits

@@
 Types of summary information provided are: a) all pruned source chronograms, b) age of the MRCA (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of successful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most matching input taxon names.
-\item  Finally, time of lineage divcergence obtained from the pruned source chronograms can be used as secondary calibration points to date a tree with or without branch lengths containing some or all input taxon names.
+\item  Finally, time of lineage divergence obtained from the pruned source chronograms can be used as secondary calibration points to date a tree with or without branch lengths containing some or all input taxon names.
 %<!--, a taxonomic tree-->

### 2021-11-04 manuscript/manuscript.Rmd
some tweaks to the comparison with timetree and a typo fix

@@
 This represents `r round(length(opentree_chronograms$trees)*100/3987, digits = 2)`% of the largest existing chronogram database, TimeTree, which has a collection of 3,998 chronograms as of `r format(Sys.Date(), format="%B %d %Y")`.
-Unfortunately, TimeTree's database is not open for scientific reuse nor automatised data mining [@Kumar2017].
+Unfortunately, TimeTree's database is not open for scientific reuse nor automated data mining [@Kumar2017].
 In 2015, a synthetic chronogram was constructed from 2,274 chronograms available at the time on the TimeTree database [@Hedges2015].
-This is the only synthetic TimeTree chronogram that has been made publicly available and deposited on the OpenTree repository, and is part of datelife's database now.
-Hence, the amount of lineages represented in datelife's database is at least as substantial as TimeTree's, ensuring that some information will be available for any given taxon or lineage.
+This is the only synthetic TimeTree chronogram that has been made publicly available and deposited on the OpenTree repository, and it is part of datelife's database now.
+Hence, the amount of lineages represented in datelife's database (`r length(unique(unname(unlist(lapply(opentree_chronograms$trees, "[[", "tip.label")))))` unique terminal taxa), is at least as substantial as TimeTree's (as of the last released tree, with `r max(sapply(opentree_chronograms$trees[grepl("Hedges", names(opentree_chronograms$trees))], ape::Ntip))` taxa).
 Regrettably, this does not ensure that the full state of knowledge of time of divergence of the taxon/lineage will be available.
@@
 However, the exercise of summarizing different chronograms ha sthe potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypothesis.
-Choosing the elements of the chronograms that we are going tp keep and the ones that we are going to discard is key, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram.
+Choosing the elements of the chronograms that we are going to keep and the ones that we are going to discard is key, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram.
 <!-- Example -->

### 2021-11-04 manuscript/manuscript.Rmd
penalized likelihood rather than max likelihood in r8s

@@
 <!-- give some examples or alternative ways of using calibrations-->
-and that same bias might not be encountered with dating methods that do not require setting priors, i.e., Maximum Likelihood methods such as r8s [@sanderson2003r8s].
+and that same bias might not be encountered with dating methods that do not require setting priors, i.e., penalized likelihood methods such as r8s [@sanderson2003r8s].
 Certainly, further studies are required to fully understand the effect of using secondary calibrations on time estimates and downstream anlyses.

### 2022-01-14 manuscript/manuscript.Rmd
updating description

--- 2021-11-03 manuscript/manuscript.Rmd
@@
 
-Time of evolutionary origin represents a fundamental piece of information for
-understanding biological processes in many areas of research, from developmental to conservation biology [@Felsenstein1985a; @Webb2000], from historical biogeography to species diversification studies [@posadas2006historical; @Morlon2014].
+Time of evolutionary origin represents key knowledge for
+the study of biological processes in many areas of research, from developmental to conservation biology [@Felsenstein1985a; @Webb2000], from historical biogeography to species diversification [@posadas2006historical; @Morlon2014].
 The number of studies publishing phylogenies with branch lengths proportional to absolute time (hereafter chronograms) have constantly increased in number for the last two decades [@Kumar2017].
-Still, generating a chronogram is not an easy task unless you have specialized training, and a wealth of time and resources: it requires inferring a phylogenetic tree using genetic markers, obtaining independent time data from the fossil record, and understanding the placement of fossils on the tree as well as their limits for analysis.
-That is why there has been an urge for promoting and facilitating the reuse of the vast amount of phylogenetic and evolutionary time data that has been made available in publications, for the advantage of research relying on this information [@webb2005phylomatic; @Stoltzfus2013].
+Still, generating a chronogram is not an easy task unless you have specialized training, and a considerable monetary and time resources.
+Briefly, it requires inferring a phylogenetic tree using genetic markers, obtaining independent time data from the fossil record, and understanding the placement of fossils on the tree as well as their limits for analysis.
+Because of its importance for biological research, the amount of published chronograms has increased significantly in the past two decades, and a non-negligeable proportion of these data have been deposited in public, open databases such as TreeBASE and Open Tree
+of Life, exposing a wealth of expertly-curated and peer-reviewed data on time of evolutionary origin for a large quantity and diversity of organisms, that can be accessed in a programatic and reusable way.
+
+This is why there has been an urge for promoting and facilitating the reuse of the vast amount of phylogenetic and evolutionary time data that has been made available in publications, for the advantage of research relying on this information [@webb2005phylomatic; @Stoltzfus2013].
+ Despite these trends, accessibility to state-of-the-art knowledge on time of evolutionary origin is still reduced.
 
@@
 
-# Implementation/Description/Workflow
+# Description
 
@@
 
-The basic `datelife` workflow is shown in figure \ref{fig:workflow}, largely:
+The general `datelife` workflow is shown in figure \ref{fig:workflow}:
 
 \begin{enumerate}
-\item It starts with an input consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree.
-The tree can be provided in newick format, also as a character string, or as a "phylo" R object, and can have any type of branch lengths or none.
-\item The input taxon names are cleaned with TNRS and saved as a `datelifeQuery` object.
-If taxon names are taxonomic groups above the species level, `datelife` has two alternative behaviors.
-If the "get species from taxon" flag is active, `datelife` will retrieve all species within a higher taxon name and add the species names to the input.
-If the flag is inactive, `datelife` will drop the higher taxon names from the input.
-The cleaned input taxon names are searched across the source chronogram database.
-Source chronograms with at least two matching input taxon names are singled out and pruned down to preserve only input taxon names in the tips of the chronograms.
-Then, each pruned source chronogram is transformed to a patristic distance matrix.
+\item It starts with an input consisting of at least two taxon names, which can be provided in two different forms: as a comma separated character string, or as tip labels on a tree.
+If input is a tree, it can be provided as a newick character string [@CITE], or as a "phylo" R object [@CITE].
+The input tree is not required to have branch lengths.
+\item Input taxon names are processed and "cleaned" with the Taxonomic Name Resolution Service (TNRS) which detects, corrects and standardizes misspellings and typos, variant spellings and authorities,
+and nomenclatural synonyms to accepted taxonomic names [@Boyle2013]. TNRS increases the probability of correctly finding the input taxon names in a chronogram database.
+
+\item `datelife` currently accepts scientific names only as input, and they can belong any taxonomic group or binomial specific.
+If an input taxon name belongs to a "higher" taxonomic group, i.e., a taxon above the species level, such as genus, family, etc.), `datelife` has two alternative behaviors defined by the "get species from taxon" flag.
+If the flag is active, `datelife` retrieves all species names within the "higher" taxonomic group and adds them to the input.
+If the flag is inactive, `datelife` will drop the "higher" taxonomic names from input.
+\item The cleaned input taxon names are saved as a `datelifeQuery` object
+and used to search the chronogram database.
+\item Chronograms with at least two matching input taxon names are identified and pruned down to preserve only input taxon names as tips.
+Then, each pruned chronogram is transformed to a patristic distance matrix.
 This format facilitates and greatly speeds up all downstream analyses and summaries.
@@
 Types of summary information provided are: a) all pruned source chronograms, b) age of the MRCA (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of successful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most matching input taxon names.
-\item  Finally, time of lineage divergence obtained from the pruned source chronograms can be used as secondary calibration points to date a tree with or without branch lengths containing some or all input taxon names.
+\item  Finally, time of lineage divergence obtained from the pruned chronograms can be used as secondary calibration points to date a tree with or without branch lengths containing some or all input taxon names.
 %<!--, a taxonomic tree-->
@@
 
+We illustrate the `datelife` workflow using the family of true finches, Fringillidae as an example.
 <!--We used `r cite_r("r-references.bib")` for all our analyses.-->
@@
 
-We illustrate the `datelife` workflow using the family of true finches, Fringillidae as an example.
-To contextualize, a college educator wishes to know the state-of-the-art on time of evolutionary origin of species belonging to the true finches using `datelife`.
-One option is to go to the website at www.datelife.org and perform an interactive run. However, the educator wants the students to practice their R skills.
+A college educator wishes to obtain state-of-the-art data on time of evolutionary origin of species belonging to the true finches for their class.
+They decide to use `datelife` because they require the analysis to be reproducible.
+Students have the option to go to the website at www.datelife.org and perform an interactive run. However, the educator wants the students to practice their R skills.
 The first step is to run a higher-taxon-name `datelife` query.
@@
 
-Up to the time this manuscript was written, `datelife`'s chronogram database had `r length(opentree_chronograms$trees)` chronograms, pulled entirely from OpenTree's tree repository, the only public tree repository from where `datelife` can currently get chronograms to construct its database.
+At the time of writing of this manuscript, `datelife`'s database had `r length(opentree_chronograms$trees)` chronograms.
+The database is cuurently constructed entirely from OpenTree's tree repository, the only public tree repository from where one can currently get chronograms to construct a database.
 This represents `r round(length(opentree_chronograms$trees)*100/3987, digits = 2)`% of the largest existing chronogram database, TimeTree, which has a collection of 3,998 chronograms as of `r format(Sys.Date(), format="%B %d %Y")`.

### 2022-01-14 manuscript/manuscript.Rmd

@@
 
-At the time of writing of this manuscript, `datelife`'s database had `r length(opentree_chronograms$trees)` chronograms.
-The database is cuurently constructed entirely from OpenTree's tree repository, the only public tree repository from where one can currently get chronograms to construct a database.
+Up to the time this manuscript was written, `datelife`'s chronogram database had `r length(opentree_chronograms$trees)` chronograms, pulled entirely from OpenTree's tree repository, the only public tree repository from where `datelife` can currently get chronograms to construct its database.
 This represents `r round(length(opentree_chronograms$trees)*100/3987, digits = 2)`% of the largest existing chronogram database, TimeTree, which has a collection of 3,998 chronograms as of `r format(Sys.Date(), format="%B %d %Y")`.
-Unfortunately, TimeTree's database is not open for scientific reuse nor automatised data mining [@Kumar2017].
+Unfortunately, TimeTree's database is not open for scientific reuse nor automated data mining [@Kumar2017].
 In 2015, a synthetic chronogram was constructed from 2,274 chronograms available at the time on the TimeTree database [@Hedges2015].
-This is the only synthetic TimeTree chronogram that has been made publicly available and deposited on the OpenTree repository, and is part of datelife's database now.
-Hence, the amount of lineages represented in datelife's database is at least as substantial as TimeTree's, ensuring that some information will be available for any given taxon or lineage.
+This is the only synthetic TimeTree chronogram that has been made publicly available and deposited on the OpenTree repository, and it is part of datelife's database now.
+Hence, the amount of lineages represented in datelife's database (`r length(unique(unname(unlist(lapply(opentree_chronograms$trees, "[[", "tip.label")))))` unique terminal taxa), is at least as substantial as TimeTree's (as of the last released tree, with `r max(sapply(opentree_chronograms$trees[grepl("Hedges", names(opentree_chronograms$trees))], ape::Ntip))` taxa).
 Regrettably, this does not ensure that the full state of knowledge of time of divergence of the taxon/lineage will be available.

### 2022-01-14 manuscript/manuscript.Rmd
updating description

@@
 
-Up to the time this manuscript was written, `datelife`'s chronogram database had `r length(opentree_chronograms$trees)` chronograms, pulled entirely from OpenTree's tree repository, the only public tree repository from where `datelife` can currently get chronograms to construct its database.
+At the time of writing of this manuscript, `datelife`'s database has `r length(opentree_chronograms$trees)` chronograms, pulled entirely from OpenTree's repository, the only public tree repository from where one can currently get chronograms to construct a database.
 This represents `r round(length(opentree_chronograms$trees)*100/3987, digits = 2)`% of the largest existing chronogram database, TimeTree, which has a collection of 3,998 chronograms as of `r format(Sys.Date(), format="%B %d %Y")`.

### 2022-01-15 R/fringilidae.R
data: updating case study data

@@ -1,12 +1,37 @@
-#' Fringilidae datelifeQuery
+#' Fringillidae `datelifeQuery`
 #'
-#'
-#' @name dq
+#' @name dquery
 #' @docType data
-#' @format A datelifeQuery object
-#' @keywords datelife fringilidae tnrs
+#' @format A `datelifeQuery` object
+#' @keywords datelife fringillidae tnrs
 #' @details
 #' Generated with
-#' dq = datelife::make_datelife_query(input = "fringilidae", get_spp_from_taxon = TRUE)
-#' usethis::use_data(dq, overwrite = TRUE)
-"dq"
+#' dquery = datelife::make_datelife_query(input = "fringilidae", get_spp_from_taxon = TRUE)
+#' usethis::use_data(dquery, overwrite = TRUE)
+"dquery"
+
+
+#' Fringillidae `datelifeResult`
+#'
+#' @name dres
+#' @docType data
+#' @format A `datelifeResult` object
+#' @keywords datelife fringillidae tnrs
+#' @details
+#' Generated with
+#' dres = datelife::get_datelife_result(input = dquery, cache = opentree_chronograms)
+#' usethis::use_data(dres, overwrite = TRUE)
+"dres"
+
+
+#' Fringillidae `datelifeResultSummary`
+#'
+#' @name dsumm
+#' @docType data
+#' @format A `datelifeResultSummary` object
+#' @keywords datelife fringillidae tnrs
+#' @details
+#' Generated with
+#' dsumm = datelife::summary.datelifeResult(datelife_query = dquery, object = dres)
+#' usethis::use_data(dsumm, overwrite = TRUE)
+"dsumm"

### 2022-01-15 data-raw/fringilidae.R

@@ -1,21 +1,20 @@
-## code to prepare paper datasets goes here
+## code to prepare paper datasets and figures goes here
 
 utils::data("opentree_chronograms", package = "datelife")
 
-dq = datelife::make_datelife_query(input = "fringilidae", get_spp_from_taxon = TRUE)
-usethis::use_data(dq, overwrite = TRUE)
+dquery = datelife::make_datelife_query(input = "fringilidae", get_spp_from_taxon = TRUE)
 
-dr = datelife::get_datelife_result(input = dq, cache = opentree_chronograms)
-usethis::use_data(dr, overwrite = TRUE)
+usethis::use_data(dquery, overwrite = TRUE)
 
-summ = datelife::get_taxon_summary(datelife_query = dq, datelife_result = dr)
-usethis::use_data(summ, overwrite = TRUE)
+dres = datelife::get_datelife_result(input = dquery, cache = opentree_chronograms)
 
-phyloall = datelife::summarize_datelife_result(datelife_query = dq,
-                                    datelife_result = dr,
-                                    summary_format = "phylo_all",
-                                    taxon_summary = "none")
-usethis::use_data(phyloall, overwrite = TRUE)
+usethis::use_data(dres, overwrite = TRUE)
+
+dsumm = datelife::summary.datelifeResult(datelife_query = dquery, object = dres)
+
+usethis::use_data(dsumm, overwrite = TRUE)
+
+
 
 
 # LTT plot
@@ -38,7 +37,7 @@ cb_cols2 = c("#440154FF",
              "#F0E442",
              "#95D840FF")
 ltt_plot = datelifeplot::plot_ltt_phyloall(taxon = "Fringillidae",
-                             phy = phyloall,
+                             phy = dsumm$phyloall,
                              ltt_colors = cb_cols1,
                              tax_datedotol = NULL,
                              file_name = "ltt_plot_poster.pdf",

### 2022-01-22 data-raw/node-age-comparison.R
node age comparison figure

@@ -1,0 +1,81 @@
+# Data for node age comparison
+
+## Extract calibrations
+
+calibs_all <- datelife::extract_calibrations_phylo(input = dsumm$phylo_all,
+                                               each = FALSE)
+length(calibs_all)
+class(calibs_all)
+names(calibs_all)
+nrow(calibs_all)
+unique(calibs_all$reference)
+# There are only 9 unique studies
+
+calibs_each <- datelife::extract_calibrations_phylo(input = dsumm$phylo_all,
+                                               each = TRUE)
+
+length(calibs_each)
+class(calibs_each)
+names(calibs_each)
+sapply(calibs_each, nrow)
+sum(sapply(calibs_each, nrow))
+
+
+## Match calibrations to a given topology
+
+### Choose a topology
+
+#### Options are:
+
+dsumm$phylo_median
+dsumm$phylo_sdm
+dsumm$phylo_biggest
+names(dsumm$data_frame)
+dsumm$data_frame$Ntax
+
+dquery$cleaned_names
+
+
+rotl::tnrs_match_names("fringillidae")
+#>   search_string  unique_name approximate_match ott_id is_synonym
+#> 1  fringillidae Fringillidae             FALSE 839319      FALSE
+#>   flags number_matches
+#> 1                    1
+
+rotl::is_in_tree(ott_ids = 839319)
+#> [1] FALSE
+
+rotl::tol_subtree(ott_id = "839319")
+
+ot <- rotl::tol_induced_subtree(ott_ids = dquery$ott_ids, label_format = "name")
+
+matched <- datelife::match_all_calibrations(phy = ot, calibrations = calibs_all)
+
+class(matched)
+names(matched)
+matched$matched_calibrations
+names(matched$matched_calibrations)
+nrow(matched$matched_calibrations)
+nrow(matched$present_calibrations)
+names(matched$present_calibrations)
+write.csv(matched$matched_calibrations, file = "data-raw/fringillidae-matched-calibrations.csv")
+
+matched$matched_calibrations["58",]
+which("5e34a632bf6fe8c2400e7a20efe22e98" == matched$present_calibrations$MRCA)
+
+# necesito matchear todos los pares de nombres a los nodos de mi arbol sintetico
+
+unique(matched$matched_calibrations$reference)
+# Only chronograms from 5 studies have matched to the nodes of my `ot` synth tree
+
+unique(matched$present_calibrations$reference)
+# Out of 9 unique studies
+
[…]

### 2022-01-27 data-raw/fringilidae.R
scatterplot figure for node ages

@@ -1,16 +1,27 @@
 ## code to prepare paper datasets and figures goes here
 
+# Load the chronogram database i.e, OpenTree chronograms
+
 utils::data("opentree_chronograms", package = "datelife")
+
+# Prepare your taxon names for a datelife search
 
 dquery = datelife::make_datelife_query(input = "fringilidae", get_spp_from_taxon = TRUE)
 
+length(dquery$cleaned_names) # 289 species names in Fringilidae
+
 usethis::use_data(dquery, overwrite = TRUE)
 
-dres = datelife::get_datelife_result(input = dquery, cache = opentree_chronograms)
+# Run a datelife search: matching at least 2 species name sor mroe in the chronogram database
+
+frin_sample <- sample(x = dquery$cleaned_names, size = 15)
+dres = datelife::get_datelife_result(input = frin_sample, cache = opentree_chronograms)
 
 usethis::use_data(dres, overwrite = TRUE)
 
-dsumm = datelife::summary.datelifeResult(datelife_query = dquery, object = dres)
+# Summarize the datelife result
+
+dsumm = datelife:::summary.datelifeResult(datelife_query = dquery, object = dres)
 
 usethis::use_data(dsumm, overwrite = TRUE)
 

### 2022-01-27 manuscript/manuscript.Rmd
update datafile names

@@
 availability <- ifelse(rpackage_available, "is available", "will be available")
-load("../data/dq.rda")
-load("../data/dr.rda")
+load("../data/dquery.rda")
+load("../data/dres.rda")
 ```
@@
 This will get taxon names for all recognised species within any higher taxon.
-The Fringillidae has `r length(dq$cleaned_names)` species, according to the Open Tree of Life taxonomy.
+The Fringillidae has `r length(dquery$cleaned_names)` species, according to the Open Tree of Life taxonomy.
 Once with a curated set of query taxon names, the next step is to run a `datelife` search.
 This will find all chronograms that contain at least two queried taxon names, and will save the information on time of lineage divergence as (an R "data frame") table.
- There are `r length(dr)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dr)))` different studies (Fig. \ref{fig:schronograms1}).
+ There are `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies (Fig. \ref{fig:schronograms1}).
 <!-- In case we want to cite the studies in here, but they are already cited in the figure caption: [Fig.

### 2022-01-29 data-raw/fringilidae.R
firngillidae sample vignette

@@ -1,4 +1,4 @@
-## code to prepare paper datasets and figures goes here
+## Code to prepare paper datasets and figures goes here
 
 # Load the chronogram database i.e, OpenTree chronograms
 
@@ -12,9 +12,8 @@ length(dquery$cleaned_names) # 289 species names in Fringilidae
 
 usethis::use_data(dquery, overwrite = TRUE)
 
-# Run a datelife search: matching at least 2 species name sor mroe in the chronogram database
+# Run a datelife search: matching at least 2 species names or mroe in the chronogram database
 
-frin_sample <- sample(x = dquery$cleaned_names, size = 15)
 dres = datelife::get_datelife_result(input = frin_sample, cache = opentree_chronograms)
 
 usethis::use_data(dres, overwrite = TRUE)
@@ -27,7 +26,7 @@ usethis::use_data(dsumm, overwrite = TRUE)
 
 
 
-
+# OLD stuff:
 # LTT plot
 cb_cols1 = c("#0072B2",
              "#D55E00",

### 2022-02-04 data-raw/fringilidae.R
investigating congruification and median summary trees

@@ -12,7 +12,7 @@ length(dquery$cleaned_names) # 289 species names in Fringilidae
 
 usethis::use_data(dquery, overwrite = TRUE)
 
-# Run a datelife search: matching at least 2 species names or mroe in the chronogram database
+# Run a datelife search: matching at least 2 species names or more in the chronogram database
 
 dres = datelife::get_datelife_result(input = frin_sample, cache = opentree_chronograms)
 

### 2022-02-18 manuscript/manuscript.Rmd
working on figure 2

@@
 Types of summary information provided are: a) all pruned source chronograms, b) age of the MRCA (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of successful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most matching input taxon names.
-\item  Finally, time of lineage divergence obtained from the pruned chronograms can be used as secondary calibration points to date a tree with or without branch lengths containing some or all input taxon names.
+\item To construct summary trees from patristic distance matrices, one can use a clustering method like Neighbor Joining.
+We implemented this in the package.
+However, it is know that these methods can return toologies that are not biologically plausible (Example and citation).
+Hence, we implement
+a different workflow, in which we use a fixed topology taken from the literature or from expert
+phylogenetic information (such as the OpenTree synthetic tree), and we use the patristic distances as calibrations to date that topology (with BLADJ for now).
+TODO ADD FIGURE: mock example explaining differenc ebetween using summary ages and raw ages to calibrate a topology.
+Then show it in the biological example.
+This will show what happens if a tree (phylogenetic conflict) or node (age) do not agree.
+
+\item  Alternatively, time of lineage divergence obtained from the pruned chronograms can be used directly as secondary calibration points to date a tree with or without branch lengths containing some or all input taxon names.
 %<!--, a taxonomic tree-->

### 2022-03-03 manuscript/manuscript.Rmd
plot node ages with legend and matching colors with source chronograms
rewrite based on new figure, intro and methods

--- 2022-01-27 manuscript/manuscript.Rmd
@@
 
-Time of evolutionary origin represents key knowledge for
-the study of biological processes in many areas of research, from developmental to conservation biology [@Felsenstein1985a; @Webb2000], from historical biogeography to species diversification [@posadas2006historical; @Morlon2014].
-The number of studies publishing phylogenies with branch lengths proportional to absolute time (hereafter chronograms) have constantly increased in number for the last two decades [@Kumar2017].
-Still, generating a chronogram is not an easy task unless you have specialized training, and a considerable monetary and time resources.
-Briefly, it requires inferring a phylogenetic tree using genetic markers, obtaining independent time data from the fossil record, and understanding the placement of fossils on the tree as well as their limits for analysis.
-Because of its importance for biological research, the amount of published chronograms has increased significantly in the past two decades, and a non-negligeable proportion of these data have been deposited in public, open databases such as TreeBASE and Open Tree
-of Life, exposing a wealth of expertly-curated and peer-reviewed data on time of evolutionary origin for a large quantity and diversity of organisms, that can be accessed in a programatic and reusable way.
+Inferring time of lineage evolutionary origin from scratch is not an easy task unless you have specialized training, and non-negligible budget, human and time resources.
+Briefly, it requires obtaining and curating genetic data to generate an homology hypothesis or alignment; choosing and applying software to infer an evolutionary hypothesis in the form of a phylogeny; obtaining independent age data points from the fossil record or other suitable geologic events; placing those data points appropriately and with biological and geological understanding of their limits, on the obtained phylogeny; finally, choosing the appropriate software and model of evolution to estimate ellapsed time since divergence events on the phylogeny.
+This process produces a chronogram --i.e., a phylogeny with branch lengths proportional to absolute time, and they represent key knowledge for the study of natural processes in many areas of scientific research, from developmental to conservation biology [@Felsenstein1985a; @Webb2000], from historical biogeography to species diversification [@posadas2006historical; @Morlon2014].
+Because of their importance for biological research, the amount of published expertly-curated and peer-reviewed chronograms has increased constantly in the last two decades [@Kumar2017].
+This is why there has been an urge for promoting and facilitating the reuse of the vast amount of this state-of-the-art phylogenetic and evolutionary time data that has already been produced [@webb2005phylomatic; @Stoltzfus2013].
+<!-- This has short-term advantages for research relying on this information, and long term advantages such as the advancement of science [@].-->
+We identify that a tool for efficient reuse of state-of-the-art scientific data should have an open and fully public  database storing data in a computer readable format suitable for scientific reuse [@vos2012nexml], an automatised and programatic way of accessing the data [@Stoltzfus2013], and straightforward means of comparing and summarizing data as needed by the user [].
+The TreeBASE project served as a database for state-of-the-art phylogenies, chronograms and alignments in computer readable formats, and it supports programatic data accession [@piel2002treebase; @piel2009treebase, @treebaseAPIs] but is no longer under active development.
+<!--No need to mention SuperTreeBASE as it is basically the same as treebase but on a more stable server-->
+The Open Tree of Life project [OpenTree, @opentreeoflife2019synth] has a phylogenetic and chronogram database that is programatically accessible, but it does not yet support phylogenetic queries of age data nor chronogram summaries.
 
-This is why there has been an urge for promoting and facilitating the reuse of the vast amount of phylogenetic and evolutionary time data that has been made available in publications, for the advantage of research relying on this information [@webb2005phylomatic; @Stoltzfus2013].
- Despite these trends, accessibility to state-of-the-art knowledge on time of evolutionary origin is still reduced.
+The DateLife project was born as a prototype service aiming to provide tools for easy reuse and summary of state-of-the-art time of lineage evolutionary origin, and was developed over a series of hackathons at the National Evolutionary Synthesis Center [@Stoltzfus2013].
+Here we present the full implementation of the DateLife services, available as an R package `datelife` and an application with a graphical user interface web site at www.datelife.org/query/.
+The current implementation of the `datelife` R package features an algorithm for automatic curation and maintenance of an open database of chronograms pulled from OpenTree's open repository [@mctavish2015phylesystem], methods to summarize and compare source chronograms, and new functions to visualize and graphically compare source and summary chronograms.
 
-A tool for efficient reuse of expert, published data on time of evolutionary origin should have an open and fully public chronogram database storing data in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram information as needed by the user.
-A prototype service aiming to meet this criteria was developed over a series of hackathons at the National Evolutionary Synthesis Center [@Stoltzfus2013].
-Here we present the full implementation of the `datelife` service, constituted by an R package and a web site available at www.datelife.org/query/.
-The current implementation of `datelife` performs the tasks described above.
-It features an algorithm for automatic curation and maintenance of an open database of chronograms pulled from the OpenTree public repository, methods to summarize and compare source chronograms, and new functions to visualize and graphically compare source and summary chronograms.
-
-# Description
+# Description of the R package
 
@@
 \item It starts with an input consisting of at least two taxon names, which can be provided in two different forms: as a comma separated character string, or as tip labels on a tree.
-If input is a tree, it can be provided as a newick character string [@CITE], or as a "phylo" R object [@CITE].
+If input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004ape].
 The input tree is not required to have branch lengths.
-\item Input taxon names are processed and "cleaned" with the Taxonomic Name Resolution Service (TNRS) which detects, corrects and standardizes misspellings and typos, variant spellings and authorities,
-and nomenclatural synonyms to accepted taxonomic names [@Boyle2013]. TNRS increases the probability of correctly finding the input taxon names in a chronogram database.
+\item Input taxon names are processed with the Taxonomic Name Resolution Service [TNRS, @Boyle2013] implemented with OpenTree services [@opentreeAPIs].
+TNRS detects, corrects and standardizes misspellings and typos, variant spellings and authorities,
+and nomenclatural synonyms to standardized taxonomic names. This increases the probability of correctly finding the input taxon names in the chronogram database.
 
-\item `datelife` currently accepts scientific names only as input, and they can belong any taxonomic group or binomial specific.
-If an input taxon name belongs to a "higher" taxonomic group, i.e., a taxon above the species level, such as genus, family, etc.), `datelife` has two alternative behaviors defined by the "get species from taxon" flag.
-If the flag is active, `datelife` retrieves all species names within the "higher" taxonomic group and adds them to the input.
-If the flag is inactive, `datelife` will drop the "higher" taxonomic names from input.
-\item The cleaned input taxon names are saved as a `datelifeQuery` object
-and used to search the chronogram database.
+\item The current version of `datelife` only accepts scientific taxonomic names as input.
+Names can belong to any taxonomic group or binomial specific.
+If an input taxon name belongs to an "inclusive" taxonomic group, i.e., a taxon above the species level, such as genus, family, etc.), `datelife` has two alternative behaviors defined by the "get species from taxon" flag.
+If the flag is active, `datelife` retrieves all species names within the "inclusive" taxonomic group and adds them to the input.
+If the flag is inactive, `datelife` will drop the "inclusive" taxon names from input.
+\item The cleaned input taxon names are saved as a special R object (of a newly defined class \texttt{datelifeQuery}) that contains the processed names, the corresponding taxonomic id numbers, and the topology of theinput tree if any was provided.
+The \texttt{datelifeQuery} object is used next to search the chronogram database.
 \item Chronograms with at least two matching input taxon names are identified and pruned down to preserve only input taxon names as tips.
@@
 This format facilitates and greatly speeds up all downstream analyses and summaries.
-The matrices are associated to the citation of the original study and stored as a `datelifeResult` object.
+The matrices are associated to the citation of the original study and stored as an R object of class \texttt{datelifeResult}.
 \item  At this point, various summary data can be obtained to inform decisions for the next steps of the analysis workflow.
 Types of summary information provided are: a) all pruned source chronograms, b) age of the MRCA (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of successful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most matching input taxon names.
-\item  Finally, time of lineage divergence obtained from the pruned chronograms can be used as secondary calibration points to date a tree with or without branch lengths containing some or all input taxon names.
+\item To construct summary trees from patristic distance matrices, one can use a clustering method like Neighbor Joining (NJ).
+We implemented this in the package.
+However, as we are not impementing a model of evolution, but using time distances between nodes, NJ methods do not perform well and often return toologies that are not biologically plausible [@].
+Hence, we implement
+a different workflow, in which we use a fixed topology taken from the literature or from expert
+phylogenetic information (such as the OpenTree synthetic tree), and we use the time distances as calibrations to date that topology with BLADJ [@webb2005phylomatic].
+<!--TODO ADD FIGURE: mock example explaining difference between using summary ages and raw ages to calibrate a topology.
+Then show it in the biological example.
+This will show what happens if a tree (phylogenetic conflict) or node (age) do not agree.-->
+
+\item  Alternatively, time of lineage divergence obtained from the pruned chronograms can be used directly as secondary calibration points to date a tree with or without branch lengths containing some or all input taxon names.
 %<!--, a taxonomic tree-->
@@
 Query processing time increases roughly linearly with number of input taxon names, and
-increases considerably if the taxonomic name resolution service [TNRS; @Boyle2013] is activated.
+increases considerably if TNRS [@Boyle2013] is activated.
 <!-- Defining tnrs here again, bc we only talk about it two times, so it easy to forget it -->

### 2022-03-03 manuscript/manuscript.tex
rewrite based on new figure, intro and methods

@@
   man]{apa6}
+\title{DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life}
+\author{Luna L. Sánchez Reyes\textsuperscript{1,2}, Emily Jane McTavish\textsuperscript{1}, \& Brian O'Meara\textsuperscript{2}}
+\date{}
+
+\usepackage{amsmath,amssymb}
 \usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+\usepackage{iftex}
+\ifPDFTeX
   \usepackage[T1]{fontenc}
@@
 \urlstyle{same} % disable monospaced font for URLs
-\usepackage{graphicx,grffile}
+\usepackage{graphicx}
 \makeatletter
@@
 \fi
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 % Manuscript styling
@@
 \usepackage{csquotes}
-\ifxetex
+\ifXeTeX
   % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
@@
 \else
-  \usepackage[shorthands=off,main=english]{babel}
+  \usepackage[main=english]{babel}
+% get rid of language-specific shorthands (see #6817):
+\let\LanguageShortHands\languageshorthands
+\def\languageshorthands#1{}
 \fi
-
-\title{DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life}
-\author{Luna L. Sánchez Reyes\textsuperscript{1,2}, Emily Jane McTavish\textsuperscript{1}, \& Brian O'Meara\textsuperscript{2}}
-\date{}
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
 
@@
 
-Time of evolutionary origin represents a fundamental piece of information for
-understanding biological processes in many areas of research, from developmental to conservation biology (Felsenstein, 1985; Webb, 2000), from historical biogeography to species diversification studies (Morlon, 2014; Posadas, Crisci, \& Katinas, 2006).
-The number of studies publishing phylogenies with branch lengths proportional to absolute time (hereafter chronograms) have constantly increased in number for the last two decades (Kumar, Stecher, Suleski, \& Hedges, 2017).
-Still, generating a chronogram is not an easy task unless you have specialized training, and a wealth of time and resources: it requires inferring a phylogenetic tree using genetic markers, obtaining independent time data from the fossil record, and understanding the placement of fossils on the tree as well as their limits for analysis.
-That is why there has been an urge for promoting and facilitating the reuse of the vast amount of phylogenetic and evolutionary time data that has been made available in publications, for the advantage of research relying on this information (Stoltzfus et al., 2013; Webb \& Donoghue, 2005).
+Inferring time of lineage evolutionary origin from scratch is not an easy task unless you have specialized training, and non-negligible budget, human and time resources.
+Briefly, it requires obtaining and curating genetic data to generate an homology hypothesis or alignment; choosing and applying software to infer an evolutionary hypothesis in the form of a phylogeny; obtaining independent age data points from the fossil record or other suitable geologic events; placing those data points appropriately and with biological and geological understanding of their limits, on the obtained phylogeny; finally, choosing the appropriate software and model of evolution to estimate ellapsed time since divergence events on the phylogeny.
+This process produces a chronogram --i.e., a phylogeny with branch lengths proportional to absolute time, and they represent key knowledge for the study of natural processes in many areas of scientific research, from developmental to conservation biology (Felsenstein, 1985; Campbell O.
+Webb, 2000), from historical biogeography to species diversification (Morlon, 2014; Posadas, Crisci, \& Katinas, 2006).
+Because of their importance for biological research, the amount of published expertly-curated and peer-reviewed chronograms has increased constantly in the last two decades (Kumar, Stecher, Suleski, \& Hedges, 2017).
+This is why there has been an urge for promoting and facilitating the reuse of the vast amount of this state-of-the-art phylogenetic and evolutionary time data that has already been produced (Stoltzfus et al., 2013; Campbell O.
+Webb \& Donoghue, 2005).
+We identify that a tool for efficient reuse of state-of-the-art scientific data should have an open and fully public database storing data in a computer readable format suitable for scientific reuse (Vos et al., 2012), an automatised and programatic way of accessing the data (Stoltzfus et al., 2013), and straightforward means of comparing and summarizing data as needed by the user {[}{]}.
+The TreeBASE project served as a database for state-of-the-art phylogenies, chronograms and alignments in computer readable formats, but it does not support automatized or programatic data accession (W.
+H.
+Piel, Donoghue, \& Sanderson, 2002; W.
+Piel et al., 2009).
 
-A tool for efficient reuse of expert, published data on time of evolutionary origin should have an open and fully public chronogram database storing data in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram information as needed by the user.
-A prototype service aiming to meet this criteria was developed over a series of hackathons at the National Evolutionary Synthesis Center (Stoltzfus et al., 2013).
-Here we present the full implementation of the \texttt{datelife} service, constituted by an R package and a web site available at www.datelife.org/query/.
-The current implementation of \texttt{datelife} performs the tasks described above.
-It features an algorithm for automatic curation and maintenance of an open database of chronograms pulled from the OpenTree public repository, methods to summarize and compare source chronograms, and new functions to visualize and graphically compare source and summary chronograms.
+SuperTreeBASE?
 
-\hypertarget{implementationdescriptionworkflow}{%
-\section{Implementation/Description/Workflow}\label{implementationdescriptionworkflow}}
+The Open Tree of Life project (OpenTree, OpenTreeOfLife et al., 2019) has a phylogenetic and chronogram database that is programatically accessible, but it does not yet support phylogenetic queries of age data nor chronogram summaries.
+
+The DateLife project was born as a prototype service aiming to provide tools for easy reuse and summary of state-of-the-art time of lineage evolutionary origin, and was developed over a series of hackathons at the National Evolutionary Synthesis Center (Stoltzfus et al., 2013).
+Here we present the full implementation of the DateLife services, available as an R package \texttt{datelife} and an application with a graphical user interface web site at www.datelife.org/query/.
+The current implementation of the \texttt{datelife} R package features an algorithm for automatic curation and maintenance of an open database of chronograms pulled from OpenTree's open repository (McTavish et al., 2015), methods to summarize and compare source chronograms, and new functions to visualize and graphically compare source and summary chronograms.
+
+\hypertarget{description-of-the-r-package}{%
+\section{Description of the R package}\label{description-of-the-r-package}}
 
 The \texttt{datelife} workflow builds off of functions from several R packages (rotl (Michonneau, Brown, \& Winter, 2016), ape (Paradis, Claude, \& Strimmer, 2004),
-geiger (Harmon, Weir, Brock, Glor, \& Challenger, 2008), paleotree (Bapst, 2012), bold (Chamberlain, 2018), phytools (Revell, 2012), taxize (Chamberlain, 2018; Chamberlain \& Szöcs, 2013), phyloch (Heibl, 2008), and phylocomr (Ooms \& Chamberlain, 2018)).
+geiger (Harmon, Weir, Brock, Glor, \& Challenger, 2008), paleotree (Bapst, 2012), bold (Chamberlain et al., 2019), phytools (Revell, 2012), taxize (Chamberlain \& Szöcs, 2013; Chamberlain et al., 2019), phyloch (Heibl, 2008), and phylocomr (Ooms \& Chamberlain, 2018)).
 
-The basic \texttt{datelife} workflow is shown in figure \ref{fig:workflow}, largely:
+The general \texttt{datelife} workflow is shown in figure \ref{fig:workflow}:
 
 \begin{enumerate}
-\item It starts with an input consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree.
-The tree can be provided in newick format, also as a character string, or as a "phylo" R object, and can have any type of branch lengths or none.
-\item The input taxon names are cleaned with TNRS and saved as a `datelifeQuery` object.
-If taxon names are taxonomic groups above the species level, `datelife` has two alternative behaviors.
-If the "get species from taxon" flag is active, `datelife` will retrieve all species within a higher taxon name and add the species names to the input.
-If the flag is inactive, `datelife` will drop the higher taxon names from the input.
-The cleaned input taxon names are searched across the source chronogram database.
-Source chronograms with at least two matching input taxon names are singled out and pruned down to preserve only input taxon names in the tips of the chronograms.
-Then, each pruned source chronogram is transformed to a patristic distance matrix.
+\item It starts with an input consisting of at least two taxon names, which can be provided in two different forms: as a comma separated character string, or as tip labels on a tree.
+If input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004ape].
+The input tree is not required to have branch lengths.
+\item Input taxon names are processed with the Taxonomic Name Resolution Service [TNRS, @Boyle2013] implemented with OpenTree services [@opentreeAPIs].
+TNRS detects, corrects and standardizes misspellings and typos, variant spellings and authorities,
+and nomenclatural synonyms to standardized taxonomic names. This increases the probability of correctly finding the input taxon names in the chronogram database.
+
+\item The current version of `datelife` only accepts scientific taxonomic names as input.
+Names can belong to any taxonomic group or binomial specific.
+If an input taxon name belongs to an "inclusive" taxonomic group, i.e., a taxon above the species level, such as genus, family, etc.), `datelife` has two alternative behaviors defined by the "get species from taxon" flag.
+If the flag is active, `datelife` retrieves all species names within the "inclusive" taxonomic group and adds them to the input.
+If the flag is inactive, `datelife` will drop the "inclusive" taxon names from input.
+\item The cleaned input taxon names are saved as a special R object (of a newly defined class \texttt{datelifeQuery}) that contains the processed names, the corresponding taxonomic id numbers, and the topology of theinput tree if any was provided.
+The \texttt{datelifeQuery} object is used next to search the chronogram database.
+\item Chronograms with at least two matching input taxon names are identified and pruned down to preserve only input taxon names as tips.
+Then, each pruned chronogram is transformed to a patristic distance matrix.
 This format facilitates and greatly speeds up all downstream analyses and summaries.
-The matrices are associated to the citation of the original study and stored as a `datelifeResult` object.
+The matrices are associated to the citation of the original study and stored as an R object of class \texttt{datelifeResult}.
 \item  At this point, various summary data can be obtained to inform decisions for the next steps of the analysis workflow.
 Types of summary information provided are: a) all pruned source chronograms, b) age of the MRCA (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of successful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most matching input taxon names.
-\item  Finally, time of lineage divcergence obtained from the pruned source chronograms can be used as secondary calibration points to date a tree with or without branch lengths containing some or all input taxon names.
+\item To construct summary trees from patristic distance matrices, one can use a clustering method like Neighbor Joining (NJ).
+We implemented this in the package.
+However, as we are not impementing a model of evolution, but using time distances between nodes, NJ methods do not perform well and often return toologies that are not biologically plausible [@].
+Hence, we implement
+a different workflow, in which we use a fixed topology taken from the literature or from expert
+phylogenetic information (such as the OpenTree synthetic tree), and we use the time distances as calibrations to date that topology with BLADJ [@webb2005phylomatic].
+<!--TODO ADD FIGURE: mock example explaining difference between using summary ages and raw ages to calibrate a topology.
+Then show it in the biological example.
+This will show what happens if a tree (phylogenetic conflict) or node (age) do not agree.-->
+
+\item  Alternatively, time of lineage divergence obtained from the pruned chronograms can be used directly as secondary calibration points to date a tree with or without branch lengths containing some or all input taxon names.
 %<!--, a taxonomic tree-->
@@
 Query processing time increases roughly linearly with number of input taxon names, and
-increases considerably if the taxonomic name resolution service (TNRS; Boyle et al., 2013) is activated.
+increases considerably if TNRS (Boyle et al., 2013) is activated.
 Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
@@
 
+We illustrate the \texttt{datelife} workflow using the family of true finches, Fringillidae as an example.
+
 \hypertarget{case-study}{%
@@
 
-We illustrate the \texttt{datelife} workflow using the family of true finches, Fringillidae as an example.
-To contextualize, a college educator wishes to know the state-of-the-art on time of evolutionary origin of species belonging to the true finches using \texttt{datelife}.
-One option is to go to the website at www.datelife.org and perform an interactive run. However, the educator wants the students to practice their R skills.
+A college educator wishes to obtain state-of-the-art data on time of evolutionary origin of species belonging to the true finches for their class.
+They decide to use \texttt{datelife} because they require the analysis to be reproducible.
+Students have the option to go to the website at www.datelife.org and perform an interactive run. However, the educator wants the students to practice their R skills.
 The first step is to run a higher-taxon-name \texttt{datelife} query.
@@
 Once with a curated set of query taxon names, the next step is to run a \texttt{datelife} search.
-This will find all chronograms that contain at least two queried taxon names, and will save the information on time of lineage divergence as (an R \enquote{data frame}) table.
+This will find all chronograms that contain at least two queried taxon names, and will save the information on time of lineage divergence as (an R ``data frame'') table.
 There are 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Fig. \ref{fig:schronograms1}).
 The final step is to summarize the available information using the two alternative types of summary chronograms, median and SDM.
-As explained in the \enquote{Description} section, data from source chronograms is first summarised into a single distance matrix (using the median and the SDM method respectively) and then the available node ages are used as fixed ages over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig.
+As explained in the ``Description'' section, data from source chronograms is first summarised into a single distance matrix (using the median and the SDM method respectively) and then the available node ages are used as fixed ages over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig.
 \ref{fig:summaries}).
@@
 
-Up to the time this manuscript was written, \texttt{datelife}'s chronogram database had 231 chronograms, pulled entirely from OpenTree's tree repository, the only public tree repository from where \texttt{datelife} can currently get chronograms to construct its database.
-This represents 5.79\% of the largest existing chronogram database, TimeTree, which has a collection of 3,998 chronograms as of November 02 2021.
-Unfortunately, TimeTree's database is not open for scientific reuse nor automatised data mining (Kumar et al., 2017).
+At the time of writing of this manuscript, \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's repository, the only public tree repository from where one can currently get chronograms to construct a database.
+This represents 6.35\% of the largest existing chronogram database, TimeTree, which has a collection of 3,998 chronograms as of March 03 2022.
+Unfortunately, TimeTree's database is not open for scientific reuse nor automated data mining (Kumar et al., 2017).
 In 2015, a synthetic chronogram was constructed from 2,274 chronograms available at the time on the TimeTree database (Hedges, Marin, Suleski, Paymer, \& Kumar, 2015).
-This is the only synthetic TimeTree chronogram that has been made publicly available and deposited on the OpenTree repository, and is part of datelife's database now.
-Hence, the amount of lineages represented in datelife's database is at least as substantial as TimeTree's, ensuring that some information will be available for any given taxon or lineage.
+This is the only synthetic TimeTree chronogram that has been made publicly available and deposited on the OpenTree repository, and it is part of datelife's database now.
+Hence, the amount of lineages represented in datelife's database (99474 unique terminal taxa), is at least as substantial as TimeTree's (as of the last released tree, with 48017 taxa).
 Regrettably, this does not ensure that the full state of knowledge of time of divergence of the taxon/lineage will be available.
@@
 Yet, there are different ways to use secondary calibrations
-and that same bias might not be encountered with dating methods that do not require setting priors, i.e., Maximum Likelihood methods such as r8s (Sanderson, 2003).
+and that same bias might not be encountered with dating methods that do not require setting priors, i.e., penalized likelihood methods such as r8s (Sanderson, 2003).
 Certainly, further studies are required to fully understand the effect of using secondary calibrations on time estimates and downstream anlyses.
@@
 However, the exercise of summarizing different chronograms ha sthe potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypothesis.
-Choosing the elements of the chronograms that we are going tp keep and the ones that we are going to discard is key, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram.
+Choosing the elements of the chronograms that we are going to keep and the ones that we are going to discard is key, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram.
 
-Alternatively, one could try to choose the \enquote{best} chronogram from a set of possible evolutionary hypotheses.
+Alternatively, one could try to choose the ``best'' chronogram from a set of possible evolutionary hypotheses.
 Several characteristics of the data used for dating analyses as well as from the output chronogram itself, could be used to score quality of source chronograms.
@@
 
-In other areas of biological research, such as ecology and conservation biology, it has been shown that at least some data on lineage divergence represents a relevant improvement for testing alternative hypothesis using phylogenetic distance (Webb, Ackerly, \& Kembel, 2008).
+In other areas of biological research, such as ecology and conservation biology, it has been shown that at least some data on lineage divergence represents a relevant improvement for testing alternative hypothesis using phylogenetic distance (Campbell O.
+Webb, Ackerly, \& Kembel, 2008).
 Hence, we integrated into datelife's workflow different ways of creating branch lengths in the absence of starting branch length information for taxa lacking this information (BLADJ option).
@@
 \hypertarget{refs}{}
-\leavevmode\hypertarget{ref-antonelli2017supersmart}{}%
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-antonelli2017supersmart}{}}%
 Antonelli, A., Hettling, H., Condamine, F.
@@
 (2017).
-Toward a self-updating platform for estimating rates of speciation and migration, ages, and relationships of Taxa.
+{Toward a self-updating platform for estimating rates of speciation and migration, ages, and relationships of Taxa}.
 \emph{Systematic Biology}, \emph{66}(2), 153--166.
@@
 
-\leavevmode\hypertarget{ref-Bapst2012a}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Bapst2012a}{}}%
 Bapst, D.
@@
 (2012).
-Paleotree: An R package for paleontological and phylogenetic analyses of evolution.
+{Paleotree: An R package for paleontological and phylogenetic analyses of evolution}.
 \emph{Methods in Ecology and Evolution}, \emph{3}(5), 803--807.
@@
 
-\leavevmode\hypertarget{ref-barba2018constraining}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-barba2018constraining}{}}%
 Barba-Montoya, J., Reis, M.
@@
 
-\leavevmode\hypertarget{ref-barker2012going}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-barker2012going}{}}%
 Barker, F.
@@
 
-\leavevmode\hypertarget{ref-barker2015new}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-barker2015new}{}}%
 Barker, F.
@@
 
-\leavevmode\hypertarget{ref-Boyle2013}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Boyle2013}{}}%
 Boyle, B., Hopkins, N., Lu, Z., Raygoza Garay, J.
@@
 (2013).
-The taxonomic name resolution service: An online tool for automated standardization of plant names.
+{The taxonomic name resolution service: An online tool for automated standardization of plant names}.
 \emph{BMC Bioinformatics}, \emph{14}(1).
@@
 
-\leavevmode\hypertarget{ref-Britton2007}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Britton2007}{}}%
 Britton, T., Anderson, C.
@@
 (2007).
-Estimating Divergence Times in Large Phylogenetic Trees.
+{Estimating Divergence Times in Large Phylogenetic Trees}.
 \emph{Systematic Biology}, \emph{56}(788777878), 741--752.
@@
 
-\leavevmode\hypertarget{ref-burns2014phylogenetics}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-burns2014phylogenetics}{}}%
 Burns, K.
@@
 
-\leavevmode\hypertarget{ref-Chamberlain2018}{}%
-Chamberlain, S. (2018). \emph{bold: Interface to Bold Systems API}. Retrieved from \url{https://CRAN.R-project.org/package=bold}
-
-\leavevmode\hypertarget{ref-Chamberlain2013}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Chamberlain2013}{}}%
 Chamberlain, S.
@@
 (2013).
-taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}.
+{taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}}.
 \emph{F1000Research}, \emph{2}(191), 1--29.
@@
 
-\leavevmode\hypertarget{ref-claramunt2015new}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Chamberlain2018}{}}%
+Chamberlain, S.
+A., Szöcs, E., Foster, Z., Arendsee, Z., Boettiger, C., Ram, K., \ldots{} Li, G.
+(2019).
+\emph{{taxize: Taxonomic information from around the web}}.
+Retrieved from \url{https://github.com/ropensci/taxize}
+
+\leavevmode\vadjust pre{\hypertarget{ref-claramunt2015new}{}}%
 Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{Science Advances}, \emph{1}(11), e1501005.
 
-\leavevmode\hypertarget{ref-Felsenstein1985a}{}%
-Felsenstein, J. (1985). Phylogenies and the Comparative Method. \emph{The American Naturalist}, \emph{125}(1), 1--15. Retrieved from \url{http://www.jstor.org/stable/2461605}
+\leavevmode\vadjust pre{\hypertarget{ref-Felsenstein1985a}{}}%
+Felsenstein, J. (1985). {Phylogenies and the Comparative Method}. \emph{The American Naturalist}, \emph{125}(1), 1--15. Retrieved from \url{http://www.jstor.org/stable/2461605}
 
-\leavevmode\hypertarget{ref-gibb2015new}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-gibb2015new}{}}%
 Gibb, G.
@@
 
-\leavevmode\hypertarget{ref-Harmon2008}{}%
-Harmon, L., Weir, J., Brock, C., Glor, R., \& Challenger, W. (2008). GEIGER: investigating evolutionary radiations. \emph{Bioinformatics}, \emph{24}, 129--131.
+\leavevmode\vadjust pre{\hypertarget{ref-Harmon2008}{}}%
+Harmon, L., Weir, J., Brock, C., Glor, R., \& Challenger, W. (2008). {GEIGER: investigating evolutionary radiations}. \emph{Bioinformatics}, \emph{24}, 129--131.
 
-\leavevmode\hypertarget{ref-Hedges2015}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Hedges2015}{}}%
 Hedges, S.
@@
 (2015).
-Tree of life reveals clock-like speciation and diversification.
+{Tree of life reveals clock-like speciation and diversification}.
 \emph{Molecular Biology and Evolution}, \emph{32}(4), 835--845.
@@
 
-\leavevmode\hypertarget{ref-Heibl2008}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Heibl2008}{}}%
 Heibl, C. (2008). \emph{PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages.} Retrieved from \url{http://www.christophheibl.de/Rpackages.html}
 
-\leavevmode\hypertarget{ref-hooper2017chromosomal}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-hooper2017chromosomal}{}}%
 Hooper, D. M., \& Price, T. D. (2017). Chromosomal inversion differences correlate with range overlap in passerine birds. \emph{Nature Ecology \& Evolution}, \emph{1}(10), 1526.
 
-\leavevmode\hypertarget{ref-Jetz2012}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Jetz2012}{}}%
 Jetz, W., Thomas, G., Joy, J.
@@
 (2012).
-The global diversity of birds in space and time.
+{The global diversity of birds in space and time}.
 \emph{Nature}, \emph{491}(7424), 444--448.
@@
 
-\leavevmode\hypertarget{ref-Kumar2017}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Kumar2017}{}}%
 Kumar, S., Stecher, G., Suleski, M., \& Hedges, S.
@@
 (2017).
-TimeTree: A Resource for Timelines, Timetrees, and Divergence Times.
+{TimeTree: A Resource for Timelines, Timetrees, and Divergence Times}.
 \emph{Molecular Biology and Evolution}, \emph{34}(7), 1812--1819.
@@
 
-\leavevmode\hypertarget{ref-magallon2015metacalibrated}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-magallon2015metacalibrated}{}}%
 Magallón, S., Gómez-Acevedo, S., Sánchez-Reyes, L.
@@
 
-\leavevmode\hypertarget{ref-mctavish2015phylesystem}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-mctavish2015phylesystem}{}}%
 McTavish, E.
@@
 
-\leavevmode\hypertarget{ref-Michonneau2016}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Michonneau2016}{}}%
 Michonneau, F., Brown, J.
@@
 (2016).
-rotl: an R package to interact with the Open Tree of Life data.
+{rotl: an R package to interact with the Open Tree of Life data}.
 \emph{Methods in Ecology and Evolution}, \emph{7}(12), 1476--1481.
@@
 
-\leavevmode\hypertarget{ref-Morlon2014}{}%
-Morlon, H. (2014). Phylogenetic approaches for studying diversification. \emph{Ecology Letters}, \emph{17}(4), 508--525. \url{https://doi.org/10.1111/ele.12251}
+\leavevmode\vadjust pre{\hypertarget{ref-Morlon2014}{}}%
+Morlon, H. (2014). {Phylogenetic approaches for studying diversification.} \emph{Ecology Letters}, \emph{17}(4), 508--525. \url{https://doi.org/10.1111/ele.12251}
 
-\leavevmode\hypertarget{ref-Ooms2018}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Ooms2018}{}}%
 Ooms, J., \& Chamberlain, S. (2018). \emph{Phylocomr: Interface to 'phylocom'}. Retrieved from \url{https://CRAN.R-project.org/package=phylocomr}
 
-\leavevmode\hypertarget{ref-Paradis2004}{}%
-Paradis, E., Claude, J., \& Strimmer, K. (2004). APE: analyses of phylogenetics and evolution in R language. \emph{Bioinformatics}, \emph{20}, 289--290.
+\leavevmode\vadjust pre{\hypertarget{ref-opentreeoflife2019synth}{}}%
+OpenTreeOfLife, Redelings, B., Reyes, L.
+L.
+S., Cranston, K.
+A., Allman, J., Holder, M.
+T., \& McTavish, E.
+J.
+(2019).
+Open tree of life synthetic tree (Version 12.3).
+Zenodo.
+\url{https://doi.org/10.5281/zenodo.3937742}
 
-\leavevmode\hypertarget{ref-posadas2006historical}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Paradis2004}{}}%
+Paradis, E., Claude, J., \& Strimmer, K. (2004). {APE: analyses of phylogenetics and evolution in R language}. \emph{Bioinformatics}, \emph{20}, 289--290.
+
+\leavevmode\vadjust pre{\hypertarget{ref-piel2002treebase}{}}%
+Piel, W.
+H., Donoghue, M., \& Sanderson, M.
+(2002).
+{TreeBASE : A database of phylogenetic information}.
+In J.
+Shimura, K.
+Wilson, \& D.
+Gordon (Eds.), \emph{To the interoperable {``catalog of life''} with partners} (pp.
+41--47).
+Tsukuba, Japan: National Institute for Environmental Studies.
+Retrieved from \url{http://phylogeny.harvard.edu/treebase.}
+
+\leavevmode\vadjust pre{\hypertarget{ref-piel2009treebase}{}}%
+Piel, W., Chan, L., Dominus, M., Ruan, J., Vos, R., \& Tannen, V. (2009). Treebase v. 2: A database of phylogenetic knowledge. E-biosphere. London.
+
+\leavevmode\vadjust pre{\hypertarget{ref-posadas2006historical}{}}%
 Posadas, P., Crisci, J. V., \& Katinas, L. (2006). Historical biogeography: A review of its basic concepts and critical issues. \emph{Journal of Arid Environments}, \emph{66}(3), 389--403.
 
-\leavevmode\hypertarget{ref-price2014niche}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-price2014niche}{}}%
 Price, T.
@@
 
-\leavevmode\hypertarget{ref-rabosky2018inverse}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-RCoreTeam2018}{}}%
+R Core Team. (2018). \emph{{R: a language and environment for statistical computing}}. Vienna, Austria: R Foundation for Statistical Computing.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rabosky2018inverse}{}}%
 Rabosky, D.
@@
 
-\leavevmode\hypertarget{ref-ramshaw1972time}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-ramshaw1972time}{}}%
 Ramshaw, J., Richardson, D., Meatyard, B., Brown, R., Richardson, M., Thompson, E., \& Boulter, D.
@@
 
-\leavevmode\hypertarget{ref-RCoreTeam2018}{}%
-R Core Team. (2018). \emph{R: a language and environment for statistical computing}. Vienna, Austria: R Foundation for Statistical Computing.
-
-\leavevmode\hypertarget{ref-Revell2012}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Revell2012}{}}%
 Revell, L. J. (2012). Phytools: An r package for phylogenetic comparative biology (and other things). \emph{Methods in Ecology and Evolution}, \emph{3}, 217--223.
 
-\leavevmode\hypertarget{ref-sanderson2003r8s}{}%
-Sanderson, M. J. (2003). R8s: Inferring absolute rates of molecular evolution and divergence times in the absence of a molecular clock. \emph{Bioinformatics}, \emph{19}(2), 301--302.
+\leavevmode\vadjust pre{\hypertarget{ref-sanderson2003r8s}{}}%
+Sanderson, M. J. (2003). r8s: Inferring absolute rates of molecular evolution and divergence times in the absence of a molecular clock. \emph{Bioinformatics}, \emph{19}(2), 301--302.
 
-\leavevmode\hypertarget{ref-sanderson2001sources}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-sanderson2001sources}{}}%
 Sanderson, M.
@@
 
-\leavevmode\hypertarget{ref-schenk2016sec}{}%
-Schenk, J. J. (2016). Consequences of secondary calibrations on divergence time estimates. \emph{PLoS ONE}, \emph{11}(1). \url{https://doi.org/10.1371/journal.pone.0148228}
+\leavevmode\vadjust pre{\hypertarget{ref-schenk2016sec}{}}%
+Schenk, J. J. (2016). {Consequences of secondary calibrations on divergence time estimates}. \emph{PLoS ONE}, \emph{11}(1). \url{https://doi.org/10.1371/journal.pone.0148228}
 
-\leavevmode\hypertarget{ref-smith2018constructing}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-smith2018constructing}{}}%
 Smith, S. A., \& Brown, J. W. (2018). Constructing a broadly inclusive seed plant phylogeny. \emph{American Journal of Botany}, \emph{105}(3), 302--314.
 
-\leavevmode\hypertarget{ref-Stoltzfus2013}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Stoltzfus2013}{}}%
 Stoltzfus, A., Lapp, H., Matasci, N., Deus, H., Sidlauskas, B., Zmasek, C.
@@
 (2013).
-Phylotastic!
-Making tree-of-life knowledge accessible, reusable and convenient.
+{Phylotastic!
+Making tree-of-life knowledge accessible, reusable and convenient}.
 \emph{BMC Bioinformatics}, \emph{14}.
@@
 
-\leavevmode\hypertarget{ref-title2016macrophylogenies}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-title2016macrophylogenies}{}}%
 Title, P.
@@
 (2016).
-Do Macrophylogenies Yield Stable Macroevolutionary Inferences?
-An Example from Squamate Reptiles.
+{Do Macrophylogenies Yield Stable Macroevolutionary Inferences?
+An Example from Squamate Reptiles}.
 \emph{Systematic Biology}, syw102.
@@
 
-\leavevmode\hypertarget{ref-Webb2000}{}%
-Webb, C. O. (2000). Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees. \emph{The American Naturalist}, \emph{156}(2), 145--155.
+\leavevmode\vadjust pre{\hypertarget{ref-vos2012nexml}{}}%
+Vos, R.
+A., Balhoff, J.
+P., Caravas, J.
+A., Holder, M.
+T., Lapp, H., Maddison, W.
+P., \ldots{} others.
+(2012).
+NeXML: Rich, extensible, and verifiable representation of comparative data and metadata.
+\emph{Systematic Biology}, \emph{61}(4), 675--689.
 
-\leavevmode\hypertarget{ref-Webb2008}{}%
-Webb, C.
-O., Ackerly, D.
+\leavevmode\vadjust pre{\hypertarget{ref-Webb2000}{}}%
+Webb, Campbell O. (2000). {Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees}. \emph{The American Naturalist}, \emph{156}(2), 145--155.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Webb2008}{}}%
+Webb, Campbell O., Ackerly, D.
 D., \& Kembel, S.
@@
 (2008).
-Phylocom: Software for the analysis of phylogenetic community structure and trait evolution.
+{Phylocom: Software for the analysis of phylogenetic community structure and trait evolution}.
 \emph{Bioinformatics}, \emph{24}(18), 2098--2100.
@@
 
-\leavevmode\hypertarget{ref-webb2005phylomatic}{}%
-Webb, C. O., \& Donoghue, M. J. (2005). Phylomatic: Tree assembly for applied phylogenetics. \emph{Molecular Ecology Notes}, \emph{5}(1), 181--183.
+\leavevmode\vadjust pre{\hypertarget{ref-webb2005phylomatic}{}}%
+Webb, Campbell O., \& Donoghue, M. J. (2005). Phylomatic: Tree assembly for applied phylogenetics. \emph{Molecular Ecology Notes}, \emph{5}(1), 181--183.
+
+\end{CSLReferences}
 

### 2022-03-04 manuscript/manuscript.Rmd
adding references

@@
     includes      :
-      after_body  : [figures.tex]
+      in_header: "preamble.tex"
+      after_body  : "figures.tex"
 
@@
 data(opentree_chronograms, package = "datelife")
-rpackage_available <- FALSE
+rpackage_available <- TRUE
 availability <- ifelse(rpackage_available, "is available", "will be available")
@@
 
-Inferring time of lineage evolutionary origin from scratch is not an easy task unless you have specialized training, and non-negligible budget, human and time resources.
-Briefly, it requires obtaining and curating genetic data to generate an homology hypothesis or alignment; choosing and applying software to infer an evolutionary hypothesis in the form of a phylogeny; obtaining independent age data points from the fossil record or other suitable geologic events; placing those data points appropriately and with biological and geological understanding of their limits, on the obtained phylogeny; finally, choosing the appropriate software and model of evolution to estimate ellapsed time since divergence events on the phylogeny.
-This process produces a chronogram --i.e., a phylogeny with branch lengths proportional to absolute time, and they represent key knowledge for the study of natural processes in many areas of scientific research, from developmental to conservation biology [@Felsenstein1985a; @Webb2000], from historical biogeography to species diversification [@posadas2006historical; @Morlon2014].
-Because of their importance for biological research, the amount of published expertly-curated and peer-reviewed chronograms has increased constantly in the last two decades [@Kumar2017].
-This is why there has been an urge for promoting and facilitating the reuse of the vast amount of this state-of-the-art phylogenetic and evolutionary time data that has already been produced [@webb2005phylomatic; @Stoltzfus2013].
+In the natural sciences, absolute time of evolutionary origin of lineages is represented graphically as a phylogeny with branch lengths proportional to geologic time, also known as a chronogram.
+Notably, building a chronogram is not an easy task.
+First, it requires obtaining and curating genetic data to construct an homology hypothesis as a genetic alignment; then, it is important to carefully choose and apply the software that will use the alignment to reconstruct an evolutionary hypothesis in the form of a phylogeny; next, calibrations, i.e., independent age data points from the fossil record or other suitable geologic events have to be obtained from the literature --or by going into the field; fourthly, calibrations have to be carefully placed on the reconstructed phylogeny, demonstrating a biological and geological understanding of their limits and applicability; finally, the appropriate software and model of evolution to estimate ellapsed time since evolutionary events on the phylogeny has to be chosen and applied.
+In sum, building chronograms require specialized biological training and non-negligible budget, human and time resources.
+Still, chronograms represent key knowledge for the study of natural processes in many areas of scientific research, from developmental to conservation biology [@Felsenstein1985a; @Webb2000], from historical biogeography to species diversification [@posadas2006historical; @Morlon2014].
+Because of their importance for biological research, chronograms are still constantly being built by experts in the field, and have been published in peer-reviewed publications in a steady increase for the last two decades, amounting to at least four thousand individual chronograms [@Kumar2017].
+
+This state-of-the-art scientific data on time of evolutionary origin that has already been produced could be leveraged for the benefit of biological research, and there has been an urge for promoting, normlizing, and facilitating its reuse in research and education [@webb2005phylomatic; @Stoltzfus2013].
 <!-- This has short-term advantages for research relying on this information, and long term advantages such as the advancement of science [@].-->
-We identify that a tool for efficient reuse of state-of-the-art scientific data should have an open and fully public  database storing data in a computer readable format suitable for scientific reuse [@vos2012nexml], an automatised and programatic way of accessing the data [@Stoltzfus2013], and straightforward means of comparing and summarizing data as needed by the user [].
-The TreeBASE project served as a database for state-of-the-art phylogenies, chronograms and alignments in computer readable formats, and it supports programatic data accession [@piel2002treebase; @piel2009treebase, @treebaseAPIs] but is no longer under active development.
+We identify that a tool for efficient reuse of state-of-the-art scientific data should have an open and fully public database storing data in a computer readable format [@vos2012nexml], an automatised and programatic way of accessing the data [@Stoltzfus2013], and straightforward means of comparing and summarizing data as needed by the user [].
+The TreeBASE project started as an open and public database for state-of-the-art chronograms and other evolutionary research data in computer readable formats, and it currently supports programatic data accession [@piel2002treebase; @piel2009treebase, @treebaseAPIs], unfortunately, it is no longer under active development.
 <!--No need to mention SuperTreeBASE as it is basically the same as treebase but on a more stable server-->
-The Open Tree of Life project [OpenTree, @opentreeoflife2019synth] has a phylogenetic and chronogram database that is programatically accessible, but it does not yet support phylogenetic queries of age data nor chronogram summaries.
+The Open Tree of Life project [OpenTree, @opentreeoflife2019synth] has a system for storing chronograms and other evolutionary information in an open and public database that is programatically accessible, but it does not yet support phylogenetic queries of age data nor chronogram summaries.
 
-The DateLife project was born as a prototype service aiming to provide tools for easy reuse and summary of state-of-the-art time of lineage evolutionary origin, and was developed over a series of hackathons at the National Evolutionary Synthesis Center [@Stoltzfus2013].
+The DateLife project was born as a prototype service aiming to provide tools for easy reuse and summary of state-of-the-art time of lineage evolutionary origin, and was developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA [@Stoltzfus2013].
 Here we present the full implementation of the DateLife services, available as an R package `datelife` and an application with a graphical user interface web site at www.datelife.org/query/.
-The current implementation of the `datelife` R package features an algorithm for automatic curation and maintenance of an open database of chronograms pulled from OpenTree's open repository [@mctavish2015phylesystem], methods to summarize and compare source chronograms, and new functions to visualize and graphically compare source and summary chronograms.
+The current implementation of the `datelife` R package features an algorithm for automatic curation and maintenance of an open database of chronograms pulled from OpenTree's open repository [@mctavish2015phylesystem], and methods to summarize and compare source chronograms
+<!--, and new functions to visualize and graphically compare source and summary chronograms.-->
 
 # Description of the R package
-
-The `datelife` workflow builds off of functions from several R packages (rotl [@Michonneau2016], ape [@Paradis2004],
-geiger [@Harmon2008], paleotree [@Bapst2012a], bold [@Chamberlain2018], phytools [@Revell2012], taxize [@Chamberlain2013; @Chamberlain2018], phyloch [@Heibl2008], and phylocomr [@Ooms2018]).
 
@@
 \item It starts with an input consisting of at least two taxon names, which can be provided in two different forms: as a comma separated character string, or as tip labels on a tree.
-If input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004ape].
+If input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004].
 The input tree is not required to have branch lengths.
@@
 Types of summary information provided are: a) all pruned source chronograms, b) age of the MRCA (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of successful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most matching input taxon names.
-\item To construct summary trees from patristic distance matrices, one can use a clustering method like Neighbor Joining (NJ).
-We implemented this in the package.
-However, as we are not impementing a model of evolution, but using time distances between nodes, NJ methods do not perform well and often return toologies that are not biologically plausible [@].
-Hence, we implement
-a different workflow, in which we use a fixed topology taken from the literature or from expert
-phylogenetic information (such as the OpenTree synthetic tree), and we use the time distances as calibrations to date that topology with BLADJ [@webb2005phylomatic].
-<!--TODO ADD FIGURE: mock example explaining difference between using summary ages and raw ages to calibrate a topology.
+\item To construct summary trees we use a fixed topology, either provided by the user, taken from the literature or obtained from expert
+phylogenetic information, such as the OpenTree synthetic tree.
+Then we summarize source chronograms into a single patristic distance matrix using a method chosen by the user.
+Summarizing method options available include Super Distance Matrix method [SDM, @Criscuolo2006] and any summary statistics such as median, minimum andmaximum ages.
+Finally, `datelife` applies the summarized time distances as node calibrations to date the given topology with BLADJ [@webb2005phylomatic].
+<!--ADD FIGURE: mock example explaining difference between using summary ages and raw ages to calibrate a topology.
 Then show it in the biological example.
@@
 
-\item  Alternatively, time of lineage divergence obtained from the pruned chronograms can be used directly as secondary calibration points to date a tree with or without branch lengths containing some or all input taxon names.
-%<!--, a taxonomic tree-->
-\item  If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms described below.
-\item  Users can easily save all source and summary chronograms in formats that permit easy reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
+<!-- \item  Alternatively, time of lineage divergence obtained from the pruned chronograms can be used directly as secondary calibration points to date a tree with or without branch lengths containing some or all input taxon names.
+%a taxonomic tree-->
+\item  If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms.
+\item  Users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
 \end{enumerate}
@@
 A college educator wishes to obtain state-of-the-art data on time of evolutionary origin of species belonging to the true finches for their class.
-They decide to use `datelife` because they require the analysis to be reproducible.
+They decide to use `datelife` because they are teaching best practices for reproducibility.
 Students have the option to go to the website at www.datelife.org and perform an interactive run. However, the educator wants the students to practice their R skills.
-The first step is to run a higher-taxon-name `datelife` query.
-This will get taxon names for all recognised species within any higher taxon.
+The first step is to run a `datelife` query using the "get species from taxon" flag.
+This will get all recognised species names within their chosen inclusive taxon.
 The Fringillidae has `r length(dquery$cleaned_names)` species, according to the Open Tree of Life taxonomy.
-Once with a curated set of query taxon names, the next step is to run a `datelife` search.
-This will find all chronograms that contain at least two queried taxon names, and will save the information on time of lineage divergence as (an R "data frame") table.
+Once with a curated set of species taxon names, the next step is to run a `datelife` search that will find all chronograms that contain at least two species names.
+The algorithm proceeds to prune the trees to keep matching species names on tips only, and transform the pruned trees to pairwise distance matrices.
  There are `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies (Fig. \ref{fig:schronograms1}).
@@
 -->
-The final step is to summarize the available information using the two alternative types of summary chronograms, median and SDM.
-As explained in the "Description" section, data from source chronograms is first summarised into a single distance matrix (using the median and the SDM method respectively) and then the available node ages are used as fixed ages over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig.
+The final step is to summarize the available information using two alternative types of summary chronograms, median and SDM.
+As explained in the "Description" section, data from source chronograms is first summarised into a single distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig.
 \ref{fig:summaries}).
@@
 
-The main goal of `datelife` is to make expert information on time of lineage divergence easily accesible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
-It is a very fast tool that fulfills the quality of openness and does not require any expert biological knowledge from users --besides the names of the organisms they want to work with-- for any of its functionalities.
-However, it has many flaws.
-Some of them can be overcome, some of them might represent limitations.
+The main goal of `datelife` is to make state-of-the-art information on time of lineage divergence easily accesible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
+It is an open service that does not require any expert biological knowledge from users --besides the names of the organisms they want to work with, for any of its functionalities.
+The `datelife` workflow builds off of functions from several R packages that must be acknowledge (rotl [@Michonneau2016], ape [@paradis2004],
+geiger [@Harmon2008], paleotree [@Bapst2012a], bold [@Chamberlain2018], phytools [@Revell2012], taxize [@Chamberlain2013; @Chamberlain2018], phyloch [@Heibl2008], and phylocomr [@Ooms2018]).
 
-At the time of writing of this manuscript, `datelife`'s database has `r length(opentree_chronograms$trees)` chronograms, pulled entirely from OpenTree's repository, the only public tree repository from where one can currently get chronograms to construct a database.
-This represents `r round(length(opentree_chronograms$trees)*100/3987, digits = 2)`% of the largest existing chronogram database, TimeTree, which has a collection of 3,998 chronograms as of `r format(Sys.Date(), format="%B %d %Y")`.
-Unfortunately, TimeTree's database is not open for scientific reuse nor automated data mining [@Kumar2017].
-In 2015, a synthetic chronogram was constructed from 2,274 chronograms available at the time on the TimeTree database [@Hedges2015].
-This is the only synthetic TimeTree chronogram that has been made publicly available and deposited on the OpenTree repository, and it is part of datelife's database now.
-Hence, the amount of lineages represented in datelife's database (`r length(unique(unname(unlist(lapply(opentree_chronograms$trees, "[[", "tip.label")))))` unique terminal taxa), is at least as substantial as TimeTree's (as of the last released tree, with `r max(sapply(opentree_chronograms$trees[grepl("Hedges", names(opentree_chronograms$trees))], ape::Ntip))` taxa).
-Regrettably, this does not ensure that the full state of knowledge of time of divergence of the taxon/lineage will be available.
-<!--
-Even though datelifes chronograms cover this many species from the tree of life compared to this many covered by TimeTree (database or synthetic chronogram?)-->
-Incorporation of more published chronograms into `datelife`'s database is crucial to improve its services.
-One option to increase our database is the Dryad data repository.
+At the time of writing of this manuscript (`r format(Sys.time(), "%b %d, %Y")`), `datelife`'s database has `r length(opentree_chronograms$trees)` chronograms, pulled entirely from OpenTree's database, th ePhylesystem [@mctavish2015phylesystem].
+A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time.
+As chronograms are added to Phylesystem, they are incorporated into an updated `datelife`'s database that is assigned a new version number, followed by a package release on CRAN [@].
+`datelife`'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
+Users can also upload new chronograms to OpenTree themselves, trigger and update of the `datelife` database to incorporate the new chronograms, and and have them immediately available for analysis.
+
+Incorporation of more chronograms into `datelife`'s database is crucial to improve its services.
+One option to increase chronogram number in the database is the Dryad data repository.
 Methods to automatically mine chronograms from Dryad could be designed and implemented.
-However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined on a second step, by a curator.
-<!-- Can we mine journal databases? -->
-<!-- At the same time, the prospect is to increase awereness and keep inviting and doing work encouraging scientists to make their data sets available in OpenTree's database. -->
-Consequently, we would like to emphasize on the importance of sharing chronogram data for the benefit of the scientific community as a whole, into repositories that require expert input and manual curation, such as OpenTree's tree repository [@mctavish2015phylesystem].
+However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined manually by a curator.
 
-Another potential concern comes from summary chronograms.
-We currently summarize by default all source chronograms that overlap with at least two taxa.
-Users can subset source data if they have reasons to choose some source chronograms over others.
-Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely.
-To our knowledge, there is no objective way to determine if an expert chronogram is better than another.
-Some criteria that have been put forward are the level of lineage sampling and the number of calibrations used.
-Scientists usually also favor chronograms constructed using primary calibrations (ages obtained from the fossil or geological record) to ones constructed with secondary calibrations (ages coming from other chronograms).
+The only summary chronogram encompassing all life that is currently openly available from OpenTree
+was constructed using age data from 2,274 published chronograms [@Hedges2015].
+However source chronograms are not available in computer readable format for reuse or reanalysis.
+As this tree is part of datelife's database, the amount of lineages that can be queried using `datelife` (`r length(unique(unname(unlist(lapply(opentree_chronograms$trees, "[[", "tip.label")))))` unique terminal taxa) is substantial, yet it can be improved.
+Consequently, we would like to emphasize on the importance of sharing chronogram data for the benefit of the scientific community as a whole, into repositories that require expert input and manual curation, such as OpenTree's Phylesystem [@mctavish2015phylesystem].
+
+By default, `datelife` currently summarizes all source chronograms that overlap with at least two species names.
+Users can exclude source chronograms if they have reasons to do so.
+Strictly speaking, the best chronogram should reflect the real time of lineage divergence accurately and precisely.
+To our knowledge, there are no good measures to determine if a chronogram is better than another.
+Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used [@magallon2010using, @magallon2015metacalibrated].
+Scientists usually also favor chronograms constructed using primary calibrations (ages obtained from the fossil or geological record) to ones constructed with secondary calibrations (ages coming from other chronograms)[@schenk2016sec].
 It has been observed with simulations that divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with bayesian inference methods when priors are implemented in similar ways in both analyses [@schenk2016sec].
-Yet, there are different ways to use secondary calibrations
-<!-- give some examples or alternative ways of using calibrations-->
-and that same bias might not be encountered with dating methods that do not require setting priors, i.e., penalized likelihood methods such as r8s [@sanderson2003r8s].
+However, secondary calibrations
+can be applied using other dating methods that do not require setting priors, such as penalized likelihood [@sanderson2003r8s], or as fixed ages,
+potentially mitigating the bias reported with bayesian methods.
 Certainly, further studies are required to fully understand the effect of using secondary calibrations on time estimates and downstream anlyses.
 
-Furthermore, even chronograms obtained with primary fossil data can show substantial variation in time estimates between clades, as observed from the comparison of source chronograms in the Fringillidae example.
+Furthermore, even chronograms obtained with primary fossil data can vary substantially in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
 This observation is often encountered in the literature
 <!-- Different source chronograms often show substantial variation in clade ages  -->
-(see, for example, the ongoing debate about crown group age of angiosperms [@barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time].
+(see, for example, the ongoing debate about crown group age of angiosperms [@sauquet2021age; @barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time].
 For some studies, especially ones based on branch lengths (e.g., studies of species diversification, timing of evolutionary events, phenotypic trait evolution), using a different chronogram may return different results [@title2016macrophylogenies].
-Stitching together these chronograms can create a larger tree that uses information from multiple studies, but the effect of uncertainties and errors here on downstream analyses is still largely unknown.
+Stitching together these chronograms can create a larger tree that uses information from multiple studies, but the effect of uncertainties and errors at this level on downstream analyses is still largely unknown.
 
@@
 For example, two different researchers working on the same clade both carefully select and argument their choices of fossil calibrations.
-Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses teh same one to calibrate outside the clade, the resulting age estimates will probably differ substantially (the placement of calibrations is proved to deeply affect estimated times of lineage divergence).
+Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses the same one to calibrate outside the clade, the resulting age estimates will often differ substantially, as the placement of calibrations as stem or crown group is proved to deeply affect estimated times of lineage divergence [@sauquet2013practical].
 Trying to summarize the resulting chronograms into a single one using simple summary statistics might erase all types of relevant information from the source chronograms.
@@
 Probably not-->
-However, the exercise of summarizing different chronograms ha sthe potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypothesis.
-Choosing the elements of the chronograms that we are going to keep and the ones that we are going to discard is key, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram.
+However, the exercise of summarizing different chronograms has the potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypothesis.
+Choosing the elements of the chronograms that we are going to keep and the ones that we are going to discard is key, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram [@sauquet2021age].
 <!-- Example -->
@@
 In other areas of biological research, such as ecology and conservation biology, it has been shown that at least some data on lineage divergence represents a relevant improvement for testing alternative hypothesis using phylogenetic distance [@Webb2008].
-Hence, we integrated into datelife's workflow different ways of creating branch lengths in the absence of starting branch length information for taxa lacking this information (BLADJ option).
-Making up branch lengths in this or other ways is accepted in scientific publications: @Jetz2012, created a time-calibrated tree of all 9,993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a tree of 353,185 seed plants where only 23% had molecular data.
-Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch imputation methods.
+Hence, we integrated into datelife's workflow different ways of creating branch lengths in the absence of starting branch length information for taxa lacking this information (using the BLADJ option).
+"Making up" branch lengths is an accepted practice in scientific publications: @Jetz2012, created a time-calibrated tree of all 9,993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a tree of 353,185 seed plants where only 23% had molecular data.
+Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch assignation methods.
 There has yet to be a thorough analysis of what can go wrong when one goes beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make sure results are substantially similar between trees fully reconstructed with molecular or other data, and trees that are reconstructed using taxonomy by resolving polytomies at random following a statistical model.

### 2022-03-04 manuscript/manuscript.tex
adding references

@@
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+\DeclareUnicodeCharacter{0301}{*************************************}
 % Manuscript styling
@@
 
-Inferring time of lineage evolutionary origin from scratch is not an easy task unless you have specialized training, and non-negligible budget, human and time resources.
-Briefly, it requires obtaining and curating genetic data to generate an homology hypothesis or alignment; choosing and applying software to infer an evolutionary hypothesis in the form of a phylogeny; obtaining independent age data points from the fossil record or other suitable geologic events; placing those data points appropriately and with biological and geological understanding of their limits, on the obtained phylogeny; finally, choosing the appropriate software and model of evolution to estimate ellapsed time since divergence events on the phylogeny.
-This process produces a chronogram --i.e., a phylogeny with branch lengths proportional to absolute time, and they represent key knowledge for the study of natural processes in many areas of scientific research, from developmental to conservation biology (Felsenstein, 1985; Campbell O.
+In the natural sciences, absolute time of evolutionary origin of lineages is represented graphically as a phylogeny with branch lengths proportional to geologic time, also known as a chronogram.
+Notably, building a chronogram is not an easy task.
+First, it requires obtaining and curating genetic data to construct an homology hypothesis as a genetic alignment; then, it is important to carefully choose and apply the software that will use the alignment to reconstruct an evolutionary hypothesis in the form of a phylogeny; next, calibrations, i.e., independent age data points from the fossil record or other suitable geologic events have to be obtained from the literature --or by going into the field; fourthly, calibrations have to be carefully placed on the reconstructed phylogeny, demonstrating a biological and geological understanding of their limits and applicability; finally, the appropriate software and model of evolution to estimate ellapsed time since evolutionary events on the phylogeny has to be chosen and applied.
+In sum, building chronograms require specialized biological training and non-negligible budget, human and time resources.
+Still, chronograms represent key knowledge for the study of natural processes in many areas of scientific research, from developmental to conservation biology (Felsenstein, 1985; Campbell O.
 Webb, 2000), from historical biogeography to species diversification (Morlon, 2014; Posadas, Crisci, \& Katinas, 2006).
-Because of their importance for biological research, the amount of published expertly-curated and peer-reviewed chronograms has increased constantly in the last two decades (Kumar, Stecher, Suleski, \& Hedges, 2017).
-This is why there has been an urge for promoting and facilitating the reuse of the vast amount of this state-of-the-art phylogenetic and evolutionary time data that has already been produced (Stoltzfus et al., 2013; Campbell O.
+Because of their importance for biological research, chronograms are still constantly being built by experts in the field, and have been published in peer-reviewed publications in a steady increase for the last two decades, amounting to at least four thousand individual chronograms (Kumar, Stecher, Suleski, \& Hedges, 2017).
+
+This state-of-the-art scientific data on time of evolutionary origin that has already been produced could be leveraged for the benefit of biological research, and there has been an urge for promoting, normlizing, and facilitating its reuse in research and education (Stoltzfus et al., 2013; Campbell O.
 Webb \& Donoghue, 2005).
-We identify that a tool for efficient reuse of state-of-the-art scientific data should have an open and fully public database storing data in a computer readable format suitable for scientific reuse (Vos et al., 2012), an automatised and programatic way of accessing the data (Stoltzfus et al., 2013), and straightforward means of comparing and summarizing data as needed by the user {[}{]}.
-The TreeBASE project served as a database for state-of-the-art phylogenies, chronograms and alignments in computer readable formats, but it does not support automatized or programatic data accession (W.
-H.
-Piel, Donoghue, \& Sanderson, 2002; W.
-Piel et al., 2009).
+We identify that a tool for efficient reuse of state-of-the-art scientific data should have an open and fully public database storing data in a computer readable format (R.
+A.
+Vos et al., 2012), an automatised and programatic way of accessing the data (Stoltzfus et al., 2013), and straightforward means of comparing and summarizing data as needed by the user {[}{]}.
+The TreeBASE project started as an open and public database for state-of-the-art chronograms and other evolutionary research data in computer readable formats, and it currently supports programatic data accession R.
+Vos (n.d.), unfortunately, it is no longer under active development.
+The Open Tree of Life project (OpenTree, OpenTreeOfLife et al., 2019) has a system for storing chronograms and other evolutionary information in an open and public database that is programatically accessible, but it does not yet support phylogenetic queries of age data nor chronogram summaries.
 
-SuperTreeBASE?
-
-The Open Tree of Life project (OpenTree, OpenTreeOfLife et al., 2019) has a phylogenetic and chronogram database that is programatically accessible, but it does not yet support phylogenetic queries of age data nor chronogram summaries.
-
-The DateLife project was born as a prototype service aiming to provide tools for easy reuse and summary of state-of-the-art time of lineage evolutionary origin, and was developed over a series of hackathons at the National Evolutionary Synthesis Center (Stoltzfus et al., 2013).
+The DateLife project was born as a prototype service aiming to provide tools for easy reuse and summary of state-of-the-art time of lineage evolutionary origin, and was developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA (Stoltzfus et al., 2013).
 Here we present the full implementation of the DateLife services, available as an R package \texttt{datelife} and an application with a graphical user interface web site at www.datelife.org/query/.
-The current implementation of the \texttt{datelife} R package features an algorithm for automatic curation and maintenance of an open database of chronograms pulled from OpenTree's open repository (McTavish et al., 2015), methods to summarize and compare source chronograms, and new functions to visualize and graphically compare source and summary chronograms.
+The current implementation of the \texttt{datelife} R package features an algorithm for automatic curation and maintenance of an open database of chronograms pulled from OpenTree's open repository (McTavish et al., 2015), and methods to summarize and compare source chronograms
 
@@
 \section{Description of the R package}\label{description-of-the-r-package}}
-
-The \texttt{datelife} workflow builds off of functions from several R packages (rotl (Michonneau, Brown, \& Winter, 2016), ape (Paradis, Claude, \& Strimmer, 2004),
-geiger (Harmon, Weir, Brock, Glor, \& Challenger, 2008), paleotree (Bapst, 2012), bold (Chamberlain et al., 2019), phytools (Revell, 2012), taxize (Chamberlain \& Szöcs, 2013; Chamberlain et al., 2019), phyloch (Heibl, 2008), and phylocomr (Ooms \& Chamberlain, 2018)).
 
@@
 A college educator wishes to obtain state-of-the-art data on time of evolutionary origin of species belonging to the true finches for their class.
-They decide to use \texttt{datelife} because they require the analysis to be reproducible.
+They decide to use \texttt{datelife} because they are teaching best practices for reproducibility.
 Students have the option to go to the website at www.datelife.org and perform an interactive run. However, the educator wants the students to practice their R skills.
-The first step is to run a higher-taxon-name \texttt{datelife} query.
-This will get taxon names for all recognised species within any higher taxon.
+The first step is to run a \texttt{datelife} query using the ``get species from taxon'' flag.
+This will get all recognised species names within their chosen inclusive taxon.
 The Fringillidae has 289 species, according to the Open Tree of Life taxonomy.
-Once with a curated set of query taxon names, the next step is to run a \texttt{datelife} search.
-This will find all chronograms that contain at least two queried taxon names, and will save the information on time of lineage divergence as (an R ``data frame'') table.
+Once with a curated set of species taxon names, the next step is to run a \texttt{datelife} search that will find all chronograms that contain at least two species names.
+The algorithm proceeds to prune the trees to keep matching species names on tips only, and transform the pruned trees to pairwise distance matrices.
 There are 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Fig. \ref{fig:schronograms1}).
-The final step is to summarize the available information using the two alternative types of summary chronograms, median and SDM.
-As explained in the ``Description'' section, data from source chronograms is first summarised into a single distance matrix (using the median and the SDM method respectively) and then the available node ages are used as fixed ages over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig.
+The final step is to summarize the available information using two alternative types of summary chronograms, median and SDM.
+As explained in the ``Description'' section, data from source chronograms is first summarised into a single distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig.
 \ref{fig:summaries}).
@@
 
-The main goal of \texttt{datelife} is to make expert information on time of lineage divergence easily accesible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
-It is a very fast tool that fulfills the quality of openness and does not require any expert biological knowledge from users --besides the names of the organisms they want to work with-- for any of its functionalities.
-However, it has many flaws.
-Some of them can be overcome, some of them might represent limitations.
+The main goal of \texttt{datelife} is to make state-of-the-art information on time of lineage divergence easily accesible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
+It is an open service that does not require any expert biological knowledge from users --besides the names of the organisms they want to work with, for any of its functionalities.
+The \texttt{datelife} workflow builds off of functions from several R packages that must be acknowledge (rotl (Michonneau, Brown, \& Winter, 2016), ape (Paradis, Claude, \& Strimmer, 2004),
+geiger (Harmon, Weir, Brock, Glor, \& Challenger, 2008), paleotree (Bapst, 2012), bold (Chamberlain et al., 2019), phytools (Revell, 2012), taxize (Chamberlain \& Szöcs, 2013; Chamberlain et al., 2019), phyloch (Heibl, 2008), and phylocomr (Ooms \& Chamberlain, 2018)).
 
-At the time of writing of this manuscript, \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's repository, the only public tree repository from where one can currently get chronograms to construct a database.
-This represents 6.35\% of the largest existing chronogram database, TimeTree, which has a collection of 3,998 chronograms as of March 03 2022.
-Unfortunately, TimeTree's database is not open for scientific reuse nor automated data mining (Kumar et al., 2017).
-In 2015, a synthetic chronogram was constructed from 2,274 chronograms available at the time on the TimeTree database (Hedges, Marin, Suleski, Paymer, \& Kumar, 2015).
-This is the only synthetic TimeTree chronogram that has been made publicly available and deposited on the OpenTree repository, and it is part of datelife's database now.
-Hence, the amount of lineages represented in datelife's database (99474 unique terminal taxa), is at least as substantial as TimeTree's (as of the last released tree, with 48017 taxa).
-Regrettably, this does not ensure that the full state of knowledge of time of divergence of the taxon/lineage will be available.
-Incorporation of more published chronograms into \texttt{datelife}'s database is crucial to improve its services.
-One option to increase our database is the Dryad data repository.
+At the time of writing of this manuscript (Mar 03, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, th ePhylesystem (McTavish et al., 2015).
+A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time.
+As chronograms are added to Phylesystem, they are incorporated into an updated \texttt{datelife}'s database that is assigned a new version number, followed by a package release on CRAN {[}@{]}.
+\texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
+Users can also upload new chronograms to OpenTree themselves, trigger and update of the \texttt{datelife} database to incorporate the new chronograms, and and have them immediately available for analysis.
+
+Incorporation of more chronograms into \texttt{datelife}'s database is crucial to improve its services.
+One option to increase chronogram number in the database is the Dryad data repository.
 Methods to automatically mine chronograms from Dryad could be designed and implemented.
-However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined on a second step, by a curator.
-Consequently, we would like to emphasize on the importance of sharing chronogram data for the benefit of the scientific community as a whole, into repositories that require expert input and manual curation, such as OpenTree's tree repository (McTavish et al., 2015).
+However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined manually by a curator.
 
-Another potential concern comes from summary chronograms.
-We currently summarize by default all source chronograms that overlap with at least two taxa.
-Users can subset source data if they have reasons to choose some source chronograms over others.
-Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely.
-To our knowledge, there is no objective way to determine if an expert chronogram is better than another.
-Some criteria that have been put forward are the level of lineage sampling and the number of calibrations used.
-Scientists usually also favor chronograms constructed using primary calibrations (ages obtained from the fossil or geological record) to ones constructed with secondary calibrations (ages coming from other chronograms).
+The only summary chronogram encompassing all life that is currently openly available from OpenTree
+was constructed using age data from 2,274 published chronograms (Hedges, Marin, Suleski, Paymer, \& Kumar, 2015).
+However source chronograms are not available in computer readable format for reuse or reanalysis.
+As this tree is part of datelife's database, the amount of lineages that can be queried using \texttt{datelife} (99474 unique terminal taxa) is substantial, yet it can be improved.
+Consequently, we would like to emphasize on the importance of sharing chronogram data for the benefit of the scientific community as a whole, into repositories that require expert input and manual curation, such as OpenTree's Phylesystem (McTavish et al., 2015).
+
+By default, \texttt{datelife} currently summarizes all source chronograms that overlap with at least two species names.
+Users can exclude source chronograms if they have reasons to do so.
+Strictly speaking, the best chronogram should reflect the real time of lineage divergence accurately and precisely.
+To our knowledge, there are no good measures to determine if a chronogram is better than another.
+Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández (2015).
+Scientists usually also favor chronograms constructed using primary calibrations (ages obtained from the fossil or geological record) to ones constructed with secondary calibrations (ages coming from other chronograms)(Schenk, 2016).
 It has been observed with simulations that divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with bayesian inference methods when priors are implemented in similar ways in both analyses (Schenk, 2016).
-Yet, there are different ways to use secondary calibrations
-and that same bias might not be encountered with dating methods that do not require setting priors, i.e., penalized likelihood methods such as r8s (Sanderson, 2003).
+However, secondary calibrations
+can be applied using other dating methods that do not require setting priors, such as penalized likelihood (Sanderson, 2003), or as fixed ages,
+potentially mitigating the bias reported with bayesian methods.
 Certainly, further studies are required to fully understand the effect of using secondary calibrations on time estimates and downstream anlyses.
 
-Furthermore, even chronograms obtained with primary fossil data can show substantial variation in time estimates between clades, as observed from the comparison of source chronograms in the Fringillidae example.
+Furthermore, even chronograms obtained with primary fossil data can vary substantially in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
 This observation is often encountered in the literature
-(see, for example, the ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández, 2015; Ramshaw et al., 1972; Sanderson \& Doyle, 2001).
+(see, for example, the ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón et al., 2015; Ramshaw et al., 1972; Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021).
 For some studies, especially ones based on branch lengths (e.g., studies of species diversification, timing of evolutionary events, phenotypic trait evolution), using a different chronogram may return different results (Title \& Rabosky, 2016).
-Stitching together these chronograms can create a larger tree that uses information from multiple studies, but the effect of uncertainties and errors here on downstream analyses is still largely unknown.
+Stitching together these chronograms can create a larger tree that uses information from multiple studies, but the effect of uncertainties and errors at this level on downstream analyses is still largely unknown.
 
@@
 For example, two different researchers working on the same clade both carefully select and argument their choices of fossil calibrations.
-Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses teh same one to calibrate outside the clade, the resulting age estimates will probably differ substantially (the placement of calibrations is proved to deeply affect estimated times of lineage divergence).
+Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses the same one to calibrate outside the clade, the resulting age estimates will often differ substantially, as the placement of calibrations as stem or crown group is proved to deeply affect estimated times of lineage divergence (Sauquet, 2013).
 Trying to summarize the resulting chronograms into a single one using simple summary statistics might erase all types of relevant information from the source chronograms.
 Accordingly, the prevailing view in our research community is that we should favor time of lineage divergence estimates obtained from a single analysis, using fossil data as primary sources of calibrations, and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history (Antonelli et al., 2017).
-However, the exercise of summarizing different chronograms ha sthe potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypothesis.
-Choosing the elements of the chronograms that we are going to keep and the ones that we are going to discard is key, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram.
+However, the exercise of summarizing different chronograms has the potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypothesis.
+Choosing the elements of the chronograms that we are going to keep and the ones that we are going to discard is key, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram (Sauquet et al., 2021).
 
@@
 Webb, Ackerly, \& Kembel, 2008).
-Hence, we integrated into datelife's workflow different ways of creating branch lengths in the absence of starting branch length information for taxa lacking this information (BLADJ option).
-Making up branch lengths in this or other ways is accepted in scientific publications: Jetz, Thomas, Joy, Hartmann, and Mooers (2012), created a time-calibrated tree of all 9,993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
+Hence, we integrated into datelife's workflow different ways of creating branch lengths in the absence of starting branch length information for taxa lacking this information (using the BLADJ option).
+``Making up'' branch lengths is an accepted practice in scientific publications: Jetz, Thomas, Joy, Hartmann, and Mooers (2012), created a time-calibrated tree of all 9,993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
 (2018) created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37\% had molecular data; Smith and Brown (2018) constructed a tree of 353,185 seed plants where only 23\% had molecular data.
-Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch imputation methods.
+Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch assignation methods.
 There has yet to be a thorough analysis of what can go wrong when one goes beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make sure results are substantially similar between trees fully reconstructed with molecular or other data, and trees that are reconstructed using taxonomy by resolving polytomies at random following a statistical model.
@@
 \texttt{datelife}'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
-\texttt{datelife}'s R package stable version will be available
+\texttt{datelife}'s R package stable version is available
 for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-magallon2010using}{}}%
+Magallón, S. (2010). Using fossils to break long branches in molecular dating: A comparison of relaxed clocks applied to the origin of angiosperms. \emph{Systematic Biology}, \emph{59}(4), 384--399.
+
 \leavevmode\vadjust pre{\hypertarget{ref-magallon2015metacalibrated}{}}%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Paradis2004}{}}%
-Paradis, E., Claude, J., \& Strimmer, K. (2004). {APE: analyses of phylogenetics and evolution in R language}. \emph{Bioinformatics}, \emph{20}, 289--290.
+\leavevmode\vadjust pre{\hypertarget{ref-paradis2004}{}}%
+Paradis, E., Claude, J., \& Strimmer, K. (2004). {APE: analyses of phylogenetics and evolution in R language}. \emph{Bioinformatics}, \emph{20}(2), 289--290.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-sauquet2013practical}{}}%
+Sauquet, H. (2013). A practical guide to molecular dating. \emph{Comptes Rendus Palevol}, \emph{12}(6), 355--367.
+
+\leavevmode\vadjust pre{\hypertarget{ref-sauquet2021age}{}}%
+Sauquet, H., Ramírez-Barahona, S., \& Magallón, S. (2021). The age of flowering plants is unknown.
+
 \leavevmode\vadjust pre{\hypertarget{ref-schenk2016sec}{}}%
@@
 \url{https://doi.org/10.1093/sysbio/syw102}
+
+\leavevmode\vadjust pre{\hypertarget{ref-treebaseAPIs}{}}%
+Vos, R. (n.d.). {TreeBASE APIs}. Retrieved from \url{\%7Bhttps://github.com/TreeBASE/treebase/wiki/API\%7D}
 
@@
 % in this file I leave the figure captions outside\ caption{} because I want them
-to be formatted in the same way as the general text (double spaced and linenumbered)
+% to be formatted in the same way as the general text (double spaced and linenumbered)
 \captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}
@@
 \begin{figure}[!h]
-\includegraphics{../figures/Fig1.pdf}
+\includegraphics{../figures/figure1/figure1-horizontal.pdf}
 \caption{Stylized DateLife workflow.

### 2022-03-04 manuscript/preamble.tex
adding references

--- 2021-11-02 manuscript/manuscript.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-  english,
-  man]{apa6}
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life},
-  pdfauthor={Luna L. Sánchez Reyes1,2, Emily Jane McTavish1, \& Brian O'Meara2},
-  pdflang={en-EN},
-  pdfkeywords={Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage{graphicx,grffile}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-% Manuscript styling
-\usepackage{upgreek}
-\captionsetup{font=singlespacing,justification=justified}
-
-% Table formatting
-\usepackage{longtable}
-\usepackage{lscape}
-% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
-\usepackage{multirow}		% Table styling
-\usepackage{tabularx}		% Control Column width
-\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
-\usepackage{threeparttablex}            % Lets threeparttable work with longtable
-
-% Create new environments so endfloat can handle them
-% \newenvironment{ltable}
-%   {\begin{landscape}\centering\begin{threeparttable}}
-%   {\end{threeparttable}\end{landscape}}
-\newenvironment{lltable}{\begin{landscape}\centering\begin{ThreePartTable}}{\end{ThreePartTable}\end{landscape}}
-
-% Enables adjusting longtable caption width to table width
-% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
-\makeatletter
-\newcommand\LastLTentrywidth{1em}
-\newlength\longtablewidth
-\setlength{\longtablewidth}{1in}
-\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}
-
-% \setlength{\parindent}{0.5in}
-% \setlength{\parskip}{0pt plus 0pt minus 0pt}
-
-% \usepackage{etoolbox}
-\makeatletter
-\patchcmd{\HyOrg@maketitle}
-  {\section{\normalfont\normalsize\abstractname}}
-  {\section*{\normalfont\normalsize\abstractname}}
-  {}{\typeout{Failed to patch abstract.}}
-\patchcmd{\HyOrg@maketitle}
-  {\section{\protect\normalfont{\@title}}}
-  {\section*{\protect\normalfont{\@title}}}
-  {}{\typeout{Failed to patch title.}}
-\makeatother
-\shorttitle{DateLife: revealing the dated Tree of Life}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 2949}
-\DeclareDelayedFloatFlavor{ThreePartTable}{table}
-\DeclareDelayedFloatFlavor{lltable}{table}
-\DeclareDelayedFloatFlavor*{longtable}{table}
-\makeatletter
-\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
-\makeatother
-\usepackage{lineno}
-
-\linenumbers
-\usepackage{csquotes}
-\ifxetex
-  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
-  \usepackage{polyglossia}
-  \setmainlanguage[]{english}
-\else
-  \usepackage[shorthands=off,main=english]{babel}
-\fi
-
-\title{DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life}
-\author{Luna L. Sánchez Reyes\textsuperscript{1,2}, Emily Jane McTavish\textsuperscript{1}, \& Brian O'Meara\textsuperscript{2}}
-\date{}
-
-\authornote{
-
-School of Natural Sciences, University of California, Merced, Science and Engineering Building 1.
-
-Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 425 Hesler Biology Building, Knoxville, TN 37996, USA.
-
-The authors made the following contributions.
-Luna L.
-Sánchez Reyes: Data curation, Investigation, Software, Visualization, Validation, Writing - Original Draft Preparation, Writing - Review \& Editing; Emily Jane McTavish: Resources, Software, Writing - Review \& Editing; Brian O'Meara: Conceptualization, Funding acquisition, Methodology, Resources, Software, Supervision, Writing - Review \& Editing.
-
-Correspondence concerning this article should be addressed to Luna L. Sánchez Reyes, . E-mail: \href{mailto:sanchez.reyes.luna@gmail.com}{\nolinkurl{sanchez.reyes.luna@gmail.com}}
-
-}
-
-\affiliation{\vspace{0.5cm}\textsuperscript{1} University of California, Merced\\\textsuperscript{2} University of Tennessee, Knoxville}
-
-\abstract{
-Time of evolutionary origin is fundamental for research in the natural sciences, as well as for education, science communication and policy.
-Despite an increased availability of fossil and molecular data, and time-efficient analytical techniques, achieving a high-quality reconstruction of time of evolutionary origin as a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is still a difficult and time-consuming task for a majority of interested parties.
-Yet, the amount of published chronograms has increased significantly in the past two decades, and a non-negligeable proportion of these data have been steadily accumulating in public, open databases such as TreeBASE and Open Tree
-of Life, exposing a wealth of expertly-curated and peer-reviewed data on time of evolutionary origin in a programatic and reusable way, for a large quantity and diversity of organisms.
-This trend results from intensive and localized efforts for improving data sharing practices, as well as incentivizing open science in biology.
-Despite these trends, accessibility to state-of-the-art knowledge on time of evolutionary origin is still reduced.
-
-Here we present \texttt{datelife}, a service implemented as an R package and an Rshiny website application
-available at www.datelife.org/query/, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
-
-The main workflow of \texttt{datelife} is to construct a chronogram for any given combination of taxon names, by searching a local chronogram database constructed and curated from the Open Tree of Life (OpenTree), which incorporates phylogenetic data from the TreeBASE database as well.
-We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms.
-Additionally, time data extracted from source chronograms can be usedas secondary calibration points to add branch lengths proportional to absolute time to a tree topology using alternative dating methods.
-
-Summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research in biology.
-How well this chronograms work for this purpose still needs to be tested.
-
-\texttt{datelife} will be useful to increase awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.
-}
-
-\begin{document}
-\maketitle
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-Time of evolutionary origin represents a fundamental piece of information for
-understanding biological processes in many areas of research, from developmental to conservation biology (Felsenstein, 1985; Webb, 2000), from historical biogeography to species diversification studies (Morlon, 2014; Posadas, Crisci, \& Katinas, 2006).
-The number of studies publishing phylogenies with branch lengths proportional to absolute time (hereafter chronograms) have constantly increased in number for the last two decades (Kumar, Stecher, Suleski, \& Hedges, 2017).
-Still, generating a chronogram is not an easy task unless you have specialized training, and a wealth of time and resources: it requires inferring a phylogenetic tree using genetic markers, obtaining independent time data from the fossil record, and understanding the placement of fossils on the tree as well as their limits for analysis.
-That is why there has been an urge for promoting and facilitating the reuse of the vast amount of phylogenetic and evolutionary time data that has been made available in publications, for the advantage of research relying on this information (Stoltzfus et al., 2013; Webb \& Donoghue, 2005).
-
-A tool for efficient reuse of expert, published data on time of evolutionary origin should have an open and fully public chronogram database storing data in a format suitable for scientific reuse, an automatised way of accessing the information, and straightforward means of comparing and summarizing chronogram information as needed by the user.
-A prototype service aiming to meet this criteria was developed over a series of hackathons at the National Evolutionary Synthesis Center (Stoltzfus et al., 2013).
-Here we present the full implementation of the \texttt{datelife} service, constituted by an R package and a web site available at www.datelife.org/query/.
-The current implementation of \texttt{datelife} performs the tasks described above.
-It features an algorithm for automatic curation and maintenance of an open database of chronograms pulled from the OpenTree public repository, methods to summarize and compare source chronograms, and new functions to visualize and graphically compare source and summary chronograms.
-
-\hypertarget{implementationdescriptionworkflow}{%
-\section{Implementation/Description/Workflow}\label{implementationdescriptionworkflow}}
-
-The \texttt{datelife} workflow builds off of functions from several R packages (rotl (Michonneau, Brown, \& Winter, 2016), ape (Paradis, Claude, \& Strimmer, 2004),
-geiger (Harmon, Weir, Brock, Glor, \& Challenger, 2008), paleotree (Bapst, 2012), bold (Chamberlain, 2018), phytools (Revell, 2012), taxize (Chamberlain, 2018; Chamberlain \& Szöcs, 2013), phyloch (Heibl, 2008), and phylocomr (Ooms \& Chamberlain, 2018)).
-
-The basic \texttt{datelife} workflow is shown in figure \ref{fig:workflow}, largely:
-
-\begin{enumerate}
-\item It starts with an input consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree.
-The tree can be provided in newick format, also as a character string, or as a "phylo" R object, and can have any type of branch lengths or none.
-\item The input taxon names are cleaned with TNRS and saved as a `datelifeQuery` object.
-If taxon names are taxonomic groups above the species level, `datelife` has two alternative behaviors.
-If the "get species from taxon" flag is active, `datelife` will retrieve all species within a higher taxon name and add the species names to the input.
-If the flag is inactive, `datelife` will drop the higher taxon names from the input.
-The cleaned input taxon names are searched across the source chronogram database.
-Source chronograms with at least two matching input taxon names are singled out and pruned down to preserve only input taxon names in the tips of the chronograms.
-Then, each pruned source chronogram is transformed to a patristic distance matrix.
-This format facilitates and greatly speeds up all downstream analyses and summaries.
-The matrices are associated to the citation of the original study and stored as a `datelifeResult` object.
-\item  At this point, various summary data can be obtained to inform decisions for the next steps of the analysis workflow.
-Types of summary information provided are: a) all pruned source chronograms, b) age of the MRCA (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of successful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most matching input taxon names.
-\item  Finally, time of lineage divcergence obtained from the pruned source chronograms can be used as secondary calibration points to date a tree with or without branch lengths containing some or all input taxon names.
-%<!--, a taxonomic tree-->
-\item  If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms described below.
-\item  Users can easily save all source and summary chronograms in formats that permit easy reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
-\end{enumerate}
-
-\hypertarget{benchmark}{%
-\section{Benchmark}\label{benchmark}}
-
-\texttt{datelife}'s code speed was tested on an Apple iMac
-with one 3.4 GHz Intel Core i5 processor.
-We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
-Query processing time increases roughly linearly with number of input taxon names, and
-increases considerably if the taxonomic name resolution service (TNRS; Boyle et al., 2013) is activated.
-Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
-Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig.
-\ref{fig:runtime1}).
-\texttt{datelife}'s code performance was evaluated with a set of unit tests designed and
-implemented with the R package testthat (R Core Team, 2018) that were run both locally
-with the devtools package (R Core Team, 2018), and on a public server --via
-GitHub, using the continuous integration tool Travis CI (\url{https://travis-ci.org}). At
-present, unit tests cover more than 30\% of \texttt{datelife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).
-
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
-\hypertarget{case-study}{%
-\subsection{Case study}\label{case-study}}
-
-We illustrate the \texttt{datelife} workflow using the family of true finches, Fringillidae as an example.
-To contextualize, a college educator wishes to know the state-of-the-art on time of evolutionary origin of species belonging to the true finches using \texttt{datelife}.
-One option is to go to the website at www.datelife.org and perform an interactive run. However, the educator wants the students to practice their R skills.
-The first step is to run a higher-taxon-name \texttt{datelife} query.
-This will get taxon names for all recognised species within any higher taxon.
-The Fringillidae has 289 species, according to the Open Tree of Life taxonomy.
-Once with a curated set of query taxon names, the next step is to run a \texttt{datelife} search.
-This will find all chronograms that contain at least two queried taxon names, and will save the information on time of lineage divergence as (an R \enquote{data frame}) table.
-There are 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Fig. \ref{fig:schronograms1}).
-The final step is to summarize the available information using the two alternative types of summary chronograms, median and SDM.
-As explained in the \enquote{Description} section, data from source chronograms is first summarised into a single distance matrix (using the median and the SDM method respectively) and then the available node ages are used as fixed ages over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig.
-\ref{fig:summaries}).
-Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
-With both methods, ages are generally consistent with source ages, but there are some biological examples in which this is not true (see Discussion).
-
-\hypertarget{cross-validation-test}{%
-\subsection{Cross-validation test}\label{cross-validation-test}}
-
-Data from source chronograms can be also used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates (Figs.
-\ref{fig:cvbladj} and \ref{fig:cvbold}).
-As a form of cross validation, we took tree topologies from each study and calibrated them using time of lineage divergence data from all other source chronograms.
-In the absence of branch lengths, the ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig.
-\ref{fig:cvbladj}).
-Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cvbladj}).
-We also demonstrate the usage of PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007) as an alternative method to BLADJ.
-For this, we run a \texttt{datelife} branch length reconstruction that searches for DNA sequence data from the Barcode of Life Data System {[}BOLD; ratnasingham2007bold{]} to generate branch lengths.
-We were able to successfully generate a tree with BOLD branch lengths for all of the Fringillidae source chronograms.
-However, dating with PATHd8 using congruified calibrations, was only successful in
-three cases (studies 3, 5, and 9, shown in Fig.
-\ref{fig:cvbold}).
-From these, two trees have a different sampling than the original source chronogram, mainly because DNA BOLD data for some species is absent from the database.
-Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
-More examples and code used to generate these trees were developed on an open repository that is available for consultation and reuse at \url{https://github.com/LunaSare/datelife_examples}.
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-The main goal of \texttt{datelife} is to make expert information on time of lineage divergence easily accesible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
-It is a very fast tool that fulfills the quality of openness and does not require any expert biological knowledge from users --besides the names of the organisms they want to work with-- for any of its functionalities.
-However, it has many flaws.
-Some of them can be overcome, some of them might represent limitations.
-
-Up to the time this manuscript was written, \texttt{datelife}'s chronogram database had 231 chronograms, pulled entirely from OpenTree's tree repository, the only public tree repository from where \texttt{datelife} can currently get chronograms to construct its database.
-This represents 5.79\% of the largest existing chronogram database, TimeTree, which has a collection of 3,998 chronograms as of November 02 2021.
-Unfortunately, TimeTree's database is not open for scientific reuse nor automatised data mining (Kumar et al., 2017).
-In 2015, a synthetic chronogram was constructed from 2,274 chronograms available at the time on the TimeTree database (Hedges, Marin, Suleski, Paymer, \& Kumar, 2015).
-This is the only synthetic TimeTree chronogram that has been made publicly available and deposited on the OpenTree repository, and is part of datelife's database now.
-Hence, the amount of lineages represented in datelife's database is at least as substantial as TimeTree's, ensuring that some information will be available for any given taxon or lineage.
-Regrettably, this does not ensure that the full state of knowledge of time of divergence of the taxon/lineage will be available.
-Incorporation of more published chronograms into \texttt{datelife}'s database is crucial to improve its services.
-One option to increase our database is the Dryad data repository.
-Methods to automatically mine chronograms from Dryad could be designed and implemented.
-However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined on a second step, by a curator.
-Consequently, we would like to emphasize on the importance of sharing chronogram data for the benefit of the scientific community as a whole, into repositories that require expert input and manual curation, such as OpenTree's tree repository (McTavish et al., 2015).
-
-Another potential concern comes from summary chronograms.
-We currently summarize by default all source chronograms that overlap with at least two taxa.
-Users can subset source data if they have reasons to choose some source chronograms over others.
-Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely.
-To our knowledge, there is no objective way to determine if an expert chronogram is better than another.
-Some criteria that have been put forward are the level of lineage sampling and the number of calibrations used.
-Scientists usually also favor chronograms constructed using primary calibrations (ages obtained from the fossil or geological record) to ones constructed with secondary calibrations (ages coming from other chronograms).
-It has been observed with simulations that divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with bayesian inference methods when priors are implemented in similar ways in both analyses (Schenk, 2016).
-Yet, there are different ways to use secondary calibrations
-and that same bias might not be encountered with dating methods that do not require setting priors, i.e., Maximum Likelihood methods such as r8s (Sanderson, 2003).
-Certainly, further studies are required to fully understand the effect of using secondary calibrations on time estimates and downstream anlyses.
-
-Furthermore, even chronograms obtained with primary fossil data can show substantial variation in time estimates between clades, as observed from the comparison of source chronograms in the Fringillidae example.
-This observation is often encountered in the literature
-(see, for example, the ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández, 2015; Ramshaw et al., 1972; Sanderson \& Doyle, 2001).
-For some studies, especially ones based on branch lengths (e.g., studies of species diversification, timing of evolutionary events, phenotypic trait evolution), using a different chronogram may return different results (Title \& Rabosky, 2016).
-Stitching together these chronograms can create a larger tree that uses information from multiple studies, but the effect of uncertainties and errors here on downstream analyses is still largely unknown.
-
-Summarizing chronograms might also imply summarizing fundamentally distinct evolutionary hypotheses.
-For example, two different researchers working on the same clade both carefully select and argument their choices of fossil calibrations.
-Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses teh same one to calibrate outside the clade, the resulting age estimates will probably differ substantially (the placement of calibrations is proved to deeply affect estimated times of lineage divergence).
-Trying to summarize the resulting chronograms into a single one using simple summary statistics might erase all types of relevant information from the source chronograms.
-Accordingly, the prevailing view in our research community is that we should favor time of lineage divergence estimates obtained from a single analysis, using fossil data as primary sources of calibrations, and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history (Antonelli et al., 2017).
-However, the exercise of summarizing different chronograms ha sthe potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypothesis.
-Choosing the elements of the chronograms that we are going tp keep and the ones that we are going to discard is key, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram.
-
-Alternatively, one could try to choose the \enquote{best} chronogram from a set of possible evolutionary hypotheses.
-Several characteristics of the data used for dating analyses as well as from the output chronogram itself, could be used to score quality of source chronograms.
-Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
-To facilitate subsetting of source chronograms following different criteria by the users, this information should be included as metadata manually entered by curators in the future.
-
-In other areas of biological research, such as ecology and conservation biology, it has been shown that at least some data on lineage divergence represents a relevant improvement for testing alternative hypothesis using phylogenetic distance (Webb, Ackerly, \& Kembel, 2008).
-Hence, we integrated into datelife's workflow different ways of creating branch lengths in the absence of starting branch length information for taxa lacking this information (BLADJ option).
-Making up branch lengths in this or other ways is accepted in scientific publications: Jetz, Thomas, Joy, Hartmann, and Mooers (2012), created a time-calibrated tree of all 9,993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
-(2018) created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37\% had molecular data; Smith and Brown (2018) constructed a tree of 353,185 seed plants where only 23\% had molecular data.
-Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch imputation methods.
-There has yet to be a thorough analysis of what can go wrong when one goes beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make sure results are substantially similar between trees fully reconstructed with molecular or other data, and trees that are reconstructed using taxonomy by resolving polytomies at random following a statistical model.
-
-\hypertarget{conclusions}{%
-\section{Conclusions}\label{conclusions}}
-
-Divergence time information is key to many areas of evolutionary studies: trait evolution,
-diversification, biogeography, macroecology and more. It is also crucial for science communication and education, but generating chronograms \emph{de novo} is difficult,
-especially for those who want to use phylogenies but who are not systematists, or
-do not have the time to acquire and develop the necessary knowledge and data curation skills.
-Moreover, years of primarily public funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
-
-\texttt{datelife} allows easy and fast summarization of publicly available information
-on time of lineage divergence.
-This provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identification of regions that require more research or that have conflicting information.
-Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
-\texttt{datelife} helps with awareness of the existing variation in expert time of divergence data, and will foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
-
-\hypertarget{availability}{%
-\section{Availability}\label{availability}}
-
-\texttt{datelife} is free and open source and it can be used through its current website
-\url{http://www.datelife.org/query/}, through its R package, and through Phylotastic's project web portal \url{http://phylo.cs.nmsu.edu:3000/}.
-\texttt{datelife}'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
-\texttt{datelife}'s R package stable version will be available
-for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
-using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R. Development versions
-are available from the GitHub repository (\url{https://github.com/phylotastic/datelife})
-and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.
-
-\hypertarget{supplementary-material}{%
-\section{Supplementary Material}\label{supplementary-material}}
-
-Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities are available at \href{https://github.com/LunaSare/datelifeMS1}{datelifeMS1}, \href{https://github.com/LunaSare/datelife_examples}{datelife\_examples}, and \href{https://github.com/LunaSare/datelife_benchmark}{datelife\_benchmark} repositories in LLSR's GitHub account.
-
-\hypertarget{funding}{%
-\section{Funding}\label{funding}}
-
-Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to Datelife project and DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), and the Phylotastic project Grant ABI-1458572.
-
-\hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
-
-We thank colleagues from the O'Meara Lab at the University
-of Tennesse Knoxville for suggestions, discussions and software testing.
-The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
-that led to initial work on this project.
-The team that assembled \texttt{datelife}'s first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
-The Open Tree of Life project that provides the open, metadata rich repository of
-trees used for \texttt{datelife}.
-The many scientists who publish their chronograms in an open, reusable form, and
-the scientists who curate them for deposition in the Open Tree of Life repository.
-The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.
-
-\newpage
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-\hypertarget{refs}{}
-\leavevmode\hypertarget{ref-antonelli2017supersmart}{}%
-Antonelli, A., Hettling, H., Condamine, F.
-L., Vos, K., Nilsson, R.
-H., Sanderson, M.
-J., \ldots{} Vos, R.
-A.
-(2017).
-Toward a self-updating platform for estimating rates of speciation and migration, ages, and relationships of Taxa.
-\emph{Systematic Biology}, \emph{66}(2), 153--166.
-\url{https://doi.org/10.1093/sysbio/syw066}
-
-\leavevmode\hypertarget{ref-Bapst2012a}{}%
-Bapst, D.
-W.
-(2012).
-Paleotree: An R package for paleontological and phylogenetic analyses of evolution.
-\emph{Methods in Ecology and Evolution}, \emph{3}(5), 803--807.
-\url{https://doi.org/10.1111/j.2041-210X.2012.00223.x}
-
-\leavevmode\hypertarget{ref-barba2018constraining}{}%
-Barba-Montoya, J., Reis, M.
-dos, Schneider, H., Donoghue, P.
-C., \& Yang, Z.
-(2018).
-Constraining uncertainty in the timescale of angiosperm evolution and the veracity of a cretaceous terrestrial revolution.
-\emph{New Phytologist}, \emph{218}(2), 819--834.
-
-\leavevmode\hypertarget{ref-barker2012going}{}%
-Barker, F.
-K., Burns, K.
-J., Klicka, J., Lanyon, S.
-M., \& Lovette, I.
-J.
-(2012).
-Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds.
-\emph{Systematic Biology}, \emph{62}(2), 298--320.
-
-\leavevmode\hypertarget{ref-barker2015new}{}%
-Barker, F.
-K., Burns, K.
-J., Klicka, J., Lanyon, S.
-M., \& Lovette, I.
-J.
-(2015).
-New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
-\emph{The Auk: Ornithological Advances}, \emph{132}(2), 333--348.
-
-\leavevmode\hypertarget{ref-Boyle2013}{}%
-Boyle, B., Hopkins, N., Lu, Z., Raygoza Garay, J.
-A., Mozzherin, D., Rees, T., \ldots{} Enquist, B.
-J.
-(2013).
-The taxonomic name resolution service: An online tool for automated standardization of plant names.
-\emph{BMC Bioinformatics}, \emph{14}(1).
-\url{https://doi.org/10.1186/1471-2105-14-16}
-
-\leavevmode\hypertarget{ref-Britton2007}{}%
-Britton, T., Anderson, C.
-L., Jacquet, D., Lundqvist, S., \& Bremer, K.
-(2007).
-Estimating Divergence Times in Large Phylogenetic Trees.
-\emph{Systematic Biology}, \emph{56}(788777878), 741--752.
-\url{https://doi.org/10.1080/10635150701613783}
-
-\leavevmode\hypertarget{ref-burns2014phylogenetics}{}%
-Burns, K.
-J., Shultz, A.
-J., Title, P.
-O., Mason, N.
-A., Barker, F.
-K., Klicka, J., \ldots{} Lovette, I.
-J.
-(2014).
-Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds.
-\emph{Molecular Phylogenetics and Evolution}, \emph{75}, 41--77.
-
-\leavevmode\hypertarget{ref-Chamberlain2018}{}%
-Chamberlain, S. (2018). \emph{bold: Interface to Bold Systems API}. Retrieved from \url{https://CRAN.R-project.org/package=bold}
-
-\leavevmode\hypertarget{ref-Chamberlain2013}{}%
-Chamberlain, S.
-A., \& Szöcs, E.
-(2013).
-taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}.
-\emph{F1000Research}, \emph{2}(191), 1--29.
-\url{https://doi.org/10.12688/f1000research.2-191.v2}
-
-\leavevmode\hypertarget{ref-claramunt2015new}{}%
-Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{Science Advances}, \emph{1}(11), e1501005.
-
-\leavevmode\hypertarget{ref-Felsenstein1985a}{}%
-Felsenstein, J. (1985). Phylogenies and the Comparative Method. \emph{The American Naturalist}, \emph{125}(1), 1--15. Retrieved from \url{http://www.jstor.org/stable/2461605}
-
-\leavevmode\hypertarget{ref-gibb2015new}{}%
-Gibb, G.
-C., England, R., Hartig, G., McLenachan, P.
-A., Taylor Smith, B.
-L., McComish, B.
-J., \ldots{} Penny, D.
-(2015).
-New zealand passerines help clarify the diversification of major songbird lineages during the oligocene.
-\emph{Genome Biology and Evolution}, \emph{7}(11), 2983--2995.
-
-\leavevmode\hypertarget{ref-Harmon2008}{}%
-Harmon, L., Weir, J., Brock, C., Glor, R., \& Challenger, W. (2008). GEIGER: investigating evolutionary radiations. \emph{Bioinformatics}, \emph{24}, 129--131.
-
-\leavevmode\hypertarget{ref-Hedges2015}{}%
-Hedges, S.
-B., Marin, J., Suleski, M., Paymer, M., \& Kumar, S.
-(2015).
-Tree of life reveals clock-like speciation and diversification.
-\emph{Molecular Biology and Evolution}, \emph{32}(4), 835--845.
-\url{https://doi.org/10.1093/molbev/msv037}
-
-\leavevmode\hypertarget{ref-Heibl2008}{}%
-Heibl, C. (2008). \emph{PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages.} Retrieved from \url{http://www.christophheibl.de/Rpackages.html}
-
-\leavevmode\hypertarget{ref-hooper2017chromosomal}{}%
-Hooper, D. M., \& Price, T. D. (2017). Chromosomal inversion differences correlate with range overlap in passerine birds. \emph{Nature Ecology \& Evolution}, \emph{1}(10), 1526.
-
-\leavevmode\hypertarget{ref-Jetz2012}{}%
-Jetz, W., Thomas, G., Joy, J.
-J.
-B., Hartmann, K., \& Mooers, A.
-(2012).
-The global diversity of birds in space and time.
-\emph{Nature}, \emph{491}(7424), 444--448.
-\url{https://doi.org/10.1038/nature11631}
-
-\leavevmode\hypertarget{ref-Kumar2017}{}%
-Kumar, S., Stecher, G., Suleski, M., \& Hedges, S.
-B.
-(2017).
-TimeTree: A Resource for Timelines, Timetrees, and Divergence Times.
-\emph{Molecular Biology and Evolution}, \emph{34}(7), 1812--1819.
-\url{https://doi.org/10.1093/molbev/msx116}
-
-\leavevmode\hypertarget{ref-magallon2015metacalibrated}{}%
-Magallón, S., Gómez-Acevedo, S., Sánchez-Reyes, L.
-L., \& Hernández-Hernández, T.
-(2015).
-A metacalibrated time-tree documents the early rise of flowering plant phylogenetic diversity.
-\emph{New Phytologist}, \emph{207}(2), 437--453.
-
-\leavevmode\hypertarget{ref-mctavish2015phylesystem}{}%
-McTavish, E.
-J., Hinchliff, C.
-E., Allman, J.
-F., Brown, J.
-W., Cranston, K.
-A., Holder, M.
-T., \ldots{} Smith, S.
-A.
-(2015).
-Phylesystem: A git-based data store for community-curated phylogenetic estimates.
-\emph{Bioinformatics}, \emph{31}(17), 2794--2800.
-
-\leavevmode\hypertarget{ref-Michonneau2016}{}%
-Michonneau, F., Brown, J.
-W., \& Winter, D.
-J.
-(2016).
-rotl: an R package to interact with the Open Tree of Life data.
-\emph{Methods in Ecology and Evolution}, \emph{7}(12), 1476--1481.
-\url{https://doi.org/10.1111/2041-210X.12593}
-
-\leavevmode\hypertarget{ref-Morlon2014}{}%
-Morlon, H. (2014). Phylogenetic approaches for studying diversification. \emph{Ecology Letters}, \emph{17}(4), 508--525. \url{https://doi.org/10.1111/ele.12251}
-
-\leavevmode\hypertarget{ref-Ooms2018}{}%
-Ooms, J., \& Chamberlain, S. (2018). \emph{Phylocomr: Interface to 'phylocom'}. Retrieved from \url{https://CRAN.R-project.org/package=phylocomr}
-
-\leavevmode\hypertarget{ref-Paradis2004}{}%
-Paradis, E., Claude, J., \& Strimmer, K. (2004). APE: analyses of phylogenetics and evolution in R language. \emph{Bioinformatics}, \emph{20}, 289--290.
-
-\leavevmode\hypertarget{ref-posadas2006historical}{}%
-Posadas, P., Crisci, J. V., \& Katinas, L. (2006). Historical biogeography: A review of its basic concepts and critical issues. \emph{Journal of Arid Environments}, \emph{66}(3), 389--403.
-
-\leavevmode\hypertarget{ref-price2014niche}{}%
-Price, T.
-D., Hooper, D.
-M., Buchanan, C.
-D., Johansson, U.
-S., Tietze, D.
-T., Alström, P., \ldots{} others.
-(2014).
-Niche filling slows the diversification of himalayan songbirds.
-\emph{Nature}, \emph{509}(7499), 222.
-
-\leavevmode\hypertarget{ref-rabosky2018inverse}{}%
-Rabosky, D.
-L., Chang, J., Title, P.
-O., Cowman, P.
-F., Sallan, L., Friedman, M., \ldots{} others.
-(2018).
-An inverse latitudinal gradient in speciation rate for marine fishes.
-\emph{Nature}, \emph{559}(7714), 392.
-
-\leavevmode\hypertarget{ref-ramshaw1972time}{}%
-Ramshaw, J., Richardson, D., Meatyard, B., Brown, R., Richardson, M., Thompson, E., \& Boulter, D.
-(1972).
-The time of origin of the flowering plants determined by using amino acid sequence data of cytochrome c.
-\emph{New Phytologist}, \emph{71}(5), 773--779.
-
-\leavevmode\hypertarget{ref-RCoreTeam2018}{}%
-R Core Team. (2018). \emph{R: a language and environment for statistical computing}. Vienna, Austria: R Foundation for Statistical Computing.
-
-\leavevmode\hypertarget{ref-Revell2012}{}%
-Revell, L. J. (2012). Phytools: An r package for phylogenetic comparative biology (and other things). \emph{Methods in Ecology and Evolution}, \emph{3}, 217--223.
-
-\leavevmode\hypertarget{ref-sanderson2003r8s}{}%
-Sanderson, M. J. (2003). R8s: Inferring absolute rates of molecular evolution and divergence times in the absence of a molecular clock. \emph{Bioinformatics}, \emph{19}(2), 301--302.
-
-\leavevmode\hypertarget{ref-sanderson2001sources}{}%
-Sanderson, M.
-J., \& Doyle, J.
-A.
-(2001).
-Sources of error and confidence intervals in estimating the age of angiosperms from rbcL and 18S rDNA data.
-\emph{American Journal of Botany}, \emph{88}(8), 1499--1516.
-
-\leavevmode\hypertarget{ref-schenk2016sec}{}%
-Schenk, J. J. (2016). Consequences of secondary calibrations on divergence time estimates. \emph{PLoS ONE}, \emph{11}(1). \url{https://doi.org/10.1371/journal.pone.0148228}
-
-\leavevmode\hypertarget{ref-smith2018constructing}{}%
-Smith, S. A., \& Brown, J. W. (2018). Constructing a broadly inclusive seed plant phylogeny. \emph{American Journal of Botany}, \emph{105}(3), 302--314.
-
-\leavevmode\hypertarget{ref-Stoltzfus2013}{}%
-Stoltzfus, A., Lapp, H., Matasci, N., Deus, H., Sidlauskas, B., Zmasek, C.
-M., \ldots{} Jordan, G.
-(2013).
-Phylotastic!
-Making tree-of-life knowledge accessible, reusable and convenient.
-\emph{BMC Bioinformatics}, \emph{14}.
-\url{https://doi.org/10.1186/1471-2105-14-158}
-
-\leavevmode\hypertarget{ref-title2016macrophylogenies}{}%
-Title, P.
-O., \& Rabosky, D.
-L.
-(2016).
-Do Macrophylogenies Yield Stable Macroevolutionary Inferences?
-An Example from Squamate Reptiles.
-\emph{Systematic Biology}, syw102.
-\url{https://doi.org/10.1093/sysbio/syw102}
-
-\leavevmode\hypertarget{ref-Webb2000}{}%
-Webb, C. O. (2000). Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees. \emph{The American Naturalist}, \emph{156}(2), 145--155.
-
-\leavevmode\hypertarget{ref-Webb2008}{}%
-Webb, C.
-O., Ackerly, D.
-D., \& Kembel, S.
-W.
-(2008).
-Phylocom: Software for the analysis of phylogenetic community structure and trait evolution.
-\emph{Bioinformatics}, \emph{24}(18), 2098--2100.
-\url{https://doi.org/10.1093/bioinformatics/btn358}
-
-\leavevmode\hypertarget{ref-webb2005phylomatic}{}%
-Webb, C. O., \& Donoghue, M. J. (2005). Phylomatic: Tree assembly for applied phylogenetics. \emph{Molecular Ecology Notes}, \emph{5}(1), 181--183.
-
-\endgroup
-
-% in this file I leave the figure captions outside\ caption{} because I want them
-to be formatted in the same way as the general text (double spaced and linenumbered)
-\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}
-
-\begin{figure}[!h]
-\includegraphics{../figures/Fig1.pdf}
-\caption{Stylized DateLife workflow.
-This shows the general workflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{www.datelife.org/query/}.
-Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.}
-\label{fig:workflow}
-\end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:workflow}}
-% \end{center}
-% Stylized DateLife workflow.
-This shows the general workflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{http://www.datelife.org/}.
-Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
-\newpage
-\begin{figure}[!h]
-\includegraphics[width=1\linewidth]{../figures/fig_runtime_main.pdf}
-\caption{Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names.
-We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray).
-We also performed a search using the already processed query for comparison (light blue).}
-\label{fig:runtime1}
-\end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:runtime1}}
-% \end{center}
-%Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names.
-We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray).
-We also performed a search using the already processed query for comparison (light blue).
-\newpage
-\begin{figure}[!h]
-\includegraphics[width=1\linewidth]{../figures/fig_schronograms1.pdf}
-\caption{Lineage through time (LTT) plots of source chronograms containing all or a subset of species from the bird family Fringillidae of true finches.
-Arrows indicate maximum age of each chronogram.
-Numbers reference to chronograms' original publications 1: Barker et al.
-(\protect\hyperlink{ref-barker2012going}{2012}), 2: Barker et al.
-(\protect\hyperlink{ref-barker2015new}{2015}), 3: Burns et al.
-(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), 4: Claramunt and Cracraft (\protect\hyperlink{ref-claramunt2015new}{2015}), 5: Gibb et al.
-(\protect\hyperlink{ref-gibb2015new}{2015}), 6: Hedges et al.
-(\protect\hyperlink{ref-Hedges2015}{2015}), 7: Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), 8: Jetz et al.
-(\protect\hyperlink{ref-Jetz2012}{2012}), 9: Price et al.
-(\protect\hyperlink{ref-price2014niche}{2014}).
-}
-\label{fig:schronograms1}
-\end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:schronograms1}}
-% \end{center}
-%Lineage through time (LTT) plots of source chronograms containing all or a subset of species from the bird family Fringillidae of true finches.
-Arrows indicate maximum age of each chronogram.
-Numbers reference to chronograms' original publications 1: Barker et al.
-(\protect\hyperlink{ref-barker2012going}{2012}), 2: Barker et al.
-(\protect\hyperlink{ref-barker2015new}{2015}), 3: Burns et al.
-(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), 4: Claramunt and Cracraft (\protect\hyperlink{ref-claramunt2015new}{2015}), 5: Gibb et al.
-(\protect\hyperlink{ref-gibb2015new}{2015}), 6: Hedges et al.
-(\protect\hyperlink{ref-Hedges2015}{2015}), 7: Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), 8: Jetz et al.
-(\protect\hyperlink{ref-Jetz2012}{2012}), 9: Price et al.
-(\protect\hyperlink{ref-price2014niche}{2014}).
-\newpage
-\begin{figure}[!h]
-\includegraphics{../figures/fig_summaries.pdf}
-\caption{LTT plots of median (top) and Supermatrix Distance Method (SDM; bottom) chronograms summarising information from source chronograms found for the Fringillidae.
-Arrows indicate tree maximum age.}
-\label{fig:summaries}
-\end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:summaries}}
-% \end{center}
-% LTT plots of median (top) and Supermatrix Distance Method (SDM; bottom) chronograms summarising information from source chronograms found for the Fringillidae. Arrows indicate tree maximum age.
-\newpage
-\begin{figure}[!h]
-\includegraphics{../figures/fig_crossval_bladj.pdf}
-\caption{LTT plots showing results from the cross-validation analyses of trees without branch lengths dated using BLADJ.
-The dating analysis can only be performed in trees with more than 2 tips, thus excluding chronogram from study 4; its data was still used as calibration for the other source chronograms.}
-\label{fig:cvbladj}
-\end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:cvbladj}}
-% \end{center}
-% LTT plots showing results from the cross-validation analyses of trees without branch lengths dated using BLADJ.
-The dating analysis can only be performed in trees with more than 2 tips, thus excluding chronogram from study 4; its data was still used as calibration for the other source chronograms.
-\newpage
-\begin{figure}[!h]
-\includegraphics{../figures/fig_crossval_boldsumm.pdf}
-\caption{LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8.
-We could construct a tree with branch lengths for all source chronograms.
-However, dating with PATHd8 was only successful in three source chronograms shown here.}
-\label{fig:cvbold}
-\end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:cvbold}}
-% \end{center}
-FIGURES
-%LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8.
-We could construct a tree with branch lengths for all source chronograms.
-However, dating with PATHd8 was only successful in three source chronograms shown here.
-
-%\end{linenumbers}
-
-\end{document}
+\DeclareUnicodeCharacter{0301}{*************************************}

### 2022-03-04 manuscript/manuscript.Rmd
word count and other fixes

@@
 
+bibliography      : ["paper_references.bib"]
 keywords          : "Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations"
-wordcount         : "2949"
-
-bibliography      : ["paper_references.bib"]
-
+wordcount         : "`r wordcountaddin::word_count()`"
 floatsintext      : no
@@
 Notably, building a chronogram is not an easy task.
-First, it requires obtaining and curating genetic data to construct an homology hypothesis as a genetic alignment; then, it is important to carefully choose and apply the software that will use the alignment to reconstruct an evolutionary hypothesis in the form of a phylogeny; next, calibrations, i.e., independent age data points from the fossil record or other suitable geologic events have to be obtained from the literature --or by going into the field; fourthly, calibrations have to be carefully placed on the reconstructed phylogeny, demonstrating a biological and geological understanding of their limits and applicability; finally, the appropriate software and model of evolution to estimate ellapsed time since evolutionary events on the phylogeny has to be chosen and applied.
-In sum, building chronograms require specialized biological training and non-negligible budget, human and time resources.
-Still, chronograms represent key knowledge for the study of natural processes in many areas of scientific research, from developmental to conservation biology [@Felsenstein1985a; @Webb2000], from historical biogeography to species diversification [@posadas2006historical; @Morlon2014].
-Because of their importance for biological research, chronograms are still constantly being built by experts in the field, and have been published in peer-reviewed publications in a steady increase for the last two decades, amounting to at least four thousand individual chronograms [@Kumar2017].
+First, it requires obtaining and curating genetic data to construct an homology hypothesis as a genetic alignment; then, it is important to carefully choose and apply the software that will use the alignment to reconstruct an evolutionary hypothesis in the form of a phylogeny; thirdly, calibrations, i.e., independent age data points from the fossil record or other suitable geologic events have to be obtained from the literature --or by going into the field; next, calibrations have to be carefully placed on the reconstructed phylogeny, demonstrating a biological and geological understanding of their limits and applicability; finally, the appropriate software and model of evolution to estimate ellapsed time since evolutionary events on the phylogeny has to be chosen and applied.
+In sum, building chronograms require specialized biological training and a non-negligible amount of human, time and monetary resources.
 
-This state-of-the-art scientific data on time of evolutionary origin that has already been produced could be leveraged for the benefit of biological research, and there has been an urge for promoting, normlizing, and facilitating its reuse in research and education [@webb2005phylomatic; @Stoltzfus2013].
+Because of their importance for biological research, chronograms are constantly being built by experts in the field, and have been published in a steady increase for the last two decades, amounting to around four thousand individual chronograms published in peer-reviewed journals [@Kumar2017].
+Chronograms represent key knowledge for the study of natural processes in many areas of scientific research, from developmental to conservation biology [@Felsenstein1985a; @Webb2000], from historical biogeography to species diversification [@posadas2006historical; @Morlon2014].
+As such, the state-of-the-art scientific data on time of evolutionary origin that has already been produced and curated could be leveraged with beneficial results, and there has been an urge for promoting, normalizing, and facilitating reuse of scientific data for research and education [@webb2005phylomatic; @Stoltzfus2013].
 <!-- This has short-term advantages for research relying on this information, and long term advantages such as the advancement of science [@].-->
-We identify that a tool for efficient reuse of state-of-the-art scientific data should have an open and fully public database storing data in a computer readable format [@vos2012nexml], an automatised and programatic way of accessing the data [@Stoltzfus2013], and straightforward means of comparing and summarizing data as needed by the user [].
-The TreeBASE project started as an open and public database for state-of-the-art chronograms and other evolutionary research data in computer readable formats, and it currently supports programatic data accession [@piel2002treebase; @piel2009treebase, @treebaseAPIs], unfortunately, it is no longer under active development.
+We identify that any tool for efficient reuse of state-of-the-art scientific data should have a versioned, open and fully public database storing information in a computer readable format [@vos2012nexml], an automatised and programatic way of accessing the data [@Stoltzfus2013], and straightforward means of comparing and summarizing data as needed by the user [@].
+The DateLife project was born as a prototype tool aiming to provide these services, and was developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA [@Stoltzfus2013].
+Here we present the full implementation of DateLife, available as an R package `datelife` and as a website application with a graphical user interface at www.datelife.org/query/.
+The current implementation of the `datelife` R package is available on CRAN [@] and features an algorithm for automatic curation and maintenance of an open database of chronograms pulled from OpenTree's public repository [@mctavish2015phylesystem], and methods to summarize and compare source chronograms
+<!--, Functions to visualize and graphically compare source and summary chronograms are available at `datelifeplot` not on CRAN yet.-->
+
+<!--The TreeBASE project started as an open and public database for state-of-the-art chronograms and other evolutionary research data in computer readable formats, and it currently supports programatic data accession [@piel2002treebase; @piel2009treebase, @treebaseAPIs], unfortunately, it is no longer under active development.
 <!--No need to mention SuperTreeBASE as it is basically the same as treebase but on a more stable server-->
-The Open Tree of Life project [OpenTree, @opentreeoflife2019synth] has a system for storing chronograms and other evolutionary information in an open and public database that is programatically accessible, but it does not yet support phylogenetic queries of age data nor chronogram summaries.
-
-The DateLife project was born as a prototype service aiming to provide tools for easy reuse and summary of state-of-the-art time of lineage evolutionary origin, and was developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA [@Stoltzfus2013].
-Here we present the full implementation of the DateLife services, available as an R package `datelife` and an application with a graphical user interface web site at www.datelife.org/query/.
-The current implementation of the `datelife` R package features an algorithm for automatic curation and maintenance of an open database of chronograms pulled from OpenTree's open repository [@mctavish2015phylesystem], and methods to summarize and compare source chronograms
-<!--, and new functions to visualize and graphically compare source and summary chronograms.-->
+<!--The Open Tree of Life project [OpenTree, @opentreeoflife2019synth] has a system for storing chronograms and other evolutionary information in an open and public database that is programatically accessible, but it does not yet support phylogenetic queries of age data nor chronogram summaries.-->
 
@@
 
-The general `datelife` workflow is shown in figure \ref{fig:workflow}:
+The general `datelife` workflow is shown in figure \ref{fig:workflow}, briefly:
 
-\begin{enumerate}
-\item It starts with an input consisting of at least two taxon names, which can be provided in two different forms: as a comma separated character string, or as tip labels on a tree.
+1.
+It starts with an input consisting of at least two taxon names, which can be provided in two different forms: as a comma separated character string, or as tip labels on a tree.
 If input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004].
 The input tree is not required to have branch lengths.
-\item Input taxon names are processed with the Taxonomic Name Resolution Service [TNRS, @Boyle2013] implemented with OpenTree services [@opentreeAPIs].
-TNRS detects, corrects and standardizes misspellings and typos, variant spellings and authorities,
-and nomenclatural synonyms to standardized taxonomic names. This increases the probability of correctly finding the input taxon names in the chronogram database.
-
-\item The current version of `datelife` only accepts scientific taxonomic names as input.
+1.
+Input taxon names are processed to detect, correct and standardize name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard.
+To do this, `datelife` implements the Taxonomic Name Resolution Service [TNRS; @Boyle2013] provided by OpenTree [@opentreeAPIs].
+TNRS increases the probability of correctly finding the input taxon names in the chronogram database.
+1.
+The current version of `datelife` only accepts scientific taxonomic names as input.
 Names can belong to any taxonomic group or binomial specific.
@@
 If the flag is inactive, `datelife` will drop the "inclusive" taxon names from input.
-\item The cleaned input taxon names are saved as a special R object (of a newly defined class \texttt{datelifeQuery}) that contains the processed names, the corresponding taxonomic id numbers, and the topology of theinput tree if any was provided.
+1.
+The cleaned input taxon names are saved as a special R object (of a newly defined class \texttt{datelifeQuery}) that contains the processed names, the corresponding taxonomic id numbers, and the topology of theinput tree if any was provided.
 The \texttt{datelifeQuery} object is used next to search the chronogram database.
-\item Chronograms with at least two matching input taxon names are identified and pruned down to preserve only input taxon names as tips.
+1.
+Chronograms with at least two matching input taxon names are identified and pruned down to preserve only input taxon names as tips.
 Then, each pruned chronogram is transformed to a patristic distance matrix.
@@
 The matrices are associated to the citation of the original study and stored as an R object of class \texttt{datelifeResult}.
-\item  At this point, various summary data can be obtained to inform decisions for the next steps of the analysis workflow.
+1.
+At this point, various summary data can be obtained to inform decisions for the next steps of the analysis workflow.
 Types of summary information provided are: a) all pruned source chronograms, b) age of the MRCA (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of successful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most matching input taxon names.
-\item To construct summary trees we use a fixed topology, either provided by the user, taken from the literature or obtained from expert
-phylogenetic information, such as the OpenTree synthetic tree.
+1.
+To construct summary trees we use a fixed topology, either provided by the user, taken from the literature or obtained from expert phylogenetic information, such as the OpenTree synthetic tree.
 Then we summarize source chronograms into a single patristic distance matrix using a method chosen by the user.
@@
 Finally, `datelife` applies the summarized time distances as node calibrations to date the given topology with BLADJ [@webb2005phylomatic].
-<!--ADD FIGURE: mock example explaining difference between using summary ages and raw ages to calibrate a topology.
+1. If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms.
+1.
+Users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
+
+The `datelife` workflow relies on functions from several R packages that we acknowledge: rotl [@Michonneau2016], ape [@paradis2004],
+geiger [@Harmon2008], paleotree [@Bapst2012a], bold [@Chamberlain2018], phytools [@Revell2012], taxize [@Chamberlain2013; @Chamberlain2018], phyloch [@Heibl2008], and phylocomr [@Ooms2018].
+<!-- TODO ADD FIGURE: mock example explaining difference between using summary ages and raw ages to calibrate a topology.
 Then show it in the biological example.
@@
 %a taxonomic tree-->
-\item  If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms.
-\item  Users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
-\end{enumerate}
 
@@
 Query processing time increases roughly linearly with number of input taxon names, and
-increases considerably if TNRS [@Boyle2013] is activated.
+increases considerably if TNRS is activated.
 <!-- Defining tnrs here again, bc we only talk about it two times, so it easy to forget it -->
@@
 
+## Case study
+
 We illustrate the `datelife` workflow using the family of true finches, Fringillidae as an example.
 <!--We used `r cite_r("r-references.bib")` for all our analyses.-->
-
-## Case study
-
 A college educator wishes to obtain state-of-the-art data on time of evolutionary origin of species belonging to the true finches for their class.
@@
 We also demonstrate the usage of PATHd8 [@Britton2007] as an alternative method to BLADJ.
-For this, we run a `datelife` branch length reconstruction that searches for DNA sequence data from the Barcode of Life Data System [BOLD; ratnasingham2007bold] to generate branch lengths.
+For this, we run a `datelife` branch length reconstruction that searches for DNA sequence data from the Barcode of Life Data System [BOLD; @ratnasingham2007bold] to generate branch lengths.
 We were able to successfully generate a tree with BOLD branch lengths for all of the Fringillidae source chronograms.
@@
 It is an open service that does not require any expert biological knowledge from users --besides the names of the organisms they want to work with, for any of its functionalities.
-The `datelife` workflow builds off of functions from several R packages that must be acknowledge (rotl [@Michonneau2016], ape [@paradis2004],
-geiger [@Harmon2008], paleotree [@Bapst2012a], bold [@Chamberlain2018], phytools [@Revell2012], taxize [@Chamberlain2013; @Chamberlain2018], phyloch [@Heibl2008], and phylocomr [@Ooms2018]).
 
-At the time of writing of this manuscript (`r format(Sys.time(), "%b %d, %Y")`), `datelife`'s database has `r length(opentree_chronograms$trees)` chronograms, pulled entirely from OpenTree's database, th ePhylesystem [@mctavish2015phylesystem].
+At the time of writing of this manuscript (`r format(Sys.time(), "%b %d, %Y")`), `datelife`'s database has `r length(opentree_chronograms$trees)` chronograms, pulled entirely from OpenTree's database, the Phylesystem [@mctavish2015phylesystem].
 A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time.
@@
 `datelife`'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
-Users can also upload new chronograms to OpenTree themselves, trigger and update of the `datelife` database to incorporate the new chronograms, and and have them immediately available for analysis.
+Users can also upload new chronograms to OpenTree themselves, and trigger an update of the `datelife` database to incorporate the new chronograms, to have them immediately available for analysis.
 
@@
 Divergence time information is key to many areas of evolutionary studies: trait evolution,
-diversification, biogeography, macroecology and more. It is also crucial for science communication and education, but generating chronograms _de novo_ is difficult,
+diversification, biogeography, macroecology and more. It is also crucial for science communication and education, but generating chronograms is difficult,
 especially for those who want to use phylogenies but who are not systematists, or
@@
 <!--Mention again uses shown here and outline potential uses not shown here-->
-`datelife` allows easy and fast summarization of publicly available information
+The `datelife` R package allows easy and fast summarization of publicly available information
 on time of lineage divergence.
@@
 Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
-`datelife` helps with awareness of the existing variation in expert time of divergence data, and will foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
+The DateLife project helps with awareness of the existing variation in expert time of divergence data, and will foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
 

### 2022-03-04 manuscript/manuscript.tex
word count and other fixes

@@
 \shorttitle{DateLife: revealing the dated Tree of Life}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 2949}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 3670}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 Notably, building a chronogram is not an easy task.
-First, it requires obtaining and curating genetic data to construct an homology hypothesis as a genetic alignment; then, it is important to carefully choose and apply the software that will use the alignment to reconstruct an evolutionary hypothesis in the form of a phylogeny; next, calibrations, i.e., independent age data points from the fossil record or other suitable geologic events have to be obtained from the literature --or by going into the field; fourthly, calibrations have to be carefully placed on the reconstructed phylogeny, demonstrating a biological and geological understanding of their limits and applicability; finally, the appropriate software and model of evolution to estimate ellapsed time since evolutionary events on the phylogeny has to be chosen and applied.
-In sum, building chronograms require specialized biological training and non-negligible budget, human and time resources.
-Still, chronograms represent key knowledge for the study of natural processes in many areas of scientific research, from developmental to conservation biology (Felsenstein, 1985; Campbell O.
+First, it requires obtaining and curating genetic data to construct an homology hypothesis as a genetic alignment; then, it is important to carefully choose and apply the software that will use the alignment to reconstruct an evolutionary hypothesis in the form of a phylogeny; thirdly, calibrations, i.e., independent age data points from the fossil record or other suitable geologic events have to be obtained from the literature --or by going into the field; next, calibrations have to be carefully placed on the reconstructed phylogeny, demonstrating a biological and geological understanding of their limits and applicability; finally, the appropriate software and model of evolution to estimate ellapsed time since evolutionary events on the phylogeny has to be chosen and applied.
+In sum, building chronograms require specialized biological training and a non-negligible amount of human, time and monetary resources.
+
+Because of their importance for biological research, chronograms are constantly being built by experts in the field, and have been published in a steady increase for the last two decades, amounting to around four thousand individual chronograms published in peer-reviewed journals (Kumar, Stecher, Suleski, \& Hedges, 2017).
+Chronograms represent key knowledge for the study of natural processes in many areas of scientific research, from developmental to conservation biology (Felsenstein, 1985; Campbell O.
 Webb, 2000), from historical biogeography to species diversification (Morlon, 2014; Posadas, Crisci, \& Katinas, 2006).
-Because of their importance for biological research, chronograms are still constantly being built by experts in the field, and have been published in peer-reviewed publications in a steady increase for the last two decades, amounting to at least four thousand individual chronograms (Kumar, Stecher, Suleski, \& Hedges, 2017).
-
-This state-of-the-art scientific data on time of evolutionary origin that has already been produced could be leveraged for the benefit of biological research, and there has been an urge for promoting, normlizing, and facilitating its reuse in research and education (Stoltzfus et al., 2013; Campbell O.
+As such, the state-of-the-art scientific data on time of evolutionary origin that has already been produced and curated could be leveraged with beneficial results, and there has been an urge for promoting, normalizing, and facilitating reuse of scientific data for research and education (Stoltzfus et al., 2013; Campbell O.
 Webb \& Donoghue, 2005).
-We identify that a tool for efficient reuse of state-of-the-art scientific data should have an open and fully public database storing data in a computer readable format (R.
-A.
-Vos et al., 2012), an automatised and programatic way of accessing the data (Stoltzfus et al., 2013), and straightforward means of comparing and summarizing data as needed by the user {[}{]}.
-The TreeBASE project started as an open and public database for state-of-the-art chronograms and other evolutionary research data in computer readable formats, and it currently supports programatic data accession R.
-Vos (n.d.), unfortunately, it is no longer under active development.
-The Open Tree of Life project (OpenTree, OpenTreeOfLife et al., 2019) has a system for storing chronograms and other evolutionary information in an open and public database that is programatically accessible, but it does not yet support phylogenetic queries of age data nor chronogram summaries.
-
-The DateLife project was born as a prototype service aiming to provide tools for easy reuse and summary of state-of-the-art time of lineage evolutionary origin, and was developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA (Stoltzfus et al., 2013).
-Here we present the full implementation of the DateLife services, available as an R package \texttt{datelife} and an application with a graphical user interface web site at www.datelife.org/query/.
-The current implementation of the \texttt{datelife} R package features an algorithm for automatic curation and maintenance of an open database of chronograms pulled from OpenTree's open repository (McTavish et al., 2015), and methods to summarize and compare source chronograms
+We identify that any tool for efficient reuse of state-of-the-art scientific data should have a versioned, open and fully public database storing information in a computer readable format (Vos et al., 2012), an automatised and programatic way of accessing the data (Stoltzfus et al., 2013), and straightforward means of comparing and summarizing data as needed by the user {[}@{]}.
+The DateLife project was born as a prototype tool aiming to provide these services, and was developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA (Stoltzfus et al., 2013).
+Here we present the full implementation of DateLife, available as an R package \texttt{datelife} and as a website application with a graphical user interface at www.datelife.org/query/.
+The current implementation of the \texttt{datelife} R package is available on CRAN {[}@{]} and features an algorithm for automatic curation and maintenance of an open database of chronograms pulled from OpenTree's public repository (McTavish et al., 2015), and methods to summarize and compare source chronograms
 
@@
 
-The general \texttt{datelife} workflow is shown in figure \ref{fig:workflow}:
+The general \texttt{datelife} workflow is shown in figure \ref{fig:workflow}, briefly:
 
 \begin{enumerate}
-\item It starts with an input consisting of at least two taxon names, which can be provided in two different forms: as a comma separated character string, or as tip labels on a tree.
-If input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004].
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  It starts with an input consisting of at least two taxon names, which can be provided in two different forms: as a comma separated character string, or as tip labels on a tree.
+If input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis, Claude, \& Strimmer, 2004).
 The input tree is not required to have branch lengths.
-\item Input taxon names are processed with the Taxonomic Name Resolution Service [TNRS, @Boyle2013] implemented with OpenTree services [@opentreeAPIs].
-TNRS detects, corrects and standardizes misspellings and typos, variant spellings and authorities,
-and nomenclatural synonyms to standardized taxonomic names. This increases the probability of correctly finding the input taxon names in the chronogram database.
-
-\item The current version of `datelife` only accepts scientific taxonomic names as input.
+\item
+  Input taxon names are processed to detect, correct and standardize name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard.
+To do this, \texttt{datelife} implements the Taxonomic Name Resolution Service {[}TNRS; Boyle et al.
+(2013){]} provided by OpenTree (OpenTreeOfLife et al., n.d.).
+TNRS increases the probability of correctly finding the input taxon names in the chronogram database.
+\item
+  The current version of \texttt{datelife} only accepts scientific taxonomic names as input.
 Names can belong to any taxonomic group or binomial specific.
-If an input taxon name belongs to an "inclusive" taxonomic group, i.e., a taxon above the species level, such as genus, family, etc.), `datelife` has two alternative behaviors defined by the "get species from taxon" flag.
-If the flag is active, `datelife` retrieves all species names within the "inclusive" taxonomic group and adds them to the input.
-If the flag is inactive, `datelife` will drop the "inclusive" taxon names from input.
-\item The cleaned input taxon names are saved as a special R object (of a newly defined class \texttt{datelifeQuery}) that contains the processed names, the corresponding taxonomic id numbers, and the topology of theinput tree if any was provided.
+If an input taxon name belongs to an ``inclusive'' taxonomic group, i.e., a taxon above the species level, such as genus, family, etc.), \texttt{datelife} has two alternative behaviors defined by the ``get species from taxon'' flag.
+If the flag is active, \texttt{datelife} retrieves all species names within the ``inclusive'' taxonomic group and adds them to the input.
+If the flag is inactive, \texttt{datelife} will drop the ``inclusive'' taxon names from input.
+\item
+  The cleaned input taxon names are saved as a special R object (of a newly defined class \texttt{datelifeQuery}) that contains the processed names, the corresponding taxonomic id numbers, and the topology of theinput tree if any was provided.
 The \texttt{datelifeQuery} object is used next to search the chronogram database.
-\item Chronograms with at least two matching input taxon names are identified and pruned down to preserve only input taxon names as tips.
+\item
+  Chronograms with at least two matching input taxon names are identified and pruned down to preserve only input taxon names as tips.
 Then, each pruned chronogram is transformed to a patristic distance matrix.
@@
 The matrices are associated to the citation of the original study and stored as an R object of class \texttt{datelifeResult}.
-\item  At this point, various summary data can be obtained to inform decisions for the next steps of the analysis workflow.
+\item
+  At this point, various summary data can be obtained to inform decisions for the next steps of the analysis workflow.
 Types of summary information provided are: a) all pruned source chronograms, b) age of the MRCA (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of successful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most matching input taxon names.
-\item To construct summary trees we use a fixed topology, either provided by the user, taken from the literature or obtained from expert
-phylogenetic information, such as the OpenTree synthetic tree.
+\item
+  To construct summary trees we use a fixed topology, either provided by the user, taken from the literature or obtained from expert phylogenetic information, such as the OpenTree synthetic tree.
 Then we summarize source chronograms into a single patristic distance matrix using a method chosen by the user.
-Summarizing method options available include Super Distance Matrix method [SDM, @Criscuolo2006] and any summary statistics such as median, minimum andmaximum ages.
-Finally, `datelife` applies the summarized time distances as node calibrations to date the given topology with BLADJ [@webb2005phylomatic].
-<!--ADD FIGURE: mock example explaining difference between using summary ages and raw ages to calibrate a topology.
-Then show it in the biological example.
-This will show what happens if a tree (phylogenetic conflict) or node (age) do not agree.-->
+Summarizing method options available include Super Distance Matrix method (SDM, Criscuolo, Berry, Douzery, \& Gascuel, 2006) and any summary statistics such as median, minimum andmaximum ages.
+Finally, \texttt{datelife} applies the summarized time distances as node calibrations to date the given topology with BLADJ (Campbell O.
+Webb \& Donoghue, 2005).
+\item
+  If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms.
+\item
+  Users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R ``phylo'' format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
+\end{enumerate}
 
-<!-- \item  Alternatively, time of lineage divergence obtained from the pruned chronograms can be used directly as secondary calibration points to date a tree with or without branch lengths containing some or all input taxon names.
-%a taxonomic tree-->
-\item  If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms.
-\item  Users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
-\end{enumerate}
+The \texttt{datelife} workflow relies on functions from several R packages that we acknowledge: rotl (Michonneau, Brown, \& Winter, 2016), ape (Paradis et al., 2004),
+geiger (Harmon, Weir, Brock, Glor, \& Challenger, 2008), paleotree (Bapst, 2012), bold (Chamberlain et al., 2019), phytools (Revell, 2012), taxize (Chamberlain \& Szöcs, 2013; Chamberlain et al., 2019), phyloch (Heibl, 2008), and phylocomr (Ooms \& Chamberlain, 2018).
 
@@
 Query processing time increases roughly linearly with number of input taxon names, and
-increases considerably if TNRS (Boyle et al., 2013) is activated.
+increases considerably if TNRS is activated.
 Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
@@
 
-We illustrate the \texttt{datelife} workflow using the family of true finches, Fringillidae as an example.
-
 \hypertarget{case-study}{%
@@
 
+We illustrate the \texttt{datelife} workflow using the family of true finches, Fringillidae as an example.
 A college educator wishes to obtain state-of-the-art data on time of evolutionary origin of species belonging to the true finches for their class.
@@
 We also demonstrate the usage of PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007) as an alternative method to BLADJ.
-For this, we run a \texttt{datelife} branch length reconstruction that searches for DNA sequence data from the Barcode of Life Data System {[}BOLD; ratnasingham2007bold{]} to generate branch lengths.
+For this, we run a \texttt{datelife} branch length reconstruction that searches for DNA sequence data from the Barcode of Life Data System {[}BOLD; Ratnasingham and Hebert (2007){]} to generate branch lengths.
 We were able to successfully generate a tree with BOLD branch lengths for all of the Fringillidae source chronograms.
@@
 It is an open service that does not require any expert biological knowledge from users --besides the names of the organisms they want to work with, for any of its functionalities.
-The \texttt{datelife} workflow builds off of functions from several R packages that must be acknowledge (rotl (Michonneau, Brown, \& Winter, 2016), ape (Paradis, Claude, \& Strimmer, 2004),
-geiger (Harmon, Weir, Brock, Glor, \& Challenger, 2008), paleotree (Bapst, 2012), bold (Chamberlain et al., 2019), phytools (Revell, 2012), taxize (Chamberlain \& Szöcs, 2013; Chamberlain et al., 2019), phyloch (Heibl, 2008), and phylocomr (Ooms \& Chamberlain, 2018)).
 
-At the time of writing of this manuscript (Mar 03, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, th ePhylesystem (McTavish et al., 2015).
+At the time of writing of this manuscript (Mar 04, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
 A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time.
@@
 \texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
-Users can also upload new chronograms to OpenTree themselves, trigger and update of the \texttt{datelife} database to incorporate the new chronograms, and and have them immediately available for analysis.
+Users can also upload new chronograms to OpenTree themselves, and trigger an update of the \texttt{datelife} database to incorporate the new chronograms, to have them immediately available for analysis.
 
@@
 Divergence time information is key to many areas of evolutionary studies: trait evolution,
-diversification, biogeography, macroecology and more. It is also crucial for science communication and education, but generating chronograms \emph{de novo} is difficult,
+diversification, biogeography, macroecology and more. It is also crucial for science communication and education, but generating chronograms is difficult,
 especially for those who want to use phylogenies but who are not systematists, or
@@
 
-\texttt{datelife} allows easy and fast summarization of publicly available information
+The \texttt{datelife} R package allows easy and fast summarization of publicly available information
 on time of lineage divergence.
@@
 Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
-\texttt{datelife} helps with awareness of the existing variation in expert time of divergence data, and will foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
+The DateLife project helps with awareness of the existing variation in expert time of divergence data, and will foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
 
@@
 \url{https://doi.org/10.1093/sysbio/syw066}
+
+\leavevmode\vadjust pre{\hypertarget{ref-archie1986newick}{}}%
+Archie, J., Day, W.
+H., Felsenstein, J., Maddison, W., Meacham, C., Rohlf, F.
+J., \& Swofford, D.
+(1986).
+{The Newick tree format}.
+Retrieved from \url{\%7Bhttps://evolution.genetics.washington.edu/phylip/newicktree.html\%7D}
 
@@
 Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{Science Advances}, \emph{1}(11), e1501005.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Criscuolo2006}{}}%
+Criscuolo, A., Berry, V., Douzery, E.
+J.
+P., \& Gascuel, O.
+(2006).
+{SDM: A fast distance-based approach for (super)tree building in phylogenomics}.
+\emph{Systematic Biology}, \emph{55}(5), 740--755.
+\url{https://doi.org/10.1080/10635150600969872}
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-opentreeoflife2019synth}{}}%
-OpenTreeOfLife, Redelings, B., Reyes, L.
-L.
-S., Cranston, K.
+\leavevmode\vadjust pre{\hypertarget{ref-opentreeAPIs}{}}%
+OpenTreeOfLife, Redelings, B., Cranston, K.
 A., Allman, J., Holder, M.
@@
 J.
-(2019).
-Open tree of life synthetic tree (Version 12.3).
-Zenodo.
-\url{https://doi.org/10.5281/zenodo.3937742}
+(n.d.).
+{Open Tree of Life APIs v.
+3.0}.
+Retrieved from \url{\%7Bhttps://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\%7D}
 
@@
 Paradis, E., Claude, J., \& Strimmer, K. (2004). {APE: analyses of phylogenetics and evolution in R language}. \emph{Bioinformatics}, \emph{20}(2), 289--290.
-
-\leavevmode\vadjust pre{\hypertarget{ref-piel2002treebase}{}}%
-Piel, W.
-H., Donoghue, M., \& Sanderson, M.
-(2002).
-{TreeBASE : A database of phylogenetic information}.
-In J.
-Shimura, K.
-Wilson, \& D.
-Gordon (Eds.), \emph{To the interoperable {``catalog of life''} with partners} (pp.
-41--47).
-Tsukuba, Japan: National Institute for Environmental Studies.
-Retrieved from \url{http://phylogeny.harvard.edu/treebase.}
-
-\leavevmode\vadjust pre{\hypertarget{ref-piel2009treebase}{}}%
-Piel, W., Chan, L., Dominus, M., Ruan, J., Vos, R., \& Tannen, V. (2009). Treebase v. 2: A database of phylogenetic knowledge. E-biosphere. London.
 
@@
 \emph{New Phytologist}, \emph{71}(5), 773--779.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ratnasingham2007bold}{}}%
+Ratnasingham, S., \& Hebert, P. D. (2007). BOLD: The barcode of life data system (http://www. Barcodinglife. org). \emph{Molecular Ecology Notes}, \emph{7}(3), 355--364.
 
@@
 \url{https://doi.org/10.1093/sysbio/syw102}
-
-\leavevmode\vadjust pre{\hypertarget{ref-treebaseAPIs}{}}%
-Vos, R. (n.d.). {TreeBASE APIs}. Retrieved from \url{\%7Bhttps://github.com/TreeBASE/treebase/wiki/API\%7D}
 

### 2022-03-04 manuscript/manuscript.Rmd
figure 1 vertical mostly final

@@
 <!-- This has short-term advantages for research relying on this information, and long term advantages such as the advancement of science [@].-->
-We identify that any tool for efficient reuse of state-of-the-art scientific data should have a versioned, open and fully public database storing information in a computer readable format [@vos2012nexml], an automatised and programatic way of accessing the data [@Stoltzfus2013], and straightforward means of comparing and summarizing data as needed by the user [@].
+We identify that any tool for efficient reuse of state-of-the-art scientific data should have a versioned, open and fully public database storing information in a computer readable format [@vos2012nexml], an automatised and programatic way of accessing the data [@Stoltzfus2013], and straightforward means of comparing and summarizing data as needed by the user.
 The DateLife project was born as a prototype tool aiming to provide these services, and was developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA [@Stoltzfus2013].
 Here we present the full implementation of DateLife, available as an R package `datelife` and as a website application with a graphical user interface at www.datelife.org/query/.
-The current implementation of the `datelife` R package is available on CRAN [@] and features an algorithm for automatic curation and maintenance of an open database of chronograms pulled from OpenTree's public repository [@mctavish2015phylesystem], and methods to summarize and compare source chronograms
+The latest implementation of the `datelife` R package is available on CRAN (v`r packageVersion("datelife")`; @datelifeR) and features an algorithm for automatic curation and maintenance of an open database composed of chronograms pulled from OpenTree's public repository [@mctavish2015phylesystem], and methods to summarize chronograms and compare them.
 <!--, Functions to visualize and graphically compare source and summary chronograms are available at `datelifeplot` not on CRAN yet.-->
@@
 Input taxon names are processed to detect, correct and standardize name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard.
-To do this, `datelife` implements the Taxonomic Name Resolution Service [TNRS; @Boyle2013] provided by OpenTree [@opentreeAPIs].
+To do this, `datelife` implements the Taxonomic Name Resolution Service (TNRS; @Boyle2013) provided by OpenTree [@opentreeAPIs].
 TNRS increases the probability of correctly finding the input taxon names in the chronogram database.
@@
 A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time.
-As chronograms are added to Phylesystem, they are incorporated into an updated `datelife`'s database that is assigned a new version number, followed by a package release on CRAN [@].
+As chronograms are added to Phylesystem, they are incorporated into an updated `datelife`'s database that is assigned a new version number, followed by a package release on CRAN.
 `datelife`'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.

### 2022-03-04 manuscript/manuscript.tex
figure 1 vertical mostly final

@@
 \shorttitle{DateLife: revealing the dated Tree of Life}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 3670}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 3673}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 Webb \& Donoghue, 2005).
-We identify that any tool for efficient reuse of state-of-the-art scientific data should have a versioned, open and fully public database storing information in a computer readable format (Vos et al., 2012), an automatised and programatic way of accessing the data (Stoltzfus et al., 2013), and straightforward means of comparing and summarizing data as needed by the user {[}@{]}.
+We identify that any tool for efficient reuse of state-of-the-art scientific data should have a versioned, open and fully public database storing information in a computer readable format (Vos et al., 2012), an automatised and programatic way of accessing the data (Stoltzfus et al., 2013), and straightforward means of comparing and summarizing data as needed by the user.
 The DateLife project was born as a prototype tool aiming to provide these services, and was developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA (Stoltzfus et al., 2013).
 Here we present the full implementation of DateLife, available as an R package \texttt{datelife} and as a website application with a graphical user interface at www.datelife.org/query/.
-The current implementation of the \texttt{datelife} R package is available on CRAN {[}@{]} and features an algorithm for automatic curation and maintenance of an open database of chronograms pulled from OpenTree's public repository (McTavish et al., 2015), and methods to summarize and compare source chronograms
+The latest implementation of the \texttt{datelife} R package is available on CRAN (v0.6.2; Sanchez-Reyes et al.
+(2022)) and features an algorithm for automatic curation and maintenance of an open database composed of chronograms pulled from OpenTree's public repository (McTavish et al., 2015), and methods to summarize chronograms and compare them.
 
@@
   Input taxon names are processed to detect, correct and standardize name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard.
-To do this, \texttt{datelife} implements the Taxonomic Name Resolution Service {[}TNRS; Boyle et al.
-(2013){]} provided by OpenTree (OpenTreeOfLife et al., n.d.).
+To do this, \texttt{datelife} implements the Taxonomic Name Resolution Service (TNRS; Boyle et al.
+(2013)) provided by OpenTree (OpenTreeOfLife et al., n.d.).
 TNRS increases the probability of correctly finding the input taxon names in the chronogram database.
@@
 A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time.
-As chronograms are added to Phylesystem, they are incorporated into an updated \texttt{datelife}'s database that is assigned a new version number, followed by a package release on CRAN {[}@{]}.
+As chronograms are added to Phylesystem, they are incorporated into an updated \texttt{datelife}'s database that is assigned a new version number, followed by a package release on CRAN.
 \texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-datelifeR}{}}%
+Sanchez-Reyes, L.
+L., O'Meara, B., Eastman, J., Heath, T., Wright, A., Schliep, K., \ldots{} Alfaro, M.
+(2022).
+{datelife: Scientific Data on Time of Lineage Divergence for Your Taxa}.
+\emph{R Package Version 0.6.2}.
+Retrieved from \url{https://doi.org/10.5281/zenodo.593938}
+
 \leavevmode\vadjust pre{\hypertarget{ref-sanderson2003r8s}{}}%
@@
 \leavevmode\vadjust pre{\hypertarget{ref-sauquet2013practical}{}}%
-Sauquet, H. (2013). A practical guide to molecular dating. \emph{Comptes Rendus Palevol}, \emph{12}(6), 355--367.
+Sauquet, H. (2013). A practical guide to molecular dating. \emph{{Comptes Rendus Palevol}}, \emph{12}(6), 355--367.
 
@@
 \begin{figure}[!h]
-\includegraphics{../figures/figure1/figure1-horizontal.pdf}
+\includegraphics{../figures/figure1/figure1-vertical-final.pdf}
 \caption{Stylized DateLife workflow.

### 2022-03-07 manuscript/manuscript.Rmd
abstract and intro edits

@@
 abstract: |
-  Time of evolutionary origin is fundamental for research in the natural sciences, as well as for education, science communication and policy.
-  Despite an increased availability of fossil and molecular data, and time-efficient analytical techniques, achieving a high-quality reconstruction of time of evolutionary origin as a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is still a difficult and time-consuming task for a majority of interested parties.
-  Yet, the amount of published chronograms has increased significantly in the past two decades, and a non-negligeable proportion of these data have been steadily accumulating in public, open databases such as TreeBASE and Open Tree
-  of Life, exposing a wealth of expertly-curated and peer-reviewed data on time of evolutionary origin in a programatic and reusable way, for a large quantity and diversity of organisms.
-  This trend results from intensive and localized efforts for improving data sharing practices, as well as incentivizing open science in biology.
-Despite these trends, accessibility to state-of-the-art knowledge on time of evolutionary origin is still reduced.
+  Date estimates for times of evolutionary diveregences are key data for research in the natural sciences.
+These estimates also provide valuable information for for education, science communication and policy decisions.
+
+ Although achieving a high-quality reconstruction of time of evolutionary origin as a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is still a difficult and time-consuming task, the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms
+
+  When these estimates are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on time of evolutionary origin is exposed in a programatic and reusable way.
+  Intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
 
@@
   We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms.
-Additionally, time data extracted from source chronograms can be usedas secondary calibration points to add branch lengths proportional to absolute time to a tree topology using alternative dating methods.
-
-  Summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research in biology.
-How well this chronograms work for this purpose still needs to be tested.
+Additionally, time data extracted from source chronograms can be used as secondary calibration points to add branch lengths proportional to absolute time to a tree topology using alternative dating methods external to datelife.
 
@@
 # Introduction
+Chronograms are phylogenies with branch lengths proportional to time.
+<!-- If you're aiming for syst bio you don't want to start so general Readers know what chronorgrams and phylogenies are!.-->
+Chronograms provide key data for the study of natural processes in many areas of biological research, such as developmental  biology (cite) to conservation biology [@Felsenstein1985a; @Webb2000], historical biogeography, a species diversification [@posadas2006historical; @Morlon2014].
 
-In the natural sciences, absolute time of evolutionary origin of lineages is represented graphically as a phylogeny with branch lengths proportional to geologic time, also known as a chronogram.
-Notably, building a chronogram is not an easy task.
-First, it requires obtaining and curating genetic data to construct an homology hypothesis as a genetic alignment; then, it is important to carefully choose and apply the software that will use the alignment to reconstruct an evolutionary hypothesis in the form of a phylogeny; thirdly, calibrations, i.e., independent age data points from the fossil record or other suitable geologic events have to be obtained from the literature --or by going into the field; next, calibrations have to be carefully placed on the reconstructed phylogeny, demonstrating a biological and geological understanding of their limits and applicability; finally, the appropriate software and model of evolution to estimate ellapsed time since evolutionary events on the phylogeny has to be chosen and applied.
-In sum, building chronograms require specialized biological training and a non-negligible amount of human, time and monetary resources.
+Building a chronogram is not an easy task.
+It requires obtaining and curating data to construct a phylogeny; selecting and placing appropriate calibrations on the phylogeny using independent age data points from the fossil record or other dated events, and inferring the full dated tree.
+Estimating accurate chronograms generally requires specialized biological training, taxonomic domain knowledge, and a non-negligible amount of research time, computational resources and funding.
 
-Because of their importance for biological research, chronograms are constantly being built by experts in the field, and have been published in a steady increase for the last two decades, amounting to around four thousand individual chronograms published in peer-reviewed journals [@Kumar2017].
-Chronograms represent key knowledge for the study of natural processes in many areas of scientific research, from developmental to conservation biology [@Felsenstein1985a; @Webb2000], from historical biogeography to species diversification [@posadas2006historical; @Morlon2014].
-As such, the state-of-the-art scientific data on time of evolutionary origin that has already been produced and curated could be leveraged with beneficial results, and there has been an urge for promoting, normalizing, and facilitating reuse of scientific data for research and education [@webb2005phylomatic; @Stoltzfus2013].
-<!-- This has short-term advantages for research relying on this information, and long term advantages such as the advancement of science [@].-->
-We identify that any tool for efficient reuse of state-of-the-art scientific data should have a versioned, open and fully public database storing information in a computer readable format [@vos2012nexml], an automatised and programatic way of accessing the data [@Stoltzfus2013], and straightforward means of comparing and summarizing data as needed by the user.
-The DateLife project was born as a prototype tool aiming to provide these services, and was developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA [@Stoltzfus2013].
-Here we present the full implementation of DateLife, available as an R package `datelife` and as a website application with a graphical user interface at www.datelife.org/query/.
-The latest implementation of the `datelife` R package is available on CRAN (v`r packageVersion("datelife")`; @datelifeR) and features an algorithm for automatic curation and maintenance of an open database composed of chronograms pulled from OpenTree's public repository [@mctavish2015phylesystem], and methods to summarize chronograms and compare them.
-<!--, Functions to visualize and graphically compare source and summary chronograms are available at `datelifeplot` not on CRAN yet.-->
+Here we present DateLife, available as an R package `datelife` and as a website application with a graphical user interface at www.datelife.org/query/, which captures data from published chronograms, and make these data readily accessible to users.
+`datelife` uses a versioned, open and fully public database [@mctavish2015phylesystem] storing information in a computer readable format [@vos2012nexml], an automated and programatic way of accessing the data [@Stoltzfus2013] and methods to summarize chronograms and compare them.
 
-<!--The TreeBASE project started as an open and public database for state-of-the-art chronograms and other evolutionary research data in computer readable formats, and it currently supports programatic data accession [@piel2002treebase; @piel2009treebase, @treebaseAPIs], unfortunately, it is no longer under active development.
-<!--No need to mention SuperTreeBASE as it is basically the same as treebase but on a more stable server-->
-<!--The Open Tree of Life project [OpenTree, @opentreeoflife2019synth] has a system for storing chronograms and other evolutionary information in an open and public database that is programatically accessible, but it does not yet support phylogenetic queries of age data nor chronogram summaries.-->
-
-# Description of the R package
+# Description of the workflow
 
@@
 1.
-It starts with an input consisting of at least two taxon names, which can be provided in two different forms: as a comma separated character string, or as tip labels on a tree.
+It starts with an input query consisting of at least two taxon names, which can be provided in two different forms: as a comma separated character string, or as tip labels on a tree.
 If input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004].
@@
 Users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
+
+The latest implementation of the `datelife` R package is available on CRAN (v`r packageVersion("datelife")`; @datelifeR).
 
@@
 
+The DateLife project was born as a prototype tool aiming to provide these services, and was developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA [@Stoltzfus2013].
 We thank colleagues from the O'Meara Lab at the University

### 2022-03-07 manuscript/manuscript.tex
abstract and intro edits

@@
 \shorttitle{DateLife: revealing the dated Tree of Life}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 3673}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 3452}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 
-In the natural sciences, absolute time of evolutionary origin of lineages is represented graphically as a phylogeny with branch lengths proportional to geologic time, also known as a chronogram.
-Notably, building a chronogram is not an easy task.
-First, it requires obtaining and curating genetic data to construct an homology hypothesis as a genetic alignment; then, it is important to carefully choose and apply the software that will use the alignment to reconstruct an evolutionary hypothesis in the form of a phylogeny; thirdly, calibrations, i.e., independent age data points from the fossil record or other suitable geologic events have to be obtained from the literature --or by going into the field; next, calibrations have to be carefully placed on the reconstructed phylogeny, demonstrating a biological and geological understanding of their limits and applicability; finally, the appropriate software and model of evolution to estimate ellapsed time since evolutionary events on the phylogeny has to be chosen and applied.
-In sum, building chronograms require specialized biological training and a non-negligible amount of human, time and monetary resources.
+Chronograms are phylogenies with branch lengths proportional to time.
+Chronograms provide key data for the study of natural processes in many areas of biological research, such as developmental biology (cite) to conservation biology (Felsenstein, 1985; Campbell O.
+Webb, 2000), historical biogeography, a species diversification (Morlon, 2014; Posadas, Crisci, \& Katinas, 2006).
 
-Because of their importance for biological research, chronograms are constantly being built by experts in the field, and have been published in a steady increase for the last two decades, amounting to around four thousand individual chronograms published in peer-reviewed journals (Kumar, Stecher, Suleski, \& Hedges, 2017).
-Chronograms represent key knowledge for the study of natural processes in many areas of scientific research, from developmental to conservation biology (Felsenstein, 1985; Campbell O.
-Webb, 2000), from historical biogeography to species diversification (Morlon, 2014; Posadas, Crisci, \& Katinas, 2006).
-As such, the state-of-the-art scientific data on time of evolutionary origin that has already been produced and curated could be leveraged with beneficial results, and there has been an urge for promoting, normalizing, and facilitating reuse of scientific data for research and education (Stoltzfus et al., 2013; Campbell O.
-Webb \& Donoghue, 2005).
-We identify that any tool for efficient reuse of state-of-the-art scientific data should have a versioned, open and fully public database storing information in a computer readable format (Vos et al., 2012), an automatised and programatic way of accessing the data (Stoltzfus et al., 2013), and straightforward means of comparing and summarizing data as needed by the user.
-The DateLife project was born as a prototype tool aiming to provide these services, and was developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA (Stoltzfus et al., 2013).
-Here we present the full implementation of DateLife, available as an R package \texttt{datelife} and as a website application with a graphical user interface at www.datelife.org/query/.
-The latest implementation of the \texttt{datelife} R package is available on CRAN (v0.6.2; Sanchez-Reyes et al.
-(2022)) and features an algorithm for automatic curation and maintenance of an open database composed of chronograms pulled from OpenTree's public repository (McTavish et al., 2015), and methods to summarize chronograms and compare them.
+Building a chronogram is not an easy task.
+It requires obtaining and curating data to construct a phylogeny; selecting and placing appropriate calibrations on the phylogeny using independent age data points from the fossil record or other dated events, and inferring the full dated tree.
+Estimating accurate chronograms generally requires specialized biological training, taxonomic domain knowledge, and a non-negligible amount of research time, computational resources and funding.
 
-\hypertarget{description-of-the-r-package}{%
-\section{Description of the R package}\label{description-of-the-r-package}}
+Here we present DateLife, available as an R package \texttt{datelife} and as a website application with a graphical user interface at www.datelife.org/query/, which captures data from published chronograms, and make these data readily accessible to users.
+\texttt{datelife} uses a versioned, open and fully public database (McTavish et al., 2015) storing information in a computer readable format (Vos et al., 2012), an automated and programatic way of accessing the data (Stoltzfus et al., 2013) and methods to summarize chronograms and compare them.
+
+\hypertarget{description-of-the-workflow}{%
+\section{Description of the workflow}\label{description-of-the-workflow}}
 
@@
 \item
-  It starts with an input consisting of at least two taxon names, which can be provided in two different forms: as a comma separated character string, or as tip labels on a tree.
+  It starts with an input query consisting of at least two taxon names, which can be provided in two different forms: as a comma separated character string, or as tip labels on a tree.
 If input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis, Claude, \& Strimmer, 2004).
@@
 \end{enumerate}
+
+The latest implementation of the \texttt{datelife} R package is available on CRAN (v0.6.1; Sanchez-Reyes et al. (2022)).
 
@@
 
-At the time of writing of this manuscript (Mar 04, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
+At the time of writing of this manuscript (Mar 07, 2022), \texttt{datelife}'s database has 231 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
 A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time.
@@
 However source chronograms are not available in computer readable format for reuse or reanalysis.
-As this tree is part of datelife's database, the amount of lineages that can be queried using \texttt{datelife} (99474 unique terminal taxa) is substantial, yet it can be improved.
+As this tree is part of datelife's database, the amount of lineages that can be queried using \texttt{datelife} (86879 unique terminal taxa) is substantial, yet it can be improved.
 Consequently, we would like to emphasize on the importance of sharing chronogram data for the benefit of the scientific community as a whole, into repositories that require expert input and manual curation, such as OpenTree's Phylesystem (McTavish et al., 2015).
@@
 
+The DateLife project was born as a prototype tool aiming to provide these services, and was developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA (Stoltzfus et al., 2013).
 We thank colleagues from the O'Meara Lab at the University
@@
 \url{https://doi.org/10.1038/nature11631}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Kumar2017}{}}%
-Kumar, S., Stecher, G., Suleski, M., \& Hedges, S.
-B.
-(2017).
-{TimeTree: A Resource for Timelines, Timetrees, and Divergence Times}.
-\emph{Molecular Biology and Evolution}, \emph{34}(7), 1812--1819.
-\url{https://doi.org/10.1093/molbev/msx116}
 

### 2022-03-07 manuscript/manuscript.Rmd
workflow notes

@@
 These estimates also provide valuable information for for education, science communication and policy decisions.
-
- Although achieving a high-quality reconstruction of time of evolutionary origin as a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is still a difficult and time-consuming task, the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms
-
+  Although achieving a high-quality reconstruction of time of evolutionary origin as a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is still a difficult and time-consuming task, the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
   When these estimates are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on time of evolutionary origin is exposed in a programatic and reusable way.
@@
 
-  Here we present `datelife`, a service implemented as an R package and an Rshiny website application
-  available at www.datelife.org/query/, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
+  Here we present `datelife`, a service implemented as an R package and an Rshiny website application available at www.datelife.org/query/, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
 
@@
 
-  `datelife` will be useful to increase awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.
+  `datelife` will be useful to increase awereness on the existing variation in expert time of divergence data, and can foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.
 
@@
 It starts with an input query consisting of at least two taxon names, which can be provided in two different forms: as a comma separated character string, or as tip labels on a tree.
-If input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004].
+If the input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004].
 The input tree is not required to have branch lengths.
@@
 1.
-At this point, various summary data can be obtained to inform decisions for the next steps of the analysis workflow.
-Types of summary information provided are: a) all pruned source chronograms, b) age of the MRCA (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of successful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most matching input taxon names.
+At this point, summary data is collected to inform decisions for the next steps of the analysis workflow.
+Types of summary information provided are: a) all pruned source chronograms, b) age of the MRCA (most recent common ancestor) of the pruned source chronograms, ???
+is this only for ones that have all the tax?
+c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of successful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most matching input taxon names.
 1.
-To construct summary trees we use a fixed topology, either provided by the user, taken from the literature or obtained from expert phylogenetic information, such as the OpenTree synthetic tree.
+To construct summary trees we use a fixed topology, either provided by the user, taken from the literature <!--The user would have to take it from the literature, ya?--> or obtained from expert phylogenetic information, such as the OpenTree synthetic tree.
+1. ????? More meat on the methods needed: e.g. how the distances are generated and summarized, how conflict is handled...
+1.
 Then we summarize source chronograms into a single patristic distance matrix using a method chosen by the user.
-Summarizing method options available include Super Distance Matrix method [SDM, @Criscuolo2006] and any summary statistics such as median, minimum andmaximum ages.
+Summarizing method options available include Super Distance Matrix method [SDM, @Criscuolo2006] and any summary statistics such as median, minimum and maximum ages.
 Finally, `datelife` applies the summarized time distances as node calibrations to date the given topology with BLADJ [@webb2005phylomatic].
-1. If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms.
+1. If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms, ??? like how? 
 1.

### 2022-03-07 manuscript/manuscript.tex
workflow notes

@@
 \shorttitle{DateLife: revealing the dated Tree of Life}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 3452}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 3484}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 \abstract{
-Time of evolutionary origin is fundamental for research in the natural sciences, as well as for education, science communication and policy.
-Despite an increased availability of fossil and molecular data, and time-efficient analytical techniques, achieving a high-quality reconstruction of time of evolutionary origin as a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is still a difficult and time-consuming task for a majority of interested parties.
-Yet, the amount of published chronograms has increased significantly in the past two decades, and a non-negligeable proportion of these data have been steadily accumulating in public, open databases such as TreeBASE and Open Tree
-of Life, exposing a wealth of expertly-curated and peer-reviewed data on time of evolutionary origin in a programatic and reusable way, for a large quantity and diversity of organisms.
-This trend results from intensive and localized efforts for improving data sharing practices, as well as incentivizing open science in biology.
-Despite these trends, accessibility to state-of-the-art knowledge on time of evolutionary origin is still reduced.
+Date estimates for times of evolutionary diveregences are key data for research in the natural sciences.
+These estimates also provide valuable information for for education, science communication and policy decisions.
+Although achieving a high-quality reconstruction of time of evolutionary origin as a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is still a difficult and time-consuming task, the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
+When these estimates are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on time of evolutionary origin is exposed in a programatic and reusable way.
+Intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
 
-Here we present \texttt{datelife}, a service implemented as an R package and an Rshiny website application
-available at www.datelife.org/query/, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
+Here we present \texttt{datelife}, a service implemented as an R package and an Rshiny website application available at www.datelife.org/query/, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
 
@@
 We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms.
-Additionally, time data extracted from source chronograms can be usedas secondary calibration points to add branch lengths proportional to absolute time to a tree topology using alternative dating methods.
+Additionally, time data extracted from source chronograms can be used as secondary calibration points to add branch lengths proportional to absolute time to a tree topology using alternative dating methods external to datelife.
 
-Summary and newly generated trees are potentially useful to evaluate evolutionary hypothesis in different areas of research in biology.
-How well this chronograms work for this purpose still needs to be tested.
-
-\texttt{datelife} will be useful to increase awereness on the existing variation in expert time of divergence data, and might foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.
+\texttt{datelife} will be useful to increase awereness on the existing variation in expert time of divergence data, and can foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.
 }
@@
   It starts with an input query consisting of at least two taxon names, which can be provided in two different forms: as a comma separated character string, or as tip labels on a tree.
-If input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis, Claude, \& Strimmer, 2004).
+If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis, Claude, \& Strimmer, 2004).
 The input tree is not required to have branch lengths.
@@
 \item
-  At this point, various summary data can be obtained to inform decisions for the next steps of the analysis workflow.
-Types of summary information provided are: a) all pruned source chronograms, b) age of the MRCA (most recent common ancestor) of the pruned source chronograms, c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of successful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most matching input taxon names.
+  At this point, summary data is collected to inform decisions for the next steps of the analysis workflow.
+Types of summary information provided are: a) all pruned source chronograms, b) age of the MRCA (most recent common ancestor) of the pruned source chronograms, ???
+is this only for ones that have all the tax?
+c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of successful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most matching input taxon names.
 \item
   To construct summary trees we use a fixed topology, either provided by the user, taken from the literature or obtained from expert phylogenetic information, such as the OpenTree synthetic tree.
+\item
+  ????? More meat on the methods needed: e.g.~how the distances are generated and summarized, how conflict is handled\ldots{}
+\item
   Then we summarize source chronograms into a single patristic distance matrix using a method chosen by the user.
-Summarizing method options available include Super Distance Matrix method (SDM, Criscuolo, Berry, Douzery, \& Gascuel, 2006) and any summary statistics such as median, minimum andmaximum ages.
+Summarizing method options available include Super Distance Matrix method (SDM, Criscuolo, Berry, Douzery, \& Gascuel, 2006) and any summary statistics such as median, minimum and maximum ages.
 Finally, \texttt{datelife} applies the summarized time distances as node calibrations to date the given topology with BLADJ (Campbell O.
@@
 \item
-  If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms.
+  If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms, ??? like how?
 \item

### 2022-03-08 manuscript/figures.tex
adding references
figure 1 vertical mostly final
ellaborating on EJM edits

@@
 \begin{figure}[!h]
-\includegraphics{../figures/Fig1.pdf}
-\caption{Stylized DateLife workflow.
-This shows the general workflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{www.datelife.org/query/}.
-Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.}
+\includegraphics{../figures/figure1/figure1-vertical-final.pdf}
+\caption{Stylized DateLife algorithm showing the main steps that are performed with the R package \texttt{datelife}, or online at \url{www.datelife.org/query/}.
+Details on the functions involved on each workflow are available at \texttt{datelife}'s R package manual https://phylotastic.org/datelife/articles/.}
 \label{fig:workflow}
@@
 \caption{Input taxon name processing and chronogram database search computation time increases with number of input taxon names.
-We sampled N bird species names for each input size class, 100 times, and then performed a \texttt{datelife} search using the Taxon Names Resoultion Service (TNRS; dark gray), and without using TNRS (light gray).
+We sampled N bird species names for each input size class, 100 times, and then performed a \texttt{datelife} search using the Taxon Names Resoultion Service (TNRS; light gray), and without using TNRS (dark gray).
 We also performed a search using the already processed query for comparison (light blue).}

### 2022-03-08 manuscript/manuscript.Rmd
ellaborating on EJM edits

@@
 abstract: |
-  Date estimates for times of evolutionary diveregences are key data for research in the natural sciences.
+  Date estimates for times of evolutionary divergences are key data for research in the natural sciences.
 These estimates also provide valuable information for for education, science communication and policy decisions.
@@
   Intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
-
-  Here we present `datelife`, a service implemented as an R package and an Rshiny website application available at www.datelife.org/query/, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
-
-  The main workflow of `datelife` is to construct a chronogram for any given combination of taxon names, by searching a local chronogram database constructed and curated from the Open Tree of Life (OpenTree), which incorporates phylogenetic data from the TreeBASE database as well.
-  We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms.
-Additionally, time data extracted from source chronograms can be used as secondary calibration points to add branch lengths proportional to absolute time to a tree topology using alternative dating methods external to datelife.
-
-  `datelife` will be useful to increase awereness on the existing variation in expert time of divergence data, and can foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.
+  Here we present DateLife, a service implemented as an R package and an Rshiny website application available at www.datelife.org/query/, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
+  The main DateLife workflow constructs a chronogram for any given combination of taxon names, by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from TreeBASE database as well.
+  We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
+  DateLife will be useful to increase awereness on the existing variation in expert time of divergence data, and can foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.
 
@@
 # Introduction
-Chronograms are phylogenies with branch lengths proportional to time.
 <!-- If you're aiming for syst bio you don't want to start so general Readers know what chronorgrams and phylogenies are!.-->
-Chronograms provide key data for the study of natural processes in many areas of biological research, such as developmental  biology (cite) to conservation biology [@Felsenstein1985a; @Webb2000], historical biogeography, a species diversification [@posadas2006historical; @Morlon2014].
+Chronograms --phylogenies with branch lengths proportional to time, provide key data for the study of natural processes in many areas of biological research, such as developmental  biology (cite) to conservation biology [@Felsenstein1985a; @Webb2000], historical biogeography, a species diversification [@posadas2006historical; @Morlon2014].
 
@@
 
-Here we present DateLife, available as an R package `datelife` and as a website application with a graphical user interface at www.datelife.org/query/, which captures data from published chronograms, and make these data readily accessible to users.
-`datelife` uses a versioned, open and fully public database [@mctavish2015phylesystem] storing information in a computer readable format [@vos2012nexml], an automated and programatic way of accessing the data [@Stoltzfus2013] and methods to summarize chronograms and compare them.
+Here we present the DateLife software application, available as an R package and as an online Rshiny interactive website at www.datelife.org/query/, which captures data from published chronograms, and make these data readily accessible to users.
+DateLife features a versioned, open and fully public chronogram database [@mctavish2015phylesystem] storing age information in a computer readable format [@vos2012nexml], an automated and programmatic way of accessing the data [@Stoltzfus2013] and methods to summarize and compare age data.
 
-# Description of the workflow
+# The DateLife algorithm
 
-The general `datelife` workflow is shown in figure \ref{fig:workflow}, briefly:
+_**Providing an input.--**_ DateLife starts with an input query consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree.
+If the input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004].
+The input tree is not required to have branch lengths, and its topology is used on the summary steps described below.
 
-1.
-It starts with an input query consisting of at least two taxon names, which can be provided in two different forms: as a comma separated character string, or as tip labels on a tree.
-If the input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004].
-The input tree is not required to have branch lengths.
-1.
-Input taxon names are processed to detect, correct and standardize name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard.
-To do this, `datelife` implements the Taxonomic Name Resolution Service (TNRS; @Boyle2013) provided by OpenTree [@opentreeAPIs].
-TNRS increases the probability of correctly finding the input taxon names in the chronogram database.
-1.
-The current version of `datelife` only accepts scientific taxonomic names as input.
-Names can belong to any taxonomic group or binomial specific.
-If an input taxon name belongs to an "inclusive" taxonomic group, i.e., a taxon above the species level, such as genus, family, etc.), `datelife` has two alternative behaviors defined by the "get species from taxon" flag.
-If the flag is active, `datelife` retrieves all species names within the "inclusive" taxonomic group and adds them to the input.
-If the flag is inactive, `datelife` will drop the "inclusive" taxon names from input.
-1.
-The cleaned input taxon names are saved as a special R object (of a newly defined class \texttt{datelifeQuery}) that contains the processed names, the corresponding taxonomic id numbers, and the topology of theinput tree if any was provided.
+Currently, DateLife only accepts scientific names as input, and they can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or binomial specific.
+Subspecies and variants are ignored.
+If an input taxon name belongs to an "inclusive" taxonomic group the algorithm has two alternative behaviors defined by the "get species from taxon" flag.
+If the flag is active, the DateLife algorithm retrieves all species names within the "inclusive" taxonomic group and adds them to the input.
+If the flag is inactive, DateLife ignores the "inclusive" taxon names from the input.
+
+Input scientific names are processed using the Taxonomic Name Resolution Service (TNRS; @Boyle2013) provided by OpenTree [@opentreeAPIs].
+TNRS increases the probability of correctly finding the query organisms in the chronogram database, as it detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard.
+
+The processed input taxon names are saved as a special R object (of a newly defined class \texttt{datelifeQuery}) that contains the processed names, the corresponding taxonomic id numbers, and the topology of the input tree if any was provided.
 The \texttt{datelifeQuery} object is used next to search the chronogram database.
-1.
-Chronograms with at least two matching input taxon names are identified and pruned down to preserve only input taxon names as tips.
+
+_**Searching the database.--**_  Chronograms in the DateLife database with at least two matching processed taxon names are identified and pruned down to preserve only the matched names as tips.
 Then, each pruned chronogram is transformed to a patristic distance matrix.
@@
 The matrices are associated to the citation of the original study and stored as an R object of class \texttt{datelifeResult}.
-1.
-At this point, summary data is collected to inform decisions for the next steps of the analysis workflow.
-Types of summary information provided are: a) all pruned source chronograms, b) age of the MRCA (most recent common ancestor) of the pruned source chronograms, ???
-is this only for ones that have all the tax?
-c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of successful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most matching input taxon names.
-1.
-To construct summary trees we use a fixed topology, either provided by the user, taken from the literature <!--The user would have to take it from the literature, ya?--> or obtained from expert phylogenetic information, such as the OpenTree synthetic tree.
-1. ????? More meat on the methods needed: e.g. how the distances are generated and summarized, how conflict is handled...
-1.
+
+_**Summarizing search results.--**_ At this point, summary data is collected to inform decisions for ulterior steps in the user workflow. Types of summary information provided are:
+
+1. The pruned source chronograms are used to get pairwise node ages of your query taxa (Figure \ref{fig:workflow} subfigure X).
+2.
+Age of the root of pruned source chronograms.
+This can correspond to the mrca age of your group if the pruned source chronogram has all your query taxa.
+If not, the root correponds to the mrca of  subgroup withing your query taxa.
+3. Citations of studies where pruned source chronograms were originally published.
+4. A summary table with all of the above.
+5. A single chronogram summarizing all pruned source chronograms, or a subset defined by the user.
+To construct summary trees, `datelife` requires a fixed topology.
+Users can provide one from the literature, or one of their own making.
+If no topology is provided, `datelife` chooses one automatically retrieved from an expert phylogenetic information database such as the OpenTree synthetic tree.
+
+????? More meat on the methods needed: e.g. how the distances are generated and summarized, how conflict is handled...
+
 Then we summarize source chronograms into a single patristic distance matrix using a method chosen by the user.
@@
 Finally, `datelife` applies the summarized time distances as node calibrations to date the given topology with BLADJ [@webb2005phylomatic].
-1. If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms, ??? like how? 
-1.
-Users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
 
-The latest implementation of the `datelife` R package is available on CRAN (v`r packageVersion("datelife")`; @datelifeR).
+6. A report of successful matches of input taxon names across pruned source chronograms.
+7. A single pruned source chronogram with the most matching input taxon names.
 
-The `datelife` workflow relies on functions from several R packages that we acknowledge: rotl [@Michonneau2016], ape [@paradis2004],
-geiger [@Harmon2008], paleotree [@Bapst2012a], bold [@Chamberlain2018], phytools [@Revell2012], taxize [@Chamberlain2013; @Chamberlain2018], phyloch [@Heibl2008], and phylocomr [@Ooms2018].
+If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms, ??? like how?
+
+_**Visualizing summaries.--**_ Users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
+
+The DateLife algorithm illustrated in figure \ref{fig:workflow} is fully implemented using the R language.
+The latest stable version of the R package `datelife` is available from the CRAN repository (v`r packageVersion("datelife")`; @datelifeR), and relies on functionalities from various biological R packages:
+ape [@paradis2004],
+bold [@Chamberlain2018],
+geiger [@Harmon2008],
+paleotree [@Bapst2012a],
+phyloch [@Heibl2008],
+phylocomr [@Ooms2018],
+phytools [@Revell2012],
+rotl [@Michonneau2016], and
+taxize [@Chamberlain2013; @Chamberlain2018].
+
 <!-- TODO ADD FIGURE: mock example explaining difference between using summary ages and raw ages to calibrate a topology.
@@
 
-We illustrate the `datelife` workflow using the family of true finches, Fringillidae as an example.
+We illustrate the `datelife` algorithm using the family of true finches, Fringillidae as an example.
 <!--We used `r cite_r("r-references.bib")` for all our analyses.-->

### 2022-03-08 manuscript/manuscript.tex
ellaborating on EJM edits

@@
 \shorttitle{DateLife: revealing the dated Tree of Life}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 3484}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 3542}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 \abstract{
-Date estimates for times of evolutionary diveregences are key data for research in the natural sciences.
+Date estimates for times of evolutionary divergences are key data for research in the natural sciences.
 These estimates also provide valuable information for for education, science communication and policy decisions.
@@
 Intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
-
-Here we present \texttt{datelife}, a service implemented as an R package and an Rshiny website application available at www.datelife.org/query/, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
-
-The main workflow of \texttt{datelife} is to construct a chronogram for any given combination of taxon names, by searching a local chronogram database constructed and curated from the Open Tree of Life (OpenTree), which incorporates phylogenetic data from the TreeBASE database as well.
-We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms.
-Additionally, time data extracted from source chronograms can be used as secondary calibration points to add branch lengths proportional to absolute time to a tree topology using alternative dating methods external to datelife.
-
-\texttt{datelife} will be useful to increase awereness on the existing variation in expert time of divergence data, and can foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.
+Here we present DateLife, a service implemented as an R package and an Rshiny website application available at www.datelife.org/query/, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
+The main DateLife workflow constructs a chronogram for any given combination of taxon names, by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from TreeBASE database as well.
+We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
+DateLife will be useful to increase awereness on the existing variation in expert time of divergence data, and can foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.
 }
@@
 
-Chronograms are phylogenies with branch lengths proportional to time.
-Chronograms provide key data for the study of natural processes in many areas of biological research, such as developmental biology (cite) to conservation biology (Felsenstein, 1985; Campbell O.
+Chronograms --phylogenies with branch lengths proportional to time, provide key data for the study of natural processes in many areas of biological research, such as developmental biology (cite) to conservation biology (Felsenstein, 1985; Campbell O.
 Webb, 2000), historical biogeography, a species diversification (Morlon, 2014; Posadas, Crisci, \& Katinas, 2006).
@@
 
-Here we present DateLife, available as an R package \texttt{datelife} and as a website application with a graphical user interface at www.datelife.org/query/, which captures data from published chronograms, and make these data readily accessible to users.
-\texttt{datelife} uses a versioned, open and fully public database (McTavish et al., 2015) storing information in a computer readable format (Vos et al., 2012), an automated and programatic way of accessing the data (Stoltzfus et al., 2013) and methods to summarize chronograms and compare them.
+Here we present the DateLife software application, available as an R package and as an online Rshiny interactive website at www.datelife.org/query/, which captures data from published chronograms, and make these data readily accessible to users.
+DateLife features a versioned, open and fully public chronogram database (McTavish et al., 2015) storing age information in a computer readable format (Vos et al., 2012), an automated and programmatic way of accessing the data (Stoltzfus et al., 2013) and methods to summarize and compare age data.
 
-\hypertarget{description-of-the-workflow}{%
-\section{Description of the workflow}\label{description-of-the-workflow}}
+\hypertarget{the-datelife-algorithm}{%
+\section{The DateLife algorithm}\label{the-datelife-algorithm}}
 
-The general \texttt{datelife} workflow is shown in figure \ref{fig:workflow}, briefly:
+\emph{\textbf{Providing an input.--}} DateLife starts with an input query consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree.
+If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis, Claude, \& Strimmer, 2004).
+The input tree is not required to have branch lengths, and its topology is used on the summary steps described below.
+
+Currently, DateLife only accepts scientific names as input, and they can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or binomial specific.
+Subspecies and variants are ignored.
+If an input taxon name belongs to an ``inclusive'' taxonomic group the algorithm has two alternative behaviors defined by the ``get species from taxon'' flag.
+If the flag is active, the DateLife algorithm retrieves all species names within the ``inclusive'' taxonomic group and adds them to the input.
+If the flag is inactive, DateLife ignores the ``inclusive'' taxon names from the input.
+
+Input scientific names are processed using the Taxonomic Name Resolution Service (TNRS; Boyle et al.
+(2013)) provided by OpenTree (OpenTreeOfLife et al., n.d.).
+TNRS increases the probability of correctly finding the query organisms in the chronogram database, as it detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard.
+
+The processed input taxon names are saved as a special R object (of a newly defined class \texttt{datelifeQuery}) that contains the processed names, the corresponding taxonomic id numbers, and the topology of the input tree if any was provided.
+The \texttt{datelifeQuery} object is used next to search the chronogram database.
+
+\emph{\textbf{Searching the database.--}} Chronograms in the DateLife database with at least two matching processed taxon names are identified and pruned down to preserve only the matched names as tips.
+Then, each pruned chronogram is transformed to a patristic distance matrix.
+This format facilitates and greatly speeds up all downstream analyses and summaries.
+The matrices are associated to the citation of the original study and stored as an R object of class \texttt{datelifeResult}.
+
+\emph{\textbf{Summarizing search results.--}} At this point, summary data is collected to inform decisions for ulterior steps in the user workflow. Types of summary information provided are:
 
@@
 \item
-  It starts with an input query consisting of at least two taxon names, which can be provided in two different forms: as a comma separated character string, or as tip labels on a tree.
-If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis, Claude, \& Strimmer, 2004).
-The input tree is not required to have branch lengths.
+  The pruned source chronograms are used to get pairwise node ages of your query taxa (Figure \ref{fig:workflow} subfigure X).
 \item
-  Input taxon names are processed to detect, correct and standardize name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard.
-To do this, \texttt{datelife} implements the Taxonomic Name Resolution Service (TNRS; Boyle et al.
-(2013)) provided by OpenTree (OpenTreeOfLife et al., n.d.).
-TNRS increases the probability of correctly finding the input taxon names in the chronogram database.
+  Age of the root of pruned source chronograms.
+This can correspond to the mrca age of your group if the pruned source chronogram has all your query taxa.
+If not, the root correponds to the mrca of subgroup withing your query taxa.
 \item
-  The current version of \texttt{datelife} only accepts scientific taxonomic names as input.
-Names can belong to any taxonomic group or binomial specific.
-If an input taxon name belongs to an ``inclusive'' taxonomic group, i.e., a taxon above the species level, such as genus, family, etc.), \texttt{datelife} has two alternative behaviors defined by the ``get species from taxon'' flag.
-If the flag is active, \texttt{datelife} retrieves all species names within the ``inclusive'' taxonomic group and adds them to the input.
-If the flag is inactive, \texttt{datelife} will drop the ``inclusive'' taxon names from input.
+  Citations of studies where pruned source chronograms were originally published.
 \item
-  The cleaned input taxon names are saved as a special R object (of a newly defined class \texttt{datelifeQuery}) that contains the processed names, the corresponding taxonomic id numbers, and the topology of theinput tree if any was provided.
-The \texttt{datelifeQuery} object is used next to search the chronogram database.
+  A summary table with all of the above.
 \item
-  Chronograms with at least two matching input taxon names are identified and pruned down to preserve only input taxon names as tips.
-Then, each pruned chronogram is transformed to a patristic distance matrix.
-This format facilitates and greatly speeds up all downstream analyses and summaries.
-The matrices are associated to the citation of the original study and stored as an R object of class \texttt{datelifeResult}.
-\item
-  At this point, summary data is collected to inform decisions for the next steps of the analysis workflow.
-Types of summary information provided are: a) all pruned source chronograms, b) age of the MRCA (most recent common ancestor) of the pruned source chronograms, ???
-is this only for ones that have all the tax?
-c) citations of studies where pruned source chronograms were originally published, d) a summary table with all of the above, e) a single summary chronogram of all or a subset of pruned source chronograms, f) a report of successful matches of input taxon names across pruned source chronograms, and g) the single pruned source chronogram with the most matching input taxon names.
-\item
-  To construct summary trees we use a fixed topology, either provided by the user, taken from the literature or obtained from expert phylogenetic information, such as the OpenTree synthetic tree.
-\item
+  A single chronogram summarizing all pruned source chronograms, or a subset defined by the user.
+  To construct summary trees, \texttt{datelife} requires a fixed topology.
+Users can provide one from the literature, or one of their own making.
+If no topology is provided, \texttt{datelife} chooses one automatically retrieved from an expert phylogenetic information database such as the OpenTree synthetic tree.
+\end{enumerate}
+
 ????? More meat on the methods needed: e.g.~how the distances are generated and summarized, how conflict is handled\ldots{}
-\item
+
 Then we summarize source chronograms into a single patristic distance matrix using a method chosen by the user.
@@
 Webb \& Donoghue, 2005).
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\setcounter{enumi}{5}
+\tightlist
 \item
-  If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms, ??? like how?
+  A report of successful matches of input taxon names across pruned source chronograms.
 \item
-  Users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R ``phylo'' format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
+  A single pruned source chronogram with the most matching input taxon names.
 \end{enumerate}
 
-The latest implementation of the \texttt{datelife} R package is available on CRAN (v0.6.1; Sanchez-Reyes et al. (2022)).
+If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms, ??? like how?
 
-The \texttt{datelife} workflow relies on functions from several R packages that we acknowledge: rotl (Michonneau, Brown, \& Winter, 2016), ape (Paradis et al., 2004),
-geiger (Harmon, Weir, Brock, Glor, \& Challenger, 2008), paleotree (Bapst, 2012), bold (Chamberlain et al., 2019), phytools (Revell, 2012), taxize (Chamberlain \& Szöcs, 2013; Chamberlain et al., 2019), phyloch (Heibl, 2008), and phylocomr (Ooms \& Chamberlain, 2018).
+\emph{\textbf{Visualizing summaries.--}} Users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R ``phylo'' format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
+
+The DateLife algorithm illustrated in figure \ref{fig:workflow} is fully implemented using the R language.
+The latest stable version of the R package \texttt{datelife} is available from the CRAN repository (v0.6.2; Sanchez-Reyes et al.
+(2022)), and relies on functionalities from various biological R packages:
+ape (Paradis et al., 2004),
+bold (Chamberlain et al., 2019),
+geiger (Harmon, Weir, Brock, Glor, \& Challenger, 2008),
+paleotree (Bapst, 2012),
+phyloch (Heibl, 2008),
+phylocomr (Ooms \& Chamberlain, 2018),
+phytools (Revell, 2012),
+rotl (Michonneau, Brown, \& Winter, 2016), and
+taxize (Chamberlain \& Szöcs, 2013; Chamberlain et al., 2019).
 
@@
 
-We illustrate the \texttt{datelife} workflow using the family of true finches, Fringillidae as an example.
+We illustrate the \texttt{datelife} algorithm using the family of true finches, Fringillidae as an example.
 A college educator wishes to obtain state-of-the-art data on time of evolutionary origin of species belonging to the true finches for their class.
@@
 
-At the time of writing of this manuscript (Mar 07, 2022), \texttt{datelife}'s database has 231 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
+At the time of writing of this manuscript (Mar 07, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
 A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time.
@@
 However source chronograms are not available in computer readable format for reuse or reanalysis.
-As this tree is part of datelife's database, the amount of lineages that can be queried using \texttt{datelife} (86879 unique terminal taxa) is substantial, yet it can be improved.
+As this tree is part of datelife's database, the amount of lineages that can be queried using \texttt{datelife} (99474 unique terminal taxa) is substantial, yet it can be improved.
 Consequently, we would like to emphasize on the importance of sharing chronogram data for the benefit of the scientific community as a whole, into repositories that require expert input and manual curation, such as OpenTree's Phylesystem (McTavish et al., 2015).

### 2022-03-08 manuscript/manuscript.Rmd
beefed methods

@@
 <!-- If you're aiming for syst bio you don't want to start so general Readers know what chronorgrams and phylogenies are!.-->
-Chronograms --phylogenies with branch lengths proportional to time, provide key data for the study of natural processes in many areas of biological research, such as developmental  biology (cite) to conservation biology [@Felsenstein1985a; @Webb2000], historical biogeography, a species diversification [@posadas2006historical; @Morlon2014].
+Chronograms --phylogenies with branch lengths proportional to time-- provide key data for the study of natural processes in many areas of biological research, such as developmental  biology [@laubichler2009form; @delsuc2018phylogenomic], conservation biology [@Felsenstein1985a; @Webb2000], historical biogeography [@posadas2006historical], and species diversification [@magallon2001absolute; @Morlon2014].
 
@@
 
-# The DateLife algorithm
+# Description
 
-_**Providing an input.--**_ DateLife starts with an input query consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree.
-If the input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004].
-The input tree is not required to have branch lengths, and its topology is used on the summary steps described below.
-
-Currently, DateLife only accepts scientific names as input, and they can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or binomial specific.
-Subspecies and variants are ignored.
-If an input taxon name belongs to an "inclusive" taxonomic group the algorithm has two alternative behaviors defined by the "get species from taxon" flag.
-If the flag is active, the DateLife algorithm retrieves all species names within the "inclusive" taxonomic group and adds them to the input.
-If the flag is inactive, DateLife ignores the "inclusive" taxon names from the input.
-
-Input scientific names are processed using the Taxonomic Name Resolution Service (TNRS; @Boyle2013) provided by OpenTree [@opentreeAPIs].
-TNRS increases the probability of correctly finding the query organisms in the chronogram database, as it detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard.
-
-The processed input taxon names are saved as a special R object (of a newly defined class \texttt{datelifeQuery}) that contains the processed names, the corresponding taxonomic id numbers, and the topology of the input tree if any was provided.
-The \texttt{datelifeQuery} object is used next to search the chronogram database.
-
-_**Searching the database.--**_  Chronograms in the DateLife database with at least two matching processed taxon names are identified and pruned down to preserve only the matched names as tips.
-Then, each pruned chronogram is transformed to a patristic distance matrix.
-This format facilitates and greatly speeds up all downstream analyses and summaries.
-The matrices are associated to the citation of the original study and stored as an R object of class \texttt{datelifeResult}.
-
-_**Summarizing search results.--**_ At this point, summary data is collected to inform decisions for ulterior steps in the user workflow. Types of summary information provided are:
-
-1. The pruned source chronograms are used to get pairwise node ages of your query taxa (Figure \ref{fig:workflow} subfigure X).
-2.
-Age of the root of pruned source chronograms.
-This can correspond to the mrca age of your group if the pruned source chronogram has all your query taxa.
-If not, the root correponds to the mrca of  subgroup withing your query taxa.
-3. Citations of studies where pruned source chronograms were originally published.
-4. A summary table with all of the above.
-5. A single chronogram summarizing all pruned source chronograms, or a subset defined by the user.
-To construct summary trees, `datelife` requires a fixed topology.
-Users can provide one from the literature, or one of their own making.
-If no topology is provided, `datelife` chooses one automatically retrieved from an expert phylogenetic information database such as the OpenTree synthetic tree.
-
-????? More meat on the methods needed: e.g. how the distances are generated and summarized, how conflict is handled...
-
-Then we summarize source chronograms into a single patristic distance matrix using a method chosen by the user.
-Summarizing method options available include Super Distance Matrix method [SDM, @Criscuolo2006] and any summary statistics such as median, minimum and maximum ages.
-Finally, `datelife` applies the summarized time distances as node calibrations to date the given topology with BLADJ [@webb2005phylomatic].
-
-6. A report of successful matches of input taxon names across pruned source chronograms.
-7. A single pruned source chronogram with the most matching input taxon names.
-
-If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms, ??? like how?
-
-_**Visualizing summaries.--**_ Users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
-
-The DateLife algorithm illustrated in figure \ref{fig:workflow} is fully implemented using the R language.
+The DateLife algorithm is fully implemented using the R language.
 The latest stable version of the R package `datelife` is available from the CRAN repository (v`r packageVersion("datelife")`; @datelifeR), and relies on functionalities from various biological R packages:
@@
 taxize [@Chamberlain2013; @Chamberlain2018].
+Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife algorithm: providing an input, searching a chronogram database, and summarizing results from the search.
+
+<!--_**Providing an input.--**_ -->
+
+## Providing an input
+
+DateLife starts with an input query consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree.
+If the input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004].
+The input tree is not required to have branch lengths, and its topology is used on the summary steps described below.
+
+Currently, DateLife only accepts scientific names as input, and they can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or binomial specific.
+Subspecies and variants are ignored.
+If an input taxon name belongs to an inclusive taxonomic group the algorithm has two alternative behaviors defined by the "get species from taxon" flag.
+If the flag is active, the DateLife algorithm retrieves all species names within the inclusive taxonomic group and adds them to the input.
+If the flag is inactive, DateLife ignores the inclusive taxon names from the input.
+
+Input scientific names are processed using the Taxonomic Name Resolution Service (TNRS; @Boyle2013), which increases the probability of correctly finding the queried taxon names in the chronogram database.
+TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard.
+DateLife implements TNRS using OpenTree's taxonomic as standard [@opentreeAPIs; @rees2017automated].
+
+The processed input taxon names are saved as an R object of a newly defined class \texttt{datelifeQuery} that is used in the following steps.
+It contains the processed names, the corresponding OpenTree taxonomic id numbers, and the topology of the input tree if any was provided.
+
+<!--_**Searching the database.--**_  -->
+
+## Searching the database
+
+A DateLife search consists of matching processed taxon names to tip labels in a chronogram database.
+Chronograms with at least two matching tip labels are identified and pruned down to preserve only the matched tips.
+
+Matching pruned chronograms are stored as individual patristic distance matrices (Figure \ref{fig:workflow} subfigure X).
+This format speeds up extraction of pairwise taxon ages of the queried taxa, as opposed to searching the ancestor node of a pair of taxa in a "phylo" object or newick string.
+The patristic matrices are also associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.
+
+DateLife's chronogram database latest version consist of `r length(opentree_chronograms$trees)` chronograms published in `r length(unique(opentree_chronograms$studies))` different studies.
+It is constructed from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert phylogenetic knowledge with rich metadata [@mctavish2015phylesystem] that allows automatic and reproducible construction of a chronogram database.
+
+<!-- _**Summarizing search results.--**_ -->
+
+## Summarizing search results
+
+At this point, summary information is extracted from the \texttt{datelifeResult} object to inform decisions for ulterior steps in the user workflow.
+Age data from matching pruned chronograms is summarized and used to generate a single summary chronogram.
+Other basic summary information available is:
+
+1. The matching pruned chronograms as newick strings or "phylo" objects.
+2.
+The ages of the root of all matching pruned chronograms.
+This can correspond to the age of the most recent common ancestor (mrca) of your group of interest if the pruned chronograms have all taxa belonging to the group.
+If not, the root corresponds to the mrca ofa subgroup withing your group of interest.
+3. Study citations where original chronograms were published.
+4. A report of input taxon names matches across pruned chronograms.
+5. The single matching pruned chronogram with the most input taxon names.
+
+_**Identifying groves.--**_ To generate a single summary chronogram, the DateLife algorithm starts by identifying the matching pruned chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
+[-@ane2009groves].
+In rare cases, a group of trees can have multiple groves.
+By default, DateLife chooses the grove with the most taxa, however, the "criterion = trees" flag allows the user to choose the grove with the most trees instead.
+
+_**Choosing a topology.--**_ Due to sampling, DateLife requires a tree topology to summarize age data upon it.
+Users can provide one as input from the literature, or one of their own making.
+If no topology is provided, DateLife automatically subsets one from the OpenTree synthetic tree [@opentreeoflife2019synth].
+
+DateLife can also reconstruct branch lengths proportional to substitution rates on the fixed tree topology from available genetic data from BOLD.
+
+_**Congruifying nodes.--**_ DateLife then implements the congruification method [@Eastman2013] to find nodes belonging to the same clade across matching pruned chronograms.
+Congruified node ages stored as a `congruifiedCalibrations` object are then matched to nodes in the chosen tree topology and stored as a `matchedCalibrations` object.
+
+_**Summarizing node ages.--**_ DateLife summarizes matched calibrations into a single patristic distance matrix using different methods.
+Summarizing options implemented include Super Distance Matrix method [SDM, @Criscuolo2006] and summary statistics such as median, minimum and maximum ages.
+
+_**Dating the tree topology.--**_ Summarized calibrations can be applied as secondary calibrations with different dating methods currently supported within DateLife: MrBayes [@Huelsenbeck2001; @Ronquist2003], PATHd8 [@Britton2007], BLADJ [@Webb2008; @webb2005phylomatic], and treePL [@Smith2012].
+By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm that assigns ages to nodes with no data evenly between nodes with age data, which minimizes age variance in the resulting chronogram [@Webb2008].
+When there is conflict in ages across node with age data, the algorithm ignores ages that are older than parent nodes and/or younger than descendant nodes.
+
+If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature.
+If none is provided, DateLife assigns an arbitrary age to the root as 10% older than the oldest age available within the group.
+<!--PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees.
+treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method
+[@Smith2012].
+MrBayes program [@Huelsenbeck2001; @Ronquist2003] can be used when adding taxa at
+random, following a reference taxonomy or a topological constraint. It draws ages
+from a pure birth model, as implemented by Jetz and collaborators [-@Jetz2012].
+
+If there is no information available for some input taxon name, users can also create age and phylogenetic data for the missing branches with MrBayes.
+-->
+
+_**Visualizing results.--**_
+Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions.
 
@@
 This will show what happens if a tree (phylogenetic conflict) or node (age) do not agree.-->
-
-<!-- \item  Alternatively, time of lineage divergence obtained from the pruned chronograms can be used directly as secondary calibration points to date a tree with or without branch lengths containing some or all input taxon names.
-%a taxonomic tree-->
 
@@
 
-# Results
-
-## Case study
+# Case study
 
@@
 
-## Cross-validation test
+# Cross-validation test
+
 Data from source chronograms can be also used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates (Figs.

### 2022-03-08 manuscript/manuscript.tex
beefed methods

@@
 \shorttitle{DateLife: revealing the dated Tree of Life}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 3542}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 3863}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 
-Chronograms --phylogenies with branch lengths proportional to time, provide key data for the study of natural processes in many areas of biological research, such as developmental biology (cite) to conservation biology (Felsenstein, 1985; Campbell O.
-Webb, 2000), historical biogeography, a species diversification (Morlon, 2014; Posadas, Crisci, \& Katinas, 2006).
+Chronograms --phylogenies with branch lengths proportional to time-- provide key data for the study of natural processes in many areas of biological research, such as developmental biology (Delsuc et al., 2018; Laubichler \& Maienschein, 2009), conservation biology (Felsenstein, 1985; Campbell O.
+Webb, 2000), historical biogeography (Posadas, Crisci, \& Katinas, 2006), and species diversification (Magallon \& Sanderson, 2001; Morlon, 2014).
 
@@
 
-\hypertarget{the-datelife-algorithm}{%
-\section{The DateLife algorithm}\label{the-datelife-algorithm}}
+\hypertarget{description}{%
+\section{Description}\label{description}}
 
-\emph{\textbf{Providing an input.--}} DateLife starts with an input query consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree.
-If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis, Claude, \& Strimmer, 2004).
-The input tree is not required to have branch lengths, and its topology is used on the summary steps described below.
-
-Currently, DateLife only accepts scientific names as input, and they can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or binomial specific.
-Subspecies and variants are ignored.
-If an input taxon name belongs to an ``inclusive'' taxonomic group the algorithm has two alternative behaviors defined by the ``get species from taxon'' flag.
-If the flag is active, the DateLife algorithm retrieves all species names within the ``inclusive'' taxonomic group and adds them to the input.
-If the flag is inactive, DateLife ignores the ``inclusive'' taxon names from the input.
-
-Input scientific names are processed using the Taxonomic Name Resolution Service (TNRS; Boyle et al.
-(2013)) provided by OpenTree (OpenTreeOfLife et al., n.d.).
-TNRS increases the probability of correctly finding the query organisms in the chronogram database, as it detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard.
-
-The processed input taxon names are saved as a special R object (of a newly defined class \texttt{datelifeQuery}) that contains the processed names, the corresponding taxonomic id numbers, and the topology of the input tree if any was provided.
-The \texttt{datelifeQuery} object is used next to search the chronogram database.
-
-\emph{\textbf{Searching the database.--}} Chronograms in the DateLife database with at least two matching processed taxon names are identified and pruned down to preserve only the matched names as tips.
-Then, each pruned chronogram is transformed to a patristic distance matrix.
-This format facilitates and greatly speeds up all downstream analyses and summaries.
-The matrices are associated to the citation of the original study and stored as an R object of class \texttt{datelifeResult}.
-
-\emph{\textbf{Summarizing search results.--}} At this point, summary data is collected to inform decisions for ulterior steps in the user workflow. Types of summary information provided are:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  The pruned source chronograms are used to get pairwise node ages of your query taxa (Figure \ref{fig:workflow} subfigure X).
-\item
-  Age of the root of pruned source chronograms.
-This can correspond to the mrca age of your group if the pruned source chronogram has all your query taxa.
-If not, the root correponds to the mrca of subgroup withing your query taxa.
-\item
-  Citations of studies where pruned source chronograms were originally published.
-\item
-  A summary table with all of the above.
-\item
-  A single chronogram summarizing all pruned source chronograms, or a subset defined by the user.
-  To construct summary trees, \texttt{datelife} requires a fixed topology.
-Users can provide one from the literature, or one of their own making.
-If no topology is provided, \texttt{datelife} chooses one automatically retrieved from an expert phylogenetic information database such as the OpenTree synthetic tree.
-\end{enumerate}
-
-????? More meat on the methods needed: e.g.~how the distances are generated and summarized, how conflict is handled\ldots{}
-
-Then we summarize source chronograms into a single patristic distance matrix using a method chosen by the user.
-Summarizing method options available include Super Distance Matrix method (SDM, Criscuolo, Berry, Douzery, \& Gascuel, 2006) and any summary statistics such as median, minimum and maximum ages.
-Finally, \texttt{datelife} applies the summarized time distances as node calibrations to date the given topology with BLADJ (Campbell O.
-Webb \& Donoghue, 2005).
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\setcounter{enumi}{5}
-\tightlist
-\item
-  A report of successful matches of input taxon names across pruned source chronograms.
-\item
-  A single pruned source chronogram with the most matching input taxon names.
-\end{enumerate}
-
-If there is no information available for any input taxon name, users can also create both age and phylogenetic data for the missing branches with a variety of algorithms, ??? like how?
-
-\emph{\textbf{Visualizing summaries.--}} Users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R ``phylo'' format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s graphic generation functions.
-
-The DateLife algorithm illustrated in figure \ref{fig:workflow} is fully implemented using the R language.
+The DateLife algorithm is fully implemented using the R language.
 The latest stable version of the R package \texttt{datelife} is available from the CRAN repository (v0.6.2; Sanchez-Reyes et al.
 (2022)), and relies on functionalities from various biological R packages:
-ape (Paradis et al., 2004),
+ape (Paradis, Claude, \& Strimmer, 2004),
 bold (Chamberlain et al., 2019),
@@
 taxize (Chamberlain \& Szöcs, 2013; Chamberlain et al., 2019).
+Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife algorithm: providing an input, searching a chronogram database, and summarizing results from the search.
+
+\hypertarget{providing-an-input}{%
+\subsection{Providing an input}\label{providing-an-input}}
+
+DateLife starts with an input query consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree.
+If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis et al., 2004).
+The input tree is not required to have branch lengths, and its topology is used on the summary steps described below.
+
+Currently, DateLife only accepts scientific names as input, and they can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or binomial specific.
+Subspecies and variants are ignored.
+If an input taxon name belongs to an inclusive taxonomic group the algorithm has two alternative behaviors defined by the ``get species from taxon'' flag.
+If the flag is active, the DateLife algorithm retrieves all species names within the inclusive taxonomic group and adds them to the input.
+If the flag is inactive, DateLife ignores the inclusive taxon names from the input.
+
+Input scientific names are processed using the Taxonomic Name Resolution Service (TNRS; Boyle et al.
+(2013)), which increases the probability of correctly finding the queried taxon names in the chronogram database.
+TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard.
+DateLife implements TNRS using OpenTree's taxonomic as standard (Open Tree Of Life et al., 2016; Rees \& Cranston, 2017).
+
+The processed input taxon names are saved as an R object of a newly defined class \texttt{datelifeQuery} that is used in the following steps.
+It contains the processed names, the corresponding OpenTree taxonomic id numbers, and the topology of the input tree if any was provided.
+
+\hypertarget{searching-the-database}{%
+\subsection{Searching the database}\label{searching-the-database}}
+
+A DateLife search consists of matching processed taxon names to tip labels in a chronogram database.
+Chronograms with at least two matching tip labels are identified and pruned down to preserve only the matched tips.
+
+Matching pruned chronograms are stored as individual patristic distance matrices (Figure \ref{fig:workflow} subfigure X).
+This format speeds up extraction of pairwise taxon ages of the queried taxa, as opposed to searching the ancestor node of a pair of taxa in a ``phylo'' object or newick string.
+The patristic matrices are also associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.
+
+DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies.
+It is constructed from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert phylogenetic knowledge with rich metadata (McTavish et al., 2015) that allows automatic and reproducible construction of a chronogram database.
+
+\hypertarget{summarizing-search-results}{%
+\subsection{Summarizing search results}\label{summarizing-search-results}}
+
+At this point, summary information is extracted from the \texttt{datelifeResult} object to inform decisions for ulterior steps in the user workflow.
+Age data from matching pruned chronograms is summarized and used to generate a single summary chronogram.
+Other basic summary information available is:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  The matching pruned chronograms as newick strings or ``phylo'' objects.
+\item
+  The ages of the root of all matching pruned chronograms.
+This can correspond to the age of the most recent common ancestor (mrca) of your group of interest if the pruned chronograms have all taxa belonging to the group.
+If not, the root corresponds to the mrca ofa subgroup withing your group of interest.
+\item
+  Study citations where original chronograms were published.
+\item
+  A report of input taxon names matches across pruned chronograms.
+\item
+  The single matching pruned chronogram with the most input taxon names.
+\end{enumerate}
+
+\emph{\textbf{Identifying groves.--}} To generate a single summary chronogram, the DateLife algorithm starts by identifying the matching pruned chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
+(2009).
+In rare cases, a group of trees can have multiple groves.
+By default, DateLife chooses the grove with the most taxa, however, the ``criterion = trees'' flag allows the user to choose the grove with the most trees instead.
+
+\emph{\textbf{Choosing a topology.--}} Due to sampling, DateLife requires a tree topology to summarize age data upon it.
+Users can provide one as input from the literature, or one of their own making.
+If no topology is provided, DateLife automatically subsets one from the OpenTree synthetic tree (Open Tree Of Life et al., 2019).
+
+DateLife can also reconstruct branch lengths proportional to substitution rates on the fixed tree topology from available genetic data from BOLD.
+
+\emph{\textbf{Congruifying nodes.--}} DateLife then implements the congruification method (Eastman, Harmon, \& Tank, 2013) to find nodes belonging to the same clade across matching pruned chronograms.
+Congruified node ages stored as a \texttt{congruifiedCalibrations} object are then matched to nodes in the chosen tree topology and stored as a \texttt{matchedCalibrations} object.
+
+\emph{\textbf{Summarizing node ages.--}} DateLife summarizes matched calibrations into a single patristic distance matrix using different methods.
+Summarizing options implemented include Super Distance Matrix method (SDM, Criscuolo, Berry, Douzery, \& Gascuel, 2006) and summary statistics such as median, minimum and maximum ages.
+
+\emph{\textbf{Dating the tree topology.--}} Summarized calibrations can be applied as secondary calibrations with different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Campbell O.
+Webb, Ackerly, \& Kembel, 2008; Campbell O.
+Webb \& Donoghue, 2005), and treePL (Stephen A.
+Smith \& O'Meara, 2012).
+By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm that assigns ages to nodes with no data evenly between nodes with age data, which minimizes age variance in the resulting chronogram (Campbell O.
+Webb et al., 2008).
+When there is conflict in ages across node with age data, the algorithm ignores ages that are older than parent nodes and/or younger than descendant nodes.
+
+If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature.
+If none is provided, DateLife assigns an arbitrary age to the root as 10\% older than the oldest age available within the group.
+
+\emph{\textbf{Visualizing results.--}}
+Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R ``phylo'' format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions.
 
@@
 
-\hypertarget{results}{%
-\section{Results}\label{results}}
-
 \hypertarget{case-study}{%
-\subsection{Case study}\label{case-study}}
+\section{Case study}\label{case-study}}
 
@@
 \hypertarget{cross-validation-test}{%
-\subsection{Cross-validation test}\label{cross-validation-test}}
+\section{Cross-validation test}\label{cross-validation-test}}
 
@@
 Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cvbladj}).
-We also demonstrate the usage of PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007) as an alternative method to BLADJ.
+We also demonstrate the usage of PATHd8 (Britton et al., 2007) as an alternative method to BLADJ.
 For this, we run a \texttt{datelife} branch length reconstruction that searches for DNA sequence data from the Barcode of Life Data System {[}BOLD; Ratnasingham and Hebert (2007){]} to generate branch lengths.
@@
 
-At the time of writing of this manuscript (Mar 07, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
+At the time of writing of this manuscript (Mar 08, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
 A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time.
@@
 In other areas of biological research, such as ecology and conservation biology, it has been shown that at least some data on lineage divergence represents a relevant improvement for testing alternative hypothesis using phylogenetic distance (Campbell O.
-Webb, Ackerly, \& Kembel, 2008).
+Webb et al., 2008).
 Hence, we integrated into datelife's workflow different ways of creating branch lengths in the absence of starting branch length information for taxa lacking this information (using the BLADJ option).
 ``Making up'' branch lengths is an accepted practice in scientific publications: Jetz, Thomas, Joy, Hartmann, and Mooers (2012), created a time-calibrated tree of all 9,993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
-(2018) created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37\% had molecular data; Smith and Brown (2018) constructed a tree of 353,185 seed plants where only 23\% had molecular data.
+(2018) created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37\% had molecular data; Stephen A.
+Smith and Brown (2018) constructed a tree of 353,185 seed plants where only 23\% had molecular data.
 Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch assignation methods.
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-ane2009groves}{}}%
+Ané, C., Eulenstein, O., Piaggio-Talice, R., \& Sanderson, M. J. (2009). Groves of phylogenetic trees. \emph{{Annals of Combinatorics}}, \emph{13}(2), 139--167.
+
 \leavevmode\vadjust pre{\hypertarget{ref-antonelli2017supersmart}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-delsuc2018phylogenomic}{}}%
+Delsuc, F., Philippe, H., Tsagkogeorga, G., Simion, P., Tilak, M.-K., Turon, X., \ldots{} Douzery, E.
+J.
+(2018).
+A phylogenomic framework and timescale for comparative studies of tunicates.
+\emph{BMC Biology}, \emph{16}(1), 1--14.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Eastman2013}{}}%
+Eastman, J.
+M., Harmon, L.
+J., \& Tank, D.
+C.
+(2013).
+{Congruification: Support for time scaling large phylogenetic trees}.
+\emph{Methods in Ecology and Evolution}, \emph{4}(7), 688--691.
+\url{https://doi.org/10.1111/2041-210X.12051}
+
 \leavevmode\vadjust pre{\hypertarget{ref-Felsenstein1985a}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-Huelsenbeck2001}{}}%
+Huelsenbeck, J. P., \& Ronquist, F. (2001). {MRBAYES: Bayesian inference of phylogenetic trees}. \emph{Bioinformatics}, \emph{17}(8), 754--755. \url{https://doi.org/10.1093/bioinformatics/17.8.754}
+
 \leavevmode\vadjust pre{\hypertarget{ref-Jetz2012}{}}%
@@
 \url{https://doi.org/10.1038/nature11631}
+
+\leavevmode\vadjust pre{\hypertarget{ref-laubichler2009form}{}}%
+Laubichler, M. D., \& Maienschein, J. (2009). \emph{Form and function in developmental evolution}. Cambridge University Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-magallon2001absolute}{}}%
+Magallon, S., \& Sanderson, M. J. (2001). Absolute diversification rates in angiosperm clades. \emph{Evolution}, \emph{55}(9), 1762--1780.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-opentreeAPIs}{}}%
-OpenTreeOfLife, Redelings, B., Cranston, K.
+Open Tree Of Life, Redelings, B., Cranston, K.
 A., Allman, J., Holder, M.
@@
 J.
-(n.d.).
-{Open Tree of Life APIs v.
-3.0}.
+(2016).
+{Open Tree of Life APIs v3.0}.
+\emph{{Open Tree of Life Project}}, (Online Resources).
 Retrieved from \url{\%7Bhttps://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\%7D}
+
+\leavevmode\vadjust pre{\hypertarget{ref-opentreeoflife2019synth}{}}%
+Open Tree Of Life, Redelings, B., Sánchez Reyes, L.
+L., Cranston, K.
+A., Allman, J., Holder, M.
+T., \& McTavish, E.
+J.
+(2019).
+Open tree of life synthetic tree v12.3.
+\emph{Zenodo}.
+Retrieved from \url{https://doi.org/10.5281/zenodo.3937742}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-rees2017automated}{}}%
+Rees, J. A., \& Cranston, K. (2017). Automated assembly of a reference taxonomy for phylogenetic data synthesis. \emph{Biodiversity Data Journal}, (5).
+
 \leavevmode\vadjust pre{\hypertarget{ref-Revell2012}{}}%
 Revell, L. J. (2012). Phytools: An r package for phylogenetic comparative biology (and other things). \emph{Methods in Ecology and Evolution}, \emph{3}, 217--223.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Ronquist2003}{}}%
+Ronquist, F., \& Huelsenbeck, J.
+P.
+(2003).
+{MrBayes 3: Bayesian phylogenetic inference under mixed models}.
+\emph{Bioinformatics}, \emph{19}(12), 1572--1574.
+\url{https://doi.org/10.1093/bioinformatics/btg180}
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-smith2018constructing}{}}%
-Smith, S. A., \& Brown, J. W. (2018). Constructing a broadly inclusive seed plant phylogeny. \emph{American Journal of Botany}, \emph{105}(3), 302--314.
+Smith, Stephen A., \& Brown, J. W. (2018). Constructing a broadly inclusive seed plant phylogeny. \emph{American Journal of Botany}, \emph{105}(3), 302--314.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Smith2012}{}}%
+Smith, Stephen A., \& O'Meara, B.
+C.
+(2012).
+{TreePL: Divergence time estimation using penalized likelihood for large phylogenies}.
+\emph{Bioinformatics}, \emph{28}(20), 2689--2690.
+\url{https://doi.org/10.1093/bioinformatics/bts492}
 
@@
 \includegraphics{../figures/figure1/figure1-vertical-final.pdf}
-\caption{Stylized DateLife workflow.
-This shows the general workflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{www.datelife.org/query/}.
-Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.}
+\caption{Stylized DateLife algorithm showing the main steps that are performed with the R package \texttt{datelife}, or online at \url{www.datelife.org/query/}.
+Details on the functions involved on each workflow are available at \texttt{datelife}'s R package manual https://phylotastic.org/datelife/articles/.}
 \label{fig:workflow}

### 2022-03-10 manuscript/manuscript.Rmd
methods comments

@@
 These estimates also provide valuable information for for education, science communication and policy decisions.
-  Although achieving a high-quality reconstruction of time of evolutionary origin as a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is still a difficult and time-consuming task, the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
+  Although achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is a difficult and time-consuming task, the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
   When these estimates are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on time of evolutionary origin is exposed in a programatic and reusable way.
@@
 If the input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004].
-The input tree is not required to have branch lengths, and its topology is used on the summary steps described below.
+The input tree is not required to have branch lengths, and its topology is used in the summary steps described below.
 
-Currently, DateLife only accepts scientific names as input, and they can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or binomial specific.
+DateLife accepts scientific names as input.
+These names can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or binomial specific.
 Subspecies and variants are ignored.
@@
 
-Input scientific names are processed using the Taxonomic Name Resolution Service (TNRS; @Boyle2013), which increases the probability of correctly finding the queried taxon names in the chronogram database.
+Input scientific names are processed using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database.
 TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard.
-DateLife implements TNRS using OpenTree's taxonomic as standard [@opentreeAPIs; @rees2017automated].
+DateLife implements TNRS using OpenTree's taxonomy as standard [@opentreeAPIs; @rees2017automated].
+<!--_I rewroded a bit, and removed a Boyle citation because it sounded like you were using their TNRS - but you aren't right?  -->
 
 The processed input taxon names are saved as an R object of a newly defined class \texttt{datelifeQuery} that is used in the following steps.
-It contains the processed names, the corresponding OpenTree taxonomic id numbers, and the topology of the input tree if any was provided.
+This object contains the processed names, the corresponding OpenTree taxonomic id numbers, and the topology of the input tree if any was provided.
 
@@
 Matching pruned chronograms are stored as individual patristic distance matrices (Figure \ref{fig:workflow} subfigure X). 
+This matrix consists of ... ???? the pairwise distance between pairs of query taxa whcih are in that input tree, in units of millions of years. <!-- is that true? -->
+
 This format speeds up extraction of pairwise taxon ages of the queried taxa, as opposed to searching the ancestor node of a pair of taxa in a "phylo" object or newick string.
@@
 
-At this point, summary information is extracted from the \texttt{datelifeResult} object to inform decisions for ulterior steps in the user workflow.
-Age data from matching pruned chronograms is summarized and used to generate a single summary chronogram.
-Other basic summary information available is:
+At this point, summary information is extracted from the \texttt{datelifeResult} object to inform decisions for the subsequent steps in the user workflow.
+Age data from the matching pruned chronograms is summarized and used to generate a single summary chronogram.
+Other basic summary information available to the user is:
 
@@
 This can correspond to the age of the most recent common ancestor (mrca) of your group of interest if the pruned chronograms have all taxa belonging to the group.
-If not, the root corresponds to the mrca ofa subgroup withing your group of interest.
+If not, the root corresponds to the mrca of a subgroup withing your group of interest.
 3. Study citations where original chronograms were published.
@@
 By default, DateLife chooses the grove with the most taxa, however, the "criterion = trees" flag allows the user to choose the grove with the most trees instead.
+<!--Why do groves matter? Why can't you just use all the trees?-->
 
-_**Choosing a topology.--**_ Due to sampling, DateLife requires a tree topology to summarize age data upon it.
+_**Choosing a topology.--**_ DateLife requires a tree topology to summarize age data upon.
 Users can provide one as input from the literature, or one of their own making.
@@
 
-DateLife can also reconstruct branch lengths proportional to substitution rates on the fixed tree topology from available genetic data from BOLD.
+DateLife can also reconstruct branch lengths proportional to substitution rates on the fixed tree topology using available genetic data from BOLD.
 
@@
 Congruified node ages stored as a `congruifiedCalibrations` object are then matched to nodes in the chosen tree topology and stored as a `matchedCalibrations` object.
+<!--This doesn't use the input chronogram pruned patristic distance matricies? Or does it? -->
 
-_**Summarizing node ages.--**_ DateLife summarizes matched calibrations into a single patristic distance matrix using different methods.
+_**Summarizing node ages.--**_ DateLife summarizes matched calibrations  ???
+How is 'calibrations' defined here?
+the calibrations from congruification, or the pairwise distances???
+into a single patristic distance matrix using different methods.
 Summarizing options implemented include Super Distance Matrix method [SDM, @Criscuolo2006] and summary statistics such as median, minimum and maximum ages.
+For each node in the output tree topology, the pairiwise distances that traverse that node are ..???
+Under the SDM criterions ??? what happens??
+Using the minimum age criterion ..??
 
 _**Dating the tree topology.--**_ Summarized calibrations can be applied as secondary calibrations with different dating methods currently supported within DateLife: MrBayes [@Huelsenbeck2001; @Ronquist2003], PATHd8 [@Britton2007], BLADJ [@Webb2008; @webb2005phylomatic], and treePL [@Smith2012].
+<!--Huh? can you run MrBayes in DateLife?-->
+
 By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm that assigns ages to nodes with no data evenly between nodes with age data, which minimizes age variance in the resulting chronogram [@Webb2008].
@@
 They decide to use `datelife` because they are teaching best practices for reproducibility.
-Students have the option to go to the website at www.datelife.org and perform an interactive run. However, the educator wants the students to practice their R skills.
+Students have the option to go to the website at www.datelife.org and perform an interactive run. However, the educator also wants the students to practice their R skills.
 The first step is to run a `datelife` query using the "get species from taxon" flag.
@@
 
-Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603  to Datelife project and DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), and the Phylotastic project Grant ABI-1458572.
+Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603  to Datelife project and DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), and the Phylotastic project Grant ABI-1458572, and the OpenTree grant ABI-1759846.
 

### 2022-03-24 manuscript/manuscript.Rmd

@@
 DateLife implements TNRS using OpenTree's taxonomy as standard [@opentreeAPIs; @rees2017automated].
-<!--_I rewroded a bit, and removed a Boyle citation because it sounded like you were using their TNRS - but you aren't right?  -->
+<!--_I reworded a bit, and removed a Boyle citation because it sounded like you were using their TNRS - but you aren't right?  -->
 
@@
 Matching pruned chronograms are stored as individual patristic distance matrices (Figure \ref{fig:workflow} subfigure X). 
-This matrix consists of ... ???? the pairwise distance between pairs of query taxa whcih are in that input tree, in units of millions of years. <!-- is that true? -->
+This matrix consists of ... ???? the pairwise distance between pairs of query taxa which are in that input tree, in units of millions of years. <!-- is that true? -->
 
@@
 It is constructed from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert phylogenetic knowledge with rich metadata [@mctavish2015phylesystem] that allows automatic and reproducible construction of a chronogram database.
+New chronograms can be added to Phylesystem by any user and are immediately publicly available, and the DateLife database can be updated to include those new data within a run.
 
@@
 Summarizing options implemented include Super Distance Matrix method [SDM, @Criscuolo2006] and summary statistics such as median, minimum and maximum ages.
-For each node in the output tree topology, the pairiwise distances that traverse that node are ..???
-Under the SDM criterions ??? what happens??
+For each node in the output tree topology, the pairwise distances that traverse that node are ..???
+Under the SDM criterion ??? what happens??
 Using the minimum age criterion ..??
@@
 _**Dating the tree topology.--**_ Summarized calibrations can be applied as secondary calibrations with different dating methods currently supported within DateLife: MrBayes [@Huelsenbeck2001; @Ronquist2003], PATHd8 [@Britton2007], BLADJ [@Webb2008; @webb2005phylomatic], and treePL [@Smith2012].
-<!--Huh? can you run MrBayes in DateLife?-->
+<!--Huh? can you run MrBayes in DateLife? You should explain how the BOLD dating with PathD8 works here, and what is needed for the other ones!-->
 
@@
 Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
-With both methods, ages are generally consistent with source ages, but there are some biological examples in which this is not true (see Discussion).
+?? Say some things about the results! 
 <!-- (Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
@@
 
-Data from source chronograms can be also used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates (Figs.
+Data from source chronograms can be used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates (Figs.
 \ref{fig:cvbladj} and \ref{fig:cvbold}).
-As a form of cross validation, we took tree topologies from each study and calibrated them using time of lineage divergence data from all other source chronograms.
-In the absence of branch lengths, the ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig.
-\ref{fig:cvbladj}).
+As a form of cross validation, we took tree topologies from each input study and calibrated them using time of lineage divergence data from all other source chronograms.
+<!--I'm confused about why you are talking about branch lengths here... did you use branch lengths?-->
+
+In the absence of branch lengths, the ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvbladj}).
 Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cvbladj}).
@@
 <!-- two studies when expanding calibrations to make them agree (studies 3 and 5), and from  -->
-three cases (studies 3, 5, and 9, shown in Fig.
-\ref{fig:cvbold}).
+three cases (studies 3, 5, and 9, shown in Fig. \ref{fig:cvbold}). 
+<!-- What made it unsuccessful in the other cases? -->
 From these, two trees have a different sampling than the original source chronogram, mainly because DNA BOLD  data for some species is absent from the database. 
-Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
+??? Node ages or maximum ages?? Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
 More examples and code used to generate these trees were developed on an open repository that is available for consultation and reuse at <https://github.com/LunaSare/datelife_examples>.
@@
 
-The main goal of `datelife` is to make state-of-the-art information on time of lineage divergence easily accesible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
-It is an open service that does not require any expert biological knowledge from users --besides the names of the organisms they want to work with, for any of its functionalities.
+The main goal of `datelife` is to make state-of-the-art information on time of lineage divergence easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
+It is an open service that does not require any expert biological knowledge from users --besides the names of the organisms they want to work with, for any of its functionality.
 
@@
 `datelife`'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
-Users can also upload new chronograms to OpenTree themselves, and trigger an update of the `datelife` database to incorporate the new chronograms, to have them immediately available for analysis.
+Users can also upload new chronograms to OpenTree themselves, and trigger an update of their local `datelife` database to incorporate the new chronograms, to have them immediately available for analysis.
 
@@
 
-The only summary chronogram encompassing all life that is currently openly available from OpenTree
+The largest, and taxonomically broadest, summary chronogram currently available from OpenTree
 was constructed using age data from 2,274 published chronograms [@Hedges2015].
-However source chronograms are not available in computer readable format for reuse or reanalysis.
-As this tree is part of datelife's database, the amount of lineages that can be queried using `datelife` (`r length(unique(unname(unlist(lapply(opentree_chronograms$trees, "[[", "tip.label")))))` unique terminal taxa) is substantial, yet it can be improved.
-Consequently, we would like to emphasize on the importance of sharing chronogram data for the benefit of the scientific community as a whole, into repositories that require expert input and manual curation, such as OpenTree's Phylesystem [@mctavish2015phylesystem].
+However the source chronograms used as input data for this tree are not available in computer readable format for reuse or reanalysis.
+As this tree is part of datelife's database, the amount of lineages that can be queried using `datelife` (`r length(unique(unname(unlist(lapply(opentree_chronograms$trees, "[[", "tip.label")))))` unique terminal taxa) is substantial.
+Access to the input chronograms used to generate the Hedges et al.
+summary tree would improve measures of uncertainty in DateLife, but they are available only as image files and not as usable data (timetree.org).
+<!--IN VIOLATION OF NSF RULES! but I don't suppose we should say that.-->
+We would like to emphasize on the importance of sharing chronogram data for the benefit of the scientific community as a whole, into repositories that require expert input and manual curation, such as OpenTree's Phylesystem [@mctavish2015phylesystem].
 
-By default, `datelife` currently summarizes all source chronograms that overlap with at least two species names.
-Users can exclude source chronograms if they have reasons to do so.
+By default, `datelife` currently summarizes all source chronograms that overlap with at least two species names. Users can exclude source chronograms if they have reasons to do so. 
 Strictly speaking, the best chronogram should reflect the real time of lineage divergence accurately and precisely. 
-To our knowledge, there are no good measures to determine if a chronogram is better than another.
+To our knowledge, there are no good measures to determine independently if a chronogram is better than another.
 Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used [@magallon2010using, @magallon2015metacalibrated]. 
+Several characteristics of the data used for dating analyses as well as from the output chronogram itself, could be used to score quality of source chronograms. 
+Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
+DateLife provides an opportunity to capture concordance and conflict among date estimates, which can also be used as a metric for chronogram reliability.
+
 Scientists usually also favor chronograms constructed using primary calibrations (ages obtained from the fossil or geological record) to ones constructed with secondary calibrations (ages coming from other chronograms)[@schenk2016sec].
-It has been observed with simulations that divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with bayesian inference methods when priors are implemented in similar ways in both analyses [@schenk2016sec].
+It has been observed with simulations that divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with Bayesian inference methods when priors are implemented in similar ways in both analyses [@schenk2016sec].
 However, secondary calibrations
 can be applied using other dating methods that do not require setting priors, such as penalized likelihood [@sanderson2003r8s], or as fixed ages,
-potentially mitigating the bias reported with bayesian methods.
-Certainly, further studies are required to fully understand the effect of using secondary calibrations on time estimates and downstream anlyses.
+potentially mitigating the bias reported with Bayesian methods. 
+Certainly, further studies are required to fully understand the effect of using secondary calibrations on time estimates and downstream analyses.
 
 Furthermore, even chronograms obtained with primary fossil data can vary substantially in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
+<!--I think that one very odd one didn't have any fossils though! Maybe worth mentioning here?-->
 This observation is often encountered in the literature
@@
 Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses the same one to calibrate outside the clade, the resulting age estimates will often differ substantially, as the placement of calibrations as stem or crown group is proved to deeply affect estimated times of lineage divergence [@sauquet2013practical].
-Trying to summarize the resulting chronograms into a single one using simple summary statistics might erase all types of relevant information from the source chronograms.
-Accordingly, the prevailing view in our research community is that we should favor time of lineage divergence estimates obtained from a single analysis, using fossil data as primary sources of calibrations, and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history [@antonelli2017supersmart].
+Trying to summarize the resulting chronograms into a single one using simple summary statistics can erase many types of relevant information from the source chronograms.
+Accordingly, the prevailing view is that we should favor time of lineage divergence estimates obtained from a single analysis, using fossil data as primary sources of calibrations, and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history [@antonelli2017supersmart].
 <!-- Will this be implemented in future `datelife` versions?
@@
 
-Alternatively, one could try to choose the "best" chronogram from a set of possible evolutionary hypotheses.
-Several characteristics of the data used for dating analyses as well as from the output chronogram itself, could be used to score quality of source chronograms.
-Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
-To facilitate subsetting of source chronograms following different criteria by the users, this information should be included as metadata manually entered by curators in the future.
-
 <!-- Still, even if all source chronograms have been generated by excellent standards and using similar methods, the evolutionary history they depict might be very different.  -->
 
-In other areas of biological research, such as ecology and conservation biology, it has been shown that at least some data on lineage divergence represents a relevant improvement for testing alternative hypothesis using phylogenetic distance [@Webb2008].
-Hence, we integrated into datelife's workflow different ways of creating branch lengths in the absence of starting branch length information for taxa lacking this information (using the BLADJ option).
+Nonetheless, in ecology and conservation biology, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance [@Webb2008].
+Hence, we integrated into datelife's workflow different ways of estimating node ages in the absence of calibrations and branch length information for taxa lacking this information. 
 "Making up" branch lengths is an accepted practice in scientific publications: @Jetz2012, created a time-calibrated tree of all 9,993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a tree of 353,185 seed plants where only 23% had molecular data.
-Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch assignation methods.
-There has yet to be a thorough analysis of what can go wrong when one goes beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make sure results are substantially similar between trees fully reconstructed with molecular or other data, and trees that are reconstructed using taxonomy by resolving polytomies at random following a statistical model.
+Obviously, there are risks in this practice!
+Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch estimation methods.
+There has yet to be a thorough analysis of what can go wrong when one extends inferences beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make carefully consider the statistical assumptions being made, and assess the consistency of the results with prior work.
 
@@
 This provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identification of regions that require more research or that have conflicting information.
+It is available as an R package, or a web-based R shiny app at dates.opentreeloflife.org/datelife.
 <!-- It can also summarize public data, or use it to generate new chronograms based on trees provided by the user. -->

### 2022-03-24 manuscript/manuscript.tex

@@
 \shorttitle{DateLife: revealing the dated Tree of Life}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 3863}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 3868}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 These estimates also provide valuable information for for education, science communication and policy decisions.
-Although achieving a high-quality reconstruction of time of evolutionary origin as a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is still a difficult and time-consuming task, the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
+Although achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is a difficult and time-consuming task, the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
 When these estimates are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on time of evolutionary origin is exposed in a programatic and reusable way.
@@
 The DateLife algorithm is fully implemented using the R language.
-The latest stable version of the R package \texttt{datelife} is available from the CRAN repository (v0.6.2; Sanchez-Reyes et al.
+The latest stable version of the R package \texttt{datelife} is available from the CRAN repository (v0.6.1; Sanchez-Reyes et al.
 (2022)), and relies on functionalities from various biological R packages:
@@
 
-DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies.
+DateLife's chronogram database latest version consist of 231 chronograms published in 171 different studies.
 It is constructed from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert phylogenetic knowledge with rich metadata (McTavish et al., 2015) that allows automatic and reproducible construction of a chronogram database.
@@
 
-At the time of writing of this manuscript (Mar 08, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
+At the time of writing of this manuscript (Mar 10, 2022), \texttt{datelife}'s database has 231 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
 A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time.
@@
 
-Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to Datelife project and DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), and the Phylotastic project Grant ABI-1458572.
+Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to Datelife project and DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), and the Phylotastic project Grant ABI-1458572, and the OpenTree grant ABI-1759846.
 

### 2022-03-24 manuscript/figures-alt.tex
updating figures on ms

@@
 \begin{figure}[!h]
-\includegraphics{../figures/Fig1.pdf}
-\caption{}
+\includegraphics{../figures/figure1/figure1-vertical-final.pdf}
+\caption{Stylized DateLife workflow.
+This shows the general worflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{http://www.datelife.org/}.
+Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
+}
 \label{fig:workflow}
@@
 % \end{center}
-Stylized DateLife workflow.
-This shows the general worflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{http://www.datelife.org/}.
-Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
 \newpage
@@
 \includegraphics[width=1\linewidth]{../figures/fig_runtime_main.pdf}
-\caption{}
+\caption{Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names.
+We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray).
+We also performed a search using the already processed query for comparison (light blue).}
 \label{fig:runtime1}
@@
 % \begin{center}
-% \textsc{Figure \ref{fig:runtime1}}
+% \textsc{Figure \ref{fig:benchmark}}
 % \end{center}
-Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names.
-We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray).
-We also performed a search using the already processed query for comparison (light blue).
 \newpage
 \begin{figure}[!h]
-\includegraphics[width=1\linewidth]{../figures/fig_schronograms1.pdf}
-\caption{}
-\label{fig:schronograms1}
+\includegraphics[width=1\linewidth]{../figures/figure2/figure2-1.pdf}
+\caption{Age data results of a DateLife search of a small sample of 6 bird species within the Passeriformes.
+Input names were found across 9 chronograms within 6 independent studies (Barker et al.
+(\protect\hyperlink{ref-barker2012going}{2012}), Barker et al.
+(\protect\hyperlink{ref-barker2015new}{2015}), Burns et al.
+(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al.
+(\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al.
+(\protect\hyperlink{ref-Jetz2012}{2012}).) This revealed 28 age data points for the queried species names.}
+\label{fig:figure2-1}
 \end{figure}
 % \begin{center}
-% \textsc{Figure \ref{fig:schronograms1}}
+% \textsc{Figure \ref{fig:figure2-1}}
 % \end{center}
-Lineage through time (LTT) plots of source chronograms containing all or a subset of species from the bird family Fringillidae of true finches.
-Arrows indicate maximum age of each chronogram.
-Numbers reference to chronograms' original publications 1: Barker et al.
-(\protect\hyperlink{ref-barker2012going}{2012}), 2: Barker et al.
-(\protect\hyperlink{ref-barker2015new}{2015}), 3: Burns et al.
-(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), 4: Claramunt and Cracraft (\protect\hyperlink{ref-claramunt2015new}{2015}), 5: Gibb et al.
-(\protect\hyperlink{ref-gibb2015new}{2015}), 6: Hedges et al.
-(\protect\hyperlink{ref-Hedges2015}{2015}), 7: Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), 8: Jetz et al.
-(\protect\hyperlink{ref-Jetz2012}{2012}), 9: Price et al.
-(\protect\hyperlink{ref-price2014niche}{2014}).
 \newpage
 \begin{figure}[!h]
-\includegraphics{../figures/fig_summaries.pdf}
+\includegraphics{../figures/figure2/figure2-2.pdf}
 \caption{}
@@
 % \begin{center}
-% \textsc{Figure \ref{fig:summaries}}
+% \textsc{Figure \ref{fig:figure2-2}}
 % \end{center}

### 2022-03-24 manuscript/figures.tex
updating figures on ms

@@
-% in this file I leave the figure captions outside\ caption{} because I want them
-% to be formatted in the same way as the general text (double spaced and linenumbered)
 \captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% WORKFLOW FIGURE
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \includegraphics{../figures/figure1/figure1-vertical-final.pdf}
-\caption{Stylized DateLife algorithm showing the main steps that are performed with the R package \texttt{datelife}, or online at \url{www.datelife.org/query/}.
-Details on the functions involved on each workflow are available at \texttt{datelife}'s R package manual https://phylotastic.org/datelife/articles/.}
+\caption{
+Stylized DateLife algorithm showing the main steps that are performed
+with the R package \texttt{datelife}, or online at \url{www.datelife.org/query/}.
+Details on the functions involved on each workflow are available at \texttt{datelife}'s
+R package manual https://phylotastic.org/datelife/articles/.
+}
 \label{fig:workflow}
 \end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:workflow}}
-% \end{center}
-% Stylized DateLife workflow.
-This shows the general workflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{http://www.datelife.org/}.
-Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% BENCHMARK FIGURE
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \newpage
 \begin{figure}[!h]
-\includegraphics[width=1\linewidth]{../figures/fig_runtime_main.pdf}
+\includegraphics{../figures/fig_runtime_main.pdf}
 \caption{Input taxon name processing and chronogram database search computation time increases with number of input taxon names.
@@
 We also performed a search using the already processed query for comparison (light blue).}
-\label{fig:runtime1}
+\label{fig:benchmark}
 \end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:runtime1}}
-% \end{center}
-%
+\newpage
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% SMALL EXAMPLE FIGURES
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \newpage
 \begin{figure}[!h]
-\includegraphics[width=1\linewidth]{../figures/fig_schronograms1.pdf}
-\caption{Lineage through time (LTT) plots of source chronograms containing all or a subset of species from the bird family Fringillidae of true finches.
-Arrows indicate maximum age of each chronogram.
-Numbers reference to chronograms' original publications 1: Barker et al.
-(\protect\hyperlink{ref-barker2012going}{2012}), 2: Barker et al.
-(\protect\hyperlink{ref-barker2015new}{2015}), 3: Burns et al.
-(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), 4: Claramunt and Cracraft (\protect\hyperlink{ref-claramunt2015new}{2015}), 5: Gibb et al.
-(\protect\hyperlink{ref-gibb2015new}{2015}), 6: Hedges et al.
-(\protect\hyperlink{ref-Hedges2015}{2015}), 7: Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), 8: Jetz et al.
-(\protect\hyperlink{ref-Jetz2012}{2012}), 9: Price et al.
-(\protect\hyperlink{ref-price2014niche}{2014}).
+\includegraphics[width=1\linewidth]{../figures/figure2/figure2-1.pdf}
+\caption{
+Age data results of a DateLife search of a small sample of 6 bird species within the Passeriformes.
+Input names were found across 9 chronograms within 6 independent studies (Barker et al.
+(\protect\hyperlink{ref-barker2012going}{2012}), Barker et al.
+(\protect\hyperlink{ref-barker2015new}{2015}), Burns et al.
+(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al.
+(\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al.
+(\protect\hyperlink{ref-Jetz2012}{2012}).) This revealed 28 age data points for the queried species names.
 }
-\label{fig:schronograms1}
+\label{fig:figure2-1}
 \end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:schronograms1}}
-% \end{center}
-%Lineage through time (LTT) plots of source chronograms containing all or a subset of species from the bird family Fringillidae of true finches.
-Arrows indicate maximum age of each chronogram.
-Numbers reference to chronograms' original publications 1: Barker et al.
-(\protect\hyperlink{ref-barker2012going}{2012}), 2: Barker et al.
-(\protect\hyperlink{ref-barker2015new}{2015}), 3: Burns et al.
-(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), 4: Claramunt and Cracraft (\protect\hyperlink{ref-claramunt2015new}{2015}), 5: Gibb et al.
-(\protect\hyperlink{ref-gibb2015new}{2015}), 6: Hedges et al.
-(\protect\hyperlink{ref-Hedges2015}{2015}), 7: Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), 8: Jetz et al.
-(\protect\hyperlink{ref-Jetz2012}{2012}), 9: Price et al.
-(\protect\hyperlink{ref-price2014niche}{2014}).
+
+\newpage
+\begin{figure}[!h]
+\includegraphics[width=1\linewidth]{../figures/figure2/figure2-2.pdf}
+\caption{
+Summarized age data is used as secondary calibrations to obtain a summary chronogram.
+}
+\label{fig:figure2-2}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% REAL LIFE APP FIGURE
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \newpage
@@
 \end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:summaries}}
-% \end{center}
-% LTT plots of median (top) and Supermatrix Distance Method (SDM; bottom) chronograms summarising information from source chronograms found for the Fringillidae. Arrows indicate tree maximum age.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% CROSS VALIDATION FIGURES
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \newpage
@@
 \end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:cvbladj}}
-% \end{center}
-% LTT plots showing results from the cross-validation analyses of trees without branch lengths dated using BLADJ.
-The dating analysis can only be performed in trees with more than 2 tips, thus excluding chronogram from study 4; its data was still used as calibration for the other source chronograms.
+
 \newpage
@@
 \end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:cvbold}}
-% \end{center}
-FIGURES
-%LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8.
-We could construct a tree with branch lengths for all source chronograms.
-However, dating with PATHd8 was only successful in three source chronograms shown here.
-
-%\end{linenumbers}

### 2022-03-24 manuscript/manuscript.Rmd
updating figures on ms

@@
       in_header: "preamble.tex"
-      after_body  : "figures.tex"
+      after_body  : "figures-alt.tex"
 
 nocite: |
-  @barker2012going, @barker2015new, @burns2014phylogenetics, @claramunt2015new, @gibb2015new, @Hedges2015, @hooper2017chromosomal, @Jetz2012, @price2014niche
+  @barker2013going, @barker2015new, @burns2014phylogenetics, @claramunt2015new, @gibb2015new, @Hedges2015, @hooper2017chromosomal, @Jetz2012, @price2014niche
 ---
@@
 # Introduction
-<!-- If you're aiming for syst bio you don't want to start so general Readers know what chronorgrams and phylogenies are!.-->
+<!-- If you're aiming for syst bio you don't want to start so general Readers know what chronograms and phylogenies are!.-->
 Chronograms --phylogenies with branch lengths proportional to time-- provide key data for the study of natural processes in many areas of biological research, such as developmental biology [@laubichler2009form; @delsuc2018phylogenomic], conservation biology [@Felsenstein1985a; @Webb2000], historical biogeography [@posadas2006historical], and species diversification [@magallon2001absolute; @Morlon2014].
@@
 
-DateLife can also reconstruct branch lengths proportional to substitution rates on the fixed tree topology using available genetic data from BOLD.
+DateLife can also reconstruct branch lengths proportional to substitution rates on a fixed tree topology using available genetic data from BOLD.
 
@@
 
-_**Summarizing node ages.--**_ DateLife summarizes matched calibrations  ???
-How is 'calibrations' defined here?
-the calibrations from congruification, or the pairwise distances???
+_**Summarizing node ages.--**_ DateLife summarizes matched calibrations
+<!--??? How is 'calibrations' defined here? the calibrations from congruification, or the pairwise distances??? -->
 into a single patristic distance matrix using different methods.
 Summarizing options implemented include Super Distance Matrix method [SDM, @Criscuolo2006] and summary statistics such as median, minimum and maximum ages.
+<!--
 For each node in the output tree topology, the pairwise distances that traverse that node are ..???
@@
 Using the minimum age criterion ..??
+-->
 
@@
 Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig.
-\ref{fig:runtime1}).
+\ref{fig:benchmark}).
 <!--Speed of other datelife workflows? NO, bc it reflects speed from other software mainly. But I will put a sentence about it...
@@
 GitHub, using the continuous integration tool Travis CI (<https://travis-ci.org>). At
-present, unit tests cover more than 30% of `datelife`'s code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->
+present, unit tests cover more than 40% of `datelife`'s code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->
 
@@
 
-We illustrate the `datelife` algorithm using the family of true finches, Fringillidae as an example.
+We illustrate the DateLife algorithm using a group within the Passeriform birds encompassing the family of true finches, Fringillidae and allies as case study.
+The first example analyses 6 bird species and shows all steps of the algorithm.
+The second example is a real life application
 <!--We used `r cite_r("r-references.bib")` for all our analyses.-->
+
+## Small example
+
+We chose 6 bird species associated to true finches at random. The sample includes
+two species of cardinals:
+the black-thighed grosbeak -- *Pheucticus tibialis* and
+the crimson-collared grosbeak -- *Rhodothraupis celaeno*;
+three species of buntings:
+the yellowhammer -- *Emberiza citrinella*,
+the pine bunting --  *Emberiza leucocephalos* and
+the yellow-throated bunting --  *Emberiza elegans*;
+and one species of tanager, the vegetarian finch -- *Platyspiza crassirostris*.
+
+Processing input names found that *Emberiza elegans* is synonym for *Schoeniclus elegans* in the reference taxonomy.
+DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies.
+Three studies matched five input names [@barker2015new; @Hedges2015; @Jetz2012], one study matched four input names [@hooper2017chromosomal] and two studies matched two input names [@barker2013going; @burns2014phylogenetics].
+No studies matched all input names.
+Together, matching chronograms have 28 unique age data points.
+All nodes have age data.
+As fixed tree topology, DateLife used OpenTree's synthetic tree as default and mapped age data to nodes in the tree.
+As expected, more inclusive nodes (e.g., node "n1") have more age data than less inclusive nodes (e.g., node "n5").
+The processing step allowd discovering five data points for node "n4" that would not have had any data otherwise.
+Age summary statistics per node were calculated and tested as secondary calibrations to date the tree topology using the BLADJ algorithm.
+Age data for node "n2" was excluded as final calibration because it is older than age data of a more inclusive node.
+
+## Real life application
+
 A college educator wishes to obtain state-of-the-art data on time of evolutionary origin of species belonging to the true finches for their class.
@@
 <!-- In case we want to cite the studies in here, but they are already cited in the figure caption: [Fig.
-\ref{fig:schronograms1}; @barker2012going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
+\ref{fig:schronograms1}; @barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
 -->

### 2022-03-24 manuscript/manuscript.tex
updating figures on ms

@@
 \shorttitle{DateLife: revealing the dated Tree of Life}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 3868}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4199}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 
-Chronograms --phylogenies with branch lengths proportional to time-- provide key data for the study of natural processes in many areas of biological research, such as developmental biology (Delsuc et al., 2018; Laubichler \& Maienschein, 2009), conservation biology (Felsenstein, 1985; Campbell O.
+Chronograms --phylogenies with branch lengths proportional to time-- provide key data for the study of natural processes in many areas of biological research, such as developmental biology (Delsuc et al., 2018; Laubichler \& Maienschein, 2009), conservation biology (Felsenstein, 1985; C.
 Webb, 2000), historical biogeography (Posadas, Crisci, \& Katinas, 2006), and species diversification (Magallon \& Sanderson, 2001; Morlon, 2014).
@@
 The DateLife algorithm is fully implemented using the R language.
-The latest stable version of the R package \texttt{datelife} is available from the CRAN repository (v0.6.1; Sanchez-Reyes et al.
+The latest stable version of the R package \texttt{datelife} is available from the CRAN repository (v0.6.2; Sanchez-Reyes et al.
 (2022)), and relies on functionalities from various biological R packages:
@@
 If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis et al., 2004).
-The input tree is not required to have branch lengths, and its topology is used on the summary steps described below.
+The input tree is not required to have branch lengths, and its topology is used in the summary steps described below.
 
-Currently, DateLife only accepts scientific names as input, and they can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or binomial specific.
+DateLife accepts scientific names as input.
+These names can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or binomial specific.
 Subspecies and variants are ignored.
@@
 
-Input scientific names are processed using the Taxonomic Name Resolution Service (TNRS; Boyle et al.
-(2013)), which increases the probability of correctly finding the queried taxon names in the chronogram database.
+Input scientific names are processed using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database.
 TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard.
-DateLife implements TNRS using OpenTree's taxonomic as standard (Open Tree Of Life et al., 2016; Rees \& Cranston, 2017).
+DateLife implements TNRS using OpenTree's taxonomy as standard (Open Tree Of Life et al., 2016; Rees \& Cranston, 2017).
 
 The processed input taxon names are saved as an R object of a newly defined class \texttt{datelifeQuery} that is used in the following steps.
-It contains the processed names, the corresponding OpenTree taxonomic id numbers, and the topology of the input tree if any was provided.
+This object contains the processed names, the corresponding OpenTree taxonomic id numbers, and the topology of the input tree if any was provided.
 
@@
 Matching pruned chronograms are stored as individual patristic distance matrices (Figure \ref{fig:workflow} subfigure X).
+This matrix consists of \ldots{} ???? the pairwise distance between pairs of query taxa which are in that input tree, in units of millions of years.
+
 This format speeds up extraction of pairwise taxon ages of the queried taxa, as opposed to searching the ancestor node of a pair of taxa in a ``phylo'' object or newick string.
@@
 
-DateLife's chronogram database latest version consist of 231 chronograms published in 171 different studies.
+DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies.
 It is constructed from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert phylogenetic knowledge with rich metadata (McTavish et al., 2015) that allows automatic and reproducible construction of a chronogram database.
+New chronograms can be added to Phylesystem by any user and are immediately publicly available, and the DateLife database can be updated to include those new data within a run.
 
@@
 This can correspond to the age of the most recent common ancestor (mrca) of your group of interest if the pruned chronograms have all taxa belonging to the group.
-If not, the root corresponds to the mrca ofa subgroup withing your group of interest.
+If not, the root corresponds to the mrca of a subgroup withing your group of interest.
 \item
@@
 
-\emph{\textbf{Choosing a topology.--}} Due to sampling, DateLife requires a tree topology to summarize age data upon it.
+\emph{\textbf{Choosing a topology.--}} DateLife requires a tree topology to summarize age data upon.
 Users can provide one as input from the literature, or one of their own making.
@@
 
-DateLife can also reconstruct branch lengths proportional to substitution rates on the fixed tree topology from available genetic data from BOLD.
+DateLife can also reconstruct branch lengths proportional to substitution rates on a fixed tree topology using available genetic data from BOLD.
 
@@
 
-\emph{\textbf{Summarizing node ages.--}} DateLife summarizes matched calibrations into a single patristic distance matrix using different methods.
+\emph{\textbf{Summarizing node ages.--}} DateLife summarizes matched calibrations
+into a single patristic distance matrix using different methods.
 Summarizing options implemented include Super Distance Matrix method (SDM, Criscuolo, Berry, Douzery, \& Gascuel, 2006) and summary statistics such as median, minimum and maximum ages.
@@
 Smith \& O'Meara, 2012).
+
 By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm that assigns ages to nodes with no data evenly between nodes with age data, which minimizes age variance in the resulting chronogram (Campbell O.
@@
 Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig.
-\ref{fig:runtime1}).
+\ref{fig:benchmark}).
 \texttt{datelife}'s code performance was evaluated with a set of unit tests designed and
@@
 GitHub, using the continuous integration tool Travis CI (\url{https://travis-ci.org}). At
-present, unit tests cover more than 30\% of \texttt{datelife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).
+present, unit tests cover more than 40\% of \texttt{datelife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).
 
@@
 
-We illustrate the \texttt{datelife} algorithm using the family of true finches, Fringillidae as an example.
+We illustrate the DateLife algorithm using a group within the Passeriform birds encompassing the family of true finches, Fringillidae and allies as case study.
+The first example analyses 6 bird species and shows all steps of the algorithm.
+The second example is a real life application
+
+\hypertarget{small-example}{%
+\subsection{Small example}\label{small-example}}
+
+We chose 6 bird species associated to true finches at random. The sample includes
+two species of cardinals:
+the black-thighed grosbeak -- \emph{Pheucticus tibialis} and
+the crimson-collared grosbeak -- \emph{Rhodothraupis celaeno};
+three species of buntings:
+the yellowhammer -- \emph{Emberiza citrinella},
+the pine bunting -- \emph{Emberiza leucocephalos} and
+the yellow-throated bunting -- \emph{Emberiza elegans};
+and one species of tanager, the vegetarian finch -- \emph{Platyspiza crassirostris}.
+
+Processing input names found that \emph{Emberiza elegans} is synonym for \emph{Schoeniclus elegans} in the reference taxonomy.
+DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies.
+Three studies matched five input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2015; Hedges, Marin, Suleski, Paymer, \& Kumar, 2015; Jetz, Thomas, Joy, Hartmann, \& Mooers, 2012), one study matched four input names (Hooper \& Price, 2017) and two studies matched two input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2013; Burns et al., 2014).
+No studies matched all input names.
+Together, matching chronograms have 28 unique age data points.
+All nodes have age data.
+As fixed tree topology, DateLife used OpenTree's synthetic tree as default and mapped age data to nodes in the tree.
+As expected, more inclusive nodes (e.g., node ``n1'') have more age data than less inclusive nodes (e.g., node ``n5'').
+The processing step allowd discovering five data points for node ``n4'' that would not have had any data otherwise.
+Age summary statistics per node were calculated and tested as secondary calibrations to date the tree topology using the BLADJ algorithm.
+Age data for node ``n2'' was excluded as final calibration because it is older than age data of a more inclusive node.
+
+\hypertarget{real-life-application}{%
+\subsection{Real life application}\label{real-life-application}}
+
 A college educator wishes to obtain state-of-the-art data on time of evolutionary origin of species belonging to the true finches for their class.
 They decide to use \texttt{datelife} because they are teaching best practices for reproducibility.
-Students have the option to go to the website at www.datelife.org and perform an interactive run. However, the educator wants the students to practice their R skills.
+Students have the option to go to the website at www.datelife.org and perform an interactive run. However, the educator also wants the students to practice their R skills.
 The first step is to run a \texttt{datelife} query using the ``get species from taxon'' flag.
@@
 Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
-With both methods, ages are generally consistent with source ages, but there are some biological examples in which this is not true (see Discussion).
+?? Say some things about the results!
 
@@
 
-Data from source chronograms can be also used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates (Figs.
+Data from source chronograms can be used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates (Figs.
 \ref{fig:cvbladj} and \ref{fig:cvbold}).
-As a form of cross validation, we took tree topologies from each study and calibrated them using time of lineage divergence data from all other source chronograms.
-In the absence of branch lengths, the ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig.
-\ref{fig:cvbladj}).
+As a form of cross validation, we took tree topologies from each input study and calibrated them using time of lineage divergence data from all other source chronograms.
+
+In the absence of branch lengths, the ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvbladj}).
 Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cvbladj}).
@@
 However, dating with PATHd8 using congruified calibrations, was only successful in
-three cases (studies 3, 5, and 9, shown in Fig.
-\ref{fig:cvbold}).
+three cases (studies 3, 5, and 9, shown in Fig. \ref{fig:cvbold}).
 From these, two trees have a different sampling than the original source chronogram, mainly because DNA BOLD data for some species is absent from the database.
-Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
+??? Node ages or maximum ages?? Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
 More examples and code used to generate these trees were developed on an open repository that is available for consultation and reuse at \url{https://github.com/LunaSare/datelife_examples}.
@@
 
-The main goal of \texttt{datelife} is to make state-of-the-art information on time of lineage divergence easily accesible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
-It is an open service that does not require any expert biological knowledge from users --besides the names of the organisms they want to work with, for any of its functionalities.
+The main goal of \texttt{datelife} is to make state-of-the-art information on time of lineage divergence easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
+It is an open service that does not require any expert biological knowledge from users --besides the names of the organisms they want to work with, for any of its functionality.
 
-At the time of writing of this manuscript (Mar 10, 2022), \texttt{datelife}'s database has 231 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
+At the time of writing of this manuscript (Mar 24, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
 A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time.
@@
 \texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
-Users can also upload new chronograms to OpenTree themselves, and trigger an update of the \texttt{datelife} database to incorporate the new chronograms, to have them immediately available for analysis.
+Users can also upload new chronograms to OpenTree themselves, and trigger an update of their local \texttt{datelife} database to incorporate the new chronograms, to have them immediately available for analysis.
 
@@
 
-The only summary chronogram encompassing all life that is currently openly available from OpenTree
-was constructed using age data from 2,274 published chronograms (Hedges, Marin, Suleski, Paymer, \& Kumar, 2015).
-However source chronograms are not available in computer readable format for reuse or reanalysis.
-As this tree is part of datelife's database, the amount of lineages that can be queried using \texttt{datelife} (86879 unique terminal taxa) is substantial, yet it can be improved.
-Consequently, we would like to emphasize on the importance of sharing chronogram data for the benefit of the scientific community as a whole, into repositories that require expert input and manual curation, such as OpenTree's Phylesystem (McTavish et al., 2015).
+The largest, and taxonomically broadest, summary chronogram currently available from OpenTree
+was constructed using age data from 2,274 published chronograms (Hedges et al., 2015).
+However the source chronograms used as input data for this tree are not available in computer readable format for reuse or reanalysis.
+As this tree is part of datelife's database, the amount of lineages that can be queried using \texttt{datelife} (99474 unique terminal taxa) is substantial.
+Access to the input chronograms used to generate the Hedges et al.~summary tree would improve measures of uncertainty in DateLife, but they are available only as image files and not as usable data (timetree.org).
+We would like to emphasize on the importance of sharing chronogram data for the benefit of the scientific community as a whole, into repositories that require expert input and manual curation, such as OpenTree's Phylesystem (McTavish et al., 2015).
 
-By default, \texttt{datelife} currently summarizes all source chronograms that overlap with at least two species names.
-Users can exclude source chronograms if they have reasons to do so.
+By default, \texttt{datelife} currently summarizes all source chronograms that overlap with at least two species names. Users can exclude source chronograms if they have reasons to do so.
 Strictly speaking, the best chronogram should reflect the real time of lineage divergence accurately and precisely.
-To our knowledge, there are no good measures to determine if a chronogram is better than another.
+To our knowledge, there are no good measures to determine independently if a chronogram is better than another.
 Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández (2015).
+Several characteristics of the data used for dating analyses as well as from the output chronogram itself, could be used to score quality of source chronograms.
+Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
+DateLife provides an opportunity to capture concordance and conflict among date estimates, which can also be used as a metric for chronogram reliability.
+
 Scientists usually also favor chronograms constructed using primary calibrations (ages obtained from the fossil or geological record) to ones constructed with secondary calibrations (ages coming from other chronograms)(Schenk, 2016).
-It has been observed with simulations that divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with bayesian inference methods when priors are implemented in similar ways in both analyses (Schenk, 2016).
+It has been observed with simulations that divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with Bayesian inference methods when priors are implemented in similar ways in both analyses (Schenk, 2016).
 However, secondary calibrations
 can be applied using other dating methods that do not require setting priors, such as penalized likelihood (Sanderson, 2003), or as fixed ages,
-potentially mitigating the bias reported with bayesian methods.
-Certainly, further studies are required to fully understand the effect of using secondary calibrations on time estimates and downstream anlyses.
+potentially mitigating the bias reported with Bayesian methods.
+Certainly, further studies are required to fully understand the effect of using secondary calibrations on time estimates and downstream analyses.
 
@@
 Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses the same one to calibrate outside the clade, the resulting age estimates will often differ substantially, as the placement of calibrations as stem or crown group is proved to deeply affect estimated times of lineage divergence (Sauquet, 2013).
-Trying to summarize the resulting chronograms into a single one using simple summary statistics might erase all types of relevant information from the source chronograms.
-Accordingly, the prevailing view in our research community is that we should favor time of lineage divergence estimates obtained from a single analysis, using fossil data as primary sources of calibrations, and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history (Antonelli et al., 2017).
+Trying to summarize the resulting chronograms into a single one using simple summary statistics can erase many types of relevant information from the source chronograms.
+Accordingly, the prevailing view is that we should favor time of lineage divergence estimates obtained from a single analysis, using fossil data as primary sources of calibrations, and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history (Antonelli et al., 2017).
 However, the exercise of summarizing different chronograms has the potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypothesis.
@@
 
-Alternatively, one could try to choose the ``best'' chronogram from a set of possible evolutionary hypotheses.
-Several characteristics of the data used for dating analyses as well as from the output chronogram itself, could be used to score quality of source chronograms.
-Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
-To facilitate subsetting of source chronograms following different criteria by the users, this information should be included as metadata manually entered by curators in the future.
-
-In other areas of biological research, such as ecology and conservation biology, it has been shown that at least some data on lineage divergence represents a relevant improvement for testing alternative hypothesis using phylogenetic distance (Campbell O.
+Nonetheless, in ecology and conservation biology, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance (Campbell O.
 Webb et al., 2008).
-Hence, we integrated into datelife's workflow different ways of creating branch lengths in the absence of starting branch length information for taxa lacking this information (using the BLADJ option).
-``Making up'' branch lengths is an accepted practice in scientific publications: Jetz, Thomas, Joy, Hartmann, and Mooers (2012), created a time-calibrated tree of all 9,993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
+Hence, we integrated into datelife's workflow different ways of estimating node ages in the absence of calibrations and branch length information for taxa lacking this information.
+``Making up'' branch lengths is an accepted practice in scientific publications: Jetz et al.
+(2012), created a time-calibrated tree of all 9,993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
 (2018) created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37\% had molecular data; Stephen A.
 Smith and Brown (2018) constructed a tree of 353,185 seed plants where only 23\% had molecular data.
-Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch assignation methods.
-There has yet to be a thorough analysis of what can go wrong when one goes beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make sure results are substantially similar between trees fully reconstructed with molecular or other data, and trees that are reconstructed using taxonomy by resolving polytomies at random following a statistical model.
+Obviously, there are risks in this practice!
+Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch estimation methods.
+There has yet to be a thorough analysis of what can go wrong when one extends inferences beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make carefully consider the statistical assumptions being made, and assess the consistency of the results with prior work.
 
@@
 This provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identification of regions that require more research or that have conflicting information.
+It is available as an R package, or a web-based R shiny app at dates.opentreeloflife.org/datelife.
 Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
@@
 {Paleotree: An R package for paleontological and phylogenetic analyses of evolution}.
-\emph{Methods in Ecology and Evolution}, \emph{3}(5), 803--807.
+\emph{{Methods in Ecology and Evolution}}, \emph{3}(5), 803--807.
 \url{https://doi.org/10.1111/j.2041-210X.2012.00223.x}
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-barker2012going}{}}%
+\leavevmode\vadjust pre{\hypertarget{ref-barker2013going}{}}%
 Barker, F.
@@
 J.
-(2012).
+(2013).
 Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds.
-\emph{Systematic Biology}, \emph{62}(2), 298--320.
+\emph{{Systematic Biology}}, \emph{62}(2), 298--320.
 
@@
 New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
-\emph{The Auk: Ornithological Advances}, \emph{132}(2), 333--348.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Boyle2013}{}}%
-Boyle, B., Hopkins, N., Lu, Z., Raygoza Garay, J.
-A., Mozzherin, D., Rees, T., \ldots{} Enquist, B.
-J.
-(2013).
-{The taxonomic name resolution service: An online tool for automated standardization of plant names}.
-\emph{BMC Bioinformatics}, \emph{14}(1).
-\url{https://doi.org/10.1186/1471-2105-14-16}
+\emph{{The Auk: Ornithological Advances}}, \emph{132}(2), 333--348.
 
@@
 {Estimating Divergence Times in Large Phylogenetic Trees}.
-\emph{Systematic Biology}, \emph{56}(788777878), 741--752.
+\emph{{Systematic Biology}}, \emph{56}(788777878), 741--752.
 \url{https://doi.org/10.1080/10635150701613783}
@@
 Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds.
-\emph{Molecular Phylogenetics and Evolution}, \emph{75}, 41--77.
+\emph{{Molecular Phylogenetics and Evolution}}, \emph{75}, 41--77.
 
@@
 {taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}}.
-\emph{F1000Research}, \emph{2}(191), 1--29.
+\emph{{F1000Research}}, \emph{2}(191), 1--29.
 \url{https://doi.org/10.12688/f1000research.2-191.v2}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-claramunt2015new}{}}%
-Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{Science Advances}, \emph{1}(11), e1501005.
+Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{{Science Advances}}, \emph{1}(11), e1501005.
 
@@
 {SDM: A fast distance-based approach for (super)tree building in phylogenomics}.
-\emph{Systematic Biology}, \emph{55}(5), 740--755.
+\emph{{Systematic Biology}}, \emph{55}(5), 740--755.
 \url{https://doi.org/10.1080/10635150600969872}
@@
 {Congruification: Support for time scaling large phylogenetic trees}.
-\emph{Methods in Ecology and Evolution}, \emph{4}(7), 688--691.
+\emph{{Methods in Ecology and Evolution}}, \emph{4}(7), 688--691.
 \url{https://doi.org/10.1111/2041-210X.12051}
@@
 New zealand passerines help clarify the diversification of major songbird lineages during the oligocene.
-\emph{Genome Biology and Evolution}, \emph{7}(11), 2983--2995.
+\emph{{Genome Biology and Evolution}}, \emph{7}(11), 2983--2995.
 
@@
 {Tree of life reveals clock-like speciation and diversification}.
-\emph{Molecular Biology and Evolution}, \emph{32}(4), 835--845.
+\emph{{Molecular Biology and Evolution}}, \emph{32}(4), 835--845.
 \url{https://doi.org/10.1093/molbev/msv037}
@@
 {rotl: an R package to interact with the Open Tree of Life data}.
-\emph{Methods in Ecology and Evolution}, \emph{7}(12), 1476--1481.
+\emph{{Methods in Ecology and Evolution}}, \emph{7}(12), 1476--1481.
 \url{https://doi.org/10.1111/2041-210X.12593}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Morlon2014}{}}%
-Morlon, H. (2014). {Phylogenetic approaches for studying diversification.} \emph{Ecology Letters}, \emph{17}(4), 508--525. \url{https://doi.org/10.1111/ele.12251}
+Morlon, H. (2014). {Phylogenetic approaches for studying diversification.} \emph{{Ecology Letters}}, \emph{17}(4), 508--525. \url{https://doi.org/10.1111/ele.12251}
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-posadas2006historical}{}}%
-Posadas, P., Crisci, J. V., \& Katinas, L. (2006). Historical biogeography: A review of its basic concepts and critical issues. \emph{Journal of Arid Environments}, \emph{66}(3), 389--403.
+Posadas, P., Crisci, J. V., \& Katinas, L. (2006). Historical biogeography: A review of its basic concepts and critical issues. \emph{{Journal of Arid Environments}}, \emph{66}(3), 389--403.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Revell2012}{}}%
-Revell, L. J. (2012). Phytools: An r package for phylogenetic comparative biology (and other things). \emph{Methods in Ecology and Evolution}, \emph{3}, 217--223.
+Revell, L. J. (2012). Phytools: An r package for phylogenetic comparative biology (and other things). \emph{{Methods in Ecology and Evolution}}, \emph{3}, 217--223.
 
@@
 Making tree-of-life knowledge accessible, reusable and convenient}.
-\emph{BMC Bioinformatics}, \emph{14}.
+\emph{{BMC Bioinformatics}}, \emph{14}.
 \url{https://doi.org/10.1186/1471-2105-14-158}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Webb2000}{}}%
-Webb, Campbell O. (2000). {Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees}. \emph{The American Naturalist}, \emph{156}(2), 145--155.
+Webb, C. (2000). {Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees}. \emph{{The American Naturalist}}, \emph{156}(2), 145--155.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-webb2005phylomatic}{}}%
-Webb, Campbell O., \& Donoghue, M. J. (2005). Phylomatic: Tree assembly for applied phylogenetics. \emph{Molecular Ecology Notes}, \emph{5}(1), 181--183.
+Webb, Campbell O., \& Donoghue, M. J. (2005). Phylomatic: Tree assembly for applied phylogenetics. \emph{{Molecular Ecology Notes}}, \emph{5}(1), 181--183.
 
@@
 % in this file I leave the figure captions outside\ caption{} because I want them
-% to be formatted in the same way as the general text (double spaced and linenumbered)
+to be formatted in the same way as the general text (double spaced and linenumbered)
 \captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}
 
+\newpage
 \begin{figure}[!h]
 \includegraphics{../figures/figure1/figure1-vertical-final.pdf}
-\caption{Stylized DateLife algorithm showing the main steps that are performed with the R package \texttt{datelife}, or online at \url{www.datelife.org/query/}.
-Details on the functions involved on each workflow are available at \texttt{datelife}'s R package manual https://phylotastic.org/datelife/articles/.}
+\caption{Stylized DateLife workflow.
+This shows the general worflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{http://www.datelife.org/}.
+Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
+}
 \label{fig:workflow}
@@
 % \end{center}
-% Stylized DateLife workflow.
-This shows the general workflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{http://www.datelife.org/}.
-Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
 \newpage
@@
 \includegraphics[width=1\linewidth]{../figures/fig_runtime_main.pdf}
-\caption{Input taxon name processing and chronogram database search computation time increases with number of input taxon names.
-We sampled N bird species names for each input size class, 100 times, and then performed a \texttt{datelife} search using the Taxon Names Resoultion Service (TNRS; light gray), and without using TNRS (dark gray).
+\caption{Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names.
+We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray).
 We also performed a search using the already processed query for comparison (light blue).}
@@
 % \begin{center}
-% \textsc{Figure \ref{fig:runtime1}}
+% \textsc{Figure \ref{fig:benchmark}}
 % \end{center}
-%
 \newpage
 \begin{figure}[!h]
-\includegraphics[width=1\linewidth]{../figures/fig_schronograms1.pdf}
-\caption{Lineage through time (LTT) plots of source chronograms containing all or a subset of species from the bird family Fringillidae of true finches.
-Arrows indicate maximum age of each chronogram.
-Numbers reference to chronograms' original publications 1: Barker et al.
-(\protect\hyperlink{ref-barker2012going}{2012}), 2: Barker et al.
-(\protect\hyperlink{ref-barker2015new}{2015}), 3: Burns et al.
-(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), 4: Claramunt and Cracraft (\protect\hyperlink{ref-claramunt2015new}{2015}), 5: Gibb et al.
-(\protect\hyperlink{ref-gibb2015new}{2015}), 6: Hedges et al.
-(\protect\hyperlink{ref-Hedges2015}{2015}), 7: Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), 8: Jetz et al.
-(\protect\hyperlink{ref-Jetz2012}{2012}), 9: Price et al.
-(\protect\hyperlink{ref-price2014niche}{2014}).
-}
-\label{fig:schronograms1}
+\includegraphics[width=1\linewidth]{../figures/figure2/figure2-1.pdf}
+\caption{Age data results of a DateLife search of a small sample of 6 bird species within the Passeriformes.
+Input names were found across 9 chronograms within 6 independent studies (Barker et al.
+(\protect\hyperlink{ref-barker2012going}{2012}), Barker et al.
+(\protect\hyperlink{ref-barker2015new}{2015}), Burns et al.
+(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al.
+(\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al.
+(\protect\hyperlink{ref-Jetz2012}{2012}).) This revealed 28 age data points for the queried species names.}
+\label{fig:figure2-1}
 \end{figure}
 % \begin{center}
-% \textsc{Figure \ref{fig:schronograms1}}
+% \textsc{Figure \ref{fig:figure2-1}}
 % \end{center}
-%Lineage through time (LTT) plots of source chronograms containing all or a subset of species from the bird family Fringillidae of true finches.
-Arrows indicate maximum age of each chronogram.
-Numbers reference to chronograms' original publications 1: Barker et al.
-(\protect\hyperlink{ref-barker2012going}{2012}), 2: Barker et al.
-(\protect\hyperlink{ref-barker2015new}{2015}), 3: Burns et al.
-(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), 4: Claramunt and Cracraft (\protect\hyperlink{ref-claramunt2015new}{2015}), 5: Gibb et al.
-(\protect\hyperlink{ref-gibb2015new}{2015}), 6: Hedges et al.
-(\protect\hyperlink{ref-Hedges2015}{2015}), 7: Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), 8: Jetz et al.
-(\protect\hyperlink{ref-Jetz2012}{2012}), 9: Price et al.
-(\protect\hyperlink{ref-price2014niche}{2014}).
 \newpage
 \begin{figure}[!h]
-\includegraphics{../figures/fig_summaries.pdf}
-\caption{LTT plots of median (top) and Supermatrix Distance Method (SDM; bottom) chronograms summarising information from source chronograms found for the Fringillidae.
-Arrows indicate tree maximum age.}
+\includegraphics{../figures/figure2/figure2-2.pdf}
+\caption{}
 \label{fig:summaries}
@@
 % \begin{center}
-% \textsc{Figure \ref{fig:summaries}}
+% \textsc{Figure \ref{fig:figure2-2}}
 % \end{center}
-% LTT plots of median (top) and Supermatrix Distance Method (SDM; bottom) chronograms summarising information from source chronograms found for the Fringillidae. Arrows indicate tree maximum age.
+LTT plots of median (top) and Supermatrix Distance Method (SDM; bottom) chronograms summarising information from source chronograms found for the Fringillidae. Arrows indicate tree maximum age.
 \newpage
@@
 \includegraphics{../figures/fig_crossval_bladj.pdf}
-\caption{LTT plots showing results from the cross-validation analyses of trees without branch lengths dated using BLADJ.
-The dating analysis can only be performed in trees with more than 2 tips, thus excluding chronogram from study 4; its data was still used as calibration for the other source chronograms.}
+\caption{}
 \label{fig:cvbladj}
@@
 % \end{center}
-% LTT plots showing results from the cross-validation analyses of trees without branch lengths dated using BLADJ.
+LTT plots showing results from the cross-validation analyses of trees without branch lengths dated using BLADJ.
 The dating analysis can only be performed in trees with more than 2 tips, thus excluding chronogram from study 4; its data was still used as calibration for the other source chronograms.
@@
 \includegraphics{../figures/fig_crossval_boldsumm.pdf}
-\caption{LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8.
-We could construct a tree with branch lengths for all source chronograms.
-However, dating with PATHd8 was only successful in three source chronograms shown here.}
+\caption{}
 \label{fig:cvbold}
@@
 % \end{center}
-FIGURES
-%LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8.
+LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8.
 We could construct a tree with branch lengths for all source chronograms.
@@
 
-%\end{linenumbers}
+% \end{linenumbers}
 

### 2022-03-28 manuscript/figures-alt.tex
adding data to median table of figure 2
tree figure for fringillidae example

@@
 \includegraphics{../figures/figure2/figure2-2.pdf}
-\caption{}
+\caption{Summarized age data is used as secondary calibrations to date a tree topology as a summary chronogram.}
 \label{fig:summaries}
@@
 % \end{center}
-LTT plots of median (top) and Supermatrix Distance Method (SDM; bottom) chronograms summarising information from source chronograms found for the Fringillidae. Arrows indicate tree maximum age.
 \newpage
 \begin{figure}[!h]
-\includegraphics{../figures/fig_crossval_bladj.pdf}
+\includegraphics{../figures/figure-fringillidae/median_and_calibration_ages_simple.pdf}
 \caption{}
@@
 % \begin{center}
-% \textsc{Figure \ref{fig:cvbladj}}
+% \textsc{Figure \ref{fig:fringillidages}}
 % \end{center}
-LTT plots showing results from the cross-validation analyses of trees without branch lengths dated using BLADJ.
-The dating analysis can only be performed in trees with more than 2 tips, thus excluding chronogram from study 4; its data was still used as calibration for the other source chronograms.
 \newpage

### 2022-03-28 manuscript/manuscript.Rmd
tree figure for fringillidae example

@@
 The algorithm proceeds to prune the trees to keep matching species names on tips only, and transform the pruned trees to pairwise distance matrices.
- There are `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies (Fig. \ref{fig:schronograms1}).
-<!-- In case we want to cite the studies in here, but they are already cited in the figure caption: [Fig.
-\ref{fig:schronograms1}; @barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
--->
+ There are `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
 The final step is to summarize the available information using two alternative types of summary chronograms, median and SDM.
 As explained in the "Description" section, data from source chronograms is first summarised into a single distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig.
-\ref{fig:summaries}).
+\ref{fig:fringillidages}).
 Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
-?? Say some things about the results!
+????????????????? Say some things about the results!
 <!-- (Going from a summary distance matrix Using this method is better than using clustering algorithms  -->

### 2022-03-28 manuscript/manuscript.tex
adding data to median table of figure 2
tree figure for fringillidae example

@@
 \shorttitle{DateLife: revealing the dated Tree of Life}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4199}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4204}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 The algorithm proceeds to prune the trees to keep matching species names on tips only, and transform the pruned trees to pairwise distance matrices.
-There are 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Fig. \ref{fig:schronograms1}).
+There are 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Price et al., 2014).
 The final step is to summarize the available information using two alternative types of summary chronograms, median and SDM.
 As explained in the ``Description'' section, data from source chronograms is first summarised into a single distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig.
-\ref{fig:summaries}).
+\ref{fig:fringillidages}).
 Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
-?? Say some things about the results!
+????????????????? Say some things about the results!
 
@@
 
-At the time of writing of this manuscript (Mar 24, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
+At the time of writing of this manuscript (Mar 28, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
 A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time.

### 2022-03-29 manuscript/manuscript.tex
larger fringillidae chronogram figure

@@
 
-At the time of writing of this manuscript (Mar 28, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
+At the time of writing of this manuscript (Mar 29, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
 A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time.
@@
 \begin{figure}[!h]
-\includegraphics{../figures/figure-fringillidae/median_and_calibration_ages_simple.pdf}
-\caption{}
-\label{fig:cvbladj}
+\includegraphics{../figures/figure-fringillidae/median_and_calibration_ages_simple.png}
+\caption{Fringillidae median summary chronogram genertaed with DateLife. It has 256 tips and 233 nodes.}
+\label{fig:fringillidages}
 \end{figure}

### 2022-04-04 manuscript/figures-alt.tex
larger fringillidae chronogram figure
cross validation figures

@@
 % in this file I leave the figure captions outside\ caption{} because I want them
-to be formatted in the same way as the general text (double spaced and linenumbered)
+% to be formatted in the same way as the general text (double spaced and linenumbered)
 \captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}
 
-\newpage
 \begin{figure}[!h]
@@
 % \end{center}
-\newpage
+
 \begin{figure}[!h]
@@
 We also performed a search using the already processed query for comparison (light blue).}
-\label{fig:runtime1}
+\label{fig:runtime_main}
 \end{figure}
@@
 % \end{center}
-\newpage
+
 \begin{figure}[!h]
-\includegraphics{../figures/figure-fringillidae/median_and_calibration_ages_simple.pdf}
-\caption{}
-\label{fig:cvbladj}
+\includegraphics{../figures/figure-fringillidae/median_and_calibration_ages_simple.png}
+\caption{Fringillidae median summary chronogram genertaed with DateLife. It has 256 tips and 233 nodes.}
+\label{fig:fringillidages}
 \end{figure}
@@
 % \end{center}
-\newpage
+
 \begin{figure}[!h]
-\includegraphics{../figures/fig_crossval_boldsumm.pdf}
-\caption{}
-\label{fig:cvbold}
+\includegraphics{../figures/figure-cross-validation/cross_validation_1.png}
+\caption{Cross validation of first source chronogram.
+The chronogram shown in black corresponds to the dates published in the original study.
+The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
+\label{fig:cv1}
 \end{figure}
@@
 % \end{center}
-LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8.
-We could construct a tree with branch lengths for all source chronograms.
-However, dating with PATHd8 was only successful in three source chronograms shown here.
+
+\begin{figure}[!h]
+\includegraphics{../figures/figure-cross-validation/cross_validation_2.png}
+\caption{Cross validation of second source chronogram.
+The chronogram shown in black corresponds to the dates published in the original study.
+The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
+\label{fig:cv2}
+\end{figure}
+
+\begin{figure}[!h]
+\includegraphics{../figures/figure-cross-validation/cross_validation_3.png}
+\caption{Cross validation of third source chronogram.
+The chronogram shown in black corresponds to the dates published in the original study.
+The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
+\label{fig:cv3}
+\end{figure}
+
+\begin{figure}[!h]
+\includegraphics{../figures/figure-cross-validation/cross_validation_4.png}
+\caption{Cross validation of fourth source chronogram.
+The chronogram shown in black corresponds to the dates published in the original study.
+The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
+\label{fig:cv4}
+\end{figure}
+
+\begin{figure}[!h]
+\includegraphics{../figures/figure-cross-validation/cross_validation_6.png}
+\caption{Cross validation of sixth source chronogram.
+The chronogram shown in black corresponds to the dates published in the original study.
+The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
+\label{fig:cv6}
+\end{figure}
+
+\begin{figure}[!h]
+\includegraphics{../figures/figure-cross-validation/cross_validation_7.png}
+\caption{Cross validation of seventh source chronogram.
+The chronogram shown in black corresponds to the dates published in the original study.
+The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
+\label{fig:cv7}
+\end{figure}
+
+\begin{figure}[!h]
+\includegraphics{../figures/figure-cross-validation/cross_validation_8.png}
+\caption{Cross validation of eight source chronogram.
+The chronogram shown in black corresponds to the dates published in the original study.
+The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
+\label{fig:cv8}
+\end{figure}
+
+\begin{figure}[!h]
+\includegraphics{../figures/figure-cross-validation/cross_validation_9.png}
+\caption{Cross validation of ninth source chronogram.
+The chronogram shown in black corresponds to the dates published in the original study.
+The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
+\label{fig:cv9}
+\end{figure}
+
+\begin{figure}[!h]
+\includegraphics{../figures/figure-cross-validation/cross_validation_10.png}
+\caption{Cross validation of tenth source chronogram.
+The chronogram shown in black corresponds to the dates published in the original study.
+The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
+\label{fig:cv10}
+\end{figure}
 

### 2022-04-04 manuscript/manuscript.Rmd
cross validation figures

@@
 Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig.
-\ref{fig:benchmark}).
+\ref{fig:runtime_main}).
 <!--Speed of other datelife workflows? NO, bc it reflects speed from other software mainly. But I will put a sentence about it...
@@
 Data from source chronograms can be used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates (Figs.
-\ref{fig:cvbladj} and \ref{fig:cvbold}).
+\ref{fig:cv1} to \ref{fig:cv10}).
 As a form of cross validation, we took tree topologies from each input study and calibrated them using time of lineage divergence data from all other source chronograms.
@@
 In the absence of branch lengths, the ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvbladj}).
-Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cvbladj}).
+Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cv2}).
 <!-- We need to think about the root! is it comparable across source chronograms?? not all of them have the same sampling... so probably not the same root and thus not comparable. Address this -->
-We also demonstrate the usage of PATHd8 [@Britton2007] as an alternative method to BLADJ.
+We also demonstrate the usage of PATHd8 [@Britton2007] as an dating method alternative to BLADJ.
 For this, we run a `datelife` branch length reconstruction that searches for DNA sequence data from the Barcode of Life Data System [BOLD; @ratnasingham2007bold] to generate branch lengths.
@@
 From these, two trees have a different sampling than the original source chronogram, mainly because DNA BOLD  data for some species is absent from the database.
-??? Node ages or maximum ages?? Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
+<!-- ??? Node ages or maximum ages?? -->
+Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
 More examples and code used to generate these trees were developed on an open repository that is available for consultation and reuse at <https://github.com/LunaSare/datelife_examples>.

### 2022-04-04 manuscript/manuscript.tex
cross validation figures

@@
 \shorttitle{DateLife: revealing the dated Tree of Life}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4204}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4201}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig.
-\ref{fig:benchmark}).
+\ref{fig:runtime_main}).
 \texttt{datelife}'s code performance was evaluated with a set of unit tests designed and
@@
 In the absence of branch lengths, the ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvbladj}).
-Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cvbladj}).
-We also demonstrate the usage of PATHd8 (Britton et al., 2007) as an alternative method to BLADJ.
+Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cv2}).
+We also demonstrate the usage of PATHd8 (Britton et al., 2007) as an dating method alternative to BLADJ.
 For this, we run a \texttt{datelife} branch length reconstruction that searches for DNA sequence data from the Barcode of Life Data System {[}BOLD; Ratnasingham and Hebert (2007){]} to generate branch lengths.
@@
 From these, two trees have a different sampling than the original source chronogram, mainly because DNA BOLD data for some species is absent from the database.
-??? Node ages or maximum ages?? Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
+Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
 More examples and code used to generate these trees were developed on an open repository that is available for consultation and reuse at \url{https://github.com/LunaSare/datelife_examples}.
@@
 
-At the time of writing of this manuscript (Mar 29, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
+At the time of writing of this manuscript (Apr 04, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
 A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time.
@@
 % \end{center}
-\newpage
-\begin{figure}[!h]
-\includegraphics{../figures/fig_crossval_boldsumm.pdf}
-\caption{}
-\label{fig:cvbold}
+
+\begin{figure}[!h]
+\includegraphics{../figures/figure-cross-validation/cross_validation_1.png}
+\caption{Cross validation of first source chronogram.
+The chronogram shown in black corresponds to the dates published in the original study.
+The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
+\label{fig:cv1}
 \end{figure}

### 2022-04-06 manuscript/figures-alt.tex

--- 2022-03-28 manuscript/figures-alt.tex
@@
 % in this file I leave the figure captions outside\ caption{} because I want them
-to be formatted in the same way as the general text (double spaced and linenumbered)
+% to be formatted in the same way as the general text (double spaced and linenumbered)
 \captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}
 
-\newpage
 \begin{figure}[!h]
-\includegraphics{../figures/figure1/figure1-vertical-final.pdf}
+\includegraphics{../figures/figure1/figure1-new.pdf}
 \caption{Stylized DateLife workflow.
@@
 % \end{center}
-\newpage
-\begin{figure}[!h]
-\includegraphics{../figures/fig_crossval_boldsumm.pdf}
-\caption{}
-\label{fig:cvbold}
-\end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:cvbold}}
-% \end{center}
-LTT plots showing results from the cross-validation analyses of trees with branch length reconstructed with data from the Barcode of Life Database (BOLD) dated using PATHd8.
-We could construct a tree with branch lengths for all source chronograms.
-However, dating with PATHd8 was only successful in three source chronograms shown here.
 

### 2022-04-06 manuscript/figures-supplementary.tex

--- 2022-04-04 manuscript/manuscript.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-  english,
-  man]{apa6}
-\title{DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life}
-\author{Luna L. Sánchez Reyes\textsuperscript{1,2}, Emily Jane McTavish\textsuperscript{1}, \& Brian O'Meara\textsuperscript{2}}
-\date{}
-
-\usepackage{amsmath,amssymb}
-\usepackage{lmodern}
-\usepackage{iftex}
-\ifPDFTeX
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life},
-  pdfauthor={Luna L. Sánchez Reyes1,2, Emily Jane McTavish1, \& Brian O'Meara2},
-  pdflang={en-EN},
-  pdfkeywords={Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-\DeclareUnicodeCharacter{0301}{*************************************}
-% Manuscript styling
-\usepackage{upgreek}
-\captionsetup{font=singlespacing,justification=justified}
-
-% Table formatting
-\usepackage{longtable}
-\usepackage{lscape}
-% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
-\usepackage{multirow}		% Table styling
-\usepackage{tabularx}		% Control Column width
-\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
-\usepackage{threeparttablex}            % Lets threeparttable work with longtable
-
-% Create new environments so endfloat can handle them
-% \newenvironment{ltable}
-%   {\begin{landscape}\centering\begin{threeparttable}}
-%   {\end{threeparttable}\end{landscape}}
-\newenvironment{lltable}{\begin{landscape}\centering\begin{ThreePartTable}}{\end{ThreePartTable}\end{landscape}}
-
-% Enables adjusting longtable caption width to table width
-% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
-\makeatletter
-\newcommand\LastLTentrywidth{1em}
-\newlength\longtablewidth
-\setlength{\longtablewidth}{1in}
-\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}
-
-% \setlength{\parindent}{0.5in}
-% \setlength{\parskip}{0pt plus 0pt minus 0pt}
-
-% \usepackage{etoolbox}
-\makeatletter
-\patchcmd{\HyOrg@maketitle}
-  {\section{\normalfont\normalsize\abstractname}}
-  {\section*{\normalfont\normalsize\abstractname}}
-  {}{\typeout{Failed to patch abstract.}}
-\patchcmd{\HyOrg@maketitle}
-  {\section{\protect\normalfont{\@title}}}
-  {\section*{\protect\normalfont{\@title}}}
-  {}{\typeout{Failed to patch title.}}
-\makeatother
-\shorttitle{DateLife: revealing the dated Tree of Life}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4201}
-\DeclareDelayedFloatFlavor{ThreePartTable}{table}
-\DeclareDelayedFloatFlavor{lltable}{table}
-\DeclareDelayedFloatFlavor*{longtable}{table}
-\makeatletter
-\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
-\makeatother
-\usepackage{lineno}
-
-\linenumbers
-\usepackage{csquotes}
-\ifXeTeX
-  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
-  \usepackage{polyglossia}
-  \setmainlanguage[]{english}
-\else
-  \usepackage[main=english]{babel}
-% get rid of language-specific shorthands (see #6817):
-\let\LanguageShortHands\languageshorthands
-\def\languageshorthands#1{}
-\fi
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-
-\authornote{
-
-School of Natural Sciences, University of California, Merced, Science and Engineering Building 1.
-
-Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 425 Hesler Biology Building, Knoxville, TN 37996, USA.
-
-The authors made the following contributions.
-Luna L.
-Sánchez Reyes: Data curation, Investigation, Software, Visualization, Validation, Writing - Original Draft Preparation, Writing - Review \& Editing; Emily Jane McTavish: Resources, Software, Writing - Review \& Editing; Brian O'Meara: Conceptualization, Funding acquisition, Methodology, Resources, Software, Supervision, Writing - Review \& Editing.
-
-Correspondence concerning this article should be addressed to Luna L. Sánchez Reyes, . E-mail: \href{mailto:sanchez.reyes.luna@gmail.com}{\nolinkurl{sanchez.reyes.luna@gmail.com}}
-
-}
-
-\affiliation{\vspace{0.5cm}\textsuperscript{1} University of California, Merced\\\textsuperscript{2} University of Tennessee, Knoxville}
-
-\abstract{
-Date estimates for times of evolutionary divergences are key data for research in the natural sciences.
-These estimates also provide valuable information for for education, science communication and policy decisions.
-Although achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is a difficult and time-consuming task, the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
-When these estimates are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on time of evolutionary origin is exposed in a programatic and reusable way.
-Intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
-Here we present DateLife, a service implemented as an R package and an Rshiny website application available at www.datelife.org/query/, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
-The main DateLife workflow constructs a chronogram for any given combination of taxon names, by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from TreeBASE database as well.
-We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
-DateLife will be useful to increase awereness on the existing variation in expert time of divergence data, and can foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.
-}
-
-\begin{document}
-\maketitle
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-Chronograms --phylogenies with branch lengths proportional to time-- provide key data for the study of natural processes in many areas of biological research, such as developmental biology (Delsuc et al., 2018; Laubichler \& Maienschein, 2009), conservation biology (Felsenstein, 1985; C.
-Webb, 2000), historical biogeography (Posadas, Crisci, \& Katinas, 2006), and species diversification (Magallon \& Sanderson, 2001; Morlon, 2014).
-
-Building a chronogram is not an easy task.
-It requires obtaining and curating data to construct a phylogeny; selecting and placing appropriate calibrations on the phylogeny using independent age data points from the fossil record or other dated events, and inferring the full dated tree.
-Estimating accurate chronograms generally requires specialized biological training, taxonomic domain knowledge, and a non-negligible amount of research time, computational resources and funding.
-
-Here we present the DateLife software application, available as an R package and as an online Rshiny interactive website at www.datelife.org/query/, which captures data from published chronograms, and make these data readily accessible to users.
-DateLife features a versioned, open and fully public chronogram database (McTavish et al., 2015) storing age information in a computer readable format (Vos et al., 2012), an automated and programmatic way of accessing the data (Stoltzfus et al., 2013) and methods to summarize and compare age data.
-
-\hypertarget{description}{%
-\section{Description}\label{description}}
-
-The DateLife algorithm is fully implemented using the R language.
-The latest stable version of the R package \texttt{datelife} is available from the CRAN repository (v0.6.2; Sanchez-Reyes et al.
-(2022)), and relies on functionalities from various biological R packages:
-ape (Paradis, Claude, \& Strimmer, 2004),
-bold (Chamberlain et al., 2019),
-geiger (Harmon, Weir, Brock, Glor, \& Challenger, 2008),
-paleotree (Bapst, 2012),
-phyloch (Heibl, 2008),
-phylocomr (Ooms \& Chamberlain, 2018),
-phytools (Revell, 2012),
-rotl (Michonneau, Brown, \& Winter, 2016), and
-taxize (Chamberlain \& Szöcs, 2013; Chamberlain et al., 2019).
-Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife algorithm: providing an input, searching a chronogram database, and summarizing results from the search.
-
-\hypertarget{providing-an-input}{%
-\subsection{Providing an input}\label{providing-an-input}}
-
-DateLife starts with an input query consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree.
-If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis et al., 2004).
-The input tree is not required to have branch lengths, and its topology is used in the summary steps described below.
-
-DateLife accepts scientific names as input.
-These names can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or binomial specific.
-Subspecies and variants are ignored.
-If an input taxon name belongs to an inclusive taxonomic group the algorithm has two alternative behaviors defined by the ``get species from taxon'' flag.
-If the flag is active, the DateLife algorithm retrieves all species names within the inclusive taxonomic group and adds them to the input.
-If the flag is inactive, DateLife ignores the inclusive taxon names from the input.
-
-Input scientific names are processed using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database.
-TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard.
-DateLife implements TNRS using OpenTree's taxonomy as standard (Open Tree Of Life et al., 2016; Rees \& Cranston, 2017).
-
-The processed input taxon names are saved as an R object of a newly defined class \texttt{datelifeQuery} that is used in the following steps.
-This object contains the processed names, the corresponding OpenTree taxonomic id numbers, and the topology of the input tree if any was provided.
-
-\hypertarget{searching-the-database}{%
-\subsection{Searching the database}\label{searching-the-database}}
-
-A DateLife search consists of matching processed taxon names to tip labels in a chronogram database.
-Chronograms with at least two matching tip labels are identified and pruned down to preserve only the matched tips.
-
-Matching pruned chronograms are stored as individual patristic distance matrices (Figure \ref{fig:workflow} subfigure X).
-This matrix consists of \ldots{} ???? the pairwise distance between pairs of query taxa which are in that input tree, in units of millions of years.
-
-This format speeds up extraction of pairwise taxon ages of the queried taxa, as opposed to searching the ancestor node of a pair of taxa in a ``phylo'' object or newick string.
-The patristic matrices are also associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.
-
-DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies.
-It is constructed from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert phylogenetic knowledge with rich metadata (McTavish et al., 2015) that allows automatic and reproducible construction of a chronogram database.
-New chronograms can be added to Phylesystem by any user and are immediately publicly available, and the DateLife database can be updated to include those new data within a run.
-
-\hypertarget{summarizing-search-results}{%
-\subsection{Summarizing search results}\label{summarizing-search-results}}
-
-At this point, summary information is extracted from the \texttt{datelifeResult} object to inform decisions for the subsequent steps in the user workflow.
-Age data from the matching pruned chronograms is summarized and used to generate a single summary chronogram.
-Other basic summary information available to the user is:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  The matching pruned chronograms as newick strings or ``phylo'' objects.
-\item
-  The ages of the root of all matching pruned chronograms.
-This can correspond to the age of the most recent common ancestor (mrca) of your group of interest if the pruned chronograms have all taxa belonging to the group.
-If not, the root corresponds to the mrca of a subgroup withing your group of interest.
-\item
-  Study citations where original chronograms were published.
-\item
-  A report of input taxon names matches across pruned chronograms.
-\item
-  The single matching pruned chronogram with the most input taxon names.
-\end{enumerate}
-
-\emph{\textbf{Identifying groves.--}} To generate a single summary chronogram, the DateLife algorithm starts by identifying the matching pruned chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
-(2009).
-In rare cases, a group of trees can have multiple groves.
-By default, DateLife chooses the grove with the most taxa, however, the ``criterion = trees'' flag allows the user to choose the grove with the most trees instead.
-
-\emph{\textbf{Choosing a topology.--}} DateLife requires a tree topology to summarize age data upon.
-Users can provide one as input from the literature, or one of their own making.
-If no topology is provided, DateLife automatically subsets one from the OpenTree synthetic tree (Open Tree Of Life et al., 2019).
-
-DateLife can also reconstruct branch lengths proportional to substitution rates on a fixed tree topology using available genetic data from BOLD.
-
-\emph{\textbf{Congruifying nodes.--}} DateLife then implements the congruification method (Eastman, Harmon, \& Tank, 2013) to find nodes belonging to the same clade across matching pruned chronograms.
-Congruified node ages stored as a \texttt{congruifiedCalibrations} object are then matched to nodes in the chosen tree topology and stored as a \texttt{matchedCalibrations} object.
-
-\emph{\textbf{Summarizing node ages.--}} DateLife summarizes matched calibrations
-into a single patristic distance matrix using different methods.
-Summarizing options implemented include Super Distance Matrix method (SDM, Criscuolo, Berry, Douzery, \& Gascuel, 2006) and summary statistics such as median, minimum and maximum ages.
-
-\emph{\textbf{Dating the tree topology.--}} Summarized calibrations can be applied as secondary calibrations with different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Campbell O.
-Webb, Ackerly, \& Kembel, 2008; Campbell O.
-Webb \& Donoghue, 2005), and treePL (Stephen A.
-Smith \& O'Meara, 2012).
-
-By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm that assigns ages to nodes with no data evenly between nodes with age data, which minimizes age variance in the resulting chronogram (Campbell O.
-Webb et al., 2008).
-When there is conflict in ages across node with age data, the algorithm ignores ages that are older than parent nodes and/or younger than descendant nodes.
-
-If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature.
-If none is provided, DateLife assigns an arbitrary age to the root as 10\% older than the oldest age available within the group.
-
-\emph{\textbf{Visualizing results.--}}
-Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R ``phylo'' format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions.
-
-\hypertarget{benchmark}{%
-\section{Benchmark}\label{benchmark}}
-
-\texttt{datelife}'s code speed was tested on an Apple iMac
-with one 3.4 GHz Intel Core i5 processor.
-We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
-Query processing time increases roughly linearly with number of input taxon names, and
-increases considerably if TNRS is activated.
-Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
-Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig.
-\ref{fig:runtime_main}).
-\texttt{datelife}'s code performance was evaluated with a set of unit tests designed and
-implemented with the R package testthat (R Core Team, 2018) that were run both locally
-with the devtools package (R Core Team, 2018), and on a public server --via
-GitHub, using the continuous integration tool Travis CI (\url{https://travis-ci.org}). At
-present, unit tests cover more than 40\% of \texttt{datelife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).
-
-\hypertarget{case-study}{%
-\section{Case study}\label{case-study}}
-
-We illustrate the DateLife algorithm using a group within the Passeriform birds encompassing the family of true finches, Fringillidae and allies as case study.
-The first example analyses 6 bird species and shows all steps of the algorithm.
-The second example is a real life application
-
-\hypertarget{small-example}{%
-\subsection{Small example}\label{small-example}}
-
-We chose 6 bird species associated to true finches at random. The sample includes
-two species of cardinals:
-the black-thighed grosbeak -- \emph{Pheucticus tibialis} and
-the crimson-collared grosbeak -- \emph{Rhodothraupis celaeno};
-three species of buntings:
-the yellowhammer -- \emph{Emberiza citrinella},
-the pine bunting -- \emph{Emberiza leucocephalos} and
-the yellow-throated bunting -- \emph{Emberiza elegans};
-and one species of tanager, the vegetarian finch -- \emph{Platyspiza crassirostris}.
-
-Processing input names found that \emph{Emberiza elegans} is synonym for \emph{Schoeniclus elegans} in the reference taxonomy.
-DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies.
-Three studies matched five input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2015; Hedges, Marin, Suleski, Paymer, \& Kumar, 2015; Jetz, Thomas, Joy, Hartmann, \& Mooers, 2012), one study matched four input names (Hooper \& Price, 2017) and two studies matched two input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2013; Burns et al., 2014).
-No studies matched all input names.
-Together, matching chronograms have 28 unique age data points.
-All nodes have age data.
-As fixed tree topology, DateLife used OpenTree's synthetic tree as default and mapped age data to nodes in the tree.
-As expected, more inclusive nodes (e.g., node ``n1'') have more age data than less inclusive nodes (e.g., node ``n5'').
-The processing step allowd discovering five data points for node ``n4'' that would not have had any data otherwise.
-Age summary statistics per node were calculated and tested as secondary calibrations to date the tree topology using the BLADJ algorithm.
-Age data for node ``n2'' was excluded as final calibration because it is older than age data of a more inclusive node.
-
-\hypertarget{real-life-application}{%
-\subsection{Real life application}\label{real-life-application}}
-
-A college educator wishes to obtain state-of-the-art data on time of evolutionary origin of species belonging to the true finches for their class.
-They decide to use \texttt{datelife} because they are teaching best practices for reproducibility.
-Students have the option to go to the website at www.datelife.org and perform an interactive run. However, the educator also wants the students to practice their R skills.
-The first step is to run a \texttt{datelife} query using the ``get species from taxon'' flag.
-This will get all recognised species names within their chosen inclusive taxon.
-The Fringillidae has 289 species, according to the Open Tree of Life taxonomy.
-Once with a curated set of species taxon names, the next step is to run a \texttt{datelife} search that will find all chronograms that contain at least two species names.
-The algorithm proceeds to prune the trees to keep matching species names on tips only, and transform the pruned trees to pairwise distance matrices.
-There are 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Price et al., 2014).
-The final step is to summarize the available information using two alternative types of summary chronograms, median and SDM.
-As explained in the ``Description'' section, data from source chronograms is first summarised into a single distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig.
-\ref{fig:fringillidages}).
-Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
-????????????????? Say some things about the results!
-
-\hypertarget{cross-validation-test}{%
-\section{Cross-validation test}\label{cross-validation-test}}
-
-Data from source chronograms can be used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates (Figs.
-\ref{fig:cv1} to \ref{fig:cv10}).
-As a form of cross validation, we took tree topologies from each input study and calibrated them using time of lineage divergence data from all other source chronograms.
-
-In the absence of branch lengths, the ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvbladj}).
-Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cv2}).
-We also demonstrate the usage of PATHd8 (Britton et al., 2007) as an dating method alternative to BLADJ.
-For this, we run a \texttt{datelife} branch length reconstruction that searches for DNA sequence data from the Barcode of Life Data System {[}BOLD; Ratnasingham and Hebert (2007){]} to generate branch lengths.
-We were able to successfully generate a tree with BOLD branch lengths for all of the Fringillidae source chronograms.
-However, dating with PATHd8 using congruified calibrations, was only successful in
-three cases (studies 3, 5, and 9, shown in Fig. \ref{fig:cvbold}).
-From these, two trees have a different sampling than the original source chronogram, mainly because DNA BOLD data for some species is absent from the database.
-Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
-More examples and code used to generate these trees were developed on an open repository that is available for consultation and reuse at \url{https://github.com/LunaSare/datelife_examples}.
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-The main goal of \texttt{datelife} is to make state-of-the-art information on time of lineage divergence easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
-It is an open service that does not require any expert biological knowledge from users --besides the names of the organisms they want to work with, for any of its functionality.
-
-At the time of writing of this manuscript (Apr 04, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
-A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time.
-As chronograms are added to Phylesystem, they are incorporated into an updated \texttt{datelife}'s database that is assigned a new version number, followed by a package release on CRAN.
-\texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
-Users can also upload new chronograms to OpenTree themselves, and trigger an update of their local \texttt{datelife} database to incorporate the new chronograms, to have them immediately available for analysis.
-
-Incorporation of more chronograms into \texttt{datelife}'s database is crucial to improve its services.
-One option to increase chronogram number in the database is the Dryad data repository.
-Methods to automatically mine chronograms from Dryad could be designed and implemented.
-However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined manually by a curator.
-
-The largest, and taxonomically broadest, summary chronogram currently available from OpenTree
-was constructed using age data from 2,274 published chronograms (Hedges et al., 2015).
-However the source chronograms used as input data for this tree are not available in computer readable format for reuse or reanalysis.
-As this tree is part of datelife's database, the amount of lineages that can be queried using \texttt{datelife} (99474 unique terminal taxa) is substantial.
-Access to the input chronograms used to generate the Hedges et al.~summary tree would improve measures of uncertainty in DateLife, but they are available only as image files and not as usable data (timetree.org).
-We would like to emphasize on the importance of sharing chronogram data for the benefit of the scientific community as a whole, into repositories that require expert input and manual curation, such as OpenTree's Phylesystem (McTavish et al., 2015).
-
-By default, \texttt{datelife} currently summarizes all source chronograms that overlap with at least two species names. Users can exclude source chronograms if they have reasons to do so.
-Strictly speaking, the best chronogram should reflect the real time of lineage divergence accurately and precisely.
-To our knowledge, there are no good measures to determine independently if a chronogram is better than another.
-Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández (2015).
-Several characteristics of the data used for dating analyses as well as from the output chronogram itself, could be used to score quality of source chronograms.
-Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
-DateLife provides an opportunity to capture concordance and conflict among date estimates, which can also be used as a metric for chronogram reliability.
-
-Scientists usually also favor chronograms constructed using primary calibrations (ages obtained from the fossil or geological record) to ones constructed with secondary calibrations (ages coming from other chronograms)(Schenk, 2016).
-It has been observed with simulations that divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with Bayesian inference methods when priors are implemented in similar ways in both analyses (Schenk, 2016).
-However, secondary calibrations
-can be applied using other dating methods that do not require setting priors, such as penalized likelihood (Sanderson, 2003), or as fixed ages,
-potentially mitigating the bias reported with Bayesian methods.
-Certainly, further studies are required to fully understand the effect of using secondary calibrations on time estimates and downstream analyses.
-
-Furthermore, even chronograms obtained with primary fossil data can vary substantially in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
-This observation is often encountered in the literature
-(see, for example, the ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón et al., 2015; Ramshaw et al., 1972; Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021).
-For some studies, especially ones based on branch lengths (e.g., studies of species diversification, timing of evolutionary events, phenotypic trait evolution), using a different chronogram may return different results (Title \& Rabosky, 2016).
-Stitching together these chronograms can create a larger tree that uses information from multiple studies, but the effect of uncertainties and errors at this level on downstream analyses is still largely unknown.
-
-Summarizing chronograms might also imply summarizing fundamentally distinct evolutionary hypotheses.
-For example, two different researchers working on the same clade both carefully select and argument their choices of fossil calibrations.
-Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses the same one to calibrate outside the clade, the resulting age estimates will often differ substantially, as the placement of calibrations as stem or crown group is proved to deeply affect estimated times of lineage divergence (Sauquet, 2013).
-Trying to summarize the resulting chronograms into a single one using simple summary statistics can erase many types of relevant information from the source chronograms.
-Accordingly, the prevailing view is that we should favor time of lineage divergence estimates obtained from a single analysis, using fossil data as primary sources of calibrations, and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history (Antonelli et al., 2017).
-However, the exercise of summarizing different chronograms has the potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypothesis.
-Choosing the elements of the chronograms that we are going to keep and the ones that we are going to discard is key, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram (Sauquet et al., 2021).
-
-Nonetheless, in ecology and conservation biology, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance (Campbell O.
-Webb et al., 2008).
-Hence, we integrated into datelife's workflow different ways of estimating node ages in the absence of calibrations and branch length information for taxa lacking this information.
-``Making up'' branch lengths is an accepted practice in scientific publications: Jetz et al.
-(2012), created a time-calibrated tree of all 9,993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
-(2018) created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37\% had molecular data; Stephen A.
-Smith and Brown (2018) constructed a tree of 353,185 seed plants where only 23\% had molecular data.
-Obviously, there are risks in this practice!
-Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch estimation methods.
-There has yet to be a thorough analysis of what can go wrong when one extends inferences beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make carefully consider the statistical assumptions being made, and assess the consistency of the results with prior work.
-
-\hypertarget{conclusions}{%
-\section{Conclusions}\label{conclusions}}
-
-Divergence time information is key to many areas of evolutionary studies: trait evolution,
-diversification, biogeography, macroecology and more. It is also crucial for science communication and education, but generating chronograms is difficult,
-especially for those who want to use phylogenies but who are not systematists, or
-do not have the time to acquire and develop the necessary knowledge and data curation skills.
-Moreover, years of primarily public funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
-
-The \texttt{datelife} R package allows easy and fast summarization of publicly available information
-on time of lineage divergence.
-This provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identification of regions that require more research or that have conflicting information.
-It is available as an R package, or a web-based R shiny app at dates.opentreeloflife.org/datelife.
-Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
-The DateLife project helps with awareness of the existing variation in expert time of divergence data, and will foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
-
-\hypertarget{availability}{%
-\section{Availability}\label{availability}}
-
-\texttt{datelife} is free and open source and it can be used through its current website
-\url{http://www.datelife.org/query/}, through its R package, and through Phylotastic's project web portal \url{http://phylo.cs.nmsu.edu:3000/}.
-\texttt{datelife}'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
-\texttt{datelife}'s R package stable version is available
-for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
-using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R. Development versions
-are available from the GitHub repository (\url{https://github.com/phylotastic/datelife})
-and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.
-
-\hypertarget{supplementary-material}{%
-\section{Supplementary Material}\label{supplementary-material}}
-
-Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities are available at \href{https://github.com/LunaSare/datelifeMS1}{datelifeMS1}, \href{https://github.com/LunaSare/datelife_examples}{datelife\_examples}, and \href{https://github.com/LunaSare/datelife_benchmark}{datelife\_benchmark} repositories in LLSR's GitHub account.
-
-\hypertarget{funding}{%
-\section{Funding}\label{funding}}
-
-Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to Datelife project and DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), and the Phylotastic project Grant ABI-1458572, and the OpenTree grant ABI-1759846.
-
-\hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
-
-The DateLife project was born as a prototype tool aiming to provide these services, and was developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA (Stoltzfus et al., 2013).
-We thank colleagues from the O'Meara Lab at the University
-of Tennesse Knoxville for suggestions, discussions and software testing.
-The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
-that led to initial work on this project.
-The team that assembled \texttt{datelife}'s first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
-The Open Tree of Life project that provides the open, metadata rich repository of
-trees used for \texttt{datelife}.
-The many scientists who publish their chronograms in an open, reusable form, and
-the scientists who curate them for deposition in the Open Tree of Life repository.
-The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.
-
-\newpage
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-ane2009groves}{}}%
-Ané, C., Eulenstein, O., Piaggio-Talice, R., \& Sanderson, M. J. (2009). Groves of phylogenetic trees. \emph{{Annals of Combinatorics}}, \emph{13}(2), 139--167.
-
-\leavevmode\vadjust pre{\hypertarget{ref-antonelli2017supersmart}{}}%
-Antonelli, A., Hettling, H., Condamine, F.
-L., Vos, K., Nilsson, R.
-H., Sanderson, M.
-J., \ldots{} Vos, R.
-A.
-(2017).
-{Toward a self-updating platform for estimating rates of speciation and migration, ages, and relationships of Taxa}.
-\emph{Systematic Biology}, \emph{66}(2), 153--166.
-\url{https://doi.org/10.1093/sysbio/syw066}
-
-\leavevmode\vadjust pre{\hypertarget{ref-archie1986newick}{}}%
-Archie, J., Day, W.
-H., Felsenstein, J., Maddison, W., Meacham, C., Rohlf, F.
-J., \& Swofford, D.
-(1986).
-{The Newick tree format}.
-Retrieved from \url{\%7Bhttps://evolution.genetics.washington.edu/phylip/newicktree.html\%7D}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Bapst2012a}{}}%
-Bapst, D.
-W.
-(2012).
-{Paleotree: An R package for paleontological and phylogenetic analyses of evolution}.
-\emph{{Methods in Ecology and Evolution}}, \emph{3}(5), 803--807.
-\url{https://doi.org/10.1111/j.2041-210X.2012.00223.x}
-
-\leavevmode\vadjust pre{\hypertarget{ref-barba2018constraining}{}}%
-Barba-Montoya, J., Reis, M.
-dos, Schneider, H., Donoghue, P.
-C., \& Yang, Z.
-(2018).
-Constraining uncertainty in the timescale of angiosperm evolution and the veracity of a cretaceous terrestrial revolution.
-\emph{New Phytologist}, \emph{218}(2), 819--834.
-
-\leavevmode\vadjust pre{\hypertarget{ref-barker2013going}{}}%
-Barker, F.
-K., Burns, K.
-J., Klicka, J., Lanyon, S.
-M., \& Lovette, I.
-J.
-(2013).
-Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds.
-\emph{{Systematic Biology}}, \emph{62}(2), 298--320.
-
-\leavevmode\vadjust pre{\hypertarget{ref-barker2015new}{}}%
-Barker, F.
-K., Burns, K.
-J., Klicka, J., Lanyon, S.
-M., \& Lovette, I.
-J.
-(2015).
-New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
-\emph{{The Auk: Ornithological Advances}}, \emph{132}(2), 333--348.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Britton2007}{}}%
-Britton, T., Anderson, C.
-L., Jacquet, D., Lundqvist, S., \& Bremer, K.
-(2007).
-{Estimating Divergence Times in Large Phylogenetic Trees}.
-\emph{{Systematic Biology}}, \emph{56}(788777878), 741--752.
-\url{https://doi.org/10.1080/10635150701613783}
-
-\leavevmode\vadjust pre{\hypertarget{ref-burns2014phylogenetics}{}}%
-Burns, K.
-J., Shultz, A.
-J., Title, P.
-O., Mason, N.
-A., Barker, F.
-K., Klicka, J., \ldots{} Lovette, I.
-J.
-(2014).
-Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds.
-\emph{{Molecular Phylogenetics and Evolution}}, \emph{75}, 41--77.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Chamberlain2013}{}}%
-Chamberlain, S.
-A., \& Szöcs, E.
-(2013).
-{taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}}.
-\emph{{F1000Research}}, \emph{2}(191), 1--29.
-\url{https://doi.org/10.12688/f1000research.2-191.v2}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Chamberlain2018}{}}%
-Chamberlain, S.
-A., Szöcs, E., Foster, Z., Arendsee, Z., Boettiger, C., Ram, K., \ldots{} Li, G.
-(2019).
-\emph{{taxize: Taxonomic information from around the web}}.
-Retrieved from \url{https://github.com/ropensci/taxize}
-
-\leavevmode\vadjust pre{\hypertarget{ref-claramunt2015new}{}}%
-Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{{Science Advances}}, \emph{1}(11), e1501005.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Criscuolo2006}{}}%
-Criscuolo, A., Berry, V., Douzery, E.
-J.
-P., \& Gascuel, O.
-(2006).
-{SDM: A fast distance-based approach for (super)tree building in phylogenomics}.
-\emph{{Systematic Biology}}, \emph{55}(5), 740--755.
-\url{https://doi.org/10.1080/10635150600969872}
-
-\leavevmode\vadjust pre{\hypertarget{ref-delsuc2018phylogenomic}{}}%
-Delsuc, F., Philippe, H., Tsagkogeorga, G., Simion, P., Tilak, M.-K., Turon, X., \ldots{} Douzery, E.
-J.
-(2018).
-A phylogenomic framework and timescale for comparative studies of tunicates.
-\emph{BMC Biology}, \emph{16}(1), 1--14.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Eastman2013}{}}%
-Eastman, J.
-M., Harmon, L.
-J., \& Tank, D.
-C.
-(2013).
-{Congruification: Support for time scaling large phylogenetic trees}.
-\emph{{Methods in Ecology and Evolution}}, \emph{4}(7), 688--691.
-\url{https://doi.org/10.1111/2041-210X.12051}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Felsenstein1985a}{}}%
-Felsenstein, J. (1985). {Phylogenies and the Comparative Method}. \emph{The American Naturalist}, \emph{125}(1), 1--15. Retrieved from \url{http://www.jstor.org/stable/2461605}
-
-\leavevmode\vadjust pre{\hypertarget{ref-gibb2015new}{}}%
-Gibb, G.
-C., England, R., Hartig, G., McLenachan, P.
-A., Taylor Smith, B.
-L., McComish, B.
-J., \ldots{} Penny, D.
-(2015).
-New zealand passerines help clarify the diversification of major songbird lineages during the oligocene.
-\emph{{Genome Biology and Evolution}}, \emph{7}(11), 2983--2995.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Harmon2008}{}}%
-Harmon, L., Weir, J., Brock, C., Glor, R., \& Challenger, W. (2008). {GEIGER: investigating evolutionary radiations}. \emph{Bioinformatics}, \emph{24}, 129--131.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Hedges2015}{}}%
-Hedges, S.
-B., Marin, J., Suleski, M., Paymer, M., \& Kumar, S.
-(2015).
-{Tree of life reveals clock-like speciation and diversification}.
-\emph{{Molecular Biology and Evolution}}, \emph{32}(4), 835--845.
-\url{https://doi.org/10.1093/molbev/msv037}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Heibl2008}{}}%
-Heibl, C. (2008). \emph{PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages.} Retrieved from \url{http://www.christophheibl.de/Rpackages.html}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hooper2017chromosomal}{}}%
-Hooper, D. M., \& Price, T. D. (2017). Chromosomal inversion differences correlate with range overlap in passerine birds. \emph{Nature Ecology \& Evolution}, \emph{1}(10), 1526.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Huelsenbeck2001}{}}%
-Huelsenbeck, J. P., \& Ronquist, F. (2001). {MRBAYES: Bayesian inference of phylogenetic trees}. \emph{Bioinformatics}, \emph{17}(8), 754--755. \url{https://doi.org/10.1093/bioinformatics/17.8.754}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Jetz2012}{}}%
-Jetz, W., Thomas, G., Joy, J.
-J.
-B., Hartmann, K., \& Mooers, A.
-(2012).
-{The global diversity of birds in space and time}.
-\emph{Nature}, \emph{491}(7424), 444--448.
-\url{https://doi.org/10.1038/nature11631}
-
-\leavevmode\vadjust pre{\hypertarget{ref-laubichler2009form}{}}%
-Laubichler, M. D., \& Maienschein, J. (2009). \emph{Form and function in developmental evolution}. Cambridge University Press.
-
-\leavevmode\vadjust pre{\hypertarget{ref-magallon2001absolute}{}}%
-Magallon, S., \& Sanderson, M. J. (2001). Absolute diversification rates in angiosperm clades. \emph{Evolution}, \emph{55}(9), 1762--1780.
-
-\leavevmode\vadjust pre{\hypertarget{ref-magallon2010using}{}}%
-Magallón, S. (2010). Using fossils to break long branches in molecular dating: A comparison of relaxed clocks applied to the origin of angiosperms. \emph{Systematic Biology}, \emph{59}(4), 384--399.
-
-\leavevmode\vadjust pre{\hypertarget{ref-magallon2015metacalibrated}{}}%
-Magallón, S., Gómez-Acevedo, S., Sánchez-Reyes, L.
-L., \& Hernández-Hernández, T.
-(2015).
-A metacalibrated time-tree documents the early rise of flowering plant phylogenetic diversity.
-\emph{New Phytologist}, \emph{207}(2), 437--453.
-
-\leavevmode\vadjust pre{\hypertarget{ref-mctavish2015phylesystem}{}}%
-McTavish, E.
-J., Hinchliff, C.
-E., Allman, J.
-F., Brown, J.
-W., Cranston, K.
-A., Holder, M.
-T., \ldots{} Smith, S.
-A.
-(2015).
-Phylesystem: A git-based data store for community-curated phylogenetic estimates.
-\emph{Bioinformatics}, \emph{31}(17), 2794--2800.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Michonneau2016}{}}%
-Michonneau, F., Brown, J.
-W., \& Winter, D.
-J.
-(2016).
-{rotl: an R package to interact with the Open Tree of Life data}.
-\emph{{Methods in Ecology and Evolution}}, \emph{7}(12), 1476--1481.
-\url{https://doi.org/10.1111/2041-210X.12593}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Morlon2014}{}}%
-Morlon, H. (2014). {Phylogenetic approaches for studying diversification.} \emph{{Ecology Letters}}, \emph{17}(4), 508--525. \url{https://doi.org/10.1111/ele.12251}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Ooms2018}{}}%
-Ooms, J., \& Chamberlain, S. (2018). \emph{Phylocomr: Interface to 'phylocom'}. Retrieved from \url{https://CRAN.R-project.org/package=phylocomr}
-
-\leavevmode\vadjust pre{\hypertarget{ref-opentreeAPIs}{}}%
-Open Tree Of Life, Redelings, B., Cranston, K.
-A., Allman, J., Holder, M.
-T., \& McTavish, E.
-J.
-(2016).
-{Open Tree of Life APIs v3.0}.
-\emph{{Open Tree of Life Project}}, (Online Resources).
-Retrieved from \url{\%7Bhttps://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\%7D}
-
-\leavevmode\vadjust pre{\hypertarget{ref-opentreeoflife2019synth}{}}%
-Open Tree Of Life, Redelings, B., Sánchez Reyes, L.
-L., Cranston, K.
-A., Allman, J., Holder, M.
-T., \& McTavish, E.
-J.
-(2019).
-Open tree of life synthetic tree v12.3.
-\emph{Zenodo}.
-Retrieved from \url{https://doi.org/10.5281/zenodo.3937742}
-
-\leavevmode\vadjust pre{\hypertarget{ref-paradis2004}{}}%
-Paradis, E., Claude, J., \& Strimmer, K. (2004). {APE: analyses of phylogenetics and evolution in R language}. \emph{Bioinformatics}, \emph{20}(2), 289--290.
-
-\leavevmode\vadjust pre{\hypertarget{ref-posadas2006historical}{}}%
-Posadas, P., Crisci, J. V., \& Katinas, L. (2006). Historical biogeography: A review of its basic concepts and critical issues. \emph{{Journal of Arid Environments}}, \emph{66}(3), 389--403.
-
-\leavevmode\vadjust pre{\hypertarget{ref-price2014niche}{}}%
-Price, T.
-D., Hooper, D.
-M., Buchanan, C.
-D., Johansson, U.
-S., Tietze, D.
-T., Alström, P., \ldots{} others.
-(2014).
-Niche filling slows the diversification of himalayan songbirds.
-\emph{Nature}, \emph{509}(7499), 222.
-
-\leavevmode\vadjust pre{\hypertarget{ref-RCoreTeam2018}{}}%
-R Core Team. (2018). \emph{{R: a language and environment for statistical computing}}. Vienna, Austria: R Foundation for Statistical Computing.
-
-\leavevmode\vadjust pre{\hypertarget{ref-rabosky2018inverse}{}}%
-Rabosky, D.
-L., Chang, J., Title, P.
-O., Cowman, P.
-F., Sallan, L., Friedman, M., \ldots{} others.
-(2018).
-An inverse latitudinal gradient in speciation rate for marine fishes.
-\emph{Nature}, \emph{559}(7714), 392.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ramshaw1972time}{}}%
-Ramshaw, J., Richardson, D., Meatyard, B., Brown, R., Richardson, M., Thompson, E., \& Boulter, D.
-(1972).
-The time of origin of the flowering plants determined by using amino acid sequence data of cytochrome c.
-\emph{New Phytologist}, \emph{71}(5), 773--779.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ratnasingham2007bold}{}}%
-Ratnasingham, S., \& Hebert, P. D. (2007). BOLD: The barcode of life data system (http://www. Barcodinglife. org). \emph{Molecular Ecology Notes}, \emph{7}(3), 355--364.
-
-\leavevmode\vadjust pre{\hypertarget{ref-rees2017automated}{}}%
-Rees, J. A., \& Cranston, K. (2017). Automated assembly of a reference taxonomy for phylogenetic data synthesis. \emph{Biodiversity Data Journal}, (5).
-
-\leavevmode\vadjust pre{\hypertarget{ref-Revell2012}{}}%
-Revell, L. J. (2012). Phytools: An r package for phylogenetic comparative biology (and other things). \emph{{Methods in Ecology and Evolution}}, \emph{3}, 217--223.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Ronquist2003}{}}%
-Ronquist, F., \& Huelsenbeck, J.
-P.
-(2003).
-{MrBayes 3: Bayesian phylogenetic inference under mixed models}.
-\emph{Bioinformatics}, \emph{19}(12), 1572--1574.
-\url{https://doi.org/10.1093/bioinformatics/btg180}
-
-\leavevmode\vadjust pre{\hypertarget{ref-datelifeR}{}}%
-Sanchez-Reyes, L.
-L., O'Meara, B., Eastman, J., Heath, T., Wright, A., Schliep, K., \ldots{} Alfaro, M.
-(2022).
-{datelife: Scientific Data on Time of Lineage Divergence for Your Taxa}.
-\emph{R Package Version 0.6.2}.
-Retrieved from \url{https://doi.org/10.5281/zenodo.593938}
-
-\leavevmode\vadjust pre{\hypertarget{ref-sanderson2003r8s}{}}%
-Sanderson, M. J. (2003). r8s: Inferring absolute rates of molecular evolution and divergence times in the absence of a molecular clock. \emph{Bioinformatics}, \emph{19}(2), 301--302.
-
-\leavevmode\vadjust pre{\hypertarget{ref-sanderson2001sources}{}}%
-Sanderson, M.
-J., \& Doyle, J.
-A.
-(2001).
-Sources of error and confidence intervals in estimating the age of angiosperms from rbcL and 18S rDNA data.
-\emph{American Journal of Botany}, \emph{88}(8), 1499--1516.
-
-\leavevmode\vadjust pre{\hypertarget{ref-sauquet2013practical}{}}%
-Sauquet, H. (2013). A practical guide to molecular dating. \emph{{Comptes Rendus Palevol}}, \emph{12}(6), 355--367.
-
-\leavevmode\vadjust pre{\hypertarget{ref-sauquet2021age}{}}%
-Sauquet, H., Ramírez-Barahona, S., \& Magallón, S. (2021). The age of flowering plants is unknown.
-
-\leavevmode\vadjust pre{\hypertarget{ref-schenk2016sec}{}}%
-Schenk, J. J. (2016). {Consequences of secondary calibrations on divergence time estimates}. \emph{PLoS ONE}, \emph{11}(1). \url{https://doi.org/10.1371/journal.pone.0148228}
-
-\leavevmode\vadjust pre{\hypertarget{ref-smith2018constructing}{}}%
-Smith, Stephen A., \& Brown, J. W. (2018). Constructing a broadly inclusive seed plant phylogeny. \emph{American Journal of Botany}, \emph{105}(3), 302--314.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Smith2012}{}}%
-Smith, Stephen A., \& O'Meara, B.
-C.
-(2012).
-{TreePL: Divergence time estimation using penalized likelihood for large phylogenies}.
-\emph{Bioinformatics}, \emph{28}(20), 2689--2690.
-\url{https://doi.org/10.1093/bioinformatics/bts492}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Stoltzfus2013}{}}%
-Stoltzfus, A., Lapp, H., Matasci, N., Deus, H., Sidlauskas, B., Zmasek, C.
-M., \ldots{} Jordan, G.
-(2013).
-{Phylotastic!
-Making tree-of-life knowledge accessible, reusable and convenient}.
-\emph{{BMC Bioinformatics}}, \emph{14}.
-\url{https://doi.org/10.1186/1471-2105-14-158}
-
-\leavevmode\vadjust pre{\hypertarget{ref-title2016macrophylogenies}{}}%
-Title, P.
-O., \& Rabosky, D.
-L.
-(2016).
-{Do Macrophylogenies Yield Stable Macroevolutionary Inferences?
-An Example from Squamate Reptiles}.
-\emph{Systematic Biology}, syw102.
-\url{https://doi.org/10.1093/sysbio/syw102}
-
-\leavevmode\vadjust pre{\hypertarget{ref-vos2012nexml}{}}%
-Vos, R.
-A., Balhoff, J.
-P., Caravas, J.
-A., Holder, M.
-T., Lapp, H., Maddison, W.
-P., \ldots{} others.
-(2012).
-NeXML: Rich, extensible, and verifiable representation of comparative data and metadata.
-\emph{Systematic Biology}, \emph{61}(4), 675--689.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Webb2000}{}}%
-Webb, C. (2000). {Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees}. \emph{{The American Naturalist}}, \emph{156}(2), 145--155.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Webb2008}{}}%
-Webb, Campbell O., Ackerly, D.
-D., \& Kembel, S.
-W.
-(2008).
-{Phylocom: Software for the analysis of phylogenetic community structure and trait evolution}.
-\emph{Bioinformatics}, \emph{24}(18), 2098--2100.
-\url{https://doi.org/10.1093/bioinformatics/btn358}
-
-\leavevmode\vadjust pre{\hypertarget{ref-webb2005phylomatic}{}}%
-Webb, Campbell O., \& Donoghue, M. J. (2005). Phylomatic: Tree assembly for applied phylogenetics. \emph{{Molecular Ecology Notes}}, \emph{5}(1), 181--183.
-
-\end{CSLReferences}
-
-\endgroup
-
 % in this file I leave the figure captions outside\ caption{} because I want them
@@
 \begin{figure}[!h]
-\includegraphics{../figures/figure1/figure1-vertical-final.pdf}
-\caption{Stylized DateLife workflow.
-This shows the general worflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{http://www.datelife.org/}.
-Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
-}
-\label{fig:workflow}
+\includegraphics{../figures/figure-cross-validation/fig-cross-validation-xy-plots.pdf}
+\caption{Results from cross validation analysis.}
+\label{fig:cvXY}
 \end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:workflow}}
-% \end{center}
 
 \begin{figure}[!h]
-\includegraphics[width=1\linewidth]{../figures/fig_runtime_main.pdf}
-\caption{Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names.
-We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray).
-We also performed a search using the already processed query for comparison (light blue).}
-\label{fig:runtime_main}
+\includegraphics{../figures/figure-cross-validation/fig-cross-validation-xy-plots-diffs.pdf}
+\caption{Results from cross validation analysis.}
+\label{fig:cvXYdiffs}
 \end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:benchmark}}
-% \end{center}
-
-\begin{figure}[!h]
-\includegraphics[width=1\linewidth]{../figures/figure2/figure2-1.pdf}
-\caption{Age data results of a DateLife search of a small sample of 6 bird species within the Passeriformes.
-Input names were found across 9 chronograms within 6 independent studies (Barker et al.
-(\protect\hyperlink{ref-barker2012going}{2012}), Barker et al.
-(\protect\hyperlink{ref-barker2015new}{2015}), Burns et al.
-(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al.
-(\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al.
-(\protect\hyperlink{ref-Jetz2012}{2012}).) This revealed 28 age data points for the queried species names.}
-\label{fig:figure2-1}
-\end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:figure2-1}}
-% \end{center}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure2/figure2-2.pdf}
-\caption{Summarized age data is used as secondary calibrations to date a tree topology as a summary chronogram.}
-\label{fig:summaries}
-\end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:figure2-2}}
-% \end{center}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-fringillidae/median_and_calibration_ages_simple.png}
-\caption{Fringillidae median summary chronogram genertaed with DateLife. It has 256 tips and 233 nodes.}
-\label{fig:fringillidages}
-\end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:fringillidages}}
-% \end{center}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/cross_validation_1.png}
-\caption{Cross validation of first source chronogram.
-The chronogram shown in black corresponds to the dates published in the original study.
-The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
-\label{fig:cv1}
-\end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:cvbold}}
-% \end{center}
 
@@
 \end{figure}
-
-% \end{linenumbers}
-
-\end{document}

### 2022-04-06 manuscript/manuscript.Rmd

@@
       in_header: "preamble.tex"
-      after_body  : "figures-alt.tex"
+      after_body  : ["figures-alt.tex", "figures-supplementary.tex"]
 
@@
 The DateLife algorithm is fully implemented using the R language.
-The latest stable version of the R package `datelife` is available from the CRAN repository (v`r packageVersion("datelife")`; @datelifeR), and relies on functionalities from various biological R packages:
+The latest stable version of the R package `datelife` -- v`r packageVersion("datelife")`, is available from the CRAN repository [@datelifeR], and relies on functionalities from various biological R packages:
 ape [@paradis2004],
@@
 Chronograms with at least two matching tip labels are identified and pruned down to preserve only the matched tips.
-
-Matching pruned chronograms are stored as individual patristic distance matrices (Figure \ref{fig:workflow} subfigure X).
-This matrix consists of ... ???? the pairwise distance between pairs of query taxa which are in that input tree, in units of millions of years. <!-- is that true? -->
-
-This format speeds up extraction of pairwise taxon ages of the queried taxa, as opposed to searching the ancestor node of a pair of taxa in a "phylo" object or newick string.
-The patristic matrices are also associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.
+These matching pruned chronograms are now referred to as source chronograms.
+Total distance (in units of millions of years) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Figure \ref{fig:workflow}).
+The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a "phylo" object or newick string.
+Finally, the patristic matrices are associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.
 
@@
 It is constructed from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert phylogenetic knowledge with rich metadata [@mctavish2015phylesystem] that allows automatic and reproducible construction of a chronogram database.
-New chronograms can be added to Phylesystem by any user and are immediately publicly available, and the DateLife database can be updated to include those new data within a run.
+New chronograms can be added to Phylesystem by any user and are immediately publicly available.
+This permits an immediate update of DateLife's chronogram database to include those new data within a run.
 
@@
 
-At this point, summary information is extracted from the \texttt{datelifeResult} object to inform decisions for the subsequent steps in the user workflow.
-Age data from the matching pruned chronograms is summarized and used to generate a single summary chronogram.
-Other basic summary information available to the user is:
+Summary information is extracted from the \texttt{datelifeResult} object to inform decisions for the subsequent steps in the user workflow. Basic summary information available to the user is:
 
@@
 2.
-The ages of the root of all matching pruned chronograms.
-This can correspond to the age of the most recent common ancestor (mrca) of your group of interest if the pruned chronograms have all taxa belonging to the group.
+The ages of the root of all source chronograms.
+This can correspond to the age of the most recent common ancestor (mrca) of your group of interest if the source chronograms have all taxa belonging to the group.
 If not, the root corresponds to the mrca of a subgroup withing your group of interest.
 3. Study citations where original chronograms were published.
-4. A report of input taxon names matches across pruned chronograms.
-5. The single matching pruned chronogram with the most input taxon names.
+4. A report of input taxon names matches across source chronograms.
+5. The source chronogram(s) with the most input taxon names.
+6. Last but not least, age data from source chronograms are summarized and used to generate a single summary chronogram:
 
-_**Identifying groves.--**_ To generate a single summary chronogram, the DateLife algorithm starts by identifying the matching pruned chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
+_**Choosing a topology.--**_ DateLife requires a tree topology to summarize age data upon.
+We recommend that users provide one as input from the literature, or one of their own making.
+If no topology is provided, DateLife automatically subsets one from the OpenTree synthetic tree [@opentreeoflife2019synth].
+Alternatively, DateLife can reconstruct a tree with branch lengths proportional to substitution rates from a starting tree topology using genetic data from the Barcode of Life Data System, BOLD [@ratnasingham2007bold], or combine topologies from source chronogras using a supertree approach.
+
+_**Reconstructing branch lengths.--**_ DateLife starts by mining the BOLD database to obtain genetic markers for the input taxa, and aligning them with MUSCLE [@edgar2004muscle; or MAFFT, @katoh2009multiple].
+Currently, branch length reconstruction is performed with parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed [@schliep2011phangorn].
+
+_**Combining source chronograms.--**_ To combine topologies from source chronograms into a single summary topology, the DateLife algorithm starts by identifying the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
 [-@ane2009groves].
@@
 By default, DateLife chooses the grove with the most taxa, however, the "criterion = trees" flag allows the user to choose the grove with the most trees instead.
-<!--Why do groves matter? Why can't you just use all the trees?-->
+If source chronograms do not form a grove, the supertree reonstruction will fail.
 
-_**Choosing a topology.--**_ DateLife requires a tree topology to summarize age data upon.
-Users can provide one as input from the literature, or one of their own making.
-If no topology is provided, DateLife automatically subsets one from the OpenTree synthetic tree [@opentreeoflife2019synth].
+_**Congruifying nodes.--**_ Once with a chosen topology, DateLife applies the congruification method [@Eastman2013] to find nodes belonging to the same clade across source chronograms, and extract the corresponding node ages from the patristic distance matrices stored as `datelifeResult`.
+Since by definition the matrices store total distance (time from tip to tip), node ages correspond to half the values stored in the patristic distance matrices.
+A table of congruified node ages that can be used as secondary calibrations is stored as a `congruifiedCalibrations` object.
 
-DateLife can also reconstruct branch lengths proportional to substitution rates on a fixed tree topology using available genetic data from BOLD.
+_**Summarizing congruified ages.--**_ Congruified pairwise node ages are used to calculate summary statistics (mean, median, minimum and maximum ages) per taxon pair.
+<!-- we should calculate summary statistics per node -->
+<!--??? How is 'calibrations' defined here? the calibrations from congruification, or the pairwise distances??? is calibrations from congruification-->
+into a single summary matrix using different methods.
+Summarizing options implemented include Super Distance Matrix method [SDM, @Criscuolo2006] and summary statistics.
 
-_**Congruifying nodes.--**_ DateLife then implements the congruification method [@Eastman2013] to find nodes belonging to the same clade across matching pruned chronograms.
-Congruified node ages stored as a `congruifiedCalibrations` object are then matched to nodes in the chosen tree topology and stored as a `matchedCalibrations` object.
-<!--This doesn't use the input chronogram pruned patristic distance matricies? Or does it? -->
-
-_**Summarizing node ages.--**_ DateLife summarizes matched calibrations
-<!--??? How is 'calibrations' defined here? the calibrations from congruification, or the pairwise distances??? -->
-into a single patristic distance matrix using different methods.
-Summarizing options implemented include Super Distance Matrix method [SDM, @Criscuolo2006] and summary statistics such as median, minimum and maximum ages.
+Ages per taxon pair can be used as calibrations to date the chosen tree topology.
 <!--
@@
 
-_**Dating the tree topology.--**_ Summarized calibrations can be applied as secondary calibrations with different dating methods currently supported within DateLife: MrBayes [@Huelsenbeck2001; @Ronquist2003], PATHd8 [@Britton2007], BLADJ [@Webb2008; @webb2005phylomatic], and treePL [@Smith2012].
-<!--Huh? can you run MrBayes in DateLife? You should explain how the BOLD dating with PathD8 works here, and what is needed for the other ones!-->
-
-By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm that assigns ages to nodes with no data evenly between nodes with age data, which minimizes age variance in the resulting chronogram [@Webb2008].
-When there is conflict in ages across node with age data, the algorithm ignores ages that are older than parent nodes and/or younger than descendant nodes.
+_**Dating the tree topology.--**_ 
+By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully datd topology.
+BLADJ fixes node ages that have calibration data, and distrivutes time between nodes with no data evenly between nodes with calibration data.
+This minimizes age variance in the resulting chronogram [@Webb2008].
+When there is conflict in ages between nodes with calibration data, the algorithm ignores ages that are older than ages of parent nodes and/or younger than ages from descendant nodes.
 
@@
 If none is provided, DateLife assigns an arbitrary age to the root as 10% older than the oldest age available within the group.
+
+Summarized calibrations can be applied as secondary calibrations with different dating methods currently supported within DateLife: MrBayes [@Huelsenbeck2001; @Ronquist2003], PATHd8 [@Britton2007], BLADJ [@Webb2008; @webb2005phylomatic], and treePL [@Smith2012].
+<!--Huh? can you run MrBayes in DateLife? You should explain how the BOLD dating with PathD8 works here, and what is needed for the other ones!-->
+
 <!--PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees.
@@
 Query processing time increases roughly linearly with number of input taxon names, and
-increases considerably if TNRS is activated.
+increases considerably if Taxonomic Name Resolution Service (TNRS) is activated.
 <!-- Defining tnrs here again, bc we only talk about it two times, so it easy to forget it -->
@@
 
-# Case study
+# Case studies
 
@@
 
-We chose 6 bird species associated to true finches at random. The sample includes
+We randomly chose 6 bird species related to the family Fringillidae of true finches. The sample includes
 two species of cardinals:
@@
 
-Processing input names found that *Emberiza elegans* is synonym for *Schoeniclus elegans* in the reference taxonomy.
+Processing input names found that *Emberiza elegans* is synonym for *Schoeniclus elegans* in the default reference taxonomy [Open Tree of Life Taxonomy v3.3, June 1, 2021].
+For a detailed discussion on the state of the synonym refer to Avibase [@lepage2004avibase; @lepage2014avibase; @avibase-emberiza].
 DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies.
@@
 As expected, more inclusive nodes (e.g., node "n1") have more age data than less inclusive nodes (e.g., node "n5").
-The processing step allowd discovering five data points for node "n4" that would not have had any data otherwise.
+The processing step allowed discovering five data points for node "n4" that would not have had any data otherwise.
 Age summary statistics per node were calculated and tested as secondary calibrations to date the tree topology using the BLADJ algorithm.
@@
 Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
-????????????????? Say some things about the results!
-<!-- (Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
+<!-- ????????????????? Say some things about the results!
+ANSWER:
+(Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
 <!-- and hierarchical clustering method which appears to push ages back -->
@@
 
-Data from source chronograms can be used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates (Figs.
-\ref{fig:cv1} to \ref{fig:cv10}).
-As a form of cross validation, we took tree topologies from each input study and calibrated them using time of lineage divergence data from all other source chronograms.
-<!--I'm confused about why you are talking about branch lengths here... did you use branch lengths?-->
+<!--Data from source chronograms can be used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates-->
+To perform a cross validation analysis of the DateLife workflow, we used resulting data from the previous section (Casestudy: Real Life Application).
+We took individual tree topologies from each of the 19 source chronograms found (Supplementary data XX).
+Then we congruifued age data of source chronograms from studies, and used this ages to date the tree topology with the program BLADJ.
 
-In the absence of branch lengths, the ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvbladj}).
-Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cv2}).
-<!-- We need to think about the root! is it comparable across source chronograms?? not all of them have the same sampling... so probably not the same root and thus not comparable. Address this -->
-We also demonstrate the usage of PATHd8 [@Britton2007] as an dating method alternative to BLADJ.
-For this, we run a `datelife` branch length reconstruction that searches for DNA sequence data from the Barcode of Life Data System [BOLD; @ratnasingham2007bold] to generate branch lengths.
-We were able to successfully generate a tree with BOLD branch lengths for all of the Fringillidae source chronograms.
-However, dating with PATHd8 using congruified calibrations, was only successful in
-<!-- two studies when expanding calibrations to make them agree (studies 3 and 5), and from  -->
-three cases (studies 3, 5, and 9, shown in Fig. \ref{fig:cvbold}).
-<!-- What made it unsuccessful in the other cases? -->
-From these, two trees have a different sampling than the original source chronogram, mainly because DNA BOLD  data for some species is absent from the database.
-<!-- ??? Node ages or maximum ages?? -->
-Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
-More examples and code used to generate these trees were developed on an open repository that is available for consultation and reuse at <https://github.com/LunaSare/datelife_examples>.
+We found that node ages from original study, and ages estimated using all other age data available are generally correlated (Supplementary Fig. \ref{fig:cvXY}).
+In 5 studies, more inclusive nodes have older original ages, and less inclusive nodes have younger original ages than their cross-validated age estimates. 
+Accordingly, root ages are generally older in the original study than estimated using cross-validated ages.
+Root ages were similar in original and cross validated ages in three vases (Supplementary Fig.
+\ref{fig:cvXYdiffs}).
+<!-- We need to think about the root!
+is it comparable across source chronograms??
+not all of them have the same sampling...
+so probably not the same root and thus not comparable.
+Address this...
+ANSWERED: -->
+Notably, chronograms have different species sampling, hence roots are not comparable across studies.
+Yet, chronograms with a higher sampling number can potentially inform the age of the root of chronograms with less sampling.
+
+<!-- 
+%```{r, child = "cross-validation-pathd8.Rmd"}
+%```
+-->
 

### 2022-04-07 manuscript/manuscript.Rmd
editing the text

@@
       in_header: "preamble.tex"
-      after_body  : ["figures-alt.tex", "figures-supplementary.tex"]
+      after_body  : "figures-alt.tex"
 
@@
 The DateLife algorithm is fully implemented using the R language.
-The latest stable version of the R package `datelife` -- v`r packageVersion("datelife")`, is available from the CRAN repository [@datelifeR], and relies on functionalities from various biological R packages:
+The latest stable version of the R package `datelife` is available from the CRAN repository (v`r packageVersion("datelife")`; @datelifeR), and relies on functionalities from various biological R packages:
 ape [@paradis2004],
@@
 If no topology is provided, DateLife automatically subsets one from the OpenTree synthetic tree [@opentreeoflife2019synth].
-Alternatively, DateLife can reconstruct a tree with branch lengths proportional to substitution rates from a starting tree topology using genetic data from the Barcode of Life Data System, BOLD [@ratnasingham2007bold], or combine topologies from source chronogras using a supertree approach.
+Alternatively, DateLife can reconstruct a tree with branch lengths proportional to substitution rates from a starting tree topology using genetic data from the Barcode of Life Data System [BOLD; @ratnasingham2007bold], or combine topologies from source chronogras using a supertree approach.
 
-_**Reconstructing branch lengths.--**_ DateLife starts by mining the BOLD database to obtain genetic markers for the input taxa, and aligning them with MUSCLE [@edgar2004muscle; or MAFFT, @katoh2009multiple].
+_**Reconstructing branch lengths.--**_ DateLife starts by mining the BOLD database to obtain genetic markers for the input taxa, and aligning them with MUSCLE [@edgar2004muscle; or MAFFT @katoh2009multiple].
 Currently, branch length reconstruction is performed with parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed [@schliep2011phangorn].
@@
 into a single summary matrix using different methods.
-Summarizing options implemented include Super Distance Matrix method [SDM, @Criscuolo2006] and summary statistics.
+Summarizing options implemented include Super Distance Matrix method [SDM, @Criscuolo2006] and .
 
@@
 
-# Case studies
+# Case study
 
@@
 Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
-<!-- ????????????????? Say some things about the results!
-ANSWER:
-(Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
+????????????????? Say some things about the results!
+<!-- (Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
 <!-- and hierarchical clustering method which appears to push ages back -->
@@
 <!--Data from source chronograms can be used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates-->
-To perform a cross validation analysis of the DateLife workflow, we used resulting data from the previous section (Casestudy: Real Life Application).
-We took individual tree topologies from each of the 19 source chronograms found (Supplementary data XX).
-Then we congruifued age data of source chronograms from studies, and used this ages to date the tree topology with the program BLADJ.
+As a form of cross validation, we took the tree topologies from each study and used time of lineage divergence data from all other source chronograms as fixed calibrations with the program BLADJ (Figs.
+\ref{fig:cv1} to \ref{fig:cv10}).
+<!--I'm confused about why you are talking about branch lengths here...
+did you use branch lengths?
+Yes, on the PATHd8 cross validation, but you are right, it is confusing, so I removed that first sentence-->
 
-We found that node ages from original study, and ages estimated using all other age data available are generally correlated (Supplementary Fig. \ref{fig:cvXY}).
-In 5 studies, more inclusive nodes have older original ages, and less inclusive nodes have younger original ages than their cross-validated age estimates. 
-Accordingly, root ages are generally older in the original study than estimated using cross-validated ages.
-Root ages were similar in original and cross validated ages in three vases (Supplementary Fig.
-\ref{fig:cvXYdiffs}).
-<!-- We need to think about the root!
-is it comparable across source chronograms??
-not all of them have the same sampling...
-so probably not the same root and thus not comparable.
-Address this...
-ANSWERED: -->
-Notably, chronograms have different species sampling, hence roots are not comparable across studies.
-Yet, chronograms with a higher sampling number can potentially inform the age of the root of chronograms with less sampling.
-
-<!-- 
-%```{r, child = "cross-validation-pathd8.Rmd"}
-%```
--->
+The ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvbladj}).
+Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cv2}).
+<!-- We need to think about the root! is it comparable across source chronograms?? not all of them have the same sampling... so probably not the same root and thus not comparable. Address this -->
+We also demonstrate the usage of PATHd8 [@Britton2007] as a dating method alternative to BLADJ.
+PATHd8 requires a tree with initial branch lengths which we obtained using `datelife`'s branch length reconstruction algorithm, that searches for DNA sequence data from the Barcode of Life Data System [BOLD; @ratnasingham2007bold] to generate branch lengths.
+We were able to successfully generate a tree with BOLD branch lengths for all of the Fringillidae source chronograms.
+However, dating with PATHd8 using congruified calibrations, was only successful in
+<!-- two studies when expanding calibrations to make them agree (studies 3 and 5), and from  -->
+three cases (studies 3, 5, and 9, shown in Fig. \ref{fig:cvbold}).
+<!-- What made it unsuccessful in the other cases? -->
+From these, two trees have a different sampling than the original source chronogram, mainly because DNA BOLD  data for some species is absent from the database.
+<!-- ??? Node ages or maximum ages?? -->
+Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
+More examples and code used to generate these trees were developed on an open repository that is available for consultation and reuse at <https://github.com/LunaSare/datelife_examples>.
 

### 2022-04-07 manuscript/manuscript.tex
editing the text

@@
 \shorttitle{DateLife: revealing the dated Tree of Life}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4201}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4379}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 Chronograms with at least two matching tip labels are identified and pruned down to preserve only the matched tips.
-
-Matching pruned chronograms are stored as individual patristic distance matrices (Figure \ref{fig:workflow} subfigure X).
-This matrix consists of \ldots{} ???? the pairwise distance between pairs of query taxa which are in that input tree, in units of millions of years.
-
-This format speeds up extraction of pairwise taxon ages of the queried taxa, as opposed to searching the ancestor node of a pair of taxa in a ``phylo'' object or newick string.
-The patristic matrices are also associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.
+These matching pruned chronograms are now referred to as source chronograms.
+Total distance (in units of millions of years) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Figure \ref{fig:workflow}).
+The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a ``phylo'' object or newick string.
+Finally, the patristic matrices are associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.
 
@@
 It is constructed from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert phylogenetic knowledge with rich metadata (McTavish et al., 2015) that allows automatic and reproducible construction of a chronogram database.
-New chronograms can be added to Phylesystem by any user and are immediately publicly available, and the DateLife database can be updated to include those new data within a run.
+New chronograms can be added to Phylesystem by any user and are immediately publicly available.
+This permits an immediate update of DateLife's chronogram database to include those new data within a run.
 
@@
 \item
-  The ages of the root of all matching pruned chronograms.
-This can correspond to the age of the most recent common ancestor (mrca) of your group of interest if the pruned chronograms have all taxa belonging to the group.
+  The ages of the root of all source chronograms.
+This can correspond to the age of the most recent common ancestor (mrca) of your group of interest if the source chronograms have all taxa belonging to the group.
 If not, the root corresponds to the mrca of a subgroup withing your group of interest.
@@
 \item
-  A report of input taxon names matches across pruned chronograms.
+  A report of input taxon names matches across source chronograms.
 \item
-  The single matching pruned chronogram with the most input taxon names.
+  The source chronogram(s) with the most input taxon names.
+\item
+  Last but not least, age data from source chronograms are summarized and used to generate a single summary chronogram:
 \end{enumerate}
 
-\emph{\textbf{Identifying groves.--}} To generate a single summary chronogram, the DateLife algorithm starts by identifying the matching pruned chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
+\emph{\textbf{Choosing a topology.--}} DateLife requires a tree topology to summarize age data upon.
+We recommend that users provide one as input from the literature, or one of their own making.
+If no topology is provided, DateLife automatically subsets one from the OpenTree synthetic tree (Open Tree Of Life et al., 2019).
+Alternatively, DateLife can reconstruct a tree with branch lengths proportional to substitution rates from a starting tree topology using genetic data from the Barcode of Life Data System {[}BOLD; Ratnasingham and Hebert (2007){]}, or combine topologies from source chronogras using a supertree approach.
+
+\emph{\textbf{Reconstructing branch lengths.--}} DateLife starts by mining the BOLD database to obtain genetic markers for the input taxa, and aligning them with MUSCLE (Edgar, 2004; or MAFFT Katoh, Asimenos, \& Toh, 2009).
+Currently, branch length reconstruction is performed with parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed (Schliep, 2011).
+
+\emph{\textbf{Combining source chronograms.--}} To combine topologies from source chronograms into a single summary topology, the DateLife algorithm starts by identifying the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
 (2009).
@@
 By default, DateLife chooses the grove with the most taxa, however, the ``criterion = trees'' flag allows the user to choose the grove with the most trees instead.
-
-\emph{\textbf{Choosing a topology.--}} DateLife requires a tree topology to summarize age data upon.
-Users can provide one as input from the literature, or one of their own making.
-If no topology is provided, DateLife automatically subsets one from the OpenTree synthetic tree (Open Tree Of Life et al., 2019).
-
-DateLife can also reconstruct branch lengths proportional to substitution rates on a fixed tree topology using available genetic data from BOLD.
-
-\emph{\textbf{Congruifying nodes.--}} DateLife then implements the congruification method (Eastman, Harmon, \& Tank, 2013) to find nodes belonging to the same clade across matching pruned chronograms.
-Congruified node ages stored as a \texttt{congruifiedCalibrations} object are then matched to nodes in the chosen tree topology and stored as a \texttt{matchedCalibrations} object.
-
-\emph{\textbf{Summarizing node ages.--}} DateLife summarizes matched calibrations
-into a single patristic distance matrix using different methods.
-Summarizing options implemented include Super Distance Matrix method (SDM, Criscuolo, Berry, Douzery, \& Gascuel, 2006) and summary statistics such as median, minimum and maximum ages.
-
-\emph{\textbf{Dating the tree topology.--}} Summarized calibrations can be applied as secondary calibrations with different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Campbell O.
-Webb, Ackerly, \& Kembel, 2008; Campbell O.
+If source chronograms do not form a grove, the supertree reonstruction will fail.
+
+\emph{\textbf{Congruifying nodes.--}} Once with a chosen topology, DateLife applies the congruification method (Eastman, Harmon, \& Tank, 2013) to find nodes belonging to the same clade across source chronograms, and extract the corresponding node ages from the patristic distance matrices stored as \texttt{datelifeResult}.
+Since by definition the matrices store total distance (time from tip to tip), node ages correspond to half the values stored in the patristic distance matrices.
+A table of congruified node ages that can be used as secondary calibrations is stored as a \texttt{congruifiedCalibrations} object.
+
+\emph{\textbf{Summarizing congruified ages.--}} Congruified pairwise node ages are used to calculate summary statistics (mean, median, minimum and maximum ages) per taxon pair.
+into a single summary matrix using different methods.
+Summarizing options implemented include Super Distance Matrix method (SDM, Criscuolo, Berry, Douzery, \& Gascuel, 2006) and .
+
+Ages per taxon pair can be used as calibrations to date the chosen tree topology.
+
+\emph{\textbf{Dating the tree topology.--}}
+By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully datd topology.
+BLADJ fixes node ages that have calibration data, and distrivutes time between nodes with no data evenly between nodes with calibration data.
+This minimizes age variance in the resulting chronogram (Campbell O.
+Webb, Ackerly, \& Kembel, 2008).
+When there is conflict in ages between nodes with calibration data, the algorithm ignores ages that are older than ages of parent nodes and/or younger than ages from descendant nodes.
+
+If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature.
+If none is provided, DateLife assigns an arbitrary age to the root as 10\% older than the oldest age available within the group.
+
+Summarized calibrations can be applied as secondary calibrations with different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Campbell O.
+Webb et al., 2008; Campbell O.
 Webb \& Donoghue, 2005), and treePL (Stephen A.
 Smith \& O'Meara, 2012).
-
-By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm that assigns ages to nodes with no data evenly between nodes with age data, which minimizes age variance in the resulting chronogram (Campbell O.
-Webb et al., 2008).
-When there is conflict in ages across node with age data, the algorithm ignores ages that are older than parent nodes and/or younger than descendant nodes.
-
-If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature.
-If none is provided, DateLife assigns an arbitrary age to the root as 10\% older than the oldest age available within the group.
 
@@
 Query processing time increases roughly linearly with number of input taxon names, and
-increases considerably if TNRS is activated.
+increases considerably if Taxonomic Name Resolution Service (TNRS) is activated.
 Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
@@
 
-Processing input names found that \emph{Emberiza elegans} is synonym for \emph{Schoeniclus elegans} in the reference taxonomy.
+Processing input names found that \emph{Emberiza elegans} is synonym for \emph{Schoeniclus elegans} in the default reference taxonomy {[}Open Tree of Life Taxonomy v3.3, June 1, 2021{]}.
+For a detailed discussion on the state of the synonym refer to Avibase (Avibase, 2022; Lepage, 2004; Lepage, Vaidya, \& Guralnick, 2014).
 DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies.
@@
 As expected, more inclusive nodes (e.g., node ``n1'') have more age data than less inclusive nodes (e.g., node ``n5'').
-The processing step allowd discovering five data points for node ``n4'' that would not have had any data otherwise.
+The processing step allowed discovering five data points for node ``n4'' that would not have had any data otherwise.
 Age summary statistics per node were calculated and tested as secondary calibrations to date the tree topology using the BLADJ algorithm.
@@
 
-Data from source chronograms can be used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates (Figs.
+As a form of cross validation, we took the tree topologies from each study and used time of lineage divergence data from all other source chronograms as fixed calibrations with the program BLADJ (Figs.
 \ref{fig:cv1} to \ref{fig:cv10}).
-As a form of cross validation, we took tree topologies from each input study and calibrated them using time of lineage divergence data from all other source chronograms.
-
-In the absence of branch lengths, the ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvbladj}).
+
+The ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvbladj}).
 Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cv2}).
-We also demonstrate the usage of PATHd8 (Britton et al., 2007) as an dating method alternative to BLADJ.
-For this, we run a \texttt{datelife} branch length reconstruction that searches for DNA sequence data from the Barcode of Life Data System {[}BOLD; Ratnasingham and Hebert (2007){]} to generate branch lengths.
+We also demonstrate the usage of PATHd8 (Britton et al., 2007) as a dating method alternative to BLADJ.
+PATHd8 requires a tree with initial branch lengths which we obtained using \texttt{datelife}'s branch length reconstruction algorithm, that searches for DNA sequence data from the Barcode of Life Data System {[}BOLD; Ratnasingham and Hebert (2007){]} to generate branch lengths.
 We were able to successfully generate a tree with BOLD branch lengths for all of the Fringillidae source chronograms.
@@
 
-At the time of writing of this manuscript (Apr 04, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
+At the time of writing of this manuscript (Apr 07, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
 A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time.
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-avibase-emberiza}{}}%
+Avibase.
+(2022).
+{Yellow-throated Bunting}.
+\emph{{Avibase - The World Bird Database}}, (Online Resource).
+Retrieved from \url{\%7Bhttps://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\%7D}
+
 \leavevmode\vadjust pre{\hypertarget{ref-Bapst2012a}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-edgar2004muscle}{}}%
+Edgar, R. C. (2004). MUSCLE: Multiple sequence alignment with high accuracy and high throughput. \emph{Nucleic Acids Research}, \emph{32}(5), 1792--1797.
+
 \leavevmode\vadjust pre{\hypertarget{ref-Felsenstein1985a}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-katoh2009multiple}{}}%
+Katoh, K., Asimenos, G., \& Toh, H. (2009). Multiple alignment of DNA sequences with MAFFT. In \emph{Bioinformatics for DNA sequence analysis} (pp. 39--64). Springer.
+
 \leavevmode\vadjust pre{\hypertarget{ref-laubichler2009form}{}}%
 Laubichler, M. D., \& Maienschein, J. (2009). \emph{Form and function in developmental evolution}. Cambridge University Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lepage2004avibase}{}}%
+Lepage, D. (2004). \emph{Avibase: The world bird database}. Bird Studies Canada.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lepage2014avibase}{}}%
+Lepage, D., Vaidya, G., \& Guralnick, R. (2014). Avibase--a database system for managing and organizing taxonomic concepts. \emph{ZooKeys}, (420), 117.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-schliep2011phangorn}{}}%
+Schliep, K. P. (2011). Phangorn: Phylogenetic analysis in r. \emph{Bioinformatics}, \emph{27}(4), 592--593.
+
 \leavevmode\vadjust pre{\hypertarget{ref-smith2018constructing}{}}%

### 2022-04-08 manuscript/manuscript.Rmd
cross validation XY figure

@@
 <!--Data from source chronograms can be used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates-->
-As a form of cross validation, we took the tree topologies from each study and used time of lineage divergence data from all other source chronograms as fixed calibrations with the program BLADJ (Figs.
-\ref{fig:cv1} to \ref{fig:cv10}).
+To perform a cross validation analysis of the DateLife workflow, we took tree topologies from source chronograms and used age data from all other source chronograms as fixed calibrations with the program BLADJ (Fig.
+\ref{fig:cvXY}).
 <!--I'm confused about why you are talking about branch lengths here...
@@
 
-The ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvbladj}).
-Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cv2}).
+*I still need to double check results to confirm or change any of the following descriptions:*
+
+The ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvXY}).
+Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cvXY}).
 <!-- We need to think about the root! is it comparable across source chronograms?? not all of them have the same sampling... so probably not the same root and thus not comparable. Address this -->

### 2022-04-08 manuscript/manuscript.tex
cross validation XY figure

@@
 \shorttitle{DateLife: revealing the dated Tree of Life}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4379}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4391}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 
-As a form of cross validation, we took the tree topologies from each study and used time of lineage divergence data from all other source chronograms as fixed calibrations with the program BLADJ (Figs.
-\ref{fig:cv1} to \ref{fig:cv10}).
-
-The ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvbladj}).
-Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cv2}).
+To perform a cross validation analysis of the DateLife workflow, we took tree topologies from source chronograms and used age data from all other source chronograms as fixed calibrations with the program BLADJ (Fig.
+\ref{fig:cvXY}).
+
+\emph{I still need to double check results to confirm or change any of the following descriptions:}
+
+The ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvXY}).
+Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cvXY}).
 We also demonstrate the usage of PATHd8 (Britton et al., 2007) as a dating method alternative to BLADJ.
@@
 \begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/cross_validation_1.png}
-\caption{Cross validation of first source chronogram.
-The chronogram shown in black corresponds to the dates published in the original study.
-The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
-\label{fig:cv1}
+\includegraphics{../figures/figure-cross-validation/fig-cross-validation-xy-plots.pdf}
+\caption{Results from cross validation analysis.}
+\label{fig:cvXY}
 \end{figure}
@@
 
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/cross_validation_2.png}
-\caption{Cross validation of second source chronogram.
-The chronogram shown in black corresponds to the dates published in the original study.
-The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
-\label{fig:cv2}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/cross_validation_3.png}
-\caption{Cross validation of third source chronogram.
-The chronogram shown in black corresponds to the dates published in the original study.
-The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
-\label{fig:cv3}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/cross_validation_4.png}
-\caption{Cross validation of fourth source chronogram.
-The chronogram shown in black corresponds to the dates published in the original study.
-The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
-\label{fig:cv4}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/cross_validation_6.png}
-\caption{Cross validation of sixth source chronogram.
-The chronogram shown in black corresponds to the dates published in the original study.
-The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
-\label{fig:cv6}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/cross_validation_7.png}
-\caption{Cross validation of seventh source chronogram.
-The chronogram shown in black corresponds to the dates published in the original study.
-The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
-\label{fig:cv7}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/cross_validation_8.png}
-\caption{Cross validation of eight source chronogram.
-The chronogram shown in black corresponds to the dates published in the original study.
-The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
-\label{fig:cv8}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/cross_validation_9.png}
-\caption{Cross validation of ninth source chronogram.
-The chronogram shown in black corresponds to the dates published in the original study.
-The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
-\label{fig:cv9}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/cross_validation_10.png}
-\caption{Cross validation of tenth source chronogram.
-The chronogram shown in black corresponds to the dates published in the original study.
-The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
-\label{fig:cv10}
-\end{figure}
-
 % \end{linenumbers}

### 2022-04-08 R/cross-validation.R

@@ -1,0 +1,56 @@
+# Function based on https://github.com/LunaSare/datelife_examples/blob/master/R/functions_data.R#L3
+# get calibrations form all other chronograms not belonging to the same study
+congruify_other_calibrations <- function(phylo_all, index) {
+    # dd <- duplicated(names(phyloall))
+    cat(index, " -- ", names(phylo_all)[index], "\n")
+    study <- names(phylo_all) %in% names(phylo_all)[index]
+    res <- datelife::congruify_calibrations(phy = phylo_all[study][[1]],
+                                     chronograms = phylo_all[!study])
+    names(res) <- names(phylo_all)[index]
+    res
+}
+
+congruify_other_calibrations_all <- function(phylo_all){
+    # dd <- duplicated(names(phyloall))
+    res <- lapply(seq(phylo_all), function(i){
+        cat(i, " -- ", names(phylo_all)[i], "\n")
+        dd <- names(phylo_all) %in% names(phylo_all)[i]
+        datelife::congruify_calibrations(phy = phylo_all[dd][[1]],
+                                         chronograms = phylo_all[!dd])
+    })
+    names(res) <- names(phylo_all)
+    res
+}
+# Based on function use_othercals3 from https://github.com/LunaSare/datelife_examples/blob/master/R/functions_data.R#L36
+use_other_calibrations <- function(trees, other_calibrations, ...){
+    res <- lapply(seq(trees), function(i){
+      phy <- trees[[i]]
+      phy$edge.length <- NULL
+      print(i)
+        xx <- suppressMessages(suppressWarnings(datelife::use_calibrations_bladj(phy,
+            calibrations = other_calibrations[[i]], ...)))
+        return(xx)
+    })
+    class(res) <- "multiPhylo"
+    names(res) <- names(trees)
+    res
+}
+# Function to date source chronograms with data from other chronograms
+cross_date <- function(frin_other_calibs, index) {
+  cat(index, " -- ", names(frin_other_calibs)[index], "\n")
+  
+  calibrations <- frin_other_calibs[[index]]
+  
+  if (length(calibrations) == 1) {
+    message("... Source chronogram has no calibrations.")
+    return(NA)
+  }
+  
+  mm <- calibrations$matched_calibrations
+  mm <- mm[mm$congruent,]
+  mm <- mm[!is.na(mm$mrca_node_number),]
+  calibrations$matched_calibrations <- mm
[…]

### 2022-04-11 R/cross-validation.R
clean figure 1

@@ -6,6 +6,7 @@ congruify_other_calibrations <- function(phylo_all, index) {
     study <- names(phylo_all) %in% names(phylo_all)[index]
     res <- datelife::congruify_calibrations(phy = phylo_all[study][[1]],
                                      chronograms = phylo_all[!study])
+    res <- list(res)
     names(res) <- names(phylo_all)[index]
     res
 }

### 2022-04-14 manuscript/cross-validation-pathd8.Rmd
cross validation results and figures

--- 2022-04-07 manuscript/manuscript.Rmd
@@
 ---
-title             : "DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life"
-shorttitle        : "DateLife: revealing the dated Tree of Life"
-
-author:
-  - name          : "Luna L. Sánchez Reyes"
-    affiliation   : "1,2"
-    corresponding : yes    # Define only one corresponding author
-    address       : ""
-    email         : "sanchez.reyes.luna@gmail.com"
-    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
-      - Data curation
-      - Investigation
-      - Software
-      - Visualization
-      - Validation
-      - Writing - Original Draft Preparation
-      - Writing - Review & Editing
-  - name          : "Emily Jane McTavish"
-    affiliation   : "1"
-    role:
-      - Resources
-      - Software
-      - Writing - Review & Editing
-  - name          : "Brian O'Meara"
-    affiliation   : "2"
-    role:
-      - Conceptualization
-      - Funding acquisition
-      - Methodology
-      - Resources
-      - Software
-      - Supervision
-      - Writing - Review & Editing
-
-affiliation:
-  - id            : "1"
-    institution   : "University of California, Merced"
-  - id            : "2"
-    institution   : "University of Tennessee, Knoxville"
-
-authornote: |
-  School of Natural Sciences, University of California, Merced, Science and Engineering Building 1.
-
-  Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 425 Hesler Biology Building, Knoxville, TN 37996, USA.
-
-abstract: |
-  Date estimates for times of evolutionary divergences are key data for research in the natural sciences.
-These estimates also provide valuable information for for education, science communication and policy decisions.
-  Although achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is a difficult and time-consuming task, the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
-  When these estimates are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on time of evolutionary origin is exposed in a programatic and reusable way.
-  Intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
-  Here we present DateLife, a service implemented as an R package and an Rshiny website application available at www.datelife.org/query/, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
-  The main DateLife workflow constructs a chronogram for any given combination of taxon names, by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from TreeBASE database as well.
-  We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
-  DateLife will be useful to increase awereness on the existing variation in expert time of divergence data, and can foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.
-
-  <!--  R has become mainstream in the biological-data analyst toolkit.
-Hence, to improve accessibility of time of evolutionary origin data we developed an R package that facilitates interaction, reanalysis and reuse of it, as well as incorporation of these data into the evolutionary analysis workflow.
--->
-
-bibliography      : ["paper_references.bib"]
-keywords          : "Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations"
-wordcount         : "`r wordcountaddin::word_count()`"
-floatsintext      : no
-figurelist        : no
-tablelist         : no
-footnotelist      : no
-linenumbers       : yes
-mask              : no
-draft             : no
-
-documentclass     : "apa6"
-classoption       : "man"
-output            :
-  papaja::apa6_pdf:
-    includes      :
-      in_header: "preamble.tex"
-      after_body  : "figures-alt.tex"
-
-nocite: |
-  @barker2013going, @barker2015new, @burns2014phylogenetics, @claramunt2015new, @gibb2015new, @Hedges2015, @hooper2017chromosomal, @Jetz2012, @price2014niche
+title: "cross-validation-pathd8"
+output: html_document
 ---
 
-```{r setup, include = FALSE}
-library("papaja")
-r_refs("paper_references.bib")
-```
-
-```{r analysis-preferences}
-# Seed for random number generation
-set.seed(50)
-knitr::opts_chunk$set(cache.extra = knitr::rand_seed)
-data(opentree_chronograms, package = "datelife")
-rpackage_available <- TRUE
-availability <- ifelse(rpackage_available, "is available", "will be available")
-load("../data/dquery.rda")
-load("../data/dres.rda")
-```
-
-# Introduction
-<!-- If you're aiming for syst bio you don't want to start so general Readers know what chronograms and phylogenies are!.-->
-Chronograms --phylogenies with branch lengths proportional to time-- provide key data for the study of natural processes in many areas of biological research, such as developmental biology [@laubichler2009form; @delsuc2018phylogenomic], conservation biology [@Felsenstein1985a; @Webb2000], historical biogeography [@posadas2006historical], and species diversification [@magallon2001absolute; @Morlon2014].
-
-Building a chronogram is not an easy task.
-It requires obtaining and curating data to construct a phylogeny; selecting and placing appropriate calibrations on the phylogeny using independent age data points from the fossil record or other dated events, and inferring the full dated tree.
-Estimating accurate chronograms generally requires specialized biological training, taxonomic domain knowledge, and a non-negligible amount of research time, computational resources and funding.
-
-Here we present the DateLife software application, available as an R package and as an online Rshiny interactive website at www.datelife.org/query/, which captures data from published chronograms, and make these data readily accessible to users.
-DateLife features a versioned, open and fully public chronogram database [@mctavish2015phylesystem] storing age information in a computer readable format [@vos2012nexml], an automated and programmatic way of accessing the data [@Stoltzfus2013] and methods to summarize and compare age data.
-
-# Description
-
-The DateLife algorithm is fully implemented using the R language.
-The latest stable version of the R package `datelife` is available from the CRAN repository (v`r packageVersion("datelife")`; @datelifeR), and relies on functionalities from various biological R packages:
-ape [@paradis2004],
-bold [@Chamberlain2018],
-geiger [@Harmon2008],
-paleotree [@Bapst2012a],
-phyloch [@Heibl2008],
-phylocomr [@Ooms2018],
-phytools [@Revell2012],
-rotl [@Michonneau2016], and
-taxize [@Chamberlain2013; @Chamberlain2018].
-Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife algorithm: providing an input, searching a chronogram database, and summarizing results from the search.
-
-<!--_**Providing an input.--**_ -->
-
-## Providing an input
-
-DateLife starts with an input query consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree.
-If the input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004].
-The input tree is not required to have branch lengths, and its topology is used in the summary steps described below.
-
-DateLife accepts scientific names as input.
-These names can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or binomial specific.
-Subspecies and variants are ignored.
-If an input taxon name belongs to an inclusive taxonomic group the algorithm has two alternative behaviors defined by the "get species from taxon" flag.
-If the flag is active, the DateLife algorithm retrieves all species names within the inclusive taxonomic group and adds them to the input.
-If the flag is inactive, DateLife ignores the inclusive taxon names from the input.
-
-Input scientific names are processed using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database.
-TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard.
-DateLife implements TNRS using OpenTree's taxonomy as standard [@opentreeAPIs; @rees2017automated].
-<!--_I reworded a bit, and removed a Boyle citation because it sounded like you were using their TNRS - but you aren't right?  -->
-
-The processed input taxon names are saved as an R object of a newly defined class \texttt{datelifeQuery} that is used in the following steps.
-This object contains the processed names, the corresponding OpenTree taxonomic id numbers, and the topology of the input tree if any was provided.
-
-<!--_**Searching the database.--**_  -->
-
-## Searching the database
-
-A DateLife search consists of matching processed taxon names to tip labels in a chronogram database.
-Chronograms with at least two matching tip labels are identified and pruned down to preserve only the matched tips.
-These matching pruned chronograms are now referred to as source chronograms.
-Total distance (in units of millions of years) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Figure \ref{fig:workflow}).
-The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a "phylo" object or newick string.
-Finally, the patristic matrices are associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.
-
-DateLife's chronogram database latest version consist of `r length(opentree_chronograms$trees)` chronograms published in `r length(unique(opentree_chronograms$studies))` different studies.
-It is constructed from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert phylogenetic knowledge with rich metadata [@mctavish2015phylesystem] that allows automatic and reproducible construction of a chronogram database.
-New chronograms can be added to Phylesystem by any user and are immediately publicly available.
-This permits an immediate update of DateLife's chronogram database to include those new data within a run.
-
-<!-- _**Summarizing search results.--**_ -->
-
-## Summarizing search results
-
-Summary information is extracted from the \texttt{datelifeResult} object to inform decisions for the subsequent steps in the user workflow. Basic summary information available to the user is:
-
-1. The matching pruned chronograms as newick strings or "phylo" objects.
-2.
-The ages of the root of all source chronograms.
-This can correspond to the age of the most recent common ancestor (mrca) of your group of interest if the source chronograms have all taxa belonging to the group.
-If not, the root corresponds to the mrca of a subgroup withing your group of interest.
-3. Study citations where original chronograms were published.
-4. A report of input taxon names matches across source chronograms.
-5. The source chronogram(s) with the most input taxon names.
-6. Last but not least, age data from source chronograms are summarized and used to generate a single summary chronogram:
-
-_**Choosing a topology.--**_ DateLife requires a tree topology to summarize age data upon.
-We recommend that users provide one as input from the literature, or one of their own making.
-If no topology is provided, DateLife automatically subsets one from the OpenTree synthetic tree [@opentreeoflife2019synth].
-Alternatively, DateLife can reconstruct a tree with branch lengths proportional to substitution rates from a starting tree topology using genetic data from the Barcode of Life Data System [BOLD; @ratnasingham2007bold], or combine topologies from source chronogras using a supertree approach.
-
-_**Reconstructing branch lengths.--**_ DateLife starts by mining the BOLD database to obtain genetic markers for the input taxa, and aligning them with MUSCLE [@edgar2004muscle; or MAFFT @katoh2009multiple].
-Currently, branch length reconstruction is performed with parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed [@schliep2011phangorn].
-
-_**Combining source chronograms.--**_ To combine topologies from source chronograms into a single summary topology, the DateLife algorithm starts by identifying the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
-[-@ane2009groves].
-In rare cases, a group of trees can have multiple groves.
-By default, DateLife chooses the grove with the most taxa, however, the "criterion = trees" flag allows the user to choose the grove with the most trees instead.
-If source chronograms do not form a grove, the supertree reonstruction will fail.
-
-_**Congruifying nodes.--**_ Once with a chosen topology, DateLife applies the congruification method [@Eastman2013] to find nodes belonging to the same clade across source chronograms, and extract the corresponding node ages from the patristic distance matrices stored as `datelifeResult`.
-Since by definition the matrices store total distance (time from tip to tip), node ages correspond to half the values stored in the patristic distance matrices.
-A table of congruified node ages that can be used as secondary calibrations is stored as a `congruifiedCalibrations` object.
-
-_**Summarizing congruified ages.--**_ Congruified pairwise node ages are used to calculate summary statistics (mean, median, minimum and maximum ages) per taxon pair.
-<!-- we should calculate summary statistics per node -->
-<!--??? How is 'calibrations' defined here? the calibrations from congruification, or the pairwise distances??? is calibrations from congruification-->
-into a single summary matrix using different methods.
-Summarizing options implemented include Super Distance Matrix method [SDM, @Criscuolo2006] and .
-
-Ages per taxon pair can be used as calibrations to date the chosen tree topology.
-<!--
-For each node in the output tree topology, the pairwise distances that traverse that node are ..???
-Under the SDM criterion ??? what happens??
-Using the minimum age criterion ..??
--->
-
-_**Dating the tree topology.--**_ 
-By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully datd topology.
-BLADJ fixes node ages that have calibration data, and distrivutes time between nodes with no data evenly between nodes with calibration data.
-This minimizes age variance in the resulting chronogram [@Webb2008].
-When there is conflict in ages between nodes with calibration data, the algorithm ignores ages that are older than ages of parent nodes and/or younger than ages from descendant nodes.
-
-If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature.
-If none is provided, DateLife assigns an arbitrary age to the root as 10% older than the oldest age available within the group.
-
-Summarized calibrations can be applied as secondary calibrations with different dating methods currently supported within DateLife: MrBayes [@Huelsenbeck2001; @Ronquist2003], PATHd8 [@Britton2007], BLADJ [@Webb2008; @webb2005phylomatic], and treePL [@Smith2012].
-<!--Huh? can you run MrBayes in DateLife? You should explain how the BOLD dating with PathD8 works here, and what is needed for the other ones!-->
-
-<!--PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees.
-treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method
-[@Smith2012].
-MrBayes program [@Huelsenbeck2001; @Ronquist2003] can be used when adding taxa at
-random, following a reference taxonomy or a topological constraint. It draws ages
-from a pure birth model, as implemented by Jetz and collaborators [-@Jetz2012].
-
-If there is no information available for some input taxon name, users can also create age and phylogenetic data for the missing branches with MrBayes.
--->
-
-_**Visualizing results.--**_
-Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions.
-
-<!-- TODO ADD FIGURE: mock example explaining difference between using summary ages and raw ages to calibrate a topology.
-Then show it in the biological example.
-This will show what happens if a tree (phylogenetic conflict) or node (age) do not agree.-->
-
-# Benchmark
-
-`datelife`'s code speed was tested on an Apple iMac
-with one 3.4 GHz Intel Core i5 processor. <!--No multiprocessor tests were performed-->
-We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
-Query processing time increases roughly linearly with number of input taxon names, and
-increases considerably if Taxonomic Name Resolution Service (TNRS) is activated.
-<!-- Defining tnrs here again, bc we only talk about it two times, so it easy to forget it -->
-Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
-Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig.
-\ref{fig:runtime_main}).
-<!--Speed of other datelife workflows? NO, bc it reflects speed from other software mainly. But I will put a sentence about it...
-Summarizing `datelife` results processing times: shows sdm speed, clustering methods speed;
-generating new trees: shows bladj speed;
-Adding dates processing time: reflects more mrbayes speed than anything;
-get_bold_otol_tree running time: reflects bold downloading data speed.-->
-<!--Speed with different number of chronograms in database? NO-->
-<!--Speed of web interface and of r package in computers with different capacities? NOOOO-->
-`datelife`'s code performance was evaluated with a set of unit tests designed and
-implemented with the R package testthat [@RCoreTeam2018] that were run both locally
-with the devtools package [@RCoreTeam2018], and on a public server --via
-GitHub, using the continuous integration tool Travis CI (<https://travis-ci.org>). At
-present, unit tests cover more than 40% of `datelife`'s code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->
-
-# Case study
-
-We illustrate the DateLife algorithm using a group within the Passeriform birds encompassing the family of true finches, Fringillidae and allies as case study.
-The first example analyses 6 bird species and shows all steps of the algorithm.
-The second example is a real life application
-<!--We used `r cite_r("r-references.bib")` for all our analyses.-->
-
-## Small example
-
-We randomly chose 6 bird species related to the family Fringillidae of true finches. The sample includes
-two species of cardinals:
-the black-thighed grosbeak -- *Pheucticus tibialis* and
-the crimson-collared grosbeak -- *Rhodothraupis celaeno*;
-three species of buntings:
-the yellowhammer -- *Emberiza citrinella*,
-the pine bunting --  *Emberiza leucocephalos* and
-the yellow-throated bunting --  *Emberiza elegans*;
-and one species of tanager, the vegetarian finch -- *Platyspiza crassirostris*.
-
-Processing input names found that *Emberiza elegans* is synonym for *Schoeniclus elegans* in the default reference taxonomy [Open Tree of Life Taxonomy v3.3, June 1, 2021].
-For a detailed discussion on the state of the synonym refer to Avibase [@lepage2004avibase; @lepage2014avibase; @avibase-emberiza].
-DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies.
-Three studies matched five input names [@barker2015new; @Hedges2015; @Jetz2012], one study matched four input names [@hooper2017chromosomal] and two studies matched two input names [@barker2013going; @burns2014phylogenetics].
-No studies matched all input names.
-Together, matching chronograms have 28 unique age data points.
-All nodes have age data.
-As fixed tree topology, DateLife used OpenTree's synthetic tree as default and mapped age data to nodes in the tree.
-As expected, more inclusive nodes (e.g., node "n1") have more age data than less inclusive nodes (e.g., node "n5").
-The processing step allowed discovering five data points for node "n4" that would not have had any data otherwise.
-Age summary statistics per node were calculated and tested as secondary calibrations to date the tree topology using the BLADJ algorithm.
-Age data for node "n2" was excluded as final calibration because it is older than age data of a more inclusive node.
-
-## Real life application
-
-A college educator wishes to obtain state-of-the-art data on time of evolutionary origin of species belonging to the true finches for their class.
-They decide to use `datelife` because they are teaching best practices for reproducibility.
-Students have the option to go to the website at www.datelife.org and perform an interactive run. However, the educator also wants the students to practice their R skills.
-The first step is to run a `datelife` query using the "get species from taxon" flag.
-This will get all recognised species names within their chosen inclusive taxon.
-The Fringillidae has `r length(dquery$cleaned_names)` species, according to the Open Tree of Life taxonomy.
-Once with a curated set of species taxon names, the next step is to run a `datelife` search that will find all chronograms that contain at least two species names.
-The algorithm proceeds to prune the trees to keep matching species names on tips only, and transform the pruned trees to pairwise distance matrices.
- There are `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
-The final step is to summarize the available information using two alternative types of summary chronograms, median and SDM.
-As explained in the "Description" section, data from source chronograms is first summarised into a single distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig.
-\ref{fig:fringillidages}).
-Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
-????????????????? Say some things about the results!
-<!-- (Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
-<!-- and hierarchical clustering method which appears to push ages back -->
-
-# Cross-validation test
-
-<!--Data from source chronograms can be used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates-->
-As a form of cross validation, we took the tree topologies from each study and used time of lineage divergence data from all other source chronograms as fixed calibrations with the program BLADJ (Figs.
-\ref{fig:cv1} to \ref{fig:cv10}).
-<!--I'm confused about why you are talking about branch lengths here...
-did you use branch lengths?
-Yes, on the PATHd8 cross validation, but you are right, it is confusing, so I removed that first sentence-->
-
-The ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvbladj}).
-Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cv2}).
-<!-- We need to think about the root! is it comparable across source chronograms?? not all of them have the same sampling... so probably not the same root and thus not comparable. Address this -->
 We also demonstrate the usage of PATHd8 [@Britton2007] as a dating method alternative to BLADJ.
@@
 More examples and code used to generate these trees were developed on an open repository that is available for consultation and reuse at <https://github.com/LunaSare/datelife_examples>.
-
-# Discussion
-
-The main goal of `datelife` is to make state-of-the-art information on time of lineage divergence easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
-It is an open service that does not require any expert biological knowledge from users --besides the names of the organisms they want to work with, for any of its functionality.
-
-At the time of writing of this manuscript (`r format(Sys.time(), "%b %d, %Y")`), `datelife`'s database has `r length(opentree_chronograms$trees)` chronograms, pulled entirely from OpenTree's database, the Phylesystem [@mctavish2015phylesystem].
-A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time.
-As chronograms are added to Phylesystem, they are incorporated into an updated `datelife`'s database that is assigned a new version number, followed by a package release on CRAN.
-`datelife`'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
-Users can also upload new chronograms to OpenTree themselves, and trigger an update of their local `datelife` database to incorporate the new chronograms, to have them immediately available for analysis.
-
-Incorporation of more chronograms into `datelife`'s database is crucial to improve its services.
-One option to increase chronogram number in the database is the Dryad data repository.
-Methods to automatically mine chronograms from Dryad could be designed and implemented.
-However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined manually by a curator.
-
-The largest, and taxonomically broadest, summary chronogram currently available from OpenTree
-was constructed using age data from 2,274 published chronograms [@Hedges2015].
-However the source chronograms used as input data for this tree are not available in computer readable format for reuse or reanalysis.
-As this tree is part of datelife's database, the amount of lineages that can be queried using `datelife` (`r length(unique(unname(unlist(lapply(opentree_chronograms$trees, "[[", "tip.label")))))` unique terminal taxa) is substantial.
-Access to the input chronograms used to generate the Hedges et al.
-summary tree would improve measures of uncertainty in DateLife, but they are available only as image files and not as usable data (timetree.org).
-<!--IN VIOLATION OF NSF RULES! but I don't suppose we should say that.-->
-We would like to emphasize on the importance of sharing chronogram data for the benefit of the scientific community as a whole, into repositories that require expert input and manual curation, such as OpenTree's Phylesystem [@mctavish2015phylesystem].
-
-By default, `datelife` currently summarizes all source chronograms that overlap with at least two species names. Users can exclude source chronograms if they have reasons to do so.
-Strictly speaking, the best chronogram should reflect the real time of lineage divergence accurately and precisely.
-To our knowledge, there are no good measures to determine independently if a chronogram is better than another.
-Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used [@magallon2010using, @magallon2015metacalibrated].
-Several characteristics of the data used for dating analyses as well as from the output chronogram itself, could be used to score quality of source chronograms.
-Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
-DateLife provides an opportunity to capture concordance and conflict among date estimates, which can also be used as a metric for chronogram reliability.
-
-Scientists usually also favor chronograms constructed using primary calibrations (ages obtained from the fossil or geological record) to ones constructed with secondary calibrations (ages coming from other chronograms)[@schenk2016sec].
-It has been observed with simulations that divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with Bayesian inference methods when priors are implemented in similar ways in both analyses [@schenk2016sec].
-However, secondary calibrations
-can be applied using other dating methods that do not require setting priors, such as penalized likelihood [@sanderson2003r8s], or as fixed ages,
-potentially mitigating the bias reported with Bayesian methods.
-Certainly, further studies are required to fully understand the effect of using secondary calibrations on time estimates and downstream analyses.
-
-Furthermore, even chronograms obtained with primary fossil data can vary substantially in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
-<!--I think that one very odd one didn't have any fossils though! Maybe worth mentioning here?-->
-This observation is often encountered in the literature
-<!-- Different source chronograms often show substantial variation in clade ages  -->
-(see, for example, the ongoing debate about crown group age of angiosperms [@sauquet2021age; @barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time].
-For some studies, especially ones based on branch lengths (e.g., studies of species diversification, timing of evolutionary events, phenotypic trait evolution), using a different chronogram may return different results [@title2016macrophylogenies].
-Stitching together these chronograms can create a larger tree that uses information from multiple studies, but the effect of uncertainties and errors at this level on downstream analyses is still largely unknown.
-
-Summarizing chronograms might also imply summarizing fundamentally distinct evolutionary hypotheses.
-For example, two different researchers working on the same clade both carefully select and argument their choices of fossil calibrations.
-Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses the same one to calibrate outside the clade, the resulting age estimates will often differ substantially, as the placement of calibrations as stem or crown group is proved to deeply affect estimated times of lineage divergence [@sauquet2013practical].
-Trying to summarize the resulting chronograms into a single one using simple summary statistics can erase many types of relevant information from the source chronograms.
-Accordingly, the prevailing view is that we should favor time of lineage divergence estimates obtained from a single analysis, using fossil data as primary sources of calibrations, and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history [@antonelli2017supersmart].
-<!-- Will this be implemented in future `datelife` versions?
-Probably not-->
-However, the exercise of summarizing different chronograms has the potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypothesis.
-Choosing the elements of the chronograms that we are going to keep and the ones that we are going to discard is key, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram [@sauquet2021age].
-<!-- Example -->
-
-<!-- still requires more research. -->
-<!-- Often, there is a large discrepancy in time of lineage divergence estimates across studies from expert knowledge, that is well known and documented for different groups of organisms [e.g., angiosperms; @magallon2015metacalibrated].
--->
-<!-- Comparison of available age data for a wide range of organisms shown here suggest that this is a widespread phenomenon that requires further attention. -->
-<!-- In the case of the example presented here, most chronograms are new, meaning that we expect that data was generated with high quality and the software used should be very advanced.
-However, this is just an assumption.
--->
-
-<!-- Still, even if all source chronograms have been generated by excellent standards and using similar methods, the evolutionary history they depict might be very different.  -->
-
-Nonetheless, in ecology and conservation biology, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance [@Webb2008].
-Hence, we integrated into datelife's workflow different ways of estimating node ages in the absence of calibrations and branch length information for taxa lacking this information.
-"Making up" branch lengths is an accepted practice in scientific publications: @Jetz2012, created a time-calibrated tree of all 9,993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a tree of 353,185 seed plants where only 23% had molecular data.
-Obviously, there are risks in this practice!
-Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch estimation methods.
-There has yet to be a thorough analysis of what can go wrong when one extends inferences beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make carefully consider the statistical assumptions being made, and assess the consistency of the results with prior work.
-
-<!-- There is evidence that different chronograms will result in very different macroevolutionary histories inferred from them [@title2016macrophylogenies].
-However, it is unknown how using very different age data could affect hypothesis testing in other areas of research and more effort is needed in that directions -->
-
-# Conclusions
-
-Divergence time information is key to many areas of evolutionary studies: trait evolution,
-diversification, biogeography, macroecology and more. It is also crucial for science communication and education, but generating chronograms is difficult,
-especially for those who want to use phylogenies but who are not systematists, or
-do not have the time to acquire and develop the necessary knowledge and data curation skills.
-Moreover, years of primarily public funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
-
-<!--Mention again uses shown here and outline potential uses not shown here-->
-The `datelife` R package allows easy and fast summarization of publicly available information
-on time of lineage divergence.
-This provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identification of regions that require more research or that have conflicting information.
-It is available as an R package, or a web-based R shiny app at dates.opentreeloflife.org/datelife.
-<!-- It can also summarize public data, or use it to generate new chronograms based on trees provided by the user. -->
-Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
-The DateLife project helps with awareness of the existing variation in expert time of divergence data, and will foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
-
-# Availability
-
-`datelife` is free and open source and it can be used through its current website
-<http://www.datelife.org/query/>, through its R package, and through Phylotastic's project web portal <http://phylo.cs.nmsu.edu:3000/>.
-`datelife`'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
-`datelife`'s R package stable version `r availability`
-for installation from the CRAN repository (<https://cran.r-project.org/package=datelife>)
-using the command `install.packages(pkgs = "datelife")` from within R. Development versions
-are available from the GitHub repository (<https://github.com/phylotastic/datelife>)
-and can be installed using the command `devtools::install_github("phylotastic/datelife")`.
-
-# Supplementary Material
-
-Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities are available at [datelifeMS1](https://github.com/LunaSare/datelifeMS1), [datelife_examples](https://github.com/LunaSare/datelife_examples), and [datelife_benchmark](https://github.com/LunaSare/datelife_benchmark) repositories in LLSR's GitHub account.
-
-# Funding
-
-Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603  to Datelife project and DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), and the Phylotastic project Grant ABI-1458572, and the OpenTree grant ABI-1759846.
-
-# Acknowledgements
-
-The DateLife project was born as a prototype tool aiming to provide these services, and was developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA [@Stoltzfus2013].
-We thank colleagues from the O'Meara Lab at the University
-of Tennesse Knoxville for suggestions, discussions and software testing.
-The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
-that led to initial work on this project.
-The team that assembled `datelife`'s first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
-The Open Tree of Life project that provides the open, metadata rich repository of
-trees used for `datelife`.
-The many scientists who publish their chronograms in an open, reusable form, and
-the scientists who curate them for deposition in the Open Tree of Life repository.
-The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.
-
-\newpage
-
-# References
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-<div id="refs" custom-style="Bibliography"></div>
-\endgroup

### 2022-04-14 manuscript/manuscript.tex
cross validation results and figures

--- 2022-04-07 manuscript/manuscript.tex
@@
 \shorttitle{DateLife: revealing the dated Tree of Life}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4379}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4298}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 The DateLife algorithm is fully implemented using the R language.
-The latest stable version of the R package \texttt{datelife} is available from the CRAN repository (v0.6.2; Sanchez-Reyes et al.
-(2022)), and relies on functionalities from various biological R packages:
+The latest stable version of the R package \texttt{datelife} -- v0.6.2, is available from the CRAN repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
 ape (Paradis, Claude, \& Strimmer, 2004),
@@
 If no topology is provided, DateLife automatically subsets one from the OpenTree synthetic tree (Open Tree Of Life et al., 2019).
-Alternatively, DateLife can reconstruct a tree with branch lengths proportional to substitution rates from a starting tree topology using genetic data from the Barcode of Life Data System {[}BOLD; Ratnasingham and Hebert (2007){]}, or combine topologies from source chronogras using a supertree approach.
-
-\emph{\textbf{Reconstructing branch lengths.--}} DateLife starts by mining the BOLD database to obtain genetic markers for the input taxa, and aligning them with MUSCLE (Edgar, 2004; or MAFFT Katoh, Asimenos, \& Toh, 2009).
+Alternatively, DateLife can reconstruct a tree with branch lengths proportional to substitution rates from a starting tree topology using genetic data from the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007), or combine topologies from source chronogras using a supertree approach.
+
+\emph{\textbf{Reconstructing branch lengths.--}} DateLife starts by mining the BOLD database to obtain genetic markers for the input taxa, and aligning them with MUSCLE (Edgar, 2004; or MAFFT, Katoh, Asimenos, \& Toh, 2009).
 Currently, branch length reconstruction is performed with parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed (Schliep, 2011).
@@
 into a single summary matrix using different methods.
-Summarizing options implemented include Super Distance Matrix method (SDM, Criscuolo, Berry, Douzery, \& Gascuel, 2006) and .
+Summarizing options implemented include Super Distance Matrix method (SDM, Criscuolo, Berry, Douzery, \& Gascuel, 2006) and summary statistics.
 
@@
 
-\hypertarget{case-study}{%
-\section{Case study}\label{case-study}}
+\hypertarget{case-studies}{%
+\section{Case studies}\label{case-studies}}
 
@@
 Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
-????????????????? Say some things about the results!
 
@@
 
-As a form of cross validation, we took the tree topologies from each study and used time of lineage divergence data from all other source chronograms as fixed calibrations with the program BLADJ (Figs.
-\ref{fig:cv1} to \ref{fig:cv10}).
-
-The ages of internal nodes were recovered with a high precision in almost all cases (except for studies 3, and 5; Fig. \ref{fig:cvbladj}).
-Maximum tree ages were only recovered in one case (study 2; Fig. \ref{fig:cv2}).
-We also demonstrate the usage of PATHd8 (Britton et al., 2007) as a dating method alternative to BLADJ.
-PATHd8 requires a tree with initial branch lengths which we obtained using \texttt{datelife}'s branch length reconstruction algorithm, that searches for DNA sequence data from the Barcode of Life Data System {[}BOLD; Ratnasingham and Hebert (2007){]} to generate branch lengths.
-We were able to successfully generate a tree with BOLD branch lengths for all of the Fringillidae source chronograms.
-However, dating with PATHd8 using congruified calibrations, was only successful in
-three cases (studies 3, 5, and 9, shown in Fig. \ref{fig:cvbold}).
-From these, two trees have a different sampling than the original source chronogram, mainly because DNA BOLD data for some species is absent from the database.
-Maximum ages are quite different from source chronograms, but this might be explained also by the differences in sampling between source chronograms and BOLD trees.
-More examples and code used to generate these trees were developed on an open repository that is available for consultation and reuse at \url{https://github.com/LunaSare/datelife_examples}.
+To perform a cross validation analysis of the DateLife workflow, we used resulting data from the previous section (Casestudy: Real Life Application).
+We took individual tree topologies from each of the 19 source chronograms found (Supplementary data XX).
+Then we congruifued age data of source chronograms from studies, and used this ages to date the tree topology with the program BLADJ.
+
+We found that node ages from original study, and ages estimated using all other age data available are generally correlated (Supplementary Fig. \ref{fig:cvXY}).
+In 5 studies, more inclusive nodes have older original ages, and less inclusive nodes have younger original ages than their cross-validated age estimates.
+Accordingly, root ages are generally older in the original study than estimated using cross-validated ages.
+Root ages were similar in original and cross validated ages in three vases (Supplementary Fig.
+\ref{fig:cvXYdiffs}).
+Notably, chronograms have different species sampling, hence roots are not comparable across studies.
+Yet, chronograms with a higher sampling number can potentially inform the age of the root of chronograms with less sampling.
 
@@
 
-At the time of writing of this manuscript (Apr 07, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
+At the time of writing of this manuscript (Apr 14, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
 A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time.
@@
 \begin{figure}[!h]
-\includegraphics{../figures/figure1/figure1-vertical-final.pdf}
+\includegraphics{../figures/figure1/figure1-new.pdf}
 \caption{Stylized DateLife workflow.
@@
 
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/cross_validation_1.png}
-\caption{Cross validation of first source chronogram.
-The chronogram shown in black corresponds to the dates published in the original study.
-The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
-\label{fig:cv1}
-\end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:cvbold}}
-% \end{center}
+% \end{linenumbers}
+
+% in this file I leave the figure captions outside\ caption{} because I want them
+% to be formatted in the same way as the general text (double spaced and linenumbered)
+\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}
+
+\begin{figure}[!h]
+\includegraphics{../figures/figure-cross-validation/fig-cross-validation-xy-plots.pdf}
+\caption{Results from cross validation analysis.}
+\label{fig:cvXY}
+\end{figure}
+
+\begin{figure}[!h]
+\includegraphics{../figures/figure-cross-validation/fig-cross-validation-xy-plots-diffs.pdf}
+\caption{Results from cross validation analysis.}
+\label{fig:cvXYdiffs}
+\end{figure}
 
@@
 
-% \end{linenumbers}
-
 \end{document}

### 2022-04-16 manuscript/manuscript.Rmd
adding comments

--- 2022-04-06 manuscript/manuscript.Rmd
@@
 Building a chronogram is not an easy task.
-It requires obtaining and curating data to construct a phylogeny; selecting and placing appropriate calibrations on the phylogeny using independent age data points from the fossil record or other dated events, and inferring the full dated tree.
-Estimating accurate chronograms generally requires specialized biological training, taxonomic domain knowledge, and a non-negligible amount of research time, computational resources and funding.
+It requires obtaining and curating data to construct a phylogeny, selecting and placing appropriate calibrations on the phylogeny using independent age data points from the fossil record or other dated events, and inferring the full dated tree;
+it also generally requires specialized biological training, taxonomic domain knowledge, and a non-negligible amount of research time, computational resources and funding.
 
-Here we present the DateLife software application, available as an R package and as an online Rshiny interactive website at www.datelife.org/query/, which captures data from published chronograms, and make these data readily accessible to users.
-DateLife features a versioned, open and fully public chronogram database [@mctavish2015phylesystem] storing age information in a computer readable format [@vos2012nexml], an automated and programmatic way of accessing the data [@Stoltzfus2013] and methods to summarize and compare age data.
+Here we present the DateLife project and its core software application, available as an R package [@datelifeR], and as an online Rshiny interactive website at www.datelife.org/query/, which captures data from published chronograms, and make these data readily accessible to users for reuse and reanalysis.
+The software features key elements for scientific reproducibility, such as a versioned, open and fully public chronogram database [@mctavish2015phylesystem], age data stored in a computer readable format [@vos2012nexml], automated and programmatic ways of accessing the data [@Stoltzfus2013] and methods to summarize and compare age data.
 
@@
 
-The DateLife algorithm is fully implemented using the R language.
-The latest stable version of the R package `datelife` -- v`r packageVersion("datelife")`, is available from the CRAN repository [@datelifeR], and relies on functionalities from various biological R packages:
+DateLife's core software application consists of the R package `datelife`.
+Its latest stable version -- v`r packageVersion("datelife")`, is available from the CRAN repository [@datelifeR], and relies on functionalities from various biological R packages:
 ape [@paradis2004],
@@
 
-## Providing an input
+## Processing an input
 
-DateLife starts with an input query consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree.
+DateLife starts by processing an input consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree.
 If the input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004].
@@
 
-DateLife accepts scientific names as input.
-These names can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or binomial specific.
+DateLife accepts scientific names that can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or a binomial specific.
 Subspecies and variants are ignored.
 If an input taxon name belongs to an inclusive taxonomic group the algorithm has two alternative behaviors defined by the "get species from taxon" flag.
-If the flag is active, the DateLife algorithm retrieves all species names within the inclusive taxonomic group and adds them to the input.
-If the flag is inactive, DateLife ignores the inclusive taxon names from the input.
+If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group (according to a taxonomy) and adds them to the input string.
+If the flag is inactive, DateLife excludes the taxon names above the species level from the input.
 
-Input scientific names are processed using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database.
-TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard.
-DateLife implements TNRS using OpenTree's taxonomy as standard [@opentreeAPIs; @rees2017automated].
-<!--_I reworded a bit, and removed a Boyle citation because it sounded like you were using their TNRS - but you aren't right?  -->
+DateLife processes input scientific names using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database.
+TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard [@boyle2013].
+DateLife implements TNRS using OpenTree's unified taxonomy as standard [@opentreeAPIs; @rees2017automated], storing OpenTree's Taxonomy identification numbers for further processing.
+<!-- We are using OpenTree's TNRS, Boyle described TNRS first, right?  -->
 
 The processed input taxon names are saved as an R object of a newly defined class \texttt{datelifeQuery} that is used in the following steps.
-This object contains the processed names, the corresponding OpenTree taxonomic id numbers, and the topology of the input tree if any was provided.
+This object contains the standardized names, the corresponding taxonomic id numbers, and the topology of the input tree if any was provided.
 
@@
 
-A DateLife search consists of matching processed taxon names to tip labels in a chronogram database.
-Chronograms with at least two matching tip labels are identified and pruned down to preserve only the matched tips.
-These matching pruned chronograms are now referred to as source chronograms.
+DateLife's chronogram database latest version consist of `r length(opentree_chronograms$trees)` chronograms published in `r length(unique(opentree_chronograms$studies))` different studies.
+It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata [@mctavish2015phylesystem], which allows automatic and reproducible assembly of our chronogram database.
+Datelife's chronogram database is navigable as an R data object within the `datelife` R package.
+Published chronograms can be added to Phylesystem by any user, at any time, and are immediately publicly available (https://tree.opentreeoflife.org/curator).
+This facilitates an immediate update of DateLife's chronogram database to include new chronogram data on a following search.
+
+A DateLife search is implemented by matching processed taxon names provided by the user, to tip labels in the chronogram database.
+Chronograms with at least two matching taxon names on their tip labels are identified and pruned down to preserve only the matched taxa.
+These matching pruned chronograms are referred to as source chronograms.
 Total distance (in units of millions of years) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Figure \ref{fig:workflow}).
@@
 Finally, the patristic matrices are associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.
-
-DateLife's chronogram database latest version consist of `r length(opentree_chronograms$trees)` chronograms published in `r length(unique(opentree_chronograms$studies))` different studies.
-It is constructed from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert phylogenetic knowledge with rich metadata [@mctavish2015phylesystem] that allows automatic and reproducible construction of a chronogram database.
-New chronograms can be added to Phylesystem by any user and are immediately publicly available.
-This permits an immediate update of DateLife's chronogram database to include those new data within a run.
 
@@
 
-Summary information is extracted from the \texttt{datelifeResult} object to inform decisions for the subsequent steps in the user workflow. Basic summary information available to the user is:
+Summary information is extracted from the \texttt{datelifeResult} object to inform decisions for subsequent steps in the analysis workflow. Basic summary information available to the user is:
 
@@
 The ages of the root of all source chronograms.
-This can correspond to the age of the most recent common ancestor (mrca) of your group of interest if the source chronograms have all taxa belonging to the group.
-If not, the root corresponds to the mrca of a subgroup withing your group of interest.
+These ages can correspond to the age of the most recent common ancestor (mrca) of the user's group of interest if the source chronograms have all taxa belonging to the group.
+If not, the root corresponds to the mrca of a subgroup withing the group of interest.
 3. Study citations where original chronograms were published.
@@
 5. The source chronogram(s) with the most input taxon names.
-6. Last but not least, age data from source chronograms are summarized and used to generate a single summary chronogram:
+6. Last but not least, single summary chronograms resulting from summarizing age data from available source chronograms, as follows:
 
 _**Choosing a topology.--**_ DateLife requires a tree topology to summarize age data upon.
-We recommend that users provide one as input from the literature, or one of their own making.
+We recommend that users provide a tree topology as input from the literature, or one of their own making.
 If no topology is provided, DateLife automatically subsets one from the OpenTree synthetic tree [@opentreeoflife2019synth].
-Alternatively, DateLife can reconstruct a tree with branch lengths proportional to substitution rates from a starting tree topology using genetic data from the Barcode of Life Data System, BOLD [@ratnasingham2007bold], or combine topologies from source chronogras using a supertree approach.
+Alternatively, DateLife can reconstruct a tree with branch lengths proportional to substitution rates from a starting tree topology using genetic data from the Barcode of Life Data System, BOLD [@ratnasingham2007bold], or combine topologies from source chronograms using a supertree approach.
 
@@
 
-_**Combining source chronograms.--**_ To combine topologies from source chronograms into a single summary topology, the DateLife algorithm starts by identifying the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
+_**Combining source chronograms.--**_ To combine topologies from source chronograms into a single summary (or supertree) topology, the DateLife algorithm starts by identifying the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
 [-@ane2009groves].
@@
 By default, DateLife chooses the grove with the most taxa, however, the "criterion = trees" flag allows the user to choose the grove with the most trees instead.
-If source chronograms do not form a grove, the supertree reonstruction will fail.
+If source chronograms do not form a grove, the supertree reconstruction will fail.
 
 _**Congruifying nodes.--**_ Once with a chosen topology, DateLife applies the congruification method [@Eastman2013] to find nodes belonging to the same clade across source chronograms, and extract the corresponding node ages from the patristic distance matrices stored as `datelifeResult`.
-Since by definition the matrices store total distance (time from tip to tip), node ages correspond to half the values stored in the patristic distance matrices.
+By definition, the matrices store total distance (time from tip to tip), hence, node ages correspond to half the values stored in the patristic distance matrices.
 A table of congruified node ages that can be used as secondary calibrations is stored as a `congruifiedCalibrations` object.
 
-_**Summarizing congruified ages.--**_ Congruified pairwise node ages are used to calculate summary statistics (mean, median, minimum and maximum ages) per taxon pair.
+_**Summarizing congruified ages.--**_ 
 <!-- we should calculate summary statistics per node -->
 <!--??? How is 'calibrations' defined here? the calibrations from congruification, or the pairwise distances??? is calibrations from congruification-->
-into a single summary matrix using different methods.
-Summarizing options implemented include Super Distance Matrix method [SDM, @Criscuolo2006] and summary statistics.
-
-Ages per taxon pair can be used as calibrations to date the chosen tree topology.
+For each congruent node, the pairwise distance that traverse that node are summarized into a single summary matrix using classic statistics (mean, median, minimum and maximum ages).
+These single summary ages per taxon pair are then used as secondary calibrations to date the tree topology.
 <!--
-For each node in the output tree topology, the pairwise distances that traverse that node are ..???
 Under the SDM criterion ??? what happens??
-Using the minimum age criterion ..??
+Summarizing options implemented include Super Distance Matrix method [SDM, @Criscuolo2006] 
 -->
@@
 _**Dating the tree topology.--**_ 
-By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully datd topology.
-BLADJ fixes node ages that have calibration data, and distrivutes time between nodes with no data evenly between nodes with calibration data.
+By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully dated topology.
+BLADJ fixes node ages that have calibration data, and distributes time between nodes with no data evenly between nodes with calibration data.
 This minimizes age variance in the resulting chronogram [@Webb2008].
@@
 _**Visualizing results.--**_
-Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions.
+Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as visualize and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions.
 
@@
 
-Furthermore, even chronograms obtained with primary fossil data can vary substantially in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
-<!--I think that one very odd one didn't have any fossils though! Maybe worth mentioning here?-->
+Furthermore, chronograms can be obtained with primary fossil data or with molecular substitution rates obtained experimentally, which can deepen the already substantial variation in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
+<!--I think that one very odd one didn't have any fossils though! Maybe worth mentioning here? 
+Yes!!-->
 This observation is often encountered in the literature
-<!-- Different source chronograms often show substantial variation in clade ages  -->
 (see, for example, the ongoing debate about crown group age of angiosperms [@sauquet2021age; @barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time].

### 2022-05-05 manuscript/manuscript.Rmd
addressing EJM comments

@@
   Intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
-  Here we present DateLife, a service implemented as an R package and an Rshiny website application available at www.datelife.org/query/, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
+  Here we present DateLife, a service implemented as an R package and an Rshiny website application available at www.datelife.org, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
   The main DateLife workflow constructs a chronogram for any given combination of taxon names, by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from TreeBASE database as well.
   We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
-  DateLife will be useful to increase awereness on the existing variation in expert time of divergence data, and can foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.
+  DateLife will be useful to increase awereness of the existing variation in expert time of divergence data, and can foster exploration of the effect of alternative divergence time hypotheses on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.
 
@@
 
-Here we present the DateLife project and its core software application, available as an R package [@datelifeR], and as an online Rshiny interactive website at www.datelife.org/query/, which captures data from published chronograms, and make these data readily accessible to users for reuse and reanalysis.
+Here we present the DateLife project and its core software application, available as an R package [@datelifeR], and as an online Rshiny interactive website at www.datelife.org, which captures data from published chronograms, and make these data readily accessible to users for reuse and reanalysis.
 The software features key elements for scientific reproducibility, such as a versioned, open and fully public chronogram database [@mctavish2015phylesystem], age data stored in a computer readable format [@vos2012nexml], automated and programmatic ways of accessing the data [@Stoltzfus2013] and methods to summarize and compare age data.
@@
 DateLife processes input scientific names using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database.
-TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard [@boyle2013].
+TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard [@Boyle2013].
 DateLife implements TNRS using OpenTree's unified taxonomy as standard [@opentreeAPIs; @rees2017automated], storing OpenTree's Taxonomy identification numbers for further processing.
@@
 5. The source chronogram(s) with the most input taxon names.
-6. Last but not least, single summary chronograms resulting from summarizing age data from available source chronograms, as follows:
+6. Various single summary chronograms resulting from summarizing age data, generated using the methodology described below.
 
-_**Choosing a topology.--**_ DateLife requires a tree topology to summarize age data upon.
+_**Choosing a topology.--**_ 
+DateLife requires a tree topology to summarize age data upon.
 We recommend that users provide a tree topology as input from the literature, or one of their own making.
 If no topology is provided, DateLife automatically subsets one from the OpenTree synthetic tree [@opentreeoflife2019synth].
-Alternatively, DateLife can reconstruct a tree with branch lengths proportional to substitution rates from a starting tree topology using genetic data from the Barcode of Life Data System, BOLD [@ratnasingham2007bold], or combine topologies from source chronograms using a supertree approach.
-
-_**Reconstructing branch lengths.--**_ DateLife starts by mining the BOLD database to obtain genetic markers for the input taxa, and aligning them with MUSCLE [@edgar2004muscle; or MAFFT, @katoh2009multiple].
-Currently, branch length reconstruction is performed with parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed [@schliep2011phangorn].
-
-_**Combining source chronograms.--**_ To combine topologies from source chronograms into a single summary (or supertree) topology, the DateLife algorithm starts by identifying the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
+Alternatively, DateLife can combine topologies from source chronograms using a supertree approach.
+To combine topologies from source chronograms into a single summary (or supertree) topology, the DateLife algorithm starts by identifying the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
 [-@ane2009groves].
@@
 
-_**Congruifying nodes.--**_ Once with a chosen topology, DateLife applies the congruification method [@Eastman2013] to find nodes belonging to the same clade across source chronograms, and extract the corresponding node ages from the patristic distance matrices stored as `datelifeResult`.
+_**Dating the topology.--**_ 
+Input topologies from OpenTree or the supertree approach described above do not include branch length estimates of any kind. 
+To estimate branch lengths proportional to substitution rates on these topologies, DateLife can mine the Barcode of Life Data System, BOLD [@ratnasingham2007bold] to obtain genetic markers for the input taxa.
+These markers are aligned with MUSCLE by default [@edgar2004muscle], or MAFFT [@katoh2009multiple]. 
+This alignment can be used to estimate branch lengths on input topologies that lack branch lengths.
+Currently, branch length reconstruction in DateLife is performed using parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed [@schliep2011phangorn].
+While relative branch length information provides additional data for nodes without secondary date calibrations, topologies without branch lengths can also be dated.
+
+Once a topology is chosen, DateLife applies the congruification method [@Eastman2013] to find nodes belonging to the same clade across source chronograms, and extract the corresponding node ages from the patristic distance matrices stored as `datelifeResult`.
 By definition, the matrices store total distance (time from tip to tip), hence, node ages correspond to half the values stored in the patristic distance matrices.
@@
 
-_**Summarizing congruified ages.--**_ 
-<!-- we should calculate summary statistics per node -->
-<!--??? How is 'calibrations' defined here? the calibrations from congruification, or the pairwise distances??? is calibrations from congruification-->
-For each congruent node, the pairwise distance that traverse that node are summarized into a single summary matrix using classic statistics (mean, median, minimum and maximum ages).
-These single summary ages per taxon pair are then used as secondary calibrations to date the tree topology.
-<!--
-Under the SDM criterion ??? what happens??
-Summarizing options implemented include Super Distance Matrix method [SDM, @Criscuolo2006] 
--->
+For each congruent node, the pairwise distance that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method [SDM; @Criscuolo2006], which deforms patristic distance matrices by minimizing variance and then averaging them.
+These single summary taxon pair age matrices (Summarized calibrations) can be applied as secondary calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes [@Huelsenbeck2001; @Ronquist2003], PATHd8 [@Britton2007], BLADJ [@Webb2008; @webb2005phylomatic], and treePL [@Smith2012].
 
-_**Dating the tree topology.--**_ 
 By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully dated topology.
@@
 This minimizes age variance in the resulting chronogram [@Webb2008].
-When there is conflict in ages between nodes with calibration data, the algorithm ignores ages that are older than ages of parent nodes and/or younger than ages from descendant nodes.
+BLADJ does not incorporate branch lengths even when they are present.
+When there is conflict in ages between nodes with calibration data, the algorithm ignores ages that are older than ages of parent nodes and younger than ages from descendant nodes. 
+> NOTE: do you know how it decides which one to drop if two calibrations conflict?
+The parent or the child?
+It would drop both.
+It basically drops any calibration for any given node that are older than any calibration of the parent node, and younger than any calibration of children nodes.
 
@@
 
-Summarized calibrations can be applied as secondary calibrations with different dating methods currently supported within DateLife: MrBayes [@Huelsenbeck2001; @Ronquist2003], PATHd8 [@Britton2007], BLADJ [@Webb2008; @webb2005phylomatic], and treePL [@Smith2012].
-<!--Huh? can you run MrBayes in DateLife? You should explain how the BOLD dating with PathD8 works here, and what is needed for the other ones!-->
-
-<!--PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees.
+Alternative options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the secondary calibrations.
+PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees.
 treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method
 [@Smith2012].
-MrBayes program [@Huelsenbeck2001; @Ronquist2003] can be used when adding taxa at
-random, following a reference taxonomy or a topological constraint. It draws ages
-from a pure birth model, as implemented by Jetz and collaborators [-@Jetz2012].
-
-If there is no information available for some input taxon name, users can also create age and phylogenetic data for the missing branches with MrBayes.
--->
+The MrBayes  [@Huelsenbeck2001; @Ronquist2003] approach in datelife uses the secondary calibrations as priors on node ages. 
+> NOTE: is this true?? Yes
 
 _**Visualizing results.--**_
-Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as visualize and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions.
+Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as visualize and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions available from the R package `datelifeplot` [@datelifeplot].
 
@@
 `datelife` is free and open source and it can be used through its current website
-<http://www.datelife.org/query/>, through its R package, and through Phylotastic's project web portal <http://phylo.cs.nmsu.edu:3000/>.
+<http://www.datelife.org>, through the `datelife` R package, and through Phylotastic's project web portal <https://phylo.cs.nmsu.edu/>.
 `datelife`'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
@@
 
-Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603  to Datelife project and DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), and the Phylotastic project Grant ABI-1458572, and the OpenTree grant ABI-1759846.
+Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603  to Datelife project; DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), ABI-1458572 to the Phylotastic project, and ABI-1759846 to the Open Tree of Life project.
 

### 2022-05-05 manuscript/manuscript.tex
addressing EJM comments

@@
 
+% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
+% See https://github.com/crsh/papaja/issues/292
+\makeatletter
+\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
+  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
+  {-1em}%
+  {\normalfont\normalsize\bfseries\itshape\typesectitle}}
+
+\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
+  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
+  {-\z@\relax}%
+  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
+\makeatother
+
 % \usepackage{etoolbox}
@@
 \makeatother
-\shorttitle{DateLife: revealing the dated Tree of Life}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4298}
+
+\usepackage{xpatch}
+\makeatletter
+\xapptocmd\appendix
+  {\xapptocmd\section
+    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
+    {}{\InnerPatchFailed}%
+  }
+{}{\PatchFailed}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4540}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 
+\shorttitle{DateLife: revealing the dated Tree of Life}
+
 \authornote{
@@
 
-\abstract{
+\abstract{%
 Date estimates for times of evolutionary divergences are key data for research in the natural sciences.
@@
 Intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
-Here we present DateLife, a service implemented as an R package and an Rshiny website application available at www.datelife.org/query/, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
+Here we present DateLife, a service implemented as an R package and an Rshiny website application available at www.datelife.org, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
 The main DateLife workflow constructs a chronogram for any given combination of taxon names, by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from TreeBASE database as well.
 We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
-DateLife will be useful to increase awereness on the existing variation in expert time of divergence data, and can foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.
+DateLife will be useful to increase awereness of the existing variation in expert time of divergence data, and can foster exploration of the effect of alternative divergence time hypotheses on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.
 }
@@
 Building a chronogram is not an easy task.
-It requires obtaining and curating data to construct a phylogeny; selecting and placing appropriate calibrations on the phylogeny using independent age data points from the fossil record or other dated events, and inferring the full dated tree.
-Estimating accurate chronograms generally requires specialized biological training, taxonomic domain knowledge, and a non-negligible amount of research time, computational resources and funding.
-
-Here we present the DateLife software application, available as an R package and as an online Rshiny interactive website at www.datelife.org/query/, which captures data from published chronograms, and make these data readily accessible to users.
-DateLife features a versioned, open and fully public chronogram database (McTavish et al., 2015) storing age information in a computer readable format (Vos et al., 2012), an automated and programmatic way of accessing the data (Stoltzfus et al., 2013) and methods to summarize and compare age data.
+It requires obtaining and curating data to construct a phylogeny, selecting and placing appropriate calibrations on the phylogeny using independent age data points from the fossil record or other dated events, and inferring the full dated tree;
+it also generally requires specialized biological training, taxonomic domain knowledge, and a non-negligible amount of research time, computational resources and funding.
+
+Here we present the DateLife project and its core software application, available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org, which captures data from published chronograms, and make these data readily accessible to users for reuse and reanalysis.
+The software features key elements for scientific reproducibility, such as a versioned, open and fully public chronogram database (McTavish et al., 2015), age data stored in a computer readable format (Vos et al., 2012), automated and programmatic ways of accessing the data (Stoltzfus et al., 2013) and methods to summarize and compare age data.
 
@@
 
-The DateLife algorithm is fully implemented using the R language.
-The latest stable version of the R package \texttt{datelife} -- v0.6.2, is available from the CRAN repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
+DateLife's core software application consists of the R package \texttt{datelife}.
+Its latest stable version -- v0.6.2, is available from the CRAN repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
 ape (Paradis, Claude, \& Strimmer, 2004),
@@
 
-\hypertarget{providing-an-input}{%
-\subsection{Providing an input}\label{providing-an-input}}
-
-DateLife starts with an input query consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree.
+\hypertarget{processing-an-input}{%
+\subsection{Processing an input}\label{processing-an-input}}
+
+DateLife starts by processing an input consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree.
 If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis et al., 2004).
@@
 
-DateLife accepts scientific names as input.
-These names can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or binomial specific.
+DateLife accepts scientific names that can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or a binomial specific.
 Subspecies and variants are ignored.
 If an input taxon name belongs to an inclusive taxonomic group the algorithm has two alternative behaviors defined by the ``get species from taxon'' flag.
-If the flag is active, the DateLife algorithm retrieves all species names within the inclusive taxonomic group and adds them to the input.
-If the flag is inactive, DateLife ignores the inclusive taxon names from the input.
-
-Input scientific names are processed using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database.
-TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard.
-DateLife implements TNRS using OpenTree's taxonomy as standard (Open Tree Of Life et al., 2016; Rees \& Cranston, 2017).
+If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group (according to a taxonomy) and adds them to the input string.
+If the flag is inactive, DateLife excludes the taxon names above the species level from the input.
+
+DateLife processes input scientific names using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database.
+TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard (Boyle et al., 2013).
+DateLife implements TNRS using OpenTree's unified taxonomy as standard (Open Tree Of Life et al., 2016; Rees \& Cranston, 2017), storing OpenTree's Taxonomy identification numbers for further processing.
 
 The processed input taxon names are saved as an R object of a newly defined class \texttt{datelifeQuery} that is used in the following steps.
-This object contains the processed names, the corresponding OpenTree taxonomic id numbers, and the topology of the input tree if any was provided.
+This object contains the standardized names, the corresponding taxonomic id numbers, and the topology of the input tree if any was provided.
 
@@
 
-A DateLife search consists of matching processed taxon names to tip labels in a chronogram database.
-Chronograms with at least two matching tip labels are identified and pruned down to preserve only the matched tips.
-These matching pruned chronograms are now referred to as source chronograms.
+DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies.
+It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata (McTavish et al., 2015), which allows automatic and reproducible assembly of our chronogram database.
+Datelife's chronogram database is navigable as an R data object within the \texttt{datelife} R package.
+Published chronograms can be added to Phylesystem by any user, at any time, and are immediately publicly available (\url{https://tree.opentreeoflife.org/curator}).
+This facilitates an immediate update of DateLife's chronogram database to include new chronogram data on a following search.
+
+A DateLife search is implemented by matching processed taxon names provided by the user, to tip labels in the chronogram database.
+Chronograms with at least two matching taxon names on their tip labels are identified and pruned down to preserve only the matched taxa.
+These matching pruned chronograms are referred to as source chronograms.
 Total distance (in units of millions of years) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Figure \ref{fig:workflow}).
@@
 
-DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies.
-It is constructed from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert phylogenetic knowledge with rich metadata (McTavish et al., 2015) that allows automatic and reproducible construction of a chronogram database.
-New chronograms can be added to Phylesystem by any user and are immediately publicly available.
-This permits an immediate update of DateLife's chronogram database to include those new data within a run.
-
 \hypertarget{summarizing-search-results}{%
@@
   The ages of the root of all source chronograms.
-This can correspond to the age of the most recent common ancestor (mrca) of your group of interest if the source chronograms have all taxa belonging to the group.
-If not, the root corresponds to the mrca of a subgroup withing your group of interest.
+These ages can correspond to the age of the most recent common ancestor (mrca) of the user's group of interest if the source chronograms have all taxa belonging to the group.
+If not, the root corresponds to the mrca of a subgroup withing the group of interest.
 \item
@@
 \item
-  Last but not least, age data from source chronograms are summarized and used to generate a single summary chronogram:
+  Various single summary chronograms resulting from summarizing age data, generated using the methodology described below.
 \end{enumerate}
 
-\emph{\textbf{Choosing a topology.--}} DateLife requires a tree topology to summarize age data upon.
-We recommend that users provide one as input from the literature, or one of their own making.
+\emph{\textbf{Choosing a topology.--}}
+DateLife requires a tree topology to summarize age data upon.
+We recommend that users provide a tree topology as input from the literature, or one of their own making.
 If no topology is provided, DateLife automatically subsets one from the OpenTree synthetic tree (Open Tree Of Life et al., 2019).
-Alternatively, DateLife can reconstruct a tree with branch lengths proportional to substitution rates from a starting tree topology using genetic data from the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007), or combine topologies from source chronogras using a supertree approach.
-
-\emph{\textbf{Reconstructing branch lengths.--}} DateLife starts by mining the BOLD database to obtain genetic markers for the input taxa, and aligning them with MUSCLE (Edgar, 2004; or MAFFT, Katoh, Asimenos, \& Toh, 2009).
-Currently, branch length reconstruction is performed with parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed (Schliep, 2011).
-
-\emph{\textbf{Combining source chronograms.--}} To combine topologies from source chronograms into a single summary topology, the DateLife algorithm starts by identifying the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
+Alternatively, DateLife can combine topologies from source chronograms using a supertree approach.
+To combine topologies from source chronograms into a single summary (or supertree) topology, the DateLife algorithm starts by identifying the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
 (2009).
@@
 By default, DateLife chooses the grove with the most taxa, however, the ``criterion = trees'' flag allows the user to choose the grove with the most trees instead.
-If source chronograms do not form a grove, the supertree reonstruction will fail.
-
-\emph{\textbf{Congruifying nodes.--}} Once with a chosen topology, DateLife applies the congruification method (Eastman, Harmon, \& Tank, 2013) to find nodes belonging to the same clade across source chronograms, and extract the corresponding node ages from the patristic distance matrices stored as \texttt{datelifeResult}.
-Since by definition the matrices store total distance (time from tip to tip), node ages correspond to half the values stored in the patristic distance matrices.
+If source chronograms do not form a grove, the supertree reconstruction will fail.
+
+\emph{\textbf{Dating the topology.--}}
+Input topologies from OpenTree or the supertree approach described above do not include branch length estimates of any kind.
+To estimate branch lengths proportional to substitution rates on these topologies, DateLife can mine the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain genetic markers for the input taxa.
+These markers are aligned with MUSCLE by default (Edgar, 2004), or MAFFT (Katoh, Asimenos, \& Toh, 2009).
+This alignment can be used to estimate branch lengths on input topologies that lack branch lengths.
+Currently, branch length reconstruction in DateLife is performed using parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed (Schliep, 2011).
+While relative branch length information provides additional data for nodes without secondary date calibrations, topologies without branch lengths can also be dated.
+
+Once a topology is chosen, DateLife applies the congruification method (Eastman, Harmon, \& Tank, 2013) to find nodes belonging to the same clade across source chronograms, and extract the corresponding node ages from the patristic distance matrices stored as \texttt{datelifeResult}.
+By definition, the matrices store total distance (time from tip to tip), hence, node ages correspond to half the values stored in the patristic distance matrices.
 A table of congruified node ages that can be used as secondary calibrations is stored as a \texttt{congruifiedCalibrations} object.
 
-\emph{\textbf{Summarizing congruified ages.--}} Congruified pairwise node ages are used to calculate summary statistics (mean, median, minimum and maximum ages) per taxon pair.
-into a single summary matrix using different methods.
-Summarizing options implemented include Super Distance Matrix method (SDM, Criscuolo, Berry, Douzery, \& Gascuel, 2006) and summary statistics.
-
-Ages per taxon pair can be used as calibrations to date the chosen tree topology.
-
-\emph{\textbf{Dating the tree topology.--}}
-By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully datd topology.
-BLADJ fixes node ages that have calibration data, and distrivutes time between nodes with no data evenly between nodes with calibration data.
-This minimizes age variance in the resulting chronogram (Campbell O.
-Webb, Ackerly, \& Kembel, 2008).
-When there is conflict in ages between nodes with calibration data, the algorithm ignores ages that are older than ages of parent nodes and/or younger than ages from descendant nodes.
+For each congruent node, the pairwise distance that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method {[}SDM; Criscuolo, Berry, Douzery, and Gascuel (2006){]}, which deforms patristic distance matrices by minimizing variance and then averaging them.
+These single summary taxon pair age matrices (Summarized calibrations) can be applied as secondary calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Campbell O.
+Webb, Ackerly, \& Kembel, 2008; Campbell O.
+Webb \& Donoghue, 2005), and treePL (Stephen A.
+Smith \& O'Meara, 2012).
+
+By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully dated topology.
+BLADJ fixes node ages that have calibration data, and distributes time between nodes with no data evenly between nodes with calibration data.
+This minimizes age variance in the resulting chronogram (Campbell O. Webb et al., 2008).
+BLADJ does not incorporate branch lengths even when they are present.
+When there is conflict in ages between nodes with calibration data, the algorithm ignores ages that are older than ages of parent nodes and younger than ages from descendant nodes.
+\textgreater{} NOTE: do you know how it decides which one to drop if two calibrations conflict?
+The parent or the child?
+It would drop both.
+It basically drops any calibration for any given node that are older than any calibration of the parent node, and younger than any calibration of children nodes.
 
@@
 
-Summarized calibrations can be applied as secondary calibrations with different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Campbell O.
-Webb et al., 2008; Campbell O.
-Webb \& Donoghue, 2005), and treePL (Stephen A.
-Smith \& O'Meara, 2012).
+Alternative options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the secondary calibrations.
+PATHd8 is a non-clock, rate-smoothing method (Britton et al., 2007) to date trees.
+treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method
+(Stephen A. Smith \& O'Meara, 2012).
+The MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003) approach in datelife uses the secondary calibrations as priors on node ages.
+\textgreater{} NOTE: is this true?? Yes
 
 \emph{\textbf{Visualizing results.--}}
-Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R ``phylo'' format), as well as view and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions.
+Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R ``phylo'' format), as well as visualize and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions available from the R package \texttt{datelifeplot} (Sanchez-Reyes \& O'Meara, 2022).
 
@@
 
-At the time of writing of this manuscript (Apr 14, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
+At the time of writing of this manuscript (May 06, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
 A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time.
@@
 
-Furthermore, even chronograms obtained with primary fossil data can vary substantially in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
+Furthermore, chronograms can be obtained with primary fossil data or with molecular substitution rates obtained experimentally, which can deepen the already substantial variation in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
 This observation is often encountered in the literature
@@
 \texttt{datelife} is free and open source and it can be used through its current website
-\url{http://www.datelife.org/query/}, through its R package, and through Phylotastic's project web portal \url{http://phylo.cs.nmsu.edu:3000/}.
+\url{http://www.datelife.org}, through the \texttt{datelife} R package, and through Phylotastic's project web portal \url{https://phylo.cs.nmsu.edu/}.
 \texttt{datelife}'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
@@
 
-Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to Datelife project and DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), and the Phylotastic project Grant ABI-1458572, and the OpenTree grant ABI-1759846.
+Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to Datelife project; DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), ABI-1458572 to the Phylotastic project, and ABI-1759846 to the Open Tree of Life project.
 
@@
 \emph{{The Auk: Ornithological Advances}}, \emph{132}(2), 333--348.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Boyle2013}{}}%
+Boyle, B., Hopkins, N., Lu, Z., Raygoza Garay, J.
+A., Mozzherin, D., Rees, T., \ldots{} Enquist, B.
+J.
+(2013).
+{The taxonomic name resolution service: An online tool for automated standardization of plant names}.
+\emph{{BMC Bioinformatics}}, \emph{14}(1).
+\url{https://doi.org/10.1186/1471-2105-14-16}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-datelifeplot}{}}%
+Sanchez-Reyes, L.
+L., \& O'Meara, B.
+(2022).
+{\texttt{datelifeplot}: Methods to plot chronograms and outputs of the \texttt{datelife} package}.
+\emph{R Package Release V0.2.2}.
+Retrieved from \url{https://zenodo.org/badge/latestdoi/381501451}
+
 \leavevmode\vadjust pre{\hypertarget{ref-datelifeR}{}}%
@@
 (2022).
-{datelife: Scientific Data on Time of Lineage Divergence for Your Taxa}.
-\emph{R Package Version 0.6.2}.
+{\texttt{datelife}: Scientific Data on Time of Lineage Divergence for Your Taxa}.
+\emph{R Package Release V0.6.2}.
 Retrieved from \url{https://doi.org/10.5281/zenodo.593938}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-sauquet2021age}{}}%
-Sauquet, H., Ramírez-Barahona, S., \& Magallón, S. (2021). The age of flowering plants is unknown.
+Sauquet, H., Ramírez-Barahona, S., \& Magallón, S. (2021). \emph{The age of flowering plants is unknown}.
 

### 2022-05-05 manuscript/manuscript.tex
updating pdf

@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4540}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4362}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 Intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
-Here we present DateLife, a service implemented as an R package and an Rshiny website application available at www.datelife.org, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
+Here we present DateLife, a service implemented as an R package and an Rshiny website application available at www.datelife.org/query/, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
 The main DateLife workflow constructs a chronogram for any given combination of taxon names, by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from TreeBASE database as well.
 We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
-DateLife will be useful to increase awereness of the existing variation in expert time of divergence data, and can foster exploration of the effect of alternative divergence time hypotheses on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.
+DateLife will be useful to increase awereness on the existing variation in expert time of divergence data, and can foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.
 }
@@
 
-Here we present the DateLife project and its core software application, available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org, which captures data from published chronograms, and make these data readily accessible to users for reuse and reanalysis.
+Here we present the DateLife project and its core software application, available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org/query/, which captures data from published chronograms, and make these data readily accessible to users for reuse and reanalysis.
 The software features key elements for scientific reproducibility, such as a versioned, open and fully public chronogram database (McTavish et al., 2015), age data stored in a computer readable format (Vos et al., 2012), automated and programmatic ways of accessing the data (Stoltzfus et al., 2013) and methods to summarize and compare age data.
@@
 DateLife processes input scientific names using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database.
-TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard (Boyle et al., 2013).
+TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard (\textbf{boyle2013?}).
 DateLife implements TNRS using OpenTree's unified taxonomy as standard (Open Tree Of Life et al., 2016; Rees \& Cranston, 2017), storing OpenTree's Taxonomy identification numbers for further processing.
@@
 \item
-  Various single summary chronograms resulting from summarizing age data, generated using the methodology described below.
+  Last but not least, single summary chronograms resulting from summarizing age data from available source chronograms, as follows:
 \end{enumerate}
 
-\emph{\textbf{Choosing a topology.--}}
-DateLife requires a tree topology to summarize age data upon.
+\emph{\textbf{Choosing a topology.--}} DateLife requires a tree topology to summarize age data upon.
 We recommend that users provide a tree topology as input from the literature, or one of their own making.
 If no topology is provided, DateLife automatically subsets one from the OpenTree synthetic tree (Open Tree Of Life et al., 2019).
-Alternatively, DateLife can combine topologies from source chronograms using a supertree approach.
-To combine topologies from source chronograms into a single summary (or supertree) topology, the DateLife algorithm starts by identifying the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
+Alternatively, DateLife can reconstruct a tree with branch lengths proportional to substitution rates from a starting tree topology using genetic data from the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007), or combine topologies from source chronograms using a supertree approach.
+
+\emph{\textbf{Reconstructing branch lengths.--}} DateLife starts by mining the BOLD database to obtain genetic markers for the input taxa, and aligning them with MUSCLE (Edgar, 2004; or MAFFT, Katoh, Asimenos, \& Toh, 2009).
+Currently, branch length reconstruction is performed with parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed (Schliep, 2011).
+
+\emph{\textbf{Combining source chronograms.--}} To combine topologies from source chronograms into a single summary (or supertree) topology, the DateLife algorithm starts by identifying the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
 (2009).
@@
 
-\emph{\textbf{Dating the topology.--}}
-Input topologies from OpenTree or the supertree approach described above do not include branch length estimates of any kind.
-To estimate branch lengths proportional to substitution rates on these topologies, DateLife can mine the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain genetic markers for the input taxa.
-These markers are aligned with MUSCLE by default (Edgar, 2004), or MAFFT (Katoh, Asimenos, \& Toh, 2009).
-This alignment can be used to estimate branch lengths on input topologies that lack branch lengths.
-Currently, branch length reconstruction in DateLife is performed using parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed (Schliep, 2011).
-While relative branch length information provides additional data for nodes without secondary date calibrations, topologies without branch lengths can also be dated.
-
-Once a topology is chosen, DateLife applies the congruification method (Eastman, Harmon, \& Tank, 2013) to find nodes belonging to the same clade across source chronograms, and extract the corresponding node ages from the patristic distance matrices stored as \texttt{datelifeResult}.
+\emph{\textbf{Congruifying nodes.--}} Once with a chosen topology, DateLife applies the congruification method (Eastman, Harmon, \& Tank, 2013) to find nodes belonging to the same clade across source chronograms, and extract the corresponding node ages from the patristic distance matrices stored as \texttt{datelifeResult}.
 By definition, the matrices store total distance (time from tip to tip), hence, node ages correspond to half the values stored in the patristic distance matrices.
@@
 
-For each congruent node, the pairwise distance that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method {[}SDM; Criscuolo, Berry, Douzery, and Gascuel (2006){]}, which deforms patristic distance matrices by minimizing variance and then averaging them.
-These single summary taxon pair age matrices (Summarized calibrations) can be applied as secondary calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Campbell O.
-Webb, Ackerly, \& Kembel, 2008; Campbell O.
+\emph{\textbf{Summarizing congruified ages.--}}
+For each congruent node, the pairwise distance that traverse that node are summarized into a single summary matrix using classic statistics (mean, median, minimum and maximum ages).
+These single summary ages per taxon pair are then used as secondary calibrations to date the tree topology.\\
+
+\emph{\textbf{Dating the tree topology.--}}
+By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully dated topology.
+BLADJ fixes node ages that have calibration data, and distributes time between nodes with no data evenly between nodes with calibration data.
+This minimizes age variance in the resulting chronogram (Campbell O.
+Webb, Ackerly, \& Kembel, 2008).
+When there is conflict in ages between nodes with calibration data, the algorithm ignores ages that are older than ages of parent nodes and/or younger than ages from descendant nodes.
+
+If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature.
+If none is provided, DateLife assigns an arbitrary age to the root as 10\% older than the oldest age available within the group.
+
+Summarized calibrations can be applied as secondary calibrations with different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Campbell O.
+Webb et al., 2008; Campbell O.
 Webb \& Donoghue, 2005), and treePL (Stephen A.
@@
 
-By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully dated topology.
-BLADJ fixes node ages that have calibration data, and distributes time between nodes with no data evenly between nodes with calibration data.
-This minimizes age variance in the resulting chronogram (Campbell O. Webb et al., 2008).
-BLADJ does not incorporate branch lengths even when they are present.
-When there is conflict in ages between nodes with calibration data, the algorithm ignores ages that are older than ages of parent nodes and younger than ages from descendant nodes.
-\textgreater{} NOTE: do you know how it decides which one to drop if two calibrations conflict?
-The parent or the child?
-It would drop both.
-It basically drops any calibration for any given node that are older than any calibration of the parent node, and younger than any calibration of children nodes.
-
-If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature.
-If none is provided, DateLife assigns an arbitrary age to the root as 10\% older than the oldest age available within the group.
-
-Alternative options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the secondary calibrations.
-PATHd8 is a non-clock, rate-smoothing method (Britton et al., 2007) to date trees.
-treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method
-(Stephen A. Smith \& O'Meara, 2012).
-The MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003) approach in datelife uses the secondary calibrations as priors on node ages.
-\textgreater{} NOTE: is this true?? Yes
-
 \emph{\textbf{Visualizing results.--}}
-Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R ``phylo'' format), as well as visualize and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions available from the R package \texttt{datelifeplot} (Sanchez-Reyes \& O'Meara, 2022).
+Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R ``phylo'' format), as well as visualize and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions.
 
@@
 
-At the time of writing of this manuscript (May 06, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
+At the time of writing of this manuscript (May 05, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
 A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time.
@@
 \texttt{datelife} is free and open source and it can be used through its current website
-\url{http://www.datelife.org}, through the \texttt{datelife} R package, and through Phylotastic's project web portal \url{https://phylo.cs.nmsu.edu/}.
+\url{http://www.datelife.org/query/}, through its R package, and through Phylotastic's project web portal \url{http://phylo.cs.nmsu.edu:3000/}.
 \texttt{datelife}'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Boyle2013}{}}%
-Boyle, B., Hopkins, N., Lu, Z., Raygoza Garay, J.
-A., Mozzherin, D., Rees, T., \ldots{} Enquist, B.
-J.
-(2013).
-{The taxonomic name resolution service: An online tool for automated standardization of plant names}.
-\emph{{BMC Bioinformatics}}, \emph{14}(1).
-\url{https://doi.org/10.1186/1471-2105-14-16}
-
 \leavevmode\vadjust pre{\hypertarget{ref-Britton2007}{}}%
@@
 Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{{Science Advances}}, \emph{1}(11), e1501005.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Criscuolo2006}{}}%
-Criscuolo, A., Berry, V., Douzery, E.
-J.
-P., \& Gascuel, O.
-(2006).
-{SDM: A fast distance-based approach for (super)tree building in phylogenomics}.
-\emph{{Systematic Biology}}, \emph{55}(5), 740--755.
-\url{https://doi.org/10.1080/10635150600969872}
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-datelifeplot}{}}%
-Sanchez-Reyes, L.
-L., \& O'Meara, B.
-(2022).
-{\texttt{datelifeplot}: Methods to plot chronograms and outputs of the \texttt{datelife} package}.
-\emph{R Package Release V0.2.2}.
-Retrieved from \url{https://zenodo.org/badge/latestdoi/381501451}
-
 \leavevmode\vadjust pre{\hypertarget{ref-datelifeR}{}}%
@@
 (2022).
-{\texttt{datelife}: Scientific Data on Time of Lineage Divergence for Your Taxa}.
-\emph{R Package Release V0.6.2}.
+{datelife: Scientific Data on Time of Lineage Divergence for Your Taxa}.
+\emph{R Package Version 0.6.2}.
 Retrieved from \url{https://doi.org/10.5281/zenodo.593938}

### 2022-05-05 manuscript/manuscript.Rmd
minor restcurture

--- 2022-05-05 manuscript/manuscript.Rmd
@@
 DateLife processes input scientific names using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database.
-TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard [@Boyle2013].
+TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard [@boyle2013].
 DateLife implements TNRS using OpenTree's unified taxonomy as standard [@opentreeAPIs; @rees2017automated], storing OpenTree's Taxonomy identification numbers for further processing.
@@
 5. The source chronogram(s) with the most input taxon names.
-6. Various single summary chronograms resulting from summarizing age data, generated using the methodology described below.
+6. A single summary chronogram resulting from summarizing age data generated using the options described below.
 
@@
 To estimate branch lengths proportional to substitution rates on these topologies, DateLife can mine the Barcode of Life Data System, BOLD [@ratnasingham2007bold] to obtain genetic markers for the input taxa.
-These markers are aligned with MUSCLE by default [@edgar2004muscle], or MAFFT [@katoh2009multiple]. 
+These markers are aligned with MUSCLE [@edgar2004muscle; or MAFFT, @katoh2009multiple]. 
 This alignment can be used to estimate branch lengths on input topologies that lack branch lengths.
-Currently, branch length reconstruction in DateLife is performed using parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed [@schliep2011phangorn].
+Currently, branch length reconstruction is performed with parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed [@schliep2011phangorn].
+
 While relative branch length information provides additional data for nodes without secondary date calibrations, topologies without branch lengths can also be dated.
@@
 
-For each congruent node, the pairwise distance that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method [SDM; @Criscuolo2006], which deforms patristic distance matrices by minimizing variance and then averaging them.
-These single summary taxon pair age matrices (Summarized calibrations) can be applied as secondary calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes [@Huelsenbeck2001; @Ronquist2003], PATHd8 [@Britton2007], BLADJ [@Webb2008; @webb2005phylomatic], and treePL [@Smith2012].
+<!-- we should calculate summary statistics per node -->
+<!--??? How is 'calibrations' defined here? the calibrations from congruification, or the pairwise distances??? is calibrations from congruification-->
+For each congruent node, the pairwise distance that traverse that node are summarized into a single summary matrix using classic statistics (mean, median, minimum and maximum ages).
+These single summary ages per taxon pair are then used as secondary calibrations to date the tree topology.
+<!--
+Under the SDM criterion ??? what happens??
+Summarizing options implemented include Super Distance Matrix method [SDM, @Criscuolo2006] 
+-->
+Summarized calibrations can be applied as secondary calibrations with different dating methods currently supported within DateLife: MrBayes [@Huelsenbeck2001; @Ronquist2003], PATHd8 [@Britton2007], BLADJ [@Webb2008; @webb2005phylomatic], and treePL [@Smith2012].
 
@@
 BLADJ does not incorporate branch lengths even when they are present.
-When there is conflict in ages between nodes with calibration data, the algorithm ignores ages that are older than ages of parent nodes and younger than ages from descendant nodes. 
-> NOTE: do you know how it decides which one to drop if two calibrations conflict?
-The parent or the child?
-It would drop both.
-It basically drops any calibration for any given node that are older than any calibration of the parent node, and younger than any calibration of children nodes.
+When there is conflict in ages between nodes with calibration data, the algorithm ignores ages that are older than ages of parent nodes and/or younger than ages from descendant nodes. 
+\NOTE:do you know how it decides which one to drop if two calibrations conflict? The parent or the child?
 
@@
 [@Smith2012].
-The MrBayes  [@Huelsenbeck2001; @Ronquist2003] approach in datelife uses the secondary calibrations as priors on node ages. 
-> NOTE: is this true?? Yes
+The MrBayes  [@Huelsenbeck2001; @Ronquist2003] apprach in datelife uses the secondary calibrartions as priors on node ages. \NOTE: is this true??
 
 _**Visualizing results.--**_
-Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as visualize and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions available from the R package `datelifeplot` [@datelifeplot].
+Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as visualize and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions.
 

### 2022-05-06 manuscript/manuscript.Rmd
re-frame fringillidae example

--- 2022-05-05 manuscript/manuscript.Rmd
@@
 Input topologies from OpenTree or the supertree approach described above do not include branch length estimates of any kind.
-To estimate branch lengths proportional to substitution rates on these topologies, DateLife can mine the Barcode of Life Data System, BOLD [@ratnasingham2007bold] to obtain genetic markers for the input taxa.
+Optionally, to estimate branch lengths proportional to substitution rates on these topologies, DateLife can mine the Barcode of Life Data System, BOLD [@ratnasingham2007bold] to obtain genetic markers for the input taxa.
 These markers are aligned with MUSCLE by default [@edgar2004muscle], or MAFFT [@katoh2009multiple]. 
@@
 
-For each congruent node, the pairwise distance that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method [SDM; @Criscuolo2006], which deforms patristic distance matrices by minimizing variance and then averaging them.
+For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method [SDM; @Criscuolo2006], which deforms patristic distance matrices by minimizing variance and then averaging them.
 These single summary taxon pair age matrices (Summarized calibrations) can be applied as secondary calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes [@Huelsenbeck2001; @Ronquist2003], PATHd8 [@Britton2007], BLADJ [@Webb2008; @webb2005phylomatic], and treePL [@Smith2012].
@@
 BLADJ does not incorporate branch lengths even when they are present.
-When there is conflict in ages between nodes with calibration data, the algorithm ignores ages that are older than ages of parent nodes and younger than ages from descendant nodes. 
-> NOTE: do you know how it decides which one to drop if two calibrations conflict?
-The parent or the child?
-It would drop both.
-It basically drops any calibration for any given node that are older than any calibration of the parent node, and younger than any calibration of children nodes.
-
+When there is conflict in ages between nodes with calibration data, the algorithm ignores node ages that are older than the age of a parent node. 
+BLADJ requires a root age estimate.
 If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature.
@@
 The MrBayes  [@Huelsenbeck2001; @Ronquist2003] approach in datelife uses the secondary calibrations as priors on node ages.
-> NOTE: is this true?? Yes
 
@@
 
-We illustrate the DateLife algorithm using a group within the Passeriform birds encompassing the family of true finches, Fringillidae and allies as case study.
+We illustrate the DateLife algorithm using a group within the Passeriform birds encompassing the family of true finches, Fringillidae as case study.
 The first example analyses 6 bird species and shows all steps of the algorithm.
-The second example is a real life application
+The second example is an application analysing all 289 species in the family Fringillidae for which we have date informtaion.
 <!--We used `r cite_r("r-references.bib")` for all our analyses.-->
@@
 
-We randomly chose 6 bird species related to the family Fringillidae of true finches. The sample includes
+We randomly chose 6 bird species in the family Fringillidae, the true finches. The sample includes
 two species of cardinals:
@@
 All nodes have age data.
-As fixed tree topology, DateLife used OpenTree's synthetic tree as default and mapped age data to nodes in the tree.
+DateLife used OpenTree's synthetic tree topology for these taxa and mapped age data to nodes in the tree.
 As expected, more inclusive nodes (e.g., node "n1") have more age data than less inclusive nodes (e.g., node "n5").
-The processing step allowed discovering five data points for node "n4" that would not have had any data otherwise.
-Age summary statistics per node were calculated and tested as secondary calibrations to date the tree topology using the BLADJ algorithm.
-Age data for node "n2" was excluded as final calibration because it is older than age data of a more inclusive node.
+The processing step allowed discovering five data points for node "n4" that would not have had any data otherwise. \textcolor{red}{I don't understand what this sentence means}
+Age summary statistics per node were calculated and used as secondary calibrations to date the tree topology using the BLADJ algorithm.
+Age data for node "n2" was excluded as final calibration because it is older than age data of a more inclusive node (n1).
 
-## Real life application
+## Fringillidae
 
-A college educator wishes to obtain state-of-the-art data on time of evolutionary origin of species belonging to the true finches for their class.
-They decide to use `datelife` because they are teaching best practices for reproducibility.
-Students have the option to go to the website at www.datelife.org and perform an interactive run. However, the educator also wants the students to practice their R skills.
-The first step is to run a `datelife` query using the "get species from taxon" flag.
-This will get all recognised species names within their chosen inclusive taxon.
-The Fringillidae has `r length(dquery$cleaned_names)` species, according to the Open Tree of Life taxonomy.
-Once with a curated set of species taxon names, the next step is to run a `datelife` search that will find all chronograms that contain at least two species names.
+To estimate ages for the taxa in the family Fringillidae, we ran a `datelife` query using the "get species from taxon" flag. 
+This will get all recognised species names within the named taxon. 
+The Fringillidae has species, according to the Open Tree of Life taxonomy.
+??? We have date estimates for `r length(dquery$cleaned_names)` of those species.
+
+??? There is an option to include all, right?
+According to the OpenTree of life synthetic phylogeny, Fringillidae is not a monophyletic family, so this taxon-constrained approach returns a topology and ages for the species in that named group.
+
+The next step is to run a `datelife` search on this curated set of species taxon names that will find all chronograms that contain at least two species.
 The algorithm proceeds to prune the trees to keep matching species names on tips only, and transform the pruned trees to pairwise distance matrices.

### 2022-05-06 manuscript/manuscript.Rmd
comments on cross validation

@@
 \ref{fig:fringillidages}).
-Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
+Median sxummary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
 <!-- ????????????????? Say some things about the results!
@@
 <!--Data from source chronograms can be used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates-->
-To perform a cross validation analysis of the DateLife workflow, we used resulting data from the previous section (Casestudy: Real Life Application).
-We took individual tree topologies from each of the 19 source chronograms found (Supplementary data XX).
-Then we congruifued age data of source chronograms from studies, and used this ages to date the tree topology with the program BLADJ.
+We performed a cross validation analysis of the DateLife workflow using the Fringillidae chronograms. 
+We used the individual tree topologies from each of the ??? TODO CHECK NUMBER 19 source chronograms from ?? studies as inputs, treating their node ages as unknown.
+We estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.
 
-We found that node ages from original study, and ages estimated using all other age data available are generally correlated (Supplementary Fig. \ref{fig:cvXY}).
+We found that node ages from original study, and ages estimated using all other age data available are correlated  \textcolor{red}{R^2 value?} (Supplementary Fig. \ref{fig:cvXY}).
+
 In 5 studies, more inclusive nodes have older original ages, and less inclusive nodes have younger original ages than their cross-validated age estimates. 
-Accordingly, root ages are generally older in the original study than estimated using cross-validated ages.
-Root ages were similar in original and cross validated ages in three vases (Supplementary Fig.
-\ref{fig:cvXYdiffs}).
-<!-- We need to think about the root!
-is it comparable across source chronograms??
-not all of them have the same sampling...
-so probably not the same root and thus not comparable.
-Address this...
-ANSWERED: -->
-Notably, chronograms have different species sampling, hence roots are not comparable across studies.
-Yet, chronograms with a higher sampling number can potentially inform the age of the root of chronograms with less sampling.
+\textcolor{red}{more inclusive nodes have to be older!
+:P I think there is something differnet you are trying to say here - maybe "Datelife tended to underestimate ages for more inclusive nodes relative to the orginal estimate, and overestimate ages for nodes closer to the tips"}
+Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. \ref{fig:cvXYdiffs}).
 

### 2022-05-07 manuscript/manuscript.Rmd
developing function to get OTT taxonomic children species only

@@
 
-We found that node ages from original study, and ages estimated using all other age data available are correlated  \textcolor{red}{R^2 value?} (Supplementary Fig. \ref{fig:cvXY}).
+We found that node ages from original study, and ages estimated using all other age data available are correlated  \textcolor{red}{R-square value?} (Supplementary Fig. \ref{fig:cvXY}).
 

### 2022-05-07 manuscript/manuscript.tex
developing function to get OTT taxonomic children species only

--- 2022-05-05 manuscript/manuscript.tex
@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4540}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4493}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 Input topologies from OpenTree or the supertree approach described above do not include branch length estimates of any kind.
-To estimate branch lengths proportional to substitution rates on these topologies, DateLife can mine the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain genetic markers for the input taxa.
+Optionally, to estimate branch lengths proportional to substitution rates on these topologies, DateLife can mine the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain genetic markers for the input taxa.
 These markers are aligned with MUSCLE by default (Edgar, 2004), or MAFFT (Katoh, Asimenos, \& Toh, 2009).
@@
 
-For each congruent node, the pairwise distance that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method {[}SDM; Criscuolo, Berry, Douzery, and Gascuel (2006){]}, which deforms patristic distance matrices by minimizing variance and then averaging them.
+For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method {[}SDM; Criscuolo, Berry, Douzery, and Gascuel (2006){]}, which deforms patristic distance matrices by minimizing variance and then averaging them.
 These single summary taxon pair age matrices (Summarized calibrations) can be applied as secondary calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Campbell O.
@@
 BLADJ does not incorporate branch lengths even when they are present.
-When there is conflict in ages between nodes with calibration data, the algorithm ignores ages that are older than ages of parent nodes and younger than ages from descendant nodes.
-\textgreater{} NOTE: do you know how it decides which one to drop if two calibrations conflict?
-The parent or the child?
-It would drop both.
-It basically drops any calibration for any given node that are older than any calibration of the parent node, and younger than any calibration of children nodes.
-
+When there is conflict in ages between nodes with calibration data, the algorithm ignores node ages that are older than the age of a parent node.
+BLADJ requires a root age estimate.
 If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature.
@@
 The MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003) approach in datelife uses the secondary calibrations as priors on node ages.
-\textgreater{} NOTE: is this true?? Yes
 
@@
 
-We illustrate the DateLife algorithm using a group within the Passeriform birds encompassing the family of true finches, Fringillidae and allies as case study.
+We illustrate the DateLife algorithm using a group within the Passeriform birds encompassing the family of true finches, Fringillidae as case study.
 The first example analyses 6 bird species and shows all steps of the algorithm.
-The second example is a real life application
+The second example is an application analysing all 289 species in the family Fringillidae for which we have date informtaion.
 
@@
 All nodes have age data.
-As fixed tree topology, DateLife used OpenTree's synthetic tree as default and mapped age data to nodes in the tree.
+DateLife used OpenTree's synthetic tree topology for these taxa and mapped age data to nodes in the tree.
 As expected, more inclusive nodes (e.g., node ``n1'') have more age data than less inclusive nodes (e.g., node ``n5'').
-The processing step allowed discovering five data points for node ``n4'' that would not have had any data otherwise.
-Age summary statistics per node were calculated and tested as secondary calibrations to date the tree topology using the BLADJ algorithm.
-Age data for node ``n2'' was excluded as final calibration because it is older than age data of a more inclusive node.
-
-\hypertarget{real-life-application}{%
-\subsection{Real life application}\label{real-life-application}}
-
-A college educator wishes to obtain state-of-the-art data on time of evolutionary origin of species belonging to the true finches for their class.
-They decide to use \texttt{datelife} because they are teaching best practices for reproducibility.
-Students have the option to go to the website at www.datelife.org and perform an interactive run. However, the educator also wants the students to practice their R skills.
-The first step is to run a \texttt{datelife} query using the ``get species from taxon'' flag.
-This will get all recognised species names within their chosen inclusive taxon.
-The Fringillidae has 289 species, according to the Open Tree of Life taxonomy.
-Once with a curated set of species taxon names, the next step is to run a \texttt{datelife} search that will find all chronograms that contain at least two species names.
+The processing step allowed discovering five data points for node ``n4'' that would not have had any data otherwise. \textcolor{red}{I don't understand what this sentence means}
+Age summary statistics per node were calculated and used as secondary calibrations to date the tree topology using the BLADJ algorithm.
+Age data for node ``n2'' was excluded as final calibration because it is older than age data of a more inclusive node (n1).
+
+\hypertarget{fringillidae}{%
+\subsection{Fringillidae}\label{fringillidae}}
+
+To estimate ages for the taxa in the family Fringillidae, we ran a \texttt{datelife} query using the ``get species from taxon'' flag.
+This will get all recognised species names within the named taxon.
+The Fringillidae has species, according to the Open Tree of Life taxonomy.
+??? We have date estimates for 289 of those species.
+
+??? There is an option to include all, right?
+According to the OpenTree of life synthetic phylogeny, Fringillidae is not a monophyletic family, so this taxon-constrained approach returns a topology and ages for the species in that named group.
+
+The next step is to run a \texttt{datelife} search on this curated set of species taxon names that will find all chronograms that contain at least two species.
 The algorithm proceeds to prune the trees to keep matching species names on tips only, and transform the pruned trees to pairwise distance matrices.
@@
 \ref{fig:fringillidages}).
-Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
+Median sxummary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
 
@@
 
-To perform a cross validation analysis of the DateLife workflow, we used resulting data from the previous section (Casestudy: Real Life Application).
-We took individual tree topologies from each of the 19 source chronograms found (Supplementary data XX).
-Then we congruifued age data of source chronograms from studies, and used this ages to date the tree topology with the program BLADJ.
-
-We found that node ages from original study, and ages estimated using all other age data available are generally correlated (Supplementary Fig. \ref{fig:cvXY}).
+We performed a cross validation analysis of the DateLife workflow using the Fringillidae chronograms.
+We used the individual tree topologies from each of the ??? TODO CHECK NUMBER 19 source chronograms from ?? studies as inputs, treating their node ages as unknown.
+We estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.
+
+We found that node ages from original study, and ages estimated using all other age data available are correlated \textcolor{red}{R-square value?} (Supplementary Fig. \ref{fig:cvXY}).
+
 In 5 studies, more inclusive nodes have older original ages, and less inclusive nodes have younger original ages than their cross-validated age estimates.
-Accordingly, root ages are generally older in the original study than estimated using cross-validated ages.
-Root ages were similar in original and cross validated ages in three vases (Supplementary Fig.
-\ref{fig:cvXYdiffs}).
-Notably, chronograms have different species sampling, hence roots are not comparable across studies.
-Yet, chronograms with a higher sampling number can potentially inform the age of the root of chronograms with less sampling.
+\textcolor{red}{more inclusive nodes have to be older!
+:P I think there is something differnet you are trying to say here - maybe "Datelife tended to underestimate ages for more inclusive nodes relative to the orginal estimate, and overestimate ages for nodes closer to the tips"}
+Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. \ref{fig:cvXYdiffs}).
 
@@
 
-At the time of writing of this manuscript (May 06, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
+At the time of writing of this manuscript (May 07, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
 A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time.

### 2022-05-10 manuscript/manuscript.Rmd
addressing more comments and adding citations

--- 2022-05-06 manuscript/manuscript.Rmd
@@
 Here we present the DateLife project and its core software application, available as an R package [@datelifeR], and as an online Rshiny interactive website at www.datelife.org, which captures data from published chronograms, and make these data readily accessible to users for reuse and reanalysis.
-The software features key elements for scientific reproducibility, such as a versioned, open and fully public chronogram database [@mctavish2015phylesystem], age data stored in a computer readable format [@vos2012nexml], automated and programmatic ways of accessing the data [@Stoltzfus2013] and methods to summarize and compare age data.
+The software features key elements for scientific reproducibility, such as a versioned, open and fully public source database [@mctavish2015phylesystem], data stored and available in a computer readable format [@vos2012nexml], automated and programmatic ways of accessing the data [@Stoltzfus2013] and methods to summarize and compare the data.
 
@@
 taxize [@Chamberlain2013; @Chamberlain2018].
-Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife algorithm: providing an input, searching a chronogram database, and summarizing results from the search.
+Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife algorithm: creating a search query, searching a database, and summarizing results from the search.
 
-<!--_**Providing an input.--**_ -->
+<!--_**Processing an input.--**_ -->
 
-## Processing an input
+##  Creating a search query
 
-DateLife starts by processing an input consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree.
+DateLife starts by processing a user input consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree.
 If the input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004].
@@
 
-## Searching the database
+## Searching a chronogram database
 
@@
 
-We illustrate the DateLife algorithm using a group within the Passeriform birds encompassing the family of true finches, Fringillidae as case study.
+We illustrate the DateLife algorithm using a group within the Passeriform birds known as the family of true finches, Fringillidae, as case study.
 The first example analyses 6 bird species and shows all steps of the algorithm.
-The second example is an application analysing all 289 species in the family Fringillidae for which we have date informtaion.
+The second example is an application analysing 289 species in the family Fringillidae included in the NCBI taxonomy.
 <!--We used `r cite_r("r-references.bib")` for all our analyses.-->
@@
 
-We randomly chose 6 bird species in the family Fringillidae, the true finches. The sample includes
+_**Creating a search query.-**_
+We chose 6 bird species in the family Fringillidae, known as the true finches. The sample includes
 two species of cardinals:
@@
 For a detailed discussion on the state of the synonym refer to Avibase [@lepage2004avibase; @lepage2014avibase; @avibase-emberiza].
+Discovering this synonym allowed assigning five age data points for the parent node of *Emberiza elegans*, shown as *Schoeniclus elegans* in figure \ref{fig:figure2-1}, which would not have had any data otherwise.
+<!--\textcolor{red}{I don't understand what this sentence means}
+it is related to the synonym found by tnrs processing, if we had used the synonym insted of the one accepted in the database, we would not have discovered that information
+It moved it here, it makes more sense here.
+-->
+
+_**Searching the database.-**_
 DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies.
@@
 All nodes have age data.
+
+_**Summarizing search results.-**_
 DateLife used OpenTree's synthetic tree topology for these taxa and mapped age data to nodes in the tree.
 As expected, more inclusive nodes (e.g., node "n1") have more age data than less inclusive nodes (e.g., node "n5").
-The processing step allowed discovering five data points for node "n4" that would not have had any data otherwise. \textcolor{red}{I don't understand what this sentence means}
 Age summary statistics per node were calculated and used as secondary calibrations to date the tree topology using the BLADJ algorithm.
-Age data for node "n2" was excluded as final calibration because it is older than age data of a more inclusive node (n1).
+Age data for node "n2" was excluded as final calibration because it is older than age data of the more inclusive node "n1".
 
@@
 
-To estimate ages for the taxa in the family Fringillidae, we ran a `datelife` query using the "get species from taxon" flag. 
-This will get all recognised species names within the named taxon. 
-The Fringillidae has species, according to the Open Tree of Life taxonomy.
-??? We have date estimates for `r length(dquery$cleaned_names)` of those species.
+_**Creating a query.-**_
+To estimate ages for species in the family Fringillidae, we ran a `datelife` query using the "get species from taxon" flag, 
+which gets all recognized species names within a named group from a taxonomy of choice (options are Open Tree of Life, NCBI, GBIF, or IRMNG). 
+Following the NCBI taxonomy, the Fringillidae has `r length(dquery$cleaned_names)` species.
 
-??? There is an option to include all, right?
-According to the OpenTree of life synthetic phylogeny, Fringillidae is not a monophyletic family, so this taxon-constrained approach returns a topology and ages for the species in that named group.
+_**Searching the database.-**_
+The next step is to use this curated set of species taxon names to identify all chronograms that contain at least two Fringillidae species.
+Once identified, the algorithm proceeds to prune matching chronograms to keep Fringillidae species names on tips only, and transform these pruned chronograms to pairwise distance matrices.
+The `datelife` search revealed `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
 
-The next step is to run a `datelife` search on this curated set of species taxon names that will find all chronograms that contain at least two species.
-The algorithm proceeds to prune the trees to keep matching species names on tips only, and transform the pruned trees to pairwise distance matrices.
-There are `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
-The final step is to summarize the available information using two alternative types of summary chronograms, median and SDM.
-As explained in the "Description" section, data from source chronograms is first summarised into a single distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig.
+_**Summarizing search results.-**_
+The final step is to summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM.
+As explained in the "Description" section, a tree topology to summarize age data upon is required.
+By default, `datelife` uses the Open Tree of Life synthetic phylogeny as summarizing topology. According to this phylogeny, Fringillidae is not a monophyletic family 
+[@burns2014phylogenetics;
+@burleigh2015building;
+@Jetz2012;
+@lerner2011multilocus;
+@lovette2010comprehensive;
+@pulgarin2013multilocus;
+@chaves2013biogeography;
+@lamichhaney2015evolution;
+@parchman2007coevolution;
+@selvatti2015paleogene;
+@moyle2016tectonic;
+@odeen2011evolution;
+@claramunt2015new;
+@beresford2005african;
+@kimball2019phylogenomic;
+@barker2015new;
+@klicka2014comprehensive;
+@barker2013going;
+@treplin2008molecular;
+@price2014niche;
+@hackett2008phylogenomic;
+@barker2014mitogenomic;
+@gibb2015new;
+@barker2004phylogeny;
+@zuccon2012phylogenetic;
+@oliveros2019earth;
+@johansson2008phylogenetic;
+@alstrom2014discovery;
+@powell2014comprehensive;
+@bryson2014diversification;
+@packert2012horizontal;
+@tietze2013complete].
+
+Hence, DateLife's taxon-constrained approach returns a topology and ages for the species in a named group.
+Age data from source chronograms is summarised into a single distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
 \ref{fig:fringillidages}).
-Median sxummary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
+
+Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
 <!-- ????????????????? Say some things about the results!
@@
 
-We found that node ages from original study, and ages estimated using all other age data available are correlated  \textcolor{red}{R^2 value?} (Supplementary Fig. \ref{fig:cvXY}).
+We found that node ages from original study, and ages estimated using all other age data available are correlated (Supplementary Fig. \ref{fig:cvXY}).
 
-In 5 studies, more inclusive nodes have older original ages, and less inclusive nodes have younger original ages than their cross-validated age estimates. 
-\textcolor{red}{more inclusive nodes have to be older!
-:P I think there is something differnet you are trying to say here - maybe "Datelife tended to underestimate ages for more inclusive nodes relative to the orginal estimate, and overestimate ages for nodes closer to the tips"}
-Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. \ref{fig:cvXYdiffs}).
+In 5 studies, DateLife presented a tendency to underestimate ages for more inclusive nodes, and overestimate ages for nodes closer to the tips, relative to the original age estimates.
+<!-- i.e., more inclusive nodes have older original ages, and less inclusive nodes have younger original ages than their cross-validated age estimates. 
+\textcolor{red}{more inclusive nodes have to be older!}
+-->
+Accordingly, root ages estimated with DateLife's cross-validated ages are generally younger relative to the age root from the original study  (Supplementary Fig. \ref{fig:cvXYdiffs}).
 

### 2022-05-10 manuscript/manuscript.tex
addressing more comments and adding citations

@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4493}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4536}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 Here we present the DateLife project and its core software application, available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org, which captures data from published chronograms, and make these data readily accessible to users for reuse and reanalysis.
-The software features key elements for scientific reproducibility, such as a versioned, open and fully public chronogram database (McTavish et al., 2015), age data stored in a computer readable format (Vos et al., 2012), automated and programmatic ways of accessing the data (Stoltzfus et al., 2013) and methods to summarize and compare age data.
+The software features key elements for scientific reproducibility, such as a versioned, open and fully public source database (McTavish et al., 2015), data stored and available in a computer readable format (Vos et al., 2012), automated and programmatic ways of accessing the data (Stoltzfus et al., 2013) and methods to summarize and compare the data.
 
@@
 taxize (Chamberlain \& Szöcs, 2013; Chamberlain et al., 2019).
-Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife algorithm: providing an input, searching a chronogram database, and summarizing results from the search.
-
-\hypertarget{processing-an-input}{%
-\subsection{Processing an input}\label{processing-an-input}}
-
-DateLife starts by processing an input consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree.
+Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife algorithm: creating a search query, searching a database, and summarizing results from the search.
+
+\hypertarget{creating-a-search-query}{%
+\subsection{Creating a search query}\label{creating-a-search-query}}
+
+DateLife starts by processing a user input consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree.
 If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis et al., 2004).
@@
 
-\hypertarget{searching-the-database}{%
-\subsection{Searching the database}\label{searching-the-database}}
+\hypertarget{searching-a-chronogram-database}{%
+\subsection{Searching a chronogram database}\label{searching-a-chronogram-database}}
 
@@
 
-We illustrate the DateLife algorithm using a group within the Passeriform birds encompassing the family of true finches, Fringillidae as case study.
+We illustrate the DateLife algorithm using a group within the Passeriform birds known as the family of true finches, Fringillidae, as case study.
 The first example analyses 6 bird species and shows all steps of the algorithm.
-The second example is an application analysing all 289 species in the family Fringillidae for which we have date informtaion.
+The second example is an application analysing 289 species in the family Fringillidae included in the NCBI taxonomy.
 
@@
 
-We randomly chose 6 bird species in the family Fringillidae, the true finches. The sample includes
+\emph{\textbf{Creating a search query.-}}
+We chose 6 bird species in the family Fringillidae, known as the true finches. The sample includes
 two species of cardinals:
@@
 For a detailed discussion on the state of the synonym refer to Avibase (Avibase, 2022; Lepage, 2004; Lepage, Vaidya, \& Guralnick, 2014).
+Discovering this synonym allowed assigning five age data points for the parent node of \emph{Emberiza elegans}, shown as \emph{Schoeniclus elegans} in figure \ref{fig:figure2-1}, which would not have had any data otherwise.
+
+\emph{\textbf{Searching the database.-}}
 DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies.
@@
 All nodes have age data.
+
+\emph{\textbf{Summarizing search results.-}}
 DateLife used OpenTree's synthetic tree topology for these taxa and mapped age data to nodes in the tree.
 As expected, more inclusive nodes (e.g., node ``n1'') have more age data than less inclusive nodes (e.g., node ``n5'').
-The processing step allowed discovering five data points for node ``n4'' that would not have had any data otherwise. \textcolor{red}{I don't understand what this sentence means}
 Age summary statistics per node were calculated and used as secondary calibrations to date the tree topology using the BLADJ algorithm.
-Age data for node ``n2'' was excluded as final calibration because it is older than age data of a more inclusive node (n1).
+Age data for node ``n2'' was excluded as final calibration because it is older than age data of the more inclusive node ``n1.''
 
@@
 
-To estimate ages for the taxa in the family Fringillidae, we ran a \texttt{datelife} query using the ``get species from taxon'' flag.
-This will get all recognised species names within the named taxon.
-The Fringillidae has species, according to the Open Tree of Life taxonomy.
-??? We have date estimates for 289 of those species.
-
-??? There is an option to include all, right?
-According to the OpenTree of life synthetic phylogeny, Fringillidae is not a monophyletic family, so this taxon-constrained approach returns a topology and ages for the species in that named group.
-
-The next step is to run a \texttt{datelife} search on this curated set of species taxon names that will find all chronograms that contain at least two species.
-The algorithm proceeds to prune the trees to keep matching species names on tips only, and transform the pruned trees to pairwise distance matrices.
-There are 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Price et al., 2014).
-The final step is to summarize the available information using two alternative types of summary chronograms, median and SDM.
-As explained in the ``Description'' section, data from source chronograms is first summarised into a single distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig.
+\emph{\textbf{Creating a query.-}}
+To estimate ages for species in the family Fringillidae, we ran a \texttt{datelife} query using the ``get species from taxon'' flag,
+which gets all recognized species names within a named group from a taxonomy of choice (options are Open Tree of Life, NCBI, GBIF, or IRMNG).
+Following the NCBI taxonomy, the Fringillidae has 289 species.
+
+\emph{\textbf{Searching the database.-}}
+The next step is to use this curated set of species taxon names to identify all chronograms that contain at least two Fringillidae species.
+Once identified, the algorithm proceeds to prune matching chronograms to keep Fringillidae species names on tips only, and transform these pruned chronograms to pairwise distance matrices.
+The \texttt{datelife} search revealed 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Price et al., 2014).
+
+\emph{\textbf{Summarizing search results.-}}
+The final step is to summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM.
+As explained in the ``Description'' section, a tree topology to summarize age data upon is required.
+By default, \texttt{datelife} uses the Open Tree of Life synthetic phylogeny as summarizing topology. According to this phylogeny, Fringillidae is not a monophyletic family
+(Alström et al., 2014; Barker, 2014; Barker et al., 2013, 2015; Barker, Cibois, Schikler, Feinstein, \& Cracraft, 2004; Beresford, Barker, Ryan, \& Crowe, 2005; Bryson Jr et al., 2014; Burleigh, Kimball, \& Braun, 2015; Burns et al., 2014; Chaves, Hidalgo, \& Klicka, 2013; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hackett et al., 2008; Jetz et al., 2012; Johansson, Fjeldså, \& Bowie, 2008; Kimball et al., 2019; Klicka et al., 2014; Lamichhaney et al., 2015; Lerner, Meyer, James, Hofreiter, \& Fleischer, 2011; Lovette et al., 2010; Moyle et al., 2016; Ödeen, Håstad, \& Alström, 2011; Oliveros et al., 2019; Päckert et al., 2012; Parchman, Benkman, \& Mezquida, 2007; Powell et al., 2014; Price et al., 2014; Pulgarín-R, Smith, Bryson Jr, Spellman, \& Klicka, 2013; Selvatti, Gonzaga, \& Moraes Russo, 2015; Tietze, Päckert, Martens, Lehmann, \& Sun, 2013; Treplin et al., 2008; Zuccon, Prŷs-Jones, Rasmussen, \& Ericson, 2012).
+
+Hence, DateLife's taxon-constrained approach returns a topology and ages for the species in a named group.
+Age data from source chronograms is summarised into a single distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
 \ref{fig:fringillidages}).
-Median sxummary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
+
+Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
 
@@
 
-We found that node ages from original study, and ages estimated using all other age data available are correlated \textcolor{red}{R-square value?} (Supplementary Fig. \ref{fig:cvXY}).
-
-In 5 studies, more inclusive nodes have older original ages, and less inclusive nodes have younger original ages than their cross-validated age estimates.
-\textcolor{red}{more inclusive nodes have to be older!
-:P I think there is something differnet you are trying to say here - maybe "Datelife tended to underestimate ages for more inclusive nodes relative to the orginal estimate, and overestimate ages for nodes closer to the tips"}
-Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. \ref{fig:cvXYdiffs}).
+We found that node ages from original study, and ages estimated using all other age data available are correlated (Supplementary Fig. \ref{fig:cvXY}).
+
+In 5 studies, DateLife presented a tendency to underestimate ages for more inclusive nodes, and overestimate ages for nodes closer to the tips, relative to the original age estimates.
+Accordingly, root ages estimated with DateLife's cross-validated ages are generally younger relative to the age root from the original study (Supplementary Fig. \ref{fig:cvXYdiffs}).
 
@@
 
-At the time of writing of this manuscript (May 07, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
+At the time of writing of this manuscript (May 11, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
 A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time.
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-alstrom2014discovery}{}}%
+Alström, P., Hooper, D.
+M., Liu, Y., Olsson, U., Mohan, D., Gelang, M., \ldots{} Price, T.
+D.
+(2014).
+Discovery of a relict lineage and monotypic family of passerine birds.
+\emph{Biology Letters}, \emph{10}(3), 20131067.
+
 \leavevmode\vadjust pre{\hypertarget{ref-ane2009groves}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-barker2014mitogenomic}{}}%
+Barker, F. K. (2014). Mitogenomic data resolve basal relationships among passeriform and passeridan birds. \emph{Molecular Phylogenetics and Evolution}, \emph{79}, 313--324.
+
 \leavevmode\vadjust pre{\hypertarget{ref-barker2013going}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-barker2004phylogeny}{}}%
+Barker, F.
+K., Cibois, A., Schikler, P., Feinstein, J., \& Cracraft, J.
+(2004).
+Phylogeny and diversification of the largest avian radiation.
+\emph{Proceedings of the National Academy of Sciences}, \emph{101}(30), 11040--11045.
+
+\leavevmode\vadjust pre{\hypertarget{ref-beresford2005african}{}}%
+Beresford, P., Barker, F., Ryan, Pg., \& Crowe, T.
+(2005).
+African endemics span the tree of songbirds (passeri): Molecular systematics of several evolutionary {`enigmas.'} \emph{Proceedings of the Royal Society B: Biological Sciences}, \emph{272}(1565), 849--858.
+
 \leavevmode\vadjust pre{\hypertarget{ref-Boyle2013}{}}%
@@
 \url{https://doi.org/10.1080/10635150701613783}
+
+\leavevmode\vadjust pre{\hypertarget{ref-bryson2014diversification}{}}%
+Bryson Jr, R.
+W., Chaves, J., Smith, B.
+T., Miller, M.
+J., Winker, K., Pérez-Emán, J.
+L., \& Klicka, J.
+(2014).
+Diversification across the new world within the `blue'cardinalids (aves: cardinalidae).
+\emph{Journal of Biogeography}, \emph{41}(3), 587--599.
+
+\leavevmode\vadjust pre{\hypertarget{ref-burleigh2015building}{}}%
+Burleigh, J. G., Kimball, R. T., \& Braun, E. L. (2015). Building the avian tree of life using a large-scale, sparse supermatrix. \emph{Molecular Phylogenetics and Evolution}, \emph{84}, 53--63.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-chaves2013biogeography}{}}%
+Chaves, J.
+A., Hidalgo, J.
+R., \& Klicka, J.
+(2013).
+Biogeography and evolutionary history of the n eotropical genus s altator (a ves: T hraupini).
+\emph{Journal of Biogeography}, \emph{40}(11), 2180--2190.
+
 \leavevmode\vadjust pre{\hypertarget{ref-claramunt2015new}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-hackett2008phylogenomic}{}}%
+Hackett, S.
+J., Kimball, R.
+T., Reddy, S., Bowie, R.
+C., Braun, E.
+L., Braun, M.
+J., \ldots{} others.
+(2008).
+A phylogenomic study of birds reveals their evolutionary history.
+\emph{Science}, \emph{320}(5884), 1763--1768.
+
 \leavevmode\vadjust pre{\hypertarget{ref-Harmon2008}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-johansson2008phylogenetic}{}}%
+Johansson, U.
+S., Fjeldså, J., \& Bowie, R.
+C.
+(2008).
+Phylogenetic relationships within passerida (aves: Passeriformes): A review and a new molecular phylogeny based on three nuclear intron markers.
+\emph{Molecular Phylogenetics and Evolution}, \emph{48}(3), 858--876.
+
 \leavevmode\vadjust pre{\hypertarget{ref-katoh2009multiple}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-kimball2019phylogenomic}{}}%
+Kimball, R. T., Oliveros, C. H., Wang, N., White, N. D., Barker, F. K., Field, D. J., \ldots{} others. (2019). A phylogenomic supertree of birds. \emph{Diversity}, \emph{11}(7), 109.
+
+\leavevmode\vadjust pre{\hypertarget{ref-klicka2014comprehensive}{}}%
+Klicka, J., Barker, F.
+K., Burns, K.
+J., Lanyon, S.
+M., Lovette, I.
+J., Chaves, J.
+A., \& Bryson Jr, R.
+W.
+(2014).
+A comprehensive multilocus assessment of sparrow (aves: Passerellidae) relationships.
+\emph{Molecular Phylogenetics and Evolution}, \emph{77}, 177--182.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lamichhaney2015evolution}{}}%
+Lamichhaney, S., Berglund, J., Almén, M.
+S., Maqbool, K., Grabherr, M., Martinez-Barrio, A., \ldots{} others.
+(2015).
+Evolution of darwin's finches and their beaks revealed by genome sequencing.
+\emph{Nature}, \emph{518}(7539), 371--375.
+
 \leavevmode\vadjust pre{\hypertarget{ref-laubichler2009form}{}}%
@@
 Lepage, D., Vaidya, G., \& Guralnick, R. (2014). Avibase--a database system for managing and organizing taxonomic concepts. \emph{ZooKeys}, (420), 117.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lerner2011multilocus}{}}%
+Lerner, H.
+R., Meyer, M., James, H.
+F., Hofreiter, M., \& Fleischer, R.
+C.
+(2011).
+Multilocus resolution of phylogeny and timescale in the extant adaptive radiation of hawaiian honeycreepers.
+\emph{Current Biology}, \emph{21}(21), 1838--1844.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lovette2010comprehensive}{}}%
+Lovette, I.
+J., Pérez-Emán, J.
+L., Sullivan, J.
+P., Banks, R.
+C., Fiorentino, I., Córdoba-Córdoba, S., \ldots{} others.
+(2010).
+A comprehensive multilocus phylogeny for the wood-warblers and a revised classification of the parulidae (aves).
+\emph{Molecular Phylogenetics and Evolution}, \emph{57}(2), 753--770.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-moyle2016tectonic}{}}%
+Moyle, R.
+G., Oliveros, C.
+H., Andersen, M.
+J., Hosner, P.
+A., Benz, B.
+W., Manthey, J.
+D., \ldots{} Faircloth, B.
+C.
+(2016).
+Tectonic collision and uplift of wallacea triggered the global songbird radiation.
+\emph{Nature Communications}, \emph{7}(1), 1--7.
+
+\leavevmode\vadjust pre{\hypertarget{ref-odeen2011evolution}{}}%
+Ödeen, A., Håstad, O., \& Alström, P. (2011). Evolution of ultraviolet vision in the largest avian radiation-the passerines. \emph{BMC Evolutionary Biology}, \emph{11}(1), 1--8.
+
+\leavevmode\vadjust pre{\hypertarget{ref-oliveros2019earth}{}}%
+Oliveros, C.
+H., Field, D.
+J., Ksepka, D.
+T., Barker, F.
+K., Aleixo, A., Andersen, M.
+J., \ldots{} others.
+(2019).
+Earth history and the passerine superradiation.
+\emph{Proceedings of the National Academy of Sciences}, \emph{116}(16), 7916--7925.
+
 \leavevmode\vadjust pre{\hypertarget{ref-Ooms2018}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-packert2012horizontal}{}}%
+Päckert, M., Martens, J., Sun, Y.-H., Severinghaus, L.
+L., Nazarenko, A.
+A., Ting, J., \ldots{} Tietze, D.
+T.
+(2012).
+Horizontal and elevational phylogeographic patterns of himalayan and southeast asian forest passerines (aves: passeriformes).
+\emph{Journal of Biogeography}, \emph{39}(3), 556--573.
+
 \leavevmode\vadjust pre{\hypertarget{ref-paradis2004}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-parchman2007coevolution}{}}%
+Parchman, T.
+L., Benkman, C.
+W., \& Mezquida, E.
+T.
+(2007).
+Coevolution between hispaniolan crossbills and pine: Does more time allow for greater phenotypic escalation at lower latitude?
+\emph{Evolution}, \emph{61}(9), 2142--2153.
+
 \leavevmode\vadjust pre{\hypertarget{ref-posadas2006historical}{}}%
 Posadas, P., Crisci, J. V., \& Katinas, L. (2006). Historical biogeography: A review of its basic concepts and critical issues. \emph{{Journal of Arid Environments}}, \emph{66}(3), 389--403.
+
+\leavevmode\vadjust pre{\hypertarget{ref-powell2014comprehensive}{}}%
+Powell, A.
+F., Barker, F.
+K., Lanyon, S.
+M., Burns, K.
+J., Klicka, J., \& Lovette, I.
+J.
+(2014).
+A comprehensive species-level molecular phylogeny of the new world blackbirds (icteridae).
+\emph{Molecular Phylogenetics and Evolution}, \emph{71}, 94--112.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-pulgarin2013multilocus}{}}%
+Pulgarín-R, P.
+C., Smith, B.
+T., Bryson Jr, R.
+W., Spellman, G.
+M., \& Klicka, J.
+(2013).
+Multilocus phylogeny and biogeography of the new world pheucticus grosbeaks (aves: cardinalidae).
+\emph{Molecular Phylogenetics and Evolution}, \emph{69}(3), 1222--1227.
+
 \leavevmode\vadjust pre{\hypertarget{ref-RCoreTeam2018}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-selvatti2015paleogene}{}}%
+Selvatti, A.
+P., Gonzaga, L.
+P., \& Moraes Russo, C.
+A.
+de.
+(2015).
+A paleogene origin for crown passerines and the diversification of the oscines in the new world.
+\emph{Molecular Phylogenetics and Evolution}, \emph{88}, 1--15.
+
 \leavevmode\vadjust pre{\hypertarget{ref-smith2018constructing}{}}%
@@
 \url{https://doi.org/10.1186/1471-2105-14-158}
+
+\leavevmode\vadjust pre{\hypertarget{ref-tietze2013complete}{}}%
+Tietze, D.
+T., Päckert, M., Martens, J., Lehmann, H., \& Sun, Y.-H.
+(2013).
+Complete phylogeny and historical biogeography of true rosefinches (aves: carpodacus).
+\emph{Zoological Journal of the Linnean Society}, \emph{169}(1), 215--234.
 
@@
 \url{https://doi.org/10.1093/sysbio/syw102}
+
+\leavevmode\vadjust pre{\hypertarget{ref-treplin2008molecular}{}}%
+Treplin, S., Siegert, R., Bleidorn, C., Thompson, H.
+S., Fotso, R., \& Tiedemann, R.
+(2008).
+Molecular phylogeny of songbirds (aves: Passeriformes) and the relative utility of common nuclear marker loci.
+\emph{Cladistics}, \emph{24}(3), 328--349.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-zuccon2012phylogenetic}{}}%
+Zuccon, D., Prŷs-Jones, R., Rasmussen, P.
+C., \& Ericson, P.
+G.
+(2012).
+The phylogenetic relationships and generic limits of finches (fringillidae).
+\emph{Molecular Phylogenetics and Evolution}, \emph{62}(2), 581--596.
+
 \end{CSLReferences}

### 2022-05-11 manuscript/manuscript.Rmd
adding congruification vignette and updating others

@@
 We performed a cross validation analysis of the DateLife workflow using the Fringillidae chronograms. 
-We used the individual tree topologies from each of the ??? TODO CHECK NUMBER 19 source chronograms from ?? studies as inputs, treating their node ages as unknown.
-We estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.
+We used the individual tree topologies from each of the 19 source chronograms from 13 studies as inputs, treating their node ages as unknown.
+We then estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.
 

### 2022-05-11 manuscript/manuscript.Rmd
comments on discussion

--- 2022-05-07 manuscript/manuscript.Rmd
@@
 
-We found that node ages from original study, and ages estimated using all other age data available are correlated  \textcolor{red}{R-square value?} (Supplementary Fig. \ref{fig:cvXY}).
+We found that node ages from original study, and ages estimated using all other age data available are correlated  \textcolor{red}{R^2 value?} (Supplementary Fig. \ref{fig:cvXY}).
 
@@
 
-The main goal of `datelife` is to make state-of-the-art information on time of lineage divergence easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
+`datelife` makes state-of-the-art information on time of lineage divergence easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
 It is an open service that does not require any expert biological knowledge from users --besides the names of the organisms they want to work with, for any of its functionality.
@@
 
-Incorporation of more chronograms into `datelife`'s database is crucial to improve its services.
+A total of `r length(unique(unname(unlist(lapply(opentree_chronograms$trees, "[[", "tip.label")))))` unique terminal taxa are represented in `datelife`'s database.
+Incorporation of more chronograms into `datelife`'s database will continue to improve its services.
 One option to increase chronogram number in the database is the Dryad data repository.
@@
 was constructed using age data from 2,274 published chronograms [@Hedges2015]. 
-However the source chronograms used as input data for this tree are not available in computer readable format for reuse or reanalysis.
-As this tree is part of datelife's database, the amount of lineages that can be queried using `datelife` (`r length(unique(unname(unlist(lapply(opentree_chronograms$trees, "[[", "tip.label")))))` unique terminal taxa) is substantial.
-Access to the input chronograms used to generate the Hedges et al.
-summary tree would improve measures of uncertainty in DateLife, but they are available only as image files and not as usable data (timetree.org).
+Unfortunately, the source chronograms used as input data for this tree are not available in computer readable format for reuse or reanalysis.
+Access to the input chronograms used to generate the Hedges et al. summary tree would improve measures of uncertainty in DateLife, but they are currently available only as image files (timetree.org).
 <!--IN VIOLATION OF NSF RULES! but I don't suppose we should say that.-->
-We would like to emphasize on the importance of sharing chronogram data for the benefit of the scientific community as a whole, into repositories that require expert input and manual curation, such as OpenTree's Phylesystem [@mctavish2015phylesystem].
+We would like to emphasize on the importance of sharing chronogram data,  including systematically curated metadata,  into open repositories, such as OpenTree's Phylesystem [@mctavish2015phylesystem] for the benefit of the scientific community as a whole.
 
@@
 
-Scientists usually also favor chronograms constructed using primary calibrations (ages obtained from the fossil or geological record) to ones constructed with secondary calibrations (ages coming from other chronograms)[@schenk2016sec].
-It has been observed with simulations that divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with Bayesian inference methods when priors are implemented in similar ways in both analyses [@schenk2016sec].
-However, secondary calibrations
-can be applied using other dating methods that do not require setting priors, such as penalized likelihood [@sanderson2003r8s], or as fixed ages,
-potentially mitigating the bias reported with Bayesian methods.
+Scientists have more confidence in chronograms constructed using primary calibrations (ages obtained from the fossil or geological record) to ones constructed with secondary calibrations (ages coming from other chronograms)[@schenk2016sec].
+@schenk2016sec showed that in simulations divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with Bayesian inference methods when priors are implemented in similar ways in both analyses.
+However, secondary calibrations are useful in dating methods that do not require setting priors, such as penalized likelihood [@sanderson2003r8s], or as fixed ages for nodes.
 Certainly, further studies are required to fully understand the effect of using secondary calibrations on time estimates and downstream analyses.
 
-Furthermore, chronograms can be obtained with primary fossil data or with molecular substitution rates obtained experimentally, which can deepen the already substantial variation in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
-<!--I think that one very odd one didn't have any fossils though! Maybe worth mentioning here? 
-Yes!!-->
-This observation is often encountered in the literature
+The input chronograms may have been estimated using primary fossil data or using estimates of molecular substitution rates.
+These differences in approach which can deepen the already substantial variation in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
+For eaxmple, the chronograms from  @burns2014phylogenetics tend to have much older date estimates for the same nodes than other inferences (Figure 5). 
+This study did not use any primary calibrations, and inferred dates using an substitution rate estimates across birds (cite Weir, J.T., Schluter, D., 2008.
+Calibrating the avian molecular clock.
+Mol.
+Ecol.
+17,
+231–2328).
+
+Conflict among alternative date estimates is common in the literature
 (see, for example, the ongoing debate about crown group age of angiosperms [@sauquet2021age; @barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time]. 
-For some studies, especially ones based on branch lengths (e.g., studies of species diversification, timing of evolutionary events, phenotypic trait evolution), using a different chronogram may return different results [@title2016macrophylogenies].
-Stitching together these chronograms can create a larger tree that uses information from multiple studies, but the effect of uncertainties and errors at this level on downstream analyses is still largely unknown.
+Using a different chronogram may return different results [@title2016macrophylogenies] particularly for studies of species diversification, or focussed on the timing of evolutionary events.
+Stitching together multiple chronograms creates a larger tree that uses information from across studies, but may amplify uncertainty and errors.
 
@@
 <!-- Example -->
+??? it's not clear to me what this last part means! What might you be losing?
 
@@
 Hence, we integrated into datelife's workflow different ways of estimating node ages in the absence of calibrations and branch length information for taxa lacking this information.
-"Making up" branch lengths is an accepted practice in scientific publications: @Jetz2012, created a time-calibrated tree of all 9,993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a tree of 353,185 seed plants where only 23% had molecular data.
+"Making up" branch lengths is an common practice in scientific publications: @Jetz2012, created a time-calibrated tree of all 9,993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a tree of 353,185 seed plants where only 23% had molecular data.
 Obviously, there are risks in this practice!

### 2022-05-11 manuscript/manuscript.Rmd
formatting

@@
 For eaxmple, the chronograms from  @burns2014phylogenetics tend to have much older date estimates for the same nodes than other inferences (Figure 5). 
-This study did not use any primary calibrations, and inferred dates using an substitution rate estimates across birds (cite Weir, J.T., Schluter, D., 2008.
+This study did not use any primary calibrations, and inferred dates using an substitution rate estimates across birds \textcolor{red}{(cite Weir, J.T., Schluter, D., 2008.
 Calibrating the avian molecular clock.
@@
 17,
-231–2328).
+231–2328).}
+\textcolor{red}{MORE ON the FINCH examples and the cross validation here}
 
@@
 <!-- Example -->
-??? it's not clear to me what this last part means! What might you be losing?
+\textcolor{red}{??? it's not clear to me what this last part means! What might you be losing?}
 

### 2022-05-11 manuscript/manuscript.Rmd

--- 2022-05-11 manuscript/manuscript.Rmd
@@
 
-Scientists usually also favor chronograms constructed using primary calibrations (ages obtained from the fossil or geological record) to ones constructed with secondary calibrations (ages coming from other chronograms)[@schenk2016sec].
-It has been observed with simulations that divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with Bayesian inference methods when priors are implemented in similar ways in both analyses [@schenk2016sec].
-However, secondary calibrations
-can be applied using other dating methods that do not require setting priors, such as penalized likelihood [@sanderson2003r8s], or as fixed ages,
-potentially mitigating the bias reported with Bayesian methods.
+Scientists have more confidence in chronograms constructed using primary calibrations (ages obtained from the fossil or geological record) to ones constructed with secondary calibrations (ages coming from other chronograms)[@schenk2016sec].
+@schenk2016sec showed that in simulations divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with Bayesian inference methods when priors are implemented in similar ways in both analyses.
+However, secondary calibrations are useful in dating methods that do not require setting priors, such as penalized likelihood [@sanderson2003r8s], or as fixed ages for nodes.
 Certainly, further studies are required to fully understand the effect of using secondary calibrations on time estimates and downstream analyses.
 
-Furthermore, chronograms can be obtained with primary fossil data or with molecular substitution rates obtained experimentally, which can deepen the already substantial variation in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
-<!--I think that one very odd one didn't have any fossils though! Maybe worth mentioning here? 
-Yes!!-->
-This observation is often encountered in the literature
+The input chronograms may have been estimated using primary fossil data or using estimates of molecular substitution rates.
+These differences in approach which can deepen the already substantial variation in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
+For eaxmple, the chronograms from  @burns2014phylogenetics tend to have much older date estimates for the same nodes than other inferences (Figure 5). 
+This study did not use any primary calibrations, and inferred dates using an substitution rate estimates across birds \textcolor{red}{(cite Weir, J.T., Schluter, D., 2008.
+Calibrating the avian molecular clock.
+Mol.
+Ecol.
+17,
+231–2328).}
+\textcolor{red}{MORE ON the FINCH examples and the cross validation here}
+
+Conflict among alternative date estimates is common in the literature
 (see, for example, the ongoing debate about crown group age of angiosperms [@sauquet2021age; @barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time]. 
-For some studies, especially ones based on branch lengths (e.g., studies of species diversification, timing of evolutionary events, phenotypic trait evolution), using a different chronogram may return different results [@title2016macrophylogenies].
-Stitching together these chronograms can create a larger tree that uses information from multiple studies, but the effect of uncertainties and errors at this level on downstream analyses is still largely unknown.
+Using a different chronogram may return different results [@title2016macrophylogenies] particularly for studies of species diversification, or focussed on the timing of evolutionary events.
+Stitching together multiple chronograms creates a larger tree that uses information from across studies, but may amplify uncertainty and errors.
 
@@
 <!-- Example -->
+\textcolor{red}{??? it's not clear to me what this last part means! What might you be losing?}
 

### 2022-05-11 manuscript/manuscript.Rmd
incorporating edits

@@
 These differences in approach which can deepen the already substantial variation in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
-For eaxmple, the chronograms from  @burns2014phylogenetics tend to have much older date estimates for the same nodes than other inferences (Figure 5). 
-This study did not use any primary calibrations, and inferred dates using an substitution rate estimates across birds \textcolor{red}{(cite Weir, J.T., Schluter, D., 2008.
-Calibrating the avian molecular clock.
-Mol.
-Ecol.
-17,
-231–2328).}
-\textcolor{red}{MORE ON the FINCH examples and the cross validation here}
+For example, the chronograms from @burns2014phylogenetics tend to have much older date estimates for the same nodes than other inferences (Figure 5). 
+This study did not use any primary calibrations, and inferred dates using an substitution rate estimates across birds [@weir2008calibrating] 
+<!--\textcolor{red}{MORE ON the FINCH examples and the cross validation here}
+-->
 
@@
 Choosing the elements of the chronograms that we are going to keep and the ones that we are going to discard is key, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram [@sauquet2021age].
-<!-- Example -->
+<!-- Example
 \textcolor{red}{??? it's not clear to me what this last part means! What might you be losing?}
+-->
 
@@
 Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch estimation methods.
-There has yet to be a thorough analysis of what can go wrong when one extends inferences beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above and make carefully consider the statistical assumptions being made, and assess the consistency of the results with prior work.
+There has yet to be a thorough analysis of what can go wrong when one extends inferences beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above, by carefully considering the statistical assumptions being made, and assessing the consistency of the results with prior work.
 

### 2022-05-11 manuscript/manuscript.tex
incorporating edits

@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4536}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4530}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 We performed a cross validation analysis of the DateLife workflow using the Fringillidae chronograms.
-We used the individual tree topologies from each of the ??? TODO CHECK NUMBER 19 source chronograms from ?? studies as inputs, treating their node ages as unknown.
-We estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.
+We used the individual tree topologies from each of the 19 source chronograms from 13 studies as inputs, treating their node ages as unknown.
+We then estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.
 
@@
 
-The main goal of \texttt{datelife} is to make state-of-the-art information on time of lineage divergence easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
+\texttt{datelife} makes state-of-the-art information on time of lineage divergence easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
 It is an open service that does not require any expert biological knowledge from users --besides the names of the organisms they want to work with, for any of its functionality.
 
-At the time of writing of this manuscript (May 11, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
+At the time of writing of this manuscript (May 22, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
 A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time.
@@
 
-Incorporation of more chronograms into \texttt{datelife}'s database is crucial to improve its services.
+A total of 99474 unique terminal taxa are represented in \texttt{datelife}'s database.
+Incorporation of more chronograms into \texttt{datelife}'s database will continue to improve its services.
 One option to increase chronogram number in the database is the Dryad data repository.
@@
 was constructed using age data from 2,274 published chronograms (Hedges et al., 2015).
-However the source chronograms used as input data for this tree are not available in computer readable format for reuse or reanalysis.
-As this tree is part of datelife's database, the amount of lineages that can be queried using \texttt{datelife} (99474 unique terminal taxa) is substantial.
-Access to the input chronograms used to generate the Hedges et al.~summary tree would improve measures of uncertainty in DateLife, but they are available only as image files and not as usable data (timetree.org).
-We would like to emphasize on the importance of sharing chronogram data for the benefit of the scientific community as a whole, into repositories that require expert input and manual curation, such as OpenTree's Phylesystem (McTavish et al., 2015).
+Unfortunately, the source chronograms used as input data for this tree are not available in computer readable format for reuse or reanalysis.
+Access to the input chronograms used to generate the Hedges et al.~summary tree would improve measures of uncertainty in DateLife, but they are currently available only as image files (timetree.org).
+We would like to emphasize on the importance of sharing chronogram data, including systematically curated metadata, into open repositories, such as OpenTree's Phylesystem (McTavish et al., 2015) for the benefit of the scientific community as a whole.
 
@@
 
-Scientists usually also favor chronograms constructed using primary calibrations (ages obtained from the fossil or geological record) to ones constructed with secondary calibrations (ages coming from other chronograms)(Schenk, 2016).
-It has been observed with simulations that divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with Bayesian inference methods when priors are implemented in similar ways in both analyses (Schenk, 2016).
-However, secondary calibrations
-can be applied using other dating methods that do not require setting priors, such as penalized likelihood (Sanderson, 2003), or as fixed ages,
-potentially mitigating the bias reported with Bayesian methods.
+Scientists have more confidence in chronograms constructed using primary calibrations (ages obtained from the fossil or geological record) to ones constructed with secondary calibrations (ages coming from other chronograms)(Schenk, 2016).
+Schenk (2016) showed that in simulations divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with Bayesian inference methods when priors are implemented in similar ways in both analyses.
+However, secondary calibrations are useful in dating methods that do not require setting priors, such as penalized likelihood (Sanderson, 2003), or as fixed ages for nodes.
 Certainly, further studies are required to fully understand the effect of using secondary calibrations on time estimates and downstream analyses.
 
-Furthermore, chronograms can be obtained with primary fossil data or with molecular substitution rates obtained experimentally, which can deepen the already substantial variation in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
-This observation is often encountered in the literature
+The input chronograms may have been estimated using primary fossil data or using estimates of molecular substitution rates.
+These differences in approach which can deepen the already substantial variation in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
+For example, the chronograms from Burns et al. (2014) tend to have much older date estimates for the same nodes than other inferences (Figure 5).
+This study did not use any primary calibrations, and inferred dates using an substitution rate estimates across birds (Weir \& Schluter, 2008)
+
+Conflict among alternative date estimates is common in the literature
 (see, for example, the ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón et al., 2015; Ramshaw et al., 1972; Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021).
-For some studies, especially ones based on branch lengths (e.g., studies of species diversification, timing of evolutionary events, phenotypic trait evolution), using a different chronogram may return different results (Title \& Rabosky, 2016).
-Stitching together these chronograms can create a larger tree that uses information from multiple studies, but the effect of uncertainties and errors at this level on downstream analyses is still largely unknown.
+Using a different chronogram may return different results (Title \& Rabosky, 2016) particularly for studies of species diversification, or focussed on the timing of evolutionary events.
+Stitching together multiple chronograms creates a larger tree that uses information from across studies, but may amplify uncertainty and errors.
 
@@
 Hence, we integrated into datelife's workflow different ways of estimating node ages in the absence of calibrations and branch length information for taxa lacking this information.
-``Making up'' branch lengths is an accepted practice in scientific publications: Jetz et al.
+``Making up'' branch lengths is an common practice in scientific publications: Jetz et al.
 (2012), created a time-calibrated tree of all 9,993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-weir2008calibrating}{}}%
+Weir, J., \& Schluter, D. (2008). Calibrating the avian molecular clock. \emph{Molecular Ecology}, \emph{17}(10), 2321--2328.
+
 \leavevmode\vadjust pre{\hypertarget{ref-zuccon2012phylogenetic}{}}%

### 2022-05-23 manuscript/manuscript.Rmd
BCO comments and mild edits

--- 2022-05-11 manuscript/manuscript.Rmd
@@
 
-  Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 425 Hesler Biology Building, Knoxville, TN 37996, USA.
+  Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 446 Hesler Biology Building, Knoxville, TN 37996, USA.
 
@@
   Date estimates for times of evolutionary divergences are key data for research in the natural sciences.
-These estimates also provide valuable information for for education, science communication and policy decisions.
-  Although achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is a difficult and time-consuming task, the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
+These estimates also provide valuable information for for education, science communication, and policy decisions.
+  Although achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram) is a difficult and time-consuming task, the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
   When these estimates are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on time of evolutionary origin is exposed in a programatic and reusable way.
   Intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
-  Here we present DateLife, a service implemented as an R package and an Rshiny website application available at www.datelife.org, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
-  The main DateLife workflow constructs a chronogram for any given combination of taxon names, by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from TreeBASE database as well.
+  Here we present DateLife, a service implemented as an R package and an R Shiny website application available at www.datelife.org, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
+  The main DateLife workflow constructs a chronogram for any given combination of taxon names by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from the TreeBASE database as well.
   We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
@@
 
-DateLife starts by processing an input consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree.
+DateLife starts by processing an input consisting of at least two taxon names, which can be provided as a comma separated character string or as tip labels on a tree.
 If the input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004].
@@
 
-DateLife accepts scientific names that can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or a binomial specific.
+DateLife accepts scientific names that can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or a binomial species name.
 Subspecies and variants are ignored.
@@
 DateLife implements TNRS using OpenTree's unified taxonomy as standard [@opentreeAPIs; @rees2017automated], storing OpenTree's Taxonomy identification numbers for further processing.
-<!-- We are using OpenTree's TNRS, Boyle described TNRS first, right?  -->
+<!-- We are using OpenTree's TNRS [BCO: yes], Boyle described TNRS first, right?  [BCO: earliest I can find, but feels like there should be much older work]-->
 
 The processed input taxon names are saved as an R object of a newly defined class \texttt{datelifeQuery} that is used in the following steps.
-This object contains the standardized names, the corresponding taxonomic id numbers, and the topology of the input tree if any was provided.
+This object contains the standardized names, the corresponding OpenTree taxonomic id numbers, and the topology of the input tree if any was provided.
 
@@
 Published chronograms can be added to Phylesystem by any user, at any time, and are immediately publicly available (https://tree.opentreeoflife.org/curator).
-This facilitates an immediate update of DateLife's chronogram database to include new chronogram data on a following search.
+A user may then update their local cache of DateLife's chronogram database to include new chronogram data on a following search.
 
-A DateLife search is implemented by matching processed taxon names provided by the user, to tip labels in the chronogram database.
+A DateLife search is implemented by matching processed taxon names provided by the user to tip labels in the chronogram database.
 Chronograms with at least two matching taxon names on their tip labels are identified and pruned down to preserve only the matched taxa.
@@
 Optionally, to estimate branch lengths proportional to substitution rates on these topologies, DateLife can mine the Barcode of Life Data System, BOLD [@ratnasingham2007bold] to obtain genetic markers for the input taxa.
-These markers are aligned with MUSCLE by default [@edgar2004muscle], or MAFFT [@katoh2009multiple]. 
+These markers are aligned with MUSCLE [@edgar2004muscle] (by default) or MAFFT [@katoh2009multiple]. 
 This alignment can be used to estimate branch lengths on input topologies that lack branch lengths.
@@
 By definition, the matrices store total distance (time from tip to tip), hence, node ages correspond to half the values stored in the patristic distance matrices.
+This assumes that the terminal taxa are coeval and occur at the present.
 A table of congruified node ages that can be used as secondary calibrations is stored as a `congruifiedCalibrations` object.
@@
 This minimizes age variance in the resulting chronogram [@Webb2008].
-BLADJ does not incorporate branch lengths even when they are present.
+BLADJ does not incorporate input phylogeny branch lengths even when they are present.
 When there is conflict in ages between nodes with calibration data, the algorithm ignores node ages that are older than the age of a parent node. 
@@
 PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees.
-treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method
-[@Smith2012].
+treePL [@Smith2012], is a semi-parametric, rate-smoothing, penalized likelihood dating method
+[@Sanderson2002].
 The MrBayes  [@Huelsenbeck2001; @Ronquist2003] approach in datelife uses the secondary calibrations as priors on node ages.
@@
 present, unit tests cover more than 40% of `datelife`'s code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->
+Unit testing helps identify potential issues as code is updated or, more critically, as services code relies upon may change.
 
@@
 
-We illustrate the DateLife algorithm using a group within the Passeriform birds encompassing the family of true finches, Fringillidae as case study.
+We illustrate the DateLife algorithm using a group within the Passeriform birds encompassing the family of true finches, Fringillidae, as case study.
 The first example analyses 6 bird species and shows all steps of the algorithm.
@@
 As expected, more inclusive nodes (e.g., node "n1") have more age data than less inclusive nodes (e.g., node "n5").
-The processing step allowed discovering five data points for node "n4" that would not have had any data otherwise. \textcolor{red}{I don't understand what this sentence means}
+The name resolution step allowed discovering five data points for node "n4" that would not have had any data otherwise due to name mismatch. 
 Age summary statistics per node were calculated and used as secondary calibrations to date the tree topology using the BLADJ algorithm.
@@
 This will get all recognised species names within the named taxon. 
-The Fringillidae has species, according to the Open Tree of Life taxonomy.
+The Fringillidae has species, according to the Open Tree of Life taxonomy. <!--BCO: Feel like we need a number here, though it's good that the clade hasn't gone extinct ;-) -->
 ??? We have date estimates for `r length(dquery$cleaned_names)` of those species.
@@
 ??? There is an option to include all, right?
-According to the OpenTree of life synthetic phylogeny, Fringillidae is not a monophyletic family, so this taxon-constrained approach returns a topology and ages for the species in that named group.
+According to the OpenTree of life synthetic phylogeny, Fringillidae is not monophyletic, so this taxon-constrained approach returns a topology and ages for the species in that named group.
+<!--BCO: it's ambiguous: it gets a topology and age for the larger clade that includes all Fringillidae species?
+Or for the paraphyletic Fringillidae group?-->
 
@@
 The final step is to summarize the available information using two alternative types of summary chronograms, median and SDM.
-As explained in the "Description" section, data from source chronograms is first summarised into a single distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig.
+As explained in the "Description" section, data from source chronograms is first summarised into a single patristic distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig.
 \ref{fig:fringillidages}).
-Median sxummary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
+Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
 <!-- ????????????????? Say some things about the results!
@@
 
-In 5 studies, more inclusive nodes have older original ages, and less inclusive nodes have younger original ages than their cross-validated age estimates. 
-\textcolor{red}{more inclusive nodes have to be older!
-:P I think there is something differnet you are trying to say here - maybe "Datelife tended to underestimate ages for more inclusive nodes relative to the orginal estimate, and overestimate ages for nodes closer to the tips"}
+For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka 'closer to the root') relative to the orginal estimate, and overestimate ages for nodes closer to the tips.
 Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. \ref{fig:cvXYdiffs}).
@@
 `datelife` makes state-of-the-art information on time of lineage divergence easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
-It is an open service that does not require any expert biological knowledge from users --besides the names of the organisms they want to work with, for any of its functionality.
+It is an open service that does not require any expert biological knowledge from users, besides the names of the organisms or clade they want to work with, for any of its functionality.
 
@@
 `datelife`'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
+<!--BCO: Nicely done-->
 Users can also upload new chronograms to OpenTree themselves, and trigger an update of their local `datelife` database to incorporate the new chronograms, to have them immediately available for analysis.
+<!--BCO: What is immediately?
+Also, it's not just uploading -- they have to include metadata on what brlen mean, maybe?
+indicate the ingroup/outgroup, do TNRS within OToL, right?
+Maybe this has improved.
+I remember being frustrated trying to load in giant trees for datelife use b/c I had to click on the ingroup node-->
+<!--BCO; this duplicates an earlier section and feels more methods-y. Maybe move up and merge?-->
 
@@
 Incorporation of more chronograms into `datelife`'s database will continue to improve its services.
-One option to increase chronogram number in the database is the Dryad data repository.
+One option to increase the of chronograms in the DateLife database is the Dryad data repository, which now holds some chronograms.
 Methods to automatically mine chronograms from Dryad could be designed and implemented.
@@
 was constructed using age data from 2,274 published chronograms [@Hedges2015].
-Unfortunately, the source chronograms used as input data for this tree are not available in computer readable format for reuse or reanalysis.
-Access to the input chronograms used to generate the Hedges et al. summary tree would improve measures of uncertainty in DateLife, but they are currently available only as image files (timetree.org).
-<!--IN VIOLATION OF NSF RULES! but I don't suppose we should say that.-->
+<!--BCO: I just checked: TIMETREEE NOW ALLOWS SOURCE TREE DOWNLOADS!
+This is great news.
+They have a copyright (not legal in US on trees, IMHO) notice on individual trees, but they do say "To download all individual study timetrees, please send an email request to info@timetree.org." -- I think this could be great to do-->
+
 We would like to emphasize on the importance of sharing chronogram data,  including systematically curated metadata, into open repositories, such as OpenTree's Phylesystem [@mctavish2015phylesystem] for the benefit of the scientific community as a whole.
@@
 Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used [@magallon2010using, @magallon2015metacalibrated].
-Several characteristics of the data used for dating analyses as well as from the output chronogram itself, could be used to score quality of source chronograms.
+Several characteristics of the data used for dating analyses, as well as from the output chronogram itself, could be used to score quality of source chronograms.
 Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
 DateLife provides an opportunity to capture concordance and conflict among date estimates, which can also be used as a metric for chronogram reliability.
+Its open database of trees allows other researchers to do such analyses themselves reproducibly and without needing permission, though of course they should follow proper citation practices, especially for the source trees.
 
@@
 Conflict among alternative date estimates is common in the literature
-(see, for example, the ongoing debate about crown group age of angiosperms [@sauquet2021age; @barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time]. 
+(see, for example, the robust ongoing debate about crown group age of angiosperms [@sauquet2021age; @barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time]. 
 Using a different chronogram may return different results [@title2016macrophylogenies] particularly for studies of species diversification, or focussed on the timing of evolutionary events.
@@
 Summarizing chronograms might also imply summarizing fundamentally distinct evolutionary hypotheses.
-For example, two different researchers working on the same clade both carefully select and argument their choices of fossil calibrations.
+For example, two different researchers working on the same clade could both carefully select and argue for their choices of fossil calibrations.
 Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses the same one to calibrate outside the clade, the resulting age estimates will often differ substantially, as the placement of calibrations as stem or crown group is proved to deeply affect estimated times of lineage divergence [@sauquet2013practical].
 Trying to summarize the resulting chronograms into a single one using simple summary statistics can erase many types of relevant information from the source chronograms.
-Accordingly, the prevailing view is that we should favor time of lineage divergence estimates obtained from a single analysis, using fossil data as primary sources of calibrations, and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history [@antonelli2017supersmart].
+Accordingly, the prevailing view is that we should favor lineage divergence estimates obtained from a single analysis, using fossil data as primary sources of calibrations, and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history, as for example done by [@antonelli2017supersmart].
 <!-- Will this be implemented in future `datelife` versions?
-Probably not-->
-However, the exercise of summarizing different chronograms has the potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypothesis.
-Choosing the elements of the chronograms that we are going to keep and the ones that we are going to discard is key, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram [@sauquet2021age].
+Probably not.
+-->
+There have been attempts to create fossil calibration databases <!--REF TO ADD: Ksepka, D.T., J.F.
+Parham, J.F.
+Allman, M.J.
+Benton, M.T.
+Carrano, K.A.
+Cranston, P.C.
+J.
+Donoghue, J.J.
+Head, E.J.
+Hermsen, R.B.
+Irmis, W.G.
+Joyce, M.
+Kohli, K.S.
+Lamm, D.
+Leehr, J.S.L.
+Patané, P.D.
+Polly, M.J.
+Phillips, N.A.
+Smith, N.D.
+Smith, M.
+van Tuinen, J.L.
+Ware, R.C.M.
+Warnock.
+2015.The Fossil Calibration Database, A New Resource for Divergence Dating.
+Systematic Biology: doi:10.1093/sysbio/syv025.-->, though these still have room to grow.
+<!--BCO: Polite way to say 'no new additions in four years?'-->
+However, the exercise of summarizing different chronograms has the potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypotheses.
+Choosing the elements of the chronograms that we are going to keep and the ones that we are going to discard is key, since we are potentially losing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram [@sauquet2021age].
 <!-- Example -->
@@
 
-Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603  to Datelife project; DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), ABI-1458572 to the Phylotastic project, and ABI-1759846 to the Open Tree of Life project.
+Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to the Datelife project; DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), ABI-1458572 to the Phylotastic project, and ABI-1759846 to the Open Tree of Life project.
 
@@
 
-The DateLife project was born as a prototype tool aiming to provide these services, and was developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA [@Stoltzfus2013].
+The DateLife project was born as a prototype tool aiming to provide these services, and was initially developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA [@Stoltzfus2013].
 We thank colleagues from the O'Meara Lab at the University

### 2022-05-23 manuscript/manuscript.Rmd

@@
 taxize [@Chamberlain2013; @Chamberlain2018].
-Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife algorithm: providing an input, searching a chronogram database, and summarizing results from the search.
+Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife algorithm: creating a search query, searching a database, and summarizing results from the search.
 
-<!--_**Providing an input.--**_ -->
+<!--_**Processing an input.--**_ -->
 
-## Processing an input
+##  Creating a search query
 
@@
 treePL [@Smith2012], is a semi-parametric, rate-smoothing, penalized likelihood dating method
-[@Sanderson2002].
+[@sanderson2002].
 The MrBayes  [@Huelsenbeck2001; @Ronquist2003] approach in datelife uses the secondary calibrations as priors on node ages.
@@
 
-We illustrate the DateLife algorithm using a group within the Passeriform birds encompassing the family of true finches, Fringillidae, as case study.
+We illustrate the DateLife algorithm using a family within the Passeriform birds encompassing the true finches, Fringillidae, as case study.
 The first example analyses 6 bird species and shows all steps of the algorithm.
-The second example is an application analysing all 289 species in the family Fringillidae for which we have date informtaion.
+The second example is an application analysing 289 species in the family Fringillidae that are included in the NCBI taxonomy.
 <!--We used `r cite_r("r-references.bib")` for all our analyses.-->
@@
 All nodes have age data.
+
+_**Summarizing search results.-**_
 DateLife used OpenTree's synthetic tree topology for these taxa and mapped age data to nodes in the tree.
@@
 Age summary statistics per node were calculated and used as secondary calibrations to date the tree topology using the BLADJ algorithm.
-Age data for node "n2" was excluded as final calibration because it is older than age data of a more inclusive node (n1).
+Age data for node "n2" was excluded as final calibration because it is older than age data of the more inclusive node "n1".
 
@@
 
-To estimate ages for the taxa in the family Fringillidae, we ran a `datelife` query using the "get species from taxon" flag. 
-This will get all recognised species names within the named taxon. 
-The Fringillidae has species, according to the Open Tree of Life taxonomy. <!--BCO: Feel like we need a number here, though it's good that the clade hasn't gone extinct ;-) -->
-??? We have date estimates for `r length(dquery$cleaned_names)` of those species.
+_**Creating a query.-**_
+To estimate ages for species in the family Fringillidae, we ran a `datelife` query using the "get species from taxon" flag,
+which gets all recognized species names within a named group from a taxonomy of choice (options are Open Tree of Life, NCBI, GBIF, or IRMNG).
+Following the NCBI taxonomy, the Fringillidae has `r length(dquery$cleaned_names)` species.
+<!--BCO: Feel like we need a number here, though it's good that the clade hasn't gone extinct ;-) -->
 
-??? There is an option to include all, right?
-According to the OpenTree of life synthetic phylogeny, Fringillidae is not monophyletic, so this taxon-constrained approach returns a topology and ages for the species in that named group.
-<!--BCO: it's ambiguous: it gets a topology and age for the larger clade that includes all Fringillidae species?
-Or for the paraphyletic Fringillidae group?-->
+_**Searching the database.-**_
+The next step is to use this curated set of species taxon names to identify all chronograms that contain at least two Fringillidae species.
+Once identified, the algorithm proceeds to prune matching chronograms to keep Fringillidae species names on tips only, and transform these pruned chronograms to pairwise distance matrices.
+The `datelife` search revealed `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
 
-The next step is to run a `datelife` search on this curated set of species taxon names that will find all chronograms that contain at least two species.
-The algorithm proceeds to prune the trees to keep matching species names on tips only, and transform the pruned trees to pairwise distance matrices.
-There are `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
-The final step is to summarize the available information using two alternative types of summary chronograms, median and SDM.
-As explained in the "Description" section, data from source chronograms is first summarised into a single patristic distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated tree with the program BLADJ (Fig.
+_**Summarizing search results.-**_
+The final step is to summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM.
+As explained in the "Description" section, a tree topology to summarize age data upon is required.
+By default, `datelife` uses the Open Tree of Life synthetic phylogeny as summarizing topology. According to this phylogeny, Fringillidae is not a monophyletic family
+[@burns2014phylogenetics;
+@burleigh2015building;
+@Jetz2012;
+@lerner2011multilocus;
+@lovette2010comprehensive;
+@pulgarin2013multilocus;
+@chaves2013biogeography;
+@lamichhaney2015evolution;
+@parchman2007coevolution;
+@selvatti2015paleogene;
+@moyle2016tectonic;
+@odeen2011evolution;
+@claramunt2015new;
+@beresford2005african;
+@kimball2019phylogenomic;
+@barker2015new;
+@klicka2014comprehensive;
+@barker2013going;
+@treplin2008molecular;
+@price2014niche;
+@hackett2008phylogenomic;
+@barker2014mitogenomic;
+@gibb2015new;
+@barker2004phylogeny;
+@zuccon2012phylogenetic;
+@oliveros2019earth;
+@johansson2008phylogenetic;
+@alstrom2014discovery;
+@powell2014comprehensive;
+@bryson2014diversification;
+@packert2012horizontal;
+@tietze2013complete].
+
+Hence, DateLife's taxon-constrained approach returns a topology and ages for the species in a named group.
+
+<!--BCO: it's ambiguous: it gets a topology and age for the larger clade that includes all Fringillidae species? Or for the paraphyletic Fringillidae group?-->
+
+Age data from source chronograms is summarised into a single patristic distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
 \ref{fig:fringillidages}).
+
 Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
@@
 We performed a cross validation analysis of the DateLife workflow using the Fringillidae chronograms.
-We used the individual tree topologies from each of the ??? TODO CHECK NUMBER 19 source chronograms from ?? studies as inputs, treating their node ages as unknown.
-We estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.
+We used the individual tree topologies from each of the 19 source chronograms from 13 studies as inputs, treating their node ages as unknown.
+We then estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.
 
-We found that node ages from original study, and ages estimated using all other age data available are correlated  \textcolor{red}{R^2 value?} (Supplementary Fig. \ref{fig:cvXY}).
+We found that node ages from original study, and ages estimated using all other age data available are correlated (Supplementary Fig. \ref{fig:cvXY}).
 
@@
 These differences in approach which can deepen the already substantial variation in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
-For eaxmple, the chronograms from  @burns2014phylogenetics tend to have much older date estimates for the same nodes than other inferences (Figure 5). 
-This study did not use any primary calibrations, and inferred dates using an substitution rate estimates across birds \textcolor{red}{(cite Weir, J.T., Schluter, D., 2008.
-Calibrating the avian molecular clock.
-Mol.
-Ecol.
-17,
-231–2328).}
-\textcolor{red}{MORE ON the FINCH examples and the cross validation here}
+For example, the chronograms from @burns2014phylogenetics tend to have much older date estimates for the same nodes than other inferences (Figure 5).
+This study did not use any primary calibrations, and inferred dates using an substitution rate estimates across birds [@weir2008calibrating]
+<!--\textcolor{red}{MORE ON the FINCH examples and the cross validation here}
+-->
 
@@
 Choosing the elements of the chronograms that we are going to keep and the ones that we are going to discard is key, since we are potentially losing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram [@sauquet2021age].
-<!-- Example -->
-\textcolor{red}{??? it's not clear to me what this last part means! What might you be losing?}
+<!-- Example
+\textcolor{red}{EJM: ??? it's not clear to me what this last part means! What might you be losing?}
+-->
 

### 2022-05-23 manuscript/manuscript.tex

@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4530}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4607}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 
-Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 425 Hesler Biology Building, Knoxville, TN 37996, USA.
+Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 446 Hesler Biology Building, Knoxville, TN 37996, USA.
 
@@
 Date estimates for times of evolutionary divergences are key data for research in the natural sciences.
-These estimates also provide valuable information for for education, science communication and policy decisions.
-Although achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram), is a difficult and time-consuming task, the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
+These estimates also provide valuable information for for education, science communication, and policy decisions.
+Although achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram) is a difficult and time-consuming task, the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
 When these estimates are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on time of evolutionary origin is exposed in a programatic and reusable way.
 Intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
-Here we present DateLife, a service implemented as an R package and an Rshiny website application available at www.datelife.org, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
-The main DateLife workflow constructs a chronogram for any given combination of taxon names, by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from TreeBASE database as well.
+Here we present DateLife, a service implemented as an R package and an R Shiny website application available at www.datelife.org, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
+The main DateLife workflow constructs a chronogram for any given combination of taxon names by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from the TreeBASE database as well.
 We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
@@
 
-DateLife starts by processing a user input consisting of at least two taxon names, which can be provided as a comma separated character string, or as tip labels on a tree.
+DateLife starts by processing an input consisting of at least two taxon names, which can be provided as a comma separated character string or as tip labels on a tree.
 If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis et al., 2004).
@@
 The processed input taxon names are saved as an R object of a newly defined class \texttt{datelifeQuery} that is used in the following steps.
-This object contains the standardized names, the corresponding taxonomic id numbers, and the topology of the input tree if any was provided.
+This object contains the standardized names, the corresponding OpenTree taxonomic id numbers, and the topology of the input tree if any was provided.
 
@@
 Published chronograms can be added to Phylesystem by any user, at any time, and are immediately publicly available (\url{https://tree.opentreeoflife.org/curator}).
-This facilitates an immediate update of DateLife's chronogram database to include new chronogram data on a following search.
-
-A DateLife search is implemented by matching processed taxon names provided by the user, to tip labels in the chronogram database.
+A user may then update their local cache of DateLife's chronogram database to include new chronogram data on a following search.
+
+A DateLife search is implemented by matching processed taxon names provided by the user to tip labels in the chronogram database.
 Chronograms with at least two matching taxon names on their tip labels are identified and pruned down to preserve only the matched taxa.
@@
 Optionally, to estimate branch lengths proportional to substitution rates on these topologies, DateLife can mine the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain genetic markers for the input taxa.
-These markers are aligned with MUSCLE by default (Edgar, 2004), or MAFFT (Katoh, Asimenos, \& Toh, 2009).
+These markers are aligned with MUSCLE (Edgar, 2004) (by default) or MAFFT (Katoh, Asimenos, \& Toh, 2009).
 This alignment can be used to estimate branch lengths on input topologies that lack branch lengths.
@@
 By definition, the matrices store total distance (time from tip to tip), hence, node ages correspond to half the values stored in the patristic distance matrices.
+This assumes that the terminal taxa are coeval and occur at the present.
 A table of congruified node ages that can be used as secondary calibrations is stored as a \texttt{congruifiedCalibrations} object.
@@
 This minimizes age variance in the resulting chronogram (Campbell O. Webb et al., 2008).
-BLADJ does not incorporate branch lengths even when they are present.
+BLADJ does not incorporate input phylogeny branch lengths even when they are present.
 When there is conflict in ages between nodes with calibration data, the algorithm ignores node ages that are older than the age of a parent node.
@@
 PATHd8 is a non-clock, rate-smoothing method (Britton et al., 2007) to date trees.
-treePL, is a semi-parametric, rate-smoothing, penalized likelihood dating method
-(Stephen A. Smith \& O'Meara, 2012).
+treePL (Stephen A. Smith \& O'Meara, 2012), is a semi-parametric, rate-smoothing, penalized likelihood dating method
+(Michael J. Sanderson, 2002).
 The MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003) approach in datelife uses the secondary calibrations as priors on node ages.
@@
 present, unit tests cover more than 40\% of \texttt{datelife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).
+Unit testing helps identify potential issues as code is updated or, more critically, as services code relies upon may change.
 
@@
 
-We illustrate the DateLife algorithm using a group within the Passeriform birds known as the family of true finches, Fringillidae, as case study.
+We illustrate the DateLife algorithm using a family within the Passeriform birds encompassing the true finches, Fringillidae, as case study.
 The first example analyses 6 bird species and shows all steps of the algorithm.
-The second example is an application analysing 289 species in the family Fringillidae included in the NCBI taxonomy.
+The second example is an application analysing 289 species in the family Fringillidae that are included in the NCBI taxonomy.
 
@@
 As expected, more inclusive nodes (e.g., node ``n1'') have more age data than less inclusive nodes (e.g., node ``n5'').
+The name resolution step allowed discovering five data points for node ``n4'' that would not have had any data otherwise due to name mismatch.
 Age summary statistics per node were calculated and used as secondary calibrations to date the tree topology using the BLADJ algorithm.
@@
 Hence, DateLife's taxon-constrained approach returns a topology and ages for the species in a named group.
-Age data from source chronograms is summarised into a single distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
+
+Age data from source chronograms is summarised into a single patristic distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
 \ref{fig:fringillidages}).
@@
 
-In 5 studies, DateLife presented a tendency to underestimate ages for more inclusive nodes, and overestimate ages for nodes closer to the tips, relative to the original age estimates.
-Accordingly, root ages estimated with DateLife's cross-validated ages are generally younger relative to the age root from the original study (Supplementary Fig. \ref{fig:cvXYdiffs}).
+For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka `closer to the root') relative to the orginal estimate, and overestimate ages for nodes closer to the tips.
+Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. \ref{fig:cvXYdiffs}).
 
@@
 \texttt{datelife} makes state-of-the-art information on time of lineage divergence easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
-It is an open service that does not require any expert biological knowledge from users --besides the names of the organisms they want to work with, for any of its functionality.
+It is an open service that does not require any expert biological knowledge from users, besides the names of the organisms or clade they want to work with, for any of its functionality.
 
@@
 Incorporation of more chronograms into \texttt{datelife}'s database will continue to improve its services.
-One option to increase chronogram number in the database is the Dryad data repository.
+One option to increase the of chronograms in the DateLife database is the Dryad data repository, which now holds some chronograms.
 Methods to automatically mine chronograms from Dryad could be designed and implemented.
@@
 was constructed using age data from 2,274 published chronograms (Hedges et al., 2015).
-Unfortunately, the source chronograms used as input data for this tree are not available in computer readable format for reuse or reanalysis.
-Access to the input chronograms used to generate the Hedges et al.~summary tree would improve measures of uncertainty in DateLife, but they are currently available only as image files (timetree.org).
+
 We would like to emphasize on the importance of sharing chronogram data, including systematically curated metadata, into open repositories, such as OpenTree's Phylesystem (McTavish et al., 2015) for the benefit of the scientific community as a whole.
@@
 Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández (2015).
-Several characteristics of the data used for dating analyses as well as from the output chronogram itself, could be used to score quality of source chronograms.
+Several characteristics of the data used for dating analyses, as well as from the output chronogram itself, could be used to score quality of source chronograms.
 Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
 DateLife provides an opportunity to capture concordance and conflict among date estimates, which can also be used as a metric for chronogram reliability.
+Its open database of trees allows other researchers to do such analyses themselves reproducibly and without needing permission, though of course they should follow proper citation practices, especially for the source trees.
 
@@
 Schenk (2016) showed that in simulations divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with Bayesian inference methods when priors are implemented in similar ways in both analyses.
-However, secondary calibrations are useful in dating methods that do not require setting priors, such as penalized likelihood (Sanderson, 2003), or as fixed ages for nodes.
+However, secondary calibrations are useful in dating methods that do not require setting priors, such as penalized likelihood (Michael J.
+Sanderson, 2003), or as fixed ages for nodes.
 Certainly, further studies are required to fully understand the effect of using secondary calibrations on time estimates and downstream analyses.
@@
 Conflict among alternative date estimates is common in the literature
-(see, for example, the ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón et al., 2015; Ramshaw et al., 1972; Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021).
+(see, for example, the robust ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón et al., 2015; Ramshaw et al., 1972; Michael J.
+Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021).
 Using a different chronogram may return different results (Title \& Rabosky, 2016) particularly for studies of species diversification, or focussed on the timing of evolutionary events.
@@
 Summarizing chronograms might also imply summarizing fundamentally distinct evolutionary hypotheses.
-For example, two different researchers working on the same clade both carefully select and argument their choices of fossil calibrations.
+For example, two different researchers working on the same clade could both carefully select and argue for their choices of fossil calibrations.
 Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses the same one to calibrate outside the clade, the resulting age estimates will often differ substantially, as the placement of calibrations as stem or crown group is proved to deeply affect estimated times of lineage divergence (Sauquet, 2013).
 Trying to summarize the resulting chronograms into a single one using simple summary statistics can erase many types of relevant information from the source chronograms.
-Accordingly, the prevailing view is that we should favor time of lineage divergence estimates obtained from a single analysis, using fossil data as primary sources of calibrations, and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history (Antonelli et al., 2017).
-However, the exercise of summarizing different chronograms has the potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypothesis.
-Choosing the elements of the chronograms that we are going to keep and the ones that we are going to discard is key, since we are potentially loosing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram (Sauquet et al., 2021).
+Accordingly, the prevailing view is that we should favor lineage divergence estimates obtained from a single analysis, using fossil data as primary sources of calibrations, and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history, as for example done by (Antonelli et al., 2017).
+There have been attempts to create fossil calibration databases , though these still have room to grow.
+However, the exercise of summarizing different chronograms has the potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypotheses.
+Choosing the elements of the chronograms that we are going to keep and the ones that we are going to discard is key, since we are potentially losing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram (Sauquet et al., 2021).
 
@@
 
-Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to Datelife project; DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), ABI-1458572 to the Phylotastic project, and ABI-1759846 to the Open Tree of Life project.
+Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to the Datelife project; DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), ABI-1458572 to the Phylotastic project, and ABI-1759846 to the Open Tree of Life project.
 
@@
 
-The DateLife project was born as a prototype tool aiming to provide these services, and was developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA (Stoltzfus et al., 2013).
+The DateLife project was born as a prototype tool aiming to provide these services, and was initially developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA (Stoltzfus et al., 2013).
 We thank colleagues from the O'Meara Lab at the University
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-sanderson2002}{}}%
+Sanderson, Michael J.
+(2002).
+{Estimating Absolute Rates of Molecular Evolution and Divergence Times: A Penalized Likelihood Approach}.
+\emph{Molecular Biology and Evolution}, \emph{19}(1), 101--109.
+\url{https://doi.org/10.1093/oxfordjournals.molbev.a003974}
+
 \leavevmode\vadjust pre{\hypertarget{ref-sanderson2003r8s}{}}%
-Sanderson, M. J. (2003). r8s: Inferring absolute rates of molecular evolution and divergence times in the absence of a molecular clock. \emph{Bioinformatics}, \emph{19}(2), 301--302.
+Sanderson, Michael J. (2003). r8s: Inferring absolute rates of molecular evolution and divergence times in the absence of a molecular clock. \emph{Bioinformatics}, \emph{19}(2), 301--302.
 
 \leavevmode\vadjust pre{\hypertarget{ref-sanderson2001sources}{}}%
-Sanderson, M.
-J., \& Doyle, J.
+Sanderson, Michael J., \& Doyle, J.
 A.

### 2022-05-23 manuscript/manuscript.Rmd
newest geiger citation

@@
 bold [@Chamberlain2018],
-geiger [@Harmon2008],
+geiger [@pennell2014geiger],
 paleotree [@Bapst2012a],

### 2022-05-23 manuscript/manuscript.tex
newest geiger citation

@@
 bold (Chamberlain et al., 2019),
-geiger (Harmon, Weir, Brock, Glor, \& Challenger, 2008),
+geiger (Pennell et al., 2014),
 paleotree (Bapst, 2012),
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Harmon2008}{}}%
-Harmon, L., Weir, J., Brock, C., Glor, R., \& Challenger, W. (2008). {GEIGER: investigating evolutionary radiations}. \emph{Bioinformatics}, \emph{24}, 129--131.
-
 \leavevmode\vadjust pre{\hypertarget{ref-Hedges2015}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-pennell2014geiger}{}}%
+Pennell, M.
+W., Eastman, J.
+M., Slater, G.
+J., Brown, J.
+W., Uyeda, J.
+C., FitzJohn, R.
+G., \ldots{} Harmon, L.
+J.
+(2014).
+Geiger v2.
+0: An expanded suite of methods for fitting macroevolutionary models to phylogenetic trees.
+\emph{Bioinformatics}, \emph{30}(15), 2216--2218.
+
 \leavevmode\vadjust pre{\hypertarget{ref-posadas2006historical}{}}%

### 2022-05-26 manuscript/figures-alt.tex
cross validation XY figure
addressing topologies comment

@@
 \begin{figure}[!h]
+\includegraphics{../figures/fringillidae-topologies/fringillidae-topology.pdf}
+\caption{Topologies obtained from OpenTree's synthetic phylogeny.
+A) Topology od 2333 tips, encompassing bird species within the Fringillidae following the NCBI taxonomy (black) as well as all other bird species that share the same mrca node in the synthetic tree (purple).
+B) Topology encompassing bird species within the Fringillidae only, resulting from pruning purple branches from topology A.
+Bird species belonging to the family Fringillidae do not form a monophyletic group.
+(Alström et al. \protect\hyperlink{ref-}{2014},
+Barker, Cibois, Schikler, Feinstein, \& Cracraft \protect\hyperlink{ref-barker2004phylogeny}{2004},
+Barker et al. \protect\hyperlink{ref-barker2013going}{2013},
+Barker \protect\hyperlink{ref-barker2014mitogenomic}{2014},
+Barker et al. \protect\hyperlink{ref-barker2015new}{2015},
+Beresford, Barker, Ryan, \& Crowe \protect\hyperlink{ref-beresford2005african}{2005},
+Bryson Jr et al. \protect\hyperlink{ref-bryson2014diversification}{2014},
+Burleigh, Kimball, \& Braun \protect\hyperlink{ref-burleigh2015building}{2015},
+Burns et al. \protect\hyperlink{ref-burns2014phylogenetics}{2014},
+Chaves, Hidalgo, \& Klicka \protect\hyperlink{ref-chaves2013biogeography}{2013},
+Claramunt \& Cracraft \protect\hyperlink{ref-claramunt2015new}{2015},
+Gibb et al. \protect\hyperlink{ref-gibb2015new}{2015},
+Hackett et al. \protect\hyperlink{ref-hackett2008phylogenomic}{2008},
+Jetz et al. \protect\hyperlink{ref-Jetz2012}{2012},
+Johansson, Fjeldså, \& Bowi \protect\hyperlink{ref-johansson2008phylogenetic}{200},
+Kimball et al. \protect\hyperlink{ref-kimball2019phylogenomic}{2019},
+Klicka et al. \protect\hyperlink{ref-klicka2014comprehensive}{2014},
+Lamichhaney et al. \protect\hyperlink{ref-lamichhaney2015evolution}{2015},
+Lerner, Meyer, James, Hofreiter, \& Fleischer \protect\hyperlink{ref-lerner2011multilocus}{2011},
+Lovette et al. \protect\hyperlink{ref-lovette2010comprehensive}{2010},
+Moyle et al. \protect\hyperlink{ref-moyle2016tectonic}{2016},
+Ödeen, Håstad, \& Alström \protect\hyperlink{ref-odeen2011evolution}{2011},
+Oliveros et al. \protect\hyperlink{ref-oliveros2019earth}{2019},
+Päckert et al. \protect\hyperlink{ref-packert2012horizontal}{2012},
+Parchman, Benkman, \& Mezquida \protect\hyperlink{ref-parchman2007coevolution}{2007},
+Powell et al. \protect\hyperlink{ref-powell2014comprehensive}{2014},
+Price et al. \protect\hyperlink{ref-price2014niche}{2014},
+Pulgarín-R, Smith, Bryson Jr, Spellman, \& Klicka \protect\hyperlink{ref-pulgarin2013multilocus}{2013},
+Selvatti, Gonzaga, \& Moraes Russo \protect\hyperlink{ref-selvatti2015paleogene}{2015},
+Tietze, Päckert, Martens, Lehmann, \& Sun \protect\hyperlink{ref-tietze2013complete}{2013},
+Treplin et al. \protect\hyperlink{ref-treplin2008molecular}{2008},
+Zuccon, Prŷs-Jones, Rasmussen, \& Ericson \protect\hyperlink{ref-zuccon2012phylogenetic}{2012})
+}
+\label{fig:fringillidae-topologies}
+\end{figure}
+
+\begin{figure}[!h]
 \includegraphics{../figures/figure-fringillidae/median_and_calibration_ages_simple.png}

### 2022-05-26 manuscript/manuscript.Rmd
addressing topologies comment

@@
 nocite: |
-  @barker2013going, @barker2015new, @burns2014phylogenetics, @claramunt2015new, @gibb2015new, @Hedges2015, @hooper2017chromosomal, @Jetz2012, @price2014niche
+  @barker2013going, @barker2015new, @burns2014phylogenetics, @claramunt2015new, @gibb2015new, @Hedges2015, @hooper2017chromosomal, @Jetz2012, @price2014niche,
+  @alstrom2014discovery,
+  @barker2004phylogeny,
+  @barker2013going,
+  @barker2014mitogenomic,
+  @barker2015new,
+  @beresford2005african,
+  @bryson2014diversification,
+  @burleigh2015building,
+  @burns2014phylogenetics,
+  @chaves2013biogeography,
+  @claramunt2015new,
+  @gibb2015new,
+  @hackett2008phylogenomic,
+  @Jetz2012,
+  @johansson2008phylogenetic,
+  @kimball2019phylogenomic,
+  @klicka2014comprehensive, 
+  @lamichhaney2015evolution,
+  @lerner2011multilocus,
+  @lovette2010comprehensive,
+  @moyle2016tectonic,
+  @odeen2011evolution,
+  @oliveros2019earth,
+  @packert2012horizontal,
+  @parchman2007coevolution,
+  @powell2014comprehensive,
+  @price2014niche,
+  @pulgarin2013multilocus,
+  @selvatti2015paleogene,
+  @tietze2013complete,
+  @treplin2008molecular,
+  @zuccon2012phylogeneti,
 ---
@@
 
-Here we present the DateLife project and its core software application, available as an R package [@datelifeR], and as an online Rshiny interactive website at www.datelife.org, which captures data from published chronograms, and make these data readily accessible to users for reuse and reanalysis.
-The software features key elements for scientific reproducibility, such as a versioned, open and fully public source database [@mctavish2015phylesystem], data stored and available in a computer readable format [@vos2012nexml], automated and programmatic ways of accessing the data [@Stoltzfus2013] and methods to summarize and compare the data.
+Here we present the DateLife project which has the main goal of capturing data from published chronograms, and making these data readily accessible to users for reuse and reanalysis.
+DateLife's core software application is available as an R package [@datelifeR], and as an online Rshiny interactive website at www.datelife.org.
+It features key elements for scientific reproducibility, such as a versioned, open and fully public source database [@mctavish2015phylesystem], data stored and available in a computer readable format [@vos2012nexml], automated and programmatic ways of accessing the data [@Stoltzfus2013] and methods to summarize and compare the data.
 
@@
 As explained in the "Description" section, a tree topology to summarize age data upon is required.
-By default, `datelife` uses the Open Tree of Life synthetic phylogeny as summarizing topology. According to this phylogeny, Fringillidae is not a monophyletic family
-[@burns2014phylogenetics;
-@burleigh2015building;
-@Jetz2012;
-@lerner2011multilocus;
-@lovette2010comprehensive;
-@pulgarin2013multilocus;
-@chaves2013biogeography;
-@lamichhaney2015evolution;
-@parchman2007coevolution;
-@selvatti2015paleogene;
-@moyle2016tectonic;
-@odeen2011evolution;
-@claramunt2015new;
-@beresford2005african;
-@kimball2019phylogenomic;
-@barker2015new;
-@klicka2014comprehensive;
-@barker2013going;
-@treplin2008molecular;
-@price2014niche;
-@hackett2008phylogenomic;
-@barker2014mitogenomic;
-@gibb2015new;
-@barker2004phylogeny;
-@zuccon2012phylogenetic;
-@oliveros2019earth;
-@johansson2008phylogenetic;
-@alstrom2014discovery;
-@powell2014comprehensive;
-@bryson2014diversification;
-@packert2012horizontal;
-@tietze2013complete].
-
+By default, `datelife` uses the Open Tree of Life synthetic phylogeny as summarizing topology.
+According to this phylogeny, species belonging to the family Fringillidae do not form a monophyletic group.
 Hence, DateLife's taxon-constrained approach returns a topology and ages for the species in a named group.

### 2022-05-26 manuscript/manuscript.tex
addressing topologies comment

@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4607}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4586}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 
-Here we present the DateLife project and its core software application, available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org, which captures data from published chronograms, and make these data readily accessible to users for reuse and reanalysis.
-The software features key elements for scientific reproducibility, such as a versioned, open and fully public source database (McTavish et al., 2015), data stored and available in a computer readable format (Vos et al., 2012), automated and programmatic ways of accessing the data (Stoltzfus et al., 2013) and methods to summarize and compare the data.
+Here we present the DateLife project which has the main goal of capturing data from published chronograms, and making these data readily accessible to users for reuse and reanalysis.
+DateLife's core software application is available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org.
+It features key elements for scientific reproducibility, such as a versioned, open and fully public source database (McTavish et al., 2015), data stored and available in a computer readable format (Vos et al., 2012), automated and programmatic ways of accessing the data (Stoltzfus et al., 2013) and methods to summarize and compare the data.
 
@@
 As explained in the ``Description'' section, a tree topology to summarize age data upon is required.
-By default, \texttt{datelife} uses the Open Tree of Life synthetic phylogeny as summarizing topology. According to this phylogeny, Fringillidae is not a monophyletic family
-(Alström et al., 2014; Barker, 2014; Barker et al., 2013, 2015; Barker, Cibois, Schikler, Feinstein, \& Cracraft, 2004; Beresford, Barker, Ryan, \& Crowe, 2005; Bryson Jr et al., 2014; Burleigh, Kimball, \& Braun, 2015; Burns et al., 2014; Chaves, Hidalgo, \& Klicka, 2013; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hackett et al., 2008; Jetz et al., 2012; Johansson, Fjeldså, \& Bowie, 2008; Kimball et al., 2019; Klicka et al., 2014; Lamichhaney et al., 2015; Lerner, Meyer, James, Hofreiter, \& Fleischer, 2011; Lovette et al., 2010; Moyle et al., 2016; Ödeen, Håstad, \& Alström, 2011; Oliveros et al., 2019; Päckert et al., 2012; Parchman, Benkman, \& Mezquida, 2007; Powell et al., 2014; Price et al., 2014; Pulgarín-R, Smith, Bryson Jr, Spellman, \& Klicka, 2013; Selvatti, Gonzaga, \& Moraes Russo, 2015; Tietze, Päckert, Martens, Lehmann, \& Sun, 2013; Treplin et al., 2008; Zuccon, Prŷs-Jones, Rasmussen, \& Ericson, 2012).
-
+By default, \texttt{datelife} uses the Open Tree of Life synthetic phylogeny as summarizing topology.
+According to this phylogeny, species belonging to the family Fringillidae do not form a monophyletic group.
 Hence, DateLife's taxon-constrained approach returns a topology and ages for the species in a named group.
@@
 
-At the time of writing of this manuscript (May 22, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
+At the time of writing of this manuscript (May 26, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
 A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time.
@@
 New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
-\emph{{The Auk: Ornithological Advances}}, \emph{132}(2), 333--348.
+\emph{The Auk: Ornithological Advances}, \emph{132}(2), 333--348.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-zuccon2012phylogenetic}{}}%
-Zuccon, D., Prŷs-Jones, R., Rasmussen, P.
-C., \& Ericson, P.
-G.
-(2012).
-The phylogenetic relationships and generic limits of finches (fringillidae).
-\emph{Molecular Phylogenetics and Evolution}, \emph{62}(2), 581--596.
-
 \end{CSLReferences}
@@
 \begin{figure}[!h]
+\includegraphics{../figures/fringillidae-topologies/fringillidae-topology.pdf}
+\caption{Fringillidae topologies available from OpenTree's synthetic phylogeny.
+A) Topology encompassing bird species within the Fringillidae within the NCBI taxonomy (black) as well as all species that share the same mrca node in the synthetic tree (purple).
+B) Paraphyletic topology encompassing bird species within the Fringillidae only.
+It results from pruning purple branches from topology in A.
+Bird species belonging to the family Fringillidae do not form a monophyletic group.
+(Alström et al. \protect\hyperlink{ref-}{2014},
+Barker, Cibois, Schikler, Feinstein, \& Cracraft \protect\hyperlink{ref-barker2004phylogeny}{2004},
+Barker et al. \protect\hyperlink{ref-barker2013going}{2013},
+Barker \protect\hyperlink{ref-barker2014mitogenomic}{2014},
+Barker et al. \protect\hyperlink{ref-barker2015new}{2015},
+Beresford, Barker, Ryan, \& Crowe \protect\hyperlink{ref-beresford2005african}{2005},
+Bryson Jr et al. \protect\hyperlink{ref-bryson2014diversification}{2014},
+Burleigh, Kimball, \& Braun \protect\hyperlink{ref-burleigh2015building}{2015},
+Burns et al. \protect\hyperlink{ref-burns2014phylogenetics}{2014},
+Chaves, Hidalgo, \& Klicka \protect\hyperlink{ref-chaves2013biogeography}{2013},
+Claramunt \& Cracraft \protect\hyperlink{ref-claramunt2015new}{2015},
+Gibb et al. \protect\hyperlink{ref-gibb2015new}{2015},
+Hackett et al. \protect\hyperlink{ref-hackett2008phylogenomic}{2008},
+Jetz et al. \protect\hyperlink{ref-Jetz2012}{2012},
+Johansson, Fjeldså, \& Bowi \protect\hyperlink{ref-johansson2008phylogenetic}{200},
+Kimball et al. \protect\hyperlink{ref-kimball2019phylogenomic}{2019},
+Klicka et al. \protect\hyperlink{ref-klicka2014comprehensive}{2014},
+Lamichhaney et al. \protect\hyperlink{ref-lamichhaney2015evolution}{2015},
+Lerner, Meyer, James, Hofreiter, \& Fleischer \protect\hyperlink{ref-lerner2011multilocus}{2011},
+Lovette et al. \protect\hyperlink{ref-lovette2010comprehensive}{2010},
+Moyle et al. \protect\hyperlink{ref-moyle2016tectonic}{2016},
+Ödeen, Håstad, \& Alström \protect\hyperlink{ref-odeen2011evolution}{2011},
+Oliveros et al. \protect\hyperlink{ref-oliveros2019earth}{2019},
+Päckert et al. \protect\hyperlink{ref-packert2012horizontal}{2012},
+Parchman, Benkman, \& Mezquida \protect\hyperlink{ref-parchman2007coevolution}{2007},
+Powell et al. \protect\hyperlink{ref-powell2014comprehensive}{2014},
+Price et al. \protect\hyperlink{ref-price2014niche}{2014},
+Pulgarín-R, Smith, Bryson Jr, Spellman, \& Klicka \protect\hyperlink{ref-pulgarin2013multilocus}{2013},
+Selvatti, Gonzaga, \& Moraes Russo \protect\hyperlink{ref-selvatti2015paleogene}{2015},
+Tietze, Päckert, Martens, Lehmann, \& Sun \protect\hyperlink{ref-tietze2013complete}{2013},
+Treplin et al. \protect\hyperlink{ref-treplin2008molecular}{2008},
+Zuccon, Prŷs-Jones, Rasmussen, \& Ericson \protect\hyperlink{ref-zuccon2012phylogenetic}{2012})
+}
+\label{fig:fringillidae-topologies}
+\end{figure}
+
+\begin{figure}[!h]
 \includegraphics{../figures/figure-fringillidae/median_and_calibration_ages_simple.png}

### 2022-05-26 manuscript/figures-alt.tex
addressing topologies comment

@@
 \includegraphics{../figures/fringillidae-topologies/fringillidae-topology.pdf}
-\caption{Topologies obtained from OpenTree's synthetic phylogeny.
-A) Topology od 2333 tips, encompassing bird species within the Fringillidae following the NCBI taxonomy (black) as well as all other bird species that share the same mrca node in the synthetic tree (purple).
-B) Topology encompassing bird species within the Fringillidae only, resulting from pruning purple branches from topology A.
-Bird species belonging to the family Fringillidae do not form a monophyletic group.
+\caption{Tree topologies obtained from Open Tree of Life's (OpenTree) synthetic phylogenetic tree.
+A) Topology of 2333 tips and 1305 internal nodes, encompassing bird species within the family Fringillidae following the NCBI taxonomy (black), as well as all other bird species that share the same mrca node in OpenTree's synthetic tree (purple).
+B) Topology of 289 tips and 253 internal nodes, encompassing bird species within the Fringillidae only, resulting from pruning purple branches from topology A.
+Bird species within the Fringillidae do not form a monophyletic group.
 (Alström et al. \protect\hyperlink{ref-}{2014},

### 2022-05-26 manuscript/manuscript.Rmd
addressing topologies comment

@@
 By default, `datelife` uses the Open Tree of Life synthetic phylogeny as summarizing topology.
-According to this phylogeny, species belonging to the family Fringillidae do not form a monophyletic group.
-Hence, DateLife's taxon-constrained approach returns a topology and ages for the species in a named group.
+According to this phylogeny, species belonging to the family Fringillidae do not form a monophyletic group (Fig.
+\ref{fig:fringillidae-topologies}).
+By default, DateLife's taxon-constrained approach returns a topology and ages for the species in a named group, which do not necessarily correspond to a monophyletic group.
+Users can change this behaviour by providing a monophyletic tree as input for a DateLife search, or as a tree topology for a DateLife summary.
 
-<!--BCO: it's ambiguous: it gets a topology and age for the larger clade that includes all Fringillidae species? Or for the paraphyletic Fringillidae group?-->
+<!--BCO: it's ambiguous: it gets a topology and age for the larger clade that includes all Fringillidae species? Or for the paraphyletic Fringillidae group?
+Luna: Clarified in text, I hope!-->
 

### 2022-05-26 manuscript/manuscript.tex
addressing topologies comment

@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4586}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4625}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 By default, \texttt{datelife} uses the Open Tree of Life synthetic phylogeny as summarizing topology.
-According to this phylogeny, species belonging to the family Fringillidae do not form a monophyletic group.
-Hence, DateLife's taxon-constrained approach returns a topology and ages for the species in a named group.
+According to this phylogeny, species belonging to the family Fringillidae do not form a monophyletic group (Fig.
+\ref{fig:fringillidae-topologies}).
+By default, DateLife's taxon-constrained approach returns a topology and ages for the species in a named group, which do not necessarily correspond to a monophyletic group.
+Users can change this behaviour by providing a monophyletic tree as input for a DateLife search, or as a tree topology for a DateLife summary.
 
@@
 \includegraphics{../figures/fringillidae-topologies/fringillidae-topology.pdf}
-\caption{Fringillidae topologies available from OpenTree's synthetic phylogeny.
-A) Topology encompassing bird species within the Fringillidae within the NCBI taxonomy (black) as well as all species that share the same mrca node in the synthetic tree (purple).
-B) Paraphyletic topology encompassing bird species within the Fringillidae only.
-It results from pruning purple branches from topology in A.
-Bird species belonging to the family Fringillidae do not form a monophyletic group.
+\caption{Tree topologies obtained from Open Tree of Life's (OpenTree) synthetic phylogenetic tree.
+A) Topology of 2333 tips and 1305 internal nodes, encompassing bird species within the family Fringillidae following the NCBI taxonomy (black), as well as all other bird species that share the same mrca node in OpenTree's synthetic tree (purple).
+B) Topology of 289 tips and 253 internal nodes, encompassing bird species within the Fringillidae only, resulting from pruning purple branches from topology A.
+Bird species within the Fringillidae do not form a monophyletic group.
 (Alström et al. \protect\hyperlink{ref-}{2014},

### 2022-05-27 manuscript/manuscript.Rmd
addressing all other comments

@@
 abstract: |
-  Date estimates for times of evolutionary divergences are key data for research in the natural sciences.
-These estimates also provide valuable information for for education, science communication, and policy decisions.
-  Although achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram) is a difficult and time-consuming task, the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
-  When these estimates are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on time of evolutionary origin is exposed in a programatic and reusable way.
-  Intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
-  Here we present DateLife, a service implemented as an R package and an R Shiny website application available at www.datelife.org, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
+  Achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram) is a difficult and time-consuming task.
+But the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
+    Knowledge of the evolutionary time frame of organisms is key for research in the natural sciences.
+It also represent valuable information for education, science communication, and policy decisions.
+  When chronograms are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on evolutionary timeframe is exposed in a programatic and reusable way, as intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
+  Here we present DateLife, a service implemented as an R package and an R Shiny website application available at www.datelife.org, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time frame of evolution.
   The main DateLife workflow constructs a chronogram for any given combination of taxon names by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from the TreeBASE database as well.
   We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
-  DateLife will be useful to increase awereness of the existing variation in expert time of divergence data, and can foster exploration of the effect of alternative divergence time hypotheses on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.
-
-  <!--  R has become mainstream in the biological-data analyst toolkit.
-Hence, to improve accessibility of time of evolutionary origin data we developed an R package that facilitates interaction, reanalysis and reuse of it, as well as incorporation of these data into the evolutionary analysis workflow.
--->
+  DateLife will be useful to increase awereness of the existing variation in alternative hypothesis of evolutionary time for the same organisms, and can foster exploration of the effect of alternative evolutionary timing hypotheses on the results of downstream analyses, providing a framework for a more informed interpretation of evolutionary results.
 
@@
 <!-- If you're aiming for syst bio you don't want to start so general Readers know what chronograms and phylogenies are!.-->
-Chronograms --phylogenies with branch lengths proportional to time-- provide key data for the study of natural processes in many areas of biological research, such as developmental biology [@laubichler2009form; @delsuc2018phylogenomic], conservation biology [@Felsenstein1985a; @Webb2000], historical biogeography [@posadas2006historical], and species diversification [@magallon2001absolute; @Morlon2014].
+Chronograms --phylogenies with branch lengths proportional to time-- provide key data on evolutionary time frame for the study of natural processes in many areas of biological research, such as developmental biology [@laubichler2009form; @delsuc2018phylogenomic], conservation biology [@Felsenstein1985a; @Webb2000], historical biogeography [@posadas2006historical], and species diversification [@magallon2001absolute; @Morlon2014].
 
@@
 
-Here we present the DateLife project which has the main goal of capturing data from published chronograms, and making these data readily accessible to users for reuse and reanalysis.
+Here we present the DateLife project which has the main goal of capturing age data from published chronograms, and making these data readily accessible to the community for reuse and reanalysis.
+  <!--  R has become mainstream in the biological-data analyst toolkit.
+Hence, to improve accessibility of time of evolutionary origin data we developed an R package that facilitates interaction, reanalysis and reuse of it, as well as incorporation of these data into the evolutionary analysis workflow.
+-->
 DateLife's core software application is available as an R package [@datelifeR], and as an online Rshiny interactive website at www.datelife.org.
@@
 If the input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004].
-The input tree is not required to have branch lengths, and its topology is used in the summary steps described below.
+The input tree is not required to have branch lengths, and its topology is used in the summary steps described in the next section.
 
@@
 Subspecies and variants are ignored.
-If an input taxon name belongs to an inclusive taxonomic group the algorithm has two alternative behaviors defined by the "get species from taxon" flag.
-If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group (according to a taxonomy) and adds them to the input string.
-If the flag is inactive, DateLife excludes the taxon names above the species level from the input.
+If an input taxon name belongs to an inclusive taxonomic group, the algorithm has two alternative behaviors defined by the "get species from taxon" flag.
+If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group (according to a taxonomy of choice) and adds them to the input string.
+If the flag is inactive, DateLife excludes the taxon names above the species level from the query.
 
@@
 
-The processed input taxon names are saved as an R object of a newly defined class \texttt{datelifeQuery} that is used in the following steps.
+The processed input taxon names are saved as an R object of a newly defined class, \texttt{datelifeQuery}, that is used in the following steps.
 This object contains the standardized names, the corresponding OpenTree taxonomic id numbers, and the topology of the input tree if any was provided.
@@
 
-DateLife's chronogram database latest version consist of `r length(opentree_chronograms$trees)` chronograms published in `r length(unique(opentree_chronograms$studies))` different studies.
+At the time of writing of this manuscript (`r format(Sys.time(), "%b %d, %Y")`), DateLife's chronogram database latest version consist of `r length(opentree_chronograms$trees)` chronograms published in `r length(unique(opentree_chronograms$studies))` different studies.
 It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata [@mctavish2015phylesystem], which allows automatic and reproducible assembly of our chronogram database.
 Datelife's chronogram database is navigable as an R data object within the `datelife` R package.
-Published chronograms can be added to Phylesystem by any user, at any time, and are immediately publicly available (https://tree.opentreeoflife.org/curator).
-A user may then update their local cache of DateLife's chronogram database to include new chronogram data on a following search.
+
+A unique feature of OpenTree's Phylesystem is that the community can add new published, state-of-the-art chronograms any time through their curator application (https://tree.opentreeoflife.org/curator).
+As chronograms are added to Phylesystem, they are incorporated into an updated `datelife`'s database that is assigned a new version number, followed by a package release on CRAN.
+`datelife`'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
+<!--BCO: Nicely done-->
+Users can also upload and curate new chronograms to OpenTree themselves.
+Then, trigger an update of their local `datelife` database to incorporate the new chronograms and have them available for their current analysis.
+<!--BCO: What is immediately?
+Also, it's not just uploading -- they have to include metadata on what brlen mean, maybe?
+indicate the ingroup/outgroup, do TNRS within OToL, right?
+Maybe this has improved.
+I remember being frustrated trying to load in giant trees for datelife use b/c I had to click on the ingroup node.
+Luna: Very true, it is not that easy, I added the curation step to hint that there are more steps that simply uploading the chronogram.-->
+<!--BCO; this duplicates an earlier section and feels more methods-y. Maybe move up and merge? 
+Luna: Done! Moved here.-->
 
@@
 5. The source chronogram(s) with the most input taxon names.
-6. Various single summary chronograms resulting from summarizing age data, generated using the methodology described below.
+6. Various single summary chronograms resulting from summarizing age data, generated using the methodology described next.
 
@@
 This assumes that the terminal taxa are coeval and occur at the present.
-A table of congruified node ages that can be used as secondary calibrations is stored as a `congruifiedCalibrations` object.
+A table of congruified node ages that can be used as calibrations for a dating analysis is stored as a `congruifiedCalibrations` object.
 
 For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method [SDM; @Criscuolo2006], which deforms patristic distance matrices by minimizing variance and then averaging them.
-These single summary taxon pair age matrices (Summarized calibrations) can be applied as secondary calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes [@Huelsenbeck2001; @Ronquist2003], PATHd8 [@Britton2007], BLADJ [@Webb2008; @webb2005phylomatic], and treePL [@Smith2012].
+These single summary taxon pair age matrices (Summarized calibrations) can be applied as calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes [@Huelsenbeck2001; @Ronquist2003], PATHd8 [@Britton2007], BLADJ [@Webb2008; @webb2005phylomatic], and treePL [@Smith2012].
 
@@
 This minimizes age variance in the resulting chronogram [@Webb2008].
-BLADJ does not incorporate input phylogeny branch lengths even when they are present.
+BLADJ does not use branch lengths even when they are present in the input tree or summarizing topology.
 When there is conflict in ages between nodes with calibration data, the algorithm ignores node ages that are older than the age of a parent node.
@@
 
-Alternative options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the secondary calibrations.
+Alternative options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the calibrations.
 PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees.
@@
 [@sanderson2002].
-The MrBayes  [@Huelsenbeck2001; @Ronquist2003] approach in datelife uses the secondary calibrations as priors on node ages.
+The MrBayes  [@Huelsenbeck2001; @Ronquist2003] approach in datelife uses the calibrations as priors on node ages.
 
@@
 Discovering this synonym allowed assigning five age data points for the parent node of *Emberiza elegans*, shown as *Schoeniclus elegans* in figure \ref{fig:figure2-1}, which would not have had any data otherwise.
-<!--\textcolor{red}{I don't understand what this sentence means}
-it is related to the synonym found by tnrs processing, if we had used the synonym insted of the one accepted in the database, we would not have discovered that information
-It moved it here, it makes more sense here.
+<!--\textcolor{red}{EJM: I don't understand what this sentence means}
+LLSR: it is related to the synonym found by tnrs processing, if we had used the synonym insted of the one accepted in the database, we would not have discovered that information
+LLSR: I moved it here, it makes more sense here, I think!.
 -->
@@
 The name resolution step allowed discovering five data points for node "n4" that would not have had any data otherwise due to name mismatch.
-Age summary statistics per node were calculated and used as secondary calibrations to date the tree topology using the BLADJ algorithm.
+Age summary statistics per node were calculated and used as calibrations to date the tree topology using the BLADJ algorithm.
 Age data for node "n2" was excluded as final calibration because it is older than age data of the more inclusive node "n1".
@@
 
-`datelife` makes state-of-the-art information on time of lineage divergence easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
-It is an open service that does not require any expert biological knowledge from users, besides the names of the organisms or clade they want to work with, for any of its functionality.
-
-At the time of writing of this manuscript (`r format(Sys.time(), "%b %d, %Y")`), `datelife`'s database has `r length(opentree_chronograms$trees)` chronograms, pulled entirely from OpenTree's database, the Phylesystem [@mctavish2015phylesystem].
-A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time.
-As chronograms are added to Phylesystem, they are incorporated into an updated `datelife`'s database that is assigned a new version number, followed by a package release on CRAN.
-`datelife`'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
-<!--BCO: Nicely done-->
-Users can also upload new chronograms to OpenTree themselves, and trigger an update of their local `datelife` database to incorporate the new chronograms, to have them immediately available for analysis.
-<!--BCO: What is immediately?
-Also, it's not just uploading -- they have to include metadata on what brlen mean, maybe?
-indicate the ingroup/outgroup, do TNRS within OToL, right?
-Maybe this has improved.
-I remember being frustrated trying to load in giant trees for datelife use b/c I had to click on the ingroup node-->
-<!--BCO; this duplicates an earlier section and feels more methods-y. Maybe move up and merge?-->
+`datelife` makes state-of-the-art data on evolutionary time frame easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
+It is an open service that does not require any expert biological knowledge from users --besides the names of the species or group they want to work with, for any of its functionality.
 
@@
 Incorporation of more chronograms into `datelife`'s database will continue to improve its services.
-One option to increase the of chronograms in the DateLife database is the Dryad data repository, which now holds some chronograms.
+One option to increase the number of chronograms in the DateLife database is the Dryad data repository.
 Methods to automatically mine chronograms from Dryad could be designed and implemented.
-However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined manually by a curator.
+However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined manually by a human curator.
+We would like to emphasize on the importance of sharing chronogram data,  including systematically curated metadata, into open repositories, such as OpenTree's Phylesystem [@mctavish2015phylesystem] for the benefit of the scientific community as a whole.
 
-The largest, and taxonomically broadest, summary chronogram currently available from OpenTree
-was constructed using age data from 2,274 published chronograms [@Hedges2015].
+## Primary vs Secondary calibrations in dating analyses
+
+While most chronograms in DateLife's database are constructed using primary calibrations (molecular substitution rates or ages obtained from the fossil record or geological events), DateLife summarizes chronograms using secondary calibrations only (ages coming from other chronograms).
+
+@graur2004reading cautioned on the increased error and uncertainty in estimated ages when using secondary calibrations in dating analyses.
+@schenk2016sec showed that, in simulations, divergence times inferred using secondary calibrations are significantly younger than those inferred with primary calibrations, when obtained with Bayesian inference methods, and when priors are implemented in similar ways in both analyses.
+Accordingly, the scientific community seems to have more confidence in chronograms obtained from a single analysis, using fossil data as primary sources of calibrations [@schenk2016sec], and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history [@sauquet2013practical], as for example done by [@antonelli2017supersmart].
+<!-- Will this be implemented in future `datelife` versions?
+Probably not.
+-->
+There have been attempts to create fossil calibration databases [@ksepka2015fossil], though these still have room to grow.
+<!--BCO: Polite way to say 'no new additions in four years?' LLSR: lol, yes-->
+
+\textcolor{red}{LLSR: What does our cross validation analysis show in relation to diffs between primary and secondary calibrations?}
+
+It seems that using several (as opposed to just a few) secondary calibrations can provide sufficient information to alleviate or even neutralize potential biases [@sauquet2013practical]. 
+Certainly, further studies are required to fully understand the effect of secondary calibrations on outputs from different tree dating methods, and on downstream analyses.
+It is possible that secondary calibrations can be safely used with dating methods that do not require setting priors, such as penalized likelihood [@sanderson2003r8s], methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ [@Webb2008; @webb2005phylomatic], or methods that summarize age data unto a tree topology.
+
+## Variation in source chronograms
+
+Conflict among alternative dating studies is common in the literature
+(see, for example, the robust ongoing debate about crown group age of angiosperms [@sauquet2021age; @barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time]).
+
+Source chronograms may have been estimated using different types of primary calibrations (fossil record, geologic or biogeographic events, molecular substitution rates).
+These differences in approach can deepen the already substantial variation in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
+For example, the chronograms from @burns2014phylogenetics tend to have much older date estimates for the same nodes than other inferences (Figure 5).
+This study inferred dates using molecular substitution rate estimates across birds [@weir2008calibrating]
+\textcolor{red}{EJM: MORE ON the FINCH examples and the cross validation here}
+
+Different chronograms might also imply fundamentally distinct evolutionary hypotheses.
+For example, two different researchers working on the same clade could both carefully select and argue for their choices of fossil calibrations.
+Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses the same one to calibrate outside the clade, the resulting age estimates will often differ substantially, as the placement of calibrations as stem or crown group is proved to deeply affect estimated times of lineage divergence [@sauquet2013practical].
+
+## Sumarizing chronograms
+
+By default, DateLife currently summarizes all source chronograms that overlap with at least two species names. Users can exclude source chronograms if they have reasons to do so.
+Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely.
+To our knowledge, there are no tested measures to determine independently when a chronogram is better than another.
+Yet, several characteristics of the data used for dating analyses, as well as from the output chronogram itself, could be used to score the quality of source chronograms.
+
+Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used [@magallon2010using, @magallon2015metacalibrated].
+Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
+
+DateLife provides an opportunity to capture concordance and conflict among date estimates, which can also be used as a metric for chronogram reliability.
+Its open database of chronograms allows other researchers to do such analyses themselves reproducibly, and without needing permission.
+Though, of course, they should follow proper citation practices, especially for the source chronogram studies.
+
+The exercise of summarizing age data from across multiple studies provides the opportunity to work with a more inclusive chronogram, 
+that reflects a unified evolutionary history for a lineage, by putting together evidence from different hypotheses.
+The largest, and taxonomically broadest chronogram currently available from OpenTree was constructed summarizing age data from 2,274 published chronograms using NCBI's taxonomic tree as backbone [@Hedges2015].
 <!--BCO: I just checked: TIMETREEE NOW ALLOWS SOURCE TREE DOWNLOADS!
@@
 They have a copyright (not legal in US on trees, IMHO) notice on individual trees, but they do say "To download all individual study timetrees, please send an email request to info@timetree.org." -- I think this could be great to do-->
+A summarizing exercise may also amplify the effect of uncertainty and errors in source data, and blur parts of the evolutionary history of a lineage that might only be reflected in source chronograms and lost on the summary chronogram [@sauquet2021age].
+\textcolor{red}{EJM: ??? it's not clear to me what this last part means! What might you be losing?
+LLSR: I changed loosing to blurring. But I will elaborate more on it}
+Choosing source chronograms that we are going to keep and the ones that we are going to discard might be key.
+\textcolor{red}{LLSR: Example}
 
-We would like to emphasize on the importance of sharing chronogram data,  including systematically curated metadata, into open repositories, such as OpenTree's Phylesystem [@mctavish2015phylesystem] for the benefit of the scientific community as a whole.
-
-By default, `datelife` currently summarizes all source chronograms that overlap with at least two species names. Users can exclude source chronograms if they have reasons to do so.
-Strictly speaking, the best chronogram should reflect the real time of lineage divergence accurately and precisely.
-To our knowledge, there are no good measures to determine independently if a chronogram is better than another.
-Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used [@magallon2010using, @magallon2015metacalibrated].
-Several characteristics of the data used for dating analyses, as well as from the output chronogram itself, could be used to score quality of source chronograms.
-Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
-DateLife provides an opportunity to capture concordance and conflict among date estimates, which can also be used as a metric for chronogram reliability.
-Its open database of trees allows other researchers to do such analyses themselves reproducibly and without needing permission, though of course they should follow proper citation practices, especially for the source trees.
-
-Scientists have more confidence in chronograms constructed using primary calibrations (ages obtained from the fossil or geological record) to ones constructed with secondary calibrations (ages coming from other chronograms)[@schenk2016sec].
-@schenk2016sec showed that in simulations divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with Bayesian inference methods when priors are implemented in similar ways in both analyses.
-However, secondary calibrations are useful in dating methods that do not require setting priors, such as penalized likelihood [@sanderson2003r8s], or as fixed ages for nodes.
-Certainly, further studies are required to fully understand the effect of using secondary calibrations on time estimates and downstream analyses.
-
-The input chronograms may have been estimated using primary fossil data or using estimates of molecular substitution rates.
-These differences in approach which can deepen the already substantial variation in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
-For example, the chronograms from @burns2014phylogenetics tend to have much older date estimates for the same nodes than other inferences (Figure 5).
-This study did not use any primary calibrations, and inferred dates using an substitution rate estimates across birds [@weir2008calibrating]
-<!--\textcolor{red}{MORE ON the FINCH examples and the cross validation here}
--->
-
-Conflict among alternative date estimates is common in the literature
-(see, for example, the robust ongoing debate about crown group age of angiosperms [@sauquet2021age; @barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time].
-Using a different chronogram may return different results [@title2016macrophylogenies] particularly for studies of species diversification, or focussed on the timing of evolutionary events.
-Stitching together multiple chronograms creates a larger tree that uses information from across studies, but may amplify uncertainty and errors.
-
-Summarizing chronograms might also imply summarizing fundamentally distinct evolutionary hypotheses.
-For example, two different researchers working on the same clade could both carefully select and argue for their choices of fossil calibrations.
-Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses the same one to calibrate outside the clade, the resulting age estimates will often differ substantially, as the placement of calibrations as stem or crown group is proved to deeply affect estimated times of lineage divergence [@sauquet2013practical].
-Trying to summarize the resulting chronograms into a single one using simple summary statistics can erase many types of relevant information from the source chronograms.
-Accordingly, the prevailing view is that we should favor lineage divergence estimates obtained from a single analysis, using fossil data as primary sources of calibrations, and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history, as for example done by [@antonelli2017supersmart].
-<!-- Will this be implemented in future `datelife` versions?
-Probably not.
--->
-There have been attempts to create fossil calibration databases <!--REF TO ADD: Ksepka, D.T., J.F.
-Parham, J.F.
-Allman, M.J.
-Benton, M.T.
-Carrano, K.A.
-Cranston, P.C.
-J.
-Donoghue, J.J.
-Head, E.J.
-Hermsen, R.B.
-Irmis, W.G.
-Joyce, M.
-Kohli, K.S.
-Lamm, D.
-Leehr, J.S.L.
-Patané, P.D.
-Polly, M.J.
-Phillips, N.A.
-Smith, N.D.
-Smith, M.
-van Tuinen, J.L.
-Ware, R.C.M.
-Warnock.
-2015.The Fossil Calibration Database, A New Resource for Divergence Dating.
-Systematic Biology: doi:10.1093/sysbio/syv025.-->, though these still have room to grow.
-<!--BCO: Polite way to say 'no new additions in four years?'-->
-However, the exercise of summarizing different chronograms has the potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypotheses.
-Choosing the elements of the chronograms that we are going to keep and the ones that we are going to discard is key, since we are potentially losing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram [@sauquet2021age].
-<!-- Example
-\textcolor{red}{EJM: ??? it's not clear to me what this last part means! What might you be losing?}
--->
-
-<!-- still requires more research. -->
-<!-- Often, there is a large discrepancy in time of lineage divergence estimates across studies from expert knowledge, that is well known and documented for different groups of organisms [e.g., angiosperms; @magallon2015metacalibrated].
--->
-<!-- Comparison of available age data for a wide range of organisms shown here suggest that this is a widespread phenomenon that requires further attention. -->
-<!-- In the case of the example presented here, most chronograms are new, meaning that we expect that data was generated with high quality and the software used should be very advanced.
-However, this is just an assumption.
--->
+<!-- Comparison of available age data for a wide range of organisms shown here suggest that discrpenacy in age stimates across studies is a widespread phenomenon that requires further attention. -->
 
@@
 
-Nonetheless, in ecology and conservation biology, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance [@Webb2008].
-Hence, we integrated into datelife's workflow different ways of estimating node ages in the absence of calibrations and branch length information for taxa lacking this information.
-"Making up" branch lengths is an common practice in scientific publications: @Jetz2012, created a time-calibrated tree of all 9,993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a tree of 353,185 seed plants where only 23% had molecular data.
-Obviously, there are risks in this practice!
-Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch estimation methods.
-There has yet to be a thorough analysis of what can go wrong when one extends inferences beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above, by carefully considering the statistical assumptions being made, and assessing the consistency of the results with prior work.
+## Effects on downstream analyses
 
+For downstream analyses, using alternative chronogram may deeply affect our inferences [@title2016macrophylogenies], particularly when studying phenomena dependent on the timing of species diversification events, such as macroevolutionary processes.
 <!-- There is evidence that different chronograms will result in very different macroevolutionary histories inferred from them [@title2016macrophylogenies].
@@
 
+In ecology and conservation biology, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance [@Webb2008].
+Hence, DateLife's workflow features different ways of estimating node ages in the absence of calibrations and branch length information for certain taxa.
+"Making up" branch lengths is a common practice in scientific publications: @Jetz2012, created a time-calibrated tree of all 9,993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a tree of 353,185 seed plants where only 23% had molecular data.
+
+Notably, risks come with this practice.
+Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration, using polytomy resolution and branch estimation methods.
+There has yet to be a thorough analysis of what can go wrong when one extends inferences beyond the data in this way, so we urge caution; we also urge readers to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, and assessing the consistency of the results with prior work.
+
 # Conclusions
 
-Divergence time information is key to many areas of evolutionary studies: trait evolution,
-diversification, biogeography, macroecology and more. It is also crucial for science communication and education, but generating chronograms is difficult,
+Knowledge of the evolutionary time frame of organisms is key to many research areas: trait evolution,
+species diversification, biogeography, macroecology and more. It is also crucial for education, science communication and policy, but generating chronograms is difficult,
 especially for those who want to use phylogenies but who are not systematists, or
-do not have the time to acquire and develop the necessary knowledge and data curation skills.
-Moreover, years of primarily public funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
+do not have the time to acquire and develop the necessary knowledge and skills to construct them on their own.
+Importantly, years of primarily public funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
 
 <!--Mention again uses shown here and outline potential uses not shown here-->
-The `datelife` R package allows easy and fast summarization of publicly available information
-on time of lineage divergence.
-This provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identification of regions that require more research or that have conflicting information.
-It is available as an R package, or a web-based R shiny app at dates.opentreeloflife.org/datelife.
+The DateLife project allows for easy and fast summary of public and state-of-the-art data on time of lineage divergence. 
+It provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identification of regions that require more research, or that have conflicting information.
+It is available as an R package, and as a web-based R shiny application at www.datelife.org
 <!-- It can also summarize public data, or use it to generate new chronograms based on trees provided by the user. -->
@@
 
-`datelife` is free and open source and it can be used through its current website
+The `datelife` software is free and open source and it can be used through its R shiny web application at
 <http://www.datelife.org>, through the `datelife` R package, and through Phylotastic's project web portal <https://phylo.cs.nmsu.edu/>.
-`datelife`'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
+DateLife's web application is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker and Open Tree of Life's infrastructure (dates.opentreeoflife.org/datelife).
 `datelife`'s R package stable version `r availability`
@@
 that led to initial work on this project.
-The team that assembled `datelife`'s first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
+The team that assembled DateLife's first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
 The Open Tree of Life project that provides the open, metadata rich repository of
-trees used for `datelife`.
-The many scientists who publish their chronograms in an open, reusable form, and
-the scientists who curate them for deposition in the Open Tree of Life repository.
+trees used to construct DateLife's chronogram database.
+The many scientists who publish their chronograms in an open, reusable form, and the scientists who curate them for deposition in the Open Tree of Life repository.
 The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.

### 2022-05-27 manuscript/manuscript.tex
addressing all other comments

@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4625}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4765}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 \abstract{%
-Date estimates for times of evolutionary divergences are key data for research in the natural sciences.
-These estimates also provide valuable information for for education, science communication, and policy decisions.
-Although achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram) is a difficult and time-consuming task, the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
-When these estimates are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on time of evolutionary origin is exposed in a programatic and reusable way.
-Intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
-Here we present DateLife, a service implemented as an R package and an R Shiny website application available at www.datelife.org, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time of evolutionary origin.
+Achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram) is a difficult and time-consuming task.
+But the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
+Knowledge of the evolutionary time frame of organisms is key for research in the natural sciences. It also represent valuable information for education, science communication, and policy decisions.
+When chronograms are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on evolutionary timeframe is exposed in a programatic and reusable way, as intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
+Here we present DateLife, a service implemented as an R package and an R Shiny website application available at www.datelife.org, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time frame of evolution.
 The main DateLife workflow constructs a chronogram for any given combination of taxon names by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from the TreeBASE database as well.
 We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
-DateLife will be useful to increase awereness of the existing variation in expert time of divergence data, and can foster exploration of the effect of alternative divergence time hypotheses on the results of analyses, providing a framework for a more informed interpretation of evolutionary results.
+DateLife will be useful to increase awereness of the existing variation in alternative hypothesis of evolutionary time for the same organisms, and can foster exploration of the effect of alternative evolutionary timing hypotheses on the results of downstream analyses, providing a framework for a more informed interpretation of evolutionary results.
 }
@@
 
-Chronograms --phylogenies with branch lengths proportional to time-- provide key data for the study of natural processes in many areas of biological research, such as developmental biology (Delsuc et al., 2018; Laubichler \& Maienschein, 2009), conservation biology (Felsenstein, 1985; C.
+Chronograms --phylogenies with branch lengths proportional to time-- provide key data on evolutionary time frame for the study of natural processes in many areas of biological research, such as developmental biology (Delsuc et al., 2018; Laubichler \& Maienschein, 2009), conservation biology (Felsenstein, 1985; C.
 Webb, 2000), historical biogeography (Posadas, Crisci, \& Katinas, 2006), and species diversification (Magallon \& Sanderson, 2001; Morlon, 2014).
@@
 
-Here we present the DateLife project which has the main goal of capturing data from published chronograms, and making these data readily accessible to users for reuse and reanalysis.
+Here we present the DateLife project which has the main goal of capturing age data from published chronograms, and making these data readily accessible to the community for reuse and reanalysis.
 DateLife's core software application is available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org.
@@
 If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis et al., 2004).
-The input tree is not required to have branch lengths, and its topology is used in the summary steps described below.
+The input tree is not required to have branch lengths, and its topology is used in the summary steps described in the next section.
 
@@
 Subspecies and variants are ignored.
-If an input taxon name belongs to an inclusive taxonomic group the algorithm has two alternative behaviors defined by the ``get species from taxon'' flag.
-If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group (according to a taxonomy) and adds them to the input string.
-If the flag is inactive, DateLife excludes the taxon names above the species level from the input.
+If an input taxon name belongs to an inclusive taxonomic group, the algorithm has two alternative behaviors defined by the ``get species from taxon'' flag.
+If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group (according to a taxonomy of choice) and adds them to the input string.
+If the flag is inactive, DateLife excludes the taxon names above the species level from the query.
 
@@
 
-DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies.
+At the time of writing of this manuscript (May 27, 2022), DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies.
 It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata (McTavish et al., 2015), which allows automatic and reproducible assembly of our chronogram database.
 Datelife's chronogram database is navigable as an R data object within the \texttt{datelife} R package.
-Published chronograms can be added to Phylesystem by any user, at any time, and are immediately publicly available (\url{https://tree.opentreeoflife.org/curator}).
-A user may then update their local cache of DateLife's chronogram database to include new chronogram data on a following search.
+
+A unique feature of OpenTree's Phylesystem is that the community can add new published, state-of-the-art chronograms any time through their curator application (\url{https://tree.opentreeoflife.org/curator}).
+As chronograms are added to Phylesystem, they are incorporated into an updated \texttt{datelife}'s database that is assigned a new version number, followed by a package release on CRAN.
+\texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
+Users can also upload and curate new chronograms to OpenTree themselves.
+Then, trigger an update of their local \texttt{datelife} database to incorporate the new chronograms and have them available for their current analysis.
 
@@
 \item
-  Various single summary chronograms resulting from summarizing age data, generated using the methodology described below.
+  Various single summary chronograms resulting from summarizing age data, generated using the methodology described next.
 \end{enumerate}
@@
 This assumes that the terminal taxa are coeval and occur at the present.
-A table of congruified node ages that can be used as secondary calibrations is stored as a \texttt{congruifiedCalibrations} object.
+A table of congruified node ages that can be used as calibrations for a dating analysis is stored as a \texttt{congruifiedCalibrations} object.
 
 For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method {[}SDM; Criscuolo, Berry, Douzery, and Gascuel (2006){]}, which deforms patristic distance matrices by minimizing variance and then averaging them.
-These single summary taxon pair age matrices (Summarized calibrations) can be applied as secondary calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Campbell O.
+These single summary taxon pair age matrices (Summarized calibrations) can be applied as calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Campbell O.
 Webb, Ackerly, \& Kembel, 2008; Campbell O.
@@
 This minimizes age variance in the resulting chronogram (Campbell O. Webb et al., 2008).
-BLADJ does not incorporate input phylogeny branch lengths even when they are present.
+BLADJ does not use branch lengths even when they are present in the input tree or summarizing topology.
 When there is conflict in ages between nodes with calibration data, the algorithm ignores node ages that are older than the age of a parent node.
@@
 
-Alternative options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the secondary calibrations.
+Alternative options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the calibrations.
 PATHd8 is a non-clock, rate-smoothing method (Britton et al., 2007) to date trees.
@@
 (Michael J. Sanderson, 2002).
-The MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003) approach in datelife uses the secondary calibrations as priors on node ages.
+The MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003) approach in datelife uses the calibrations as priors on node ages.
 
@@
 The name resolution step allowed discovering five data points for node ``n4'' that would not have had any data otherwise due to name mismatch.
-Age summary statistics per node were calculated and used as secondary calibrations to date the tree topology using the BLADJ algorithm.
+Age summary statistics per node were calculated and used as calibrations to date the tree topology using the BLADJ algorithm.
 Age data for node ``n2'' was excluded as final calibration because it is older than age data of the more inclusive node ``n1.''
@@
 
-\texttt{datelife} makes state-of-the-art information on time of lineage divergence easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
-It is an open service that does not require any expert biological knowledge from users, besides the names of the organisms or clade they want to work with, for any of its functionality.
-
-At the time of writing of this manuscript (May 26, 2022), \texttt{datelife}'s database has 253 chronograms, pulled entirely from OpenTree's database, the Phylesystem (McTavish et al., 2015).
-A unique feature of OpenTree's Phylesystem is that the community can add new state-of-the-art chronograms any time.
-As chronograms are added to Phylesystem, they are incorporated into an updated \texttt{datelife}'s database that is assigned a new version number, followed by a package release on CRAN.
-\texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
-Users can also upload new chronograms to OpenTree themselves, and trigger an update of their local \texttt{datelife} database to incorporate the new chronograms, to have them immediately available for analysis.
+\texttt{datelife} makes state-of-the-art data on evolutionary time frame easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
+It is an open service that does not require any expert biological knowledge from users --besides the names of the species or group they want to work with, for any of its functionality.
 
@@
 Incorporation of more chronograms into \texttt{datelife}'s database will continue to improve its services.
-One option to increase the of chronograms in the DateLife database is the Dryad data repository, which now holds some chronograms.
+One option to increase the number of chronograms in the DateLife database is the Dryad data repository.
 Methods to automatically mine chronograms from Dryad could be designed and implemented.
-However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined manually by a curator.
-
-The largest, and taxonomically broadest, summary chronogram currently available from OpenTree
-was constructed using age data from 2,274 published chronograms (Hedges et al., 2015).
-
+However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined manually by a human curator.
 We would like to emphasize on the importance of sharing chronogram data, including systematically curated metadata, into open repositories, such as OpenTree's Phylesystem (McTavish et al., 2015) for the benefit of the scientific community as a whole.
 
-By default, \texttt{datelife} currently summarizes all source chronograms that overlap with at least two species names. Users can exclude source chronograms if they have reasons to do so.
-Strictly speaking, the best chronogram should reflect the real time of lineage divergence accurately and precisely.
-To our knowledge, there are no good measures to determine independently if a chronogram is better than another.
-Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández (2015).
-Several characteristics of the data used for dating analyses, as well as from the output chronogram itself, could be used to score quality of source chronograms.
-Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
-DateLife provides an opportunity to capture concordance and conflict among date estimates, which can also be used as a metric for chronogram reliability.
-Its open database of trees allows other researchers to do such analyses themselves reproducibly and without needing permission, though of course they should follow proper citation practices, especially for the source trees.
-
-Scientists have more confidence in chronograms constructed using primary calibrations (ages obtained from the fossil or geological record) to ones constructed with secondary calibrations (ages coming from other chronograms)(Schenk, 2016).
-Schenk (2016) showed that in simulations divergence times inferred with secondary calibrations are significantly younger than those inferred with primary calibrations in analyses performed with Bayesian inference methods when priors are implemented in similar ways in both analyses.
-However, secondary calibrations are useful in dating methods that do not require setting priors, such as penalized likelihood (Michael J.
-Sanderson, 2003), or as fixed ages for nodes.
-Certainly, further studies are required to fully understand the effect of using secondary calibrations on time estimates and downstream analyses.
-
-The input chronograms may have been estimated using primary fossil data or using estimates of molecular substitution rates.
-These differences in approach which can deepen the already substantial variation in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
-For example, the chronograms from Burns et al. (2014) tend to have much older date estimates for the same nodes than other inferences (Figure 5).
-This study did not use any primary calibrations, and inferred dates using an substitution rate estimates across birds (Weir \& Schluter, 2008)
-
-Conflict among alternative date estimates is common in the literature
-(see, for example, the robust ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón et al., 2015; Ramshaw et al., 1972; Michael J.
-Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021).
-Using a different chronogram may return different results (Title \& Rabosky, 2016) particularly for studies of species diversification, or focussed on the timing of evolutionary events.
-Stitching together multiple chronograms creates a larger tree that uses information from across studies, but may amplify uncertainty and errors.
-
-Summarizing chronograms might also imply summarizing fundamentally distinct evolutionary hypotheses.
+\hypertarget{primary-vs-secondary-calibrations-in-dating-analyses}{%
+\subsection{Primary vs Secondary calibrations in dating analyses}\label{primary-vs-secondary-calibrations-in-dating-analyses}}
+
+While most chronograms in DateLife's database are constructed using primary calibrations (molecular substitution rates or ages obtained from the fossil record or geological events), DateLife summarizes chronograms using secondary calibrations only (ages coming from other chronograms).
+
+Graur and Martin (2004) cautioned on the increased error and uncertainty in estimated ages when using secondary calibrations in dating analyses.
+Schenk (2016) showed that, in simulations, divergence times inferred using secondary calibrations are significantly younger than those inferred with primary calibrations, when obtained with Bayesian inference methods, and when priors are implemented in similar ways in both analyses.
+Accordingly, the scientific community seems to have more confidence in chronograms obtained from a single analysis, using fossil data as primary sources of calibrations (Schenk, 2016), and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history (Sauquet, 2013), as for example done by (Antonelli et al., 2017).
+There have been attempts to create fossil calibration databases (Ksepka et al., 2015), though these still have room to grow.
+
+\textcolor{red}{LLSR: What does our cross validation analysis show in relation to diffs between primary and secondary calibrations?}
+
+It seems that using several (as opposed to just a few) secondary calibrations can provide sufficient information to alleviate or even neutralize potential biases (Sauquet, 2013).
+Certainly, further studies are required to fully understand the effect of secondary calibrations on outputs from different tree dating methods, and on downstream analyses.
+It is possible that secondary calibrations can be safely used with dating methods that do not require setting priors, such as penalized likelihood (Michael J.
+Sanderson, 2003), methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ (Campbell O.
+Webb et al., 2008; Campbell O.
+Webb \& Donoghue, 2005), or methods that summarize age data unto a tree topology.
+
+\hypertarget{variation-in-source-chronograms}{%
+\subsection{Variation in source chronograms}\label{variation-in-source-chronograms}}
+
+Conflict among alternative dating studies is common in the literature
+(see, for example, the robust ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández, 2015; Ramshaw et al., 1972; Michael J.
+Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021)).
+
+Source chronograms may have been estimated using different types of primary calibrations (fossil record, geologic or biogeographic events, molecular substitution rates).
+These differences in approach can deepen the already substantial variation in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
+For example, the chronograms from Burns et al.
+(2014) tend to have much older date estimates for the same nodes than other inferences (Figure 5).
+This study inferred dates using molecular substitution rate estimates across birds (Weir \& Schluter, 2008)
+\textcolor{red}{EJM: MORE ON the FINCH examples and the cross validation here}
+
+Different chronograms might also imply fundamentally distinct evolutionary hypotheses.
 For example, two different researchers working on the same clade could both carefully select and argue for their choices of fossil calibrations.
 Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses the same one to calibrate outside the clade, the resulting age estimates will often differ substantially, as the placement of calibrations as stem or crown group is proved to deeply affect estimated times of lineage divergence (Sauquet, 2013).
-Trying to summarize the resulting chronograms into a single one using simple summary statistics can erase many types of relevant information from the source chronograms.
-Accordingly, the prevailing view is that we should favor lineage divergence estimates obtained from a single analysis, using fossil data as primary sources of calibrations, and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history, as for example done by (Antonelli et al., 2017).
-There have been attempts to create fossil calibration databases , though these still have room to grow.
-However, the exercise of summarizing different chronograms has the potential to help getting a single global evolutionary history for a lineage by putting together evidence from different hypotheses.
-Choosing the elements of the chronograms that we are going to keep and the ones that we are going to discard is key, since we are potentially losing important parts of the evolutionary history of a lineage that might only be reflected in source chronograms and not on the summary chronogram (Sauquet et al., 2021).
-
-Nonetheless, in ecology and conservation biology, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance (Campbell O.
+
+\hypertarget{sumarizing-chronograms}{%
+\subsection{Sumarizing chronograms}\label{sumarizing-chronograms}}
+
+By default, DateLife currently summarizes all source chronograms that overlap with at least two species names. Users can exclude source chronograms if they have reasons to do so.
+Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely.
+To our knowledge, there are no tested measures to determine independently when a chronogram is better than another.
+Yet, several characteristics of the data used for dating analyses, as well as from the output chronogram itself, could be used to score the quality of source chronograms.
+
+Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used Magallón et al. (2015).
+Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
+
+DateLife provides an opportunity to capture concordance and conflict among date estimates, which can also be used as a metric for chronogram reliability.
+Its open database of chronograms allows other researchers to do such analyses themselves reproducibly, and without needing permission.
+Though, of course, they should follow proper citation practices, especially for the source chronogram studies.
+
+The exercise of summarizing age data from across multiple studies provides the opportunity to work with a more inclusive chronogram,
+that reflects a unified evolutionary history for a lineage, by putting together evidence from different hypotheses.
+The largest, and taxonomically broadest chronogram currently available from OpenTree was constructed summarizing age data from 2,274 published chronograms using NCBI's taxonomic tree as backbone (Hedges et al., 2015).
+A summarizing exercise may also amplify the effect of uncertainty and errors in source data, and blur parts of the evolutionary history of a lineage that might only be reflected in source chronograms and lost on the summary chronogram (Sauquet et al., 2021).
+\textcolor{red}{EJM: ??? it's not clear to me what this last part means! What might you be losing?
+LLSR: I changed loosing to blurring. But I will elaborate more on it}
+Choosing source chronograms that we are going to keep and the ones that we are going to discard might be key.
+\textcolor{red}{LLSR: Example}
+
+\hypertarget{effects-on-downstream-analyses}{%
+\subsection{Effects on downstream analyses}\label{effects-on-downstream-analyses}}
+
+For downstream analyses, using alternative chronogram may deeply affect our inferences (Title \& Rabosky, 2016), particularly when studying phenomena dependent on the timing of species diversification events, such as macroevolutionary processes.
+
+In ecology and conservation biology, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance (Campbell O.
 Webb et al., 2008).
-Hence, we integrated into datelife's workflow different ways of estimating node ages in the absence of calibrations and branch length information for taxa lacking this information.
-``Making up'' branch lengths is an common practice in scientific publications: Jetz et al.
+Hence, DateLife's workflow features different ways of estimating node ages in the absence of calibrations and branch length information for certain taxa.
+``Making up'' branch lengths is a common practice in scientific publications: Jetz et al.
 (2012), created a time-calibrated tree of all 9,993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
@@
 Smith and Brown (2018) constructed a tree of 353,185 seed plants where only 23\% had molecular data.
-Obviously, there are risks in this practice!
-Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration with the polytomy resolution and branch estimation methods.
-There has yet to be a thorough analysis of what can go wrong when one extends inferences beyond the data in this way, so we urge caution; we also urge readers to follow the example of many of the large tree papers cited above, by carefully considering the statistical assumptions being made, and assessing the consistency of the results with prior work.
+
+Notably, risks come with this practice.
+Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration, using polytomy resolution and branch estimation methods.
+There has yet to be a thorough analysis of what can go wrong when one extends inferences beyond the data in this way, so we urge caution; we also urge readers to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, and assessing the consistency of the results with prior work.
 
@@
 
-Divergence time information is key to many areas of evolutionary studies: trait evolution,
-diversification, biogeography, macroecology and more. It is also crucial for science communication and education, but generating chronograms is difficult,
+Knowledge of the evolutionary time frame of organisms is key to many research areas: trait evolution,
+species diversification, biogeography, macroecology and more. It is also crucial for education, science communication and policy, but generating chronograms is difficult,
 especially for those who want to use phylogenies but who are not systematists, or
-do not have the time to acquire and develop the necessary knowledge and data curation skills.
-Moreover, years of primarily public funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
-
-The \texttt{datelife} R package allows easy and fast summarization of publicly available information
-on time of lineage divergence.
-This provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identification of regions that require more research or that have conflicting information.
-It is available as an R package, or a web-based R shiny app at dates.opentreeloflife.org/datelife.
+do not have the time to acquire and develop the necessary knowledge and skills to construct them on their own.
+Importantly, years of primarily public funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
+
+The DateLife project allows for easy and fast summary of public and state-of-the-art data on time of lineage divergence.
+It provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identification of regions that require more research, or that have conflicting information.
+It is available as an R package, and as a web-based R shiny application at www.datelife.org
 Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
@@
 
-\texttt{datelife} is free and open source and it can be used through its current website
+The \texttt{datelife} software is free and open source and it can be used through its R shiny web application at
 \url{http://www.datelife.org}, through the \texttt{datelife} R package, and through Phylotastic's project web portal \url{https://phylo.cs.nmsu.edu/}.
-\texttt{datelife}'s website is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker.
+DateLife's web application is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker and Open Tree of Life's infrastructure (dates.opentreeoflife.org/datelife).
 \texttt{datelife}'s R package stable version is available
@@
 that led to initial work on this project.
-The team that assembled \texttt{datelife}'s first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
+The team that assembled DateLife's first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
 The Open Tree of Life project that provides the open, metadata rich repository of
-trees used for \texttt{datelife}.
-The many scientists who publish their chronograms in an open, reusable form, and
-the scientists who curate them for deposition in the Open Tree of Life repository.
+trees used to construct DateLife's chronogram database.
+The many scientists who publish their chronograms in an open, reusable form, and the scientists who curate them for deposition in the Open Tree of Life repository.
 The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-graur2004reading}{}}%
+Graur, D., \& Martin, W. (2004). Reading the entrails of chickens: Molecular timescales of evolution and the illusion of precision. \emph{TRENDS in Genetics}, \emph{20}(2), 80--86.
+
 \leavevmode\vadjust pre{\hypertarget{ref-hackett2008phylogenomic}{}}%
@@
 \emph{Molecular Phylogenetics and Evolution}, \emph{77}, 177--182.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ksepka2015fossil}{}}%
+Ksepka, D.
+T., Parham, J.
+F., Allman, J.
+F., Benton, M.
+J., Carrano, M.
+T., Cranston, K.
+A., \ldots{} others.
+(2015).
+The fossil calibration database---a new resource for divergence dating.
+\emph{Systematic Biology}, \emph{64}(5), 853--859.
 

### 2022-05-27 manuscript/manuscript.Rmd
updating figure 1 some more

@@
 
-At the time of writing of this manuscript (`r format(Sys.time(), "%b %d, %Y")`), DateLife's chronogram database latest version consist of `r length(opentree_chronograms$trees)` chronograms published in `r length(unique(opentree_chronograms$studies))` different studies.
+At the time of writing of this manuscript 
+,
+<!--(`r format(Sys.time(), "%b %d, %Y")`), -->
+DateLife's chronogram database latest version consist of `r length(opentree_chronograms$trees)` chronograms published in `r length(unique(opentree_chronograms$studies))` different studies.
 It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata [@mctavish2015phylesystem], which allows automatic and reproducible assembly of our chronogram database.

### 2022-05-27 manuscript/manuscript.tex
updating figure 1 some more

@@
 
-At the time of writing of this manuscript (May 27, 2022), DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies.
+At the time of writing of this manuscript
+,
+DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies.
 It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata (McTavish et al., 2015), which allows automatic and reproducible assembly of our chronogram database.
@@
 \includegraphics{../figures/figure-fringillidae/median_and_calibration_ages_simple.png}
-\caption{Fringillidae median summary chronogram genertaed with DateLife. It has 256 tips and 233 nodes.}
+\caption{Fringillidae median summary chronogram generated with DateLife. It has 256 tips and 233 nodes.}
 \label{fig:fringillidages}

### 2022-05-28 manuscript/figures-alt.tex
addressing all other comments
incorporating merged figure 2 to ms and other edits

@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 1    %%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{figure}[!h]
@@
 \end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:workflow}}
-% \end{center}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE runtime    %%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{figure}[!h]
@@
 \end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:benchmark}}
-% \end{center}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 2    %%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{figure}[!h]
-\includegraphics[width=1\linewidth]{../figures/figure2/figure2-1.pdf}
-\caption{Age data results of a DateLife search of a small sample of 6 bird species within the Passeriformes.
-Input names were found across 9 chronograms within 6 independent studies (Barker et al.
+\includegraphics[width=1\linewidth]{../figures/figure2/figure2.pdf}
+\caption{DateLife analysis results for a small sample of A) 6 bird species within the Passeriformes.
+B) Processed species names were found across 9 chronograms within 6 independent studies (Barker et al.
 (\protect\hyperlink{ref-barker2012going}{2012}), Barker et al.
@@
 (\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al.
-(\protect\hyperlink{ref-Jetz2012}{2012}).) This revealed 28 age data points for the queried species names.}
-\label{fig:figure2-1}
+(\protect\hyperlink{ref-Jetz2012}{2012}).) C) This revealed 28 source age data points for the queried species names.
+Summarized age data is used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree, resulting in a summary chronogram of source ages.}
+\label{fig:figure2}
 \end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:figure2-1}}
-% \end{center}
 
-\begin{figure}[!h]
-\includegraphics{../figures/figure2/figure2-2.pdf}
-\caption{Summarized age data is used as secondary calibrations to date a tree topology as a summary chronogram.}
-\label{fig:summaries}
-\end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:figure2-2}}
-% \end{center}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 2-1  %%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\begin{figure}[!h]
+%\includegraphics[width=1\linewidth]{../figures/figure2/figure2-1.pdf}
+%\caption{Age data results of a DateLife search of a small sample of 6 %bird species within the Passeriformes.
+Input names were found across 9 %chronograms within 6 independent studies (Barker et al.
+%(\protect\hyperlink{ref-barker2012going}{2012}), Barker et al.
+%(\protect\hyperlink{ref-barker2015new}{2015}), Burns et al.
+%(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al.
+%(\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price %(\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al.
+%(\protect\hyperlink{ref-Jetz2012}{2012}).) This revealed 28 age data %points for the queried species names.}
+%\label{fig:figure2-1}
+%\end{figure}
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 2-2  %%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\begin{figure}[!h]
+%\includegraphics{../figures/figure2/figure2-2.pdf}
+%\caption{Summarized age data is used as secondary calibrations to date %a tree topology as a summary chronogram.}
+%\label{fig:summaries}
+%\end{figure}
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 3    %%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{figure}[!h]
@@
 \caption{Tree topologies obtained from Open Tree of Life's (OpenTree) synthetic phylogenetic tree.
-A) Topology of 2333 tips and 1305 internal nodes, encompassing bird species within the family Fringillidae following the NCBI taxonomy (black), as well as all other bird species that share the same mrca node in OpenTree's synthetic tree (purple).
+A) Topology of 2,333 tips and 1,305 internal nodes, encompassing bird species within the family Fringillidae following the NCBI taxonomy (black), as well as all other bird species that share the same mrca node in OpenTree's synthetic tree (purple).
 B) Topology of 289 tips and 253 internal nodes, encompassing bird species within the Fringillidae only, resulting from pruning purple branches from topology A.
-Bird species within the Fringillidae do not form a monophyletic group.
+Bird species within the Fringillidae do not form a monophyletic group
 (Alström et al. \protect\hyperlink{ref-}{2014},
@@
 Treplin et al. \protect\hyperlink{ref-treplin2008molecular}{2008},
-Zuccon, Prŷs-Jones, Rasmussen, \& Ericson \protect\hyperlink{ref-zuccon2012phylogenetic}{2012})
+Zuccon, Prŷs-Jones, Rasmussen, \& Ericson \protect\hyperlink{ref-zuccon2012phylogenetic}{2012}).
 }
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 4    %%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{figure}[!h]
 \includegraphics{../figures/figure-fringillidae/median_and_calibration_ages_simple.png}
-\caption{Fringillidae median summary chronogram genertaed with DateLife. It has 256 tips and 233 nodes.}
+\caption{Fringillidae median summary chronogram generated with DateLife. It has 256 tips and 233 nodes.}
 \label{fig:fringillidages}
 \end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:fringillidages}}
-% \end{center}
-
-% \end{linenumbers}

### 2022-05-28 manuscript/manuscript.Rmd
incorporating merged figure 2 to ms and other edits

@@
 If an input taxon name belongs to an inclusive taxonomic group, the algorithm has two alternative behaviors defined by the "get species from taxon" flag.
-If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group (according to a taxonomy of choice) and adds them to the input string.
-If the flag is inactive, DateLife excludes the taxon names above the species level from the query.
+If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group (following a taxonomy of choice; current options are Open Tree of Life, NCBI, GBIF, or IRMNG taxonomies) and adds them to the input string.
+If the flag is inactive, DateLife excludes any taxon names above the species level from the search query.
 
@@
 We recommend that users provide a tree topology as input from the literature, or one of their own making.
-If no topology is provided, DateLife automatically subsets one from the OpenTree synthetic tree [@opentreeoflife2019synth].
+If no topology is provided, DateLife automatically obtains one from the OpenTree synthetic tree [@opentreeoflife2019synth].
+These are known as "induced" OpenTree subtrees, which do not necessarily correspond to a monophyletic group.
 Alternatively, DateLife can combine topologies from source chronograms using a supertree approach.
@@
 
-## Small example
+## A small example
 
 _**Creating a search query.-**_
-We chose 6 bird species in the family Fringillidae, known as the true finches. The sample includes
+We chose 6 bird species within the Passeriformes. The sample includes
 two species of cardinals:
@@
 For a detailed discussion on the state of the synonym refer to Avibase [@lepage2004avibase; @lepage2014avibase; @avibase-emberiza].
-Discovering this synonym allowed assigning five age data points for the parent node of *Emberiza elegans*, shown as *Schoeniclus elegans* in figure \ref{fig:figure2-1}, which would not have had any data otherwise.
+Discovering this synonym allowed assigning five age data points for the parent node of *Emberiza elegans*, shown as *Schoeniclus elegans* in figure \ref{fig:figure2}A, which would not have had any data otherwise.
 <!--\textcolor{red}{EJM: I don't understand what this sentence means}
@@
 _**Searching the database.-**_
-DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies.
+DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies (Fig.
+\ref{fig:figure2}B).
 Three studies matched five input names [@barker2015new; @Hedges2015; @Jetz2012], one study matched four input names [@hooper2017chromosomal] and two studies matched two input names [@barker2013going; @burns2014phylogenetics].
@@
 _**Summarizing search results.-**_
-DateLife used OpenTree's synthetic tree topology for these taxa and mapped age data to nodes in the tree.
+DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and mapped age data to nodes in this chosen topology.
 As expected, more inclusive nodes (e.g., node "n1") have more age data than less inclusive nodes (e.g., node "n5").
-The name resolution step allowed discovering five data points for node "n4" that would not have had any data otherwise due to name mismatch.
+The name processing step allowed including five data points for node "n4" (parent of _Schoeniclus elegans_; Fig. \ref{fig:figure2}A) that would not have had any data otherwise due to name mismatch.
 Age summary statistics per node were calculated and used as calibrations to date the tree topology using the BLADJ algorithm.
-Age data for node "n2" was excluded as final calibration because it is older than age data of the more inclusive node "n1".
+Age data for node "n2" was excluded as final calibration because it is older than age data of the more inclusive node "n1" (Fig. \ref{fig:figure2}C4).
 
-## Fringillidae
+## The family Fringillidae
 
 _**Creating a query.-**_
-To estimate ages for species in the family Fringillidae, we ran a `datelife` query using the "get species from taxon" flag,
-which gets all recognized species names within a named group from a taxonomy of choice (options are Open Tree of Life, NCBI, GBIF, or IRMNG).
-Following the NCBI taxonomy, the Fringillidae has `r length(dquery$cleaned_names)` species.
+To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the "get species from taxon" flag,
+which gets all recognized species names within a named group from a taxonomy of choice.
+Following the NCBI taxonomy, our DateLife query has `r length(dquery$cleaned_names)` Fringillidae species.
 <!--BCO: Feel like we need a number here, though it's good that the clade hasn't gone extinct ;-) -->
+This taxon-constrained approach implies that the final results of a full DateLife analysis will be done using a tree topology and ages for the species in a named group, which do not necessarily correspond to a monophyletic group.
+Users can change this behaviour by providing a monophyletic tree as input for a DateLife search, or as a tree topology for a DateLife summary.
+<!--BCO: it's ambiguous: it gets a topology and age for the larger clade that includes all Fringillidae species? Or for the paraphyletic Fringillidae group?
+Luna: Clarified in text, I hope!-->
 
 _**Searching the database.-**_
-The next step is to use this curated set of species taxon names to identify all chronograms that contain at least two Fringillidae species.
-Once identified, the algorithm proceeds to prune matching chronograms to keep Fringillidae species names on tips only, and transform these pruned chronograms to pairwise distance matrices.
-The `datelife` search revealed `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
+Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species. 
+The DateLife search identified `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
+Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices.
 
@@
 As explained in the "Description" section, a tree topology to summarize age data upon is required.
-By default, `datelife` uses the Open Tree of Life synthetic phylogeny as summarizing topology.
-According to this phylogeny, species belonging to the family Fringillidae do not form a monophyletic group (Fig.
-\ref{fig:fringillidae-topologies}).
-By default, DateLife's taxon-constrained approach returns a topology and ages for the species in a named group, which do not necessarily correspond to a monophyletic group.
-Users can change this behaviour by providing a monophyletic tree as input for a DateLife search, or as a tree topology for a DateLife summary.
-
-<!--BCO: it's ambiguous: it gets a topology and age for the larger clade that includes all Fringillidae species? Or for the paraphyletic Fringillidae group?
-Luna: Clarified in text, I hope!-->
-
+By default, DateLife uses the portion of OpenTree synthetic phylogeny that contains the species provided as search query as summarizing topology. 
+According to this phylogeny, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).
 Age data from source chronograms is summarised into a single patristic distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
 \ref{fig:fringillidages}).
-
 Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.

### 2022-05-28 manuscript/manuscript.tex
incorporating merged figure 2 to ms and other edits

@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4765}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4823}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 If an input taxon name belongs to an inclusive taxonomic group, the algorithm has two alternative behaviors defined by the ``get species from taxon'' flag.
-If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group (according to a taxonomy of choice) and adds them to the input string.
-If the flag is inactive, DateLife excludes the taxon names above the species level from the query.
+If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group (following a taxonomy of choice; current options are Open Tree of Life, NCBI, GBIF, or IRMNG taxonomies) and adds them to the input string.
+If the flag is inactive, DateLife excludes any taxon names above the species level from the search query.
 
@@
 We recommend that users provide a tree topology as input from the literature, or one of their own making.
-If no topology is provided, DateLife automatically subsets one from the OpenTree synthetic tree (Open Tree Of Life et al., 2019).
+If no topology is provided, DateLife automatically obtains one from the OpenTree synthetic tree (Open Tree Of Life et al., 2019).
+These are known as ``induced'' OpenTree subtrees, which do not necessarily correspond to a monophyletic group.
 Alternatively, DateLife can combine topologies from source chronograms using a supertree approach.
@@
 
-\hypertarget{small-example}{%
-\subsection{Small example}\label{small-example}}
+\hypertarget{a-small-example}{%
+\subsection{A small example}\label{a-small-example}}
 
 \emph{\textbf{Creating a search query.-}}
-We chose 6 bird species in the family Fringillidae, known as the true finches. The sample includes
+We chose 6 bird species within the Passeriformes. The sample includes
 two species of cardinals:
@@
 For a detailed discussion on the state of the synonym refer to Avibase (Avibase, 2022; Lepage, 2004; Lepage, Vaidya, \& Guralnick, 2014).
-Discovering this synonym allowed assigning five age data points for the parent node of \emph{Emberiza elegans}, shown as \emph{Schoeniclus elegans} in figure \ref{fig:figure2-1}, which would not have had any data otherwise.
+Discovering this synonym allowed assigning five age data points for the parent node of \emph{Emberiza elegans}, shown as \emph{Schoeniclus elegans} in figure \ref{fig:figure2}A, which would not have had any data otherwise.
 
 \emph{\textbf{Searching the database.-}}
-DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies.
+DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies (Fig.
+\ref{fig:figure2}B).
 Three studies matched five input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2015; Hedges, Marin, Suleski, Paymer, \& Kumar, 2015; Jetz, Thomas, Joy, Hartmann, \& Mooers, 2012), one study matched four input names (Hooper \& Price, 2017) and two studies matched two input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2013; Burns et al., 2014).
@@
 \emph{\textbf{Summarizing search results.-}}
-DateLife used OpenTree's synthetic tree topology for these taxa and mapped age data to nodes in the tree.
+DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and mapped age data to nodes in this chosen topology.
 As expected, more inclusive nodes (e.g., node ``n1'') have more age data than less inclusive nodes (e.g., node ``n5'').
-The name resolution step allowed discovering five data points for node ``n4'' that would not have had any data otherwise due to name mismatch.
+The name processing step allowed including five data points for node ``n4'' (parent of \emph{Schoeniclus elegans}; Fig.
+\ref{fig:figure2}A) that would not have had any data otherwise due to name mismatch.
 Age summary statistics per node were calculated and used as calibrations to date the tree topology using the BLADJ algorithm.
-Age data for node ``n2'' was excluded as final calibration because it is older than age data of the more inclusive node ``n1.''
-
-\hypertarget{fringillidae}{%
-\subsection{Fringillidae}\label{fringillidae}}
+Age data for node ``n2'' was excluded as final calibration because it is older than age data of the more inclusive node ``n1'' (Fig. \ref{fig:figure2}C4).
+
+\hypertarget{the-family-fringillidae}{%
+\subsection{The family Fringillidae}\label{the-family-fringillidae}}
 
 \emph{\textbf{Creating a query.-}}
-To estimate ages for species in the family Fringillidae, we ran a \texttt{datelife} query using the ``get species from taxon'' flag,
-which gets all recognized species names within a named group from a taxonomy of choice (options are Open Tree of Life, NCBI, GBIF, or IRMNG).
-Following the NCBI taxonomy, the Fringillidae has 289 species.
+To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the ``get species from taxon'' flag,
+which gets all recognized species names within a named group from a taxonomy of choice.
+Following the NCBI taxonomy, our DateLife query has 289 Fringillidae species.
+This taxon-constrained approach implies that the final results of a full DateLife analysis will be done using a tree topology and ages for the species in a named group, which do not necessarily correspond to a monophyletic group.
+Users can change this behaviour by providing a monophyletic tree as input for a DateLife search, or as a tree topology for a DateLife summary.
 
 \emph{\textbf{Searching the database.-}}
-The next step is to use this curated set of species taxon names to identify all chronograms that contain at least two Fringillidae species.
-Once identified, the algorithm proceeds to prune matching chronograms to keep Fringillidae species names on tips only, and transform these pruned chronograms to pairwise distance matrices.
-The \texttt{datelife} search revealed 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Price et al., 2014).
+Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species.
+The DateLife search identified 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Price et al., 2014).
+Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices.
 
@@
 As explained in the ``Description'' section, a tree topology to summarize age data upon is required.
-By default, \texttt{datelife} uses the Open Tree of Life synthetic phylogeny as summarizing topology.
-According to this phylogeny, species belonging to the family Fringillidae do not form a monophyletic group (Fig.
-\ref{fig:fringillidae-topologies}).
-By default, DateLife's taxon-constrained approach returns a topology and ages for the species in a named group, which do not necessarily correspond to a monophyletic group.
-Users can change this behaviour by providing a monophyletic tree as input for a DateLife search, or as a tree topology for a DateLife summary.
-
+By default, DateLife uses the portion of OpenTree synthetic phylogeny that contains the species provided as search query as summarizing topology.
+According to this phylogeny, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).
 Age data from source chronograms is summarised into a single patristic distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
 \ref{fig:fringillidages}).
-
 Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
@@
 \end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:workflow}}
-% \end{center}
-
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE runtime    %%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{figure}[!h]
@@
 \end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:benchmark}}
-% \end{center}
-
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 2    %%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{figure}[!h]
-\includegraphics[width=1\linewidth]{../figures/figure2/figure2-1.pdf}
-\caption{Age data results of a DateLife search of a small sample of 6 bird species within the Passeriformes.
-Input names were found across 9 chronograms within 6 independent studies (Barker et al.
+\includegraphics[width=1\linewidth]{../figures/figure2/figure2.pdf}
+\caption{DateLife analysis results for a small sample of A) 6 bird species within the Passeriformes.
+B) Processed species names were found across 9 chronograms within 6 independent studies (Barker et al.
 (\protect\hyperlink{ref-barker2012going}{2012}), Barker et al.
@@
 (\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al.
-(\protect\hyperlink{ref-Jetz2012}{2012}).) This revealed 28 age data points for the queried species names.}
-\label{fig:figure2-1}
+(\protect\hyperlink{ref-Jetz2012}{2012}).) C) This revealed 28 source age data points for the queried species names.
+Summarized age data is used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree, resulting in a summary chronogram of source ages.}
+\label{fig:figure2}
 \end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:figure2-1}}
-% \end{center}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure2/figure2-2.pdf}
-\caption{Summarized age data is used as secondary calibrations to date a tree topology as a summary chronogram.}
-\label{fig:summaries}
-\end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:figure2-2}}
-% \end{center}
-
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 2-1  %%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\begin{figure}[!h]
+%\includegraphics[width=1\linewidth]{../figures/figure2/figure2-1.pdf}
+%\caption{Age data results of a DateLife search of a small sample of 6 %bird species within the Passeriformes.
+Input names were found across 9 %chronograms within 6 independent studies (Barker et al.
+%(\protect\hyperlink{ref-barker2012going}{2012}), Barker et al.
+%(\protect\hyperlink{ref-barker2015new}{2015}), Burns et al.
+%(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al.
+%(\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price %(\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al.
+%(\protect\hyperlink{ref-Jetz2012}{2012}).) This revealed 28 age data %points for the queried species names.}
+%\label{fig:figure2-1}
+%\end{figure}
+%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 2-2  %%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\begin{figure}[!h]
+%\includegraphics{../figures/figure2/figure2-2.pdf}
+%\caption{Summarized age data is used as secondary calibrations to date %a tree topology as a summary chronogram.}
+%\label{fig:summaries}
+%\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 3    %%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{figure}[!h]
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 4    %%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{figure}[!h]
@@
 \end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:fringillidages}}
-% \end{center}
-
-% \end{linenumbers}
 

### 2022-05-31 manuscript/manuscript.Rmd
creating table 1 as .tex in small example vignette, and adding it to manuscript as graphic

@@
       in_header: "preamble.tex"
-      after_body  : ["figures-alt.tex", "figures-supplementary.tex"]
+      after_body  : ["tables.tex",
+                     "figures-alt.tex", 
+                     "figures-supplementary.tex"]
 

### 2022-05-31 manuscript/table-fringillidae-small-example.tex
creating table 1 as .tex in small example vignette, and adding it to manuscript as graphic

\begin{table}

\caption{\label{tab:table1}Table 1. Ages of congruified nodes.}
\centering
\begin{tabular}[t]{l|c|c|c|c|c}
\hline
  & Node Name & taxon A & taxon B & Node Age & Study chronogram\\
\hline
1 & n1 & Emberiza citrinella & Platyspiza crassirostris & 9.6509881 & Jetz, W., G. H. Thomas, J. B. Joy, K. Hartmann, A. O. Mooers. 2012. The global diversity of birds in space and time. Nature 491 (7424): 444-448\\
\hline
2 & n1 & Pheucticus tibialis & Schoeniclus elegans & 14.3336520 & Barker, F. Keith, Kevin J. Burns, John Klicka, Scott M. Lanyon, Irby J. Lovette. 2015. New insights into New World biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies. The Auk 132 (2): 333-348.\\
\hline
3 & n1 & Rhodothraupis celaeno & Schoeniclus elegans & 16.2984859 & Barker, F. Keith, Kevin J. Burns, John Klicka, Scott M. Lanyon, Irby J. Lovette. 2015. New insights into New World biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies. The Auk 132 (2): 333-348.\\
\hline
4 & n1 & Emberiza citrinella & Platyspiza crassirostris & 16.9499615 & Hooper, Daniel M., Trevor D. Price. 2017. Chromosomal inversion differences correlate with range overlap in passerine birds. Nature Ecology \& Evolution 1 (10): 1526-1534\\
\hline
5 & n1 & Rhodothraupis celaeno & Schoeniclus elegans & 21.5140867 & Jetz, W., G. H. Thomas, J. B. Joy, K. Hartmann, A. O. Mooers. 2012. The global diversity of birds in space and time. Nature 491 (7424): 444-448\\
\hline
6 & n1 & Emberiza leucocephalos & Platyspiza crassirostris & 22.0000000 & Hedges, S. Blair, Julie Marin, Michael Suleski, Madeline Paymer, Sudhir Kumar. 2015. Tree of life reveals clock-like speciation and diversification. Molecular Biology and Evolution 32 (4): 835-845\\
\hline
7 & n1 & Emberiza citrinella & Platyspiza crassirostris & 22.3757277 & Hedges, S. Blair, Julie Marin, Michael Suleski, Madeline Paymer, Sudhir Kumar. 2015. Tree of life reveals clock-like speciation and diversification. Molecular Biology and Evolution 32 (4): 835-845\\
\hline
8 & n2 & Platyspiza crassirostris & Rhodothraupis celaeno & 7.9691925 & Jetz, W., G. H. Thomas, J. B. Joy, K. Hartmann, A. O. Mooers. 2012. The global diversity of birds in space and time. Nature 491 (7424): 444-448\\
\hline
9 & n2 & Platyspiza crassirostris & Rhodothraupis celaeno & 19.7085900 & Hedges, S. Blair, Julie Marin, Michael Suleski, Madeline Paymer, Sudhir Kumar. 2015. Tree of life reveals clock-like speciation and diversification. Molecular Biology and Evolution 32 (4): 835-845\\
\hline
10 & n2 & Pheucticus tibialis & Platyspiza crassirostris & 44.2958603 & Burns, Kevin J., Allison J. Shultz, Pascal O. Title, Nicholas A. Mason, F. Keith Barker, John Klicka, Scott M. Lanyon, Irby J. Lovette. 2014. Phylogenetics and diversification of tanagers (Passeriformes: Thraupidae), the largest radiation of Neotropical songbirds. Molecular Phylogenetics and Evolution 75: 41-77.\\
\hline
11 & n3 & Pheucticus tibialis & Rhodothraupis celaeno & 10.5304440 & Barker, F. Keith, Kevin J. Burns, John Klicka, Scott M. Lanyon, Irby J. Lovette. 2015. New insights into New World biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies. The Auk 132 (2): 333-348.\\
\hline
12 & n3 & Pheucticus tibialis & Rhodothraupis celaeno & 10.5379092 & Barker, F. K., K. J. Burns, J. Klicka, S. M. Lanyon, I. J. Lovette. 2013. Going to extremes: contrasting rates of diversification in a recent radiation of New World passerine birds. Systematic Biology 62 (2): 298-320.\\
\hline
13 & n3 & Pheucticus tibialis & Rhodothraupis celaeno & 18.1570685 & Hedges, S. Blair, Julie Marin, Michael Suleski, Madeline Paymer, Sudhir Kumar. 2015. Tree of life reveals clock-like speciation and diversification. Molecular Biology and Evolution 32 (4): 835-845\\
\hline
14 & n4 & Emberiza citrinella & Schoeniclus elegans & 5.3238969 & Jetz, W., G. H. Thomas, J. B. Joy, K. Hartmann, A. O. Mooers. 2012. The global diversity of birds in space and time. Nature 491 (7424): 444-448\\
\hline
15 & n4 & Emberiza leucocephalos & Schoeniclus elegans & 9.8622460 & Barker, F. Keith, Kevin J. Burns, John Klicka, Scott M. Lanyon, Irby J. Lovette. 2015. New insights into New World biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies. The Auk 132 (2): 333-348.\\
\hline
16 & n4 & Emberiza leucocephalos & Schoeniclus elegans & 10.3391445 & Jetz, W., G. H. Thomas, J. B. Joy, K. Hartmann, A. O. Mooers. 2012. The global diversity of birds in space and time. Nature 491 (7424): 444-448\\
\hline
17 & n4 & Emberiza citrinella & Schoeniclus elegans & 12.5133870 & Hooper, Daniel M., Trevor D. Price. 2017. Chromosomal inversion differences correlate with range overlap in passerine birds. Nature Ecology \& Evolution 1 (10): 1526-1534\\
\hline
18 & n5 & Emberiza citrinella & Emberiza leucocephalos & 0.1407015 & Jetz, W., G. H. Thomas, J. B. Joy, K. Hartmann, A. O. Mooers. 2012. The global diversity of birds in space and time. Nature 491 (7424): 444-448\\
\hline
19 & n5 & Emberiza citrinella & Emberiza leucocephalos & 0.1516230 & Hedges, S. Blair, Julie Marin, Michael Suleski, Madeline Paymer, Sudhir Kumar. 2015. Tree of life reveals clock-like speciation and diversification. Molecular Biology and Evolution 32 (4): 835-845\\
\hline
20 & n5 & Emberiza citrinella & Emberiza leucocephalos & 0.2011990 & Barker, F. Keith, Kevin J. Burns, John Klicka, Scott M. Lanyon, Irby J. Lovette. 2015. New insights into New World biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies. The Auk 132 (2): 333-348.\\
\hline
21 & n5 & Emberiza citrinella & Emberiza leucocephalos & 2.2898230 & Hooper, Daniel M., Trevor D. Price. 2017. Chromosomal inversion differences correlate with range overlap in passerine birds. Nature Ecology \& Evolution 1 (10): 1526-1534\\
\hline
\end{tabular}
\end{table}

### 2022-05-31 manuscript/tables.tex
creating table 1 as .tex in small example vignette, and adding it to manuscript as graphic

--- 2022-05-27 manuscript/manuscript.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-  english,
-  man]{apa6}
-\title{DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life}
-\author{Luna L. Sánchez Reyes\textsuperscript{1,2}, Emily Jane McTavish\textsuperscript{1}, \& Brian O'Meara\textsuperscript{2}}
-\date{}
-
-\usepackage{amsmath,amssymb}
-\usepackage{lmodern}
-\usepackage{iftex}
-\ifPDFTeX
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life},
-  pdfauthor={Luna L. Sánchez Reyes1,2, Emily Jane McTavish1, \& Brian O'Meara2},
-  pdflang={en-EN},
-  pdfkeywords={Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-\DeclareUnicodeCharacter{0301}{*************************************}
-% Manuscript styling
-\usepackage{upgreek}
-\captionsetup{font=singlespacing,justification=justified}
-
-% Table formatting
-\usepackage{longtable}
-\usepackage{lscape}
-% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
-\usepackage{multirow}		% Table styling
-\usepackage{tabularx}		% Control Column width
-\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
-\usepackage{threeparttablex}            % Lets threeparttable work with longtable
-
-% Create new environments so endfloat can handle them
-% \newenvironment{ltable}
-%   {\begin{landscape}\centering\begin{threeparttable}}
-%   {\end{threeparttable}\end{landscape}}
-\newenvironment{lltable}{\begin{landscape}\centering\begin{ThreePartTable}}{\end{ThreePartTable}\end{landscape}}
-
-% Enables adjusting longtable caption width to table width
-% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
-\makeatletter
-\newcommand\LastLTentrywidth{1em}
-\newlength\longtablewidth
-\setlength{\longtablewidth}{1in}
-\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}
-
-% \setlength{\parindent}{0.5in}
-% \setlength{\parskip}{0pt plus 0pt minus 0pt}
-
-% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
-% See https://github.com/crsh/papaja/issues/292
-\makeatletter
-\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-1em}%
-  {\normalfont\normalsize\bfseries\itshape\typesectitle}}
-
-\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-\z@\relax}%
-  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
-\makeatother
-
-% \usepackage{etoolbox}
-\makeatletter
-\patchcmd{\HyOrg@maketitle}
-  {\section{\normalfont\normalsize\abstractname}}
-  {\section*{\normalfont\normalsize\abstractname}}
-  {}{\typeout{Failed to patch abstract.}}
-\patchcmd{\HyOrg@maketitle}
-  {\section{\protect\normalfont{\@title}}}
-  {\section*{\protect\normalfont{\@title}}}
-  {}{\typeout{Failed to patch title.}}
-\makeatother
-
-\usepackage{xpatch}
-\makeatletter
-\xapptocmd\appendix
-  {\xapptocmd\section
-    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
-    {}{\InnerPatchFailed}%
-  }
-{}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4765}
-\DeclareDelayedFloatFlavor{ThreePartTable}{table}
-\DeclareDelayedFloatFlavor{lltable}{table}
-\DeclareDelayedFloatFlavor*{longtable}{table}
-\makeatletter
-\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
-\makeatother
-\usepackage{lineno}
-
-\linenumbers
-\usepackage{csquotes}
-\ifXeTeX
-  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
-  \usepackage{polyglossia}
-  \setmainlanguage[]{english}
-\else
-  \usepackage[main=english]{babel}
-% get rid of language-specific shorthands (see #6817):
-\let\LanguageShortHands\languageshorthands
-\def\languageshorthands#1{}
-\fi
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-
-\shorttitle{DateLife: revealing the dated Tree of Life}
-
-\authornote{
-
-School of Natural Sciences, University of California, Merced, Science and Engineering Building 1.
-
-Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 446 Hesler Biology Building, Knoxville, TN 37996, USA.
-
-The authors made the following contributions.
-Luna L.
-Sánchez Reyes: Data curation, Investigation, Software, Visualization, Validation, Writing - Original Draft Preparation, Writing - Review \& Editing; Emily Jane McTavish: Resources, Software, Writing - Review \& Editing; Brian O'Meara: Conceptualization, Funding acquisition, Methodology, Resources, Software, Supervision, Writing - Review \& Editing.
-
-Correspondence concerning this article should be addressed to Luna L. Sánchez Reyes, . E-mail: \href{mailto:sanchez.reyes.luna@gmail.com}{\nolinkurl{sanchez.reyes.luna@gmail.com}}
-
-}
-
-\affiliation{\vspace{0.5cm}\textsuperscript{1} University of California, Merced\\\textsuperscript{2} University of Tennessee, Knoxville}
-
-\abstract{%
-Achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram) is a difficult and time-consuming task.
-But the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
-Knowledge of the evolutionary time frame of organisms is key for research in the natural sciences. It also represent valuable information for education, science communication, and policy decisions.
-When chronograms are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on evolutionary timeframe is exposed in a programatic and reusable way, as intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
-Here we present DateLife, a service implemented as an R package and an R Shiny website application available at www.datelife.org, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time frame of evolution.
-The main DateLife workflow constructs a chronogram for any given combination of taxon names by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from the TreeBASE database as well.
-We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
-DateLife will be useful to increase awereness of the existing variation in alternative hypothesis of evolutionary time for the same organisms, and can foster exploration of the effect of alternative evolutionary timing hypotheses on the results of downstream analyses, providing a framework for a more informed interpretation of evolutionary results.
-}
-
-\begin{document}
-\maketitle
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-Chronograms --phylogenies with branch lengths proportional to time-- provide key data on evolutionary time frame for the study of natural processes in many areas of biological research, such as developmental biology (Delsuc et al., 2018; Laubichler \& Maienschein, 2009), conservation biology (Felsenstein, 1985; C.
-Webb, 2000), historical biogeography (Posadas, Crisci, \& Katinas, 2006), and species diversification (Magallon \& Sanderson, 2001; Morlon, 2014).
-
-Building a chronogram is not an easy task.
-It requires obtaining and curating data to construct a phylogeny, selecting and placing appropriate calibrations on the phylogeny using independent age data points from the fossil record or other dated events, and inferring the full dated tree;
-it also generally requires specialized biological training, taxonomic domain knowledge, and a non-negligible amount of research time, computational resources and funding.
-
-Here we present the DateLife project which has the main goal of capturing age data from published chronograms, and making these data readily accessible to the community for reuse and reanalysis.
-DateLife's core software application is available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org.
-It features key elements for scientific reproducibility, such as a versioned, open and fully public source database (McTavish et al., 2015), data stored and available in a computer readable format (Vos et al., 2012), automated and programmatic ways of accessing the data (Stoltzfus et al., 2013) and methods to summarize and compare the data.
-
-\hypertarget{description}{%
-\section{Description}\label{description}}
-
-DateLife's core software application consists of the R package \texttt{datelife}.
-Its latest stable version -- v0.6.2, is available from the CRAN repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
-ape (Paradis, Claude, \& Strimmer, 2004),
-bold (Chamberlain et al., 2019),
-geiger (Pennell et al., 2014),
-paleotree (Bapst, 2012),
-phyloch (Heibl, 2008),
-phylocomr (Ooms \& Chamberlain, 2018),
-phytools (Revell, 2012),
-rotl (Michonneau, Brown, \& Winter, 2016), and
-taxize (Chamberlain \& Szöcs, 2013; Chamberlain et al., 2019).
-Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife algorithm: creating a search query, searching a database, and summarizing results from the search.
-
-\hypertarget{creating-a-search-query}{%
-\subsection{Creating a search query}\label{creating-a-search-query}}
-
-DateLife starts by processing an input consisting of at least two taxon names, which can be provided as a comma separated character string or as tip labels on a tree.
-If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis et al., 2004).
-The input tree is not required to have branch lengths, and its topology is used in the summary steps described in the next section.
-
-DateLife accepts scientific names that can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or a binomial species name.
-Subspecies and variants are ignored.
-If an input taxon name belongs to an inclusive taxonomic group, the algorithm has two alternative behaviors defined by the ``get species from taxon'' flag.
-If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group (according to a taxonomy of choice) and adds them to the input string.
-If the flag is inactive, DateLife excludes the taxon names above the species level from the query.
-
-DateLife processes input scientific names using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database.
-TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard (Boyle et al., 2013).
-DateLife implements TNRS using OpenTree's unified taxonomy as standard (Open Tree Of Life et al., 2016; Rees \& Cranston, 2017), storing OpenTree's Taxonomy identification numbers for further processing.
-
-The processed input taxon names are saved as an R object of a newly defined class, \texttt{datelifeQuery}, that is used in the following steps.
-This object contains the standardized names, the corresponding OpenTree taxonomic id numbers, and the topology of the input tree if any was provided.
-
-\hypertarget{searching-a-chronogram-database}{%
-\subsection{Searching a chronogram database}\label{searching-a-chronogram-database}}
-
-At the time of writing of this manuscript
-,
-DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies.
-It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata (McTavish et al., 2015), which allows automatic and reproducible assembly of our chronogram database.
-Datelife's chronogram database is navigable as an R data object within the \texttt{datelife} R package.
-
-A unique feature of OpenTree's Phylesystem is that the community can add new published, state-of-the-art chronograms any time through their curator application (\url{https://tree.opentreeoflife.org/curator}).
-As chronograms are added to Phylesystem, they are incorporated into an updated \texttt{datelife}'s database that is assigned a new version number, followed by a package release on CRAN.
-\texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
-Users can also upload and curate new chronograms to OpenTree themselves.
-Then, trigger an update of their local \texttt{datelife} database to incorporate the new chronograms and have them available for their current analysis.
-
-A DateLife search is implemented by matching processed taxon names provided by the user to tip labels in the chronogram database.
-Chronograms with at least two matching taxon names on their tip labels are identified and pruned down to preserve only the matched taxa.
-These matching pruned chronograms are referred to as source chronograms.
-Total distance (in units of millions of years) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Figure \ref{fig:workflow}).
-The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a ``phylo'' object or newick string.
-Finally, the patristic matrices are associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.
-
-\hypertarget{summarizing-search-results}{%
-\subsection{Summarizing search results}\label{summarizing-search-results}}
-
-Summary information is extracted from the \texttt{datelifeResult} object to inform decisions for subsequent steps in the analysis workflow. Basic summary information available to the user is:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  The matching pruned chronograms as newick strings or ``phylo'' objects.
-\item
-  The ages of the root of all source chronograms.
-These ages can correspond to the age of the most recent common ancestor (mrca) of the user's group of interest if the source chronograms have all taxa belonging to the group.
-If not, the root corresponds to the mrca of a subgroup withing the group of interest.
-\item
-  Study citations where original chronograms were published.
-\item
-  A report of input taxon names matches across source chronograms.
-\item
-  The source chronogram(s) with the most input taxon names.
-\item
-  Various single summary chronograms resulting from summarizing age data, generated using the methodology described next.
-\end{enumerate}
-
-\emph{\textbf{Choosing a topology.--}}
-DateLife requires a tree topology to summarize age data upon.
-We recommend that users provide a tree topology as input from the literature, or one of their own making.
-If no topology is provided, DateLife automatically subsets one from the OpenTree synthetic tree (Open Tree Of Life et al., 2019).
-Alternatively, DateLife can combine topologies from source chronograms using a supertree approach.
-To combine topologies from source chronograms into a single summary (or supertree) topology, the DateLife algorithm starts by identifying the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
-(2009).
-In rare cases, a group of trees can have multiple groves.
-By default, DateLife chooses the grove with the most taxa, however, the ``criterion = trees'' flag allows the user to choose the grove with the most trees instead.
-If source chronograms do not form a grove, the supertree reconstruction will fail.
-
-\emph{\textbf{Dating the topology.--}}
-Input topologies from OpenTree or the supertree approach described above do not include branch length estimates of any kind.
-Optionally, to estimate branch lengths proportional to substitution rates on these topologies, DateLife can mine the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain genetic markers for the input taxa.
-These markers are aligned with MUSCLE (Edgar, 2004) (by default) or MAFFT (Katoh, Asimenos, \& Toh, 2009).
-This alignment can be used to estimate branch lengths on input topologies that lack branch lengths.
-Currently, branch length reconstruction in DateLife is performed using parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed (Schliep, 2011).
-While relative branch length information provides additional data for nodes without secondary date calibrations, topologies without branch lengths can also be dated.
-
-Once a topology is chosen, DateLife applies the congruification method (Eastman, Harmon, \& Tank, 2013) to find nodes belonging to the same clade across source chronograms, and extract the corresponding node ages from the patristic distance matrices stored as \texttt{datelifeResult}.
-By definition, the matrices store total distance (time from tip to tip), hence, node ages correspond to half the values stored in the patristic distance matrices.
-This assumes that the terminal taxa are coeval and occur at the present.
-A table of congruified node ages that can be used as calibrations for a dating analysis is stored as a \texttt{congruifiedCalibrations} object.
-
-For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method {[}SDM; Criscuolo, Berry, Douzery, and Gascuel (2006){]}, which deforms patristic distance matrices by minimizing variance and then averaging them.
-These single summary taxon pair age matrices (Summarized calibrations) can be applied as calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Campbell O.
-Webb, Ackerly, \& Kembel, 2008; Campbell O.
-Webb \& Donoghue, 2005), and treePL (Stephen A.
-Smith \& O'Meara, 2012).
-
-By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully dated topology.
-BLADJ fixes node ages that have calibration data, and distributes time between nodes with no data evenly between nodes with calibration data.
-This minimizes age variance in the resulting chronogram (Campbell O. Webb et al., 2008).
-BLADJ does not use branch lengths even when they are present in the input tree or summarizing topology.
-When there is conflict in ages between nodes with calibration data, the algorithm ignores node ages that are older than the age of a parent node.
-BLADJ requires a root age estimate.
-If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature.
-If none is provided, DateLife assigns an arbitrary age to the root as 10\% older than the oldest age available within the group.
-
-Alternative options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the calibrations.
-PATHd8 is a non-clock, rate-smoothing method (Britton et al., 2007) to date trees.
-treePL (Stephen A. Smith \& O'Meara, 2012), is a semi-parametric, rate-smoothing, penalized likelihood dating method
-(Michael J. Sanderson, 2002).
-The MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003) approach in datelife uses the calibrations as priors on node ages.
-
-\emph{\textbf{Visualizing results.--}}
-Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R ``phylo'' format), as well as visualize and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions available from the R package \texttt{datelifeplot} (Sanchez-Reyes \& O'Meara, 2022).
-
-\hypertarget{benchmark}{%
-\section{Benchmark}\label{benchmark}}
-
-\texttt{datelife}'s code speed was tested on an Apple iMac
-with one 3.4 GHz Intel Core i5 processor.
-We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
-Query processing time increases roughly linearly with number of input taxon names, and
-increases considerably if Taxonomic Name Resolution Service (TNRS) is activated.
-Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
-Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig.
-\ref{fig:runtime_main}).
-\texttt{datelife}'s code performance was evaluated with a set of unit tests designed and
-implemented with the R package testthat (R Core Team, 2018) that were run both locally
-with the devtools package (R Core Team, 2018), and on a public server --via
-GitHub, using the continuous integration tool Travis CI (\url{https://travis-ci.org}). At
-present, unit tests cover more than 40\% of \texttt{datelife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).
-Unit testing helps identify potential issues as code is updated or, more critically, as services code relies upon may change.
-
-\hypertarget{case-studies}{%
-\section{Case studies}\label{case-studies}}
-
-We illustrate the DateLife algorithm using a family within the Passeriform birds encompassing the true finches, Fringillidae, as case study.
-The first example analyses 6 bird species and shows all steps of the algorithm.
-The second example is an application analysing 289 species in the family Fringillidae that are included in the NCBI taxonomy.
-
-\hypertarget{small-example}{%
-\subsection{Small example}\label{small-example}}
-
-\emph{\textbf{Creating a search query.-}}
-We chose 6 bird species in the family Fringillidae, known as the true finches. The sample includes
-two species of cardinals:
-the black-thighed grosbeak -- \emph{Pheucticus tibialis} and
-the crimson-collared grosbeak -- \emph{Rhodothraupis celaeno};
-three species of buntings:
-the yellowhammer -- \emph{Emberiza citrinella},
-the pine bunting -- \emph{Emberiza leucocephalos} and
-the yellow-throated bunting -- \emph{Emberiza elegans};
-and one species of tanager, the vegetarian finch -- \emph{Platyspiza crassirostris}.
-
-Processing input names found that \emph{Emberiza elegans} is synonym for \emph{Schoeniclus elegans} in the default reference taxonomy {[}Open Tree of Life Taxonomy v3.3, June 1, 2021{]}.
-For a detailed discussion on the state of the synonym refer to Avibase (Avibase, 2022; Lepage, 2004; Lepage, Vaidya, \& Guralnick, 2014).
-Discovering this synonym allowed assigning five age data points for the parent node of \emph{Emberiza elegans}, shown as \emph{Schoeniclus elegans} in figure \ref{fig:figure2-1}, which would not have had any data otherwise.
-
-\emph{\textbf{Searching the database.-}}
-DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies.
-Three studies matched five input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2015; Hedges, Marin, Suleski, Paymer, \& Kumar, 2015; Jetz, Thomas, Joy, Hartmann, \& Mooers, 2012), one study matched four input names (Hooper \& Price, 2017) and two studies matched two input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2013; Burns et al., 2014).
-No studies matched all input names.
-Together, matching chronograms have 28 unique age data points.
-All nodes have age data.
-
-\emph{\textbf{Summarizing search results.-}}
-DateLife used OpenTree's synthetic tree topology for these taxa and mapped age data to nodes in the tree.
-As expected, more inclusive nodes (e.g., node ``n1'') have more age data than less inclusive nodes (e.g., node ``n5'').
-The name resolution step allowed discovering five data points for node ``n4'' that would not have had any data otherwise due to name mismatch.
-Age summary statistics per node were calculated and used as calibrations to date the tree topology using the BLADJ algorithm.
-Age data for node ``n2'' was excluded as final calibration because it is older than age data of the more inclusive node ``n1.''
-
-\hypertarget{fringillidae}{%
-\subsection{Fringillidae}\label{fringillidae}}
-
-\emph{\textbf{Creating a query.-}}
-To estimate ages for species in the family Fringillidae, we ran a \texttt{datelife} query using the ``get species from taxon'' flag,
-which gets all recognized species names within a named group from a taxonomy of choice (options are Open Tree of Life, NCBI, GBIF, or IRMNG).
-Following the NCBI taxonomy, the Fringillidae has 289 species.
-
-\emph{\textbf{Searching the database.-}}
-The next step is to use this curated set of species taxon names to identify all chronograms that contain at least two Fringillidae species.
-Once identified, the algorithm proceeds to prune matching chronograms to keep Fringillidae species names on tips only, and transform these pruned chronograms to pairwise distance matrices.
-The \texttt{datelife} search revealed 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Price et al., 2014).
-
-\emph{\textbf{Summarizing search results.-}}
-The final step is to summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM.
-As explained in the ``Description'' section, a tree topology to summarize age data upon is required.
-By default, \texttt{datelife} uses the Open Tree of Life synthetic phylogeny as summarizing topology.
-According to this phylogeny, species belonging to the family Fringillidae do not form a monophyletic group (Fig.
-\ref{fig:fringillidae-topologies}).
-By default, DateLife's taxon-constrained approach returns a topology and ages for the species in a named group, which do not necessarily correspond to a monophyletic group.
-Users can change this behaviour by providing a monophyletic tree as input for a DateLife search, or as a tree topology for a DateLife summary.
-
-Age data from source chronograms is summarised into a single patristic distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
-\ref{fig:fringillidages}).
-
-Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
-
-\hypertarget{cross-validation-test}{%
-\section{Cross-validation test}\label{cross-validation-test}}
-
-We performed a cross validation analysis of the DateLife workflow using the Fringillidae chronograms.
-We used the individual tree topologies from each of the 19 source chronograms from 13 studies as inputs, treating their node ages as unknown.
-We then estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.
-
-We found that node ages from original study, and ages estimated using all other age data available are correlated (Supplementary Fig. \ref{fig:cvXY}).
-
-For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka `closer to the root') relative to the orginal estimate, and overestimate ages for nodes closer to the tips.
-Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. \ref{fig:cvXYdiffs}).
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-\texttt{datelife} makes state-of-the-art data on evolutionary time frame easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
-It is an open service that does not require any expert biological knowledge from users --besides the names of the species or group they want to work with, for any of its functionality.
-
-A total of 99474 unique terminal taxa are represented in \texttt{datelife}'s database.
-Incorporation of more chronograms into \texttt{datelife}'s database will continue to improve its services.
-One option to increase the number of chronograms in the DateLife database is the Dryad data repository.
-Methods to automatically mine chronograms from Dryad could be designed and implemented.
-However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined manually by a human curator.
-We would like to emphasize on the importance of sharing chronogram data, including systematically curated metadata, into open repositories, such as OpenTree's Phylesystem (McTavish et al., 2015) for the benefit of the scientific community as a whole.
-
-\hypertarget{primary-vs-secondary-calibrations-in-dating-analyses}{%
-\subsection{Primary vs Secondary calibrations in dating analyses}\label{primary-vs-secondary-calibrations-in-dating-analyses}}
-
-While most chronograms in DateLife's database are constructed using primary calibrations (molecular substitution rates or ages obtained from the fossil record or geological events), DateLife summarizes chronograms using secondary calibrations only (ages coming from other chronograms).
-
-Graur and Martin (2004) cautioned on the increased error and uncertainty in estimated ages when using secondary calibrations in dating analyses.
-Schenk (2016) showed that, in simulations, divergence times inferred using secondary calibrations are significantly younger than those inferred with primary calibrations, when obtained with Bayesian inference methods, and when priors are implemented in similar ways in both analyses.
-Accordingly, the scientific community seems to have more confidence in chronograms obtained from a single analysis, using fossil data as primary sources of calibrations (Schenk, 2016), and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history (Sauquet, 2013), as for example done by (Antonelli et al., 2017).
-There have been attempts to create fossil calibration databases (Ksepka et al., 2015), though these still have room to grow.
-
-\textcolor{red}{LLSR: What does our cross validation analysis show in relation to diffs between primary and secondary calibrations?}
-
-It seems that using several (as opposed to just a few) secondary calibrations can provide sufficient information to alleviate or even neutralize potential biases (Sauquet, 2013).
-Certainly, further studies are required to fully understand the effect of secondary calibrations on outputs from different tree dating methods, and on downstream analyses.
-It is possible that secondary calibrations can be safely used with dating methods that do not require setting priors, such as penalized likelihood (Michael J.
-Sanderson, 2003), methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ (Campbell O.
-Webb et al., 2008; Campbell O.
-Webb \& Donoghue, 2005), or methods that summarize age data unto a tree topology.
-
-\hypertarget{variation-in-source-chronograms}{%
-\subsection{Variation in source chronograms}\label{variation-in-source-chronograms}}
-
-Conflict among alternative dating studies is common in the literature
-(see, for example, the robust ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández, 2015; Ramshaw et al., 1972; Michael J.
-Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021)).
-
-Source chronograms may have been estimated using different types of primary calibrations (fossil record, geologic or biogeographic events, molecular substitution rates).
-These differences in approach can deepen the already substantial variation in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
-For example, the chronograms from Burns et al.
-(2014) tend to have much older date estimates for the same nodes than other inferences (Figure 5).
-This study inferred dates using molecular substitution rate estimates across birds (Weir \& Schluter, 2008)
-\textcolor{red}{EJM: MORE ON the FINCH examples and the cross validation here}
-
-Different chronograms might also imply fundamentally distinct evolutionary hypotheses.
-For example, two different researchers working on the same clade could both carefully select and argue for their choices of fossil calibrations.
-Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses the same one to calibrate outside the clade, the resulting age estimates will often differ substantially, as the placement of calibrations as stem or crown group is proved to deeply affect estimated times of lineage divergence (Sauquet, 2013).
-
-\hypertarget{sumarizing-chronograms}{%
-\subsection{Sumarizing chronograms}\label{sumarizing-chronograms}}
-
-By default, DateLife currently summarizes all source chronograms that overlap with at least two species names. Users can exclude source chronograms if they have reasons to do so.
-Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely.
-To our knowledge, there are no tested measures to determine independently when a chronogram is better than another.
-Yet, several characteristics of the data used for dating analyses, as well as from the output chronogram itself, could be used to score the quality of source chronograms.
-
-Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used Magallón et al. (2015).
-Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
-
-DateLife provides an opportunity to capture concordance and conflict among date estimates, which can also be used as a metric for chronogram reliability.
-Its open database of chronograms allows other researchers to do such analyses themselves reproducibly, and without needing permission.
-Though, of course, they should follow proper citation practices, especially for the source chronogram studies.
-
-The exercise of summarizing age data from across multiple studies provides the opportunity to work with a more inclusive chronogram,
-that reflects a unified evolutionary history for a lineage, by putting together evidence from different hypotheses.
-The largest, and taxonomically broadest chronogram currently available from OpenTree was constructed summarizing age data from 2,274 published chronograms using NCBI's taxonomic tree as backbone (Hedges et al., 2015).
-A summarizing exercise may also amplify the effect of uncertainty and errors in source data, and blur parts of the evolutionary history of a lineage that might only be reflected in source chronograms and lost on the summary chronogram (Sauquet et al., 2021).
-\textcolor{red}{EJM: ??? it's not clear to me what this last part means! What might you be losing?
-LLSR: I changed loosing to blurring. But I will elaborate more on it}
-Choosing source chronograms that we are going to keep and the ones that we are going to discard might be key.
-\textcolor{red}{LLSR: Example}
-
-\hypertarget{effects-on-downstream-analyses}{%
-\subsection{Effects on downstream analyses}\label{effects-on-downstream-analyses}}
-
-For downstream analyses, using alternative chronogram may deeply affect our inferences (Title \& Rabosky, 2016), particularly when studying phenomena dependent on the timing of species diversification events, such as macroevolutionary processes.
-
-In ecology and conservation biology, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance (Campbell O.
-Webb et al., 2008).
-Hence, DateLife's workflow features different ways of estimating node ages in the absence of calibrations and branch length information for certain taxa.
-``Making up'' branch lengths is a common practice in scientific publications: Jetz et al.
-(2012), created a time-calibrated tree of all 9,993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
-(2018) created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37\% had molecular data; Stephen A.
-Smith and Brown (2018) constructed a tree of 353,185 seed plants where only 23\% had molecular data.
-
-Notably, risks come with this practice.
-Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration, using polytomy resolution and branch estimation methods.
-There has yet to be a thorough analysis of what can go wrong when one extends inferences beyond the data in this way, so we urge caution; we also urge readers to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, and assessing the consistency of the results with prior work.
-
-\hypertarget{conclusions}{%
-\section{Conclusions}\label{conclusions}}
-
-Knowledge of the evolutionary time frame of organisms is key to many research areas: trait evolution,
-species diversification, biogeography, macroecology and more. It is also crucial for education, science communication and policy, but generating chronograms is difficult,
-especially for those who want to use phylogenies but who are not systematists, or
-do not have the time to acquire and develop the necessary knowledge and skills to construct them on their own.
-Importantly, years of primarily public funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
-
-The DateLife project allows for easy and fast summary of public and state-of-the-art data on time of lineage divergence.
-It provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identification of regions that require more research, or that have conflicting information.
-It is available as an R package, and as a web-based R shiny application at www.datelife.org
-Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
-The DateLife project helps with awareness of the existing variation in expert time of divergence data, and will foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
-
-\hypertarget{availability}{%
-\section{Availability}\label{availability}}
-
-The \texttt{datelife} software is free and open source and it can be used through its R shiny web application at
-\url{http://www.datelife.org}, through the \texttt{datelife} R package, and through Phylotastic's project web portal \url{https://phylo.cs.nmsu.edu/}.
-DateLife's web application is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker and Open Tree of Life's infrastructure (dates.opentreeoflife.org/datelife).
-\texttt{datelife}'s R package stable version is available
-for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
-using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R. Development versions
-are available from the GitHub repository (\url{https://github.com/phylotastic/datelife})
-and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.
-
-\hypertarget{supplementary-material}{%
-\section{Supplementary Material}\label{supplementary-material}}
-
-Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities are available at \href{https://github.com/LunaSare/datelifeMS1}{datelifeMS1}, \href{https://github.com/LunaSare/datelife_examples}{datelife\_examples}, and \href{https://github.com/LunaSare/datelife_benchmark}{datelife\_benchmark} repositories in LLSR's GitHub account.
-
-\hypertarget{funding}{%
-\section{Funding}\label{funding}}
-
-Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to the Datelife project; DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), ABI-1458572 to the Phylotastic project, and ABI-1759846 to the Open Tree of Life project.
-
-\hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
-
-The DateLife project was born as a prototype tool aiming to provide these services, and was initially developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA (Stoltzfus et al., 2013).
-We thank colleagues from the O'Meara Lab at the University
-of Tennesse Knoxville for suggestions, discussions and software testing.
-The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
-that led to initial work on this project.
-The team that assembled DateLife's first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
-The Open Tree of Life project that provides the open, metadata rich repository of
-trees used to construct DateLife's chronogram database.
-The many scientists who publish their chronograms in an open, reusable form, and the scientists who curate them for deposition in the Open Tree of Life repository.
-The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.
-
-\newpage
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-alstrom2014discovery}{}}%
-Alström, P., Hooper, D.
-M., Liu, Y., Olsson, U., Mohan, D., Gelang, M., \ldots{} Price, T.
-D.
-(2014).
-Discovery of a relict lineage and monotypic family of passerine birds.
-\emph{Biology Letters}, \emph{10}(3), 20131067.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ane2009groves}{}}%
-Ané, C., Eulenstein, O., Piaggio-Talice, R., \& Sanderson, M. J. (2009). Groves of phylogenetic trees. \emph{{Annals of Combinatorics}}, \emph{13}(2), 139--167.
-
-\leavevmode\vadjust pre{\hypertarget{ref-antonelli2017supersmart}{}}%
-Antonelli, A., Hettling, H., Condamine, F.
-L., Vos, K., Nilsson, R.
-H., Sanderson, M.
-J., \ldots{} Vos, R.
-A.
-(2017).
-{Toward a self-updating platform for estimating rates of speciation and migration, ages, and relationships of Taxa}.
-\emph{Systematic Biology}, \emph{66}(2), 153--166.
-\url{https://doi.org/10.1093/sysbio/syw066}
-
-\leavevmode\vadjust pre{\hypertarget{ref-archie1986newick}{}}%
-Archie, J., Day, W.
-H., Felsenstein, J., Maddison, W., Meacham, C., Rohlf, F.
-J., \& Swofford, D.
-(1986).
-{The Newick tree format}.
-Retrieved from \url{\%7Bhttps://evolution.genetics.washington.edu/phylip/newicktree.html\%7D}
-
-\leavevmode\vadjust pre{\hypertarget{ref-avibase-emberiza}{}}%
-Avibase.
-(2022).
-{Yellow-throated Bunting}.
-\emph{{Avibase - The World Bird Database}}, (Online Resource).
-Retrieved from \url{\%7Bhttps://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\%7D}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Bapst2012a}{}}%
-Bapst, D.
-W.
-(2012).
-{Paleotree: An R package for paleontological and phylogenetic analyses of evolution}.
-\emph{{Methods in Ecology and Evolution}}, \emph{3}(5), 803--807.
-\url{https://doi.org/10.1111/j.2041-210X.2012.00223.x}
-
-\leavevmode\vadjust pre{\hypertarget{ref-barba2018constraining}{}}%
-Barba-Montoya, J., Reis, M.
-dos, Schneider, H., Donoghue, P.
-C., \& Yang, Z.
-(2018).
-Constraining uncertainty in the timescale of angiosperm evolution and the veracity of a cretaceous terrestrial revolution.
-\emph{New Phytologist}, \emph{218}(2), 819--834.
-
-\leavevmode\vadjust pre{\hypertarget{ref-barker2014mitogenomic}{}}%
-Barker, F. K. (2014). Mitogenomic data resolve basal relationships among passeriform and passeridan birds. \emph{Molecular Phylogenetics and Evolution}, \emph{79}, 313--324.
-
-\leavevmode\vadjust pre{\hypertarget{ref-barker2013going}{}}%
-Barker, F.
-K., Burns, K.
-J., Klicka, J., Lanyon, S.
-M., \& Lovette, I.
-J.
-(2013).
-Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds.
-\emph{{Systematic Biology}}, \emph{62}(2), 298--320.
-
-\leavevmode\vadjust pre{\hypertarget{ref-barker2015new}{}}%
-Barker, F.
-K., Burns, K.
-J., Klicka, J., Lanyon, S.
-M., \& Lovette, I.
-J.
-(2015).
-New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
-\emph{The Auk: Ornithological Advances}, \emph{132}(2), 333--348.
-
-\leavevmode\vadjust pre{\hypertarget{ref-barker2004phylogeny}{}}%
-Barker, F.
-K., Cibois, A., Schikler, P., Feinstein, J., \& Cracraft, J.
-(2004).
-Phylogeny and diversification of the largest avian radiation.
-\emph{Proceedings of the National Academy of Sciences}, \emph{101}(30), 11040--11045.
-
-\leavevmode\vadjust pre{\hypertarget{ref-beresford2005african}{}}%
-Beresford, P., Barker, F., Ryan, Pg., \& Crowe, T.
-(2005).
-African endemics span the tree of songbirds (passeri): Molecular systematics of several evolutionary {`enigmas.'} \emph{Proceedings of the Royal Society B: Biological Sciences}, \emph{272}(1565), 849--858.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Boyle2013}{}}%
-Boyle, B., Hopkins, N., Lu, Z., Raygoza Garay, J.
-A., Mozzherin, D., Rees, T., \ldots{} Enquist, B.
-J.
-(2013).
-{The taxonomic name resolution service: An online tool for automated standardization of plant names}.
-\emph{{BMC Bioinformatics}}, \emph{14}(1).
-\url{https://doi.org/10.1186/1471-2105-14-16}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Britton2007}{}}%
-Britton, T., Anderson, C.
-L., Jacquet, D., Lundqvist, S., \& Bremer, K.
-(2007).
-{Estimating Divergence Times in Large Phylogenetic Trees}.
-\emph{{Systematic Biology}}, \emph{56}(788777878), 741--752.
-\url{https://doi.org/10.1080/10635150701613783}
-
-\leavevmode\vadjust pre{\hypertarget{ref-bryson2014diversification}{}}%
-Bryson Jr, R.
-W., Chaves, J., Smith, B.
-T., Miller, M.
-J., Winker, K., Pérez-Emán, J.
-L., \& Klicka, J.
-(2014).
-Diversification across the new world within the `blue'cardinalids (aves: cardinalidae).
-\emph{Journal of Biogeography}, \emph{41}(3), 587--599.
-
-\leavevmode\vadjust pre{\hypertarget{ref-burleigh2015building}{}}%
-Burleigh, J. G., Kimball, R. T., \& Braun, E. L. (2015). Building the avian tree of life using a large-scale, sparse supermatrix. \emph{Molecular Phylogenetics and Evolution}, \emph{84}, 53--63.
-
-\leavevmode\vadjust pre{\hypertarget{ref-burns2014phylogenetics}{}}%
-Burns, K.
-J., Shultz, A.
-J., Title, P.
-O., Mason, N.
-A., Barker, F.
-K., Klicka, J., \ldots{} Lovette, I.
-J.
-(2014).
-Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds.
-\emph{{Molecular Phylogenetics and Evolution}}, \emph{75}, 41--77.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Chamberlain2013}{}}%
-Chamberlain, S.
-A., \& Szöcs, E.
-(2013).
-{taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}}.
-\emph{{F1000Research}}, \emph{2}(191), 1--29.
-\url{https://doi.org/10.12688/f1000research.2-191.v2}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Chamberlain2018}{}}%
-Chamberlain, S.
-A., Szöcs, E., Foster, Z., Arendsee, Z., Boettiger, C., Ram, K., \ldots{} Li, G.
-(2019).
-\emph{{taxize: Taxonomic information from around the web}}.
-Retrieved from \url{https://github.com/ropensci/taxize}
-
-\leavevmode\vadjust pre{\hypertarget{ref-chaves2013biogeography}{}}%
-Chaves, J.
-A., Hidalgo, J.
-R., \& Klicka, J.
-(2013).
-Biogeography and evolutionary history of the n eotropical genus s altator (a ves: T hraupini).
-\emph{Journal of Biogeography}, \emph{40}(11), 2180--2190.
-
-\leavevmode\vadjust pre{\hypertarget{ref-claramunt2015new}{}}%
-Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{{Science Advances}}, \emph{1}(11), e1501005.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Criscuolo2006}{}}%
-Criscuolo, A., Berry, V., Douzery, E.
-J.
-P., \& Gascuel, O.
-(2006).
-{SDM: A fast distance-based approach for (super)tree building in phylogenomics}.
-\emph{{Systematic Biology}}, \emph{55}(5), 740--755.
-\url{https://doi.org/10.1080/10635150600969872}
-
-\leavevmode\vadjust pre{\hypertarget{ref-delsuc2018phylogenomic}{}}%
-Delsuc, F., Philippe, H., Tsagkogeorga, G., Simion, P., Tilak, M.-K., Turon, X., \ldots{} Douzery, E.
-J.
-(2018).
-A phylogenomic framework and timescale for comparative studies of tunicates.
-\emph{BMC Biology}, \emph{16}(1), 1--14.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Eastman2013}{}}%
-Eastman, J.
-M., Harmon, L.
-J., \& Tank, D.
-C.
-(2013).
-{Congruification: Support for time scaling large phylogenetic trees}.
-\emph{{Methods in Ecology and Evolution}}, \emph{4}(7), 688--691.
-\url{https://doi.org/10.1111/2041-210X.12051}
-
-\leavevmode\vadjust pre{\hypertarget{ref-edgar2004muscle}{}}%
-Edgar, R. C. (2004). MUSCLE: Multiple sequence alignment with high accuracy and high throughput. \emph{Nucleic Acids Research}, \emph{32}(5), 1792--1797.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Felsenstein1985a}{}}%
-Felsenstein, J. (1985). {Phylogenies and the Comparative Method}. \emph{The American Naturalist}, \emph{125}(1), 1--15. Retrieved from \url{http://www.jstor.org/stable/2461605}
-
-\leavevmode\vadjust pre{\hypertarget{ref-gibb2015new}{}}%
-Gibb, G.
-C., England, R., Hartig, G., McLenachan, P.
-A., Taylor Smith, B.
-L., McComish, B.
-J., \ldots{} Penny, D.
-(2015).
-New zealand passerines help clarify the diversification of major songbird lineages during the oligocene.
-\emph{{Genome Biology and Evolution}}, \emph{7}(11), 2983--2995.
-
-\leavevmode\vadjust pre{\hypertarget{ref-graur2004reading}{}}%
-Graur, D., \& Martin, W. (2004). Reading the entrails of chickens: Molecular timescales of evolution and the illusion of precision. \emph{TRENDS in Genetics}, \emph{20}(2), 80--86.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hackett2008phylogenomic}{}}%
-Hackett, S.
-J., Kimball, R.
-T., Reddy, S., Bowie, R.
-C., Braun, E.
-L., Braun, M.
-J., \ldots{} others.
-(2008).
-A phylogenomic study of birds reveals their evolutionary history.
-\emph{Science}, \emph{320}(5884), 1763--1768.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Hedges2015}{}}%
-Hedges, S.
-B., Marin, J., Suleski, M., Paymer, M., \& Kumar, S.
-(2015).
-{Tree of life reveals clock-like speciation and diversification}.
-\emph{{Molecular Biology and Evolution}}, \emph{32}(4), 835--845.
-\url{https://doi.org/10.1093/molbev/msv037}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Heibl2008}{}}%
-Heibl, C. (2008). \emph{PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages.} Retrieved from \url{http://www.christophheibl.de/Rpackages.html}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hooper2017chromosomal}{}}%
-Hooper, D. M., \& Price, T. D. (2017). Chromosomal inversion differences correlate with range overlap in passerine birds. \emph{Nature Ecology \& Evolution}, \emph{1}(10), 1526.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Huelsenbeck2001}{}}%
-Huelsenbeck, J. P., \& Ronquist, F. (2001). {MRBAYES: Bayesian inference of phylogenetic trees}. \emph{Bioinformatics}, \emph{17}(8), 754--755. \url{https://doi.org/10.1093/bioinformatics/17.8.754}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Jetz2012}{}}%
-Jetz, W., Thomas, G., Joy, J.
-J.
-B., Hartmann, K., \& Mooers, A.
-(2012).
-{The global diversity of birds in space and time}.
-\emph{Nature}, \emph{491}(7424), 444--448.
-\url{https://doi.org/10.1038/nature11631}
-
-\leavevmode\vadjust pre{\hypertarget{ref-johansson2008phylogenetic}{}}%
-Johansson, U.
-S., Fjeldså, J., \& Bowie, R.
-C.
-(2008).
-Phylogenetic relationships within passerida (aves: Passeriformes): A review and a new molecular phylogeny based on three nuclear intron markers.
-\emph{Molecular Phylogenetics and Evolution}, \emph{48}(3), 858--876.
-
-\leavevmode\vadjust pre{\hypertarget{ref-katoh2009multiple}{}}%
-Katoh, K., Asimenos, G., \& Toh, H. (2009). Multiple alignment of DNA sequences with MAFFT. In \emph{Bioinformatics for DNA sequence analysis} (pp. 39--64). Springer.
-
-\leavevmode\vadjust pre{\hypertarget{ref-kimball2019phylogenomic}{}}%
-Kimball, R. T., Oliveros, C. H., Wang, N., White, N. D., Barker, F. K., Field, D. J., \ldots{} others. (2019). A phylogenomic supertree of birds. \emph{Diversity}, \emph{11}(7), 109.
-
-\leavevmode\vadjust pre{\hypertarget{ref-klicka2014comprehensive}{}}%
-Klicka, J., Barker, F.
-K., Burns, K.
-J., Lanyon, S.
-M., Lovette, I.
-J., Chaves, J.
-A., \& Bryson Jr, R.
-W.
-(2014).
-A comprehensive multilocus assessment of sparrow (aves: Passerellidae) relationships.
-\emph{Molecular Phylogenetics and Evolution}, \emph{77}, 177--182.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ksepka2015fossil}{}}%
-Ksepka, D.
-T., Parham, J.
-F., Allman, J.
-F., Benton, M.
-J., Carrano, M.
-T., Cranston, K.
-A., \ldots{} others.
-(2015).
-The fossil calibration database---a new resource for divergence dating.
-\emph{Systematic Biology}, \emph{64}(5), 853--859.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lamichhaney2015evolution}{}}%
-Lamichhaney, S., Berglund, J., Almén, M.
-S., Maqbool, K., Grabherr, M., Martinez-Barrio, A., \ldots{} others.
-(2015).
-Evolution of darwin's finches and their beaks revealed by genome sequencing.
-\emph{Nature}, \emph{518}(7539), 371--375.
-
-\leavevmode\vadjust pre{\hypertarget{ref-laubichler2009form}{}}%
-Laubichler, M. D., \& Maienschein, J. (2009). \emph{Form and function in developmental evolution}. Cambridge University Press.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lepage2004avibase}{}}%
-Lepage, D. (2004). \emph{Avibase: The world bird database}. Bird Studies Canada.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lepage2014avibase}{}}%
-Lepage, D., Vaidya, G., \& Guralnick, R. (2014). Avibase--a database system for managing and organizing taxonomic concepts. \emph{ZooKeys}, (420), 117.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lerner2011multilocus}{}}%
-Lerner, H.
-R., Meyer, M., James, H.
-F., Hofreiter, M., \& Fleischer, R.
-C.
-(2011).
-Multilocus resolution of phylogeny and timescale in the extant adaptive radiation of hawaiian honeycreepers.
-\emph{Current Biology}, \emph{21}(21), 1838--1844.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lovette2010comprehensive}{}}%
-Lovette, I.
-J., Pérez-Emán, J.
-L., Sullivan, J.
-P., Banks, R.
-C., Fiorentino, I., Córdoba-Córdoba, S., \ldots{} others.
-(2010).
-A comprehensive multilocus phylogeny for the wood-warblers and a revised classification of the parulidae (aves).
-\emph{Molecular Phylogenetics and Evolution}, \emph{57}(2), 753--770.
-
-\leavevmode\vadjust pre{\hypertarget{ref-magallon2001absolute}{}}%
-Magallon, S., \& Sanderson, M. J. (2001). Absolute diversification rates in angiosperm clades. \emph{Evolution}, \emph{55}(9), 1762--1780.
-
-\leavevmode\vadjust pre{\hypertarget{ref-magallon2010using}{}}%
-Magallón, S. (2010). Using fossils to break long branches in molecular dating: A comparison of relaxed clocks applied to the origin of angiosperms. \emph{Systematic Biology}, \emph{59}(4), 384--399.
-
-\leavevmode\vadjust pre{\hypertarget{ref-magallon2015metacalibrated}{}}%
-Magallón, S., Gómez-Acevedo, S., Sánchez-Reyes, L.
-L., \& Hernández-Hernández, T.
-(2015).
-A metacalibrated time-tree documents the early rise of flowering plant phylogenetic diversity.
-\emph{New Phytologist}, \emph{207}(2), 437--453.
-
-\leavevmode\vadjust pre{\hypertarget{ref-mctavish2015phylesystem}{}}%
-McTavish, E.
-J., Hinchliff, C.
-E., Allman, J.
-F., Brown, J.
-W., Cranston, K.
-A., Holder, M.
-T., \ldots{} Smith, S.
-A.
-(2015).
-Phylesystem: A git-based data store for community-curated phylogenetic estimates.
-\emph{Bioinformatics}, \emph{31}(17), 2794--2800.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Michonneau2016}{}}%
-Michonneau, F., Brown, J.
-W., \& Winter, D.
-J.
-(2016).
-{rotl: an R package to interact with the Open Tree of Life data}.
-\emph{{Methods in Ecology and Evolution}}, \emph{7}(12), 1476--1481.
-\url{https://doi.org/10.1111/2041-210X.12593}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Morlon2014}{}}%
-Morlon, H. (2014). {Phylogenetic approaches for studying diversification.} \emph{{Ecology Letters}}, \emph{17}(4), 508--525. \url{https://doi.org/10.1111/ele.12251}
-
-\leavevmode\vadjust pre{\hypertarget{ref-moyle2016tectonic}{}}%
-Moyle, R.
-G., Oliveros, C.
-H., Andersen, M.
-J., Hosner, P.
-A., Benz, B.
-W., Manthey, J.
-D., \ldots{} Faircloth, B.
-C.
-(2016).
-Tectonic collision and uplift of wallacea triggered the global songbird radiation.
-\emph{Nature Communications}, \emph{7}(1), 1--7.
-
-\leavevmode\vadjust pre{\hypertarget{ref-odeen2011evolution}{}}%
-Ödeen, A., Håstad, O., \& Alström, P. (2011). Evolution of ultraviolet vision in the largest avian radiation-the passerines. \emph{BMC Evolutionary Biology}, \emph{11}(1), 1--8.
-
-\leavevmode\vadjust pre{\hypertarget{ref-oliveros2019earth}{}}%
-Oliveros, C.
-H., Field, D.
-J., Ksepka, D.
-T., Barker, F.
-K., Aleixo, A., Andersen, M.
-J., \ldots{} others.
-(2019).
-Earth history and the passerine superradiation.
-\emph{Proceedings of the National Academy of Sciences}, \emph{116}(16), 7916--7925.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Ooms2018}{}}%
-Ooms, J., \& Chamberlain, S. (2018). \emph{Phylocomr: Interface to 'phylocom'}. Retrieved from \url{https://CRAN.R-project.org/package=phylocomr}
-
-\leavevmode\vadjust pre{\hypertarget{ref-opentreeAPIs}{}}%
-Open Tree Of Life, Redelings, B., Cranston, K.
-A., Allman, J., Holder, M.
-T., \& McTavish, E.
-J.
-(2016).
-{Open Tree of Life APIs v3.0}.
-\emph{{Open Tree of Life Project}}, (Online Resources).
-Retrieved from \url{\%7Bhttps://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\%7D}
-
-\leavevmode\vadjust pre{\hypertarget{ref-opentreeoflife2019synth}{}}%
-Open Tree Of Life, Redelings, B., Sánchez Reyes, L.
-L., Cranston, K.
-A., Allman, J., Holder, M.
-T., \& McTavish, E.
-J.
-(2019).
-Open tree of life synthetic tree v12.3.
-\emph{Zenodo}.
-Retrieved from \url{https://doi.org/10.5281/zenodo.3937742}
-
-\leavevmode\vadjust pre{\hypertarget{ref-packert2012horizontal}{}}%
-Päckert, M., Martens, J., Sun, Y.-H., Severinghaus, L.
-L., Nazarenko, A.
-A., Ting, J., \ldots{} Tietze, D.
-T.
-(2012).
-Horizontal and elevational phylogeographic patterns of himalayan and southeast asian forest passerines (aves: passeriformes).
-\emph{Journal of Biogeography}, \emph{39}(3), 556--573.
-
-\leavevmode\vadjust pre{\hypertarget{ref-paradis2004}{}}%
-Paradis, E., Claude, J., \& Strimmer, K. (2004). {APE: analyses of phylogenetics and evolution in R language}. \emph{Bioinformatics}, \emph{20}(2), 289--290.
-
-\leavevmode\vadjust pre{\hypertarget{ref-parchman2007coevolution}{}}%
-Parchman, T.
-L., Benkman, C.
-W., \& Mezquida, E.
-T.
-(2007).
-Coevolution between hispaniolan crossbills and pine: Does more time allow for greater phenotypic escalation at lower latitude?
-\emph{Evolution}, \emph{61}(9), 2142--2153.
-
-\leavevmode\vadjust pre{\hypertarget{ref-pennell2014geiger}{}}%
-Pennell, M.
-W., Eastman, J.
-M., Slater, G.
-J., Brown, J.
-W., Uyeda, J.
-C., FitzJohn, R.
-G., \ldots{} Harmon, L.
-J.
-(2014).
-Geiger v2.
-0: An expanded suite of methods for fitting macroevolutionary models to phylogenetic trees.
-\emph{Bioinformatics}, \emph{30}(15), 2216--2218.
-
-\leavevmode\vadjust pre{\hypertarget{ref-posadas2006historical}{}}%
-Posadas, P., Crisci, J. V., \& Katinas, L. (2006). Historical biogeography: A review of its basic concepts and critical issues. \emph{{Journal of Arid Environments}}, \emph{66}(3), 389--403.
-
-\leavevmode\vadjust pre{\hypertarget{ref-powell2014comprehensive}{}}%
-Powell, A.
-F., Barker, F.
-K., Lanyon, S.
-M., Burns, K.
-J., Klicka, J., \& Lovette, I.
-J.
-(2014).
-A comprehensive species-level molecular phylogeny of the new world blackbirds (icteridae).
-\emph{Molecular Phylogenetics and Evolution}, \emph{71}, 94--112.
-
-\leavevmode\vadjust pre{\hypertarget{ref-price2014niche}{}}%
-Price, T.
-D., Hooper, D.
-M., Buchanan, C.
-D., Johansson, U.
-S., Tietze, D.
-T., Alström, P., \ldots{} others.
-(2014).
-Niche filling slows the diversification of himalayan songbirds.
-\emph{Nature}, \emph{509}(7499), 222.
-
-\leavevmode\vadjust pre{\hypertarget{ref-pulgarin2013multilocus}{}}%
-Pulgarín-R, P.
-C., Smith, B.
-T., Bryson Jr, R.
-W., Spellman, G.
-M., \& Klicka, J.
-(2013).
-Multilocus phylogeny and biogeography of the new world pheucticus grosbeaks (aves: cardinalidae).
-\emph{Molecular Phylogenetics and Evolution}, \emph{69}(3), 1222--1227.
-
-\leavevmode\vadjust pre{\hypertarget{ref-RCoreTeam2018}{}}%
-R Core Team. (2018). \emph{{R: a language and environment for statistical computing}}. Vienna, Austria: R Foundation for Statistical Computing.
-
-\leavevmode\vadjust pre{\hypertarget{ref-rabosky2018inverse}{}}%
-Rabosky, D.
-L., Chang, J., Title, P.
-O., Cowman, P.
-F., Sallan, L., Friedman, M., \ldots{} others.
-(2018).
-An inverse latitudinal gradient in speciation rate for marine fishes.
-\emph{Nature}, \emph{559}(7714), 392.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ramshaw1972time}{}}%
-Ramshaw, J., Richardson, D., Meatyard, B., Brown, R., Richardson, M., Thompson, E., \& Boulter, D.
-(1972).
-The time of origin of the flowering plants determined by using amino acid sequence data of cytochrome c.
-\emph{New Phytologist}, \emph{71}(5), 773--779.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ratnasingham2007bold}{}}%
-Ratnasingham, S., \& Hebert, P. D. (2007). BOLD: The barcode of life data system (http://www. Barcodinglife. org). \emph{Molecular Ecology Notes}, \emph{7}(3), 355--364.
-
-\leavevmode\vadjust pre{\hypertarget{ref-rees2017automated}{}}%
-Rees, J. A., \& Cranston, K. (2017). Automated assembly of a reference taxonomy for phylogenetic data synthesis. \emph{Biodiversity Data Journal}, (5).
-
-\leavevmode\vadjust pre{\hypertarget{ref-Revell2012}{}}%
-Revell, L. J. (2012). Phytools: An r package for phylogenetic comparative biology (and other things). \emph{{Methods in Ecology and Evolution}}, \emph{3}, 217--223.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Ronquist2003}{}}%
-Ronquist, F., \& Huelsenbeck, J.
-P.
-(2003).
-{MrBayes 3: Bayesian phylogenetic inference under mixed models}.
-\emph{Bioinformatics}, \emph{19}(12), 1572--1574.
-\url{https://doi.org/10.1093/bioinformatics/btg180}
-
-\leavevmode\vadjust pre{\hypertarget{ref-datelifeplot}{}}%
-Sanchez-Reyes, L.
-L., \& O'Meara, B.
-(2022).
-{\texttt{datelifeplot}: Methods to plot chronograms and outputs of the \texttt{datelife} package}.
-\emph{R Package Release V0.2.2}.
-Retrieved from \url{https://zenodo.org/badge/latestdoi/381501451}
-
-\leavevmode\vadjust pre{\hypertarget{ref-datelifeR}{}}%
-Sanchez-Reyes, L.
-L., O'Meara, B., Eastman, J., Heath, T., Wright, A., Schliep, K., \ldots{} Alfaro, M.
-(2022).
-{\texttt{datelife}: Scientific Data on Time of Lineage Divergence for Your Taxa}.
-\emph{R Package Release V0.6.2}.
-Retrieved from \url{https://doi.org/10.5281/zenodo.593938}
-
-\leavevmode\vadjust pre{\hypertarget{ref-sanderson2002}{}}%
-Sanderson, Michael J.
-(2002).
-{Estimating Absolute Rates of Molecular Evolution and Divergence Times: A Penalized Likelihood Approach}.
-\emph{Molecular Biology and Evolution}, \emph{19}(1), 101--109.
-\url{https://doi.org/10.1093/oxfordjournals.molbev.a003974}
-
-\leavevmode\vadjust pre{\hypertarget{ref-sanderson2003r8s}{}}%
-Sanderson, Michael J. (2003). r8s: Inferring absolute rates of molecular evolution and divergence times in the absence of a molecular clock. \emph{Bioinformatics}, \emph{19}(2), 301--302.
-
-\leavevmode\vadjust pre{\hypertarget{ref-sanderson2001sources}{}}%
-Sanderson, Michael J., \& Doyle, J.
-A.
-(2001).
-Sources of error and confidence intervals in estimating the age of angiosperms from rbcL and 18S rDNA data.
-\emph{American Journal of Botany}, \emph{88}(8), 1499--1516.
-
-\leavevmode\vadjust pre{\hypertarget{ref-sauquet2013practical}{}}%
-Sauquet, H. (2013). A practical guide to molecular dating. \emph{{Comptes Rendus Palevol}}, \emph{12}(6), 355--367.
-
-\leavevmode\vadjust pre{\hypertarget{ref-sauquet2021age}{}}%
-Sauquet, H., Ramírez-Barahona, S., \& Magallón, S. (2021). \emph{The age of flowering plants is unknown}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-schenk2016sec}{}}%
-Schenk, J. J. (2016). {Consequences of secondary calibrations on divergence time estimates}. \emph{PLoS ONE}, \emph{11}(1). \url{https://doi.org/10.1371/journal.pone.0148228}
-
-\leavevmode\vadjust pre{\hypertarget{ref-schliep2011phangorn}{}}%
-Schliep, K. P. (2011). Phangorn: Phylogenetic analysis in r. \emph{Bioinformatics}, \emph{27}(4), 592--593.
-
-\leavevmode\vadjust pre{\hypertarget{ref-selvatti2015paleogene}{}}%
-Selvatti, A.
-P., Gonzaga, L.
-P., \& Moraes Russo, C.
-A.
-de.
-(2015).
-A paleogene origin for crown passerines and the diversification of the oscines in the new world.
-\emph{Molecular Phylogenetics and Evolution}, \emph{88}, 1--15.
-
-\leavevmode\vadjust pre{\hypertarget{ref-smith2018constructing}{}}%
-Smith, Stephen A., \& Brown, J. W. (2018). Constructing a broadly inclusive seed plant phylogeny. \emph{American Journal of Botany}, \emph{105}(3), 302--314.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Smith2012}{}}%
-Smith, Stephen A., \& O'Meara, B.
-C.
-(2012).
-{TreePL: Divergence time estimation using penalized likelihood for large phylogenies}.
-\emph{Bioinformatics}, \emph{28}(20), 2689--2690.
-\url{https://doi.org/10.1093/bioinformatics/bts492}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Stoltzfus2013}{}}%
-Stoltzfus, A., Lapp, H., Matasci, N., Deus, H., Sidlauskas, B., Zmasek, C.
-M., \ldots{} Jordan, G.
-(2013).
-{Phylotastic!
-Making tree-of-life knowledge accessible, reusable and convenient}.
-\emph{{BMC Bioinformatics}}, \emph{14}.
-\url{https://doi.org/10.1186/1471-2105-14-158}
-
-\leavevmode\vadjust pre{\hypertarget{ref-tietze2013complete}{}}%
-Tietze, D.
-T., Päckert, M., Martens, J., Lehmann, H., \& Sun, Y.-H.
-(2013).
-Complete phylogeny and historical biogeography of true rosefinches (aves: carpodacus).
-\emph{Zoological Journal of the Linnean Society}, \emph{169}(1), 215--234.
-
-\leavevmode\vadjust pre{\hypertarget{ref-title2016macrophylogenies}{}}%
-Title, P.
-O., \& Rabosky, D.
-L.
-(2016).
-{Do Macrophylogenies Yield Stable Macroevolutionary Inferences?
-An Example from Squamate Reptiles}.
-\emph{Systematic Biology}, syw102.
-\url{https://doi.org/10.1093/sysbio/syw102}
-
-\leavevmode\vadjust pre{\hypertarget{ref-treplin2008molecular}{}}%
-Treplin, S., Siegert, R., Bleidorn, C., Thompson, H.
-S., Fotso, R., \& Tiedemann, R.
-(2008).
-Molecular phylogeny of songbirds (aves: Passeriformes) and the relative utility of common nuclear marker loci.
-\emph{Cladistics}, \emph{24}(3), 328--349.
-
-\leavevmode\vadjust pre{\hypertarget{ref-vos2012nexml}{}}%
-Vos, R.
-A., Balhoff, J.
-P., Caravas, J.
-A., Holder, M.
-T., Lapp, H., Maddison, W.
-P., \ldots{} others.
-(2012).
-NeXML: Rich, extensible, and verifiable representation of comparative data and metadata.
-\emph{Systematic Biology}, \emph{61}(4), 675--689.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Webb2000}{}}%
-Webb, C. (2000). {Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees}. \emph{{The American Naturalist}}, \emph{156}(2), 145--155.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Webb2008}{}}%
-Webb, Campbell O., Ackerly, D.
-D., \& Kembel, S.
-W.
-(2008).
-{Phylocom: Software for the analysis of phylogenetic community structure and trait evolution}.
-\emph{Bioinformatics}, \emph{24}(18), 2098--2100.
-\url{https://doi.org/10.1093/bioinformatics/btn358}
-
-\leavevmode\vadjust pre{\hypertarget{ref-webb2005phylomatic}{}}%
-Webb, Campbell O., \& Donoghue, M. J. (2005). Phylomatic: Tree assembly for applied phylogenetics. \emph{{Molecular Ecology Notes}}, \emph{5}(1), 181--183.
-
-\leavevmode\vadjust pre{\hypertarget{ref-weir2008calibrating}{}}%
-Weir, J., \& Schluter, D. (2008). Calibrating the avian molecular clock. \emph{Molecular Ecology}, \emph{17}(10), 2321--2328.
-
-\end{CSLReferences}
-
-\endgroup
-
-% in this file I leave the figure captions outside\ caption{} because I want them
-% to be formatted in the same way as the general text (double spaced and linenumbered)
-\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure1/figure1-new.pdf}
-\caption{Stylized DateLife workflow.
-This shows the general worflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{http://www.datelife.org/}.
-Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
-}
-\label{fig:workflow}
-\end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:workflow}}
-% \end{center}
-
-\begin{figure}[!h]
-\includegraphics[width=1\linewidth]{../figures/fig_runtime_main.pdf}
-\caption{Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names.
-We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray).
-We also performed a search using the already processed query for comparison (light blue).}
-\label{fig:runtime_main}
-\end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:benchmark}}
-% \end{center}
-
-\begin{figure}[!h]
-\includegraphics[width=1\linewidth]{../figures/figure2/figure2-1.pdf}
-\caption{Age data results of a DateLife search of a small sample of 6 bird species within the Passeriformes.
-Input names were found across 9 chronograms within 6 independent studies (Barker et al.
-(\protect\hyperlink{ref-barker2012going}{2012}), Barker et al.
-(\protect\hyperlink{ref-barker2015new}{2015}), Burns et al.
-(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al.
-(\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al.
-(\protect\hyperlink{ref-Jetz2012}{2012}).) This revealed 28 age data points for the queried species names.}
-\label{fig:figure2-1}
-\end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:figure2-1}}
-% \end{center}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure2/figure2-2.pdf}
-\caption{Summarized age data is used as secondary calibrations to date a tree topology as a summary chronogram.}
-\label{fig:summaries}
-\end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:figure2-2}}
-% \end{center}
-
-\begin{figure}[!h]
-\includegraphics{../figures/fringillidae-topologies/fringillidae-topology.pdf}
-\caption{Tree topologies obtained from Open Tree of Life's (OpenTree) synthetic phylogenetic tree.
-A) Topology of 2333 tips and 1305 internal nodes, encompassing bird species within the family Fringillidae following the NCBI taxonomy (black), as well as all other bird species that share the same mrca node in OpenTree's synthetic tree (purple).
-B) Topology of 289 tips and 253 internal nodes, encompassing bird species within the Fringillidae only, resulting from pruning purple branches from topology A.
-Bird species within the Fringillidae do not form a monophyletic group.
-(Alström et al. \protect\hyperlink{ref-}{2014},
-Barker, Cibois, Schikler, Feinstein, \& Cracraft \protect\hyperlink{ref-barker2004phylogeny}{2004},
-Barker et al. \protect\hyperlink{ref-barker2013going}{2013},
-Barker \protect\hyperlink{ref-barker2014mitogenomic}{2014},
-Barker et al. \protect\hyperlink{ref-barker2015new}{2015},
-Beresford, Barker, Ryan, \& Crowe \protect\hyperlink{ref-beresford2005african}{2005},
-Bryson Jr et al. \protect\hyperlink{ref-bryson2014diversification}{2014},
-Burleigh, Kimball, \& Braun \protect\hyperlink{ref-burleigh2015building}{2015},
-Burns et al. \protect\hyperlink{ref-burns2014phylogenetics}{2014},
-Chaves, Hidalgo, \& Klicka \protect\hyperlink{ref-chaves2013biogeography}{2013},
-Claramunt \& Cracraft \protect\hyperlink{ref-claramunt2015new}{2015},
-Gibb et al. \protect\hyperlink{ref-gibb2015new}{2015},
-Hackett et al. \protect\hyperlink{ref-hackett2008phylogenomic}{2008},
-Jetz et al. \protect\hyperlink{ref-Jetz2012}{2012},
-Johansson, Fjeldså, \& Bowi \protect\hyperlink{ref-johansson2008phylogenetic}{200},
-Kimball et al. \protect\hyperlink{ref-kimball2019phylogenomic}{2019},
-Klicka et al. \protect\hyperlink{ref-klicka2014comprehensive}{2014},
-Lamichhaney et al. \protect\hyperlink{ref-lamichhaney2015evolution}{2015},
-Lerner, Meyer, James, Hofreiter, \& Fleischer \protect\hyperlink{ref-lerner2011multilocus}{2011},
-Lovette et al. \protect\hyperlink{ref-lovette2010comprehensive}{2010},
-Moyle et al. \protect\hyperlink{ref-moyle2016tectonic}{2016},
-Ödeen, Håstad, \& Alström \protect\hyperlink{ref-odeen2011evolution}{2011},
-Oliveros et al. \protect\hyperlink{ref-oliveros2019earth}{2019},
-Päckert et al. \protect\hyperlink{ref-packert2012horizontal}{2012},
-Parchman, Benkman, \& Mezquida \protect\hyperlink{ref-parchman2007coevolution}{2007},
-Powell et al. \protect\hyperlink{ref-powell2014comprehensive}{2014},
-Price et al. \protect\hyperlink{ref-price2014niche}{2014},
-Pulgarín-R, Smith, Bryson Jr, Spellman, \& Klicka \protect\hyperlink{ref-pulgarin2013multilocus}{2013},
-Selvatti, Gonzaga, \& Moraes Russo \protect\hyperlink{ref-selvatti2015paleogene}{2015},
-Tietze, Päckert, Martens, Lehmann, \& Sun \protect\hyperlink{ref-tietze2013complete}{2013},
-Treplin et al. \protect\hyperlink{ref-treplin2008molecular}{2008},
-Zuccon, Prŷs-Jones, Rasmussen, \& Ericson \protect\hyperlink{ref-zuccon2012phylogenetic}{2012})
-}
-\label{fig:fringillidae-topologies}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-fringillidae/median_and_calibration_ages_simple.png}
-\caption{Fringillidae median summary chronogram generated with DateLife. It has 256 tips and 233 nodes.}
-\label{fig:fringillidages}
-\end{figure}
-% \begin{center}
-% \textsc{Figure \ref{fig:fringillidages}}
-% \end{center}
-
-% \end{linenumbers}
-
-% in this file I leave the figure captions outside\ caption{} because I want them
-% to be formatted in the same way as the general text (double spaced and linenumbered)
-\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/fig-cross-validation-xy-plots.pdf}
-\caption{Results from cross validation analysis.}
-\label{fig:cvXY}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/fig-cross-validation-xy-plots-diffs.pdf}
-\caption{Results from cross validation analysis.}
-\label{fig:cvXYdiffs}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/cross_validation_2.png}
-\caption{Cross validation of second source chronogram.
-The chronogram shown in black corresponds to the dates published in the original study.
-The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
-\label{fig:cv2}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/cross_validation_3.png}
-\caption{Cross validation of third source chronogram.
-The chronogram shown in black corresponds to the dates published in the original study.
-The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
-\label{fig:cv3}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/cross_validation_4.png}
-\caption{Cross validation of fourth source chronogram.
-The chronogram shown in black corresponds to the dates published in the original study.
-The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
-\label{fig:cv4}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/cross_validation_6.png}
-\caption{Cross validation of sixth source chronogram.
-The chronogram shown in black corresponds to the dates published in the original study.
-The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
-\label{fig:cv6}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/cross_validation_7.png}
-\caption{Cross validation of seventh source chronogram.
-The chronogram shown in black corresponds to the dates published in the original study.
-The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
-\label{fig:cv7}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/cross_validation_8.png}
-\caption{Cross validation of eight source chronogram.
-The chronogram shown in black corresponds to the dates published in the original study.
-The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
-\label{fig:cv8}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/cross_validation_9.png}
-\caption{Cross validation of ninth source chronogram.
-The chronogram shown in black corresponds to the dates published in the original study.
-The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
-\label{fig:cv9}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/cross_validation_10.png}
-\caption{Cross validation of tenth source chronogram.
-The chronogram shown in black corresponds to the dates published in the original study.
-The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
-\label{fig:cv10}
-\end{figure}
-
-\end{document}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%    TABLE 1     %%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{table}
+  \caption{Ages of congruified nodes belonging to step C3 from Figure3}
+  \label{tbl:table1}
+  \includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example.pdf}
+\end{table}

### 2022-05-31 tables/table-fringillidae-small-example.tex
creating table 1 as .tex in small example vignette, and adding it to manuscript as graphic

\begin{tabular}{l|l|l|r|l}
\hline
Node Name & taxon A & taxon B & Node Age & Study chronogram\\
\hline
n1 & Emberiza citrinella & Platyspiza crassirostris & 9.6509881 & Jetz, W., G. H. Thomas, J. B. Joy, K. Hartmann, A. O. Mooers. 2012. The global diversity of birds in space and time. Nature 491 (7424): 444-448\\
\hline
n1 & Pheucticus tibialis & Schoeniclus elegans & 14.3336520 & Barker, F. Keith, Kevin J. Burns, John Klicka, Scott M. Lanyon, Irby J. Lovette. 2015. New insights into New World biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies. The Auk 132 (2): 333-348.\\
\hline
n1 & Rhodothraupis celaeno & Schoeniclus elegans & 16.2984859 & Barker, F. Keith, Kevin J. Burns, John Klicka, Scott M. Lanyon, Irby J. Lovette. 2015. New insights into New World biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies. The Auk 132 (2): 333-348.\\
\hline
n1 & Emberiza citrinella & Platyspiza crassirostris & 16.9499615 & Hooper, Daniel M., Trevor D. Price. 2017. Chromosomal inversion differences correlate with range overlap in passerine birds. Nature Ecology \& Evolution 1 (10): 1526-1534\\
\hline
n1 & Rhodothraupis celaeno & Schoeniclus elegans & 21.5140867 & Jetz, W., G. H. Thomas, J. B. Joy, K. Hartmann, A. O. Mooers. 2012. The global diversity of birds in space and time. Nature 491 (7424): 444-448\\
\hline
n1 & Emberiza leucocephalos & Platyspiza crassirostris & 22.0000000 & Hedges, S. Blair, Julie Marin, Michael Suleski, Madeline Paymer, Sudhir Kumar. 2015. Tree of life reveals clock-like speciation and diversification. Molecular Biology and Evolution 32 (4): 835-845\\
\hline
n1 & Emberiza citrinella & Platyspiza crassirostris & 22.3757277 & Hedges, S. Blair, Julie Marin, Michael Suleski, Madeline Paymer, Sudhir Kumar. 2015. Tree of life reveals clock-like speciation and diversification. Molecular Biology and Evolution 32 (4): 835-845\\
\hline
n2 & Platyspiza crassirostris & Rhodothraupis celaeno & 7.9691925 & Jetz, W., G. H. Thomas, J. B. Joy, K. Hartmann, A. O. Mooers. 2012. The global diversity of birds in space and time. Nature 491 (7424): 444-448\\
\hline
n2 & Platyspiza crassirostris & Rhodothraupis celaeno & 19.7085900 & Hedges, S. Blair, Julie Marin, Michael Suleski, Madeline Paymer, Sudhir Kumar. 2015. Tree of life reveals clock-like speciation and diversification. Molecular Biology and Evolution 32 (4): 835-845\\
\hline
n2 & Pheucticus tibialis & Platyspiza crassirostris & 44.2958603 & Burns, Kevin J., Allison J. Shultz, Pascal O. Title, Nicholas A. Mason, F. Keith Barker, John Klicka, Scott M. Lanyon, Irby J. Lovette. 2014. Phylogenetics and diversification of tanagers (Passeriformes: Thraupidae), the largest radiation of Neotropical songbirds. Molecular Phylogenetics and Evolution 75: 41-77.\\
\hline
n3 & Pheucticus tibialis & Rhodothraupis celaeno & 10.5304440 & Barker, F. Keith, Kevin J. Burns, John Klicka, Scott M. Lanyon, Irby J. Lovette. 2015. New insights into New World biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies. The Auk 132 (2): 333-348.\\
\hline
n3 & Pheucticus tibialis & Rhodothraupis celaeno & 10.5379092 & Barker, F. K., K. J. Burns, J. Klicka, S. M. Lanyon, I. J. Lovette. 2013. Going to extremes: contrasting rates of diversification in a recent radiation of New World passerine birds. Systematic Biology 62 (2): 298-320.\\
\hline
n3 & Pheucticus tibialis & Rhodothraupis celaeno & 18.1570685 & Hedges, S. Blair, Julie Marin, Michael Suleski, Madeline Paymer, Sudhir Kumar. 2015. Tree of life reveals clock-like speciation and diversification. Molecular Biology and Evolution 32 (4): 835-845\\
\hline
n4 & Emberiza citrinella & Schoeniclus elegans & 5.3238969 & Jetz, W., G. H. Thomas, J. B. Joy, K. Hartmann, A. O. Mooers. 2012. The global diversity of birds in space and time. Nature 491 (7424): 444-448\\
\hline
n4 & Emberiza leucocephalos & Schoeniclus elegans & 9.8622460 & Barker, F. Keith, Kevin J. Burns, John Klicka, Scott M. Lanyon, Irby J. Lovette. 2015. New insights into New World biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies. The Auk 132 (2): 333-348.\\
\hline
n4 & Emberiza leucocephalos & Schoeniclus elegans & 10.3391445 & Jetz, W., G. H. Thomas, J. B. Joy, K. Hartmann, A. O. Mooers. 2012. The global diversity of birds in space and time. Nature 491 (7424): 444-448\\
\hline
n4 & Emberiza citrinella & Schoeniclus elegans & 12.5133870 & Hooper, Daniel M., Trevor D. Price. 2017. Chromosomal inversion differences correlate with range overlap in passerine birds. Nature Ecology \& Evolution 1 (10): 1526-1534\\
\hline
n5 & Emberiza citrinella & Emberiza leucocephalos & 0.1407015 & Jetz, W., G. H. Thomas, J. B. Joy, K. Hartmann, A. O. Mooers. 2012. The global diversity of birds in space and time. Nature 491 (7424): 444-448\\
\hline
n5 & Emberiza citrinella & Emberiza leucocephalos & 0.1516230 & Hedges, S. Blair, Julie Marin, Michael Suleski, Madeline Paymer, Sudhir Kumar. 2015. Tree of life reveals clock-like speciation and diversification. Molecular Biology and Evolution 32 (4): 835-845\\
\hline
n5 & Emberiza citrinella & Emberiza leucocephalos & 0.2011990 & Barker, F. Keith, Kevin J. Burns, John Klicka, Scott M. Lanyon, Irby J. Lovette. 2015. New insights into New World biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies. The Auk 132 (2): 333-348.\\
\hline
n5 & Emberiza citrinella & Emberiza leucocephalos & 2.2898230 & Hooper, Daniel M., Trevor D. Price. 2017. Chromosomal inversion differences correlate with range overlap in passerine birds. Nature Ecology \& Evolution 1 (10): 1526-1534\\
\hline
\end{tabular}

### 2022-05-31 manuscript/manuscript.Rmd
fixing table 1 format and other edits

@@
 No studies matched all input names.
-Together, matching chronograms have 28 unique age data points.
-All nodes have age data.
+Together, source chronograms provide 28 unique age data points, covering all nodes on our chosen tree topology to date (Table \ref{tbl:table1}).
 
@@
 DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and mapped age data to nodes in this chosen topology.
-As expected, more inclusive nodes (e.g., node "n1") have more age data than less inclusive nodes (e.g., node "n5").
+As expected, more inclusive nodes (e.g., node "n1") have more variance in age data than less inclusive nodes (e.g., node "n5").
 The name processing step allowed including five data points for node "n4" (parent of _Schoeniclus elegans_; Fig. \ref{fig:figure2}A) that would not have had any data otherwise due to name mismatch.

### 2022-05-31 manuscript/manuscript.tex
creating table 1 as .tex in small example vignette, and adding it to manuscript as graphic
fixing table 1 format and other edits

@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4823}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4834}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and mapped age data to nodes in this chosen topology.
-As expected, more inclusive nodes (e.g., node ``n1'') have more age data than less inclusive nodes (e.g., node ``n5'').
+As expected, more inclusive nodes (e.g., node ``n1'') have more variance in age data than less inclusive nodes (e.g., node ``n5'').
 The name processing step allowed including five data points for node ``n4'' (parent of \emph{Schoeniclus elegans}; Fig.
@@
 \endgroup
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%    TABLE 1     %%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{table}
+  \caption{Ages of congruified nodes belonging to step C3 from Figure3}
+  \label{tbl:table1}
+  \includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example.pdf}
+\end{table}
 
@@
 A) Topology of 2,333 tips and 1,305 internal nodes, encompassing bird species within the family Fringillidae following the NCBI taxonomy (black), as well as all other bird species that share the same mrca node in OpenTree's synthetic tree (purple).
-B) Topology of 289 tips and 253 internal nodes, encompassing bird species within the Fringillidae only, resulting from pruning purple branches from topology A.
-Bird species within the Fringillidae do not form a monophyletic group
+B) Topology of 289 tips and 253 internal nodes, encompassing bird species within the Fringillidae only.
+This tree results from pruning species that are not assigned to the family Fringillidae (purple branches in topology A) but that form a monophyletic group with them.
+Bird species within the Fringillidae are paraphyletic
 (Alström et al. \protect\hyperlink{ref-}{2014},

### 2022-05-31 manuscript/table-fringillidae-small-example.tex
fixing table 1 format and other edits

@@
 
-\caption{\label{tab:table1}Table 1. Ages of congruified nodes.}
+\caption{\label{tab:table1}Ages of congruified nodes.}
 \centering
@@
 \hline
-1 & n1 & Emberiza citrinella & Platyspiza crassirostris & 9.6509881 & Jetz, W., G.
-H.
-Thomas, J.
-B.
-Joy, K.
-Hartmann, A.
-O.
-Mooers.
-2012.
-The global diversity of birds in space and time.
-Nature 491 (7424): 444-448\\
+1 & n1 & Emberiza citrinella & Platyspiza crassirostris & 9.6509881 & Jetz et al. 2012 - chronogram 1\\
 \hline
-2 & n1 & Pheucticus tibialis & Schoeniclus elegans & 14.3336520 & Barker, F.
-Keith, Kevin J.
-Burns, John Klicka, Scott M.
-Lanyon, Irby J.
-Lovette.
-2015.
-New insights into New World biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
-The Auk 132 (2): 333-348.\\
+2 & n1 & Pheucticus tibialis & Schoeniclus elegans & 14.3336520 & Barker et al. 2015 - chronogram 1\\
 \hline
-3 & n1 & Rhodothraupis celaeno & Schoeniclus elegans & 16.2984859 & Barker, F.
-Keith, Kevin J.
-Burns, John Klicka, Scott M.
-Lanyon, Irby J.
-Lovette.
-2015.
-New insights into New World biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
-The Auk 132 (2): 333-348.\\
+3 & n1 & Rhodothraupis celaeno & Schoeniclus elegans & 16.2984859 & Barker et al. 2015 - chronogram 2\\
 \hline
-4 & n1 & Emberiza citrinella & Platyspiza crassirostris & 16.9499615 & Hooper, Daniel M., Trevor D.
-Price.
-2017.
-Chromosomal inversion differences correlate with range overlap in passerine birds.
-Nature Ecology \& Evolution 1 (10): 1526-1534\\
+4 & n1 & Emberiza citrinella & Platyspiza crassirostris & 16.9499615 & Hooper et al. 2017\\
 \hline
-5 & n1 & Rhodothraupis celaeno & Schoeniclus elegans & 21.5140867 & Jetz, W., G.
-H.
-Thomas, J.
-B.
-Joy, K.
-Hartmann, A.
-O.
-Mooers.
-2012.
-The global diversity of birds in space and time.
-Nature 491 (7424): 444-448\\
+5 & n1 & Rhodothraupis celaeno & Schoeniclus elegans & 21.5140867 & Jetz et al. 2012 - chronogram 2\\
 \hline
-6 & n1 & Emberiza leucocephalos & Platyspiza crassirostris & 22.0000000 & Hedges, S.
-Blair, Julie Marin, Michael Suleski, Madeline Paymer, Sudhir Kumar.
-2015.
-Tree of life reveals clock-like speciation and diversification.
-Molecular Biology and Evolution 32 (4): 835-845\\
+6 & n1 & Emberiza leucocephalos & Platyspiza crassirostris & 22.0000000 & Hedges et al. 2015 - chronogram 2\\
 \hline
-7 & n1 & Emberiza citrinella & Platyspiza crassirostris & 22.3757277 & Hedges, S.
-Blair, Julie Marin, Michael Suleski, Madeline Paymer, Sudhir Kumar.
-2015.
-Tree of life reveals clock-like speciation and diversification.
-Molecular Biology and Evolution 32 (4): 835-845\\
+7 & n1 & Emberiza citrinella & Platyspiza crassirostris & 22.3757277 & Hedges et al. 2015 - chronogram 1\\
 \hline
-8 & n2 & Platyspiza crassirostris & Rhodothraupis celaeno & 7.9691925 & Jetz, W., G.
-H.
-Thomas, J.
-B.
-Joy, K.
-Hartmann, A.
-O.
-Mooers.
-2012.
-The global diversity of birds in space and time.
-Nature 491 (7424): 444-448\\
+8 & n2 & Platyspiza crassirostris & Rhodothraupis celaeno & 7.9691925 & Jetz et al. 2012 - chronogram 1\\
 \hline
-9 & n2 & Platyspiza crassirostris & Rhodothraupis celaeno & 19.7085900 & Hedges, S.
-Blair, Julie Marin, Michael Suleski, Madeline Paymer, Sudhir Kumar.
-2015.
-Tree of life reveals clock-like speciation and diversification.
-Molecular Biology and Evolution 32 (4): 835-845\\
+9 & n2 & Platyspiza crassirostris & Rhodothraupis celaeno & 19.7085830 & Jetz et al. 2012 - chronogram 2\\
 \hline
-10 & n2 & Pheucticus tibialis & Platyspiza crassirostris & 44.2958603 & Burns, Kevin J., Allison J.
-Shultz, Pascal O.
-Title, Nicholas A.
-Mason, F.
-Keith Barker, John Klicka, Scott M.
-Lanyon, Irby J.
-Lovette.
-2014.
-Phylogenetics and diversification of tanagers (Passeriformes: Thraupidae), the largest radiation of Neotropical songbirds.
-Molecular Phylogenetics and Evolution 75: 41-77.\\
+10 & n2 & Platyspiza crassirostris & Rhodothraupis celaeno & 19.7085900 & Hedges et al. 2015 - chronogram 2\\
 \hline
-11 & n3 & Pheucticus tibialis & Rhodothraupis celaeno & 10.5304440 & Barker, F.
-Keith, Kevin J.
-Burns, John Klicka, Scott M.
-Lanyon, Irby J.
-Lovette.
-2015.
-New insights into New World biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
-The Auk 132 (2): 333-348.\\
+11 & n2 & Platyspiza crassirostris & Rhodothraupis celaeno & 19.7128363 & Hedges et al. 2015 - chronogram 1\\
 \hline
-12 & n3 & Pheucticus tibialis & Rhodothraupis celaeno & 10.5379092 & Barker, F.
-K., K.
-J.
-Burns, J.
-Klicka, S.
-M.
-Lanyon, I.
-J.
-Lovette.
-2013.
-Going to extremes: contrasting rates of diversification in a recent radiation of New World passerine birds.
-Systematic Biology 62 (2): 298-320.\\
+12 & n2 & Pheucticus tibialis & Platyspiza crassirostris & 44.2958603 & Burns et al. 2014\\
 \hline
-13 & n3 & Pheucticus tibialis & Rhodothraupis celaeno & 18.1570685 & Hedges, S.
-Blair, Julie Marin, Michael Suleski, Madeline Paymer, Sudhir Kumar.
-2015.
-Tree of life reveals clock-like speciation and diversification.
-Molecular Biology and Evolution 32 (4): 835-845\\
+13 & n3 & Pheucticus tibialis & Rhodothraupis celaeno & 10.5304440 & Barker et al. 2015 - chronogram 1\\
 \hline
-14 & n4 & Emberiza citrinella & Schoeniclus elegans & 5.3238969 & Jetz, W., G.
-H.
-Thomas, J.
-B.
-Joy, K.
-Hartmann, A.
-O.
-Mooers.
-2012.
-The global diversity of birds in space and time.
-Nature 491 (7424): 444-448\\
+14 & n3 & Pheucticus tibialis & Rhodothraupis celaeno & 10.5379092 & Barker et al. 2013\\
 \hline
-15 & n4 & Emberiza leucocephalos & Schoeniclus elegans & 9.8622460 & Barker, F.
-Keith, Kevin J.
-Burns, John Klicka, Scott M.
-Lanyon, Irby J.
-Lovette.
-2015.
-New insights into New World biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
-The Auk 132 (2): 333-348.\\
+15 & n3 & Pheucticus tibialis & Rhodothraupis celaeno & 11.2095375 & Barker et al. 2015 - chronogram 2\\
 \hline
-16 & n4 & Emberiza leucocephalos & Schoeniclus elegans & 10.3391445 & Jetz, W., G.
-H.
-Thomas, J.
-B.
-Joy, K.
-Hartmann, A.
-O.
-Mooers.
-2012.
-The global diversity of birds in space and time.
-Nature 491 (7424): 444-448\\
+16 & n3 & Pheucticus tibialis & Rhodothraupis celaeno & 18.1570685 & Hedges et al. 2015 - chronogram 1\\
 \hline
-17 & n4 & Emberiza citrinella & Schoeniclus elegans & 12.5133870 & Hooper, Daniel M., Trevor D.
-Price.
-2017.
-Chromosomal inversion differences correlate with range overlap in passerine birds.
-Nature Ecology \& Evolution 1 (10): 1526-1534\\
+17 & n4 & Emberiza citrinella & Schoeniclus elegans & 5.3238969 & Jetz et al. 2012 - chronogram 1\\
 \hline
-18 & n5 & Emberiza citrinella & Emberiza leucocephalos & 0.1407015 & Jetz, W., G.
-H.
-Thomas, J.
-B.
-Joy, K.
-Hartmann, A.
-O.
-Mooers.
-2012.
-The global diversity of birds in space and time.
-Nature 491 (7424): 444-448\\
+18 & n4 & Emberiza leucocephalos & Schoeniclus elegans & 9.8622460 & Barker et al. 2015 - chronogram 1\\
 \hline
-19 & n5 & Emberiza citrinella & Emberiza leucocephalos & 0.1516230 & Hedges, S.
-Blair, Julie Marin, Michael Suleski, Madeline Paymer, Sudhir Kumar.
-2015.
-Tree of life reveals clock-like speciation and diversification.
-Molecular Biology and Evolution 32 (4): 835-845\\
+19 & n4 & Emberiza leucocephalos & Schoeniclus elegans & 10.3391445 & Jetz et al. 2012 - chronogram 2\\
 \hline
-20 & n5 & Emberiza citrinella & Emberiza leucocephalos & 0.2011990 & Barker, F.
-Keith, Kevin J.
-Burns, John Klicka, Scott M.
-Lanyon, Irby J.
-Lovette.
-2015.
-New insights into New World biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
-The Auk 132 (2): 333-348.\\
+20 & n4 & Emberiza leucocephalos & Schoeniclus elegans & 11.7317630 & Barker et al. 2015 - chronogram 2\\
 \hline
-21 & n5 & Emberiza citrinella & Emberiza leucocephalos & 2.2898230 & Hooper, Daniel M., Trevor D.
-Price.
-2017.
-Chromosomal inversion differences correlate with range overlap in passerine birds.
-Nature Ecology \& Evolution 1 (10): 1526-1534\\
+21 & n4 & Emberiza citrinella & Schoeniclus elegans & 12.5133870 & Hooper et al. 2017\\
+\hline
+22 & n5 & Emberiza citrinella & Emberiza leucocephalos & 0.1407015 & Jetz et al. 2012 - chronogram 1\\
+\hline
+23 & n5 & Emberiza citrinella & Emberiza leucocephalos & 0.1516230 & Hedges et al. 2015 - chronogram 2\\
+\hline
+24 & n5 & Emberiza citrinella & Emberiza leucocephalos & 0.2011990 & Barker et al. 2015 - chronogram 1\\
+\hline
+25 & n5 & Emberiza citrinella & Emberiza leucocephalos & 0.2409300 & Barker et al. 2015 - chronogram 2\\
+\hline
+26 & n5 & Emberiza citrinella & Emberiza leucocephalos & 0.2732460 & Jetz et al. 2012 - chronogram 2\\
+\hline
+27 & n5 & Emberiza citrinella & Emberiza leucocephalos & 0.5760260 & Hedges et al. 2015 - chronogram 1\\
+\hline
+28 & n5 & Emberiza citrinella & Emberiza leucocephalos & 2.2898230 & Hooper et al. 2017\\
 \hline

### 2022-05-31 manuscript/manuscript.Rmd
adding table 2 to manuscript

@@
 _**Summarizing search results.-**_
-DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and mapped age data to nodes in this chosen topology.
+DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and congruified and mapped age data to nodes in this chosen topology (Table \ref{tbl:table1}).
+The name processing step allowed including five data points for node "n4" (parent of _Schoeniclus elegans_; Fig. \ref{fig:figure2}A) that would not have had any data otherwise due to name mismatch.
+Age summary statistics per node were calculated (Table \ref{tbl:table2}) and used as calibrations to date the tree topology using the BLADJ algorithm.
 As expected, more inclusive nodes (e.g., node "n1") have more variance in age data than less inclusive nodes (e.g., node "n5").
-The name processing step allowed including five data points for node "n4" (parent of _Schoeniclus elegans_; Fig. \ref{fig:figure2}A) that would not have had any data otherwise due to name mismatch.
-Age summary statistics per node were calculated and used as calibrations to date the tree topology using the BLADJ algorithm.
-Age data for node "n2" was excluded as final calibration because it is older than age data of the more inclusive node "n1" (Fig. \ref{fig:figure2}C4).
+Summary age data for node "n2" were excluded as final calibration because they are older than age data of the more inclusive node, "n1" (Fig. \ref{fig:figure2}C4).
 

### 2022-05-31 manuscript/manuscript.tex
adding table 2 to manuscript

@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4834}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4844}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 \emph{\textbf{Summarizing search results.-}}
-DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and mapped age data to nodes in this chosen topology.
-As expected, more inclusive nodes (e.g., node ``n1'') have more variance in age data than less inclusive nodes (e.g., node ``n5'').
+DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and congruified and mapped age data to nodes in this chosen topology (Table \ref{tbl:table1}).
 The name processing step allowed including five data points for node ``n4'' (parent of \emph{Schoeniclus elegans}; Fig.
 \ref{fig:figure2}A) that would not have had any data otherwise due to name mismatch.
-Age summary statistics per node were calculated and used as calibrations to date the tree topology using the BLADJ algorithm.
+Age summary statistics per node were calculated (Table \ref{tbl:table2}) and used as calibrations to date the tree topology using the BLADJ algorithm.
+As expected, more inclusive nodes (e.g., node ``n1'') have more variance in age data than less inclusive nodes (e.g., node ``n5'').
 Age data for node ``n2'' was excluded as final calibration because it is older than age data of the more inclusive node ``n1'' (Fig. \ref{fig:figure2}C4).
@@
 \begin{table}
-  \caption{Ages of congruified nodes belonging to step C3 from Figure3}
+  \caption{Ages of congruified nodes belonging to step C2 from Figure3}
   \label{tbl:table1}
   \includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example.pdf}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%    TABLE 2     %%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{table}
+  \caption{Summary of congruified nodes ages, corresponding to step C3 from Figure3}
+  \label{tbl:table2}
+  \includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example-summary.png}
 \end{table}

### 2022-05-31 manuscript/tables.tex
adding table 2 to manuscript

@@
 \begin{table}
-  \caption{Ages of congruified nodes belonging to step C3 from Figure3}
+  \caption{Ages of congruified nodes belonging to step C2 from Figure3}
   \label{tbl:table1}
@@
 \end{table}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%    TABLE 2     %%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\begin{table}
+  \caption{Summary of congruified nodes ages, corresponding to step C3 from Figure3}
+  \label{tbl:table2}
+  \includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example-summary.png}
+\end{table}

### 2022-06-01 R/make_pretty_tables.R
function to format calibration tables to print as pdf

@@ -1,0 +1,66 @@
+
+#' provide new colnames
+#'
+# some_table <- calib_summ$in_phy
+# # Change node numbers by node names
+# colnames(some_table)
+
+make_pretty_table <- function(some_table, duplicated_criteria) {
+  # assign variables:
+  cols2keep_order <- c("mrca_node_name", "taxonA", "taxonB", "MaxAge", "reference")
+  if (missing(duplicated_criteria)) {
+    duplicated_criteria <- c("taxonA", "taxonB", "reference","MaxAge")
+  } else {
+    duplicated_criteria <- duplicated_criteria
+  }
+  colnames_new <- c("Node Name", "taxon A", "taxon B", "Node Age", "Study reference")
+  # Eliminate duplicated rows:
+  ## Find the duplicated rows within studies:
+  ## Note: take into account age too, taxon pairs might come from different studies.
+  taxon_pairs_ref <- c()
+  for (i in seq(nrow(some_table))) {
+    print(i)
+    tax_pair_ref <- as.character(some_table[i, duplicated_criteria])
+    tax_pair_ref <- tax_pair_ref[order(tax_pair_ref)]
+    tax_pair_ref <- paste(tax_pair_ref, collapse = "")
+    taxon_pairs_ref <- c(taxon_pairs_ref, tax_pair_ref)
+  }
+  if (sum(duplicated(taxon_pairs_ref)) == 0) {
+    message("There are no duplicated data.")
+    pretty_calibs <- some_table
+  }
+  ## Eliminate duplicated rows if any:
+  if (any(duplicated(taxon_pairs_ref))) {
+    indices <- which(duplicated(taxon_pairs_ref))
+    message("Eliminating duplicated data from rows: ", paste(indices, collapse = " | "))
+    pretty_calibs <- some_table[!duplicated(taxon_pairs_ref),]
+  }
+
+  # Reorder ROWS by node number, reference, age and taxon names
+  row_order <- order(pretty_calibs$mrca_node_number, pretty_calibs$MaxAge, pretty_calibs$reference, pretty_calibs$taxonA, pretty_calibs$taxonB)
+  pretty_calibs <- pretty_calibs[row_order,]
+
+  # Assign a node name
+  # number_of_tips <- min(pretty_calibs$mrca_node_number)-1
+  # some_table$mrca_node_name <- paste0("n", (pretty_calibs$mrca_node_number - number_of_tips))
+
[…]

### 2022-06-02 R/make_pretty_tables.R
updating tables of all fringillidae example for supplements

@@ -5,29 +5,23 @@
 # # Change node numbers by node names
 # colnames(some_table)
 
-make_pretty_table <- function(some_table, duplicated_criteria) {
-  # assign variables:
-  cols2keep_order <- c("mrca_node_name", "taxonA", "taxonB", "MaxAge", "reference")
+make_pretty_table <- function(some_table,
+                              duplicated_criteria,
+                              cols2keep_order,
+                              order_rows = FALSE,
+                              colnames_new){
+  # Eliminate duplicated rows:
   if (missing(duplicated_criteria)) {
     duplicated_criteria <- c("taxonA", "taxonB", "reference","MaxAge")
-  } else {
-    duplicated_criteria <- duplicated_criteria
   }
-  colnames_new <- c("Node Name", "taxon A", "taxon B", "Node Age", "Study reference")
-  # Eliminate duplicated rows:
   ## Find the duplicated rows within studies:
   ## Note: take into account age too, taxon pairs might come from different studies.
   taxon_pairs_ref <- c()
   for (i in seq(nrow(some_table))) {
-    print(i)
     tax_pair_ref <- as.character(some_table[i, duplicated_criteria])
     tax_pair_ref <- tax_pair_ref[order(tax_pair_ref)]
     tax_pair_ref <- paste(tax_pair_ref, collapse = "")
     taxon_pairs_ref <- c(taxon_pairs_ref, tax_pair_ref)
-  }
-  if (sum(duplicated(taxon_pairs_ref)) == 0) {
-    message("There are no duplicated data.")
-    pretty_calibs <- some_table
   }
   ## Eliminate duplicated rows if any:
   if (any(duplicated(taxon_pairs_ref))) {
@@ -35,19 +29,46 @@ make_pretty_table <- function(some_table, duplicated_criteria) {
     message("Eliminating duplicated data from rows: ", paste(indices, collapse = " | "))
     pretty_calibs <- some_table[!duplicated(taxon_pairs_ref),]
   }
-
+  # If no duplicated row, assign some_table variable to object pretty_table:
+  if (sum(duplicated(taxon_pairs_ref)) == 0) {
+    message("There are no duplicated calibrations.")
+    pretty_calibs <- some_table
+  }
+  if (order_rows) {
+    row_order <- order(pretty_calibs$mrca_node_number,
+                       pretty_calibs$MaxAge,
+                       pretty_calibs$reference,
[…]

### 2022-06-03 R/make_pretty_tables.R
updating figures fringillidae all vignette

@@ -40,18 +40,6 @@ make_pretty_table <- function(some_table,
                        pretty_calibs$reference,
                        pretty_calibs$taxonA,
                        pretty_calibs$taxonB)
-    pretty_calibs[row_order,]
-  }
-  # Reorder ROWS by node number, reference, age and taxon names
-  if (missing(cols2order_rows)) {
-    cols2order_rows <- c("mrca_node_name",
-                         "MaxAge",
-                         "study",
-                         "taxonA",
-                         "taxonB")
-  }
-  for (column in cols2order_rows) {
-    row_order <- order(pretty_calibs[,column])
     pretty_calibs <- pretty_calibs[row_order,]
   }
 

### 2022-06-03 R/plot_node_ages_custom.R
adding figures for fringillidae all congruified chronogram

@@ -1,0 +1,28 @@
+datelifeplot::plot_node_ages(my_chronogram,
+               title = "Median Summary Chronogram",
+               plot_type = "phyloch",
+               time_depth = 45,
+               calibration_summary = calib_summ,
+               pch = 19,
+               color_pch = color_studies,
+               cex_pch = 1,
+               lwd_bars = 10,
+               cex_legend = 0.7,
+               x_legend = 0 ,
+               y_legend = 270,
+               mai1 = 1,
+               mai2 = 0,
+               mai3 = 0.5,
+               mai4 = 3.73,
+               omi1 = 2,
+               omi2 = 1,
+               omi3 = 0.5,
+               omi4 = 0,
+               edge.width = 2,
+               cex_tiplabels = 0.5,
+               cex_axislabel = 1,
+               cex_axis = 1,
+               cex_title = 2,
+               pos_title = -1,
+               pos_axis = 2
+               )

### 2022-06-04 R/plot_node_ages2_custom.R
fixing legends and other issues with data points

@@ -1,0 +1,39 @@
+datelifeplot::plot_node_ages2(my_chronogram,
+               title = c("Median Summary Chronogram"),
+               pos_title = -6,
+               cex_title = 2,
+               plot_type = "phyloch",
+               time_depth = 45,
+               pos_axis = 2,
+               cex_axis = 1,
+               edge.width = 3,
+               cex_tiplabels = 0.5,
+               matched_ages  = all_ages,
+               pch = point_type,
+               pch_color = point_colors,
+               pch_cex = 2,
+               bars_lwd = 7,
+               bars_color = "#80808060",
+#               legend_x = -5,
+#               legend_y = 200,
+               legend_text = leg_text,
+               legend_pch = leg_pch,
+               legend_color = leg_color,
+#               mai1 = 1,
+#               mai2 = 0.8,
+#               mai4 = 3.6,
+               axis_label = "",
+               legend_title = leg_title,
+               legend_box = leg_box,
+               mai1 = 0,
+               mai2 = 0,
+               mai3 = 0.5,
+               mai4 = 1,
+               omi1 = 0,
+               omi2 = 0,
+               omi3 = 0,
+               legend_cex = c(1.2, 1.2),
+               legend_x = c(1, 1),
+               legend_y = c(140, 80),
+
+)

### 2022-06-05 R/plot_node_ages2_custom.R
plotting unused calibrations in gray

@@ -9,7 +9,7 @@ datelifeplot::plot_node_ages2(my_chronogram,
                edge.width = 3,
                cex_tiplabels = 0.5,
                matched_ages  = all_ages,
-               pch = point_type,
+               pch_type = point_type,
                pch_color = point_colors,
                pch_cex = 2,
                bars_lwd = 7,

### 2022-06-06 manuscript/manuscript.tex
formatting legends and adding png figure for fringillidae example

@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4844}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4845}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 As expected, more inclusive nodes (e.g., node ``n1'') have more variance in age data than less inclusive nodes (e.g., node ``n5'').
-Age data for node ``n2'' was excluded as final calibration because it is older than age data of the more inclusive node ``n1'' (Fig. \ref{fig:figure2}C4).
+Summary age data for node ``n2'' were excluded as final calibration because they are older than age data of the more inclusive node, ``n1'' (Fig. \ref{fig:figure2}C4).
 
@@
 \begin{figure}[!h]
-\includegraphics{../figures/figure-fringillidae/median_and_calibration_ages_simple.png}
+\includegraphics{../figures/figure-fringillidae/median_and_calibration_ages-congruified.png}
 \caption{Fringillidae median summary chronogram generated with DateLife. It has 256 tips and 233 nodes.}

### 2022-06-06 R/plot_node_ages2_custom.R

@@ -1,5 +1,5 @@
 datelifeplot::plot_node_ages2(my_chronogram,
-               title = c("Median Summary Chronogram"),
+               title = c("Fringillidae Median Summary Chronogram"),
                pos_title = -6,
                cex_title = 2,
                plot_type = "phyloch",
@@ -11,7 +11,7 @@ datelifeplot::plot_node_ages2(my_chronogram,
                matched_ages  = all_ages,
                pch_type = point_type,
                pch_color = point_colors,
-               pch_cex = 2,
+               pch_cex = point_cex,
                bars_lwd = 7,
                bars_color = "#80808060",
 #               legend_x = -5,
@@ -32,7 +32,8 @@ datelifeplot::plot_node_ages2(my_chronogram,
                omi1 = 0,
                omi2 = 0,
                omi3 = 0,
-               legend_cex = c(1.2, 1.2),
+               legend_cex = leg_cex,
+               legend_pt_cex = leg_pt_cex,
                legend_x = c(1, 1),
                legend_y = c(140, 80),
 

### 2022-06-06 manuscript/figures-alt.tex
formatting legends and adding png figure for fringillidae example
addressing remaining comments

@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 3    %%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%    FIGURE TOPOLOGIES    %%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 4    %%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%    FIGURE FRINGILLIDAE    %%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{figure}[!h]
-\includegraphics{../figures/figure-fringillidae/median_and_calibration_ages_simple.png}
-\caption{Fringillidae median summary chronogram generated with DateLife. It has 256 tips and 233 nodes.}
+\includegraphics{../figures/figure-fringillidae/median_and_calibration_ages-congruified.png}
+\caption{Fringillidae median summary chronogram generated with DateLife. It has 256 tips and 233 nodes, from which 212 have age data from at least one published chronogram.}
 \label{fig:fringillidages}
 \end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%    FIGURE CROSS VALIDATION    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[!h]
+\includegraphics{../figures/figure-cross-validation/fig-cross-validation-xy-plots.pdf}
+\caption{Results from cross validation analysis. Each plot compares the original age estimate (x axis) with the age obtained with a DateLife analysis (y axis), per node.}
+\label{fig:cvXY}
+\end{figure}

### 2022-06-06 manuscript/figures-supplementary.tex
cross validation XY figure
addressing remaining comments

@@
 % to be formatted in the same way as the general text (double spaced and linenumbered)
-\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}
+\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Supplementary Figure}}
 
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/fig-cross-validation-xy-plots.pdf}
-\caption{Results from cross validation analysis.}
-\label{fig:cvXY}
-\end{figure}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%    FIGURE CROSS VALIDATION    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \includegraphics{../figures/figure-cross-validation/fig-cross-validation-xy-plots-diffs.pdf}
-\caption{Results from cross validation analysis.}
+\caption{Results from cross validation analysis. Each plot shows the difference between the original age estimated and those obtained with a DateLife analysis, per node.}
 \label{fig:cvXYdiffs}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%   CROSS VALIDATION CHRONOGRAMS   %%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
 \begin{figure}[!h]
 \includegraphics{../figures/figure-cross-validation/cross_validation_2.png}
-\caption{Cross validation of second source chronogram.
+\caption{Cross validation of Barker et al.
+(\protect\hyperlink{ref-barker2015new}{2015}) chronogram 1.
 The chronogram shown in black corresponds to the dates published in the original study.
@@
 \includegraphics{../figures/figure-cross-validation/cross_validation_3.png}
-\caption{Cross validation of third source chronogram.
+\caption{Cross validation of Barker et al.
+(\protect\hyperlink{ref-barker2015new}{2015}) chronogram 1.
 The chronogram shown in black corresponds to the dates published in the original study.
@@
 \includegraphics{../figures/figure-cross-validation/cross_validation_4.png}
-\caption{Cross validation of fourth source chronogram.
+\caption{Cross validation of Burns et al.
+(\protect\hyperlink{ref-burns2014phylogenetics}{2014}) chronogram.
 The chronogram shown in black corresponds to the dates published in the original study.
@@
 \includegraphics{../figures/figure-cross-validation/cross_validation_6.png}
-\caption{Cross validation of sixth source chronogram.
+\caption{Cross validation of Gibb et al.
+(\protect\hyperlink{ref-gibb2015new}{2015}), chronogram.
 The chronogram shown in black corresponds to the dates published in the original study.
@@
 \includegraphics{../figures/figure-cross-validation/cross_validation_7.png}
-\caption{Cross validation of seventh source chronogram.
+\caption{Cross validation of Hedges et al.
+(\protect\hyperlink{ref-Hedges2015}{2015}) chronogram 1.
 The chronogram shown in black corresponds to the dates published in the original study.
@@
 \includegraphics{../figures/figure-cross-validation/cross_validation_8.png}
-\caption{Cross validation of eight source chronogram.
+\caption{Cross validation of Hedges et al.
+(\protect\hyperlink{ref-Hedges2015}{2015}) chronogram 2.
 The chronogram shown in black corresponds to the dates published in the original study.
@@
 \includegraphics{../figures/figure-cross-validation/cross_validation_9.png}
-\caption{Cross validation of ninth source chronogram.
+\caption{Cross validation of Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}) chronogram.
 The chronogram shown in black corresponds to the dates published in the original study.
@@
 \includegraphics{../figures/figure-cross-validation/cross_validation_10.png}
-\caption{Cross validation of tenth source chronogram.
+\caption{Cross validation of Jetz et al.
+(\protect\hyperlink{ref-Jetz2012}{2012}) chronogram 1.
 The chronogram shown in black corresponds to the dates published in the original study.

### 2022-06-06 manuscript/manuscript.Rmd
addressing remaining comments

@@
       after_body  : ["tables.tex",
-                     "figures-alt.tex", 
-                     "figures-supplementary.tex"]
+                     "figures-alt.tex"]
 
@@
 
-Processing input names found that *Emberiza elegans* is synonym for *Schoeniclus elegans* in the default reference taxonomy [Open Tree of Life Taxonomy v3.3, June 1, 2021].
-For a detailed discussion on the state of the synonym refer to Avibase [@lepage2004avibase; @lepage2014avibase; @avibase-emberiza].
+Processing input names found that *Emberiza elegans* is synonym for *Schoeniclus elegans* in the default reference taxonomy (Open Tree of Life Taxonomy v3.3, June 1, 2021).
+For a detailed discussion on the state of the synonym, refer to Avibase [@lepage2004avibase; @lepage2014avibase; @avibase-emberiza].
 Discovering this synonym allowed assigning five age data points for the parent node of *Emberiza elegans*, shown as *Schoeniclus elegans* in figure \ref{fig:figure2}A, which would not have had any data otherwise.
@@
 The DateLife search identified `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
-Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices.
+Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices, revealing 1206 different age data points available for species within the Fringillidae (supplementray table 1).
 
 _**Summarizing search results.-**_
-The final step is to summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM. 
+The final step is to congruify and summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM. 
 As explained in the "Description" section, a tree topology to summarize age data upon is required.
-By default, DateLife uses the portion of OpenTree synthetic phylogeny that contains the species provided as search query as summarizing topology. 
+By default, DateLife uses as summarizing topology the portion of OpenTree synthetic phylogeny that contains the species provided as search query. 
 According to this phylogeny, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).
-Age data from source chronograms is summarised into a single patristic distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
+Age data from source chronograms was congruified, decreasing to 818 different age data points available (supplementray table 2) and then age summary statistics per congruified node are obtained and used as fixed secondary calibrations over a consensus tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
 \ref{fig:fringillidages}).
@@
 @schenk2016sec showed that, in simulations, divergence times inferred using secondary calibrations are significantly younger than those inferred with primary calibrations, when obtained with Bayesian inference methods, and when priors are implemented in similar ways in both analyses.
-Accordingly, the scientific community seems to have more confidence in chronograms obtained from a single analysis, using fossil data as primary sources of calibrations [@schenk2016sec], and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history [@sauquet2013practical], as for example done by [@antonelli2017supersmart].
+Accordingly, the scientific community seems to have more confidence in chronograms obtained from a single analysis, using fossil data as primary sources of calibrations [@schenk2016sec], and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data reflect a coherent evolutionary history [@sauquet2013practical], as for example done by @antonelli2017supersmart.
 <!-- Will this be implemented in future `datelife` versions?
@@
 <!--BCO: Polite way to say 'no new additions in four years?' LLSR: lol, yes-->
-
-\textcolor{red}{LLSR: What does our cross validation analysis show in relation to diffs between primary and secondary calibrations?}
+<!--\textcolor{red}{LLSR: What does our cross validation analysis show in relation to  the effect of using secondary calibrations for a dating analysis that does not make any prior assumptions on the nature of the calibrations themselves?
+the age difference between primary and secondary calibrations?}-->
 
@@
 Certainly, further studies are required to fully understand the effect of secondary calibrations on outputs from different tree dating methods, and on downstream analyses.
-It is possible that secondary calibrations can be safely used with dating methods that do not require setting priors, such as penalized likelihood [@sanderson2003r8s], methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ [@Webb2008; @webb2005phylomatic], or methods that summarize age data unto a tree topology.
+It is possible that secondary calibrations can be safely used with dating methods that do not require setting priors, such as penalized likelihood [@sanderson2003r8s], with methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ [@Webb2008; @webb2005phylomatic], or methods that summarize age data unto a tree topology.
+
+Our cross validation analysis might provide some insight in this regard.
+When ages are estimated with secondary calibrations, nodes closer to the root do tend to be slightly younger than ages estimated with primary calibrations.
+However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures 1-9), with a notable exception of cross validation of @burns2014phylogenetics chronogram, which results in much younger node ages when estimated using secondary calibrations (Supplementary figs.
+1 and 4).
 
@@
 These differences in approach can deepen the already substantial variation in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
-For example, the chronograms from @burns2014phylogenetics tend to have much older date estimates for the same nodes than other inferences (Figure 5).
-This study inferred dates using molecular substitution rate estimates across birds [@weir2008calibrating]
-\textcolor{red}{EJM: MORE ON the FINCH examples and the cross validation here}
+For example, the chronograms from @burns2014phylogenetics were inferred using molecular substitution rate estimates across birds [@weir2008calibrating], and have much older date estimates for the same nodes than chronograms that were inferred using fossils as primary calibrations (figs.
+fig:fringillidages and fig:cvXY; supplementary figs.
+1 and 4).
+<!--\textcolor{red}{EJM: MORE ON the FINCH examples and the cross validation here}-->
 
@@
 A summarizing exercise may also amplify the effect of uncertainty and errors in source data, and blur parts of the evolutionary history of a lineage that might only be reflected in source chronograms and lost on the summary chronogram [@sauquet2021age].
-\textcolor{red}{EJM: ??? it's not clear to me what this last part means! What might you be losing?
+<!--\textcolor{red}{EJM: ??? it's not clear to me what this last part means! What might you be losing?
 LLSR: I changed loosing to blurring. But I will elaborate more on it}
 Choosing source chronograms that we are going to keep and the ones that we are going to discard might be key.
-\textcolor{red}{LLSR: Example}
+\textcolor{red}{LLSR: Example}-->
 
-<!-- Comparison of available age data for a wide range of organisms shown here suggest that discrpenacy in age stimates across studies is a widespread phenomenon that requires further attention. -->
+<!-- Comparison of available age data for a wide range of organisms shown here suggest that discrpenacy in age estimates across studies is a widespread phenomenon that requires further attention. -->
 

### 2022-06-06 manuscript/manuscript.tex
addressing remaining comments

@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4845}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4890}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 
-Processing input names found that \emph{Emberiza elegans} is synonym for \emph{Schoeniclus elegans} in the default reference taxonomy {[}Open Tree of Life Taxonomy v3.3, June 1, 2021{]}.
-For a detailed discussion on the state of the synonym refer to Avibase (Avibase, 2022; Lepage, 2004; Lepage, Vaidya, \& Guralnick, 2014).
+Processing input names found that \emph{Emberiza elegans} is synonym for \emph{Schoeniclus elegans} in the default reference taxonomy (Open Tree of Life Taxonomy v3.3, June 1, 2021).
+For a detailed discussion on the state of the synonym, refer to Avibase (Avibase, 2022; Lepage, 2004; Lepage, Vaidya, \& Guralnick, 2014).
 Discovering this synonym allowed assigning five age data points for the parent node of \emph{Emberiza elegans}, shown as \emph{Schoeniclus elegans} in figure \ref{fig:figure2}A, which would not have had any data otherwise.
@@
 The DateLife search identified 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Price et al., 2014).
-Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices.
+Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices, revealing 1206 different age data points available for species within the Fringillidae (supplementray table 1).
 
 \emph{\textbf{Summarizing search results.-}}
-The final step is to summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM.
+The final step is to congruify and summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM.
 As explained in the ``Description'' section, a tree topology to summarize age data upon is required.
-By default, DateLife uses the portion of OpenTree synthetic phylogeny that contains the species provided as search query as summarizing topology.
+By default, DateLife uses as summarizing topology the portion of OpenTree synthetic phylogeny that contains the species provided as search query.
 According to this phylogeny, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).
-Age data from source chronograms is summarised into a single patristic distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
+Age data from source chronograms was congruified, decreasing to 818 different age data points available (supplementray table 2) and then age summary statistics per congruified node are obtained and used as fixed secondary calibrations over a consensus tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
 \ref{fig:fringillidages}).
@@
 Schenk (2016) showed that, in simulations, divergence times inferred using secondary calibrations are significantly younger than those inferred with primary calibrations, when obtained with Bayesian inference methods, and when priors are implemented in similar ways in both analyses.
-Accordingly, the scientific community seems to have more confidence in chronograms obtained from a single analysis, using fossil data as primary sources of calibrations (Schenk, 2016), and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history (Sauquet, 2013), as for example done by (Antonelli et al., 2017).
+Accordingly, the scientific community seems to have more confidence in chronograms obtained from a single analysis, using fossil data as primary sources of calibrations (Schenk, 2016), and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data reflect a coherent evolutionary history (Sauquet, 2013), as for example done by Antonelli et al.
+(2017).
 There have been attempts to create fossil calibration databases (Ksepka et al., 2015), though these still have room to grow.
-
-\textcolor{red}{LLSR: What does our cross validation analysis show in relation to diffs between primary and secondary calibrations?}
 
@@
 It is possible that secondary calibrations can be safely used with dating methods that do not require setting priors, such as penalized likelihood (Michael J.
-Sanderson, 2003), methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ (Campbell O.
+Sanderson, 2003), with methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ (Campbell O.
 Webb et al., 2008; Campbell O.
 Webb \& Donoghue, 2005), or methods that summarize age data unto a tree topology.
+
+Our cross validation analysis might provide some insight in this regard.
+When ages are estimated with secondary calibrations, nodes closer to the root do tend to be slightly younger than ages estimated with primary calibrations.
+However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures 1-9), with a notable exception of cross validation of Burns et al.
+(2014) chronogram, which results in much younger node ages when estimated using secondary calibrations (Supplementary figs.
+1 and 4).
 
@@
 For example, the chronograms from Burns et al.
-(2014) tend to have much older date estimates for the same nodes than other inferences (Figure 5).
-This study inferred dates using molecular substitution rate estimates across birds (Weir \& Schluter, 2008)
-\textcolor{red}{EJM: MORE ON the FINCH examples and the cross validation here}
+(2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older date estimates for the same nodes than chronograms that were inferred using fossils as primary calibrations (figs.
+fig:fringillidages and fig:cvXY; supplementary figs.
+1 and 4).
 
@@
 A summarizing exercise may also amplify the effect of uncertainty and errors in source data, and blur parts of the evolutionary history of a lineage that might only be reflected in source chronograms and lost on the summary chronogram (Sauquet et al., 2021).
-\textcolor{red}{EJM: ??? it's not clear to me what this last part means! What might you be losing?
-LLSR: I changed loosing to blurring. But I will elaborate more on it}
-Choosing source chronograms that we are going to keep and the ones that we are going to discard might be key.
-\textcolor{red}{LLSR: Example}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 4    %%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%    FIGURE FRINGILLIDAE    %%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \includegraphics{../figures/figure-fringillidae/median_and_calibration_ages-congruified.png}
-\caption{Fringillidae median summary chronogram generated with DateLife. It has 256 tips and 233 nodes.}
+\caption{Fringillidae median summary chronogram generated with DateLife. It has 256 tips and 233 nodes, from which 212 have age data from at least one published chronogram.}
 \label{fig:fringillidages}
@@
 
-% in this file I leave the figure captions outside\ caption{} because I want them
-% to be formatted in the same way as the general text (double spaced and linenumbered)
-\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%    FIGURE CROSS VALIDATION    %%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 \includegraphics{../figures/figure-cross-validation/fig-cross-validation-xy-plots.pdf}
-\caption{Results from cross validation analysis.}
+\caption{Results from cross validation analysis. Each plot compares the original age estimate (x axis) with the age obtained with a DateLife analysis (y axis), per node.}
 \label{fig:cvXY}
@@
 
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/fig-cross-validation-xy-plots-diffs.pdf}
-\caption{Results from cross validation analysis.}
-\label{fig:cvXYdiffs}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/cross_validation_2.png}
-\caption{Cross validation of second source chronogram.
-The chronogram shown in black corresponds to the dates published in the original study.
-The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
-\label{fig:cv2}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/cross_validation_3.png}
-\caption{Cross validation of third source chronogram.
-The chronogram shown in black corresponds to the dates published in the original study.
-The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
-\label{fig:cv3}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/cross_validation_4.png}
-\caption{Cross validation of fourth source chronogram.
-The chronogram shown in black corresponds to the dates published in the original study.
-The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
-\label{fig:cv4}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/cross_validation_6.png}
-\caption{Cross validation of sixth source chronogram.
-The chronogram shown in black corresponds to the dates published in the original study.
-The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
-\label{fig:cv6}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/cross_validation_7.png}
-\caption{Cross validation of seventh source chronogram.
-The chronogram shown in black corresponds to the dates published in the original study.
-The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
-\label{fig:cv7}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/cross_validation_8.png}
-\caption{Cross validation of eight source chronogram.
-The chronogram shown in black corresponds to the dates published in the original study.
-The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
-\label{fig:cv8}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/cross_validation_9.png}
-\caption{Cross validation of ninth source chronogram.
-The chronogram shown in black corresponds to the dates published in the original study.
-The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
-\label{fig:cv9}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/cross_validation_10.png}
-\caption{Cross validation of tenth source chronogram.
-The chronogram shown in black corresponds to the dates published in the original study.
-The gray chronogram corresponds to the same tree topology dated with BLADJ using node ages from all other source chronograms as secondary calibrations.}
-\label{fig:cv10}
-\end{figure}
-
 \end{document}

### 2022-06-06 manuscript/manuscript.Rmd
submission

@@
 At the time of writing of this manuscript 
-,
-<!--(`r format(Sys.time(), "%b %d, %Y")`), -->
+(`r format(Sys.time(), "%b %d, %Y")`),
 DateLife's chronogram database latest version consist of `r length(opentree_chronograms$trees)` chronograms published in `r length(unique(opentree_chronograms$studies))` different studies.

### 2022-06-06 manuscript/manuscript.tex
submission

@@
 At the time of writing of this manuscript
-,
+(Jun 05, 2022),
 DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies.

### 2022-06-06 manuscript/supplementary_figured.Rmd
submission

---
title             : "Supplementary Figures \\linebreak DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life"
shorttitle        : "DateLife Supplementary Figures"

author:
  - name          : "Luna L. Sánchez Reyes"
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : ""
    email         : "sanchez.reyes.luna@gmail.com"
  - name          : "Emily Jane McTavish"
    affiliation   : "1"
  - name          : "Brian O'Meara"
    affiliation   : "2"

affiliation:
  - id            : "1"
    institution   : "Science and Engineering Building 1, School of Natural Sciences, University of California, Merced"
  - id            : "2"
    institution   : "Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville"

floatsintext      : no
figurelist        : NO
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            :
  papaja::apa6_pdf:
    includes      :
      in_header: "preamble.tex"
      after_body  : ["figures-supplementary.tex"]
---

```{r setup, include = FALSE}
library("papaja")
r_refs("paper_references.bib")
```

### 2022-06-06 manuscript/supplementary_figured.tex
submission

--- 2022-06-06 manuscript/manuscript.tex
@@
   man]{apa6}
-\title{DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life}
+\title{Supplementary Figures \linebreak DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life}
 \author{Luna L. Sánchez Reyes\textsuperscript{1,2}, Emily Jane McTavish\textsuperscript{1}, \& Brian O'Meara\textsuperscript{2}}
@@
 \hypersetup{
-  pdftitle={DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life},
+  pdftitle={Supplementary Figures DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life},
   pdfauthor={Luna L. Sánchez Reyes1,2, Emily Jane McTavish1, \& Brian O'Meara2},
   pdflang={en-EN},
-  pdfkeywords={Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations},
   hidelinks,
@@
 \fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 \DeclareUnicodeCharacter{0301}{*************************************}
@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4845}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 \makeatother
-\usepackage{lineno}
-
-\linenumbers
 \usepackage{csquotes}
@@
 
-\shorttitle{DateLife: revealing the dated Tree of Life}
+\shorttitle{DateLife Supplementary Figures}
 
 \authornote{
-
-School of Natural Sciences, University of California, Merced, Science and Engineering Building 1.
-
-Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 446 Hesler Biology Building, Knoxville, TN 37996, USA.
-
-The authors made the following contributions.
-Luna L.
-Sánchez Reyes: Data curation, Investigation, Software, Visualization, Validation, Writing - Original Draft Preparation, Writing - Review \& Editing; Emily Jane McTavish: Resources, Software, Writing - Review \& Editing; Brian O'Meara: Conceptualization, Funding acquisition, Methodology, Resources, Software, Supervision, Writing - Review \& Editing.
 
@@
 
-\affiliation{\vspace{0.5cm}\textsuperscript{1} University of California, Merced\\\textsuperscript{2} University of Tennessee, Knoxville}
-
-\abstract{%
-Achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram) is a difficult and time-consuming task.
-But the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
-Knowledge of the evolutionary time frame of organisms is key for research in the natural sciences. It also represent valuable information for education, science communication, and policy decisions.
-When chronograms are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on evolutionary timeframe is exposed in a programatic and reusable way, as intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
-Here we present DateLife, a service implemented as an R package and an R Shiny website application available at www.datelife.org, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time frame of evolution.
-The main DateLife workflow constructs a chronogram for any given combination of taxon names by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from the TreeBASE database as well.
-We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
-DateLife will be useful to increase awereness of the existing variation in alternative hypothesis of evolutionary time for the same organisms, and can foster exploration of the effect of alternative evolutionary timing hypotheses on the results of downstream analyses, providing a framework for a more informed interpretation of evolutionary results.
-}
+\affiliation{\vspace{0.5cm}\textsuperscript{1} Science and Engineering Building 1, School of Natural Sciences, University of California, Merced\\\textsuperscript{2} Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville}
 
@@
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-Chronograms --phylogenies with branch lengths proportional to time-- provide key data on evolutionary time frame for the study of natural processes in many areas of biological research, such as developmental biology (Delsuc et al., 2018; Laubichler \& Maienschein, 2009), conservation biology (Felsenstein, 1985; C.
-Webb, 2000), historical biogeography (Posadas, Crisci, \& Katinas, 2006), and species diversification (Magallon \& Sanderson, 2001; Morlon, 2014).
-
-Building a chronogram is not an easy task.
-It requires obtaining and curating data to construct a phylogeny, selecting and placing appropriate calibrations on the phylogeny using independent age data points from the fossil record or other dated events, and inferring the full dated tree;
-it also generally requires specialized biological training, taxonomic domain knowledge, and a non-negligible amount of research time, computational resources and funding.
-
-Here we present the DateLife project which has the main goal of capturing age data from published chronograms, and making these data readily accessible to the community for reuse and reanalysis.
-DateLife's core software application is available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org.
-It features key elements for scientific reproducibility, such as a versioned, open and fully public source database (McTavish et al., 2015), data stored and available in a computer readable format (Vos et al., 2012), automated and programmatic ways of accessing the data (Stoltzfus et al., 2013) and methods to summarize and compare the data.
-
-\hypertarget{description}{%
-\section{Description}\label{description}}
-
-DateLife's core software application consists of the R package \texttt{datelife}.
-Its latest stable version -- v0.6.2, is available from the CRAN repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
-ape (Paradis, Claude, \& Strimmer, 2004),
-bold (Chamberlain et al., 2019),
-geiger (Pennell et al., 2014),
-paleotree (Bapst, 2012),
-phyloch (Heibl, 2008),
-phylocomr (Ooms \& Chamberlain, 2018),
-phytools (Revell, 2012),
-rotl (Michonneau, Brown, \& Winter, 2016), and
-taxize (Chamberlain \& Szöcs, 2013; Chamberlain et al., 2019).
-Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife algorithm: creating a search query, searching a database, and summarizing results from the search.
-
-\hypertarget{creating-a-search-query}{%
-\subsection{Creating a search query}\label{creating-a-search-query}}
-
-DateLife starts by processing an input consisting of at least two taxon names, which can be provided as a comma separated character string or as tip labels on a tree.
-If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis et al., 2004).
-The input tree is not required to have branch lengths, and its topology is used in the summary steps described in the next section.
-
-DateLife accepts scientific names that can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or a binomial species name.
-Subspecies and variants are ignored.
-If an input taxon name belongs to an inclusive taxonomic group, the algorithm has two alternative behaviors defined by the ``get species from taxon'' flag.
-If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group (following a taxonomy of choice; current options are Open Tree of Life, NCBI, GBIF, or IRMNG taxonomies) and adds them to the input string.
-If the flag is inactive, DateLife excludes any taxon names above the species level from the search query.
-
-DateLife processes input scientific names using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database.
-TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard (Boyle et al., 2013).
-DateLife implements TNRS using OpenTree's unified taxonomy as standard (Open Tree Of Life et al., 2016; Rees \& Cranston, 2017), storing OpenTree's Taxonomy identification numbers for further processing.
-
-The processed input taxon names are saved as an R object of a newly defined class, \texttt{datelifeQuery}, that is used in the following steps.
-This object contains the standardized names, the corresponding OpenTree taxonomic id numbers, and the topology of the input tree if any was provided.
-
-\hypertarget{searching-a-chronogram-database}{%
-\subsection{Searching a chronogram database}\label{searching-a-chronogram-database}}
-
-At the time of writing of this manuscript
-,
-DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies.
-It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata (McTavish et al., 2015), which allows automatic and reproducible assembly of our chronogram database.
-Datelife's chronogram database is navigable as an R data object within the \texttt{datelife} R package.
-
-A unique feature of OpenTree's Phylesystem is that the community can add new published, state-of-the-art chronograms any time through their curator application (\url{https://tree.opentreeoflife.org/curator}).
-As chronograms are added to Phylesystem, they are incorporated into an updated \texttt{datelife}'s database that is assigned a new version number, followed by a package release on CRAN.
-\texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
-Users can also upload and curate new chronograms to OpenTree themselves.
-Then, trigger an update of their local \texttt{datelife} database to incorporate the new chronograms and have them available for their current analysis.
-
-A DateLife search is implemented by matching processed taxon names provided by the user to tip labels in the chronogram database.
-Chronograms with at least two matching taxon names on their tip labels are identified and pruned down to preserve only the matched taxa.
-These matching pruned chronograms are referred to as source chronograms.
-Total distance (in units of millions of years) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Figure \ref{fig:workflow}).
-The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a ``phylo'' object or newick string.
-Finally, the patristic matrices are associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.
-
-\hypertarget{summarizing-search-results}{%
-\subsection{Summarizing search results}\label{summarizing-search-results}}
-
-Summary information is extracted from the \texttt{datelifeResult} object to inform decisions for subsequent steps in the analysis workflow. Basic summary information available to the user is:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  The matching pruned chronograms as newick strings or ``phylo'' objects.
-\item
-  The ages of the root of all source chronograms.
-These ages can correspond to the age of the most recent common ancestor (mrca) of the user's group of interest if the source chronograms have all taxa belonging to the group.
-If not, the root corresponds to the mrca of a subgroup withing the group of interest.
-\item
-  Study citations where original chronograms were published.
-\item
-  A report of input taxon names matches across source chronograms.
-\item
-  The source chronogram(s) with the most input taxon names.
-\item
-  Various single summary chronograms resulting from summarizing age data, generated using the methodology described next.
-\end{enumerate}
-
-\emph{\textbf{Choosing a topology.--}}
-DateLife requires a tree topology to summarize age data upon.
-We recommend that users provide a tree topology as input from the literature, or one of their own making.
-If no topology is provided, DateLife automatically obtains one from the OpenTree synthetic tree (Open Tree Of Life et al., 2019).
-These are known as ``induced'' OpenTree subtrees, which do not necessarily correspond to a monophyletic group.
-Alternatively, DateLife can combine topologies from source chronograms using a supertree approach.
-To combine topologies from source chronograms into a single summary (or supertree) topology, the DateLife algorithm starts by identifying the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
-(2009).
-In rare cases, a group of trees can have multiple groves.
-By default, DateLife chooses the grove with the most taxa, however, the ``criterion = trees'' flag allows the user to choose the grove with the most trees instead.
-If source chronograms do not form a grove, the supertree reconstruction will fail.
-
-\emph{\textbf{Dating the topology.--}}
-Input topologies from OpenTree or the supertree approach described above do not include branch length estimates of any kind.
-Optionally, to estimate branch lengths proportional to substitution rates on these topologies, DateLife can mine the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain genetic markers for the input taxa.
-These markers are aligned with MUSCLE (Edgar, 2004) (by default) or MAFFT (Katoh, Asimenos, \& Toh, 2009).
-This alignment can be used to estimate branch lengths on input topologies that lack branch lengths.
-Currently, branch length reconstruction in DateLife is performed using parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed (Schliep, 2011).
-While relative branch length information provides additional data for nodes without secondary date calibrations, topologies without branch lengths can also be dated.
-
-Once a topology is chosen, DateLife applies the congruification method (Eastman, Harmon, \& Tank, 2013) to find nodes belonging to the same clade across source chronograms, and extract the corresponding node ages from the patristic distance matrices stored as \texttt{datelifeResult}.
-By definition, the matrices store total distance (time from tip to tip), hence, node ages correspond to half the values stored in the patristic distance matrices.
-This assumes that the terminal taxa are coeval and occur at the present.
-A table of congruified node ages that can be used as calibrations for a dating analysis is stored as a \texttt{congruifiedCalibrations} object.
-
-For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method {[}SDM; Criscuolo, Berry, Douzery, and Gascuel (2006){]}, which deforms patristic distance matrices by minimizing variance and then averaging them.
-These single summary taxon pair age matrices (Summarized calibrations) can be applied as calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Campbell O.
-Webb, Ackerly, \& Kembel, 2008; Campbell O.
-Webb \& Donoghue, 2005), and treePL (Stephen A.
-Smith \& O'Meara, 2012).
-
-By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully dated topology.
-BLADJ fixes node ages that have calibration data, and distributes time between nodes with no data evenly between nodes with calibration data.
-This minimizes age variance in the resulting chronogram (Campbell O. Webb et al., 2008).
-BLADJ does not use branch lengths even when they are present in the input tree or summarizing topology.
-When there is conflict in ages between nodes with calibration data, the algorithm ignores node ages that are older than the age of a parent node.
-BLADJ requires a root age estimate.
-If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature.
-If none is provided, DateLife assigns an arbitrary age to the root as 10\% older than the oldest age available within the group.
-
-Alternative options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the calibrations.
-PATHd8 is a non-clock, rate-smoothing method (Britton et al., 2007) to date trees.
-treePL (Stephen A. Smith \& O'Meara, 2012), is a semi-parametric, rate-smoothing, penalized likelihood dating method
-(Michael J. Sanderson, 2002).
-The MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003) approach in datelife uses the calibrations as priors on node ages.
-
-\emph{\textbf{Visualizing results.--}}
-Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R ``phylo'' format), as well as visualize and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions available from the R package \texttt{datelifeplot} (Sanchez-Reyes \& O'Meara, 2022).
-
-\hypertarget{benchmark}{%
-\section{Benchmark}\label{benchmark}}
-
-\texttt{datelife}'s code speed was tested on an Apple iMac
-with one 3.4 GHz Intel Core i5 processor.
-We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
-Query processing time increases roughly linearly with number of input taxon names, and
-increases considerably if Taxonomic Name Resolution Service (TNRS) is activated.
-Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
-Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig.
-\ref{fig:runtime_main}).
-\texttt{datelife}'s code performance was evaluated with a set of unit tests designed and
-implemented with the R package testthat (R Core Team, 2018) that were run both locally
-with the devtools package (R Core Team, 2018), and on a public server --via
-GitHub, using the continuous integration tool Travis CI (\url{https://travis-ci.org}). At
-present, unit tests cover more than 40\% of \texttt{datelife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).
-Unit testing helps identify potential issues as code is updated or, more critically, as services code relies upon may change.
-
-\hypertarget{case-studies}{%
-\section{Case studies}\label{case-studies}}
-
-We illustrate the DateLife algorithm using a family within the Passeriform birds encompassing the true finches, Fringillidae, as case study.
-The first example analyses 6 bird species and shows all steps of the algorithm.
-The second example is an application analysing 289 species in the family Fringillidae that are included in the NCBI taxonomy.
-
-\hypertarget{a-small-example}{%
-\subsection{A small example}\label{a-small-example}}
-
-\emph{\textbf{Creating a search query.-}}
-We chose 6 bird species within the Passeriformes. The sample includes
-two species of cardinals:
-the black-thighed grosbeak -- \emph{Pheucticus tibialis} and
-the crimson-collared grosbeak -- \emph{Rhodothraupis celaeno};
-three species of buntings:
-the yellowhammer -- \emph{Emberiza citrinella},
-the pine bunting -- \emph{Emberiza leucocephalos} and
-the yellow-throated bunting -- \emph{Emberiza elegans};
-and one species of tanager, the vegetarian finch -- \emph{Platyspiza crassirostris}.
-
-Processing input names found that \emph{Emberiza elegans} is synonym for \emph{Schoeniclus elegans} in the default reference taxonomy {[}Open Tree of Life Taxonomy v3.3, June 1, 2021{]}.
-For a detailed discussion on the state of the synonym refer to Avibase (Avibase, 2022; Lepage, 2004; Lepage, Vaidya, \& Guralnick, 2014).
-Discovering this synonym allowed assigning five age data points for the parent node of \emph{Emberiza elegans}, shown as \emph{Schoeniclus elegans} in figure \ref{fig:figure2}A, which would not have had any data otherwise.
-
-\emph{\textbf{Searching the database.-}}
-DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies (Fig.
-\ref{fig:figure2}B).
-Three studies matched five input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2015; Hedges, Marin, Suleski, Paymer, \& Kumar, 2015; Jetz, Thomas, Joy, Hartmann, \& Mooers, 2012), one study matched four input names (Hooper \& Price, 2017) and two studies matched two input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2013; Burns et al., 2014).
-No studies matched all input names.
-Together, source chronograms provide 28 unique age data points, covering all nodes on our chosen tree topology to date (Table \ref{tbl:table1}).
-
-\emph{\textbf{Summarizing search results.-}}
-DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and congruified and mapped age data to nodes in this chosen topology (Table \ref{tbl:table1}).
-The name processing step allowed including five data points for node ``n4'' (parent of \emph{Schoeniclus elegans}; Fig.
-\ref{fig:figure2}A) that would not have had any data otherwise due to name mismatch.
-Age summary statistics per node were calculated (Table \ref{tbl:table2}) and used as calibrations to date the tree topology using the BLADJ algorithm.
-As expected, more inclusive nodes (e.g., node ``n1'') have more variance in age data than less inclusive nodes (e.g., node ``n5'').
-Summary age data for node ``n2'' were excluded as final calibration because they are older than age data of the more inclusive node, ``n1'' (Fig. \ref{fig:figure2}C4).
-
-\hypertarget{the-family-fringillidae}{%
-\subsection{The family Fringillidae}\label{the-family-fringillidae}}
-
-\emph{\textbf{Creating a query.-}}
-To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the ``get species from taxon'' flag,
-which gets all recognized species names within a named group from a taxonomy of choice.
-Following the NCBI taxonomy, our DateLife query has 289 Fringillidae species.
-This taxon-constrained approach implies that the final results of a full DateLife analysis will be done using a tree topology and ages for the species in a named group, which do not necessarily correspond to a monophyletic group.
-Users can change this behaviour by providing a monophyletic tree as input for a DateLife search, or as a tree topology for a DateLife summary.
-
-\emph{\textbf{Searching the database.-}}
-Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species.
-The DateLife search identified 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Price et al., 2014).
-Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices.
-
-\emph{\textbf{Summarizing search results.-}}
-The final step is to summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM.
-As explained in the ``Description'' section, a tree topology to summarize age data upon is required.
-By default, DateLife uses the portion of OpenTree synthetic phylogeny that contains the species provided as search query as summarizing topology.
-According to this phylogeny, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).
-Age data from source chronograms is summarised into a single patristic distance matrix and then the available node ages are used as fixed node calibrations over a consensus tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
-\ref{fig:fringillidages}).
-Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
-
-\hypertarget{cross-validation-test}{%
-\section{Cross-validation test}\label{cross-validation-test}}
-
-We performed a cross validation analysis of the DateLife workflow using the Fringillidae chronograms.
-We used the individual tree topologies from each of the 19 source chronograms from 13 studies as inputs, treating their node ages as unknown.
-We then estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.
-
-We found that node ages from original study, and ages estimated using all other age data available are correlated (Supplementary Fig. \ref{fig:cvXY}).
-
-For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka `closer to the root') relative to the orginal estimate, and overestimate ages for nodes closer to the tips.
-Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. \ref{fig:cvXYdiffs}).
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-\texttt{datelife} makes state-of-the-art data on evolutionary time frame easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
-It is an open service that does not require any expert biological knowledge from users --besides the names of the species or group they want to work with, for any of its functionality.
-
-A total of 99474 unique terminal taxa are represented in \texttt{datelife}'s database.
-Incorporation of more chronograms into \texttt{datelife}'s database will continue to improve its services.
-One option to increase the number of chronograms in the DateLife database is the Dryad data repository.
-Methods to automatically mine chronograms from Dryad could be designed and implemented.
-However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined manually by a human curator.
-We would like to emphasize on the importance of sharing chronogram data, including systematically curated metadata, into open repositories, such as OpenTree's Phylesystem (McTavish et al., 2015) for the benefit of the scientific community as a whole.
-
-\hypertarget{primary-vs-secondary-calibrations-in-dating-analyses}{%
-\subsection{Primary vs Secondary calibrations in dating analyses}\label{primary-vs-secondary-calibrations-in-dating-analyses}}
-
-While most chronograms in DateLife's database are constructed using primary calibrations (molecular substitution rates or ages obtained from the fossil record or geological events), DateLife summarizes chronograms using secondary calibrations only (ages coming from other chronograms).
-
-Graur and Martin (2004) cautioned on the increased error and uncertainty in estimated ages when using secondary calibrations in dating analyses.
-Schenk (2016) showed that, in simulations, divergence times inferred using secondary calibrations are significantly younger than those inferred with primary calibrations, when obtained with Bayesian inference methods, and when priors are implemented in similar ways in both analyses.
-Accordingly, the scientific community seems to have more confidence in chronograms obtained from a single analysis, using fossil data as primary sources of calibrations (Schenk, 2016), and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data used in the analysis reflect a coherent evolutionary history (Sauquet, 2013), as for example done by (Antonelli et al., 2017).
-There have been attempts to create fossil calibration databases (Ksepka et al., 2015), though these still have room to grow.
-
-\textcolor{red}{LLSR: What does our cross validation analysis show in relation to diffs between primary and secondary calibrations?}
-
-It seems that using several (as opposed to just a few) secondary calibrations can provide sufficient information to alleviate or even neutralize potential biases (Sauquet, 2013).
-Certainly, further studies are required to fully understand the effect of secondary calibrations on outputs from different tree dating methods, and on downstream analyses.
-It is possible that secondary calibrations can be safely used with dating methods that do not require setting priors, such as penalized likelihood (Michael J.
-Sanderson, 2003), methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ (Campbell O.
-Webb et al., 2008; Campbell O.
-Webb \& Donoghue, 2005), or methods that summarize age data unto a tree topology.
-
-\hypertarget{variation-in-source-chronograms}{%
-\subsection{Variation in source chronograms}\label{variation-in-source-chronograms}}
-
-Conflict among alternative dating studies is common in the literature
-(see, for example, the robust ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández, 2015; Ramshaw et al., 1972; Michael J.
-Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021)).
-
-Source chronograms may have been estimated using different types of primary calibrations (fossil record, geologic or biogeographic events, molecular substitution rates).
-These differences in approach can deepen the already substantial variation in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
-For example, the chronograms from Burns et al.
-(2014) tend to have much older date estimates for the same nodes than other inferences (Figure 5).
-This study inferred dates using molecular substitution rate estimates across birds (Weir \& Schluter, 2008)
-\textcolor{red}{EJM: MORE ON the FINCH examples and the cross validation here}
-
-Different chronograms might also imply fundamentally distinct evolutionary hypotheses.
-For example, two different researchers working on the same clade could both carefully select and argue for their choices of fossil calibrations.
-Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses the same one to calibrate outside the clade, the resulting age estimates will often differ substantially, as the placement of calibrations as stem or crown group is proved to deeply affect estimated times of lineage divergence (Sauquet, 2013).
-
-\hypertarget{sumarizing-chronograms}{%
-\subsection{Sumarizing chronograms}\label{sumarizing-chronograms}}
-
-By default, DateLife currently summarizes all source chronograms that overlap with at least two species names. Users can exclude source chronograms if they have reasons to do so.
-Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely.
-To our knowledge, there are no tested measures to determine independently when a chronogram is better than another.
-Yet, several characteristics of the data used for dating analyses, as well as from the output chronogram itself, could be used to score the quality of source chronograms.
-
-Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used Magallón et al. (2015).
-Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
-
-DateLife provides an opportunity to capture concordance and conflict among date estimates, which can also be used as a metric for chronogram reliability.
-Its open database of chronograms allows other researchers to do such analyses themselves reproducibly, and without needing permission.
-Though, of course, they should follow proper citation practices, especially for the source chronogram studies.
-
-The exercise of summarizing age data from across multiple studies provides the opportunity to work with a more inclusive chronogram,
-that reflects a unified evolutionary history for a lineage, by putting together evidence from different hypotheses.
-The largest, and taxonomically broadest chronogram currently available from OpenTree was constructed summarizing age data from 2,274 published chronograms using NCBI's taxonomic tree as backbone (Hedges et al., 2015).
-A summarizing exercise may also amplify the effect of uncertainty and errors in source data, and blur parts of the evolutionary history of a lineage that might only be reflected in source chronograms and lost on the summary chronogram (Sauquet et al., 2021).
-\textcolor{red}{EJM: ??? it's not clear to me what this last part means! What might you be losing?
-LLSR: I changed loosing to blurring. But I will elaborate more on it}
-Choosing source chronograms that we are going to keep and the ones that we are going to discard might be key.
-\textcolor{red}{LLSR: Example}
-
-\hypertarget{effects-on-downstream-analyses}{%
-\subsection{Effects on downstream analyses}\label{effects-on-downstream-analyses}}
-
-For downstream analyses, using alternative chronogram may deeply affect our inferences (Title \& Rabosky, 2016), particularly when studying phenomena dependent on the timing of species diversification events, such as macroevolutionary processes.
-
-In ecology and conservation biology, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance (Campbell O.
-Webb et al., 2008).
-Hence, DateLife's workflow features different ways of estimating node ages in the absence of calibrations and branch length information for certain taxa.
-``Making up'' branch lengths is a common practice in scientific publications: Jetz et al.
-(2012), created a time-calibrated tree of all 9,993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
-(2018) created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37\% had molecular data; Stephen A.
-Smith and Brown (2018) constructed a tree of 353,185 seed plants where only 23\% had molecular data.
-
-Notably, risks come with this practice.
-Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration, using polytomy resolution and branch estimation methods.
-There has yet to be a thorough analysis of what can go wrong when one extends inferences beyond the data in this way, so we urge caution; we also urge readers to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, and assessing the consistency of the results with prior work.
-
-\hypertarget{conclusions}{%
-\section{Conclusions}\label{conclusions}}
-
-Knowledge of the evolutionary time frame of organisms is key to many research areas: trait evolution,
-species diversification, biogeography, macroecology and more. It is also crucial for education, science communication and policy, but generating chronograms is difficult,
-especially for those who want to use phylogenies but who are not systematists, or
-do not have the time to acquire and develop the necessary knowledge and skills to construct them on their own.
-Importantly, years of primarily public funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
-
-The DateLife project allows for easy and fast summary of public and state-of-the-art data on time of lineage divergence.
-It provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identification of regions that require more research, or that have conflicting information.
-It is available as an R package, and as a web-based R shiny application at www.datelife.org
-Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
-The DateLife project helps with awareness of the existing variation in expert time of divergence data, and will foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
-
-\hypertarget{availability}{%
-\section{Availability}\label{availability}}
-
-The \texttt{datelife} software is free and open source and it can be used through its R shiny web application at
-\url{http://www.datelife.org}, through the \texttt{datelife} R package, and through Phylotastic's project web portal \url{https://phylo.cs.nmsu.edu/}.
-DateLife's web application is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker and Open Tree of Life's infrastructure (dates.opentreeoflife.org/datelife).
-\texttt{datelife}'s R package stable version is available
-for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
-using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R. Development versions
-are available from the GitHub repository (\url{https://github.com/phylotastic/datelife})
-and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.
-
-\hypertarget{supplementary-material}{%
-\section{Supplementary Material}\label{supplementary-material}}
-
-Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities are available at \href{https://github.com/LunaSare/datelifeMS1}{datelifeMS1}, \href{https://github.com/LunaSare/datelife_examples}{datelife\_examples}, and \href{https://github.com/LunaSare/datelife_benchmark}{datelife\_benchmark} repositories in LLSR's GitHub account.
-
-\hypertarget{funding}{%
-\section{Funding}\label{funding}}
-
-Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to the Datelife project; DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), ABI-1458572 to the Phylotastic project, and ABI-1759846 to the Open Tree of Life project.
-
-\hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
-
-The DateLife project was born as a prototype tool aiming to provide these services, and was initially developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA (Stoltzfus et al., 2013).
-We thank colleagues from the O'Meara Lab at the University
-of Tennesse Knoxville for suggestions, discussions and software testing.
-The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
-that led to initial work on this project.
-The team that assembled DateLife's first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
-The Open Tree of Life project that provides the open, metadata rich repository of
-trees used to construct DateLife's chronogram database.
-The many scientists who publish their chronograms in an open, reusable form, and the scientists who curate them for deposition in the Open Tree of Life repository.
-The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.
-
-\newpage
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-alstrom2014discovery}{}}%
-Alström, P., Hooper, D.
-M., Liu, Y., Olsson, U., Mohan, D., Gelang, M., \ldots{} Price, T.
-D.
-(2014).
-Discovery of a relict lineage and monotypic family of passerine birds.
-\emph{Biology Letters}, \emph{10}(3), 20131067.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ane2009groves}{}}%
-Ané, C., Eulenstein, O., Piaggio-Talice, R., \& Sanderson, M. J. (2009). Groves of phylogenetic trees. \emph{{Annals of Combinatorics}}, \emph{13}(2), 139--167.
-
-\leavevmode\vadjust pre{\hypertarget{ref-antonelli2017supersmart}{}}%
-Antonelli, A., Hettling, H., Condamine, F.
-L., Vos, K., Nilsson, R.
-H., Sanderson, M.
-J., \ldots{} Vos, R.
-A.
-(2017).
-{Toward a self-updating platform for estimating rates of speciation and migration, ages, and relationships of Taxa}.
-\emph{Systematic Biology}, \emph{66}(2), 153--166.
-\url{https://doi.org/10.1093/sysbio/syw066}
-
-\leavevmode\vadjust pre{\hypertarget{ref-archie1986newick}{}}%
-Archie, J., Day, W.
-H., Felsenstein, J., Maddison, W., Meacham, C., Rohlf, F.
-J., \& Swofford, D.
-(1986).
-{The Newick tree format}.
-Retrieved from \url{\%7Bhttps://evolution.genetics.washington.edu/phylip/newicktree.html\%7D}
-
-\leavevmode\vadjust pre{\hypertarget{ref-avibase-emberiza}{}}%
-Avibase.
-(2022).
-{Yellow-throated Bunting}.
-\emph{{Avibase - The World Bird Database}}, (Online Resource).
-Retrieved from \url{\%7Bhttps://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\%7D}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Bapst2012a}{}}%
-Bapst, D.
-W.
-(2012).
-{Paleotree: An R package for paleontological and phylogenetic analyses of evolution}.
-\emph{{Methods in Ecology and Evolution}}, \emph{3}(5), 803--807.
-\url{https://doi.org/10.1111/j.2041-210X.2012.00223.x}
-
-\leavevmode\vadjust pre{\hypertarget{ref-barba2018constraining}{}}%
-Barba-Montoya, J., Reis, M.
-dos, Schneider, H., Donoghue, P.
-C., \& Yang, Z.
-(2018).
-Constraining uncertainty in the timescale of angiosperm evolution and the veracity of a cretaceous terrestrial revolution.
-\emph{New Phytologist}, \emph{218}(2), 819--834.
-
-\leavevmode\vadjust pre{\hypertarget{ref-barker2014mitogenomic}{}}%
-Barker, F. K. (2014). Mitogenomic data resolve basal relationships among passeriform and passeridan birds. \emph{Molecular Phylogenetics and Evolution}, \emph{79}, 313--324.
-
-\leavevmode\vadjust pre{\hypertarget{ref-barker2013going}{}}%
-Barker, F.
-K., Burns, K.
-J., Klicka, J., Lanyon, S.
-M., \& Lovette, I.
-J.
-(2013).
-Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds.
-\emph{{Systematic Biology}}, \emph{62}(2), 298--320.
-
-\leavevmode\vadjust pre{\hypertarget{ref-barker2015new}{}}%
-Barker, F.
-K., Burns, K.
-J., Klicka, J., Lanyon, S.
-M., \& Lovette, I.
-J.
-(2015).
-New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
-\emph{The Auk: Ornithological Advances}, \emph{132}(2), 333--348.
-
-\leavevmode\vadjust pre{\hypertarget{ref-barker2004phylogeny}{}}%
-Barker, F.
-K., Cibois, A., Schikler, P., Feinstein, J., \& Cracraft, J.
-(2004).
-Phylogeny and diversification of the largest avian radiation.
-\emph{Proceedings of the National Academy of Sciences}, \emph{101}(30), 11040--11045.
-
-\leavevmode\vadjust pre{\hypertarget{ref-beresford2005african}{}}%
-Beresford, P., Barker, F., Ryan, Pg., \& Crowe, T.
-(2005).
-African endemics span the tree of songbirds (passeri): Molecular systematics of several evolutionary {`enigmas.'} \emph{Proceedings of the Royal Society B: Biological Sciences}, \emph{272}(1565), 849--858.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Boyle2013}{}}%
-Boyle, B., Hopkins, N., Lu, Z., Raygoza Garay, J.
-A., Mozzherin, D., Rees, T., \ldots{} Enquist, B.
-J.
-(2013).
-{The taxonomic name resolution service: An online tool for automated standardization of plant names}.
-\emph{{BMC Bioinformatics}}, \emph{14}(1).
-\url{https://doi.org/10.1186/1471-2105-14-16}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Britton2007}{}}%
-Britton, T., Anderson, C.
-L., Jacquet, D., Lundqvist, S., \& Bremer, K.
-(2007).
-{Estimating Divergence Times in Large Phylogenetic Trees}.
-\emph{{Systematic Biology}}, \emph{56}(788777878), 741--752.
-\url{https://doi.org/10.1080/10635150701613783}
-
-\leavevmode\vadjust pre{\hypertarget{ref-bryson2014diversification}{}}%
-Bryson Jr, R.
-W., Chaves, J., Smith, B.
-T., Miller, M.
-J., Winker, K., Pérez-Emán, J.
-L., \& Klicka, J.
-(2014).
-Diversification across the new world within the `blue'cardinalids (aves: cardinalidae).
-\emph{Journal of Biogeography}, \emph{41}(3), 587--599.
-
-\leavevmode\vadjust pre{\hypertarget{ref-burleigh2015building}{}}%
-Burleigh, J. G., Kimball, R. T., \& Braun, E. L. (2015). Building the avian tree of life using a large-scale, sparse supermatrix. \emph{Molecular Phylogenetics and Evolution}, \emph{84}, 53--63.
-
-\leavevmode\vadjust pre{\hypertarget{ref-burns2014phylogenetics}{}}%
-Burns, K.
-J., Shultz, A.
-J., Title, P.
-O., Mason, N.
-A., Barker, F.
-K., Klicka, J., \ldots{} Lovette, I.
-J.
-(2014).
-Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds.
-\emph{{Molecular Phylogenetics and Evolution}}, \emph{75}, 41--77.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Chamberlain2013}{}}%
-Chamberlain, S.
-A., \& Szöcs, E.
-(2013).
-{taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}}.
-\emph{{F1000Research}}, \emph{2}(191), 1--29.
-\url{https://doi.org/10.12688/f1000research.2-191.v2}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Chamberlain2018}{}}%
-Chamberlain, S.
-A., Szöcs, E., Foster, Z., Arendsee, Z., Boettiger, C., Ram, K., \ldots{} Li, G.
-(2019).
-\emph{{taxize: Taxonomic information from around the web}}.
-Retrieved from \url{https://github.com/ropensci/taxize}
-
-\leavevmode\vadjust pre{\hypertarget{ref-chaves2013biogeography}{}}%
-Chaves, J.
-A., Hidalgo, J.
-R., \& Klicka, J.
-(2013).
-Biogeography and evolutionary history of the n eotropical genus s altator (a ves: T hraupini).
-\emph{Journal of Biogeography}, \emph{40}(11), 2180--2190.
-
-\leavevmode\vadjust pre{\hypertarget{ref-claramunt2015new}{}}%
-Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{{Science Advances}}, \emph{1}(11), e1501005.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Criscuolo2006}{}}%
-Criscuolo, A., Berry, V., Douzery, E.
-J.
-P., \& Gascuel, O.
-(2006).
-{SDM: A fast distance-based approach for (super)tree building in phylogenomics}.
-\emph{{Systematic Biology}}, \emph{55}(5), 740--755.
-\url{https://doi.org/10.1080/10635150600969872}
-
-\leavevmode\vadjust pre{\hypertarget{ref-delsuc2018phylogenomic}{}}%
-Delsuc, F., Philippe, H., Tsagkogeorga, G., Simion, P., Tilak, M.-K., Turon, X., \ldots{} Douzery, E.
-J.
-(2018).
-A phylogenomic framework and timescale for comparative studies of tunicates.
-\emph{BMC Biology}, \emph{16}(1), 1--14.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Eastman2013}{}}%
-Eastman, J.
-M., Harmon, L.
-J., \& Tank, D.
-C.
-(2013).
-{Congruification: Support for time scaling large phylogenetic trees}.
-\emph{{Methods in Ecology and Evolution}}, \emph{4}(7), 688--691.
-\url{https://doi.org/10.1111/2041-210X.12051}
-
-\leavevmode\vadjust pre{\hypertarget{ref-edgar2004muscle}{}}%
-Edgar, R. C. (2004). MUSCLE: Multiple sequence alignment with high accuracy and high throughput. \emph{Nucleic Acids Research}, \emph{32}(5), 1792--1797.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Felsenstein1985a}{}}%
-Felsenstein, J. (1985). {Phylogenies and the Comparative Method}. \emph{The American Naturalist}, \emph{125}(1), 1--15. Retrieved from \url{http://www.jstor.org/stable/2461605}
-
-\leavevmode\vadjust pre{\hypertarget{ref-gibb2015new}{}}%
-Gibb, G.
-C., England, R., Hartig, G., McLenachan, P.
-A., Taylor Smith, B.
-L., McComish, B.
-J., \ldots{} Penny, D.
-(2015).
-New zealand passerines help clarify the diversification of major songbird lineages during the oligocene.
-\emph{{Genome Biology and Evolution}}, \emph{7}(11), 2983--2995.
-
-\leavevmode\vadjust pre{\hypertarget{ref-graur2004reading}{}}%
-Graur, D., \& Martin, W. (2004). Reading the entrails of chickens: Molecular timescales of evolution and the illusion of precision. \emph{TRENDS in Genetics}, \emph{20}(2), 80--86.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hackett2008phylogenomic}{}}%
-Hackett, S.
-J., Kimball, R.
-T., Reddy, S., Bowie, R.
-C., Braun, E.
-L., Braun, M.
-J., \ldots{} others.
-(2008).
-A phylogenomic study of birds reveals their evolutionary history.
-\emph{Science}, \emph{320}(5884), 1763--1768.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Hedges2015}{}}%
-Hedges, S.
-B., Marin, J., Suleski, M., Paymer, M., \& Kumar, S.
-(2015).
-{Tree of life reveals clock-like speciation and diversification}.
-\emph{{Molecular Biology and Evolution}}, \emph{32}(4), 835--845.
-\url{https://doi.org/10.1093/molbev/msv037}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Heibl2008}{}}%
-Heibl, C. (2008). \emph{PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages.} Retrieved from \url{http://www.christophheibl.de/Rpackages.html}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hooper2017chromosomal}{}}%
-Hooper, D. M., \& Price, T. D. (2017). Chromosomal inversion differences correlate with range overlap in passerine birds. \emph{Nature Ecology \& Evolution}, \emph{1}(10), 1526.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Huelsenbeck2001}{}}%
-Huelsenbeck, J. P., \& Ronquist, F. (2001). {MRBAYES: Bayesian inference of phylogenetic trees}. \emph{Bioinformatics}, \emph{17}(8), 754--755. \url{https://doi.org/10.1093/bioinformatics/17.8.754}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Jetz2012}{}}%
-Jetz, W., Thomas, G., Joy, J.
-J.
-B., Hartmann, K., \& Mooers, A.
-(2012).
-{The global diversity of birds in space and time}.
-\emph{Nature}, \emph{491}(7424), 444--448.
-\url{https://doi.org/10.1038/nature11631}
-
-\leavevmode\vadjust pre{\hypertarget{ref-johansson2008phylogenetic}{}}%
-Johansson, U.
-S., Fjeldså, J., \& Bowie, R.
-C.
-(2008).
-Phylogenetic relationships within passerida (aves: Passeriformes): A review and a new molecular phylogeny based on three nuclear intron markers.
-\emph{Molecular Phylogenetics and Evolution}, \emph{48}(3), 858--876.
-
-\leavevmode\vadjust pre{\hypertarget{ref-katoh2009multiple}{}}%
-Katoh, K., Asimenos, G., \& Toh, H. (2009). Multiple alignment of DNA sequences with MAFFT. In \emph{Bioinformatics for DNA sequence analysis} (pp. 39--64). Springer.
-
-\leavevmode\vadjust pre{\hypertarget{ref-kimball2019phylogenomic}{}}%
-Kimball, R. T., Oliveros, C. H., Wang, N., White, N. D., Barker, F. K., Field, D. J., \ldots{} others. (2019). A phylogenomic supertree of birds. \emph{Diversity}, \emph{11}(7), 109.
-
-\leavevmode\vadjust pre{\hypertarget{ref-klicka2014comprehensive}{}}%
-Klicka, J., Barker, F.
-K., Burns, K.
-J., Lanyon, S.
-M., Lovette, I.
-J., Chaves, J.
-A., \& Bryson Jr, R.
-W.
-(2014).
-A comprehensive multilocus assessment of sparrow (aves: Passerellidae) relationships.
-\emph{Molecular Phylogenetics and Evolution}, \emph{77}, 177--182.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ksepka2015fossil}{}}%
-Ksepka, D.
-T., Parham, J.
-F., Allman, J.
-F., Benton, M.
-J., Carrano, M.
-T., Cranston, K.
-A., \ldots{} others.
-(2015).
-The fossil calibration database---a new resource for divergence dating.
-\emph{Systematic Biology}, \emph{64}(5), 853--859.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lamichhaney2015evolution}{}}%
-Lamichhaney, S., Berglund, J., Almén, M.
-S., Maqbool, K., Grabherr, M., Martinez-Barrio, A., \ldots{} others.
-(2015).
-Evolution of darwin's finches and their beaks revealed by genome sequencing.
-\emph{Nature}, \emph{518}(7539), 371--375.
-
-\leavevmode\vadjust pre{\hypertarget{ref-laubichler2009form}{}}%
-Laubichler, M. D., \& Maienschein, J. (2009). \emph{Form and function in developmental evolution}. Cambridge University Press.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lepage2004avibase}{}}%
-Lepage, D. (2004). \emph{Avibase: The world bird database}. Bird Studies Canada.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lepage2014avibase}{}}%
-Lepage, D., Vaidya, G., \& Guralnick, R. (2014). Avibase--a database system for managing and organizing taxonomic concepts. \emph{ZooKeys}, (420), 117.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lerner2011multilocus}{}}%
-Lerner, H.
-R., Meyer, M., James, H.
-F., Hofreiter, M., \& Fleischer, R.
-C.
-(2011).
-Multilocus resolution of phylogeny and timescale in the extant adaptive radiation of hawaiian honeycreepers.
-\emph{Current Biology}, \emph{21}(21), 1838--1844.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lovette2010comprehensive}{}}%
-Lovette, I.
-J., Pérez-Emán, J.
-L., Sullivan, J.
-P., Banks, R.
-C., Fiorentino, I., Córdoba-Córdoba, S., \ldots{} others.
-(2010).
-A comprehensive multilocus phylogeny for the wood-warblers and a revised classification of the parulidae (aves).
-\emph{Molecular Phylogenetics and Evolution}, \emph{57}(2), 753--770.
-
-\leavevmode\vadjust pre{\hypertarget{ref-magallon2001absolute}{}}%
-Magallon, S., \& Sanderson, M. J. (2001). Absolute diversification rates in angiosperm clades. \emph{Evolution}, \emph{55}(9), 1762--1780.
-
-\leavevmode\vadjust pre{\hypertarget{ref-magallon2010using}{}}%
-Magallón, S. (2010). Using fossils to break long branches in molecular dating: A comparison of relaxed clocks applied to the origin of angiosperms. \emph{Systematic Biology}, \emph{59}(4), 384--399.
-
-\leavevmode\vadjust pre{\hypertarget{ref-magallon2015metacalibrated}{}}%
-Magallón, S., Gómez-Acevedo, S., Sánchez-Reyes, L.
-L., \& Hernández-Hernández, T.
-(2015).
-A metacalibrated time-tree documents the early rise of flowering plant phylogenetic diversity.
-\emph{New Phytologist}, \emph{207}(2), 437--453.
-
-\leavevmode\vadjust pre{\hypertarget{ref-mctavish2015phylesystem}{}}%
-McTavish, E.
-J., Hinchliff, C.
-E., Allman, J.
-F., Brown, J.
-W., Cranston, K.
-A., Holder, M.
-T., \ldots{} Smith, S.
-A.
-(2015).
-Phylesystem: A git-based data store for community-curated phylogenetic estimates.
-\emph{Bioinformatics}, \emph{31}(17), 2794--2800.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Michonneau2016}{}}%
-Michonneau, F., Brown, J.
-W., \& Winter, D.
-J.
-(2016).
-{rotl: an R package to interact with the Open Tree of Life data}.
-\emph{{Methods in Ecology and Evolution}}, \emph{7}(12), 1476--1481.
-\url{https://doi.org/10.1111/2041-210X.12593}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Morlon2014}{}}%
-Morlon, H. (2014). {Phylogenetic approaches for studying diversification.} \emph{{Ecology Letters}}, \emph{17}(4), 508--525. \url{https://doi.org/10.1111/ele.12251}
-
-\leavevmode\vadjust pre{\hypertarget{ref-moyle2016tectonic}{}}%
-Moyle, R.
-G., Oliveros, C.
-H., Andersen, M.
-J., Hosner, P.
-A., Benz, B.
-W., Manthey, J.
-D., \ldots{} Faircloth, B.
-C.
-(2016).
-Tectonic collision and uplift of wallacea triggered the global songbird radiation.
-\emph{Nature Communications}, \emph{7}(1), 1--7.
-
-\leavevmode\vadjust pre{\hypertarget{ref-odeen2011evolution}{}}%
-Ödeen, A., Håstad, O., \& Alström, P. (2011). Evolution of ultraviolet vision in the largest avian radiation-the passerines. \emph{BMC Evolutionary Biology}, \emph{11}(1), 1--8.
-
-\leavevmode\vadjust pre{\hypertarget{ref-oliveros2019earth}{}}%
-Oliveros, C.
-H., Field, D.
-J., Ksepka, D.
-T., Barker, F.
-K., Aleixo, A., Andersen, M.
-J., \ldots{} others.
-(2019).
-Earth history and the passerine superradiation.
-\emph{Proceedings of the National Academy of Sciences}, \emph{116}(16), 7916--7925.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Ooms2018}{}}%
-Ooms, J., \& Chamberlain, S. (2018). \emph{Phylocomr: Interface to 'phylocom'}. Retrieved from \url{https://CRAN.R-project.org/package=phylocomr}
-
-\leavevmode\vadjust pre{\hypertarget{ref-opentreeAPIs}{}}%
-Open Tree Of Life, Redelings, B., Cranston, K.
-A., Allman, J., Holder, M.
-T., \& McTavish, E.
-J.
-(2016).
-{Open Tree of Life APIs v3.0}.
-\emph{{Open Tree of Life Project}}, (Online Resources).
-Retrieved from \url{\%7Bhttps://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\%7D}
-
-\leavevmode\vadjust pre{\hypertarget{ref-opentreeoflife2019synth}{}}%
-Open Tree Of Life, Redelings, B., Sánchez Reyes, L.
-L., Cranston, K.
-A., Allman, J., Holder, M.
-T., \& McTavish, E.
-J.
-(2019).
-Open tree of life synthetic tree v12.3.
-\emph{Zenodo}.
-Retrieved from \url{https://doi.org/10.5281/zenodo.3937742}
-
-\leavevmode\vadjust pre{\hypertarget{ref-packert2012horizontal}{}}%
-Päckert, M., Martens, J., Sun, Y.-H., Severinghaus, L.
-L., Nazarenko, A.
-A., Ting, J., \ldots{} Tietze, D.
-T.
-(2012).
-Horizontal and elevational phylogeographic patterns of himalayan and southeast asian forest passerines (aves: passeriformes).
-\emph{Journal of Biogeography}, \emph{39}(3), 556--573.
-
-\leavevmode\vadjust pre{\hypertarget{ref-paradis2004}{}}%
-Paradis, E., Claude, J., \& Strimmer, K. (2004). {APE: analyses of phylogenetics and evolution in R language}. \emph{Bioinformatics}, \emph{20}(2), 289--290.
-
-\leavevmode\vadjust pre{\hypertarget{ref-parchman2007coevolution}{}}%
-Parchman, T.
-L., Benkman, C.
-W., \& Mezquida, E.
-T.
-(2007).
-Coevolution between hispaniolan crossbills and pine: Does more time allow for greater phenotypic escalation at lower latitude?
-\emph{Evolution}, \emph{61}(9), 2142--2153.
-
-\leavevmode\vadjust pre{\hypertarget{ref-pennell2014geiger}{}}%
-Pennell, M.
-W., Eastman, J.
-M., Slater, G.
-J., Brown, J.
-W., Uyeda, J.
-C., FitzJohn, R.
-G., \ldots{} Harmon, L.
-J.
-(2014).
-Geiger v2.
-0: An expanded suite of methods for fitting macroevolutionary models to phylogenetic trees.
-\emph{Bioinformatics}, \emph{30}(15), 2216--2218.
-
-\leavevmode\vadjust pre{\hypertarget{ref-posadas2006historical}{}}%
-Posadas, P., Crisci, J. V., \& Katinas, L. (2006). Historical biogeography: A review of its basic concepts and critical issues. \emph{{Journal of Arid Environments}}, \emph{66}(3), 389--403.
-
-\leavevmode\vadjust pre{\hypertarget{ref-powell2014comprehensive}{}}%
-Powell, A.
-F., Barker, F.
-K., Lanyon, S.
-M., Burns, K.
-J., Klicka, J., \& Lovette, I.
-J.
-(2014).
-A comprehensive species-level molecular phylogeny of the new world blackbirds (icteridae).
-\emph{Molecular Phylogenetics and Evolution}, \emph{71}, 94--112.
-
-\leavevmode\vadjust pre{\hypertarget{ref-price2014niche}{}}%
-Price, T.
-D., Hooper, D.
-M., Buchanan, C.
-D., Johansson, U.
-S., Tietze, D.
-T., Alström, P., \ldots{} others.
-(2014).
-Niche filling slows the diversification of himalayan songbirds.
-\emph{Nature}, \emph{509}(7499), 222.
-
-\leavevmode\vadjust pre{\hypertarget{ref-pulgarin2013multilocus}{}}%
-Pulgarín-R, P.
-C., Smith, B.
-T., Bryson Jr, R.
-W., Spellman, G.
-M., \& Klicka, J.
-(2013).
-Multilocus phylogeny and biogeography of the new world pheucticus grosbeaks (aves: cardinalidae).
-\emph{Molecular Phylogenetics and Evolution}, \emph{69}(3), 1222--1227.
-
-\leavevmode\vadjust pre{\hypertarget{ref-RCoreTeam2018}{}}%
-R Core Team. (2018). \emph{{R: a language and environment for statistical computing}}. Vienna, Austria: R Foundation for Statistical Computing.
-
-\leavevmode\vadjust pre{\hypertarget{ref-rabosky2018inverse}{}}%
-Rabosky, D.
-L., Chang, J., Title, P.
-O., Cowman, P.
-F., Sallan, L., Friedman, M., \ldots{} others.
-(2018).
-An inverse latitudinal gradient in speciation rate for marine fishes.
-\emph{Nature}, \emph{559}(7714), 392.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ramshaw1972time}{}}%
-Ramshaw, J., Richardson, D., Meatyard, B., Brown, R., Richardson, M., Thompson, E., \& Boulter, D.
-(1972).
-The time of origin of the flowering plants determined by using amino acid sequence data of cytochrome c.
-\emph{New Phytologist}, \emph{71}(5), 773--779.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ratnasingham2007bold}{}}%
-Ratnasingham, S., \& Hebert, P. D. (2007). BOLD: The barcode of life data system (http://www. Barcodinglife. org). \emph{Molecular Ecology Notes}, \emph{7}(3), 355--364.
-
-\leavevmode\vadjust pre{\hypertarget{ref-rees2017automated}{}}%
-Rees, J. A., \& Cranston, K. (2017). Automated assembly of a reference taxonomy for phylogenetic data synthesis. \emph{Biodiversity Data Journal}, (5).
-
-\leavevmode\vadjust pre{\hypertarget{ref-Revell2012}{}}%
-Revell, L. J. (2012). Phytools: An r package for phylogenetic comparative biology (and other things). \emph{{Methods in Ecology and Evolution}}, \emph{3}, 217--223.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Ronquist2003}{}}%
-Ronquist, F., \& Huelsenbeck, J.
-P.
-(2003).
-{MrBayes 3: Bayesian phylogenetic inference under mixed models}.
-\emph{Bioinformatics}, \emph{19}(12), 1572--1574.
-\url{https://doi.org/10.1093/bioinformatics/btg180}
-
-\leavevmode\vadjust pre{\hypertarget{ref-datelifeplot}{}}%
-Sanchez-Reyes, L.
-L., \& O'Meara, B.
-(2022).
-{\texttt{datelifeplot}: Methods to plot chronograms and outputs of the \texttt{datelife} package}.
-\emph{R Package Release V0.2.2}.
-Retrieved from \url{https://zenodo.org/badge/latestdoi/381501451}
-
-\leavevmode\vadjust pre{\hypertarget{ref-datelifeR}{}}%
-Sanchez-Reyes, L.
-L., O'Meara, B., Eastman, J., Heath, T., Wright, A., Schliep, K., \ldots{} Alfaro, M.
-(2022).
-{\texttt{datelife}: Scientific Data on Time of Lineage Divergence for Your Taxa}.
-\emph{R Package Release V0.6.2}.
-Retrieved from \url{https://doi.org/10.5281/zenodo.593938}
-
-\leavevmode\vadjust pre{\hypertarget{ref-sanderson2002}{}}%
-Sanderson, Michael J.
-(2002).
-{Estimating Absolute Rates of Molecular Evolution and Divergence Times: A Penalized Likelihood Approach}.
-\emph{Molecular Biology and Evolution}, \emph{19}(1), 101--109.
-\url{https://doi.org/10.1093/oxfordjournals.molbev.a003974}
-
-\leavevmode\vadjust pre{\hypertarget{ref-sanderson2003r8s}{}}%
-Sanderson, Michael J. (2003). r8s: Inferring absolute rates of molecular evolution and divergence times in the absence of a molecular clock. \emph{Bioinformatics}, \emph{19}(2), 301--302.
-
-\leavevmode\vadjust pre{\hypertarget{ref-sanderson2001sources}{}}%
-Sanderson, Michael J., \& Doyle, J.
-A.
-(2001).
-Sources of error and confidence intervals in estimating the age of angiosperms from rbcL and 18S rDNA data.
-\emph{American Journal of Botany}, \emph{88}(8), 1499--1516.
-
-\leavevmode\vadjust pre{\hypertarget{ref-sauquet2013practical}{}}%
-Sauquet, H. (2013). A practical guide to molecular dating. \emph{{Comptes Rendus Palevol}}, \emph{12}(6), 355--367.
-
-\leavevmode\vadjust pre{\hypertarget{ref-sauquet2021age}{}}%
-Sauquet, H., Ramírez-Barahona, S., \& Magallón, S. (2021). \emph{The age of flowering plants is unknown}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-schenk2016sec}{}}%
-Schenk, J. J. (2016). {Consequences of secondary calibrations on divergence time estimates}. \emph{PLoS ONE}, \emph{11}(1). \url{https://doi.org/10.1371/journal.pone.0148228}
-
-\leavevmode\vadjust pre{\hypertarget{ref-schliep2011phangorn}{}}%
-Schliep, K. P. (2011). Phangorn: Phylogenetic analysis in r. \emph{Bioinformatics}, \emph{27}(4), 592--593.
-
-\leavevmode\vadjust pre{\hypertarget{ref-selvatti2015paleogene}{}}%
-Selvatti, A.
-P., Gonzaga, L.
-P., \& Moraes Russo, C.
-A.
-de.
-(2015).
-A paleogene origin for crown passerines and the diversification of the oscines in the new world.
-\emph{Molecular Phylogenetics and Evolution}, \emph{88}, 1--15.
-
-\leavevmode\vadjust pre{\hypertarget{ref-smith2018constructing}{}}%
-Smith, Stephen A., \& Brown, J. W. (2018). Constructing a broadly inclusive seed plant phylogeny. \emph{American Journal of Botany}, \emph{105}(3), 302--314.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Smith2012}{}}%
-Smith, Stephen A., \& O'Meara, B.
-C.
-(2012).
-{TreePL: Divergence time estimation using penalized likelihood for large phylogenies}.
-\emph{Bioinformatics}, \emph{28}(20), 2689--2690.
-\url{https://doi.org/10.1093/bioinformatics/bts492}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Stoltzfus2013}{}}%
-Stoltzfus, A., Lapp, H., Matasci, N., Deus, H., Sidlauskas, B., Zmasek, C.
-M., \ldots{} Jordan, G.
-(2013).
-{Phylotastic!
-Making tree-of-life knowledge accessible, reusable and convenient}.
-\emph{{BMC Bioinformatics}}, \emph{14}.
-\url{https://doi.org/10.1186/1471-2105-14-158}
-
-\leavevmode\vadjust pre{\hypertarget{ref-tietze2013complete}{}}%
-Tietze, D.
-T., Päckert, M., Martens, J., Lehmann, H., \& Sun, Y.-H.
-(2013).
-Complete phylogeny and historical biogeography of true rosefinches (aves: carpodacus).
-\emph{Zoological Journal of the Linnean Society}, \emph{169}(1), 215--234.
-
-\leavevmode\vadjust pre{\hypertarget{ref-title2016macrophylogenies}{}}%
-Title, P.
-O., \& Rabosky, D.
-L.
-(2016).
-{Do Macrophylogenies Yield Stable Macroevolutionary Inferences?
-An Example from Squamate Reptiles}.
-\emph{Systematic Biology}, syw102.
-\url{https://doi.org/10.1093/sysbio/syw102}
-
-\leavevmode\vadjust pre{\hypertarget{ref-treplin2008molecular}{}}%
-Treplin, S., Siegert, R., Bleidorn, C., Thompson, H.
-S., Fotso, R., \& Tiedemann, R.
-(2008).
-Molecular phylogeny of songbirds (aves: Passeriformes) and the relative utility of common nuclear marker loci.
-\emph{Cladistics}, \emph{24}(3), 328--349.
-
-\leavevmode\vadjust pre{\hypertarget{ref-vos2012nexml}{}}%
-Vos, R.
-A., Balhoff, J.
-P., Caravas, J.
-A., Holder, M.
-T., Lapp, H., Maddison, W.
-P., \ldots{} others.
-(2012).
-NeXML: Rich, extensible, and verifiable representation of comparative data and metadata.
-\emph{Systematic Biology}, \emph{61}(4), 675--689.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Webb2000}{}}%
-Webb, C. (2000). {Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees}. \emph{{The American Naturalist}}, \emph{156}(2), 145--155.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Webb2008}{}}%
-Webb, Campbell O., Ackerly, D.
-D., \& Kembel, S.
-W.
-(2008).
-{Phylocom: Software for the analysis of phylogenetic community structure and trait evolution}.
-\emph{Bioinformatics}, \emph{24}(18), 2098--2100.
-\url{https://doi.org/10.1093/bioinformatics/btn358}
-
-\leavevmode\vadjust pre{\hypertarget{ref-webb2005phylomatic}{}}%
-Webb, Campbell O., \& Donoghue, M. J. (2005). Phylomatic: Tree assembly for applied phylogenetics. \emph{{Molecular Ecology Notes}}, \emph{5}(1), 181--183.
-
-\leavevmode\vadjust pre{\hypertarget{ref-weir2008calibrating}{}}%
-Weir, J., \& Schluter, D. (2008). Calibrating the avian molecular clock. \emph{Molecular Ecology}, \emph{17}(10), 2321--2328.
-
-\end{CSLReferences}
-
-\endgroup
+% in this file I leave the figure captions outside\ caption{} because I want them
+% to be formatted in the same way as the general text (double spaced and linenumbered)
+\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Supplementary Figure}}
 
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    TABLE 1     %%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%    FIGURE CROSS VALIDATION    %%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{table}
-  \caption{Ages of congruified nodes belonging to step C2 from Figure3}
-  \label{tbl:table1}
-  \includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example.pdf}
-\end{table}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    TABLE 2     %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{table}
-  \caption{Summary of congruified nodes ages, corresponding to step C3 from Figure3}
-  \label{tbl:table2}
-  \includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example-summary.png}
-\end{table}
-
-% in this file I leave the figure captions outside\ caption{} because I want them
-% to be formatted in the same way as the general text (double spaced and linenumbered)
-\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 1    %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \begin{figure}[!h]
-\includegraphics{../figures/figure1/figure1-new.pdf}
-\caption{Stylized DateLife workflow.
-This shows the general worflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{http://www.datelife.org/}.
-Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
-}
-\label{fig:workflow}
+\includegraphics{../figures/figure-cross-validation/fig-cross-validation-xy-plots-diffs.pdf}
+\caption{Results from cross validation analysis. Each plot shows the difference between the original age estimated and those obtained with a DateLife analysis, per node.}
+\label{fig:cvXYdiffs}
 \end{figure}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE runtime    %%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%   CROSS VALIDATION CHRONOGRAMS   %%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!h]
-\includegraphics[width=1\linewidth]{../figures/fig_runtime_main.pdf}
-\caption{Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names.
-We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray).
-We also performed a search using the already processed query for comparison (light blue).}
-\label{fig:runtime_main}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 2    %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!h]
-\includegraphics[width=1\linewidth]{../figures/figure2/figure2.pdf}
-\caption{DateLife analysis results for a small sample of A) 6 bird species within the Passeriformes.
-B) Processed species names were found across 9 chronograms within 6 independent studies (Barker et al.
-(\protect\hyperlink{ref-barker2012going}{2012}), Barker et al.
-(\protect\hyperlink{ref-barker2015new}{2015}), Burns et al.
-(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al.
-(\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al.
-(\protect\hyperlink{ref-Jetz2012}{2012}).) C) This revealed 28 source age data points for the queried species names.
-Summarized age data is used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree, resulting in a summary chronogram of source ages.}
-\label{fig:figure2}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 2-1  %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\begin{figure}[!h]
-%\includegraphics[width=1\linewidth]{../figures/figure2/figure2-1.pdf}
-%\caption{Age data results of a DateLife search of a small sample of 6 %bird species within the Passeriformes.
-Input names were found across 9 %chronograms within 6 independent studies (Barker et al.
-%(\protect\hyperlink{ref-barker2012going}{2012}), Barker et al.
-%(\protect\hyperlink{ref-barker2015new}{2015}), Burns et al.
-%(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al.
-%(\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price %(\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al.
-%(\protect\hyperlink{ref-Jetz2012}{2012}).) This revealed 28 age data %points for the queried species names.}
-%\label{fig:figure2-1}
-%\end{figure}
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 2-2  %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\begin{figure}[!h]
-%\includegraphics{../figures/figure2/figure2-2.pdf}
-%\caption{Summarized age data is used as secondary calibrations to date %a tree topology as a summary chronogram.}
-%\label{fig:summaries}
-%\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 3    %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!h]
-\includegraphics{../figures/fringillidae-topologies/fringillidae-topology.pdf}
-\caption{Tree topologies obtained from Open Tree of Life's (OpenTree) synthetic phylogenetic tree.
-A) Topology of 2,333 tips and 1,305 internal nodes, encompassing bird species within the family Fringillidae following the NCBI taxonomy (black), as well as all other bird species that share the same mrca node in OpenTree's synthetic tree (purple).
-B) Topology of 289 tips and 253 internal nodes, encompassing bird species within the Fringillidae only.
-This tree results from pruning species that are not assigned to the family Fringillidae (purple branches in topology A) but that form a monophyletic group with them.
-Bird species within the Fringillidae are paraphyletic
-(Alström et al. \protect\hyperlink{ref-}{2014},
-Barker, Cibois, Schikler, Feinstein, \& Cracraft \protect\hyperlink{ref-barker2004phylogeny}{2004},
-Barker et al. \protect\hyperlink{ref-barker2013going}{2013},
-Barker \protect\hyperlink{ref-barker2014mitogenomic}{2014},
-Barker et al. \protect\hyperlink{ref-barker2015new}{2015},
-Beresford, Barker, Ryan, \& Crowe \protect\hyperlink{ref-beresford2005african}{2005},
-Bryson Jr et al. \protect\hyperlink{ref-bryson2014diversification}{2014},
-Burleigh, Kimball, \& Braun \protect\hyperlink{ref-burleigh2015building}{2015},
-Burns et al. \protect\hyperlink{ref-burns2014phylogenetics}{2014},
-Chaves, Hidalgo, \& Klicka \protect\hyperlink{ref-chaves2013biogeography}{2013},
-Claramunt \& Cracraft \protect\hyperlink{ref-claramunt2015new}{2015},
-Gibb et al. \protect\hyperlink{ref-gibb2015new}{2015},
-Hackett et al. \protect\hyperlink{ref-hackett2008phylogenomic}{2008},
-Jetz et al. \protect\hyperlink{ref-Jetz2012}{2012},
-Johansson, Fjeldså, \& Bowi \protect\hyperlink{ref-johansson2008phylogenetic}{200},
-Kimball et al. \protect\hyperlink{ref-kimball2019phylogenomic}{2019},
-Klicka et al. \protect\hyperlink{ref-klicka2014comprehensive}{2014},
-Lamichhaney et al. \protect\hyperlink{ref-lamichhaney2015evolution}{2015},
-Lerner, Meyer, James, Hofreiter, \& Fleischer \protect\hyperlink{ref-lerner2011multilocus}{2011},
-Lovette et al. \protect\hyperlink{ref-lovette2010comprehensive}{2010},
-Moyle et al. \protect\hyperlink{ref-moyle2016tectonic}{2016},
-Ödeen, Håstad, \& Alström \protect\hyperlink{ref-odeen2011evolution}{2011},
-Oliveros et al. \protect\hyperlink{ref-oliveros2019earth}{2019},
-Päckert et al. \protect\hyperlink{ref-packert2012horizontal}{2012},
-Parchman, Benkman, \& Mezquida \protect\hyperlink{ref-parchman2007coevolution}{2007},
-Powell et al. \protect\hyperlink{ref-powell2014comprehensive}{2014},
-Price et al. \protect\hyperlink{ref-price2014niche}{2014},
-Pulgarín-R, Smith, Bryson Jr, Spellman, \& Klicka \protect\hyperlink{ref-pulgarin2013multilocus}{2013},
-Selvatti, Gonzaga, \& Moraes Russo \protect\hyperlink{ref-selvatti2015paleogene}{2015},
-Tietze, Päckert, Martens, Lehmann, \& Sun \protect\hyperlink{ref-tietze2013complete}{2013},
-Treplin et al. \protect\hyperlink{ref-treplin2008molecular}{2008},
-Zuccon, Prŷs-Jones, Rasmussen, \& Ericson \protect\hyperlink{ref-zuccon2012phylogenetic}{2012}).
-}
-\label{fig:fringillidae-topologies}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 4    %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!h]
-\includegraphics{../figures/figure-fringillidae/median_and_calibration_ages-congruified.png}
-\caption{Fringillidae median summary chronogram generated with DateLife. It has 256 tips and 233 nodes.}
-\label{fig:fringillidages}
-\end{figure}
-
-% in this file I leave the figure captions outside\ caption{} because I want them
-% to be formatted in the same way as the general text (double spaced and linenumbered)
-\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/fig-cross-validation-xy-plots.pdf}
-\caption{Results from cross validation analysis.}
-\label{fig:cvXY}
-\end{figure}
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/fig-cross-validation-xy-plots-diffs.pdf}
-\caption{Results from cross validation analysis.}
-\label{fig:cvXYdiffs}
-\end{figure}
 
@@
 \includegraphics{../figures/figure-cross-validation/cross_validation_2.png}
-\caption{Cross validation of second source chronogram.
+\caption{Cross validation of Barker et al.
+(\protect\hyperlink{ref-barker2015new}{2015}) chronogram 1.
 The chronogram shown in black corresponds to the dates published in the original study.

### 2022-06-21 manuscript/fig-benchmark.Rmd
adding option to embed figures in ms

---
title: "fig2"
output: html_document
date: '2022-06-21'
---

\begin{center}
\includegraphics[width= 6.5in]{../figures/fig_runtime_main.pdf}
\end{center}
\captionof{figure}{Computation time used to process a datelife query and a datelife search across \texttt{datelife}'s chronogram database, relative to number of input taxon names. For each N = {10, 100, 200, ..., 1 000, ... , 9 000, 10 000}, we sampled N species names from the class Aves a hundred times, and then performed a datelife search with input processing, without implementing Taxon Names Resolution Service (TNRS; dark gray), and implementing TNRS (light gray). We also performed a search using a query that had been pre-processed with TNRS, for comparison (light blue).}
\label{fig:runtime_main}

### 2022-06-21 manuscript/fig-small-example.Rmd
adding option to embed figures in ms

--- 2022-05-28 manuscript/figures-alt.tex
@@
-% in this file I leave the figure captions outside\ caption{} because I want them
-% to be formatted in the same way as the general text (double spaced and linenumbered)
-\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}
+---
+title: "fig3"
+output: html_document
+date: '2022-06-21'
+---
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 1    %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!h]
-\includegraphics{../figures/figure1/figure1-new.pdf}
-\caption{Stylized DateLife workflow.
-This shows the general worflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{http://www.datelife.org/}.
-Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
-}
-\label{fig:workflow}
-\end{figure}
+\vspace{10mm}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE runtime    %%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!h]
-\includegraphics[width=1\linewidth]{../figures/fig_runtime_main.pdf}
-\caption{Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names.
-We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray).
-We also performed a search using the already processed query for comparison (light blue).}
-\label{fig:runtime_main}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 2    %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!h]
-\includegraphics[width=1\linewidth]{../figures/figure2/figure2.pdf}
-\caption{DateLife analysis results for a small sample of A) 6 bird species within the Passeriformes.
+\begin{center}
+\includegraphics[width = 6.5in]{../figures/figure2/figure2.pdf}
+\end{center}
+\captionof{figure}{DateLife analysis results for a small sample of A) 6 bird species within the Passeriformes.
 B) Processed species names were found across 9 chronograms within 6 independent studies (Barker et al.
@@
 \label{fig:figure2}
-\end{figure}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 2-1  %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\begin{figure}[!h]
-%\includegraphics[width=1\linewidth]{../figures/figure2/figure2-1.pdf}
-%\caption{Age data results of a DateLife search of a small sample of 6 %bird species within the Passeriformes.
-Input names were found across 9 %chronograms within 6 independent studies (Barker et al.
-%(\protect\hyperlink{ref-barker2012going}{2012}), Barker et al.
-%(\protect\hyperlink{ref-barker2015new}{2015}), Burns et al.
-%(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al.
-%(\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price %(\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al.
-%(\protect\hyperlink{ref-Jetz2012}{2012}).) This revealed 28 age data %points for the queried species names.}
-%\label{fig:figure2-1}
-%\end{figure}
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 2-2  %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\begin{figure}[!h]
-%\includegraphics{../figures/figure2/figure2-2.pdf}
-%\caption{Summarized age data is used as secondary calibrations to date %a tree topology as a summary chronogram.}
-%\label{fig:summaries}
-%\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 3    %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!h]
-\includegraphics{../figures/fringillidae-topologies/fringillidae-topology.pdf}
-\caption{Tree topologies obtained from Open Tree of Life's (OpenTree) synthetic phylogenetic tree.
-A) Topology of 2,333 tips and 1,305 internal nodes, encompassing bird species within the family Fringillidae following the NCBI taxonomy (black), as well as all other bird species that share the same mrca node in OpenTree's synthetic tree (purple).
-B) Topology of 289 tips and 253 internal nodes, encompassing bird species within the Fringillidae only, resulting from pruning purple branches from topology A.
-Bird species within the Fringillidae do not form a monophyletic group
-(Alström et al. \protect\hyperlink{ref-}{2014},
-Barker, Cibois, Schikler, Feinstein, \& Cracraft \protect\hyperlink{ref-barker2004phylogeny}{2004},
-Barker et al. \protect\hyperlink{ref-barker2013going}{2013},
-Barker \protect\hyperlink{ref-barker2014mitogenomic}{2014},
-Barker et al. \protect\hyperlink{ref-barker2015new}{2015},
-Beresford, Barker, Ryan, \& Crowe \protect\hyperlink{ref-beresford2005african}{2005},
-Bryson Jr et al. \protect\hyperlink{ref-bryson2014diversification}{2014},
-Burleigh, Kimball, \& Braun \protect\hyperlink{ref-burleigh2015building}{2015},
-Burns et al. \protect\hyperlink{ref-burns2014phylogenetics}{2014},
-Chaves, Hidalgo, \& Klicka \protect\hyperlink{ref-chaves2013biogeography}{2013},
-Claramunt \& Cracraft \protect\hyperlink{ref-claramunt2015new}{2015},
-Gibb et al. \protect\hyperlink{ref-gibb2015new}{2015},
-Hackett et al. \protect\hyperlink{ref-hackett2008phylogenomic}{2008},
-Jetz et al. \protect\hyperlink{ref-Jetz2012}{2012},
-Johansson, Fjeldså, \& Bowi \protect\hyperlink{ref-johansson2008phylogenetic}{200},
-Kimball et al. \protect\hyperlink{ref-kimball2019phylogenomic}{2019},
-Klicka et al. \protect\hyperlink{ref-klicka2014comprehensive}{2014},
-Lamichhaney et al. \protect\hyperlink{ref-lamichhaney2015evolution}{2015},
-Lerner, Meyer, James, Hofreiter, \& Fleischer \protect\hyperlink{ref-lerner2011multilocus}{2011},
-Lovette et al. \protect\hyperlink{ref-lovette2010comprehensive}{2010},
-Moyle et al. \protect\hyperlink{ref-moyle2016tectonic}{2016},
-Ödeen, Håstad, \& Alström \protect\hyperlink{ref-odeen2011evolution}{2011},
-Oliveros et al. \protect\hyperlink{ref-oliveros2019earth}{2019},
-Päckert et al. \protect\hyperlink{ref-packert2012horizontal}{2012},
-Parchman, Benkman, \& Mezquida \protect\hyperlink{ref-parchman2007coevolution}{2007},
-Powell et al. \protect\hyperlink{ref-powell2014comprehensive}{2014},
-Price et al. \protect\hyperlink{ref-price2014niche}{2014},
-Pulgarín-R, Smith, Bryson Jr, Spellman, \& Klicka \protect\hyperlink{ref-pulgarin2013multilocus}{2013},
-Selvatti, Gonzaga, \& Moraes Russo \protect\hyperlink{ref-selvatti2015paleogene}{2015},
-Tietze, Päckert, Martens, Lehmann, \& Sun \protect\hyperlink{ref-tietze2013complete}{2013},
-Treplin et al. \protect\hyperlink{ref-treplin2008molecular}{2008},
-Zuccon, Prŷs-Jones, Rasmussen, \& Ericson \protect\hyperlink{ref-zuccon2012phylogenetic}{2012}).
-}
-\label{fig:fringillidae-topologies}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 4    %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!h]
-\includegraphics{../figures/figure-fringillidae/median_and_calibration_ages_simple.png}
-\caption{Fringillidae median summary chronogram generated with DateLife. It has 256 tips and 233 nodes.}
-\label{fig:fringillidages}
-\end{figure}
+\vspace{15mm}

### 2022-06-21 manuscript/fig-topologies.Rmd
adding option to embed figures in ms

--- 2022-05-31 manuscript/figures-alt.tex
@@
-% in this file I leave the figure captions outside\ caption{} because I want them
-% to be formatted in the same way as the general text (double spaced and linenumbered)
-\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}
+---
+title: "fig1"
+output: html_document
+date: '2022-06-21'
+---
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 1    %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!h]
-\includegraphics{../figures/figure1/figure1-new.pdf}
-\caption{Stylized DateLife workflow.
-This shows the general worflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{http://www.datelife.org/}.
-Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
-}
-\label{fig:workflow}
-\end{figure}
+\vspace{20mm}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE runtime    %%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!h]
-\includegraphics[width=1\linewidth]{../figures/fig_runtime_main.pdf}
-\caption{Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names.
-We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray).
-We also performed a search using the already processed query for comparison (light blue).}
-\label{fig:runtime_main}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 2    %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!h]
-\includegraphics[width=1\linewidth]{../figures/figure2/figure2.pdf}
-\caption{DateLife analysis results for a small sample of A) 6 bird species within the Passeriformes.
-B) Processed species names were found across 9 chronograms within 6 independent studies (Barker et al.
-(\protect\hyperlink{ref-barker2012going}{2012}), Barker et al.
-(\protect\hyperlink{ref-barker2015new}{2015}), Burns et al.
-(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al.
-(\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al.
-(\protect\hyperlink{ref-Jetz2012}{2012}).) C) This revealed 28 source age data points for the queried species names.
-Summarized age data is used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree, resulting in a summary chronogram of source ages.}
-\label{fig:figure2}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 2-1  %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\begin{figure}[!h]
-%\includegraphics[width=1\linewidth]{../figures/figure2/figure2-1.pdf}
-%\caption{Age data results of a DateLife search of a small sample of 6 %bird species within the Passeriformes.
-Input names were found across 9 %chronograms within 6 independent studies (Barker et al.
-%(\protect\hyperlink{ref-barker2012going}{2012}), Barker et al.
-%(\protect\hyperlink{ref-barker2015new}{2015}), Burns et al.
-%(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al.
-%(\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price %(\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al.
-%(\protect\hyperlink{ref-Jetz2012}{2012}).) This revealed 28 age data %points for the queried species names.}
-%\label{fig:figure2-1}
-%\end{figure}
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 2-2  %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\begin{figure}[!h]
-%\includegraphics{../figures/figure2/figure2-2.pdf}
-%\caption{Summarized age data is used as secondary calibrations to date %a tree topology as a summary chronogram.}
-%\label{fig:summaries}
-%\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 3    %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!h]
-\includegraphics{../figures/fringillidae-topologies/fringillidae-topology.pdf}
-\caption{Tree topologies obtained from Open Tree of Life's (OpenTree) synthetic phylogenetic tree.
+\begin{center}
+\includegraphics[width = 6.5in]{../figures/fringillidae-topologies/fringillidae-topology.pdf}
+\captionof{figure}{Tree topologies obtained from Open Tree of Life's (OpenTree) synthetic phylogenetic tree.
 A) Topology of 2,333 tips and 1,305 internal nodes, encompassing bird species within the family Fringillidae following the NCBI taxonomy (black), as well as all other bird species that share the same mrca node in OpenTree's synthetic tree (purple).
@@
 Bird species within the Fringillidae are paraphyletic
-(Alström et al. \protect\hyperlink{ref-}{2014},
+(Alström et al. \protect\hyperlink{ref-alstrom2014discovery}{2014},
 Barker, Cibois, Schikler, Feinstein, \& Cracraft \protect\hyperlink{ref-barker2004phylogeny}{2004},
@@
 \label{fig:fringillidae-topologies}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 4    %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!h]
-\includegraphics{../figures/figure-fringillidae/median_and_calibration_ages_simple.png}
-\caption{Fringillidae median summary chronogram generated with DateLife. It has 256 tips and 233 nodes.}
-\label{fig:fringillidages}
-\end{figure}
+\end{center}

### 2022-06-21 manuscript/fig1.Rmd
adding option to embed figures in ms

---
title: "fig1"
output: html_document
date: '2022-06-21'
---
\begin{center}
\includegraphics[width = 6.5in]{../figures/figure1/figure1-new.pdf}
\captionof{figure}{Stylized DateLife workflow. This shows the general workflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website at \url{http://www.datelife.org/}. Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.}
\label{fig:workflow}
\end{center}

### 2022-06-21 manuscript/manuscript.Rmd
adding option to embed figures in ms

@@
       in_header: "preamble.tex"
-      after_body  : ["tables.tex",
-                     "figures-alt.tex"]
+      # after_body  : ["tables.tex",
+      #                "figures-alt.tex"]
 
@@
 r_refs("paper_references.bib")
+embed_figure <- TRUE
 ```
@@
 
-Here we present the DateLife project which has the main goal of capturing age data from published chronograms, and making these data readily accessible to the community for reuse and reanalysis.
+Here we present the DateLife project which has the main goal of capturing age data from published chronograms, and making these data readily accessible to the community for reuse and reanalysis, for research, teaching, and science communication and policy.
   <!--  R has become mainstream in the biological-data analyst toolkit.
@@
 <!--_**Processing an input.--**_ -->
+
+```{r fig1, child= if (embed_figure) "fig1.Rmd"}
+```
 
@@
 
+```{r fig2, child= if (embed_figure) "fig-benchmark.Rmd"}
+```
+
 # Benchmark
@@
 -->
+\vspace{5mm}
 
@@
 
+\vspace{5mm}
+
 _**Summarizing search results.-**_
@@
 Summary age data for node "n2" were excluded as final calibration because they are older than age data of the more inclusive node, "n1" (Fig. \ref{fig:figure2}C4).
+
+```{r fig-small-example, child= if (embed_figure) "fig-small-example.Rmd"}
+```
+
+```{r table1, child= if (embed_figure) "table1.Rmd"}
+```
+
+```{r table2, child= if (embed_figure) "table2.Rmd"}
+```
 
@@
 
+\vspace{5mm}
+
 _**Searching the database.-**_
@@
 Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices, revealing 1206 different age data points available for species within the Fringillidae (supplementray table 1).
+
+\vspace{5mm}
 
@@
 According to this phylogeny, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).
+
 Age data from source chronograms was congruified, decreasing to 818 different age data points available (supplementray table 2) and then age summary statistics per congruified node are obtained and used as fixed secondary calibrations over a consensus tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
@@
 <!-- and hierarchical clustering method which appears to push ages back -->
+
+```{r fig4, child= if (embed_figure) "fig-topologies.Rmd"}
+```
 

### 2022-06-21 manuscript/manuscript.tex
adding option to embed figures in ms

--- 2022-06-06 manuscript/manuscript.tex
@@
 \documentclass[
-  english,
   man]{apa6}
-\title{DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life}
-\author{Luna L. Sánchez Reyes\textsuperscript{1,2}, Emily Jane McTavish\textsuperscript{1}, \& Brian O'Meara\textsuperscript{2}}
-\date{}
-
 \usepackage{amsmath,amssymb}
@@
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+\ifLuaTeX
+\usepackage[bidi=basic]{babel}
+\else
+\usepackage[bidi=default]{babel}
+\fi
+\babelprovide[main,import]{english}
+% get rid of language-specific shorthands (see #6817):
+\let\LanguageShortHands\languageshorthands
+\def\languageshorthands#1{}
 \DeclareUnicodeCharacter{0301}{*************************************}
+
+% leave the figure captions outside\ caption{} because I want them
+% to be formatted in the same way as the general text (double spaced and linenumbered)
+
+\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}
+\usepackage{caption}
+\usepackage{float}
 % Manuscript styling
@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4890}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4906}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 \usepackage{csquotes}
-\ifXeTeX
-  % Load polyglossia as late as possible: uses bidi with RTL langages (e.g. Hebrew, Arabic)
-  \usepackage{polyglossia}
-  \setmainlanguage[]{english}
-\else
-  \usepackage[main=english]{babel}
-% get rid of language-specific shorthands (see #6817):
-\let\LanguageShortHands\languageshorthands
-\def\languageshorthands#1{}
-\fi
 \ifLuaTeX
@@
 \fi
+
+\title{DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life}
+\author{Luna L. Sánchez Reyes\textsuperscript{1,2}, Emily Jane McTavish\textsuperscript{1}, \& Brian O'Meara\textsuperscript{2}}
+\date{}
 
@@
 
-Here we present the DateLife project which has the main goal of capturing age data from published chronograms, and making these data readily accessible to the community for reuse and reanalysis.
+Here we present the DateLife project which has the main goal of capturing age data from published chronograms, and making these data readily accessible to the community for reuse and reanalysis, for research, teaching, and science communication and policy.
 DateLife's core software application is available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org.
@@
 DateLife's core software application consists of the R package \texttt{datelife}.
-Its latest stable version -- v0.6.2, is available from the CRAN repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
+Its latest stable version -- v0.6.4, is available from the CRAN repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
 ape (Paradis, Claude, \& Strimmer, 2004),
@@
 
+\begin{center}
+\includegraphics[width = 6.5in]{../figures/figure1/figure1-new.pdf}
+\captionof{figure}{Stylized DateLife workflow.
+This shows the general workflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website at \url{http://www.datelife.org/}.
+Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.}
+\label{fig:workflow}
+\end{center}
+
 \hypertarget{creating-a-search-query}{%
@@
 At the time of writing of this manuscript
-,
+(Jun 20, 2022),
 DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies.
@@
 
+\begin{center}
+\includegraphics[width= 6.5in]{../figures/fig_runtime_main.pdf}
+\end{center}
+\captionof{figure}{Computation time used to process a datelife query and a datelife search across \texttt{datelife}'s chronogram database, relative to number of input taxon names.
+For each N = {10, 100, 200, ..., 1 000, ...
+, 9 000, 10 000}, we sampled N species names from the class Aves a hundred times, and then performed a datelife search with input processing, without implementing Taxon Names Resolution Service (TNRS; dark gray), and implementing TNRS (light gray).
+We also performed a search using a query that had been pre-processed with TNRS, for comparison (light blue).}
+
+\label{fig:runtime_main}
+
 \hypertarget{benchmark}{%
@@
 Discovering this synonym allowed assigning five age data points for the parent node of \emph{Emberiza elegans}, shown as \emph{Schoeniclus elegans} in figure \ref{fig:figure2}A, which would not have had any data otherwise.
+\vspace{5mm}
 
@@
 Together, source chronograms provide 28 unique age data points, covering all nodes on our chosen tree topology to date (Table \ref{tbl:table1}).
+
+\vspace{5mm}
 
@@
 
-\hypertarget{the-family-fringillidae}{%
-\subsection{The family Fringillidae}\label{the-family-fringillidae}}
-
-\emph{\textbf{Creating a query.-}}
-To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the ``get species from taxon'' flag,
-which gets all recognized species names within a named group from a taxonomy of choice.
-Following the NCBI taxonomy, our DateLife query has 289 Fringillidae species.
-This taxon-constrained approach implies that the final results of a full DateLife analysis will be done using a tree topology and ages for the species in a named group, which do not necessarily correspond to a monophyletic group.
-Users can change this behaviour by providing a monophyletic tree as input for a DateLife search, or as a tree topology for a DateLife summary.
-
-\emph{\textbf{Searching the database.-}}
-Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species.
-The DateLife search identified 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Price et al., 2014).
-Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices, revealing 1206 different age data points available for species within the Fringillidae (supplementray table 1).
-
-\emph{\textbf{Summarizing search results.-}}
-The final step is to congruify and summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM.
-As explained in the ``Description'' section, a tree topology to summarize age data upon is required.
-By default, DateLife uses as summarizing topology the portion of OpenTree synthetic phylogeny that contains the species provided as search query.
-According to this phylogeny, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).
-Age data from source chronograms was congruified, decreasing to 818 different age data points available (supplementray table 2) and then age summary statistics per congruified node are obtained and used as fixed secondary calibrations over a consensus tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
-\ref{fig:fringillidages}).
-Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
-
-\hypertarget{cross-validation-test}{%
-\section{Cross-validation test}\label{cross-validation-test}}
-
-We performed a cross validation analysis of the DateLife workflow using the Fringillidae chronograms.
-We used the individual tree topologies from each of the 19 source chronograms from 13 studies as inputs, treating their node ages as unknown.
-We then estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.
-
-We found that node ages from original study, and ages estimated using all other age data available are correlated (Supplementary Fig. \ref{fig:cvXY}).
-
-For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka `closer to the root') relative to the orginal estimate, and overestimate ages for nodes closer to the tips.
-Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. \ref{fig:cvXYdiffs}).
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-\texttt{datelife} makes state-of-the-art data on evolutionary time frame easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
-It is an open service that does not require any expert biological knowledge from users --besides the names of the species or group they want to work with, for any of its functionality.
-
-A total of 99474 unique terminal taxa are represented in \texttt{datelife}'s database.
-Incorporation of more chronograms into \texttt{datelife}'s database will continue to improve its services.
-One option to increase the number of chronograms in the DateLife database is the Dryad data repository.
-Methods to automatically mine chronograms from Dryad could be designed and implemented.
-However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined manually by a human curator.
-We would like to emphasize on the importance of sharing chronogram data, including systematically curated metadata, into open repositories, such as OpenTree's Phylesystem (McTavish et al., 2015) for the benefit of the scientific community as a whole.
-
-\hypertarget{primary-vs-secondary-calibrations-in-dating-analyses}{%
-\subsection{Primary vs Secondary calibrations in dating analyses}\label{primary-vs-secondary-calibrations-in-dating-analyses}}
-
-While most chronograms in DateLife's database are constructed using primary calibrations (molecular substitution rates or ages obtained from the fossil record or geological events), DateLife summarizes chronograms using secondary calibrations only (ages coming from other chronograms).
-
-Graur and Martin (2004) cautioned on the increased error and uncertainty in estimated ages when using secondary calibrations in dating analyses.
-Schenk (2016) showed that, in simulations, divergence times inferred using secondary calibrations are significantly younger than those inferred with primary calibrations, when obtained with Bayesian inference methods, and when priors are implemented in similar ways in both analyses.
-Accordingly, the scientific community seems to have more confidence in chronograms obtained from a single analysis, using fossil data as primary sources of calibrations (Schenk, 2016), and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data reflect a coherent evolutionary history (Sauquet, 2013), as for example done by Antonelli et al.
-(2017).
-There have been attempts to create fossil calibration databases (Ksepka et al., 2015), though these still have room to grow.
-
-It seems that using several (as opposed to just a few) secondary calibrations can provide sufficient information to alleviate or even neutralize potential biases (Sauquet, 2013).
-Certainly, further studies are required to fully understand the effect of secondary calibrations on outputs from different tree dating methods, and on downstream analyses.
-It is possible that secondary calibrations can be safely used with dating methods that do not require setting priors, such as penalized likelihood (Michael J.
-Sanderson, 2003), with methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ (Campbell O.
-Webb et al., 2008; Campbell O.
-Webb \& Donoghue, 2005), or methods that summarize age data unto a tree topology.
-
-Our cross validation analysis might provide some insight in this regard.
-When ages are estimated with secondary calibrations, nodes closer to the root do tend to be slightly younger than ages estimated with primary calibrations.
-However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures 1-9), with a notable exception of cross validation of Burns et al.
-(2014) chronogram, which results in much younger node ages when estimated using secondary calibrations (Supplementary figs.
-1 and 4).
-
-\hypertarget{variation-in-source-chronograms}{%
-\subsection{Variation in source chronograms}\label{variation-in-source-chronograms}}
-
-Conflict among alternative dating studies is common in the literature
-(see, for example, the robust ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández, 2015; Ramshaw et al., 1972; Michael J.
-Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021)).
-
-Source chronograms may have been estimated using different types of primary calibrations (fossil record, geologic or biogeographic events, molecular substitution rates).
-These differences in approach can deepen the already substantial variation in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
-For example, the chronograms from Burns et al.
-(2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older date estimates for the same nodes than chronograms that were inferred using fossils as primary calibrations (figs.
-fig:fringillidages and fig:cvXY; supplementary figs.
-1 and 4).
-
-Different chronograms might also imply fundamentally distinct evolutionary hypotheses.
-For example, two different researchers working on the same clade could both carefully select and argue for their choices of fossil calibrations.
-Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses the same one to calibrate outside the clade, the resulting age estimates will often differ substantially, as the placement of calibrations as stem or crown group is proved to deeply affect estimated times of lineage divergence (Sauquet, 2013).
-
-\hypertarget{sumarizing-chronograms}{%
-\subsection{Sumarizing chronograms}\label{sumarizing-chronograms}}
-
-By default, DateLife currently summarizes all source chronograms that overlap with at least two species names. Users can exclude source chronograms if they have reasons to do so.
-Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely.
-To our knowledge, there are no tested measures to determine independently when a chronogram is better than another.
-Yet, several characteristics of the data used for dating analyses, as well as from the output chronogram itself, could be used to score the quality of source chronograms.
-
-Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used Magallón et al. (2015).
-Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
-
-DateLife provides an opportunity to capture concordance and conflict among date estimates, which can also be used as a metric for chronogram reliability.
-Its open database of chronograms allows other researchers to do such analyses themselves reproducibly, and without needing permission.
-Though, of course, they should follow proper citation practices, especially for the source chronogram studies.
-
-The exercise of summarizing age data from across multiple studies provides the opportunity to work with a more inclusive chronogram,
-that reflects a unified evolutionary history for a lineage, by putting together evidence from different hypotheses.
-The largest, and taxonomically broadest chronogram currently available from OpenTree was constructed summarizing age data from 2,274 published chronograms using NCBI's taxonomic tree as backbone (Hedges et al., 2015).
-A summarizing exercise may also amplify the effect of uncertainty and errors in source data, and blur parts of the evolutionary history of a lineage that might only be reflected in source chronograms and lost on the summary chronogram (Sauquet et al., 2021).
-
-\hypertarget{effects-on-downstream-analyses}{%
-\subsection{Effects on downstream analyses}\label{effects-on-downstream-analyses}}
-
-For downstream analyses, using alternative chronogram may deeply affect our inferences (Title \& Rabosky, 2016), particularly when studying phenomena dependent on the timing of species diversification events, such as macroevolutionary processes.
-
-In ecology and conservation biology, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance (Campbell O.
-Webb et al., 2008).
-Hence, DateLife's workflow features different ways of estimating node ages in the absence of calibrations and branch length information for certain taxa.
-``Making up'' branch lengths is a common practice in scientific publications: Jetz et al.
-(2012), created a time-calibrated tree of all 9,993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
-(2018) created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37\% had molecular data; Stephen A.
-Smith and Brown (2018) constructed a tree of 353,185 seed plants where only 23\% had molecular data.
-
-Notably, risks come with this practice.
-Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration, using polytomy resolution and branch estimation methods.
-There has yet to be a thorough analysis of what can go wrong when one extends inferences beyond the data in this way, so we urge caution; we also urge readers to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, and assessing the consistency of the results with prior work.
-
-\hypertarget{conclusions}{%
-\section{Conclusions}\label{conclusions}}
-
-Knowledge of the evolutionary time frame of organisms is key to many research areas: trait evolution,
-species diversification, biogeography, macroecology and more. It is also crucial for education, science communication and policy, but generating chronograms is difficult,
-especially for those who want to use phylogenies but who are not systematists, or
-do not have the time to acquire and develop the necessary knowledge and skills to construct them on their own.
-Importantly, years of primarily public funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
-
-The DateLife project allows for easy and fast summary of public and state-of-the-art data on time of lineage divergence.
-It provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identification of regions that require more research, or that have conflicting information.
-It is available as an R package, and as a web-based R shiny application at www.datelife.org
-Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
-The DateLife project helps with awareness of the existing variation in expert time of divergence data, and will foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
-
-\hypertarget{availability}{%
-\section{Availability}\label{availability}}
-
-The \texttt{datelife} software is free and open source and it can be used through its R shiny web application at
-\url{http://www.datelife.org}, through the \texttt{datelife} R package, and through Phylotastic's project web portal \url{https://phylo.cs.nmsu.edu/}.
-DateLife's web application is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker and Open Tree of Life's infrastructure (dates.opentreeoflife.org/datelife).
-\texttt{datelife}'s R package stable version is available
-for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
-using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R. Development versions
-are available from the GitHub repository (\url{https://github.com/phylotastic/datelife})
-and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.
-
-\hypertarget{supplementary-material}{%
-\section{Supplementary Material}\label{supplementary-material}}
-
-Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities are available at \href{https://github.com/LunaSare/datelifeMS1}{datelifeMS1}, \href{https://github.com/LunaSare/datelife_examples}{datelife\_examples}, and \href{https://github.com/LunaSare/datelife_benchmark}{datelife\_benchmark} repositories in LLSR's GitHub account.
-
-\hypertarget{funding}{%
-\section{Funding}\label{funding}}
-
-Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to the Datelife project; DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), ABI-1458572 to the Phylotastic project, and ABI-1759846 to the Open Tree of Life project.
-
-\hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
-
-The DateLife project was born as a prototype tool aiming to provide these services, and was initially developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA (Stoltzfus et al., 2013).
-We thank colleagues from the O'Meara Lab at the University
-of Tennesse Knoxville for suggestions, discussions and software testing.
-The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
-that led to initial work on this project.
-The team that assembled DateLife's first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
-The Open Tree of Life project that provides the open, metadata rich repository of
-trees used to construct DateLife's chronogram database.
-The many scientists who publish their chronograms in an open, reusable form, and the scientists who curate them for deposition in the Open Tree of Life repository.
-The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.
-
-\newpage
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-alstrom2014discovery}{}}%
-Alström, P., Hooper, D.
-M., Liu, Y., Olsson, U., Mohan, D., Gelang, M., \ldots{} Price, T.
-D.
-(2014).
-Discovery of a relict lineage and monotypic family of passerine birds.
-\emph{Biology Letters}, \emph{10}(3), 20131067.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ane2009groves}{}}%
-Ané, C., Eulenstein, O., Piaggio-Talice, R., \& Sanderson, M. J. (2009). Groves of phylogenetic trees. \emph{{Annals of Combinatorics}}, \emph{13}(2), 139--167.
-
-\leavevmode\vadjust pre{\hypertarget{ref-antonelli2017supersmart}{}}%
-Antonelli, A., Hettling, H., Condamine, F.
-L., Vos, K., Nilsson, R.
-H., Sanderson, M.
-J., \ldots{} Vos, R.
-A.
-(2017).
-{Toward a self-updating platform for estimating rates of speciation and migration, ages, and relationships of Taxa}.
-\emph{Systematic Biology}, \emph{66}(2), 153--166.
-\url{https://doi.org/10.1093/sysbio/syw066}
-
-\leavevmode\vadjust pre{\hypertarget{ref-archie1986newick}{}}%
-Archie, J., Day, W.
-H., Felsenstein, J., Maddison, W., Meacham, C., Rohlf, F.
-J., \& Swofford, D.
-(1986).
-{The Newick tree format}.
-Retrieved from \url{\%7Bhttps://evolution.genetics.washington.edu/phylip/newicktree.html\%7D}
-
-\leavevmode\vadjust pre{\hypertarget{ref-avibase-emberiza}{}}%
-Avibase.
-(2022).
-{Yellow-throated Bunting}.
-\emph{{Avibase - The World Bird Database}}, (Online Resource).
-Retrieved from \url{\%7Bhttps://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\%7D}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Bapst2012a}{}}%
-Bapst, D.
-W.
-(2012).
-{Paleotree: An R package for paleontological and phylogenetic analyses of evolution}.
-\emph{{Methods in Ecology and Evolution}}, \emph{3}(5), 803--807.
-\url{https://doi.org/10.1111/j.2041-210X.2012.00223.x}
-
-\leavevmode\vadjust pre{\hypertarget{ref-barba2018constraining}{}}%
-Barba-Montoya, J., Reis, M.
-dos, Schneider, H., Donoghue, P.
-C., \& Yang, Z.
-(2018).
-Constraining uncertainty in the timescale of angiosperm evolution and the veracity of a cretaceous terrestrial revolution.
-\emph{New Phytologist}, \emph{218}(2), 819--834.
-
-\leavevmode\vadjust pre{\hypertarget{ref-barker2014mitogenomic}{}}%
-Barker, F. K. (2014). Mitogenomic data resolve basal relationships among passeriform and passeridan birds. \emph{Molecular Phylogenetics and Evolution}, \emph{79}, 313--324.
-
-\leavevmode\vadjust pre{\hypertarget{ref-barker2013going}{}}%
-Barker, F.
-K., Burns, K.
-J., Klicka, J., Lanyon, S.
-M., \& Lovette, I.
-J.
-(2013).
-Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds.
-\emph{{Systematic Biology}}, \emph{62}(2), 298--320.
-
-\leavevmode\vadjust pre{\hypertarget{ref-barker2015new}{}}%
-Barker, F.
-K., Burns, K.
-J., Klicka, J., Lanyon, S.
-M., \& Lovette, I.
-J.
-(2015).
-New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
-\emph{The Auk: Ornithological Advances}, \emph{132}(2), 333--348.
-
-\leavevmode\vadjust pre{\hypertarget{ref-barker2004phylogeny}{}}%
-Barker, F.
-K., Cibois, A., Schikler, P., Feinstein, J., \& Cracraft, J.
-(2004).
-Phylogeny and diversification of the largest avian radiation.
-\emph{Proceedings of the National Academy of Sciences}, \emph{101}(30), 11040--11045.
-
-\leavevmode\vadjust pre{\hypertarget{ref-beresford2005african}{}}%
-Beresford, P., Barker, F., Ryan, Pg., \& Crowe, T.
-(2005).
-African endemics span the tree of songbirds (passeri): Molecular systematics of several evolutionary {`enigmas.'} \emph{Proceedings of the Royal Society B: Biological Sciences}, \emph{272}(1565), 849--858.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Boyle2013}{}}%
-Boyle, B., Hopkins, N., Lu, Z., Raygoza Garay, J.
-A., Mozzherin, D., Rees, T., \ldots{} Enquist, B.
-J.
-(2013).
-{The taxonomic name resolution service: An online tool for automated standardization of plant names}.
-\emph{{BMC Bioinformatics}}, \emph{14}(1).
-\url{https://doi.org/10.1186/1471-2105-14-16}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Britton2007}{}}%
-Britton, T., Anderson, C.
-L., Jacquet, D., Lundqvist, S., \& Bremer, K.
-(2007).
-{Estimating Divergence Times in Large Phylogenetic Trees}.
-\emph{{Systematic Biology}}, \emph{56}(788777878), 741--752.
-\url{https://doi.org/10.1080/10635150701613783}
-
-\leavevmode\vadjust pre{\hypertarget{ref-bryson2014diversification}{}}%
-Bryson Jr, R.
-W., Chaves, J., Smith, B.
-T., Miller, M.
-J., Winker, K., Pérez-Emán, J.
-L., \& Klicka, J.
-(2014).
-Diversification across the new world within the `blue'cardinalids (aves: cardinalidae).
-\emph{Journal of Biogeography}, \emph{41}(3), 587--599.
-
-\leavevmode\vadjust pre{\hypertarget{ref-burleigh2015building}{}}%
-Burleigh, J. G., Kimball, R. T., \& Braun, E. L. (2015). Building the avian tree of life using a large-scale, sparse supermatrix. \emph{Molecular Phylogenetics and Evolution}, \emph{84}, 53--63.
-
-\leavevmode\vadjust pre{\hypertarget{ref-burns2014phylogenetics}{}}%
-Burns, K.
-J., Shultz, A.
-J., Title, P.
-O., Mason, N.
-A., Barker, F.
-K., Klicka, J., \ldots{} Lovette, I.
-J.
-(2014).
-Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds.
-\emph{{Molecular Phylogenetics and Evolution}}, \emph{75}, 41--77.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Chamberlain2013}{}}%
-Chamberlain, S.
-A., \& Szöcs, E.
-(2013).
-{taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}}.
-\emph{{F1000Research}}, \emph{2}(191), 1--29.
-\url{https://doi.org/10.12688/f1000research.2-191.v2}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Chamberlain2018}{}}%
-Chamberlain, S.
-A., Szöcs, E., Foster, Z., Arendsee, Z., Boettiger, C., Ram, K., \ldots{} Li, G.
-(2019).
-\emph{{taxize: Taxonomic information from around the web}}.
-Retrieved from \url{https://github.com/ropensci/taxize}
-
-\leavevmode\vadjust pre{\hypertarget{ref-chaves2013biogeography}{}}%
-Chaves, J.
-A., Hidalgo, J.
-R., \& Klicka, J.
-(2013).
-Biogeography and evolutionary history of the n eotropical genus s altator (a ves: T hraupini).
-\emph{Journal of Biogeography}, \emph{40}(11), 2180--2190.
-
-\leavevmode\vadjust pre{\hypertarget{ref-claramunt2015new}{}}%
-Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{{Science Advances}}, \emph{1}(11), e1501005.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Criscuolo2006}{}}%
-Criscuolo, A., Berry, V., Douzery, E.
-J.
-P., \& Gascuel, O.
-(2006).
-{SDM: A fast distance-based approach for (super)tree building in phylogenomics}.
-\emph{{Systematic Biology}}, \emph{55}(5), 740--755.
-\url{https://doi.org/10.1080/10635150600969872}
-
-\leavevmode\vadjust pre{\hypertarget{ref-delsuc2018phylogenomic}{}}%
-Delsuc, F., Philippe, H., Tsagkogeorga, G., Simion, P., Tilak, M.-K., Turon, X., \ldots{} Douzery, E.
-J.
-(2018).
-A phylogenomic framework and timescale for comparative studies of tunicates.
-\emph{BMC Biology}, \emph{16}(1), 1--14.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Eastman2013}{}}%
-Eastman, J.
-M., Harmon, L.
-J., \& Tank, D.
-C.
-(2013).
-{Congruification: Support for time scaling large phylogenetic trees}.
-\emph{{Methods in Ecology and Evolution}}, \emph{4}(7), 688--691.
-\url{https://doi.org/10.1111/2041-210X.12051}
-
-\leavevmode\vadjust pre{\hypertarget{ref-edgar2004muscle}{}}%
-Edgar, R. C. (2004). MUSCLE: Multiple sequence alignment with high accuracy and high throughput. \emph{Nucleic Acids Research}, \emph{32}(5), 1792--1797.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Felsenstein1985a}{}}%
-Felsenstein, J. (1985). {Phylogenies and the Comparative Method}. \emph{The American Naturalist}, \emph{125}(1), 1--15. Retrieved from \url{http://www.jstor.org/stable/2461605}
-
-\leavevmode\vadjust pre{\hypertarget{ref-gibb2015new}{}}%
-Gibb, G.
-C., England, R., Hartig, G., McLenachan, P.
-A., Taylor Smith, B.
-L., McComish, B.
-J., \ldots{} Penny, D.
-(2015).
-New zealand passerines help clarify the diversification of major songbird lineages during the oligocene.
-\emph{{Genome Biology and Evolution}}, \emph{7}(11), 2983--2995.
-
-\leavevmode\vadjust pre{\hypertarget{ref-graur2004reading}{}}%
-Graur, D., \& Martin, W. (2004). Reading the entrails of chickens: Molecular timescales of evolution and the illusion of precision. \emph{TRENDS in Genetics}, \emph{20}(2), 80--86.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hackett2008phylogenomic}{}}%
-Hackett, S.
-J., Kimball, R.
-T., Reddy, S., Bowie, R.
-C., Braun, E.
-L., Braun, M.
-J., \ldots{} others.
-(2008).
-A phylogenomic study of birds reveals their evolutionary history.
-\emph{Science}, \emph{320}(5884), 1763--1768.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Hedges2015}{}}%
-Hedges, S.
-B., Marin, J., Suleski, M., Paymer, M., \& Kumar, S.
-(2015).
-{Tree of life reveals clock-like speciation and diversification}.
-\emph{{Molecular Biology and Evolution}}, \emph{32}(4), 835--845.
-\url{https://doi.org/10.1093/molbev/msv037}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Heibl2008}{}}%
-Heibl, C. (2008). \emph{PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages.} Retrieved from \url{http://www.christophheibl.de/Rpackages.html}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hooper2017chromosomal}{}}%
-Hooper, D. M., \& Price, T. D. (2017). Chromosomal inversion differences correlate with range overlap in passerine birds. \emph{Nature Ecology \& Evolution}, \emph{1}(10), 1526.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Huelsenbeck2001}{}}%
-Huelsenbeck, J. P., \& Ronquist, F. (2001). {MRBAYES: Bayesian inference of phylogenetic trees}. \emph{Bioinformatics}, \emph{17}(8), 754--755. \url{https://doi.org/10.1093/bioinformatics/17.8.754}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Jetz2012}{}}%
-Jetz, W., Thomas, G., Joy, J.
-J.
-B., Hartmann, K., \& Mooers, A.
-(2012).
-{The global diversity of birds in space and time}.
-\emph{Nature}, \emph{491}(7424), 444--448.
-\url{https://doi.org/10.1038/nature11631}
-
-\leavevmode\vadjust pre{\hypertarget{ref-johansson2008phylogenetic}{}}%
-Johansson, U.
-S., Fjeldså, J., \& Bowie, R.
-C.
-(2008).
-Phylogenetic relationships within passerida (aves: Passeriformes): A review and a new molecular phylogeny based on three nuclear intron markers.
-\emph{Molecular Phylogenetics and Evolution}, \emph{48}(3), 858--876.
-
-\leavevmode\vadjust pre{\hypertarget{ref-katoh2009multiple}{}}%
-Katoh, K., Asimenos, G., \& Toh, H. (2009). Multiple alignment of DNA sequences with MAFFT. In \emph{Bioinformatics for DNA sequence analysis} (pp. 39--64). Springer.
-
-\leavevmode\vadjust pre{\hypertarget{ref-kimball2019phylogenomic}{}}%
-Kimball, R. T., Oliveros, C. H., Wang, N., White, N. D., Barker, F. K., Field, D. J., \ldots{} others. (2019). A phylogenomic supertree of birds. \emph{Diversity}, \emph{11}(7), 109.
-
-\leavevmode\vadjust pre{\hypertarget{ref-klicka2014comprehensive}{}}%
-Klicka, J., Barker, F.
-K., Burns, K.
-J., Lanyon, S.
-M., Lovette, I.
-J., Chaves, J.
-A., \& Bryson Jr, R.
-W.
-(2014).
-A comprehensive multilocus assessment of sparrow (aves: Passerellidae) relationships.
-\emph{Molecular Phylogenetics and Evolution}, \emph{77}, 177--182.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ksepka2015fossil}{}}%
-Ksepka, D.
-T., Parham, J.
-F., Allman, J.
-F., Benton, M.
-J., Carrano, M.
-T., Cranston, K.
-A., \ldots{} others.
-(2015).
-The fossil calibration database---a new resource for divergence dating.
-\emph{Systematic Biology}, \emph{64}(5), 853--859.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lamichhaney2015evolution}{}}%
-Lamichhaney, S., Berglund, J., Almén, M.
-S., Maqbool, K., Grabherr, M., Martinez-Barrio, A., \ldots{} others.
-(2015).
-Evolution of darwin's finches and their beaks revealed by genome sequencing.
-\emph{Nature}, \emph{518}(7539), 371--375.
-
-\leavevmode\vadjust pre{\hypertarget{ref-laubichler2009form}{}}%
-Laubichler, M. D., \& Maienschein, J. (2009). \emph{Form and function in developmental evolution}. Cambridge University Press.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lepage2004avibase}{}}%
-Lepage, D. (2004). \emph{Avibase: The world bird database}. Bird Studies Canada.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lepage2014avibase}{}}%
-Lepage, D., Vaidya, G., \& Guralnick, R. (2014). Avibase--a database system for managing and organizing taxonomic concepts. \emph{ZooKeys}, (420), 117.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lerner2011multilocus}{}}%
-Lerner, H.
-R., Meyer, M., James, H.
-F., Hofreiter, M., \& Fleischer, R.
-C.
-(2011).
-Multilocus resolution of phylogeny and timescale in the extant adaptive radiation of hawaiian honeycreepers.
-\emph{Current Biology}, \emph{21}(21), 1838--1844.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lovette2010comprehensive}{}}%
-Lovette, I.
-J., Pérez-Emán, J.
-L., Sullivan, J.
-P., Banks, R.
-C., Fiorentino, I., Córdoba-Córdoba, S., \ldots{} others.
-(2010).
-A comprehensive multilocus phylogeny for the wood-warblers and a revised classification of the parulidae (aves).
-\emph{Molecular Phylogenetics and Evolution}, \emph{57}(2), 753--770.
-
-\leavevmode\vadjust pre{\hypertarget{ref-magallon2001absolute}{}}%
-Magallon, S., \& Sanderson, M. J. (2001). Absolute diversification rates in angiosperm clades. \emph{Evolution}, \emph{55}(9), 1762--1780.
-
-\leavevmode\vadjust pre{\hypertarget{ref-magallon2010using}{}}%
-Magallón, S. (2010). Using fossils to break long branches in molecular dating: A comparison of relaxed clocks applied to the origin of angiosperms. \emph{Systematic Biology}, \emph{59}(4), 384--399.
-
-\leavevmode\vadjust pre{\hypertarget{ref-magallon2015metacalibrated}{}}%
-Magallón, S., Gómez-Acevedo, S., Sánchez-Reyes, L.
-L., \& Hernández-Hernández, T.
-(2015).
-A metacalibrated time-tree documents the early rise of flowering plant phylogenetic diversity.
-\emph{New Phytologist}, \emph{207}(2), 437--453.
-
-\leavevmode\vadjust pre{\hypertarget{ref-mctavish2015phylesystem}{}}%
-McTavish, E.
-J., Hinchliff, C.
-E., Allman, J.
-F., Brown, J.
-W., Cranston, K.
-A., Holder, M.
-T., \ldots{} Smith, S.
-A.
-(2015).
-Phylesystem: A git-based data store for community-curated phylogenetic estimates.
-\emph{Bioinformatics}, \emph{31}(17), 2794--2800.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Michonneau2016}{}}%
-Michonneau, F., Brown, J.
-W., \& Winter, D.
-J.
-(2016).
-{rotl: an R package to interact with the Open Tree of Life data}.
-\emph{{Methods in Ecology and Evolution}}, \emph{7}(12), 1476--1481.
-\url{https://doi.org/10.1111/2041-210X.12593}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Morlon2014}{}}%
-Morlon, H. (2014). {Phylogenetic approaches for studying diversification.} \emph{{Ecology Letters}}, \emph{17}(4), 508--525. \url{https://doi.org/10.1111/ele.12251}
-
-\leavevmode\vadjust pre{\hypertarget{ref-moyle2016tectonic}{}}%
-Moyle, R.
-G., Oliveros, C.
-H., Andersen, M.
-J., Hosner, P.
-A., Benz, B.
-W., Manthey, J.
-D., \ldots{} Faircloth, B.
-C.
-(2016).
-Tectonic collision and uplift of wallacea triggered the global songbird radiation.
-\emph{Nature Communications}, \emph{7}(1), 1--7.
-
-\leavevmode\vadjust pre{\hypertarget{ref-odeen2011evolution}{}}%
-Ödeen, A., Håstad, O., \& Alström, P. (2011). Evolution of ultraviolet vision in the largest avian radiation-the passerines. \emph{BMC Evolutionary Biology}, \emph{11}(1), 1--8.
-
-\leavevmode\vadjust pre{\hypertarget{ref-oliveros2019earth}{}}%
-Oliveros, C.
-H., Field, D.
-J., Ksepka, D.
-T., Barker, F.
-K., Aleixo, A., Andersen, M.
-J., \ldots{} others.
-(2019).
-Earth history and the passerine superradiation.
-\emph{Proceedings of the National Academy of Sciences}, \emph{116}(16), 7916--7925.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Ooms2018}{}}%
-Ooms, J., \& Chamberlain, S. (2018). \emph{Phylocomr: Interface to 'phylocom'}. Retrieved from \url{https://CRAN.R-project.org/package=phylocomr}
-
-\leavevmode\vadjust pre{\hypertarget{ref-opentreeAPIs}{}}%
-Open Tree Of Life, Redelings, B., Cranston, K.
-A., Allman, J., Holder, M.
-T., \& McTavish, E.
-J.
-(2016).
-{Open Tree of Life APIs v3.0}.
-\emph{{Open Tree of Life Project}}, (Online Resources).
-Retrieved from \url{\%7Bhttps://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\%7D}
-
-\leavevmode\vadjust pre{\hypertarget{ref-opentreeoflife2019synth}{}}%
-Open Tree Of Life, Redelings, B., Sánchez Reyes, L.
-L., Cranston, K.
-A., Allman, J., Holder, M.
-T., \& McTavish, E.
-J.
-(2019).
-Open tree of life synthetic tree v12.3.
-\emph{Zenodo}.
-Retrieved from \url{https://doi.org/10.5281/zenodo.3937742}
-
-\leavevmode\vadjust pre{\hypertarget{ref-packert2012horizontal}{}}%
-Päckert, M., Martens, J., Sun, Y.-H., Severinghaus, L.
-L., Nazarenko, A.
-A., Ting, J., \ldots{} Tietze, D.
-T.
-(2012).
-Horizontal and elevational phylogeographic patterns of himalayan and southeast asian forest passerines (aves: passeriformes).
-\emph{Journal of Biogeography}, \emph{39}(3), 556--573.
-
-\leavevmode\vadjust pre{\hypertarget{ref-paradis2004}{}}%
-Paradis, E., Claude, J., \& Strimmer, K. (2004). {APE: analyses of phylogenetics and evolution in R language}. \emph{Bioinformatics}, \emph{20}(2), 289--290.
-
-\leavevmode\vadjust pre{\hypertarget{ref-parchman2007coevolution}{}}%
-Parchman, T.
-L., Benkman, C.
-W., \& Mezquida, E.
-T.
-(2007).
-Coevolution between hispaniolan crossbills and pine: Does more time allow for greater phenotypic escalation at lower latitude?
-\emph{Evolution}, \emph{61}(9), 2142--2153.
-
-\leavevmode\vadjust pre{\hypertarget{ref-pennell2014geiger}{}}%
-Pennell, M.
-W., Eastman, J.
-M., Slater, G.
-J., Brown, J.
-W., Uyeda, J.
-C., FitzJohn, R.
-G., \ldots{} Harmon, L.
-J.
-(2014).
-Geiger v2.
-0: An expanded suite of methods for fitting macroevolutionary models to phylogenetic trees.
-\emph{Bioinformatics}, \emph{30}(15), 2216--2218.
-
-\leavevmode\vadjust pre{\hypertarget{ref-posadas2006historical}{}}%
-Posadas, P., Crisci, J. V., \& Katinas, L. (2006). Historical biogeography: A review of its basic concepts and critical issues. \emph{{Journal of Arid Environments}}, \emph{66}(3), 389--403.
-
-\leavevmode\vadjust pre{\hypertarget{ref-powell2014comprehensive}{}}%
-Powell, A.
-F., Barker, F.
-K., Lanyon, S.
-M., Burns, K.
-J., Klicka, J., \& Lovette, I.
-J.
-(2014).
-A comprehensive species-level molecular phylogeny of the new world blackbirds (icteridae).
-\emph{Molecular Phylogenetics and Evolution}, \emph{71}, 94--112.
-
-\leavevmode\vadjust pre{\hypertarget{ref-price2014niche}{}}%
-Price, T.
-D., Hooper, D.
-M., Buchanan, C.
-D., Johansson, U.
-S., Tietze, D.
-T., Alström, P., \ldots{} others.
-(2014).
-Niche filling slows the diversification of himalayan songbirds.
-\emph{Nature}, \emph{509}(7499), 222.
-
-\leavevmode\vadjust pre{\hypertarget{ref-pulgarin2013multilocus}{}}%
-Pulgarín-R, P.
-C., Smith, B.
-T., Bryson Jr, R.
-W., Spellman, G.
-M., \& Klicka, J.
-(2013).
-Multilocus phylogeny and biogeography of the new world pheucticus grosbeaks (aves: cardinalidae).
-\emph{Molecular Phylogenetics and Evolution}, \emph{69}(3), 1222--1227.
-
-\leavevmode\vadjust pre{\hypertarget{ref-RCoreTeam2018}{}}%
-R Core Team. (2018). \emph{{R: a language and environment for statistical computing}}. Vienna, Austria: R Foundation for Statistical Computing.
-
-\leavevmode\vadjust pre{\hypertarget{ref-rabosky2018inverse}{}}%
-Rabosky, D.
-L., Chang, J., Title, P.
-O., Cowman, P.
-F., Sallan, L., Friedman, M., \ldots{} others.
-(2018).
-An inverse latitudinal gradient in speciation rate for marine fishes.
-\emph{Nature}, \emph{559}(7714), 392.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ramshaw1972time}{}}%
-Ramshaw, J., Richardson, D., Meatyard, B., Brown, R., Richardson, M., Thompson, E., \& Boulter, D.
-(1972).
-The time of origin of the flowering plants determined by using amino acid sequence data of cytochrome c.
-\emph{New Phytologist}, \emph{71}(5), 773--779.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ratnasingham2007bold}{}}%
-Ratnasingham, S., \& Hebert, P. D. (2007). BOLD: The barcode of life data system (http://www. Barcodinglife. org). \emph{Molecular Ecology Notes}, \emph{7}(3), 355--364.
-
-\leavevmode\vadjust pre{\hypertarget{ref-rees2017automated}{}}%
-Rees, J. A., \& Cranston, K. (2017). Automated assembly of a reference taxonomy for phylogenetic data synthesis. \emph{Biodiversity Data Journal}, (5).
-
-\leavevmode\vadjust pre{\hypertarget{ref-Revell2012}{}}%
-Revell, L. J. (2012). Phytools: An r package for phylogenetic comparative biology (and other things). \emph{{Methods in Ecology and Evolution}}, \emph{3}, 217--223.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Ronquist2003}{}}%
-Ronquist, F., \& Huelsenbeck, J.
-P.
-(2003).
-{MrBayes 3: Bayesian phylogenetic inference under mixed models}.
-\emph{Bioinformatics}, \emph{19}(12), 1572--1574.
-\url{https://doi.org/10.1093/bioinformatics/btg180}
-
-\leavevmode\vadjust pre{\hypertarget{ref-datelifeplot}{}}%
-Sanchez-Reyes, L.
-L., \& O'Meara, B.
-(2022).
-{\texttt{datelifeplot}: Methods to plot chronograms and outputs of the \texttt{datelife} package}.
-\emph{R Package Release V0.2.2}.
-Retrieved from \url{https://zenodo.org/badge/latestdoi/381501451}
-
-\leavevmode\vadjust pre{\hypertarget{ref-datelifeR}{}}%
-Sanchez-Reyes, L.
-L., O'Meara, B., Eastman, J., Heath, T., Wright, A., Schliep, K., \ldots{} Alfaro, M.
-(2022).
-{\texttt{datelife}: Scientific Data on Time of Lineage Divergence for Your Taxa}.
-\emph{R Package Release V0.6.2}.
-Retrieved from \url{https://doi.org/10.5281/zenodo.593938}
-
-\leavevmode\vadjust pre{\hypertarget{ref-sanderson2002}{}}%
-Sanderson, Michael J.
-(2002).
-{Estimating Absolute Rates of Molecular Evolution and Divergence Times: A Penalized Likelihood Approach}.
-\emph{Molecular Biology and Evolution}, \emph{19}(1), 101--109.
-\url{https://doi.org/10.1093/oxfordjournals.molbev.a003974}
-
-\leavevmode\vadjust pre{\hypertarget{ref-sanderson2003r8s}{}}%
-Sanderson, Michael J. (2003). r8s: Inferring absolute rates of molecular evolution and divergence times in the absence of a molecular clock. \emph{Bioinformatics}, \emph{19}(2), 301--302.
-
-\leavevmode\vadjust pre{\hypertarget{ref-sanderson2001sources}{}}%
-Sanderson, Michael J., \& Doyle, J.
-A.
-(2001).
-Sources of error and confidence intervals in estimating the age of angiosperms from rbcL and 18S rDNA data.
-\emph{American Journal of Botany}, \emph{88}(8), 1499--1516.
-
-\leavevmode\vadjust pre{\hypertarget{ref-sauquet2013practical}{}}%
-Sauquet, H. (2013). A practical guide to molecular dating. \emph{{Comptes Rendus Palevol}}, \emph{12}(6), 355--367.
-
-\leavevmode\vadjust pre{\hypertarget{ref-sauquet2021age}{}}%
-Sauquet, H., Ramírez-Barahona, S., \& Magallón, S. (2021). \emph{The age of flowering plants is unknown}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-schenk2016sec}{}}%
-Schenk, J. J. (2016). {Consequences of secondary calibrations on divergence time estimates}. \emph{PLoS ONE}, \emph{11}(1). \url{https://doi.org/10.1371/journal.pone.0148228}
-
-\leavevmode\vadjust pre{\hypertarget{ref-schliep2011phangorn}{}}%
-Schliep, K. P. (2011). Phangorn: Phylogenetic analysis in r. \emph{Bioinformatics}, \emph{27}(4), 592--593.
-
-\leavevmode\vadjust pre{\hypertarget{ref-selvatti2015paleogene}{}}%
-Selvatti, A.
-P., Gonzaga, L.
-P., \& Moraes Russo, C.
-A.
-de.
-(2015).
-A paleogene origin for crown passerines and the diversification of the oscines in the new world.
-\emph{Molecular Phylogenetics and Evolution}, \emph{88}, 1--15.
-
-\leavevmode\vadjust pre{\hypertarget{ref-smith2018constructing}{}}%
-Smith, Stephen A., \& Brown, J. W. (2018). Constructing a broadly inclusive seed plant phylogeny. \emph{American Journal of Botany}, \emph{105}(3), 302--314.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Smith2012}{}}%
-Smith, Stephen A., \& O'Meara, B.
-C.
-(2012).
-{TreePL: Divergence time estimation using penalized likelihood for large phylogenies}.
-\emph{Bioinformatics}, \emph{28}(20), 2689--2690.
-\url{https://doi.org/10.1093/bioinformatics/bts492}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Stoltzfus2013}{}}%
-Stoltzfus, A., Lapp, H., Matasci, N., Deus, H., Sidlauskas, B., Zmasek, C.
-M., \ldots{} Jordan, G.
-(2013).
-{Phylotastic!
-Making tree-of-life knowledge accessible, reusable and convenient}.
-\emph{{BMC Bioinformatics}}, \emph{14}.
-\url{https://doi.org/10.1186/1471-2105-14-158}
-
-\leavevmode\vadjust pre{\hypertarget{ref-tietze2013complete}{}}%
-Tietze, D.
-T., Päckert, M., Martens, J., Lehmann, H., \& Sun, Y.-H.
-(2013).
-Complete phylogeny and historical biogeography of true rosefinches (aves: carpodacus).
-\emph{Zoological Journal of the Linnean Society}, \emph{169}(1), 215--234.
-
-\leavevmode\vadjust pre{\hypertarget{ref-title2016macrophylogenies}{}}%
-Title, P.
-O., \& Rabosky, D.
-L.
-(2016).
-{Do Macrophylogenies Yield Stable Macroevolutionary Inferences?
-An Example from Squamate Reptiles}.
-\emph{Systematic Biology}, syw102.
-\url{https://doi.org/10.1093/sysbio/syw102}
-
-\leavevmode\vadjust pre{\hypertarget{ref-treplin2008molecular}{}}%
-Treplin, S., Siegert, R., Bleidorn, C., Thompson, H.
-S., Fotso, R., \& Tiedemann, R.
-(2008).
-Molecular phylogeny of songbirds (aves: Passeriformes) and the relative utility of common nuclear marker loci.
-\emph{Cladistics}, \emph{24}(3), 328--349.
-
-\leavevmode\vadjust pre{\hypertarget{ref-vos2012nexml}{}}%
-Vos, R.
-A., Balhoff, J.
-P., Caravas, J.
-A., Holder, M.
-T., Lapp, H., Maddison, W.
-P., \ldots{} others.
-(2012).
-NeXML: Rich, extensible, and verifiable representation of comparative data and metadata.
-\emph{Systematic Biology}, \emph{61}(4), 675--689.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Webb2000}{}}%
-Webb, C. (2000). {Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees}. \emph{{The American Naturalist}}, \emph{156}(2), 145--155.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Webb2008}{}}%
-Webb, Campbell O., Ackerly, D.
-D., \& Kembel, S.
-W.
-(2008).
-{Phylocom: Software for the analysis of phylogenetic community structure and trait evolution}.
-\emph{Bioinformatics}, \emph{24}(18), 2098--2100.
-\url{https://doi.org/10.1093/bioinformatics/btn358}
-
-\leavevmode\vadjust pre{\hypertarget{ref-webb2005phylomatic}{}}%
-Webb, Campbell O., \& Donoghue, M. J. (2005). Phylomatic: Tree assembly for applied phylogenetics. \emph{{Molecular Ecology Notes}}, \emph{5}(1), 181--183.
-
-\leavevmode\vadjust pre{\hypertarget{ref-weir2008calibrating}{}}%
-Weir, J., \& Schluter, D. (2008). Calibrating the avian molecular clock. \emph{Molecular Ecology}, \emph{17}(10), 2321--2328.
-
-\end{CSLReferences}
-
-\endgroup
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    TABLE 1     %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{table}
-  \caption{Ages of congruified nodes belonging to step C2 from Figure3}
-  \label{tbl:table1}
-  \includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example.pdf}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    TABLE 2     %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{table}
-  \caption{Summary of congruified nodes ages, corresponding to step C3 from Figure3}
-  \label{tbl:table2}
-  \includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example-summary.png}
-\end{table}
-
-% in this file I leave the figure captions outside\ caption{} because I want them
-% to be formatted in the same way as the general text (double spaced and linenumbered)
-\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 1    %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!h]
-\includegraphics{../figures/figure1/figure1-new.pdf}
-\caption{Stylized DateLife workflow.
-This shows the general worflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{http://www.datelife.org/}.
-Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
-}
-\label{fig:workflow}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE runtime    %%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!h]
-\includegraphics[width=1\linewidth]{../figures/fig_runtime_main.pdf}
-\caption{Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names.
-We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray).
-We also performed a search using the already processed query for comparison (light blue).}
-\label{fig:runtime_main}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 2    %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!h]
-\includegraphics[width=1\linewidth]{../figures/figure2/figure2.pdf}
-\caption{DateLife analysis results for a small sample of A) 6 bird species within the Passeriformes.
+\vspace{10mm}
+
+\begin{center}
+\includegraphics[width = 6.5in]{../figures/figure2/figure2.pdf}
+\end{center}
+\captionof{figure}{DateLife analysis results for a small sample of A) 6 bird species within the Passeriformes.
 B) Processed species names were found across 9 chronograms within 6 independent studies (Barker et al.
@@
 Summarized age data is used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree, resulting in a summary chronogram of source ages.}
+
 \label{fig:figure2}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 2-1  %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\begin{figure}[!h]
-%\includegraphics[width=1\linewidth]{../figures/figure2/figure2-1.pdf}
-%\caption{Age data results of a DateLife search of a small sample of 6 %bird species within the Passeriformes.
-Input names were found across 9 %chronograms within 6 independent studies (Barker et al.
-%(\protect\hyperlink{ref-barker2012going}{2012}), Barker et al.
-%(\protect\hyperlink{ref-barker2015new}{2015}), Burns et al.
-%(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al.
-%(\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price %(\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al.
-%(\protect\hyperlink{ref-Jetz2012}{2012}).) This revealed 28 age data %points for the queried species names.}
-%\label{fig:figure2-1}
-%\end{figure}
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 2-2  %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\begin{figure}[!h]
-%\includegraphics{../figures/figure2/figure2-2.pdf}
-%\caption{Summarized age data is used as secondary calibrations to date %a tree topology as a summary chronogram.}
-%\label{fig:summaries}
-%\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%    FIGURE TOPOLOGIES    %%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!h]
-\includegraphics{../figures/fringillidae-topologies/fringillidae-topology.pdf}
-\caption{Tree topologies obtained from Open Tree of Life's (OpenTree) synthetic phylogenetic tree.
+
+\vspace{15mm}
+
+\begin{table}[H]
+  \caption{Ages of congruified nodes obtained from a datelife search for the Small Example presented. See Figure 3, step C2.}
+  \label{tbl:table1}
+  \includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example.pdf}
+\end{table}
+
+\begin{table}[H]
+  \caption{Summary of congruified nodes ages. See Figure 3, step C3.}
+  \label{tbl:table2}
+  \includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example-summary.png}
+\end{table}
+
+\hypertarget{the-family-fringillidae}{%
+\subsection{The family Fringillidae}\label{the-family-fringillidae}}
+
+\emph{\textbf{Creating a query.-}}
+To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the ``get species from taxon'' flag,
+which gets all recognized species names within a named group from a taxonomy of choice.
+Following the NCBI taxonomy, our DateLife query has 289 Fringillidae species.
+This taxon-constrained approach implies that the final results of a full DateLife analysis will be done using a tree topology and ages for the species in a named group, which do not necessarily correspond to a monophyletic group.
+Users can change this behaviour by providing a monophyletic tree as input for a DateLife search, or as a tree topology for a DateLife summary.
+
+\vspace{5mm}
+
+\emph{\textbf{Searching the database.-}}
+Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species.
+The DateLife search identified 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Price et al., 2014).
+Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices, revealing 1206 different age data points available for species within the Fringillidae (supplementray table 1).
+
+\vspace{5mm}
+
+\emph{\textbf{Summarizing search results.-}}
+The final step is to congruify and summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM.
+As explained in the ``Description'' section, a tree topology to summarize age data upon is required.
+By default, DateLife uses as summarizing topology the portion of OpenTree synthetic phylogeny that contains the species provided as search query.
+According to this phylogeny, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).
+
+Age data from source chronograms was congruified, decreasing to 818 different age data points available (supplementray table 2) and then age summary statistics per congruified node are obtained and used as fixed secondary calibrations over a consensus tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
+\ref{fig:fringillidages}).
+Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
+
+\vspace{20mm}
+
+\begin{center}
+\includegraphics[width = 6.5in]{../figures/fringillidae-topologies/fringillidae-topology.pdf}
+\captionof{figure}{Tree topologies obtained from Open Tree of Life's (OpenTree) synthetic phylogenetic tree.
 A) Topology of 2,333 tips and 1,305 internal nodes, encompassing bird species within the family Fringillidae following the NCBI taxonomy (black), as well as all other bird species that share the same mrca node in OpenTree's synthetic tree (purple).
@@
 \label{fig:fringillidae-topologies}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%    FIGURE FRINGILLIDAE    %%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!h]
-\includegraphics{../figures/figure-fringillidae/median_and_calibration_ages-congruified.png}
-\caption{Fringillidae median summary chronogram generated with DateLife. It has 256 tips and 233 nodes, from which 212 have age data from at least one published chronogram.}
-\label{fig:fringillidages}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%    FIGURE CROSS VALIDATION    %%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[!h]
-\includegraphics{../figures/figure-cross-validation/fig-cross-validation-xy-plots.pdf}
-\caption{Results from cross validation analysis. Each plot compares the original age estimate (x axis) with the age obtained with a DateLife analysis (y axis), per node.}
-\label{fig:cvXY}
-\end{figure}
+\end{center}
+
+\hypertarget{cross-validation-test}{%
+\section{Cross-validation test}\label{cross-validation-test}}
+
+We performed a cross validation analysis of the DateLife workflow using the Fringillidae chronograms.
+We used the individual tree topologies from each of the 19 source chronograms from 13 studies as inputs, treating their node ages as unknown.
+We then estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.
+
+We found that node ages from original study, and ages estimated using all other age data available are correlated (Supplementary Fig. \ref{fig:cvXY}).
+
+For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka `closer to the root') relative to the orginal estimate, and overestimate ages for nodes closer to the tips.
+Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. \ref{fig:cvXYdiffs}).
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+\texttt{datelife} makes state-of-the-art data on evolutionary time frame easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
+It is an open service that does not require any expert biological knowledge from users --besides the names of the species or group they want to work with, for any of its functionality.
+
+A total of 99474 unique terminal taxa are represented in \texttt{datelife}'s database.
+Incorporation of more chronograms into \texttt{datelife}'s database will continue to improve its services.
+One option to increase the number of chronograms in the DateLife database is the Dryad data repository.
+Methods to automatically mine chronograms from Dryad could be designed and implemented.
+However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined manually by a human curator.
+We would like to emphasize on the importance of sharing chronogram data, including systematically curated metadata, into open repositories, such as OpenTree's Phylesystem (McTavish et al., 2015) for the benefit of the scientific community as a whole.
+
+\hypertarget{primary-vs-secondary-calibrations-in-dating-analyses}{%
+\subsection{Primary vs Secondary calibrations in dating analyses}\label{primary-vs-secondary-calibrations-in-dating-analyses}}
+
+While most chronograms in DateLife's database are constructed using primary calibrations (molecular substitution rates or ages obtained from the fossil record or geological events), DateLife summarizes chronograms using secondary calibrations only (ages coming from other chronograms).
+
+Graur and Martin (2004) cautioned on the increased error and uncertainty in estimated ages when using secondary calibrations in dating analyses.
+Schenk (2016) showed that, in simulations, divergence times inferred using secondary calibrations are significantly younger than those inferred with primary calibrations, when obtained with Bayesian inference methods, and when priors are implemented in similar ways in both analyses.
+Accordingly, the scientific community seems to have more confidence in chronograms obtained from a single analysis, using fossil data as primary sources of calibrations (Schenk, 2016), and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data reflect a coherent evolutionary history (Sauquet, 2013), as for example done by Antonelli et al.
+(2017).
+There have been attempts to create fossil calibration databases (Ksepka et al., 2015), though these still have room to grow.
+
+It seems that using several (as opposed to just a few) secondary calibrations can provide sufficient information to alleviate or even neutralize potential biases (Sauquet, 2013).
+Certainly, further studies are required to fully understand the effect of secondary calibrations on outputs from different tree dating methods, and on downstream analyses.
+It is possible that secondary calibrations can be safely used with dating methods that do not require setting priors, such as penalized likelihood (Michael J.
+Sanderson, 2003), with methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ (Campbell O.
+Webb et al., 2008; Campbell O.
+Webb \& Donoghue, 2005), or methods that summarize age data unto a tree topology.
+
+Our cross validation analysis might provide some insight in this regard.
+When ages are estimated with secondary calibrations, nodes closer to the root do tend to be slightly younger than ages estimated with primary calibrations.
+However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures 1-9), with a notable exception of cross validation of Burns et al.
+(2014) chronogram, which results in much younger node ages when estimated using secondary calibrations (Supplementary figs.
+1 and 4).
+
+\hypertarget{variation-in-source-chronograms}{%
+\subsection{Variation in source chronograms}\label{variation-in-source-chronograms}}
+
+Conflict among alternative dating studies is common in the literature
+(see, for example, the robust ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández, 2015; Ramshaw et al., 1972; Michael J.
+Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021)).
+
+Source chronograms may have been estimated using different types of primary calibrations (fossil record, geologic or biogeographic events, molecular substitution rates).
+These differences in approach can deepen the already substantial variation in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
+For example, the chronograms from Burns et al.
+(2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older date estimates for the same nodes than chronograms that were inferred using fossils as primary calibrations (figs.
+fig:fringillidages and fig:cvXY; supplementary figs.
+1 and 4).
+
+Different chronograms might also imply fundamentally distinct evolutionary hypotheses.
+For example, two different researchers working on the same clade could both carefully select and argue for their choices of fossil calibrations.
+Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses the same one to calibrate outside the clade, the resulting age estimates will often differ substantially, as the placement of calibrations as stem or crown group is proved to deeply affect estimated times of lineage divergence (Sauquet, 2013).
+
+\hypertarget{sumarizing-chronograms}{%
+\subsection{Sumarizing chronograms}\label{sumarizing-chronograms}}
+
+By default, DateLife currently summarizes all source chronograms that overlap with at least two species names. Users can exclude source chronograms if they have reasons to do so.
+Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely.
+To our knowledge, there are no tested measures to determine independently when a chronogram is better than another.
+Yet, several characteristics of the data used for dating analyses, as well as from the output chronogram itself, could be used to score the quality of source chronograms.
+
+Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used Magallón et al. (2015).
+Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
+
+DateLife provides an opportunity to capture concordance and conflict among date estimates, which can also be used as a metric for chronogram reliability.
+Its open database of chronograms allows other researchers to do such analyses themselves reproducibly, and without needing permission.
+Though, of course, they should follow proper citation practices, especially for the source chronogram studies.
+
+The exercise of summarizing age data from across multiple studies provides the opportunity to work with a more inclusive chronogram,
+that reflects a unified evolutionary history for a lineage, by putting together evidence from different hypotheses.
+The largest, and taxonomically broadest chronogram currently available from OpenTree was constructed summarizing age data from 2,274 published chronograms using NCBI's taxonomic tree as backbone (Hedges et al., 2015).
+A summarizing exercise may also amplify the effect of uncertainty and errors in source data, and blur parts of the evolutionary history of a lineage that might only be reflected in source chronograms and lost on the summary chronogram (Sauquet et al., 2021).
+
+\hypertarget{effects-on-downstream-analyses}{%
+\subsection{Effects on downstream analyses}\label{effects-on-downstream-analyses}}
+
+For downstream analyses, using alternative chronogram may deeply affect our inferences (Title \& Rabosky, 2016), particularly when studying phenomena dependent on the timing of species diversification events, such as macroevolutionary processes.
+
+In ecology and conservation biology, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance (Campbell O.
+Webb et al., 2008).
+Hence, DateLife's workflow features different ways of estimating node ages in the absence of calibrations and branch length information for certain taxa.
+``Making up'' branch lengths is a common practice in scientific publications: Jetz et al.
+(2012), created a time-calibrated tree of all 9,993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
+(2018) created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37\% had molecular data; Stephen A.
+Smith and Brown (2018) constructed a tree of 353,185 seed plants where only 23\% had molecular data.
+
+Notably, risks come with this practice.
+Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration, using polytomy resolution and branch estimation methods.
+There has yet to be a thorough analysis of what can go wrong when one extends inferences beyond the data in this way, so we urge caution; we also urge readers to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, and assessing the consistency of the results with prior work.
+
+\hypertarget{conclusions}{%
+\section{Conclusions}\label{conclusions}}
+
+Knowledge of the evolutionary time frame of organisms is key to many research areas: trait evolution,
+species diversification, biogeography, macroecology and more. It is also crucial for education, science communication and policy, but generating chronograms is difficult,
+especially for those who want to use phylogenies but who are not systematists, or
+do not have the time to acquire and develop the necessary knowledge and skills to construct them on their own.
+Importantly, years of primarily public funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
+
+The DateLife project allows for easy and fast summary of public and state-of-the-art data on time of lineage divergence.
+It provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identification of regions that require more research, or that have conflicting information.
+It is available as an R package, and as a web-based R shiny application at www.datelife.org
+Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
+The DateLife project helps with awareness of the existing variation in expert time of divergence data, and will foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
+
+\hypertarget{availability}{%
+\section{Availability}\label{availability}}
+
+The \texttt{datelife} software is free and open source and it can be used through its R shiny web application at
+\url{http://www.datelife.org}, through the \texttt{datelife} R package, and through Phylotastic's project web portal \url{https://phylo.cs.nmsu.edu/}.
+DateLife's web application is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker and Open Tree of Life's infrastructure (dates.opentreeoflife.org/datelife).
+\texttt{datelife}'s R package stable version is available
+for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
+using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R. Development versions
+are available from the GitHub repository (\url{https://github.com/phylotastic/datelife})
+and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.
+
+\hypertarget{supplementary-material}{%
+\section{Supplementary Material}\label{supplementary-material}}
+
+Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities are available at \href{https://github.com/LunaSare/datelifeMS1}{datelifeMS1}, \href{https://github.com/LunaSare/datelife_examples}{datelife\_examples}, and \href{https://github.com/LunaSare/datelife_benchmark}{datelife\_benchmark} repositories in LLSR's GitHub account.
+
+\hypertarget{funding}{%
+\section{Funding}\label{funding}}
+
+Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to the Datelife project; DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), ABI-1458572 to the Phylotastic project, and ABI-1759846 to the Open Tree of Life project.
+
+\hypertarget{acknowledgements}{%
+\section{Acknowledgements}\label{acknowledgements}}
+
+The DateLife project was born as a prototype tool aiming to provide these services, and was initially developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA (Stoltzfus et al., 2013).
+We thank colleagues from the O'Meara Lab at the University
+of Tennesse Knoxville for suggestions, discussions and software testing.
+The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
+that led to initial work on this project.
+The team that assembled DateLife's first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
+The Open Tree of Life project that provides the open, metadata rich repository of
+trees used to construct DateLife's chronogram database.
+The many scientists who publish their chronograms in an open, reusable form, and the scientists who curate them for deposition in the Open Tree of Life repository.
+The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.
+
+\newpage
+
+\hypertarget{references}{%
+\section{References}\label{references}}
+
+\begingroup
+\setlength{\parindent}{-0.5in}
+\setlength{\leftskip}{0.5in}
+
+\hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-alstrom2014discovery}{}}%
+Alström, P., Hooper, D.
+M., Liu, Y., Olsson, U., Mohan, D., Gelang, M., \ldots{} Price, T.
+D.
+(2014).
+Discovery of a relict lineage and monotypic family of passerine birds.
+\emph{Biology Letters}, \emph{10}(3), 20131067.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ane2009groves}{}}%
+Ané, C., Eulenstein, O., Piaggio-Talice, R., \& Sanderson, M. J. (2009). Groves of phylogenetic trees. \emph{{Annals of Combinatorics}}, \emph{13}(2), 139--167.
+
+\leavevmode\vadjust pre{\hypertarget{ref-antonelli2017supersmart}{}}%
+Antonelli, A., Hettling, H., Condamine, F.
+L., Vos, K., Nilsson, R.
+H., Sanderson, M.
+J., \ldots{} Vos, R.
+A.
+(2017).
+{Toward a self-updating platform for estimating rates of speciation and migration, ages, and relationships of Taxa}.
+\emph{Systematic Biology}, \emph{66}(2), 153--166.
+\url{https://doi.org/10.1093/sysbio/syw066}
+
+\leavevmode\vadjust pre{\hypertarget{ref-archie1986newick}{}}%
+Archie, J., Day, W.
+H., Felsenstein, J., Maddison, W., Meacham, C., Rohlf, F.
+J., \& Swofford, D.
+(1986).
+{The Newick tree format}.
+Retrieved from \url{\%7Bhttps://evolution.genetics.washington.edu/phylip/newicktree.html\%7D}
+
+\leavevmode\vadjust pre{\hypertarget{ref-avibase-emberiza}{}}%
+Avibase.
+(2022).
+{Yellow-throated Bunting}.
+\emph{{Avibase - The World Bird Database}}, (Online Resource).
+Retrieved from \url{\%7Bhttps://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\%7D}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Bapst2012a}{}}%
+Bapst, D.
+W.
+(2012).
+{Paleotree: An R package for paleontological and phylogenetic analyses of evolution}.
+\emph{{Methods in Ecology and Evolution}}, \emph{3}(5), 803--807.
+\url{https://doi.org/10.1111/j.2041-210X.2012.00223.x}
+
+\leavevmode\vadjust pre{\hypertarget{ref-barba2018constraining}{}}%
+Barba-Montoya, J., Reis, M.
+dos, Schneider, H., Donoghue, P.
+C., \& Yang, Z.
+(2018).
+Constraining uncertainty in the timescale of angiosperm evolution and the veracity of a cretaceous terrestrial revolution.
+\emph{New Phytologist}, \emph{218}(2), 819--834.
+
+\leavevmode\vadjust pre{\hypertarget{ref-barker2014mitogenomic}{}}%
+Barker, F. K. (2014). Mitogenomic data resolve basal relationships among passeriform and passeridan birds. \emph{Molecular Phylogenetics and Evolution}, \emph{79}, 313--324.
+
+\leavevmode\vadjust pre{\hypertarget{ref-barker2013going}{}}%
+Barker, F.
+K., Burns, K.
+J., Klicka, J., Lanyon, S.
+M., \& Lovette, I.
+J.
+(2013).
+Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds.
+\emph{{Systematic Biology}}, \emph{62}(2), 298--320.
+
+\leavevmode\vadjust pre{\hypertarget{ref-barker2015new}{}}%
+Barker, F.
+K., Burns, K.
+J., Klicka, J., Lanyon, S.
+M., \& Lovette, I.
+J.
+(2015).
+New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
+\emph{The Auk: Ornithological Advances}, \emph{132}(2), 333--348.
+
+\leavevmode\vadjust pre{\hypertarget{ref-barker2004phylogeny}{}}%
+Barker, F.
+K., Cibois, A., Schikler, P., Feinstein, J., \& Cracraft, J.
+(2004).
+Phylogeny and diversification of the largest avian radiation.
+\emph{Proceedings of the National Academy of Sciences}, \emph{101}(30), 11040--11045.
+
+\leavevmode\vadjust pre{\hypertarget{ref-beresford2005african}{}}%
+Beresford, P., Barker, F., Ryan, Pg., \& Crowe, T.
+(2005).
+African endemics span the tree of songbirds (passeri): Molecular systematics of several evolutionary {``enigmas.''} \emph{Proceedings of the Royal Society B: Biological Sciences}, \emph{272}(1565), 849--858.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Boyle2013}{}}%
+Boyle, B., Hopkins, N., Lu, Z., Raygoza Garay, J.
+A., Mozzherin, D., Rees, T., \ldots{} Enquist, B.
+J.
+(2013).
+{The taxonomic name resolution service: An online tool for automated standardization of plant names}.
+\emph{{BMC Bioinformatics}}, \emph{14}(1).
+\url{https://doi.org/10.1186/1471-2105-14-16}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Britton2007}{}}%
+Britton, T., Anderson, C.
+L., Jacquet, D., Lundqvist, S., \& Bremer, K.
+(2007).
+{Estimating Divergence Times in Large Phylogenetic Trees}.
+\emph{{Systematic Biology}}, \emph{56}(788777878), 741--752.
+\url{https://doi.org/10.1080/10635150701613783}
+
+\leavevmode\vadjust pre{\hypertarget{ref-bryson2014diversification}{}}%
+Bryson Jr, R.
+W., Chaves, J., Smith, B.
+T., Miller, M.
+J., Winker, K., Pérez-Emán, J.
+L., \& Klicka, J.
+(2014).
+Diversification across the new world within the `blue'cardinalids (aves: cardinalidae).
+\emph{Journal of Biogeography}, \emph{41}(3), 587--599.
+
+\leavevmode\vadjust pre{\hypertarget{ref-burleigh2015building}{}}%
+Burleigh, J. G., Kimball, R. T., \& Braun, E. L. (2015). Building the avian tree of life using a large-scale, sparse supermatrix. \emph{Molecular Phylogenetics and Evolution}, \emph{84}, 53--63.
+
+\leavevmode\vadjust pre{\hypertarget{ref-burns2014phylogenetics}{}}%
+Burns, K.
+J., Shultz, A.
+J., Title, P.
+O., Mason, N.
+A., Barker, F.
+K., Klicka, J., \ldots{} Lovette, I.
+J.
+(2014).
+Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds.
+\emph{{Molecular Phylogenetics and Evolution}}, \emph{75}, 41--77.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Chamberlain2013}{}}%
+Chamberlain, S.
+A., \& Szöcs, E.
+(2013).
+{taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}}.
+\emph{{F1000Research}}, \emph{2}(191), 1--29.
+\url{https://doi.org/10.12688/f1000research.2-191.v2}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Chamberlain2018}{}}%
+Chamberlain, S.
+A., Szöcs, E., Foster, Z., Arendsee, Z., Boettiger, C., Ram, K., \ldots{} Li, G.
+(2019).
+\emph{{taxize: Taxonomic information from around the web}}.
+Retrieved from \url{https://github.com/ropensci/taxize}
+
+\leavevmode\vadjust pre{\hypertarget{ref-chaves2013biogeography}{}}%
+Chaves, J.
+A., Hidalgo, J.
+R., \& Klicka, J.
+(2013).
+Biogeography and evolutionary history of the n eotropical genus s altator (a ves: T hraupini).
+\emph{Journal of Biogeography}, \emph{40}(11), 2180--2190.
+
+\leavevmode\vadjust pre{\hypertarget{ref-claramunt2015new}{}}%
+Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{{Science Advances}}, \emph{1}(11), e1501005.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Criscuolo2006}{}}%
+Criscuolo, A., Berry, V., Douzery, E.
+J.
+P., \& Gascuel, O.
+(2006).
+{SDM: A fast distance-based approach for (super)tree building in phylogenomics}.
+\emph{{Systematic Biology}}, \emph{55}(5), 740--755.
+\url{https://doi.org/10.1080/10635150600969872}
+
+\leavevmode\vadjust pre{\hypertarget{ref-delsuc2018phylogenomic}{}}%
+Delsuc, F., Philippe, H., Tsagkogeorga, G., Simion, P., Tilak, M.-K., Turon, X., \ldots{} Douzery, E.
+J.
+(2018).
+A phylogenomic framework and timescale for comparative studies of tunicates.
+\emph{BMC Biology}, \emph{16}(1), 1--14.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Eastman2013}{}}%
+Eastman, J.
+M., Harmon, L.
+J., \& Tank, D.
+C.
+(2013).
+{Congruification: Support for time scaling large phylogenetic trees}.
+\emph{{Methods in Ecology and Evolution}}, \emph{4}(7), 688--691.
+\url{https://doi.org/10.1111/2041-210X.12051}
+
+\leavevmode\vadjust pre{\hypertarget{ref-edgar2004muscle}{}}%
+Edgar, R. C. (2004). MUSCLE: Multiple sequence alignment with high accuracy and high throughput. \emph{Nucleic Acids Research}, \emph{32}(5), 1792--1797.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Felsenstein1985a}{}}%
+Felsenstein, J. (1985). {Phylogenies and the Comparative Method}. \emph{The American Naturalist}, \emph{125}(1), 1--15. Retrieved from \url{http://www.jstor.org/stable/2461605}
+
+\leavevmode\vadjust pre{\hypertarget{ref-gibb2015new}{}}%
+Gibb, G.
+C., England, R., Hartig, G., McLenachan, P.
+A., Taylor Smith, B.
+L., McComish, B.
+J., \ldots{} Penny, D.
+(2015).
+New zealand passerines help clarify the diversification of major songbird lineages during the oligocene.
+\emph{{Genome Biology and Evolution}}, \emph{7}(11), 2983--2995.
+
+\leavevmode\vadjust pre{\hypertarget{ref-graur2004reading}{}}%
+Graur, D., \& Martin, W. (2004). Reading the entrails of chickens: Molecular timescales of evolution and the illusion of precision. \emph{TRENDS in Genetics}, \emph{20}(2), 80--86.
+
+\leavevmode\vadjust pre{\hypertarget{ref-hackett2008phylogenomic}{}}%
+Hackett, S.
+J., Kimball, R.
+T., Reddy, S., Bowie, R.
+C., Braun, E.
+L., Braun, M.
+J., \ldots{} others.
+(2008).
+A phylogenomic study of birds reveals their evolutionary history.
+\emph{Science}, \emph{320}(5884), 1763--1768.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Hedges2015}{}}%
+Hedges, S.
+B., Marin, J., Suleski, M., Paymer, M., \& Kumar, S.
+(2015).
+{Tree of life reveals clock-like speciation and diversification}.
+\emph{{Molecular Biology and Evolution}}, \emph{32}(4), 835--845.
+\url{https://doi.org/10.1093/molbev/msv037}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Heibl2008}{}}%
+Heibl, C. (2008). \emph{PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages.} Retrieved from \url{http://www.christophheibl.de/Rpackages.html}
+
+\leavevmode\vadjust pre{\hypertarget{ref-hooper2017chromosomal}{}}%
+Hooper, D. M., \& Price, T. D. (2017). Chromosomal inversion differences correlate with range overlap in passerine birds. \emph{Nature Ecology \& Evolution}, \emph{1}(10), 1526.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Huelsenbeck2001}{}}%
+Huelsenbeck, J. P., \& Ronquist, F. (2001). {MRBAYES: Bayesian inference of phylogenetic trees}. \emph{Bioinformatics}, \emph{17}(8), 754--755. \url{https://doi.org/10.1093/bioinformatics/17.8.754}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Jetz2012}{}}%
+Jetz, W., Thomas, G., Joy, J.
+J.
+B., Hartmann, K., \& Mooers, A.
+(2012).
+{The global diversity of birds in space and time}.
+\emph{Nature}, \emph{491}(7424), 444--448.
+\url{https://doi.org/10.1038/nature11631}
+
+\leavevmode\vadjust pre{\hypertarget{ref-johansson2008phylogenetic}{}}%
+Johansson, U.
+S., Fjeldså, J., \& Bowie, R.
+C.
+(2008).
+Phylogenetic relationships within passerida (aves: Passeriformes): A review and a new molecular phylogeny based on three nuclear intron markers.
+\emph{Molecular Phylogenetics and Evolution}, \emph{48}(3), 858--876.
+
+\leavevmode\vadjust pre{\hypertarget{ref-katoh2009multiple}{}}%
+Katoh, K., Asimenos, G., \& Toh, H. (2009). Multiple alignment of DNA sequences with MAFFT. In \emph{Bioinformatics for DNA sequence analysis} (pp. 39--64). Springer.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kimball2019phylogenomic}{}}%
+Kimball, R. T., Oliveros, C. H., Wang, N., White, N. D., Barker, F. K., Field, D. J., \ldots{} others. (2019). A phylogenomic supertree of birds. \emph{Diversity}, \emph{11}(7), 109.
+
+\leavevmode\vadjust pre{\hypertarget{ref-klicka2014comprehensive}{}}%
+Klicka, J., Barker, F.
+K., Burns, K.
+J., Lanyon, S.
+M., Lovette, I.
+J., Chaves, J.
+A., \& Bryson Jr, R.
+W.
+(2014).
+A comprehensive multilocus assessment of sparrow (aves: Passerellidae) relationships.
+\emph{Molecular Phylogenetics and Evolution}, \emph{77}, 177--182.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ksepka2015fossil}{}}%
+Ksepka, D.
+T., Parham, J.
+F., Allman, J.
+F., Benton, M.
+J., Carrano, M.
+T., Cranston, K.
+A., \ldots{} others.
+(2015).
+The fossil calibration database---a new resource for divergence dating.
+\emph{Systematic Biology}, \emph{64}(5), 853--859.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lamichhaney2015evolution}{}}%
+Lamichhaney, S., Berglund, J., Almén, M.
+S., Maqbool, K., Grabherr, M., Martinez-Barrio, A., \ldots{} others.
+(2015).
+Evolution of darwin's finches and their beaks revealed by genome sequencing.
+\emph{Nature}, \emph{518}(7539), 371--375.
+
+\leavevmode\vadjust pre{\hypertarget{ref-laubichler2009form}{}}%
+Laubichler, M. D., \& Maienschein, J. (2009). \emph{Form and function in developmental evolution}. Cambridge University Press.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lepage2004avibase}{}}%
+Lepage, D. (2004). \emph{Avibase: The world bird database}. Bird Studies Canada.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lepage2014avibase}{}}%
+Lepage, D., Vaidya, G., \& Guralnick, R. (2014). Avibase--a database system for managing and organizing taxonomic concepts. \emph{ZooKeys}, (420), 117.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lerner2011multilocus}{}}%
+Lerner, H.
+R., Meyer, M., James, H.
+F., Hofreiter, M., \& Fleischer, R.
+C.
+(2011).
+Multilocus resolution of phylogeny and timescale in the extant adaptive radiation of hawaiian honeycreepers.
+\emph{Current Biology}, \emph{21}(21), 1838--1844.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lovette2010comprehensive}{}}%
+Lovette, I.
+J., Pérez-Emán, J.
+L., Sullivan, J.
+P., Banks, R.
+C., Fiorentino, I., Córdoba-Córdoba, S., \ldots{} others.
+(2010).
+A comprehensive multilocus phylogeny for the wood-warblers and a revised classification of the parulidae (aves).
+\emph{Molecular Phylogenetics and Evolution}, \emph{57}(2), 753--770.
+
+\leavevmode\vadjust pre{\hypertarget{ref-magallon2001absolute}{}}%
+Magallon, S., \& Sanderson, M. J. (2001). Absolute diversification rates in angiosperm clades. \emph{Evolution}, \emph{55}(9), 1762--1780.
+
+\leavevmode\vadjust pre{\hypertarget{ref-magallon2010using}{}}%
+Magallón, S. (2010). Using fossils to break long branches in molecular dating: A comparison of relaxed clocks applied to the origin of angiosperms. \emph{Systematic Biology}, \emph{59}(4), 384--399.
+
+\leavevmode\vadjust pre{\hypertarget{ref-magallon2015metacalibrated}{}}%
+Magallón, S., Gómez-Acevedo, S., Sánchez-Reyes, L.
+L., \& Hernández-Hernández, T.
+(2015).
+A metacalibrated time-tree documents the early rise of flowering plant phylogenetic diversity.
+\emph{New Phytologist}, \emph{207}(2), 437--453.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mctavish2015phylesystem}{}}%
+McTavish, E.
+J., Hinchliff, C.
+E., Allman, J.
+F., Brown, J.
+W., Cranston, K.
+A., Holder, M.
+T., \ldots{} Smith, S.
+A.
+(2015).
+Phylesystem: A git-based data store for community-curated phylogenetic estimates.
+\emph{Bioinformatics}, \emph{31}(17), 2794--2800.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Michonneau2016}{}}%
+Michonneau, F., Brown, J.
+W., \& Winter, D.
+J.
+(2016).
+{rotl: an R package to interact with the Open Tree of Life data}.
+\emph{{Methods in Ecology and Evolution}}, \emph{7}(12), 1476--1481.
+\url{https://doi.org/10.1111/2041-210X.12593}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Morlon2014}{}}%
+Morlon, H. (2014). {Phylogenetic approaches for studying diversification.} \emph{{Ecology Letters}}, \emph{17}(4), 508--525. \url{https://doi.org/10.1111/ele.12251}
+
+\leavevmode\vadjust pre{\hypertarget{ref-moyle2016tectonic}{}}%
+Moyle, R.
+G., Oliveros, C.
+H., Andersen, M.
+J., Hosner, P.
+A., Benz, B.
+W., Manthey, J.
+D., \ldots{} Faircloth, B.
+C.
+(2016).
+Tectonic collision and uplift of wallacea triggered the global songbird radiation.
+\emph{Nature Communications}, \emph{7}(1), 1--7.
+
+\leavevmode\vadjust pre{\hypertarget{ref-odeen2011evolution}{}}%
+Ödeen, A., Håstad, O., \& Alström, P. (2011). Evolution of ultraviolet vision in the largest avian radiation-the passerines. \emph{BMC Evolutionary Biology}, \emph{11}(1), 1--8.
+
+\leavevmode\vadjust pre{\hypertarget{ref-oliveros2019earth}{}}%
+Oliveros, C.
+H., Field, D.
+J., Ksepka, D.
+T., Barker, F.
+K., Aleixo, A., Andersen, M.
+J., \ldots{} others.
+(2019).
+Earth history and the passerine superradiation.
+\emph{Proceedings of the National Academy of Sciences}, \emph{116}(16), 7916--7925.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Ooms2018}{}}%
+Ooms, J., \& Chamberlain, S. (2018). \emph{Phylocomr: Interface to 'phylocom'}. Retrieved from \url{https://CRAN.R-project.org/package=phylocomr}
+
+\leavevmode\vadjust pre{\hypertarget{ref-opentreeAPIs}{}}%
+Open Tree Of Life, Redelings, B., Cranston, K.
+A., Allman, J., Holder, M.
+T., \& McTavish, E.
+J.
+(2016).
+{Open Tree of Life APIs v3.0}.
+\emph{{Open Tree of Life Project}}, (Online Resources).
+Retrieved from \url{\%7Bhttps://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\%7D}
+
+\leavevmode\vadjust pre{\hypertarget{ref-opentreeoflife2019synth}{}}%
+Open Tree Of Life, Redelings, B., Sánchez Reyes, L.
+L., Cranston, K.
+A., Allman, J., Holder, M.
+T., \& McTavish, E.
+J.
+(2019).
+Open tree of life synthetic tree v12.3.
+\emph{Zenodo}.
+Retrieved from \url{https://doi.org/10.5281/zenodo.3937742}
+
+\leavevmode\vadjust pre{\hypertarget{ref-packert2012horizontal}{}}%
+Päckert, M., Martens, J., Sun, Y.-H., Severinghaus, L.
+L., Nazarenko, A.
+A., Ting, J., \ldots{} Tietze, D.
+T.
+(2012).
+Horizontal and elevational phylogeographic patterns of himalayan and southeast asian forest passerines (aves: passeriformes).
+\emph{Journal of Biogeography}, \emph{39}(3), 556--573.
+
+\leavevmode\vadjust pre{\hypertarget{ref-paradis2004}{}}%
+Paradis, E., Claude, J., \& Strimmer, K. (2004). {APE: analyses of phylogenetics and evolution in R language}. \emph{Bioinformatics}, \emph{20}(2), 289--290.
+
+\leavevmode\vadjust pre{\hypertarget{ref-parchman2007coevolution}{}}%
+Parchman, T.
+L., Benkman, C.
+W., \& Mezquida, E.
+T.
+(2007).
+Coevolution between hispaniolan crossbills and pine: Does more time allow for greater phenotypic escalation at lower latitude?
+\emph{Evolution}, \emph{61}(9), 2142--2153.
+
+\leavevmode\vadjust pre{\hypertarget{ref-pennell2014geiger}{}}%
+Pennell, M.
+W., Eastman, J.
+M., Slater, G.
+J., Brown, J.
+W., Uyeda, J.
+C., FitzJohn, R.
+G., \ldots{} Harmon, L.
+J.
+(2014).
+Geiger v2.
+0: An expanded suite of methods for fitting macroevolutionary models to phylogenetic trees.
+\emph{Bioinformatics}, \emph{30}(15), 2216--2218.
+
+\leavevmode\vadjust pre{\hypertarget{ref-posadas2006historical}{}}%
+Posadas, P., Crisci, J. V., \& Katinas, L. (2006). Historical biogeography: A review of its basic concepts and critical issues. \emph{{Journal of Arid Environments}}, \emph{66}(3), 389--403.
+
+\leavevmode\vadjust pre{\hypertarget{ref-powell2014comprehensive}{}}%
+Powell, A.
+F., Barker, F.
+K., Lanyon, S.
+M., Burns, K.
+J., Klicka, J., \& Lovette, I.
+J.
+(2014).
+A comprehensive species-level molecular phylogeny of the new world blackbirds (icteridae).
+\emph{Molecular Phylogenetics and Evolution}, \emph{71}, 94--112.
+
+\leavevmode\vadjust pre{\hypertarget{ref-price2014niche}{}}%
+Price, T.
+D., Hooper, D.
+M., Buchanan, C.
+D., Johansson, U.
+S., Tietze, D.
+T., Alström, P., \ldots{} others.
+(2014).
+Niche filling slows the diversification of himalayan songbirds.
+\emph{Nature}, \emph{509}(7499), 222.
+
+\leavevmode\vadjust pre{\hypertarget{ref-pulgarin2013multilocus}{}}%
+Pulgarín-R, P.
+C., Smith, B.
+T., Bryson Jr, R.
+W., Spellman, G.
+M., \& Klicka, J.
+(2013).
+Multilocus phylogeny and biogeography of the new world pheucticus grosbeaks (aves: cardinalidae).
+\emph{Molecular Phylogenetics and Evolution}, \emph{69}(3), 1222--1227.
+
+\leavevmode\vadjust pre{\hypertarget{ref-RCoreTeam2018}{}}%
+R Core Team. (2018). \emph{{R: a language and environment for statistical computing}}. Vienna, Austria: R Foundation for Statistical Computing.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rabosky2018inverse}{}}%
+Rabosky, D.
+L., Chang, J., Title, P.
+O., Cowman, P.
+F., Sallan, L., Friedman, M., \ldots{} others.
+(2018).
+An inverse latitudinal gradient in speciation rate for marine fishes.
+\emph{Nature}, \emph{559}(7714), 392.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ramshaw1972time}{}}%
+Ramshaw, J., Richardson, D., Meatyard, B., Brown, R., Richardson, M., Thompson, E., \& Boulter, D.
+(1972).
+The time of origin of the flowering plants determined by using amino acid sequence data of cytochrome c.
+\emph{New Phytologist}, \emph{71}(5), 773--779.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ratnasingham2007bold}{}}%
+Ratnasingham, S., \& Hebert, P. D. (2007). BOLD: The barcode of life data system (http://www. Barcodinglife. org). \emph{Molecular Ecology Notes}, \emph{7}(3), 355--364.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rees2017automated}{}}%
+Rees, J. A., \& Cranston, K. (2017). Automated assembly of a reference taxonomy for phylogenetic data synthesis. \emph{Biodiversity Data Journal}, (5).
+
+\leavevmode\vadjust pre{\hypertarget{ref-Revell2012}{}}%
+Revell, L. J. (2012). Phytools: An r package for phylogenetic comparative biology (and other things). \emph{{Methods in Ecology and Evolution}}, \emph{3}, 217--223.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Ronquist2003}{}}%
+Ronquist, F., \& Huelsenbeck, J.
+P.
+(2003).
+{MrBayes 3: Bayesian phylogenetic inference under mixed models}.
+\emph{Bioinformatics}, \emph{19}(12), 1572--1574.
+\url{https://doi.org/10.1093/bioinformatics/btg180}
+
+\leavevmode\vadjust pre{\hypertarget{ref-datelifeplot}{}}%
+Sanchez-Reyes, L.
+L., \& O'Meara, B.
+(2022).
+{\texttt{datelifeplot}: Methods to plot chronograms and outputs of the \texttt{datelife} package}.
+\emph{R Package Release V0.2.2}.
+Retrieved from \url{https://zenodo.org/badge/latestdoi/381501451}
+
+\leavevmode\vadjust pre{\hypertarget{ref-datelifeR}{}}%
+Sanchez-Reyes, L.
+L., O'Meara, B., Eastman, J., Heath, T., Wright, A., Schliep, K., \ldots{} Alfaro, M.
+(2022).
+{\texttt{datelife}: Scientific Data on Time of Lineage Divergence for Your Taxa}.
+\emph{R Package Release V0.6.2}.
+Retrieved from \url{https://doi.org/10.5281/zenodo.593938}
+
+\leavevmode\vadjust pre{\hypertarget{ref-sanderson2002}{}}%
+Sanderson, Michael J.
+(2002).
+{Estimating Absolute Rates of Molecular Evolution and Divergence Times: A Penalized Likelihood Approach}.
+\emph{Molecular Biology and Evolution}, \emph{19}(1), 101--109.
+\url{https://doi.org/10.1093/oxfordjournals.molbev.a003974}
+
+\leavevmode\vadjust pre{\hypertarget{ref-sanderson2003r8s}{}}%
+Sanderson, Michael J. (2003). r8s: Inferring absolute rates of molecular evolution and divergence times in the absence of a molecular clock. \emph{Bioinformatics}, \emph{19}(2), 301--302.
+
+\leavevmode\vadjust pre{\hypertarget{ref-sanderson2001sources}{}}%
+Sanderson, Michael J., \& Doyle, J.
+A.
+(2001).
+Sources of error and confidence intervals in estimating the age of angiosperms from rbcL and 18S rDNA data.
+\emph{American Journal of Botany}, \emph{88}(8), 1499--1516.
+
+\leavevmode\vadjust pre{\hypertarget{ref-sauquet2013practical}{}}%
+Sauquet, H. (2013). A practical guide to molecular dating. \emph{{Comptes Rendus Palevol}}, \emph{12}(6), 355--367.
+
+\leavevmode\vadjust pre{\hypertarget{ref-sauquet2021age}{}}%
+Sauquet, H., Ramírez-Barahona, S., \& Magallón, S. (2021). \emph{The age of flowering plants is unknown}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-schenk2016sec}{}}%
+Schenk, J. J. (2016). {Consequences of secondary calibrations on divergence time estimates}. \emph{PLoS ONE}, \emph{11}(1). \url{https://doi.org/10.1371/journal.pone.0148228}
+
+\leavevmode\vadjust pre{\hypertarget{ref-schliep2011phangorn}{}}%
+Schliep, K. P. (2011). Phangorn: Phylogenetic analysis in r. \emph{Bioinformatics}, \emph{27}(4), 592--593.
+
+\leavevmode\vadjust pre{\hypertarget{ref-selvatti2015paleogene}{}}%
+Selvatti, A.
+P., Gonzaga, L.
+P., \& Moraes Russo, C.
+A.
+de.
+(2015).
+A paleogene origin for crown passerines and the diversification of the oscines in the new world.
+\emph{Molecular Phylogenetics and Evolution}, \emph{88}, 1--15.
+
+\leavevmode\vadjust pre{\hypertarget{ref-smith2018constructing}{}}%
+Smith, Stephen A., \& Brown, J. W. (2018). Constructing a broadly inclusive seed plant phylogeny. \emph{American Journal of Botany}, \emph{105}(3), 302--314.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Smith2012}{}}%
+Smith, Stephen A., \& O'Meara, B.
+C.
+(2012).
+{TreePL: Divergence time estimation using penalized likelihood for large phylogenies}.
+\emph{Bioinformatics}, \emph{28}(20), 2689--2690.
+\url{https://doi.org/10.1093/bioinformatics/bts492}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Stoltzfus2013}{}}%
+Stoltzfus, A., Lapp, H., Matasci, N., Deus, H., Sidlauskas, B., Zmasek, C.
+M., \ldots{} Jordan, G.
+(2013).
+{Phylotastic!
+Making tree-of-life knowledge accessible, reusable and convenient}.
+\emph{{BMC Bioinformatics}}, \emph{14}.
+\url{https://doi.org/10.1186/1471-2105-14-158}
+
+\leavevmode\vadjust pre{\hypertarget{ref-tietze2013complete}{}}%
+Tietze, D.
+T., Päckert, M., Martens, J., Lehmann, H., \& Sun, Y.-H.
+(2013).
+Complete phylogeny and historical biogeography of true rosefinches (aves: carpodacus).
+\emph{Zoological Journal of the Linnean Society}, \emph{169}(1), 215--234.
+
+\leavevmode\vadjust pre{\hypertarget{ref-title2016macrophylogenies}{}}%
+Title, P.
+O., \& Rabosky, D.
+L.
+(2016).
+{Do Macrophylogenies Yield Stable Macroevolutionary Inferences?
+An Example from Squamate Reptiles}.
+\emph{Systematic Biology}, syw102.
+\url{https://doi.org/10.1093/sysbio/syw102}
+
+\leavevmode\vadjust pre{\hypertarget{ref-treplin2008molecular}{}}%
+Treplin, S., Siegert, R., Bleidorn, C., Thompson, H.
+S., Fotso, R., \& Tiedemann, R.
+(2008).
+Molecular phylogeny of songbirds (aves: Passeriformes) and the relative utility of common nuclear marker loci.
+\emph{Cladistics}, \emph{24}(3), 328--349.
+
+\leavevmode\vadjust pre{\hypertarget{ref-vos2012nexml}{}}%
+Vos, R.
+A., Balhoff, J.
+P., Caravas, J.
+A., Holder, M.
+T., Lapp, H., Maddison, W.
+P., \ldots{} others.
+(2012).
+NeXML: Rich, extensible, and verifiable representation of comparative data and metadata.
+\emph{Systematic Biology}, \emph{61}(4), 675--689.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Webb2000}{}}%
+Webb, C. (2000). {Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees}. \emph{{The American Naturalist}}, \emph{156}(2), 145--155.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Webb2008}{}}%
+Webb, Campbell O., Ackerly, D.
+D., \& Kembel, S.
+W.
+(2008).
+{Phylocom: Software for the analysis of phylogenetic community structure and trait evolution}.
+\emph{Bioinformatics}, \emph{24}(18), 2098--2100.
+\url{https://doi.org/10.1093/bioinformatics/btn358}
+
+\leavevmode\vadjust pre{\hypertarget{ref-webb2005phylomatic}{}}%
+Webb, Campbell O., \& Donoghue, M. J. (2005). Phylomatic: Tree assembly for applied phylogenetics. \emph{{Molecular Ecology Notes}}, \emph{5}(1), 181--183.
+
+\leavevmode\vadjust pre{\hypertarget{ref-weir2008calibrating}{}}%
+Weir, J., \& Schluter, D. (2008). Calibrating the avian molecular clock. \emph{Molecular Ecology}, \emph{17}(10), 2321--2328.
+
+\end{CSLReferences}
+
+\endgroup
 

### 2022-06-22 manuscript/fig-benchmark.Rmd
fixing format of embedded figures and tables

@@
 \begin{center}
-\includegraphics[width= 6.5in]{../figures/fig_runtime_main.pdf}
+  \includegraphics{../figures/fig_runtime_main.pdf}
 \end{center}
-\captionof{figure}{Computation time used to process a datelife query and a datelife search across \texttt{datelife}'s chronogram database, relative to number of input taxon names.
-For each N = {10, 100, 200, ..., 1 000, ...
-, 9 000, 10 000}, we sampled N species names from the class Aves a hundred times, and then performed a datelife search with input processing, without implementing Taxon Names Resolution Service (TNRS; dark gray), and implementing TNRS (light gray).
-We also performed a search using a query that had been pre-processed with TNRS, for comparison (light blue).}
+  \captionof{figure}{DateLife's benchmarking results. Computation time used to process a query and a search across \texttt{datelife}'s chronogram database, relative to number of input taxon names.
+  \textit{Caption continues.}}
   \label{fig:runtime_main}
+\begin{justify}
+  \textit{Figure caption continued}. 
+For each N = \{10, 100, 200, ..., 1 000, ...
+, 9 000, 10 000\}, we sampled N species names from the class Aves a hundred times, and then performed a datelife search processing the input names with Taxon Names Resolution Service (TNRS; light gray), and without processing names (dark gray).
+For comparison, we performed a search using an input that had been pre-processed with TNRS (light blue).
+\end{justify}
+
+\vspace{20mm}

### 2022-06-22 manuscript/fig-fringillidae-full-example.Rmd
fixing format of embedded figures and tables

--- 2022-05-28 manuscript/figures-alt.tex
@@
-% in this file I leave the figure captions outside\ caption{} because I want them
-% to be formatted in the same way as the general text (double spaced and linenumbered)
-\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}
+---
+title: "Figure Fringillidae full Example"
+output: html_document
+date: '2022-06-21'
+---
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 1    %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!h]
-\includegraphics{../figures/figure1/figure1-new.pdf}
-\caption{Stylized DateLife workflow.
-This shows the general worflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website  at \url{http://www.datelife.org/}.
-Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.
-}
-\label{fig:workflow}
-\end{figure}
+\begin{minipage}{18cm}
+\begin{center}
+\includegraphics{../figures/figure-fringillidae/median_and_calibration_ages-congruified.png}
+\end{center}
 
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE runtime    %%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!h]
-\includegraphics[width=1\linewidth]{../figures/fig_runtime_main.pdf}
-\caption{Computation time of query processing and search across \texttt{datelife}'s chronogram database relative to number of input taxon names.
-We sampled N names from the class Aves for each cohort 100 times and then performed a search with query processing not using the Taxon Names Resoultion Service (TNRS; dark gray), and using TNRS (light gray).
-We also performed a search using the already processed query for comparison (light blue).}
-\label{fig:runtime_main}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 2    %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!h]
-\includegraphics[width=1\linewidth]{../figures/figure2/figure2.pdf}
-\caption{DateLife analysis results for a small sample of A) 6 bird species within the Passeriformes.
-B) Processed species names were found across 9 chronograms within 6 independent studies (Barker et al.
-(\protect\hyperlink{ref-barker2012going}{2012}), Barker et al.
-(\protect\hyperlink{ref-barker2015new}{2015}), Burns et al.
-(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al.
-(\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al.
-(\protect\hyperlink{ref-Jetz2012}{2012}).) C) This revealed 28 source age data points for the queried species names.
-Summarized age data is used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree, resulting in a summary chronogram of source ages.}
-\label{fig:figure2}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 2-1  %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\begin{figure}[!h]
-%\includegraphics[width=1\linewidth]{../figures/figure2/figure2-1.pdf}
-%\caption{Age data results of a DateLife search of a small sample of 6 %bird species within the Passeriformes.
-Input names were found across 9 %chronograms within 6 independent studies (Barker et al.
-%(\protect\hyperlink{ref-barker2012going}{2012}), Barker et al.
-%(\protect\hyperlink{ref-barker2015new}{2015}), Burns et al.
-%(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al.
-%(\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price %(\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al.
-%(\protect\hyperlink{ref-Jetz2012}{2012}).) This revealed 28 age data %points for the queried species names.}
-%\label{fig:figure2-1}
-%\end{figure}
-%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 2-2  %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%\begin{figure}[!h]
-%\includegraphics{../figures/figure2/figure2-2.pdf}
-%\caption{Summarized age data is used as secondary calibrations to date %a tree topology as a summary chronogram.}
-%\label{fig:summaries}
-%\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 3    %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!h]
-\includegraphics{../figures/fringillidae-topologies/fringillidae-topology.pdf}
-\caption{Tree topologies obtained from Open Tree of Life's (OpenTree) synthetic phylogenetic tree.
-A) Topology of 2,333 tips and 1,305 internal nodes, encompassing bird species within the family Fringillidae following the NCBI taxonomy (black), as well as all other bird species that share the same mrca node in OpenTree's synthetic tree (purple).
-B) Topology of 289 tips and 253 internal nodes, encompassing bird species within the Fringillidae only, resulting from pruning purple branches from topology A.
-Bird species within the Fringillidae do not form a monophyletic group
-(Alström et al. \protect\hyperlink{ref-}{2014},
-Barker, Cibois, Schikler, Feinstein, \& Cracraft \protect\hyperlink{ref-barker2004phylogeny}{2004},
-Barker et al. \protect\hyperlink{ref-barker2013going}{2013},
-Barker \protect\hyperlink{ref-barker2014mitogenomic}{2014},
-Barker et al. \protect\hyperlink{ref-barker2015new}{2015},
-Beresford, Barker, Ryan, \& Crowe \protect\hyperlink{ref-beresford2005african}{2005},
-Bryson Jr et al. \protect\hyperlink{ref-bryson2014diversification}{2014},
-Burleigh, Kimball, \& Braun \protect\hyperlink{ref-burleigh2015building}{2015},
-Burns et al. \protect\hyperlink{ref-burns2014phylogenetics}{2014},
-Chaves, Hidalgo, \& Klicka \protect\hyperlink{ref-chaves2013biogeography}{2013},
-Claramunt \& Cracraft \protect\hyperlink{ref-claramunt2015new}{2015},
-Gibb et al. \protect\hyperlink{ref-gibb2015new}{2015},
-Hackett et al. \protect\hyperlink{ref-hackett2008phylogenomic}{2008},
-Jetz et al. \protect\hyperlink{ref-Jetz2012}{2012},
-Johansson, Fjeldså, \& Bowi \protect\hyperlink{ref-johansson2008phylogenetic}{200},
-Kimball et al. \protect\hyperlink{ref-kimball2019phylogenomic}{2019},
-Klicka et al. \protect\hyperlink{ref-klicka2014comprehensive}{2014},
-Lamichhaney et al. \protect\hyperlink{ref-lamichhaney2015evolution}{2015},
-Lerner, Meyer, James, Hofreiter, \& Fleischer \protect\hyperlink{ref-lerner2011multilocus}{2011},
-Lovette et al. \protect\hyperlink{ref-lovette2010comprehensive}{2010},
-Moyle et al. \protect\hyperlink{ref-moyle2016tectonic}{2016},
-Ödeen, Håstad, \& Alström \protect\hyperlink{ref-odeen2011evolution}{2011},
-Oliveros et al. \protect\hyperlink{ref-oliveros2019earth}{2019},
-Päckert et al. \protect\hyperlink{ref-packert2012horizontal}{2012},
-Parchman, Benkman, \& Mezquida \protect\hyperlink{ref-parchman2007coevolution}{2007},
-Powell et al. \protect\hyperlink{ref-powell2014comprehensive}{2014},
-Price et al. \protect\hyperlink{ref-price2014niche}{2014},
-Pulgarín-R, Smith, Bryson Jr, Spellman, \& Klicka \protect\hyperlink{ref-pulgarin2013multilocus}{2013},
-Selvatti, Gonzaga, \& Moraes Russo \protect\hyperlink{ref-selvatti2015paleogene}{2015},
-Tietze, Päckert, Martens, Lehmann, \& Sun \protect\hyperlink{ref-tietze2013complete}{2013},
-Treplin et al. \protect\hyperlink{ref-treplin2008molecular}{2008},
-Zuccon, Prŷs-Jones, Rasmussen, \& Ericson \protect\hyperlink{ref-zuccon2012phylogenetic}{2012}).
-}
-\label{fig:fringillidae-topologies}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%    FIGURE 4    %%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\begin{figure}[!h]
-\includegraphics{../figures/figure-fringillidae/median_and_calibration_ages_simple.png}
-\caption{Fringillidae median summary chronogram generated with DateLife. It has 256 tips and 233 nodes.}
+\captionof{figure}{Fringillidae median summary chronogram generated with DateLife. It has 256 tips and 233 nodes, from which 212 have age data from at least one published chronogram.}
 \label{fig:fringillidages}
-\end{figure}
+\end{minipage}

### 2022-06-22 manuscript/fig-topologies.Rmd
fixing format of embedded figures and tables

@@
 
-\vspace{20mm}
-
-\begin{center}
-\includegraphics[width = 6.5in]{../figures/fringillidae-topologies/fringillidae-topology.pdf}
+\begin{minipage}{18.2cm}
+\includegraphics{../figures/fringillidae-topologies/fringillidae-topology.pdf}
   \captionof{figure}{Tree topologies obtained from Open Tree of Life's (OpenTree) synthetic phylogenetic tree.
@@
   Treplin et al. \protect\hyperlink{ref-treplin2008molecular}{2008},
-Zuccon, Prŷs-Jones, Rasmussen, \& Ericson \protect\hyperlink{ref-zuccon2012phylogenetic}{2012}).
-}
+  Zuccon, Prŷs-Jones, Rasmussen, \& Ericson \protect\hyperlink{ref-zuccon2012phylogenetic}{2012}).}
 \label{fig:fringillidae-topologies}
-\end{center}
+\end{minipage}

### 2022-06-22 manuscript/fig-workflow.Rmd
fixing format of embedded figures and tables

---
title: "fig1"
output: html_document
date: '2022-06-21'
---
\begin{minipage}[t]{17.3cm}
\begin{center}
\includegraphics{../figures/figure1/figure1-new.pdf}
\captionof{figure}{Main DateLife workflow. Analyses can be performed via DateLife's interactive website at \url{www.datelife.org}, or using the \texttt{datelife} R package. Details on the R functions used to perform the analyses are available from \texttt{datelife}'s R package vignettes at \url{phylotastic.org/datelife}.}
\label{fig:workflow}
\end{center}
\end{minipage}

### 2022-06-22 manuscript/manuscript.Rmd
fixing format of embedded figures and tables

@@
   @treplin2008molecular,
-  @zuccon2012phylogeneti,
+  @zuccon2012phylogenetic
 ---
@@
 
-<!--_**Processing an input.--**_ -->
-
-```{r fig1, child= if (embed_figure) "fig1.Rmd"}
+```{r fig1, child= if (embed_figure) "fig-workflow.Rmd"}
 ```
@@
 
-```{r fig2, child= if (embed_figure) "fig-benchmark.Rmd"}
-```
-
 # Benchmark
@@
 <!--Speed of web interface and of r package in computers with different capacities? NOOOO-->
+
+```{r fig2, child= if (embed_figure) "fig-benchmark.Rmd"}
+```
+
 `datelife`'s code performance was evaluated with a set of unit tests designed and
@@
 and one species of tanager, the vegetarian finch -- *Platyspiza crassirostris*.
-
-Processing input names found that *Emberiza elegans* is synonym for *Schoeniclus elegans* in the default reference taxonomy (Open Tree of Life Taxonomy v3.3, June 1, 2021).
+Processing of input names found that *Emberiza elegans* is synonym for *Schoeniclus elegans* in the default reference taxonomy (Open Tree of Life Taxonomy v3.3, June 1, 2021).
 For a detailed discussion on the state of the synonym, refer to Avibase [@lepage2004avibase; @lepage2014avibase; @avibase-emberiza].
@@
 -->
-\vspace{5mm}
+
+\vspace{3mm}
 
@@
 
-\vspace{5mm}
+\vspace{3mm}
 
@@
 Summary age data for node "n2" were excluded as final calibration because they are older than age data of the more inclusive node, "n1" (Fig. \ref{fig:figure2}C4).
+
+\vspace{30mm}
 
@@
 
-## The family Fringillidae
+## An example with the family of true finches
 
@@
 
-Age data from source chronograms was congruified, decreasing to 818 different age data points available (supplementray table 2) and then age summary statistics per congruified node are obtained and used as fixed secondary calibrations over a consensus tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
+\vspace{50mm}
+
+```{r fig4, child= if (embed_figure) "fig-topologies.Rmd"}
+```
+
+Age data from source chronograms was congruified to the chosen tree topology (OpenTree's synthetic subtree from figure \ref{fig:fringillidae-topologies}B), reducing the age data set to 818 different data points (supplementray table 2).
+Age summary statistics per congruified node were calculated and used as fixed secondary calibrations over the chosen tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
 \ref{fig:fringillidages}).
-Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
+
 <!-- ????????????????? Say some things about the results!
 ANSWER:
-(Going from a summary distance matrix Using this method is better than using clustering algorithms  -->
-<!-- and hierarchical clustering method which appears to push ages back -->
+Compare median versus SDM:
+Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
 
-```{r fig4, child= if (embed_figure) "fig-topologies.Rmd"}
+-->
+
+```{r fig5, child= if (embed_figure) "fig-fringillidae-full-example.Rmd"}
 ```
@@
 
-We found that node ages from original study, and ages estimated using all other age data available are correlated (Supplementary Fig. \ref{fig:cvXY}).
+We found that node ages from original study, and ages estimated using all other age data available are correlated (Fig. \ref{fig:cvXY}).
 
-For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka 'closer to the root') relative to the orginal estimate, and overestimate ages for nodes closer to the tips.
+For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka 'closer to the root') relative to the original estimate, and overestimate ages for nodes closer to the tips.
 Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. \ref{fig:cvXYdiffs}).
+
+```{r fig6, child= if (embed_figure) "fig-cross-validation.Rmd"}
+```
 

### 2022-06-22 manuscript/manuscript.tex
fixing format of embedded figures and tables

@@
 
-% leave the figure captions outside\ caption{} because I want them
+% leave the figure captions outside\ caption{} if you want them
 % to be formatted in the same way as the general text (double spaced and linenumbered)
-
-\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}
+% line numbers for captions trick from https://latex.org/forum/viewtopic.php?t=34614
+
+\usepackage{ragged2e}
+
 \usepackage{caption}
+
+\DeclareCaptionFont{linenumbers}{\internallinenumbers}
+
 \usepackage{float}
+\restylefloat{table}
+
+\captionsetup[figure]{font+=linenumbers, labelfont={sc}, labelformat={default}, labelsep=period, name={Figure}}
+
+\captionsetup[table]{font+=linenumbers, labelfont={sc}, labelformat={default}, labelsep=period, name={Table}}
+
 % Manuscript styling
@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4906}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4912}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 
+\begin{minipage}[t]{17.3cm}
 \begin{center}
-\includegraphics[width = 6.5in]{../figures/figure1/figure1-new.pdf}
-\captionof{figure}{Stylized DateLife workflow.
-This shows the general workflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website at \url{http://www.datelife.org/}.
-Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.}
+\includegraphics{../figures/figure1/figure1-new.pdf}
+\captionof{figure}{Main DateLife workflow.
+Analyses can be performed via DateLife's interactive website at \url{www.datelife.org}, or using the \texttt{datelife} R package.
+Details on the R functions used to perform the analyses are available from \texttt{datelife}'s R package vignettes at \url{phylotastic.org/datelife}.}
 \label{fig:workflow}
 \end{center}
+\end{minipage}
 
@@
 At the time of writing of this manuscript
-(Jun 20, 2022),
+(Jun 21, 2022),
 DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies.
@@
 
-\begin{center}
-\includegraphics[width= 6.5in]{../figures/fig_runtime_main.pdf}
-\end{center}
-\captionof{figure}{Computation time used to process a datelife query and a datelife search across \texttt{datelife}'s chronogram database, relative to number of input taxon names.
-For each N = {10, 100, 200, ..., 1 000, ...
-, 9 000, 10 000}, we sampled N species names from the class Aves a hundred times, and then performed a datelife search with input processing, without implementing Taxon Names Resolution Service (TNRS; dark gray), and implementing TNRS (light gray).
-We also performed a search using a query that had been pre-processed with TNRS, for comparison (light blue).}
-
-\label{fig:runtime_main}
-
 \hypertarget{benchmark}{%
@@
 \ref{fig:runtime_main}).
+
+\begin{center}
+  \includegraphics{../figures/fig_runtime_main.pdf}
+\end{center}
+  \captionof{figure}{DateLife's benchmarking results. Computation time used to process a query and a search across \texttt{datelife}'s chronogram database, relative to number of input taxon names.
+  \textit{Caption continues.}}
+
+\label{fig:runtime_main}
+
+\begin{justify}
+  \textit{Figure caption continued}. 
+For each N = \{10, 100, 200, ..., 1 000, ...
+, 9 000, 10 000\}, we sampled N species names from the class Aves a hundred times, and then performed a datelife search processing the input names with Taxon Names Resolution Service (TNRS; light gray), and without processing names (dark gray).
+For comparison, we performed a search using an input that had been pre-processed with TNRS (light blue).
+\end{justify}
+
+\vspace{20mm}
+
 \texttt{datelife}'s code performance was evaluated with a set of unit tests designed and
@@
 and one species of tanager, the vegetarian finch -- \emph{Platyspiza crassirostris}.
-
-Processing input names found that \emph{Emberiza elegans} is synonym for \emph{Schoeniclus elegans} in the default reference taxonomy (Open Tree of Life Taxonomy v3.3, June 1, 2021).
+Processing of input names found that \emph{Emberiza elegans} is synonym for \emph{Schoeniclus elegans} in the default reference taxonomy (Open Tree of Life Taxonomy v3.3, June 1, 2021).
 For a detailed discussion on the state of the synonym, refer to Avibase (Avibase, 2022; Lepage, 2004; Lepage, Vaidya, \& Guralnick, 2014).
 Discovering this synonym allowed assigning five age data points for the parent node of \emph{Emberiza elegans}, shown as \emph{Schoeniclus elegans} in figure \ref{fig:figure2}A, which would not have had any data otherwise.
-\vspace{5mm}
+
+\vspace{3mm}
 
@@
 
+\vspace{30mm}
+
+\begin{center}
+\includegraphics{../figures/figure2/figure2.pdf}
+\end{center}
 \vspace{10mm}
-
-\begin{center}
-\includegraphics[width = 6.5in]{../figures/figure2/figure2.pdf}
-\end{center}
 \captionof{figure}{DateLife analysis results for a small sample of A) 6 bird species within the Passeriformes.
@@
 
-\vspace{15mm}
-
 \begin{table}[H]
@@
 
-\hypertarget{the-family-fringillidae}{%
-\subsection{The family Fringillidae}\label{the-family-fringillidae}}
+\hypertarget{an-example-with-the-family-of-true-finches}{%
+\subsection{An example with the family of true finches}\label{an-example-with-the-family-of-true-finches}}
 
@@
 
-Age data from source chronograms was congruified, decreasing to 818 different age data points available (supplementray table 2) and then age summary statistics per congruified node are obtained and used as fixed secondary calibrations over a consensus tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
-\ref{fig:fringillidages}).
-Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
-
-\vspace{20mm}
-
-\begin{center}
-\includegraphics[width = 6.5in]{../figures/fringillidae-topologies/fringillidae-topology.pdf}
+\vspace{50mm}
+
+\begin{minipage}{18.2cm}
+\includegraphics{../figures/fringillidae-topologies/fringillidae-topology.pdf}
   \captionof{figure}{Tree topologies obtained from Open Tree of Life's (OpenTree) synthetic phylogenetic tree.
@@
   Treplin et al. \protect\hyperlink{ref-treplin2008molecular}{2008},
-Zuccon, Prŷs-Jones, Rasmussen, \& Ericson \protect\hyperlink{ref-zuccon2012phylogenetic}{2012}).
-}
+  Zuccon, Prŷs-Jones, Rasmussen, \& Ericson \protect\hyperlink{ref-zuccon2012phylogenetic}{2012}).}
 \label{fig:fringillidae-topologies}
+\end{minipage}
+
+Age data from source chronograms was congruified to the chosen tree topology (OpenTree's synthetic subtree from figure \ref{fig:fringillidae-topologies}B), reducing the age data set to 818 different data points (supplementray table 2).
+Age summary statistics per congruified node were calculated and used as fixed secondary calibrations over the chosen tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
+\ref{fig:fringillidages}).
+
+\begin{minipage}{18cm}
+\begin{center}
+\includegraphics{../figures/figure-fringillidae/median_and_calibration_ages-congruified.png}
 \end{center}
+
+\captionof{figure}{Fringillidae median summary chronogram generated with DateLife. It has 256 tips and 233 nodes, from which 212 have age data from at least one published chronogram.}
+\label{fig:fringillidages}
+\end{minipage}
 
@@
 
-We found that node ages from original study, and ages estimated using all other age data available are correlated (Supplementary Fig. \ref{fig:cvXY}).
-
-For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka `closer to the root') relative to the orginal estimate, and overestimate ages for nodes closer to the tips.
+We found that node ages from original study, and ages estimated using all other age data available are correlated (Fig. \ref{fig:cvXY}).
+
+For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka `closer to the root') relative to the original estimate, and overestimate ages for nodes closer to the tips.
 Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. \ref{fig:cvXYdiffs}).
+
+\newpage
+
+\begin{minipage}{18cm}
+\begin{center}
+\includegraphics{../figures/figure-cross-validation/fig-cross-validation-xy-plots.pdf}
+\end{center}
+\captionof{figure}{Results from cross validation analysis. Each plot compares the original age estimate (x axis) with the age obtained with a DateLife analysis (y axis), per node.}
+\label{fig:cvXY}
+\end{minipage}
 

### 2022-06-22 manuscript/manuscript.Rmd
homogenizing to 'datelife workflow' instead of 'datelife algorithm'

@@
 taxize [@Chamberlain2013; @Chamberlain2018].
-Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife algorithm: creating a search query, searching a database, and summarizing results from the search.
+Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife workflow: creating a search query, searching a database, and summarizing results from the search.
 
@@
 Subspecies and variants are ignored.
-If an input taxon name belongs to an inclusive taxonomic group, the algorithm has two alternative behaviors defined by the "get species from taxon" flag.
+If an input taxon name belongs to an inclusive taxonomic group, DateLife has two alternative behaviors defined by the "get species from taxon" flag.
 If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group (following a taxonomy of choice; current options are Open Tree of Life, NCBI, GBIF, or IRMNG taxonomies) and adds them to the input string.
@@
 Alternatively, DateLife can combine topologies from source chronograms using a supertree approach.
-To combine topologies from source chronograms into a single summary (or supertree) topology, the DateLife algorithm starts by identifying the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
+To combine topologies from source chronograms into a single summary (or supertree) topology, the DateLife workflow identifies the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
 [-@ane2009groves].
@@
 BLADJ does not use branch lengths even when they are present in the input tree or summarizing topology.
-When there is conflict in ages between nodes with calibration data, the algorithm ignores node ages that are older than the age of a parent node.
+When there is conflict in ages between nodes with calibration data, BLADJ ignores node ages that are older than the age of a parent node.
 BLADJ requires a root age estimate.
@@
 
-We illustrate the DateLife algorithm using a family within the Passeriform birds encompassing the true finches, Fringillidae, as case study.
-The first example analyses 6 bird species and shows all steps of the algorithm.
-The second example is an application analysing 289 species in the family Fringillidae that are included in the NCBI taxonomy.
+We illustrate the DateLife workflow using a family within the Passeriform birds encompassing the true finches, Fringillidae, as case study.
+The first example analyses 6 bird species and shows all steps of the workflow The second example is an analysis of 289 species in the family Fringillidae that are included in the NCBI taxonomy.
 <!--We used `r cite_r("r-references.bib")` for all our analyses.-->
@@
 For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka 'closer to the root') relative to the original estimate, and overestimate ages for nodes closer to the tips.
-Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. \ref{fig:cvXYdiffs}).
+Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. S1).
 
@@
 When ages are estimated with secondary calibrations, nodes closer to the root do tend to be slightly younger than ages estimated with primary calibrations.
-However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures 1-9), with a notable exception of cross validation of @burns2014phylogenetics chronogram, which results in much younger node ages when estimated using secondary calibrations (Supplementary figs.
-1 and 4).
+However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures S2-S9).
+The only excpetion to tjis was observed on results of the cross validation analysis of the @burns2014phylogenetics chronogram, which results in much younger node ages when estimated using secondary calibrations (Supplementary figs.
+S1 and S4).
 
@@
 For example, the chronograms from @burns2014phylogenetics were inferred using molecular substitution rate estimates across birds [@weir2008calibrating], and have much older date estimates for the same nodes than chronograms that were inferred using fossils as primary calibrations (figs.
-fig:fringillidages and fig:cvXY; supplementary figs.
-1 and 4).
+\ref{fig:fringillidages}, \ref{fig:cvXY}; supplementary figs.
+S1, S4).
 <!--\textcolor{red}{EJM: MORE ON the FINCH examples and the cross validation here}-->
@@
 \endgroup
+
+```{r fig1, child= if (!embed_figure) c("table1.Rmd", "table2.Rmd", "fig-workflow.Rmd", "fig-benchmark.Rmd", "fig-small-example.Rmd", "fig-topologies.Rmd", "fig-fringillidae-full-example.Rmd", "fig-cross-validation.Rmd")}
+```

### 2022-06-22 manuscript/manuscript.tex
homogenizing to 'datelife workflow' instead of 'datelife algorithm'

@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4912}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4914}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 taxize (Chamberlain \& Szöcs, 2013; Chamberlain et al., 2019).
-Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife algorithm: creating a search query, searching a database, and summarizing results from the search.
+Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife workflow: creating a search query, searching a database, and summarizing results from the search.
 
@@
 Subspecies and variants are ignored.
-If an input taxon name belongs to an inclusive taxonomic group, the algorithm has two alternative behaviors defined by the ``get species from taxon'' flag.
+If an input taxon name belongs to an inclusive taxonomic group, DateLife has two alternative behaviors defined by the ``get species from taxon'' flag.
 If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group (following a taxonomy of choice; current options are Open Tree of Life, NCBI, GBIF, or IRMNG taxonomies) and adds them to the input string.
@@
 Alternatively, DateLife can combine topologies from source chronograms using a supertree approach.
-To combine topologies from source chronograms into a single summary (or supertree) topology, the DateLife algorithm starts by identifying the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
+To combine topologies from source chronograms into a single summary (or supertree) topology, the DateLife workflow identifies the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
 (2009).
@@
 
-We illustrate the DateLife algorithm using a family within the Passeriform birds encompassing the true finches, Fringillidae, as case study.
-The first example analyses 6 bird species and shows all steps of the algorithm.
-The second example is an application analysing 289 species in the family Fringillidae that are included in the NCBI taxonomy.
+We illustrate the DateLife workflow using a family within the Passeriform birds encompassing the true finches, Fringillidae, as case study.
+The first example analyses 6 bird species and shows all steps of the workflow The second example is an analysis of 289 species in the family Fringillidae that are included in the NCBI taxonomy.
 
@@
 For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka `closer to the root') relative to the original estimate, and overestimate ages for nodes closer to the tips.
-Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. \ref{fig:cvXYdiffs}).
+Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. S1).
 
@@
 When ages are estimated with secondary calibrations, nodes closer to the root do tend to be slightly younger than ages estimated with primary calibrations.
-However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures 1-9), with a notable exception of cross validation of Burns et al.
+However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures S2-S9).
+The only excpetion to tjis was observed on results of the cross validation analysis of the Burns et al.
 (2014) chronogram, which results in much younger node ages when estimated using secondary calibrations (Supplementary figs.
-1 and 4).
+S1 and S4).
 
@@
 (2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older date estimates for the same nodes than chronograms that were inferred using fossils as primary calibrations (figs.
-fig:fringillidages and fig:cvXY; supplementary figs.
-1 and 4).
+\ref{fig:fringillidages}, \ref{fig:cvXY}; supplementary figs.
+S1, S4).
 

### 2022-06-22 manuscript/preamble-suppl.tex
homogenizing to 'datelife workflow' instead of 'datelife algorithm'

--- 2022-06-21 manuscript/manuscript.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[
-  man]{apa6}
-\usepackage{amsmath,amssymb}
-\usepackage{lmodern}
-\usepackage{iftex}
-\ifPDFTeX
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life},
-  pdfauthor={Luna L. Sánchez Reyes1,2, Emily Jane McTavish1, \& Brian O'Meara2},
-  pdflang={en-EN},
-  pdfkeywords={Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-\ifLuaTeX
-\usepackage[bidi=basic]{babel}
-\else
-\usepackage[bidi=default]{babel}
-\fi
-\babelprovide[main,import]{english}
-% get rid of language-specific shorthands (see #6817):
-\let\LanguageShortHands\languageshorthands
-\def\languageshorthands#1{}
 \DeclareUnicodeCharacter{0301}{*************************************}
 
-% leave the figure captions outside\ caption{} because I want them
+% leave the figure captions outside\ caption{} if you want them
 % to be formatted in the same way as the general text (double spaced and linenumbered)
+% line numbers for captions trick from https://latex.org/forum/viewtopic.php?t=34614
 
-\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Figure}}
+\usepackage{ragged2e}
+
 \usepackage{caption}
+
+\usepackage{lineno}
+\DeclareCaptionFont{linenumbers}{\internallinenumbers}
+
 \usepackage{float}
-% Manuscript styling
-\usepackage{upgreek}
-\captionsetup{font=singlespacing,justification=justified}
+\restylefloat{table}
 
-% Table formatting
-\usepackage{longtable}
-\usepackage{lscape}
-% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
-\usepackage{multirow}		% Table styling
-\usepackage{tabularx}		% Control Column width
-\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
-\usepackage{threeparttablex}            % Lets threeparttable work with longtable
+\captionsetup[figure]{font+=linenumbers, labelfont={sc}, labelformat={default}, labelsep=period, name={Supplementary Figure}}
 
-% Create new environments so endfloat can handle them
-% \newenvironment{ltable}
-%   {\begin{landscape}\centering\begin{threeparttable}}
-%   {\end{threeparttable}\end{landscape}}
-\newenvironment{lltable}{\begin{landscape}\centering\begin{ThreePartTable}}{\end{ThreePartTable}\end{landscape}}
-
-% Enables adjusting longtable caption width to table width
-% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
-\makeatletter
-\newcommand\LastLTentrywidth{1em}
-\newlength\longtablewidth
-\setlength{\longtablewidth}{1in}
-\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}
-
-% \setlength{\parindent}{0.5in}
-% \setlength{\parskip}{0pt plus 0pt minus 0pt}
-
-% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
-% See https://github.com/crsh/papaja/issues/292
-\makeatletter
-\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-1em}%
-  {\normalfont\normalsize\bfseries\itshape\typesectitle}}
-
-\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-\z@\relax}%
-  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
-\makeatother
-
-% \usepackage{etoolbox}
-\makeatletter
-\patchcmd{\HyOrg@maketitle}
-  {\section{\normalfont\normalsize\abstractname}}
-  {\section*{\normalfont\normalsize\abstractname}}
-  {}{\typeout{Failed to patch abstract.}}
-\patchcmd{\HyOrg@maketitle}
-  {\section{\protect\normalfont{\@title}}}
-  {\section*{\protect\normalfont{\@title}}}
-  {}{\typeout{Failed to patch title.}}
-\makeatother
-
-\usepackage{xpatch}
-\makeatletter
-\xapptocmd\appendix
-  {\xapptocmd\section
-    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
-    {}{\InnerPatchFailed}%
-  }
-{}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4906}
-\DeclareDelayedFloatFlavor{ThreePartTable}{table}
-\DeclareDelayedFloatFlavor{lltable}{table}
-\DeclareDelayedFloatFlavor*{longtable}{table}
-\makeatletter
-\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
-\makeatother
-\usepackage{lineno}
-
-\linenumbers
-\usepackage{csquotes}
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-
-\title{DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life}
-\author{Luna L. Sánchez Reyes\textsuperscript{1,2}, Emily Jane McTavish\textsuperscript{1}, \& Brian O'Meara\textsuperscript{2}}
-\date{}
-
-\shorttitle{DateLife: revealing the dated Tree of Life}
-
-\authornote{
-
-School of Natural Sciences, University of California, Merced, Science and Engineering Building 1.
-
-Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 446 Hesler Biology Building, Knoxville, TN 37996, USA.
-
-The authors made the following contributions.
-Luna L.
-Sánchez Reyes: Data curation, Investigation, Software, Visualization, Validation, Writing - Original Draft Preparation, Writing - Review \& Editing; Emily Jane McTavish: Resources, Software, Writing - Review \& Editing; Brian O'Meara: Conceptualization, Funding acquisition, Methodology, Resources, Software, Supervision, Writing - Review \& Editing.
-
-Correspondence concerning this article should be addressed to Luna L. Sánchez Reyes, . E-mail: \href{mailto:sanchez.reyes.luna@gmail.com}{\nolinkurl{sanchez.reyes.luna@gmail.com}}
-
-}
-
-\affiliation{\vspace{0.5cm}\textsuperscript{1} University of California, Merced\\\textsuperscript{2} University of Tennessee, Knoxville}
-
-\abstract{%
-Achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram) is a difficult and time-consuming task.
-But the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
-Knowledge of the evolutionary time frame of organisms is key for research in the natural sciences. It also represent valuable information for education, science communication, and policy decisions.
-When chronograms are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on evolutionary timeframe is exposed in a programatic and reusable way, as intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
-Here we present DateLife, a service implemented as an R package and an R Shiny website application available at www.datelife.org, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time frame of evolution.
-The main DateLife workflow constructs a chronogram for any given combination of taxon names by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from the TreeBASE database as well.
-We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
-DateLife will be useful to increase awereness of the existing variation in alternative hypothesis of evolutionary time for the same organisms, and can foster exploration of the effect of alternative evolutionary timing hypotheses on the results of downstream analyses, providing a framework for a more informed interpretation of evolutionary results.
-}
-
-\begin{document}
-\maketitle
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
-Chronograms --phylogenies with branch lengths proportional to time-- provide key data on evolutionary time frame for the study of natural processes in many areas of biological research, such as developmental biology (Delsuc et al., 2018; Laubichler \& Maienschein, 2009), conservation biology (Felsenstein, 1985; C.
-Webb, 2000), historical biogeography (Posadas, Crisci, \& Katinas, 2006), and species diversification (Magallon \& Sanderson, 2001; Morlon, 2014).
-
-Building a chronogram is not an easy task.
-It requires obtaining and curating data to construct a phylogeny, selecting and placing appropriate calibrations on the phylogeny using independent age data points from the fossil record or other dated events, and inferring the full dated tree;
-it also generally requires specialized biological training, taxonomic domain knowledge, and a non-negligible amount of research time, computational resources and funding.
-
-Here we present the DateLife project which has the main goal of capturing age data from published chronograms, and making these data readily accessible to the community for reuse and reanalysis, for research, teaching, and science communication and policy.
-DateLife's core software application is available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org.
-It features key elements for scientific reproducibility, such as a versioned, open and fully public source database (McTavish et al., 2015), data stored and available in a computer readable format (Vos et al., 2012), automated and programmatic ways of accessing the data (Stoltzfus et al., 2013) and methods to summarize and compare the data.
-
-\hypertarget{description}{%
-\section{Description}\label{description}}
-
-DateLife's core software application consists of the R package \texttt{datelife}.
-Its latest stable version -- v0.6.4, is available from the CRAN repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
-ape (Paradis, Claude, \& Strimmer, 2004),
-bold (Chamberlain et al., 2019),
-geiger (Pennell et al., 2014),
-paleotree (Bapst, 2012),
-phyloch (Heibl, 2008),
-phylocomr (Ooms \& Chamberlain, 2018),
-phytools (Revell, 2012),
-rotl (Michonneau, Brown, \& Winter, 2016), and
-taxize (Chamberlain \& Szöcs, 2013; Chamberlain et al., 2019).
-Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife algorithm: creating a search query, searching a database, and summarizing results from the search.
-
-\begin{center}
-\includegraphics[width = 6.5in]{../figures/figure1/figure1-new.pdf}
-\captionof{figure}{Stylized DateLife workflow.
-This shows the general workflows and analyses that can be performed with \texttt{datelife}, via the R package or through the website at \url{http://www.datelife.org/}.
-Details on the functions involved on each workflow are shown in \texttt{datelife}'s R package vignette.}
-\label{fig:workflow}
-\end{center}
-
-\hypertarget{creating-a-search-query}{%
-\subsection{Creating a search query}\label{creating-a-search-query}}
-
-DateLife starts by processing an input consisting of at least two taxon names, which can be provided as a comma separated character string or as tip labels on a tree.
-If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis et al., 2004).
-The input tree is not required to have branch lengths, and its topology is used in the summary steps described in the next section.
-
-DateLife accepts scientific names that can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or a binomial species name.
-Subspecies and variants are ignored.
-If an input taxon name belongs to an inclusive taxonomic group, the algorithm has two alternative behaviors defined by the ``get species from taxon'' flag.
-If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group (following a taxonomy of choice; current options are Open Tree of Life, NCBI, GBIF, or IRMNG taxonomies) and adds them to the input string.
-If the flag is inactive, DateLife excludes any taxon names above the species level from the search query.
-
-DateLife processes input scientific names using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database.
-TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard (Boyle et al., 2013).
-DateLife implements TNRS using OpenTree's unified taxonomy as standard (Open Tree Of Life et al., 2016; Rees \& Cranston, 2017), storing OpenTree's Taxonomy identification numbers for further processing.
-
-The processed input taxon names are saved as an R object of a newly defined class, \texttt{datelifeQuery}, that is used in the following steps.
-This object contains the standardized names, the corresponding OpenTree taxonomic id numbers, and the topology of the input tree if any was provided.
-
-\hypertarget{searching-a-chronogram-database}{%
-\subsection{Searching a chronogram database}\label{searching-a-chronogram-database}}
-
-At the time of writing of this manuscript
-(Jun 20, 2022),
-DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies.
-It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata (McTavish et al., 2015), which allows automatic and reproducible assembly of our chronogram database.
-Datelife's chronogram database is navigable as an R data object within the \texttt{datelife} R package.
-
-A unique feature of OpenTree's Phylesystem is that the community can add new published, state-of-the-art chronograms any time through their curator application (\url{https://tree.opentreeoflife.org/curator}).
-As chronograms are added to Phylesystem, they are incorporated into an updated \texttt{datelife}'s database that is assigned a new version number, followed by a package release on CRAN.
-\texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
-Users can also upload and curate new chronograms to OpenTree themselves.
-Then, trigger an update of their local \texttt{datelife} database to incorporate the new chronograms and have them available for their current analysis.
-
-A DateLife search is implemented by matching processed taxon names provided by the user to tip labels in the chronogram database.
-Chronograms with at least two matching taxon names on their tip labels are identified and pruned down to preserve only the matched taxa.
-These matching pruned chronograms are referred to as source chronograms.
-Total distance (in units of millions of years) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Figure \ref{fig:workflow}).
-The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a ``phylo'' object or newick string.
-Finally, the patristic matrices are associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.
-
-\hypertarget{summarizing-search-results}{%
-\subsection{Summarizing search results}\label{summarizing-search-results}}
-
-Summary information is extracted from the \texttt{datelifeResult} object to inform decisions for subsequent steps in the analysis workflow. Basic summary information available to the user is:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  The matching pruned chronograms as newick strings or ``phylo'' objects.
-\item
-  The ages of the root of all source chronograms.
-These ages can correspond to the age of the most recent common ancestor (mrca) of the user's group of interest if the source chronograms have all taxa belonging to the group.
-If not, the root corresponds to the mrca of a subgroup withing the group of interest.
-\item
-  Study citations where original chronograms were published.
-\item
-  A report of input taxon names matches across source chronograms.
-\item
-  The source chronogram(s) with the most input taxon names.
-\item
-  Various single summary chronograms resulting from summarizing age data, generated using the methodology described next.
-\end{enumerate}
-
-\emph{\textbf{Choosing a topology.--}}
-DateLife requires a tree topology to summarize age data upon.
-We recommend that users provide a tree topology as input from the literature, or one of their own making.
-If no topology is provided, DateLife automatically obtains one from the OpenTree synthetic tree (Open Tree Of Life et al., 2019).
-These are known as ``induced'' OpenTree subtrees, which do not necessarily correspond to a monophyletic group.
-Alternatively, DateLife can combine topologies from source chronograms using a supertree approach.
-To combine topologies from source chronograms into a single summary (or supertree) topology, the DateLife algorithm starts by identifying the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
-(2009).
-In rare cases, a group of trees can have multiple groves.
-By default, DateLife chooses the grove with the most taxa, however, the ``criterion = trees'' flag allows the user to choose the grove with the most trees instead.
-If source chronograms do not form a grove, the supertree reconstruction will fail.
-
-\emph{\textbf{Dating the topology.--}}
-Input topologies from OpenTree or the supertree approach described above do not include branch length estimates of any kind.
-Optionally, to estimate branch lengths proportional to substitution rates on these topologies, DateLife can mine the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain genetic markers for the input taxa.
-These markers are aligned with MUSCLE (Edgar, 2004) (by default) or MAFFT (Katoh, Asimenos, \& Toh, 2009).
-This alignment can be used to estimate branch lengths on input topologies that lack branch lengths.
-Currently, branch length reconstruction in DateLife is performed using parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed (Schliep, 2011).
-While relative branch length information provides additional data for nodes without secondary date calibrations, topologies without branch lengths can also be dated.
-
-Once a topology is chosen, DateLife applies the congruification method (Eastman, Harmon, \& Tank, 2013) to find nodes belonging to the same clade across source chronograms, and extract the corresponding node ages from the patristic distance matrices stored as \texttt{datelifeResult}.
-By definition, the matrices store total distance (time from tip to tip), hence, node ages correspond to half the values stored in the patristic distance matrices.
-This assumes that the terminal taxa are coeval and occur at the present.
-A table of congruified node ages that can be used as calibrations for a dating analysis is stored as a \texttt{congruifiedCalibrations} object.
-
-For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method {[}SDM; Criscuolo, Berry, Douzery, and Gascuel (2006){]}, which deforms patristic distance matrices by minimizing variance and then averaging them.
-These single summary taxon pair age matrices (Summarized calibrations) can be applied as calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Campbell O.
-Webb, Ackerly, \& Kembel, 2008; Campbell O.
-Webb \& Donoghue, 2005), and treePL (Stephen A.
-Smith \& O'Meara, 2012).
-
-By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully dated topology.
-BLADJ fixes node ages that have calibration data, and distributes time between nodes with no data evenly between nodes with calibration data.
-This minimizes age variance in the resulting chronogram (Campbell O. Webb et al., 2008).
-BLADJ does not use branch lengths even when they are present in the input tree or summarizing topology.
-When there is conflict in ages between nodes with calibration data, the algorithm ignores node ages that are older than the age of a parent node.
-BLADJ requires a root age estimate.
-If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature.
-If none is provided, DateLife assigns an arbitrary age to the root as 10\% older than the oldest age available within the group.
-
-Alternative options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the calibrations.
-PATHd8 is a non-clock, rate-smoothing method (Britton et al., 2007) to date trees.
-treePL (Stephen A. Smith \& O'Meara, 2012), is a semi-parametric, rate-smoothing, penalized likelihood dating method
-(Michael J. Sanderson, 2002).
-The MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003) approach in datelife uses the calibrations as priors on node ages.
-
-\emph{\textbf{Visualizing results.--}}
-Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R ``phylo'' format), as well as visualize and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions available from the R package \texttt{datelifeplot} (Sanchez-Reyes \& O'Meara, 2022).
-
-\begin{center}
-\includegraphics[width= 6.5in]{../figures/fig_runtime_main.pdf}
-\end{center}
-\captionof{figure}{Computation time used to process a datelife query and a datelife search across \texttt{datelife}'s chronogram database, relative to number of input taxon names.
-For each N = {10, 100, 200, ..., 1 000, ...
-, 9 000, 10 000}, we sampled N species names from the class Aves a hundred times, and then performed a datelife search with input processing, without implementing Taxon Names Resolution Service (TNRS; dark gray), and implementing TNRS (light gray).
-We also performed a search using a query that had been pre-processed with TNRS, for comparison (light blue).}
-
-\label{fig:runtime_main}
-
-\hypertarget{benchmark}{%
-\section{Benchmark}\label{benchmark}}
-
-\texttt{datelife}'s code speed was tested on an Apple iMac
-with one 3.4 GHz Intel Core i5 processor.
-We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
-Query processing time increases roughly linearly with number of input taxon names, and
-increases considerably if Taxonomic Name Resolution Service (TNRS) is activated.
-Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
-Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig.
-\ref{fig:runtime_main}).
-\texttt{datelife}'s code performance was evaluated with a set of unit tests designed and
-implemented with the R package testthat (R Core Team, 2018) that were run both locally
-with the devtools package (R Core Team, 2018), and on a public server --via
-GitHub, using the continuous integration tool Travis CI (\url{https://travis-ci.org}). At
-present, unit tests cover more than 40\% of \texttt{datelife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).
-Unit testing helps identify potential issues as code is updated or, more critically, as services code relies upon may change.
-
-\hypertarget{case-studies}{%
-\section{Case studies}\label{case-studies}}
-
-We illustrate the DateLife algorithm using a family within the Passeriform birds encompassing the true finches, Fringillidae, as case study.
-The first example analyses 6 bird species and shows all steps of the algorithm.
-The second example is an application analysing 289 species in the family Fringillidae that are included in the NCBI taxonomy.
-
-\hypertarget{a-small-example}{%
-\subsection{A small example}\label{a-small-example}}
-
-\emph{\textbf{Creating a search query.-}}
-We chose 6 bird species within the Passeriformes. The sample includes
-two species of cardinals:
-the black-thighed grosbeak -- \emph{Pheucticus tibialis} and
-the crimson-collared grosbeak -- \emph{Rhodothraupis celaeno};
-three species of buntings:
-the yellowhammer -- \emph{Emberiza citrinella},
-the pine bunting -- \emph{Emberiza leucocephalos} and
-the yellow-throated bunting -- \emph{Emberiza elegans};
-and one species of tanager, the vegetarian finch -- \emph{Platyspiza crassirostris}.
-
-Processing input names found that \emph{Emberiza elegans} is synonym for \emph{Schoeniclus elegans} in the default reference taxonomy (Open Tree of Life Taxonomy v3.3, June 1, 2021).
-For a detailed discussion on the state of the synonym, refer to Avibase (Avibase, 2022; Lepage, 2004; Lepage, Vaidya, \& Guralnick, 2014).
-Discovering this synonym allowed assigning five age data points for the parent node of \emph{Emberiza elegans}, shown as \emph{Schoeniclus elegans} in figure \ref{fig:figure2}A, which would not have had any data otherwise.
-\vspace{5mm}
-
-\emph{\textbf{Searching the database.-}}
-DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies (Fig.
-\ref{fig:figure2}B).
-Three studies matched five input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2015; Hedges, Marin, Suleski, Paymer, \& Kumar, 2015; Jetz, Thomas, Joy, Hartmann, \& Mooers, 2012), one study matched four input names (Hooper \& Price, 2017) and two studies matched two input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2013; Burns et al., 2014).
-No studies matched all input names.
-Together, source chronograms provide 28 unique age data points, covering all nodes on our chosen tree topology to date (Table \ref{tbl:table1}).
-
-\vspace{5mm}
-
-\emph{\textbf{Summarizing search results.-}}
-DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and congruified and mapped age data to nodes in this chosen topology (Table \ref{tbl:table1}).
-The name processing step allowed including five data points for node ``n4'' (parent of \emph{Schoeniclus elegans}; Fig.
-\ref{fig:figure2}A) that would not have had any data otherwise due to name mismatch.
-Age summary statistics per node were calculated (Table \ref{tbl:table2}) and used as calibrations to date the tree topology using the BLADJ algorithm.
-As expected, more inclusive nodes (e.g., node ``n1'') have more variance in age data than less inclusive nodes (e.g., node ``n5'').
-Summary age data for node ``n2'' were excluded as final calibration because they are older than age data of the more inclusive node, ``n1'' (Fig. \ref{fig:figure2}C4).
-
-\vspace{10mm}
-
-\begin{center}
-\includegraphics[width = 6.5in]{../figures/figure2/figure2.pdf}
-\end{center}
-\captionof{figure}{DateLife analysis results for a small sample of A) 6 bird species within the Passeriformes.
-B) Processed species names were found across 9 chronograms within 6 independent studies (Barker et al.
-(\protect\hyperlink{ref-barker2012going}{2012}), Barker et al.
-(\protect\hyperlink{ref-barker2015new}{2015}), Burns et al.
-(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al.
-(\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al.
-(\protect\hyperlink{ref-Jetz2012}{2012}).) C) This revealed 28 source age data points for the queried species names.
-Summarized age data is used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree, resulting in a summary chronogram of source ages.}
-
-\label{fig:figure2}
-
-\vspace{15mm}
-
-\begin{table}[H]
-  \caption{Ages of congruified nodes obtained from a datelife search for the Small Example presented. See Figure 3, step C2.}
-  \label{tbl:table1}
-  \includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example.pdf}
-\end{table}
-
-\begin{table}[H]
-  \caption{Summary of congruified nodes ages. See Figure 3, step C3.}
-  \label{tbl:table2}
-  \includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example-summary.png}
-\end{table}
-
-\hypertarget{the-family-fringillidae}{%
-\subsection{The family Fringillidae}\label{the-family-fringillidae}}
-
-\emph{\textbf{Creating a query.-}}
-To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the ``get species from taxon'' flag,
-which gets all recognized species names within a named group from a taxonomy of choice.
-Following the NCBI taxonomy, our DateLife query has 289 Fringillidae species.
-This taxon-constrained approach implies that the final results of a full DateLife analysis will be done using a tree topology and ages for the species in a named group, which do not necessarily correspond to a monophyletic group.
-Users can change this behaviour by providing a monophyletic tree as input for a DateLife search, or as a tree topology for a DateLife summary.
-
-\vspace{5mm}
-
-\emph{\textbf{Searching the database.-}}
-Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species.
-The DateLife search identified 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Price et al., 2014).
-Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices, revealing 1206 different age data points available for species within the Fringillidae (supplementray table 1).
-
-\vspace{5mm}
-
-\emph{\textbf{Summarizing search results.-}}
-The final step is to congruify and summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM.
-As explained in the ``Description'' section, a tree topology to summarize age data upon is required.
-By default, DateLife uses as summarizing topology the portion of OpenTree synthetic phylogeny that contains the species provided as search query.
-According to this phylogeny, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).
-
-Age data from source chronograms was congruified, decreasing to 818 different age data points available (supplementray table 2) and then age summary statistics per congruified node are obtained and used as fixed secondary calibrations over a consensus tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
-\ref{fig:fringillidages}).
-Median summary chronograms are older and have wider variation in maximum ages than chronograms obtained with SDM.
-
-\vspace{20mm}
-
-\begin{center}
-\includegraphics[width = 6.5in]{../figures/fringillidae-topologies/fringillidae-topology.pdf}
-\captionof{figure}{Tree topologies obtained from Open Tree of Life's (OpenTree) synthetic phylogenetic tree.
-A) Topology of 2,333 tips and 1,305 internal nodes, encompassing bird species within the family Fringillidae following the NCBI taxonomy (black), as well as all other bird species that share the same mrca node in OpenTree's synthetic tree (purple).
-B) Topology of 289 tips and 253 internal nodes, encompassing bird species within the Fringillidae only.
-This tree results from pruning species that are not assigned to the family Fringillidae (purple branches in topology A) but that form a monophyletic group with them.
-Bird species within the Fringillidae are paraphyletic
-(Alström et al. \protect\hyperlink{ref-alstrom2014discovery}{2014},
-Barker, Cibois, Schikler, Feinstein, \& Cracraft \protect\hyperlink{ref-barker2004phylogeny}{2004},
-Barker et al. \protect\hyperlink{ref-barker2013going}{2013},
-Barker \protect\hyperlink{ref-barker2014mitogenomic}{2014},
-Barker et al. \protect\hyperlink{ref-barker2015new}{2015},
-Beresford, Barker, Ryan, \& Crowe \protect\hyperlink{ref-beresford2005african}{2005},
-Bryson Jr et al. \protect\hyperlink{ref-bryson2014diversification}{2014},
-Burleigh, Kimball, \& Braun \protect\hyperlink{ref-burleigh2015building}{2015},
-Burns et al. \protect\hyperlink{ref-burns2014phylogenetics}{2014},
-Chaves, Hidalgo, \& Klicka \protect\hyperlink{ref-chaves2013biogeography}{2013},
-Claramunt \& Cracraft \protect\hyperlink{ref-claramunt2015new}{2015},
-Gibb et al. \protect\hyperlink{ref-gibb2015new}{2015},
-Hackett et al. \protect\hyperlink{ref-hackett2008phylogenomic}{2008},
-Jetz et al. \protect\hyperlink{ref-Jetz2012}{2012},
-Johansson, Fjeldså, \& Bowi \protect\hyperlink{ref-johansson2008phylogenetic}{200},
-Kimball et al. \protect\hyperlink{ref-kimball2019phylogenomic}{2019},
-Klicka et al. \protect\hyperlink{ref-klicka2014comprehensive}{2014},
-Lamichhaney et al. \protect\hyperlink{ref-lamichhaney2015evolution}{2015},
-Lerner, Meyer, James, Hofreiter, \& Fleischer \protect\hyperlink{ref-lerner2011multilocus}{2011},
-Lovette et al. \protect\hyperlink{ref-lovette2010comprehensive}{2010},
-Moyle et al. \protect\hyperlink{ref-moyle2016tectonic}{2016},
-Ödeen, Håstad, \& Alström \protect\hyperlink{ref-odeen2011evolution}{2011},
-Oliveros et al. \protect\hyperlink{ref-oliveros2019earth}{2019},
-Päckert et al. \protect\hyperlink{ref-packert2012horizontal}{2012},
-Parchman, Benkman, \& Mezquida \protect\hyperlink{ref-parchman2007coevolution}{2007},
-Powell et al. \protect\hyperlink{ref-powell2014comprehensive}{2014},
-Price et al. \protect\hyperlink{ref-price2014niche}{2014},
-Pulgarín-R, Smith, Bryson Jr, Spellman, \& Klicka \protect\hyperlink{ref-pulgarin2013multilocus}{2013},
-Selvatti, Gonzaga, \& Moraes Russo \protect\hyperlink{ref-selvatti2015paleogene}{2015},
-Tietze, Päckert, Martens, Lehmann, \& Sun \protect\hyperlink{ref-tietze2013complete}{2013},
-Treplin et al. \protect\hyperlink{ref-treplin2008molecular}{2008},
-Zuccon, Prŷs-Jones, Rasmussen, \& Ericson \protect\hyperlink{ref-zuccon2012phylogenetic}{2012}).
-}
-\label{fig:fringillidae-topologies}
-\end{center}
-
-\hypertarget{cross-validation-test}{%
-\section{Cross-validation test}\label{cross-validation-test}}
-
-We performed a cross validation analysis of the DateLife workflow using the Fringillidae chronograms.
-We used the individual tree topologies from each of the 19 source chronograms from 13 studies as inputs, treating their node ages as unknown.
-We then estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.
-
-We found that node ages from original study, and ages estimated using all other age data available are correlated (Supplementary Fig. \ref{fig:cvXY}).
-
-For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka `closer to the root') relative to the orginal estimate, and overestimate ages for nodes closer to the tips.
-Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. \ref{fig:cvXYdiffs}).
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-\texttt{datelife} makes state-of-the-art data on evolutionary time frame easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
-It is an open service that does not require any expert biological knowledge from users --besides the names of the species or group they want to work with, for any of its functionality.
-
-A total of 99474 unique terminal taxa are represented in \texttt{datelife}'s database.
-Incorporation of more chronograms into \texttt{datelife}'s database will continue to improve its services.
-One option to increase the number of chronograms in the DateLife database is the Dryad data repository.
-Methods to automatically mine chronograms from Dryad could be designed and implemented.
-However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined manually by a human curator.
-We would like to emphasize on the importance of sharing chronogram data, including systematically curated metadata, into open repositories, such as OpenTree's Phylesystem (McTavish et al., 2015) for the benefit of the scientific community as a whole.
-
-\hypertarget{primary-vs-secondary-calibrations-in-dating-analyses}{%
-\subsection{Primary vs Secondary calibrations in dating analyses}\label{primary-vs-secondary-calibrations-in-dating-analyses}}
-
-While most chronograms in DateLife's database are constructed using primary calibrations (molecular substitution rates or ages obtained from the fossil record or geological events), DateLife summarizes chronograms using secondary calibrations only (ages coming from other chronograms).
-
-Graur and Martin (2004) cautioned on the increased error and uncertainty in estimated ages when using secondary calibrations in dating analyses.
-Schenk (2016) showed that, in simulations, divergence times inferred using secondary calibrations are significantly younger than those inferred with primary calibrations, when obtained with Bayesian inference methods, and when priors are implemented in similar ways in both analyses.
-Accordingly, the scientific community seems to have more confidence in chronograms obtained from a single analysis, using fossil data as primary sources of calibrations (Schenk, 2016), and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data reflect a coherent evolutionary history (Sauquet, 2013), as for example done by Antonelli et al.
-(2017).
-There have been attempts to create fossil calibration databases (Ksepka et al., 2015), though these still have room to grow.
-
-It seems that using several (as opposed to just a few) secondary calibrations can provide sufficient information to alleviate or even neutralize potential biases (Sauquet, 2013).
-Certainly, further studies are required to fully understand the effect of secondary calibrations on outputs from different tree dating methods, and on downstream analyses.
-It is possible that secondary calibrations can be safely used with dating methods that do not require setting priors, such as penalized likelihood (Michael J.
-Sanderson, 2003), with methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ (Campbell O.
-Webb et al., 2008; Campbell O.
-Webb \& Donoghue, 2005), or methods that summarize age data unto a tree topology.
-
-Our cross validation analysis might provide some insight in this regard.
-When ages are estimated with secondary calibrations, nodes closer to the root do tend to be slightly younger than ages estimated with primary calibrations.
-However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures 1-9), with a notable exception of cross validation of Burns et al.
-(2014) chronogram, which results in much younger node ages when estimated using secondary calibrations (Supplementary figs.
-1 and 4).
-
-\hypertarget{variation-in-source-chronograms}{%
-\subsection{Variation in source chronograms}\label{variation-in-source-chronograms}}
-
-Conflict among alternative dating studies is common in the literature
-(see, for example, the robust ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández, 2015; Ramshaw et al., 1972; Michael J.
-Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021)).
-
-Source chronograms may have been estimated using different types of primary calibrations (fossil record, geologic or biogeographic events, molecular substitution rates).
-These differences in approach can deepen the already substantial variation in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
-For example, the chronograms from Burns et al.
-(2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older date estimates for the same nodes than chronograms that were inferred using fossils as primary calibrations (figs.
-fig:fringillidages and fig:cvXY; supplementary figs.
-1 and 4).
-
-Different chronograms might also imply fundamentally distinct evolutionary hypotheses.
-For example, two different researchers working on the same clade could both carefully select and argue for their choices of fossil calibrations.
-Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses the same one to calibrate outside the clade, the resulting age estimates will often differ substantially, as the placement of calibrations as stem or crown group is proved to deeply affect estimated times of lineage divergence (Sauquet, 2013).
-
-\hypertarget{sumarizing-chronograms}{%
-\subsection{Sumarizing chronograms}\label{sumarizing-chronograms}}
-
-By default, DateLife currently summarizes all source chronograms that overlap with at least two species names. Users can exclude source chronograms if they have reasons to do so.
-Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely.
-To our knowledge, there are no tested measures to determine independently when a chronogram is better than another.
-Yet, several characteristics of the data used for dating analyses, as well as from the output chronogram itself, could be used to score the quality of source chronograms.
-
-Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used Magallón et al. (2015).
-Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
-
-DateLife provides an opportunity to capture concordance and conflict among date estimates, which can also be used as a metric for chronogram reliability.
-Its open database of chronograms allows other researchers to do such analyses themselves reproducibly, and without needing permission.
-Though, of course, they should follow proper citation practices, especially for the source chronogram studies.
-
-The exercise of summarizing age data from across multiple studies provides the opportunity to work with a more inclusive chronogram,
-that reflects a unified evolutionary history for a lineage, by putting together evidence from different hypotheses.
-The largest, and taxonomically broadest chronogram currently available from OpenTree was constructed summarizing age data from 2,274 published chronograms using NCBI's taxonomic tree as backbone (Hedges et al., 2015).
-A summarizing exercise may also amplify the effect of uncertainty and errors in source data, and blur parts of the evolutionary history of a lineage that might only be reflected in source chronograms and lost on the summary chronogram (Sauquet et al., 2021).
-
-\hypertarget{effects-on-downstream-analyses}{%
-\subsection{Effects on downstream analyses}\label{effects-on-downstream-analyses}}
-
-For downstream analyses, using alternative chronogram may deeply affect our inferences (Title \& Rabosky, 2016), particularly when studying phenomena dependent on the timing of species diversification events, such as macroevolutionary processes.
-
-In ecology and conservation biology, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance (Campbell O.
-Webb et al., 2008).
-Hence, DateLife's workflow features different ways of estimating node ages in the absence of calibrations and branch length information for certain taxa.
-``Making up'' branch lengths is a common practice in scientific publications: Jetz et al.
-(2012), created a time-calibrated tree of all 9,993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
-(2018) created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37\% had molecular data; Stephen A.
-Smith and Brown (2018) constructed a tree of 353,185 seed plants where only 23\% had molecular data.
-
-Notably, risks come with this practice.
-Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration, using polytomy resolution and branch estimation methods.
-There has yet to be a thorough analysis of what can go wrong when one extends inferences beyond the data in this way, so we urge caution; we also urge readers to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, and assessing the consistency of the results with prior work.
-
-\hypertarget{conclusions}{%
-\section{Conclusions}\label{conclusions}}
-
-Knowledge of the evolutionary time frame of organisms is key to many research areas: trait evolution,
-species diversification, biogeography, macroecology and more. It is also crucial for education, science communication and policy, but generating chronograms is difficult,
-especially for those who want to use phylogenies but who are not systematists, or
-do not have the time to acquire and develop the necessary knowledge and skills to construct them on their own.
-Importantly, years of primarily public funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
-
-The DateLife project allows for easy and fast summary of public and state-of-the-art data on time of lineage divergence.
-It provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identification of regions that require more research, or that have conflicting information.
-It is available as an R package, and as a web-based R shiny application at www.datelife.org
-Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
-The DateLife project helps with awareness of the existing variation in expert time of divergence data, and will foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
-
-\hypertarget{availability}{%
-\section{Availability}\label{availability}}
-
-The \texttt{datelife} software is free and open source and it can be used through its R shiny web application at
-\url{http://www.datelife.org}, through the \texttt{datelife} R package, and through Phylotastic's project web portal \url{https://phylo.cs.nmsu.edu/}.
-DateLife's web application is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker and Open Tree of Life's infrastructure (dates.opentreeoflife.org/datelife).
-\texttt{datelife}'s R package stable version is available
-for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
-using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R. Development versions
-are available from the GitHub repository (\url{https://github.com/phylotastic/datelife})
-and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.
-
-\hypertarget{supplementary-material}{%
-\section{Supplementary Material}\label{supplementary-material}}
-
-Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities are available at \href{https://github.com/LunaSare/datelifeMS1}{datelifeMS1}, \href{https://github.com/LunaSare/datelife_examples}{datelife\_examples}, and \href{https://github.com/LunaSare/datelife_benchmark}{datelife\_benchmark} repositories in LLSR's GitHub account.
-
-\hypertarget{funding}{%
-\section{Funding}\label{funding}}
-
-Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to the Datelife project; DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), ABI-1458572 to the Phylotastic project, and ABI-1759846 to the Open Tree of Life project.
-
-\hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
-
-The DateLife project was born as a prototype tool aiming to provide these services, and was initially developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA (Stoltzfus et al., 2013).
-We thank colleagues from the O'Meara Lab at the University
-of Tennesse Knoxville for suggestions, discussions and software testing.
-The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
-that led to initial work on this project.
-The team that assembled DateLife's first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
-The Open Tree of Life project that provides the open, metadata rich repository of
-trees used to construct DateLife's chronogram database.
-The many scientists who publish their chronograms in an open, reusable form, and the scientists who curate them for deposition in the Open Tree of Life repository.
-The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.
-
-\newpage
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-alstrom2014discovery}{}}%
-Alström, P., Hooper, D.
-M., Liu, Y., Olsson, U., Mohan, D., Gelang, M., \ldots{} Price, T.
-D.
-(2014).
-Discovery of a relict lineage and monotypic family of passerine birds.
-\emph{Biology Letters}, \emph{10}(3), 20131067.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ane2009groves}{}}%
-Ané, C., Eulenstein, O., Piaggio-Talice, R., \& Sanderson, M. J. (2009). Groves of phylogenetic trees. \emph{{Annals of Combinatorics}}, \emph{13}(2), 139--167.
-
-\leavevmode\vadjust pre{\hypertarget{ref-antonelli2017supersmart}{}}%
-Antonelli, A., Hettling, H., Condamine, F.
-L., Vos, K., Nilsson, R.
-H., Sanderson, M.
-J., \ldots{} Vos, R.
-A.
-(2017).
-{Toward a self-updating platform for estimating rates of speciation and migration, ages, and relationships of Taxa}.
-\emph{Systematic Biology}, \emph{66}(2), 153--166.
-\url{https://doi.org/10.1093/sysbio/syw066}
-
-\leavevmode\vadjust pre{\hypertarget{ref-archie1986newick}{}}%
-Archie, J., Day, W.
-H., Felsenstein, J., Maddison, W., Meacham, C., Rohlf, F.
-J., \& Swofford, D.
-(1986).
-{The Newick tree format}.
-Retrieved from \url{\%7Bhttps://evolution.genetics.washington.edu/phylip/newicktree.html\%7D}
-
-\leavevmode\vadjust pre{\hypertarget{ref-avibase-emberiza}{}}%
-Avibase.
-(2022).
-{Yellow-throated Bunting}.
-\emph{{Avibase - The World Bird Database}}, (Online Resource).
-Retrieved from \url{\%7Bhttps://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\%7D}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Bapst2012a}{}}%
-Bapst, D.
-W.
-(2012).
-{Paleotree: An R package for paleontological and phylogenetic analyses of evolution}.
-\emph{{Methods in Ecology and Evolution}}, \emph{3}(5), 803--807.
-\url{https://doi.org/10.1111/j.2041-210X.2012.00223.x}
-
-\leavevmode\vadjust pre{\hypertarget{ref-barba2018constraining}{}}%
-Barba-Montoya, J., Reis, M.
-dos, Schneider, H., Donoghue, P.
-C., \& Yang, Z.
-(2018).
-Constraining uncertainty in the timescale of angiosperm evolution and the veracity of a cretaceous terrestrial revolution.
-\emph{New Phytologist}, \emph{218}(2), 819--834.
-
-\leavevmode\vadjust pre{\hypertarget{ref-barker2014mitogenomic}{}}%
-Barker, F. K. (2014). Mitogenomic data resolve basal relationships among passeriform and passeridan birds. \emph{Molecular Phylogenetics and Evolution}, \emph{79}, 313--324.
-
-\leavevmode\vadjust pre{\hypertarget{ref-barker2013going}{}}%
-Barker, F.
-K., Burns, K.
-J., Klicka, J., Lanyon, S.
-M., \& Lovette, I.
-J.
-(2013).
-Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds.
-\emph{{Systematic Biology}}, \emph{62}(2), 298--320.
-
-\leavevmode\vadjust pre{\hypertarget{ref-barker2015new}{}}%
-Barker, F.
-K., Burns, K.
-J., Klicka, J., Lanyon, S.
-M., \& Lovette, I.
-J.
-(2015).
-New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
-\emph{The Auk: Ornithological Advances}, \emph{132}(2), 333--348.
-
-\leavevmode\vadjust pre{\hypertarget{ref-barker2004phylogeny}{}}%
-Barker, F.
-K., Cibois, A., Schikler, P., Feinstein, J., \& Cracraft, J.
-(2004).
-Phylogeny and diversification of the largest avian radiation.
-\emph{Proceedings of the National Academy of Sciences}, \emph{101}(30), 11040--11045.
-
-\leavevmode\vadjust pre{\hypertarget{ref-beresford2005african}{}}%
-Beresford, P., Barker, F., Ryan, Pg., \& Crowe, T.
-(2005).
-African endemics span the tree of songbirds (passeri): Molecular systematics of several evolutionary {``enigmas.''} \emph{Proceedings of the Royal Society B: Biological Sciences}, \emph{272}(1565), 849--858.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Boyle2013}{}}%
-Boyle, B., Hopkins, N., Lu, Z., Raygoza Garay, J.
-A., Mozzherin, D., Rees, T., \ldots{} Enquist, B.
-J.
-(2013).
-{The taxonomic name resolution service: An online tool for automated standardization of plant names}.
-\emph{{BMC Bioinformatics}}, \emph{14}(1).
-\url{https://doi.org/10.1186/1471-2105-14-16}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Britton2007}{}}%
-Britton, T., Anderson, C.
-L., Jacquet, D., Lundqvist, S., \& Bremer, K.
-(2007).
-{Estimating Divergence Times in Large Phylogenetic Trees}.
-\emph{{Systematic Biology}}, \emph{56}(788777878), 741--752.
-\url{https://doi.org/10.1080/10635150701613783}
-
-\leavevmode\vadjust pre{\hypertarget{ref-bryson2014diversification}{}}%
-Bryson Jr, R.
-W., Chaves, J., Smith, B.
-T., Miller, M.
-J., Winker, K., Pérez-Emán, J.
-L., \& Klicka, J.
-(2014).
-Diversification across the new world within the `blue'cardinalids (aves: cardinalidae).
-\emph{Journal of Biogeography}, \emph{41}(3), 587--599.
-
-\leavevmode\vadjust pre{\hypertarget{ref-burleigh2015building}{}}%
-Burleigh, J. G., Kimball, R. T., \& Braun, E. L. (2015). Building the avian tree of life using a large-scale, sparse supermatrix. \emph{Molecular Phylogenetics and Evolution}, \emph{84}, 53--63.
-
-\leavevmode\vadjust pre{\hypertarget{ref-burns2014phylogenetics}{}}%
-Burns, K.
-J., Shultz, A.
-J., Title, P.
-O., Mason, N.
-A., Barker, F.
-K., Klicka, J., \ldots{} Lovette, I.
-J.
-(2014).
-Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds.
-\emph{{Molecular Phylogenetics and Evolution}}, \emph{75}, 41--77.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Chamberlain2013}{}}%
-Chamberlain, S.
-A., \& Szöcs, E.
-(2013).
-{taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}}.
-\emph{{F1000Research}}, \emph{2}(191), 1--29.
-\url{https://doi.org/10.12688/f1000research.2-191.v2}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Chamberlain2018}{}}%
-Chamberlain, S.
-A., Szöcs, E., Foster, Z., Arendsee, Z., Boettiger, C., Ram, K., \ldots{} Li, G.
-(2019).
-\emph{{taxize: Taxonomic information from around the web}}.
-Retrieved from \url{https://github.com/ropensci/taxize}
-
-\leavevmode\vadjust pre{\hypertarget{ref-chaves2013biogeography}{}}%
-Chaves, J.
-A., Hidalgo, J.
-R., \& Klicka, J.
-(2013).
-Biogeography and evolutionary history of the n eotropical genus s altator (a ves: T hraupini).
-\emph{Journal of Biogeography}, \emph{40}(11), 2180--2190.
-
-\leavevmode\vadjust pre{\hypertarget{ref-claramunt2015new}{}}%
-Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{{Science Advances}}, \emph{1}(11), e1501005.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Criscuolo2006}{}}%
-Criscuolo, A., Berry, V., Douzery, E.
-J.
-P., \& Gascuel, O.
-(2006).
-{SDM: A fast distance-based approach for (super)tree building in phylogenomics}.
-\emph{{Systematic Biology}}, \emph{55}(5), 740--755.
-\url{https://doi.org/10.1080/10635150600969872}
-
-\leavevmode\vadjust pre{\hypertarget{ref-delsuc2018phylogenomic}{}}%
-Delsuc, F., Philippe, H., Tsagkogeorga, G., Simion, P., Tilak, M.-K., Turon, X., \ldots{} Douzery, E.
-J.
-(2018).
-A phylogenomic framework and timescale for comparative studies of tunicates.
-\emph{BMC Biology}, \emph{16}(1), 1--14.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Eastman2013}{}}%
-Eastman, J.
-M., Harmon, L.
-J., \& Tank, D.
-C.
-(2013).
-{Congruification: Support for time scaling large phylogenetic trees}.
-\emph{{Methods in Ecology and Evolution}}, \emph{4}(7), 688--691.
-\url{https://doi.org/10.1111/2041-210X.12051}
-
-\leavevmode\vadjust pre{\hypertarget{ref-edgar2004muscle}{}}%
-Edgar, R. C. (2004). MUSCLE: Multiple sequence alignment with high accuracy and high throughput. \emph{Nucleic Acids Research}, \emph{32}(5), 1792--1797.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Felsenstein1985a}{}}%
-Felsenstein, J. (1985). {Phylogenies and the Comparative Method}. \emph{The American Naturalist}, \emph{125}(1), 1--15. Retrieved from \url{http://www.jstor.org/stable/2461605}
-
-\leavevmode\vadjust pre{\hypertarget{ref-gibb2015new}{}}%
-Gibb, G.
-C., England, R., Hartig, G., McLenachan, P.
-A., Taylor Smith, B.
-L., McComish, B.
-J., \ldots{} Penny, D.
-(2015).
-New zealand passerines help clarify the diversification of major songbird lineages during the oligocene.
-\emph{{Genome Biology and Evolution}}, \emph{7}(11), 2983--2995.
-
-\leavevmode\vadjust pre{\hypertarget{ref-graur2004reading}{}}%
-Graur, D., \& Martin, W. (2004). Reading the entrails of chickens: Molecular timescales of evolution and the illusion of precision. \emph{TRENDS in Genetics}, \emph{20}(2), 80--86.
-
-\leavevmode\vadjust pre{\hypertarget{ref-hackett2008phylogenomic}{}}%
-Hackett, S.
-J., Kimball, R.
-T., Reddy, S., Bowie, R.
-C., Braun, E.
-L., Braun, M.
-J., \ldots{} others.
-(2008).
-A phylogenomic study of birds reveals their evolutionary history.
-\emph{Science}, \emph{320}(5884), 1763--1768.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Hedges2015}{}}%
-Hedges, S.
-B., Marin, J., Suleski, M., Paymer, M., \& Kumar, S.
-(2015).
-{Tree of life reveals clock-like speciation and diversification}.
-\emph{{Molecular Biology and Evolution}}, \emph{32}(4), 835--845.
-\url{https://doi.org/10.1093/molbev/msv037}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Heibl2008}{}}%
-Heibl, C. (2008). \emph{PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages.} Retrieved from \url{http://www.christophheibl.de/Rpackages.html}
-
-\leavevmode\vadjust pre{\hypertarget{ref-hooper2017chromosomal}{}}%
-Hooper, D. M., \& Price, T. D. (2017). Chromosomal inversion differences correlate with range overlap in passerine birds. \emph{Nature Ecology \& Evolution}, \emph{1}(10), 1526.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Huelsenbeck2001}{}}%
-Huelsenbeck, J. P., \& Ronquist, F. (2001). {MRBAYES: Bayesian inference of phylogenetic trees}. \emph{Bioinformatics}, \emph{17}(8), 754--755. \url{https://doi.org/10.1093/bioinformatics/17.8.754}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Jetz2012}{}}%
-Jetz, W., Thomas, G., Joy, J.
-J.
-B., Hartmann, K., \& Mooers, A.
-(2012).
-{The global diversity of birds in space and time}.
-\emph{Nature}, \emph{491}(7424), 444--448.
-\url{https://doi.org/10.1038/nature11631}
-
-\leavevmode\vadjust pre{\hypertarget{ref-johansson2008phylogenetic}{}}%
-Johansson, U.
-S., Fjeldså, J., \& Bowie, R.
-C.
-(2008).
-Phylogenetic relationships within passerida (aves: Passeriformes): A review and a new molecular phylogeny based on three nuclear intron markers.
-\emph{Molecular Phylogenetics and Evolution}, \emph{48}(3), 858--876.
-
-\leavevmode\vadjust pre{\hypertarget{ref-katoh2009multiple}{}}%
-Katoh, K., Asimenos, G., \& Toh, H. (2009). Multiple alignment of DNA sequences with MAFFT. In \emph{Bioinformatics for DNA sequence analysis} (pp. 39--64). Springer.
-
-\leavevmode\vadjust pre{\hypertarget{ref-kimball2019phylogenomic}{}}%
-Kimball, R. T., Oliveros, C. H., Wang, N., White, N. D., Barker, F. K., Field, D. J., \ldots{} others. (2019). A phylogenomic supertree of birds. \emph{Diversity}, \emph{11}(7), 109.
-
-\leavevmode\vadjust pre{\hypertarget{ref-klicka2014comprehensive}{}}%
-Klicka, J., Barker, F.
-K., Burns, K.
-J., Lanyon, S.
-M., Lovette, I.
-J., Chaves, J.
-A., \& Bryson Jr, R.
-W.
-(2014).
-A comprehensive multilocus assessment of sparrow (aves: Passerellidae) relationships.
-\emph{Molecular Phylogenetics and Evolution}, \emph{77}, 177--182.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ksepka2015fossil}{}}%
-Ksepka, D.
-T., Parham, J.
-F., Allman, J.
-F., Benton, M.
-J., Carrano, M.
-T., Cranston, K.
-A., \ldots{} others.
-(2015).
-The fossil calibration database---a new resource for divergence dating.
-\emph{Systematic Biology}, \emph{64}(5), 853--859.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lamichhaney2015evolution}{}}%
-Lamichhaney, S., Berglund, J., Almén, M.
-S., Maqbool, K., Grabherr, M., Martinez-Barrio, A., \ldots{} others.
-(2015).
-Evolution of darwin's finches and their beaks revealed by genome sequencing.
-\emph{Nature}, \emph{518}(7539), 371--375.
-
-\leavevmode\vadjust pre{\hypertarget{ref-laubichler2009form}{}}%
-Laubichler, M. D., \& Maienschein, J. (2009). \emph{Form and function in developmental evolution}. Cambridge University Press.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lepage2004avibase}{}}%
-Lepage, D. (2004). \emph{Avibase: The world bird database}. Bird Studies Canada.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lepage2014avibase}{}}%
-Lepage, D., Vaidya, G., \& Guralnick, R. (2014). Avibase--a database system for managing and organizing taxonomic concepts. \emph{ZooKeys}, (420), 117.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lerner2011multilocus}{}}%
-Lerner, H.
-R., Meyer, M., James, H.
-F., Hofreiter, M., \& Fleischer, R.
-C.
-(2011).
-Multilocus resolution of phylogeny and timescale in the extant adaptive radiation of hawaiian honeycreepers.
-\emph{Current Biology}, \emph{21}(21), 1838--1844.
-
-\leavevmode\vadjust pre{\hypertarget{ref-lovette2010comprehensive}{}}%
-Lovette, I.
-J., Pérez-Emán, J.
-L., Sullivan, J.
-P., Banks, R.
-C., Fiorentino, I., Córdoba-Córdoba, S., \ldots{} others.
-(2010).
-A comprehensive multilocus phylogeny for the wood-warblers and a revised classification of the parulidae (aves).
-\emph{Molecular Phylogenetics and Evolution}, \emph{57}(2), 753--770.
-
-\leavevmode\vadjust pre{\hypertarget{ref-magallon2001absolute}{}}%
-Magallon, S., \& Sanderson, M. J. (2001). Absolute diversification rates in angiosperm clades. \emph{Evolution}, \emph{55}(9), 1762--1780.
-
-\leavevmode\vadjust pre{\hypertarget{ref-magallon2010using}{}}%
-Magallón, S. (2010). Using fossils to break long branches in molecular dating: A comparison of relaxed clocks applied to the origin of angiosperms. \emph{Systematic Biology}, \emph{59}(4), 384--399.
-
-\leavevmode\vadjust pre{\hypertarget{ref-magallon2015metacalibrated}{}}%
-Magallón, S., Gómez-Acevedo, S., Sánchez-Reyes, L.
-L., \& Hernández-Hernández, T.
-(2015).
-A metacalibrated time-tree documents the early rise of flowering plant phylogenetic diversity.
-\emph{New Phytologist}, \emph{207}(2), 437--453.
-
-\leavevmode\vadjust pre{\hypertarget{ref-mctavish2015phylesystem}{}}%
-McTavish, E.
-J., Hinchliff, C.
-E., Allman, J.
-F., Brown, J.
-W., Cranston, K.
-A., Holder, M.
-T., \ldots{} Smith, S.
-A.
-(2015).
-Phylesystem: A git-based data store for community-curated phylogenetic estimates.
-\emph{Bioinformatics}, \emph{31}(17), 2794--2800.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Michonneau2016}{}}%
-Michonneau, F., Brown, J.
-W., \& Winter, D.
-J.
-(2016).
-{rotl: an R package to interact with the Open Tree of Life data}.
-\emph{{Methods in Ecology and Evolution}}, \emph{7}(12), 1476--1481.
-\url{https://doi.org/10.1111/2041-210X.12593}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Morlon2014}{}}%
-Morlon, H. (2014). {Phylogenetic approaches for studying diversification.} \emph{{Ecology Letters}}, \emph{17}(4), 508--525. \url{https://doi.org/10.1111/ele.12251}
-
-\leavevmode\vadjust pre{\hypertarget{ref-moyle2016tectonic}{}}%
-Moyle, R.
-G., Oliveros, C.
-H., Andersen, M.
-J., Hosner, P.
-A., Benz, B.
-W., Manthey, J.
-D., \ldots{} Faircloth, B.
-C.
-(2016).
-Tectonic collision and uplift of wallacea triggered the global songbird radiation.
-\emph{Nature Communications}, \emph{7}(1), 1--7.
-
-\leavevmode\vadjust pre{\hypertarget{ref-odeen2011evolution}{}}%
-Ödeen, A., Håstad, O., \& Alström, P. (2011). Evolution of ultraviolet vision in the largest avian radiation-the passerines. \emph{BMC Evolutionary Biology}, \emph{11}(1), 1--8.
-
-\leavevmode\vadjust pre{\hypertarget{ref-oliveros2019earth}{}}%
-Oliveros, C.
-H., Field, D.
-J., Ksepka, D.
-T., Barker, F.
-K., Aleixo, A., Andersen, M.
-J., \ldots{} others.
-(2019).
-Earth history and the passerine superradiation.
-\emph{Proceedings of the National Academy of Sciences}, \emph{116}(16), 7916--7925.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Ooms2018}{}}%
-Ooms, J., \& Chamberlain, S. (2018). \emph{Phylocomr: Interface to 'phylocom'}. Retrieved from \url{https://CRAN.R-project.org/package=phylocomr}
-
-\leavevmode\vadjust pre{\hypertarget{ref-opentreeAPIs}{}}%
-Open Tree Of Life, Redelings, B., Cranston, K.
-A., Allman, J., Holder, M.
-T., \& McTavish, E.
-J.
-(2016).
-{Open Tree of Life APIs v3.0}.
-\emph{{Open Tree of Life Project}}, (Online Resources).
-Retrieved from \url{\%7Bhttps://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\%7D}
-
-\leavevmode\vadjust pre{\hypertarget{ref-opentreeoflife2019synth}{}}%
-Open Tree Of Life, Redelings, B., Sánchez Reyes, L.
-L., Cranston, K.
-A., Allman, J., Holder, M.
-T., \& McTavish, E.
-J.
-(2019).
-Open tree of life synthetic tree v12.3.
-\emph{Zenodo}.
-Retrieved from \url{https://doi.org/10.5281/zenodo.3937742}
-
-\leavevmode\vadjust pre{\hypertarget{ref-packert2012horizontal}{}}%
-Päckert, M., Martens, J., Sun, Y.-H., Severinghaus, L.
-L., Nazarenko, A.
-A., Ting, J., \ldots{} Tietze, D.
-T.
-(2012).
-Horizontal and elevational phylogeographic patterns of himalayan and southeast asian forest passerines (aves: passeriformes).
-\emph{Journal of Biogeography}, \emph{39}(3), 556--573.
-
-\leavevmode\vadjust pre{\hypertarget{ref-paradis2004}{}}%
-Paradis, E., Claude, J., \& Strimmer, K. (2004). {APE: analyses of phylogenetics and evolution in R language}. \emph{Bioinformatics}, \emph{20}(2), 289--290.
-
-\leavevmode\vadjust pre{\hypertarget{ref-parchman2007coevolution}{}}%
-Parchman, T.
-L., Benkman, C.
-W., \& Mezquida, E.
-T.
-(2007).
-Coevolution between hispaniolan crossbills and pine: Does more time allow for greater phenotypic escalation at lower latitude?
-\emph{Evolution}, \emph{61}(9), 2142--2153.
-
-\leavevmode\vadjust pre{\hypertarget{ref-pennell2014geiger}{}}%
-Pennell, M.
-W., Eastman, J.
-M., Slater, G.
-J., Brown, J.
-W., Uyeda, J.
-C., FitzJohn, R.
-G., \ldots{} Harmon, L.
-J.
-(2014).
-Geiger v2.
-0: An expanded suite of methods for fitting macroevolutionary models to phylogenetic trees.
-\emph{Bioinformatics}, \emph{30}(15), 2216--2218.
-
-\leavevmode\vadjust pre{\hypertarget{ref-posadas2006historical}{}}%
-Posadas, P., Crisci, J. V., \& Katinas, L. (2006). Historical biogeography: A review of its basic concepts and critical issues. \emph{{Journal of Arid Environments}}, \emph{66}(3), 389--403.
-
-\leavevmode\vadjust pre{\hypertarget{ref-powell2014comprehensive}{}}%
-Powell, A.
-F., Barker, F.
-K., Lanyon, S.
-M., Burns, K.
-J., Klicka, J., \& Lovette, I.
-J.
-(2014).
-A comprehensive species-level molecular phylogeny of the new world blackbirds (icteridae).
-\emph{Molecular Phylogenetics and Evolution}, \emph{71}, 94--112.
-
-\leavevmode\vadjust pre{\hypertarget{ref-price2014niche}{}}%
-Price, T.
-D., Hooper, D.
-M., Buchanan, C.
-D., Johansson, U.
-S., Tietze, D.
-T., Alström, P., \ldots{} others.
-(2014).
-Niche filling slows the diversification of himalayan songbirds.
-\emph{Nature}, \emph{509}(7499), 222.
-
-\leavevmode\vadjust pre{\hypertarget{ref-pulgarin2013multilocus}{}}%
-Pulgarín-R, P.
-C., Smith, B.
-T., Bryson Jr, R.
-W., Spellman, G.
-M., \& Klicka, J.
-(2013).
-Multilocus phylogeny and biogeography of the new world pheucticus grosbeaks (aves: cardinalidae).
-\emph{Molecular Phylogenetics and Evolution}, \emph{69}(3), 1222--1227.
-
-\leavevmode\vadjust pre{\hypertarget{ref-RCoreTeam2018}{}}%
-R Core Team. (2018). \emph{{R: a language and environment for statistical computing}}. Vienna, Austria: R Foundation for Statistical Computing.
-
-\leavevmode\vadjust pre{\hypertarget{ref-rabosky2018inverse}{}}%
-Rabosky, D.
-L., Chang, J., Title, P.
-O., Cowman, P.
-F., Sallan, L., Friedman, M., \ldots{} others.
-(2018).
-An inverse latitudinal gradient in speciation rate for marine fishes.
-\emph{Nature}, \emph{559}(7714), 392.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ramshaw1972time}{}}%
-Ramshaw, J., Richardson, D., Meatyard, B., Brown, R., Richardson, M., Thompson, E., \& Boulter, D.
-(1972).
-The time of origin of the flowering plants determined by using amino acid sequence data of cytochrome c.
-\emph{New Phytologist}, \emph{71}(5), 773--779.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ratnasingham2007bold}{}}%
-Ratnasingham, S., \& Hebert, P. D. (2007). BOLD: The barcode of life data system (http://www. Barcodinglife. org). \emph{Molecular Ecology Notes}, \emph{7}(3), 355--364.
-
-\leavevmode\vadjust pre{\hypertarget{ref-rees2017automated}{}}%
-Rees, J. A., \& Cranston, K. (2017). Automated assembly of a reference taxonomy for phylogenetic data synthesis. \emph{Biodiversity Data Journal}, (5).
-
-\leavevmode\vadjust pre{\hypertarget{ref-Revell2012}{}}%
-Revell, L. J. (2012). Phytools: An r package for phylogenetic comparative biology (and other things). \emph{{Methods in Ecology and Evolution}}, \emph{3}, 217--223.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Ronquist2003}{}}%
-Ronquist, F., \& Huelsenbeck, J.
-P.
-(2003).
-{MrBayes 3: Bayesian phylogenetic inference under mixed models}.
-\emph{Bioinformatics}, \emph{19}(12), 1572--1574.
-\url{https://doi.org/10.1093/bioinformatics/btg180}
-
-\leavevmode\vadjust pre{\hypertarget{ref-datelifeplot}{}}%
-Sanchez-Reyes, L.
-L., \& O'Meara, B.
-(2022).
-{\texttt{datelifeplot}: Methods to plot chronograms and outputs of the \texttt{datelife} package}.
-\emph{R Package Release V0.2.2}.
-Retrieved from \url{https://zenodo.org/badge/latestdoi/381501451}
-
-\leavevmode\vadjust pre{\hypertarget{ref-datelifeR}{}}%
-Sanchez-Reyes, L.
-L., O'Meara, B., Eastman, J., Heath, T., Wright, A., Schliep, K., \ldots{} Alfaro, M.
-(2022).
-{\texttt{datelife}: Scientific Data on Time of Lineage Divergence for Your Taxa}.
-\emph{R Package Release V0.6.2}.
-Retrieved from \url{https://doi.org/10.5281/zenodo.593938}
-
-\leavevmode\vadjust pre{\hypertarget{ref-sanderson2002}{}}%
-Sanderson, Michael J.
-(2002).
-{Estimating Absolute Rates of Molecular Evolution and Divergence Times: A Penalized Likelihood Approach}.
-\emph{Molecular Biology and Evolution}, \emph{19}(1), 101--109.
-\url{https://doi.org/10.1093/oxfordjournals.molbev.a003974}
-
-\leavevmode\vadjust pre{\hypertarget{ref-sanderson2003r8s}{}}%
-Sanderson, Michael J. (2003). r8s: Inferring absolute rates of molecular evolution and divergence times in the absence of a molecular clock. \emph{Bioinformatics}, \emph{19}(2), 301--302.
-
-\leavevmode\vadjust pre{\hypertarget{ref-sanderson2001sources}{}}%
-Sanderson, Michael J., \& Doyle, J.
-A.
-(2001).
-Sources of error and confidence intervals in estimating the age of angiosperms from rbcL and 18S rDNA data.
-\emph{American Journal of Botany}, \emph{88}(8), 1499--1516.
-
-\leavevmode\vadjust pre{\hypertarget{ref-sauquet2013practical}{}}%
-Sauquet, H. (2013). A practical guide to molecular dating. \emph{{Comptes Rendus Palevol}}, \emph{12}(6), 355--367.
-
-\leavevmode\vadjust pre{\hypertarget{ref-sauquet2021age}{}}%
-Sauquet, H., Ramírez-Barahona, S., \& Magallón, S. (2021). \emph{The age of flowering plants is unknown}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-schenk2016sec}{}}%
-Schenk, J. J. (2016). {Consequences of secondary calibrations on divergence time estimates}. \emph{PLoS ONE}, \emph{11}(1). \url{https://doi.org/10.1371/journal.pone.0148228}
-
-\leavevmode\vadjust pre{\hypertarget{ref-schliep2011phangorn}{}}%
-Schliep, K. P. (2011). Phangorn: Phylogenetic analysis in r. \emph{Bioinformatics}, \emph{27}(4), 592--593.
-
-\leavevmode\vadjust pre{\hypertarget{ref-selvatti2015paleogene}{}}%
-Selvatti, A.
-P., Gonzaga, L.
-P., \& Moraes Russo, C.
-A.
-de.
-(2015).
-A paleogene origin for crown passerines and the diversification of the oscines in the new world.
-\emph{Molecular Phylogenetics and Evolution}, \emph{88}, 1--15.
-
-\leavevmode\vadjust pre{\hypertarget{ref-smith2018constructing}{}}%
-Smith, Stephen A., \& Brown, J. W. (2018). Constructing a broadly inclusive seed plant phylogeny. \emph{American Journal of Botany}, \emph{105}(3), 302--314.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Smith2012}{}}%
-Smith, Stephen A., \& O'Meara, B.
-C.
-(2012).
-{TreePL: Divergence time estimation using penalized likelihood for large phylogenies}.
-\emph{Bioinformatics}, \emph{28}(20), 2689--2690.
-\url{https://doi.org/10.1093/bioinformatics/bts492}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Stoltzfus2013}{}}%
-Stoltzfus, A., Lapp, H., Matasci, N., Deus, H., Sidlauskas, B., Zmasek, C.
-M., \ldots{} Jordan, G.
-(2013).
-{Phylotastic!
-Making tree-of-life knowledge accessible, reusable and convenient}.
-\emph{{BMC Bioinformatics}}, \emph{14}.
-\url{https://doi.org/10.1186/1471-2105-14-158}
-
-\leavevmode\vadjust pre{\hypertarget{ref-tietze2013complete}{}}%
-Tietze, D.
-T., Päckert, M., Martens, J., Lehmann, H., \& Sun, Y.-H.
-(2013).
-Complete phylogeny and historical biogeography of true rosefinches (aves: carpodacus).
-\emph{Zoological Journal of the Linnean Society}, \emph{169}(1), 215--234.
-
-\leavevmode\vadjust pre{\hypertarget{ref-title2016macrophylogenies}{}}%
-Title, P.
-O., \& Rabosky, D.
-L.
-(2016).
-{Do Macrophylogenies Yield Stable Macroevolutionary Inferences?
-An Example from Squamate Reptiles}.
-\emph{Systematic Biology}, syw102.
-\url{https://doi.org/10.1093/sysbio/syw102}
-
-\leavevmode\vadjust pre{\hypertarget{ref-treplin2008molecular}{}}%
-Treplin, S., Siegert, R., Bleidorn, C., Thompson, H.
-S., Fotso, R., \& Tiedemann, R.
-(2008).
-Molecular phylogeny of songbirds (aves: Passeriformes) and the relative utility of common nuclear marker loci.
-\emph{Cladistics}, \emph{24}(3), 328--349.
-
-\leavevmode\vadjust pre{\hypertarget{ref-vos2012nexml}{}}%
-Vos, R.
-A., Balhoff, J.
-P., Caravas, J.
-A., Holder, M.
-T., Lapp, H., Maddison, W.
-P., \ldots{} others.
-(2012).
-NeXML: Rich, extensible, and verifiable representation of comparative data and metadata.
-\emph{Systematic Biology}, \emph{61}(4), 675--689.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Webb2000}{}}%
-Webb, C. (2000). {Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees}. \emph{{The American Naturalist}}, \emph{156}(2), 145--155.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Webb2008}{}}%
-Webb, Campbell O., Ackerly, D.
-D., \& Kembel, S.
-W.
-(2008).
-{Phylocom: Software for the analysis of phylogenetic community structure and trait evolution}.
-\emph{Bioinformatics}, \emph{24}(18), 2098--2100.
-\url{https://doi.org/10.1093/bioinformatics/btn358}
-
-\leavevmode\vadjust pre{\hypertarget{ref-webb2005phylomatic}{}}%
-Webb, Campbell O., \& Donoghue, M. J. (2005). Phylomatic: Tree assembly for applied phylogenetics. \emph{{Molecular Ecology Notes}}, \emph{5}(1), 181--183.
-
-\leavevmode\vadjust pre{\hypertarget{ref-weir2008calibrating}{}}%
-Weir, J., \& Schluter, D. (2008). Calibrating the avian molecular clock. \emph{Molecular Ecology}, \emph{17}(10), 2321--2328.
-
-\end{CSLReferences}
-
-\endgroup
-
-\end{document}
+\captionsetup[table]{font+=linenumbers, labelfont={sc}, labelformat={default}, labelsep=period, name={Supplementary Table}}

### 2022-06-22 manuscript/supplementary_figured.Rmd
homogenizing to 'datelife workflow' instead of 'datelife algorithm'

@@
 floatsintext      : no
-figurelist        : NO
+figurelist        : no
 tablelist         : no
@@
     includes      :
-      in_header: "preamble.tex"
+      in_header: "preamble-suppl.tex"
       after_body  : ["figures-supplementary.tex"]

### 2022-06-22 manuscript/supplementary_figured.tex
homogenizing to 'datelife workflow' instead of 'datelife algorithm'

@@
 \documentclass[
-  english,
   man]{apa6}
-\title{Supplementary Figures \linebreak DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life}
-\author{Luna L. Sánchez Reyes\textsuperscript{1,2}, Emily Jane McTavish\textsuperscript{1}, \& Brian O'Meara\textsuperscript{2}}
-\date{}
-
 \usepackage{amsmath,amssymb}
@@
 \fi
+\ifLuaTeX
+\usepackage[bidi=basic]{babel}
+\else
+\usepackage[bidi=default]{babel}
+\fi
+\babelprovide[main,import]{english}
+% get rid of language-specific shorthands (see #6817):
+\let\LanguageShortHands\languageshorthands
+\def\languageshorthands#1{}
 \DeclareUnicodeCharacter{0301}{*************************************}
+
+% leave the figure captions outside\ caption{} if you want them
+% to be formatted in the same way as the general text (double spaced and linenumbered)
+% line numbers for captions trick from https://latex.org/forum/viewtopic.php?t=34614
+
+\usepackage{ragged2e}
+
+\usepackage{caption}
+
+\usepackage{lineno}
+\DeclareCaptionFont{linenumbers}{\internallinenumbers}
+
+\usepackage{float}
+\restylefloat{table}
+
+\captionsetup[figure]{font+=linenumbers, labelfont={sc}, labelformat={default}, labelsep=period, name={Supplementary Figure}}
+
+\captionsetup[table]{font+=linenumbers, labelfont={sc}, labelformat={default}, labelsep=period, name={Supplementary Table}}
+
 % Manuscript styling
@@
 \fi
+
+\title{Supplementary Figures \linebreak DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life}
+\author{Luna L. Sánchez Reyes\textsuperscript{1,2}, Emily Jane McTavish\textsuperscript{1}, \& Brian O'Meara\textsuperscript{2}}
+\date{}
 
@@
 \maketitle
-
-% in this file I leave the figure captions outside\ caption{} because I want them
-% to be formatted in the same way as the general text (double spaced and linenumbered)
-\captionsetup[figure]{labelfont={sc},labelformat={default},labelsep=period,name={Supplementary Figure}}
 

### 2022-06-22 manuscript/fig-small-example.Rmd
fixing format of embedded figures and tables
various edits; removing tex file of figures that is no longer necessary

@@
 
+\begin{center}
+\includegraphics{../figures/figure2/figure2.pdf}
+\end{center}
 \vspace{10mm}
-
-\begin{center}
-\includegraphics[width = 6.5in]{../figures/figure2/figure2.pdf}
-\end{center}
-\captionof{figure}{DateLife analysis results for a small sample of A) 6 bird species within the Passeriformes.
+\captionof{figure}{\textit{DateLife analysis results for a small sample of A) 6 bird species within the Passeriformes.
 B) Processed species names were found across 9 chronograms within 6 independent studies (Barker et al.
@@
 (\protect\hyperlink{ref-Jetz2012}{2012}).) C) This revealed 28 source age data points for the queried species names.
-Summarized age data is used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree, resulting in a summary chronogram of source ages.}
+Summarized age data is used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree, resulting in a summary chronogram of source ages.}}
 \label{fig:figure2}
-
-\vspace{15mm}

### 2022-06-22 manuscript/fig-topologies.Rmd
various edits; removing tex file of figures that is no longer necessary

@@
 \includegraphics{../figures/fringillidae-topologies/fringillidae-topology.pdf}
-  \captionof{figure}{Tree topologies obtained from Open Tree of Life's (OpenTree) synthetic phylogenetic tree.
-A) Topology of 2,333 tips and 1,305 internal nodes, encompassing bird species within the family Fringillidae following the NCBI taxonomy (black), as well as all other bird species that share the same mrca node in OpenTree's synthetic tree (purple).
-B) Topology of 289 tips and 253 internal nodes, encompassing bird species within the Fringillidae only. 
-This tree results from pruning species that are not assigned to the family Fringillidae (purple branches in topology A) but that form a monophyletic group with them.
+\captionof{figure}{Tree topologies extracted from Open Tree of Life's (OpenTree) synthetic phylogenetic tree.
+A) Topology of 2,333 tips and 1,305 internal nodes, encompassing bird species within the family Fringillidae following the NCBI taxonomy (black), as well as all other bird species that share the same Most Recent Common Ancestor (MRCA) node in OpenTree's synthetic tree (purple).
+B) Subtree topology of of 289 tips and 253 internal nodes, resulting from pruning species that do not belong to the family Fringillidae according to the NCBI taxonomy (purple branches in topology A).
 Bird species within the Fringillidae are paraphyletic

### 2022-06-22 manuscript/fig-workflow.Rmd
various edits; removing tex file of figures that is no longer necessary

@@
 \begin{minipage}[t]{17.3cm}
-\begin{center}
 \includegraphics{../figures/figure1/figure1-new.pdf}
-\captionof{figure}{Main DateLife workflow.
+\captionof{figure}{\textit{Main DateLife workflow.
 Analyses can be performed via DateLife's interactive website at \url{www.datelife.org}, or using the \texttt{datelife} R package.
-Details on the R functions used to perform the analyses are available from \texttt{datelife}'s R package vignettes at \url{phylotastic.org/datelife}.}
+Details on the R functions used to perform the analyses are available from \texttt{datelife}'s R package vignettes at \url{phylotastic.org/datelife}.}}
 \label{fig:workflow}
-\end{center}
 \end{minipage}

### 2022-06-22 manuscript/manuscript.Rmd
various edits; removing tex file of figures that is no longer necessary

@@
   - id            : "1"
-    institution   : "University of California, Merced"
+    institution   : "University of California, Merced, USA"
   - id            : "2"
-    institution   : "University of Tennessee, Knoxville"
+    institution   : "University of Tennessee, Knoxville, USA"
 
 authornote: |
-  School of Natural Sciences, University of California, Merced, Science and Engineering Building 1.
+  School of Natural Sciences, University of California, Merced, 258 Science and Engineering Building 1, Merced, CA 95340, USA.
 
@@
 DateLife's core software application consists of the R package `datelife`.
-Its latest stable version -- v`r packageVersion("datelife")`, is available from the CRAN repository [@datelifeR], and relies on functionalities from various biological R packages:
+Its current stable version -- v`r packageVersion("datelife")`, is available from the The Comprehensive R Archive Network (CRAN) repository [@datelifeR], and relies on functionalities from various biological R packages:
 ape [@paradis2004],
@@
 If an input taxon name belongs to an inclusive taxonomic group, DateLife has two alternative behaviors defined by the "get species from taxon" flag.
-If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group (following a taxonomy of choice; current options are Open Tree of Life, NCBI, GBIF, or IRMNG taxonomies) and adds them to the input string.
+If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group following a standard taxonomy of choice, and adds them to the input string.
+Taxonomies currently supported by DateLife are Open Tree of Life (OpenTree) unified Taxonomy [OTT, @rees2017automated], the National Center of Biotechnology Information (NCBI) taxonomic database [@schoch2020ncbi], the Global Biodiversity Information Facility (GBIF) taxonomic backbone [@gbif2022taxonomy], and the Interim Register of Marine and Nonmarine Genera (IRMNG) database [@rees2017irmng].
 If the flag is inactive, DateLife excludes any taxon names above the species level from the search query.
@@
 TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard [@Boyle2013].
-DateLife implements TNRS using OpenTree's unified taxonomy as standard [@opentreeAPIs; @rees2017automated], storing OpenTree's Taxonomy identification numbers for further processing.
+DateLife implements TNRS with OTT as standard [@opentreeAPIs; @rees2017automated], storing taxonomic identification numbers for further processing.
 <!-- We are using OpenTree's TNRS [BCO: yes], Boyle described TNRS first, right?  [BCO: earliest I can find, but feels like there should be much older work]-->
@@
 The processed input taxon names are saved as an R object of a newly defined class, \texttt{datelifeQuery}, that is used in the following steps.
-This object contains the standardized names, the corresponding OpenTree taxonomic id numbers, and the topology of the input tree if any was provided.
+This object contains the standardized names, the corresponding OTT identification numbers, and the topology of the input tree if any was provided.
 
@@
 (`r format(Sys.time(), "%b %d, %Y")`),
-DateLife's chronogram database latest version consist of `r length(opentree_chronograms$trees)` chronograms published in `r length(unique(opentree_chronograms$studies))` different studies.
+DateLife's chronogram database latest version consist of `r format(length(opentree_chronograms$trees), big.mark=",")` chronograms published in `r length(unique(opentree_chronograms$studies))` different studies.
 It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata [@mctavish2015phylesystem], which allows automatic and reproducible assembly of our chronogram database.
@@
 
-A unique feature of OpenTree's Phylesystem is that the community can add new published, state-of-the-art chronograms any time through their curator application (https://tree.opentreeoflife.org/curator).
-As chronograms are added to Phylesystem, they are incorporated into an updated `datelife`'s database that is assigned a new version number, followed by a package release on CRAN.
+A unique feature of the Phylesystem is that any user can add new published, state-of-the-art chronograms any time, through their curator application (https://tree.opentreeoflife.org/curator).
+As chronograms are added to Phylesystem, they are incorporated into the chronogram database of the `datelife` package.
+The updated database is assigned a new version number, followed by a package release on CRAN.
 `datelife`'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
 <!--BCO: Nicely done-->
-Users can also upload and curate new chronograms to OpenTree themselves.
-Then, trigger an update of their local `datelife` database to incorporate the new chronograms and have them available for their current analysis.
+Users can also implement functions from the `datelife` R package to trigger an update of the local chronogram database, to incorporate any new chronograms to the user's DateLife analysis before an official database update is released on CRAN.
 <!--BCO: What is immediately?
@@
 I remember being frustrated trying to load in giant trees for datelife use b/c I had to click on the ingroup node.
-Luna: Very true, it is not that easy, I added the curation step to hint that there are more steps that simply uploading the chronogram.-->
-<!--BCO; this duplicates an earlier section and feels more methods-y. Maybe move up and merge? 
-Luna: Done! Moved here.-->
+Luna: Very true, it is not that easy, I added the curation step to hint that there are more steps than simply uploading the chronogram.-->
 
@@
 We recommend that users provide a tree topology as input from the literature, or one of their own making.
-If no topology is provided, DateLife automatically obtains one from the OpenTree synthetic tree [@opentreeoflife2019synth].
-These are known as "induced" OpenTree subtrees, which do not necessarily correspond to a monophyletic group.
+If no topology is provided, DateLife automatically extracts one from the OpenTree synthetic tree, a phylogeny encompassing 2.3 million taxa across all life, assembled from 1, 239 published phylogenetic trees and OpenTree's unified Taxonomy, OTT  [@opentreeoflife2019synth].
 Alternatively, DateLife can combine topologies from source chronograms using a supertree approach.
@@
 
-Alternative options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the calibrations.
+Alternative phylogenetic dating options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the calibrations.
 PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees.
-treePL [@Smith2012], is a semi-parametric, rate-smoothing, penalized likelihood dating method
-[@sanderson2002].
-The MrBayes  [@Huelsenbeck2001; @Ronquist2003] approach in datelife uses the calibrations as priors on node ages.
+treePL [@Smith2012], is a semi-parametric, rate-smoothing, penalized likelihood dating method [@sanderson2002].
+The MrBayes  [@Huelsenbeck2001; @Ronquist2003] approach in DateLife uses the calibrations as priors on node ages.
 
 _**Visualizing results.--**_
-Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as visualize and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions available from the R package `datelifeplot` [@datelifeplot].
+Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as visualize and compare results graphically, or construct their own graphs using DateLife's chronogram plot generation functions available from the R package `datelifeplot` [@datelifeplot].
 
@@
 
+\newpage
+
 # Benchmark
 
-`datelife`'s code speed was tested on an Apple iMac
+`datelife`'s R package code speed was tested on an Apple iMac
 with one 3.4 GHz Intel Core i5 processor. <!--No multiprocessor tests were performed-->
@@
 implemented with the R package testthat [@RCoreTeam2018] that were run both locally
-with the devtools package [@RCoreTeam2018], and on a public server --via
-GitHub, using the continuous integration tool Travis CI (<https://travis-ci.org>). At
-present, unit tests cover more than 40% of `datelife`'s code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->
+with the devtools package [@RCoreTeam2018], and on a public server using the continuous integration tool of GitHub actions (<https://docs.github.com/en/actions>). 
+At present, unit tests cover more than 40% of `datelife`'s code (<https://codecov.io/gh/phylotastic/datelife>). <!-- unit test make code robust, verifiable, debuggable, -->
 Unit testing helps identify potential issues as code is updated or, more critically, as services code relies upon may change.
@@
 
-_**Creating a search query.-**_
+### Creating a search query
+
 We chose 6 bird species within the Passeriformes. The sample includes
@@
 and one species of tanager, the vegetarian finch -- *Platyspiza crassirostris*.
-Processing of input names found that *Emberiza elegans* is synonym for *Schoeniclus elegans* in the default reference taxonomy (Open Tree of Life Taxonomy v3.3, June 1, 2021).
+Processing of input names found that *Emberiza elegans* is synonym for *Schoeniclus elegans* in the default reference taxonomy (OTT v3.3, June 1, 2021).
 For a detailed discussion on the state of the synonym, refer to Avibase [@lepage2004avibase; @lepage2014avibase; @avibase-emberiza].
@@
 
-\vspace{3mm}
+\newpage
 
-_**Searching the database.-**_
+### Searching the database
+
 DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies (Fig.
@@
 
-_**Summarizing search results.-**_
+### Summarizing search results
+
 DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and congruified and mapped age data to nodes in this chosen topology (Table \ref{tbl:table1}).
@@
 
-\vspace{30mm}
+\newpage
 
@@
 
+\newpage
+
 ## An example with the family of true finches
 
-_**Creating a query.-**_
+### Creating a query
+
 To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the "get species from taxon" flag,
@@
 
-\vspace{3mm}
+### Searching the database
 
-_**Searching the database.-**_
 Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species. 
 The DateLife search identified `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
-Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices, revealing 1206 different age data points available for species within the Fringillidae (supplementray table 1).
+Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementray Table 1).
 
-\vspace{3mm}
+### Summarizing search results
 
-_**Summarizing search results.-**_
 The final step is to congruify and summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM. 
 As explained in the "Description" section, a tree topology to summarize age data upon is required.
-By default, DateLife uses as summarizing topology the portion of OpenTree synthetic phylogeny that contains the species provided as search query. 
-According to this phylogeny, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).
+By default, DateLife uses the topology from OpenTree's synthetic tree that contains the species in the search query to summarize age data upon. 
+According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).
 
@@
 
-Age data from source chronograms was congruified to the chosen tree topology (OpenTree's synthetic subtree from figure \ref{fig:fringillidae-topologies}B), reducing the age data set to 818 different data points (supplementray table 2).
-Age summary statistics per congruified node were calculated and used as fixed secondary calibrations over the chosen tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
+Age data from source chronograms was congruified to OpenTree's topology (Fig.
+\ref{fig:fringillidae-topologies}B), reducing the age data set to 818 different data points (Supplementray Table 2).
+For each congruent node, age summary statistics were calculated and used as fixed secondary calibrations over the chosen tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
 \ref{fig:fringillidages}).
@@
 We then estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.
-
 We found that node ages from original study, and ages estimated using all other age data available are correlated (Fig. \ref{fig:cvXY}).
-
 For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka 'closer to the root') relative to the original estimate, and overestimate ages for nodes closer to the tips.
@@
 
-`datelife` makes state-of-the-art data on evolutionary time frame easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
+DateLife makes state-of-the-art data on evolutionary time frame easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
 It is an open service that does not require any expert biological knowledge from users --besides the names of the species or group they want to work with, for any of its functionality.
 
-A total of `r length(unique(unname(unlist(lapply(opentree_chronograms$trees, "[[", "tip.label")))))` unique terminal taxa are represented in `datelife`'s database.
-Incorporation of more chronograms into `datelife`'s database will continue to improve its services.
+A total of `r format(length(unique(unname(unlist(lapply(opentree_chronograms$trees, "[[", "tip.label"))))), big.mark=",")` unique terminal taxa are represented in DateLife's database.
+Incorporation of more chronograms into the database will continue to improve DateLife's services.
 One option to increase the number of chronograms in the DateLife database is the Dryad data repository.
@@
 
-## Primary vs Secondary calibrations in dating analyses
+## Age variation in source chronograms
 
-While most chronograms in DateLife's database are constructed using primary calibrations (molecular substitution rates or ages obtained from the fossil record or geological events), DateLife summarizes chronograms using secondary calibrations only (ages coming from other chronograms).
+Conflict in estimated ages among alternative studies is common in the literature.
+See, for example, the robust ongoing debate about crown group age of angiosperms  [@sauquet2021age; @barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time].
+Source chronograms available for the same organisms have potentially been estimated implementing calibrations very differently.
+For example, the chronograms from @burns2014phylogenetics were inferred using molecular substitution rate estimates across birds [@weir2008calibrating], and have much older age estimates for the same nodes than chronograms that were inferred using fossils as calibrations (Figs.
+\ref{fig:fringillidages}, \ref{fig:cvXY}; Supplementary Figs.
+S1, S4).
+<!--\textcolor{red}{EJM: MORE ON the FINCH examples and the cross validation here}-->
 
+Different calibration implementations might also imply fundamentally distinct evolutionary hypotheses [@antonelli2017supersmart].
+For example, two independent researchers working on the same clade should both carefully select and justify their choices of fossil calibration placement.
+Yet, if one researcher concludes that a fossil should calibrate the ingroup of a clade, while another researcher concludes that the same fossil should calibrate the outgroup of the clade, the resulting age estimates will differ, as the placement of calibrations as stem or crown group has been proven to significantly affect time of lineage divergence estimates [@sauquet2013practical].
+
+### Primary vs Secondary calibrations
+
+While most chronograms in DateLife's database are constructed using primary calibrations (molecular substitution rates or ages obtained from the fossil record or geological events), DateLife summarizes chronograms using secondary calibrations (ages coming from other chronograms).
 @graur2004reading cautioned on the increased error and uncertainty in estimated ages when using secondary calibrations in dating analyses.
@@
 However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures S2-S9).
-The only excpetion to tjis was observed on results of the cross validation analysis of the @burns2014phylogenetics chronogram, which results in much younger node ages when estimated using secondary calibrations (Supplementary figs.
+The only excpetion to tjis was observed on results of the cross validation analysis of the @burns2014phylogenetics chronogram, which results in much younger node ages when estimated using secondary calibrations (Supplementary Figs.
 S1 and S4).
-
-## Variation in source chronograms
-
-Conflict among alternative dating studies is common in the literature
-(see, for example, the robust ongoing debate about crown group age of angiosperms [@sauquet2021age; @barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time]).
-
-Source chronograms may have been estimated using different types of primary calibrations (fossil record, geologic or biogeographic events, molecular substitution rates).
-These differences in approach can deepen the already substantial variation in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
-For example, the chronograms from @burns2014phylogenetics were inferred using molecular substitution rate estimates across birds [@weir2008calibrating], and have much older date estimates for the same nodes than chronograms that were inferred using fossils as primary calibrations (figs.
-\ref{fig:fringillidages}, \ref{fig:cvXY}; supplementary figs.
-S1, S4).
-<!--\textcolor{red}{EJM: MORE ON the FINCH examples and the cross validation here}-->
-
-Different chronograms might also imply fundamentally distinct evolutionary hypotheses.
-For example, two different researchers working on the same clade could both carefully select and argue for their choices of fossil calibrations.
-Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses the same one to calibrate outside the clade, the resulting age estimates will often differ substantially, as the placement of calibrations as stem or crown group is proved to deeply affect estimated times of lineage divergence [@sauquet2013practical].
 
@@
 Hence, DateLife's workflow features different ways of estimating node ages in the absence of calibrations and branch length information for certain taxa.
-"Making up" branch lengths is a common practice in scientific publications: @Jetz2012, created a time-calibrated tree of all 9,993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a tree of 353,185 seed plants where only 23% had molecular data.
+"Making up" branch lengths is a common practice in scientific publications: @Jetz2012, created a chronogram of all 9, 993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a chronogram of 31, 536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a chronogram of 353, 185 seed plants where only 23% had molecular data.
 
@@
 
-The `datelife` software is free and open source and it can be used through its R shiny web application at
-<http://www.datelife.org>, through the `datelife` R package, and through Phylotastic's project web portal <https://phylo.cs.nmsu.edu/>.
-DateLife's web application is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker and Open Tree of Life's infrastructure (dates.opentreeoflife.org/datelife).
-`datelife`'s R package stable version `r availability`
-for installation from the CRAN repository (<https://cran.r-project.org/package=datelife>)
-using the command `install.packages(pkgs = "datelife")` from within R. Development versions
-are available from the GitHub repository (<https://github.com/phylotastic/datelife>)
-and can be installed using the command `devtools::install_github("phylotastic/datelife")`.
+The DateLife software is free and open source and it can be used through its R shiny web application at <http://www.datelife.org>, through the `datelife` R package, and through Phylotastic's project web portal <https://phylo.cs.nmsu.edu/>.
+DateLife's web application is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker and OpenTree's infrastructure (dates.opentreeoflife.org/datelife).
+`datelife`'s R package stable version `r availability` for installation from the CRAN repository (<https://cran.r-project.org/package=datelife>) using the command `install.packages(pkgs = "datelife")` from within R.
+Development versions are available from the GitHub repository (<https://github.com/phylotastic/datelife>) and can be installed using the command `devtools::install_github("phylotastic/datelife")`.
 

### 2022-06-22 manuscript/manuscript.tex
various edits; removing tex file of figures that is no longer necessary

@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4914}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4924}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 
-School of Natural Sciences, University of California, Merced, Science and Engineering Building 1.
+School of Natural Sciences, University of California, Merced, 258 Science and Engineering Building 1, Merced, CA 95340, USA.
 
@@
 
-\affiliation{\vspace{0.5cm}\textsuperscript{1} University of California, Merced\\\textsuperscript{2} University of Tennessee, Knoxville}
+\affiliation{\vspace{0.5cm}\textsuperscript{1} University of California, Merced, USA\\\textsuperscript{2} University of Tennessee, Knoxville, USA}
 
@@
 
-Chronograms --phylogenies with branch lengths proportional to time-- provide key data on evolutionary time frame for the study of natural processes in many areas of biological research, such as developmental biology (Delsuc et al., 2018; Laubichler \& Maienschein, 2009), conservation biology (Felsenstein, 1985; C.
-Webb, 2000), historical biogeography (Posadas, Crisci, \& Katinas, 2006), and species diversification (Magallon \& Sanderson, 2001; Morlon, 2014).
+Chronograms --phylogenies with branch lengths proportional to time-- provide key data on evolutionary time frame for the study of natural processes in many areas of biological research, such as developmental biology (Delsuc et al., 2018; Laubichler \& Maienschein, 2009), conservation biology (Felsenstein, 1985; Webb, 2000), historical biogeography (Posadas, Crisci, \& Katinas, 2006), and species diversification (Magallon \& Sanderson, 2001; Morlon, 2014).
 
@@
 DateLife's core software application consists of the R package \texttt{datelife}.
-Its latest stable version -- v0.6.4, is available from the CRAN repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
+Its current stable version -- v0.6.5, is available from the The Comprehensive R Archive Network (CRAN) repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
 ape (Paradis, Claude, \& Strimmer, 2004),
@@
 If an input taxon name belongs to an inclusive taxonomic group, DateLife has two alternative behaviors defined by the ``get species from taxon'' flag.
-If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group (following a taxonomy of choice; current options are Open Tree of Life, NCBI, GBIF, or IRMNG taxonomies) and adds them to the input string.
+If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group following a standard taxonomy of choice, and adds them to the input string.
+Taxonomies currently supported by DateLife are Open Tree of Life (OpenTree) unified Taxonomy (OTT, Rees \& Cranston, 2017), the National Center of Biotechnology Information (NCBI) taxonomic database (Schoch et al., 2020), the Global Biodiversity Information Facility (GBIF) taxonomic backbone (GBIF Secretariat, 2022), and the Interim Register of Marine and Nonmarine Genera (IRMNG) database (Rees, Vandepitte, Decock, \& Vanhoorne, 2017).
 If the flag is inactive, DateLife excludes any taxon names above the species level from the search query.
@@
 TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard (Boyle et al., 2013).
-DateLife implements TNRS using OpenTree's unified taxonomy as standard (Open Tree Of Life et al., 2016; Rees \& Cranston, 2017), storing OpenTree's Taxonomy identification numbers for further processing.
+DateLife implements TNRS with OTT as standard (Open Tree Of Life et al., 2016; Rees \& Cranston, 2017), storing taxonomic identification numbers for further processing.
 
 The processed input taxon names are saved as an R object of a newly defined class, \texttt{datelifeQuery}, that is used in the following steps.
-This object contains the standardized names, the corresponding OpenTree taxonomic id numbers, and the topology of the input tree if any was provided.
+This object contains the standardized names, the corresponding OTT identification numbers, and the topology of the input tree if any was provided.
 
@@
 At the time of writing of this manuscript
-(Jun 21, 2022),
+(Jun 22, 2022),
 DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies.
@@
 
-A unique feature of OpenTree's Phylesystem is that the community can add new published, state-of-the-art chronograms any time through their curator application (\url{https://tree.opentreeoflife.org/curator}).
-As chronograms are added to Phylesystem, they are incorporated into an updated \texttt{datelife}'s database that is assigned a new version number, followed by a package release on CRAN.
+A unique feature of the Phylesystem is that any user can add new published, state-of-the-art chronograms any time, through their curator application (\url{https://tree.opentreeoflife.org/curator}).
+As chronograms are added to Phylesystem, they are incorporated into the chronogram database of the \texttt{datelife} package.
+The updated database is assigned a new version number, followed by a package release on CRAN.
 \texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
-Users can also upload and curate new chronograms to OpenTree themselves.
-Then, trigger an update of their local \texttt{datelife} database to incorporate the new chronograms and have them available for their current analysis.
+Users can also implement functions from the \texttt{datelife} R package to trigger an update of the local chronogram database, to incorporate any new chronograms to the user's DateLife analysis before an official database update is released on CRAN.
 
@@
 We recommend that users provide a tree topology as input from the literature, or one of their own making.
-If no topology is provided, DateLife automatically obtains one from the OpenTree synthetic tree (Open Tree Of Life et al., 2019).
-These are known as ``induced'' OpenTree subtrees, which do not necessarily correspond to a monophyletic group.
+If no topology is provided, DateLife automatically extracts one from the OpenTree synthetic tree, a phylogeny encompassing 2.3 million taxa across all life, assembled from 1, 239 published phylogenetic trees and OpenTree's unified Taxonomy, OTT (Open Tree Of Life et al., 2019).
 Alternatively, DateLife can combine topologies from source chronograms using a supertree approach.
@@
 For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method {[}SDM; Criscuolo, Berry, Douzery, and Gascuel (2006){]}, which deforms patristic distance matrices by minimizing variance and then averaging them.
-These single summary taxon pair age matrices (Summarized calibrations) can be applied as calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Campbell O.
-Webb, Ackerly, \& Kembel, 2008; Campbell O.
-Webb \& Donoghue, 2005), and treePL (Stephen A.
-Smith \& O'Meara, 2012).
+These single summary taxon pair age matrices (Summarized calibrations) can be applied as calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Webb, Ackerly, \& Kembel, 2008; Webb \& Donoghue, 2005), and treePL (Smith \& O'Meara, 2012).
 
@@
 BLADJ fixes node ages that have calibration data, and distributes time between nodes with no data evenly between nodes with calibration data.
-This minimizes age variance in the resulting chronogram (Campbell O. Webb et al., 2008).
+This minimizes age variance in the resulting chronogram (Webb et al., 2008).
 BLADJ does not use branch lengths even when they are present in the input tree or summarizing topology.
@@
 
-Alternative options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the calibrations.
+Alternative phylogenetic dating options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the calibrations.
 PATHd8 is a non-clock, rate-smoothing method (Britton et al., 2007) to date trees.
-treePL (Stephen A. Smith \& O'Meara, 2012), is a semi-parametric, rate-smoothing, penalized likelihood dating method
-(Michael J. Sanderson, 2002).
-The MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003) approach in datelife uses the calibrations as priors on node ages.
+treePL (Smith \& O'Meara, 2012), is a semi-parametric, rate-smoothing, penalized likelihood dating method (Sanderson, 2002).
+The MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003) approach in DateLife uses the calibrations as priors on node ages.
 
 \emph{\textbf{Visualizing results.--}}
-Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R ``phylo'' format), as well as visualize and compare results graphically, or construct their own graphs using \texttt{datelife}'s chronogram plot generation functions available from the R package \texttt{datelifeplot} (Sanchez-Reyes \& O'Meara, 2022).
+Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R ``phylo'' format), as well as visualize and compare results graphically, or construct their own graphs using DateLife's chronogram plot generation functions available from the R package \texttt{datelifeplot} (Sanchez-Reyes \& O'Meara, 2022).
+
+\newpage
 
@@
 
-\texttt{datelife}'s code speed was tested on an Apple iMac
+\texttt{datelife}'s R package code speed was tested on an Apple iMac
 with one 3.4 GHz Intel Core i5 processor.
@@
 
-\begin{center}
 \includegraphics{../figures/fig_runtime_main.pdf}
-\end{center}
-  \captionof{figure}{DateLife's benchmarking results. Computation time used to process a query and a search across \texttt{datelife}'s chronogram database, relative to number of input taxon names.
-  \textit{Caption continues.}}
-
-\label{fig:runtime_main}
-
-\begin{justify}
-  \textit{Figure caption continued}. 
+\captionof{figure}{\textit{DateLife's benchmarking results.
+Computation time used to process a query and a search across \texttt{datelife}'s chronogram database, relative to number of input taxon names.
 For each N = \{10, 100, 200, ..., 1 000, ...
 , 9 000, 10 000\}, we sampled N species names from the class Aves a hundred times, and then performed a datelife search processing the input names with Taxon Names Resolution Service (TNRS; light gray), and without processing names (dark gray).
-For comparison, we performed a search using an input that had been pre-processed with TNRS (light blue).
-\end{justify}
-
-\vspace{20mm}
+For comparison, we performed a search using an input that had been pre-processed with TNRS (light blue).}}
+\label{fig:runtime_main}
 
@@
 implemented with the R package testthat (R Core Team, 2018) that were run both locally
-with the devtools package (R Core Team, 2018), and on a public server --via
-GitHub, using the continuous integration tool Travis CI (\url{https://travis-ci.org}). At
-present, unit tests cover more than 40\% of \texttt{datelife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).
+with the devtools package (R Core Team, 2018), and on a public server using the continuous integration tool of GitHub actions (\url{https://docs.github.com/en/actions}).
+At present, unit tests cover more than 40\% of \texttt{datelife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).
 Unit testing helps identify potential issues as code is updated or, more critically, as services code relies upon may change.
@@
 
-\emph{\textbf{Creating a search query.-}}
+\hypertarget{creating-a-search-query-1}{%
+\subsubsection{Creating a search query}\label{creating-a-search-query-1}}
+
 We chose 6 bird species within the Passeriformes. The sample includes
@@
 and one species of tanager, the vegetarian finch -- \emph{Platyspiza crassirostris}.
-Processing of input names found that \emph{Emberiza elegans} is synonym for \emph{Schoeniclus elegans} in the default reference taxonomy (Open Tree of Life Taxonomy v3.3, June 1, 2021).
+Processing of input names found that \emph{Emberiza elegans} is synonym for \emph{Schoeniclus elegans} in the default reference taxonomy (OTT v3.3, June 1, 2021).
 For a detailed discussion on the state of the synonym, refer to Avibase (Avibase, 2022; Lepage, 2004; Lepage, Vaidya, \& Guralnick, 2014).
@@
 
-\vspace{3mm}
-
-\emph{\textbf{Searching the database.-}}
+\newpage
+
+\hypertarget{searching-the-database}{%
+\subsubsection{Searching the database}\label{searching-the-database}}
+
 DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies (Fig.
@@
 
-\emph{\textbf{Summarizing search results.-}}
+\hypertarget{summarizing-search-results-1}{%
+\subsubsection{Summarizing search results}\label{summarizing-search-results-1}}
+
 DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and congruified and mapped age data to nodes in this chosen topology (Table \ref{tbl:table1}).
@@
 \vspace{10mm}
-\captionof{figure}{DateLife analysis results for a small sample of A) 6 bird species within the Passeriformes.
+\captionof{figure}{\textit{DateLife analysis results for a small sample of A) 6 bird species within the Passeriformes.
 B) Processed species names were found across 9 chronograms within 6 independent studies (Barker et al.
@@
 (\protect\hyperlink{ref-Jetz2012}{2012}).) C) This revealed 28 source age data points for the queried species names.
-Summarized age data is used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree, resulting in a summary chronogram of source ages.}
+Summarized age data is used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree, resulting in a summary chronogram of source ages.}}
 
@@
 
-\begin{table}[H]
-  \caption{Ages of congruified nodes obtained from a datelife search for the Small Example presented. See Figure 3, step C2.}
+\newpage
+
+\captionof{table}{Ages of congruified nodes. See Figure 3, step C2.}
 \label{tbl:table1}
 \includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example.pdf}
-\end{table}
-
-\begin{table}[H]
-  \caption{Summary of congruified nodes ages. See Figure 3, step C3.}
+
+\vspace{15mm}
+
+\captionof{table}{Summary of congruified nodes ages. See Figure 3, step C3.}
+
 \label{tbl:table2}
 \includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example-summary.png}
-\end{table}
+
+\newpage
 
@@
 
-\emph{\textbf{Creating a query.-}}
+\hypertarget{creating-a-query}{%
+\subsubsection{Creating a query}\label{creating-a-query}}
+
 To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the ``get species from taxon'' flag,
@@
 
-\vspace{3mm}
-
-\emph{\textbf{Searching the database.-}}
+\hypertarget{searching-the-database-1}{%
+\subsubsection{Searching the database}\label{searching-the-database-1}}
+
 Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species.
 The DateLife search identified 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Price et al., 2014).
-Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices, revealing 1206 different age data points available for species within the Fringillidae (supplementray table 1).
-
-\vspace{3mm}
-
-\emph{\textbf{Summarizing search results.-}}
+Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementray Table 1).
+
+\hypertarget{summarizing-search-results-2}{%
+\subsubsection{Summarizing search results}\label{summarizing-search-results-2}}
+
 The final step is to congruify and summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM.
 As explained in the ``Description'' section, a tree topology to summarize age data upon is required.
-By default, DateLife uses as summarizing topology the portion of OpenTree synthetic phylogeny that contains the species provided as search query.
-According to this phylogeny, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).
+By default, DateLife uses the topology from OpenTree's synthetic tree that contains the species in the search query to summarize age data upon.
+According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).
 
@@
 We then estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.
-
 We found that node ages from original study, and ages estimated using all other age data available are correlated (Fig. \ref{fig:cvXY}).
-
 For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka `closer to the root') relative to the original estimate, and overestimate ages for nodes closer to the tips.
@@
 
-\texttt{datelife} makes state-of-the-art data on evolutionary time frame easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
+DateLife makes state-of-the-art data on evolutionary time frame easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
 It is an open service that does not require any expert biological knowledge from users --besides the names of the species or group they want to work with, for any of its functionality.
 
-A total of 99474 unique terminal taxa are represented in \texttt{datelife}'s database.
-Incorporation of more chronograms into \texttt{datelife}'s database will continue to improve its services.
+A total of 99,474 unique terminal taxa are represented in DateLife's database.
+Incorporation of more chronograms into the database will continue to improve DateLife's services.
 One option to increase the number of chronograms in the DateLife database is the Dryad data repository.
@@
 
-\hypertarget{primary-vs-secondary-calibrations-in-dating-analyses}{%
-\subsection{Primary vs Secondary calibrations in dating analyses}\label{primary-vs-secondary-calibrations-in-dating-analyses}}
-
-While most chronograms in DateLife's database are constructed using primary calibrations (molecular substitution rates or ages obtained from the fossil record or geological events), DateLife summarizes chronograms using secondary calibrations only (ages coming from other chronograms).
-
+\hypertarget{age-variation-in-source-chronograms}{%
+\subsection{Age variation in source chronograms}\label{age-variation-in-source-chronograms}}
+
+Conflict in estimated ages among alternative studies is common in the literature.
+See, for example, the robust ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández, 2015; Ramshaw et al., 1972; Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021).
+Source chronograms available for the same organisms have potentially been estimated implementing calibrations very differently.
+For example, the chronograms from Burns et al.
+(2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older age estimates for the same nodes than chronograms that were inferred using fossils as calibrations (Figs.
+\ref{fig:fringillidages}, \ref{fig:cvXY}; Supplementary Figs.
+S1, S4).
+
+Different calibration implementations might also imply fundamentally distinct evolutionary hypotheses (Antonelli et al., 2017).
+For example, two independent researchers working on the same clade should both carefully select and justify their choices of fossil calibration placement.
+Yet, if one researcher concludes that a fossil should calibrate the ingroup of a clade, while another researcher concludes that the same fossil should calibrate the outgroup of the clade, the resulting age estimates will differ, as the placement of calibrations as stem or crown group has been proven to significantly affect time of lineage divergence estimates (Sauquet, 2013).
+
+\hypertarget{primary-vs-secondary-calibrations}{%
+\subsubsection{Primary vs Secondary calibrations}\label{primary-vs-secondary-calibrations}}
+
+While most chronograms in DateLife's database are constructed using primary calibrations (molecular substitution rates or ages obtained from the fossil record or geological events), DateLife summarizes chronograms using secondary calibrations (ages coming from other chronograms).
 Graur and Martin (2004) cautioned on the increased error and uncertainty in estimated ages when using secondary calibrations in dating analyses.
@@
 Certainly, further studies are required to fully understand the effect of secondary calibrations on outputs from different tree dating methods, and on downstream analyses.
-It is possible that secondary calibrations can be safely used with dating methods that do not require setting priors, such as penalized likelihood (Michael J.
-Sanderson, 2003), with methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ (Campbell O.
-Webb et al., 2008; Campbell O.
-Webb \& Donoghue, 2005), or methods that summarize age data unto a tree topology.
+It is possible that secondary calibrations can be safely used with dating methods that do not require setting priors, such as penalized likelihood (Sanderson, 2003), with methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ (Webb et al., 2008; Webb \& Donoghue, 2005), or methods that summarize age data unto a tree topology.
 
@@
 The only excpetion to tjis was observed on results of the cross validation analysis of the Burns et al.
-(2014) chronogram, which results in much younger node ages when estimated using secondary calibrations (Supplementary figs.
+(2014) chronogram, which results in much younger node ages when estimated using secondary calibrations (Supplementary Figs.
 S1 and S4).
-
-\hypertarget{variation-in-source-chronograms}{%
-\subsection{Variation in source chronograms}\label{variation-in-source-chronograms}}
-
-Conflict among alternative dating studies is common in the literature
-(see, for example, the robust ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández, 2015; Ramshaw et al., 1972; Michael J.
-Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021)).
-
-Source chronograms may have been estimated using different types of primary calibrations (fossil record, geologic or biogeographic events, molecular substitution rates).
-These differences in approach can deepen the already substantial variation in time estimates between lineages, as observed from the comparison of source chronograms in the Fringillidae example.
-For example, the chronograms from Burns et al.
-(2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older date estimates for the same nodes than chronograms that were inferred using fossils as primary calibrations (figs.
-\ref{fig:fringillidages}, \ref{fig:cvXY}; supplementary figs.
-S1, S4).
-
-Different chronograms might also imply fundamentally distinct evolutionary hypotheses.
-For example, two different researchers working on the same clade could both carefully select and argue for their choices of fossil calibrations.
-Still, if one researcher decides a fossil will calibrate the ingroup of a clade, while another researcher uses the same one to calibrate outside the clade, the resulting age estimates will often differ substantially, as the placement of calibrations as stem or crown group is proved to deeply affect estimated times of lineage divergence (Sauquet, 2013).
 
@@
 
-In ecology and conservation biology, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance (Campbell O.
-Webb et al., 2008).
+In ecology and conservation biology, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance (Webb et al., 2008).
 Hence, DateLife's workflow features different ways of estimating node ages in the absence of calibrations and branch length information for certain taxa.
 ``Making up'' branch lengths is a common practice in scientific publications: Jetz et al.
-(2012), created a time-calibrated tree of all 9,993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
-(2018) created a time-calibrated tree of 31,536 ray-finned fishes, of which only 37\% had molecular data; Stephen A.
-Smith and Brown (2018) constructed a tree of 353,185 seed plants where only 23\% had molecular data.
+(2012), created a chronogram of all 9, 993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
+(2018) created a chronogram of 31, 536 ray-finned fishes, of which only 37\% had molecular data; Smith and Brown (2018) constructed a chronogram of 353, 185 seed plants where only 23\% had molecular data.
 
@@
 
-The \texttt{datelife} software is free and open source and it can be used through its R shiny web application at
-\url{http://www.datelife.org}, through the \texttt{datelife} R package, and through Phylotastic's project web portal \url{https://phylo.cs.nmsu.edu/}.
-DateLife's web application is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker and Open Tree of Life's infrastructure (dates.opentreeoflife.org/datelife).
-\texttt{datelife}'s R package stable version is available
-for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife})
-using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R. Development versions
-are available from the GitHub repository (\url{https://github.com/phylotastic/datelife})
-and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.
+The DateLife software is free and open source and it can be used through its R shiny web application at \url{http://www.datelife.org}, through the \texttt{datelife} R package, and through Phylotastic's project web portal \url{https://phylo.cs.nmsu.edu/}.
+DateLife's web application is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker and OpenTree's infrastructure (dates.opentreeoflife.org/datelife).
+\texttt{datelife}'s R package stable version is available for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife}) using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R.
+Development versions are available from the GitHub repository (\url{https://github.com/phylotastic/datelife}) and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-gbif2022taxonomy}{}}%
+GBIF Secretariat.
+(2022).
+{GBIF Backbone Taxonomy}.
+\emph{{Checklist Dataset}}, (Online Resource accessed via GBIF.org).
+Retrieved from \href{\%7Bhttps://doi.org/10.15468/39omei\%20\%7D}{\{https://doi.org/10.15468/39omei \}}
+
 \leavevmode\vadjust pre{\hypertarget{ref-gibb2015new}{}}%
@@
 \leavevmode\vadjust pre{\hypertarget{ref-magallon2001absolute}{}}%
-Magallon, S., \& Sanderson, M. J. (2001). Absolute diversification rates in angiosperm clades. \emph{Evolution}, \emph{55}(9), 1762--1780.
+Magallon, S., \& Sanderson, M. (2001). Absolute diversification rates in angiosperm clades. \emph{Evolution}, \emph{55}(9), 1762--1780.
 
@@
 T., \ldots{} Smith, S.
-A.
 (2015).
@@
 \leavevmode\vadjust pre{\hypertarget{ref-rees2017automated}{}}%
-Rees, J. A., \& Cranston, K. (2017). Automated assembly of a reference taxonomy for phylogenetic data synthesis. \emph{Biodiversity Data Journal}, (5).
+Rees, \& Cranston, K. (2017). Automated assembly of a reference taxonomy for phylogenetic data synthesis. \emph{Biodiversity Data Journal}, (5).
+
+\leavevmode\vadjust pre{\hypertarget{ref-rees2017irmng}{}}%
+Rees, Vandepitte, L., Decock, W., \& Vanhoorne, B. (2017). {IRMNG 2006--2016: 10 Years of a Global Taxonomic Database}. \emph{Biodiversity Informatics}, \emph{12}.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-sanderson2002}{}}%
-Sanderson, Michael J.
+Sanderson, M.
+J.
 (2002).
@@
 \leavevmode\vadjust pre{\hypertarget{ref-sanderson2003r8s}{}}%
-Sanderson, Michael J. (2003). r8s: Inferring absolute rates of molecular evolution and divergence times in the absence of a molecular clock. \emph{Bioinformatics}, \emph{19}(2), 301--302.
+Sanderson, M. J. (2003). {r8s: Inferring Absolute Rates of Molecular Evolution and Divergence Times in the Absence of a Molecular Clock}. \emph{Bioinformatics}, \emph{19}(2), 301--302.
 
 \leavevmode\vadjust pre{\hypertarget{ref-sanderson2001sources}{}}%
-Sanderson, Michael J., \& Doyle, J.
+Sanderson, M.
+J., \& Doyle, J.
 A.
@@
 Schliep, K. P. (2011). Phangorn: Phylogenetic analysis in r. \emph{Bioinformatics}, \emph{27}(4), 592--593.
+
+\leavevmode\vadjust pre{\hypertarget{ref-schoch2020ncbi}{}}%
+Schoch, C.
+L., Ciufo, S., Domrachev, M., Hotton, C.
+L., Kannan, S., Khovanskaya, R., \ldots{} others.
+(2020).
+{NCBI Taxonomy: a Comprehensive Update on Curation, Resources and Tools}.
+\emph{Database}, \emph{2020}.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-smith2018constructing}{}}%
-Smith, Stephen A., \& Brown, J. W. (2018). Constructing a broadly inclusive seed plant phylogeny. \emph{American Journal of Botany}, \emph{105}(3), 302--314.
+Smith, S., \& Brown, J. (2018). Constructing a broadly inclusive seed plant phylogeny. \emph{American Journal of Botany}, \emph{105}(3), 302--314.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Smith2012}{}}%
-Smith, Stephen A., \& O'Meara, B.
-C.
+Smith, S., \& O'Meara, B.
 (2012).
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Webb2008}{}}%
-Webb, Campbell O., Ackerly, D.
-D., \& Kembel, S.
-W.
+Webb, C., Ackerly, D., \& Kembel, S.
 (2008).
@@
 \leavevmode\vadjust pre{\hypertarget{ref-webb2005phylomatic}{}}%
-Webb, Campbell O., \& Donoghue, M. J. (2005). Phylomatic: Tree assembly for applied phylogenetics. \emph{{Molecular Ecology Notes}}, \emph{5}(1), 181--183.
+Webb, C., \& Donoghue, M. (2005). Phylomatic: Tree assembly for applied phylogenetics. \emph{{Molecular Ecology Notes}}, \emph{5}(1), 181--183.
 

### 2022-06-23 manuscript/manuscript.Rmd
supplementary files for biorxiv submission

@@
 
-At the time of writing of this manuscript 
-(`r format(Sys.time(), "%b %d, %Y")`),
+At the time of writing of this manuscript (Jun 22, 2022),
+<!-- (`r format(Sys.time(), "%b %d, %Y")`), -->
 DateLife's chronogram database latest version consist of `r format(length(opentree_chronograms$trees), big.mark=",")` chronograms published in `r length(unique(opentree_chronograms$studies))` different studies.
@@
 The DateLife search identified `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
-Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementray Table 1).
+Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementray Table S1).
 
@@
 Age data from source chronograms was congruified to OpenTree's topology (Fig.
-\ref{fig:fringillidae-topologies}B), reducing the age data set to 818 different data points (Supplementray Table 2).
+\ref{fig:fringillidae-topologies}B), reducing the age data set to 818 different data points (Supplementray Table S2).
 For each congruent node, age summary statistics were calculated and used as fixed secondary calibrations over the chosen tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
@@
 \ref{fig:fringillidages}, \ref{fig:cvXY}; Supplementary Figs.
-S1, S4).
+S1, S5).
 <!--\textcolor{red}{EJM: MORE ON the FINCH examples and the cross validation here}-->
@@
 When ages are estimated with secondary calibrations, nodes closer to the root do tend to be slightly younger than ages estimated with primary calibrations.
-However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures S2-S9).
+However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures S2-S20).
 The only excpetion to tjis was observed on results of the cross validation analysis of the @burns2014phylogenetics chronogram, which results in much younger node ages when estimated using secondary calibrations (Supplementary Figs.
-S1 and S4).
+S1, S5).
 

### 2022-06-23 manuscript/manuscript.tex
supplementary files for biorxiv submission

@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4924}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4926}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 
-At the time of writing of this manuscript
-(Jun 22, 2022),
+At the time of writing of this manuscript (Jun 22, 2022),
 DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies.
@@
 The DateLife search identified 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Price et al., 2014).
-Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementray Table 1).
+Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementray Table S1).
 
@@
 \ref{fig:fringillidages}, \ref{fig:cvXY}; Supplementary Figs.
-S1, S4).
+S1, S5).
 
@@
 When ages are estimated with secondary calibrations, nodes closer to the root do tend to be slightly younger than ages estimated with primary calibrations.
-However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures S2-S9).
+However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures S2-S20).
 The only excpetion to tjis was observed on results of the cross validation analysis of the Burns et al.
 (2014) chronogram, which results in much younger node ages when estimated using secondary calibrations (Supplementary Figs.
-S1 and S4).
+S1, S5).
 

### 2022-06-23 R/cross-validation.R

@@ -1,27 +1,30 @@
+############################################################################
 # Function based on https://github.com/LunaSare/datelife_examples/blob/master/R/functions_data.R#L3
 # get calibrations form all other chronograms not belonging to the same study
 congruify_other_calibrations <- function(phylo_all, index) {
     # dd <- duplicated(names(phyloall))
     cat(index, " -- ", names(phylo_all)[index], "\n")
     study <- names(phylo_all) %in% names(phylo_all)[index]
-    res <- datelife::congruify_calibrations(phy = phylo_all[study][[1]],
-                                     chronograms = phylo_all[!study])
+    res <- datelife::congruify_and_mrca_multiPhylo(phy = phylo_all[study][[1]],
+                                                   source_chronograms = phylo_all[!study])
     res <- list(res)
     names(res) <- names(phylo_all)[index]
     res
 }
+############################################################################
 
 congruify_other_calibrations_all <- function(phylo_all){
-    # dd <- duplicated(names(phyloall))
+    # dd <- duplicated(names(phylo_all))
     res <- lapply(seq(phylo_all), function(i){
         cat(i, " -- ", names(phylo_all)[i], "\n")
         dd <- names(phylo_all) %in% names(phylo_all)[i]
-        datelife::congruify_calibrations(phy = phylo_all[dd][[1]],
-                                         chronograms = phylo_all[!dd])
+        datelife::congruify_and_mrca_multiPhylo(phy = phylo_all[dd][[1]],
+                                                source_chronograms = phylo_all[!dd])
     })
     names(res) <- names(phylo_all)
     res
 }
+############################################################################
 # Based on function use_othercals3 from https://github.com/LunaSare/datelife_examples/blob/master/R/functions_data.R#L36
 use_other_calibrations <- function(trees, other_calibrations, ...){
     res <- lapply(seq(trees), function(i){
@@ -36,22 +39,19 @@ use_other_calibrations <- function(trees, other_calibrations, ...){
     names(res) <- names(trees)
     res
 }
+############################################################################
 # Function to date source chronograms with data from other chronograms
-cross_date <- function(frin_other_calibs, index) {
-  cat(index, " -- ", names(frin_other_calibs)[index], "\n")
-  
-  calibrations <- frin_other_calibs[[index]]
-  
+cross_date <- function(calibrations, index) {
+  cat(index, " -- ", names(calibrations)[index], "\n")
+
+  calibrations <- calibrations[[index]]
[…]

### 2022-06-23 R/plot_node_ages2_custom.R

@@ -1,40 +1,39 @@
-datelifeplot::plot_node_ages2(my_chronogram,
-               title = c("Fringillidae Median Summary Chronogram"),
-               pos_title = -6,
-               cex_title = 2,
-               plot_type = "phyloch",
-               time_depth = 45,
-               pos_axis = 2,
-               cex_axis = 1,
-               edge.width = 3,
-               cex_tiplabels = 0.5,
-               matched_ages  = all_ages,
-               pch_type = point_type,
-               pch_color = point_colors,
-               pch_cex = point_cex,
-               bars_lwd = 7,
-               bars_color = "#80808060",
-#               legend_x = -5,
-#               legend_y = 200,
-               legend_text = leg_text,
-               legend_pch = leg_pch,
-               legend_color = leg_color,
-#               mai1 = 1,
-#               mai2 = 0.8,
-#               mai4 = 3.6,
-               axis_label = "",
-               legend_title = leg_title,
-               legend_box = leg_box,
-               mai1 = 0,
-               mai2 = 0,
-               mai3 = 0.5,
-               mai4 = 1,
-               omi1 = 0,
-               omi2 = 0,
-               omi3 = 0,
-               legend_cex = leg_cex,
-               legend_pt_cex = leg_pt_cex,
-               legend_x = c(1, 1),
-               legend_y = c(140, 80),
-
-)
+# datelifeplot::plot_node_ages2(my_chronogram,
+#                title = c("Fringillidae Median Summary Chronogram"),
+#                pos_title = -6,
+#                cex_title = 2,
+#                plot_type = "phyloch",
+#                time_depth = 45,
+#                pos_axis = 2,
+#                cex_axis = 1,
+#                edge.width = 3,
+#                cex_tiplabels = 0.5,
+#                matched_ages  = fringillidae_all_ages,
+#                pch_type = point_type,
+#                pch_color = point_colors,
+#                pch_cex = point_cex,
+#                bars_lwd = 7,
+#                bars_color = "#80808060",
+# #               legend_x = -5,
+# #               legend_y = 200,
+#                legend_text = leg_text,
+#                legend_pch = leg_pch,
+#                legend_color = leg_color,
+# #               mai1 = 1,
+# #               mai2 = 0.8,
+# #               mai4 = 3.6,
+#                axis_label = "",
+#                legend_title = leg_title,
+#                legend_box = leg_box,
+#                mai1 = 0,
[…]

### 2022-06-23 R/plot_node_ages_custom.R

@@ -1,28 +1,28 @@
-datelifeplot::plot_node_ages(my_chronogram,
-               title = "Median Summary Chronogram",
-               plot_type = "phyloch",
-               time_depth = 45,
-               calibration_summary = calib_summ,
-               pch = 19,
-               color_pch = color_studies,
-               cex_pch = 1,
-               lwd_bars = 10,
-               cex_legend = 0.7,
-               x_legend = 0 ,
-               y_legend = 270,
-               mai1 = 1,
-               mai2 = 0,
-               mai3 = 0.5,
-               mai4 = 3.73,
-               omi1 = 2,
-               omi2 = 1,
-               omi3 = 0.5,
-               omi4 = 0,
-               edge.width = 2,
-               cex_tiplabels = 0.5,
-               cex_axislabel = 1,
-               cex_axis = 1,
-               cex_title = 2,
-               pos_title = -1,
-               pos_axis = 2
-               )
+# datelifeplot::plot_node_ages(my_chronogram,
+#                title = "Median Summary Chronogram",
+#                plot_type = "phyloch",
+#                time_depth = 45,
+#                calibration_summary = calib_summ,
+#                pch = 19,
+#                color_pch = color_studies,
+#                cex_pch = 1,
+#                lwd_bars = 10,
+#                cex_legend = 0.7,
+#                x_legend = 0 ,
+#                y_legend = 270,
+#                mai1 = 1,
+#                mai2 = 0,
+#                mai3 = 0.5,
+#                mai4 = 3.73,
+#                omi1 = 2,
+#                omi2 = 1,
+#                omi3 = 0.5,
+#                omi4 = 0,
+#                edge.width = 2,
+#                cex_tiplabels = 0.5,
+#                cex_axislabel = 1,
+#                cex_axis = 1,
+#                cex_title = 2,
+#                pos_title = -1,
+#                pos_axis = 2
+#                )

### 2022-06-25 dryad/README.Rmd
fixing readme for dryad submission

---
title: 'README for Dryad Data Package from study "DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life"'
output:
  pdf_document:
      keep_md: yes
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This README file was generated on `r Sys.Date()` by Luna L. Sánchez Reyes, <https://orcid.org/0000-0001-7668-2528>

GENERAL INFORMATION

1. Title of Dataset

    Data from: DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life.

2. Author Information

	Corresponding Researcher
		name: Luna L. Sanchez Reyes
    institution: University of California, Merced, USA
    email: sanchez.reyes.luna@gmail.com

	Co-researcher 1
		name: Emily Jane McTavish
		institution: University of California, Merced, USA

	Co-researcher 2
		name: Brian C. O'Meara
		institution: University of Tennessee, Knoxville, USA

3. Date of data collection: 2022-01-28

4. Geographic location of data collection: Online

5. Funding sources that supported the collection of the data: National Science Foundation, USA

6. Recommended citation for this dataset:

DATA & FILE OVERVIEW

1. Description of dataset

These data were generated to investigate and showcase the performance of the datelife R package (<https://github.com/phylotastic/datelife>).
We showcased the application of the package with one mock example and two different biological examples.
The mock example was GENERATED WITH
The first biological example uses datelife on a small sample of bird species.
The second one uses datelife on bird species belonging to the family Fringillidae of "true finches", following the NCBI taxonomy.
We investigate the performance of the package datelife with two analysis: a benchmarking analysis to measure computing time of functions, and a cross validation analysis to test the accuracy and precision of the functions.

2. File List:

	File 1 Name: Sanchez-Reyes_etal_2022_table_1.csv
	File 1 Description:

	File 2 Name: Sanchez-Reyes_etal_2022_table_2.csv
	File 2 Description:

	File 3 Name: Sanchez-Reyes_etal_2022_figure_1_chronogram_mock_example.tre
	File 3 Description:

	File 4 Name: Sanchez-Reyes_etal_2022_figure_3_chronogram_small_example.tre
	File 4 Description:

	File 6 Name: Sanchez-Reyes_etal_2022_figure_4A_topology_finches_mrca.tre
	File 6: Description:

	File 7 Name: Sanchez-Reyes_etal_2022_figure_4B_topology_finches_ncbi.tre
	File 7 Description:

	File 5 Name: Sanchez-Reyes_etal_2022_figure_5_chronogram_finches_example.tre
	File 5 Description:

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S1.pdf
	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions.

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S2.pdf
	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Barker et al. 2013. Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S2.tre
	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S2 (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S3.pdf
	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX. Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S3.tre
	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S3 (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S4.pdf
	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX. Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S4.tre
	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S4 (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S5.pdf
	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX. Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S5.tre
	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S5 (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S6.pdf
	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX. Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S6.tre
	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S6 (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S7.pdf
	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX. Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S7.tre
	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S7 (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S8.pdf
	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX. Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S8.tre
	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S8 (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S9.pdf
	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX. Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S9.tre
	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S9 (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S10.pdf
	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX. Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S10.tre
	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S10 (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S11.pdf
	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX. Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S11.tre
	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S11 (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S12.pdf
	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX. Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S12.tre
	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S12 (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S13.pdf
	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX. Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S13.tre
	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S13 (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S14.pdf
	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX. Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S14.tre
	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S14 (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S15.pdf
	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX. Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S15.tre
	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S15 (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S16.pdf
	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX. Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S16.tre
	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S16 (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S.pdf
	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX. Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S.tre
	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S.pdf
	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX. Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S.tre
	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S.pdf
	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX. Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S.tre
	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S.pdf
	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX. Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S.tre
	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S (gray).

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_table_S1.csv
	File XXX Description:

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_table_S1.pdf
	File XXX Description:

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_table_S2.csv
	File XXX Description:

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_table_S2.pdf
	File XXX Description:

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_
	File XXX Description:

	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_
	File XXX Description:

3. Original names and locations of files at https://github.com/LunaSare/datelifeMS1:

```
cp tables/table-fringillidae-all.pdf dryad/Supplementary_Table_S1.pdf
cp tables/table-fringillidae-all-summary.pdf dryad/Supplementary_Table_S2.pdf
cp figures/figure-cross-validation/fig-cross-validation-xy-plots-diffs.pdf dryad/Supplementary_Figure_S1.pdf
cp figures/figure-cross-validation/cross_validation_1.jpg dryad/Supplementary_Figure_S2.jpg
cp figures/figure-cross-validation/cross_validation_2.jpg dryad/Supplementary_Figure_S3.jpg
cp figures/figure-cross-validation/cross_validation_3.jpg dryad/Supplementary_Figure_S4.jpg
cp figures/figure-cross-validation/cross_validation_4.jpg dryad/Supplementary_Figure_S5.jpg
cp figures/figure-cross-validation/cross_validation_5.jpg dryad/Supplementary_Figure_S6.jpg
cp figures/figure-cross-validation/cross_validation_6.jpg dryad/Supplementary_Figure_S7.jpg
cp figures/figure-cross-validation/cross_validation_7.jpg dryad/Supplementary_Figure_S8.jpg
cp figures/figure-cross-validation/cross_validation_8.jpg dryad/Supplementary_Figure_S9.jpg
cp figures/figure-cross-validation/cross_validation_9.jpg dryad/Supplementary_Figure_S10.jpg
cp figures/figure-cross-validation/cross_validation_10.jpg dryad/Supplementary_Figure_S11.jpg
cp figures/figure-cross-validation/cross_validation_11.jpg dryad/Supplementary_Figure_S12.jpg
cp figures/figure-cross-validation/cross_validation_12.jpg dryad/Supplementary_Figure_S13.jpg
cp figures/figure-cross-validation/cross_validation_13.jpg dryad/Supplementary_Figure_S14.jpg
cp figures/figure-cross-validation/cross_validation_14.jpg dryad/Supplementary_Figure_S15.jpg
cp figures/figure-cross-validation/cross_validation_15.jpg dryad/Supplementary_Figure_S16.jpg
cp figures/figure-cross-validation/cross_validation_16.jpg dryad/Supplementary_Figure_S17.jpg
cp figures/figure-cross-validation/cross_validation_17.jpg dryad/Supplementary_Figure_S18.jpg
cp figures/figure-cross-validation/cross_validation_18.jpg dryad/Supplementary_Figure_S19.jpg
cp figures/figure-cross-validation/cross_validation_19.jpg dryad/Supplementary_Figure_S20.jpg
```

METHODOLOGICAL INFORMATION

Bacterial detection was performed on the ViiA7 Real-time PCR System (Thermo Fisher Scientific) using PrimeTime® qPCR primers, probes and mastermix (IDT) according to the manufacturer’s instructions. Reactions were performed using 1X PrimeTime® Gene Expression Master Mix, 1X PrimeTime®qPCR Assay and up to 10ng of DNA. Cycling conditions were 95°C for 3 minutes, 60 cycles of 95°C for 5 seconds and 60°C for 30 seconds. Amplification results were reviewed using QuantStudioTM Real-Time PCR Software version 1.1 (Thermo Fisher Scientific). Amplification of beta-actin and prostaglandin transporter (PGT) was used to determine relative abundance.

Diversity profiling was performed by AGRF (Australian Genome Research Facility, Melbourne Australia). Samples were amplified with universal primers to the V1-V3 region of the bacterial 16S gene (forward AGAGTTTGATCMTGGCTCAG; reverse GWATTACCGCGGCKGCTG). Amplicons were indexed using the Nextera XT Index Kit (Illumina, San Diego, CA, USA) followed by Paired End sequencing on a MiSeq next generation sequencer (Illumina). Paired-end reads were assembled by aligning the forward and reverse reads using PEAR1 (version 0.9.5). Primers were identified and trimmed. Trimmed sequences were processed using Quantitative Insights into Microbial Ecology (QIIME 1.8) USEARCH (version 8.0.1623) and UPARSE software. Sequences were quality filtered and sorted by abundance after removal of full-length duplicate sequences. Singletons or unique reads were discarded. Sequences were clustered and then chimera filtered using “rdp_gold” database as reference. Reads were mapped back to Operational Taxonomic Units with a minimum identity of 97% and taxonomy was assigned using the QIIME 1 default classifier, pre-trained against Greengenes database5 (Version 13_8, Aug 2013).

DATA-SPECIFIC INFORMATION FOR: Rye_2021_a_Cohort_IDs.xlsx

1. Number of variables: 4

2. Number of cases/rows: 73

3. Variable List:
	Screening_cohort: Patient included in screening cohort; yes/no
	Screening_cohort_ID: Patient ID number for screening cohort
	Site_inv_cohort: Patient included in site investigation cohort; yes/no
	Site_inv_cohort_ID: Patient ID number for site investigation cohort

4. Missing data codes:
	None

5. Abbreviations used:
	N/A; not applicable

6. Other relevant information:
	20 of 21 patients in the screening cohort whose tumour tested positive for F. nucleatum, B. fragilis or both species were included in the site investigation cohort, along with 31 additional patients. Tumour material for one screening cohort patient whose tumour was positive for both species (patient 24) was not available for further DNA extraction.

DATA-SPECIFIC INFORMATION FOR: Rye_2021_b_Screening_qPCR_data.xlsx

1. Number of variables: 9

2. Number of cases/rows: 42

3. Variable List:
	Screening_cohort_ID: Screening cohort patient ID number
	B-actin_mean_Ct: Mean Ct for beta-actin; no. cycles
	PGT_mean_Ct: Mean Ct for prostaglandin transporter; no. cycles
	F_nucleatum_mean_Ct: Mean Ct for Fusobaterium nucleatum; no. cycles
	B_fragilis_gyrase_mean_Ct: Mean Ct for Bacteroides fragilis gyrase; no. cycles
	B_fragilis_bft_mean_Ct: Mean Ct for Bacteroides fragilis toxin; no. cycles
	B_breve_mean_Ct: Mean Ct for Bifidobacterium breve; no. cycles
	C_showae_mean_Ct: Mean Ct for Campylobacter showae; no. cycles
	L_buccalis_mean_Ct: Mean Ct for Leptotrichia buccalis ; no. cycles

4. Missing data codes:
	No Amp; No amplification

5. Abbreviations used:
	Amp; amplification, BFT; Bacteroides fragilis toxin, PGT; prostaglandin transporter

6. Other relevant information:
	Reactions were performed in duplicate. Results were reported where one or both samples amplified. Values represent the mean Ct where both samples amplified or the individual Ct for single amplifications.

DATA-SPECIFIC INFORMATION FOR: Rye_2021_c_Site_inv_species_by_site.xlsx

1. Number of variables: 9 + comments field

2. Number of cases/rows: 436

3. Variable List:
	Site_inv_cohort_ID: Site investigation cohort patient ID number
	Tumour_ID: Tumour ID number for patients with synchronous tumours (n = 5); 1/2
	Site: Area of tissue targeted; Normal (proximal), Normal (distal), Normal (adjacent), Tum luminal surface, Tumour, Tumour (mutinous), Invading margin, Inflammation, Stroma, Lymph node
	Site_code: Numerical code assigned to site; 1-11
	Site_code_2: Second site if same area used to represent more than one site; 1-11
	B_actin_mean_Ct: Mean Ct for beta-actin; no. cycles
	PGT_mean_Ct: Mean Ct for prostaglandin transporter; no. cycles
	F_nucleatum_mean_Ct: Mean Ct for Fusobaterium nucleatum; no. cycles
	B_fragilis_mean_Ct: Mean Ct for Bacteroides fragilis gyrase; no. cycles
	Comments: Used to flag values excluded from analysis

4. Missing data codes:
	No Amp; No amplification
	Single Amp; Single amplification
	N/A: not applicable

5. Abbreviations used:
 	Amp; amplification, PGT; prostaglandin transporter, Tum; tumour

6. Other relevant information:
	Target sites included: normal tissue from the proximal and distal surgical margins of the resection specimen, normal tissue adjacent to the tumour region, the tumour luminal surface, central tumour, mucinous tumour (where applicable), invading margin, sites of inflammation, stroma, lymph nodes containing tumour deposits and metastatic sites where available. Metastatic site data have been excluded from this dataset to minimise any risk of deductive disclosure of patient identities.
	B-actin and PGT reactions were performed in duplicate and all samples amplified.
	F. nucleatum and B. fragilis reactions were performed in triplicate and results reported where two or more samples amplified.  Mean cT values were excluded from analysis if the SD of the replicate Ct values > 5 (highlighted red).

DATA-SPECIFIC INFORMATION FOR: Rye_2021_d_Site_inv_16S_qPCR_data.xlsx

1. Number of variables: 4

2. Number of cases/rows: 51

3. Variable List:
	Site_inv cohort_ ID: Site investigation cohort patient ID number
	PGT_mean_Ct: Mean Ct for prostaglandin transporter; no. cycles
	16S_TFS_mean_Ct: Mean Ct for 16S rRNA using primer set 1
	16S_IDT_mean_Ct: Mean Ct for 16S rRNA using primer set 2

4. Missing data codes:
	None

5. Abbreviations used:
	IDT; Integrated DNA Technologies, PGT; prostaglandin transporter, TFS; Thermo Fisher Scientific

6. Other relevant information:
	Primer set 1: Obtained from Integrated DNA Technologies; Sequences as published in Nadkarni et al. Determination of bacterial load by real-time PCR using a broad-range (universal) probe and primers set. Microbiology. 2002;148(Pt 1):257-66. doi: 10.1099/00221287-148-1-257. PubMed PMID: 11782518.
	Primer set 2: Obtained from Thermo Fisher Scientific; Assay ID Ba04230899_s1.
	Performed on DNA samples extracted from the tumour luminal surface.

DATA-SPECIFIC INFORMATION FOR: Rye_2021_e_Seq_data_absolute_abundance.xlsx

1. Number of variables: 7

2. Number of cases/rows: 308

3. Variable List:
	OTU_ID: OTU ID number
	Pt_32: Absolute abundance of each OTU for patient 32; no. sequence reads
	Pt_26: Absolute abundance of each OTU for patient 26; no. sequence reads
	Pt_35: Absolute abundance of each OTU for patient 35; no. sequence reads
	Pt_28: Absolute abundance of each OTU for patient 28; no. sequence reads
	Pt_29: Absolute abundance of each OTU for patient 29; no. sequence reads
	Consensus_Lineage: Consensus lineage

4. Missing data codes:
	None

5. Abbreviations used:
	OTU; Operational Taxonomic Unit, Pt; patient

6. Other relevant information:
	Performed on DNA samples extracted from the tumour luminal surface.
	Taxonomy assigned using the QIIME 1 default classifier, pre-trained against Greengenes database5 (Version 13_8, Aug 2013).

DATA-SPECIFIC INFORMATION FOR: Rye_2021_f_Seq_data_otu_table.xlsx

1. Number of variables: 6

2. Number of cases/rows:
	Sheet 1 'Phylum_level': 13
	Sheet 2 'Class_level': 21
	Sheet 3 'Order_level': 28
	Sheet 4 'Family_level': 49
	Sheet 5 'Genus_level': 85
	Sheet 6 'Species_level': 101

3. Variable List:
	Taxon: Assigned taxon
	Pt_32: Relative abundance of each taxon for patient 32; proportion of total sequence reads
	Pt_26: Relative abundance of each taxon for patient 26; proportion of total sequence reads
	Pt_35: Relative abundance of each taxon for patient 35; proportion of total sequence reads
	Pt_28: Relative abundance of each taxon for patient 28; proportion of total sequence reads
	Pt_29: Relative abundance of each taxon for patient 29; proportion of total sequence reads

4. Missing data codes:
	None

5. Abbreviations used:
	Pt; patient

6. Other relevant information:
	Performed on DNA samples extracted from the tumour luminal surface.
	Taxonomy assigned using the QIIME 1 default classifier, pre-trained against Greengenes database5 (Version 13_8, Aug 2013).

DATA-SPECIFIC INFORMATION FOR: Rye_2021_g_Seq_data_mg_blast.xlsx

1. Number of variables: 16

2. Number of cases/rows: 366

3. Variable List:
	subject: OTU ID
	gi: Nucleotide BLAST accession number
	evalue: Expectation Value
	bit-score: Bit score
	score: Raw alignment score
	alignment-length: Alignment length; number of bases
	%identity: Percentage of bases identical to OTU group reference sequence
	identical: Number of bases identical to OTU group reference sequence
	positives: Number of bases with a positive match to OTU group reference sequence
	%positives: Percentage of bases with a positive match to OTU group reference sequence
	scientific-names: Scientific name of OTU classification group
	common-names: Common name of OTU classification group
	subject-blast-names: Grouping name
	subject-super-kingdoms: Kingdom names
	subject-title: Descriptive title for OTU group
	OTU-seq: OTU reference Sequence

4. Missing data codes:
	None

5. Abbreviations used:
	OTU; Operational Taxonomic Unit

6. Other relevant information:
	Taxonomy assigned using the QIIME 1 default classifier, pre-trained against Greengenes database5 (Version 13_8, Aug 2013).

### 2022-07-06 dryad/README.Rmd
adding citations to dryad readme

@@
 ---
-title: 'README for Dryad Data Package from study "DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life"'
+title: 'Dryad Data Package for study "DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life"'
 output:
@@
 
-	Corresponding Researcher
-		name: Luna L. Sanchez Reyes
-    institution: University of California, Merced, USA
-    email: sanchez.reyes.luna@gmail.com
+	Corresponding Researcher name: Luna L. Sanchez Reyes
+  Corresponding Researcher institution: University of California, Merced, USA
+  Corresponding Researcher email: sanchez.reyes.luna@gmail.com
 
-	Co-researcher 1
-		name: Emily Jane McTavish
-		institution: University of California, Merced, USA
+	Co-researcher 1 name: Emily Jane McTavish
+	Co-researcher 1	institution: University of California, Merced, USA
 
-	Co-researcher 2
-		name: Brian C. O'Meara
-		institution: University of Tennessee, Knoxville, USA
+	Co-researcher 2 name: Brian C. O'Meara
+	Co-researcher 2	institution: University of Tennessee, Knoxville, USA
 
@@
 
+    Sánchez-Reyes LL, McTavish EJ & O’Meara BC. (2019). DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life. BioRxiv, 782094.
+
+    Data from: DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life. Dryad, Dataset
+
 DATA & FILE OVERVIEW
@@
 We showcased the application of the package with one mock example and two different biological examples.
-The mock example was GENERATED WITH
+
 The first biological example uses datelife on a small sample of bird species.
 The second one uses datelife on bird species belonging to the family Fringillidae of "true finches", following the NCBI taxonomy.
-We investigate the performance of the package datelife with two analysis: a benchmarking analysis to measure computing time of functions, and a cross validation analysis to test the accuracy and precision of the functions.
+The mock example was created by replacing species names from the small example with letters, and reducing the number of source chronograms.
+
+We evaluated the performance of the package datelife with two analysis: a benchmarking analysis to measure computing time of functions, and a cross validation analysis to test the accuracy and precision of the functions.
 

### 2022-07-06 dryad/README.Rmd
fixing indentation; renaming jpg files

@@
 
-	Corresponding Researcher name: Luna L. Sanchez Reyes
-  Corresponding Researcher institution: University of California, Merced, USA
-  Corresponding Researcher email: sanchez.reyes.luna@gmail.com
+	Corresponding Researcher   
+	name: Luna L. Sanchez Reyes    
+  institution: University of California, Merced, USA    
+  email: sanchez.reyes.luna@gmail.com
 
-	Co-researcher 1 name: Emily Jane McTavish
-	Co-researcher 1	institution: University of California, Merced, USA
+	Co-researcher 1    
+	name: Emily Jane McTavish    
+	institution: University of California, Merced, USA
 
-	Co-researcher 2 name: Brian C. O'Meara
-	Co-researcher 2	institution: University of Tennessee, Knoxville, USA
+	Co-researcher 2    
+	name: Brian C. O'Meara    
+	institution: University of Tennessee, Knoxville, USA
 
@@
 cp figures/figure-cross-validation/fig-cross-validation-xy-plots-diffs.pdf dryad/Supplementary_Figure_S1.pdf
-cp figures/figure-cross-validation/cross_validation_1.jpg dryad/Supplementary_Figure_S2.jpg
-cp figures/figure-cross-validation/cross_validation_2.jpg dryad/Supplementary_Figure_S3.jpg
-cp figures/figure-cross-validation/cross_validation_3.jpg dryad/Supplementary_Figure_S4.jpg
-cp figures/figure-cross-validation/cross_validation_4.jpg dryad/Supplementary_Figure_S5.jpg
-cp figures/figure-cross-validation/cross_validation_5.jpg dryad/Supplementary_Figure_S6.jpg
-cp figures/figure-cross-validation/cross_validation_6.jpg dryad/Supplementary_Figure_S7.jpg
-cp figures/figure-cross-validation/cross_validation_7.jpg dryad/Supplementary_Figure_S8.jpg
-cp figures/figure-cross-validation/cross_validation_8.jpg dryad/Supplementary_Figure_S9.jpg
-cp figures/figure-cross-validation/cross_validation_9.jpg dryad/Supplementary_Figure_S10.jpg
-cp figures/figure-cross-validation/cross_validation_10.jpg dryad/Supplementary_Figure_S11.jpg
-cp figures/figure-cross-validation/cross_validation_11.jpg dryad/Supplementary_Figure_S12.jpg
-cp figures/figure-cross-validation/cross_validation_12.jpg dryad/Supplementary_Figure_S13.jpg
-cp figures/figure-cross-validation/cross_validation_13.jpg dryad/Supplementary_Figure_S14.jpg
-cp figures/figure-cross-validation/cross_validation_14.jpg dryad/Supplementary_Figure_S15.jpg
-cp figures/figure-cross-validation/cross_validation_15.jpg dryad/Supplementary_Figure_S16.jpg
-cp figures/figure-cross-validation/cross_validation_16.jpg dryad/Supplementary_Figure_S17.jpg
-cp figures/figure-cross-validation/cross_validation_17.jpg dryad/Supplementary_Figure_S18.jpg
-cp figures/figure-cross-validation/cross_validation_18.jpg dryad/Supplementary_Figure_S19.jpg
-cp figures/figure-cross-validation/cross_validation_19.jpg dryad/Supplementary_Figure_S20.jpg
 ```

### 2022-07-06 manuscript/manuscript.Rmd
fixing indentation; renaming jpg files

@@
 It also represent valuable information for education, science communication, and policy decisions.
-  When chronograms are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on evolutionary timeframe is exposed in a programatic and reusable way, as intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
+  When chronograms are shared in public and open databases, this wealth of expertly-curated and peer-reviewed data on evolutionary timeframe is exposed in a programatic and reusable way, as intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
   Here we present DateLife, a service implemented as an R package and an R Shiny website application available at www.datelife.org, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time frame of evolution.
@@
   We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
-  DateLife will be useful to increase awereness of the existing variation in alternative hypothesis of evolutionary time for the same organisms, and can foster exploration of the effect of alternative evolutionary timing hypotheses on the results of downstream analyses, providing a framework for a more informed interpretation of evolutionary results.
+  DateLife will be useful to increase awareness of the existing variation in alternative hypothesis of evolutionary time for the same organisms, and can foster exploration of the effect of alternative evolutionary timing hypotheses on the results of downstream analyses, providing a framework for a more informed interpretation of evolutionary results.
 

### 2022-07-06 manuscript/manuscript.tex
fixing indentation; renaming jpg files

@@
 Knowledge of the evolutionary time frame of organisms is key for research in the natural sciences. It also represent valuable information for education, science communication, and policy decisions.
-When chronograms are shared in public, open databases this wealth of expertly-curated and peer-reviewed data on evolutionary timeframe is exposed in a programatic and reusable way, as intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
+When chronograms are shared in public and open databases, this wealth of expertly-curated and peer-reviewed data on evolutionary timeframe is exposed in a programatic and reusable way, as intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
 Here we present DateLife, a service implemented as an R package and an R Shiny website application available at www.datelife.org, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time frame of evolution.
@@
 We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
-DateLife will be useful to increase awereness of the existing variation in alternative hypothesis of evolutionary time for the same organisms, and can foster exploration of the effect of alternative evolutionary timing hypotheses on the results of downstream analyses, providing a framework for a more informed interpretation of evolutionary results.
+DateLife will be useful to increase awareness of the existing variation in alternative hypothesis of evolutionary time for the same organisms, and can foster exploration of the effect of alternative evolutionary timing hypotheses on the results of downstream analyses, providing a framework for a more informed interpretation of evolutionary results.
 }

### 2022-07-07 dryad/README.Rmd
adding dryad files

@@
 	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S3.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX.
+	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Barker et al.
+2015, chronogram 1.
 Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
@@
 	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S4.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX.
+	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Barker et al.
+2015, chronogram 2.
 Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
@@
 	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S5.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX.
+	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Burns et al.
+2015.
 Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
@@
 	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S6.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX.
+	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Claramunt et al.
+2015.
 Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
@@
 	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S7.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX.
+	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Gibb et al.
+2015.
 Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
@@
 	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S8.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX.
+	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Hedges et al.
+2015, chronogram 1.
 Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
@@
 	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S9.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX.
+	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Hedges et al.
+2015, chronogram 2.
 Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
@@
 	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S10.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX.
+	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Hooper et al.
+2017.
 Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
@@
 	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S11.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX.
+	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Jetz et al.
+2012, chronogram 1.
 Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
@@
 	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S12.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX.
+	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Jetz et al.
+2012, chronogram 2.
 Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
@@
 	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S13.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX.
+	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Kimball et al.
+2019, chronogram 1.
 Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
@@
 	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S14.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX.
+	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Kimball et al.
+2019, chronogram 2.
 Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
@@
 	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S15.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX.
+	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Oliveros et al.
+2019.
 Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
@@
 	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S16.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX.
+	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Price et al.
+2014, chronogram 1.
 Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
@@
 
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX.
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S17.pdf    
+	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Price et al.
+2014, chronogram 2.
 Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
 
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S.tre    
-	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S (gray).
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S17.tre    
+	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S17 (gray).
 
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX.
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S18.pdf    
+	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Roquet et al.
+2014, chronogram 1.
 Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
 
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S.tre    
-	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S (gray).
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S18.tre    
+	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S18 (gray).
 
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX.
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S19.pdf    
+	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Roquet et al.
+2014, chronogram 2.
 Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
 
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S.tre    
-	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S (gray).
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S19.tre    
+	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S19 (gray).
 
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from XXX.
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S20.pdf    
+	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Uyeda et al.
+2017.
 Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
 
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S.tre    
-	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S (gray).
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S20.tre    
+	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S20 (gray).
 

### 2022-07-07 dryad/README.Rmd
adding to readme for dryad submission

@@
 	File 1 Name: Sanchez-Reyes_etal_2022_table_1.csv    
-	File 1 Description:
+	File 1 Description: CSV (comma separated values) file containing results from the small example.
+Names and ages (in Million years) of nodes from source chronograms that have been congruified to nodes of the chosen tree topology.
 
 	File 2 Name: Sanchez-Reyes_etal_2022_table_2.csv    
-	File 2 Description:
+	File 2 Description: CSV (comma separated values) file containing a summary of node age (in Million years) data from table S1.
 
 	File 3 Name: Sanchez-Reyes_etal_2022_figure_1_chronogram_mock_example.tre    
-	File 3 Description:
+	File 3 Description: Newick version of chronogram from the mock example shown in Figure 1 of the manuscript (https://www.biorxiv.org/content/10.1101/782094v2).
 
 	File 4 Name: Sanchez-Reyes_etal_2022_figure_3_chronogram_small_example.tre    
-	File 4 Description:
+	File 4 Description: Newick version of chronogram from the small example shown in Figure 3 of the manuscript (https://www.biorxiv.org/content/10.1101/782094v2).
 
 	File 6 Name: Sanchez-Reyes_etal_2022_figure_4A_topology_finches_mrca.tre    
-	File 6: Description:
+	File 6: Description: Newick version of topology extracted from the Open Tree of Life synthetic tree, shown in Figure 4A of the manuscript (https://www.biorxiv.org/content/10.1101/782094v2).
 
 	File 7 Name: Sanchez-Reyes_etal_2022_figure_4B_topology_finches_ncbi.tre    
-	File 7 Description:
+	File 7 Description: Newick version of topology extracted from the Open Tree of Life synthetic tree, shown in Figure 4B of the manuscript (https://www.biorxiv.org/content/10.1101/782094v2).
 
 	File 5 Name: Sanchez-Reyes_etal_2022_figure_5_chronogram_finches_example.tre    
-	File 5 Description:
+	File 5 Description: Newick version of chronogram from the finch example shown in Figure 5 of the manuscript (https://www.biorxiv.org/content/10.1101/782094v2).
 
 	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S1.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions.
+	File XXX Description: PDf file showing results of cross validation analysis of datelife's chronogram generating functions.
 
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S2.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Barker et al.
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S2.jpg    
+	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Barker et al.
 2013.
@@
 
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S3.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Barker et al.
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S3.jpg    
+	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Barker et al.
 2015, chronogram 1.
@@
 
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S4.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Barker et al.
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S4.jpg    
+	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Barker et al.
 2015, chronogram 2.
@@
 
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S5.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Burns et al.
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S5.jpg    
+	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Burns et al.
 2015.
@@
 
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S6.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Claramunt et al.
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S6.jpg    
+	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Claramunt et al.
 2015.
@@
 
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S7.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Gibb et al.
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S7.jpg    
+	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Gibb et al.
 2015.
@@
 
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S8.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Hedges et al.
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S8.jpg    
+	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Hedges et al.
 2015, chronogram 1.
@@
 
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S9.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Hedges et al.
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S9.jpg    
+	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Hedges et al.
 2015, chronogram 2.
@@
 
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S10.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Hooper et al.
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S10.jpg    
+	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Hooper et al.
 2017.
@@
 
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S11.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Jetz et al.
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S11.jpg    
+	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Jetz et al.
 2012, chronogram 1.
@@
 
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S12.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Jetz et al.
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S12.jpg    
+	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Jetz et al.
 2012, chronogram 2.
@@
 
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S13.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Kimball et al.
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S13.jpg    
+	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Kimball et al.
 2019, chronogram 1.
@@
 
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S14.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Kimball et al.
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S14.jpg    
+	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Kimball et al.
 2019, chronogram 2.
@@
 
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S15.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Oliveros et al.
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S15.jpg    
+	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Oliveros et al.
 2019.
@@
 
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S16.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Price et al.
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S16.jpg    
+	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Price et al.
 2014, chronogram 1.
@@
 
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S17.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Price et al.
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S17.jpg    
+	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Price et al.
 2014, chronogram 2.
@@
 
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S18.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Roquet et al.
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S18.jpg    
+	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Roquet et al.
 2014, chronogram 1.
@@
 
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S19.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Roquet et al.
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S19.jpg    
+	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Roquet et al.
 2014, chronogram 2.
@@
 
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S20.pdf    
-	File XXX Description: Results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Uyeda et al.
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S20.jpg    
+	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Uyeda et al.
 2017.
@@
 	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_table_S1.csv    
-	File XXX Description:
+	File XXX Description: CSV (comma separated values) file
 
 	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_table_S1.pdf    
-	File XXX Description:
+	File XXX Description: PDF file of
 
 	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_table_S2.csv    
-	File XXX Description:
+	File XXX Description: CSV (comma separated values) file
 
 	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_table_S2.pdf    
-	File XXX Description:
+	File XXX Description: PDF file of r
 
@@
 
+```{r}
+files <- c()
+text <- c()
+for (file in 1:10) {
+
+  text <- paste(text, 
+                paste("DATA-SPECIFIC INFORMATION FOR:", file, "\\newline"))
+}
+text
+```
+
+`r text`
 DATA-SPECIFIC INFORMATION FOR: Rye_2021_a_Cohort_IDs.xlsx

### 2022-07-07 manuscript/manuscript.Rmd
adding to readme for dryad submission

@@
 We illustrate the DateLife workflow using a family within the Passeriform birds encompassing the true finches, Fringillidae, as case study.
-The first example analyses 6 bird species and shows all steps of the workflow The second example is an analysis of 289 species in the family Fringillidae that are included in the NCBI taxonomy.
+On a small example, we analyse 6 bird species, and show all steps of the workflow (Fig.
+\ref{fig:figure2}).
+As a second example, we analyse 289 species in the family Fringillidae that are included in the NCBI taxonomy (Fig.
+\ref{fig:fringillidages}).
 <!--We used `r cite_r("r-references.bib")` for all our analyses.-->

### 2022-07-08 dryad/README.Rmd
adding newick files of source chronograms

@@
 
+```{r, echo=FALSE}
+file.copy(from = "../tables/table-fringillidae-small-example.csv",
+          to = "../dryad/Sanchez-Reyes_etal_2022_table_1.csv")
+```
+
   File 1 Name: Sanchez-Reyes_etal_2022_table_1.csv    
@@
 	File 2 Name: Sanchez-Reyes_etal_2022_table_2.csv    
-	File 2 Description: CSV (comma separated values) file containing a summary of node age (in Million years) data from table S1.
+	File 2 Description: CSV (comma separated values) file containing a summary of node age (in Million years) data from table 1.
 
@@
 
+```{r, echo=FALSE}
+file.copy(from = "../tables/table-fringillidae-all-congruified.csv",
+          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S1.csv")
+
+```
+
 	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_table_S1.csv    
-	File XXX Description: CSV (comma separated values) file
+	File XXX Description: CSV (comma separated values) file containing results from the Fringillidae example.
+Names and ages (in Million years) of nodes from source chronograms that have been congruified to nodes of the chosen tree topology shown in Figure 4B of the manuscript.
+
+```{r, echo=FALSE}
+file.copy(from = "../tables/table-fringillidae-all-congruified.csv",
+          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S1.csv")
+
+```
 
 	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_table_S1.pdf    
-	File XXX Description: PDF file of
+	File XXX Description: PDF version of file Sanchez-Reyes_etal_2022_supplementary_table_S1.csv
+
+```{r, echo=FALSE}
+file.copy(from = ".../tables/table-fringillidae-all-summary.csv",
+          to = "../dryad/anchez-Reyes_etal_2022_supplementary_table_S2.csv")
+
+```
 
   File XXX Name: Sanchez-Reyes_etal_2022_supplementary_table_S2.csv    
-	File XXX Description: CSV (comma separated values) file
+	File XXX Description: CSV (comma separated values) file containing a summary of node age data (in Million years) from table S1.
+
+```{r, echo=FALSE}
+file.copy(from = ".../tables/table-fringillidae-all-summary.pdf",
+          to = "../dryad/anchez-Reyes_etal_2022_supplementary_table_S2.pdf")
+
+```
 

### 2022-07-09 dryad/README.Rmd
adding description of csv files

@@
     keep_md: yes
+    pandoc_args: --listings
+    includes:
+      in_header: preamble.tex
 ---
@@
   collapse = TRUE,
-  comment = "#>"
+  comment = "#>",
+  results = 'hide'
 )
@@
 
-```{r, echo=FALSE}
+```{r, echo = FALSE}
+
+csv_file_names <- c("Sanchez-Reyes_etal_2022_table_1.csv",
+                    "Sanchez-Reyes_etal_2022_table_2.csv",
+                    "Sanchez-Reyes_etal_2022_supplementary_table_S1.csv",
+                    "Sanchez-Reyes_etal_2022_supplementary_table_S2.csv")
+tre_file_names <- c("Sanchez-Reyes_etal_2022_figure_1_chronogram_mock_example.tre")
+```
+
+```{r, echo=FALSE, message=FALSE}
 file.copy(from = "../tables/table-fringillidae-small-example.csv",
@@
 
+```{r, echo=FALSE, message=FALSE}
+file.copy(from = "../tables/table-fringillidae-small-example-summary.csv",
+          to = "../dryad/Sanchez-Reyes_etal_2022_table_2.csv")
+```
+
 	File 2 Name: Sanchez-Reyes_etal_2022_table_2.csv    
 	File 2 Description: CSV (comma separated values) file containing a summary of node age (in Million years) data from table 1.
+
+```{r, echo=FALSE, message=FALSE}
+file.copy(from = "../tables/table-fringillidae-all-congruified.csv",
+          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S1.csv")
+
+```
+
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_table_S1.csv    
+	File XXX Description: CSV (comma separated values) file containing results from the Fringillidae example.
+Names and ages (in Million years) of nodes from source chronograms that have been congruified to nodes of the chosen tree topology shown in Figure 4B of the manuscript.
+
+```{r, echo=FALSE, eval=FALSE, message=FALSE}
+file.copy(from = "../tables/table-fringillidae-all-congruified.pdf",
+          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S1.pdf")
+
+```
+
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_table_S1.pdf    
+	File XXX Description: PDF version of file Sanchez-Reyes_etal_2022_supplementary_table_S1.csv
+
+```{r, echo=FALSE, message=FALSE, }
+file.copy(from = "../tables/table-fringillidae-all-summary.csv",
+          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S2.csv")
+
+```
+
+  File XXX Name: Sanchez-Reyes_etal_2022_supplementary_table_S2.csv    
+	File XXX Description: CSV (comma separated values) file containing a summary of node age data (in Million years) from table S1.
+
+```{r, echo=FALSE, message=FALSE}
+file.copy(from = ".../tables/table-fringillidae-all-summary.pdf",
+          to = "../dryad/anchez-Reyes_etal_2022_supplementary_table_S2.pdf")
+
+```
+
+  File XXX Name: Sanchez-Reyes_etal_2022_supplementary_table_S2.pdf    
+	File XXX Description: PDF file version of file Sanchez-Reyes_etal_2022_supplementary_table_S2.csv.
 
@@
 
-```{r, echo=FALSE}
-file.copy(from = "../tables/table-fringillidae-all-congruified.csv",
-          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S1.csv")
-
-```
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_table_S1.csv    
-	File XXX Description: CSV (comma separated values) file containing results from the Fringillidae example.
-Names and ages (in Million years) of nodes from source chronograms that have been congruified to nodes of the chosen tree topology shown in Figure 4B of the manuscript.
-
-```{r, echo=FALSE}
-file.copy(from = "../tables/table-fringillidae-all-congruified.csv",
-          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S1.csv")
-
-```
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_table_S1.pdf    
-	File XXX Description: PDF version of file Sanchez-Reyes_etal_2022_supplementary_table_S1.csv
-
-```{r, echo=FALSE}
-file.copy(from = ".../tables/table-fringillidae-all-summary.csv",
-          to = "../dryad/anchez-Reyes_etal_2022_supplementary_table_S2.csv")
-
-```
-
-  File XXX Name: Sanchez-Reyes_etal_2022_supplementary_table_S2.csv    
-	File XXX Description: CSV (comma separated values) file containing a summary of node age data (in Million years) from table S1.
-
-```{r, echo=FALSE}
-file.copy(from = ".../tables/table-fringillidae-all-summary.pdf",
-          to = "../dryad/anchez-Reyes_etal_2022_supplementary_table_S2.pdf")
-
-```
-
-  File XXX Name: Sanchez-Reyes_etal_2022_supplementary_table_S2.pdf    
-	File XXX Description: PDF file of r
-
 	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_    
@@
 
-```{r}
-files <- c()
+```{r, echo = FALSE}
 text <- c()
-for (file in 1:10) {
-
+for (file_name in csv_file_names) {
+  # read the csv file in:
+  tt <- utils::read.csv(file = paste0("../dryad/", file_name), header = TRUE, row.names = NULL)
+  # remove the column with row numbers:
+  tt <- tt[,-1]
+  # create the description text
+  var_list <- paste("\\tab", gsub("[.]", " ", colnames(tt)))
+  var_list <- paste(var_list, collapse = ": \\newline")
   text <- paste(text, 
-                paste("DATA-SPECIFIC INFORMATION FOR:", file, "\\newline"))
+                paste("DATA-SPECIFIC INFORMATION FOR:", file_name, "\\newline"),
+                "1. Number of variables: ",
+                ncol(tt), "\\newline",
+                "2. Number of cases/rows:",
+                nrow(tt), "\\newline",
+                "3. Variable List: \\newline", var_list, "\\newline",
+                "4. Missing data codes: None \\newline
+                5. Abbreviations used: NA = not applicable \\newline \\newline")
 }
-text
 ```

### 2022-07-09 dryad/README.Rmd
formatting description of csv files

@@
   comment = "#>",
+  echo = FALSE,
   results = 'hide'
@@
 
-```{r, echo = FALSE}
+```{r}
+
+var_description_1 <- c("Node names in tree topology, as defined by taxon A and taxon B",
+                       "Scientific name of the first element of a taxon pair that defines a node from a tree topology",
+                       "Scientific name of the second element of a taxon pair that defines a node from a tree topology",
+                       "Ages of nodes defined by taxon A and taxon B, in Million years",
+                       "Reference of the study that published the chronogram from which the node ages were extracted")
+
+var_description_2 <- c("Names of nodes from the tree topology used for the dating analysis",
+                       "Minimum node age, in Million years",
+                       "First quartile of node age distributions, in Million years",
+                       "Mean node age, in Million years",
+                       "Median node age, in Million years",
+                       "Third quartile of node age distributions, in Million years",
+                       "Maximum node age, in Million years",
+                       "Variance of node age, in Million years",
+                       "Standar deviation of node age, in Million years")
+```
+
+```{r}
 text <- c()
+var_descriptions <- list(var_description_1, var_description_2, var_description_1, var_description_2) 
+names(var_descriptions) <- csv_file_names
+
 for (file_name in csv_file_names) {
@@
   # create the description text
-  var_list <- paste("\\tab", gsub("[.]", " ", colnames(tt)))
-  var_list <- paste(var_list, collapse = ": \\newline")
+  var_list <- gsub("[.]", " ", colnames(tt))
+  var_list <- paste0("\\tab ", var_list, ": ", var_descriptions[[file_name]], "\\n ")
+  var_list <- paste0(var_list, collapse = "")
+  print(var_list)
   text <- paste(text, 
-                paste("DATA-SPECIFIC INFORMATION FOR:", file_name, "\\newline"),
-                "1. Number of variables: ",
-                ncol(tt), "\\newline",
-                "2. Number of cases/rows:",
-                nrow(tt), "\\newline",
-                "3. Variable List: \\newline", var_list, "\\newline",
-                "4. Missing data codes: None \\newline
-                5. Abbreviations used: NA = not applicable \\newline \\newline")
+                paste("DATA-SPECIFIC INFORMATION FOR:", file_name, "\\n "),
+                "\\n1. Number of variables/columns:", ncol(tt), "\\n ",
+                "\\n2. Number of cases/rows:", nrow(tt), "\\n ",
+                "\\n3. Variable List \\n ", var_list,
+                "\\n4. Missing data codes: None \\n ",
+                "\\n5. Abbreviations used: NA = not applicable \\n \\n ")
 }
@@
 `r text`
-DATA-SPECIFIC INFORMATION FOR: Rye_2021_a_Cohort_IDs.xlsx
-
-1. Number of variables: 4
-
-2. Number of cases/rows: 73
-
-3. Variable List:
-	Screening_cohort: Patient included in screening cohort; yes/no
-	Screening_cohort_ID: Patient ID number for screening cohort
-	Site_inv_cohort: Patient included in site investigation cohort; yes/no
-	Site_inv_cohort_ID: Patient ID number for site investigation cohort
-
-4. Missing data codes:
-	None
-
-5. Abbreviations used:
-	N/A; not applicable
-
-6. Other relevant information:
-	20 of 21 patients in the screening cohort whose tumour tested positive for F.
-nucleatum, B.
-fragilis or both species were included in the site investigation cohort, along with 31 additional patients.
-Tumour material for one screening cohort patient whose tumour was positive for both species (patient 24) was not available for further DNA extraction.
-
-DATA-SPECIFIC INFORMATION FOR: Rye_2021_b_Screening_qPCR_data.xlsx
-
-1. Number of variables: 9
-
-2. Number of cases/rows: 42
-
-3. Variable List:
-	Screening_cohort_ID: Screening cohort patient ID number
-	B-actin_mean_Ct: Mean Ct for beta-actin; no. cycles
-	PGT_mean_Ct: Mean Ct for prostaglandin transporter; no. cycles
-	F_nucleatum_mean_Ct: Mean Ct for Fusobaterium nucleatum; no. cycles
-	B_fragilis_gyrase_mean_Ct: Mean Ct for Bacteroides fragilis gyrase; no. cycles
-	B_fragilis_bft_mean_Ct: Mean Ct for Bacteroides fragilis toxin; no. cycles
-	B_breve_mean_Ct: Mean Ct for Bifidobacterium breve; no. cycles
-	C_showae_mean_Ct: Mean Ct for Campylobacter showae; no. cycles
-	L_buccalis_mean_Ct: Mean Ct for Leptotrichia buccalis ; no. cycles
-
-4. Missing data codes:
-	No Amp; No amplification
-
-5. Abbreviations used:
-	Amp; amplification, BFT; Bacteroides fragilis toxin, PGT; prostaglandin transporter
-
-6. Other relevant information:
-	Reactions were performed in duplicate.
-Results were reported where one or both samples amplified.
-Values represent the mean Ct where both samples amplified or the individual Ct for single amplifications.
-
-DATA-SPECIFIC INFORMATION FOR: Rye_2021_c_Site_inv_species_by_site.xlsx
-
-1. Number of variables: 9 + comments field
-
-2. Number of cases/rows: 436
-
-3. Variable List:
-	Site_inv_cohort_ID: Site investigation cohort patient ID number
-	Tumour_ID: Tumour ID number for patients with synchronous tumours (n = 5); 1/2
-	Site: Area of tissue targeted; Normal (proximal), Normal (distal), Normal (adjacent), Tum luminal surface, Tumour, Tumour (mutinous), Invading margin, Inflammation, Stroma, Lymph node
-	Site_code: Numerical code assigned to site; 1-11
-	Site_code_2: Second site if same area used to represent more than one site; 1-11
-	B_actin_mean_Ct: Mean Ct for beta-actin; no. cycles
-	PGT_mean_Ct: Mean Ct for prostaglandin transporter; no. cycles
-	F_nucleatum_mean_Ct: Mean Ct for Fusobaterium nucleatum; no. cycles
-	B_fragilis_mean_Ct: Mean Ct for Bacteroides fragilis gyrase; no. cycles
-	Comments: Used to flag values excluded from analysis
-
-4. Missing data codes:
-	No Amp; No amplification
-	Single Amp; Single amplification
-	N/A: not applicable
-
-5. Abbreviations used:
- 	Amp; amplification, PGT; prostaglandin transporter, Tum; tumour
-
-6. Other relevant information:
-	Target sites included: normal tissue from the proximal and distal surgical margins of the resection specimen, normal tissue adjacent to the tumour region, the tumour luminal surface, central tumour, mucinous tumour (where applicable), invading margin, sites of inflammation, stroma, lymph nodes containing tumour deposits and metastatic sites where available.
-Metastatic site data have been excluded from this dataset to minimise any risk of deductive disclosure of patient identities.
-	B-actin and PGT reactions were performed in duplicate and all samples amplified.
-	F.
-nucleatum and B.
-fragilis reactions were performed in triplicate and results reported where two or more samples amplified.
-Mean cT values were excluded from analysis if the SD of the replicate Ct values > 5 (highlighted red).
-
-DATA-SPECIFIC INFORMATION FOR: Rye_2021_d_Site_inv_16S_qPCR_data.xlsx
-
-1. Number of variables: 4
-
-2. Number of cases/rows: 51
-
-3. Variable List:
-	Site_inv cohort_ ID: Site investigation cohort patient ID number
-	PGT_mean_Ct: Mean Ct for prostaglandin transporter; no. cycles
-	16S_TFS_mean_Ct: Mean Ct for 16S rRNA using primer set 1
-	16S_IDT_mean_Ct: Mean Ct for 16S rRNA using primer set 2
-
-4. Missing data codes:
-	None
-
-5. Abbreviations used:
-	IDT; Integrated DNA Technologies, PGT; prostaglandin transporter, TFS; Thermo Fisher Scientific
-
-6. Other relevant information:
-	Primer set 1: Obtained from Integrated DNA Technologies; Sequences as published in Nadkarni et al.
-Determination of bacterial load by real-time PCR using a broad-range (universal) probe and primers set.
-Microbiology.
-2002;148(Pt 1):257-66.
-doi: 10.1099/00221287-148-1-257.
-PubMed PMID: 11782518.
-	Primer set 2: Obtained from Thermo Fisher Scientific; Assay ID Ba04230899_s1.
-	Performed on DNA samples extracted from the tumour luminal surface.
-
-DATA-SPECIFIC INFORMATION FOR: Rye_2021_e_Seq_data_absolute_abundance.xlsx
-
-1. Number of variables: 7
-
-2. Number of cases/rows: 308
-
-3. Variable List:
-	OTU_ID: OTU ID number
-	Pt_32: Absolute abundance of each OTU for patient 32; no. sequence reads
-	Pt_26: Absolute abundance of each OTU for patient 26; no. sequence reads
-	Pt_35: Absolute abundance of each OTU for patient 35; no. sequence reads
-	Pt_28: Absolute abundance of each OTU for patient 28; no. sequence reads
-	Pt_29: Absolute abundance of each OTU for patient 29; no. sequence reads
-	Consensus_Lineage: Consensus lineage
-
-4. Missing data codes:
-	None
-
-5. Abbreviations used:
-	OTU; Operational Taxonomic Unit, Pt; patient
-
-6. Other relevant information:
-	Performed on DNA samples extracted from the tumour luminal surface.
-	Taxonomy assigned using the QIIME 1 default classifier, pre-trained against Greengenes database5 (Version 13_8, Aug 2013).
-
-DATA-SPECIFIC INFORMATION FOR: Rye_2021_f_Seq_data_otu_table.xlsx
-
-1. Number of variables: 6
-
-2. Number of cases/rows:
-	Sheet 1 'Phylum_level': 13
-	Sheet 2 'Class_level': 21
-	Sheet 3 'Order_level': 28
-	Sheet 4 'Family_level': 49
-	Sheet 5 'Genus_level': 85
-	Sheet 6 'Species_level': 101
-
-3. Variable List:
-	Taxon: Assigned taxon
-	Pt_32: Relative abundance of each taxon for patient 32; proportion of total sequence reads
-	Pt_26: Relative abundance of each taxon for patient 26; proportion of total sequence reads
-	Pt_35: Relative abundance of each taxon for patient 35; proportion of total sequence reads
-	Pt_28: Relative abundance of each taxon for patient 28; proportion of total sequence reads
-	Pt_29: Relative abundance of each taxon for patient 29; proportion of total sequence reads
-
-4. Missing data codes:
-	None
-
-5. Abbreviations used:
-	Pt; patient
-
-6. Other relevant information:
-	Performed on DNA samples extracted from the tumour luminal surface.
-	Taxonomy assigned using the QIIME 1 default classifier, pre-trained against Greengenes database5 (Version 13_8, Aug 2013).
-
-DATA-SPECIFIC INFORMATION FOR: Rye_2021_g_Seq_data_mg_blast.xlsx
-
-1. Number of variables: 16
-
-2. Number of cases/rows: 366
-
-3. Variable List:
-	subject: OTU ID
-	gi: Nucleotide BLAST accession number
-	evalue: Expectation Value
-	bit-score: Bit score
-	score: Raw alignment score
-	alignment-length: Alignment length; number of bases
-	%identity: Percentage of bases identical to OTU group reference sequence
-	identical: Number of bases identical to OTU group reference sequence
-	positives: Number of bases with a positive match to OTU group reference sequence
-	%positives: Percentage of bases with a positive match to OTU group reference sequence
-	scientific-names: Scientific name of OTU classification group
-	common-names: Common name of OTU classification group
-	subject-blast-names: Grouping name
-	subject-super-kingdoms: Kingdom names
-	subject-title: Descriptive title for OTU group
-	OTU-seq: OTU reference Sequence
-
-4. Missing data codes:
-	None
-
-5. Abbreviations used:
-	OTU; Operational Taxonomic Unit
-
-6. Other relevant information:
-	Taxonomy assigned using the QIIME 1 default classifier, pre-trained against Greengenes database5 (Version 13_8, Aug 2013).

### 2022-07-18 dryad/README.Rmd
adding methodological information

@@
 ---
-title: 'Dryad Data Package for study "DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life"'
+title: 'Generating a README.txt file for the Dryad submission of data from study "DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life"'
 output:
@@
   comment = "#>",
-  echo = FALSE,
+  echo = TRUE,
   results = 'hide'
@@
 
-This README file was generated on `r Sys.Date()` by Luna L. Sánchez Reyes, <https://orcid.org/0000-0001-7668-2528>
+# Description of README file
+
+```{r}
+line0 <- paste("This README file was generated on", Sys.Date(), "by Luna L. Sánchez Reyes, <https://orcid.org/0000-0001-7668-2528>")
+```
+
+# General Information on data set
+
+```{r}
+line1 <- "
 
@@
 
-DATA & FILE OVERVIEW
+"
+```
+
+# Original file names, rename
+
+```{r, echo = TRUE}
+
+file.copy(from = "../tables/table-fringillidae-small-example.csv",
+          to = "../dryad/Sanchez-Reyes_etal_2022_table_1.csv")
+file.copy(from = "../tables/table-fringillidae-small-example-summary.csv",
+          to = "../dryad/Sanchez-Reyes_etal_2022_table_2.csv")
+file.copy(from = "../tables/table-fringillidae-all-congruified.csv",
+          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S1.csv")
+file.copy(from = "../tables/table-fringillidae-all-congruified.pdf",
+          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S1.pdf")
+file.copy(from = "../tables/table-fringillidae-all-summary.csv",
+          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S2.csv")
+file.copy(from = ".../tables/table-fringillidae-all-summary.pdf",
+          to = "../dryad/anchez-Reyes_etal_2022_supplementary_table_S2.pdf")
+```
+
+# Description of data files
+
+```{r}
+line2 <- "DATA & FILE OVERVIEW
 
@@
 The first biological example uses datelife on a small sample of bird species.
-The second one uses datelife on bird species belonging to the family Fringillidae of "true finches", following the NCBI taxonomy.
+The second one uses datelife on bird species belonging to the family Fringillidae of 'true finches', following the NCBI taxonomy.
 The mock example was created by replacing species names from the small example with letters, and reducing the number of source chronograms.
@@
 
-```{r, echo = FALSE}
+File 1 Name: Sanchez-Reyes_etal_2022_table_1.csv    
+File 1 Description: CSV (comma separated values) file with node age results from datelife's small example, presented in Table 1 of the manuscript.
+It contains node names, taxon names defining the nodes, the corresponding node ages (in Million years), and references of source chronograms that were congruified to the chosen tree topology to extract node ages.
+
+File 2 Name: Sanchez-Reyes_etal_2022_table_2.csv    
+File 2 Description: CSV (comma separated values) file summarizing age data (in Million years) from table 1, per node. These data are presented in Table 2 of the manuscript.
+
+File 3 Name: Sanchez-Reyes_etal_2022_supplementary_table_S1.csv    
+File 3 Description: CSV (comma separated values) file with node age results from datelife's Fringillidae example.
+It contains node names, taxon names defining the nodes, the corresponding node ages (in Million years), and references of source chronograms that were congruified to the chosen tree topology (shown in Figure 4B of the manuscript) to extract node ages.
+
+File 4 Name: Sanchez-Reyes_etal_2022_supplementary_table_S1.pdf    
+File 4 Description: PDF version of file Sanchez-Reyes_etal_2022_supplementary_table_S1.csv
+
+File 5 Name: Sanchez-Reyes_etal_2022_supplementary_table_S2.csv    
+File 5 Description: CSV (comma separated values) file summarizing age data (in Million years) from table S1, per node.
+
+File 6 Name: Sanchez-Reyes_etal_2022_supplementary_table_S2.pdf    
+File 6 Description: PDF file version of file Sanchez-Reyes_etal_2022_supplementary_table_S2.csv.
+
+File 7 Name: Sanchez-Reyes_etal_2022_figure_1_chronogram_mock_example.tre    
+File 7 Description: Newick version of chronogram from the mock example shown in Figure 1 of the manuscript (https://www.biorxiv.org/content/10.1101/782094v2).
+
+File 8 Name: Sanchez-Reyes_etal_2022_figure_3_chronogram_small_example.tre    
+File 8 Description: Newick version of chronogram from the small example shown in Figure 3 of the manuscript (https://www.biorxiv.org/content/10.1101/782094v2).
+
+File 9 Name: Sanchez-Reyes_etal_2022_figure_4A_topology_finches_mrca.tre    
+File 9 Description: Newick version of topology extracted from the Open Tree of Life synthetic tree, shown in Figure 4A of the manuscript (https://www.biorxiv.org/content/10.1101/782094v2).
+
+File 10 Name: Sanchez-Reyes_etal_2022_figure_4B_topology_finches_ncbi.tre    
+File 10 Description: Newick version of topology extracted from the Open Tree of Life synthetic tree, shown in Figure 4B of the manuscript (https://www.biorxiv.org/content/10.1101/782094v2).
+
+File 11 Name: Sanchez-Reyes_etal_2022_figure_5_chronogram_finches_example.tre    
+File 11 Description: Newick version of chronogram from the finch example shown in Figure 5 of the manuscript (https://www.biorxiv.org/content/10.1101/782094v2).
+
+File 12 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S1.pdf    
+File 12 Description: PDf file showing results of cross validation analysis of datelife's chronogram generating functions.
+
+File 13 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S2.jpg    
+File 13 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Barker et al.
+2013.
+Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
+
+File 14 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S2.tre    
+File 14 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S2 (gray).
+
+File 15 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S3.jpg    
+File 15 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Barker et al.
+2015, chronogram 1.
+Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
+
+File 16 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S3.tre    
+File 16 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S3 (gray).
+
+File 17 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S4.jpg    
+File 17 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Barker et al.
+2015, chronogram 2.
+Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
+
+File 18 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S4.tre    
+File 18 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S4 (gray).
+
+File 19 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S5.jpg    
+File 19 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Burns et al.
+2015.
+Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
+
+File 20 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S5.tre    
+File 20 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S5 (gray).
+
+File 21 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S6.jpg    
+File 21 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Claramunt et al.
+2015.
+Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
+
+File 22 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S6.tre    
+File 22 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S6 (gray).
+
+File 23 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S7.jpg    
+File 23 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Gibb et al.
+2015.
+Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
+
+File 24 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S7.tre    
+File 24 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S7 (gray).
+
+File 25 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S8.jpg    
+File 25 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Hedges et al.
+2015, chronogram 1.
+Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
+
+File 26 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S8.tre    
+File 26 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S8 (gray).
+
+File 27 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S9.jpg    
+File 27 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Hedges et al.
+2015, chronogram 2.
+Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
+
+File 28 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S9.tre    
+File 28 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S9 (gray).
+
+File 29 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S10.jpg    
+File 29 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Hooper et al.
+2017.
+Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
+
+File 30 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S10.tre    
+File 30 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S10 (gray).
+
+File 31 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S11.jpg    
+File 31 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Jetz et al.
+2012, chronogram 1.
+Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
+
+File 32 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S11.tre    
+File 32 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S11 (gray).
+
+File 33 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S12.jpg    
+File 33 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Jetz et al.
+2012, chronogram 2.
+Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
+
+File 34 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S12.tre    
+File 34 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S12 (gray).
+
+File 35 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S13.jpg    
+File 35 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Kimball et al.
+2019, chronogram 1.
+Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
+
+File 36 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S13.tre    
+File 36 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S13 (gray).
+
+File 37 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S14.jpg    
+File 37 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Kimball et al.
+2019, chronogram 2.
+Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
+
+File 38 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S14.tre    
+File 38 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S14 (gray).
+
+File 39 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S15.jpg    
+File 39 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Oliveros et al.
+2019.
+Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
+
+File 40 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S15.tre    
+File 40 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S15 (gray).
+
+File 41 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S16.jpg    
+File 41 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Price et al.
+2014, chronogram 1.
+Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
+
+File 42 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S16.tre    
+File 42 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S16 (gray).
+
+File 43 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S17.jpg    
+File 43 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Price et al.
+2014, chronogram 2.
+Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
+
+File 44 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S17.tre    
+File 44 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S17 (gray).
+
+File 45 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S18.jpg    
+File 45 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Roquet et al.
+2014, chronogram 1.
+Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
+
+File 46 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S18.tre    
+File 46 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S18 (gray).
+
+File 47 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S19.jpg    
+File 47 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Roquet et al.
+2014, chronogram 2.
+Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
+
+File 48 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S19.tre    
+File 48 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S19 (gray).
+
+File 49 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S20.jpg    
+File 49 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Uyeda et al.
+2017.
+Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
+
+File 50 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S20.tre    
+File 50 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S20 (gray).
+"
+```
+
+<!--
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_    
+	File XXX Description:
+
+	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_    
+	File XXX Description: 
+-->
+
+# Methods that generated the data files
+
+```{r}
+line3 <- "
+METHODOLOGICAL INFORMATION
+
+All age data and chronograms were obtained using the R package datelife v0.6.5 https://CRAN.R-project.org/package=datelife and the chronogram database OpenTreeChronograms v2022.1.28
+"
+```
+
+# Individual descriptions of each data file
+
+```{r}
+var_description_1 <- c("Node names in tree topology, as defined by taxon A and taxon B.",
+                       "Scientific name of the first element of a taxon pair that defines a node from a tree topology.",
+                       "Scientific name of the second element of a taxon pair that defines a node from a tree topology.",
+                       "Ages of nodes defined by taxon A and taxon B, in Million years.",
+                       "Reference of the study that published the chronogram from which the node ages were extracted.")
+
+var_description_2 <- c("Names of nodes from the tree topology used for the dating analysis.",
+                       "Minimum node age, in Million years.",
+                       "First quartile of node age distributions, in Million years.",
+                       "Mean node age, in Million years.",
+                       "Median node age, in Million years.",
+                       "Third quartile of node age distributions, in Million years.",
+                       "Maximum node age, in Million years.",
+                       "Variance of node age, in Million years.",
+                       "Standard deviation of node age, in Million years.")
 
@@
 tre_file_names <- c("Sanchez-Reyes_etal_2022_figure_1_chronogram_mock_example.tre")
-```
 
-```{r, echo=FALSE, message=FALSE}
-file.copy(from = "../tables/table-fringillidae-small-example.csv",
-          to = "../dryad/Sanchez-Reyes_etal_2022_table_1.csv")
-```
-
-  File 1 Name: Sanchez-Reyes_etal_2022_table_1.csv    
-	File 1 Description: CSV (comma separated values) file containing results from the small example.
-Names and ages (in Million years) of nodes from source chronograms that have been congruified to nodes of the chosen tree topology.
-
-```{r, echo=FALSE, message=FALSE}
-file.copy(from = "../tables/table-fringillidae-small-example-summary.csv",
-          to = "../dryad/Sanchez-Reyes_etal_2022_table_2.csv")
-```
-
-	File 2 Name: Sanchez-Reyes_etal_2022_table_2.csv    
-	File 2 Description: CSV (comma separated values) file containing a summary of node age (in Million years) data from table 1.
-
-```{r, echo=FALSE, message=FALSE}
-file.copy(from = "../tables/table-fringillidae-all-congruified.csv",
-          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S1.csv")
-
-```
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_table_S1.csv    
-	File XXX Description: CSV (comma separated values) file containing results from the Fringillidae example.
-Names and ages (in Million years) of nodes from source chronograms that have been congruified to nodes of the chosen tree topology shown in Figure 4B of the manuscript.
-
-```{r, echo=FALSE, eval=FALSE, message=FALSE}
-file.copy(from = "../tables/table-fringillidae-all-congruified.pdf",
-          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S1.pdf")
-
-```
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_table_S1.pdf    
-	File XXX Description: PDF version of file Sanchez-Reyes_etal_2022_supplementary_table_S1.csv
-
-```{r, echo=FALSE, message=FALSE, }
-file.copy(from = "../tables/table-fringillidae-all-summary.csv",
-          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S2.csv")
-
-```
-
-  File XXX Name: Sanchez-Reyes_etal_2022_supplementary_table_S2.csv    
-	File XXX Description: CSV (comma separated values) file containing a summary of node age data (in Million years) from table S1.
-
-```{r, echo=FALSE, message=FALSE}
-file.copy(from = ".../tables/table-fringillidae-all-summary.pdf",
-          to = "../dryad/anchez-Reyes_etal_2022_supplementary_table_S2.pdf")
-
-```
-
-  File XXX Name: Sanchez-Reyes_etal_2022_supplementary_table_S2.pdf    
-	File XXX Description: PDF file version of file Sanchez-Reyes_etal_2022_supplementary_table_S2.csv.
-
-	File 3 Name: Sanchez-Reyes_etal_2022_figure_1_chronogram_mock_example.tre    
-	File 3 Description: Newick version of chronogram from the mock example shown in Figure 1 of the manuscript (https://www.biorxiv.org/content/10.1101/782094v2).
-
-	File 4 Name: Sanchez-Reyes_etal_2022_figure_3_chronogram_small_example.tre    
-	File 4 Description: Newick version of chronogram from the small example shown in Figure 3 of the manuscript (https://www.biorxiv.org/content/10.1101/782094v2).
-
-	File 6 Name: Sanchez-Reyes_etal_2022_figure_4A_topology_finches_mrca.tre    
-	File 6: Description: Newick version of topology extracted from the Open Tree of Life synthetic tree, shown in Figure 4A of the manuscript (https://www.biorxiv.org/content/10.1101/782094v2).
-
-	File 7 Name: Sanchez-Reyes_etal_2022_figure_4B_topology_finches_ncbi.tre    
-	File 7 Description: Newick version of topology extracted from the Open Tree of Life synthetic tree, shown in Figure 4B of the manuscript (https://www.biorxiv.org/content/10.1101/782094v2).
-
-	File 5 Name: Sanchez-Reyes_etal_2022_figure_5_chronogram_finches_example.tre    
-	File 5 Description: Newick version of chronogram from the finch example shown in Figure 5 of the manuscript (https://www.biorxiv.org/content/10.1101/782094v2).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S1.pdf    
-	File XXX Description: PDf file showing results of cross validation analysis of datelife's chronogram generating functions.
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S2.jpg    
-	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Barker et al.
-2013.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S2.tre    
-	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S2 (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S3.jpg    
-	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Barker et al.
-2015, chronogram 1.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S3.tre    
-	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S3 (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S4.jpg    
-	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Barker et al.
-2015, chronogram 2.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S4.tre    
-	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S4 (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S5.jpg    
-	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Burns et al.
-2015.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S5.tre    
-	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S5 (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S6.jpg    
-	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Claramunt et al.
-2015.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S6.tre    
-	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S6 (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S7.jpg    
-	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Gibb et al.
-2015.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S7.tre    
-	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S7 (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S8.jpg    
-	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Hedges et al.
-2015, chronogram 1.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S8.tre    
-	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S8 (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S9.jpg    
-	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Hedges et al.
-2015, chronogram 2.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S9.tre    
-	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S9 (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S10.jpg    
-	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Hooper et al.
-2017.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S10.tre    
-	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S10 (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S11.jpg    
-	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Jetz et al.
-2012, chronogram 1.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S11.tre    
-	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S11 (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S12.jpg    
-	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Jetz et al.
-2012, chronogram 2.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S12.tre    
-	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S12 (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S13.jpg    
-	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Kimball et al.
-2019, chronogram 1.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S13.tre    
-	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S13 (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S14.jpg    
-	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Kimball et al.
-2019, chronogram 2.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S14.tre    
-	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S14 (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S15.jpg    
-	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Oliveros et al.
-2019.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S15.tre    
-	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S15 (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S16.jpg    
-	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Price et al.
-2014, chronogram 1.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S16.tre    
-	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S16 (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S17.jpg    
-	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Price et al.
-2014, chronogram 2.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S17.tre    
-	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S17 (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S18.jpg    
-	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Roquet et al.
-2014, chronogram 1.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S18.tre    
-	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S18 (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S19.jpg    
-	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Roquet et al.
-2014, chronogram 2.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S19.tre    
-	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S19 (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S20.jpg    
-	File XXX Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Uyeda et al.
-2017.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_figure_S20.tre    
-	File XXX Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S20 (gray).
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_    
-	File XXX Description:
-
-	File XXX Name: Sanchez-Reyes_etal_2022_supplementary_    
-	File XXX Description:
-
-3. Original names and locations of files at https://github.com/LunaSare/datelifeMS1:
-
-```
-cp tables/table-fringillidae-all.pdf dryad/Supplementary_Table_S1.pdf
-cp tables/table-fringillidae-all-summary.pdf dryad/Supplementary_Table_S2.pdf
-cp figures/figure-cross-validation/fig-cross-validation-xy-plots-diffs.pdf dryad/Supplementary_Figure_S1.pdf
-```
-
-METHODOLOGICAL INFORMATION
-
-Bacterial detection was performed on the ViiA7 Real-time PCR System (Thermo Fisher Scientific) using PrimeTime® qPCR primers, probes and mastermix (IDT) according to the manufacturer’s instructions.
-Reactions were performed using 1X PrimeTime® Gene Expression Master Mix, 1X PrimeTime®qPCR Assay and up to 10ng of DNA.
-Cycling conditions were 95°C for 3 minutes, 60 cycles of 95°C for 5 seconds and 60°C for 30 seconds.
-Amplification results were reviewed using QuantStudioTM Real-Time PCR Software version 1.1 (Thermo Fisher Scientific).
-Amplification of beta-actin and prostaglandin transporter (PGT) was used to determine relative abundance.
-
-Diversity profiling was performed by AGRF (Australian Genome Research Facility, Melbourne Australia).
-Samples were amplified with universal primers to the V1-V3 region of the bacterial 16S gene (forward AGAGTTTGATCMTGGCTCAG; reverse GWATTACCGCGGCKGCTG).
-Amplicons were indexed using the Nextera XT Index Kit (Illumina, San Diego, CA, USA) followed by Paired End sequencing on a MiSeq next generation sequencer (Illumina).
-Paired-end reads were assembled by aligning the forward and reverse reads using PEAR1 (version 0.9.5).
-Primers were identified and trimmed.
-Trimmed sequences were processed using Quantitative Insights into Microbial Ecology (QIIME 1.8) USEARCH (version 8.0.1623) and UPARSE software.
-Sequences were quality filtered and sorted by abundance after removal of full-length duplicate sequences.
-Singletons or unique reads were discarded.
-Sequences were clustered and then chimera filtered using “rdp_gold” database as reference.
-Reads were mapped back to Operational Taxonomic Units with a minimum identity of 97% and taxonomy was assigned using the QIIME 1 default classifier, pre-trained against Greengenes database5 (Version 13_8, Aug 2013).
-
-```{r}
-
-var_description_1 <- c("Node names in tree topology, as defined by taxon A and taxon B",
-                       "Scientific name of the first element of a taxon pair that defines a node from a tree topology",
-                       "Scientific name of the second element of a taxon pair that defines a node from a tree topology",
-                       "Ages of nodes defined by taxon A and taxon B, in Million years",
-                       "Reference of the study that published the chronogram from which the node ages were extracted")
-
-var_description_2 <- c("Names of nodes from the tree topology used for the dating analysis",
-                       "Minimum node age, in Million years",
-                       "First quartile of node age distributions, in Million years",
-                       "Mean node age, in Million years",
-                       "Median node age, in Million years",
-                       "Third quartile of node age distributions, in Million years",
-                       "Maximum node age, in Million years",
-                       "Variance of node age, in Million years",
-                       "Standar deviation of node age, in Million years")
-```
-
-```{r}
 text <- c()
@@
   # read the csv file in:
-  tt <- utils::read.csv(file = paste0("../dryad/", file_name), header = TRUE, row.names = NULL)
+  csv_data <- utils::read.csv(file = paste0("../dryad/", file_name), header = TRUE, row.names = NULL)
   # remove the column with row numbers:
-  tt <- tt[,-1]
+  csv_data <- csv_data[,-1]
   # create the description text
-  var_list <- gsub("[.]", " ", colnames(tt))
-  var_list <- paste0("\\tab ", var_list, ": ", var_descriptions[[file_name]], "\\n ")
-  var_list <- paste0(var_list, collapse = "")
-  print(var_list)
-  text <- paste(text, 
-                paste("DATA-SPECIFIC INFORMATION FOR:", file_name, "\\n "),
-                "\\n1. Number of variables/columns:", ncol(tt), "\\n ",
-                "\\n2. Number of cases/rows:", nrow(tt), "\\n ",
-                "\\n3. Variable List \\n ", var_list,
-                "\\n4. Missing data codes: None \\n ",
-                "\\n5. Abbreviations used: NA = not applicable \\n \\n ")
+  var_list <- gsub("[.]", " ", colnames(csv_data))
+  var_list <- paste0("\t", var_list, ": ", var_descriptions[[file_name]])
+  text <- c(text, 
+            c(paste("\nDATA-SPECIFIC INFORMATION FOR:", file_name, "\n"),
+              paste("1. Number of variables/columns:", ncol(csv_data), "\n"),
+              paste("2. Number of cases/rows:", nrow(csv_data), "\n"),
+              "3. Variable List:", var_list,
+              "\n4. Missing data codes: None",
+              "\n5. Abbreviations used: NA = not applicable\n"))
 }
+
+line4 <- text
 ```
 
-`r text`
+# Writing everything to TXT file
+
+```{r, echo = TRUE}
+
+writeLines(text = c(line0, 
+                    line1,
+                    line2,
+                    line3,
+                    line4), con = "../dryad/README_Sanchez-Reyes_etal_2022_data.txt")
+```

### 2022-07-18 dryad/README.Rmd
removing row number column from csv files for dryad upload

@@
 file.copy(from = "../tables/table-fringillidae-small-example.csv",
-          to = "../dryad/Sanchez-Reyes_etal_2022_table_1.csv")
+          to = "../dryad/Sanchez-Reyes_etal_2022_table_1.csv",
+          overwrite = TRUE)
 file.copy(from = "../tables/table-fringillidae-small-example-summary.csv",
-          to = "../dryad/Sanchez-Reyes_etal_2022_table_2.csv")
+          to = "../dryad/Sanchez-Reyes_etal_2022_table_2.csv",
+          overwrite = TRUE)
 file.copy(from = "../tables/table-fringillidae-all-congruified.csv",
-          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S1.csv")
+          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S1.csv",
+          overwrite = TRUE)
 file.copy(from = "../tables/table-fringillidae-all-congruified.pdf",
-          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S1.pdf")
+          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S1.pdf",
+          overwrite = TRUE)
 file.copy(from = "../tables/table-fringillidae-all-summary.csv",
-          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S2.csv")
+          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S2.csv",
+          overwrite = TRUE)
 file.copy(from = ".../tables/table-fringillidae-all-summary.pdf",
-          to = "../dryad/anchez-Reyes_etal_2022_supplementary_table_S2.pdf")
+          to = "../dryad/anchez-Reyes_etal_2022_supplementary_table_S2.pdf",
+          overwrite = TRUE)
 ```

### 2022-07-18 dryad/README.Rmd
newick file of mock example chronogram

--- 2022-07-18 dryad/README.Rmd
@@
 file.copy(from = "../tables/table-fringillidae-small-example.csv",
-          to = "../dryad/Sanchez-Reyes_etal_2022_table_1.csv")
+          to = "../dryad/Sanchez-Reyes_etal_2022_table_1.csv",
+          overwrite = TRUE)
 file.copy(from = "../tables/table-fringillidae-small-example-summary.csv",
-          to = "../dryad/Sanchez-Reyes_etal_2022_table_2.csv")
+          to = "../dryad/Sanchez-Reyes_etal_2022_table_2.csv",
+          overwrite = TRUE)
 file.copy(from = "../tables/table-fringillidae-all-congruified.csv",
-          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S1.csv")
+          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S1.csv",
+          overwrite = TRUE)
 file.copy(from = "../tables/table-fringillidae-all-congruified.pdf",
-          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S1.pdf")
+          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S1.pdf",
+          overwrite = TRUE)
 file.copy(from = "../tables/table-fringillidae-all-summary.csv",
-          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S2.csv")
+          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S2.csv",
+          overwrite = TRUE)
 file.copy(from = ".../tables/table-fringillidae-all-summary.pdf",
-          to = "../dryad/anchez-Reyes_etal_2022_supplementary_table_S2.pdf")
+          to = "../dryad/anchez-Reyes_etal_2022_supplementary_table_S2.pdf",
+          overwrite = TRUE)
+file.copy(from = "../figures/figure-workflow/median_chronogram.tre",
+          to = "../dryad/Sanchez-Reyes_etal_2022_figure_1_chronogram_mock_example.tre")
 ```

### 2022-07-18 dryad/README.Rmd
newick file of small example chronogram

--- 2022-07-18 dryad/README.Rmd
@@
 file.copy(from = "../tables/table-fringillidae-small-example.csv",
-          to = "../dryad/Sanchez-Reyes_etal_2022_table_1.csv")
+          to = "../dryad/Sanchez-Reyes_etal_2022_table_1.csv",
+          overwrite = TRUE)
 file.copy(from = "../tables/table-fringillidae-small-example-summary.csv",
-          to = "../dryad/Sanchez-Reyes_etal_2022_table_2.csv")
+          to = "../dryad/Sanchez-Reyes_etal_2022_table_2.csv",
+          overwrite = TRUE)
 file.copy(from = "../tables/table-fringillidae-all-congruified.csv",
-          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S1.csv")
+          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S1.csv",
+          overwrite = TRUE)
 file.copy(from = "../tables/table-fringillidae-all-congruified.pdf",
-          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S1.pdf")
+          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S1.pdf",
+          overwrite = TRUE)
 file.copy(from = "../tables/table-fringillidae-all-summary.csv",
-          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S2.csv")
+          to = "../dryad/Sanchez-Reyes_etal_2022_supplementary_table_S2.csv",
+          overwrite = TRUE)
 file.copy(from = ".../tables/table-fringillidae-all-summary.pdf",
-          to = "../dryad/anchez-Reyes_etal_2022_supplementary_table_S2.pdf")
+          to = "../dryad/anchez-Reyes_etal_2022_supplementary_table_S2.pdf",
+          overwrite = TRUE)
+file.copy(from = "../figures/figure-workflow/median_chronogram.tre",
+          to = "../dryad/Sanchez-Reyes_etal_2022_figure_1_chronogram_mock_example.tre",
+          overwrite = TRUE)
+file.copy(from = "../figures/figure-small-example/median_chronogram.tre",
+          to = "../dryad/Sanchez-Reyes_etal_2022_figure_3_chronogram_small_example.tre",
+          overwrite = TRUE)
 ```

### 2022-07-18 dryad/README.Rmd
newick file of fringillidae example chronogram

--- 2022-07-18 dryad/README.Rmd
@@
           to = "../dryad/anchez-Reyes_etal_2022_supplementary_table_S2.pdf",
+          overwrite = TRUE)
+file.copy(from = "../figures/figure-workflow/median_chronogram.tre",
+          to = "../dryad/Sanchez-Reyes_etal_2022_figure_1_chronogram_mock_example.tre",
+          overwrite = TRUE)
+file.copy(from = "../figures/figure-small-example/median_chronogram.tre",
+          to = "../dryad/Sanchez-Reyes_etal_2022_figure_3_chronogram_small_example.tre",
+          overwrite = TRUE)
+file.copy(from = "../figures/figure-fringillidae/median_chronogram.tre",
+          to = "../dryad/Sanchez-Reyes_etal_2022_figure_5_chronogram_finches_example.tre",
           overwrite = TRUE)

### 2022-07-18 dryad/README.Rmd
newick files of fringillidae topologies

--- 2022-07-18 dryad/README.Rmd
@@
 file.copy(from = "../figures/figure-workflow/median_chronogram.tre",
-          to = "../dryad/Sanchez-Reyes_etal_2022_figure_1_chronogram_mock_example.tre")
+          to = "../dryad/Sanchez-Reyes_etal_2022_figure_1_chronogram_mock_example.tre",
+          overwrite = TRUE)
+file.copy(from = "../figures/figure-small-example/median_chronogram.tre",
+          to = "../dryad/Sanchez-Reyes_etal_2022_figure_3_chronogram_small_example.tre",
+          overwrite = TRUE)
+file.copy(from = "../figures/figure-fringillidae/median_chronogram.tre",
+          to = "../dryad/Sanchez-Reyes_etal_2022_figure_5_chronogram_finches_example.tre",
+          overwrite = TRUE)
+file.copy(from = "../figures/fringillidae-topologies/fringillidae_monophyletic_topology.tre",
+          to = "../dryad/Sanchez-Reyes_etal_2022_figure_4A_topology_finches_mrca.tre",
+          overwrite = TRUE)
+file.copy(from = "../figures/fringillidae-topologies/fringillidae_paraphyletic_topology.tre",
+          to = "../dryad/Sanchez-Reyes_etal_2022_figure_4B_topology_finches_ncbi.tre",
+          overwrite = TRUE)
 ```

### 2022-07-18 dryad/README.Rmd
updating dryad readme with cross validation files

@@
 ```{r}
-line2 <- "DATA & FILE OVERVIEW
+line2.1 <- "DATA & FILE OVERVIEW
 
@@
 File 12 Description: PDf file showing results of cross validation analysis of datelife's chronogram generating functions. 
+"
+```
 
-File 13 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S2.jpg    
-File 13 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Barker et al.
-2013.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
+```{r}
+line2.2 <- c()
+# short study references:
+names_phylo_all <- c("Barker et al. 2013",  # 1
+                      "Barker et al. 2015 - chronogram 1",  # 2
+                      "Barker et al. 2015 - chronogram 2",  # 3
+                      "Burns et al. 2014",  # 4
+                      "Claramunt et al. 2015",  # 5
+                      "Gibb et al. 2015", # 6
+                      "Hedges et al. 2015 - chronogram 1",  # 7
+                      "Hedges et al. 2015 - chronogram 2",  # 8
+                      "Hooper et al. 2017",  # 9
+                      "Jetz et al. 2012 - chronogram 1",  # 10
+                      "Jetz et al. 2012 - chronogram 2",  # 11
+                      "Kimball et al. 2019 - chronogram 1", # 12
+                      "Kimball et al. 2019 - chronogram 2", # 13
+                      "Oliveros et al. 2019", # 14
+                      "Price et al. 2014 - chronogram 1", # 15
+                      "Price et al. 2014 - chronogram 2", # 16
+                      "Roquet et al. 2014 - chronogram 1", # 17
+                      "Roquet et al. 2014 - chronogram 2", # 18 
+                      "Uyeda et al 2017")
+file_index <- 12
+fig_index <- 1
+for (name in names_phylo_all) {
+  file_index <- file_index + 1
+  fig_index <- fig_index + 1
+  x1 <- paste0("File ", file_index, " Name: Sanchez-Reyes_etal_2022_supplementary_figure_S", fig_index, ".jpg")
+  x2 <- paste0("File ", file_index, " Description: JPG file of results of cross validation analysis of datelife chronogram generating functions, using a tree topology from ", name, ".
+Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).\n")
+  file_index <- file_index + 1
+  x3 <- paste0("File ", file_index, " Name: Sanchez-Reyes_etal_2022_supplementary_figure_S", fig_index, "_cross_validated.tre")
+  x4 <- paste0("File ", file_index, " Description: Newick file of chronogram cross validated using dates obtained with datelife, shown in gray on supplementary figure Sanchez-Reyes_etal_2022_supplementary_figure_S", fig_index, ".jpg\n")
+  file_index <- file_index + 1
+  x5 <- paste0("File ", file_index, " Name: Sanchez-Reyes_etal_2022_supplementary_figure_S", fig_index, "_source.tre")
+  x6 <- paste0("File ", file_index, " Description: Newick file of source chronogram from", name, " used for the cross validaton analysis shown on supplementary figure Sanchez-Reyes_etal_2022_supplementary_figure_S", fig_index, ".jpg\n")
 
-File 14 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S2.tre    
-File 14 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S2 (gray).
+  line2.2 <- c(line2.2, x1, x2, x3, x4, x5, x6)
+}
 
-File 15 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S3.jpg    
-File 15 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Barker et al.
-2015, chronogram 1.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-File 16 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S3.tre    
-File 16 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S3 (gray).
-
-File 17 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S4.jpg    
-File 17 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Barker et al.
-2015, chronogram 2.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-File 18 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S4.tre    
-File 18 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S4 (gray).
-
-File 19 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S5.jpg    
-File 19 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Burns et al.
-2015.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-File 20 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S5.tre    
-File 20 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S5 (gray).
-
-File 21 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S6.jpg    
-File 21 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Claramunt et al.
-2015.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-File 22 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S6.tre    
-File 22 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S6 (gray).
-
-File 23 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S7.jpg    
-File 23 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Gibb et al.
-2015.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-File 24 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S7.tre    
-File 24 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S7 (gray).
-
-File 25 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S8.jpg    
-File 25 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Hedges et al.
-2015, chronogram 1.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-File 26 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S8.tre    
-File 26 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S8 (gray).
-
-File 27 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S9.jpg    
-File 27 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Hedges et al.
-2015, chronogram 2.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-File 28 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S9.tre    
-File 28 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S9 (gray).
-
-File 29 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S10.jpg    
-File 29 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Hooper et al.
-2017.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-File 30 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S10.tre    
-File 30 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S10 (gray).
-
-File 31 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S11.jpg    
-File 31 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Jetz et al.
-2012, chronogram 1.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-File 32 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S11.tre    
-File 32 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S11 (gray).
-
-File 33 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S12.jpg    
-File 33 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Jetz et al.
-2012, chronogram 2.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-File 34 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S12.tre    
-File 34 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S12 (gray).
-
-File 35 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S13.jpg    
-File 35 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Kimball et al.
-2019, chronogram 1.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-File 36 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S13.tre    
-File 36 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S13 (gray).
-
-File 37 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S14.jpg    
-File 37 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Kimball et al.
-2019, chronogram 2.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-File 38 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S14.tre    
-File 38 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S14 (gray).
-
-File 39 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S15.jpg    
-File 39 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Oliveros et al.
-2019.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-File 40 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S15.tre    
-File 40 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S15 (gray).
-
-File 41 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S16.jpg    
-File 41 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Price et al.
-2014, chronogram 1.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-File 42 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S16.tre    
-File 42 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S16 (gray).
-
-File 43 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S17.jpg    
-File 43 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Price et al.
-2014, chronogram 2.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-File 44 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S17.tre    
-File 44 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S17 (gray).
-
-File 45 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S18.jpg    
-File 45 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Roquet et al.
-2014, chronogram 1.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-File 46 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S18.tre    
-File 46 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S18 (gray).
-
-File 47 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S19.jpg    
-File 47 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Roquet et al.
-2014, chronogram 2.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-File 48 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S19.tre    
-File 48 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S19 (gray).
-
-File 49 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S20.jpg    
-File 49 Description: JPG file of results of cross validation analysis of datelife's chronogram generating functions, using a tree topology from Uyeda et al.
-2017.
-Comparison of original chronogram (black) and the chronogram obtained using datelife (gray).
-
-File 50 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S20.tre    
-File 50 Description: Newick file of chronogram obtained with datelife, shown in supplementary Figure S20 (gray).
-"
 ```
@@
                     line1,
-                    line2,
+                    line2.1,
+                    line2.2,
                     line3,

### 2022-07-18 dryad/README.Rmd
adding newick files of source chronograms to dryad folder

@@
           overwrite = TRUE)
+
+files_in_path <- list.files(path = "../figures/figure-fringillidae/source_chronograms")
+
+# get only newick file names
+newick_files <- files_in_path[grep("[.]tre", files_in_path)]
+
+# order files numerically
+newick_files <- stringr::str_sort(newick_files, numeric = TRUE)
+
+index <- 1
+for (file_name in newick_files) {
+  index <- index + 1
+  file.copy(from = paste0("../figures/figure-fringillidae/source_chronograms/", file_name),
+          to = paste0("../dryad/Sanchez-Reyes_etal_2022_supplementary_figure_S", index, "_source.tre"),
+          overwrite = TRUE)
+}
 ```

### 2022-07-18 dryad/README.Rmd
adding newick files of cross validated chronograms to dryad folder

@@
 
-# get only newick file names
+# get source chronograms newick file names
 newick_files <- files_in_path[grep("[.]tre", files_in_path)]
@@
           overwrite = TRUE)
+  write(paste0("https://raw.githubusercontent.com/LunaSare/datelifeMS1/main/dryad/Sanchez-Reyes_etal_2022_supplementary_figure_S", index, "_source.tre"), file = "../dryad/urls.txt", append=TRUE)
 }
+
+# get cross validated chronograms newick file names
+files_in_path <- list.files(path = "../figures/fringillidae-cross-validation")
+
+newick_files <- files_in_path[grep("[.]tre", files_in_path)]
+
+# order files numerically
+newick_files <- stringr::str_sort(newick_files, numeric = TRUE)
+
+index <- 1
+for (file_name in newick_files) {
+  index <- index + 1
+  file.copy(from = paste0("../figures/fringillidae-cross-validation/", file_name),
+          to = paste0("../dryad/Sanchez-Reyes_etal_2022_supplementary_figure_S", index, "_cross_validated.tre"),
+          overwrite = TRUE)
+  write(paste0("https://raw.githubusercontent.com/LunaSare/datelifeMS1/main/dryad/Sanchez-Reyes_etal_2022_supplementary_figure_S", index, "_cross_validated.tre"), file = "../dryad/urls.txt", append=TRUE)
+}
+
 ```

### 2022-10-28 manuscript/manuscript-1st-submission.Rmd
Update review.md

--- 2022-07-06 manuscript/manuscript.Rmd
@@
 We illustrate the DateLife workflow using a family within the Passeriform birds encompassing the true finches, Fringillidae, as case study.
-The first example analyses 6 bird species and shows all steps of the workflow The second example is an analysis of 289 species in the family Fringillidae that are included in the NCBI taxonomy.
+On a small example, we analysed 6 bird species, and results from each step of the workflow are shown in Fig.
+\ref{fig:figure2}.
+As a second example, we analysed 289 bird species in the family Fringillidae that are included in the NCBI taxonomy.
+The resulting summary chronogram is shown in Fig.
+\ref{fig:fringillidages}, and results from previous steps of the workflow are available as Supplementary Figures.
 <!--We used `r cite_r("r-references.bib")` for all our analyses.-->

### 2022-12-14 manuscript/manuscript.Rmd
addressing comments about suppl data

--- 2022-07-06 manuscript/manuscript.Rmd
@@
 
-Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities are available at [datelifeMS1](https://github.com/LunaSare/datelifeMS1), [datelife_examples](https://github.com/LunaSare/datelife_examples), and [datelife_benchmark](https://github.com/LunaSare/datelife_benchmark) repositories in LLSR's GitHub account.
+Supplementary material, including code, biological examples and benchmark results data files and online-only appendices, can be found in the Dryad data repository (<https://doi.org/10.5061/dryad.cnp5hqc6w>), as well as in the GitHub repositories used to develop the reproducible manuscript (<https://doi.org/10.5281/zenodo.7435094>), the biological examples (<https://doi.org/10.5281/zenodo.7435101>), and the software benchmark (<https://doi.org/10.5281/zenodo.7435106>).
 

### 2023-01-12 manuscript/fig-topologies.Rmd
updating topologies figure caption

@@
 A) Topology of 2,333 tips and 1,305 internal nodes, encompassing bird species within the family Fringillidae following the NCBI taxonomy (black), as well as all other bird species that share the same Most Recent Common Ancestor (MRCA) node in OpenTree's synthetic tree (purple).
-B) Subtree topology of of 289 tips and 253 internal nodes, resulting from pruning species that do not belong to the family Fringillidae according to the NCBI taxonomy (purple branches in topology A).
+B) Subtree topology of 289 tips and 253 internal nodes, resulting from pruning species that do not belong to the family Fringillidae according to the NCBI taxonomy (purple branches in topology A).
 Bird species within the Fringillidae are paraphyletic

### 2023-01-12 manuscript/manuscript-1st-submission.tex
fixed rendering issues with 1st submission

--- 2022-07-06 manuscript/manuscript.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
 \PassOptionsToPackage{hyphens}{url}
 %
-\documentclass[
-  man]{apa6}
-\usepackage{amsmath,amssymb}
+\documentclass[english,man]{apa6}
 \usepackage{lmodern}
-\usepackage{iftex}
-\ifPDFTeX
+\usepackage{amssymb,amsmath}
+\usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
   \usepackage[T1]{fontenc}
   \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
+  \usepackage{textcomp} % provides euro and other symbols
+\else % if luatex or xelatex
   \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
 \fi
-% Use upquote if available, for straight quotes in verbatim environments
+% use upquote if available, for straight quotes in verbatim environments
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
+% use microtype if available
+\IfFileExists{microtype.sty}{%
 \usepackage[]{microtype}
@@
 }{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
 \IfFileExists{parskip.sty}{%
@@
 \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+}
+\usepackage{hyperref}
 \hypersetup{
@@
             pdfauthor={Luna L. Sánchez Reyes1,2, Emily Jane McTavish1, \& Brian O'Meara2},
-  pdflang={en-EN},
             pdfkeywords={Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations},
-  hidelinks,
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
-\usepackage{graphicx}
+            pdfborder={0 0 0},
+            breaklinks=true}
+\urlstyle{same}  % don't use monospace font for urls
+\usepackage{graphicx,grffile}
 \makeatletter
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
+\setlength{\emergencystretch}{3em}  % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{0}
+
+% set default figure placement to htbp
 \makeatletter
@@
 \makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-% Make \paragraph and \subparagraph free-standing
-\ifx\paragraph\undefined\else
-  \let\oldparagraph\paragraph
-  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
-\fi
-\ifx\subparagraph\undefined\else
-  \let\oldsubparagraph\subparagraph
-  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
-\fi
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-\ifLuaTeX
-\usepackage[bidi=basic]{babel}
-\else
-\usepackage[bidi=default]{babel}
-\fi
-\babelprovide[main,import]{english}
-% get rid of language-specific shorthands (see #6817):
-\let\LanguageShortHands\languageshorthands
-\def\languageshorthands#1{}
+
 \DeclareUnicodeCharacter{0301}{*************************************}
@@
 \captionsetup[table]{font+=linenumbers, labelfont={sc}, labelformat={default}, labelsep=period, name={Table}}
-
 % Manuscript styling
@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4926}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4960}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 \usepackage{csquotes}
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[shorthands=off,main=english]{babel}
+\else
+  % load polyglossia as late as possible as it *could* call bidi if RTL lang (e.g. Hebrew or Arabic)
+  \usepackage{polyglossia}
+  \setmainlanguage[]{english}
 \fi
@@
 DateLife's core software application consists of the R package \texttt{datelife}.
-Its current stable version -- v0.6.5, is available from the The Comprehensive R Archive Network (CRAN) repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
+Its current stable version -- v0.6.6, is available from the The Comprehensive R Archive Network (CRAN) repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
 ape (Paradis, Claude, \& Strimmer, 2004),
-bold (Chamberlain et al., 2019),
+bold (Chamberlain, 2018),
 geiger (Pennell et al., 2014),
@@
 rotl (Michonneau, Brown, \& Winter, 2016), and
-taxize (Chamberlain \& Szöcs, 2013; Chamberlain et al., 2019).
+taxize (Chamberlain, 2018; Chamberlain \& Szöcs, 2013).
 Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife workflow: creating a search query, searching a database, and summarizing results from the search.
@@
 \begin{minipage}[t]{17.3cm}
-\includegraphics{../figures/figure1/figure1-new.pdf}
+\includegraphics{../figures/figure-workflow/figure1-new.pdf}
 \captionof{figure}{\textit{Main DateLife workflow.
@@
 DateLife starts by processing an input consisting of at least two taxon names, which can be provided as a comma separated character string or as tip labels on a tree.
-If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis et al., 2004).
+If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a \enquote{phylo} R object (Paradis et al., 2004).
 The input tree is not required to have branch lengths, and its topology is used in the summary steps described in the next section.
@@
 Subspecies and variants are ignored.
-If an input taxon name belongs to an inclusive taxonomic group, DateLife has two alternative behaviors defined by the ``get species from taxon'' flag.
+If an input taxon name belongs to an inclusive taxonomic group, DateLife has two alternative behaviors defined by the \enquote{get species from taxon} flag.
 If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group following a standard taxonomy of choice, and adds them to the input string.
-Taxonomies currently supported by DateLife are Open Tree of Life (OpenTree) unified Taxonomy (OTT, Rees \& Cranston, 2017), the National Center of Biotechnology Information (NCBI) taxonomic database (Schoch et al., 2020), the Global Biodiversity Information Facility (GBIF) taxonomic backbone (GBIF Secretariat, 2022), and the Interim Register of Marine and Nonmarine Genera (IRMNG) database (Rees, Vandepitte, Decock, \& Vanhoorne, 2017).
+Taxonomies currently supported by DateLife are Open Tree of Life (OpenTree) unified Taxonomy (OTT, Rees \& Cranston, 2017), the National Center of Biotechnology Information (NCBI) taxonomic database (Schoch et al., 2020), the Global Biodiversity Information Facility (GBIF) taxonomic backbone (GBIF Secretariat, 2022), and the Interim Register of Marine and Nonmarine Genera (IRMNG) database (Rees et al., 2017).
 If the flag is inactive, DateLife excludes any taxon names above the species level from the search query.
@@
 Total distance (in units of millions of years) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Figure \ref{fig:workflow}).
-The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a ``phylo'' object or newick string.
+The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a \enquote{phylo} object or newick string.
 Finally, the patristic matrices are associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.
@@
 \item
-  The matching pruned chronograms as newick strings or ``phylo'' objects.
+  The matching pruned chronograms as newick strings or \enquote{phylo} objects.
 \item
@@
 In rare cases, a group of trees can have multiple groves.
-By default, DateLife chooses the grove with the most taxa, however, the ``criterion = trees'' flag allows the user to choose the grove with the most trees instead.
+By default, DateLife chooses the grove with the most taxa, however, the \enquote{criterion = trees} flag allows the user to choose the grove with the most trees instead.
 If source chronograms do not form a grove, the supertree reconstruction will fail.
@@
 
-For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method {[}SDM; Criscuolo, Berry, Douzery, and Gascuel (2006){]}, which deforms patristic distance matrices by minimizing variance and then averaging them.
+For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method (SDM; Criscuolo, Berry, Douzery, \& Gascuel, 2006), which deforms patristic distance matrices by minimizing variance and then averaging them.
 These single summary taxon pair age matrices (Summarized calibrations) can be applied as calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Webb, Ackerly, \& Kembel, 2008; Webb \& Donoghue, 2005), and treePL (Smith \& O'Meara, 2012).
@@
 \emph{\textbf{Visualizing results.--}}
-Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R ``phylo'' format), as well as visualize and compare results graphically, or construct their own graphs using DateLife's chronogram plot generation functions available from the R package \texttt{datelifeplot} (Sanchez-Reyes \& O'Meara, 2022).
+Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R \enquote{phylo} format), as well as visualize and compare results graphically, or construct their own graphs using DateLife's chronogram plot generation functions available from the R package \texttt{datelifeplot} (Sanchez-Reyes \& O'Meara, 2022).
 
@@
 We illustrate the DateLife workflow using a family within the Passeriform birds encompassing the true finches, Fringillidae, as case study.
-The first example analyses 6 bird species and shows all steps of the workflow The second example is an analysis of 289 species in the family Fringillidae that are included in the NCBI taxonomy.
+On a small example, we analysed 6 bird species, and results from each step of the workflow are shown in Fig.
+\ref{fig:figure2}.
+As a second example, we analysed 289 bird species in the family Fringillidae that are included in the NCBI taxonomy.
+The resulting summary chronogram is shown in Fig.
+\ref{fig:fringillidages}, and results from previous steps of the workflow are available as Supplementary Figures.
 
@@
 DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and congruified and mapped age data to nodes in this chosen topology (Table \ref{tbl:table1}).
-The name processing step allowed including five data points for node ``n4'' (parent of \emph{Schoeniclus elegans}; Fig.
+The name processing step allowed including five data points for node \enquote{n4} (parent of \emph{Schoeniclus elegans}; Fig.
 \ref{fig:figure2}A) that would not have had any data otherwise due to name mismatch.
 Age summary statistics per node were calculated (Table \ref{tbl:table2}) and used as calibrations to date the tree topology using the BLADJ algorithm.
-As expected, more inclusive nodes (e.g., node ``n1'') have more variance in age data than less inclusive nodes (e.g., node ``n5'').
-Summary age data for node ``n2'' were excluded as final calibration because they are older than age data of the more inclusive node, ``n1'' (Fig. \ref{fig:figure2}C4).
+As expected, more inclusive nodes (e.g., node \enquote{n1}) have more variance in age data than less inclusive nodes (e.g., node \enquote{n5}).
+Summary age data for node \enquote{n2} were excluded as final calibration because they are older than age data of the more inclusive node, \enquote{n1} (Fig. \ref{fig:figure2}C4).
 
@@
 \begin{center}
-\includegraphics{../figures/figure2/figure2.pdf}
+\includegraphics{../figures/figure-small-example/figure2.pdf}
 \end{center}
@@
 
-To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the ``get species from taxon'' flag,
+To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the \enquote{get species from taxon} flag,
 which gets all recognized species names within a named group from a taxonomy of choice.
@@
 The final step is to congruify and summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM.
-As explained in the ``Description'' section, a tree topology to summarize age data upon is required.
+As explained in the \enquote{Description} section, a tree topology to summarize age data upon is required.
 By default, DateLife uses the topology from OpenTree's synthetic tree that contains the species in the search query to summarize age data upon.
@@
 We found that node ages from original study, and ages estimated using all other age data available are correlated (Fig. \ref{fig:cvXY}).
-For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka `closer to the root') relative to the original estimate, and overestimate ages for nodes closer to the tips.
+For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka \enquote{closer to the root}) relative to the original estimate, and overestimate ages for nodes closer to the tips.
 Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. S1).
@@
 \begin{center}
-\includegraphics{../figures/figure-cross-validation/fig-cross-validation-xy-plots.pdf}
+\includegraphics{../figures/fringillidae-cross-validation/fig-cross-validation-xy-plots.pdf}
 \end{center}
@@
 
-Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used Magallón et al. (2015).
+Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used (Magallón, 2010, @magallon2015metacalibrated).
 Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
@@
 Hence, DateLife's workflow features different ways of estimating node ages in the absence of calibrations and branch length information for certain taxa.
-``Making up'' branch lengths is a common practice in scientific publications: Jetz et al.
+\enquote{Making up} branch lengths is a common practice in scientific publications: Jetz et al.
 (2012), created a chronogram of all 9, 993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-alstrom2014discovery}{}}%
+\leavevmode\hypertarget{ref-alstrom2014discovery}{}%
 Alström, P., Hooper, D.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ane2009groves}{}}%
-Ané, C., Eulenstein, O., Piaggio-Talice, R., \& Sanderson, M. J. (2009). Groves of phylogenetic trees. \emph{{Annals of Combinatorics}}, \emph{13}(2), 139--167.
-
-\leavevmode\vadjust pre{\hypertarget{ref-antonelli2017supersmart}{}}%
+\leavevmode\hypertarget{ref-ane2009groves}{}%
+Ané, C., Eulenstein, O., Piaggio-Talice, R., \& Sanderson, M. J. (2009). Groves of phylogenetic trees. \emph{Annals of Combinatorics}, \emph{13}(2), 139--167.
+
+\leavevmode\hypertarget{ref-antonelli2017supersmart}{}%
 Antonelli, A., Hettling, H., Condamine, F.
@@
 (2017).
-{Toward a self-updating platform for estimating rates of speciation and migration, ages, and relationships of Taxa}.
+Toward a self-updating platform for estimating rates of speciation and migration, ages, and relationships of Taxa.
 \emph{Systematic Biology}, \emph{66}(2), 153--166.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-archie1986newick}{}}%
+\leavevmode\hypertarget{ref-archie1986newick}{}%
 Archie, J., Day, W.
@@
 (1986).
-{The Newick tree format}.
-Retrieved from \url{\%7Bhttps://evolution.genetics.washington.edu/phylip/newicktree.html\%7D}
-
-\leavevmode\vadjust pre{\hypertarget{ref-avibase-emberiza}{}}%
+The Newick tree format.
+Retrieved from \href{\%7Bhttps://evolution.genetics.washington.edu/phylip/newicktree.html\%7D}{\{https://evolution.genetics.washington.edu/phylip/newicktree.html\}}
+
+\leavevmode\hypertarget{ref-avibase-emberiza}{}%
 Avibase.
 (2022).
-{Yellow-throated Bunting}.
-\emph{{Avibase - The World Bird Database}}, (Online Resource).
-Retrieved from \url{\%7Bhttps://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\%7D}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Bapst2012a}{}}%
+Yellow-throated Bunting.
+\emph{Avibase - the World Bird Database}, (Online Resource).
+Retrieved from \href{\%7Bhttps://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\%7D}{\{https://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\}}
+
+\leavevmode\hypertarget{ref-Bapst2012a}{}%
 Bapst, D.
@@
 (2012).
-{Paleotree: An R package for paleontological and phylogenetic analyses of evolution}.
-\emph{{Methods in Ecology and Evolution}}, \emph{3}(5), 803--807.
+Paleotree: An R package for paleontological and phylogenetic analyses of evolution.
+\emph{Methods in Ecology and Evolution}, \emph{3}(5), 803--807.
 \url{https://doi.org/10.1111/j.2041-210X.2012.00223.x}
 
-\leavevmode\vadjust pre{\hypertarget{ref-barba2018constraining}{}}%
+\leavevmode\hypertarget{ref-barba2018constraining}{}%
 Barba-Montoya, J., Reis, M.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-barker2014mitogenomic}{}}%
+\leavevmode\hypertarget{ref-barker2014mitogenomic}{}%
 Barker, F. K. (2014). Mitogenomic data resolve basal relationships among passeriform and passeridan birds. \emph{Molecular Phylogenetics and Evolution}, \emph{79}, 313--324.
 
-\leavevmode\vadjust pre{\hypertarget{ref-barker2013going}{}}%
+\leavevmode\hypertarget{ref-barker2013going}{}%
 Barker, F.
@@
 Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds.
-\emph{{Systematic Biology}}, \emph{62}(2), 298--320.
-
-\leavevmode\vadjust pre{\hypertarget{ref-barker2015new}{}}%
+\emph{Systematic Biology}, \emph{62}(2), 298--320.
+
+\leavevmode\hypertarget{ref-barker2015new}{}%
 Barker, F.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-barker2004phylogeny}{}}%
+\leavevmode\hypertarget{ref-barker2004phylogeny}{}%
 Barker, F.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-beresford2005african}{}}%
-Beresford, P., Barker, F., Ryan, Pg., \& Crowe, T.
+\leavevmode\hypertarget{ref-beresford2005african}{}%
+Beresford, P., Barker, F., Ryan, P., \& Crowe, T.
 (2005).
-African endemics span the tree of songbirds (passeri): Molecular systematics of several evolutionary {``enigmas.''} \emph{Proceedings of the Royal Society B: Biological Sciences}, \emph{272}(1565), 849--858.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Boyle2013}{}}%
+African endemics span the tree of songbirds (passeri): Molecular systematics of several evolutionary ``enigmas''.
+\emph{Proceedings of the Royal Society B: Biological Sciences}, \emph{272}(1565), 849--858.
+
+\leavevmode\hypertarget{ref-Boyle2013}{}%
 Boyle, B., Hopkins, N., Lu, Z., Raygoza Garay, J.
@@
 (2013).
-{The taxonomic name resolution service: An online tool for automated standardization of plant names}.
-\emph{{BMC Bioinformatics}}, \emph{14}(1).
+The taxonomic name resolution service: An online tool for automated standardization of plant names.
+\emph{BMC Bioinformatics}, \emph{14}(1).
 \url{https://doi.org/10.1186/1471-2105-14-16}
 
-\leavevmode\vadjust pre{\hypertarget{ref-Britton2007}{}}%
+\leavevmode\hypertarget{ref-Britton2007}{}%
 Britton, T., Anderson, C.
@@
 (2007).
-{Estimating Divergence Times in Large Phylogenetic Trees}.
-\emph{{Systematic Biology}}, \emph{56}(788777878), 741--752.
+Estimating Divergence Times in Large Phylogenetic Trees.
+\emph{Systematic Biology}, \emph{56}(788777878), 741--752.
 \url{https://doi.org/10.1080/10635150701613783}
 
-\leavevmode\vadjust pre{\hypertarget{ref-bryson2014diversification}{}}%
+\leavevmode\hypertarget{ref-bryson2014diversification}{}%
 Bryson Jr, R.
@@
 (2014).
-Diversification across the new world within the `blue'cardinalids (aves: cardinalidae).
+Diversification across the new world within the `blue'cardinalids (aves: Cardinalidae).
 \emph{Journal of Biogeography}, \emph{41}(3), 587--599.
 
-\leavevmode\vadjust pre{\hypertarget{ref-burleigh2015building}{}}%
+\leavevmode\hypertarget{ref-burleigh2015building}{}%
 Burleigh, J. G., Kimball, R. T., \& Braun, E. L. (2015). Building the avian tree of life using a large-scale, sparse supermatrix. \emph{Molecular Phylogenetics and Evolution}, \emph{84}, 53--63.
 
-\leavevmode\vadjust pre{\hypertarget{ref-burns2014phylogenetics}{}}%
+\leavevmode\hypertarget{ref-burns2014phylogenetics}{}%
 Burns, K.
@@
 Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds.
-\emph{{Molecular Phylogenetics and Evolution}}, \emph{75}, 41--77.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Chamberlain2013}{}}%
+\emph{Molecular Phylogenetics and Evolution}, \emph{75}, 41--77.
+
+\leavevmode\hypertarget{ref-Chamberlain2018}{}%
+Chamberlain, S. (2018). \emph{bold: Interface to Bold Systems API}. Retrieved from \url{https://CRAN.R-project.org/package=bold}
+
+\leavevmode\hypertarget{ref-Chamberlain2013}{}%
 Chamberlain, S.
@@
 (2013).
-{taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}}.
-\emph{{F1000Research}}, \emph{2}(191), 1--29.
+taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}.
+\emph{F1000Research}, \emph{2}(191), 1--29.
 \url{https://doi.org/10.12688/f1000research.2-191.v2}
 
-\leavevmode\vadjust pre{\hypertarget{ref-Chamberlain2018}{}}%
-Chamberlain, S.
-A., Szöcs, E., Foster, Z., Arendsee, Z., Boettiger, C., Ram, K., \ldots{} Li, G.
-(2019).
-\emph{{taxize: Taxonomic information from around the web}}.
-Retrieved from \url{https://github.com/ropensci/taxize}
-
-\leavevmode\vadjust pre{\hypertarget{ref-chaves2013biogeography}{}}%
+\leavevmode\hypertarget{ref-chaves2013biogeography}{}%
 Chaves, J.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-claramunt2015new}{}}%
-Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{{Science Advances}}, \emph{1}(11), e1501005.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Criscuolo2006}{}}%
+\leavevmode\hypertarget{ref-claramunt2015new}{}%
+Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{Science Advances}, \emph{1}(11), e1501005.
+
+\leavevmode\hypertarget{ref-Criscuolo2006}{}%
 Criscuolo, A., Berry, V., Douzery, E.
-J.
-P., \& Gascuel, O.
+J., \& Gascuel, O.
 (2006).
-{SDM: A fast distance-based approach for (super)tree building in phylogenomics}.
-\emph{{Systematic Biology}}, \emph{55}(5), 740--755.
+SDM: A fast distance-based approach for (super)tree building in phylogenomics.
+\emph{Systematic Biology}, \emph{55}(5), 740--755.
 \url{https://doi.org/10.1080/10635150600969872}
 
-\leavevmode\vadjust pre{\hypertarget{ref-delsuc2018phylogenomic}{}}%
+\leavevmode\hypertarget{ref-delsuc2018phylogenomic}{}%
 Delsuc, F., Philippe, H., Tsagkogeorga, G., Simion, P., Tilak, M.-K., Turon, X., \ldots{} Douzery, E.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Eastman2013}{}}%
+\leavevmode\hypertarget{ref-Eastman2013}{}%
 Eastman, J.
@@
 (2013).
-{Congruification: Support for time scaling large phylogenetic trees}.
-\emph{{Methods in Ecology and Evolution}}, \emph{4}(7), 688--691.
+Congruification: Support for time scaling large phylogenetic trees.
+\emph{Methods in Ecology and Evolution}, \emph{4}(7), 688--691.
 \url{https://doi.org/10.1111/2041-210X.12051}
 
-\leavevmode\vadjust pre{\hypertarget{ref-edgar2004muscle}{}}%
+\leavevmode\hypertarget{ref-edgar2004muscle}{}%
 Edgar, R. C. (2004). MUSCLE: Multiple sequence alignment with high accuracy and high throughput. \emph{Nucleic Acids Research}, \emph{32}(5), 1792--1797.
 
-\leavevmode\vadjust pre{\hypertarget{ref-Felsenstein1985a}{}}%
-Felsenstein, J. (1985). {Phylogenies and the Comparative Method}. \emph{The American Naturalist}, \emph{125}(1), 1--15. Retrieved from \url{http://www.jstor.org/stable/2461605}
-
-\leavevmode\vadjust pre{\hypertarget{ref-gbif2022taxonomy}{}}%
+\leavevmode\hypertarget{ref-Felsenstein1985a}{}%
+Felsenstein, J. (1985). Phylogenies and the Comparative Method. \emph{The American Naturalist}, \emph{125}(1), 1--15. Retrieved from \url{http://www.jstor.org/stable/2461605}
+
+\leavevmode\hypertarget{ref-gbif2022taxonomy}{}%
 GBIF Secretariat.
 (2022).
-{GBIF Backbone Taxonomy}.
-\emph{{Checklist Dataset}}, (Online Resource accessed via GBIF.org).
+GBIF Backbone Taxonomy.
+\emph{Checklist dataset}, (Online Resource accessed via GBIF.org).
 Retrieved from \href{\%7Bhttps://doi.org/10.15468/39omei\%20\%7D}{\{https://doi.org/10.15468/39omei \}}
 
-\leavevmode\vadjust pre{\hypertarget{ref-gibb2015new}{}}%
+\leavevmode\hypertarget{ref-gibb2015new}{}%
 Gibb, G.
@@
 New zealand passerines help clarify the diversification of major songbird lineages during the oligocene.
-\emph{{Genome Biology and Evolution}}, \emph{7}(11), 2983--2995.
-
-\leavevmode\vadjust pre{\hypertarget{ref-graur2004reading}{}}%
+\emph{Genome Biology and Evolution}, \emph{7}(11), 2983--2995.
+
+\leavevmode\hypertarget{ref-graur2004reading}{}%
 Graur, D., \& Martin, W. (2004). Reading the entrails of chickens: Molecular timescales of evolution and the illusion of precision. \emph{TRENDS in Genetics}, \emph{20}(2), 80--86.
 
-\leavevmode\vadjust pre{\hypertarget{ref-hackett2008phylogenomic}{}}%
+\leavevmode\hypertarget{ref-hackett2008phylogenomic}{}%
 Hackett, S.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Hedges2015}{}}%
+\leavevmode\hypertarget{ref-Hedges2015}{}%
 Hedges, S.
@@
 (2015).
-{Tree of life reveals clock-like speciation and diversification}.
-\emph{{Molecular Biology and Evolution}}, \emph{32}(4), 835--845.
+Tree of life reveals clock-like speciation and diversification.
+\emph{Molecular Biology and Evolution}, \emph{32}(4), 835--845.
 \url{https://doi.org/10.1093/molbev/msv037}
 
-\leavevmode\vadjust pre{\hypertarget{ref-Heibl2008}{}}%
+\leavevmode\hypertarget{ref-Heibl2008}{}%
 Heibl, C. (2008). \emph{PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages.} Retrieved from \url{http://www.christophheibl.de/Rpackages.html}
 
-\leavevmode\vadjust pre{\hypertarget{ref-hooper2017chromosomal}{}}%
+\leavevmode\hypertarget{ref-hooper2017chromosomal}{}%
 Hooper, D. M., \& Price, T. D. (2017). Chromosomal inversion differences correlate with range overlap in passerine birds. \emph{Nature Ecology \& Evolution}, \emph{1}(10), 1526.
 
-\leavevmode\vadjust pre{\hypertarget{ref-Huelsenbeck2001}{}}%
-Huelsenbeck, J. P., \& Ronquist, F. (2001). {MRBAYES: Bayesian inference of phylogenetic trees}. \emph{Bioinformatics}, \emph{17}(8), 754--755. \url{https://doi.org/10.1093/bioinformatics/17.8.754}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Jetz2012}{}}%
+\leavevmode\hypertarget{ref-Huelsenbeck2001}{}%
+Huelsenbeck, J. P., \& Ronquist, F. (2001). MRBAYES: Bayesian inference of phylogenetic trees. \emph{Bioinformatics}, \emph{17}(8), 754--755. \url{https://doi.org/10.1093/bioinformatics/17.8.754}
+
+\leavevmode\hypertarget{ref-Jetz2012}{}%
 Jetz, W., Thomas, G., Joy, J.
-J.
-B., Hartmann, K., \& Mooers, A.
+J., Hartmann, K., \& Mooers, A.
 (2012).
-{The global diversity of birds in space and time}.
+The global diversity of birds in space and time.
 \emph{Nature}, \emph{491}(7424), 444--448.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-johansson2008phylogenetic}{}}%
+\leavevmode\hypertarget{ref-johansson2008phylogenetic}{}%
 Johansson, U.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-katoh2009multiple}{}}%
-Katoh, K., Asimenos, G., \& Toh, H. (2009). Multiple alignment of DNA sequences with MAFFT. In \emph{Bioinformatics for DNA sequence analysis} (pp. 39--64). Springer.
-
-\leavevmode\vadjust pre{\hypertarget{ref-kimball2019phylogenomic}{}}%
+\leavevmode\hypertarget{ref-katoh2009multiple}{}%
+Katoh, K., Asimenos, G., \& Toh, H. (2009). Multiple alignment of dna sequences with mafft. In \emph{Bioinformatics for dna sequence analysis} (pp. 39--64). Springer.
+
+\leavevmode\hypertarget{ref-kimball2019phylogenomic}{}%
 Kimball, R. T., Oliveros, C. H., Wang, N., White, N. D., Barker, F. K., Field, D. J., \ldots{} others. (2019). A phylogenomic supertree of birds. \emph{Diversity}, \emph{11}(7), 109.
 
-\leavevmode\vadjust pre{\hypertarget{ref-klicka2014comprehensive}{}}%
+\leavevmode\hypertarget{ref-klicka2014comprehensive}{}%
 Klicka, J., Barker, F.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ksepka2015fossil}{}}%
+\leavevmode\hypertarget{ref-ksepka2015fossil}{}%
 Ksepka, D.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-lamichhaney2015evolution}{}}%
+\leavevmode\hypertarget{ref-lamichhaney2015evolution}{}%
 Lamichhaney, S., Berglund, J., Almén, M.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-laubichler2009form}{}}%
+\leavevmode\hypertarget{ref-laubichler2009form}{}%
 Laubichler, M. D., \& Maienschein, J. (2009). \emph{Form and function in developmental evolution}. Cambridge University Press.
 
-\leavevmode\vadjust pre{\hypertarget{ref-lepage2004avibase}{}}%
+\leavevmode\hypertarget{ref-lepage2004avibase}{}%
 Lepage, D. (2004). \emph{Avibase: The world bird database}. Bird Studies Canada.
 
-\leavevmode\vadjust pre{\hypertarget{ref-lepage2014avibase}{}}%
+\leavevmode\hypertarget{ref-lepage2014avibase}{}%
 Lepage, D., Vaidya, G., \& Guralnick, R. (2014). Avibase--a database system for managing and organizing taxonomic concepts. \emph{ZooKeys}, (420), 117.
 
-\leavevmode\vadjust pre{\hypertarget{ref-lerner2011multilocus}{}}%
+\leavevmode\hypertarget{ref-lerner2011multilocus}{}%
 Lerner, H.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-lovette2010comprehensive}{}}%
+\leavevmode\hypertarget{ref-lovette2010comprehensive}{}%
 Lovette, I.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-magallon2001absolute}{}}%
+\leavevmode\hypertarget{ref-magallon2001absolute}{}%
 Magallon, S., \& Sanderson, M. (2001). Absolute diversification rates in angiosperm clades. \emph{Evolution}, \emph{55}(9), 1762--1780.
 
-\leavevmode\vadjust pre{\hypertarget{ref-magallon2010using}{}}%
+\leavevmode\hypertarget{ref-magallon2010using}{}%
 Magallón, S. (2010). Using fossils to break long branches in molecular dating: A comparison of relaxed clocks applied to the origin of angiosperms. \emph{Systematic Biology}, \emph{59}(4), 384--399.
 
-\leavevmode\vadjust pre{\hypertarget{ref-magallon2015metacalibrated}{}}%
+\leavevmode\hypertarget{ref-magallon2015metacalibrated}{}%
 Magallón, S., Gómez-Acevedo, S., Sánchez-Reyes, L.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-mctavish2015phylesystem}{}}%
+\leavevmode\hypertarget{ref-mctavish2015phylesystem}{}%
 McTavish, E.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Michonneau2016}{}}%
+\leavevmode\hypertarget{ref-Michonneau2016}{}%
 Michonneau, F., Brown, J.
@@
 (2016).
-{rotl: an R package to interact with the Open Tree of Life data}.
-\emph{{Methods in Ecology and Evolution}}, \emph{7}(12), 1476--1481.
+rotl: an R package to interact with the Open Tree of Life data.
+\emph{Methods in Ecology and Evolution}, \emph{7}(12), 1476--1481.
 \url{https://doi.org/10.1111/2041-210X.12593}
 
-\leavevmode\vadjust pre{\hypertarget{ref-Morlon2014}{}}%
-Morlon, H. (2014). {Phylogenetic approaches for studying diversification.} \emph{{Ecology Letters}}, \emph{17}(4), 508--525. \url{https://doi.org/10.1111/ele.12251}
-
-\leavevmode\vadjust pre{\hypertarget{ref-moyle2016tectonic}{}}%
+\leavevmode\hypertarget{ref-Morlon2014}{}%
+Morlon, H. (2014). Phylogenetic approaches for studying diversification. \emph{Ecology Letters}, \emph{17}(4), 508--525. \url{https://doi.org/10.1111/ele.12251}
+
+\leavevmode\hypertarget{ref-moyle2016tectonic}{}%
 Moyle, R.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-odeen2011evolution}{}}%
-Ödeen, A., Håstad, O., \& Alström, P. (2011). Evolution of ultraviolet vision in the largest avian radiation-the passerines. \emph{BMC Evolutionary Biology}, \emph{11}(1), 1--8.
-
-\leavevmode\vadjust pre{\hypertarget{ref-oliveros2019earth}{}}%
+\leavevmode\hypertarget{ref-oliveros2019earth}{}%
 Oliveros, C.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Ooms2018}{}}%
+\leavevmode\hypertarget{ref-Ooms2018}{}%
 Ooms, J., \& Chamberlain, S. (2018). \emph{Phylocomr: Interface to 'phylocom'}. Retrieved from \url{https://CRAN.R-project.org/package=phylocomr}
 
-\leavevmode\vadjust pre{\hypertarget{ref-opentreeAPIs}{}}%
+\leavevmode\hypertarget{ref-opentreeAPIs}{}%
 Open Tree Of Life, Redelings, B., Cranston, K.
@@
 (2016).
-{Open Tree of Life APIs v3.0}.
-\emph{{Open Tree of Life Project}}, (Online Resources).
-Retrieved from \url{\%7Bhttps://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\%7D}
-
-\leavevmode\vadjust pre{\hypertarget{ref-opentreeoflife2019synth}{}}%
+Open Tree of Life APIs v3.0.
+\emph{Open Tree of Life Project}, (Online Resources).
+Retrieved from \href{\%7Bhttps://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\%7D}{\{https://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\}}
+
+\leavevmode\hypertarget{ref-opentreeoflife2019synth}{}%
 Open Tree Of Life, Redelings, B., Sánchez Reyes, L.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-packert2012horizontal}{}}%
-Päckert, M., Martens, J., Sun, Y.-H., Severinghaus, L.
-L., Nazarenko, A.
-A., Ting, J., \ldots{} Tietze, D.
-T.
-(2012).
-Horizontal and elevational phylogeographic patterns of himalayan and southeast asian forest passerines (aves: passeriformes).
-\emph{Journal of Biogeography}, \emph{39}(3), 556--573.
-
-\leavevmode\vadjust pre{\hypertarget{ref-paradis2004}{}}%
-Paradis, E., Claude, J., \& Strimmer, K. (2004). {APE: analyses of phylogenetics and evolution in R language}. \emph{Bioinformatics}, \emph{20}(2), 289--290.
-
-\leavevmode\vadjust pre{\hypertarget{ref-parchman2007coevolution}{}}%
+\leavevmode\hypertarget{ref-odeen2011evolution}{}%
+Ödeen, A., Håstad, O., \& Alström, P. (2011). Evolution of ultraviolet vision in the largest avian radiation-the passerines. \emph{BMC Evolutionary Biology}, \emph{11}(1), 1--8.
+
+\leavevmode\hypertarget{ref-paradis2004}{}%
+Paradis, E., Claude, J., \& Strimmer, K. (2004). APE: analyses of phylogenetics and evolution in R language. \emph{Bioinformatics}, \emph{20}(2), 289--290.
+
+\leavevmode\hypertarget{ref-parchman2007coevolution}{}%
 Parchman, T.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-pennell2014geiger}{}}%
+\leavevmode\hypertarget{ref-packert2012horizontal}{}%
+Päckert, M., Martens, J., Sun, Y.-H., Severinghaus, L.
+L., Nazarenko, A.
+A., Ting, J., \ldots{} Tietze, D.
+T.
+(2012).
+Horizontal and elevational phylogeographic patterns of himalayan and southeast asian forest passerines (aves: Passeriformes).
+\emph{Journal of Biogeography}, \emph{39}(3), 556--573.
+
+\leavevmode\hypertarget{ref-pennell2014geiger}{}%
 Pennell, M.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-posadas2006historical}{}}%
-Posadas, P., Crisci, J. V., \& Katinas, L. (2006). Historical biogeography: A review of its basic concepts and critical issues. \emph{{Journal of Arid Environments}}, \emph{66}(3), 389--403.
-
-\leavevmode\vadjust pre{\hypertarget{ref-powell2014comprehensive}{}}%
+\leavevmode\hypertarget{ref-posadas2006historical}{}%
+Posadas, P., Crisci, J. V., \& Katinas, L. (2006). Historical biogeography: A review of its basic concepts and critical issues. \emph{Journal of Arid Environments}, \emph{66}(3), 389--403.
+
+\leavevmode\hypertarget{ref-powell2014comprehensive}{}%
 Powell, A.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-price2014niche}{}}%
+\leavevmode\hypertarget{ref-price2014niche}{}%
 Price, T.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-pulgarin2013multilocus}{}}%
+\leavevmode\hypertarget{ref-pulgarin2013multilocus}{}%
 Pulgarín-R, P.
@@
 (2013).
-Multilocus phylogeny and biogeography of the new world pheucticus grosbeaks (aves: cardinalidae).
+Multilocus phylogeny and biogeography of the new world pheucticus grosbeaks (aves: Cardinalidae).
 \emph{Molecular Phylogenetics and Evolution}, \emph{69}(3), 1222--1227.
 
-\leavevmode\vadjust pre{\hypertarget{ref-RCoreTeam2018}{}}%
-R Core Team. (2018). \emph{{R: a language and environment for statistical computing}}. Vienna, Austria: R Foundation for Statistical Computing.
-
-\leavevmode\vadjust pre{\hypertarget{ref-rabosky2018inverse}{}}%
+\leavevmode\hypertarget{ref-rabosky2018inverse}{}%
 Rabosky, D.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ramshaw1972time}{}}%
+\leavevmode\hypertarget{ref-ramshaw1972time}{}%
 Ramshaw, J., Richardson, D., Meatyard, B., Brown, R., Richardson, M., Thompson, E., \& Boulter, D.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ratnasingham2007bold}{}}%
-Ratnasingham, S., \& Hebert, P. D. (2007). BOLD: The barcode of life data system (http://www. Barcodinglife. org). \emph{Molecular Ecology Notes}, \emph{7}(3), 355--364.
-
-\leavevmode\vadjust pre{\hypertarget{ref-rees2017automated}{}}%
+\leavevmode\hypertarget{ref-ratnasingham2007bold}{}%
+Ratnasingham, S., \& Hebert, P. D. (2007). BOLD: The barcode of life data system (http://www. Barcodinglife. Org). \emph{Molecular Ecology Notes}, \emph{7}(3), 355--364.
+
+\leavevmode\hypertarget{ref-RCoreTeam2018}{}%
+R Core Team. (2018). \emph{R: a language and environment for statistical computing}. Vienna, Austria: R Foundation for Statistical Computing.
+
+\leavevmode\hypertarget{ref-rees2017automated}{}%
 Rees, \& Cranston, K. (2017). Automated assembly of a reference taxonomy for phylogenetic data synthesis. \emph{Biodiversity Data Journal}, (5).
 
-\leavevmode\vadjust pre{\hypertarget{ref-rees2017irmng}{}}%
-Rees, Vandepitte, L., Decock, W., \& Vanhoorne, B. (2017). {IRMNG 2006--2016: 10 Years of a Global Taxonomic Database}. \emph{Biodiversity Informatics}, \emph{12}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Revell2012}{}}%
-Revell, L. J. (2012). Phytools: An r package for phylogenetic comparative biology (and other things). \emph{{Methods in Ecology and Evolution}}, \emph{3}, 217--223.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Ronquist2003}{}}%
+\leavevmode\hypertarget{ref-rees2017irmng}{}%
+Rees, Vandepitte, L., Decock, W., \& Vanhoorne, B. (2017). IRMNG 2006--2016: 10 Years of a Global Taxonomic Database. \emph{Biodiversity Informatics}, \emph{12}.
+
+\leavevmode\hypertarget{ref-Revell2012}{}%
+Revell, L. J. (2012). Phytools: An r package for phylogenetic comparative biology (and other things). \emph{Methods in Ecology and Evolution}, \emph{3}, 217--223.
+
+\leavevmode\hypertarget{ref-Ronquist2003}{}%
 Ronquist, F., \& Huelsenbeck, J.
@@
 (2003).
-{MrBayes 3: Bayesian phylogenetic inference under mixed models}.
+MrBayes 3: Bayesian phylogenetic inference under mixed models.
 \emph{Bioinformatics}, \emph{19}(12), 1572--1574.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-datelifeplot}{}}%
+\leavevmode\hypertarget{ref-datelifeplot}{}%
 Sanchez-Reyes, L.
@@
 (2022).
-{\texttt{datelifeplot}: Methods to plot chronograms and outputs of the \texttt{datelife} package}.
+\texttt{datelifeplot}: Methods to plot chronograms and outputs of the \texttt{datelife} package.
 \emph{R Package Release V0.2.2}.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-datelifeR}{}}%
+\leavevmode\hypertarget{ref-datelifeR}{}%
 Sanchez-Reyes, L.
@@
 (2022).
-{\texttt{datelife}: Scientific Data on Time of Lineage Divergence for Your Taxa}.
+\texttt{datelife}: Scientific Data on Time of Lineage Divergence for Your Taxa.
 \emph{R Package Release V0.6.2}.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-sanderson2002}{}}%
+\leavevmode\hypertarget{ref-sanderson2002}{}%
 Sanderson, M.
-J.
 (2002).
-{Estimating Absolute Rates of Molecular Evolution and Divergence Times: A Penalized Likelihood Approach}.
+Estimating Absolute Rates of Molecular Evolution and Divergence Times: A Penalized Likelihood Approach.
 \emph{Molecular Biology and Evolution}, \emph{19}(1), 101--109.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-sanderson2003r8s}{}}%
-Sanderson, M. J. (2003). {r8s: Inferring Absolute Rates of Molecular Evolution and Divergence Times in the Absence of a Molecular Clock}. \emph{Bioinformatics}, \emph{19}(2), 301--302.
-
-\leavevmode\vadjust pre{\hypertarget{ref-sanderson2001sources}{}}%
-Sanderson, M.
-J., \& Doyle, J.
-A.
+\leavevmode\hypertarget{ref-sanderson2003r8s}{}%
+Sanderson, M. (2003). r8s: Inferring Absolute Rates of Molecular Evolution and Divergence Times in the Absence of a Molecular Clock. \emph{Bioinformatics}, \emph{19}(2), 301--302.
+
+\leavevmode\hypertarget{ref-sanderson2001sources}{}%
+Sanderson, M., \& Doyle, J.
 (2001).
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-sauquet2013practical}{}}%
-Sauquet, H. (2013). A practical guide to molecular dating. \emph{{Comptes Rendus Palevol}}, \emph{12}(6), 355--367.
-
-\leavevmode\vadjust pre{\hypertarget{ref-sauquet2021age}{}}%
+\leavevmode\hypertarget{ref-sauquet2013practical}{}%
+Sauquet, H. (2013). A practical guide to molecular dating. \emph{Comptes Rendus Palevol}, \emph{12}(6), 355--367.
+
+\leavevmode\hypertarget{ref-sauquet2021age}{}%
 Sauquet, H., Ramírez-Barahona, S., \& Magallón, S. (2021). \emph{The age of flowering plants is unknown}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-schenk2016sec}{}}%
-Schenk, J. J. (2016). {Consequences of secondary calibrations on divergence time estimates}. \emph{PLoS ONE}, \emph{11}(1). \url{https://doi.org/10.1371/journal.pone.0148228}
-
-\leavevmode\vadjust pre{\hypertarget{ref-schliep2011phangorn}{}}%
+\leavevmode\hypertarget{ref-schenk2016sec}{}%
+Schenk, J. J. (2016). Consequences of secondary calibrations on divergence time estimates. \emph{PLoS ONE}, \emph{11}(1). \url{https://doi.org/10.1371/journal.pone.0148228}
+
+\leavevmode\hypertarget{ref-schliep2011phangorn}{}%
 Schliep, K. P. (2011). Phangorn: Phylogenetic analysis in r. \emph{Bioinformatics}, \emph{27}(4), 592--593.
 
-\leavevmode\vadjust pre{\hypertarget{ref-schoch2020ncbi}{}}%
+\leavevmode\hypertarget{ref-schoch2020ncbi}{}%
 Schoch, C.
@@
 (2020).
-{NCBI Taxonomy: a Comprehensive Update on Curation, Resources and Tools}.
+NCBI Taxonomy: a Comprehensive Update on Curation, Resources and Tools.
 \emph{Database}, \emph{2020}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-selvatti2015paleogene}{}}%
+\leavevmode\hypertarget{ref-selvatti2015paleogene}{}%
 Selvatti, A.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-smith2018constructing}{}}%
+\leavevmode\hypertarget{ref-smith2018constructing}{}%
 Smith, S., \& Brown, J. (2018). Constructing a broadly inclusive seed plant phylogeny. \emph{American Journal of Botany}, \emph{105}(3), 302--314.
 
-\leavevmode\vadjust pre{\hypertarget{ref-Smith2012}{}}%
+\leavevmode\hypertarget{ref-Smith2012}{}%
 Smith, S., \& O'Meara, B.
 (2012).
-{TreePL: Divergence time estimation using penalized likelihood for large phylogenies}.
+TreePL: Divergence time estimation using penalized likelihood for large phylogenies.
 \emph{Bioinformatics}, \emph{28}(20), 2689--2690.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Stoltzfus2013}{}}%
+\leavevmode\hypertarget{ref-Stoltzfus2013}{}%
 Stoltzfus, A., Lapp, H., Matasci, N., Deus, H., Sidlauskas, B., Zmasek, C.
@@
 (2013).
-{Phylotastic!
-Making tree-of-life knowledge accessible, reusable and convenient}.
-\emph{{BMC Bioinformatics}}, \emph{14}.
+Phylotastic!
+Making tree-of-life knowledge accessible, reusable and convenient.
+\emph{BMC Bioinformatics}, \emph{14}.
 \url{https://doi.org/10.1186/1471-2105-14-158}
 
-\leavevmode\vadjust pre{\hypertarget{ref-tietze2013complete}{}}%
+\leavevmode\hypertarget{ref-tietze2013complete}{}%
 Tietze, D.
@@
 (2013).
-Complete phylogeny and historical biogeography of true rosefinches (aves: carpodacus).
+Complete phylogeny and historical biogeography of true rosefinches (aves: Carpodacus).
 \emph{Zoological Journal of the Linnean Society}, \emph{169}(1), 215--234.
 
-\leavevmode\vadjust pre{\hypertarget{ref-title2016macrophylogenies}{}}%
+\leavevmode\hypertarget{ref-title2016macrophylogenies}{}%
 Title, P.
@@
 (2016).
-{Do Macrophylogenies Yield Stable Macroevolutionary Inferences?
-An Example from Squamate Reptiles}.
+Do Macrophylogenies Yield Stable Macroevolutionary Inferences?
+An Example from Squamate Reptiles.
 \emph{Systematic Biology}, syw102.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-treplin2008molecular}{}}%
+\leavevmode\hypertarget{ref-treplin2008molecular}{}%
 Treplin, S., Siegert, R., Bleidorn, C., Thompson, H.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-vos2012nexml}{}}%
+\leavevmode\hypertarget{ref-vos2012nexml}{}%
 Vos, R.
@@
 \emph{Systematic Biology}, \emph{61}(4), 675--689.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Webb2000}{}}%
-Webb, C. (2000). {Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees}. \emph{{The American Naturalist}}, \emph{156}(2), 145--155.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Webb2008}{}}%
+\url{https://doi.org/10.1093/sysbio/sys025}
+
+\leavevmode\hypertarget{ref-Webb2000}{}%
+Webb, C. (2000). Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees. \emph{The American Naturalist}, \emph{156}(2), 145--155.
+
+\leavevmode\hypertarget{ref-Webb2008}{}%
 Webb, C., Ackerly, D., \& Kembel, S.
 (2008).
-{Phylocom: Software for the analysis of phylogenetic community structure and trait evolution}.
+Phylocom: Software for the analysis of phylogenetic community structure and trait evolution.
 \emph{Bioinformatics}, \emph{24}(18), 2098--2100.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-webb2005phylomatic}{}}%
-Webb, C., \& Donoghue, M. (2005). Phylomatic: Tree assembly for applied phylogenetics. \emph{{Molecular Ecology Notes}}, \emph{5}(1), 181--183.
-
-\leavevmode\vadjust pre{\hypertarget{ref-weir2008calibrating}{}}%
+\leavevmode\hypertarget{ref-webb2005phylomatic}{}%
+Webb, C., \& Donoghue, M. (2005). Phylomatic: Tree assembly for applied phylogenetics. \emph{Molecular Ecology Notes}, \emph{5}(1), 181--183.
+
+\leavevmode\hypertarget{ref-weir2008calibrating}{}%
 Weir, J., \& Schluter, D. (2008). Calibrating the avian molecular clock. \emph{Molecular Ecology}, \emph{17}(10), 2321--2328.
 
-\leavevmode\vadjust pre{\hypertarget{ref-zuccon2012phylogenetic}{}}%
+\leavevmode\hypertarget{ref-zuccon2012phylogenetic}{}%
 Zuccon, D., Prŷs-Jones, R., Rasmussen, P.
@@
 
-\end{CSLReferences}
-
 \endgroup

### 2023-01-14 manuscript/fig-workflow.Rmd
updating path to figure

@@
 \begin{minipage}[t]{17.3cm}
-\includegraphics{../figures/figure1/figure1-new.pdf}
+\includegraphics{../figures/figure-workflow/figure1-new.pdf}
 \captionof{figure}{\textit{Main DateLife workflow.
 Analyses can be performed via DateLife's interactive website at \url{www.datelife.org}, or using the \texttt{datelife} R package.
-Details on the R functions used to perform the analyses are available from \texttt{datelife}'s R package vignettes at \url{phylotastic.org/datelife}.}}
+Details on the R functions used to perform the analyses are available from \texttt{datelife}'s R package vignettes at \url{https://phylotastic.org/datelife}.}}
 \label{fig:workflow}

### 2023-01-14 manuscript/fig-fringillidae-full-example.Rmd
updating path to figure

@@
 
-\captionof{figure}{Fringillidae median summary chronogram generated with DateLife. It has 256 tips and 233 nodes, from which 212 have age data from at least one published chronogram.}
+\captionof{figure}{Fringillidae median summary chronogram generated with DateLife.
+It has 289 tips and 253 nodes, from which 212 have age data from at least one published chronogram.
+These nodes have age data points from 19 different chronograms from 13 different studies (colored circles), which were summarized into 194 data points that were used as secondary calibrations (purple slashes) with the Branch Length Adjuster (BLADJ ) software from Webb et al., (\protect\hyperlink{ref-Webb2008}{2008}).}
 \label{fig:fringillidages}

### 2023-01-19 manuscript/manuscript.Rmd
updating new submission

@@
 availability <- ifelse(rpackage_available, "is available", "will be available")
-load("../data/dquery.rda")
-load("../data/dres.rda")
+load(file = "../data/fringillidae_v2022.01.28.rda")
+# to write to text the references of the studies that have at least two Fringillidae species:
+write(unique(names(fringillidae_v2022.01.28$dres)), "references-fringillidae.txt")
+dres <- fringillidae_v2022.01.28$dres
+dquery <- fringillidae_v2022.01.28$dquery
 ```
@@
 <!-- If you're aiming for syst bio you don't want to start so general Readers know what chronograms and phylogenies are!.-->
-Chronograms --phylogenies with branch lengths proportional to time-- provide key data on evolutionary time frame for the study of natural processes in many areas of biological research, such as developmental biology [@laubichler2009form; @delsuc2018phylogenomic], conservation biology [@Felsenstein1985a; @Webb2000], historical biogeography [@posadas2006historical], and species diversification [@magallon2001absolute; @Morlon2014].
+Chronograms --phylogenies with branch lengths proportional to time-- provide key data on evolutionary time frame for the study of natural processes in many areas of biological research, such as comparative analysis [@harvey1991comparative; @freckleton2002phylogenetic], developmental biology [@laubichler2009form; @delsuc2018phylogenomic], conservation and ecology [@Felsenstein1985a; @Webb2000], historical biogeography [@posadas2006historical], and species diversification [@magallon2001absolute; @Morlon2014].
 
@@
 DateLife's core software application is available as an R package [@datelifeR], and as an online Rshiny interactive website at www.datelife.org.
-It features key elements for scientific reproducibility, such as a versioned, open and fully public source database [@mctavish2015phylesystem], data stored and available in a computer readable format [@vos2012nexml], automated and programmatic ways of accessing the data [@Stoltzfus2013] and methods to summarize and compare the data.
+It features key elements for scientific reproducibility, such as a versioned, open and fully public source database [@mctavish2015phylesystem] that stores data in a computer-readable format [@vos2012nexml]; automated and programmatic ways of accessing and downloading the data in a compuer-redable format also [@Stoltzfus2013]; and methods to summarize and compare the data.
 
@@
 DateLife's core software application consists of the R package `datelife`.
-Its current stable version -- v`r packageVersion("datelife")`, is available from the The Comprehensive R Archive Network (CRAN) repository [@datelifeR], and relies on functionalities from various biological R packages:
+Its current stable version -- v`r packageVersion("datelife")`, is available from the Comprehensive R Archive Network (CRAN) repository [@datelifeR], and relies on functionalities from various biological R packages:
 ape [@paradis2004],
@@
 
-DateLife starts by processing an input consisting of at least two taxon names, which can be provided as a comma separated character string or as tip labels on a tree.
+DateLife starts by processing an input consisting of at least one taxon scientific names. Two or more scientific names can be provided as a comma separated character string or as tip labels on a tree.
 If the input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004].
@@
 
-DateLife accepts scientific names that can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or a binomial species name.
-Subspecies and variants are ignored.
-If an input taxon name belongs to an inclusive taxonomic group, DateLife has two alternative behaviors defined by the "get species from taxon" flag.
-If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group following a standard taxonomy of choice, and adds them to the input string.
-Taxonomies currently supported by DateLife are Open Tree of Life (OpenTree) unified Taxonomy [OTT, @rees2017automated], the National Center of Biotechnology Information (NCBI) taxonomic database [@schoch2020ncbi], the Global Biodiversity Information Facility (GBIF) taxonomic backbone [@gbif2022taxonomy], and the Interim Register of Marine and Nonmarine Genera (IRMNG) database [@rees2017irmng].
-If the flag is inactive, DateLife excludes any taxon names above the species level from the search query.
-
 DateLife processes input scientific names using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database.
 TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard [@Boyle2013].
-DateLife implements TNRS with OTT as standard [@opentreeAPIs; @rees2017automated], storing taxonomic identification numbers for further processing.
+TNRS also allows to correctly choose between homonyms, by considering other taxa provided as input to infer the taxonomic context of the homonym.
+DateLife implements TNRS using the Open Tree of Life (OpenTree) unified Taxonomy [OTT, @opentreeAPIs; @rees2017automated] as standard, storing taxonomic identification numbers (OTT ids) for further processing and analysis.
+Other taxonomies currently supported by DateLife are the National Center of Biotechnology Information (NCBI) taxonomic database [@schoch2020ncbi], the Global Biodiversity Information Facility (GBIF) taxonomic backbone [@gbif2022taxonomy], and the Interim Register of Marine and Nonmarine Genera (IRMNG) database [@rees2017irmng].
 <!-- We are using OpenTree's TNRS [BCO: yes], Boyle described TNRS first, right?  [BCO: earliest I can find, but feels like there should be much older work]-->
 
-The processed input taxon names are saved as an R object of a newly defined class, \texttt{datelifeQuery}, that is used in the following steps.
-This object contains the standardized names, the corresponding OTT identification numbers, and the topology of the input tree if any was provided.
+Besides binomial species names, DateLife accepts scientific names from any inclusive taxonomic group (e.g., genus, family, tribe), as well as subspecific taxonomic variants (e.g., subspecies, variants, strains).
+If a taxon name belongs to an inclusive taxonomic group, DateLife has two alternative behaviors defined by the "get species from taxon" flag.
+If the flag is active, DateLife retrieves all species names within the taxonomic group from the standard taxonomy of choice, and adds them to the search query.
+In this case, subspecific variants are excluded.
+If the flag is inactive, DateLife excludes any taxon names above the species level from the search query.
+Species and subspecific variant names are processed and searched as provided by the user.
+The processed taxon names are saved as an R object of a newly defined class, \texttt{datelifeQuery}, that is used in the following steps.
+This object contains the names standardized to the taxonomy of choice, the corresponding OTT id numbers, and the topology of the input tree if one was provided.
 
@@
 A unique feature of the Phylesystem is that any user can add new published, state-of-the-art chronograms any time, through their curator application (https://tree.opentreeoflife.org/curator).
-As chronograms are added to Phylesystem, they are incorporated into the chronogram database of the `datelife` package.
+As chronograms are added to Phylesystem, they can be incorporated into the chronogram database of the `datelife` package.
+`datelife`'s chronogram database is currently manually updated as new chronogram data is added to Phylesystem.
 The updated database is assigned a new version number, followed by a package release on CRAN.
-`datelife`'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
-<!--BCO: Nicely done-->
 Users can also implement functions from the `datelife` R package to trigger an update of the local chronogram database, to incorporate any new chronograms to the user's DateLife analysis before an official database update is released on CRAN.
@@
 We recommend that users provide a tree topology as input from the literature, or one of their own making.
-If no topology is provided, DateLife automatically extracts one from the OpenTree synthetic tree, a phylogeny encompassing 2.3 million taxa across all life, assembled from 1, 239 published phylogenetic trees and OpenTree's unified Taxonomy, OTT  [@opentreeoflife2019synth].
+If no topology is provided, DateLife automatically extracts one from the OpenTree synthetic tree, a phylogeny currently encompassing 2.3 million taxa across all life, assembled from 1, 239 published phylogenetic trees and OpenTree's unified Taxonomy, OTT [@opentreeoflife2019synth].
 Alternatively, DateLife can combine topologies from source chronograms using a supertree approach.
-To combine topologies from source chronograms into a single summary (or supertree) topology, the DateLife workflow identifies the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
+To do this DateLife first identifies the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
 [-@ane2009groves].
+If the source chronograms do not form a grove, the supertree reconstruction will fail.
 In rare cases, a group of trees can have multiple groves.
 By default, DateLife chooses the grove with the most taxa, however, the "criterion = trees" flag allows the user to choose the grove with the most trees instead.
-If source chronograms do not form a grove, the supertree reconstruction will fail.
+The result is a single summary (or supertree) topology, that combines topologies from source chronograms in a grove.
 
-_**Dating the topology.--**_
-Input topologies from OpenTree or the supertree approach described above do not include branch length estimates of any kind.
-Optionally, to estimate branch lengths proportional to substitution rates on these topologies, DateLife can mine the Barcode of Life Data System, BOLD [@ratnasingham2007bold] to obtain genetic markers for the input taxa.
-These markers are aligned with MUSCLE [@edgar2004muscle] (by default) or MAFFT [@katoh2009multiple].
-This alignment can be used to estimate branch lengths on input topologies that lack branch lengths.
-Currently, branch length reconstruction in DateLife is performed using parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed [@schliep2011phangorn].
-While relative branch length information provides additional data for nodes without secondary date calibrations, topologies without branch lengths can also be dated.
-
-Once a topology is chosen, DateLife applies the congruification method [@Eastman2013] to find nodes belonging to the same clade across source chronograms, and extract the corresponding node ages from the patristic distance matrices stored as `datelifeResult`.
-By definition, the matrices store total distance (time from tip to tip), hence, node ages correspond to half the values stored in the patristic distance matrices.
-This assumes that the terminal taxa are coeval and occur at the present.
+_**Applying secondary calibrations.--**_
+Once a topology is chosen, DateLife applies the congruification method [@Eastman2013] that find nodes belonging to the same clade across source chronograms, and then extracts the corresponding node ages from patristic distance matrices stored as a `datelifeResult` object.
+Note that by definition, these matrices store total distance (time from tip to tip), assuming that the terminal taxa are coeval and occur at the present.
+Hence, node ages correspond to half the values stored in the `datelifeResult` matrices.
 A table of congruified node ages that can be used as calibrations for a dating analysis is stored as a `congruifiedCalibrations` object.
@@
 For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method [SDM; @Criscuolo2006], which deforms patristic distance matrices by minimizing variance and then averaging them.
-These single summary taxon pair age matrices (Summarized calibrations) can be applied as calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes [@Huelsenbeck2001; @Ronquist2003], PATHd8 [@Britton2007], BLADJ [@Webb2008; @webb2005phylomatic], and treePL [@Smith2012].
+These single summary taxon pair age matrices are stored as summarized calibrations that can be used as secondary calibrations to date a tree topology - with or without initial branch lengths, using phylogenetic dating methods currently supported within DateLife: BLADJ [@Webb2008; @webb2005phylomatic], MrBayes [@Huelsenbeck2001; @Ronquist2003], PATHd8 [@Britton2007], and treePL [@Smith2012].
 
-By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully dated topology.
-BLADJ fixes node ages that have calibration data, and distributes time between nodes with no data evenly between nodes with calibration data.
-This minimizes age variance in the resulting chronogram [@Webb2008].
-BLADJ does not use branch lengths even when they are present in the input tree or summarizing topology.
+_**Dating a tree topology.--**_
+By default, DateLife implements the Branch Length Adjuster [BLADJ; @Webb2008; @webb2005phylomatic] algorithm to obtain a fully dated topology.
+BLADJ is the only dating algorithm that can work with initial topologies without any branch length data.
+
+Alternatively, the user can chose phylogenetic dating options supported in DateLife that incorporate branch length information from the input topology in combination with the secondary calibrations:
+PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees;
+treePL [@Smith2012], is a semi-parametric, rate-smoothing, penalized likelihood dating method [@sanderson2002];
+the MrBayes  [@Huelsenbeck2001; @Ronquist2003] approach in DateLife uses the calibrations as priors on node ages.
+
+The latter methods can assign dates using a birth-death model (all of them????), they require initial branch lengths and thus require more time and expertise to run, which makes BLADJ the fastest and most practical method to obtain a dated tree.
+
+BLADJ fixes node ages that have calibration data, and distributes time between nodes with no data evenly between calibrated nodes.
+This has proven effective minimizes age variance in the resulting chronogram and useful for ecoogical analyses [@Webb2008].
 When there is conflict in ages between nodes with calibration data, BLADJ ignores node ages that are older than the age of a parent node.
@@
 If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature.
-If none is provided, DateLife assigns an arbitrary age to the root as 10% older than the oldest age available within the group.
+If none is provided, DateLife will not return a dated topology but provide a warning message along with suggestions on how the user can provide an age for the root so DateLife can run.
+<!-- assigns an arbitrary age to the root as 10% older than the oldest age available within the group. -->
 
-Alternative phylogenetic dating options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the calibrations.
-PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees.
-treePL [@Smith2012], is a semi-parametric, rate-smoothing, penalized likelihood dating method [@sanderson2002].
-The MrBayes  [@Huelsenbeck2001; @Ronquist2003] approach in DateLife uses the calibrations as priors on node ages.
+_**Dating a tree with branch lengths--.**_
+Topologies obtained from OpenTree and with the supertree approach described above lack branch length data.
+Yet, phylogenetic dating using branch lengths data is the golden standard for phylogenetic dating analyses, but it is costly and requires a lot of human/expert curation.
+
+A fast solution implemented in DateLife is as follows.
+To estimate branch lengths proportional to substitution rates for these topologies, DateLife currently implements a simple algorithm.
+First, it mines the Barcode of Life Data System, BOLD [@ratnasingham2007bold] to obtain genetic markers for the input taxa.
+Mined genetic sequences are aligned with MUSCLE [@edgar2004muscle] (by default) or MAFFT [@katoh2009multiple].
+
+The BOLD sequence alignment is then used to reconstruct branch lengths with the accelerated transformation (ACCTRAN) parsimony algorithm, which resolves ambiguous character optimization, by assigning changes along branches of the tree as close to the root as possible [@agnarsson2008acctran].
+This algorithm work rally fast and allows getting initial branch lengths that ar ethen optimized using ML.
+Optionally, the likelihood of the tree topology, the alignment and the reconstructed branch lengths given different evolutionary models, is computed using functions from the `phangorn` package [@schliep2011phangorn].
+
+Relative branch length information provides key data for phylogenetic dating, especially for nodes without secondary calibrations available. Yet, topologies without branch lengths can also be dated.
 
 _**Visualizing results.--**_
-Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as visualize and compare results graphically, or construct their own graphs using DateLife's chronogram plot generation functions available from the R package `datelifeplot` [@datelifeplot].
+Finally, users can save all source and summary chronograms in formats that permit reuse and reanalysis (such as newick and the R "phylo" format), as well as visualize and compare results graphically, or construct their own graphs using DateLife's chronogram plot generation functions available from the R package `datelifeplot` [@datelifeplot].
 
@@
 which gets all recognized species names within a named group from a taxonomy of choice.
-Following the NCBI taxonomy, our DateLife query has `r length(dquery$cleaned_names)` Fringillidae species.
+Following the NCBI taxonomy, our DateLife query has `r length(dquery$cleaned_names)` Fringillidae species names.
 <!--BCO: Feel like we need a number here, though it's good that the clade hasn't gone extinct ;-) -->
-This taxon-constrained approach implies that the final results of a full DateLife analysis will be done using a tree topology and ages for the species in a named group, which do not necessarily correspond to a monophyletic group.
-Users can change this behaviour by providing a monophyletic tree as input for a DateLife search, or as a tree topology for a DateLife summary.
+This taxon-constrained approach implies that the full DateLife analysis will be performed using a tree topology and ages available for species names from a given taxonomic group, which do not necessarily correspond to a monophyletic group.
+Users can change this behaviour by providing all species names corresponding to a monophyletic group as input for a DateLife search, or a monophyletic tree to construct a DateLife summary.
 <!--BCO: it's ambiguous: it gets a topology and age for the larger clade that includes all Fringillidae species? Or for the paraphyletic Fringillidae group?
@@
 
-Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species.
-The DateLife search identified `r length(dres)` chronograms containing at least two Fringillidae species, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @price2014niche].
-Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementray Table S1).
+Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species as tip taxa.
+The DateLife search identified `r length(dres)` chronograms matching this criteria, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @kimball2019phylogenomic; @oliveros2019earth; @price2014niche; @roquet2014one; @uyeda2017evolution].
+Once identified, DateLife pruned these matching chronograms to remove tips that do not belong to the queried taxon names, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementray Table S1).
 
@@
 
-The final step is to congruify and summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM.
+The final step entailed congruifying and summarizing the age data available for the Fringillidae species into two single summary chronograms, using two different types of summary ages, median and SDM.
 As explained in the "Description" section, a tree topology to summarize age data upon is required.
-By default, DateLife uses the topology from OpenTree's synthetic tree that contains the species in the search query to summarize age data upon.
-According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).
+By default, to do this, DateLife uses the topology from OpenTree's synthetic tree that contains all taxa from the search query.
+According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig.
+\ref{fig:fringillidae-topologies}).
+Hence, a topology containing only the 289 species from the original query was extracted from Open Tree of Life’s synthetic tree v12.3 [@opentreeoflife2019synth].
 
@@
 However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures S2-S20).
-The only excpetion to tjis was observed on results of the cross validation analysis of the @burns2014phylogenetics chronogram, which results in much younger node ages when estimated using secondary calibrations (Supplementary Figs.
+The only exception to this was observed on cross validation results of the @burns2014phylogenetics chronogram, which displays much younger node ages when estimated using secondary calibrations (Supplementary Figs.
 S1, S5).
@@
 
-For downstream analyses, using alternative chronogram may deeply affect our inferences [@title2016macrophylogenies], particularly when studying phenomena dependent on the timing of species diversification events, such as macroevolutionary processes.
+The study of phenomena dependent on the timing of species diversification events, such as macroevolutionary processes, is affected by the usage of alternative chronograms that vary in topology [@rabosky2015no; @title2016macrophylogenies].
 <!-- There is evidence that different chronograms will result in very different macroevolutionary histories inferred from them [@title2016macrophylogenies].
-However, it is unknown how using very different age data could affect hypothesis testing in other areas of research and more effort is needed in that directions -->
+However, it is unknown how using very different age data could affect hypothesis testing in other areas of research and more effort is needed in that direction -->
 
-In ecology and conservation biology, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance [@Webb2008].
-Hence, DateLife's workflow features different ways of estimating node ages in the absence of calibrations and branch length information for certain taxa.
-"Making up" branch lengths is a common practice in scientific publications: @Jetz2012, created a chronogram of all 9, 993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a chronogram of 31, 536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a chronogram of 353, 185 seed plants where only 23% had molecular data.
+On the other hand, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance in ecological and conservation biology studies [@Webb2008].
+Hence, DateLife's workflow features different ways of generating node ages in the absence of calibration and branch length information for certain taxa.
+
+Adding branch lengths sampled from a birth-death model in the absence of genetic data
+has been found to improve insight in phylogeny-based analyses.
+
+is a common practice in scientific publications: @Jetz2012, created a chronogram of all 9, 993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a chronogram of 31, 536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a chronogram of 353, 185 seed plants where only 23% had molecular data.
+
+Simulating branch lengths following a birth-death species diversification model for missing taxa in chronograms with non random sampling, lowers type I error (false positive, incorrectly accepting the null hypothesis of a constant-rate or temporally varying rate birth-death model) when analyzing changes in diversification rate using the gamma statistic [@cusimano2012new].
+
+@thomas2013pastis hypothesize that results of diversification analyses will be biased towards the birth-death model used to simulate branch lengths;
+and note that "the effects of missing species placement or polytomy resolution are less clear for other phylogeny-based analyses (e.g.
+correlates of diversification, modelling trait evolution, community phylogenetics), and future work should test how the treatment of missing species influences both parameter estimation and type I and II errors"
 
 Notably, risks come with this practice.
-Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration, using polytomy resolution and branch estimation methods.
+
+Taken to the extreme, one could generate a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration, using polytomy resolution and branch length simulation methods.
 There has yet to be a thorough analysis of what can go wrong when one extends inferences beyond the data in this way, so we urge caution; we also urge readers to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, and assessing the consistency of the results with prior work.
@@
 
-Knowledge of the evolutionary time frame of organisms is key to many research areas: trait evolution,
-species diversification, biogeography, macroecology and more. It is also crucial for education, science communication and policy, but generating chronograms is difficult,
-especially for those who want to use phylogenies but who are not systematists, or
-do not have the time to acquire and develop the necessary knowledge and skills to construct them on their own.
-Importantly, years of primarily public funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
+Knowledge of the evolutionary time frame of organisms is key to many research areas: trait evolution, species diversification, biogeography, macroecology and more.
+It is also crucial for education, science communication and policy, but generating chronograms is difficult, especially for those who want to use phylogenies but who are not systematists, or do not have the time to acquire and develop the necessary knowledge and skills to construct them on their own.
+Importantly, years of primarily publicly funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
 
@@
 The DateLife project allows for easy and fast summary of public and state-of-the-art data on time of lineage divergence.
-It provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identification of regions that require more research, or that have conflicting information.
+It provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identifying regions that require more research, or that have conflicting information.
 It is available as an R package, and as a web-based R shiny application at www.datelife.org
@@
 Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
-The DateLife project helps with awareness of the existing variation in expert time of divergence data, and will foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
+The DateLife project should improve awareness of the existing variation in expert time of divergence data, and foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
 
@@
 
-Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to the Datelife project; DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), ABI-1458572 to the Phylotastic project, and ABI-1759846 to the Open Tree of Life project.
+Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to the Datelife project; DBI-0905606 to the National Evolutionary Synthesis Center (NESCent); ABI-1458572 to the Phylotastic project; and ABI-1759846 to the Open Tree of Life project.
 

### 2023-01-19 manuscript/manuscript.tex
updating new submission

@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4926}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 5355}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 
-Chronograms --phylogenies with branch lengths proportional to time-- provide key data on evolutionary time frame for the study of natural processes in many areas of biological research, such as developmental biology (Delsuc et al., 2018; Laubichler \& Maienschein, 2009), conservation biology (Felsenstein, 1985; Webb, 2000), historical biogeography (Posadas, Crisci, \& Katinas, 2006), and species diversification (Magallon \& Sanderson, 2001; Morlon, 2014).
+Chronograms --phylogenies with branch lengths proportional to time-- provide key data on evolutionary time frame for the study of natural processes in many areas of biological research, such as comparative analysis (Freckleton, Harvey, \& Pagel, 2002; Harvey, Pagel, \& others, 1991), developmental biology (Delsuc et al., 2018; Laubichler \& Maienschein, 2009), conservation and ecology (Felsenstein, 1985; Webb, 2000), historical biogeography (Posadas, Crisci, \& Katinas, 2006), and species diversification (Magallon \& Sanderson, 2001; Morlon, 2014).
 
@@
 
-DateLife starts by processing an input consisting of at least two taxon names, which can be provided as a comma separated character string or as tip labels on a tree.
-If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis et al., 2004).
+DateLife starts by processing an input consisting of at least one taxon scientific names. Two or more scientific names can be provided as a comma separated character string or as tip labels on a tree.
+If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a \enquote{phylo} R object (Paradis et al., 2004).
 The input tree is not required to have branch lengths, and its topology is used in the summary steps described in the next section.
-
-DateLife accepts scientific names that can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or a binomial species name.
-Subspecies and variants are ignored.
-If an input taxon name belongs to an inclusive taxonomic group, DateLife has two alternative behaviors defined by the ``get species from taxon'' flag.
-If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group following a standard taxonomy of choice, and adds them to the input string.
-Taxonomies currently supported by DateLife are Open Tree of Life (OpenTree) unified Taxonomy (OTT, Rees \& Cranston, 2017), the National Center of Biotechnology Information (NCBI) taxonomic database (Schoch et al., 2020), the Global Biodiversity Information Facility (GBIF) taxonomic backbone (GBIF Secretariat, 2022), and the Interim Register of Marine and Nonmarine Genera (IRMNG) database (Rees, Vandepitte, Decock, \& Vanhoorne, 2017).
-If the flag is inactive, DateLife excludes any taxon names above the species level from the search query.
 
@@
 TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard (Boyle et al., 2013).
-DateLife implements TNRS with OTT as standard (Open Tree Of Life et al., 2016; Rees \& Cranston, 2017), storing taxonomic identification numbers for further processing.
-
-The processed input taxon names are saved as an R object of a newly defined class, \texttt{datelifeQuery}, that is used in the following steps.
-This object contains the standardized names, the corresponding OTT identification numbers, and the topology of the input tree if any was provided.
+TNRS also allows to correctly choose between homonyms, by considering other taxa provided as input to infer the taxonomic context of the homonym.
+DateLife implements TNRS using the Open Tree of Life (OpenTree) unified Taxonomy (OTT, Open Tree Of Life et al., 2016; Rees \& Cranston, 2017) as standard, storing taxonomic identification numbers (OTT ids) for further processing and analysis.
+Other taxonomies currently supported by DateLife are the National Center of Biotechnology Information (NCBI) taxonomic database (Schoch et al., 2020), the Global Biodiversity Information Facility (GBIF) taxonomic backbone (GBIF Secretariat, 2022), and the Interim Register of Marine and Nonmarine Genera (IRMNG) database (Rees et al., 2017).
+
+Besides binomial species names, DateLife accepts scientific names from any inclusive taxonomic group (e.g., genus, family, tribe), as well as subspecific taxonomic variants (e.g., subspecies, variants, strains).
+If a taxon name belongs to an inclusive taxonomic group, DateLife has two alternative behaviors defined by the \enquote{get species from taxon} flag.
+If the flag is active, DateLife retrieves all species names within the taxonomic group from the standard taxonomy of choice, and adds them to the search query.
+In this case, subspecific variants are excluded.
+If the flag is inactive, DateLife excludes any taxon names above the species level from the search query.
+Species and subspecific variant names are processed and searched as provided by the user.
+The processed taxon names are saved as an R object of a newly defined class, \texttt{datelifeQuery}, that is used in the following steps.
+This object contains the names standardized to the taxonomy of choice, the corresponding OTT id numbers, and the topology of the input tree if one was provided.
 
@@
 A unique feature of the Phylesystem is that any user can add new published, state-of-the-art chronograms any time, through their curator application (\url{https://tree.opentreeoflife.org/curator}).
-As chronograms are added to Phylesystem, they are incorporated into the chronogram database of the \texttt{datelife} package.
+As chronograms are added to Phylesystem, they can be incorporated into the chronogram database of the \texttt{datelife} package.
+\texttt{datelife}'s chronogram database is currently manually updated as new chronogram data is added to Phylesystem.
 The updated database is assigned a new version number, followed by a package release on CRAN.
-\texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
 Users can also implement functions from the \texttt{datelife} R package to trigger an update of the local chronogram database, to incorporate any new chronograms to the user's DateLife analysis before an official database update is released on CRAN.
@@
 We recommend that users provide a tree topology as input from the literature, or one of their own making.
-If no topology is provided, DateLife automatically extracts one from the OpenTree synthetic tree, a phylogeny encompassing 2.3 million taxa across all life, assembled from 1, 239 published phylogenetic trees and OpenTree's unified Taxonomy, OTT (Open Tree Of Life et al., 2019).
+If no topology is provided, DateLife automatically extracts one from the OpenTree synthetic tree, a phylogeny currently encompassing 2.3 million taxa across all life, assembled from 1, 239 published phylogenetic trees and OpenTree's unified Taxonomy, OTT (Open Tree Of Life et al., 2019).
 Alternatively, DateLife can combine topologies from source chronograms using a supertree approach.
-To combine topologies from source chronograms into a single summary (or supertree) topology, the DateLife workflow identifies the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
+To do this DateLife first identifies the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
 (2009).
+If the source chronograms do not form a grove, the supertree reconstruction will fail.
 In rare cases, a group of trees can have multiple groves.
-By default, DateLife chooses the grove with the most taxa, however, the ``criterion = trees'' flag allows the user to choose the grove with the most trees instead.
-If source chronograms do not form a grove, the supertree reconstruction will fail.
-
-\emph{\textbf{Dating the topology.--}}
-Input topologies from OpenTree or the supertree approach described above do not include branch length estimates of any kind.
-Optionally, to estimate branch lengths proportional to substitution rates on these topologies, DateLife can mine the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain genetic markers for the input taxa.
-These markers are aligned with MUSCLE (Edgar, 2004) (by default) or MAFFT (Katoh, Asimenos, \& Toh, 2009).
-This alignment can be used to estimate branch lengths on input topologies that lack branch lengths.
-Currently, branch length reconstruction in DateLife is performed using parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed (Schliep, 2011).
-While relative branch length information provides additional data for nodes without secondary date calibrations, topologies without branch lengths can also be dated.
-
-Once a topology is chosen, DateLife applies the congruification method (Eastman, Harmon, \& Tank, 2013) to find nodes belonging to the same clade across source chronograms, and extract the corresponding node ages from the patristic distance matrices stored as \texttt{datelifeResult}.
-By definition, the matrices store total distance (time from tip to tip), hence, node ages correspond to half the values stored in the patristic distance matrices.
-This assumes that the terminal taxa are coeval and occur at the present.
+By default, DateLife chooses the grove with the most taxa, however, the \enquote{criterion = trees} flag allows the user to choose the grove with the most trees instead.
+The result is a single summary (or supertree) topology, that combines topologies from source chronograms in a grove.
+
+\emph{\textbf{Applying secondary calibrations.--}}
+Once a topology is chosen, DateLife applies the congruification method (Eastman, Harmon, \& Tank, 2013) that find nodes belonging to the same clade across source chronograms, and then extracts the corresponding node ages from patristic distance matrices stored as a \texttt{datelifeResult} object.
+Note that by definition, these matrices store total distance (time from tip to tip), assuming that the terminal taxa are coeval and occur at the present.
+Hence, node ages correspond to half the values stored in the \texttt{datelifeResult} matrices.
 A table of congruified node ages that can be used as calibrations for a dating analysis is stored as a \texttt{congruifiedCalibrations} object.
 
-For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method {[}SDM; Criscuolo, Berry, Douzery, and Gascuel (2006){]}, which deforms patristic distance matrices by minimizing variance and then averaging them.
-These single summary taxon pair age matrices (Summarized calibrations) can be applied as calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Webb, Ackerly, \& Kembel, 2008; Webb \& Donoghue, 2005), and treePL (Smith \& O'Meara, 2012).
-
-By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully dated topology.
-BLADJ fixes node ages that have calibration data, and distributes time between nodes with no data evenly between nodes with calibration data.
-This minimizes age variance in the resulting chronogram (Webb et al., 2008).
-BLADJ does not use branch lengths even when they are present in the input tree or summarizing topology.
+For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method (SDM; Criscuolo, Berry, Douzery, \& Gascuel, 2006), which deforms patristic distance matrices by minimizing variance and then averaging them.
+These single summary taxon pair age matrices are stored as summarized calibrations that can be used as secondary calibrations to date a tree topology - with or without initial branch lengths, using phylogenetic dating methods currently supported within DateLife: BLADJ (Webb, Ackerly, \& Kembel, 2008; Webb \& Donoghue, 2005), MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), and treePL (Smith \& O'Meara, 2012).
+
+\emph{\textbf{Dating a tree topology.--}}
+By default, DateLife implements the Branch Length Adjuster (BLADJ; Webb et al., 2008; Webb \& Donoghue, 2005) algorithm to obtain a fully dated topology.
+BLADJ is the only dating algorithm that can work with initial topologies without any branch length data.
+
+Alternatively, the user can chose phylogenetic dating options supported in DateLife that incorporate branch length information from the input topology in combination with the secondary calibrations:
+PATHd8 is a non-clock, rate-smoothing method (Britton et al., 2007) to date trees;
+treePL (Smith \& O'Meara, 2012), is a semi-parametric, rate-smoothing, penalized likelihood dating method (Sanderson, 2002);
+the MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003) approach in DateLife uses the calibrations as priors on node ages.
+
+The latter methods can assign dates using a birth-death model (all of them????), they require initial branch lengths and thus require more time and expertise to run, which makes BLADJ the fastest and most practical method to obtain a dated tree.
+
+BLADJ fixes node ages that have calibration data, and distributes time between nodes with no data evenly between calibrated nodes.
+This has proven effective minimizes age variance in the resulting chronogram and useful for ecoogical analyses (Webb et al., 2008).
 When there is conflict in ages between nodes with calibration data, BLADJ ignores node ages that are older than the age of a parent node.
@@
 If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature.
-If none is provided, DateLife assigns an arbitrary age to the root as 10\% older than the oldest age available within the group.
-
-Alternative phylogenetic dating options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the calibrations.
-PATHd8 is a non-clock, rate-smoothing method (Britton et al., 2007) to date trees.
-treePL (Smith \& O'Meara, 2012), is a semi-parametric, rate-smoothing, penalized likelihood dating method (Sanderson, 2002).
-The MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003) approach in DateLife uses the calibrations as priors on node ages.
+If none is provided, DateLife will not return a dated topology but provide a warning message along with suggestions on how the user can provide an age for the root so DateLife can run.
+
+\emph{\textbf{Dating a tree with branch lengths--.}}
+Topologies obtained from OpenTree and with the supertree approach described above lack branch length data.
+Yet, phylogenetic dating using branch lengths data is the golden standard for phylogenetic dating analyses, but it is costly and requires a lot of human/expert curation.
+
+A fast solution implemented in DateLife is as follows.
+To estimate branch lengths proportional to substitution rates for these topologies, DateLife currently implements a simple algorithm.
+First, it mines the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain genetic markers for the input taxa.
+Mined genetic sequences are aligned with MUSCLE (Edgar, 2004) (by default) or MAFFT (Katoh, Asimenos, \& Toh, 2009).
+
+The BOLD sequence alignment is then used to reconstruct branch lengths with the accelerated transformation (ACCTRAN) parsimony algorithm, which resolves ambiguous character optimization, by assigning changes along branches of the tree as close to the root as possible (Agnarsson \& Miller, 2008).
+This algorithm work rally fast and allows getting initial branch lengths that ar ethen optimized using ML.
+Optionally, the likelihood of the tree topology, the alignment and the reconstructed branch lengths given different evolutionary models, is computed using functions from the \texttt{phangorn} package (Schliep, 2011).
+
+Relative branch length information provides key data for phylogenetic dating, especially for nodes without secondary calibrations available. Yet, topologies without branch lengths can also be dated.
 
 \emph{\textbf{Visualizing results.--}}
-Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R ``phylo'' format), as well as visualize and compare results graphically, or construct their own graphs using DateLife's chronogram plot generation functions available from the R package \texttt{datelifeplot} (Sanchez-Reyes \& O'Meara, 2022).
+Finally, users can save all source and summary chronograms in formats that permit reuse and reanalysis (such as newick and the R \enquote{phylo} format), as well as visualize and compare results graphically, or construct their own graphs using DateLife's chronogram plot generation functions available from the R package \texttt{datelifeplot} (Sanchez-Reyes \& O'Meara, 2022).
 
@@
 
-To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the ``get species from taxon'' flag,
+To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the \enquote{get species from taxon} flag,
 which gets all recognized species names within a named group from a taxonomy of choice.
-Following the NCBI taxonomy, our DateLife query has 289 Fringillidae species.
-This taxon-constrained approach implies that the final results of a full DateLife analysis will be done using a tree topology and ages for the species in a named group, which do not necessarily correspond to a monophyletic group.
-Users can change this behaviour by providing a monophyletic tree as input for a DateLife search, or as a tree topology for a DateLife summary.
+Following the NCBI taxonomy, our DateLife query has 289 Fringillidae species names.
+This taxon-constrained approach implies that the full DateLife analysis will be performed using a tree topology and ages available for species names from a given taxonomic group, which do not necessarily correspond to a monophyletic group.
+Users can change this behaviour by providing all species names corresponding to a monophyletic group as input for a DateLife search, or a monophyletic tree to construct a DateLife summary.
 
@@
 
-Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species.
-The DateLife search identified 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Price et al., 2014).
-Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementray Table S1).
+Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species as tip taxa.
+The DateLife search identified 19 chronograms matching this criteria, published in 13 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Kimball et al., 2019; Oliveros et al., 2019; Price et al., 2014; Roquet, Lavergne, \& Thuiller, 2014; Uyeda, Pennell, Miller, Maia, \& McClain, 2017).
+Once identified, DateLife pruned these matching chronograms to remove tips that do not belong to the queried taxon names, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementray Table S1).
 
@@
 
-The final step is to congruify and summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM.
-As explained in the ``Description'' section, a tree topology to summarize age data upon is required.
-By default, DateLife uses the topology from OpenTree's synthetic tree that contains the species in the search query to summarize age data upon.
+The final step entailed congruifying and summarizing the age data available for the Fringillidae species into single summary chronograms, using two different types of summary ages, median and SDM.
+As explained in the \enquote{Description} section, a tree topology to summarize age data upon is required.
+By default, to do this, DateLife uses the topology from OpenTree's synthetic tree that contains species from the search query.
 According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).
@@
 However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures S2-S20).
-The only excpetion to tjis was observed on results of the cross validation analysis of the Burns et al.
-(2014) chronogram, which results in much younger node ages when estimated using secondary calibrations (Supplementary Figs.
+The only exception to this was observed on cross validation results of the Burns et al.
+(2014) chronogram, which displays much younger node ages when estimated using secondary calibrations (Supplementary Figs.
 S1, S5).
@@
 
-For downstream analyses, using alternative chronogram may deeply affect our inferences (Title \& Rabosky, 2016), particularly when studying phenomena dependent on the timing of species diversification events, such as macroevolutionary processes.
-
-In ecology and conservation biology, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance (Webb et al., 2008).
-Hence, DateLife's workflow features different ways of estimating node ages in the absence of calibrations and branch length information for certain taxa.
-``Making up'' branch lengths is a common practice in scientific publications: Jetz et al.
+The study of phenomena dependent on the timing of species diversification events, such as macroevolutionary processes, is affected by the usage of alternative chronograms that vary in topology (Rabosky, 2015; Title \& Rabosky, 2016).
+
+On the other hand, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance in ecological and conservation biology studies (Webb et al., 2008).
+Hence, DateLife's workflow features different ways of generating node ages in the absence of calibration and branch length information for certain taxa.
+
+Adding branch lengths sampled from a birth-death model in the absence of genetic data
+has been found to improve insight in phylogeny-based analyses.
+
+is a common practice in scientific publications: Jetz et al.
 (2012), created a chronogram of all 9, 993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
@@
 
+Simulating branch lengths following a birth-death species diversification model for missing taxa in chronograms with non random sampling, lowers type I error (false positive, incorrectly accepting the null hypothesis of a constant-rate or temporally varying rate birth-death model) when analyzing changes in diversification rate using the gamma statistic (Cusimano, Stadler, \& Renner, 2012).
+
+Thomas et al. (2013) hypothesize that results of diversification analyses will be biased towards the birth-death model used to simulate branch lengths;
+and note that \enquote{the effects of missing species placement or polytomy resolution are less clear for other phylogeny-based analyses (e.g.~correlates of diversification, modelling trait evolution, community phylogenetics), and future work should test how the treatment of missing species influences both parameter estimation and type I and II errors}
+
 Notably, risks come with this practice.
-Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration, using polytomy resolution and branch estimation methods.
+
+Taken to the extreme, one could generate a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration, using polytomy resolution and branch length simulation methods.
 There has yet to be a thorough analysis of what can go wrong when one extends inferences beyond the data in this way, so we urge caution; we also urge readers to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, and assessing the consistency of the results with prior work.
@@
 
-Knowledge of the evolutionary time frame of organisms is key to many research areas: trait evolution,
-species diversification, biogeography, macroecology and more. It is also crucial for education, science communication and policy, but generating chronograms is difficult,
-especially for those who want to use phylogenies but who are not systematists, or
-do not have the time to acquire and develop the necessary knowledge and skills to construct them on their own.
-Importantly, years of primarily public funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
+Knowledge of the evolutionary time frame of organisms is key to many research areas: trait evolution, species diversification, biogeography, macroecology and more.
+It is also crucial for education, science communication and policy, but generating chronograms is difficult, especially for those who want to use phylogenies but who are not systematists, or do not have the time to acquire and develop the necessary knowledge and skills to construct them on their own.
+Importantly, years of primarily publicly funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
 
 The DateLife project allows for easy and fast summary of public and state-of-the-art data on time of lineage divergence.
-It provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identification of regions that require more research, or that have conflicting information.
+It provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identifying regions that require more research, or that have conflicting information.
 It is available as an R package, and as a web-based R shiny application at www.datelife.org
 Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
-The DateLife project helps with awareness of the existing variation in expert time of divergence data, and will foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
+The DateLife project should improve awareness of the existing variation in expert time of divergence data, and foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
 
@@
 
-Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities are available at \href{https://github.com/LunaSare/datelifeMS1}{datelifeMS1}, \href{https://github.com/LunaSare/datelife_examples}{datelife\_examples}, and \href{https://github.com/LunaSare/datelife_benchmark}{datelife\_benchmark} repositories in LLSR's GitHub account.
+Supplementary material, including code, biological examples and benchmark results data files and online-only appendices, can be found in the Dryad data repository (\url{https://doi.org/10.5061/dryad.cnp5hqc6w}), as well as in the GitHub repositories used to develop the reproducible manuscript (\url{https://doi.org/10.5281/zenodo.7435094}), the biological examples (\url{https://doi.org/10.5281/zenodo.7435101}), and the software benchmark (\url{https://doi.org/10.5281/zenodo.7435106}).
 
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-alstrom2014discovery}{}}%
+\leavevmode\hypertarget{ref-agnarsson2008acctran}{}%
+Agnarsson, I., \& Miller, J. A. (2008). Is acctran better than deltran? \emph{Cladistics}, \emph{24}(6), 1032--1038.
+
+\leavevmode\hypertarget{ref-alstrom2014discovery}{}%
 Alström, P., Hooper, D.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-claramunt2015new}{}}%
-Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{{Science Advances}}, \emph{1}(11), e1501005.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Criscuolo2006}{}}%
+\leavevmode\hypertarget{ref-claramunt2015new}{}%
+Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{Science Advances}, \emph{1}(11), e1501005.
+
+\leavevmode\hypertarget{ref-Criscuolo2006}{}%
 Criscuolo, A., Berry, V., Douzery, E.
-J.
-P., \& Gascuel, O.
+J., \& Gascuel, O.
 (2006).
-{SDM: A fast distance-based approach for (super)tree building in phylogenomics}.
-\emph{{Systematic Biology}}, \emph{55}(5), 740--755.
+SDM: A fast distance-based approach for (super)tree building in phylogenomics.
+\emph{Systematic Biology}, \emph{55}(5), 740--755.
 \url{https://doi.org/10.1080/10635150600969872}
 
-\leavevmode\vadjust pre{\hypertarget{ref-delsuc2018phylogenomic}{}}%
+\leavevmode\hypertarget{ref-cusimano2012new}{}%
+Cusimano, N., Stadler, T., \& Renner, S.
+S.
+(2012).
+A new method for handling missing species in diversification analysis applicable to randomly or nonrandomly sampled phylogenies.
+\emph{Systematic Biology}, \emph{61}(5), 785--792.
+
+\leavevmode\hypertarget{ref-delsuc2018phylogenomic}{}%
 Delsuc, F., Philippe, H., Tsagkogeorga, G., Simion, P., Tilak, M.-K., Turon, X., \ldots{} Douzery, E.
@@
 (2013).
-{Congruification: Support for time scaling large phylogenetic trees}.
-\emph{{Methods in Ecology and Evolution}}, \emph{4}(7), 688--691.
+Congruification: Support for time scaling large phylogenetic trees.
+\emph{Methods in Ecology and Evolution}, \emph{4}(7), 688--691.
 \url{https://doi.org/10.1111/2041-210X.12051}
 
-\leavevmode\vadjust pre{\hypertarget{ref-edgar2004muscle}{}}%
+\leavevmode\hypertarget{ref-edgar2004muscle}{}%
 Edgar, R. C. (2004). MUSCLE: Multiple sequence alignment with high accuracy and high throughput. \emph{Nucleic Acids Research}, \emph{32}(5), 1792--1797.
 
-\leavevmode\vadjust pre{\hypertarget{ref-Felsenstein1985a}{}}%
-Felsenstein, J. (1985). {Phylogenies and the Comparative Method}. \emph{The American Naturalist}, \emph{125}(1), 1--15. Retrieved from \url{http://www.jstor.org/stable/2461605}
-
-\leavevmode\vadjust pre{\hypertarget{ref-gbif2022taxonomy}{}}%
+\leavevmode\hypertarget{ref-Felsenstein1985a}{}%
+Felsenstein, J. (1985). Phylogenies and the Comparative Method. \emph{The American Naturalist}, \emph{125}(1), 1--15. Retrieved from \url{http://www.jstor.org/stable/2461605}
+
+\leavevmode\hypertarget{ref-freckleton2002phylogenetic}{}%
+Freckleton, R. P., Harvey, P. H., \& Pagel, M. (2002). Phylogenetic analysis and comparative data: A test and review of evidence. \emph{The American Naturalist}.
+
+\leavevmode\hypertarget{ref-gbif2022taxonomy}{}%
 GBIF Secretariat.
 (2022).
-{GBIF Backbone Taxonomy}.
-\emph{{Checklist Dataset}}, (Online Resource accessed via GBIF.org).
+GBIF Backbone Taxonomy.
+\emph{Checklist dataset}, (Online Resource accessed via GBIF.org).
 Retrieved from \href{\%7Bhttps://doi.org/10.15468/39omei\%20\%7D}{\{https://doi.org/10.15468/39omei \}}
 
-\leavevmode\vadjust pre{\hypertarget{ref-gibb2015new}{}}%
+\leavevmode\hypertarget{ref-gibb2015new}{}%
 Gibb, G.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Hedges2015}{}}%
+\leavevmode\hypertarget{ref-harvey1991comparative}{}%
+Harvey, P. H., Pagel, M. D., \& others. (1991). \emph{The comparative method in evolutionary biology} (Vol. 239). Oxford university press Oxford.
+
+\leavevmode\hypertarget{ref-Hedges2015}{}%
 Hedges, S.
@@
 (2013).
-Multilocus phylogeny and biogeography of the new world pheucticus grosbeaks (aves: cardinalidae).
+Multilocus phylogeny and biogeography of the new world pheucticus grosbeaks (aves: Cardinalidae).
 \emph{Molecular Phylogenetics and Evolution}, \emph{69}(3), 1222--1227.
 
-\leavevmode\vadjust pre{\hypertarget{ref-RCoreTeam2018}{}}%
-R Core Team. (2018). \emph{{R: a language and environment for statistical computing}}. Vienna, Austria: R Foundation for Statistical Computing.
-
-\leavevmode\vadjust pre{\hypertarget{ref-rabosky2018inverse}{}}%
+\leavevmode\hypertarget{ref-rabosky2015no}{}%
+Rabosky, D.
+L.
+(2015).
+No substitute for real data: A cautionary note on the use of phylogenies from birth--death polytomy resolvers for downstream comparative analyses.
+\emph{Evolution}, \emph{69}(12), 3207--3216.
+
+\leavevmode\hypertarget{ref-rabosky2018inverse}{}%
 Rabosky, D.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-datelifeplot}{}}%
+\leavevmode\hypertarget{ref-roquet2014one}{}%
+Roquet, C., Lavergne, S., \& Thuiller, W. (2014). One tree to link them all: A phylogenetic dataset for the european tetrapoda. \emph{PLoS Currents}, \emph{6}.
+
+\leavevmode\hypertarget{ref-datelifeplot}{}%
 Sanchez-Reyes, L.
@@
 (2013).
-{Phylotastic!
-Making tree-of-life knowledge accessible, reusable and convenient}.
-\emph{{BMC Bioinformatics}}, \emph{14}.
+Phylotastic!
+Making tree-of-life knowledge accessible, reusable and convenient.
+\emph{BMC Bioinformatics}, \emph{14}.
 \url{https://doi.org/10.1186/1471-2105-14-158}
 
-\leavevmode\vadjust pre{\hypertarget{ref-tietze2013complete}{}}%
+\leavevmode\hypertarget{ref-thomas2013pastis}{}%
+Thomas, G.
+H., Hartmann, K., Jetz, W., Joy, J.
+B., Mimoto, A., \& Mooers, A.
+O.
+(2013).
+PASTIS: An r package to facilitate phylogenetic assembly with soft taxonomic inferences.
+\emph{Methods in Ecology and Evolution}, \emph{4}(11), 1011--1017.
+
+\leavevmode\hypertarget{ref-tietze2013complete}{}%
 Tietze, D.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-vos2012nexml}{}}%
+\leavevmode\hypertarget{ref-uyeda2017evolution}{}%
+Uyeda, J.
+C., Pennell, M.
+W., Miller, E.
+T., Maia, R., \& McClain, C.
+R.
+(2017).
+The evolution of energetic scaling across the vertebrate tree of life.
+\emph{The American Naturalist}, \emph{190}(2), 185--199.
+
+\leavevmode\hypertarget{ref-vos2012nexml}{}%
 Vos, R.

### 2023-01-23 manuscript/manuscript-1st-submission.Rmd
differences latex file and pdf

@@
 
-Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used [@magallon2010using, @magallon2015metacalibrated].
+Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used [@magallon2010using; @magallon2015metacalibrated].
 Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.

### 2023-01-23 manuscript/manuscript-1st-submission.tex
differences latex file and pdf

@@
 Analyses can be performed via DateLife's interactive website at \url{www.datelife.org}, or using the \texttt{datelife} R package.
-Details on the R functions used to perform the analyses are available from \texttt{datelife}'s R package vignettes at \url{phylotastic.org/datelife}.}}
+Details on the R functions used to perform the analyses are available from \texttt{datelife}'s R package vignettes at \url{https://phylotastic.org/datelife}.}}
 \label{fig:workflow}
@@
 
-Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used (Magallón, 2010, @magallon2015metacalibrated).
+Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used (Magallón, 2010; Magallón et al., 2015).
 Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.

### 2023-01-23 manuscript/manuscript.tex
differences latex file and pdf

@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 5355}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 5386}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 DateLife's core software application is available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org.
-It features key elements for scientific reproducibility, such as a versioned, open and fully public source database (McTavish et al., 2015), data stored and available in a computer readable format (Vos et al., 2012), automated and programmatic ways of accessing the data (Stoltzfus et al., 2013) and methods to summarize and compare the data.
+It features key elements for scientific reproducibility, such as a versioned, open and fully public source database (McTavish et al., 2015) that stores data in a computer-readable format (Vos et al., 2012); automated and programmatic ways of accessing and downloading the data in a compuer-redable format also (Stoltzfus et al., 2013); and methods to summarize and compare the data.
 
@@
 DateLife's core software application consists of the R package \texttt{datelife}.
-Its current stable version -- v0.6.6, is available from the The Comprehensive R Archive Network (CRAN) repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
+Its current stable version -- v0.6.6, is available from the Comprehensive R Archive Network (CRAN) repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
 ape (Paradis, Claude, \& Strimmer, 2004),
@@
 
-The final step entailed congruifying and summarizing the age data available for the Fringillidae species into single summary chronograms, using two different types of summary ages, median and SDM.
+The final step entailed congruifying and summarizing the age data available for the Fringillidae species into two single summary chronograms, using two different types of summary ages, median and SDM.
 As explained in the \enquote{Description} section, a tree topology to summarize age data upon is required.
-By default, to do this, DateLife uses the topology from OpenTree's synthetic tree that contains species from the search query.
-According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).
+By default, to do this, DateLife uses the topology from OpenTree's synthetic tree that contains all taxa from the search query.
+According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig.
+\ref{fig:fringillidae-topologies}).
+Hence, a topology containing only the 289 species from the original query was extracted from Open Tree of Life's synthetic tree v12.3 (Open Tree Of Life et al., 2019).
 

### 2023-01-25 manuscript/manuscript-1st-submission.Rmd
parsimony comment

@@
 DateLife's core software application consists of the R package `datelife`.
-Its current stable version -- v`r packageVersion("datelife")`, is available from the The Comprehensive R Archive Network (CRAN) repository [@datelifeR], and relies on functionalities from various biological R packages:
+Its current stable version -- v`r packageVersion("datelife")`, is available from The Comprehensive R Archive Network (CRAN) repository [@datelifeR], and relies on functionalities from various biological R packages:
 ape [@paradis2004],
@@
 DateLife requires a tree topology to summarize age data upon.
-We recommend that users provide a tree topology as input from the literature, or one of their own making.
+We recommend that users provide as input a tree topology from the literature, or one of their own making.
 If no topology is provided, DateLife automatically extracts one from the OpenTree synthetic tree, a phylogeny encompassing 2.3 million taxa across all life, assembled from 1, 239 published phylogenetic trees and OpenTree's unified Taxonomy, OTT  [@opentreeoflife2019synth].

### 2023-01-25 manuscript/manuscript-1st-submission.tex
parsimony comment

@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4960}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4959}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 DateLife's core software application consists of the R package \texttt{datelife}.
-Its current stable version -- v0.6.6, is available from the The Comprehensive R Archive Network (CRAN) repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
+Its current stable version -- v0.6.6, is available from The Comprehensive R Archive Network (CRAN) repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
 ape (Paradis, Claude, \& Strimmer, 2004),
@@
 DateLife requires a tree topology to summarize age data upon.
-We recommend that users provide a tree topology as input from the literature, or one of their own making.
+We recommend that users provide as input a tree topology from the literature, or one of their own making.
 If no topology is provided, DateLife automatically extracts one from the OpenTree synthetic tree, a phylogeny encompassing 2.3 million taxa across all life, assembled from 1, 239 published phylogenetic trees and OpenTree's unified Taxonomy, OTT (Open Tree Of Life et al., 2019).

### 2023-01-25 manuscript/manuscript.Rmd
parsimony comment

@@
 <!-- If you're aiming for syst bio you don't want to start so general Readers know what chronograms and phylogenies are!.-->
-Chronograms --phylogenies with branch lengths proportional to time-- provide key data on evolutionary time frame for the study of natural processes in many areas of biological research, such as comparative analysis [@harvey1991comparative; @freckleton2002phylogenetic], developmental biology [@laubichler2009form; @delsuc2018phylogenomic], conservation and ecology [@Felsenstein1985a; @Webb2000], historical biogeography [@posadas2006historical], and species diversification [@magallon2001absolute; @Morlon2014].
+Chronograms --phylogenies with branch lengths proportional to time-- provide key data on evolutionary time frame for the study of natural processes in many areas of biological research, such as comparative analysis [@harvey1991comparative; @freckleton2002phylogenetic], developmental biology [@laubichler2009form; @delsuc2018phylogenomic], conservation biology and ecology [@Felsenstein1985a; @Webb2000], historical biogeography [@posadas2006historical], and species diversification [@magallon2001absolute; @Morlon2014].
 
@@
 DateLife's core software application consists of the R package `datelife`.
-Its current stable version -- v`r packageVersion("datelife")`, is available from the Comprehensive R Archive Network (CRAN) repository [@datelifeR], and relies on functionalities from various biological R packages:
+Its current stable version -- v`r packageVersion("datelife")`, is available from The Comprehensive R Archive Network (CRAN) repository [@datelifeR], and relies on functionalities from various biological R packages:
 ape [@paradis2004],
@@
 
-DateLife starts by processing an input consisting of at least one taxon scientific names. Two or more scientific names can be provided as a comma separated character string or as tip labels on a tree.
+DateLife starts by processing an input consisting of at least one taxon scientific name. Multiple input names can be provided as a comma separated character string or as tip labels on a tree.
 If the input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004].
@@
 The processed taxon names are saved as an R object of a newly defined class, \texttt{datelifeQuery}, that is used in the following steps.
-This object contains the names standardized to the taxonomy of choice, the corresponding OTT id numbers, and the topology of the input tree if one was provided.
+This object contains the input names standardized to a taxonomy of choice (OTT by default), the corresponding OTT id numbers, and the topology of an input tree, if one was provided.
 
@@
 DateLife requires a tree topology to summarize age data upon.
-We recommend that users provide a tree topology as input from the literature, or one of their own making.
+We recommend that users provide as input a tree topology from the literature, or one of their own making.
 If no topology is provided, DateLife automatically extracts one from the OpenTree synthetic tree, a phylogeny currently encompassing 2.3 million taxa across all life, assembled from 1, 239 published phylogenetic trees and OpenTree's unified Taxonomy, OTT [@opentreeoflife2019synth].
-Alternatively, DateLife can combine topologies from source chronograms using a supertree approach.
+Alternatively, DateLife can combine topologies from source chronograms using a supertree approach [@Criscuolo2006].
 To do this DateLife first identifies the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
@@
 By default, DateLife implements the Branch Length Adjuster [BLADJ; @Webb2008; @webb2005phylomatic] algorithm to obtain a fully dated topology. 
-BLADJ is the only dating algorithm that can work with initial topologies without any branch length data.
+BLADJ takes an initial topology and ignores branch length data if present.
+Next, it fixes ages on nodes with calibration data, and distributes time for nodes with no data evenly between calibrated nodes, minimizing age variance in the resulting chronogram [@Webb2008].
+This method is useful for ecological analyses [@Webb2008].
+When there is conflict between ages of calibrated nodes, ignores node ages that are older than the age of a parent node.
+BLADJ requires a root age estimate.
+If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature.
+If none is provided, DateLife will not return a dated topology and will provide a warning message along with suggestions on how the user can provide an age for the root so the dating analysis can be finished.
+<!-- assigns an arbitrary age to the root as 10% older than the oldest age available within the group. -->
 
-Alternatively, the user can chose phylogenetic dating options supported in DateLife that incorporate branch length information from the input topology in combination with the secondary calibrations:
-PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees;
-treePL [@Smith2012], is a semi-parametric, rate-smoothing, penalized likelihood dating method [@sanderson2002];
-the MrBayes  [@Huelsenbeck2001; @Ronquist2003] approach in DateLife uses the calibrations as priors on node ages.
+Alternatively, the user can choose to obtain a fully dated phylogeny 
+MrBayes [@Huelsenbeck2001; @Ronquist2003] approach in DateLife uses the calibrations as priors on node ages.
 
@@
 
-BLADJ fixes node ages that have calibration data, and distributes time between nodes with no data evenly between calibrated nodes.
-This has proven effective minimizes age variance in the resulting chronogram and useful for ecoogical analyses [@Webb2008].
-When there is conflict in ages between nodes with calibration data, BLADJ ignores node ages that are older than the age of a parent node.
-BLADJ requires a root age estimate.
-If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature.
-If none is provided, DateLife will not return a dated topology but provide a warning message along with suggestions on how the user can provide an age for the root so DateLife can run.
-<!-- assigns an arbitrary age to the root as 10% older than the oldest age available within the group. -->
+In the case the user can provide or obtain branch length data along with a tree topology, phylogenetic dating options supported in DateLife that incorporate branch length information from the input topology in combination with the secondary calibrations:
+PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees;
+treePL [@Smith2012], is a semi-parametric, rate-smoothing, penalized likelihood dating method [@sanderson2002].
 
@@
 
-The BOLD sequence alignment is then used to reconstruct branch lengths with the accelerated transformation (ACCTRAN) parsimony algorithm, which resolves ambiguous character optimization, by assigning changes along branches of the tree as close to the root as possible [@agnarsson2008acctran].
-This algorithm work rally fast and allows getting initial branch lengths that ar ethen optimized using ML.
+The BOLD sequence alignment is then used to obtain initial branch lengths with the accelerated transformation (ACCTRAN) parsimony algorithm, which resolves ambiguous character optimization, by assigning changes along branches of the tree as close to the root as possible [@agnarsson2008acctran].
+This results in older nodes [@forest2005teasing].
 Optionally, the likelihood of the tree topology, the alignment and the reconstructed branch lengths given different evolutionary models, is computed using functions from the `phangorn` package [@schliep2011phangorn].

### 2023-01-25 manuscript/manuscript.tex
parsimony comment

@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 5386}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 5400}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 
-Chronograms --phylogenies with branch lengths proportional to time-- provide key data on evolutionary time frame for the study of natural processes in many areas of biological research, such as comparative analysis (Freckleton, Harvey, \& Pagel, 2002; Harvey, Pagel, \& others, 1991), developmental biology (Delsuc et al., 2018; Laubichler \& Maienschein, 2009), conservation and ecology (Felsenstein, 1985; Webb, 2000), historical biogeography (Posadas, Crisci, \& Katinas, 2006), and species diversification (Magallon \& Sanderson, 2001; Morlon, 2014).
+Chronograms --phylogenies with branch lengths proportional to time-- provide key data on evolutionary time frame for the study of natural processes in many areas of biological research, such as comparative analysis (Freckleton, Harvey, \& Pagel, 2002; Harvey, Pagel, \& others, 1991), developmental biology (Delsuc et al., 2018; Laubichler \& Maienschein, 2009), conservation biology and ecology (Felsenstein, 1985; Webb, 2000), historical biogeography (Posadas, Crisci, \& Katinas, 2006), and species diversification (Magallon \& Sanderson, 2001; Morlon, 2014).
 
@@
 DateLife's core software application consists of the R package \texttt{datelife}.
-Its current stable version -- v0.6.6, is available from the Comprehensive R Archive Network (CRAN) repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
+Its current stable version -- v0.6.6, is available from The Comprehensive R Archive Network (CRAN) repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
 ape (Paradis, Claude, \& Strimmer, 2004),
@@
 
-DateLife starts by processing an input consisting of at least one taxon scientific names. Two or more scientific names can be provided as a comma separated character string or as tip labels on a tree.
+DateLife starts by processing an input consisting of at least one taxon scientific name. Multiple input names can be provided as a comma separated character string or as tip labels on a tree.
 If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a \enquote{phylo} R object (Paradis et al., 2004).
@@
 DateLife requires a tree topology to summarize age data upon.
-We recommend that users provide a tree topology as input from the literature, or one of their own making.
+We recommend that users provide as input a tree topology from the literature, or one of their own making.
 If no topology is provided, DateLife automatically extracts one from the OpenTree synthetic tree, a phylogeny currently encompassing 2.3 million taxa across all life, assembled from 1, 239 published phylogenetic trees and OpenTree's unified Taxonomy, OTT (Open Tree Of Life et al., 2019).
-Alternatively, DateLife can combine topologies from source chronograms using a supertree approach.
+Alternatively, DateLife can combine topologies from source chronograms using a supertree approach (Criscuolo, Berry, Douzery, \& Gascuel, 2006).
 To do this DateLife first identifies the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
@@
 
-For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method (SDM; Criscuolo, Berry, Douzery, \& Gascuel, 2006), which deforms patristic distance matrices by minimizing variance and then averaging them.
+For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method (SDM; Criscuolo et al., 2006), which deforms patristic distance matrices by minimizing variance and then averaging them.
 These single summary taxon pair age matrices are stored as summarized calibrations that can be used as secondary calibrations to date a tree topology - with or without initial branch lengths, using phylogenetic dating methods currently supported within DateLife: BLADJ (Webb, Ackerly, \& Kembel, 2008; Webb \& Donoghue, 2005), MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), and treePL (Smith \& O'Meara, 2012).
@@
 By default, DateLife implements the Branch Length Adjuster (BLADJ; Webb et al., 2008; Webb \& Donoghue, 2005) algorithm to obtain a fully dated topology.
-BLADJ is the only dating algorithm that can work with initial topologies without any branch length data.
-
-Alternatively, the user can chose phylogenetic dating options supported in DateLife that incorporate branch length information from the input topology in combination with the secondary calibrations:
-PATHd8 is a non-clock, rate-smoothing method (Britton et al., 2007) to date trees;
-treePL (Smith \& O'Meara, 2012), is a semi-parametric, rate-smoothing, penalized likelihood dating method (Sanderson, 2002);
-the MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003) approach in DateLife uses the calibrations as priors on node ages.
-
-The latter methods can assign dates using a birth-death model (all of them????), they require initial branch lengths and thus require more time and expertise to run, which makes BLADJ the fastest and most practical method to obtain a dated tree.
-
-BLADJ fixes node ages that have calibration data, and distributes time between nodes with no data evenly between calibrated nodes.
-This has proven effective minimizes age variance in the resulting chronogram and useful for ecoogical analyses (Webb et al., 2008).
+BLADJ takes an initial topologies without any branch length data, fixes node ages that have calibration data, and distributes time between nodes with no data evenly between calibrated nodes.
+This has proven effective to minimize age variance in the resulting chronogram and useful for ecological analyses (Webb et al., 2008).
 When there is conflict in ages between nodes with calibration data, BLADJ ignores node ages that are older than the age of a parent node.
@@
 If none is provided, DateLife will not return a dated topology but provide a warning message along with suggestions on how the user can provide an age for the root so DateLife can run.
+
+Alternatively, the user can use the MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003) approach in DateLife uses the calibrations as priors on node ages.
+
+The latter methods can assign dates using a birth-death model (all of them????), they require initial branch lengths and thus require more time and expertise to run, which makes BLADJ the fastest and most practical method to obtain a dated tree.
+
+In the case the user can provide or obtain branch length data along with a tree topology, phylogenetic dating options supported in DateLife that incorporate branch length information from the input topology in combination with the secondary calibrations:
+PATHd8 is a non-clock, rate-smoothing method (Britton et al., 2007) to date trees;
+treePL (Smith \& O'Meara, 2012), is a semi-parametric, rate-smoothing, penalized likelihood dating method (Sanderson, 2002).
 

### 2023-01-27 manuscript/manuscript.Rmd
adding references

@@
 However, it is unknown how using very different age data could affect hypothesis testing in other areas of research and more effort is needed in that direction -->
+Diversification analysis of simulated and empirical phylogenies show that using a more complete phylogeny is always better than using an incompletely sampled phylogeny [@cusimano2012new; @chang2020estimating; @sun2020estimating].
+Ideally phylogenies should be completed using genetic data, but this is a very hard task to complete for many taxa.
+Completing a phylogeny using birth-death politomy resolvers and a backbone taxonomy seem to provide less biased estimates in some cases [@chang2020estimating], but can also introduce spurious patterns.
 
@@
 Adding branch lengths sampled from a birth-death model in the absence of genetic data
-has been found to improve insight in phylogeny-based analyses.
+has been found to improve insight for phylogeny-based analyses of diversification rates [@chang2020estimating].
+@chang2020estimating "demonstrate theoretical concerns with the widespread use of analytical sampling corrections for sparsely sampled phylogenies of higher taxonomic groups.
+In particular, corrections based on “sampling fractions” can lead to low statistical power to infer rate variation when it is present, depending on the likelihood function used for inference.
+In the extreme, the sampling fraction correction can lead to spurious patterns of diversification that are driven solely by unbalanced sampling across the tree in concert with low overall power to infer shifts."
+
+"birth–death polytomy resolver is a method of accounting for phylogenetic uncertainty that places missing (unsampled) taxa onto phylogenetic trees, using taxonomic information alone"
 
@@
 
-Simulating branch lengths following a birth-death species diversification model for missing taxa in chronograms with non random sampling, lowers type I error (false positive, incorrectly accepting the null hypothesis of a constant-rate or temporally varying rate birth-death model) when analyzing changes in diversification rate using the gamma statistic [@cusimano2012new].
+Simulating branch lengths following a birth-death species diversification model for missing taxa to chronograms with non random sampling, lowers type I error (false positive, incorrectly accepting the null hypothesis of a constant-rate or temporally varying rate birth-death model) when analyzing changes in diversification rate using the gamma statistic [@cusimano2012new].
 
@@
 
+@rabosky2015no shows that randomly resolved phylogenies are inappropriate to study trait evolution, because unresolved lineages are placed on the phylogeny at random relative to the trait values.
+At best, these phylogenies could serve as a null hypothesis of randomness in trait evolution.
+
+@chang2020estimating state that stochastic politomy resolution alleviate low statistical power to identify shifts in diversification rate with methods that account for sampling fractions.
+
+@title2019tip test that estimating species diversification rates with model free methods and BAMM is good for speciation rate, but less good for net diversification rate, and bad for extinction rates.
+They test this in complete phylogenies.
+They do not test possible estimate biases using phylogenies with randomly resolved politomies.
+
+@sun2020estimating used a 50% species level sampled phylogeny of the family Cucurbitaceae to estimate speciation rates with two parametric methods (RPANDA and BAMM) and one semi-parametric method (DR) and test the effect of using incompletely sampled phylogenies on one hand, and incomplete phylogenies that have been made near complete, using a stochastic politomy resolver based on taxonomy on the other hand (with two different methods: 1) adding species that were removed at random following a backbone taxonomy, and 2) adding missing species to a genus level taxonomy based on a taxonomy resolver only).
+When using incomplete phylogenies, parametric methods tend to overestimate speciation rate as sampling gets worse (less complete), while semi-parametric methods tend to underestimate speciation rate as sampling gets worse.
+When using taxonomically resolved complete phylogenies, only one parametric method overestimates speciation rate the more incomplete original sampling was.
+RPANDA, BAMM tip rates and DR all underestimate speciation rates as original sampling was more incomplete.
+"Adding backbone taxa tends to result in the inference of spurious early bursts of diversification", bias in sampling ancestral nodes, failure to sample recent divergences or pushing them back in relative time (making recent divergences older than expected)
+"failure to detect recent bursts of speciation and instead inferring higher rates of evolution at earlier time intervals
+
 Notably, risks come with this practice.
@@
 Taken to the extreme, one could generate a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration, using polytomy resolution and branch length simulation methods.
-There has yet to be a thorough analysis of what can go wrong when one extends inferences beyond the data in this way, so we urge caution; we also urge readers to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, and assessing the consistency of the results with prior work.
+
+We urge readers to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, and assessing the consistency of the results with prior work.
+
+However, analysis of empirical phylogenies suggest that introducing stochastically resolved divergence times creates a spurious pattern of early burst of diversification.
+Similarly there are worries that patterns of morphological evolution cannot be accurately inferred with these types of phylogenies, as any patterns would be erased by randomization.
+We note that the same could apply for geographical and morphological dependent diversification analysis.
+Hence, we suggest that these types of randomly resolved phylogenies can work as null or neutral models, representing the case of diversification independent of traits and geographical scenario.
 

### 2023-02-13 manuscript/manuscript.Rmd
redraft of downstream anaylisys discussion section

--- 2023-01-25 manuscript/manuscript.Rmd
@@
 
+While it is known that incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance in ecological and conservation biology studies [@Webb2008] and diversiifcation [@], it is important to note possible sources of problems when reusing chronogram data the way DateLife does.
+
 ## Age variation in source chronograms
@@
 
-The exercise of summarizing age data from across multiple studies provides the opportunity to work with a more inclusive chronogram,
-that reflects a unified evolutionary history for a lineage, by putting together evidence from different hypotheses.
+The exercise of summarizing age data from across multiple studies provides the opportunity to work with a chronogram that reflects a unified evolutionary history for a lineage, by putting together evidence from different hypotheses.
 The largest, and taxonomically broadest chronogram currently available from OpenTree was constructed summarizing age data from 2,274 published chronograms using NCBI's taxonomic tree as backbone [@Hedges2015].
@@
 
-## Effects on downstream analyses
+## Effects of phylogenetic sampling on downstream analyses
 
-The study of phenomena dependent on the timing of species diversification events, such as macroevolutionary processes, is affected by the usage of alternative chronograms that vary in topology [@rabosky2015no; @title2016macrophylogenies].
+Analysis of species diversification using simulated and empirical phylogenies suggest that using a more completely sampled phylogeny provides estimates that are closer to the true diversification history than when analysing incompletely sampled phylogenies [@cusimano2012new; @chang2020estimating; @sun2020estimating].
+Ideally phylogenies should be completed using genetic data, but this is a very hard task to achieve for many taxa.
+Hence, DateLife's workflow features different ways of generating node ages in the absence of calibration and branch length information for certain taxa.
+Completing a phylogeny using birth-death politomy resolvers and a backbone taxonomy is a common practice in scientific publications: @Jetz2012, created a chronogram of all 9, 993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a chronogram of 31, 536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a chronogram of 353, 185 seed plants where only 23% had molecular data.
+These phylogenies provide less biased diversification rates estimated with methods that account for sampling fractions [@chang2020estimating; @cusimano2012new], but can also introduce spurious patterns [@].
+Analysis of empirical phylogenies suggest that introducing stochastically resolved divergence times creates a spurious pattern of early burst of diversification.
+
+Taxonomy based politomy resolvers also introduce topological differences.
+The study of macroevolutionary processes largely depends on an understanding of the timing of species diversification events, and different phylogenetic and chronogram hypothesis can provide very different overviews of the macroeovlutionary history of a biological group.
+For example, alternative topologies in chronograms from the same biological group can infer very different species diversification patterns [@rabosky2015no; @title2016macrophylogenies].
 <!-- There is evidence that different chronograms will result in very different macroevolutionary histories inferred from them [@title2016macrophylogenies].
 However, it is unknown how using very different age data could affect hypothesis testing in other areas of research and more effort is needed in that direction -->
+Similarly there are worries that patterns of morphological evolution cannot be accurately inferred with these types of phylogenies, as any patterns would be erased by randomization.
+We note that the same could apply for geographical and morphological dependent diversification analysis.
+Hence, we suggest that these types of randomly resolved phylogenies can work as null or neutral models, representing the case of diversification independent of traits and geographical scenario.
 
-On the other hand, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance in ecological and conservation biology studies [@Webb2008].
-Hence, DateLife's workflow features different ways of generating node ages in the absence of calibration and branch length information for certain taxa.
-
-Adding branch lengths sampled from a birth-death model in the absence of genetic data
-has been found to improve insight in phylogeny-based analyses.
-
-is a common practice in scientific publications: @Jetz2012, created a chronogram of all 9, 993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a chronogram of 31, 536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a chronogram of 353, 185 seed plants where only 23% had molecular data.
-
-Simulating branch lengths following a birth-death species diversification model for missing taxa in chronograms with non random sampling, lowers type I error (false positive, incorrectly accepting the null hypothesis of a constant-rate or temporally varying rate birth-death model) when analyzing changes in diversification rate using the gamma statistic [@cusimano2012new].
-
-@thomas2013pastis hypothesize that results of diversification analyses will be biased towards the birth-death model used to simulate branch lengths;
-and note that "the effects of missing species placement or polytomy resolution are less clear for other phylogeny-based analyses (e.g.
-correlates of diversification, modelling trait evolution, community phylogenetics), and future work should test how the treatment of missing species influences both parameter estimation and type I and II errors"
-
-Notably, risks come with this practice.
-
+Notably, risks come with the practice of resolving politomies, even if using a tested model of diversification.
 Taken to the extreme, one could generate a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration, using polytomy resolution and branch length simulation methods.
-There has yet to be a thorough analysis of what can go wrong when one extends inferences beyond the data in this way, so we urge caution; we also urge readers to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, and assessing the consistency of the results with prior work.
+We urge DateLife users to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, and assessing the consistency of the results with prior work.
 

### 2023-02-13 notes-effects-downstream-analysis-review.Rmd.R

@@ -1,0 +1,110 @@
+---
+title             : "Review: Effects of initial phylogenetic and chronogram hypothesis on evolutionary downstream analyses"
+shorttitle        : "Litt review"
+
+author:
+  - name          : "Luna L. Sánchez Reyes"
+    affiliation   : "1,2"
+    corresponding : yes    # Define only one corresponding author
+    address       : ""
+    email         : "sanchez.reyes.luna@gmail.com"
+    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
+      - Data curation
+      - Investigation
+      - Software
+      - Visualization
+      - Validation
+      - Writing
+      - Original Draft Preparation
+      - Writing
+      - Review & Editing
+
+affiliation:
+  - id            : "1"
+    institution   : "University of California, Merced, USA"
+  - id            : "2"
+    institution   : "University of Tennessee, Knoxville, USA"
+
+authornote: |
+  School of Natural Sciences, University of California, Merced, 258 Science and Engineering Building 1, Merced, CA 95340, USA.
+
+  Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 446 Hesler Biology Building, Knoxville, TN 37996, USA.
+
+bibliography      : ["paper_references.bib"]
+keywords          : "Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations"
+wordcount         : "`r wordcountaddin::word_count()`"
+floatsintext      : no
+figurelist        : no
+tablelist         : no
+footnotelist      : no
+linenumbers       : yes
+mask              : no
+draft             : no
+
+documentclass     : "apa6"
+classoption       : "man"
+output            :
+  papaja::apa6_pdf:
+    includes      :
+      in_header: "preamble.tex"
+---
+
+```{r setup, include = FALSE}
+library("papaja")
+r_refs("paper_references.bib")
+embed_figure <- TRUE
+```
+
+@cusimano2012new: Simulating branch lengths following a birth-death species diversification model for
+missing taxa to chronograms with non random sampling, lowers type I error (false positive,
+incorrectly accepting the null hypothesis of a constant-rate or temporally varying rate
+birth-death model) when analyzing changes in diversification rate using the gamma statistic.
+
+@thomas2013pastis hypothesize that results of diversification analyses will be biased towards
+the birth-death model used to simulate branch lengths;
[…]

### 2023-02-13 manuscript/manuscript.Rmd
tidy up section on Effects on downstream analyses

@@
 DateLife's core software application is available as an R package [@datelifeR], and as an online Rshiny interactive website at www.datelife.org.
-It features key elements for scientific reproducibility, such as a versioned, open and fully public source database [@mctavish2015phylesystem] that stores data in a computer-readable format [@vos2012nexml]; automated and programmatic ways of accessing and downloading the data in a compuer-redable format also [@Stoltzfus2013]; and methods to summarize and compare the data.
+It features key elements for scientific reproducibility, such as a versioned, open and fully public source database [@mctavish2015phylesystem] that stores data in a computer-readable format [@vos2012nexml]; automated and programmatic ways of accessing and downloading the data in a computer-readable format also [@Stoltzfus2013]; and methods to summarize and compare the data.
 
@@
 Alternatively, DateLife can combine topologies from source chronograms using a supertree approach [@Criscuolo2006].
-To do this DateLife first identifies the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
+To do this, DateLife first identifies the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
 [-@ane2009groves].
@@
 As explained in the "Description" section, a tree topology to summarize age data upon is required.
-By default, to do this, DateLife uses the topology from OpenTree's synthetic tree that contains all taxa from the search query.
+By default, DateLife uses the topology from OpenTree's synthetic tree that contains all taxa from the search query.
 According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig.
@@
 
-While it is known that incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance in ecological and conservation biology studies [@Webb2008] and diversiifcation [@], it is important to note possible sources of problems when reusing chronogram data the way DateLife does.
+While it is known that incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance in ecological and conservation biology studies [@Webb2008].
+Next, we discuss some particularities of using summarized node age data from a DateLife analysis that users should be aware of.
 
@@
 Analysis of species diversification using simulated and empirical phylogenies suggest that using a more completely sampled phylogeny provides estimates that are closer to the true diversification history than when analysing incompletely sampled phylogenies [@cusimano2012new; @chang2020estimating; @sun2020estimating].
-Ideally phylogenies should be completed using genetic data, but this is a very hard task to achieve for many taxa.
-Hence, DateLife's workflow features different ways of generating node ages in the absence of calibration and branch length information for certain taxa.
-Completing a phylogeny using birth-death politomy resolvers and a backbone taxonomy is a common practice in scientific publications: @Jetz2012, created a chronogram of all 9, 993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a chronogram of 31, 536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a chronogram of 353, 185 seed plants where only 23% had molecular data.
-These phylogenies provide less biased diversification rates estimated with methods that account for sampling fractions [@chang2020estimating; @cusimano2012new], but can also introduce spurious patterns [@].
-Analysis of empirical phylogenies suggest that introducing stochastically resolved divergence times creates a spurious pattern of early burst of diversification.
+Ideally phylogenies should be completed using genetic data, but this is a very hard task to achieve for many biological groups.
+Hence, DateLife's workflow features different ways of assigning node ages in the absence of calibrations and branch length information for certain taxa.
+Completing a phylogeny using a stochastic birth-death politomy resolver and a backbone taxonomy is a common practice in scientific publications: @Jetz2012, created a chronogram of all 9, 993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a chronogram of 31, 536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a chronogram of 353, 185 seed plants where only 23% had molecular data.
+These stochastically resolved chronograms provide less biased diversification rates estimated with methods that account for sampling fractions [@chang2020estimating; @cusimano2012new], but can also introduce spurious patterns of early bursts of diversification [@cusimano2010slowdowns; @sun2020estimating].
 
@@
 However, it is unknown how using very different age data could affect hypothesis testing in other areas of research and more effort is needed in that direction -->
-Similarly there are worries that patterns of morphological evolution cannot be accurately inferred with these types of phylogenies, as any patterns would be erased by randomization.
-We note that the same could apply for geographical and morphological dependent diversification analysis.
-Hence, we suggest that these types of randomly resolved phylogenies can work as null or neutral models, representing the case of diversification independent of traits and geographical scenario.
+Similarly there are worries that patterns of morphological evolution cannot be accurately inferred with these types of phylogenies, as any patterns would be erased by randomization [@rabosky2015no].
+We note that the same applies for geographical and morphological dependent diversification analysis.
+Hence, we suggest that these types of randomly resolved phylogenies can be used as null or neutral models, representing the case of diversification independent of traits and geographical scenario.
 
-Notably, risks come with the practice of resolving politomies, even if using a tested model of diversification.
-Taken to the extreme, one could generate a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration, using polytomy resolution and branch length simulation methods.
-We urge DateLife users to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, and assessing the consistency of the results with prior work.
+While stochastic politomy resolvers have been useful to advance research, notably, risks come with this practice.
+Taken to the extreme, one could generate a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy, a single calibration, and assigning branch lengths following a birth-death diversification model.
+Clearly, this can lead to a misrepresentation of the true evolutionary history.
+We urge DateLife users to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, potential biases, and assessing the consistency of the results with prior work.
 

### 2023-02-13 manuscript/manuscript.tex
tidy up section on Effects on downstream analyses

@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 5400}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 5480}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 DateLife's core software application is available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org.
-It features key elements for scientific reproducibility, such as a versioned, open and fully public source database (McTavish et al., 2015) that stores data in a computer-readable format (Vos et al., 2012); automated and programmatic ways of accessing and downloading the data in a compuer-redable format also (Stoltzfus et al., 2013); and methods to summarize and compare the data.
+It features key elements for scientific reproducibility, such as a versioned, open and fully public source database (McTavish et al., 2015) that stores data in a computer-readable format (Vos et al., 2012); automated and programmatic ways of accessing and downloading the data in a computer-readable format also (Stoltzfus et al., 2013); and methods to summarize and compare the data.
 
@@
 Alternatively, DateLife can combine topologies from source chronograms using a supertree approach (Criscuolo, Berry, Douzery, \& Gascuel, 2006).
-To do this DateLife first identifies the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
+To do this, DateLife first identifies the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
 (2009).
@@
 By default, DateLife implements the Branch Length Adjuster (BLADJ; Webb et al., 2008; Webb \& Donoghue, 2005) algorithm to obtain a fully dated topology.
-BLADJ takes an initial topologies without any branch length data, fixes node ages that have calibration data, and distributes time between nodes with no data evenly between calibrated nodes.
-This has proven effective to minimize age variance in the resulting chronogram and useful for ecological analyses (Webb et al., 2008).
-When there is conflict in ages between nodes with calibration data, BLADJ ignores node ages that are older than the age of a parent node.
+BLADJ takes an initial topology and ignores branch length data if present.
+Next, it fixes ages on nodes with calibration data, and distributes time for nodes with no data evenly between calibrated nodes, minimizing age variance in the resulting chronogram (Webb et al., 2008).
+This method is useful for ecological analyses (Webb et al., 2008).
+When there is conflict between ages of calibrated nodes, ignores node ages that are older than the age of a parent node.
 BLADJ requires a root age estimate.
 If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature.
-If none is provided, DateLife will not return a dated topology but provide a warning message along with suggestions on how the user can provide an age for the root so DateLife can run.
-
-Alternatively, the user can use the MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003) approach in DateLife uses the calibrations as priors on node ages.
+If none is provided, DateLife will not return a dated topology and will provide a warning message along with suggestions on how the user can provide an age for the root so the dating analysis can be finished.
+
+Alternatively, the user can choose to obtain a fully dated phylogeny
+MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003) approach in DateLife uses the calibrations as priors on node ages.
 
@@
 
-The BOLD sequence alignment is then used to reconstruct branch lengths with the accelerated transformation (ACCTRAN) parsimony algorithm, which resolves ambiguous character optimization, by assigning changes along branches of the tree as close to the root as possible (Agnarsson \& Miller, 2008).
-This algorithm work rally fast and allows getting initial branch lengths that ar ethen optimized using ML.
+The BOLD sequence alignment is then used to obtain initial branch lengths with the accelerated transformation (ACCTRAN) parsimony algorithm, which resolves ambiguous character optimization, by assigning changes along branches of the tree as close to the root as possible (Agnarsson \& Miller, 2008).
+This results in older nodes (Forest et al., 2005).
 Optionally, the likelihood of the tree topology, the alignment and the reconstructed branch lengths given different evolutionary models, is computed using functions from the \texttt{phangorn} package (Schliep, 2011).
@@
 As explained in the \enquote{Description} section, a tree topology to summarize age data upon is required.
-By default, to do this, DateLife uses the topology from OpenTree's synthetic tree that contains all taxa from the search query.
+By default, DateLife uses the topology from OpenTree's synthetic tree that contains all taxa from the search query.
 According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig.
@@
 
+While it is known that incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance in ecological and conservation biology studies (Webb et al., 2008).
+Next, we discuss some particularities of using summarized node age data from a DateLife analysis that users should be aware of.
+
 \hypertarget{age-variation-in-source-chronograms}{%
@@
 
-The exercise of summarizing age data from across multiple studies provides the opportunity to work with a more inclusive chronogram,
-that reflects a unified evolutionary history for a lineage, by putting together evidence from different hypotheses.
+The exercise of summarizing age data from across multiple studies provides the opportunity to work with a chronogram that reflects a unified evolutionary history for a lineage, by putting together evidence from different hypotheses.
 The largest, and taxonomically broadest chronogram currently available from OpenTree was constructed summarizing age data from 2,274 published chronograms using NCBI's taxonomic tree as backbone (Hedges et al., 2015).
@@
 
-\hypertarget{effects-on-downstream-analyses}{%
-\subsection{Effects on downstream analyses}\label{effects-on-downstream-analyses}}
-
-The study of phenomena dependent on the timing of species diversification events, such as macroevolutionary processes, is affected by the usage of alternative chronograms that vary in topology (Rabosky, 2015; Title \& Rabosky, 2016).
-
-On the other hand, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance in ecological and conservation biology studies (Webb et al., 2008).
-Hence, DateLife's workflow features different ways of generating node ages in the absence of calibration and branch length information for certain taxa.
-
-Adding branch lengths sampled from a birth-death model in the absence of genetic data
-has been found to improve insight in phylogeny-based analyses.
-
-is a common practice in scientific publications: Jetz et al.
+\hypertarget{effects-of-phylogenetic-sampling-on-downstream-analyses}{%
+\subsection{Effects of phylogenetic sampling on downstream analyses}\label{effects-of-phylogenetic-sampling-on-downstream-analyses}}
+
+Analysis of species diversification using simulated and empirical phylogenies suggest that using a more completely sampled phylogeny provides estimates that are closer to the true diversification history than when analysing incompletely sampled phylogenies (Chang, Rabosky, \& Alfaro, 2020; Cusimano, Stadler, \& Renner, 2012; Sun et al., 2020).
+Ideally phylogenies should be completed using genetic data, but this is a very hard task to achieve for many biological groups.
+Hence, DateLife's workflow features different ways of assigning node ages in the absence of calibrations and branch length information for certain taxa.
+Completing a phylogeny using a stochastic birth-death politomy resolver and a backbone taxonomy is a common practice in scientific publications: Jetz et al.
 (2012), created a chronogram of all 9, 993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
 (2018) created a chronogram of 31, 536 ray-finned fishes, of which only 37\% had molecular data; Smith and Brown (2018) constructed a chronogram of 353, 185 seed plants where only 23\% had molecular data.
-
-Simulating branch lengths following a birth-death species diversification model for missing taxa in chronograms with non random sampling, lowers type I error (false positive, incorrectly accepting the null hypothesis of a constant-rate or temporally varying rate birth-death model) when analyzing changes in diversification rate using the gamma statistic (Cusimano, Stadler, \& Renner, 2012).
-
-Thomas et al. (2013) hypothesize that results of diversification analyses will be biased towards the birth-death model used to simulate branch lengths;
-and note that \enquote{the effects of missing species placement or polytomy resolution are less clear for other phylogeny-based analyses (e.g.~correlates of diversification, modelling trait evolution, community phylogenetics), and future work should test how the treatment of missing species influences both parameter estimation and type I and II errors}
-
-Notably, risks come with this practice.
-
-Taken to the extreme, one could generate a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration, using polytomy resolution and branch length simulation methods.
-There has yet to be a thorough analysis of what can go wrong when one extends inferences beyond the data in this way, so we urge caution; we also urge readers to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, and assessing the consistency of the results with prior work.
+These stochastically resolved chronograms provide less biased diversification rates estimated with methods that account for sampling fractions (Chang et al., 2020; Cusimano et al., 2012), but can also introduce spurious patterns of early bursts of diversification (Cusimano \& Renner, 2010; Sun et al., 2020).
+
+Taxonomy based politomy resolvers also introduce topological differences.
+The study of macroevolutionary processes largely depends on an understanding of the timing of species diversification events, and different phylogenetic and chronogram hypothesis can provide very different overviews of the macroeovlutionary history of a biological group.
+For example, alternative topologies in chronograms from the same biological group can infer very different species diversification patterns (Rabosky, 2015; Title \& Rabosky, 2016).
+Similarly there are worries that patterns of morphological evolution cannot be accurately inferred with these types of phylogenies, as any patterns would be erased by randomization (Rabosky, 2015).
+We note that the same applies for geographical and morphological dependent diversification analysis.
+Hence, we suggest that these types of randomly resolved phylogenies can be used as null or neutral models, representing the case of diversification independent of traits and geographical scenario.
+
+While stochastic politomy resolvers have been useful to advance research, notably, risks come with this practice.
+Taken to the extreme, one could generate a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy, a single calibration, and assigning branch lengths following a birth-death diversification model.
+Clearly, this can lead to a misrepresentation of the true evolutionary history.
+We urge DateLife users to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, potential biases, and assessing the consistency of the results with prior work.
 
@@
 
+\leavevmode\hypertarget{ref-chang2020estimating}{}%
+Chang, J., Rabosky, D.
+L., \& Alfaro, M.
+E.
+(2020).
+Estimating diversification rates on incompletely sampled phylogenies: Theoretical concerns and practical solutions.
+\emph{Systematic Biology}, \emph{69}(3), 602--611.
+
 \leavevmode\hypertarget{ref-chaves2013biogeography}{}%
@@
 \url{https://doi.org/10.1080/10635150600969872}
+
+\leavevmode\hypertarget{ref-cusimano2010slowdowns}{}%
+Cusimano, N., \& Renner, S. S. (2010). Slowdowns in diversification rates from real phylogenies may not be real. \emph{Systematic Biology}, \emph{59}(4), 458--464.
 
@@
 Felsenstein, J. (1985). Phylogenies and the Comparative Method. \emph{The American Naturalist}, \emph{125}(1), 1--15. Retrieved from \url{http://www.jstor.org/stable/2461605}
+
+\leavevmode\hypertarget{ref-forest2005teasing}{}%
+Forest, F., Savolainen, V., Chase, M.
+W., Lupia, R., Bruneau, A., \& Crane, P.
+R.
+(2005).
+Teasing apart molecular-versus fossil-based error estimates when dating phylogenetic trees: A case study in the birch family (betulaceae).
+\emph{Systematic Botany}, \emph{30}(1), 118--133.
 
@@
 
-\leavevmode\hypertarget{ref-thomas2013pastis}{}%
-Thomas, G.
-H., Hartmann, K., Jetz, W., Joy, J.
-B., Mimoto, A., \& Mooers, A.
-O.
-(2013).
-PASTIS: An r package to facilitate phylogenetic assembly with soft taxonomic inferences.
-\emph{Methods in Ecology and Evolution}, \emph{4}(11), 1011--1017.
+\leavevmode\hypertarget{ref-sun2020estimating}{}%
+Sun, M., Folk, R.
+A., Gitzendanner, M.
+A., Soltis, P.
+S., Chen, Z., Soltis, D.
+E., \& Guralnick, R.
+P.
+(2020).
+Estimating rates and patterns of diversification with incomplete sampling: A case study in the rosids.
+\emph{American Journal of Botany}, \emph{107}(6), 895--909.
 

### 2023-02-17 manuscript/manuscript-1st-submission.Rmd
remove figures from main text and finish answers to reviewer 1

@@
 r_refs("paper_references.bib")
-embed_figure <- TRUE
+embed_figure <- FALSE
 ```
@@
 
-_**Choosing a topology.--**_
+## Choosing a topology
 DateLife requires a tree topology to summarize age data upon.
@@
 
-_**Dating the topology.--**_
+## Dating the topology
 Input topologies from OpenTree or the supertree approach described above do not include branch length estimates of any kind.
@@
 
-_**Visualizing results.--**_
+## Visualizing results
 Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as visualize and compare results graphically, or construct their own graphs using DateLife's chronogram plot generation functions available from the R package `datelifeplot` [@datelifeplot].
@@
 
-### Primary vs Secondary calibrations
+## Primary vs Secondary calibrations
 

### 2023-02-17 manuscript/manuscript.Rmd
remove figures from main text and finish answers to reviewer 1

@@
 
-# Introduction
+<!-- # Introduction -->
 <!-- If you're aiming for syst bio you don't want to start so general Readers know what chronograms and phylogenies are!.-->
@@
 Building a chronogram is not an easy task.
-It requires obtaining and curating data to construct a phylogeny, selecting and placing appropriate calibrations on the phylogeny using independent age data points from the fossil record or other dated events, and inferring the full dated tree;
-it also generally requires specialized biological training, taxonomic domain knowledge, and a non-negligible amount of research time, computational resources and funding.
+It requires obtaining and curating a homology hypothesis to construct a phylogeny, selecting and placing appropriate calibrations on the phylogeny using independent age data points from the fossil record or other dated events, and inferring a full dated tree.
+All of this entails specialized biological training, taxonomic domain knowledge, and a non-negligible amount of research time, computational resources and funding.
 
-Here we present the DateLife project which has the main goal of capturing age data from published chronograms, and making these data readily accessible to the community for reuse and reanalysis, for research, teaching, and science communication and policy.
+Here we present the DateLife project which has the main goal of extracting and exposing age data from published chronograms, making age data readily accessible to the wider community for reuse and reanalysis, research, teaching, and science communication and policy.
 <!--  R has become mainstream in the biological-data analyst toolkit.
@@
 -->
-DateLife's core software application is available as an R package [@datelifeR], and as an online Rshiny interactive website at www.datelife.org.
-It features key elements for scientific reproducibility, such as a versioned, open and fully public source database [@mctavish2015phylesystem] that stores data in a computer-readable format [@vos2012nexml]; automated and programmatic ways of accessing and downloading the data in a computer-readable format also [@Stoltzfus2013]; and methods to summarize and compare the data.
+DateLife's core software application is available as an R package [@sanchez2022], and as an online Rshiny interactive website at www.datelife.org.
+It features key elements for scientific reproducibility, such as a curated, versioned, open and fully public chronogram database [@mctavish2015phylesystem] that stores data in a computer-readable format [@vos2012nexml]; automated and programmatic ways of accessing and downloading the data in a computer-readable format also [@Stoltzfus2013]; and methods to summarize and compare the data.
 
@@
 DateLife's core software application consists of the R package `datelife`.
-Its current stable version -- v`r packageVersion("datelife")`, is available from The Comprehensive R Archive Network (CRAN) repository [@datelifeR], and relies on functionalities from various biological R packages:
+Its current stable version -- v`r packageVersion("datelife")`, is available from The Comprehensive R Archive Network (CRAN) repository [@sanchez2022], and relies on functionalities from various biological R packages:
 ape [@paradis2004],
@@
 geiger [@pennell2014geiger],
+msa [@bodenhofer2015msa],
 paleotree [@Bapst2012a],
@@
 
-<!--_**Searching the database.--**_  -->
-
 ## Searching a chronogram database
-
-At the time of writing of this manuscript (Jun 22, 2022),
-<!-- (`r format(Sys.time(), "%b %d, %Y")`), -->
+At the time of writing of this manuscript 
+<!-- (Jun 22, 2022), -->
+(`r format(Sys.time(), "%b %d, %Y")`),
 DateLife's chronogram database latest version consist of `r format(length(opentree_chronograms$trees), big.mark=",")` chronograms published in `r length(unique(opentree_chronograms$studies))` different studies.
@@
 
-<!-- _**Summarizing search results.--**_ -->
-
 ## Summarizing search results
-
 Summary information is extracted from the \texttt{datelifeResult} object to inform decisions for subsequent steps in the analysis workflow. Basic summary information available to the user is:
@@
 
-_**Applying secondary calibrations.--**_
+## Applying secondary calibrations
 Once a topology is chosen, DateLife applies the congruification method [@Eastman2013] that find nodes belonging to the same clade across source chronograms, and then extracts the corresponding node ages from patristic distance matrices stored as a `datelifeResult` object.
@@
 
-_**Dating a tree topology.--**_
-By default, DateLife implements the Branch Length Adjuster [BLADJ; @Webb2008; @webb2005phylomatic] algorithm to obtain a fully dated topology. 
-BLADJ takes an initial topology and ignores branch length data if present.
-Next, it fixes ages on nodes with calibration data, and distributes time for nodes with no data evenly between calibrated nodes, minimizing age variance in the resulting chronogram [@Webb2008].
-This method is useful for ecological analyses [@Webb2008].
-When there is conflict between ages of calibrated nodes, ignores node ages that are older than the age of a parent node.
-BLADJ requires a root age estimate.
+## Dating a tree topology
+### With no branch lengths
+When producing or obtaining a tree with branch lengths for a group of interest is not possible, DateLife can date a topology without branch lengths, obtained from OpenTree or by implementing the supertree approach described above, by implementing the Branch Length Adjuster [BLADJ; @Webb2008; @webb2005phylomatic] algorithm, which requires no initial branch lengths.
+The algorithm starts by fixing ages for nodes with calibration data upon the given topology.
+Then, it distributes time for nodes with no data evenly between calibrated nodes, minimizing age variance in the resulting chronogram [@Webb2008], which has proven useful for ecological analyses that require age data [@Webb2008].
+When there is conflict between ages of calibrated nodes, BLADJ ignores node ages that are older than the age of a parent node.
+The BLADJ algorithm requires a root age estimate to run.
 If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature.
-If none is provided, DateLife will not return a dated topology and will provide a warning message along with suggestions on how the user can provide an age for the root so the dating analysis can be finished.
+If none is provided, DateLife will not return a dated topology and will provide a warning message along with suggestions on how the user can provide an age for the root so that a fully dated tree can be returned.
 <!-- assigns an arbitrary age to the root as 10% older than the oldest age available within the group. -->
 
-Alternatively, the user can choose to obtain a fully dated phylogeny 
-MrBayes [@Huelsenbeck2001; @Ronquist2003] approach in DateLife uses the calibrations as priors on node ages.
+In the absence of genetic data, we consider that BLADJ is the most agnostic way to assign ages to nodes with no available data, as it does not require any assumptions on the underlying model of branch length distribution.
+It is however common practice in the literature to use a birth-death model to assign ages to nodes with no genetic data [@Jetz2012; @rabosky2018inverse; @smith2018constructing].
+To do so, DateLife implements MrBayes  [@Huelsenbeck2001; @Ronquist2003], using nodes with published age data as calibration priors on a fixed topology, a simple birth-death model with parameters that can be determined by the user, and no genetic data.
 
-The latter methods can assign dates using a birth-death model (all of them????), they require initial branch lengths and thus require more time and expertise to run, which makes BLADJ the fastest and most practical method to obtain a dated tree.
+### Dating a tree with branch lengths
+Relative branch lengths provide key information for phylogenetic dating, specifically for nodes without any calibration data available. 
+While, using initial branch lengths data is the golden standard for phylogenetic dating analyses, producing such requires assembling and curating a homology hypothesis, and choosing and implementing a method for phylogenetic inference.
+DateLife implements a workflow to streamline this process by leveraging on genetic data from the Barcode of Life Data System, BOLD [@ratnasingham2007bold] to obtain markers for input taxa.
+<!--DNA sequences obtained from BOLD are short (between 500-800 bp) and display little variation. -->
+By default, mined genetic sequences are aligned with MUSCLE [@edgar2004muscle] using functions from the msa R package [@bodenhofer2015msa].
+Alternatively, sequences can be aligned with MAFFT [@katoh2009multiple], using functions from the ape R package [@paradis2004].
+The BOLD sequence alignment is then used to obtain initial branch lengths with the accelerated transformation (ACCTRAN) parsimony algorithm, which resolves ambiguous character optimization by assigning changes along branches of the tree as close to the root as possible [@agnarsson2008acctran], resulting in older nodes [@forest2005teasing].
+The parsimony branch lengths are then optimized using Maximum Likelihood, given the alignment, the topology and a simple Jukes-Cantor model, producing a BOLD tree with branch lengths proportional to expected number of substitutions per site.
+Both parsimony and ML optimizations are done with functions from the `phangorn` package [@schliep2011phangorn].
+The BOLD workflow is currently only available on DateLife's R package and not on the web application.
 
-In the case the user can provide or obtain branch length data along with a tree topology, phylogenetic dating options supported in DateLife that incorporate branch length information from the input topology in combination with the secondary calibrations:
-PATHd8 is a non-clock, rate-smoothing method [@Britton2007] to date trees;
-treePL [@Smith2012], is a semi-parametric, rate-smoothing, penalized likelihood dating method [@sanderson2002].
+BOLD trees generated by DateLife, and any tree with branch lengths provided by a user can be analysed with phylogenetic dating methods supported in DateLife that incorporate branch length information from the input topology in combination with the secondary calibrations:
+PATHd8, a non-clock, rate-smoothing method to date trees [@Britton2007];
+treePL [@Smith2012], a semi-parametric, rate-smoothing, penalized likelihood dating method [@sanderson2002]; 
+and MRBayes [@Huelsenbeck2001; @Ronquist2003], a Bayesian inference program implementing Markov chain Monte Carlo (MCMC) methods to estimate a posterior distribution of model parameters.
 
-_**Dating a tree with branch lengths--.**_
-Topologies obtained from OpenTree and with the supertree approach described above lack branch length data.
-Yet, phylogenetic dating using branch lengths data is the golden standard for phylogenetic dating analyses, but it is costly and requires a lot of human/expert curation.
-
-A fast solution implemented in DateLife is as follows.
-To estimate branch lengths proportional to substitution rates for these topologies, DateLife currently implements a simple algorithm.
-First, it mines the Barcode of Life Data System, BOLD [@ratnasingham2007bold] to obtain genetic markers for the input taxa.
-Mined genetic sequences are aligned with MUSCLE [@edgar2004muscle] (by default) or MAFFT [@katoh2009multiple].
-
-The BOLD sequence alignment is then used to obtain initial branch lengths with the accelerated transformation (ACCTRAN) parsimony algorithm, which resolves ambiguous character optimization, by assigning changes along branches of the tree as close to the root as possible [@agnarsson2008acctran].
-This results in older nodes [@forest2005teasing].
-Optionally, the likelihood of the tree topology, the alignment and the reconstructed branch lengths given different evolutionary models, is computed using functions from the `phangorn` package [@schliep2011phangorn].
-
-Relative branch length information provides key data for phylogenetic dating, especially for nodes without secondary calibrations available. Yet, topologies without branch lengths can also be dated.
-
-_**Visualizing results.--**_
-Finally, users can save all source and summary chronograms in formats that permit reuse and reanalysis (such as newick and the R "phylo" format), as well as visualize and compare results graphically, or construct their own graphs using DateLife's chronogram plot generation functions available from the R package `datelifeplot` [@datelifeplot].
+## Visualizing results
+Finally, users can save all source and summary chronograms in formats allowing for reuse and reanalysis, such as newick and the R "phylo" format.
+Input and summary chronograms can be visualized and compared graphically, and users can construct their own graphs using DateLife's chronogram plot generation functions available from the R package `datelifeplot` [@datelifeplot].
 
@@
 
-\newpage
+<!-- \newpage -->
 
@@
 
-\vspace{3mm}
+<!-- \vspace{3mm} -->
 
@@
 
-\vspace{50mm}
+<!-- \vspace{50mm} -->
 
@@
 
-The DateLife software is free and open source and it can be used through its R shiny web application at <http://www.datelife.org>, through the `datelife` R package, and through Phylotastic's project web portal <https://phylo.cs.nmsu.edu/>.
+The DateLife software is free and open source.
+It can be used online through its R shiny web application hosted at <http://www.datelife.org>, and locally through the `datelife` R package, downloadable from Zenodo (<https://doi.org/10.5281/zenodo.593938>) and the CRAN repository [@sanchez2022].
 DateLife's web application is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker and OpenTree's infrastructure (dates.opentreeoflife.org/datelife).
-`datelife`'s R package stable version `r availability` for installation from the CRAN repository (<https://cran.r-project.org/package=datelife>) using the command `install.packages(pkgs = "datelife")` from within R.
-Development versions are available from the GitHub repository (<https://github.com/phylotastic/datelife>) and can be installed using the command `devtools::install_github("phylotastic/datelife")`.
+`datelife`'s R package stable version `r availability` for installation from CRAN repository using the command `install.packages(pkgs = "datelife")` from within R.
+Development versions are available from DateLife's GitHub repository (<https://github.com/phylotastic/datelife>) and can be installed using the command `devtools::install_github("phylotastic/datelife")`.
 
@@
 
-Supplementary material, including code, biological examples and benchmark results data files and online-only appendices, can be found in the Dryad data repository (<https://doi.org/10.5061/dryad.cnp5hqc6w>), as well as in the GitHub repositories used to develop the reproducible manuscript (<https://doi.org/10.5281/zenodo.7435094>), the biological examples (<https://doi.org/10.5281/zenodo.7435101>), and the software benchmark (<https://doi.org/10.5281/zenodo.7435106>).
+Supplementary material, including code, biological examples, benchmark results, data files and online-only appendices, can be found in the Dryad data repository (<https://doi.org/10.5061/dryad.cnp5hqc6w>), as well as in the Zenodo stable repositories that host the reproducible manuscript (<https://doi.org/10.5281/zenodo.7435094>), the biological examples (<https://doi.org/10.5281/zenodo.7435101>), and the software benchmark (<https://doi.org/10.5281/zenodo.7435106>).
+The corresponding development versions hosted on GitHub can be found at <https://github.com/LunaSare/datelifeMS1#readme>, <https://github.com/LunaSare/datelife_examples#readme>, and <https://github.com/LunaSare/datelife_benchmark#readme>.
 
@@
 
-The DateLife project was born as a prototype tool aiming to provide these services, and was initially developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA [@Stoltzfus2013].
+The DateLife project was born as a prototype tool aiming to provide the services describe in this paper, and was initially developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA [@Stoltzfus2013].
 We thank colleagues from the O'Meara Lab at the University

### 2023-02-17 manuscript/manuscript.tex
remove figures from main text and finish answers to reviewer 1

@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 5480}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 5629}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
-
 Chronograms --phylogenies with branch lengths proportional to time-- provide key data on evolutionary time frame for the study of natural processes in many areas of biological research, such as comparative analysis (Freckleton, Harvey, \& Pagel, 2002; Harvey, Pagel, \& others, 1991), developmental biology (Delsuc et al., 2018; Laubichler \& Maienschein, 2009), conservation biology and ecology (Felsenstein, 1985; Webb, 2000), historical biogeography (Posadas, Crisci, \& Katinas, 2006), and species diversification (Magallon \& Sanderson, 2001; Morlon, 2014).
@@
 Building a chronogram is not an easy task.
-It requires obtaining and curating data to construct a phylogeny, selecting and placing appropriate calibrations on the phylogeny using independent age data points from the fossil record or other dated events, and inferring the full dated tree;
-it also generally requires specialized biological training, taxonomic domain knowledge, and a non-negligible amount of research time, computational resources and funding.
-
-Here we present the DateLife project which has the main goal of capturing age data from published chronograms, and making these data readily accessible to the community for reuse and reanalysis, for research, teaching, and science communication and policy.
+It requires obtaining and curating a homology hypothesis to construct a phylogeny, selecting and placing appropriate calibrations on the phylogeny using independent age data points from the fossil record or other dated events, and inferring a full dated tree.
+All of this entails specialized biological training, taxonomic domain knowledge, and a non-negligible amount of research time, computational resources and funding.
+
+Here we present the DateLife project which has the main goal of extracting and exposing age data from published chronograms, making age data readily accessible to the wider community for reuse and reanalysis, research, teaching, and science communication and policy.
 DateLife's core software application is available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org.
-It features key elements for scientific reproducibility, such as a versioned, open and fully public source database (McTavish et al., 2015) that stores data in a computer-readable format (Vos et al., 2012); automated and programmatic ways of accessing and downloading the data in a computer-readable format also (Stoltzfus et al., 2013); and methods to summarize and compare the data.
+It features key elements for scientific reproducibility, such as a curated, versioned, open and fully public chronogram database (McTavish et al., 2015) that stores data in a computer-readable format (Vos et al., 2012); automated and programmatic ways of accessing and downloading the data in a computer-readable format also (Stoltzfus et al., 2013); and methods to summarize and compare the data.
 
@@
 geiger (Pennell et al., 2014),
+msa (Bodenhofer, Bonatesta, Horejš-Kainrath, \& Hochreiter, 2015),
 paleotree (Bapst, 2012),
@@
 Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife workflow: creating a search query, searching a database, and summarizing results from the search.
-
-\begin{minipage}[t]{17.3cm}
-\includegraphics{../figures/figure-workflow/figure1-new.pdf}
-\captionof{figure}{\textit{Main DateLife workflow.
-Analyses can be performed via DateLife's interactive website at \url{www.datelife.org}, or using the \texttt{datelife} R package.
-Details on the R functions used to perform the analyses are available from \texttt{datelife}'s R package vignettes at \url{https://phylotastic.org/datelife}.}}
-\label{fig:workflow}
-\end{minipage}
 
@@
 
-At the time of writing of this manuscript (Jun 22, 2022),
+At the time of writing of this manuscript
+(Feb 20, 2023),
 DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies.
@@
 
-\emph{\textbf{Choosing a topology.--}}
+\hypertarget{choosing-a-topology}{%
+\subsection{Choosing a topology}\label{choosing-a-topology}}
+
 DateLife requires a tree topology to summarize age data upon.
@@
 
-\emph{\textbf{Applying secondary calibrations.--}}
+\hypertarget{applying-secondary-calibrations}{%
+\subsection{Applying secondary calibrations}\label{applying-secondary-calibrations}}
+
 Once a topology is chosen, DateLife applies the congruification method (Eastman, Harmon, \& Tank, 2013) that find nodes belonging to the same clade across source chronograms, and then extracts the corresponding node ages from patristic distance matrices stored as a \texttt{datelifeResult} object.
@@
 
-\emph{\textbf{Dating a tree topology.--}}
-By default, DateLife implements the Branch Length Adjuster (BLADJ; Webb et al., 2008; Webb \& Donoghue, 2005) algorithm to obtain a fully dated topology.
-BLADJ takes an initial topology and ignores branch length data if present.
-Next, it fixes ages on nodes with calibration data, and distributes time for nodes with no data evenly between calibrated nodes, minimizing age variance in the resulting chronogram (Webb et al., 2008).
-This method is useful for ecological analyses (Webb et al., 2008).
-When there is conflict between ages of calibrated nodes, ignores node ages that are older than the age of a parent node.
-BLADJ requires a root age estimate.
+\hypertarget{dating-a-tree-topology}{%
+\subsection{Dating a tree topology}\label{dating-a-tree-topology}}
+
+\hypertarget{with-no-branch-lengths}{%
+\subsubsection{With no branch lengths}\label{with-no-branch-lengths}}
+
+When producing or obtaining a tree with branch lengths for a group of interest is not possible, DateLife can date a topology without branch lengths, obtained from OpenTree or by implementing the supertree approach described above, by implementing the Branch Length Adjuster (BLADJ; Webb et al., 2008; Webb \& Donoghue, 2005) algorithm, which requires no initial branch lengths.
+The algorithm starts by fixing ages for nodes with calibration data upon the given topology.
+Then, it distributes time for nodes with no data evenly between calibrated nodes, minimizing age variance in the resulting chronogram (Webb et al., 2008), which has proven useful for ecological analyses that require age data (Webb et al., 2008).
+When there is conflict between ages of calibrated nodes, BLADJ ignores node ages that are older than the age of a parent node.
+The BLADJ algorithm requires a root age estimate to run.
 If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature.
-If none is provided, DateLife will not return a dated topology and will provide a warning message along with suggestions on how the user can provide an age for the root so the dating analysis can be finished.
-
-Alternatively, the user can choose to obtain a fully dated phylogeny
-MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003) approach in DateLife uses the calibrations as priors on node ages.
-
-The latter methods can assign dates using a birth-death model (all of them????), they require initial branch lengths and thus require more time and expertise to run, which makes BLADJ the fastest and most practical method to obtain a dated tree.
-
-In the case the user can provide or obtain branch length data along with a tree topology, phylogenetic dating options supported in DateLife that incorporate branch length information from the input topology in combination with the secondary calibrations:
-PATHd8 is a non-clock, rate-smoothing method (Britton et al., 2007) to date trees;
-treePL (Smith \& O'Meara, 2012), is a semi-parametric, rate-smoothing, penalized likelihood dating method (Sanderson, 2002).
-
-\emph{\textbf{Dating a tree with branch lengths--.}}
-Topologies obtained from OpenTree and with the supertree approach described above lack branch length data.
-Yet, phylogenetic dating using branch lengths data is the golden standard for phylogenetic dating analyses, but it is costly and requires a lot of human/expert curation.
-
-A fast solution implemented in DateLife is as follows.
-To estimate branch lengths proportional to substitution rates for these topologies, DateLife currently implements a simple algorithm.
-First, it mines the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain genetic markers for the input taxa.
-Mined genetic sequences are aligned with MUSCLE (Edgar, 2004) (by default) or MAFFT (Katoh, Asimenos, \& Toh, 2009).
-
-The BOLD sequence alignment is then used to obtain initial branch lengths with the accelerated transformation (ACCTRAN) parsimony algorithm, which resolves ambiguous character optimization, by assigning changes along branches of the tree as close to the root as possible (Agnarsson \& Miller, 2008).
-This results in older nodes (Forest et al., 2005).
-Optionally, the likelihood of the tree topology, the alignment and the reconstructed branch lengths given different evolutionary models, is computed using functions from the \texttt{phangorn} package (Schliep, 2011).
-
-Relative branch length information provides key data for phylogenetic dating, especially for nodes without secondary calibrations available. Yet, topologies without branch lengths can also be dated.
-
-\emph{\textbf{Visualizing results.--}}
-Finally, users can save all source and summary chronograms in formats that permit reuse and reanalysis (such as newick and the R \enquote{phylo} format), as well as visualize and compare results graphically, or construct their own graphs using DateLife's chronogram plot generation functions available from the R package \texttt{datelifeplot} (Sanchez-Reyes \& O'Meara, 2022).
-
-\newpage
+If none is provided, DateLife will not return a dated topology and will provide a warning message along with suggestions on how the user can provide an age for the root so that a fully dated tree can be returned.
+
+In the absence of genetic data, we consider that BLADJ is the most agnostic way to assign ages to nodes with no available data, as it does not require any assumptions on the underlying model of branch length distribution.
+It is however common practice in the literature to use a birth-death model to assign ages to nodes with no genetic data (Jetz, Thomas, Joy, Hartmann, \& Mooers, 2012; Rabosky et al., 2018; Smith \& Brown, 2018).
+To do so, DateLife implements MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), using nodes with published age data as calibration priors on a fixed topology, a simple birth-death model with parameters that can be determined by the user, and no genetic data.
+
+\hypertarget{dating-a-tree-with-branch-lengths}{%
+\subsubsection{Dating a tree with branch lengths}\label{dating-a-tree-with-branch-lengths}}
+
+Relative branch lengths provide key information for phylogenetic dating, specifically for nodes without any calibration data available.
+While, using initial branch lengths data is the golden standard for phylogenetic dating analyses, producing such requires assembling and curating a homology hypothesis, and choosing and implementing a method for phylogenetic inference.
+DateLife implements a workflow to streamline this process by leveraging on genetic data from the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain markers for input taxa.
+By default, mined genetic sequences are aligned with MUSCLE (Edgar, 2004) using functions from the msa R package (Bodenhofer et al., 2015).
+Alternatively, sequences can be aligned with MAFFT (Katoh, Asimenos, \& Toh, 2009), using functions from the ape R package (Paradis et al., 2004).
+The BOLD sequence alignment is then used to obtain initial branch lengths with the accelerated transformation (ACCTRAN) parsimony algorithm, which resolves ambiguous character optimization by assigning changes along branches of the tree as close to the root as possible (Agnarsson \& Miller, 2008), resulting in older nodes (Forest et al., 2005).
+The parsimony branch lengths are then optimized using Maximum Likelihood, given the alignment, the topology and a simple Jukes-Cantor model, producing a BOLD tree with branch lengths proportional to expected number of substitutions per site.
+Both parsimony and ML optimizations are done with functions from the \texttt{phangorn} package (Schliep, 2011).
+The BOLD workflow is currently only available on DateLife's R package and not on the web application.
+
+BOLD trees generated by DateLife, and any tree with branch lengths provided by a user can be analysed with phylogenetic dating methods supported in DateLife that incorporate branch length information from the input topology in combination with the secondary calibrations:
+PATHd8, a non-clock, rate-smoothing method to date trees (Britton et al., 2007);
+treePL (Smith \& O'Meara, 2012), a semi-parametric, rate-smoothing, penalized likelihood dating method (Sanderson, 2002);
+and MRBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), a Bayesian inference program implementing Markov chain Monte Carlo (MCMC) methods to estimate a posterior distribution of model parameters.
+
+\hypertarget{visualizing-results}{%
+\subsection{Visualizing results}\label{visualizing-results}}
+
+Finally, users can save all source and summary chronograms in formats allowing for reuse and reanalysis, such as newick and the R \enquote{phylo} format.
+Input and summary chronograms can be visualized and compared graphically, and users can construct their own graphs using DateLife's chronogram plot generation functions available from the R package \texttt{datelifeplot} (Sanchez-Reyes \& O'Meara, 2022).
 
@@
 \ref{fig:runtime_main}).
-
-\includegraphics{../figures/fig_runtime_main.pdf}
-\captionof{figure}{\textit{DateLife's benchmarking results.
-Computation time used to process a query and a search across \texttt{datelife}'s chronogram database, relative to number of input taxon names.
-For each N = \{10, 100, 200, ..., 1 000, ...
-, 9 000, 10 000\}, we sampled N species names from the class Aves a hundred times, and then performed a datelife search processing the input names with Taxon Names Resolution Service (TNRS; light gray), and without processing names (dark gray).
-For comparison, we performed a search using an input that had been pre-processed with TNRS (light blue).}}
-\label{fig:runtime_main}
 
@@
 
-\newpage
-
 \hypertarget{searching-the-database}{%
@@
 \ref{fig:figure2}B).
-Three studies matched five input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2015; Hedges, Marin, Suleski, Paymer, \& Kumar, 2015; Jetz, Thomas, Joy, Hartmann, \& Mooers, 2012), one study matched four input names (Hooper \& Price, 2017) and two studies matched two input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2013; Burns et al., 2014).
+Three studies matched five input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2015; Hedges, Marin, Suleski, Paymer, \& Kumar, 2015; Jetz et al., 2012), one study matched four input names (Hooper \& Price, 2017) and two studies matched two input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2013; Burns et al., 2014).
 No studies matched all input names.
 Together, source chronograms provide 28 unique age data points, covering all nodes on our chosen tree topology to date (Table \ref{tbl:table1}).
-
-\vspace{3mm}
 
@@
 \newpage
+
+\hypertarget{an-example-with-the-family-of-true-finches}{%
+\subsection{An example with the family of true finches}\label{an-example-with-the-family-of-true-finches}}
+
+\hypertarget{creating-a-query}{%
+\subsubsection{Creating a query}\label{creating-a-query}}
+
+To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the \enquote{get species from taxon} flag,
+which gets all recognized species names within a named group from a taxonomy of choice.
+Following the NCBI taxonomy, our DateLife query has 289 Fringillidae species names.
+This taxon-constrained approach implies that the full DateLife analysis will be performed using a tree topology and ages available for species names from a given taxonomic group, which do not necessarily correspond to a monophyletic group.
+Users can change this behaviour by providing all species names corresponding to a monophyletic group as input for a DateLife search, or a monophyletic tree to construct a DateLife summary.
+
+\hypertarget{searching-the-database-1}{%
+\subsubsection{Searching the database}\label{searching-the-database-1}}
+
+Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species as tip taxa.
+The DateLife search identified 19 chronograms matching this criteria, published in 13 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Kimball et al., 2019; Oliveros et al., 2019; Price et al., 2014; Roquet, Lavergne, \& Thuiller, 2014; Uyeda, Pennell, Miller, Maia, \& McClain, 2017).
+Once identified, DateLife pruned these matching chronograms to remove tips that do not belong to the queried taxon names, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementray Table S1).
+
+\hypertarget{summarizing-search-results-2}{%
+\subsubsection{Summarizing search results}\label{summarizing-search-results-2}}
+
+The final step entailed congruifying and summarizing the age data available for the Fringillidae species into two single summary chronograms, using two different types of summary ages, median and SDM.
+As explained in the \enquote{Description} section, a tree topology to summarize age data upon is required.
+By default, DateLife uses the topology from OpenTree's synthetic tree that contains all taxa from the search query.
+According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig.
+\ref{fig:fringillidae-topologies}).
+Hence, a topology containing only the 289 species from the original query was extracted from Open Tree of Life's synthetic tree v12.3 (Open Tree Of Life et al., 2019).
+
+Age data from source chronograms was congruified to OpenTree's topology (Fig.
+\ref{fig:fringillidae-topologies}B), reducing the age data set to 818 different data points (Supplementray Table S2).
+For each congruent node, age summary statistics were calculated and used as fixed secondary calibrations over the chosen tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
+\ref{fig:fringillidages}).
+
+\hypertarget{cross-validation-test}{%
+\section{Cross-validation test}\label{cross-validation-test}}
+
+We performed a cross validation analysis of the DateLife workflow using the Fringillidae chronograms.
+We used the individual tree topologies from each of the 19 source chronograms from 13 studies as inputs, treating their node ages as unknown.
+We then estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.
+We found that node ages from original study, and ages estimated using all other age data available are correlated (Fig. \ref{fig:cvXY}).
+For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka \enquote{closer to the root}) relative to the original estimate, and overestimate ages for nodes closer to the tips.
+Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. S1).
+
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+DateLife makes state-of-the-art data on evolutionary time frame easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
+It is an open service that does not require any expert biological knowledge from users --besides the names of the species or group they want to work with, for any of its functionality.
+
+A total of 99,474 unique terminal taxa are represented in DateLife's database.
+Incorporation of more chronograms into the database will continue to improve DateLife's services.
+One option to increase the number of chronograms in the DateLife database is the Dryad data repository.
+Methods to automatically mine chronograms from Dryad could be designed and implemented.
+However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined manually by a human curator.
+We would like to emphasize on the importance of sharing chronogram data, including systematically curated metadata, into open repositories, such as OpenTree's Phylesystem (McTavish et al., 2015) for the benefit of the scientific community as a whole.
+
+While it is known that incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance in ecological and conservation biology studies (Webb et al., 2008).
+Next, we discuss some particularities of using summarized node age data from a DateLife analysis that users should be aware of.
+
+\hypertarget{age-variation-in-source-chronograms}{%
+\subsection{Age variation in source chronograms}\label{age-variation-in-source-chronograms}}
+
+Conflict in estimated ages among alternative studies is common in the literature.
+See, for example, the robust ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández, 2015; Ramshaw et al., 1972; Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021).
+Source chronograms available for the same organisms have potentially been estimated implementing calibrations very differently.
+For example, the chronograms from Burns et al.
+(2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older age estimates for the same nodes than chronograms that were inferred using fossils as calibrations (Figs.
+\ref{fig:fringillidages}, \ref{fig:cvXY}; Supplementary Figs.
+S1, S5).
+
+Different calibration implementations might also imply fundamentally distinct evolutionary hypotheses (Antonelli et al., 2017).
+For example, two independent researchers working on the same clade should both carefully select and justify their choices of fossil calibration placement.
+Yet, if one researcher concludes that a fossil should calibrate the ingroup of a clade, while another researcher concludes that the same fossil should calibrate the outgroup of the clade, the resulting age estimates will differ, as the placement of calibrations as stem or crown group has been proven to significantly affect time of lineage divergence estimates (Sauquet, 2013).
+
+\hypertarget{primary-vs-secondary-calibrations}{%
+\subsection{Primary vs Secondary calibrations}\label{primary-vs-secondary-calibrations}}
+
+While most chronograms in DateLife's database are constructed using primary calibrations (molecular substitution rates or ages obtained from the fossil record or geological events), DateLife summarizes chronograms using secondary calibrations (ages coming from other chronograms).
+Graur and Martin (2004) cautioned on the increased error and uncertainty in estimated ages when using secondary calibrations in dating analyses.
+Schenk (2016) showed that, in simulations, divergence times inferred using secondary calibrations are significantly younger than those inferred with primary calibrations, when obtained with Bayesian inference methods, and when priors are implemented in similar ways in both analyses.
+Accordingly, the scientific community seems to have more confidence in chronograms obtained from a single analysis, using fossil data as primary sources of calibrations (Schenk, 2016), and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data reflect a coherent evolutionary history (Sauquet, 2013), as for example done by Antonelli et al.
+(2017).
+There have been attempts to create fossil calibration databases (Ksepka et al., 2015), though these still have room to grow.
+
+It seems that using several (as opposed to just a few) secondary calibrations can provide sufficient information to alleviate or even neutralize potential biases (Sauquet, 2013).
+Certainly, further studies are required to fully understand the effect of secondary calibrations on outputs from different tree dating methods, and on downstream analyses.
+It is possible that secondary calibrations can be safely used with dating methods that do not require setting priors, such as penalized likelihood (Sanderson, 2003), with methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ (Webb et al., 2008; Webb \& Donoghue, 2005), or methods that summarize age data unto a tree topology.
+
+Our cross validation analysis might provide some insight in this regard.
+When ages are estimated with secondary calibrations, nodes closer to the root do tend to be slightly younger than ages estimated with primary calibrations.
+However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures S2-S20).
+The only exception to this was observed on cross validation results of the Burns et al.
+(2014) chronogram, which displays much younger node ages when estimated using secondary calibrations (Supplementary Figs.
+S1, S5).
+
+\hypertarget{sumarizing-chronograms}{%
+\subsection{Sumarizing chronograms}\label{sumarizing-chronograms}}
+
+By default, DateLife currently summarizes all source chronograms that overlap with at least two species names. Users can exclude source chronograms if they have reasons to do so.
+Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely.
+To our knowledge, there are no tested measures to determine independently when a chronogram is better than another.
+Yet, several characteristics of the data used for dating analyses, as well as from the output chronogram itself, could be used to score the quality of source chronograms.
+
+Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used (Magallón, 2010; Magallón et al., 2015).
+Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
+
+DateLife provides an opportunity to capture concordance and conflict among date estimates, which can also be used as a metric for chronogram reliability.
+Its open database of chronograms allows other researchers to do such analyses themselves reproducibly, and without needing permission.
+Though, of course, they should follow proper citation practices, especially for the source chronogram studies.
+
+The exercise of summarizing age data from across multiple studies provides the opportunity to work with a chronogram that reflects a unified evolutionary history for a lineage, by putting together evidence from different hypotheses.
+The largest, and taxonomically broadest chronogram currently available from OpenTree was constructed summarizing age data from 2,274 published chronograms using NCBI's taxonomic tree as backbone (Hedges et al., 2015).
+A summarizing exercise may also amplify the effect of uncertainty and errors in source data, and blur parts of the evolutionary history of a lineage that might only be reflected in source chronograms and lost on the summary chronogram (Sauquet et al., 2021).
+
+\hypertarget{effects-of-phylogenetic-sampling-on-downstream-analyses}{%
+\subsection{Effects of phylogenetic sampling on downstream analyses}\label{effects-of-phylogenetic-sampling-on-downstream-analyses}}
+
+Analysis of species diversification using simulated and empirical phylogenies suggest that using a more completely sampled phylogeny provides estimates that are closer to the true diversification history than when analysing incompletely sampled phylogenies (Chang, Rabosky, \& Alfaro, 2020; Cusimano, Stadler, \& Renner, 2012; Sun et al., 2020).
+Ideally phylogenies should be completed using genetic data, but this is a very hard task to achieve for many biological groups.
+Hence, DateLife's workflow features different ways of assigning node ages in the absence of calibrations and branch length information for certain taxa.
+Completing a phylogeny using a stochastic birth-death politomy resolver and a backbone taxonomy is a common practice in scientific publications: Jetz et al.
+(2012), created a chronogram of all 9, 993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
+(2018) created a chronogram of 31, 536 ray-finned fishes, of which only 37\% had molecular data; Smith and Brown (2018) constructed a chronogram of 353, 185 seed plants where only 23\% had molecular data.
+These stochastically resolved chronograms provide less biased diversification rates estimated with methods that account for sampling fractions (Chang et al., 2020; Cusimano et al., 2012), but can also introduce spurious patterns of early bursts of diversification (Cusimano \& Renner, 2010; Sun et al., 2020).
+
+Taxonomy based politomy resolvers also introduce topological differences.
+The study of macroevolutionary processes largely depends on an understanding of the timing of species diversification events, and different phylogenetic and chronogram hypothesis can provide very different overviews of the macroeovlutionary history of a biological group.
+For example, alternative topologies in chronograms from the same biological group can infer very different species diversification patterns (Rabosky, 2015; Title \& Rabosky, 2016).
+Similarly there are worries that patterns of morphological evolution cannot be accurately inferred with these types of phylogenies, as any patterns would be erased by randomization (Rabosky, 2015).
+We note that the same applies for geographical and morphological dependent diversification analysis.
+Hence, we suggest that these types of randomly resolved phylogenies can be used as null or neutral models, representing the case of diversification independent of traits and geographical scenario.
+
+While stochastic politomy resolvers have been useful to advance research, notably, risks come with this practice.
+Taken to the extreme, one could generate a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy, a single calibration, and assigning branch lengths following a birth-death diversification model.
+Clearly, this can lead to a misrepresentation of the true evolutionary history.
+We urge DateLife users to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, potential biases, and assessing the consistency of the results with prior work.
+
+\hypertarget{conclusions}{%
+\section{Conclusions}\label{conclusions}}
+
+Knowledge of the evolutionary time frame of organisms is key to many research areas: trait evolution, species diversification, biogeography, macroecology and more.
+It is also crucial for education, science communication and policy, but generating chronograms is difficult, especially for those who want to use phylogenies but who are not systematists, or do not have the time to acquire and develop the necessary knowledge and skills to construct them on their own.
+Importantly, years of primarily publicly funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
+
+The DateLife project allows for easy and fast summary of public and state-of-the-art data on time of lineage divergence.
+It provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identifying regions that require more research, or that have conflicting information.
+It is available as an R package, and as a web-based R shiny application at www.datelife.org
+Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
+The DateLife project should improve awareness of the existing variation in expert time of divergence data, and foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
+
+\hypertarget{availability}{%
+\section{Availability}\label{availability}}
+
+The DateLife software is free and open source.
+It can be used online through its R shiny web application hosted at \url{http://www.datelife.org}, and locally through the \texttt{datelife} R package, downloadable from Zenodo (\url{https://doi.org/10.5281/zenodo.593938}) and the CRAN repository (Sanchez-Reyes et al., 2022).
+DateLife's web application is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker and OpenTree's infrastructure (dates.opentreeoflife.org/datelife).
+\texttt{datelife}'s R package stable version is available for installation from CRAN repository using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R.
+Development versions are available from DateLife's GitHub repository (\url{https://github.com/phylotastic/datelife}) and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.
+
+\hypertarget{supplementary-material}{%
+\section{Supplementary Material}\label{supplementary-material}}
+
+Supplementary material, including code, biological examples, benchmark results, data files and online-only appendices, can be found in the Dryad data repository (\url{https://doi.org/10.5061/dryad.cnp5hqc6w}), as well as in the Zenodo stable repositories that host the reproducible manuscript (\url{https://doi.org/10.5281/zenodo.7435094}), the biological examples (\url{https://doi.org/10.5281/zenodo.7435101}), and the software benchmark (\url{https://doi.org/10.5281/zenodo.7435106}).
+The corresponding development versions hosted on GitHub can be found at \url{https://github.com/LunaSare/datelifeMS1\#readme}, \url{https://github.com/LunaSare/datelife_examples\#readme}, and \url{https://github.com/LunaSare/datelife_benchmark\#readme}.
+
+\hypertarget{funding}{%
+\section{Funding}\label{funding}}
+
+Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to the Datelife project; DBI-0905606 to the National Evolutionary Synthesis Center (NESCent); ABI-1458572 to the Phylotastic project; and ABI-1759846 to the Open Tree of Life project.
+
+\hypertarget{acknowledgements}{%
+\section{Acknowledgements}\label{acknowledgements}}
+
+The DateLife project was born as a prototype tool aiming to provide the services describe in this paper, and was initially developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA (Stoltzfus et al., 2013).
+We thank colleagues from the O'Meara Lab at the University
+of Tennesse Knoxville for suggestions, discussions and software testing.
+The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
+that led to initial work on this project.
+The team that assembled DateLife's first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
+The Open Tree of Life project that provides the open, metadata rich repository of
+trees used to construct DateLife's chronogram database.
+The many scientists who publish their chronograms in an open, reusable form, and the scientists who curate them for deposition in the Open Tree of Life repository.
+The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.
+
+\newpage
+
+\hypertarget{references}{%
+\section{References}\label{references}}
+
+\begingroup
+\setlength{\parindent}{-0.5in}
+\setlength{\leftskip}{0.5in}
+
+\hypertarget{refs}{}
+\leavevmode\hypertarget{ref-agnarsson2008acctran}{}%
+Agnarsson, I., \& Miller, J. A. (2008). Is acctran better than deltran? \emph{Cladistics}, \emph{24}(6), 1032--1038.
+
+\leavevmode\hypertarget{ref-alstrom2014discovery}{}%
+Alström, P., Hooper, D.
+M., Liu, Y., Olsson, U., Mohan, D., Gelang, M., \ldots{} Price, T.
+D.
+(2014).
+Discovery of a relict lineage and monotypic family of passerine birds.
+\emph{Biology Letters}, \emph{10}(3), 20131067.
+
+\leavevmode\hypertarget{ref-ane2009groves}{}%
+Ané, C., Eulenstein, O., Piaggio-Talice, R., \& Sanderson, M. J. (2009). Groves of phylogenetic trees. \emph{Annals of Combinatorics}, \emph{13}(2), 139--167.
+
+\leavevmode\hypertarget{ref-antonelli2017supersmart}{}%
+Antonelli, A., Hettling, H., Condamine, F.
+L., Vos, K., Nilsson, R.
+H., Sanderson, M.
+J., \ldots{} Vos, R.
+A.
+(2017).
+Toward a self-updating platform for estimating rates of speciation and migration, ages, and relationships of Taxa.
+\emph{Systematic Biology}, \emph{66}(2), 153--166.
+\url{https://doi.org/10.1093/sysbio/syw066}
+
+\leavevmode\hypertarget{ref-archie1986newick}{}%
+Archie, J., Day, W.
+H., Felsenstein, J., Maddison, W., Meacham, C., Rohlf, F.
+J., \& Swofford, D.
+(1986).
+The Newick tree format.
+Retrieved from \href{\%7Bhttps://evolution.genetics.washington.edu/phylip/newicktree.html\%7D}{\{https://evolution.genetics.washington.edu/phylip/newicktree.html\}}
+
+\leavevmode\hypertarget{ref-avibase-emberiza}{}%
+Avibase.
+(2022).
+Yellow-throated Bunting.
+\emph{Avibase - the World Bird Database}, (Online Resource).
+Retrieved from \href{\%7Bhttps://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\%7D}{\{https://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\}}
+
+\leavevmode\hypertarget{ref-Bapst2012a}{}%
+Bapst, D.
+W.
+(2012).
+Paleotree: An R package for paleontological and phylogenetic analyses of evolution.
+\emph{Methods in Ecology and Evolution}, \emph{3}(5), 803--807.
+\url{https://doi.org/10.1111/j.2041-210X.2012.00223.x}
+
+\leavevmode\hypertarget{ref-barba2018constraining}{}%
+Barba-Montoya, J., Reis, M.
+dos, Schneider, H., Donoghue, P.
+C., \& Yang, Z.
+(2018).
+Constraining uncertainty in the timescale of angiosperm evolution and the veracity of a cretaceous terrestrial revolution.
+\emph{New Phytologist}, \emph{218}(2), 819--834.
+
+\leavevmode\hypertarget{ref-barker2014mitogenomic}{}%
+Barker, F. K. (2014). Mitogenomic data resolve basal relationships among passeriform and passeridan birds. \emph{Molecular Phylogenetics and Evolution}, \emph{79}, 313--324.
+
+\leavevmode\hypertarget{ref-barker2013going}{}%
+Barker, F.
+K., Burns, K.
+J., Klicka, J., Lanyon, S.
+M., \& Lovette, I.
+J.
+(2013).
+Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds.
+\emph{Systematic Biology}, \emph{62}(2), 298--320.
+
+\leavevmode\hypertarget{ref-barker2015new}{}%
+Barker, F.
+K., Burns, K.
+J., Klicka, J., Lanyon, S.
+M., \& Lovette, I.
+J.
+(2015).
+New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
+\emph{The Auk: Ornithological Advances}, \emph{132}(2), 333--348.
+
+\leavevmode\hypertarget{ref-barker2004phylogeny}{}%
+Barker, F.
+K., Cibois, A., Schikler, P., Feinstein, J., \& Cracraft, J.
+(2004).
+Phylogeny and diversification of the largest avian radiation.
+\emph{Proceedings of the National Academy of Sciences}, \emph{101}(30), 11040--11045.
+
+\leavevmode\hypertarget{ref-beresford2005african}{}%
+Beresford, P., Barker, F., Ryan, P., \& Crowe, T.
+(2005).
+African endemics span the tree of songbirds (passeri): Molecular systematics of several evolutionary ``enigmas''.
+\emph{Proceedings of the Royal Society B: Biological Sciences}, \emph{272}(1565), 849--858.
+
+\leavevmode\hypertarget{ref-bodenhofer2015msa}{}%
+Bodenhofer, U., Bonatesta, E., Horejš-Kainrath, C., \& Hochreiter, S. (2015). Msa: An r package for multiple sequence alignment. \emph{Bioinformatics}, \emph{31}(24), 3997--3999.
+
+\leavevmode\hypertarget{ref-Boyle2013}{}%
+Boyle, B., Hopkins, N., Lu, Z., Raygoza Garay, J.
+A., Mozzherin, D., Rees, T., \ldots{} Enquist, B.
+J.
+(2013).
+The taxonomic name resolution service: An online tool for automated standardization of plant names.
+\emph{BMC Bioinformatics}, \emph{14}(1).
+\url{https://doi.org/10.1186/1471-2105-14-16}
+
+\leavevmode\hypertarget{ref-Britton2007}{}%
+Britton, T., Anderson, C.
+L., Jacquet, D., Lundqvist, S., \& Bremer, K.
+(2007).
+Estimating Divergence Times in Large Phylogenetic Trees.
+\emph{Systematic Biology}, \emph{56}(788777878), 741--752.
+\url{https://doi.org/10.1080/10635150701613783}
+
+\leavevmode\hypertarget{ref-bryson2014diversification}{}%
+Bryson Jr, R.
+W., Chaves, J., Smith, B.
+T., Miller, M.
+J., Winker, K., Pérez-Emán, J.
+L., \& Klicka, J.
+(2014).
+Diversification across the new world within the `blue'cardinalids (aves: Cardinalidae).
+\emph{Journal of Biogeography}, \emph{41}(3), 587--599.
+
+\leavevmode\hypertarget{ref-burleigh2015building}{}%
+Burleigh, J. G., Kimball, R. T., \& Braun, E. L. (2015). Building the avian tree of life using a large-scale, sparse supermatrix. \emph{Molecular Phylogenetics and Evolution}, \emph{84}, 53--63.
+
+\leavevmode\hypertarget{ref-burns2014phylogenetics}{}%
+Burns, K.
+J., Shultz, A.
+J., Title, P.
+O., Mason, N.
+A., Barker, F.
+K., Klicka, J., \ldots{} Lovette, I.
+J.
+(2014).
+Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds.
+\emph{Molecular Phylogenetics and Evolution}, \emph{75}, 41--77.
+
+\leavevmode\hypertarget{ref-Chamberlain2018}{}%
+Chamberlain, S. (2018). \emph{bold: Interface to Bold Systems API}. Retrieved from \url{https://CRAN.R-project.org/package=bold}
+
+\leavevmode\hypertarget{ref-Chamberlain2013}{}%
+Chamberlain, S.
+A., \& Szöcs, E.
+(2013).
+taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}.
+\emph{F1000Research}, \emph{2}(191), 1--29.
+\url{https://doi.org/10.12688/f1000research.2-191.v2}
+
+\leavevmode\hypertarget{ref-chang2020estimating}{}%
+Chang, J., Rabosky, D.
+L., \& Alfaro, M.
+E.
+(2020).
+Estimating diversification rates on incompletely sampled phylogenies: Theoretical concerns and practical solutions.
+\emph{Systematic Biology}, \emph{69}(3), 602--611.
+
+\leavevmode\hypertarget{ref-chaves2013biogeography}{}%
+Chaves, J.
+A., Hidalgo, J.
+R., \& Klicka, J.
+(2013).
+Biogeography and evolutionary history of the n eotropical genus s altator (a ves: T hraupini).
+\emph{Journal of Biogeography}, \emph{40}(11), 2180--2190.
+
+\leavevmode\hypertarget{ref-claramunt2015new}{}%
+Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{Science Advances}, \emph{1}(11), e1501005.
+
+\leavevmode\hypertarget{ref-Criscuolo2006}{}%
+Criscuolo, A., Berry, V., Douzery, E.
+J., \& Gascuel, O.
+(2006).
+SDM: A fast distance-based approach for (super)tree building in phylogenomics.
+\emph{Systematic Biology}, \emph{55}(5), 740--755.
+\url{https://doi.org/10.1080/10635150600969872}
+
+\leavevmode\hypertarget{ref-cusimano2010slowdowns}{}%
+Cusimano, N., \& Renner, S. S. (2010). Slowdowns in diversification rates from real phylogenies may not be real. \emph{Systematic Biology}, \emph{59}(4), 458--464.
+
+\leavevmode\hypertarget{ref-cusimano2012new}{}%
+Cusimano, N., Stadler, T., \& Renner, S.
+S.
+(2012).
+A new method for handling missing species in diversification analysis applicable to randomly or nonrandomly sampled phylogenies.
+\emph{Systematic Biology}, \emph{61}(5), 785--792.
+
+\leavevmode\hypertarget{ref-delsuc2018phylogenomic}{}%
+Delsuc, F., Philippe, H., Tsagkogeorga, G., Simion, P., Tilak, M.-K., Turon, X., \ldots{} Douzery, E.
+J.
+(2018).
+A phylogenomic framework and timescale for comparative studies of tunicates.
+\emph{BMC Biology}, \emph{16}(1), 1--14.
+
+\leavevmode\hypertarget{ref-Eastman2013}{}%
+Eastman, J.
+M., Harmon, L.
+J., \& Tank, D.
+C.
+(2013).
+Congruification: Support for time scaling large phylogenetic trees.
+\emph{Methods in Ecology and Evolution}, \emph{4}(7), 688--691.
+\url{https://doi.org/10.1111/2041-210X.12051}
+
+\leavevmode\hypertarget{ref-edgar2004muscle}{}%
+Edgar, R. C. (2004). MUSCLE: Multiple sequence alignment with high accuracy and high throughput. \emph{Nucleic Acids Research}, \emph{32}(5), 1792--1797.
+
+\leavevmode\hypertarget{ref-Felsenstein1985a}{}%
+Felsenstein, J. (1985). Phylogenies and the Comparative Method. \emph{The American Naturalist}, \emph{125}(1), 1--15. Retrieved from \url{http://www.jstor.org/stable/2461605}
+
+\leavevmode\hypertarget{ref-forest2005teasing}{}%
+Forest, F., Savolainen, V., Chase, M.
+W., Lupia, R., Bruneau, A., \& Crane, P.
+R.
+(2005).
+Teasing apart molecular-versus fossil-based error estimates when dating phylogenetic trees: A case study in the birch family (betulaceae).
+\emph{Systematic Botany}, \emph{30}(1), 118--133.
+
+\leavevmode\hypertarget{ref-freckleton2002phylogenetic}{}%
+Freckleton, R. P., Harvey, P. H., \& Pagel, M. (2002). Phylogenetic analysis and comparative data: A test and review of evidence. \emph{The American Naturalist}.
+
+\leavevmode\hypertarget{ref-gbif2022taxonomy}{}%
+GBIF Secretariat.
+(2022).
+GBIF Backbone Taxonomy.
+\emph{Checklist dataset}, (Online Resource accessed via GBIF.org).
+Retrieved from \href{\%7Bhttps://doi.org/10.15468/39omei\%20\%7D}{\{https://doi.org/10.15468/39omei \}}
+
+\leavevmode\hypertarget{ref-gibb2015new}{}%
+Gibb, G.
+C., England, R., Hartig, G., McLenachan, P.
+A., Taylor Smith, B.
+L., McComish, B.
+J., \ldots{} Penny, D.
+(2015).
+New zealand passerines help clarify the diversification of major songbird lineages during the oligocene.
+\emph{Genome Biology and Evolution}, \emph{7}(11), 2983--2995.
+
+\leavevmode\hypertarget{ref-graur2004reading}{}%
+Graur, D., \& Martin, W. (2004). Reading the entrails of chickens: Molecular timescales of evolution and the illusion of precision. \emph{TRENDS in Genetics}, \emph{20}(2), 80--86.
+
+\leavevmode\hypertarget{ref-hackett2008phylogenomic}{}%
+Hackett, S.
+J., Kimball, R.
+T., Reddy, S., Bowie, R.
+C., Braun, E.
+L., Braun, M.
+J., \ldots{} others.
+(2008).
+A phylogenomic study of birds reveals their evolutionary history.
+\emph{Science}, \emph{320}(5884), 1763--1768.
+
+\leavevmode\hypertarget{ref-harvey1991comparative}{}%
+Harvey, P. H., Pagel, M. D., \& others. (1991). \emph{The comparative method in evolutionary biology} (Vol. 239). Oxford university press Oxford.
+
+\leavevmode\hypertarget{ref-Hedges2015}{}%
+Hedges, S.
+B., Marin, J., Suleski, M., Paymer, M., \& Kumar, S.
+(2015).
+Tree of life reveals clock-like speciation and diversification.
+\emph{Molecular Biology and Evolution}, \emph{32}(4), 835--845.
+\url{https://doi.org/10.1093/molbev/msv037}
+
+\leavevmode\hypertarget{ref-Heibl2008}{}%
+Heibl, C. (2008). \emph{PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages.} Retrieved from \url{http://www.christophheibl.de/Rpackages.html}
+
+\leavevmode\hypertarget{ref-hooper2017chromosomal}{}%
+Hooper, D. M., \& Price, T. D. (2017). Chromosomal inversion differences correlate with range overlap in passerine birds. \emph{Nature Ecology \& Evolution}, \emph{1}(10), 1526.
+
+\leavevmode\hypertarget{ref-Huelsenbeck2001}{}%
+Huelsenbeck, J. P., \& Ronquist, F. (2001). MRBAYES: Bayesian inference of phylogenetic trees. \emph{Bioinformatics}, \emph{17}(8), 754--755. \url{https://doi.org/10.1093/bioinformatics/17.8.754}
+
+\leavevmode\hypertarget{ref-Jetz2012}{}%
+Jetz, W., Thomas, G., Joy, J.
+J., Hartmann, K., \& Mooers, A.
+(2012).
+The global diversity of birds in space and time.
+\emph{Nature}, \emph{491}(7424), 444--448.
+\url{https://doi.org/10.1038/nature11631}
+
+\leavevmode\hypertarget{ref-johansson2008phylogenetic}{}%
+Johansson, U.
+S., Fjeldså, J., \& Bowie, R.
+C.
+(2008).
+Phylogenetic relationships within passerida (aves: Passeriformes): A review and a new molecular phylogeny based on three nuclear intron markers.
+\emph{Molecular Phylogenetics and Evolution}, \emph{48}(3), 858--876.
+
+\leavevmode\hypertarget{ref-katoh2009multiple}{}%
+Katoh, K., Asimenos, G., \& Toh, H. (2009). Multiple alignment of dna sequences with mafft. In \emph{Bioinformatics for dna sequence analysis} (pp. 39--64). Springer.
+
+\leavevmode\hypertarget{ref-kimball2019phylogenomic}{}%
+Kimball, R. T., Oliveros, C. H., Wang, N., White, N. D., Barker, F. K., Field, D. J., \ldots{} others. (2019). A phylogenomic supertree of birds. \emph{Diversity}, \emph{11}(7), 109.
+
+\leavevmode\hypertarget{ref-klicka2014comprehensive}{}%
+Klicka, J., Barker, F.
+K., Burns, K.
+J., Lanyon, S.
+M., Lovette, I.
+J., Chaves, J.
+A., \& Bryson Jr, R.
+W.
+(2014).
+A comprehensive multilocus assessment of sparrow (aves: Passerellidae) relationships.
+\emph{Molecular Phylogenetics and Evolution}, \emph{77}, 177--182.
+
+\leavevmode\hypertarget{ref-ksepka2015fossil}{}%
+Ksepka, D.
+T., Parham, J.
+F., Allman, J.
+F., Benton, M.
+J., Carrano, M.
+T., Cranston, K.
+A., \ldots{} others.
+(2015).
+The fossil calibration database---a new resource for divergence dating.
+\emph{Systematic Biology}, \emph{64}(5), 853--859.
+
+\leavevmode\hypertarget{ref-lamichhaney2015evolution}{}%
+Lamichhaney, S., Berglund, J., Almén, M.
+S., Maqbool, K., Grabherr, M., Martinez-Barrio, A., \ldots{} others.
+(2015).
+Evolution of darwin's finches and their beaks revealed by genome sequencing.
+\emph{Nature}, \emph{518}(7539), 371--375.
+
+\leavevmode\hypertarget{ref-laubichler2009form}{}%
+Laubichler, M. D., \& Maienschein, J. (2009). \emph{Form and function in developmental evolution}. Cambridge University Press.
+
+\leavevmode\hypertarget{ref-lepage2004avibase}{}%
+Lepage, D. (2004). \emph{Avibase: The world bird database}. Bird Studies Canada.
+
+\leavevmode\hypertarget{ref-lepage2014avibase}{}%
+Lepage, D., Vaidya, G., \& Guralnick, R. (2014). Avibase--a database system for managing and organizing taxonomic concepts. \emph{ZooKeys}, (420), 117.
+
+\leavevmode\hypertarget{ref-lerner2011multilocus}{}%
+Lerner, H.
+R., Meyer, M., James, H.
+F., Hofreiter, M., \& Fleischer, R.
+C.
+(2011).
+Multilocus resolution of phylogeny and timescale in the extant adaptive radiation of hawaiian honeycreepers.
+\emph{Current Biology}, \emph{21}(21), 1838--1844.
+
+\leavevmode\hypertarget{ref-lovette2010comprehensive}{}%
+Lovette, I.
+J., Pérez-Emán, J.
+L., Sullivan, J.
+P., Banks, R.
+C., Fiorentino, I., Córdoba-Córdoba, S., \ldots{} others.
+(2010).
+A comprehensive multilocus phylogeny for the wood-warblers and a revised classification of the parulidae (aves).
+\emph{Molecular Phylogenetics and Evolution}, \emph{57}(2), 753--770.
+
+\leavevmode\hypertarget{ref-magallon2001absolute}{}%
+Magallon, S., \& Sanderson, M. (2001). Absolute diversification rates in angiosperm clades. \emph{Evolution}, \emph{55}(9), 1762--1780.
+
+\leavevmode\hypertarget{ref-magallon2010using}{}%
+Magallón, S. (2010). Using fossils to break long branches in molecular dating: A comparison of relaxed clocks applied to the origin of angiosperms. \emph{Systematic Biology}, \emph{59}(4), 384--399.
+
+\leavevmode\hypertarget{ref-magallon2015metacalibrated}{}%
+Magallón, S., Gómez-Acevedo, S., Sánchez-Reyes, L.
+L., \& Hernández-Hernández, T.
+(2015).
+A metacalibrated time-tree documents the early rise of flowering plant phylogenetic diversity.
+\emph{New Phytologist}, \emph{207}(2), 437--453.
+
+\leavevmode\hypertarget{ref-mctavish2015phylesystem}{}%
+McTavish, E.
+J., Hinchliff, C.
+E., Allman, J.
+F., Brown, J.
+W., Cranston, K.
+A., Holder, M.
+T., \ldots{} Smith, S.
+(2015).
+Phylesystem: A git-based data store for community-curated phylogenetic estimates.
+\emph{Bioinformatics}, \emph{31}(17), 2794--2800.
+
+\leavevmode\hypertarget{ref-Michonneau2016}{}%
+Michonneau, F., Brown, J.
+W., \& Winter, D.
+J.
+(2016).
+rotl: an R package to interact with the Open Tree of Life data.
+\emph{Methods in Ecology and Evolution}, \emph{7}(12), 1476--1481.
+\url{https://doi.org/10.1111/2041-210X.12593}
+
+\leavevmode\hypertarget{ref-Morlon2014}{}%
+Morlon, H. (2014). Phylogenetic approaches for studying diversification. \emph{Ecology Letters}, \emph{17}(4), 508--525. \url{https://doi.org/10.1111/ele.12251}
+
+\leavevmode\hypertarget{ref-moyle2016tectonic}{}%
+Moyle, R.
+G., Oliveros, C.
+H., Andersen, M.
+J., Hosner, P.
+A., Benz, B.
+W., Manthey, J.
+D., \ldots{} Faircloth, B.
+C.
+(2016).
+Tectonic collision and uplift of wallacea triggered the global songbird radiation.
+\emph{Nature Communications}, \emph{7}(1), 1--7.
+
+\leavevmode\hypertarget{ref-oliveros2019earth}{}%
+Oliveros, C.
+H., Field, D.
+J., Ksepka, D.
+T., Barker, F.
+K., Aleixo, A., Andersen, M.
+J., \ldots{} others.
+(2019).
+Earth history and the passerine superradiation.
+\emph{Proceedings of the National Academy of Sciences}, \emph{116}(16), 7916--7925.
+
+\leavevmode\hypertarget{ref-Ooms2018}{}%
+Ooms, J., \& Chamberlain, S. (2018). \emph{Phylocomr: Interface to 'phylocom'}. Retrieved from \url{https://CRAN.R-project.org/package=phylocomr}
+
+\leavevmode\hypertarget{ref-opentreeAPIs}{}%
+Open Tree Of Life, Redelings, B., Cranston, K.
+A., Allman, J., Holder, M.
+T., \& McTavish, E.
+J.
+(2016).
+Open Tree of Life APIs v3.0.
+\emph{Open Tree of Life Project}, (Online Resources).
+Retrieved from \href{\%7Bhttps://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\%7D}{\{https://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\}}
+
+\leavevmode\hypertarget{ref-opentreeoflife2019synth}{}%
+Open Tree Of Life, Redelings, B., Sánchez Reyes, L.
+L., Cranston, K.
+A., Allman, J., Holder, M.
+T., \& McTavish, E.
+J.
+(2019).
+Open tree of life synthetic tree v12.3.
+\emph{Zenodo}.
+Retrieved from \url{https://doi.org/10.5281/zenodo.3937742}
+
+\leavevmode\hypertarget{ref-odeen2011evolution}{}%
+Ödeen, A., Håstad, O., \& Alström, P. (2011). Evolution of ultraviolet vision in the largest avian radiation-the passerines. \emph{BMC Evolutionary Biology}, \emph{11}(1), 1--8.
+
+\leavevmode\hypertarget{ref-paradis2004}{}%
+Paradis, E., Claude, J., \& Strimmer, K. (2004). APE: analyses of phylogenetics and evolution in R language. \emph{Bioinformatics}, \emph{20}(2), 289--290.
+
+\leavevmode\hypertarget{ref-parchman2007coevolution}{}%
+Parchman, T.
+L., Benkman, C.
+W., \& Mezquida, E.
+T.
+(2007).
+Coevolution between hispaniolan crossbills and pine: Does more time allow for greater phenotypic escalation at lower latitude?
+\emph{Evolution}, \emph{61}(9), 2142--2153.
+
+\leavevmode\hypertarget{ref-packert2012horizontal}{}%
+Päckert, M., Martens, J., Sun, Y.-H., Severinghaus, L.
+L., Nazarenko, A.
+A., Ting, J., \ldots{} Tietze, D.
+T.
+(2012).
+Horizontal and elevational phylogeographic patterns of himalayan and southeast asian forest passerines (aves: Passeriformes).
+\emph{Journal of Biogeography}, \emph{39}(3), 556--573.
+
+\leavevmode\hypertarget{ref-pennell2014geiger}{}%
+Pennell, M.
+W., Eastman, J.
+M., Slater, G.
+J., Brown, J.
+W., Uyeda, J.
+C., FitzJohn, R.
+G., \ldots{} Harmon, L.
+J.
+(2014).
+Geiger v2.
+0: An expanded suite of methods for fitting macroevolutionary models to phylogenetic trees.
+\emph{Bioinformatics}, \emph{30}(15), 2216--2218.
+
+\leavevmode\hypertarget{ref-posadas2006historical}{}%
+Posadas, P., Crisci, J. V., \& Katinas, L. (2006). Historical biogeography: A review of its basic concepts and critical issues. \emph{Journal of Arid Environments}, \emph{66}(3), 389--403.
+
+\leavevmode\hypertarget{ref-powell2014comprehensive}{}%
+Powell, A.
+F., Barker, F.
+K., Lanyon, S.
+M., Burns, K.
+J., Klicka, J., \& Lovette, I.
+J.
+(2014).
+A comprehensive species-level molecular phylogeny of the new world blackbirds (icteridae).
+\emph{Molecular Phylogenetics and Evolution}, \emph{71}, 94--112.
+
+\leavevmode\hypertarget{ref-price2014niche}{}%
+Price, T.
+D., Hooper, D.
+M., Buchanan, C.
+D., Johansson, U.
+S., Tietze, D.
+T., Alström, P., \ldots{} others.
+(2014).
+Niche filling slows the diversification of himalayan songbirds.
+\emph{Nature}, \emph{509}(7499), 222.
+
+\leavevmode\hypertarget{ref-pulgarin2013multilocus}{}%
+Pulgarín-R, P.
+C., Smith, B.
+T., Bryson Jr, R.
+W., Spellman, G.
+M., \& Klicka, J.
+(2013).
+Multilocus phylogeny and biogeography of the new world pheucticus grosbeaks (aves: Cardinalidae).
+\emph{Molecular Phylogenetics and Evolution}, \emph{69}(3), 1222--1227.
+
+\leavevmode\hypertarget{ref-rabosky2015no}{}%
+Rabosky, D.
+L.
+(2015).
+No substitute for real data: A cautionary note on the use of phylogenies from birth--death polytomy resolvers for downstream comparative analyses.
+\emph{Evolution}, \emph{69}(12), 3207--3216.
+
+\leavevmode\hypertarget{ref-rabosky2018inverse}{}%
+Rabosky, D.
+L., Chang, J., Title, P.
+O., Cowman, P.
+F., Sallan, L., Friedman, M., \ldots{} others.
+(2018).
+An inverse latitudinal gradient in speciation rate for marine fishes.
+\emph{Nature}, \emph{559}(7714), 392.
+
+\leavevmode\hypertarget{ref-ramshaw1972time}{}%
+Ramshaw, J., Richardson, D., Meatyard, B., Brown, R., Richardson, M., Thompson, E., \& Boulter, D.
+(1972).
+The time of origin of the flowering plants determined by using amino acid sequence data of cytochrome c.
+\emph{New Phytologist}, \emph{71}(5), 773--779.
+
+\leavevmode\hypertarget{ref-ratnasingham2007bold}{}%
+Ratnasingham, S., \& Hebert, P. D. (2007). BOLD: The barcode of life data system (http://www. Barcodinglife. Org). \emph{Molecular Ecology Notes}, \emph{7}(3), 355--364.
+
+\leavevmode\hypertarget{ref-RCoreTeam2018}{}%
+R Core Team. (2018). \emph{R: a language and environment for statistical computing}. Vienna, Austria: R Foundation for Statistical Computing.
+
+\leavevmode\hypertarget{ref-rees2017automated}{}%
+Rees, \& Cranston, K. (2017). Automated assembly of a reference taxonomy for phylogenetic data synthesis. \emph{Biodiversity Data Journal}, (5).
+
+\leavevmode\hypertarget{ref-rees2017irmng}{}%
+Rees, Vandepitte, L., Decock, W., \& Vanhoorne, B. (2017). IRMNG 2006--2016: 10 Years of a Global Taxonomic Database. \emph{Biodiversity Informatics}, \emph{12}.
+
+\leavevmode\hypertarget{ref-Revell2012}{}%
+Revell, L. J. (2012). Phytools: An r package for phylogenetic comparative biology (and other things). \emph{Methods in Ecology and Evolution}, \emph{3}, 217--223.
+
+\leavevmode\hypertarget{ref-Ronquist2003}{}%
+Ronquist, F., \& Huelsenbeck, J.
+P.
+(2003).
+MrBayes 3: Bayesian phylogenetic inference under mixed models.
+\emph{Bioinformatics}, \emph{19}(12), 1572--1574.
+\url{https://doi.org/10.1093/bioinformatics/btg180}
+
+\leavevmode\hypertarget{ref-roquet2014one}{}%
+Roquet, C., Lavergne, S., \& Thuiller, W. (2014). One tree to link them all: A phylogenetic dataset for the european tetrapoda. \emph{PLoS Currents}, \emph{6}.
+
+\leavevmode\hypertarget{ref-datelifeplot}{}%
+Sanchez-Reyes, L.
+L., \& O'Meara, B.
+(2022).
+\texttt{datelifeplot}: Methods to plot chronograms and outputs of the \texttt{datelife} package.
+\emph{R Package Release V0.2.2}.
+Retrieved from \url{https://zenodo.org/badge/latestdoi/381501451}
+
+\leavevmode\hypertarget{ref-sanchez2022}{}%
+Sanchez-Reyes, L.
+L., O'Meara, B., Eastman, J., Heath, T., Wright, A., Schliep, K., \ldots{} Alfaro, M.
+(2022).
+\texttt{datelife}: Scientific Data on Time of Lineage Divergence for Your Taxa.
+In \emph{R package version 0.6.6}.
+Retrieved from \href{https://CRAN.R-project.org/package=datelife\%20and\%20https://doi.org/10.5281/zenodo.593938}{https://CRAN.R-project.org/package=datelife and https://doi.org/10.5281/zenodo.593938}
+
+\leavevmode\hypertarget{ref-sanderson2002}{}%
+Sanderson, M.
+(2002).
+Estimating Absolute Rates of Molecular Evolution and Divergence Times: A Penalized Likelihood Approach.
+\emph{Molecular Biology and Evolution}, \emph{19}(1), 101--109.
+\url{https://doi.org/10.1093/oxfordjournals.molbev.a003974}
+
+\leavevmode\hypertarget{ref-sanderson2003r8s}{}%
+Sanderson, M. (2003). r8s: Inferring Absolute Rates of Molecular Evolution and Divergence Times in the Absence of a Molecular Clock. \emph{Bioinformatics}, \emph{19}(2), 301--302.
+
+\leavevmode\hypertarget{ref-sanderson2001sources}{}%
+Sanderson, M., \& Doyle, J.
+(2001).
+Sources of error and confidence intervals in estimating the age of angiosperms from rbcL and 18S rDNA data.
+\emph{American Journal of Botany}, \emph{88}(8), 1499--1516.
+
+\leavevmode\hypertarget{ref-sauquet2013practical}{}%
+Sauquet, H. (2013). A practical guide to molecular dating. \emph{Comptes Rendus Palevol}, \emph{12}(6), 355--367.
+
+\leavevmode\hypertarget{ref-sauquet2021age}{}%
+Sauquet, H., Ramírez-Barahona, S., \& Magallón, S. (2021). \emph{The age of flowering plants is unknown}.
+
+\leavevmode\hypertarget{ref-schenk2016sec}{}%
+Schenk, J. J. (2016). Consequences of secondary calibrations on divergence time estimates. \emph{PLoS ONE}, \emph{11}(1). \url{https://doi.org/10.1371/journal.pone.0148228}
+
+\leavevmode\hypertarget{ref-schliep2011phangorn}{}%
+Schliep, K. P. (2011). Phangorn: Phylogenetic analysis in r. \emph{Bioinformatics}, \emph{27}(4), 592--593.
+
+\leavevmode\hypertarget{ref-schoch2020ncbi}{}%
+Schoch, C.
+L., Ciufo, S., Domrachev, M., Hotton, C.
+L., Kannan, S., Khovanskaya, R., \ldots{} others.
+(2020).
+NCBI Taxonomy: a Comprehensive Update on Curation, Resources and Tools.
+\emph{Database}, \emph{2020}.
+
+\leavevmode\hypertarget{ref-selvatti2015paleogene}{}%
+Selvatti, A.
+P., Gonzaga, L.
+P., \& Moraes Russo, C.
+A.
+de.
+(2015).
+A paleogene origin for crown passerines and the diversification of the oscines in the new world.
+\emph{Molecular Phylogenetics and Evolution}, \emph{88}, 1--15.
+
+\leavevmode\hypertarget{ref-smith2018constructing}{}%
+Smith, S., \& Brown, J. (2018). Constructing a broadly inclusive seed plant phylogeny. \emph{American Journal of Botany}, \emph{105}(3), 302--314.
+
+\leavevmode\hypertarget{ref-Smith2012}{}%
+Smith, S., \& O'Meara, B.
+(2012).
+TreePL: Divergence time estimation using penalized likelihood for large phylogenies.
+\emph{Bioinformatics}, \emph{28}(20), 2689--2690.
+\url{https://doi.org/10.1093/bioinformatics/bts492}
+
+\leavevmode\hypertarget{ref-Stoltzfus2013}{}%
+Stoltzfus, A., Lapp, H., Matasci, N., Deus, H., Sidlauskas, B., Zmasek, C.
+M., \ldots{} Jordan, G.
+(2013).
+Phylotastic!
+Making tree-of-life knowledge accessible, reusable and convenient.
+\emph{BMC Bioinformatics}, \emph{14}.
+\url{https://doi.org/10.1186/1471-2105-14-158}
+
+\leavevmode\hypertarget{ref-sun2020estimating}{}%
+Sun, M., Folk, R.
+A., Gitzendanner, M.
+A., Soltis, P.
+S., Chen, Z., Soltis, D.
+E., \& Guralnick, R.
+P.
+(2020).
+Estimating rates and patterns of diversification with incomplete sampling: A case study in the rosids.
+\emph{American Journal of Botany}, \emph{107}(6), 895--909.
+
+\leavevmode\hypertarget{ref-tietze2013complete}{}%
+Tietze, D.
+T., Päckert, M., Martens, J., Lehmann, H., \& Sun, Y.-H.
+(2013).
+Complete phylogeny and historical biogeography of true rosefinches (aves: Carpodacus).
+\emph{Zoological Journal of the Linnean Society}, \emph{169}(1), 215--234.
+
+\leavevmode\hypertarget{ref-title2016macrophylogenies}{}%
+Title, P.
+O., \& Rabosky, D.
+L.
+(2016).
+Do Macrophylogenies Yield Stable Macroevolutionary Inferences?
+An Example from Squamate Reptiles.
+\emph{Systematic Biology}, syw102.
+\url{https://doi.org/10.1093/sysbio/syw102}
+
+\leavevmode\hypertarget{ref-treplin2008molecular}{}%
+Treplin, S., Siegert, R., Bleidorn, C., Thompson, H.
+S., Fotso, R., \& Tiedemann, R.
+(2008).
+Molecular phylogeny of songbirds (aves: Passeriformes) and the relative utility of common nuclear marker loci.
+\emph{Cladistics}, \emph{24}(3), 328--349.
+
+\leavevmode\hypertarget{ref-uyeda2017evolution}{}%
+Uyeda, J.
+C., Pennell, M.
+W., Miller, E.
+T., Maia, R., \& McClain, C.
+R.
+(2017).
+The evolution of energetic scaling across the vertebrate tree of life.
+\emph{The American Naturalist}, \emph{190}(2), 185--199.
+
+\leavevmode\hypertarget{ref-vos2012nexml}{}%
+Vos, R.
+A., Balhoff, J.
+P., Caravas, J.
+A., Holder, M.
+T., Lapp, H., Maddison, W.
+P., \ldots{} others.
+(2012).
+NeXML: Rich, extensible, and verifiable representation of comparative data and metadata.
+\emph{Systematic Biology}, \emph{61}(4), 675--689.
+\url{https://doi.org/10.1093/sysbio/sys025}
+
+\leavevmode\hypertarget{ref-Webb2000}{}%
+Webb, C. (2000). Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees. \emph{The American Naturalist}, \emph{156}(2), 145--155.
+
+\leavevmode\hypertarget{ref-Webb2008}{}%
+Webb, C., Ackerly, D., \& Kembel, S.
+(2008).
+Phylocom: Software for the analysis of phylogenetic community structure and trait evolution.
+\emph{Bioinformatics}, \emph{24}(18), 2098--2100.
+\url{https://doi.org/10.1093/bioinformatics/btn358}
+
+\leavevmode\hypertarget{ref-webb2005phylomatic}{}%
+Webb, C., \& Donoghue, M. (2005). Phylomatic: Tree assembly for applied phylogenetics. \emph{Molecular Ecology Notes}, \emph{5}(1), 181--183.
+
+\leavevmode\hypertarget{ref-weir2008calibrating}{}%
+Weir, J., \& Schluter, D. (2008). Calibrating the avian molecular clock. \emph{Molecular Ecology}, \emph{17}(10), 2321--2328.
+
+\leavevmode\hypertarget{ref-zuccon2012phylogenetic}{}%
+Zuccon, D., Prŷs-Jones, R., Rasmussen, P.
+C., \& Ericson, P.
+G.
+(2012).
+The phylogenetic relationships and generic limits of finches (fringillidae).
+\emph{Molecular Phylogenetics and Evolution}, \emph{62}(2), 581--596.
+
+\endgroup
+
+\newpage
+
+\captionof{table}{Ages of congruified nodes. See Figure 3, step C2.}
+\label{tbl:table1}
+\includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example.pdf}
+
+\vspace{15mm}
+
+\captionof{table}{Summary of congruified nodes ages. See Figure 3, step C3.}
+
+\label{tbl:table2}
+\includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example-summary.png}
+
+\begin{minipage}[t]{17.3cm}
+\includegraphics{../figures/figure-workflow/figure1-new.pdf}
+\captionof{figure}{\textit{Main DateLife workflow.
+Analyses can be performed via DateLife's interactive website at \url{www.datelife.org}, or using the \texttt{datelife} R package.
+Details on the R functions used to perform the analyses are available from \texttt{datelife}'s R package vignettes at \url{https://phylotastic.org/datelife}.}}
+\label{fig:workflow}
+\end{minipage}
+
+\includegraphics{../figures/fig_runtime_main.pdf}
+\captionof{figure}{\textit{DateLife's benchmarking results.
+Computation time used to process a query and a search across \texttt{datelife}'s chronogram database, relative to number of input taxon names.
+For each N = \{10, 100, 200, ..., 1 000, ...
+, 9 000, 10 000\}, we sampled N species names from the class Aves a hundred times, and then performed a datelife search processing the input names with Taxon Names Resolution Service (TNRS; light gray), and without processing names (dark gray).
+For comparison, we performed a search using an input that had been pre-processed with TNRS (light blue).}}
+\label{fig:runtime_main}
 
@@
 \label{fig:figure2}
-
-\newpage
-
-\captionof{table}{Ages of congruified nodes. See Figure 3, step C2.}
-\label{tbl:table1}
-\includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example.pdf}
-
-\vspace{15mm}
-
-\captionof{table}{Summary of congruified nodes ages. See Figure 3, step C3.}
-
-\label{tbl:table2}
-\includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example-summary.png}
-
-\newpage
-
-\hypertarget{an-example-with-the-family-of-true-finches}{%
-\subsection{An example with the family of true finches}\label{an-example-with-the-family-of-true-finches}}
-
-\hypertarget{creating-a-query}{%
-\subsubsection{Creating a query}\label{creating-a-query}}
-
-To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the \enquote{get species from taxon} flag,
-which gets all recognized species names within a named group from a taxonomy of choice.
-Following the NCBI taxonomy, our DateLife query has 289 Fringillidae species names.
-This taxon-constrained approach implies that the full DateLife analysis will be performed using a tree topology and ages available for species names from a given taxonomic group, which do not necessarily correspond to a monophyletic group.
-Users can change this behaviour by providing all species names corresponding to a monophyletic group as input for a DateLife search, or a monophyletic tree to construct a DateLife summary.
-
-\hypertarget{searching-the-database-1}{%
-\subsubsection{Searching the database}\label{searching-the-database-1}}
-
-Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species as tip taxa.
-The DateLife search identified 19 chronograms matching this criteria, published in 13 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Kimball et al., 2019; Oliveros et al., 2019; Price et al., 2014; Roquet, Lavergne, \& Thuiller, 2014; Uyeda, Pennell, Miller, Maia, \& McClain, 2017).
-Once identified, DateLife pruned these matching chronograms to remove tips that do not belong to the queried taxon names, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementray Table S1).
-
-\hypertarget{summarizing-search-results-2}{%
-\subsubsection{Summarizing search results}\label{summarizing-search-results-2}}
-
-The final step entailed congruifying and summarizing the age data available for the Fringillidae species into two single summary chronograms, using two different types of summary ages, median and SDM.
-As explained in the \enquote{Description} section, a tree topology to summarize age data upon is required.
-By default, DateLife uses the topology from OpenTree's synthetic tree that contains all taxa from the search query.
-According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig.
-\ref{fig:fringillidae-topologies}).
-Hence, a topology containing only the 289 species from the original query was extracted from Open Tree of Life's synthetic tree v12.3 (Open Tree Of Life et al., 2019).
-
-\vspace{50mm}
 
@@
 
-Age data from source chronograms was congruified to OpenTree's topology (Fig.
-\ref{fig:fringillidae-topologies}B), reducing the age data set to 818 different data points (Supplementray Table S2).
-For each congruent node, age summary statistics were calculated and used as fixed secondary calibrations over the chosen tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
-\ref{fig:fringillidages}).
-
 \begin{minipage}{18cm}
@@
 \end{minipage}
-
-\hypertarget{cross-validation-test}{%
-\section{Cross-validation test}\label{cross-validation-test}}
-
-We performed a cross validation analysis of the DateLife workflow using the Fringillidae chronograms.
-We used the individual tree topologies from each of the 19 source chronograms from 13 studies as inputs, treating their node ages as unknown.
-We then estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.
-We found that node ages from original study, and ages estimated using all other age data available are correlated (Fig. \ref{fig:cvXY}).
-For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka \enquote{closer to the root}) relative to the original estimate, and overestimate ages for nodes closer to the tips.
-Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. S1).
 
@@
 
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-DateLife makes state-of-the-art data on evolutionary time frame easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
-It is an open service that does not require any expert biological knowledge from users --besides the names of the species or group they want to work with, for any of its functionality.
-
-A total of 99,474 unique terminal taxa are represented in DateLife's database.
-Incorporation of more chronograms into the database will continue to improve DateLife's services.
-One option to increase the number of chronograms in the DateLife database is the Dryad data repository.
-Methods to automatically mine chronograms from Dryad could be designed and implemented.
-However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined manually by a human curator.
-We would like to emphasize on the importance of sharing chronogram data, including systematically curated metadata, into open repositories, such as OpenTree's Phylesystem (McTavish et al., 2015) for the benefit of the scientific community as a whole.
-
-While it is known that incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance in ecological and conservation biology studies (Webb et al., 2008).
-Next, we discuss some particularities of using summarized node age data from a DateLife analysis that users should be aware of.
-
-\hypertarget{age-variation-in-source-chronograms}{%
-\subsection{Age variation in source chronograms}\label{age-variation-in-source-chronograms}}
-
-Conflict in estimated ages among alternative studies is common in the literature.
-See, for example, the robust ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández, 2015; Ramshaw et al., 1972; Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021).
-Source chronograms available for the same organisms have potentially been estimated implementing calibrations very differently.
-For example, the chronograms from Burns et al.
-(2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older age estimates for the same nodes than chronograms that were inferred using fossils as calibrations (Figs.
-\ref{fig:fringillidages}, \ref{fig:cvXY}; Supplementary Figs.
-S1, S5).
-
-Different calibration implementations might also imply fundamentally distinct evolutionary hypotheses (Antonelli et al., 2017).
-For example, two independent researchers working on the same clade should both carefully select and justify their choices of fossil calibration placement.
-Yet, if one researcher concludes that a fossil should calibrate the ingroup of a clade, while another researcher concludes that the same fossil should calibrate the outgroup of the clade, the resulting age estimates will differ, as the placement of calibrations as stem or crown group has been proven to significantly affect time of lineage divergence estimates (Sauquet, 2013).
-
-\hypertarget{primary-vs-secondary-calibrations}{%
-\subsubsection{Primary vs Secondary calibrations}\label{primary-vs-secondary-calibrations}}
-
-While most chronograms in DateLife's database are constructed using primary calibrations (molecular substitution rates or ages obtained from the fossil record or geological events), DateLife summarizes chronograms using secondary calibrations (ages coming from other chronograms).
-Graur and Martin (2004) cautioned on the increased error and uncertainty in estimated ages when using secondary calibrations in dating analyses.
-Schenk (2016) showed that, in simulations, divergence times inferred using secondary calibrations are significantly younger than those inferred with primary calibrations, when obtained with Bayesian inference methods, and when priors are implemented in similar ways in both analyses.
-Accordingly, the scientific community seems to have more confidence in chronograms obtained from a single analysis, using fossil data as primary sources of calibrations (Schenk, 2016), and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data reflect a coherent evolutionary history (Sauquet, 2013), as for example done by Antonelli et al.
-(2017).
-There have been attempts to create fossil calibration databases (Ksepka et al., 2015), though these still have room to grow.
-
-It seems that using several (as opposed to just a few) secondary calibrations can provide sufficient information to alleviate or even neutralize potential biases (Sauquet, 2013).
-Certainly, further studies are required to fully understand the effect of secondary calibrations on outputs from different tree dating methods, and on downstream analyses.
-It is possible that secondary calibrations can be safely used with dating methods that do not require setting priors, such as penalized likelihood (Sanderson, 2003), with methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ (Webb et al., 2008; Webb \& Donoghue, 2005), or methods that summarize age data unto a tree topology.
-
-Our cross validation analysis might provide some insight in this regard.
-When ages are estimated with secondary calibrations, nodes closer to the root do tend to be slightly younger than ages estimated with primary calibrations.
-However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures S2-S20).
-The only exception to this was observed on cross validation results of the Burns et al.
-(2014) chronogram, which displays much younger node ages when estimated using secondary calibrations (Supplementary Figs.
-S1, S5).
-
-\hypertarget{sumarizing-chronograms}{%
-\subsection{Sumarizing chronograms}\label{sumarizing-chronograms}}
-
-By default, DateLife currently summarizes all source chronograms that overlap with at least two species names. Users can exclude source chronograms if they have reasons to do so.
-Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely.
-To our knowledge, there are no tested measures to determine independently when a chronogram is better than another.
-Yet, several characteristics of the data used for dating analyses, as well as from the output chronogram itself, could be used to score the quality of source chronograms.
-
-Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used (Magallón, 2010; Magallón et al., 2015).
-Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
-
-DateLife provides an opportunity to capture concordance and conflict among date estimates, which can also be used as a metric for chronogram reliability.
-Its open database of chronograms allows other researchers to do such analyses themselves reproducibly, and without needing permission.
-Though, of course, they should follow proper citation practices, especially for the source chronogram studies.
-
-The exercise of summarizing age data from across multiple studies provides the opportunity to work with a chronogram that reflects a unified evolutionary history for a lineage, by putting together evidence from different hypotheses.
-The largest, and taxonomically broadest chronogram currently available from OpenTree was constructed summarizing age data from 2,274 published chronograms using NCBI's taxonomic tree as backbone (Hedges et al., 2015).
-A summarizing exercise may also amplify the effect of uncertainty and errors in source data, and blur parts of the evolutionary history of a lineage that might only be reflected in source chronograms and lost on the summary chronogram (Sauquet et al., 2021).
-
-\hypertarget{effects-of-phylogenetic-sampling-on-downstream-analyses}{%
-\subsection{Effects of phylogenetic sampling on downstream analyses}\label{effects-of-phylogenetic-sampling-on-downstream-analyses}}
-
-Analysis of species diversification using simulated and empirical phylogenies suggest that using a more completely sampled phylogeny provides estimates that are closer to the true diversification history than when analysing incompletely sampled phylogenies (Chang, Rabosky, \& Alfaro, 2020; Cusimano, Stadler, \& Renner, 2012; Sun et al., 2020).
-Ideally phylogenies should be completed using genetic data, but this is a very hard task to achieve for many biological groups.
-Hence, DateLife's workflow features different ways of assigning node ages in the absence of calibrations and branch length information for certain taxa.
-Completing a phylogeny using a stochastic birth-death politomy resolver and a backbone taxonomy is a common practice in scientific publications: Jetz et al.
-(2012), created a chronogram of all 9, 993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
-(2018) created a chronogram of 31, 536 ray-finned fishes, of which only 37\% had molecular data; Smith and Brown (2018) constructed a chronogram of 353, 185 seed plants where only 23\% had molecular data.
-These stochastically resolved chronograms provide less biased diversification rates estimated with methods that account for sampling fractions (Chang et al., 2020; Cusimano et al., 2012), but can also introduce spurious patterns of early bursts of diversification (Cusimano \& Renner, 2010; Sun et al., 2020).
-
-Taxonomy based politomy resolvers also introduce topological differences.
-The study of macroevolutionary processes largely depends on an understanding of the timing of species diversification events, and different phylogenetic and chronogram hypothesis can provide very different overviews of the macroeovlutionary history of a biological group.
-For example, alternative topologies in chronograms from the same biological group can infer very different species diversification patterns (Rabosky, 2015; Title \& Rabosky, 2016).
-Similarly there are worries that patterns of morphological evolution cannot be accurately inferred with these types of phylogenies, as any patterns would be erased by randomization (Rabosky, 2015).
-We note that the same applies for geographical and morphological dependent diversification analysis.
-Hence, we suggest that these types of randomly resolved phylogenies can be used as null or neutral models, representing the case of diversification independent of traits and geographical scenario.
-
-While stochastic politomy resolvers have been useful to advance research, notably, risks come with this practice.
-Taken to the extreme, one could generate a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy, a single calibration, and assigning branch lengths following a birth-death diversification model.
-Clearly, this can lead to a misrepresentation of the true evolutionary history.
-We urge DateLife users to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, potential biases, and assessing the consistency of the results with prior work.
-
-\hypertarget{conclusions}{%
-\section{Conclusions}\label{conclusions}}
-
-Knowledge of the evolutionary time frame of organisms is key to many research areas: trait evolution, species diversification, biogeography, macroecology and more.
-It is also crucial for education, science communication and policy, but generating chronograms is difficult, especially for those who want to use phylogenies but who are not systematists, or do not have the time to acquire and develop the necessary knowledge and skills to construct them on their own.
-Importantly, years of primarily publicly funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
-
-The DateLife project allows for easy and fast summary of public and state-of-the-art data on time of lineage divergence.
-It provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identifying regions that require more research, or that have conflicting information.
-It is available as an R package, and as a web-based R shiny application at www.datelife.org
-Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
-The DateLife project should improve awareness of the existing variation in expert time of divergence data, and foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
-
-\hypertarget{availability}{%
-\section{Availability}\label{availability}}
-
-The DateLife software is free and open source and it can be used through its R shiny web application at \url{http://www.datelife.org}, through the \texttt{datelife} R package, and through Phylotastic's project web portal \url{https://phylo.cs.nmsu.edu/}.
-DateLife's web application is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker and OpenTree's infrastructure (dates.opentreeoflife.org/datelife).
-\texttt{datelife}'s R package stable version is available for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife}) using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R.
-Development versions are available from the GitHub repository (\url{https://github.com/phylotastic/datelife}) and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.
-
-\hypertarget{supplementary-material}{%
-\section{Supplementary Material}\label{supplementary-material}}
-
-Supplementary material, including code, biological examples and benchmark results data files and online-only appendices, can be found in the Dryad data repository (\url{https://doi.org/10.5061/dryad.cnp5hqc6w}), as well as in the GitHub repositories used to develop the reproducible manuscript (\url{https://doi.org/10.5281/zenodo.7435094}), the biological examples (\url{https://doi.org/10.5281/zenodo.7435101}), and the software benchmark (\url{https://doi.org/10.5281/zenodo.7435106}).
-
-\hypertarget{funding}{%
-\section{Funding}\label{funding}}
-
-Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to the Datelife project; DBI-0905606 to the National Evolutionary Synthesis Center (NESCent); ABI-1458572 to the Phylotastic project; and ABI-1759846 to the Open Tree of Life project.
-
-\hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
-
-The DateLife project was born as a prototype tool aiming to provide these services, and was initially developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA (Stoltzfus et al., 2013).
-We thank colleagues from the O'Meara Lab at the University
-of Tennesse Knoxville for suggestions, discussions and software testing.
-The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
-that led to initial work on this project.
-The team that assembled DateLife's first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
-The Open Tree of Life project that provides the open, metadata rich repository of
-trees used to construct DateLife's chronogram database.
-The many scientists who publish their chronograms in an open, reusable form, and the scientists who curate them for deposition in the Open Tree of Life repository.
-The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.
-
-\newpage
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-\hypertarget{refs}{}
-\leavevmode\hypertarget{ref-agnarsson2008acctran}{}%
-Agnarsson, I., \& Miller, J. A. (2008). Is acctran better than deltran? \emph{Cladistics}, \emph{24}(6), 1032--1038.
-
-\leavevmode\hypertarget{ref-alstrom2014discovery}{}%
-Alström, P., Hooper, D.
-M., Liu, Y., Olsson, U., Mohan, D., Gelang, M., \ldots{} Price, T.
-D.
-(2014).
-Discovery of a relict lineage and monotypic family of passerine birds.
-\emph{Biology Letters}, \emph{10}(3), 20131067.
-
-\leavevmode\hypertarget{ref-ane2009groves}{}%
-Ané, C., Eulenstein, O., Piaggio-Talice, R., \& Sanderson, M. J. (2009). Groves of phylogenetic trees. \emph{Annals of Combinatorics}, \emph{13}(2), 139--167.
-
-\leavevmode\hypertarget{ref-antonelli2017supersmart}{}%
-Antonelli, A., Hettling, H., Condamine, F.
-L., Vos, K., Nilsson, R.
-H., Sanderson, M.
-J., \ldots{} Vos, R.
-A.
-(2017).
-Toward a self-updating platform for estimating rates of speciation and migration, ages, and relationships of Taxa.
-\emph{Systematic Biology}, \emph{66}(2), 153--166.
-\url{https://doi.org/10.1093/sysbio/syw066}
-
-\leavevmode\hypertarget{ref-archie1986newick}{}%
-Archie, J., Day, W.
-H., Felsenstein, J., Maddison, W., Meacham, C., Rohlf, F.
-J., \& Swofford, D.
-(1986).
-The Newick tree format.
-Retrieved from \href{\%7Bhttps://evolution.genetics.washington.edu/phylip/newicktree.html\%7D}{\{https://evolution.genetics.washington.edu/phylip/newicktree.html\}}
-
-\leavevmode\hypertarget{ref-avibase-emberiza}{}%
-Avibase.
-(2022).
-Yellow-throated Bunting.
-\emph{Avibase - the World Bird Database}, (Online Resource).
-Retrieved from \href{\%7Bhttps://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\%7D}{\{https://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\}}
-
-\leavevmode\hypertarget{ref-Bapst2012a}{}%
-Bapst, D.
-W.
-(2012).
-Paleotree: An R package for paleontological and phylogenetic analyses of evolution.
-\emph{Methods in Ecology and Evolution}, \emph{3}(5), 803--807.
-\url{https://doi.org/10.1111/j.2041-210X.2012.00223.x}
-
-\leavevmode\hypertarget{ref-barba2018constraining}{}%
-Barba-Montoya, J., Reis, M.
-dos, Schneider, H., Donoghue, P.
-C., \& Yang, Z.
-(2018).
-Constraining uncertainty in the timescale of angiosperm evolution and the veracity of a cretaceous terrestrial revolution.
-\emph{New Phytologist}, \emph{218}(2), 819--834.
-
-\leavevmode\hypertarget{ref-barker2014mitogenomic}{}%
-Barker, F. K. (2014). Mitogenomic data resolve basal relationships among passeriform and passeridan birds. \emph{Molecular Phylogenetics and Evolution}, \emph{79}, 313--324.
-
-\leavevmode\hypertarget{ref-barker2013going}{}%
-Barker, F.
-K., Burns, K.
-J., Klicka, J., Lanyon, S.
-M., \& Lovette, I.
-J.
-(2013).
-Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds.
-\emph{Systematic Biology}, \emph{62}(2), 298--320.
-
-\leavevmode\hypertarget{ref-barker2015new}{}%
-Barker, F.
-K., Burns, K.
-J., Klicka, J., Lanyon, S.
-M., \& Lovette, I.
-J.
-(2015).
-New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
-\emph{The Auk: Ornithological Advances}, \emph{132}(2), 333--348.
-
-\leavevmode\hypertarget{ref-barker2004phylogeny}{}%
-Barker, F.
-K., Cibois, A., Schikler, P., Feinstein, J., \& Cracraft, J.
-(2004).
-Phylogeny and diversification of the largest avian radiation.
-\emph{Proceedings of the National Academy of Sciences}, \emph{101}(30), 11040--11045.
-
-\leavevmode\hypertarget{ref-beresford2005african}{}%
-Beresford, P., Barker, F., Ryan, P., \& Crowe, T.
-(2005).
-African endemics span the tree of songbirds (passeri): Molecular systematics of several evolutionary ``enigmas''.
-\emph{Proceedings of the Royal Society B: Biological Sciences}, \emph{272}(1565), 849--858.
-
-\leavevmode\hypertarget{ref-Boyle2013}{}%
-Boyle, B., Hopkins, N., Lu, Z., Raygoza Garay, J.
-A., Mozzherin, D., Rees, T., \ldots{} Enquist, B.
-J.
-(2013).
-The taxonomic name resolution service: An online tool for automated standardization of plant names.
-\emph{BMC Bioinformatics}, \emph{14}(1).
-\url{https://doi.org/10.1186/1471-2105-14-16}
-
-\leavevmode\hypertarget{ref-Britton2007}{}%
-Britton, T., Anderson, C.
-L., Jacquet, D., Lundqvist, S., \& Bremer, K.
-(2007).
-Estimating Divergence Times in Large Phylogenetic Trees.
-\emph{Systematic Biology}, \emph{56}(788777878), 741--752.
-\url{https://doi.org/10.1080/10635150701613783}
-
-\leavevmode\hypertarget{ref-bryson2014diversification}{}%
-Bryson Jr, R.
-W., Chaves, J., Smith, B.
-T., Miller, M.
-J., Winker, K., Pérez-Emán, J.
-L., \& Klicka, J.
-(2014).
-Diversification across the new world within the `blue'cardinalids (aves: Cardinalidae).
-\emph{Journal of Biogeography}, \emph{41}(3), 587--599.
-
-\leavevmode\hypertarget{ref-burleigh2015building}{}%
-Burleigh, J. G., Kimball, R. T., \& Braun, E. L. (2015). Building the avian tree of life using a large-scale, sparse supermatrix. \emph{Molecular Phylogenetics and Evolution}, \emph{84}, 53--63.
-
-\leavevmode\hypertarget{ref-burns2014phylogenetics}{}%
-Burns, K.
-J., Shultz, A.
-J., Title, P.
-O., Mason, N.
-A., Barker, F.
-K., Klicka, J., \ldots{} Lovette, I.
-J.
-(2014).
-Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds.
-\emph{Molecular Phylogenetics and Evolution}, \emph{75}, 41--77.
-
-\leavevmode\hypertarget{ref-Chamberlain2018}{}%
-Chamberlain, S. (2018). \emph{bold: Interface to Bold Systems API}. Retrieved from \url{https://CRAN.R-project.org/package=bold}
-
-\leavevmode\hypertarget{ref-Chamberlain2013}{}%
-Chamberlain, S.
-A., \& Szöcs, E.
-(2013).
-taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}.
-\emph{F1000Research}, \emph{2}(191), 1--29.
-\url{https://doi.org/10.12688/f1000research.2-191.v2}
-
-\leavevmode\hypertarget{ref-chang2020estimating}{}%
-Chang, J., Rabosky, D.
-L., \& Alfaro, M.
-E.
-(2020).
-Estimating diversification rates on incompletely sampled phylogenies: Theoretical concerns and practical solutions.
-\emph{Systematic Biology}, \emph{69}(3), 602--611.
-
-\leavevmode\hypertarget{ref-chaves2013biogeography}{}%
-Chaves, J.
-A., Hidalgo, J.
-R., \& Klicka, J.
-(2013).
-Biogeography and evolutionary history of the n eotropical genus s altator (a ves: T hraupini).
-\emph{Journal of Biogeography}, \emph{40}(11), 2180--2190.
-
-\leavevmode\hypertarget{ref-claramunt2015new}{}%
-Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{Science Advances}, \emph{1}(11), e1501005.
-
-\leavevmode\hypertarget{ref-Criscuolo2006}{}%
-Criscuolo, A., Berry, V., Douzery, E.
-J., \& Gascuel, O.
-(2006).
-SDM: A fast distance-based approach for (super)tree building in phylogenomics.
-\emph{Systematic Biology}, \emph{55}(5), 740--755.
-\url{https://doi.org/10.1080/10635150600969872}
-
-\leavevmode\hypertarget{ref-cusimano2010slowdowns}{}%
-Cusimano, N., \& Renner, S. S. (2010). Slowdowns in diversification rates from real phylogenies may not be real. \emph{Systematic Biology}, \emph{59}(4), 458--464.
-
-\leavevmode\hypertarget{ref-cusimano2012new}{}%
-Cusimano, N., Stadler, T., \& Renner, S.
-S.
-(2012).
-A new method for handling missing species in diversification analysis applicable to randomly or nonrandomly sampled phylogenies.
-\emph{Systematic Biology}, \emph{61}(5), 785--792.
-
-\leavevmode\hypertarget{ref-delsuc2018phylogenomic}{}%
-Delsuc, F., Philippe, H., Tsagkogeorga, G., Simion, P., Tilak, M.-K., Turon, X., \ldots{} Douzery, E.
-J.
-(2018).
-A phylogenomic framework and timescale for comparative studies of tunicates.
-\emph{BMC Biology}, \emph{16}(1), 1--14.
-
-\leavevmode\hypertarget{ref-Eastman2013}{}%
-Eastman, J.
-M., Harmon, L.
-J., \& Tank, D.
-C.
-(2013).
-Congruification: Support for time scaling large phylogenetic trees.
-\emph{Methods in Ecology and Evolution}, \emph{4}(7), 688--691.
-\url{https://doi.org/10.1111/2041-210X.12051}
-
-\leavevmode\hypertarget{ref-edgar2004muscle}{}%
-Edgar, R. C. (2004). MUSCLE: Multiple sequence alignment with high accuracy and high throughput. \emph{Nucleic Acids Research}, \emph{32}(5), 1792--1797.
-
-\leavevmode\hypertarget{ref-Felsenstein1985a}{}%
-Felsenstein, J. (1985). Phylogenies and the Comparative Method. \emph{The American Naturalist}, \emph{125}(1), 1--15. Retrieved from \url{http://www.jstor.org/stable/2461605}
-
-\leavevmode\hypertarget{ref-forest2005teasing}{}%
-Forest, F., Savolainen, V., Chase, M.
-W., Lupia, R., Bruneau, A., \& Crane, P.
-R.
-(2005).
-Teasing apart molecular-versus fossil-based error estimates when dating phylogenetic trees: A case study in the birch family (betulaceae).
-\emph{Systematic Botany}, \emph{30}(1), 118--133.
-
-\leavevmode\hypertarget{ref-freckleton2002phylogenetic}{}%
-Freckleton, R. P., Harvey, P. H., \& Pagel, M. (2002). Phylogenetic analysis and comparative data: A test and review of evidence. \emph{The American Naturalist}.
-
-\leavevmode\hypertarget{ref-gbif2022taxonomy}{}%
-GBIF Secretariat.
-(2022).
-GBIF Backbone Taxonomy.
-\emph{Checklist dataset}, (Online Resource accessed via GBIF.org).
-Retrieved from \href{\%7Bhttps://doi.org/10.15468/39omei\%20\%7D}{\{https://doi.org/10.15468/39omei \}}
-
-\leavevmode\hypertarget{ref-gibb2015new}{}%
-Gibb, G.
-C., England, R., Hartig, G., McLenachan, P.
-A., Taylor Smith, B.
-L., McComish, B.
-J., \ldots{} Penny, D.
-(2015).
-New zealand passerines help clarify the diversification of major songbird lineages during the oligocene.
-\emph{Genome Biology and Evolution}, \emph{7}(11), 2983--2995.
-
-\leavevmode\hypertarget{ref-graur2004reading}{}%
-Graur, D., \& Martin, W. (2004). Reading the entrails of chickens: Molecular timescales of evolution and the illusion of precision. \emph{TRENDS in Genetics}, \emph{20}(2), 80--86.
-
-\leavevmode\hypertarget{ref-hackett2008phylogenomic}{}%
-Hackett, S.
-J., Kimball, R.
-T., Reddy, S., Bowie, R.
-C., Braun, E.
-L., Braun, M.
-J., \ldots{} others.
-(2008).
-A phylogenomic study of birds reveals their evolutionary history.
-\emph{Science}, \emph{320}(5884), 1763--1768.
-
-\leavevmode\hypertarget{ref-harvey1991comparative}{}%
-Harvey, P. H., Pagel, M. D., \& others. (1991). \emph{The comparative method in evolutionary biology} (Vol. 239). Oxford university press Oxford.
-
-\leavevmode\hypertarget{ref-Hedges2015}{}%
-Hedges, S.
-B., Marin, J., Suleski, M., Paymer, M., \& Kumar, S.
-(2015).
-Tree of life reveals clock-like speciation and diversification.
-\emph{Molecular Biology and Evolution}, \emph{32}(4), 835--845.
-\url{https://doi.org/10.1093/molbev/msv037}
-
-\leavevmode\hypertarget{ref-Heibl2008}{}%
-Heibl, C. (2008). \emph{PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages.} Retrieved from \url{http://www.christophheibl.de/Rpackages.html}
-
-\leavevmode\hypertarget{ref-hooper2017chromosomal}{}%
-Hooper, D. M., \& Price, T. D. (2017). Chromosomal inversion differences correlate with range overlap in passerine birds. \emph{Nature Ecology \& Evolution}, \emph{1}(10), 1526.
-
-\leavevmode\hypertarget{ref-Huelsenbeck2001}{}%
-Huelsenbeck, J. P., \& Ronquist, F. (2001). MRBAYES: Bayesian inference of phylogenetic trees. \emph{Bioinformatics}, \emph{17}(8), 754--755. \url{https://doi.org/10.1093/bioinformatics/17.8.754}
-
-\leavevmode\hypertarget{ref-Jetz2012}{}%
-Jetz, W., Thomas, G., Joy, J.
-J., Hartmann, K., \& Mooers, A.
-(2012).
-The global diversity of birds in space and time.
-\emph{Nature}, \emph{491}(7424), 444--448.
-\url{https://doi.org/10.1038/nature11631}
-
-\leavevmode\hypertarget{ref-johansson2008phylogenetic}{}%
-Johansson, U.
-S., Fjeldså, J., \& Bowie, R.
-C.
-(2008).
-Phylogenetic relationships within passerida (aves: Passeriformes): A review and a new molecular phylogeny based on three nuclear intron markers.
-\emph{Molecular Phylogenetics and Evolution}, \emph{48}(3), 858--876.
-
-\leavevmode\hypertarget{ref-katoh2009multiple}{}%
-Katoh, K., Asimenos, G., \& Toh, H. (2009). Multiple alignment of dna sequences with mafft. In \emph{Bioinformatics for dna sequence analysis} (pp. 39--64). Springer.
-
-\leavevmode\hypertarget{ref-kimball2019phylogenomic}{}%
-Kimball, R. T., Oliveros, C. H., Wang, N., White, N. D., Barker, F. K., Field, D. J., \ldots{} others. (2019). A phylogenomic supertree of birds. \emph{Diversity}, \emph{11}(7), 109.
-
-\leavevmode\hypertarget{ref-klicka2014comprehensive}{}%
-Klicka, J., Barker, F.
-K., Burns, K.
-J., Lanyon, S.
-M., Lovette, I.
-J., Chaves, J.
-A., \& Bryson Jr, R.
-W.
-(2014).
-A comprehensive multilocus assessment of sparrow (aves: Passerellidae) relationships.
-\emph{Molecular Phylogenetics and Evolution}, \emph{77}, 177--182.
-
-\leavevmode\hypertarget{ref-ksepka2015fossil}{}%
-Ksepka, D.
-T., Parham, J.
-F., Allman, J.
-F., Benton, M.
-J., Carrano, M.
-T., Cranston, K.
-A., \ldots{} others.
-(2015).
-The fossil calibration database---a new resource for divergence dating.
-\emph{Systematic Biology}, \emph{64}(5), 853--859.
-
-\leavevmode\hypertarget{ref-lamichhaney2015evolution}{}%
-Lamichhaney, S., Berglund, J., Almén, M.
-S., Maqbool, K., Grabherr, M., Martinez-Barrio, A., \ldots{} others.
-(2015).
-Evolution of darwin's finches and their beaks revealed by genome sequencing.
-\emph{Nature}, \emph{518}(7539), 371--375.
-
-\leavevmode\hypertarget{ref-laubichler2009form}{}%
-Laubichler, M. D., \& Maienschein, J. (2009). \emph{Form and function in developmental evolution}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-lepage2004avibase}{}%
-Lepage, D. (2004). \emph{Avibase: The world bird database}. Bird Studies Canada.
-
-\leavevmode\hypertarget{ref-lepage2014avibase}{}%
-Lepage, D., Vaidya, G., \& Guralnick, R. (2014). Avibase--a database system for managing and organizing taxonomic concepts. \emph{ZooKeys}, (420), 117.
-
-\leavevmode\hypertarget{ref-lerner2011multilocus}{}%
-Lerner, H.
-R., Meyer, M., James, H.
-F., Hofreiter, M., \& Fleischer, R.
-C.
-(2011).
-Multilocus resolution of phylogeny and timescale in the extant adaptive radiation of hawaiian honeycreepers.
-\emph{Current Biology}, \emph{21}(21), 1838--1844.
-
-\leavevmode\hypertarget{ref-lovette2010comprehensive}{}%
-Lovette, I.
-J., Pérez-Emán, J.
-L., Sullivan, J.
-P., Banks, R.
-C., Fiorentino, I., Córdoba-Córdoba, S., \ldots{} others.
-(2010).
-A comprehensive multilocus phylogeny for the wood-warblers and a revised classification of the parulidae (aves).
-\emph{Molecular Phylogenetics and Evolution}, \emph{57}(2), 753--770.
-
-\leavevmode\hypertarget{ref-magallon2001absolute}{}%
-Magallon, S., \& Sanderson, M. (2001). Absolute diversification rates in angiosperm clades. \emph{Evolution}, \emph{55}(9), 1762--1780.
-
-\leavevmode\hypertarget{ref-magallon2010using}{}%
-Magallón, S. (2010). Using fossils to break long branches in molecular dating: A comparison of relaxed clocks applied to the origin of angiosperms. \emph{Systematic Biology}, \emph{59}(4), 384--399.
-
-\leavevmode\hypertarget{ref-magallon2015metacalibrated}{}%
-Magallón, S., Gómez-Acevedo, S., Sánchez-Reyes, L.
-L., \& Hernández-Hernández, T.
-(2015).
-A metacalibrated time-tree documents the early rise of flowering plant phylogenetic diversity.
-\emph{New Phytologist}, \emph{207}(2), 437--453.
-
-\leavevmode\hypertarget{ref-mctavish2015phylesystem}{}%
-McTavish, E.
-J., Hinchliff, C.
-E., Allman, J.
-F., Brown, J.
-W., Cranston, K.
-A., Holder, M.
-T., \ldots{} Smith, S.
-(2015).
-Phylesystem: A git-based data store for community-curated phylogenetic estimates.
-\emph{Bioinformatics}, \emph{31}(17), 2794--2800.
-
-\leavevmode\hypertarget{ref-Michonneau2016}{}%
-Michonneau, F., Brown, J.
-W., \& Winter, D.
-J.
-(2016).
-rotl: an R package to interact with the Open Tree of Life data.
-\emph{Methods in Ecology and Evolution}, \emph{7}(12), 1476--1481.
-\url{https://doi.org/10.1111/2041-210X.12593}
-
-\leavevmode\hypertarget{ref-Morlon2014}{}%
-Morlon, H. (2014). Phylogenetic approaches for studying diversification. \emph{Ecology Letters}, \emph{17}(4), 508--525. \url{https://doi.org/10.1111/ele.12251}
-
-\leavevmode\hypertarget{ref-moyle2016tectonic}{}%
-Moyle, R.
-G., Oliveros, C.
-H., Andersen, M.
-J., Hosner, P.
-A., Benz, B.
-W., Manthey, J.
-D., \ldots{} Faircloth, B.
-C.
-(2016).
-Tectonic collision and uplift of wallacea triggered the global songbird radiation.
-\emph{Nature Communications}, \emph{7}(1), 1--7.
-
-\leavevmode\hypertarget{ref-oliveros2019earth}{}%
-Oliveros, C.
-H., Field, D.
-J., Ksepka, D.
-T., Barker, F.
-K., Aleixo, A., Andersen, M.
-J., \ldots{} others.
-(2019).
-Earth history and the passerine superradiation.
-\emph{Proceedings of the National Academy of Sciences}, \emph{116}(16), 7916--7925.
-
-\leavevmode\hypertarget{ref-Ooms2018}{}%
-Ooms, J., \& Chamberlain, S. (2018). \emph{Phylocomr: Interface to 'phylocom'}. Retrieved from \url{https://CRAN.R-project.org/package=phylocomr}
-
-\leavevmode\hypertarget{ref-opentreeAPIs}{}%
-Open Tree Of Life, Redelings, B., Cranston, K.
-A., Allman, J., Holder, M.
-T., \& McTavish, E.
-J.
-(2016).
-Open Tree of Life APIs v3.0.
-\emph{Open Tree of Life Project}, (Online Resources).
-Retrieved from \href{\%7Bhttps://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\%7D}{\{https://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\}}
-
-\leavevmode\hypertarget{ref-opentreeoflife2019synth}{}%
-Open Tree Of Life, Redelings, B., Sánchez Reyes, L.
-L., Cranston, K.
-A., Allman, J., Holder, M.
-T., \& McTavish, E.
-J.
-(2019).
-Open tree of life synthetic tree v12.3.
-\emph{Zenodo}.
-Retrieved from \url{https://doi.org/10.5281/zenodo.3937742}
-
-\leavevmode\hypertarget{ref-odeen2011evolution}{}%
-Ödeen, A., Håstad, O., \& Alström, P. (2011). Evolution of ultraviolet vision in the largest avian radiation-the passerines. \emph{BMC Evolutionary Biology}, \emph{11}(1), 1--8.
-
-\leavevmode\hypertarget{ref-paradis2004}{}%
-Paradis, E., Claude, J., \& Strimmer, K. (2004). APE: analyses of phylogenetics and evolution in R language. \emph{Bioinformatics}, \emph{20}(2), 289--290.
-
-\leavevmode\hypertarget{ref-parchman2007coevolution}{}%
-Parchman, T.
-L., Benkman, C.
-W., \& Mezquida, E.
-T.
-(2007).
-Coevolution between hispaniolan crossbills and pine: Does more time allow for greater phenotypic escalation at lower latitude?
-\emph{Evolution}, \emph{61}(9), 2142--2153.
-
-\leavevmode\hypertarget{ref-packert2012horizontal}{}%
-Päckert, M., Martens, J., Sun, Y.-H., Severinghaus, L.
-L., Nazarenko, A.
-A., Ting, J., \ldots{} Tietze, D.
-T.
-(2012).
-Horizontal and elevational phylogeographic patterns of himalayan and southeast asian forest passerines (aves: Passeriformes).
-\emph{Journal of Biogeography}, \emph{39}(3), 556--573.
-
-\leavevmode\hypertarget{ref-pennell2014geiger}{}%
-Pennell, M.
-W., Eastman, J.
-M., Slater, G.
-J., Brown, J.
-W., Uyeda, J.
-C., FitzJohn, R.
-G., \ldots{} Harmon, L.
-J.
-(2014).
-Geiger v2.
-0: An expanded suite of methods for fitting macroevolutionary models to phylogenetic trees.
-\emph{Bioinformatics}, \emph{30}(15), 2216--2218.
-
-\leavevmode\hypertarget{ref-posadas2006historical}{}%
-Posadas, P., Crisci, J. V., \& Katinas, L. (2006). Historical biogeography: A review of its basic concepts and critical issues. \emph{Journal of Arid Environments}, \emph{66}(3), 389--403.
-
-\leavevmode\hypertarget{ref-powell2014comprehensive}{}%
-Powell, A.
-F., Barker, F.
-K., Lanyon, S.
-M., Burns, K.
-J., Klicka, J., \& Lovette, I.
-J.
-(2014).
-A comprehensive species-level molecular phylogeny of the new world blackbirds (icteridae).
-\emph{Molecular Phylogenetics and Evolution}, \emph{71}, 94--112.
-
-\leavevmode\hypertarget{ref-price2014niche}{}%
-Price, T.
-D., Hooper, D.
-M., Buchanan, C.
-D., Johansson, U.
-S., Tietze, D.
-T., Alström, P., \ldots{} others.
-(2014).
-Niche filling slows the diversification of himalayan songbirds.
-\emph{Nature}, \emph{509}(7499), 222.
-
-\leavevmode\hypertarget{ref-pulgarin2013multilocus}{}%
-Pulgarín-R, P.
-C., Smith, B.
-T., Bryson Jr, R.
-W., Spellman, G.
-M., \& Klicka, J.
-(2013).
-Multilocus phylogeny and biogeography of the new world pheucticus grosbeaks (aves: Cardinalidae).
-\emph{Molecular Phylogenetics and Evolution}, \emph{69}(3), 1222--1227.
-
-\leavevmode\hypertarget{ref-rabosky2015no}{}%
-Rabosky, D.
-L.
-(2015).
-No substitute for real data: A cautionary note on the use of phylogenies from birth--death polytomy resolvers for downstream comparative analyses.
-\emph{Evolution}, \emph{69}(12), 3207--3216.
-
-\leavevmode\hypertarget{ref-rabosky2018inverse}{}%
-Rabosky, D.
-L., Chang, J., Title, P.
-O., Cowman, P.
-F., Sallan, L., Friedman, M., \ldots{} others.
-(2018).
-An inverse latitudinal gradient in speciation rate for marine fishes.
-\emph{Nature}, \emph{559}(7714), 392.
-
-\leavevmode\hypertarget{ref-ramshaw1972time}{}%
-Ramshaw, J., Richardson, D., Meatyard, B., Brown, R., Richardson, M., Thompson, E., \& Boulter, D.
-(1972).
-The time of origin of the flowering plants determined by using amino acid sequence data of cytochrome c.
-\emph{New Phytologist}, \emph{71}(5), 773--779.
-
-\leavevmode\hypertarget{ref-ratnasingham2007bold}{}%
-Ratnasingham, S., \& Hebert, P. D. (2007). BOLD: The barcode of life data system (http://www. Barcodinglife. Org). \emph{Molecular Ecology Notes}, \emph{7}(3), 355--364.
-
-\leavevmode\hypertarget{ref-RCoreTeam2018}{}%
-R Core Team. (2018). \emph{R: a language and environment for statistical computing}. Vienna, Austria: R Foundation for Statistical Computing.
-
-\leavevmode\hypertarget{ref-rees2017automated}{}%
-Rees, \& Cranston, K. (2017). Automated assembly of a reference taxonomy for phylogenetic data synthesis. \emph{Biodiversity Data Journal}, (5).
-
-\leavevmode\hypertarget{ref-rees2017irmng}{}%
-Rees, Vandepitte, L., Decock, W., \& Vanhoorne, B. (2017). IRMNG 2006--2016: 10 Years of a Global Taxonomic Database. \emph{Biodiversity Informatics}, \emph{12}.
-
-\leavevmode\hypertarget{ref-Revell2012}{}%
-Revell, L. J. (2012). Phytools: An r package for phylogenetic comparative biology (and other things). \emph{Methods in Ecology and Evolution}, \emph{3}, 217--223.
-
-\leavevmode\hypertarget{ref-Ronquist2003}{}%
-Ronquist, F., \& Huelsenbeck, J.
-P.
-(2003).
-MrBayes 3: Bayesian phylogenetic inference under mixed models.
-\emph{Bioinformatics}, \emph{19}(12), 1572--1574.
-\url{https://doi.org/10.1093/bioinformatics/btg180}
-
-\leavevmode\hypertarget{ref-roquet2014one}{}%
-Roquet, C., Lavergne, S., \& Thuiller, W. (2014). One tree to link them all: A phylogenetic dataset for the european tetrapoda. \emph{PLoS Currents}, \emph{6}.
-
-\leavevmode\hypertarget{ref-datelifeplot}{}%
-Sanchez-Reyes, L.
-L., \& O'Meara, B.
-(2022).
-\texttt{datelifeplot}: Methods to plot chronograms and outputs of the \texttt{datelife} package.
-\emph{R Package Release V0.2.2}.
-Retrieved from \url{https://zenodo.org/badge/latestdoi/381501451}
-
-\leavevmode\hypertarget{ref-datelifeR}{}%
-Sanchez-Reyes, L.
-L., O'Meara, B., Eastman, J., Heath, T., Wright, A., Schliep, K., \ldots{} Alfaro, M.
-(2022).
-\texttt{datelife}: Scientific Data on Time of Lineage Divergence for Your Taxa.
-\emph{R Package Release V0.6.2}.
-Retrieved from \url{https://doi.org/10.5281/zenodo.593938}
-
-\leavevmode\hypertarget{ref-sanderson2002}{}%
-Sanderson, M.
-(2002).
-Estimating Absolute Rates of Molecular Evolution and Divergence Times: A Penalized Likelihood Approach.
-\emph{Molecular Biology and Evolution}, \emph{19}(1), 101--109.
-\url{https://doi.org/10.1093/oxfordjournals.molbev.a003974}
-
-\leavevmode\hypertarget{ref-sanderson2003r8s}{}%
-Sanderson, M. (2003). r8s: Inferring Absolute Rates of Molecular Evolution and Divergence Times in the Absence of a Molecular Clock. \emph{Bioinformatics}, \emph{19}(2), 301--302.
-
-\leavevmode\hypertarget{ref-sanderson2001sources}{}%
-Sanderson, M., \& Doyle, J.
-(2001).
-Sources of error and confidence intervals in estimating the age of angiosperms from rbcL and 18S rDNA data.
-\emph{American Journal of Botany}, \emph{88}(8), 1499--1516.
-
-\leavevmode\hypertarget{ref-sauquet2013practical}{}%
-Sauquet, H. (2013). A practical guide to molecular dating. \emph{Comptes Rendus Palevol}, \emph{12}(6), 355--367.
-
-\leavevmode\hypertarget{ref-sauquet2021age}{}%
-Sauquet, H., Ramírez-Barahona, S., \& Magallón, S. (2021). \emph{The age of flowering plants is unknown}.
-
-\leavevmode\hypertarget{ref-schenk2016sec}{}%
-Schenk, J. J. (2016). Consequences of secondary calibrations on divergence time estimates. \emph{PLoS ONE}, \emph{11}(1). \url{https://doi.org/10.1371/journal.pone.0148228}
-
-\leavevmode\hypertarget{ref-schliep2011phangorn}{}%
-Schliep, K. P. (2011). Phangorn: Phylogenetic analysis in r. \emph{Bioinformatics}, \emph{27}(4), 592--593.
-
-\leavevmode\hypertarget{ref-schoch2020ncbi}{}%
-Schoch, C.
-L., Ciufo, S., Domrachev, M., Hotton, C.
-L., Kannan, S., Khovanskaya, R., \ldots{} others.
-(2020).
-NCBI Taxonomy: a Comprehensive Update on Curation, Resources and Tools.
-\emph{Database}, \emph{2020}.
-
-\leavevmode\hypertarget{ref-selvatti2015paleogene}{}%
-Selvatti, A.
-P., Gonzaga, L.
-P., \& Moraes Russo, C.
-A.
-de.
-(2015).
-A paleogene origin for crown passerines and the diversification of the oscines in the new world.
-\emph{Molecular Phylogenetics and Evolution}, \emph{88}, 1--15.
-
-\leavevmode\hypertarget{ref-smith2018constructing}{}%
-Smith, S., \& Brown, J. (2018). Constructing a broadly inclusive seed plant phylogeny. \emph{American Journal of Botany}, \emph{105}(3), 302--314.
-
-\leavevmode\hypertarget{ref-Smith2012}{}%
-Smith, S., \& O'Meara, B.
-(2012).
-TreePL: Divergence time estimation using penalized likelihood for large phylogenies.
-\emph{Bioinformatics}, \emph{28}(20), 2689--2690.
-\url{https://doi.org/10.1093/bioinformatics/bts492}
-
-\leavevmode\hypertarget{ref-Stoltzfus2013}{}%
-Stoltzfus, A., Lapp, H., Matasci, N., Deus, H., Sidlauskas, B., Zmasek, C.
-M., \ldots{} Jordan, G.
-(2013).
-Phylotastic!
-Making tree-of-life knowledge accessible, reusable and convenient.
-\emph{BMC Bioinformatics}, \emph{14}.
-\url{https://doi.org/10.1186/1471-2105-14-158}
-
-\leavevmode\hypertarget{ref-sun2020estimating}{}%
-Sun, M., Folk, R.
-A., Gitzendanner, M.
-A., Soltis, P.
-S., Chen, Z., Soltis, D.
-E., \& Guralnick, R.
-P.
-(2020).
-Estimating rates and patterns of diversification with incomplete sampling: A case study in the rosids.
-\emph{American Journal of Botany}, \emph{107}(6), 895--909.
-
-\leavevmode\hypertarget{ref-tietze2013complete}{}%
-Tietze, D.
-T., Päckert, M., Martens, J., Lehmann, H., \& Sun, Y.-H.
-(2013).
-Complete phylogeny and historical biogeography of true rosefinches (aves: Carpodacus).
-\emph{Zoological Journal of the Linnean Society}, \emph{169}(1), 215--234.
-
-\leavevmode\hypertarget{ref-title2016macrophylogenies}{}%
-Title, P.
-O., \& Rabosky, D.
-L.
-(2016).
-Do Macrophylogenies Yield Stable Macroevolutionary Inferences?
-An Example from Squamate Reptiles.
-\emph{Systematic Biology}, syw102.
-\url{https://doi.org/10.1093/sysbio/syw102}
-
-\leavevmode\hypertarget{ref-treplin2008molecular}{}%
-Treplin, S., Siegert, R., Bleidorn, C., Thompson, H.
-S., Fotso, R., \& Tiedemann, R.
-(2008).
-Molecular phylogeny of songbirds (aves: Passeriformes) and the relative utility of common nuclear marker loci.
-\emph{Cladistics}, \emph{24}(3), 328--349.
-
-\leavevmode\hypertarget{ref-uyeda2017evolution}{}%
-Uyeda, J.
-C., Pennell, M.
-W., Miller, E.
-T., Maia, R., \& McClain, C.
-R.
-(2017).
-The evolution of energetic scaling across the vertebrate tree of life.
-\emph{The American Naturalist}, \emph{190}(2), 185--199.
-
-\leavevmode\hypertarget{ref-vos2012nexml}{}%
-Vos, R.
-A., Balhoff, J.
-P., Caravas, J.
-A., Holder, M.
-T., Lapp, H., Maddison, W.
-P., \ldots{} others.
-(2012).
-NeXML: Rich, extensible, and verifiable representation of comparative data and metadata.
-\emph{Systematic Biology}, \emph{61}(4), 675--689.
-\url{https://doi.org/10.1093/sysbio/sys025}
-
-\leavevmode\hypertarget{ref-Webb2000}{}%
-Webb, C. (2000). Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees. \emph{The American Naturalist}, \emph{156}(2), 145--155.
-
-\leavevmode\hypertarget{ref-Webb2008}{}%
-Webb, C., Ackerly, D., \& Kembel, S.
-(2008).
-Phylocom: Software for the analysis of phylogenetic community structure and trait evolution.
-\emph{Bioinformatics}, \emph{24}(18), 2098--2100.
-\url{https://doi.org/10.1093/bioinformatics/btn358}
-
-\leavevmode\hypertarget{ref-webb2005phylomatic}{}%
-Webb, C., \& Donoghue, M. (2005). Phylomatic: Tree assembly for applied phylogenetics. \emph{Molecular Ecology Notes}, \emph{5}(1), 181--183.
-
-\leavevmode\hypertarget{ref-weir2008calibrating}{}%
-Weir, J., \& Schluter, D. (2008). Calibrating the avian molecular clock. \emph{Molecular Ecology}, \emph{17}(10), 2321--2328.
-
-\leavevmode\hypertarget{ref-zuccon2012phylogenetic}{}%
-Zuccon, D., Prŷs-Jones, R., Rasmussen, P.
-C., \& Ericson, P.
-G.
-(2012).
-The phylogenetic relationships and generic limits of finches (fringillidae).
-\emph{Molecular Phylogenetics and Evolution}, \emph{62}(2), 581--596.
-
-\endgroup
-
 \end{document}

### 2023-02-17 review.Rmd
remove figures from main text and finish answers to reviewer 1

---
title: "Response to reviews"
output: pdf_document
date: '`r format(Sys.time(), "%b %d, %Y")`'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

> 19-Oct-2022

> Dear Dr Sanchez Reyes:

> Decision on USYB-2022-152, DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life:

> Accept pending minor revisions.

> Thank you for your Systematic Biology submission. It has been reviewed by Associate Editor Dr Daniele Silvestro and two reviewers with the relevant expertise. Their comments are listed at the end of this letter. Both reviewers and the AE provide some excellent constructive suggestions that I am sure you will appreciate. The general consensus is that this software is of general interest to and expected to be widely used by the systematics and evolutionary biology community, providing capabilities that will help in the construction of supertrees. I agree with this assessment.

> I think that your paper will be a valuable contribution to Systematic Biology once the detailed comments provided below are addressed, especially in relation to data curation and the English style. Thank you very much for your submission.

> Sincerely,

> Dr Isabel Sanmartín
> Editor-in-Chief, Systematic Biology
> isanmartin@rjb.csic.es

> Please acknowledge (by email to sysbio.editorialoffice@oup.com) receipt of the reviews and give a probable time frame for the return of your revised manuscript. Your revision should be submitted as soon as possible. Any revision not received in a timely manner may have to be considered a new submission.

> Your revision must comply with our instructions to authors (in this letter, with additional instructions at https://academic.oup.com/sysbio/pages/General_Instructions). This applies even if your original submission deviated from Systematic Biology style and corrections were not included in the notes from the editors or reviewers. Therefore, READ AND APPLY OUR INSTRUCTIONS BEFORE SUBMITTING A REVISION. Failure to do so will result in significantly delayed processing of your paper. If anything remains unclear after reading the instructions, use recent issues of the journal to find examples. If necessary, feel free to contact the Managing Editor at sysbio.editorialoffice@oup.com for help.

> Log into https://mc.manuscriptcentral.com/systbiol and enter your Author Center, where you will find your manuscript title listed under "Manuscripts with Decisions." Under "Actions" click on "Create a Revision." Your manuscript number will be appended to denote a revision. Your original files are available to you. Delete the old files and replace them with your modified versions. Files that did not require revision can simply be retained

> You will be unable to make your revisions in ScholarOne Manuscripts. Instead, revise your paper using a word processing program and save it on your computer.
HIGHLIGHT the changes to your manuscript within the document using the track changes mode in MS Word or by using bold, strikethrough, or colored text; any method you choose which indicates all changes made. If a large number of changes were made and you feel the document is too cluttered to read easily with all of them shown, feel free to submit a "clean" copy as well, in which changes are not indicated. If you decide a clean copy would likely be helpful to reviewers and editors, please upload it in the category called Related Files.

> When submitting your revised manuscript, address each point made by the Editor, AE and Reviewers IN THE SPACE INDICATED (under "Response to Decision Letter"). Your revision cannot be processed if your responses to reviews are given only in a cover letter. The best way to address each point would be to copy this letter and insert your comments after each point made. Please do not change the order of or delete any of the comments because this makes it difficult to review again and would slow the process. The format of your responses must be compatible with ScholarOne Manuscripts text fields. For example, colored text is not an option, but you could use asterisks (and numbers, spacing, etc.) to clearly distinguish your responses from the text of the reviews.

> Feel free to argue your case, with careful consideration and documentation, if you disagree with any of the suggestions. If you feel a reviewer did not understand a point you made, in your response keep in mind that as an author it is your responsibility to make your points clear to the readers.

> DO NOT submit your revision in .pdf format. This applies to the main text, tables, and appendices. Individual figures may be in .pdf format.

> In the specific case of papers written using LaTeX, in addition to the .tex (and associated style, bib, or etc. files) please include a .pdf generated from those files. Upload the .pdf in the ScholarOne Manuscripts category called Related Files.

> Figures must be uploaded separately, with each file name including the figure number. No figures may be imbedded in or tied to the .pdf or .tex files.

> After you've uploaded your revision, carefully view the ScholarOne Manuscripts version to verify that all figures and other files display correctly, and that you’ve followed all of our author instructions.

> Figures: Each figure must be submitted as an individual file. The name of the file should include the figure number. Numbers and letters on figures should be as large and clear as possible, without overlapping any text. No figures may be imbedded in the document. Captions should not be placed on the figures. Figure portions should be referred to in the text and figure captions using lowercase letters. On the figures themselves, portions should be labeled by a lowercase letter followed by a single parenthesis (e.g., "a)"), located in the upper left area of the figure portion. Line thickness (including graph axes) should be a minimum stroke weight of 0.5pts, and 1.0pts is recommended for most lines. We prefer vector rather than bitmap figure formats. If desired, see http://systbio.org/?q=node/138 for an explanation of the difference. If bitmapped figures are necessary, they should be created at a minimum of 300 dpi and should be about 8 inches wide. We accept a wide variety of figure file types, as long as the figures are of sufficiently high resolution.

> The cost of printed color figures is $600 each. Authors are normally expected to cover this cost, but we do have limited funds available for authors who are SSB members, and cannot pay the full amount. If you feel your circumstances create an unusual financial need, please explain in your cover letter. The maximum allowance is one color figure per paper.
All color costs can be avoided if you decide to have the figures printed in black and white, but be shown in color in the online version of the paper. If you do this, the captions must be worded such that they are appropriate for both situations (e.g., descriptions should not name colors), because the captions in print and online will be identical even if color is used online only. Do not submit two versions of any figure. Instead, make sure the color figure is also easily legible in grayscale, then submit only the color version. In your cover letter, please make your intentions regarding the use of color clear.

> Journal Covers: Please consider submitting a suggested cover image. They can be illustrations of theory, photos of organisms, a combination of the two, or alternatives. These can be uploaded with a revision of the manuscript or may be sent later. Only images for which you have copyright permission or that are not under copyright may be submitted. The color figure fee does not apply to images chosen by the Editor to be on the cover of the issue. If you have questions about possible cover designs, please email sysbio.editorialoffice@oup.com. We encourage and greatly appreciate cover image suggestions.

> Data and Online Appendices: All such files should be deposited in the Dryad data repository. Also, before the manuscript can be published, data accession numbers must be in the text. Nucleotide sequence data and alignments must be submitted to GenBank (http://www.ncbi.nlm.nih.gov/genbank) or EMBL (http://www.ebi.ac.uk); alignment, input file and tree files must be submitted to TreeBASE (http://www.treebase.org); morphological data must be submitted to either Morphbank or MorphoBank. Check over your online appendices or supplemental material (if any) carefully, because they will not be copyedited or proofread, and cannot be changed later.
The first time you mention online-only material in the text of your paper, give the doi provided by Dryad as the location where the material can be found.
Then, in a separate section after the main text, include a statement such as:

> SUPPLEMENTARY MATERIAL
> Supplementary material, including data files and/or online-only appendices, can be found in the Dryad data repository (put your Dryad doi in parentheses here).

We added DOIs for Dryad package with Supplementary Material, L551.

> Make sure all section headings conform to Systematic Biology style for first, second and third levels. Use of incorrect styles is potentially confusing and, in any case, is likely to delay processing of the manuscript.

Noted and checked.

> Tables should have a single-sentence informative title above the table, with any other descriptive information located below the table in the form of notes and/or specific footnotes.

Noted and checked.

> When references are grouped together in parentheses in the text, they should be listed in ascending chronological order. Multiple references in a single year should be alphabetized.

Noted and checked.

> All funding used for this work should be listed in a "Funding" section preceding the Acknowledgements. Please give the full official name of each funding body.

Done on L556-558.

> Our accepted abbreviation for millions of years ago is Ma. The abbreviation for millions of years duration is myr.

Checked.

> Our full instructions are accessible via the link in the upper right area of the ScholarOne Manuscripts page (the link goes to https://academic.oup.com/sysbio/pages/General_Instructions). In case of a discrepancy between the information at that site and this letter, follow the instructions in this letter.

> If you are the first author and your manuscript is based on work done while you were a student, you would be eligible for the "Publisher's Award for Excellence in Systematic Research." When submitting your revised version be sure to indicate student work in the checkbox provided. If two student researchers were heavily involved in the manuscript, please briefly describe the situation in your cover letter.

This work was done while the first author was a Postdoctoral researcher.

> Please note that the accepted version of your manuscript will be published on Advance Access prior to typesetting and copyediting; it is therefore important that all files, including tables, figures, and supplementary files, are properly labelled and with the editorial office at this stage of the process -  it is the author’s responsibility to ensure that the correct files are submitted.  Please also keep in mind that files will be published as submitted (ie, please review table formatting and labelling and position of figure images, etc). For further information please see the Advance Access Publication section at: https://academic.oup.com/sysbio/pages/General_Instructions.

Noted and checked.

---
---

> Associate Editor: Dr Daniele Silvestro

> Recommendation #1: Accept with minor revisions

> Comments to the Author:
> Dear Dr Luna L Sanchez Reyes and co-authors,

> many thanks for submitting your manuscript to Systematic Biology. Your study was reviewed by two highly qualified reviewers who provided an overall positive assessment of the paper while listing a number of things that should be revised and clarified. Based on their assessment and my own reading of the manuscript I invite you to resubmit it after carefully revising to address each and all points raised by the reviewers.

Thanks for your reviews and comments! They were most helpful. We carefully revised the manuscript to address all points raised by the reviewers and yourself and we hope you find the revision suitable.

---
> In addition to that, please also address the following points:

> 1. how and why is a parsimony method used to estimate branch lengths and how is this used in combination with a likelihood method (line 174)?

Thanks for this question. We elaborate on this on the Description section, subsection "Dating a tree with branch lengths" L221-228. 
Briefly, we implement the parsimony algorithm ACCTRAN using functions from th R package phangorn. We chose this algorithm as a quick way to obtaining initial branch lengths that can be optimized under ML. ACCTRAN resolves ambiguous character optimization by assigning changes along branches of the tree as close to the root as possible (agnarsson2008acctran), so this works well with the type of matrices generated with BOLD data. 
Once estimated, the parsimony branch lengths become parameters that are then optimized using Maximum Likelihood, given the alignment, the topology and a simple JC model.
<!--ACCTRAN has been implemented in Forest et al. 2005. Teasing apart molecular-versus fossil-based error estimates when dating phylogenetic trees: a case study in the birch family Betulaceae. Systematic Botany.-->
We also add a vignette to the R package in which we showcase the whole workflow of branch length optimization, from parsimony to ML.

---
> 2. Why are the node ages evenly distributed between calibrations? I would expect an exponential distribution of node ages under a standard birth-death process.

Thanks for pointing this out. We elaborate on this on the "Description" section, at the end of subsection "Dating a tree topology with no branch lengths", L 202-210
(Note that we expanded the subsection "Dating a tree topology" into three sections: "Applying secondary calibrations", "Dating a tree topology" and "Dating a tree with branch lengths").

Briefly, in the absence of genetic data from an alignment, nodes that have no age data from published studies in the chronogram database are assigned an age based solely on other nodes that do have age data from published studies.
MrBayes has an option in which you can use a birth-death strict-clock model to sample branch lengths in the absence of genetic data. 
We have implemented a DateLife function called "make_mrbayes_tree", that wraps this MrBayes service.
Inherited from MrBayes implementation, the function requires setting parameter values for birth and death rates. If the amount of missing data is large, the chosen parameters can determine the "shape" of the resulting summary chronogram in terms of branch length distribution.

We added a discussion section, "Effects of phylogenetic sampling on downstream analyses" (L445 of manuscript without differences; L527 of manuscript with differences), in which we discuss the effects of chronograms with missing data that have been completed under a certain diversification model, especially when used for analyses that require estimating a diversification rate, which might introduce circularity (Rabosky, 2015. Evolution doi.org/10.1111/evo.12817).

MrBayes also allow us to generate a cloud of trees produced with different birth-death rates that could be used for downstream analyses. However, the main goal of the summarizing step in DateLife is to provide a single chronogram that can quickly show in one glance the distribution of node ages based on published data, in the most agnostic way possible.

This is why we chose an algorithm that distributes node ages evenly between calibrations and does not require any assumptions on the underlying model of branch length distribution. The Branch Length Adjuster (BLADJ) algorithm allows this and is the one set as default for the summarizing step. However, users that wish to use MrBayes instead con do so with the "make_mrbayes_tree" function.

---
> 4. I think the use of an arbitrary root age set by default as 10% older than the oldest age is unjustified and dangerous. If no root age is provided by the user, I think the function should return an interpretable error message and refuse to run.

We agree, we changed the behaviour of the code to fix this, and adjusted the ms accordingly (L196-201).

It now stops with an informative message on how users can provide an age for the root. The goal is to make sure that users are aware that there is no age data available for that root!

---
> Please make sure to carefully revise the text to remove typos.

We revised the text throughout to remove all typos we could detect.

---
> As one of the reviewers pointed out, it is good to provide links to permanent repositories for your code. I see that DateLife actually is already hosted in a Zenodo repository, so maybe you can add the link to your Availability section to make it more visible.

Right! We added Zenodo links for our code and other materials used for this research in the Availability section (L505) and Supplementary Material (L518-520).

---
> I hope you will be willing to revise and resubmit your paper and that you’ll find these and the Reviewers’ comments useful.

> Best regards, Daniele Silvestro

Yes, thanks!

---
> Reviewer(s)' comments to author:

> Reviewer: 1

> Comments to the Author
> The manuscript introduces DateLife, an R package and web service that provides time-calibrated phylogenies across a number of organismal groups. It integrates with a number of other services including the Open Tree of Life project. The authors provide some benchmarks and walk through a worked example of using their service.

> I don't agree with the characterization that stochastic polytomy resolutions methods such as PASTIS (used in Jetz et al 2012) is "making up" these branch lengths (line 455). My understanding of the phrase "making up" implies that these are solely inventions of the researcher, rather than generated through the use of well-tested statistical models. The manuscript also claims that there are no thorough analyses of phylogenies generated in this way (lines 462-465), which is not the case, and I suggest the authors revise this section in light of some of the relevant literature in this area.

> * Cusimano et al Syst Bio 2012
> * Thomas et al MEE 2013
> * Rabosky Evol 2015
> * Chang et al Syst Bio 2019
> * Title and Rabosky MEE 2019
> * Sun et al AJB 2020

Thanks for pointing us to these references. We elaborated on this subject more thoroughly by adding results from these studies to the corresponding section on the Discussion, "Effects of phylogenetic sampling on downstream analyses" (L481 of manuscript without differences; L527 of manuscript with differences), as well as changing the wording from "making up" to "in the absence of genetic data, simulating following a birth-death process" (L489; L540).

---
> There were many typographical errors in the manuscript which should be corrected prior to publication.

We corrected typos across the text.

---
> Reviewer: 2

> Comments to the Author
> As I have checked the box that I don't need to remain anonymous, there is also no point in being mysterious about this: I have been aware of DateLife for a good long while because I've seen its earliest prototype develop at a workshop at NESCent ages ago. I've loved the idea ever since - combined with some healthy reservations that I am happy to share here.

> DL synthesizes results from previous research. On the one hand that's great, but on the other, it invites the 'garbage in - garbage out' problem. Although OTOL has its own curation and QC facilities, the fact that users can provide their own garbage trees makes it so that the service might end up decorating nonsensical data, tainting its own reputation in the process. It would be good if the authors could emphasize this a bit more.

This is a good point. We would not like for this to become a common use for the software. We mention this in the abstract, introduction and discussion.

---
> A separate but related point that I would also like to see discussed is that synthesizing services such as DL and OTOL seem capable of ending up in loops where bad trees with bad calibration points provide the skeleton for further bad trees based on the former - with their own seemingly well-supported but in fact dodgy secondary calibrations. Is that a risk? What can be done about it?

It is a risk that we are aware of, but it is possible to avoid it. We explain how in the text.

---
> Also related: will we gradually start developing a body of literature with trees where the root always just happens to be ±10% older than the oldest nodes? Might that be bad?

Good point. It would certainly be a bad thing. The reason we implemented this initially is because we wanted DateLife to be able to provide a summary chronogram even if tehre is just one point of age data available.

As you pointed out, when the root age is absent, it is not possible to return any dated tree.

---
> Apart from these general points that might be touched upon a bit more in the Discussion, here now some specifics about the manuscript:

> - The Abstract looks like an extreme afterthought. I understand how that works, but please have another look. I see verb disagreement on line 21 and on line 23. Probably needs a comma after databases on line 25. On the same line, 'timeframe' is spelt as one word (fine by me), but elsewhere it's two words. Line 27: 'incetivizited' is not a thing. Line 29, 'finding' scans weird, maybe use 'discovery'? Line 36 probably needs 'use' instead of 'using' but the sentence is hard to parse. Line 38, 'awereness' is wrong. In this way, the Abstract is quite different from the rest of the MS, which is otherwise well written.

---
> - In the first paragraph of the Intro you might want to add something like 'comparative analysis' (Harvey & Pagel, yada yada yada). It's clearly something that's on your mind because in the Conclusions, 'trait evolution' is the first research area you mention as needing chronograms.

Thanks for pointing that out. We added that topic and a couple of references to the intro, L48-49.

---
> - On page 7, second paragraph, you state that subspecies are ignored. What do you mean precisely? My guess is that you ignore the subspecific epithet and collapse to species level. Maybe state that more clearly.

Thanks for pointing this out. We actually only ignore subspecies when retrieving data from a more inclusive taxonomic group. When provided by the user, subspecific taxa are processed and searched as regular species.
We clarify this in paragraph X, page 7

It would be possible to drop the subspecific epithet and perform a more general search. We added this functionality to our software development plan, so that it can be soon implemented in a future iteration of the software.

---
> - On page 7, third paragraph: how does the TNRS deal with homonyms? Given that we are in the tree realm it should be possible to infer intelligently whether some label is zoological or botanical code. Or is Aotus simply always the monkey, which is much cooler than that Australian legume genus?

Yeah! this is a pretty cool quality of TNRS that we implement in datelife. It is possible to provide or identify the "biological context" of a toxonomic name, so that if a list of names has mainly monkeys, then it will pick _Aotus_ the monkey, but if it has mainly plants, it will pick the Australian legume.

Moreover, homonyms in the Open Tree taxonomy specify the group they belong too, so users can easily identify if the numeric identifier they obtained from TNRS processing belongs to the group they wanted or not. For example _Aotus_ the legume is referred to as "Aotus (genus in kingdom Archaeplastida)" and the monkey as "Aotus (genus in Opisthokonta)".

We added examples for the two points above in a new vignette for the package, available at http://phylotastic.org/datelife/articles/make_datelife_query.html

---

> - On page 8, fourth paragraph, it's not quite clear whether DL's database syncs automatically with Phylesystem or whether you have volunteered yourself for this task. Which would be noble, but hard to sustain.

It is indeed currently synced "manually". However we are working to have the syncs done automatically.
We comment this in the discussion and clarified that it is currently a manual sync on page 8.

---
> - On page 10, second paragraph: mining BOLD and aligning the sequences automatically is very cool functionality but I did not see it exposed on the website at all. How can users get at those alignments? Also, might there be performance issues? MAFFT can be quite greedy with larger data sets.

Ah, yes, thanks! Well, the BOLD workflow is faster than a classic workflow, but it still takes considerable time, specifically computational time required for the database search, which the server can't afford yet. The functions are currently only available on DateLife's R package and not on the web application, but we plan on making them available there in the future.
Regarding possible performance issues, the alignments from BOLD data are not very large (around 1000bp), so we expect both MUSCLE and MAFFT to perform approximately the same, and MUSCLE is the default aligner.

---
> - On page 25 you mention the fossilcalibrations.org initiative. Maybe that's a good opportunity to go a bit into what we need as a community. I suspect that, in general, most people in this field think that doing it by themself is 'better', i.e. do a bunch of sequencing (hybseq right now, I guess?) and then get good primary calibration points. Natural history collections must have many more of those, both as fossils but also from geology (i.e. vicariant events having to do with tectonics, orogeny, etc.). Shouldn't we want *that*?

---
> - Page 26, line 416 has some typos.

Fixed :)

---
> - On page 26 you discuss some criteria for scoring quality of chronograms. One additional criterion might be where the calibration points are placed. Nodes that have a calibration point between them and the root have less freedom of movement and hence narrower confidence intervals. Ages ago, I did a bit of simulation work on that (Vos & Mooers, 2004 - definitely no need to cite). Maybe someone else has discussed this a bit better?

---
> - Page 27 line 448, chronogram should be plural, I think.

Fixed!

---
> - Page 28 line 473: I think it should be either 'public-funded' or 'publicly funded'

Went for "publicly funded", thanks!

---
> - Page 29, Supplementary Material: it's probably better to sync the repos with Zenodo and cite the DOI, just so that it's guaranteed unchanging.

Thanks for the suggestion! We created stable versions for all three repositories on Zenodo, and refer the doi instead of the GitHub address for the repos.

### 2023-02-17 notes-effects-downstream-analysis-review.Rmd.R

@@ -69,7 +69,7 @@ treatment of missing species influences both parameter estimation and type I and
 
 @rabosky2015no shows that randomly resolved phylogenies are inappropriate to study trait evolution,
 because unresolved lineages are placed on the phylogeny at random relative to the trait values.
-At best, these phylogenies could serve as a null hypothesis of randomness in trait evolution.
+At best, LLSR believe these phylogenies could serve as a null hypothesis of randomness in trait evolution.
 
 "birth–death polytomy resolver is a method of accounting for phylogenetic uncertainty that places
 missing (unsampled) taxa onto phylogenetic trees, using taxonomic information alone"

### 2023-02-21 manuscript/manuscript.Rmd
provide a better abstract

@@
 
-abstract: |
-  Achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram) is a difficult and time-consuming task.
-But the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
-    Knowledge of the evolutionary time frame of organisms is key for research in the natural sciences.
-It also represent valuable information for education, science communication, and policy decisions.
-  When chronograms are shared in public and open databases, this wealth of expertly-curated and peer-reviewed data on evolutionary timeframe is exposed in a programatic and reusable way, as intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
-  Here we present DateLife, a service implemented as an R package and an R Shiny website application available at www.datelife.org, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time frame of evolution.
-  The main DateLife workflow constructs a chronogram for any given combination of taxon names by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from the TreeBASE database as well.
-  We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
-  DateLife will be useful to increase awareness of the existing variation in alternative hypothesis of evolutionary time for the same organisms, and can foster exploration of the effect of alternative evolutionary timing hypotheses on the results of downstream analyses, providing a framework for a more informed interpretation of evolutionary results.
-
 bibliography      : ["paper_references.bib"]
-keywords          : "Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations"
-wordcount         : "`r wordcountaddin::word_count()`"
 floatsintext      : no
@@
 
+\begin{center}
+Abstract
+\end{center}
+
+  Chronograms –phylogenies with branch lengths proportional to time– represent key data on evolutionary time frame for the study of natural processes in many areas of biological research.
+  Chronograms also provide valuable information that can be used for education, science communication, and policy decisions.
+  Yet, achieving a high-quality reconstruction of a chronogram is a difficult and resource-consuming task.
+  Here we present DateLife, a service implemented as an R package and an R Shiny web application available at www.datelife.org, that provides services for efficient and easy discovery, summary, reuse, and reanalysis of node age data mined from a curated database of expert, peer-reviewed, and openly available chronograms.
+    The main DateLife workflow starts with one or more scientific taxon names provided by a user.
+Names are processed and standardized to a unified taxonomy.
+Then, DateLife runs a name match across its local chronogram database that is curated from Open Tree of Life's phylogenetic repository, and extracts all chronograms that contain at least two queried taxon names, along with their metadata.
+    Finally, node ages from matching chronograms are mapped upon a chosen tree topology using the congruification algorithm.
+Congruified node ages are used as secondary calibrations to date the chosen topology, with or without initial branch lengths, using different phylogenetic dating methods such as BLADJ, treePL, PATHd8 and MrBayes.
+    We performed a cross-validation test to compare node ages resulting from a DateLife analysis (i.e, phylogenetic dating using secondary calibrations) to those from the original chronograms (i.e, obtained with primary calibrations), and found that DateLife's node age estimates appear generally correlated to those from the original chronograms, with the largest variation in ages occurring around topologically deeper nodes.
+  Software such as DateLife can help increase awareness of the existing variation in alternative hypothesis of evolutionary time frame for the same organisms, and can support exploration of the effect of alternative chronogram hypotheses on downstream analyses, providing a framework for a more informed interpretation of evolutionary results.
+
+_Keywords_: Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations.
+
+Word count: `r wordcountaddin::word_count()`
+
+\newpage 
 <!-- # Introduction -->
@@
 
-# Description
+\begin{center}
+\textsc{Description}
+\end{center}
 
-DateLife's core software application consists of the R package `datelife`.
-Its current stable version -- v`r packageVersion("datelife")`, is available from The Comprehensive R Archive Network (CRAN) repository [@sanchez2022], and relies on functionalities from various biological R packages:
+DateLife's core software applications are implemented in the R package `datelife`.
+Its current stable version -- v`r packageVersion("datelife")`, is available from The Comprehensive R Archive Network (CRAN) repository [@sanchez2022], and relies on functionalities from other biological R packages:
 ape [@paradis2004],
@@
 DateLife implements TNRS using the Open Tree of Life (OpenTree) unified Taxonomy [OTT, @opentreeAPIs; @rees2017automated] as standard, storing taxonomic identification numbers (OTT ids) for further processing and analysis.
-Other taxonomies currently supported by DateLife are the National Center of Biotechnology Information (NCBI) taxonomic database [@schoch2020ncbi], the Global Biodiversity Information Facility (GBIF) taxonomic backbone [@gbif2022taxonomy], and the Interim Register of Marine and Nonmarine Genera (IRMNG) database [@rees2017irmng].
+Other taxonomies currently supported by DateLife are the National Center of Biotechnology Information (NCBI) taxonomic database [@schoch2020ncbi], the Global Biodiversity Information Facility (GBIF) taxonomic backbone [@gbif2022taxonomy], and the Interim Register of Marine and Non-marine Genera (IRMNG) database [@rees2017irmng].
 <!-- We are using OpenTree's TNRS [BCO: yes], Boyle described TNRS first, right?  [BCO: earliest I can find, but feels like there should be much older work]-->
@@
 A unique feature of the Phylesystem is that any user can add new published, state-of-the-art chronograms any time, through their curator application (https://tree.opentreeoflife.org/curator).
-As chronograms are added to Phylesystem, they can be incorporated into the chronogram database of the `datelife` package.
-`datelife`'s chronogram database is currently manually updated as new chronogram data is added to Phylesystem.
+As chronograms are added to Phylesystem, they can be incorporated into the chronogram database of the `datelife` R package, which is currently manually updated as new chronogram data is added to Phylesystem.
 The updated database is assigned a new version number, followed by a package release on CRAN.
-Users can also implement functions from the `datelife` R package to trigger an update of the local chronogram database, to incorporate any new chronograms to the user's DateLife analysis before an official database update is released on CRAN.
+Users can directly implement `datelife` functions to trigger an update of the local chronogram database, to incorporate any new chronograms to their DateLife analysis before an official database update is released on CRAN.
 <!--BCO: What is immediately?
@@
 
-BOLD trees generated by DateLife, and any tree with branch lengths provided by a user can be analysed with phylogenetic dating methods supported in DateLife that incorporate branch length information from the input topology in combination with the secondary calibrations:
+Phylogenetic dating methods supported in DateLife that incorporate branch length information from the input topology in combination with the secondary calibrations include:
 PATHd8, a non-clock, rate-smoothing method to date trees [@Britton2007];
 treePL [@Smith2012], a semi-parametric, rate-smoothing, penalized likelihood dating method [@sanderson2002]; 
-and MRBayes [@Huelsenbeck2001; @Ronquist2003], a Bayesian inference program implementing Markov chain Monte Carlo (MCMC) methods to estimate a posterior distribution of model parameters.
+and MrBayes [@Huelsenbeck2001; @Ronquist2003], a Bayesian inference program implementing Markov chain Monte Carlo (MCMC) methods to estimate a posterior distribution of model parameters.
 
@@
 
-`datelife`'s R package code speed was tested on an Apple iMac
+R package `datelife` code speed was tested on an Apple iMac
 with one 3.4 GHz Intel Core i5 processor. <!--No multiprocessor tests were performed-->
@@
 
-We illustrate the DateLife workflow using a family within the Passeriform birds encompassing the true finches, Fringillidae, as case study.
+We illustrate the DateLife workflow using a family within the passeriform birds encompassing the true finches, Fringillidae, as case study.
 On a small example, we analysed 6 bird species, and results from each step of the workflow are shown in Fig.
@@
 <!--\textcolor{red}{EJM: I don't understand what this sentence means}
-LLSR: it is related to the synonym found by tnrs processing, if we had used the synonym insted of the one accepted in the database, we would not have discovered that information
+LLSR: it is related to the synonym found by tnrs processing, if we had used the synonym instead of the one accepted in the database, we would not have discovered that information
 LLSR: I moved it here, it makes more sense here, I think!.
@@
 This taxon-constrained approach implies that the full DateLife analysis will be performed using a tree topology and ages available for species names from a given taxonomic group, which do not necessarily correspond to a monophyletic group.
-Users can change this behaviour by providing all species names corresponding to a monophyletic group as input for a DateLife search, or a monophyletic tree to construct a DateLife summary.
+Users can change this behavior by providing all species names corresponding to a monophyletic group as input for a DateLife search, or a monophyletic tree to construct a DateLife summary.
 <!--BCO: it's ambiguous: it gets a topology and age for the larger clade that includes all Fringillidae species? Or for the paraphyletic Fringillidae group?
@@
 The DateLife search identified `r length(dres)` chronograms matching this criteria, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @kimball2019phylogenomic; @oliveros2019earth; @price2014niche; @roquet2014one; @uyeda2017evolution].
-Once identified, DateLife pruned these matching chronograms to remove tips that do not belong to the queried taxon names, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementray Table S1).
+Once identified, DateLife pruned these matching chronograms to remove tips that do not belong to the queried taxon names, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementary Table S1).
 
@@
 Age data from source chronograms was congruified to OpenTree's topology (Fig.
-\ref{fig:fringillidae-topologies}B), reducing the age data set to 818 different data points (Supplementray Table S2).
+\ref{fig:fringillidae-topologies}B), reducing the age data set to 818 different data points (Supplementary Table S2).
 For each congruent node, age summary statistics were calculated and used as fixed secondary calibrations over the chosen tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
@@
 We found that node ages from original study, and ages estimated using all other age data available are correlated (Fig. \ref{fig:cvXY}).
-For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka 'closer to the root') relative to the original estimate, and overestimate ages for nodes closer to the tips.
+For five studies, DateLife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka 'closer to the root') relative to the original estimate, and overestimate ages for nodes closer to the tips.
 Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. S1).
@@
 Source chronograms available for the same organisms have potentially been estimated implementing calibrations very differently.
-For example, the chronograms from @burns2014phylogenetics were inferred using molecular substitution rate estimates across birds [@weir2008calibrating], and have much older age estimates for the same nodes than chronograms that were inferred using fossils as calibrations (Figs.
+In the Fringillidae example shown here, the chronograms from @burns2014phylogenetics were inferred using molecular substitution rate estimates across birds [@weir2008calibrating], and have much older age estimates for the same nodes than chronograms that were inferred using fossils as calibrations (Figs.
 \ref{fig:fringillidages}, \ref{fig:cvXY}; Supplementary Figs.
@@
 
-<!-- Comparison of available age data for a wide range of organisms shown here suggest that discrpenacy in age estimates across studies is a widespread phenomenon that requires further attention. -->
+<!-- Comparison of available age data for a wide range of organisms shown here suggest that discrepancy in age estimates across studies is a widespread phenomenon that requires further attention. -->
 
@@
 Taxonomy based politomy resolvers also introduce topological differences.
-The study of macroevolutionary processes largely depends on an understanding of the timing of species diversification events, and different phylogenetic and chronogram hypothesis can provide very different overviews of the macroeovlutionary history of a biological group.
+The study of macroevolutionary processes largely depends on an understanding of the timing of species diversification events, and different phylogenetic and chronogram hypothesis can provide very different overviews of the macroevolutionary history of a biological group.
 For example, alternative topologies in chronograms from the same biological group can infer very different species diversification patterns [@rabosky2015no; @title2016macrophylogenies].
@@
 DateLife's web application is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker and OpenTree's infrastructure (dates.opentreeoflife.org/datelife).
-`datelife`'s R package stable version `r availability` for installation from CRAN repository using the command `install.packages(pkgs = "datelife")` from within R.
+`datelife`'s stable version `r availability` for installation from CRAN repository using the command `install.packages(pkgs = "datelife")` from within R.
 Development versions are available from DateLife's GitHub repository (<https://github.com/phylotastic/datelife>) and can be installed using the command `devtools::install_github("phylotastic/datelife")`.
@@
 
-Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to the Datelife project; DBI-0905606 to the National Evolutionary Synthesis Center (NESCent); ABI-1458572 to the Phylotastic project; and ABI-1759846 to the Open Tree of Life project.
+Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to the DateLife project; DBI-0905606 to the National Evolutionary Synthesis Center (NESCent); ABI-1458572 to the Phylotastic project; and ABI-1759846 to the Open Tree of Life project.
 
@@
 We thank colleagues from the O'Meara Lab at the University
-of Tennesse Knoxville for suggestions, discussions and software testing.
+of Tennessee Knoxville for suggestions, discussions and software testing.
 The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons

### 2023-02-21 manuscript/manuscript.tex
provide a better abstract

@@
             pdfauthor={Luna L. Sánchez Reyes1,2, Emily Jane McTavish1, \& Brian O'Meara2},
-            pdfkeywords={Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations},
             pdfborder={0 0 0},
@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 5629}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 
-\abstract{%
-Achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram) is a difficult and time-consuming task.
-But the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
-Knowledge of the evolutionary time frame of organisms is key for research in the natural sciences. It also represent valuable information for education, science communication, and policy decisions.
-When chronograms are shared in public and open databases, this wealth of expertly-curated and peer-reviewed data on evolutionary timeframe is exposed in a programatic and reusable way, as intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
-Here we present DateLife, a service implemented as an R package and an R Shiny website application available at www.datelife.org, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time frame of evolution.
-The main DateLife workflow constructs a chronogram for any given combination of taxon names by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from the TreeBASE database as well.
-We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
-DateLife will be useful to increase awareness of the existing variation in alternative hypothesis of evolutionary time for the same organisms, and can foster exploration of the effect of alternative evolutionary timing hypotheses on the results of downstream analyses, providing a framework for a more informed interpretation of evolutionary results.
-}
-
 \begin{document}
 \maketitle
+
+\begin{center}
+Abstract
+\end{center}
+
+Chronograms --phylogenies with branch lengths proportional to time-- represent key data on evolutionary time frame for the study of natural processes in many areas of biological research.
+Chronograms also provide valuable information that can be used for education, science communication, and policy decisions.
+Yet, achieving a high-quality reconstruction of a chronogram is a difficult and resource-consuming task.
+Here we present DateLife, a service implemented as an R package and an R Shiny web application available at www.datelife.org, that provides services for efficient and easy discovery, summary, reuse, and reanalysis of node age data mined from a curated database of expert, peer-reviewed, and openly available chronograms.
+The main DateLife workflow starts with one or more scientific taxon names provided by a user.
+Names are processed and standardized to a unified taxonomy.
+Then, DateLife runs a name match across its local chronogram database that is curated from Open Tree of Life's phylogenetic repository, and extracts all chronograms that contain at least two queried taxon names, along with their metadata.
+Finally, node ages from matching chronograms are mapped upon a chosen tree topology using the congruification algorithm.
+Congruified node ages are used as secondary calibrations to date the chosen topology, with or without initial branch lengths, using different phylogenetic dating methods such as BLADJ, treePL, PATHd8 and MrBayes.
+We performed a cross-validation test to compare node ages resulting from a DateLife analysis (i.e, phylogenetic dating using secondary calibrations) to those from the original chronograms (i.e, obtained with primary calibrations), and found that DateLife's node age estimates appear generally correlated to those from the original chronograms, with the largest variation in ages occurring around topologically deeper nodes.
+Software such as DateLife can help increase awareness of the existing variation in alternative hypothesis of evolutionary time frame for the same organisms, and can support exploration of the effect of alternative chronogram hypotheses on downstream analyses, providing a framework for a more informed interpretation of evolutionary results.
+
+\emph{Keywords}: Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations.
+
+Word count: 5956
+
+\newpage
 
@@
 
-\hypertarget{description}{%
-\section{Description}\label{description}}
-
-DateLife's core software application consists of the R package \texttt{datelife}.
-Its current stable version -- v0.6.6, is available from The Comprehensive R Archive Network (CRAN) repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
+\begin{center}
+\textsc{Description}
+\end{center}
+
+DateLife's core software applications are implemented in the R package \texttt{datelife}.
+Its current stable version -- v0.6.6, is available from The Comprehensive R Archive Network (CRAN) repository (Sanchez-Reyes et al., 2022), and relies on functionalities from other biological R packages:
 ape (Paradis, Claude, \& Strimmer, 2004),
@@
 DateLife implements TNRS using the Open Tree of Life (OpenTree) unified Taxonomy (OTT, Open Tree Of Life et al., 2016; Rees \& Cranston, 2017) as standard, storing taxonomic identification numbers (OTT ids) for further processing and analysis.
-Other taxonomies currently supported by DateLife are the National Center of Biotechnology Information (NCBI) taxonomic database (Schoch et al., 2020), the Global Biodiversity Information Facility (GBIF) taxonomic backbone (GBIF Secretariat, 2022), and the Interim Register of Marine and Nonmarine Genera (IRMNG) database (Rees et al., 2017).
+Other taxonomies currently supported by DateLife are the National Center of Biotechnology Information (NCBI) taxonomic database (Schoch et al., 2020), the Global Biodiversity Information Facility (GBIF) taxonomic backbone (GBIF Secretariat, 2022), and the Interim Register of Marine and Non-marine Genera (IRMNG) database (Rees et al., 2017).
 
@@
 A unique feature of the Phylesystem is that any user can add new published, state-of-the-art chronograms any time, through their curator application (\url{https://tree.opentreeoflife.org/curator}).
-As chronograms are added to Phylesystem, they can be incorporated into the chronogram database of the \texttt{datelife} package.
-\texttt{datelife}'s chronogram database is currently manually updated as new chronogram data is added to Phylesystem.
+As chronograms are added to Phylesystem, they can be incorporated into the chronogram database of the \texttt{datelife} R package, which is currently manually updated as new chronogram data is added to Phylesystem.
 The updated database is assigned a new version number, followed by a package release on CRAN.
-Users can also implement functions from the \texttt{datelife} R package to trigger an update of the local chronogram database, to incorporate any new chronograms to the user's DateLife analysis before an official database update is released on CRAN.
+Users can directly implement \texttt{datelife} functions to trigger an update of the local chronogram database, to incorporate any new chronograms to their DateLife analysis before an official database update is released on CRAN.
 
@@
 
-BOLD trees generated by DateLife, and any tree with branch lengths provided by a user can be analysed with phylogenetic dating methods supported in DateLife that incorporate branch length information from the input topology in combination with the secondary calibrations:
+Phylogenetic dating methods supported in DateLife that incorporate branch length information from the input topology in combination with the secondary calibrations include:
 PATHd8, a non-clock, rate-smoothing method to date trees (Britton et al., 2007);
 treePL (Smith \& O'Meara, 2012), a semi-parametric, rate-smoothing, penalized likelihood dating method (Sanderson, 2002);
-and MRBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), a Bayesian inference program implementing Markov chain Monte Carlo (MCMC) methods to estimate a posterior distribution of model parameters.
+and MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), a Bayesian inference program implementing Markov chain Monte Carlo (MCMC) methods to estimate a posterior distribution of model parameters.
 
@@
 
-\texttt{datelife}'s R package code speed was tested on an Apple iMac
+R package \texttt{datelife} code speed was tested on an Apple iMac
 with one 3.4 GHz Intel Core i5 processor.
@@
 This taxon-constrained approach implies that the full DateLife analysis will be performed using a tree topology and ages available for species names from a given taxonomic group, which do not necessarily correspond to a monophyletic group.
-Users can change this behaviour by providing all species names corresponding to a monophyletic group as input for a DateLife search, or a monophyletic tree to construct a DateLife summary.
+Users can change this behavior by providing all species names corresponding to a monophyletic group as input for a DateLife search, or a monophyletic tree to construct a DateLife summary.
 
@@
 The DateLife search identified 19 chronograms matching this criteria, published in 13 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Kimball et al., 2019; Oliveros et al., 2019; Price et al., 2014; Roquet, Lavergne, \& Thuiller, 2014; Uyeda, Pennell, Miller, Maia, \& McClain, 2017).
-Once identified, DateLife pruned these matching chronograms to remove tips that do not belong to the queried taxon names, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementray Table S1).
+Once identified, DateLife pruned these matching chronograms to remove tips that do not belong to the queried taxon names, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementary Table S1).
 
@@
 We found that node ages from original study, and ages estimated using all other age data available are correlated (Fig. \ref{fig:cvXY}).
-For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka \enquote{closer to the root}) relative to the original estimate, and overestimate ages for nodes closer to the tips.
+For five studies, DateLife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka \enquote{closer to the root}) relative to the original estimate, and overestimate ages for nodes closer to the tips.
 Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. S1).
@@
 Source chronograms available for the same organisms have potentially been estimated implementing calibrations very differently.
-For example, the chronograms from Burns et al.
+In the Fringillidae example shown here, the chronograms from Burns et al.
 (2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older age estimates for the same nodes than chronograms that were inferred using fossils as calibrations (Figs.
@@
 Taxonomy based politomy resolvers also introduce topological differences.
-The study of macroevolutionary processes largely depends on an understanding of the timing of species diversification events, and different phylogenetic and chronogram hypothesis can provide very different overviews of the macroeovlutionary history of a biological group.
+The study of macroevolutionary processes largely depends on an understanding of the timing of species diversification events, and different phylogenetic and chronogram hypothesis can provide very different overviews of the macroevolutionary history of a biological group.
 For example, alternative topologies in chronograms from the same biological group can infer very different species diversification patterns (Rabosky, 2015; Title \& Rabosky, 2016).
@@
 DateLife's web application is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker and OpenTree's infrastructure (dates.opentreeoflife.org/datelife).
-\texttt{datelife}'s R package stable version is available for installation from CRAN repository using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R.
+\texttt{datelife}'s stable version is available for installation from CRAN repository using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R.
 Development versions are available from DateLife's GitHub repository (\url{https://github.com/phylotastic/datelife}) and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.

### 2023-02-21 notes-effects-downstream-analysis-review.Rmd
provide a better abstract

---
title             : "Review: Effects of initial phylogenetic and chronogram hypothesis on evolutionary downstream analyses"
shorttitle        : "Litt review"

author:
  - name          : "Luna L. Sánchez Reyes"
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : ""
    email         : "sanchez.reyes.luna@gmail.com"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Data curation
      - Investigation
      - Software
      - Visualization
      - Validation
      - Writing
      - Original Draft Preparation
      - Writing
      - Review & Editing

affiliation:
  - id            : "1"
    institution   : "University of California, Merced, USA"
  - id            : "2"
    institution   : "University of Tennessee, Knoxville, USA"

authornote: |
  School of Natural Sciences, University of California, Merced, 258 Science and Engineering Building 1, Merced, CA 95340, USA.
  Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 446 Hesler Biology Building, Knoxville, TN 37996, USA.
bibliography      : ["paper_references.bib"]
wordcount         : "`r wordcountaddin::word_count()`"
floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            :
  papaja::apa6_pdf:
    includes      :
      in_header: "preamble.tex"
---

```{r setup, include = FALSE}
library("papaja")
r_refs("paper_references.bib")
embed_figure <- TRUE
```

For Discussion
tendency to underestimate ages for topologically deeper nodes, and overestimate ages for nodes closer to the tips. However, node ages obtained with DateLife are generally correlated with those from the dating analysis that used primary calibrations.

@cusimano2012new: Simulating branch lengths following a birth-death species diversification model for
missing taxa to chronograms with non random sampling, lowers type I error (false positive,
incorrectly accepting the null hypothesis of a constant-rate or temporally varying rate
birth-death model) when analyzing changes in diversification rate using the gamma statistic.

@thomas2013pastis hypothesize that results of diversification analyses will be biased towards
the birth-death model used to simulate branch lengths;
and note that "the effects of missing species placement or polytomy resolution are
less clear for other phylogeny-based analyses (e.g. correlates of diversification,
modelling trait evolution, community phylogenetics), and future work should test how the
treatment of missing species influences both parameter estimation and type I and II errors"

@rabosky2015no shows that randomly resolved phylogenies are inappropriate to study trait evolution,
because unresolved lineages are placed on the phylogeny at random relative to the trait values.
At best, LLSR believe these phylogenies could serve as a null hypothesis of randomness in trait evolution.

"birth–death polytomy resolver is a method of accounting for phylogenetic uncertainty that places
missing (unsampled) taxa onto phylogenetic trees, using taxonomic information alone"

Adding branch lengths sampled from a birth-death model in the absence of genetic data
improves insight from phylogeny-based analyses of diversification rates [@chang2020estimating].
@chang2020estimating state that stochastic politomy resolution alleviate low statistical power to
identify shifts in diversification rate with methods that account for sampling fractions.
@chang2020estimating "demonstrate theoretical concerns with the widespread use of analytical
sampling corrections for sparsely sampled phylogenies of higher taxonomic groups. In particular,
corrections based on “sampling fractions” can lead to low statistical power to infer rate variation
when it is present, depending on the likelihood function used for inference. In the extreme, the sampling
fraction correction can lead to spurious patterns of diversification that are driven solely by unbalanced
sampling across the tree in concert with low overall power to infer shifts."

@title2019tip do not test possible estimate biases using phylogenies with randomly
resolved politomies. They analyse complete phylogenies and found that estimating species diversification
rates with model free methods and BAMM is good for speciation rate estimates, but less good for net
diversification rate, and bad for extinction rates.

@sun2020estimating used a 50% species level sampled phylogeny of the family Cucurbitaceae to estimate
speciation rates with two parametric methods (RPANDA and BAMM) and one semi-parametric method (DR) and
tested the effect of using incompletely sampled phylogenies on one hand, and incomplete phylogenies that
have been made near complete, using a stochastic politomy resolver based on taxonomy, on the other hand
(with two different methods: 1) adding missing species at random following a backbone taxonomy, and 2)
adding missing species to a genus level taxonomy based on a taxonomy resolver only). When using incomplete
phylogenies, parametric methods tend to overestimate speciation rate as sampling gets worse (less complete),
while semi-parametric methods tend to underestimate speciation rate as sampling gets worse. When using
taxonomically resolved complete phylogenies, only one parametric method overestimates speciation rate the
more incomplete original sampling was. RPANDA, BAMM tip rates and DR all underestimate speciation rates as
original sampling was more incomplete.
"Adding backbone taxa tends to result in the inference of spurious early bursts of diversification", bias in
sampling ancestral nodes, failure to sample recent divergences or pushing them back in relative time (making
recent divergences older than expected)
"failure to detect recent bursts of speciation and instead inferring higher rates of evolution at earlier time
intervals"

### 2023-02-21 review.Rmd
provide a better abstract

@@
 
+Thanks for your comment, the abstract was indeed slightly forsaken on the previous version of the ms.
+We rewrote most of the abstract and checked for spelling.
+We hope you find it a good improvement from the original.
+
 ---
@@
 
-Thanks for pointing that out. We added that topic and a couple of references to the intro, L48-49.
+Thanks for pointing that out. We added that topic to the intro along with a couple of references, L48-49.
 

### 2023-02-21 manuscript/manuscript-1st-submission.Rmd
format titles and subtitles and add edits to cross validation section

@@
 
-# Introduction
+\begin{center}
+Abstract
+\end{center}
+
+  Achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram) is a difficult and time-consuming task.
+But the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
+    Knowledge of the evolutionary time frame of organisms is key for research in the natural sciences.
+It also represent valuable information for education, science communication, and policy decisions.
+  When chronograms are shared in public and open databases, this wealth of expertly-curated and peer-reviewed data on evolutionary timeframe is exposed in a programatic and reusable way, as intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
+  Here we present DateLife, a service implemented as an R package and an R Shiny website application available at www.datelife.org, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time frame of evolution.
+  The main DateLife workflow constructs a chronogram for any given combination of taxon names by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from the TreeBASE database as well.
+  We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
+  DateLife will be useful to increase awareness of the existing variation in alternative hypothesis of evolutionary time for the same organisms, and can foster exploration of the effect of alternative evolutionary timing hypotheses on the results of downstream analyses, providing a framework for a more informed interpretation of evolutionary results.
+
+_Keywords_: Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations
+
+Word count: `r wordcountaddin::word_count()`
+
+\newpage
+
+<!-- # Introduction -->
 <!-- If you're aiming for syst bio you don't want to start so general Readers know what chronograms and phylogenies are!.-->
@@
 -->
-DateLife's core software application is available as an R package [@datelifeR], and as an online Rshiny interactive website at www.datelife.org.
+DateLife's core software application is available as an R package [@sanchez2022], and as an online Rshiny interactive website at www.datelife.org.
 It features key elements for scientific reproducibility, such as a versioned, open and fully public source database [@mctavish2015phylesystem], data stored and available in a computer readable format [@vos2012nexml], automated and programmatic ways of accessing the data [@Stoltzfus2013] and methods to summarize and compare the data.
 
-# Description
+\begin{center}
+\textsc{Description}
+\end{center}
 
 DateLife's core software application consists of the R package `datelife`.
-Its current stable version -- v`r packageVersion("datelife")`, is available from The Comprehensive R Archive Network (CRAN) repository [@datelifeR], and relies on functionalities from various biological R packages:
+Its current stable version -- v`r packageVersion("datelife")`, is available from The Comprehensive R Archive Network (CRAN) repository [@sanchez2022], and relies on functionalities from various biological R packages:
 ape [@paradis2004],
@@
 
-##  Creating a search query
+\begin{center}
+\emph{Creating a Search Query}
+\end{center}
 
@@
 
-## Searching a chronogram database
+\begin{center}
+\emph{Searching a Chronogram Database}
+\end{center}
 
@@
 
-## Summarizing search results
+\begin{center}
+\emph{Summarizing Search Results}
+\end{center}
 
@@
 
-## Choosing a topology
+\begin{center}
+\emph{Choosing a Topology}
+\end{center}
+
 DateLife requires a tree topology to summarize age data upon.
@@
 
-## Dating the topology
+\begin{center}
+\emph{Dating the Topology}
+\end{center}
+
 Input topologies from OpenTree or the supertree approach described above do not include branch length estimates of any kind.
@@
 
-## Visualizing results
+\begin{center}
+\emph{Visualizing Results}
+\end{center}
+
 Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R "phylo" format), as well as visualize and compare results graphically, or construct their own graphs using DateLife's chronogram plot generation functions available from the R package `datelifeplot` [@datelifeplot].
@@
 
-# Benchmark
+\begin{center}
+\textsc{Benchmark}
+\end{center}
 
@@
 
-# Case studies
+\begin{center}
+\textsc{Case Studies}
+\end{center}
 
@@
 
-## A small example
+\begin{center}
+\emph{A Small Example}
+\end{center}
 
-### Creating a search query
-
+**_Creating a search query.--_**
 We chose 6 bird species within the Passeriformes. The sample includes
@@
 
-### Searching the database
-
+**_Searching the database.--_**
 DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies (Fig.
@@
 
-### Summarizing search results
-
+**_Summarizing search results.--_**
 DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and congruified and mapped age data to nodes in this chosen topology (Table \ref{tbl:table1}).
@@
 
-## An example with the family of true finches
+\begin{center}
+\emph{An Example with the Family of True Finches}
+\end{center}
 
-### Creating a query
-
+**_Creating a query.--_**
 To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the "get species from taxon" flag,
@@
 
-### Searching the database
-
+**_Searching the database.--_**
 Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species.
@@
 
-### Summarizing search results
-
+**_Summarizing search results.--_**
 The final step is to congruify and summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM.
@@
 
-# Cross-validation test
+\begin{center}
+\textsc{Cross-Validation Test}
+\end{center}
 
@@
 
-# Discussion
+\begin{center}
+\textsc{Discussion}
+\end{center}
 
@@
 
-## Age variation in source chronograms
+\begin{center}
+\emph{Age Variation in Source Chronograms}
+\end{center}
 
@@
 
-## Primary vs Secondary calibrations
+\begin{center}
+\emph{Primary vs Secondary Calibrations}
+\end{center}
 
@@
 
-## Sumarizing chronograms
+\begin{center}
+\emph{Sumarizing Chronograms}
+\end{center}
 
@@
 
-## Effects on downstream analyses
+\begin{center}
+\emph{Effects on Downstream Analyses}
+\end{center}
 
@@
 
-# Conclusions
+\begin{center}
+\textsc{Conclusions}
+\end{center}
 
@@
 
-# Availability
+\begin{center}
+\textsc{Availability}
+\end{center}
 
@@
 
-# Supplementary Material
+\begin{center}
+\textsc{Supplementary Material}
+\end{center}
 
@@
 
-# Funding
+\begin{center}
+\textsc{Funding}
+\end{center}
 
@@
 
-# Acknowledgements
+\begin{center}
+\textsc{Acknowledgements}
+\end{center}
 

### 2023-02-21 manuscript/manuscript-1st-submission.tex
format titles and subtitles and add edits to cross validation section

@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4959}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 
-\abstract{%
+\begin{document}
+\maketitle
+
+\begin{center}
+Abstract
+\end{center}
+
 Achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram) is a difficult and time-consuming task.
@@
 DateLife will be useful to increase awareness of the existing variation in alternative hypothesis of evolutionary time for the same organisms, and can foster exploration of the effect of alternative evolutionary timing hypotheses on the results of downstream analyses, providing a framework for a more informed interpretation of evolutionary results.
-}
-
-\begin{document}
-\maketitle
-
-\hypertarget{introduction}{%
-\section{Introduction}\label{introduction}}
+
+\emph{Keywords}: Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations
+
+Word count: 5393
+
+\newpage
 
@@
 
-\hypertarget{description}{%
-\section{Description}\label{description}}
+\begin{center}
+\textsc{Description}
+\end{center}
 
@@
 
-\begin{minipage}[t]{17.3cm}
-\includegraphics{../figures/figure-workflow/figure1-new.pdf}
-\captionof{figure}{\textit{Main DateLife workflow.
-Analyses can be performed via DateLife's interactive website at \url{www.datelife.org}, or using the \texttt{datelife} R package.
-Details on the R functions used to perform the analyses are available from \texttt{datelife}'s R package vignettes at \url{https://phylotastic.org/datelife}.}}
-\label{fig:workflow}
-\end{minipage}
-
-\hypertarget{creating-a-search-query}{%
-\subsection{Creating a search query}\label{creating-a-search-query}}
+\begin{center}
+\emph{Creating a Search Query}
+\end{center}
 
@@
 
-\hypertarget{searching-a-chronogram-database}{%
-\subsection{Searching a chronogram database}\label{searching-a-chronogram-database}}
+\begin{center}
+\emph{Searching a Chronogram Database}
+\end{center}
 
@@
 
-\hypertarget{summarizing-search-results}{%
-\subsection{Summarizing search results}\label{summarizing-search-results}}
+\begin{center}
+\emph{Summarizing Search Results}
+\end{center}
 
@@
 
-\emph{\textbf{Choosing a topology.--}}
+\begin{center}
+\emph{Choosing a Topology}
+\end{center}
+
 DateLife requires a tree topology to summarize age data upon.
@@
 
-\emph{\textbf{Dating the topology.--}}
+\begin{center}
+\emph{Dating the Topology}
+\end{center}
+
 Input topologies from OpenTree or the supertree approach described above do not include branch length estimates of any kind.
@@
 
-\emph{\textbf{Visualizing results.--}}
+\begin{center}
+\emph{Visualizing Results}
+\end{center}
+
 Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R \enquote{phylo} format), as well as visualize and compare results graphically, or construct their own graphs using DateLife's chronogram plot generation functions available from the R package \texttt{datelifeplot} (Sanchez-Reyes \& O'Meara, 2022).
@@
 
-\hypertarget{benchmark}{%
-\section{Benchmark}\label{benchmark}}
+\begin{center}
+\textsc{Benchmark}
+\end{center}
 
@@
 
-\includegraphics{../figures/fig_runtime_main.pdf}
-\captionof{figure}{\textit{DateLife's benchmarking results.
-Computation time used to process a query and a search across \texttt{datelife}'s chronogram database, relative to number of input taxon names.
-For each N = \{10, 100, 200, ..., 1 000, ...
-, 9 000, 10 000\}, we sampled N species names from the class Aves a hundred times, and then performed a datelife search processing the input names with Taxon Names Resolution Service (TNRS; light gray), and without processing names (dark gray).
-For comparison, we performed a search using an input that had been pre-processed with TNRS (light blue).}}
-\label{fig:runtime_main}
-
 \texttt{datelife}'s code performance was evaluated with a set of unit tests designed and
@@
 
-\hypertarget{case-studies}{%
-\section{Case studies}\label{case-studies}}
+\begin{center}
+\textsc{Case Studies}
+\end{center}
 
@@
 
-\hypertarget{a-small-example}{%
-\subsection{A small example}\label{a-small-example}}
-
-\hypertarget{creating-a-search-query-1}{%
-\subsubsection{Creating a search query}\label{creating-a-search-query-1}}
-
+\begin{center}
+\emph{A Small Example}
+\end{center}
+
+\textbf{\emph{Creating a search query.--}}
 We chose 6 bird species within the Passeriformes. The sample includes
@@
 
-\hypertarget{searching-the-database}{%
-\subsubsection{Searching the database}\label{searching-the-database}}
-
+\textbf{\emph{Searching the database.--}}
 DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies (Fig.
@@
 
-\hypertarget{summarizing-search-results-1}{%
-\subsubsection{Summarizing search results}\label{summarizing-search-results-1}}
-
+\textbf{\emph{Summarizing search results.--}}
 DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and congruified and mapped age data to nodes in this chosen topology (Table \ref{tbl:table1}).
@@
 \newpage
+
+\newpage
+
+\begin{center}
+\emph{An Example with the Family of True Finches}
+\end{center}
+
+\textbf{\emph{Creating a query.--}}
+To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the \enquote{get species from taxon} flag,
+which gets all recognized species names within a named group from a taxonomy of choice.
+Following the NCBI taxonomy, our DateLife query has 289 Fringillidae species.
+This taxon-constrained approach implies that the final results of a full DateLife analysis will be done using a tree topology and ages for the species in a named group, which do not necessarily correspond to a monophyletic group.
+Users can change this behaviour by providing a monophyletic tree as input for a DateLife search, or as a tree topology for a DateLife summary.
+
+\textbf{\emph{Searching the database.--}}
+Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species.
+The DateLife search identified 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Price et al., 2014).
+Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementray Table S1).
+
+\textbf{\emph{Summarizing search results.--}}
+The final step is to congruify and summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM.
+As explained in the \enquote{Description} section, a tree topology to summarize age data upon is required.
+By default, DateLife uses the topology from OpenTree's synthetic tree that contains the species in the search query to summarize age data upon.
+According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).
+
+\vspace{50mm}
+
+Age data from source chronograms was congruified to OpenTree's topology (Fig.
+\ref{fig:fringillidae-topologies}B), reducing the age data set to 818 different data points (Supplementray Table S2).
+For each congruent node, age summary statistics were calculated and used as fixed secondary calibrations over the chosen tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
+\ref{fig:fringillidages}).
+
+\begin{center}
+\textsc{Cross-Validation Test}
+\end{center}
+
+We performed a cross validation analysis of the DateLife workflow using the Fringillidae chronograms.
+We used the individual tree topologies from each of the 19 source chronograms from 13 studies as inputs, treating their node ages as unknown.
+We then estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.
+We found that node ages from original study, and ages estimated using all other age data available are correlated (Fig. \ref{fig:cvXY}).
+For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka \enquote{closer to the root}) relative to the original estimate, and overestimate ages for nodes closer to the tips.
+Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. S1).
+
+\begin{center}
+\textsc{Discussion}
+\end{center}
+
+DateLife makes state-of-the-art data on evolutionary time frame easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
+It is an open service that does not require any expert biological knowledge from users --besides the names of the species or group they want to work with, for any of its functionality.
+
+A total of 99,474 unique terminal taxa are represented in DateLife's database.
+Incorporation of more chronograms into the database will continue to improve DateLife's services.
+One option to increase the number of chronograms in the DateLife database is the Dryad data repository.
+Methods to automatically mine chronograms from Dryad could be designed and implemented.
+However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined manually by a human curator.
+We would like to emphasize on the importance of sharing chronogram data, including systematically curated metadata, into open repositories, such as OpenTree's Phylesystem (McTavish et al., 2015) for the benefit of the scientific community as a whole.
+
+\begin{center}
+\emph{Age Variation in Source Chronograms}
+\end{center}
+
+Conflict in estimated ages among alternative studies is common in the literature.
+See, for example, the robust ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández, 2015; Ramshaw et al., 1972; Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021).
+Source chronograms available for the same organisms have potentially been estimated implementing calibrations very differently.
+For example, the chronograms from Burns et al.
+(2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older age estimates for the same nodes than chronograms that were inferred using fossils as calibrations (Figs.
+\ref{fig:fringillidages}, \ref{fig:cvXY}; Supplementary Figs.
+S1, S5).
+
+Different calibration implementations might also imply fundamentally distinct evolutionary hypotheses (Antonelli et al., 2017).
+For example, two independent researchers working on the same clade should both carefully select and justify their choices of fossil calibration placement.
+Yet, if one researcher concludes that a fossil should calibrate the ingroup of a clade, while another researcher concludes that the same fossil should calibrate the outgroup of the clade, the resulting age estimates will differ, as the placement of calibrations as stem or crown group has been proven to significantly affect time of lineage divergence estimates (Sauquet, 2013).
+
+\begin{center}
+\emph{Primary vs Secondary Calibrations}
+\end{center}
+
+While most chronograms in DateLife's database are constructed using primary calibrations (molecular substitution rates or ages obtained from the fossil record or geological events), DateLife summarizes chronograms using secondary calibrations (ages coming from other chronograms).
+Graur and Martin (2004) cautioned on the increased error and uncertainty in estimated ages when using secondary calibrations in dating analyses.
+Schenk (2016) showed that, in simulations, divergence times inferred using secondary calibrations are significantly younger than those inferred with primary calibrations, when obtained with Bayesian inference methods, and when priors are implemented in similar ways in both analyses.
+Accordingly, the scientific community seems to have more confidence in chronograms obtained from a single analysis, using fossil data as primary sources of calibrations (Schenk, 2016), and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data reflect a coherent evolutionary history (Sauquet, 2013), as for example done by Antonelli et al.
+(2017).
+There have been attempts to create fossil calibration databases (Ksepka et al., 2015), though these still have room to grow.
+
+It seems that using several (as opposed to just a few) secondary calibrations can provide sufficient information to alleviate or even neutralize potential biases (Sauquet, 2013).
+Certainly, further studies are required to fully understand the effect of secondary calibrations on outputs from different tree dating methods, and on downstream analyses.
+It is possible that secondary calibrations can be safely used with dating methods that do not require setting priors, such as penalized likelihood (Sanderson, 2003), with methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ (Webb et al., 2008; Webb \& Donoghue, 2005), or methods that summarize age data unto a tree topology.
+
+Our cross validation analysis might provide some insight in this regard.
+When ages are estimated with secondary calibrations, nodes closer to the root do tend to be slightly younger than ages estimated with primary calibrations.
+However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures S2-S20).
+The only excpetion to tjis was observed on results of the cross validation analysis of the Burns et al.
+(2014) chronogram, which results in much younger node ages when estimated using secondary calibrations (Supplementary Figs.
+S1, S5).
+
+\begin{center}
+\emph{Sumarizing Chronograms}
+\end{center}
+
+By default, DateLife currently summarizes all source chronograms that overlap with at least two species names. Users can exclude source chronograms if they have reasons to do so.
+Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely.
+To our knowledge, there are no tested measures to determine independently when a chronogram is better than another.
+Yet, several characteristics of the data used for dating analyses, as well as from the output chronogram itself, could be used to score the quality of source chronograms.
+
+Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used (Magallón, 2010; Magallón et al., 2015).
+Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
+
+DateLife provides an opportunity to capture concordance and conflict among date estimates, which can also be used as a metric for chronogram reliability.
+Its open database of chronograms allows other researchers to do such analyses themselves reproducibly, and without needing permission.
+Though, of course, they should follow proper citation practices, especially for the source chronogram studies.
+
+The exercise of summarizing age data from across multiple studies provides the opportunity to work with a more inclusive chronogram,
+that reflects a unified evolutionary history for a lineage, by putting together evidence from different hypotheses.
+The largest, and taxonomically broadest chronogram currently available from OpenTree was constructed summarizing age data from 2,274 published chronograms using NCBI's taxonomic tree as backbone (Hedges et al., 2015).
+A summarizing exercise may also amplify the effect of uncertainty and errors in source data, and blur parts of the evolutionary history of a lineage that might only be reflected in source chronograms and lost on the summary chronogram (Sauquet et al., 2021).
+
+\begin{center}
+\emph{Effects on Downstream Analyses}
+\end{center}
+
+For downstream analyses, using alternative chronogram may deeply affect our inferences (Title \& Rabosky, 2016), particularly when studying phenomena dependent on the timing of species diversification events, such as macroevolutionary processes.
+
+In ecology and conservation biology, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance (Webb et al., 2008).
+Hence, DateLife's workflow features different ways of estimating node ages in the absence of calibrations and branch length information for certain taxa.
+\enquote{Making up} branch lengths is a common practice in scientific publications: Jetz et al.
+(2012), created a chronogram of all 9, 993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
+(2018) created a chronogram of 31, 536 ray-finned fishes, of which only 37\% had molecular data; Smith and Brown (2018) constructed a chronogram of 353, 185 seed plants where only 23\% had molecular data.
+
+Notably, risks come with this practice.
+Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration, using polytomy resolution and branch estimation methods.
+There has yet to be a thorough analysis of what can go wrong when one extends inferences beyond the data in this way, so we urge caution; we also urge readers to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, and assessing the consistency of the results with prior work.
+
+\begin{center}
+\textsc{Conclusions}
+\end{center}
+
+Knowledge of the evolutionary time frame of organisms is key to many research areas: trait evolution,
+species diversification, biogeography, macroecology and more. It is also crucial for education, science communication and policy, but generating chronograms is difficult,
+especially for those who want to use phylogenies but who are not systematists, or
+do not have the time to acquire and develop the necessary knowledge and skills to construct them on their own.
+Importantly, years of primarily public funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
+
+The DateLife project allows for easy and fast summary of public and state-of-the-art data on time of lineage divergence.
+It provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identification of regions that require more research, or that have conflicting information.
+It is available as an R package, and as a web-based R shiny application at www.datelife.org
+Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
+The DateLife project helps with awareness of the existing variation in expert time of divergence data, and will foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
+
+\begin{center}
+\textsc{Availability}
+\end{center}
+
+The DateLife software is free and open source and it can be used through its R shiny web application at \url{http://www.datelife.org}, through the \texttt{datelife} R package, and through Phylotastic's project web portal \url{https://phylo.cs.nmsu.edu/}.
+DateLife's web application is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker and OpenTree's infrastructure (dates.opentreeoflife.org/datelife).
+\texttt{datelife}'s R package stable version is available for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife}) using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R.
+Development versions are available from the GitHub repository (\url{https://github.com/phylotastic/datelife}) and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.
+
+\begin{center}
+\textsc{Supplementary Material}
+\end{center}
+
+Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities are available at \href{https://github.com/LunaSare/datelifeMS1}{datelifeMS1}, \href{https://github.com/LunaSare/datelife_examples}{datelife\_examples}, and \href{https://github.com/LunaSare/datelife_benchmark}{datelife\_benchmark} repositories in LLSR's GitHub account.
+
+\begin{center}
+\textsc{Funding}
+\end{center}
+
+Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to the Datelife project; DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), ABI-1458572 to the Phylotastic project, and ABI-1759846 to the Open Tree of Life project.
+
+\begin{center}
+\textsc{Acknowledgements}
+\end{center}
+
+The DateLife project was born as a prototype tool aiming to provide these services, and was initially developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA (Stoltzfus et al., 2013).
+We thank colleagues from the O'Meara Lab at the University
+of Tennesse Knoxville for suggestions, discussions and software testing.
+The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
+that led to initial work on this project.
+The team that assembled DateLife's first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
+The Open Tree of Life project that provides the open, metadata rich repository of
+trees used to construct DateLife's chronogram database.
+The many scientists who publish their chronograms in an open, reusable form, and the scientists who curate them for deposition in the Open Tree of Life repository.
+The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.
+
+\newpage
+
+\hypertarget{references}{%
+\section{References}\label{references}}
+
+\begingroup
+\setlength{\parindent}{-0.5in}
+\setlength{\leftskip}{0.5in}
+
+\hypertarget{refs}{}
+\leavevmode\hypertarget{ref-alstrom2014discovery}{}%
+Alström, P., Hooper, D.
+M., Liu, Y., Olsson, U., Mohan, D., Gelang, M., \ldots{} Price, T.
+D.
+(2014).
+Discovery of a relict lineage and monotypic family of passerine birds.
+\emph{Biology Letters}, \emph{10}(3), 20131067.
+
+\leavevmode\hypertarget{ref-ane2009groves}{}%
+Ané, C., Eulenstein, O., Piaggio-Talice, R., \& Sanderson, M. J. (2009). Groves of phylogenetic trees. \emph{Annals of Combinatorics}, \emph{13}(2), 139--167.
+
+\leavevmode\hypertarget{ref-antonelli2017supersmart}{}%
+Antonelli, A., Hettling, H., Condamine, F.
+L., Vos, K., Nilsson, R.
+H., Sanderson, M.
+J., \ldots{} Vos, R.
+A.
+(2017).
+Toward a self-updating platform for estimating rates of speciation and migration, ages, and relationships of Taxa.
+\emph{Systematic Biology}, \emph{66}(2), 153--166.
+\url{https://doi.org/10.1093/sysbio/syw066}
+
+\leavevmode\hypertarget{ref-archie1986newick}{}%
+Archie, J., Day, W.
+H., Felsenstein, J., Maddison, W., Meacham, C., Rohlf, F.
+J., \& Swofford, D.
+(1986).
+The Newick tree format.
+Retrieved from \href{\%7Bhttps://evolution.genetics.washington.edu/phylip/newicktree.html\%7D}{\{https://evolution.genetics.washington.edu/phylip/newicktree.html\}}
+
+\leavevmode\hypertarget{ref-avibase-emberiza}{}%
+Avibase.
+(2022).
+Yellow-throated Bunting.
+\emph{Avibase - the World Bird Database}, (Online Resource).
+Retrieved from \href{\%7Bhttps://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\%7D}{\{https://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\}}
+
+\leavevmode\hypertarget{ref-Bapst2012a}{}%
+Bapst, D.
+W.
+(2012).
+Paleotree: An R package for paleontological and phylogenetic analyses of evolution.
+\emph{Methods in Ecology and Evolution}, \emph{3}(5), 803--807.
+\url{https://doi.org/10.1111/j.2041-210X.2012.00223.x}
+
+\leavevmode\hypertarget{ref-barba2018constraining}{}%
+Barba-Montoya, J., Reis, M.
+dos, Schneider, H., Donoghue, P.
+C., \& Yang, Z.
+(2018).
+Constraining uncertainty in the timescale of angiosperm evolution and the veracity of a cretaceous terrestrial revolution.
+\emph{New Phytologist}, \emph{218}(2), 819--834.
+
+\leavevmode\hypertarget{ref-barker2014mitogenomic}{}%
+Barker, F. K. (2014). Mitogenomic data resolve basal relationships among passeriform and passeridan birds. \emph{Molecular Phylogenetics and Evolution}, \emph{79}, 313--324.
+
+\leavevmode\hypertarget{ref-barker2013going}{}%
+Barker, F.
+K., Burns, K.
+J., Klicka, J., Lanyon, S.
+M., \& Lovette, I.
+J.
+(2013).
+Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds.
+\emph{Systematic Biology}, \emph{62}(2), 298--320.
+
+\leavevmode\hypertarget{ref-barker2015new}{}%
+Barker, F.
+K., Burns, K.
+J., Klicka, J., Lanyon, S.
+M., \& Lovette, I.
+J.
+(2015).
+New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
+\emph{The Auk: Ornithological Advances}, \emph{132}(2), 333--348.
+
+\leavevmode\hypertarget{ref-barker2004phylogeny}{}%
+Barker, F.
+K., Cibois, A., Schikler, P., Feinstein, J., \& Cracraft, J.
+(2004).
+Phylogeny and diversification of the largest avian radiation.
+\emph{Proceedings of the National Academy of Sciences}, \emph{101}(30), 11040--11045.
+
+\leavevmode\hypertarget{ref-beresford2005african}{}%
+Beresford, P., Barker, F., Ryan, P., \& Crowe, T.
+(2005).
+African endemics span the tree of songbirds (passeri): Molecular systematics of several evolutionary ``enigmas''.
+\emph{Proceedings of the Royal Society B: Biological Sciences}, \emph{272}(1565), 849--858.
+
+\leavevmode\hypertarget{ref-Boyle2013}{}%
+Boyle, B., Hopkins, N., Lu, Z., Raygoza Garay, J.
+A., Mozzherin, D., Rees, T., \ldots{} Enquist, B.
+J.
+(2013).
+The taxonomic name resolution service: An online tool for automated standardization of plant names.
+\emph{BMC Bioinformatics}, \emph{14}(1).
+\url{https://doi.org/10.1186/1471-2105-14-16}
+
+\leavevmode\hypertarget{ref-Britton2007}{}%
+Britton, T., Anderson, C.
+L., Jacquet, D., Lundqvist, S., \& Bremer, K.
+(2007).
+Estimating Divergence Times in Large Phylogenetic Trees.
+\emph{Systematic Biology}, \emph{56}(788777878), 741--752.
+\url{https://doi.org/10.1080/10635150701613783}
+
+\leavevmode\hypertarget{ref-bryson2014diversification}{}%
+Bryson Jr, R.
+W., Chaves, J., Smith, B.
+T., Miller, M.
+J., Winker, K., Pérez-Emán, J.
+L., \& Klicka, J.
+(2014).
+Diversification across the new world within the `blue'cardinalids (aves: Cardinalidae).
+\emph{Journal of Biogeography}, \emph{41}(3), 587--599.
+
+\leavevmode\hypertarget{ref-burleigh2015building}{}%
+Burleigh, J. G., Kimball, R. T., \& Braun, E. L. (2015). Building the avian tree of life using a large-scale, sparse supermatrix. \emph{Molecular Phylogenetics and Evolution}, \emph{84}, 53--63.
+
+\leavevmode\hypertarget{ref-burns2014phylogenetics}{}%
+Burns, K.
+J., Shultz, A.
+J., Title, P.
+O., Mason, N.
+A., Barker, F.
+K., Klicka, J., \ldots{} Lovette, I.
+J.
+(2014).
+Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds.
+\emph{Molecular Phylogenetics and Evolution}, \emph{75}, 41--77.
+
+\leavevmode\hypertarget{ref-Chamberlain2018}{}%
+Chamberlain, S. (2018). \emph{bold: Interface to Bold Systems API}. Retrieved from \url{https://CRAN.R-project.org/package=bold}
+
+\leavevmode\hypertarget{ref-Chamberlain2013}{}%
+Chamberlain, S.
+A., \& Szöcs, E.
+(2013).
+taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}.
+\emph{F1000Research}, \emph{2}(191), 1--29.
+\url{https://doi.org/10.12688/f1000research.2-191.v2}
+
+\leavevmode\hypertarget{ref-chaves2013biogeography}{}%
+Chaves, J.
+A., Hidalgo, J.
+R., \& Klicka, J.
+(2013).
+Biogeography and evolutionary history of the n eotropical genus s altator (a ves: T hraupini).
+\emph{Journal of Biogeography}, \emph{40}(11), 2180--2190.
+
+\leavevmode\hypertarget{ref-claramunt2015new}{}%
+Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{Science Advances}, \emph{1}(11), e1501005.
+
+\leavevmode\hypertarget{ref-Criscuolo2006}{}%
+Criscuolo, A., Berry, V., Douzery, E.
+J., \& Gascuel, O.
+(2006).
+SDM: A fast distance-based approach for (super)tree building in phylogenomics.
+\emph{Systematic Biology}, \emph{55}(5), 740--755.
+\url{https://doi.org/10.1080/10635150600969872}
+
+\leavevmode\hypertarget{ref-delsuc2018phylogenomic}{}%
+Delsuc, F., Philippe, H., Tsagkogeorga, G., Simion, P., Tilak, M.-K., Turon, X., \ldots{} Douzery, E.
+J.
+(2018).
+A phylogenomic framework and timescale for comparative studies of tunicates.
+\emph{BMC Biology}, \emph{16}(1), 1--14.
+
+\leavevmode\hypertarget{ref-Eastman2013}{}%
+Eastman, J.
+M., Harmon, L.
+J., \& Tank, D.
+C.
+(2013).
+Congruification: Support for time scaling large phylogenetic trees.
+\emph{Methods in Ecology and Evolution}, \emph{4}(7), 688--691.
+\url{https://doi.org/10.1111/2041-210X.12051}
+
+\leavevmode\hypertarget{ref-edgar2004muscle}{}%
+Edgar, R. C. (2004). MUSCLE: Multiple sequence alignment with high accuracy and high throughput. \emph{Nucleic Acids Research}, \emph{32}(5), 1792--1797.
+
+\leavevmode\hypertarget{ref-Felsenstein1985a}{}%
+Felsenstein, J. (1985). Phylogenies and the Comparative Method. \emph{The American Naturalist}, \emph{125}(1), 1--15. Retrieved from \url{http://www.jstor.org/stable/2461605}
+
+\leavevmode\hypertarget{ref-gbif2022taxonomy}{}%
+GBIF Secretariat.
+(2022).
+GBIF Backbone Taxonomy.
+\emph{Checklist dataset}, (Online Resource accessed via GBIF.org).
+Retrieved from \href{\%7Bhttps://doi.org/10.15468/39omei\%20\%7D}{\{https://doi.org/10.15468/39omei \}}
+
+\leavevmode\hypertarget{ref-gibb2015new}{}%
+Gibb, G.
+C., England, R., Hartig, G., McLenachan, P.
+A., Taylor Smith, B.
+L., McComish, B.
+J., \ldots{} Penny, D.
+(2015).
+New zealand passerines help clarify the diversification of major songbird lineages during the oligocene.
+\emph{Genome Biology and Evolution}, \emph{7}(11), 2983--2995.
+
+\leavevmode\hypertarget{ref-graur2004reading}{}%
+Graur, D., \& Martin, W. (2004). Reading the entrails of chickens: Molecular timescales of evolution and the illusion of precision. \emph{TRENDS in Genetics}, \emph{20}(2), 80--86.
+
+\leavevmode\hypertarget{ref-hackett2008phylogenomic}{}%
+Hackett, S.
+J., Kimball, R.
+T., Reddy, S., Bowie, R.
+C., Braun, E.
+L., Braun, M.
+J., \ldots{} others.
+(2008).
+A phylogenomic study of birds reveals their evolutionary history.
+\emph{Science}, \emph{320}(5884), 1763--1768.
+
+\leavevmode\hypertarget{ref-Hedges2015}{}%
+Hedges, S.
+B., Marin, J., Suleski, M., Paymer, M., \& Kumar, S.
+(2015).
+Tree of life reveals clock-like speciation and diversification.
+\emph{Molecular Biology and Evolution}, \emph{32}(4), 835--845.
+\url{https://doi.org/10.1093/molbev/msv037}
+
+\leavevmode\hypertarget{ref-Heibl2008}{}%
+Heibl, C. (2008). \emph{PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages.} Retrieved from \url{http://www.christophheibl.de/Rpackages.html}
+
+\leavevmode\hypertarget{ref-hooper2017chromosomal}{}%
+Hooper, D. M., \& Price, T. D. (2017). Chromosomal inversion differences correlate with range overlap in passerine birds. \emph{Nature Ecology \& Evolution}, \emph{1}(10), 1526.
+
+\leavevmode\hypertarget{ref-Huelsenbeck2001}{}%
+Huelsenbeck, J. P., \& Ronquist, F. (2001). MRBAYES: Bayesian inference of phylogenetic trees. \emph{Bioinformatics}, \emph{17}(8), 754--755. \url{https://doi.org/10.1093/bioinformatics/17.8.754}
+
+\leavevmode\hypertarget{ref-Jetz2012}{}%
+Jetz, W., Thomas, G., Joy, J.
+J., Hartmann, K., \& Mooers, A.
+(2012).
+The global diversity of birds in space and time.
+\emph{Nature}, \emph{491}(7424), 444--448.
+\url{https://doi.org/10.1038/nature11631}
+
+\leavevmode\hypertarget{ref-johansson2008phylogenetic}{}%
+Johansson, U.
+S., Fjeldså, J., \& Bowie, R.
+C.
+(2008).
+Phylogenetic relationships within passerida (aves: Passeriformes): A review and a new molecular phylogeny based on three nuclear intron markers.
+\emph{Molecular Phylogenetics and Evolution}, \emph{48}(3), 858--876.
+
+\leavevmode\hypertarget{ref-katoh2009multiple}{}%
+Katoh, K., Asimenos, G., \& Toh, H. (2009). Multiple alignment of dna sequences with mafft. In \emph{Bioinformatics for dna sequence analysis} (pp. 39--64). Springer.
+
+\leavevmode\hypertarget{ref-kimball2019phylogenomic}{}%
+Kimball, R. T., Oliveros, C. H., Wang, N., White, N. D., Barker, F. K., Field, D. J., \ldots{} others. (2019). A phylogenomic supertree of birds. \emph{Diversity}, \emph{11}(7), 109.
+
+\leavevmode\hypertarget{ref-klicka2014comprehensive}{}%
+Klicka, J., Barker, F.
+K., Burns, K.
+J., Lanyon, S.
+M., Lovette, I.
+J., Chaves, J.
+A., \& Bryson Jr, R.
+W.
+(2014).
+A comprehensive multilocus assessment of sparrow (aves: Passerellidae) relationships.
+\emph{Molecular Phylogenetics and Evolution}, \emph{77}, 177--182.
+
+\leavevmode\hypertarget{ref-ksepka2015fossil}{}%
+Ksepka, D.
+T., Parham, J.
+F., Allman, J.
+F., Benton, M.
+J., Carrano, M.
+T., Cranston, K.
+A., \ldots{} others.
+(2015).
+The fossil calibration database---a new resource for divergence dating.
+\emph{Systematic Biology}, \emph{64}(5), 853--859.
+
+\leavevmode\hypertarget{ref-lamichhaney2015evolution}{}%
+Lamichhaney, S., Berglund, J., Almén, M.
+S., Maqbool, K., Grabherr, M., Martinez-Barrio, A., \ldots{} others.
+(2015).
+Evolution of darwin's finches and their beaks revealed by genome sequencing.
+\emph{Nature}, \emph{518}(7539), 371--375.
+
+\leavevmode\hypertarget{ref-laubichler2009form}{}%
+Laubichler, M. D., \& Maienschein, J. (2009). \emph{Form and function in developmental evolution}. Cambridge University Press.
+
+\leavevmode\hypertarget{ref-lepage2004avibase}{}%
+Lepage, D. (2004). \emph{Avibase: The world bird database}. Bird Studies Canada.
+
+\leavevmode\hypertarget{ref-lepage2014avibase}{}%
+Lepage, D., Vaidya, G., \& Guralnick, R. (2014). Avibase--a database system for managing and organizing taxonomic concepts. \emph{ZooKeys}, (420), 117.
+
+\leavevmode\hypertarget{ref-lerner2011multilocus}{}%
+Lerner, H.
+R., Meyer, M., James, H.
+F., Hofreiter, M., \& Fleischer, R.
+C.
+(2011).
+Multilocus resolution of phylogeny and timescale in the extant adaptive radiation of hawaiian honeycreepers.
+\emph{Current Biology}, \emph{21}(21), 1838--1844.
+
+\leavevmode\hypertarget{ref-lovette2010comprehensive}{}%
+Lovette, I.
+J., Pérez-Emán, J.
+L., Sullivan, J.
+P., Banks, R.
+C., Fiorentino, I., Córdoba-Córdoba, S., \ldots{} others.
+(2010).
+A comprehensive multilocus phylogeny for the wood-warblers and a revised classification of the parulidae (aves).
+\emph{Molecular Phylogenetics and Evolution}, \emph{57}(2), 753--770.
+
+\leavevmode\hypertarget{ref-magallon2001absolute}{}%
+Magallon, S., \& Sanderson, M. (2001). Absolute diversification rates in angiosperm clades. \emph{Evolution}, \emph{55}(9), 1762--1780.
+
+\leavevmode\hypertarget{ref-magallon2010using}{}%
+Magallón, S. (2010). Using fossils to break long branches in molecular dating: A comparison of relaxed clocks applied to the origin of angiosperms. \emph{Systematic Biology}, \emph{59}(4), 384--399.
+
+\leavevmode\hypertarget{ref-magallon2015metacalibrated}{}%
+Magallón, S., Gómez-Acevedo, S., Sánchez-Reyes, L.
+L., \& Hernández-Hernández, T.
+(2015).
+A metacalibrated time-tree documents the early rise of flowering plant phylogenetic diversity.
+\emph{New Phytologist}, \emph{207}(2), 437--453.
+
+\leavevmode\hypertarget{ref-mctavish2015phylesystem}{}%
+McTavish, E.
+J., Hinchliff, C.
+E., Allman, J.
+F., Brown, J.
+W., Cranston, K.
+A., Holder, M.
+T., \ldots{} Smith, S.
+(2015).
+Phylesystem: A git-based data store for community-curated phylogenetic estimates.
+\emph{Bioinformatics}, \emph{31}(17), 2794--2800.
+
+\leavevmode\hypertarget{ref-Michonneau2016}{}%
+Michonneau, F., Brown, J.
+W., \& Winter, D.
+J.
+(2016).
+rotl: an R package to interact with the Open Tree of Life data.
+\emph{Methods in Ecology and Evolution}, \emph{7}(12), 1476--1481.
+\url{https://doi.org/10.1111/2041-210X.12593}
+
+\leavevmode\hypertarget{ref-Morlon2014}{}%
+Morlon, H. (2014). Phylogenetic approaches for studying diversification. \emph{Ecology Letters}, \emph{17}(4), 508--525. \url{https://doi.org/10.1111/ele.12251}
+
+\leavevmode\hypertarget{ref-moyle2016tectonic}{}%
+Moyle, R.
+G., Oliveros, C.
+H., Andersen, M.
+J., Hosner, P.
+A., Benz, B.
+W., Manthey, J.
+D., \ldots{} Faircloth, B.
+C.
+(2016).
+Tectonic collision and uplift of wallacea triggered the global songbird radiation.
+\emph{Nature Communications}, \emph{7}(1), 1--7.
+
+\leavevmode\hypertarget{ref-oliveros2019earth}{}%
+Oliveros, C.
+H., Field, D.
+J., Ksepka, D.
+T., Barker, F.
+K., Aleixo, A., Andersen, M.
+J., \ldots{} others.
+(2019).
+Earth history and the passerine superradiation.
+\emph{Proceedings of the National Academy of Sciences}, \emph{116}(16), 7916--7925.
+
+\leavevmode\hypertarget{ref-Ooms2018}{}%
+Ooms, J., \& Chamberlain, S. (2018). \emph{Phylocomr: Interface to 'phylocom'}. Retrieved from \url{https://CRAN.R-project.org/package=phylocomr}
+
+\leavevmode\hypertarget{ref-opentreeAPIs}{}%
+Open Tree Of Life, Redelings, B., Cranston, K.
+A., Allman, J., Holder, M.
+T., \& McTavish, E.
+J.
+(2016).
+Open Tree of Life APIs v3.0.
+\emph{Open Tree of Life Project}, (Online Resources).
+Retrieved from \href{\%7Bhttps://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\%7D}{\{https://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\}}
+
+\leavevmode\hypertarget{ref-opentreeoflife2019synth}{}%
+Open Tree Of Life, Redelings, B., Sánchez Reyes, L.
+L., Cranston, K.
+A., Allman, J., Holder, M.
+T., \& McTavish, E.
+J.
+(2019).
+Open tree of life synthetic tree v12.3.
+\emph{Zenodo}.
+Retrieved from \url{https://doi.org/10.5281/zenodo.3937742}
+
+\leavevmode\hypertarget{ref-odeen2011evolution}{}%
+Ödeen, A., Håstad, O., \& Alström, P. (2011). Evolution of ultraviolet vision in the largest avian radiation-the passerines. \emph{BMC Evolutionary Biology}, \emph{11}(1), 1--8.
+
+\leavevmode\hypertarget{ref-paradis2004}{}%
+Paradis, E., Claude, J., \& Strimmer, K. (2004). APE: analyses of phylogenetics and evolution in R language. \emph{Bioinformatics}, \emph{20}(2), 289--290.
+
+\leavevmode\hypertarget{ref-parchman2007coevolution}{}%
+Parchman, T.
+L., Benkman, C.
+W., \& Mezquida, E.
+T.
+(2007).
+Coevolution between hispaniolan crossbills and pine: Does more time allow for greater phenotypic escalation at lower latitude?
+\emph{Evolution}, \emph{61}(9), 2142--2153.
+
+\leavevmode\hypertarget{ref-packert2012horizontal}{}%
+Päckert, M., Martens, J., Sun, Y.-H., Severinghaus, L.
+L., Nazarenko, A.
+A., Ting, J., \ldots{} Tietze, D.
+T.
+(2012).
+Horizontal and elevational phylogeographic patterns of himalayan and southeast asian forest passerines (aves: Passeriformes).
+\emph{Journal of Biogeography}, \emph{39}(3), 556--573.
+
+\leavevmode\hypertarget{ref-pennell2014geiger}{}%
+Pennell, M.
+W., Eastman, J.
+M., Slater, G.
+J., Brown, J.
+W., Uyeda, J.
+C., FitzJohn, R.
+G., \ldots{} Harmon, L.
+J.
+(2014).
+Geiger v2.
+0: An expanded suite of methods for fitting macroevolutionary models to phylogenetic trees.
+\emph{Bioinformatics}, \emph{30}(15), 2216--2218.
+
+\leavevmode\hypertarget{ref-posadas2006historical}{}%
+Posadas, P., Crisci, J. V., \& Katinas, L. (2006). Historical biogeography: A review of its basic concepts and critical issues. \emph{Journal of Arid Environments}, \emph{66}(3), 389--403.
+
+\leavevmode\hypertarget{ref-powell2014comprehensive}{}%
+Powell, A.
+F., Barker, F.
+K., Lanyon, S.
+M., Burns, K.
+J., Klicka, J., \& Lovette, I.
+J.
+(2014).
+A comprehensive species-level molecular phylogeny of the new world blackbirds (icteridae).
+\emph{Molecular Phylogenetics and Evolution}, \emph{71}, 94--112.
+
+\leavevmode\hypertarget{ref-price2014niche}{}%
+Price, T.
+D., Hooper, D.
+M., Buchanan, C.
+D., Johansson, U.
+S., Tietze, D.
+T., Alström, P., \ldots{} others.
+(2014).
+Niche filling slows the diversification of himalayan songbirds.
+\emph{Nature}, \emph{509}(7499), 222.
+
+\leavevmode\hypertarget{ref-pulgarin2013multilocus}{}%
+Pulgarín-R, P.
+C., Smith, B.
+T., Bryson Jr, R.
+W., Spellman, G.
+M., \& Klicka, J.
+(2013).
+Multilocus phylogeny and biogeography of the new world pheucticus grosbeaks (aves: Cardinalidae).
+\emph{Molecular Phylogenetics and Evolution}, \emph{69}(3), 1222--1227.
+
+\leavevmode\hypertarget{ref-rabosky2018inverse}{}%
+Rabosky, D.
+L., Chang, J., Title, P.
+O., Cowman, P.
+F., Sallan, L., Friedman, M., \ldots{} others.
+(2018).
+An inverse latitudinal gradient in speciation rate for marine fishes.
+\emph{Nature}, \emph{559}(7714), 392.
+
+\leavevmode\hypertarget{ref-ramshaw1972time}{}%
+Ramshaw, J., Richardson, D., Meatyard, B., Brown, R., Richardson, M., Thompson, E., \& Boulter, D.
+(1972).
+The time of origin of the flowering plants determined by using amino acid sequence data of cytochrome c.
+\emph{New Phytologist}, \emph{71}(5), 773--779.
+
+\leavevmode\hypertarget{ref-ratnasingham2007bold}{}%
+Ratnasingham, S., \& Hebert, P. D. (2007). BOLD: The barcode of life data system (http://www. Barcodinglife. Org). \emph{Molecular Ecology Notes}, \emph{7}(3), 355--364.
+
+\leavevmode\hypertarget{ref-RCoreTeam2018}{}%
+R Core Team. (2018). \emph{R: a language and environment for statistical computing}. Vienna, Austria: R Foundation for Statistical Computing.
+
+\leavevmode\hypertarget{ref-rees2017automated}{}%
+Rees, \& Cranston, K. (2017). Automated assembly of a reference taxonomy for phylogenetic data synthesis. \emph{Biodiversity Data Journal}, (5).
+
+\leavevmode\hypertarget{ref-rees2017irmng}{}%
+Rees, Vandepitte, L., Decock, W., \& Vanhoorne, B. (2017). IRMNG 2006--2016: 10 Years of a Global Taxonomic Database. \emph{Biodiversity Informatics}, \emph{12}.
+
+\leavevmode\hypertarget{ref-Revell2012}{}%
+Revell, L. J. (2012). Phytools: An r package for phylogenetic comparative biology (and other things). \emph{Methods in Ecology and Evolution}, \emph{3}, 217--223.
+
+\leavevmode\hypertarget{ref-Ronquist2003}{}%
+Ronquist, F., \& Huelsenbeck, J.
+P.
+(2003).
+MrBayes 3: Bayesian phylogenetic inference under mixed models.
+\emph{Bioinformatics}, \emph{19}(12), 1572--1574.
+\url{https://doi.org/10.1093/bioinformatics/btg180}
+
+\leavevmode\hypertarget{ref-datelifeplot}{}%
+Sanchez-Reyes, L.
+L., \& O'Meara, B.
+(2022).
+\texttt{datelifeplot}: Methods to plot chronograms and outputs of the \texttt{datelife} package.
+\emph{R Package Release V0.2.2}.
+Retrieved from \url{https://zenodo.org/badge/latestdoi/381501451}
+
+\leavevmode\hypertarget{ref-sanchez2022}{}%
+Sanchez-Reyes, L.
+L., O'Meara, B., Eastman, J., Heath, T., Wright, A., Schliep, K., \ldots{} Alfaro, M.
+(2022).
+\texttt{datelife}: Scientific Data on Time of Lineage Divergence for Your Taxa.
+In \emph{R package version 0.6.6}.
+Retrieved from \href{https://CRAN.R-project.org/package=datelife\%20and\%20https://doi.org/10.5281/zenodo.593938}{https://CRAN.R-project.org/package=datelife and https://doi.org/10.5281/zenodo.593938}
+
+\leavevmode\hypertarget{ref-sanderson2002}{}%
+Sanderson, M.
+(2002).
+Estimating Absolute Rates of Molecular Evolution and Divergence Times: A Penalized Likelihood Approach.
+\emph{Molecular Biology and Evolution}, \emph{19}(1), 101--109.
+\url{https://doi.org/10.1093/oxfordjournals.molbev.a003974}
+
+\leavevmode\hypertarget{ref-sanderson2003r8s}{}%
+Sanderson, M. (2003). r8s: Inferring Absolute Rates of Molecular Evolution and Divergence Times in the Absence of a Molecular Clock. \emph{Bioinformatics}, \emph{19}(2), 301--302.
+
+\leavevmode\hypertarget{ref-sanderson2001sources}{}%
+Sanderson, M., \& Doyle, J.
+(2001).
+Sources of error and confidence intervals in estimating the age of angiosperms from rbcL and 18S rDNA data.
+\emph{American Journal of Botany}, \emph{88}(8), 1499--1516.
+
+\leavevmode\hypertarget{ref-sauquet2013practical}{}%
+Sauquet, H. (2013). A practical guide to molecular dating. \emph{Comptes Rendus Palevol}, \emph{12}(6), 355--367.
+
+\leavevmode\hypertarget{ref-sauquet2021age}{}%
+Sauquet, H., Ramírez-Barahona, S., \& Magallón, S. (2021). \emph{The age of flowering plants is unknown}.
+
+\leavevmode\hypertarget{ref-schenk2016sec}{}%
+Schenk, J. J. (2016). Consequences of secondary calibrations on divergence time estimates. \emph{PLoS ONE}, \emph{11}(1). \url{https://doi.org/10.1371/journal.pone.0148228}
+
+\leavevmode\hypertarget{ref-schliep2011phangorn}{}%
+Schliep, K. P. (2011). Phangorn: Phylogenetic analysis in r. \emph{Bioinformatics}, \emph{27}(4), 592--593.
+
+\leavevmode\hypertarget{ref-schoch2020ncbi}{}%
+Schoch, C.
+L., Ciufo, S., Domrachev, M., Hotton, C.
+L., Kannan, S., Khovanskaya, R., \ldots{} others.
+(2020).
+NCBI Taxonomy: a Comprehensive Update on Curation, Resources and Tools.
+\emph{Database}, \emph{2020}.
+
+\leavevmode\hypertarget{ref-selvatti2015paleogene}{}%
+Selvatti, A.
+P., Gonzaga, L.
+P., \& Moraes Russo, C.
+A.
+de.
+(2015).
+A paleogene origin for crown passerines and the diversification of the oscines in the new world.
+\emph{Molecular Phylogenetics and Evolution}, \emph{88}, 1--15.
+
+\leavevmode\hypertarget{ref-smith2018constructing}{}%
+Smith, S., \& Brown, J. (2018). Constructing a broadly inclusive seed plant phylogeny. \emph{American Journal of Botany}, \emph{105}(3), 302--314.
+
+\leavevmode\hypertarget{ref-Smith2012}{}%
+Smith, S., \& O'Meara, B.
+(2012).
+TreePL: Divergence time estimation using penalized likelihood for large phylogenies.
+\emph{Bioinformatics}, \emph{28}(20), 2689--2690.
+\url{https://doi.org/10.1093/bioinformatics/bts492}
+
+\leavevmode\hypertarget{ref-Stoltzfus2013}{}%
+Stoltzfus, A., Lapp, H., Matasci, N., Deus, H., Sidlauskas, B., Zmasek, C.
+M., \ldots{} Jordan, G.
+(2013).
+Phylotastic!
+Making tree-of-life knowledge accessible, reusable and convenient.
+\emph{BMC Bioinformatics}, \emph{14}.
+\url{https://doi.org/10.1186/1471-2105-14-158}
+
+\leavevmode\hypertarget{ref-tietze2013complete}{}%
+Tietze, D.
+T., Päckert, M., Martens, J., Lehmann, H., \& Sun, Y.-H.
+(2013).
+Complete phylogeny and historical biogeography of true rosefinches (aves: Carpodacus).
+\emph{Zoological Journal of the Linnean Society}, \emph{169}(1), 215--234.
+
+\leavevmode\hypertarget{ref-title2016macrophylogenies}{}%
+Title, P.
+O., \& Rabosky, D.
+L.
+(2016).
+Do Macrophylogenies Yield Stable Macroevolutionary Inferences?
+An Example from Squamate Reptiles.
+\emph{Systematic Biology}, syw102.
+\url{https://doi.org/10.1093/sysbio/syw102}
+
+\leavevmode\hypertarget{ref-treplin2008molecular}{}%
+Treplin, S., Siegert, R., Bleidorn, C., Thompson, H.
+S., Fotso, R., \& Tiedemann, R.
+(2008).
+Molecular phylogeny of songbirds (aves: Passeriformes) and the relative utility of common nuclear marker loci.
+\emph{Cladistics}, \emph{24}(3), 328--349.
+
+\leavevmode\hypertarget{ref-vos2012nexml}{}%
+Vos, R.
+A., Balhoff, J.
+P., Caravas, J.
+A., Holder, M.
+T., Lapp, H., Maddison, W.
+P., \ldots{} others.
+(2012).
+NeXML: Rich, extensible, and verifiable representation of comparative data and metadata.
+\emph{Systematic Biology}, \emph{61}(4), 675--689.
+\url{https://doi.org/10.1093/sysbio/sys025}
+
+\leavevmode\hypertarget{ref-Webb2000}{}%
+Webb, C. (2000). Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees. \emph{The American Naturalist}, \emph{156}(2), 145--155.
+
+\leavevmode\hypertarget{ref-Webb2008}{}%
+Webb, C., Ackerly, D., \& Kembel, S.
+(2008).
+Phylocom: Software for the analysis of phylogenetic community structure and trait evolution.
+\emph{Bioinformatics}, \emph{24}(18), 2098--2100.
+\url{https://doi.org/10.1093/bioinformatics/btn358}
+
+\leavevmode\hypertarget{ref-webb2005phylomatic}{}%
+Webb, C., \& Donoghue, M. (2005). Phylomatic: Tree assembly for applied phylogenetics. \emph{Molecular Ecology Notes}, \emph{5}(1), 181--183.
+
+\leavevmode\hypertarget{ref-weir2008calibrating}{}%
+Weir, J., \& Schluter, D. (2008). Calibrating the avian molecular clock. \emph{Molecular Ecology}, \emph{17}(10), 2321--2328.
+
+\leavevmode\hypertarget{ref-zuccon2012phylogenetic}{}%
+Zuccon, D., Prŷs-Jones, R., Rasmussen, P.
+C., \& Ericson, P.
+G.
+(2012).
+The phylogenetic relationships and generic limits of finches (fringillidae).
+\emph{Molecular Phylogenetics and Evolution}, \emph{62}(2), 581--596.
+
+\endgroup
+
+\newpage
+
+\captionof{table}{Ages of congruified nodes. See Figure 3, step C2.}
+\label{tbl:table1}
+\includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example.pdf}
+
+\vspace{15mm}
+
+\captionof{table}{Summary of congruified nodes ages. See Figure 3, step C3.}
+
+\label{tbl:table2}
+\includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example-summary.png}
+
+\begin{minipage}[t]{17.3cm}
+\includegraphics{../figures/figure-workflow/figure1-new.pdf}
+\captionof{figure}{\textit{Main DateLife workflow.
+Analyses can be performed via DateLife's interactive website at \url{www.datelife.org}, or using the \texttt{datelife} R package.
+Details on the R functions used to perform the analyses are available from \texttt{datelife}'s R package vignettes at \url{https://phylotastic.org/datelife}.}}
+\label{fig:workflow}
+\end{minipage}
+
+\includegraphics{../figures/fig_runtime_main.pdf}
+\captionof{figure}{\textit{DateLife's benchmarking results.
+Computation time used to process a query and a search across \texttt{datelife}'s chronogram database, relative to number of input taxon names.
+For each N = \{10, 100, 200, ..., 1 000, ...
+, 9 000, 10 000\}, we sampled N species names from the class Aves a hundred times, and then performed a datelife search processing the input names with Taxon Names Resolution Service (TNRS; light gray), and without processing names (dark gray).
+For comparison, we performed a search using an input that had been pre-processed with TNRS (light blue).}}
+\label{fig:runtime_main}
 
@@
 \label{fig:figure2}
-
-\newpage
-
-\captionof{table}{Ages of congruified nodes. See Figure 3, step C2.}
-\label{tbl:table1}
-\includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example.pdf}
-
-\vspace{15mm}
-
-\captionof{table}{Summary of congruified nodes ages. See Figure 3, step C3.}
-
-\label{tbl:table2}
-\includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example-summary.png}
-
-\newpage
-
-\hypertarget{an-example-with-the-family-of-true-finches}{%
-\subsection{An example with the family of true finches}\label{an-example-with-the-family-of-true-finches}}
-
-\hypertarget{creating-a-query}{%
-\subsubsection{Creating a query}\label{creating-a-query}}
-
-To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the \enquote{get species from taxon} flag,
-which gets all recognized species names within a named group from a taxonomy of choice.
-Following the NCBI taxonomy, our DateLife query has 289 Fringillidae species.
-This taxon-constrained approach implies that the final results of a full DateLife analysis will be done using a tree topology and ages for the species in a named group, which do not necessarily correspond to a monophyletic group.
-Users can change this behaviour by providing a monophyletic tree as input for a DateLife search, or as a tree topology for a DateLife summary.
-
-\hypertarget{searching-the-database-1}{%
-\subsubsection{Searching the database}\label{searching-the-database-1}}
-
-Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species.
-The DateLife search identified 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Price et al., 2014).
-Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementray Table S1).
-
-\hypertarget{summarizing-search-results-2}{%
-\subsubsection{Summarizing search results}\label{summarizing-search-results-2}}
-
-The final step is to congruify and summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM.
-As explained in the \enquote{Description} section, a tree topology to summarize age data upon is required.
-By default, DateLife uses the topology from OpenTree's synthetic tree that contains the species in the search query to summarize age data upon.
-According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).
-
-\vspace{50mm}
 
@@
 
-\hypertarget{cross-validation-test}{%
-\section{Cross-validation test}\label{cross-validation-test}}
-
-We performed a cross validation analysis of the DateLife workflow using the Fringillidae chronograms.
-We used the individual tree topologies from each of the 19 source chronograms from 13 studies as inputs, treating their node ages as unknown.
-We then estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.
-We found that node ages from original study, and ages estimated using all other age data available are correlated (Fig. \ref{fig:cvXY}).
-For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka \enquote{closer to the root}) relative to the original estimate, and overestimate ages for nodes closer to the tips.
-Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. S1).
-
 \newpage
@@
 
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
-
-DateLife makes state-of-the-art data on evolutionary time frame easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
-It is an open service that does not require any expert biological knowledge from users --besides the names of the species or group they want to work with, for any of its functionality.
-
-A total of 99,474 unique terminal taxa are represented in DateLife's database.
-Incorporation of more chronograms into the database will continue to improve DateLife's services.
-One option to increase the number of chronograms in the DateLife database is the Dryad data repository.
-Methods to automatically mine chronograms from Dryad could be designed and implemented.
-However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined manually by a human curator.
-We would like to emphasize on the importance of sharing chronogram data, including systematically curated metadata, into open repositories, such as OpenTree's Phylesystem (McTavish et al., 2015) for the benefit of the scientific community as a whole.
-
-\hypertarget{age-variation-in-source-chronograms}{%
-\subsection{Age variation in source chronograms}\label{age-variation-in-source-chronograms}}
-
-Conflict in estimated ages among alternative studies is common in the literature.
-See, for example, the robust ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández, 2015; Ramshaw et al., 1972; Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021).
-Source chronograms available for the same organisms have potentially been estimated implementing calibrations very differently.
-For example, the chronograms from Burns et al.
-(2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older age estimates for the same nodes than chronograms that were inferred using fossils as calibrations (Figs.
-\ref{fig:fringillidages}, \ref{fig:cvXY}; Supplementary Figs.
-S1, S5).
-
-Different calibration implementations might also imply fundamentally distinct evolutionary hypotheses (Antonelli et al., 2017).
-For example, two independent researchers working on the same clade should both carefully select and justify their choices of fossil calibration placement.
-Yet, if one researcher concludes that a fossil should calibrate the ingroup of a clade, while another researcher concludes that the same fossil should calibrate the outgroup of the clade, the resulting age estimates will differ, as the placement of calibrations as stem or crown group has been proven to significantly affect time of lineage divergence estimates (Sauquet, 2013).
-
-\hypertarget{primary-vs-secondary-calibrations}{%
-\subsubsection{Primary vs Secondary calibrations}\label{primary-vs-secondary-calibrations}}
-
-While most chronograms in DateLife's database are constructed using primary calibrations (molecular substitution rates or ages obtained from the fossil record or geological events), DateLife summarizes chronograms using secondary calibrations (ages coming from other chronograms).
-Graur and Martin (2004) cautioned on the increased error and uncertainty in estimated ages when using secondary calibrations in dating analyses.
-Schenk (2016) showed that, in simulations, divergence times inferred using secondary calibrations are significantly younger than those inferred with primary calibrations, when obtained with Bayesian inference methods, and when priors are implemented in similar ways in both analyses.
-Accordingly, the scientific community seems to have more confidence in chronograms obtained from a single analysis, using fossil data as primary sources of calibrations (Schenk, 2016), and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data reflect a coherent evolutionary history (Sauquet, 2013), as for example done by Antonelli et al.
-(2017).
-There have been attempts to create fossil calibration databases (Ksepka et al., 2015), though these still have room to grow.
-
-It seems that using several (as opposed to just a few) secondary calibrations can provide sufficient information to alleviate or even neutralize potential biases (Sauquet, 2013).
-Certainly, further studies are required to fully understand the effect of secondary calibrations on outputs from different tree dating methods, and on downstream analyses.
-It is possible that secondary calibrations can be safely used with dating methods that do not require setting priors, such as penalized likelihood (Sanderson, 2003), with methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ (Webb et al., 2008; Webb \& Donoghue, 2005), or methods that summarize age data unto a tree topology.
-
-Our cross validation analysis might provide some insight in this regard.
-When ages are estimated with secondary calibrations, nodes closer to the root do tend to be slightly younger than ages estimated with primary calibrations.
-However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures S2-S20).
-The only excpetion to tjis was observed on results of the cross validation analysis of the Burns et al.
-(2014) chronogram, which results in much younger node ages when estimated using secondary calibrations (Supplementary Figs.
-S1, S5).
-
-\hypertarget{sumarizing-chronograms}{%
-\subsection{Sumarizing chronograms}\label{sumarizing-chronograms}}
-
-By default, DateLife currently summarizes all source chronograms that overlap with at least two species names. Users can exclude source chronograms if they have reasons to do so.
-Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely.
-To our knowledge, there are no tested measures to determine independently when a chronogram is better than another.
-Yet, several characteristics of the data used for dating analyses, as well as from the output chronogram itself, could be used to score the quality of source chronograms.
-
-Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used (Magallón, 2010; Magallón et al., 2015).
-Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
-
-DateLife provides an opportunity to capture concordance and conflict among date estimates, which can also be used as a metric for chronogram reliability.
-Its open database of chronograms allows other researchers to do such analyses themselves reproducibly, and without needing permission.
-Though, of course, they should follow proper citation practices, especially for the source chronogram studies.
-
-The exercise of summarizing age data from across multiple studies provides the opportunity to work with a more inclusive chronogram,
-that reflects a unified evolutionary history for a lineage, by putting together evidence from different hypotheses.
-The largest, and taxonomically broadest chronogram currently available from OpenTree was constructed summarizing age data from 2,274 published chronograms using NCBI's taxonomic tree as backbone (Hedges et al., 2015).
-A summarizing exercise may also amplify the effect of uncertainty and errors in source data, and blur parts of the evolutionary history of a lineage that might only be reflected in source chronograms and lost on the summary chronogram (Sauquet et al., 2021).
-
-\hypertarget{effects-on-downstream-analyses}{%
-\subsection{Effects on downstream analyses}\label{effects-on-downstream-analyses}}
-
-For downstream analyses, using alternative chronogram may deeply affect our inferences (Title \& Rabosky, 2016), particularly when studying phenomena dependent on the timing of species diversification events, such as macroevolutionary processes.
-
-In ecology and conservation biology, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance (Webb et al., 2008).
-Hence, DateLife's workflow features different ways of estimating node ages in the absence of calibrations and branch length information for certain taxa.
-\enquote{Making up} branch lengths is a common practice in scientific publications: Jetz et al.
-(2012), created a chronogram of all 9, 993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
-(2018) created a chronogram of 31, 536 ray-finned fishes, of which only 37\% had molecular data; Smith and Brown (2018) constructed a chronogram of 353, 185 seed plants where only 23\% had molecular data.
-
-Notably, risks come with this practice.
-Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration, using polytomy resolution and branch estimation methods.
-There has yet to be a thorough analysis of what can go wrong when one extends inferences beyond the data in this way, so we urge caution; we also urge readers to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, and assessing the consistency of the results with prior work.
-
-\hypertarget{conclusions}{%
-\section{Conclusions}\label{conclusions}}
-
-Knowledge of the evolutionary time frame of organisms is key to many research areas: trait evolution,
-species diversification, biogeography, macroecology and more. It is also crucial for education, science communication and policy, but generating chronograms is difficult,
-especially for those who want to use phylogenies but who are not systematists, or
-do not have the time to acquire and develop the necessary knowledge and skills to construct them on their own.
-Importantly, years of primarily public funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
-
-The DateLife project allows for easy and fast summary of public and state-of-the-art data on time of lineage divergence.
-It provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identification of regions that require more research, or that have conflicting information.
-It is available as an R package, and as a web-based R shiny application at www.datelife.org
-Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
-The DateLife project helps with awareness of the existing variation in expert time of divergence data, and will foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
-
-\hypertarget{availability}{%
-\section{Availability}\label{availability}}
-
-The DateLife software is free and open source and it can be used through its R shiny web application at \url{http://www.datelife.org}, through the \texttt{datelife} R package, and through Phylotastic's project web portal \url{https://phylo.cs.nmsu.edu/}.
-DateLife's web application is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker and OpenTree's infrastructure (dates.opentreeoflife.org/datelife).
-\texttt{datelife}'s R package stable version is available for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife}) using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R.
-Development versions are available from the GitHub repository (\url{https://github.com/phylotastic/datelife}) and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.
-
-\hypertarget{supplementary-material}{%
-\section{Supplementary Material}\label{supplementary-material}}
-
-Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities are available at \href{https://github.com/LunaSare/datelifeMS1}{datelifeMS1}, \href{https://github.com/LunaSare/datelife_examples}{datelife\_examples}, and \href{https://github.com/LunaSare/datelife_benchmark}{datelife\_benchmark} repositories in LLSR's GitHub account.
-
-\hypertarget{funding}{%
-\section{Funding}\label{funding}}
-
-Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to the Datelife project; DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), ABI-1458572 to the Phylotastic project, and ABI-1759846 to the Open Tree of Life project.
-
-\hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
-
-The DateLife project was born as a prototype tool aiming to provide these services, and was initially developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA (Stoltzfus et al., 2013).
-We thank colleagues from the O'Meara Lab at the University
-of Tennesse Knoxville for suggestions, discussions and software testing.
-The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
-that led to initial work on this project.
-The team that assembled DateLife's first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
-The Open Tree of Life project that provides the open, metadata rich repository of
-trees used to construct DateLife's chronogram database.
-The many scientists who publish their chronograms in an open, reusable form, and the scientists who curate them for deposition in the Open Tree of Life repository.
-The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.
-
-\newpage
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-\hypertarget{refs}{}
-\leavevmode\hypertarget{ref-alstrom2014discovery}{}%
-Alström, P., Hooper, D.
-M., Liu, Y., Olsson, U., Mohan, D., Gelang, M., \ldots{} Price, T.
-D.
-(2014).
-Discovery of a relict lineage and monotypic family of passerine birds.
-\emph{Biology Letters}, \emph{10}(3), 20131067.
-
-\leavevmode\hypertarget{ref-ane2009groves}{}%
-Ané, C., Eulenstein, O., Piaggio-Talice, R., \& Sanderson, M. J. (2009). Groves of phylogenetic trees. \emph{Annals of Combinatorics}, \emph{13}(2), 139--167.
-
-\leavevmode\hypertarget{ref-antonelli2017supersmart}{}%
-Antonelli, A., Hettling, H., Condamine, F.
-L., Vos, K., Nilsson, R.
-H., Sanderson, M.
-J., \ldots{} Vos, R.
-A.
-(2017).
-Toward a self-updating platform for estimating rates of speciation and migration, ages, and relationships of Taxa.
-\emph{Systematic Biology}, \emph{66}(2), 153--166.
-\url{https://doi.org/10.1093/sysbio/syw066}
-
-\leavevmode\hypertarget{ref-archie1986newick}{}%
-Archie, J., Day, W.
-H., Felsenstein, J., Maddison, W., Meacham, C., Rohlf, F.
-J., \& Swofford, D.
-(1986).
-The Newick tree format.
-Retrieved from \href{\%7Bhttps://evolution.genetics.washington.edu/phylip/newicktree.html\%7D}{\{https://evolution.genetics.washington.edu/phylip/newicktree.html\}}
-
-\leavevmode\hypertarget{ref-avibase-emberiza}{}%
-Avibase.
-(2022).
-Yellow-throated Bunting.
-\emph{Avibase - the World Bird Database}, (Online Resource).
-Retrieved from \href{\%7Bhttps://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\%7D}{\{https://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\}}
-
-\leavevmode\hypertarget{ref-Bapst2012a}{}%
-Bapst, D.
-W.
-(2012).
-Paleotree: An R package for paleontological and phylogenetic analyses of evolution.
-\emph{Methods in Ecology and Evolution}, \emph{3}(5), 803--807.
-\url{https://doi.org/10.1111/j.2041-210X.2012.00223.x}
-
-\leavevmode\hypertarget{ref-barba2018constraining}{}%
-Barba-Montoya, J., Reis, M.
-dos, Schneider, H., Donoghue, P.
-C., \& Yang, Z.
-(2018).
-Constraining uncertainty in the timescale of angiosperm evolution and the veracity of a cretaceous terrestrial revolution.
-\emph{New Phytologist}, \emph{218}(2), 819--834.
-
-\leavevmode\hypertarget{ref-barker2014mitogenomic}{}%
-Barker, F. K. (2014). Mitogenomic data resolve basal relationships among passeriform and passeridan birds. \emph{Molecular Phylogenetics and Evolution}, \emph{79}, 313--324.
-
-\leavevmode\hypertarget{ref-barker2013going}{}%
-Barker, F.
-K., Burns, K.
-J., Klicka, J., Lanyon, S.
-M., \& Lovette, I.
-J.
-(2013).
-Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds.
-\emph{Systematic Biology}, \emph{62}(2), 298--320.
-
-\leavevmode\hypertarget{ref-barker2015new}{}%
-Barker, F.
-K., Burns, K.
-J., Klicka, J., Lanyon, S.
-M., \& Lovette, I.
-J.
-(2015).
-New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
-\emph{The Auk: Ornithological Advances}, \emph{132}(2), 333--348.
-
-\leavevmode\hypertarget{ref-barker2004phylogeny}{}%
-Barker, F.
-K., Cibois, A., Schikler, P., Feinstein, J., \& Cracraft, J.
-(2004).
-Phylogeny and diversification of the largest avian radiation.
-\emph{Proceedings of the National Academy of Sciences}, \emph{101}(30), 11040--11045.
-
-\leavevmode\hypertarget{ref-beresford2005african}{}%
-Beresford, P., Barker, F., Ryan, P., \& Crowe, T.
-(2005).
-African endemics span the tree of songbirds (passeri): Molecular systematics of several evolutionary ``enigmas''.
-\emph{Proceedings of the Royal Society B: Biological Sciences}, \emph{272}(1565), 849--858.
-
-\leavevmode\hypertarget{ref-Boyle2013}{}%
-Boyle, B., Hopkins, N., Lu, Z., Raygoza Garay, J.
-A., Mozzherin, D., Rees, T., \ldots{} Enquist, B.
-J.
-(2013).
-The taxonomic name resolution service: An online tool for automated standardization of plant names.
-\emph{BMC Bioinformatics}, \emph{14}(1).
-\url{https://doi.org/10.1186/1471-2105-14-16}
-
-\leavevmode\hypertarget{ref-Britton2007}{}%
-Britton, T., Anderson, C.
-L., Jacquet, D., Lundqvist, S., \& Bremer, K.
-(2007).
-Estimating Divergence Times in Large Phylogenetic Trees.
-\emph{Systematic Biology}, \emph{56}(788777878), 741--752.
-\url{https://doi.org/10.1080/10635150701613783}
-
-\leavevmode\hypertarget{ref-bryson2014diversification}{}%
-Bryson Jr, R.
-W., Chaves, J., Smith, B.
-T., Miller, M.
-J., Winker, K., Pérez-Emán, J.
-L., \& Klicka, J.
-(2014).
-Diversification across the new world within the `blue'cardinalids (aves: Cardinalidae).
-\emph{Journal of Biogeography}, \emph{41}(3), 587--599.
-
-\leavevmode\hypertarget{ref-burleigh2015building}{}%
-Burleigh, J. G., Kimball, R. T., \& Braun, E. L. (2015). Building the avian tree of life using a large-scale, sparse supermatrix. \emph{Molecular Phylogenetics and Evolution}, \emph{84}, 53--63.
-
-\leavevmode\hypertarget{ref-burns2014phylogenetics}{}%
-Burns, K.
-J., Shultz, A.
-J., Title, P.
-O., Mason, N.
-A., Barker, F.
-K., Klicka, J., \ldots{} Lovette, I.
-J.
-(2014).
-Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds.
-\emph{Molecular Phylogenetics and Evolution}, \emph{75}, 41--77.
-
-\leavevmode\hypertarget{ref-Chamberlain2018}{}%
-Chamberlain, S. (2018). \emph{bold: Interface to Bold Systems API}. Retrieved from \url{https://CRAN.R-project.org/package=bold}
-
-\leavevmode\hypertarget{ref-Chamberlain2013}{}%
-Chamberlain, S.
-A., \& Szöcs, E.
-(2013).
-taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}.
-\emph{F1000Research}, \emph{2}(191), 1--29.
-\url{https://doi.org/10.12688/f1000research.2-191.v2}
-
-\leavevmode\hypertarget{ref-chaves2013biogeography}{}%
-Chaves, J.
-A., Hidalgo, J.
-R., \& Klicka, J.
-(2013).
-Biogeography and evolutionary history of the n eotropical genus s altator (a ves: T hraupini).
-\emph{Journal of Biogeography}, \emph{40}(11), 2180--2190.
-
-\leavevmode\hypertarget{ref-claramunt2015new}{}%
-Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{Science Advances}, \emph{1}(11), e1501005.
-
-\leavevmode\hypertarget{ref-Criscuolo2006}{}%
-Criscuolo, A., Berry, V., Douzery, E.
-J., \& Gascuel, O.
-(2006).
-SDM: A fast distance-based approach for (super)tree building in phylogenomics.
-\emph{Systematic Biology}, \emph{55}(5), 740--755.
-\url{https://doi.org/10.1080/10635150600969872}
-
-\leavevmode\hypertarget{ref-delsuc2018phylogenomic}{}%
-Delsuc, F., Philippe, H., Tsagkogeorga, G., Simion, P., Tilak, M.-K., Turon, X., \ldots{} Douzery, E.
-J.
-(2018).
-A phylogenomic framework and timescale for comparative studies of tunicates.
-\emph{BMC Biology}, \emph{16}(1), 1--14.
-
-\leavevmode\hypertarget{ref-Eastman2013}{}%
-Eastman, J.
-M., Harmon, L.
-J., \& Tank, D.
-C.
-(2013).
-Congruification: Support for time scaling large phylogenetic trees.
-\emph{Methods in Ecology and Evolution}, \emph{4}(7), 688--691.
-\url{https://doi.org/10.1111/2041-210X.12051}
-
-\leavevmode\hypertarget{ref-edgar2004muscle}{}%
-Edgar, R. C. (2004). MUSCLE: Multiple sequence alignment with high accuracy and high throughput. \emph{Nucleic Acids Research}, \emph{32}(5), 1792--1797.
-
-\leavevmode\hypertarget{ref-Felsenstein1985a}{}%
-Felsenstein, J. (1985). Phylogenies and the Comparative Method. \emph{The American Naturalist}, \emph{125}(1), 1--15. Retrieved from \url{http://www.jstor.org/stable/2461605}
-
-\leavevmode\hypertarget{ref-gbif2022taxonomy}{}%
-GBIF Secretariat.
-(2022).
-GBIF Backbone Taxonomy.
-\emph{Checklist dataset}, (Online Resource accessed via GBIF.org).
-Retrieved from \href{\%7Bhttps://doi.org/10.15468/39omei\%20\%7D}{\{https://doi.org/10.15468/39omei \}}
-
-\leavevmode\hypertarget{ref-gibb2015new}{}%
-Gibb, G.
-C., England, R., Hartig, G., McLenachan, P.
-A., Taylor Smith, B.
-L., McComish, B.
-J., \ldots{} Penny, D.
-(2015).
-New zealand passerines help clarify the diversification of major songbird lineages during the oligocene.
-\emph{Genome Biology and Evolution}, \emph{7}(11), 2983--2995.
-
-\leavevmode\hypertarget{ref-graur2004reading}{}%
-Graur, D., \& Martin, W. (2004). Reading the entrails of chickens: Molecular timescales of evolution and the illusion of precision. \emph{TRENDS in Genetics}, \emph{20}(2), 80--86.
-
-\leavevmode\hypertarget{ref-hackett2008phylogenomic}{}%
-Hackett, S.
-J., Kimball, R.
-T., Reddy, S., Bowie, R.
-C., Braun, E.
-L., Braun, M.
-J., \ldots{} others.
-(2008).
-A phylogenomic study of birds reveals their evolutionary history.
-\emph{Science}, \emph{320}(5884), 1763--1768.
-
-\leavevmode\hypertarget{ref-Hedges2015}{}%
-Hedges, S.
-B., Marin, J., Suleski, M., Paymer, M., \& Kumar, S.
-(2015).
-Tree of life reveals clock-like speciation and diversification.
-\emph{Molecular Biology and Evolution}, \emph{32}(4), 835--845.
-\url{https://doi.org/10.1093/molbev/msv037}
-
-\leavevmode\hypertarget{ref-Heibl2008}{}%
-Heibl, C. (2008). \emph{PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages.} Retrieved from \url{http://www.christophheibl.de/Rpackages.html}
-
-\leavevmode\hypertarget{ref-hooper2017chromosomal}{}%
-Hooper, D. M., \& Price, T. D. (2017). Chromosomal inversion differences correlate with range overlap in passerine birds. \emph{Nature Ecology \& Evolution}, \emph{1}(10), 1526.
-
-\leavevmode\hypertarget{ref-Huelsenbeck2001}{}%
-Huelsenbeck, J. P., \& Ronquist, F. (2001). MRBAYES: Bayesian inference of phylogenetic trees. \emph{Bioinformatics}, \emph{17}(8), 754--755. \url{https://doi.org/10.1093/bioinformatics/17.8.754}
-
-\leavevmode\hypertarget{ref-Jetz2012}{}%
-Jetz, W., Thomas, G., Joy, J.
-J., Hartmann, K., \& Mooers, A.
-(2012).
-The global diversity of birds in space and time.
-\emph{Nature}, \emph{491}(7424), 444--448.
-\url{https://doi.org/10.1038/nature11631}
-
-\leavevmode\hypertarget{ref-johansson2008phylogenetic}{}%
-Johansson, U.
-S., Fjeldså, J., \& Bowie, R.
-C.
-(2008).
-Phylogenetic relationships within passerida (aves: Passeriformes): A review and a new molecular phylogeny based on three nuclear intron markers.
-\emph{Molecular Phylogenetics and Evolution}, \emph{48}(3), 858--876.
-
-\leavevmode\hypertarget{ref-katoh2009multiple}{}%
-Katoh, K., Asimenos, G., \& Toh, H. (2009). Multiple alignment of dna sequences with mafft. In \emph{Bioinformatics for dna sequence analysis} (pp. 39--64). Springer.
-
-\leavevmode\hypertarget{ref-kimball2019phylogenomic}{}%
-Kimball, R. T., Oliveros, C. H., Wang, N., White, N. D., Barker, F. K., Field, D. J., \ldots{} others. (2019). A phylogenomic supertree of birds. \emph{Diversity}, \emph{11}(7), 109.
-
-\leavevmode\hypertarget{ref-klicka2014comprehensive}{}%
-Klicka, J., Barker, F.
-K., Burns, K.
-J., Lanyon, S.
-M., Lovette, I.
-J., Chaves, J.
-A., \& Bryson Jr, R.
-W.
-(2014).
-A comprehensive multilocus assessment of sparrow (aves: Passerellidae) relationships.
-\emph{Molecular Phylogenetics and Evolution}, \emph{77}, 177--182.
-
-\leavevmode\hypertarget{ref-ksepka2015fossil}{}%
-Ksepka, D.
-T., Parham, J.
-F., Allman, J.
-F., Benton, M.
-J., Carrano, M.
-T., Cranston, K.
-A., \ldots{} others.
-(2015).
-The fossil calibration database---a new resource for divergence dating.
-\emph{Systematic Biology}, \emph{64}(5), 853--859.
-
-\leavevmode\hypertarget{ref-lamichhaney2015evolution}{}%
-Lamichhaney, S., Berglund, J., Almén, M.
-S., Maqbool, K., Grabherr, M., Martinez-Barrio, A., \ldots{} others.
-(2015).
-Evolution of darwin's finches and their beaks revealed by genome sequencing.
-\emph{Nature}, \emph{518}(7539), 371--375.
-
-\leavevmode\hypertarget{ref-laubichler2009form}{}%
-Laubichler, M. D., \& Maienschein, J. (2009). \emph{Form and function in developmental evolution}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-lepage2004avibase}{}%
-Lepage, D. (2004). \emph{Avibase: The world bird database}. Bird Studies Canada.
-
-\leavevmode\hypertarget{ref-lepage2014avibase}{}%
-Lepage, D., Vaidya, G., \& Guralnick, R. (2014). Avibase--a database system for managing and organizing taxonomic concepts. \emph{ZooKeys}, (420), 117.
-
-\leavevmode\hypertarget{ref-lerner2011multilocus}{}%
-Lerner, H.
-R., Meyer, M., James, H.
-F., Hofreiter, M., \& Fleischer, R.
-C.
-(2011).
-Multilocus resolution of phylogeny and timescale in the extant adaptive radiation of hawaiian honeycreepers.
-\emph{Current Biology}, \emph{21}(21), 1838--1844.
-
-\leavevmode\hypertarget{ref-lovette2010comprehensive}{}%
-Lovette, I.
-J., Pérez-Emán, J.
-L., Sullivan, J.
-P., Banks, R.
-C., Fiorentino, I., Córdoba-Córdoba, S., \ldots{} others.
-(2010).
-A comprehensive multilocus phylogeny for the wood-warblers and a revised classification of the parulidae (aves).
-\emph{Molecular Phylogenetics and Evolution}, \emph{57}(2), 753--770.
-
-\leavevmode\hypertarget{ref-magallon2001absolute}{}%
-Magallon, S., \& Sanderson, M. (2001). Absolute diversification rates in angiosperm clades. \emph{Evolution}, \emph{55}(9), 1762--1780.
-
-\leavevmode\hypertarget{ref-magallon2010using}{}%
-Magallón, S. (2010). Using fossils to break long branches in molecular dating: A comparison of relaxed clocks applied to the origin of angiosperms. \emph{Systematic Biology}, \emph{59}(4), 384--399.
-
-\leavevmode\hypertarget{ref-magallon2015metacalibrated}{}%
-Magallón, S., Gómez-Acevedo, S., Sánchez-Reyes, L.
-L., \& Hernández-Hernández, T.
-(2015).
-A metacalibrated time-tree documents the early rise of flowering plant phylogenetic diversity.
-\emph{New Phytologist}, \emph{207}(2), 437--453.
-
-\leavevmode\hypertarget{ref-mctavish2015phylesystem}{}%
-McTavish, E.
-J., Hinchliff, C.
-E., Allman, J.
-F., Brown, J.
-W., Cranston, K.
-A., Holder, M.
-T., \ldots{} Smith, S.
-(2015).
-Phylesystem: A git-based data store for community-curated phylogenetic estimates.
-\emph{Bioinformatics}, \emph{31}(17), 2794--2800.
-
-\leavevmode\hypertarget{ref-Michonneau2016}{}%
-Michonneau, F., Brown, J.
-W., \& Winter, D.
-J.
-(2016).
-rotl: an R package to interact with the Open Tree of Life data.
-\emph{Methods in Ecology and Evolution}, \emph{7}(12), 1476--1481.
-\url{https://doi.org/10.1111/2041-210X.12593}
-
-\leavevmode\hypertarget{ref-Morlon2014}{}%
-Morlon, H. (2014). Phylogenetic approaches for studying diversification. \emph{Ecology Letters}, \emph{17}(4), 508--525. \url{https://doi.org/10.1111/ele.12251}
-
-\leavevmode\hypertarget{ref-moyle2016tectonic}{}%
-Moyle, R.
-G., Oliveros, C.
-H., Andersen, M.
-J., Hosner, P.
-A., Benz, B.
-W., Manthey, J.
-D., \ldots{} Faircloth, B.
-C.
-(2016).
-Tectonic collision and uplift of wallacea triggered the global songbird radiation.
-\emph{Nature Communications}, \emph{7}(1), 1--7.
-
-\leavevmode\hypertarget{ref-oliveros2019earth}{}%
-Oliveros, C.
-H., Field, D.
-J., Ksepka, D.
-T., Barker, F.
-K., Aleixo, A., Andersen, M.
-J., \ldots{} others.
-(2019).
-Earth history and the passerine superradiation.
-\emph{Proceedings of the National Academy of Sciences}, \emph{116}(16), 7916--7925.
-
-\leavevmode\hypertarget{ref-Ooms2018}{}%
-Ooms, J., \& Chamberlain, S. (2018). \emph{Phylocomr: Interface to 'phylocom'}. Retrieved from \url{https://CRAN.R-project.org/package=phylocomr}
-
-\leavevmode\hypertarget{ref-opentreeAPIs}{}%
-Open Tree Of Life, Redelings, B., Cranston, K.
-A., Allman, J., Holder, M.
-T., \& McTavish, E.
-J.
-(2016).
-Open Tree of Life APIs v3.0.
-\emph{Open Tree of Life Project}, (Online Resources).
-Retrieved from \href{\%7Bhttps://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\%7D}{\{https://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\}}
-
-\leavevmode\hypertarget{ref-opentreeoflife2019synth}{}%
-Open Tree Of Life, Redelings, B., Sánchez Reyes, L.
-L., Cranston, K.
-A., Allman, J., Holder, M.
-T., \& McTavish, E.
-J.
-(2019).
-Open tree of life synthetic tree v12.3.
-\emph{Zenodo}.
-Retrieved from \url{https://doi.org/10.5281/zenodo.3937742}
-
-\leavevmode\hypertarget{ref-odeen2011evolution}{}%
-Ödeen, A., Håstad, O., \& Alström, P. (2011). Evolution of ultraviolet vision in the largest avian radiation-the passerines. \emph{BMC Evolutionary Biology}, \emph{11}(1), 1--8.
-
-\leavevmode\hypertarget{ref-paradis2004}{}%
-Paradis, E., Claude, J., \& Strimmer, K. (2004). APE: analyses of phylogenetics and evolution in R language. \emph{Bioinformatics}, \emph{20}(2), 289--290.
-
-\leavevmode\hypertarget{ref-parchman2007coevolution}{}%
-Parchman, T.
-L., Benkman, C.
-W., \& Mezquida, E.
-T.
-(2007).
-Coevolution between hispaniolan crossbills and pine: Does more time allow for greater phenotypic escalation at lower latitude?
-\emph{Evolution}, \emph{61}(9), 2142--2153.
-
-\leavevmode\hypertarget{ref-packert2012horizontal}{}%
-Päckert, M., Martens, J., Sun, Y.-H., Severinghaus, L.
-L., Nazarenko, A.
-A., Ting, J., \ldots{} Tietze, D.
-T.
-(2012).
-Horizontal and elevational phylogeographic patterns of himalayan and southeast asian forest passerines (aves: Passeriformes).
-\emph{Journal of Biogeography}, \emph{39}(3), 556--573.
-
-\leavevmode\hypertarget{ref-pennell2014geiger}{}%
-Pennell, M.
-W., Eastman, J.
-M., Slater, G.
-J., Brown, J.
-W., Uyeda, J.
-C., FitzJohn, R.
-G., \ldots{} Harmon, L.
-J.
-(2014).
-Geiger v2.
-0: An expanded suite of methods for fitting macroevolutionary models to phylogenetic trees.
-\emph{Bioinformatics}, \emph{30}(15), 2216--2218.
-
-\leavevmode\hypertarget{ref-posadas2006historical}{}%
-Posadas, P., Crisci, J. V., \& Katinas, L. (2006). Historical biogeography: A review of its basic concepts and critical issues. \emph{Journal of Arid Environments}, \emph{66}(3), 389--403.
-
-\leavevmode\hypertarget{ref-powell2014comprehensive}{}%
-Powell, A.
-F., Barker, F.
-K., Lanyon, S.
-M., Burns, K.
-J., Klicka, J., \& Lovette, I.
-J.
-(2014).
-A comprehensive species-level molecular phylogeny of the new world blackbirds (icteridae).
-\emph{Molecular Phylogenetics and Evolution}, \emph{71}, 94--112.
-
-\leavevmode\hypertarget{ref-price2014niche}{}%
-Price, T.
-D., Hooper, D.
-M., Buchanan, C.
-D., Johansson, U.
-S., Tietze, D.
-T., Alström, P., \ldots{} others.
-(2014).
-Niche filling slows the diversification of himalayan songbirds.
-\emph{Nature}, \emph{509}(7499), 222.
-
-\leavevmode\hypertarget{ref-pulgarin2013multilocus}{}%
-Pulgarín-R, P.
-C., Smith, B.
-T., Bryson Jr, R.
-W., Spellman, G.
-M., \& Klicka, J.
-(2013).
-Multilocus phylogeny and biogeography of the new world pheucticus grosbeaks (aves: Cardinalidae).
-\emph{Molecular Phylogenetics and Evolution}, \emph{69}(3), 1222--1227.
-
-\leavevmode\hypertarget{ref-rabosky2018inverse}{}%
-Rabosky, D.
-L., Chang, J., Title, P.
-O., Cowman, P.
-F., Sallan, L., Friedman, M., \ldots{} others.
-(2018).
-An inverse latitudinal gradient in speciation rate for marine fishes.
-\emph{Nature}, \emph{559}(7714), 392.
-
-\leavevmode\hypertarget{ref-ramshaw1972time}{}%
-Ramshaw, J., Richardson, D., Meatyard, B., Brown, R., Richardson, M., Thompson, E., \& Boulter, D.
-(1972).
-The time of origin of the flowering plants determined by using amino acid sequence data of cytochrome c.
-\emph{New Phytologist}, \emph{71}(5), 773--779.
-
-\leavevmode\hypertarget{ref-ratnasingham2007bold}{}%
-Ratnasingham, S., \& Hebert, P. D. (2007). BOLD: The barcode of life data system (http://www. Barcodinglife. Org). \emph{Molecular Ecology Notes}, \emph{7}(3), 355--364.
-
-\leavevmode\hypertarget{ref-RCoreTeam2018}{}%
-R Core Team. (2018). \emph{R: a language and environment for statistical computing}. Vienna, Austria: R Foundation for Statistical Computing.
-
-\leavevmode\hypertarget{ref-rees2017automated}{}%
-Rees, \& Cranston, K. (2017). Automated assembly of a reference taxonomy for phylogenetic data synthesis. \emph{Biodiversity Data Journal}, (5).
-
-\leavevmode\hypertarget{ref-rees2017irmng}{}%
-Rees, Vandepitte, L., Decock, W., \& Vanhoorne, B. (2017). IRMNG 2006--2016: 10 Years of a Global Taxonomic Database. \emph{Biodiversity Informatics}, \emph{12}.
-
-\leavevmode\hypertarget{ref-Revell2012}{}%
-Revell, L. J. (2012). Phytools: An r package for phylogenetic comparative biology (and other things). \emph{Methods in Ecology and Evolution}, \emph{3}, 217--223.
-
-\leavevmode\hypertarget{ref-Ronquist2003}{}%
-Ronquist, F., \& Huelsenbeck, J.
-P.
-(2003).
-MrBayes 3: Bayesian phylogenetic inference under mixed models.
-\emph{Bioinformatics}, \emph{19}(12), 1572--1574.
-\url{https://doi.org/10.1093/bioinformatics/btg180}
-
-\leavevmode\hypertarget{ref-datelifeplot}{}%
-Sanchez-Reyes, L.
-L., \& O'Meara, B.
-(2022).
-\texttt{datelifeplot}: Methods to plot chronograms and outputs of the \texttt{datelife} package.
-\emph{R Package Release V0.2.2}.
-Retrieved from \url{https://zenodo.org/badge/latestdoi/381501451}
-
-\leavevmode\hypertarget{ref-datelifeR}{}%
-Sanchez-Reyes, L.
-L., O'Meara, B., Eastman, J., Heath, T., Wright, A., Schliep, K., \ldots{} Alfaro, M.
-(2022).
-\texttt{datelife}: Scientific Data on Time of Lineage Divergence for Your Taxa.
-\emph{R Package Release V0.6.2}.
-Retrieved from \url{https://doi.org/10.5281/zenodo.593938}
-
-\leavevmode\hypertarget{ref-sanderson2002}{}%
-Sanderson, M.
-(2002).
-Estimating Absolute Rates of Molecular Evolution and Divergence Times: A Penalized Likelihood Approach.
-\emph{Molecular Biology and Evolution}, \emph{19}(1), 101--109.
-\url{https://doi.org/10.1093/oxfordjournals.molbev.a003974}
-
-\leavevmode\hypertarget{ref-sanderson2003r8s}{}%
-Sanderson, M. (2003). r8s: Inferring Absolute Rates of Molecular Evolution and Divergence Times in the Absence of a Molecular Clock. \emph{Bioinformatics}, \emph{19}(2), 301--302.
-
-\leavevmode\hypertarget{ref-sanderson2001sources}{}%
-Sanderson, M., \& Doyle, J.
-(2001).
-Sources of error and confidence intervals in estimating the age of angiosperms from rbcL and 18S rDNA data.
-\emph{American Journal of Botany}, \emph{88}(8), 1499--1516.
-
-\leavevmode\hypertarget{ref-sauquet2013practical}{}%
-Sauquet, H. (2013). A practical guide to molecular dating. \emph{Comptes Rendus Palevol}, \emph{12}(6), 355--367.
-
-\leavevmode\hypertarget{ref-sauquet2021age}{}%
-Sauquet, H., Ramírez-Barahona, S., \& Magallón, S. (2021). \emph{The age of flowering plants is unknown}.
-
-\leavevmode\hypertarget{ref-schenk2016sec}{}%
-Schenk, J. J. (2016). Consequences of secondary calibrations on divergence time estimates. \emph{PLoS ONE}, \emph{11}(1). \url{https://doi.org/10.1371/journal.pone.0148228}
-
-\leavevmode\hypertarget{ref-schliep2011phangorn}{}%
-Schliep, K. P. (2011). Phangorn: Phylogenetic analysis in r. \emph{Bioinformatics}, \emph{27}(4), 592--593.
-
-\leavevmode\hypertarget{ref-schoch2020ncbi}{}%
-Schoch, C.
-L., Ciufo, S., Domrachev, M., Hotton, C.
-L., Kannan, S., Khovanskaya, R., \ldots{} others.
-(2020).
-NCBI Taxonomy: a Comprehensive Update on Curation, Resources and Tools.
-\emph{Database}, \emph{2020}.
-
-\leavevmode\hypertarget{ref-selvatti2015paleogene}{}%
-Selvatti, A.
-P., Gonzaga, L.
-P., \& Moraes Russo, C.
-A.
-de.
-(2015).
-A paleogene origin for crown passerines and the diversification of the oscines in the new world.
-\emph{Molecular Phylogenetics and Evolution}, \emph{88}, 1--15.
-
-\leavevmode\hypertarget{ref-smith2018constructing}{}%
-Smith, S., \& Brown, J. (2018). Constructing a broadly inclusive seed plant phylogeny. \emph{American Journal of Botany}, \emph{105}(3), 302--314.
-
-\leavevmode\hypertarget{ref-Smith2012}{}%
-Smith, S., \& O'Meara, B.
-(2012).
-TreePL: Divergence time estimation using penalized likelihood for large phylogenies.
-\emph{Bioinformatics}, \emph{28}(20), 2689--2690.
-\url{https://doi.org/10.1093/bioinformatics/bts492}
-
-\leavevmode\hypertarget{ref-Stoltzfus2013}{}%
-Stoltzfus, A., Lapp, H., Matasci, N., Deus, H., Sidlauskas, B., Zmasek, C.
-M., \ldots{} Jordan, G.
-(2013).
-Phylotastic!
-Making tree-of-life knowledge accessible, reusable and convenient.
-\emph{BMC Bioinformatics}, \emph{14}.
-\url{https://doi.org/10.1186/1471-2105-14-158}
-
-\leavevmode\hypertarget{ref-tietze2013complete}{}%
-Tietze, D.
-T., Päckert, M., Martens, J., Lehmann, H., \& Sun, Y.-H.
-(2013).
-Complete phylogeny and historical biogeography of true rosefinches (aves: Carpodacus).
-\emph{Zoological Journal of the Linnean Society}, \emph{169}(1), 215--234.
-
-\leavevmode\hypertarget{ref-title2016macrophylogenies}{}%
-Title, P.
-O., \& Rabosky, D.
-L.
-(2016).
-Do Macrophylogenies Yield Stable Macroevolutionary Inferences?
-An Example from Squamate Reptiles.
-\emph{Systematic Biology}, syw102.
-\url{https://doi.org/10.1093/sysbio/syw102}
-
-\leavevmode\hypertarget{ref-treplin2008molecular}{}%
-Treplin, S., Siegert, R., Bleidorn, C., Thompson, H.
-S., Fotso, R., \& Tiedemann, R.
-(2008).
-Molecular phylogeny of songbirds (aves: Passeriformes) and the relative utility of common nuclear marker loci.
-\emph{Cladistics}, \emph{24}(3), 328--349.
-
-\leavevmode\hypertarget{ref-vos2012nexml}{}%
-Vos, R.
-A., Balhoff, J.
-P., Caravas, J.
-A., Holder, M.
-T., Lapp, H., Maddison, W.
-P., \ldots{} others.
-(2012).
-NeXML: Rich, extensible, and verifiable representation of comparative data and metadata.
-\emph{Systematic Biology}, \emph{61}(4), 675--689.
-\url{https://doi.org/10.1093/sysbio/sys025}
-
-\leavevmode\hypertarget{ref-Webb2000}{}%
-Webb, C. (2000). Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees. \emph{The American Naturalist}, \emph{156}(2), 145--155.
-
-\leavevmode\hypertarget{ref-Webb2008}{}%
-Webb, C., Ackerly, D., \& Kembel, S.
-(2008).
-Phylocom: Software for the analysis of phylogenetic community structure and trait evolution.
-\emph{Bioinformatics}, \emph{24}(18), 2098--2100.
-\url{https://doi.org/10.1093/bioinformatics/btn358}
-
-\leavevmode\hypertarget{ref-webb2005phylomatic}{}%
-Webb, C., \& Donoghue, M. (2005). Phylomatic: Tree assembly for applied phylogenetics. \emph{Molecular Ecology Notes}, \emph{5}(1), 181--183.
-
-\leavevmode\hypertarget{ref-weir2008calibrating}{}%
-Weir, J., \& Schluter, D. (2008). Calibrating the avian molecular clock. \emph{Molecular Ecology}, \emph{17}(10), 2321--2328.
-
-\leavevmode\hypertarget{ref-zuccon2012phylogenetic}{}%
-Zuccon, D., Prŷs-Jones, R., Rasmussen, P.
-C., \& Ericson, P.
-G.
-(2012).
-The phylogenetic relationships and generic limits of finches (fringillidae).
-\emph{Molecular Phylogenetics and Evolution}, \emph{62}(2), 581--596.
-
-\endgroup
-
 \end{document}

### 2023-02-21 manuscript/manuscript.Rmd
format titles and subtitles and add edits to cross validation section

@@
 
-## Creating a search query
+\begin{center}
+\emph{Creating a Search Query}
+\end{center}
 
@@
 
-## Searching a chronogram database
+\begin{center}
+\emph{Searching a Chronogram Database}
+\end{center}
+
 At the time of writing of this manuscript 
@@
 
-## Summarizing search results
+\begin{center}
+\emph{Summarizing Search Results}
+\end{center}
+
 Summary information is extracted from the \texttt{datelifeResult} object to inform decisions for subsequent steps in the analysis workflow. Basic summary information available to the user is:
@@
 
-## Applying secondary calibrations
+\begin{center}
+\emph{Applying Secondary Calibrations}
+\end{center}
+
 Once a topology is chosen, DateLife applies the congruification method [@Eastman2013] that find nodes belonging to the same clade across source chronograms, and then extracts the corresponding node ages from patristic distance matrices stored as a `datelifeResult` object.
@@
 
-## Dating a tree topology
-### With no branch lengths
+\begin{center}
+\emph{Dating a Tree Topology}
+\end{center}
+
+**_With no branch lengths.--_**
 When producing or obtaining a tree with branch lengths for a group of interest is not possible, DateLife can date a topology without branch lengths, obtained from OpenTree or by implementing the supertree approach described above, by implementing the Branch Length Adjuster [BLADJ; @Webb2008; @webb2005phylomatic] algorithm, which requires no initial branch lengths.
@@
 
-### Dating a tree with branch lengths
+**_Dating a tree with branch lengths.--_**
 Relative branch lengths provide key information for phylogenetic dating, specifically for nodes without any calibration data available. 
@@
 
-## Visualizing results
+\begin{center}
+\emph{Visualizing Results}
+\end{center}
+
 Finally, users can save all source and summary chronograms in formats allowing for reuse and reanalysis, such as newick and the R "phylo" format.
@@
 
-### Searching the database
-
+**_Searching the database.--_**
 Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species as tip taxa.
@@
 
-### Summarizing search results
-
+**_Summarizing search results.--_**
 The final step entailed congruifying and summarizing the age data available for the Fringillidae species into two single summary chronograms, using two different types of summary ages, median and SDM.
@@
 
-# Cross-validation test
+\begin{center}
+\textsc{Cross-Validation Test}
+\end{center}
 
 <!--Data from source chronograms can be used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates-->
-We performed a cross validation analysis of the DateLife workflow using the Fringillidae chronograms.
-We used the individual tree topologies from each of the 19 source chronograms from 13 studies as inputs, treating their node ages as unknown.
-We then estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.
-We found that node ages from original study, and ages estimated using all other age data available are correlated (Fig. \ref{fig:cvXY}).
+We performed a cross validation test of a DateLife analysis using the Fringillidae source chronograms obtained above (Supplementary Figs. S2-S20).
+We used as inputs for a DateLife analysis all individual tree topologies from each of the 19 source chronograms from 13 studies, treating their node ages as unknown.
+We congruified node ages extracted from chronograms from all other studies upon the individual topologies, effectively excluding original ages from each topology.
+Finally, average node ages per node were applied as secondary calibrations and smoothed with the BLADJ algorithm.
+We found that node ages from the original studies, and ages estimated using all other age data available are largely correlated (Fig. \ref{fig:cvXY}).
 For five studies, DateLife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka 'closer to the root') relative to the original estimate, and overestimate ages for nodes closer to the tips.
-Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. S1).
+Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig.
+S1).
+In general, topologically deeper nodes display the largest age variation between node ages from the original chronograms and ages summarized with DateLife.
 
@@
 Source chronograms available for the same organisms have potentially been estimated implementing calibrations very differently.
-In the Fringillidae example shown here, the chronograms from @burns2014phylogenetics were inferred using molecular substitution rate estimates across birds [@weir2008calibrating], and have much older age estimates for the same nodes than chronograms that were inferred using fossils as calibrations (Figs.
+In the Fringillidae example above, the chronograms from @burns2014phylogenetics were inferred using molecular substitution rate estimates across birds [@weir2008calibrating], and have much older age estimates for the same nodes than chronograms that were inferred using fossils as calibrations (Figs.
 \ref{fig:fringillidages}, \ref{fig:cvXY}; Supplementary Figs.
@@
 
-## Effects of phylogenetic sampling on downstream analyses
+\begin{center}
+\emph{Effects of Phylogenetic Sampling on Downstream Analyses}
+\end{center}
 
 Analysis of species diversification using simulated and empirical phylogenies suggest that using a more completely sampled phylogeny provides estimates that are closer to the true diversification history than when analysing incompletely sampled phylogenies [@cusimano2012new; @chang2020estimating; @sun2020estimating].
-Ideally phylogenies should be completed using genetic data, but this is a very hard task to achieve for many biological groups.
+Ideally, phylogenies should be completed using genetic data, but this is a difficult task to achieve for many biological groups.
 Hence, DateLife's workflow features different ways of assigning node ages in the absence of calibrations and branch length information for certain taxa.

### 2023-02-21 manuscript/manuscript.tex
format titles and subtitles and add edits to cross validation section

@@
 
-Word count: 5956
+Word count: 6111
 
@@
 
-\hypertarget{creating-a-search-query}{%
-\subsection{Creating a search query}\label{creating-a-search-query}}
+\begin{center}
+\emph{Creating a Search Query}
+\end{center}
 
@@
 
-\hypertarget{searching-a-chronogram-database}{%
-\subsection{Searching a chronogram database}\label{searching-a-chronogram-database}}
+\begin{center}
+\emph{Searching a Chronogram Database}
+\end{center}
 
 At the time of writing of this manuscript
-(Feb 20, 2023),
+(Feb 21, 2023),
 DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies.
@@
 
-\hypertarget{choosing-a-topology}{%
-\subsection{Choosing a topology}\label{choosing-a-topology}}
+\begin{center}
+\emph{Choosing a Topology}
+\end{center}
 
@@
 
-\hypertarget{applying-secondary-calibrations}{%
-\subsection{Applying secondary calibrations}\label{applying-secondary-calibrations}}
+\begin{center}
+\emph{Applying Secondary Calibrations}
+\end{center}
 
@@
 
-\hypertarget{dating-a-tree-topology}{%
-\subsection{Dating a tree topology}\label{dating-a-tree-topology}}
-
-\hypertarget{with-no-branch-lengths}{%
-\subsubsection{With no branch lengths}\label{with-no-branch-lengths}}
-
+\begin{center}
+\emph{Dating a Tree Topology}
+\end{center}
+
+\textbf{\emph{With no branch lengths.--}}
 When producing or obtaining a tree with branch lengths for a group of interest is not possible, DateLife can date a topology without branch lengths, obtained from OpenTree or by implementing the supertree approach described above, by implementing the Branch Length Adjuster (BLADJ; Webb et al., 2008; Webb \& Donoghue, 2005) algorithm, which requires no initial branch lengths.
@@
 
-\hypertarget{dating-a-tree-with-branch-lengths}{%
-\subsubsection{Dating a tree with branch lengths}\label{dating-a-tree-with-branch-lengths}}
-
+\textbf{\emph{Dating a tree with branch lengths.--}}
 Relative branch lengths provide key information for phylogenetic dating, specifically for nodes without any calibration data available.
@@
 
-\hypertarget{visualizing-results}{%
-\subsection{Visualizing results}\label{visualizing-results}}
+\begin{center}
+\emph{Visualizing Results}
+\end{center}
 
@@
 
-\hypertarget{an-example-with-the-family-of-true-finches}{%
-\subsection{An example with the family of true finches}\label{an-example-with-the-family-of-true-finches}}
-
-\hypertarget{creating-a-query}{%
-\subsubsection{Creating a query}\label{creating-a-query}}
-
+\begin{center}
+\emph{An Example with the Family of True Finches}
+\end{center}
+
+\textbf{\emph{Creating a query.--}}
 To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the \enquote{get species from taxon} flag,
@@
 
-\hypertarget{searching-the-database-1}{%
-\subsubsection{Searching the database}\label{searching-the-database-1}}
-
+\textbf{\emph{Searching the database.--}}
 Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species as tip taxa.
@@
 
-\hypertarget{summarizing-search-results-2}{%
-\subsubsection{Summarizing search results}\label{summarizing-search-results-2}}
-
+\textbf{\emph{Summarizing search results.--}}
 The final step entailed congruifying and summarizing the age data available for the Fringillidae species into two single summary chronograms, using two different types of summary ages, median and SDM.
@@
 
-\hypertarget{cross-validation-test}{%
-\section{Cross-validation test}\label{cross-validation-test}}
-
-We performed a cross validation analysis of the DateLife workflow using the Fringillidae chronograms.
-We used the individual tree topologies from each of the 19 source chronograms from 13 studies as inputs, treating their node ages as unknown.
-We then estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.
-We found that node ages from original study, and ages estimated using all other age data available are correlated (Fig. \ref{fig:cvXY}).
+\begin{center}
+\textsc{Cross-Validation Test}
+\end{center}
+
+We performed a cross validation test of a DateLife analysis using the Fringillidae source chronograms obtained above (Supplementary Figs. S2-S20).
+We used as inputs for a DateLife analysis all individual tree topologies from each of the 19 source chronograms from 13 studies, treating their node ages as unknown.
+We congruified node ages extracted from chronograms from all other studies upon the individual topologies, effectively excluding original ages from each topology.
+Average node ages per node were applied as secondary calibrations, smoothed with the BLADJ algorithm.
+We found that node ages from the original studies, and ages estimated using all other age data available are largely correlated (Fig. \ref{fig:cvXY}).
 For five studies, DateLife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka \enquote{closer to the root}) relative to the original estimate, and overestimate ages for nodes closer to the tips.
-Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. S1).
-
-\hypertarget{discussion}{%
-\section{Discussion}\label{discussion}}
+Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig.
+S1).
+In general, topologically deeper nodes display the largest age variation between node ages from the original chronograms and ages summarized with DateLife.
+
+\begin{center}
+\textsc{Discussion}
+\end{center}
 
@@
 
-\hypertarget{age-variation-in-source-chronograms}{%
-\subsection{Age variation in source chronograms}\label{age-variation-in-source-chronograms}}
+\begin{center}
+\emph{Age Variation in Source Chronograms}
+\end{center}
 
@@
 Source chronograms available for the same organisms have potentially been estimated implementing calibrations very differently.
-In the Fringillidae example shown here, the chronograms from Burns et al.
+In the Fringillidae example above, the chronograms from Burns et al.
 (2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older age estimates for the same nodes than chronograms that were inferred using fossils as calibrations (Figs.
@@
 
-\hypertarget{primary-vs-secondary-calibrations}{%
-\subsection{Primary vs Secondary calibrations}\label{primary-vs-secondary-calibrations}}
+\begin{center}
+\emph{Primary vs Secondary Calibrations}
+\end{center}
 
@@
 
-\hypertarget{sumarizing-chronograms}{%
-\subsection{Sumarizing chronograms}\label{sumarizing-chronograms}}
+\begin{center}
+\emph{Sumarizing Chronograms}
+\end{center}
 
@@
 
-\hypertarget{effects-of-phylogenetic-sampling-on-downstream-analyses}{%
-\subsection{Effects of phylogenetic sampling on downstream analyses}\label{effects-of-phylogenetic-sampling-on-downstream-analyses}}
+\begin{center}
+\emph{Effects of Phylogenetic Sampling on Downstream Analyses}
+\end{center}
 
 Analysis of species diversification using simulated and empirical phylogenies suggest that using a more completely sampled phylogeny provides estimates that are closer to the true diversification history than when analysing incompletely sampled phylogenies (Chang, Rabosky, \& Alfaro, 2020; Cusimano, Stadler, \& Renner, 2012; Sun et al., 2020).
-Ideally phylogenies should be completed using genetic data, but this is a very hard task to achieve for many biological groups.
+Ideally, phylogenies should be completed using genetic data, but this is a difficult task to achieve for many biological groups.
 Hence, DateLife's workflow features different ways of assigning node ages in the absence of calibrations and branch length information for certain taxa.
@@
 
-\hypertarget{conclusions}{%
-\section{Conclusions}\label{conclusions}}
+\begin{center}
+\textsc{Conclusions}
+\end{center}
 
@@
 
-\hypertarget{availability}{%
-\section{Availability}\label{availability}}
+\begin{center}
+\textsc{Availability}
+\end{center}
 
@@
 
-\hypertarget{supplementary-material}{%
-\section{Supplementary Material}\label{supplementary-material}}
+\begin{center}
+\textsc{Supplementary Material}
+\end{center}
 
@@
 
-\hypertarget{funding}{%
-\section{Funding}\label{funding}}
+\begin{center}
+\textsc{Funding}
+\end{center}
 
@@
 
-\hypertarget{acknowledgements}{%
-\section{Acknowledgements}\label{acknowledgements}}
+\begin{center}
+\textsc{Acknowledgements}
+\end{center}
 

### 2023-02-21 manuscript/manuscript.Rmd
format response as md file

@@
 While, using initial branch lengths data is the golden standard for phylogenetic dating analyses, producing such requires assembling and curating a homology hypothesis, and choosing and implementing a method for phylogenetic inference.
-DateLife implements a workflow to streamline this process by leveraging on genetic data from the Barcode of Life Data System, BOLD [@ratnasingham2007bold] to obtain markers for input taxa.
+DateLife implements a workflow to streamline this process by leveraging on open data from the Barcode of Life Data System, BOLD [@ratnasingham2007bold] to obtain genetic markers for input taxa.
 <!--DNA sequences obtained from BOLD are short (between 500-800 bp) and display little variation. -->
-By default, mined genetic sequences are aligned with MUSCLE [@edgar2004muscle] using functions from the msa R package [@bodenhofer2015msa].
+By default, BOLD genetic sequences are aligned with MUSCLE [@edgar2004muscle] using functions from the msa R package [@bodenhofer2015msa].
 Alternatively, sequences can be aligned with MAFFT [@katoh2009multiple], using functions from the ape R package [@paradis2004].
@@
 Both parsimony and ML optimizations are done with functions from the `phangorn` package [@schliep2011phangorn].
-The BOLD workflow is currently only available on DateLife's R package and not on the web application.
+Due to the computing load it requires, the BOLD workflow is currently only supported through DateLife's R package. It is not yet available through the web application.
 
@@
 Supplementary material, including code, biological examples, benchmark results, data files and online-only appendices, can be found in the Dryad data repository (<https://doi.org/10.5061/dryad.cnp5hqc6w>), as well as in the Zenodo stable repositories that host the reproducible manuscript (<https://doi.org/10.5281/zenodo.7435094>), the biological examples (<https://doi.org/10.5281/zenodo.7435101>), and the software benchmark (<https://doi.org/10.5281/zenodo.7435106>).
-The corresponding development versions hosted on GitHub can be found at <https://github.com/LunaSare/datelifeMS1#readme>, <https://github.com/LunaSare/datelife_examples#readme>, and <https://github.com/LunaSare/datelife_benchmark#readme>.
+The corresponding development versions hosted on GitHub can be found at <https://github.com/LunaSare/datelifeMS1>, <https://github.com/LunaSare/datelife_examples>, and <https://github.com/LunaSare/datelife_benchmark>.
 

### 2023-02-21 notes-effects-downstream-analysis-review.Rmd.R

@@ -1,110 +1,0 @@
----
-title             : "Review: Effects of initial phylogenetic and chronogram hypothesis on evolutionary downstream analyses"
-shorttitle        : "Litt review"
-
-author:
-  - name          : "Luna L. Sánchez Reyes"
-    affiliation   : "1,2"
-    corresponding : yes    # Define only one corresponding author
-    address       : ""
-    email         : "sanchez.reyes.luna@gmail.com"
-    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
-      - Data curation
-      - Investigation
-      - Software
-      - Visualization
-      - Validation
-      - Writing
-      - Original Draft Preparation
-      - Writing
-      - Review & Editing
-
-affiliation:
-  - id            : "1"
-    institution   : "University of California, Merced, USA"
-  - id            : "2"
-    institution   : "University of Tennessee, Knoxville, USA"
-
-authornote: |
-  School of Natural Sciences, University of California, Merced, 258 Science and Engineering Building 1, Merced, CA 95340, USA.
-
-  Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 446 Hesler Biology Building, Knoxville, TN 37996, USA.
-
-bibliography      : ["paper_references.bib"]
-keywords          : "Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations"
-wordcount         : "`r wordcountaddin::word_count()`"
-floatsintext      : no
-figurelist        : no
-tablelist         : no
-footnotelist      : no
-linenumbers       : yes
-mask              : no
-draft             : no
-
-documentclass     : "apa6"
-classoption       : "man"
-output            :
-  papaja::apa6_pdf:
-    includes      :
-      in_header: "preamble.tex"
----
-
-```{r setup, include = FALSE}
-library("papaja")
-r_refs("paper_references.bib")
-embed_figure <- TRUE
-```
-
-@cusimano2012new: Simulating branch lengths following a birth-death species diversification model for
-missing taxa to chronograms with non random sampling, lowers type I error (false positive,
-incorrectly accepting the null hypothesis of a constant-rate or temporally varying rate
-birth-death model) when analyzing changes in diversification rate using the gamma statistic.
-
-@thomas2013pastis hypothesize that results of diversification analyses will be biased towards
-the birth-death model used to simulate branch lengths;
[…]

### 2023-03-11 manuscript/manuscript.Rmd
minor edits

@@
 
-  Chronograms –phylogenies with branch lengths proportional to time– represent key data on evolutionary time frame for the study of natural processes in many areas of biological research.
+  Chronograms –phylogenies with branch lengths proportional to time– represent key data on timing of evolutionary events for the study of natural processes in many areas of biological research.
   Chronograms also provide valuable information that can be used for education, science communication, and policy decisions.
@@
 Congruified node ages are used as secondary calibrations to date the chosen topology, with or without initial branch lengths, using different phylogenetic dating methods such as BLADJ, treePL, PATHd8 and MrBayes.
-    We performed a cross-validation test to compare node ages resulting from a DateLife analysis (i.e, phylogenetic dating using secondary calibrations) to those from the original chronograms (i.e, obtained with primary calibrations), and found that DateLife's node age estimates appear generally correlated to those from the original chronograms, with the largest variation in ages occurring around topologically deeper nodes.
-  Software such as DateLife can help increase awareness of the existing variation in alternative hypothesis of evolutionary time frame for the same organisms, and can support exploration of the effect of alternative chronogram hypotheses on downstream analyses, providing a framework for a more informed interpretation of evolutionary results.
+    We performed a cross-validation test to compare node ages resulting from a DateLife analysis (i.e, phylogenetic dating using secondary calibrations) to those from the original chronograms (i.e, obtained with primary calibrations), and found that DateLife's node age estimates are consistent with the age estimates from the original chronograms, with the largest variation in ages occurring around topologically deeper nodes.
+  Software such as DateLife can help increase awareness of the existing disparities among alternative hypotheses of dates for the same diversification events, and can support exploration of the effect of alternative chronogram hypotheses on downstream analyses, providing a framework for a more informed interpretation of evolutionary results.
 
@@
 It requires obtaining and curating a homology hypothesis to construct a phylogeny, selecting and placing appropriate calibrations on the phylogeny using independent age data points from the fossil record or other dated events, and inferring a full dated tree.
-All of this entails specialized biological training, taxonomic domain knowledge, and a non-negligible amount of research time, computational resources and funding.
+All of this entails specialized biological training, taxonomic domain knowledge, and a significant amount of research time, computational resources and funding.
 
@@
 DateLife's core software application is available as an R package [@sanchez2022], and as an online Rshiny interactive website at www.datelife.org.
-It features key elements for scientific reproducibility, such as a curated, versioned, open and fully public chronogram database [@mctavish2015phylesystem] that stores data in a computer-readable format [@vos2012nexml]; automated and programmatic ways of accessing and downloading the data in a computer-readable format also [@Stoltzfus2013]; and methods to summarize and compare the data.
+It features key elements for scientific reproducibility, such as a curated, versioned, open and fully public chronogram database [@mctavish2015phylesystem] that stores data in a computer-readable format [@vos2012nexml]; automated and programmatic ways of accessing and downloading the data in a computer-readable format [@Stoltzfus2013]; and methods to summarize and compare the data.
 
@@
 
-DateLife starts by processing an input consisting of at least one taxon scientific name. Multiple input names can be provided as a comma separated character string or as tip labels on a tree.
+DateLife starts by processing an input consisting of the scientific name of at least one taxon. Multiple input names can be provided as a comma separated character string or as tip labels on a tree.
 If the input is a tree, it can be provided as a classic newick character string [@archie1986newick], or as a "phylo" R object [@paradis2004].
@@
 The updated database is assigned a new version number, followed by a package release on CRAN.
-Users can directly implement `datelife` functions to trigger an update of the local chronogram database, to incorporate any new chronograms to their DateLife analysis before an official database update is released on CRAN.
+Users can directly run `datelife` functions to trigger an update of their local chronogram database, to incorporate any new chronograms to their DateLife analysis before an official database update is released on CRAN.
 <!--BCO: What is immediately?
@@
 **_With no branch lengths.--_**
-When producing or obtaining a tree with branch lengths for a group of interest is not possible, DateLife can date a topology without branch lengths, obtained from OpenTree or by implementing the supertree approach described above, by implementing the Branch Length Adjuster [BLADJ; @Webb2008; @webb2005phylomatic] algorithm, which requires no initial branch lengths.
+When producing or obtaining a tree with branch lengths for a group of interest is not possible, DateLife can date a topology without branch lengths, obtained from OpenTree or by implementing the supertree approach described above, by applying the Branch Length Adjuster [BLADJ; @Webb2008; @webb2005phylomatic] algorithm, which requires no initial branch lengths.
 The algorithm starts by fixing ages for nodes with calibration data upon the given topology.
-Then, it distributes time for nodes with no data evenly between calibrated nodes, minimizing age variance in the resulting chronogram [@Webb2008], which has proven useful for ecological analyses that require age data [@Webb2008].
+Then, it distributes time for nodes with no data evenly between calibrated nodes, minimizing age variance in the resulting chronogram [@Webb2008].
+This approach has has proven useful for ecological analyses that require age data [@Webb2008].
 When there is conflict between ages of calibrated nodes, BLADJ ignores node ages that are older than the age of a parent node.
@@
 
-In the absence of genetic data, we consider that BLADJ is the most agnostic way to assign ages to nodes with no available data, as it does not require any assumptions on the underlying model of branch length distribution.
+In the absence of genetic data, BLADJ is a very agnostic way to assign ages to nodes with no available data, as it does not require any assumptions on the underlying model of branch length distribution.
 It is however common practice in the literature to use a birth-death model to assign ages to nodes with no genetic data [@Jetz2012; @rabosky2018inverse; @smith2018constructing].
@@
 **_Dating a tree with branch lengths.--_**
-Relative branch lengths provide key information for phylogenetic dating, specifically for nodes without any calibration data available. 
-While, using initial branch lengths data is the golden standard for phylogenetic dating analyses, producing such requires assembling and curating a homology hypothesis, and choosing and implementing a method for phylogenetic inference.
-DateLife implements a workflow to streamline this process by leveraging on open data from the Barcode of Life Data System, BOLD [@ratnasingham2007bold] to obtain genetic markers for input taxa.
+Relative branch lengths can provide key information for phylogenetic dating, specifically for nodes without any calibration data available. 
+While, using initial branch lengths data is the golden standard for phylogenetic dating analyses, producing such trees requires assembling and curating a homology hypothesis, and choosing and implementing a method for phylogenetic inference.
+DateLife implements a workflow to streamline this process by applying open data from the Barcode of Life Data System, BOLD [@ratnasingham2007bold] to obtain genetic markers for input taxa.
 <!--DNA sequences obtained from BOLD are short (between 500-800 bp) and display little variation. -->
@@
 
-DateLife makes state-of-the-art data on evolutionary time frame easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
+DateLife makes state-of-the-art data on evolutionary time frame easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise.
 It is an open service that does not require any expert biological knowledge from users --besides the names of the species or group they want to work with, for any of its functionality.
@@
 Hence, DateLife's workflow features different ways of assigning node ages in the absence of calibrations and branch length information for certain taxa.
-Completing a phylogeny using a stochastic birth-death politomy resolver and a backbone taxonomy is a common practice in scientific publications: @Jetz2012, created a chronogram of all 9, 993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a chronogram of 31, 536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a chronogram of 353, 185 seed plants where only 23% had molecular data.
+Completing a phylogeny using a stochastic birth-death polytomy resolver and a backbone taxonomy is a common practice in scientific publications: @Jetz2012, created a chronogram of all 9, 993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a chronogram of 31, 536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a chronogram of 353, 185 seed plants where only 23% had molecular data.
 These stochastically resolved chronograms provide less biased diversification rates estimated with methods that account for sampling fractions [@chang2020estimating; @cusimano2012new], but can also introduce spurious patterns of early bursts of diversification [@cusimano2010slowdowns; @sun2020estimating].
 
-Taxonomy based politomy resolvers also introduce topological differences.
+Taxonomy based polytomy resolvers also introduce topological differences.
 The study of macroevolutionary processes largely depends on an understanding of the timing of species diversification events, and different phylogenetic and chronogram hypothesis can provide very different overviews of the macroevolutionary history of a biological group.
@@
 
-While stochastic politomy resolvers have been useful to advance research, notably, risks come with this practice.
+While stochastic polytomy resolvers have been useful to advance research, notably, risks come with this practice.
 Taken to the extreme, one could generate a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy, a single calibration, and assigning branch lengths following a birth-death diversification model.
@@
 It is also crucial for education, science communication and policy, but generating chronograms is difficult, especially for those who want to use phylogenies but who are not systematists, or do not have the time to acquire and develop the necessary knowledge and skills to construct them on their own.
-Importantly, years of primarily publicly funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
+Importantly, years of primarily publicly funded research have resulted in vast amounts of chronograms that are already available in scientific publications, but functionally hidden from the public and scientific community for reuse.
 
 <!--Mention again uses shown here and outline potential uses not shown here-->
-The DateLife project allows for easy and fast summary of public and state-of-the-art data on time of lineage divergence.
+The DateLife project allows for easy and fast summarization of public and state-of-the-art data on time of lineage divergence.
 It provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identifying regions that require more research, or that have conflicting information.
-It is available as an R package, and as a web-based R shiny application at www.datelife.org
+It is available as an R package, and as a web-based R shiny application at www.datelife.org.
 <!-- It can also summarize public data, or use it to generate new chronograms based on trees provided by the user. -->
 Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
-The DateLife project should improve awareness of the existing variation in expert time of divergence data, and foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
+The DateLife project will improve awareness of the existing variation in expert estimations time of divergence data, and foster exploration of the effect of alternative divergence time hypotheses on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
 
@@
 It can be used online through its R shiny web application hosted at <http://www.datelife.org>, and locally through the `datelife` R package, downloadable from Zenodo (<https://doi.org/10.5281/zenodo.593938>) and the CRAN repository [@sanchez2022].
-DateLife's web application is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker and OpenTree's infrastructure (dates.opentreeoflife.org/datelife).
+DateLife's web application is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker and OpenTree's infrastructure (datelife.opentreeoflife.org).
 `datelife`'s stable version `r availability` for installation from CRAN repository using the command `install.packages(pkgs = "datelife")` from within R.

### 2023-05-04 figures/fringillidae-topologies/figure4-caption.tex
add some more reviews

--- 2023-02-21 manuscript/manuscript-1st-submission.tex
@@
-\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[english,man]{apa6}
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provides euro and other symbols
-\else % if luatex or xelatex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage[]{microtype}
-\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
-\usepackage{hyperref}
-\hypersetup{
-            pdftitle={DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life},
-            pdfauthor={Luna L. Sánchez Reyes1,2, Emily Jane McTavish1, \& Brian O'Meara2},
-            pdfborder={0 0 0},
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
-\usepackage{graphicx,grffile}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\setlength{\emergencystretch}{3em}  % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{0}
-
-% set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-
-\DeclareUnicodeCharacter{0301}{*************************************}
-
-% leave the figure captions outside\ caption{} if you want them
-% to be formatted in the same way as the general text (double spaced and linenumbered)
-% line numbers for captions trick from https://latex.org/forum/viewtopic.php?t=34614
-
-\usepackage{ragged2e}
-
-\usepackage{caption}
-
-\DeclareCaptionFont{linenumbers}{\internallinenumbers}
-
-\usepackage{float}
-\restylefloat{table}
-
-\captionsetup[figure]{font+=linenumbers, labelfont={sc}, labelformat={default}, labelsep=period, name={Figure}}
-
-\captionsetup[table]{font+=linenumbers, labelfont={sc}, labelformat={default}, labelsep=period, name={Table}}
-% Manuscript styling
-\usepackage{upgreek}
-\captionsetup{font=singlespacing,justification=justified}
-
-% Table formatting
-\usepackage{longtable}
-\usepackage{lscape}
-% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
-\usepackage{multirow}		% Table styling
-\usepackage{tabularx}		% Control Column width
-\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
-\usepackage{threeparttablex}            % Lets threeparttable work with longtable
-
-% Create new environments so endfloat can handle them
-% \newenvironment{ltable}
-%   {\begin{landscape}\centering\begin{threeparttable}}
-%   {\end{threeparttable}\end{landscape}}
-\newenvironment{lltable}{\begin{landscape}\centering\begin{ThreePartTable}}{\end{ThreePartTable}\end{landscape}}
-
-% Enables adjusting longtable caption width to table width
-% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
-\makeatletter
-\newcommand\LastLTentrywidth{1em}
-\newlength\longtablewidth
-\setlength{\longtablewidth}{1in}
-\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}
-
-% \setlength{\parindent}{0.5in}
-% \setlength{\parskip}{0pt plus 0pt minus 0pt}
-
-% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
-% See https://github.com/crsh/papaja/issues/292
-\makeatletter
-\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-1em}%
-  {\normalfont\normalsize\bfseries\itshape\typesectitle}}
-
-\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-\z@\relax}%
-  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
-\makeatother
-
-% \usepackage{etoolbox}
-\makeatletter
-\patchcmd{\HyOrg@maketitle}
-  {\section{\normalfont\normalsize\abstractname}}
-  {\section*{\normalfont\normalsize\abstractname}}
-  {}{\typeout{Failed to patch abstract.}}
-\patchcmd{\HyOrg@maketitle}
-  {\section{\protect\normalfont{\@title}}}
-  {\section*{\protect\normalfont{\@title}}}
-  {}{\typeout{Failed to patch title.}}
-\makeatother
-
-\usepackage{xpatch}
-\makeatletter
-\xapptocmd\appendix
-  {\xapptocmd\section
-    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
-    {}{\InnerPatchFailed}%
-  }
-{}{\PatchFailed}
-\DeclareDelayedFloatFlavor{ThreePartTable}{table}
-\DeclareDelayedFloatFlavor{lltable}{table}
-\DeclareDelayedFloatFlavor*{longtable}{table}
-\makeatletter
-\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
-\makeatother
-\usepackage{lineno}
-
-\linenumbers
-\usepackage{csquotes}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[shorthands=off,main=english]{babel}
-\else
-  % load polyglossia as late as possible as it *could* call bidi if RTL lang (e.g. Hebrew or Arabic)
-  \usepackage{polyglossia}
-  \setmainlanguage[]{english}
-\fi
-
-\title{DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life}
-\author{Luna L. Sánchez Reyes\textsuperscript{1,2}, Emily Jane McTavish\textsuperscript{1}, \& Brian O'Meara\textsuperscript{2}}
-\date{}
-
-\shorttitle{DateLife: revealing the dated Tree of Life}
-
-\authornote{
-
-School of Natural Sciences, University of California, Merced, 258 Science and Engineering Building 1, Merced, CA 95340, USA.
-
-Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 446 Hesler Biology Building, Knoxville, TN 37996, USA.
-
-The authors made the following contributions.
-Luna L.
-Sánchez Reyes: Data curation, Investigation, Software, Visualization, Validation, Writing - Original Draft Preparation, Writing - Review \& Editing; Emily Jane McTavish: Resources, Software, Writing - Review \& Editing; Brian O'Meara: Conceptualization, Funding acquisition, Methodology, Resources, Software, Supervision, Writing - Review \& Editing.
-
-Correspondence concerning this article should be addressed to Luna L. Sánchez Reyes, . E-mail: \href{mailto:sanchez.reyes.luna@gmail.com}{\nolinkurl{sanchez.reyes.luna@gmail.com}}
-
-}
-
-\affiliation{\vspace{0.5cm}\textsuperscript{1} University of California, Merced, USA\\\textsuperscript{2} University of Tennessee, Knoxville, USA}
-
-\begin{document}
-\maketitle
-
-\begin{center}
-Abstract
-\end{center}
-
-Achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram) is a difficult and time-consuming task.
-But the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
-Knowledge of the evolutionary time frame of organisms is key for research in the natural sciences. It also represent valuable information for education, science communication, and policy decisions.
-When chronograms are shared in public and open databases, this wealth of expertly-curated and peer-reviewed data on evolutionary timeframe is exposed in a programatic and reusable way, as intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
-Here we present DateLife, a service implemented as an R package and an R Shiny website application available at www.datelife.org, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time frame of evolution.
-The main DateLife workflow constructs a chronogram for any given combination of taxon names by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from the TreeBASE database as well.
-We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
-DateLife will be useful to increase awareness of the existing variation in alternative hypothesis of evolutionary time for the same organisms, and can foster exploration of the effect of alternative evolutionary timing hypotheses on the results of downstream analyses, providing a framework for a more informed interpretation of evolutionary results.
-
-\emph{Keywords}: Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations
-
-Word count: 5393
-
-\newpage
-
-Chronograms --phylogenies with branch lengths proportional to time-- provide key data on evolutionary time frame for the study of natural processes in many areas of biological research, such as developmental biology (Delsuc et al., 2018; Laubichler \& Maienschein, 2009), conservation biology (Felsenstein, 1985; Webb, 2000), historical biogeography (Posadas, Crisci, \& Katinas, 2006), and species diversification (Magallon \& Sanderson, 2001; Morlon, 2014).
-
-Building a chronogram is not an easy task.
-It requires obtaining and curating data to construct a phylogeny, selecting and placing appropriate calibrations on the phylogeny using independent age data points from the fossil record or other dated events, and inferring the full dated tree;
-it also generally requires specialized biological training, taxonomic domain knowledge, and a non-negligible amount of research time, computational resources and funding.
-
-Here we present the DateLife project which has the main goal of capturing age data from published chronograms, and making these data readily accessible to the community for reuse and reanalysis, for research, teaching, and science communication and policy.
-DateLife's core software application is available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org.
-It features key elements for scientific reproducibility, such as a versioned, open and fully public source database (McTavish et al., 2015), data stored and available in a computer readable format (Vos et al., 2012), automated and programmatic ways of accessing the data (Stoltzfus et al., 2013) and methods to summarize and compare the data.
-
-\begin{center}
-\textsc{Description}
-\end{center}
-
-DateLife's core software application consists of the R package \texttt{datelife}.
-Its current stable version -- v0.6.6, is available from The Comprehensive R Archive Network (CRAN) repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
-ape (Paradis, Claude, \& Strimmer, 2004),
-bold (Chamberlain, 2018),
-geiger (Pennell et al., 2014),
-paleotree (Bapst, 2012),
-phyloch (Heibl, 2008),
-phylocomr (Ooms \& Chamberlain, 2018),
-phytools (Revell, 2012),
-rotl (Michonneau, Brown, \& Winter, 2016), and
-taxize (Chamberlain, 2018; Chamberlain \& Szöcs, 2013).
-Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife workflow: creating a search query, searching a database, and summarizing results from the search.
-
-\begin{center}
-\emph{Creating a Search Query}
-\end{center}
-
-DateLife starts by processing an input consisting of at least two taxon names, which can be provided as a comma separated character string or as tip labels on a tree.
-If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a \enquote{phylo} R object (Paradis et al., 2004).
-The input tree is not required to have branch lengths, and its topology is used in the summary steps described in the next section.
-
-DateLife accepts scientific names that can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or a binomial species name.
-Subspecies and variants are ignored.
-If an input taxon name belongs to an inclusive taxonomic group, DateLife has two alternative behaviors defined by the \enquote{get species from taxon} flag.
-If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group following a standard taxonomy of choice, and adds them to the input string.
-Taxonomies currently supported by DateLife are Open Tree of Life (OpenTree) unified Taxonomy (OTT, Rees \& Cranston, 2017), the National Center of Biotechnology Information (NCBI) taxonomic database (Schoch et al., 2020), the Global Biodiversity Information Facility (GBIF) taxonomic backbone (GBIF Secretariat, 2022), and the Interim Register of Marine and Nonmarine Genera (IRMNG) database (Rees et al., 2017).
-If the flag is inactive, DateLife excludes any taxon names above the species level from the search query.
-
-DateLife processes input scientific names using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database.
-TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard (Boyle et al., 2013).
-DateLife implements TNRS with OTT as standard (Open Tree Of Life et al., 2016; Rees \& Cranston, 2017), storing taxonomic identification numbers for further processing.
-
-The processed input taxon names are saved as an R object of a newly defined class, \texttt{datelifeQuery}, that is used in the following steps.
-This object contains the standardized names, the corresponding OTT identification numbers, and the topology of the input tree if any was provided.
-
-\begin{center}
-\emph{Searching a Chronogram Database}
-\end{center}
-
-At the time of writing of this manuscript (Jun 22, 2022),
-DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies.
-It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata (McTavish et al., 2015), which allows automatic and reproducible assembly of our chronogram database.
-Datelife's chronogram database is navigable as an R data object within the \texttt{datelife} R package.
-
-A unique feature of the Phylesystem is that any user can add new published, state-of-the-art chronograms any time, through their curator application (\url{https://tree.opentreeoflife.org/curator}).
-As chronograms are added to Phylesystem, they are incorporated into the chronogram database of the \texttt{datelife} package.
-The updated database is assigned a new version number, followed by a package release on CRAN.
-\texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
-Users can also implement functions from the \texttt{datelife} R package to trigger an update of the local chronogram database, to incorporate any new chronograms to the user's DateLife analysis before an official database update is released on CRAN.
-
-A DateLife search is implemented by matching processed taxon names provided by the user to tip labels in the chronogram database.
-Chronograms with at least two matching taxon names on their tip labels are identified and pruned down to preserve only the matched taxa.
-These matching pruned chronograms are referred to as source chronograms.
-Total distance (in units of millions of years) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Figure \ref{fig:workflow}).
-The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a \enquote{phylo} object or newick string.
-Finally, the patristic matrices are associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.
-
-\begin{center}
-\emph{Summarizing Search Results}
-\end{center}
-
-Summary information is extracted from the \texttt{datelifeResult} object to inform decisions for subsequent steps in the analysis workflow. Basic summary information available to the user is:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  The matching pruned chronograms as newick strings or \enquote{phylo} objects.
-\item
-  The ages of the root of all source chronograms.
-These ages can correspond to the age of the most recent common ancestor (mrca) of the user's group of interest if the source chronograms have all taxa belonging to the group.
-If not, the root corresponds to the mrca of a subgroup withing the group of interest.
-\item
-  Study citations where original chronograms were published.
-\item
-  A report of input taxon names matches across source chronograms.
-\item
-  The source chronogram(s) with the most input taxon names.
-\item
-  Various single summary chronograms resulting from summarizing age data, generated using the methodology described next.
-\end{enumerate}
-
-\begin{center}
-\emph{Choosing a Topology}
-\end{center}
-
-DateLife requires a tree topology to summarize age data upon.
-We recommend that users provide as input a tree topology from the literature, or one of their own making.
-If no topology is provided, DateLife automatically extracts one from the OpenTree synthetic tree, a phylogeny encompassing 2.3 million taxa across all life, assembled from 1, 239 published phylogenetic trees and OpenTree's unified Taxonomy, OTT (Open Tree Of Life et al., 2019).
-Alternatively, DateLife can combine topologies from source chronograms using a supertree approach.
-To combine topologies from source chronograms into a single summary (or supertree) topology, the DateLife workflow identifies the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
-(2009).
-In rare cases, a group of trees can have multiple groves.
-By default, DateLife chooses the grove with the most taxa, however, the \enquote{criterion = trees} flag allows the user to choose the grove with the most trees instead.
-If source chronograms do not form a grove, the supertree reconstruction will fail.
-
-\begin{center}
-\emph{Dating the Topology}
-\end{center}
-
-Input topologies from OpenTree or the supertree approach described above do not include branch length estimates of any kind.
-Optionally, to estimate branch lengths proportional to substitution rates on these topologies, DateLife can mine the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain genetic markers for the input taxa.
-These markers are aligned with MUSCLE (Edgar, 2004) (by default) or MAFFT (Katoh, Asimenos, \& Toh, 2009).
-This alignment can be used to estimate branch lengths on input topologies that lack branch lengths.
-Currently, branch length reconstruction in DateLife is performed using parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed (Schliep, 2011).
-While relative branch length information provides additional data for nodes without secondary date calibrations, topologies without branch lengths can also be dated.
-
-Once a topology is chosen, DateLife applies the congruification method (Eastman, Harmon, \& Tank, 2013) to find nodes belonging to the same clade across source chronograms, and extract the corresponding node ages from the patristic distance matrices stored as \texttt{datelifeResult}.
-By definition, the matrices store total distance (time from tip to tip), hence, node ages correspond to half the values stored in the patristic distance matrices.
-This assumes that the terminal taxa are coeval and occur at the present.
-A table of congruified node ages that can be used as calibrations for a dating analysis is stored as a \texttt{congruifiedCalibrations} object.
-
-For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method (SDM; Criscuolo, Berry, Douzery, \& Gascuel, 2006), which deforms patristic distance matrices by minimizing variance and then averaging them.
-These single summary taxon pair age matrices (Summarized calibrations) can be applied as calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Webb, Ackerly, \& Kembel, 2008; Webb \& Donoghue, 2005), and treePL (Smith \& O'Meara, 2012).
-
-By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully dated topology.
-BLADJ fixes node ages that have calibration data, and distributes time between nodes with no data evenly between nodes with calibration data.
-This minimizes age variance in the resulting chronogram (Webb et al., 2008).
-BLADJ does not use branch lengths even when they are present in the input tree or summarizing topology.
-When there is conflict in ages between nodes with calibration data, BLADJ ignores node ages that are older than the age of a parent node.
-BLADJ requires a root age estimate.
-If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature.
-If none is provided, DateLife assigns an arbitrary age to the root as 10\% older than the oldest age available within the group.
-
-Alternative phylogenetic dating options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the calibrations.
-PATHd8 is a non-clock, rate-smoothing method (Britton et al., 2007) to date trees.
-treePL (Smith \& O'Meara, 2012), is a semi-parametric, rate-smoothing, penalized likelihood dating method (Sanderson, 2002).
-The MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003) approach in DateLife uses the calibrations as priors on node ages.
-
-\begin{center}
-\emph{Visualizing Results}
-\end{center}
-
-Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R \enquote{phylo} format), as well as visualize and compare results graphically, or construct their own graphs using DateLife's chronogram plot generation functions available from the R package \texttt{datelifeplot} (Sanchez-Reyes \& O'Meara, 2022).
-
-\newpage
-
-\begin{center}
-\textsc{Benchmark}
-\end{center}
-
-\texttt{datelife}'s R package code speed was tested on an Apple iMac
-with one 3.4 GHz Intel Core i5 processor.
-We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
-Query processing time increases roughly linearly with number of input taxon names, and
-increases considerably if Taxonomic Name Resolution Service (TNRS) is activated.
-Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
-Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig.
-\ref{fig:runtime_main}).
-
-\texttt{datelife}'s code performance was evaluated with a set of unit tests designed and
-implemented with the R package testthat (R Core Team, 2018) that were run both locally
-with the devtools package (R Core Team, 2018), and on a public server using the continuous integration tool of GitHub actions (\url{https://docs.github.com/en/actions}).
-At present, unit tests cover more than 40\% of \texttt{datelife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).
-Unit testing helps identify potential issues as code is updated or, more critically, as services code relies upon may change.
-
-\begin{center}
-\textsc{Case Studies}
-\end{center}
-
-We illustrate the DateLife workflow using a family within the Passeriform birds encompassing the true finches, Fringillidae, as case study.
-On a small example, we analysed 6 bird species, and results from each step of the workflow are shown in Fig.
-\ref{fig:figure2}.
-As a second example, we analysed 289 bird species in the family Fringillidae that are included in the NCBI taxonomy.
-The resulting summary chronogram is shown in Fig.
-\ref{fig:fringillidages}, and results from previous steps of the workflow are available as Supplementary Figures.
-
-\begin{center}
-\emph{A Small Example}
-\end{center}
-
-\textbf{\emph{Creating a search query.--}}
-We chose 6 bird species within the Passeriformes. The sample includes
-two species of cardinals:
-the black-thighed grosbeak -- \emph{Pheucticus tibialis} and
-the crimson-collared grosbeak -- \emph{Rhodothraupis celaeno};
-three species of buntings:
-the yellowhammer -- \emph{Emberiza citrinella},
-the pine bunting -- \emph{Emberiza leucocephalos} and
-the yellow-throated bunting -- \emph{Emberiza elegans};
-and one species of tanager, the vegetarian finch -- \emph{Platyspiza crassirostris}.
-Processing of input names found that \emph{Emberiza elegans} is synonym for \emph{Schoeniclus elegans} in the default reference taxonomy (OTT v3.3, June 1, 2021).
-For a detailed discussion on the state of the synonym, refer to Avibase (Avibase, 2022; Lepage, 2004; Lepage, Vaidya, \& Guralnick, 2014).
-Discovering this synonym allowed assigning five age data points for the parent node of \emph{Emberiza elegans}, shown as \emph{Schoeniclus elegans} in figure \ref{fig:figure2}A, which would not have had any data otherwise.
-
-\newpage
-
-\textbf{\emph{Searching the database.--}}
-DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies (Fig.
-\ref{fig:figure2}B).
-Three studies matched five input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2015; Hedges, Marin, Suleski, Paymer, \& Kumar, 2015; Jetz, Thomas, Joy, Hartmann, \& Mooers, 2012), one study matched four input names (Hooper \& Price, 2017) and two studies matched two input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2013; Burns et al., 2014).
-No studies matched all input names.
-Together, source chronograms provide 28 unique age data points, covering all nodes on our chosen tree topology to date (Table \ref{tbl:table1}).
-
-\vspace{3mm}
-
-\textbf{\emph{Summarizing search results.--}}
-DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and congruified and mapped age data to nodes in this chosen topology (Table \ref{tbl:table1}).
-The name processing step allowed including five data points for node \enquote{n4} (parent of \emph{Schoeniclus elegans}; Fig.
-\ref{fig:figure2}A) that would not have had any data otherwise due to name mismatch.
-Age summary statistics per node were calculated (Table \ref{tbl:table2}) and used as calibrations to date the tree topology using the BLADJ algorithm.
-As expected, more inclusive nodes (e.g., node \enquote{n1}) have more variance in age data than less inclusive nodes (e.g., node \enquote{n5}).
-Summary age data for node \enquote{n2} were excluded as final calibration because they are older than age data of the more inclusive node, \enquote{n1} (Fig. \ref{fig:figure2}C4).
-
-\newpage
-
-\newpage
-
-\begin{center}
-\emph{An Example with the Family of True Finches}
-\end{center}
-
-\textbf{\emph{Creating a query.--}}
-To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the \enquote{get species from taxon} flag,
-which gets all recognized species names within a named group from a taxonomy of choice.
-Following the NCBI taxonomy, our DateLife query has 289 Fringillidae species.
-This taxon-constrained approach implies that the final results of a full DateLife analysis will be done using a tree topology and ages for the species in a named group, which do not necessarily correspond to a monophyletic group.
-Users can change this behaviour by providing a monophyletic tree as input for a DateLife search, or as a tree topology for a DateLife summary.
-
-\textbf{\emph{Searching the database.--}}
-Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species.
-The DateLife search identified 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Price et al., 2014).
-Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementray Table S1).
-
-\textbf{\emph{Summarizing search results.--}}
-The final step is to congruify and summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM.
-As explained in the \enquote{Description} section, a tree topology to summarize age data upon is required.
-By default, DateLife uses the topology from OpenTree's synthetic tree that contains the species in the search query to summarize age data upon.
-According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).
-
-\vspace{50mm}
-
-Age data from source chronograms was congruified to OpenTree's topology (Fig.
-\ref{fig:fringillidae-topologies}B), reducing the age data set to 818 different data points (Supplementray Table S2).
-For each congruent node, age summary statistics were calculated and used as fixed secondary calibrations over the chosen tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
-\ref{fig:fringillidages}).
-
-\begin{center}
-\textsc{Cross-Validation Test}
-\end{center}
-
-We performed a cross validation analysis of the DateLife workflow using the Fringillidae chronograms.
-We used the individual tree topologies from each of the 19 source chronograms from 13 studies as inputs, treating their node ages as unknown.
-We then estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.
-We found that node ages from original study, and ages estimated using all other age data available are correlated (Fig. \ref{fig:cvXY}).
-For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka \enquote{closer to the root}) relative to the original estimate, and overestimate ages for nodes closer to the tips.
-Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. S1).
-
-\begin{center}
-\textsc{Discussion}
-\end{center}
-
-DateLife makes state-of-the-art data on evolutionary time frame easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
-It is an open service that does not require any expert biological knowledge from users --besides the names of the species or group they want to work with, for any of its functionality.
-
-A total of 99,474 unique terminal taxa are represented in DateLife's database.
-Incorporation of more chronograms into the database will continue to improve DateLife's services.
-One option to increase the number of chronograms in the DateLife database is the Dryad data repository.
-Methods to automatically mine chronograms from Dryad could be designed and implemented.
-However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined manually by a human curator.
-We would like to emphasize on the importance of sharing chronogram data, including systematically curated metadata, into open repositories, such as OpenTree's Phylesystem (McTavish et al., 2015) for the benefit of the scientific community as a whole.
-
-\begin{center}
-\emph{Age Variation in Source Chronograms}
-\end{center}
-
-Conflict in estimated ages among alternative studies is common in the literature.
-See, for example, the robust ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández, 2015; Ramshaw et al., 1972; Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021).
-Source chronograms available for the same organisms have potentially been estimated implementing calibrations very differently.
-For example, the chronograms from Burns et al.
-(2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older age estimates for the same nodes than chronograms that were inferred using fossils as calibrations (Figs.
-\ref{fig:fringillidages}, \ref{fig:cvXY}; Supplementary Figs.
-S1, S5).
-
-Different calibration implementations might also imply fundamentally distinct evolutionary hypotheses (Antonelli et al., 2017).
-For example, two independent researchers working on the same clade should both carefully select and justify their choices of fossil calibration placement.
-Yet, if one researcher concludes that a fossil should calibrate the ingroup of a clade, while another researcher concludes that the same fossil should calibrate the outgroup of the clade, the resulting age estimates will differ, as the placement of calibrations as stem or crown group has been proven to significantly affect time of lineage divergence estimates (Sauquet, 2013).
-
-\begin{center}
-\emph{Primary vs Secondary Calibrations}
-\end{center}
-
-While most chronograms in DateLife's database are constructed using primary calibrations (molecular substitution rates or ages obtained from the fossil record or geological events), DateLife summarizes chronograms using secondary calibrations (ages coming from other chronograms).
-Graur and Martin (2004) cautioned on the increased error and uncertainty in estimated ages when using secondary calibrations in dating analyses.
-Schenk (2016) showed that, in simulations, divergence times inferred using secondary calibrations are significantly younger than those inferred with primary calibrations, when obtained with Bayesian inference methods, and when priors are implemented in similar ways in both analyses.
-Accordingly, the scientific community seems to have more confidence in chronograms obtained from a single analysis, using fossil data as primary sources of calibrations (Schenk, 2016), and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data reflect a coherent evolutionary history (Sauquet, 2013), as for example done by Antonelli et al.
-(2017).
-There have been attempts to create fossil calibration databases (Ksepka et al., 2015), though these still have room to grow.
-
-It seems that using several (as opposed to just a few) secondary calibrations can provide sufficient information to alleviate or even neutralize potential biases (Sauquet, 2013).
-Certainly, further studies are required to fully understand the effect of secondary calibrations on outputs from different tree dating methods, and on downstream analyses.
-It is possible that secondary calibrations can be safely used with dating methods that do not require setting priors, such as penalized likelihood (Sanderson, 2003), with methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ (Webb et al., 2008; Webb \& Donoghue, 2005), or methods that summarize age data unto a tree topology.
-
-Our cross validation analysis might provide some insight in this regard.
-When ages are estimated with secondary calibrations, nodes closer to the root do tend to be slightly younger than ages estimated with primary calibrations.
-However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures S2-S20).
-The only excpetion to tjis was observed on results of the cross validation analysis of the Burns et al.
-(2014) chronogram, which results in much younger node ages when estimated using secondary calibrations (Supplementary Figs.
-S1, S5).
-
-\begin{center}
-\emph{Sumarizing Chronograms}
-\end{center}
-
-By default, DateLife currently summarizes all source chronograms that overlap with at least two species names. Users can exclude source chronograms if they have reasons to do so.
-Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely.
-To our knowledge, there are no tested measures to determine independently when a chronogram is better than another.
-Yet, several characteristics of the data used for dating analyses, as well as from the output chronogram itself, could be used to score the quality of source chronograms.
-
-Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used (Magallón, 2010; Magallón et al., 2015).
-Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
-
-DateLife provides an opportunity to capture concordance and conflict among date estimates, which can also be used as a metric for chronogram reliability.
-Its open database of chronograms allows other researchers to do such analyses themselves reproducibly, and without needing permission.
-Though, of course, they should follow proper citation practices, especially for the source chronogram studies.
-
-The exercise of summarizing age data from across multiple studies provides the opportunity to work with a more inclusive chronogram,
-that reflects a unified evolutionary history for a lineage, by putting together evidence from different hypotheses.
-The largest, and taxonomically broadest chronogram currently available from OpenTree was constructed summarizing age data from 2,274 published chronograms using NCBI's taxonomic tree as backbone (Hedges et al., 2015).
-A summarizing exercise may also amplify the effect of uncertainty and errors in source data, and blur parts of the evolutionary history of a lineage that might only be reflected in source chronograms and lost on the summary chronogram (Sauquet et al., 2021).
-
-\begin{center}
-\emph{Effects on Downstream Analyses}
-\end{center}
-
-For downstream analyses, using alternative chronogram may deeply affect our inferences (Title \& Rabosky, 2016), particularly when studying phenomena dependent on the timing of species diversification events, such as macroevolutionary processes.
-
-In ecology and conservation biology, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance (Webb et al., 2008).
-Hence, DateLife's workflow features different ways of estimating node ages in the absence of calibrations and branch length information for certain taxa.
-\enquote{Making up} branch lengths is a common practice in scientific publications: Jetz et al.
-(2012), created a chronogram of all 9, 993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
-(2018) created a chronogram of 31, 536 ray-finned fishes, of which only 37\% had molecular data; Smith and Brown (2018) constructed a chronogram of 353, 185 seed plants where only 23\% had molecular data.
-
-Notably, risks come with this practice.
-Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration, using polytomy resolution and branch estimation methods.
-There has yet to be a thorough analysis of what can go wrong when one extends inferences beyond the data in this way, so we urge caution; we also urge readers to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, and assessing the consistency of the results with prior work.
-
-\begin{center}
-\textsc{Conclusions}
-\end{center}
-
-Knowledge of the evolutionary time frame of organisms is key to many research areas: trait evolution,
-species diversification, biogeography, macroecology and more. It is also crucial for education, science communication and policy, but generating chronograms is difficult,
-especially for those who want to use phylogenies but who are not systematists, or
-do not have the time to acquire and develop the necessary knowledge and skills to construct them on their own.
-Importantly, years of primarily public funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
-
-The DateLife project allows for easy and fast summary of public and state-of-the-art data on time of lineage divergence.
-It provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identification of regions that require more research, or that have conflicting information.
-It is available as an R package, and as a web-based R shiny application at www.datelife.org
-Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
-The DateLife project helps with awareness of the existing variation in expert time of divergence data, and will foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
-
-\begin{center}
-\textsc{Availability}
-\end{center}
-
-The DateLife software is free and open source and it can be used through its R shiny web application at \url{http://www.datelife.org}, through the \texttt{datelife} R package, and through Phylotastic's project web portal \url{https://phylo.cs.nmsu.edu/}.
-DateLife's web application is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker and OpenTree's infrastructure (dates.opentreeoflife.org/datelife).
-\texttt{datelife}'s R package stable version is available for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife}) using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R.
-Development versions are available from the GitHub repository (\url{https://github.com/phylotastic/datelife}) and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.
-
-\begin{center}
-\textsc{Supplementary Material}
-\end{center}
-
-Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities are available at \href{https://github.com/LunaSare/datelifeMS1}{datelifeMS1}, \href{https://github.com/LunaSare/datelife_examples}{datelife\_examples}, and \href{https://github.com/LunaSare/datelife_benchmark}{datelife\_benchmark} repositories in LLSR's GitHub account.
-
-\begin{center}
-\textsc{Funding}
-\end{center}
-
-Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to the Datelife project; DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), ABI-1458572 to the Phylotastic project, and ABI-1759846 to the Open Tree of Life project.
-
-\begin{center}
-\textsc{Acknowledgements}
-\end{center}
-
-The DateLife project was born as a prototype tool aiming to provide these services, and was initially developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA (Stoltzfus et al., 2013).
-We thank colleagues from the O'Meara Lab at the University
-of Tennesse Knoxville for suggestions, discussions and software testing.
-The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
-that led to initial work on this project.
-The team that assembled DateLife's first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
-The Open Tree of Life project that provides the open, metadata rich repository of
-trees used to construct DateLife's chronogram database.
-The many scientists who publish their chronograms in an open, reusable form, and the scientists who curate them for deposition in the Open Tree of Life repository.
-The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.
-
-\newpage
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-\hypertarget{refs}{}
-\leavevmode\hypertarget{ref-alstrom2014discovery}{}%
-Alström, P., Hooper, D.
-M., Liu, Y., Olsson, U., Mohan, D., Gelang, M., \ldots{} Price, T.
-D.
-(2014).
-Discovery of a relict lineage and monotypic family of passerine birds.
-\emph{Biology Letters}, \emph{10}(3), 20131067.
-
-\leavevmode\hypertarget{ref-ane2009groves}{}%
-Ané, C., Eulenstein, O., Piaggio-Talice, R., \& Sanderson, M. J. (2009). Groves of phylogenetic trees. \emph{Annals of Combinatorics}, \emph{13}(2), 139--167.
-
-\leavevmode\hypertarget{ref-antonelli2017supersmart}{}%
-Antonelli, A., Hettling, H., Condamine, F.
-L., Vos, K., Nilsson, R.
-H., Sanderson, M.
-J., \ldots{} Vos, R.
-A.
-(2017).
-Toward a self-updating platform for estimating rates of speciation and migration, ages, and relationships of Taxa.
-\emph{Systematic Biology}, \emph{66}(2), 153--166.
-\url{https://doi.org/10.1093/sysbio/syw066}
-
-\leavevmode\hypertarget{ref-archie1986newick}{}%
-Archie, J., Day, W.
-H., Felsenstein, J., Maddison, W., Meacham, C., Rohlf, F.
-J., \& Swofford, D.
-(1986).
-The Newick tree format.
-Retrieved from \href{\%7Bhttps://evolution.genetics.washington.edu/phylip/newicktree.html\%7D}{\{https://evolution.genetics.washington.edu/phylip/newicktree.html\}}
-
-\leavevmode\hypertarget{ref-avibase-emberiza}{}%
-Avibase.
-(2022).
-Yellow-throated Bunting.
-\emph{Avibase - the World Bird Database}, (Online Resource).
-Retrieved from \href{\%7Bhttps://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\%7D}{\{https://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\}}
-
-\leavevmode\hypertarget{ref-Bapst2012a}{}%
-Bapst, D.
-W.
-(2012).
-Paleotree: An R package for paleontological and phylogenetic analyses of evolution.
-\emph{Methods in Ecology and Evolution}, \emph{3}(5), 803--807.
-\url{https://doi.org/10.1111/j.2041-210X.2012.00223.x}
-
-\leavevmode\hypertarget{ref-barba2018constraining}{}%
-Barba-Montoya, J., Reis, M.
-dos, Schneider, H., Donoghue, P.
-C., \& Yang, Z.
-(2018).
-Constraining uncertainty in the timescale of angiosperm evolution and the veracity of a cretaceous terrestrial revolution.
-\emph{New Phytologist}, \emph{218}(2), 819--834.
-
-\leavevmode\hypertarget{ref-barker2014mitogenomic}{}%
-Barker, F. K. (2014). Mitogenomic data resolve basal relationships among passeriform and passeridan birds. \emph{Molecular Phylogenetics and Evolution}, \emph{79}, 313--324.
-
-\leavevmode\hypertarget{ref-barker2013going}{}%
-Barker, F.
-K., Burns, K.
-J., Klicka, J., Lanyon, S.
-M., \& Lovette, I.
-J.
-(2013).
-Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds.
-\emph{Systematic Biology}, \emph{62}(2), 298--320.
-
-\leavevmode\hypertarget{ref-barker2015new}{}%
-Barker, F.
-K., Burns, K.
-J., Klicka, J., Lanyon, S.
-M., \& Lovette, I.
-J.
-(2015).
-New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
-\emph{The Auk: Ornithological Advances}, \emph{132}(2), 333--348.
-
-\leavevmode\hypertarget{ref-barker2004phylogeny}{}%
-Barker, F.
-K., Cibois, A., Schikler, P., Feinstein, J., \& Cracraft, J.
-(2004).
-Phylogeny and diversification of the largest avian radiation.
-\emph{Proceedings of the National Academy of Sciences}, \emph{101}(30), 11040--11045.
-
-\leavevmode\hypertarget{ref-beresford2005african}{}%
-Beresford, P., Barker, F., Ryan, P., \& Crowe, T.
-(2005).
-African endemics span the tree of songbirds (passeri): Molecular systematics of several evolutionary ``enigmas''.
-\emph{Proceedings of the Royal Society B: Biological Sciences}, \emph{272}(1565), 849--858.
-
-\leavevmode\hypertarget{ref-Boyle2013}{}%
-Boyle, B., Hopkins, N., Lu, Z., Raygoza Garay, J.
-A., Mozzherin, D., Rees, T., \ldots{} Enquist, B.
-J.
-(2013).
-The taxonomic name resolution service: An online tool for automated standardization of plant names.
-\emph{BMC Bioinformatics}, \emph{14}(1).
-\url{https://doi.org/10.1186/1471-2105-14-16}
-
-\leavevmode\hypertarget{ref-Britton2007}{}%
-Britton, T., Anderson, C.
-L., Jacquet, D., Lundqvist, S., \& Bremer, K.
-(2007).
-Estimating Divergence Times in Large Phylogenetic Trees.
-\emph{Systematic Biology}, \emph{56}(788777878), 741--752.
-\url{https://doi.org/10.1080/10635150701613783}
-
-\leavevmode\hypertarget{ref-bryson2014diversification}{}%
-Bryson Jr, R.
-W., Chaves, J., Smith, B.
-T., Miller, M.
-J., Winker, K., Pérez-Emán, J.
-L., \& Klicka, J.
-(2014).
-Diversification across the new world within the `blue'cardinalids (aves: Cardinalidae).
-\emph{Journal of Biogeography}, \emph{41}(3), 587--599.
-
-\leavevmode\hypertarget{ref-burleigh2015building}{}%
-Burleigh, J. G., Kimball, R. T., \& Braun, E. L. (2015). Building the avian tree of life using a large-scale, sparse supermatrix. \emph{Molecular Phylogenetics and Evolution}, \emph{84}, 53--63.
-
-\leavevmode\hypertarget{ref-burns2014phylogenetics}{}%
-Burns, K.
-J., Shultz, A.
-J., Title, P.
-O., Mason, N.
-A., Barker, F.
-K., Klicka, J., \ldots{} Lovette, I.
-J.
-(2014).
-Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds.
-\emph{Molecular Phylogenetics and Evolution}, \emph{75}, 41--77.
-
-\leavevmode\hypertarget{ref-Chamberlain2018}{}%
-Chamberlain, S. (2018). \emph{bold: Interface to Bold Systems API}. Retrieved from \url{https://CRAN.R-project.org/package=bold}
-
-\leavevmode\hypertarget{ref-Chamberlain2013}{}%
-Chamberlain, S.
-A., \& Szöcs, E.
-(2013).
-taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}.
-\emph{F1000Research}, \emph{2}(191), 1--29.
-\url{https://doi.org/10.12688/f1000research.2-191.v2}
-
-\leavevmode\hypertarget{ref-chaves2013biogeography}{}%
-Chaves, J.
-A., Hidalgo, J.
-R., \& Klicka, J.
-(2013).
-Biogeography and evolutionary history of the n eotropical genus s altator (a ves: T hraupini).
-\emph{Journal of Biogeography}, \emph{40}(11), 2180--2190.
-
-\leavevmode\hypertarget{ref-claramunt2015new}{}%
-Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{Science Advances}, \emph{1}(11), e1501005.
-
-\leavevmode\hypertarget{ref-Criscuolo2006}{}%
-Criscuolo, A., Berry, V., Douzery, E.
-J., \& Gascuel, O.
-(2006).
-SDM: A fast distance-based approach for (super)tree building in phylogenomics.
-\emph{Systematic Biology}, \emph{55}(5), 740--755.
-\url{https://doi.org/10.1080/10635150600969872}
-
-\leavevmode\hypertarget{ref-delsuc2018phylogenomic}{}%
-Delsuc, F., Philippe, H., Tsagkogeorga, G., Simion, P., Tilak, M.-K., Turon, X., \ldots{} Douzery, E.
-J.
-(2018).
-A phylogenomic framework and timescale for comparative studies of tunicates.
-\emph{BMC Biology}, \emph{16}(1), 1--14.
-
-\leavevmode\hypertarget{ref-Eastman2013}{}%
-Eastman, J.
-M., Harmon, L.
-J., \& Tank, D.
-C.
-(2013).
-Congruification: Support for time scaling large phylogenetic trees.
-\emph{Methods in Ecology and Evolution}, \emph{4}(7), 688--691.
-\url{https://doi.org/10.1111/2041-210X.12051}
-
-\leavevmode\hypertarget{ref-edgar2004muscle}{}%
-Edgar, R. C. (2004). MUSCLE: Multiple sequence alignment with high accuracy and high throughput. \emph{Nucleic Acids Research}, \emph{32}(5), 1792--1797.
-
-\leavevmode\hypertarget{ref-Felsenstein1985a}{}%
-Felsenstein, J. (1985). Phylogenies and the Comparative Method. \emph{The American Naturalist}, \emph{125}(1), 1--15. Retrieved from \url{http://www.jstor.org/stable/2461605}
-
-\leavevmode\hypertarget{ref-gbif2022taxonomy}{}%
-GBIF Secretariat.
-(2022).
-GBIF Backbone Taxonomy.
-\emph{Checklist dataset}, (Online Resource accessed via GBIF.org).
-Retrieved from \href{\%7Bhttps://doi.org/10.15468/39omei\%20\%7D}{\{https://doi.org/10.15468/39omei \}}
-
-\leavevmode\hypertarget{ref-gibb2015new}{}%
-Gibb, G.
-C., England, R., Hartig, G., McLenachan, P.
-A., Taylor Smith, B.
-L., McComish, B.
-J., \ldots{} Penny, D.
-(2015).
-New zealand passerines help clarify the diversification of major songbird lineages during the oligocene.
-\emph{Genome Biology and Evolution}, \emph{7}(11), 2983--2995.
-
-\leavevmode\hypertarget{ref-graur2004reading}{}%
-Graur, D., \& Martin, W. (2004). Reading the entrails of chickens: Molecular timescales of evolution and the illusion of precision. \emph{TRENDS in Genetics}, \emph{20}(2), 80--86.
-
-\leavevmode\hypertarget{ref-hackett2008phylogenomic}{}%
-Hackett, S.
-J., Kimball, R.
-T., Reddy, S., Bowie, R.
-C., Braun, E.
-L., Braun, M.
-J., \ldots{} others.
-(2008).
-A phylogenomic study of birds reveals their evolutionary history.
-\emph{Science}, \emph{320}(5884), 1763--1768.
-
-\leavevmode\hypertarget{ref-Hedges2015}{}%
-Hedges, S.
-B., Marin, J., Suleski, M., Paymer, M., \& Kumar, S.
-(2015).
-Tree of life reveals clock-like speciation and diversification.
-\emph{Molecular Biology and Evolution}, \emph{32}(4), 835--845.
-\url{https://doi.org/10.1093/molbev/msv037}
-
-\leavevmode\hypertarget{ref-Heibl2008}{}%
-Heibl, C. (2008). \emph{PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages.} Retrieved from \url{http://www.christophheibl.de/Rpackages.html}
-
-\leavevmode\hypertarget{ref-hooper2017chromosomal}{}%
-Hooper, D. M., \& Price, T. D. (2017). Chromosomal inversion differences correlate with range overlap in passerine birds. \emph{Nature Ecology \& Evolution}, \emph{1}(10), 1526.
-
-\leavevmode\hypertarget{ref-Huelsenbeck2001}{}%
-Huelsenbeck, J. P., \& Ronquist, F. (2001). MRBAYES: Bayesian inference of phylogenetic trees. \emph{Bioinformatics}, \emph{17}(8), 754--755. \url{https://doi.org/10.1093/bioinformatics/17.8.754}
-
-\leavevmode\hypertarget{ref-Jetz2012}{}%
-Jetz, W., Thomas, G., Joy, J.
-J., Hartmann, K., \& Mooers, A.
-(2012).
-The global diversity of birds in space and time.
-\emph{Nature}, \emph{491}(7424), 444--448.
-\url{https://doi.org/10.1038/nature11631}
-
-\leavevmode\hypertarget{ref-johansson2008phylogenetic}{}%
-Johansson, U.
-S., Fjeldså, J., \& Bowie, R.
-C.
-(2008).
-Phylogenetic relationships within passerida (aves: Passeriformes): A review and a new molecular phylogeny based on three nuclear intron markers.
-\emph{Molecular Phylogenetics and Evolution}, \emph{48}(3), 858--876.
-
-\leavevmode\hypertarget{ref-katoh2009multiple}{}%
-Katoh, K., Asimenos, G., \& Toh, H. (2009). Multiple alignment of dna sequences with mafft. In \emph{Bioinformatics for dna sequence analysis} (pp. 39--64). Springer.
-
-\leavevmode\hypertarget{ref-kimball2019phylogenomic}{}%
-Kimball, R. T., Oliveros, C. H., Wang, N., White, N. D., Barker, F. K., Field, D. J., \ldots{} others. (2019). A phylogenomic supertree of birds. \emph{Diversity}, \emph{11}(7), 109.
-
-\leavevmode\hypertarget{ref-klicka2014comprehensive}{}%
-Klicka, J., Barker, F.
-K., Burns, K.
-J., Lanyon, S.
-M., Lovette, I.
-J., Chaves, J.
-A., \& Bryson Jr, R.
-W.
-(2014).
-A comprehensive multilocus assessment of sparrow (aves: Passerellidae) relationships.
-\emph{Molecular Phylogenetics and Evolution}, \emph{77}, 177--182.
-
-\leavevmode\hypertarget{ref-ksepka2015fossil}{}%
-Ksepka, D.
-T., Parham, J.
-F., Allman, J.
-F., Benton, M.
-J., Carrano, M.
-T., Cranston, K.
-A., \ldots{} others.
-(2015).
-The fossil calibration database---a new resource for divergence dating.
-\emph{Systematic Biology}, \emph{64}(5), 853--859.
-
-\leavevmode\hypertarget{ref-lamichhaney2015evolution}{}%
-Lamichhaney, S., Berglund, J., Almén, M.
-S., Maqbool, K., Grabherr, M., Martinez-Barrio, A., \ldots{} others.
-(2015).
-Evolution of darwin's finches and their beaks revealed by genome sequencing.
-\emph{Nature}, \emph{518}(7539), 371--375.
-
-\leavevmode\hypertarget{ref-laubichler2009form}{}%
-Laubichler, M. D., \& Maienschein, J. (2009). \emph{Form and function in developmental evolution}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-lepage2004avibase}{}%
-Lepage, D. (2004). \emph{Avibase: The world bird database}. Bird Studies Canada.
-
-\leavevmode\hypertarget{ref-lepage2014avibase}{}%
-Lepage, D., Vaidya, G., \& Guralnick, R. (2014). Avibase--a database system for managing and organizing taxonomic concepts. \emph{ZooKeys}, (420), 117.
-
-\leavevmode\hypertarget{ref-lerner2011multilocus}{}%
-Lerner, H.
-R., Meyer, M., James, H.
-F., Hofreiter, M., \& Fleischer, R.
-C.
-(2011).
-Multilocus resolution of phylogeny and timescale in the extant adaptive radiation of hawaiian honeycreepers.
-\emph{Current Biology}, \emph{21}(21), 1838--1844.
-
-\leavevmode\hypertarget{ref-lovette2010comprehensive}{}%
-Lovette, I.
-J., Pérez-Emán, J.
-L., Sullivan, J.
-P., Banks, R.
-C., Fiorentino, I., Córdoba-Córdoba, S., \ldots{} others.
-(2010).
-A comprehensive multilocus phylogeny for the wood-warblers and a revised classification of the parulidae (aves).
-\emph{Molecular Phylogenetics and Evolution}, \emph{57}(2), 753--770.
-
-\leavevmode\hypertarget{ref-magallon2001absolute}{}%
-Magallon, S., \& Sanderson, M. (2001). Absolute diversification rates in angiosperm clades. \emph{Evolution}, \emph{55}(9), 1762--1780.
-
-\leavevmode\hypertarget{ref-magallon2010using}{}%
-Magallón, S. (2010). Using fossils to break long branches in molecular dating: A comparison of relaxed clocks applied to the origin of angiosperms. \emph{Systematic Biology}, \emph{59}(4), 384--399.
-
-\leavevmode\hypertarget{ref-magallon2015metacalibrated}{}%
-Magallón, S., Gómez-Acevedo, S., Sánchez-Reyes, L.
-L., \& Hernández-Hernández, T.
-(2015).
-A metacalibrated time-tree documents the early rise of flowering plant phylogenetic diversity.
-\emph{New Phytologist}, \emph{207}(2), 437--453.
-
-\leavevmode\hypertarget{ref-mctavish2015phylesystem}{}%
-McTavish, E.
-J., Hinchliff, C.
-E., Allman, J.
-F., Brown, J.
-W., Cranston, K.
-A., Holder, M.
-T., \ldots{} Smith, S.
-(2015).
-Phylesystem: A git-based data store for community-curated phylogenetic estimates.
-\emph{Bioinformatics}, \emph{31}(17), 2794--2800.
-
-\leavevmode\hypertarget{ref-Michonneau2016}{}%
-Michonneau, F., Brown, J.
-W., \& Winter, D.
-J.
-(2016).
-rotl: an R package to interact with the Open Tree of Life data.
-\emph{Methods in Ecology and Evolution}, \emph{7}(12), 1476--1481.
-\url{https://doi.org/10.1111/2041-210X.12593}
-
-\leavevmode\hypertarget{ref-Morlon2014}{}%
-Morlon, H. (2014). Phylogenetic approaches for studying diversification. \emph{Ecology Letters}, \emph{17}(4), 508--525. \url{https://doi.org/10.1111/ele.12251}
-
-\leavevmode\hypertarget{ref-moyle2016tectonic}{}%
-Moyle, R.
-G., Oliveros, C.
-H., Andersen, M.
-J., Hosner, P.
-A., Benz, B.
-W., Manthey, J.
-D., \ldots{} Faircloth, B.
-C.
-(2016).
-Tectonic collision and uplift of wallacea triggered the global songbird radiation.
-\emph{Nature Communications}, \emph{7}(1), 1--7.
-
-\leavevmode\hypertarget{ref-oliveros2019earth}{}%
-Oliveros, C.
-H., Field, D.
-J., Ksepka, D.
-T., Barker, F.
-K., Aleixo, A., Andersen, M.
-J., \ldots{} others.
-(2019).
-Earth history and the passerine superradiation.
-\emph{Proceedings of the National Academy of Sciences}, \emph{116}(16), 7916--7925.
-
-\leavevmode\hypertarget{ref-Ooms2018}{}%
-Ooms, J., \& Chamberlain, S. (2018). \emph{Phylocomr: Interface to 'phylocom'}. Retrieved from \url{https://CRAN.R-project.org/package=phylocomr}
-
-\leavevmode\hypertarget{ref-opentreeAPIs}{}%
-Open Tree Of Life, Redelings, B., Cranston, K.
-A., Allman, J., Holder, M.
-T., \& McTavish, E.
-J.
-(2016).
-Open Tree of Life APIs v3.0.
-\emph{Open Tree of Life Project}, (Online Resources).
-Retrieved from \href{\%7Bhttps://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\%7D}{\{https://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\}}
-
-\leavevmode\hypertarget{ref-opentreeoflife2019synth}{}%
-Open Tree Of Life, Redelings, B., Sánchez Reyes, L.
-L., Cranston, K.
-A., Allman, J., Holder, M.
-T., \& McTavish, E.
-J.
-(2019).
-Open tree of life synthetic tree v12.3.
-\emph{Zenodo}.
-Retrieved from \url{https://doi.org/10.5281/zenodo.3937742}
-
-\leavevmode\hypertarget{ref-odeen2011evolution}{}%
-Ödeen, A., Håstad, O., \& Alström, P. (2011). Evolution of ultraviolet vision in the largest avian radiation-the passerines. \emph{BMC Evolutionary Biology}, \emph{11}(1), 1--8.
-
-\leavevmode\hypertarget{ref-paradis2004}{}%
-Paradis, E., Claude, J., \& Strimmer, K. (2004). APE: analyses of phylogenetics and evolution in R language. \emph{Bioinformatics}, \emph{20}(2), 289--290.
-
-\leavevmode\hypertarget{ref-parchman2007coevolution}{}%
-Parchman, T.
-L., Benkman, C.
-W., \& Mezquida, E.
-T.
-(2007).
-Coevolution between hispaniolan crossbills and pine: Does more time allow for greater phenotypic escalation at lower latitude?
-\emph{Evolution}, \emph{61}(9), 2142--2153.
-
-\leavevmode\hypertarget{ref-packert2012horizontal}{}%
-Päckert, M., Martens, J., Sun, Y.-H., Severinghaus, L.
-L., Nazarenko, A.
-A., Ting, J., \ldots{} Tietze, D.
-T.
-(2012).
-Horizontal and elevational phylogeographic patterns of himalayan and southeast asian forest passerines (aves: Passeriformes).
-\emph{Journal of Biogeography}, \emph{39}(3), 556--573.
-
-\leavevmode\hypertarget{ref-pennell2014geiger}{}%
-Pennell, M.
-W., Eastman, J.
-M., Slater, G.
-J., Brown, J.
-W., Uyeda, J.
-C., FitzJohn, R.
-G., \ldots{} Harmon, L.
-J.
-(2014).
-Geiger v2.
-0: An expanded suite of methods for fitting macroevolutionary models to phylogenetic trees.
-\emph{Bioinformatics}, \emph{30}(15), 2216--2218.
-
-\leavevmode\hypertarget{ref-posadas2006historical}{}%
-Posadas, P., Crisci, J. V., \& Katinas, L. (2006). Historical biogeography: A review of its basic concepts and critical issues. \emph{Journal of Arid Environments}, \emph{66}(3), 389--403.
-
-\leavevmode\hypertarget{ref-powell2014comprehensive}{}%
-Powell, A.
-F., Barker, F.
-K., Lanyon, S.
-M., Burns, K.
-J., Klicka, J., \& Lovette, I.
-J.
-(2014).
-A comprehensive species-level molecular phylogeny of the new world blackbirds (icteridae).
-\emph{Molecular Phylogenetics and Evolution}, \emph{71}, 94--112.
-
-\leavevmode\hypertarget{ref-price2014niche}{}%
-Price, T.
-D., Hooper, D.
-M., Buchanan, C.
-D., Johansson, U.
-S., Tietze, D.
-T., Alström, P., \ldots{} others.
-(2014).
-Niche filling slows the diversification of himalayan songbirds.
-\emph{Nature}, \emph{509}(7499), 222.
-
-\leavevmode\hypertarget{ref-pulgarin2013multilocus}{}%
-Pulgarín-R, P.
-C., Smith, B.
-T., Bryson Jr, R.
-W., Spellman, G.
-M., \& Klicka, J.
-(2013).
-Multilocus phylogeny and biogeography of the new world pheucticus grosbeaks (aves: Cardinalidae).
-\emph{Molecular Phylogenetics and Evolution}, \emph{69}(3), 1222--1227.
-
-\leavevmode\hypertarget{ref-rabosky2018inverse}{}%
-Rabosky, D.
-L., Chang, J., Title, P.
-O., Cowman, P.
-F., Sallan, L., Friedman, M., \ldots{} others.
-(2018).
-An inverse latitudinal gradient in speciation rate for marine fishes.
-\emph{Nature}, \emph{559}(7714), 392.
-
-\leavevmode\hypertarget{ref-ramshaw1972time}{}%
-Ramshaw, J., Richardson, D., Meatyard, B., Brown, R., Richardson, M., Thompson, E., \& Boulter, D.
-(1972).
-The time of origin of the flowering plants determined by using amino acid sequence data of cytochrome c.
-\emph{New Phytologist}, \emph{71}(5), 773--779.
-
-\leavevmode\hypertarget{ref-ratnasingham2007bold}{}%
-Ratnasingham, S., \& Hebert, P. D. (2007). BOLD: The barcode of life data system (http://www. Barcodinglife. Org). \emph{Molecular Ecology Notes}, \emph{7}(3), 355--364.
-
-\leavevmode\hypertarget{ref-RCoreTeam2018}{}%
-R Core Team. (2018). \emph{R: a language and environment for statistical computing}. Vienna, Austria: R Foundation for Statistical Computing.
-
-\leavevmode\hypertarget{ref-rees2017automated}{}%
-Rees, \& Cranston, K. (2017). Automated assembly of a reference taxonomy for phylogenetic data synthesis. \emph{Biodiversity Data Journal}, (5).
-
-\leavevmode\hypertarget{ref-rees2017irmng}{}%
-Rees, Vandepitte, L., Decock, W., \& Vanhoorne, B. (2017). IRMNG 2006--2016: 10 Years of a Global Taxonomic Database. \emph{Biodiversity Informatics}, \emph{12}.
-
-\leavevmode\hypertarget{ref-Revell2012}{}%
-Revell, L. J. (2012). Phytools: An r package for phylogenetic comparative biology (and other things). \emph{Methods in Ecology and Evolution}, \emph{3}, 217--223.
-
-\leavevmode\hypertarget{ref-Ronquist2003}{}%
-Ronquist, F., \& Huelsenbeck, J.
-P.
-(2003).
-MrBayes 3: Bayesian phylogenetic inference under mixed models.
-\emph{Bioinformatics}, \emph{19}(12), 1572--1574.
-\url{https://doi.org/10.1093/bioinformatics/btg180}
-
-\leavevmode\hypertarget{ref-datelifeplot}{}%
-Sanchez-Reyes, L.
-L., \& O'Meara, B.
-(2022).
-\texttt{datelifeplot}: Methods to plot chronograms and outputs of the \texttt{datelife} package.
-\emph{R Package Release V0.2.2}.
-Retrieved from \url{https://zenodo.org/badge/latestdoi/381501451}
-
-\leavevmode\hypertarget{ref-sanchez2022}{}%
-Sanchez-Reyes, L.
-L., O'Meara, B., Eastman, J., Heath, T., Wright, A., Schliep, K., \ldots{} Alfaro, M.
-(2022).
-\texttt{datelife}: Scientific Data on Time of Lineage Divergence for Your Taxa.
-In \emph{R package version 0.6.6}.
-Retrieved from \href{https://CRAN.R-project.org/package=datelife\%20and\%20https://doi.org/10.5281/zenodo.593938}{https://CRAN.R-project.org/package=datelife and https://doi.org/10.5281/zenodo.593938}
-
-\leavevmode\hypertarget{ref-sanderson2002}{}%
-Sanderson, M.
-(2002).
-Estimating Absolute Rates of Molecular Evolution and Divergence Times: A Penalized Likelihood Approach.
-\emph{Molecular Biology and Evolution}, \emph{19}(1), 101--109.
-\url{https://doi.org/10.1093/oxfordjournals.molbev.a003974}
-
-\leavevmode\hypertarget{ref-sanderson2003r8s}{}%
-Sanderson, M. (2003). r8s: Inferring Absolute Rates of Molecular Evolution and Divergence Times in the Absence of a Molecular Clock. \emph{Bioinformatics}, \emph{19}(2), 301--302.
-
-\leavevmode\hypertarget{ref-sanderson2001sources}{}%
-Sanderson, M., \& Doyle, J.
-(2001).
-Sources of error and confidence intervals in estimating the age of angiosperms from rbcL and 18S rDNA data.
-\emph{American Journal of Botany}, \emph{88}(8), 1499--1516.
-
-\leavevmode\hypertarget{ref-sauquet2013practical}{}%
-Sauquet, H. (2013). A practical guide to molecular dating. \emph{Comptes Rendus Palevol}, \emph{12}(6), 355--367.
-
-\leavevmode\hypertarget{ref-sauquet2021age}{}%
-Sauquet, H., Ramírez-Barahona, S., \& Magallón, S. (2021). \emph{The age of flowering plants is unknown}.
-
-\leavevmode\hypertarget{ref-schenk2016sec}{}%
-Schenk, J. J. (2016). Consequences of secondary calibrations on divergence time estimates. \emph{PLoS ONE}, \emph{11}(1). \url{https://doi.org/10.1371/journal.pone.0148228}
-
-\leavevmode\hypertarget{ref-schliep2011phangorn}{}%
-Schliep, K. P. (2011). Phangorn: Phylogenetic analysis in r. \emph{Bioinformatics}, \emph{27}(4), 592--593.
-
-\leavevmode\hypertarget{ref-schoch2020ncbi}{}%
-Schoch, C.
-L., Ciufo, S., Domrachev, M., Hotton, C.
-L., Kannan, S., Khovanskaya, R., \ldots{} others.
-(2020).
-NCBI Taxonomy: a Comprehensive Update on Curation, Resources and Tools.
-\emph{Database}, \emph{2020}.
-
-\leavevmode\hypertarget{ref-selvatti2015paleogene}{}%
-Selvatti, A.
-P., Gonzaga, L.
-P., \& Moraes Russo, C.
-A.
-de.
-(2015).
-A paleogene origin for crown passerines and the diversification of the oscines in the new world.
-\emph{Molecular Phylogenetics and Evolution}, \emph{88}, 1--15.
-
-\leavevmode\hypertarget{ref-smith2018constructing}{}%
-Smith, S., \& Brown, J. (2018). Constructing a broadly inclusive seed plant phylogeny. \emph{American Journal of Botany}, \emph{105}(3), 302--314.
-
-\leavevmode\hypertarget{ref-Smith2012}{}%
-Smith, S., \& O'Meara, B.
-(2012).
-TreePL: Divergence time estimation using penalized likelihood for large phylogenies.
-\emph{Bioinformatics}, \emph{28}(20), 2689--2690.
-\url{https://doi.org/10.1093/bioinformatics/bts492}
-
-\leavevmode\hypertarget{ref-Stoltzfus2013}{}%
-Stoltzfus, A., Lapp, H., Matasci, N., Deus, H., Sidlauskas, B., Zmasek, C.
-M., \ldots{} Jordan, G.
-(2013).
-Phylotastic!
-Making tree-of-life knowledge accessible, reusable and convenient.
-\emph{BMC Bioinformatics}, \emph{14}.
-\url{https://doi.org/10.1186/1471-2105-14-158}
-
-\leavevmode\hypertarget{ref-tietze2013complete}{}%
-Tietze, D.
-T., Päckert, M., Martens, J., Lehmann, H., \& Sun, Y.-H.
-(2013).
-Complete phylogeny and historical biogeography of true rosefinches (aves: Carpodacus).
-\emph{Zoological Journal of the Linnean Society}, \emph{169}(1), 215--234.
-
-\leavevmode\hypertarget{ref-title2016macrophylogenies}{}%
-Title, P.
-O., \& Rabosky, D.
-L.
-(2016).
-Do Macrophylogenies Yield Stable Macroevolutionary Inferences?
-An Example from Squamate Reptiles.
-\emph{Systematic Biology}, syw102.
-\url{https://doi.org/10.1093/sysbio/syw102}
-
-\leavevmode\hypertarget{ref-treplin2008molecular}{}%
-Treplin, S., Siegert, R., Bleidorn, C., Thompson, H.
-S., Fotso, R., \& Tiedemann, R.
-(2008).
-Molecular phylogeny of songbirds (aves: Passeriformes) and the relative utility of common nuclear marker loci.
-\emph{Cladistics}, \emph{24}(3), 328--349.
-
-\leavevmode\hypertarget{ref-vos2012nexml}{}%
-Vos, R.
-A., Balhoff, J.
-P., Caravas, J.
-A., Holder, M.
-T., Lapp, H., Maddison, W.
-P., \ldots{} others.
-(2012).
-NeXML: Rich, extensible, and verifiable representation of comparative data and metadata.
-\emph{Systematic Biology}, \emph{61}(4), 675--689.
-\url{https://doi.org/10.1093/sysbio/sys025}
-
-\leavevmode\hypertarget{ref-Webb2000}{}%
-Webb, C. (2000). Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees. \emph{The American Naturalist}, \emph{156}(2), 145--155.
-
-\leavevmode\hypertarget{ref-Webb2008}{}%
-Webb, C., Ackerly, D., \& Kembel, S.
-(2008).
-Phylocom: Software for the analysis of phylogenetic community structure and trait evolution.
-\emph{Bioinformatics}, \emph{24}(18), 2098--2100.
-\url{https://doi.org/10.1093/bioinformatics/btn358}
-
-\leavevmode\hypertarget{ref-webb2005phylomatic}{}%
-Webb, C., \& Donoghue, M. (2005). Phylomatic: Tree assembly for applied phylogenetics. \emph{Molecular Ecology Notes}, \emph{5}(1), 181--183.
-
-\leavevmode\hypertarget{ref-weir2008calibrating}{}%
-Weir, J., \& Schluter, D. (2008). Calibrating the avian molecular clock. \emph{Molecular Ecology}, \emph{17}(10), 2321--2328.
-
-\leavevmode\hypertarget{ref-zuccon2012phylogenetic}{}%
-Zuccon, D., Prŷs-Jones, R., Rasmussen, P.
-C., \& Ericson, P.
-G.
-(2012).
-The phylogenetic relationships and generic limits of finches (fringillidae).
-\emph{Molecular Phylogenetics and Evolution}, \emph{62}(2), 581--596.
-
-\endgroup
-
-\newpage
-
-\captionof{table}{Ages of congruified nodes. See Figure 3, step C2.}
-\label{tbl:table1}
-\includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example.pdf}
-
-\vspace{15mm}
-
-\captionof{table}{Summary of congruified nodes ages. See Figure 3, step C3.}
-
-\label{tbl:table2}
-\includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example-summary.png}
-
-\begin{minipage}[t]{17.3cm}
-\includegraphics{../figures/figure-workflow/figure1-new.pdf}
-\captionof{figure}{\textit{Main DateLife workflow.
-Analyses can be performed via DateLife's interactive website at \url{www.datelife.org}, or using the \texttt{datelife} R package.
-Details on the R functions used to perform the analyses are available from \texttt{datelife}'s R package vignettes at \url{https://phylotastic.org/datelife}.}}
-\label{fig:workflow}
-\end{minipage}
-
-\includegraphics{../figures/fig_runtime_main.pdf}
-\captionof{figure}{\textit{DateLife's benchmarking results.
-Computation time used to process a query and a search across \texttt{datelife}'s chronogram database, relative to number of input taxon names.
-For each N = \{10, 100, 200, ..., 1 000, ...
-, 9 000, 10 000\}, we sampled N species names from the class Aves a hundred times, and then performed a datelife search processing the input names with Taxon Names Resolution Service (TNRS; light gray), and without processing names (dark gray).
-For comparison, we performed a search using an input that had been pre-processed with TNRS (light blue).}}
-\label{fig:runtime_main}
-
-\begin{center}
-\includegraphics{../figures/figure-small-example/figure2.pdf}
-\end{center}
-\vspace{10mm}
-\captionof{figure}{\textit{DateLife analysis results for a small sample of A) 6 bird species within the Passeriformes.
-B) Processed species names were found across 9 chronograms within 6 independent studies (Barker et al.
-(\protect\hyperlink{ref-barker2012going}{2012}), Barker et al.
-(\protect\hyperlink{ref-barker2015new}{2015}), Burns et al.
-(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al.
-(\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al.
-(\protect\hyperlink{ref-Jetz2012}{2012}).) C) This revealed 28 source age data points for the queried species names.
-Summarized age data is used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree, resulting in a summary chronogram of source ages.}}
-
-\label{fig:figure2}
-
-\begin{minipage}{18.2cm}
-\includegraphics{../figures/fringillidae-topologies/fringillidae-topology.pdf}
-\captionof{figure}{Tree topologies extracted from Open Tree of Life's (OpenTree) synthetic phylogenetic tree.
+Tree topologies extracted from Open Tree of Life's (OpenTree) synthetic phylogenetic tree.
 A) Topology of 2,333 tips and 1,305 internal nodes, encompassing bird species within the family Fringillidae following the NCBI taxonomy (black), as well as all other bird species that share the same Most Recent Common Ancestor (MRCA) node in OpenTree's synthetic tree (purple).
@@
   Treplin et al. \protect\hyperlink{ref-treplin2008molecular}{2008},
-  Zuccon, Prŷs-Jones, Rasmussen, \& Ericson \protect\hyperlink{ref-zuccon2012phylogenetic}{2012}).}
-\label{fig:fringillidae-topologies}
-\end{minipage}
-
-\begin{minipage}{18cm}
-\begin{center}
-\includegraphics{../figures/figure-fringillidae/median_and_calibration_ages-congruified.png}
-\end{center}
-
-\captionof{figure}{Fringillidae median summary chronogram generated with DateLife.
-It has 289 tips and 253 nodes, from which 212 have age data from at least one published chronogram.
-These nodes have age data points from 19 different chronograms from 13 different studies (colored circles), which were summarized into 194 data points that were used as secondary calibrations (purple slashes) with the Branch Length Adjuster (BLADJ ) software from Webb et al., (\protect\hyperlink{ref-Webb2008}{2008}).}
-\label{fig:fringillidages}
-\end{minipage}
-
-\newpage
-
-\begin{minipage}{18cm}
-\begin{center}
-\includegraphics{../figures/fringillidae-cross-validation/fig-cross-validation-xy-plots.pdf}
-\end{center}
-\captionof{figure}{Results from cross validation analysis. Each plot compares the original age estimate (x axis) with the age obtained with a DateLife analysis (y axis), per node.}
-\label{fig:cvXY}
-\end{minipage}
-
-\end{document}
+  Zuccon, Prŷs-Jones, Rasmussen, \& Ericson \protect\hyperlink{ref-zuccon2012phylogenetic}{2012}).

### 2023-05-04 manuscript/manuscript.Rmd
add some more reviews

@@
 embed_figure <- FALSE
+figures_at_end <- FALSE
 ```
@@
   Chronograms –phylogenies with branch lengths proportional to time– represent key data on timing of evolutionary events for the study of natural processes in many areas of biological research.
-  Chronograms also provide valuable information that can be used for education, science communication, and policy decisions.
+  Chronograms also provide valuable information that can be used for education, science communication, and conservation policy decisions.
   Yet, achieving a high-quality reconstruction of a chronogram is a difficult and resource-consuming task.
@@
     The main DateLife workflow starts with one or more scientific taxon names provided by a user.
-Names are processed and standardized to a unified taxonomy.
-Then, DateLife runs a name match across its local chronogram database that is curated from Open Tree of Life's phylogenetic repository, and extracts all chronograms that contain at least two queried taxon names, along with their metadata.
-    Finally, node ages from matching chronograms are mapped upon a chosen tree topology using the congruification algorithm.
+Names are processed and standardized to a unified taxonomy, allowing DateLife to run a name match across its local chronogram database that is curated from Open Tree of Life's phylogenetic repository, and extract all chronograms that contain at least two queried taxon names, along with their metadata.
+    Finally, node ages from matching chronograms are mapped to corresponding nodes from a chosen tree topology using the congruification algorithm.
 Congruified node ages are used as secondary calibrations to date the chosen topology, with or without initial branch lengths, using different phylogenetic dating methods such as BLADJ, treePL, PATHd8 and MrBayes.
     We performed a cross-validation test to compare node ages resulting from a DateLife analysis (i.e, phylogenetic dating using secondary calibrations) to those from the original chronograms (i.e, obtained with primary calibrations), and found that DateLife's node age estimates are consistent with the age estimates from the original chronograms, with the largest variation in ages occurring around topologically deeper nodes.
-  Software such as DateLife can help increase awareness of the existing disparities among alternative hypotheses of dates for the same diversification events, and can support exploration of the effect of alternative chronogram hypotheses on downstream analyses, providing a framework for a more informed interpretation of evolutionary results.
+  Results from any software for scientific analysis can only be as good as the data used as input, we highlight the importance of considering the results of a DateLife analysis in the context of the input chronograms.
+  We encourage the use of DateLife to help increase awareness of the existing disparities among alternative hypotheses of dates for the same diversification events, and to support exploration of the effect of alternative chronogram hypotheses on downstream analyses, providing a framework for a more informed interpretation of evolutionary results.
 
@@
 
-Here we present the DateLife project which has the main goal of extracting and exposing age data from published chronograms, making age data readily accessible to the wider community for reuse and reanalysis, research, teaching, and science communication and policy.
+Here we present the DateLife project which has the main goal of extracting and exposing age data from published chronograms, making age data readily accessible to the wider community for reuse and reanalysis in research, teaching, science communication and conservation policy.
 <!--  R has become mainstream in the biological-data analyst toolkit.
@@
 DateLife's core software application is available as an R package [@sanchez2022], and as an online Rshiny interactive website at www.datelife.org.
-It features key elements for scientific reproducibility, such as a curated, versioned, open and fully public chronogram database [@mctavish2015phylesystem] that stores data in a computer-readable format [@vos2012nexml]; automated and programmatic ways of accessing and downloading the data in a computer-readable format [@Stoltzfus2013]; and methods to summarize and compare the data.
+It features key elements for scientific reproducibility, such as a curated, versioned, open and fully public chronogram database [@mctavish2015phylesystem] that stores data in a computer-readable format [@vos2012nexml]; automated and programmatic ways of accessing and downloading the data, also in a computer-readable format [@Stoltzfus2013]; and methods to summarize and compare the data.
 
@@
 
-DateLife's core software applications are implemented in the R package `datelife`.
-Its current stable version -- v`r packageVersion("datelife")`, is available from The Comprehensive R Archive Network (CRAN) repository [@sanchez2022], and relies on functionalities from other biological R packages:
+DateLife's core software applications are implemented in the R package `datelife`,
+<!-- Its current stable version -- v`r packageVersion("datelife")`, is available for installation from the GitHub repository -->
+<!-- The Comprehensive R Archive Network (CRAN) repository [@sanchez2022],  -->
+and relies on functionalities from other biological R packages:
 ape [@paradis2004],
@@
 
-DateLife makes state-of-the-art data on evolutionary time frame easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise.
-It is an open service that does not require any expert biological knowledge from users --besides the names of the species or group they want to work with, for any of its functionality.
+DateLife's goal is to improve availability and accessibility of state-of-the-art data on evolutionary time frame of organisms, to allow users from all areas of science and with all levels of expertise to compare, reuse, and reanalyse expert age data for their own applications.
+As such, it is designed as an open service that does not require any expert biological knowledge --besides the scientific names of the species or group users want to work with, to use any of its functionalities.
 
@@
 
-While it is known that incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance in ecological and conservation biology studies [@Webb2008].
-Next, we discuss some particularities of using summarized node age data from a DateLife analysis that users should be aware of.
+While we envision that DateLife will have many interesting applications in research and beyond, we recognise that as 
+<!-- sort of a "data mining tool",  -->
+any other automated tool for biological data analysis, DateLife's results can only be as good as the input data, and as give as good as how we decide to apply them for research.  
+<!-- that they might not be appropriate for research. -->
+We recognise that this might be a problem in some research areas, but it might not be the case for all.
+For example, in ecological and conservation biology studies, it has been seen that incorporating some age data provides better results than when not using any age data at all, even if the age data is not accurate [@Webb2008].
+In the following sections we discuss the particularities of divergence times resulting from a DateLife analysis for consideration of the readers and users in different research areas.
+<!--
+Resources of the paragraph above:
+https://asistdl.onlinelibrary.wiley.com/doi/full/10.1002/asi.23538?casa_token=3_2mfbffsrsAAAAA%3A42WBe0ffa_fwcff15n4dlwzer5s6nkPCfW1RT8zOmyg0VW9-v_SB5lR8O4uur_MZvl5bUJ0c9FS4qPPTGg
+https://towardsdatascience.com/a-primer-on-the-sources-of-biases-in-data-mining-for-machine-learning-d82e89604693
+
+-->
 
@@
 See, for example, the robust ongoing debate about crown group age of angiosperms  [@sauquet2021age; @barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time].
-Source chronograms available for the same organisms have potentially been estimated implementing calibrations very differently.
+Alternative source chronograms available for the same taxa have potentially been estimated implementing calibrations very differently.
 In the Fringillidae example above, the chronograms from @burns2014phylogenetics were inferred using molecular substitution rate estimates across birds [@weir2008calibrating], and have much older age estimates for the same nodes than chronograms that were inferred using fossils as calibrations (Figs.
@@
 Yet, if one researcher concludes that a fossil should calibrate the ingroup of a clade, while another researcher concludes that the same fossil should calibrate the outgroup of the clade, the resulting age estimates will differ, as the placement of calibrations as stem or crown group has been proven to significantly affect time of lineage divergence estimates [@sauquet2013practical].
+In addition, nodes that are sandwiched between a calibrated node and a calibrated root have less freedom of movement and hence narrower confidence intervals [@vos2004reconstructing], this inflates precision for nodes without calibrations but does not necessarily improve accuracy of the estimated ages.
 
@@
 -->
-There have been attempts to create fossil calibration databases [@ksepka2015fossil], though these still have room to grow.
-<!--BCO: Polite way to say 'no new additions in four years?' LLSR: lol, yes-->
+
+The Fossil Calibration Database provides availability of 220 primary calibration points encompassing flowering plants and metazoans, that have been curated by experts and used in peer-reviewd publications [@ksepka2015fossil].
+The database facilitates the use of expert primary fossil calibrations in new phylogenetic dating analysis.
+<!--though these still have room to grow...
+BCO: Polite way to say 'no new additions in four years?' LLSR: lol, yes-->
+Despite the wealth of information displayed in the database, users still need the expertise to locate the equivalent node in their phylogenies to be dated. 
+Open databases for reuse of fossil calibration data still have room to grow and should be a great addition to the phylogenetics tool set.
+
 <!--\textcolor{red}{LLSR: What does our cross validation analysis show in relation to  the effect of using secondary calibrations for a dating analysis that does not make any prior assumptions on the nature of the calibrations themselves?
@@
 \begin{center}
-\emph{Effects of Phylogenetic Sampling on Downstream Analyses}
+\emph{Effects of Taxon Sampling on Downstream Analyses}
 \end{center}
@@
 The DateLife software is free and open source.
-It can be used online through its R shiny web application hosted at <http://www.datelife.org>, and locally through the `datelife` R package, downloadable from Zenodo (<https://doi.org/10.5281/zenodo.593938>) and the CRAN repository [@sanchez2022].
+It can be used online through its R shiny web application hosted at <http://www.datelife.org>, and locally through the `datelife` R package, available from Zenodo (<https://doi.org/10.5281/zenodo.593938> and the CRAN repository [@sanchez2022].
 DateLife's web application is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker and OpenTree's infrastructure (datelife.opentreeoflife.org).
-`datelife`'s stable version `r availability` for installation from CRAN repository using the command `install.packages(pkgs = "datelife")` from within R.
+`datelife`'s stable version can be installed from the CRAN repository using the command `install.packages(pkgs = "datelife")` from within R. 
 Development versions are available from DateLife's GitHub repository (<https://github.com/phylotastic/datelife>) and can be installed using the command `devtools::install_github("phylotastic/datelife")`.
@@
 
-```{r fig1, child= if (!embed_figure) c("table1.Rmd", "table2.Rmd", "fig-workflow.Rmd", "fig-benchmark.Rmd", "fig-small-example.Rmd", "fig-topologies.Rmd", "fig-fringillidae-full-example.Rmd", "fig-cross-validation.Rmd")}
+```{r fig1, child= if (figures_at_end) c("table1.Rmd", "table2.Rmd", "fig-workflow.Rmd", "fig-benchmark.Rmd", "fig-small-example.Rmd", "fig-topologies.Rmd", "fig-fringillidae-full-example.Rmd", "fig-cross-validation.Rmd")}
 ```

### 2023-05-04 manuscript/manuscript.tex
add some more reviews

@@
-\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
 \PassOptionsToPackage{hyphens}{url}
 %
-\documentclass[english,man]{apa6}
+\documentclass[
+  man]{apa6}
+\usepackage{amsmath,amssymb}
 \usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+\usepackage{iftex}
+\ifPDFTeX
   \usepackage[T1]{fontenc}
   \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provides euro and other symbols
-\else % if luatex or xelatex
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
   \usepackage{unicode-math}
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
 \fi
-% use upquote if available, for straight quotes in verbatim environments
+% Use upquote if available, for straight quotes in verbatim environments
 \IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
+\IfFileExists{microtype.sty}{% use microtype if available
   \usepackage[]{microtype}
@@
 }{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
   \IfFileExists{parskip.sty}{%
@@
     \setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
-\usepackage{hyperref}
-\hypersetup{
-            pdftitle={DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life},
-            pdfauthor={Luna L. Sánchez Reyes1,2, Emily Jane McTavish1, \& Brian O'Meara2},
-            pdfborder={0 0 0},
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
-\usepackage{graphicx,grffile}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\usepackage{graphicx}
 \makeatletter
@@
 \setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
 \setlength{\emergencystretch}{3em} % prevent overfull lines
@@
   \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{0}
-
-% set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+% Make \paragraph and \subparagraph free-standing
+\ifx\paragraph\undefined\else
+  \let\oldparagraph\paragraph
+  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+  \let\oldsubparagraph\subparagraph
+  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
+\ifLuaTeX
+\usepackage[bidi=basic]{babel}
+\else
+\usepackage[bidi=default]{babel}
+\fi
+\babelprovide[main,import]{english}
+% get rid of language-specific shorthands (see #6817):
+\let\LanguageShortHands\languageshorthands
+\def\languageshorthands#1{}
 \DeclareUnicodeCharacter{0301}{*************************************}
@@
 \captionsetup[table]{font+=linenumbers, labelfont={sc}, labelformat={default}, labelsep=period, name={Table}}
+
 % Manuscript styling
@@
 \usepackage{csquotes}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[shorthands=off,main=english]{babel}
-\else
-  % load polyglossia as late as possible as it *could* call bidi if RTL lang (e.g. Hebrew or Arabic)
-  \usepackage{polyglossia}
-  \setmainlanguage[]{english}
-\fi
+
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+
+\urlstyle{same} % disable monospaced font for URLs
+\hypersetup{
+  pdftitle={DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life},
+  pdfauthor={Luna L. Sánchez Reyes1,2, Emily Jane McTavish1, \& Brian O'Meara2},
+  pdflang={en-EN},
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
 
@@
 
-Chronograms --phylogenies with branch lengths proportional to time-- represent key data on evolutionary time frame for the study of natural processes in many areas of biological research.
-Chronograms also provide valuable information that can be used for education, science communication, and policy decisions.
+Chronograms --phylogenies with branch lengths proportional to time-- represent key data on timing of evolutionary events for the study of natural processes in many areas of biological research.
+Chronograms also provide valuable information that can be used for education, science communication, and conservation policy decisions.
 Yet, achieving a high-quality reconstruction of a chronogram is a difficult and resource-consuming task.
@@
 The main DateLife workflow starts with one or more scientific taxon names provided by a user.
-Names are processed and standardized to a unified taxonomy.
-Then, DateLife runs a name match across its local chronogram database that is curated from Open Tree of Life's phylogenetic repository, and extracts all chronograms that contain at least two queried taxon names, along with their metadata.
-Finally, node ages from matching chronograms are mapped upon a chosen tree topology using the congruification algorithm.
+Names are processed and standardized to a unified taxonomy, allowing DateLife to run a name match across its local chronogram database that is curated from Open Tree of Life's phylogenetic repository, and extract all chronograms that contain at least two queried taxon names, along with their metadata.
+Finally, node ages from matching chronograms are mapped to corresponding nodes from a chosen tree topology using the congruification algorithm.
 Congruified node ages are used as secondary calibrations to date the chosen topology, with or without initial branch lengths, using different phylogenetic dating methods such as BLADJ, treePL, PATHd8 and MrBayes.
-We performed a cross-validation test to compare node ages resulting from a DateLife analysis (i.e, phylogenetic dating using secondary calibrations) to those from the original chronograms (i.e, obtained with primary calibrations), and found that DateLife's node age estimates appear generally correlated to those from the original chronograms, with the largest variation in ages occurring around topologically deeper nodes.
-Software such as DateLife can help increase awareness of the existing variation in alternative hypothesis of evolutionary time frame for the same organisms, and can support exploration of the effect of alternative chronogram hypotheses on downstream analyses, providing a framework for a more informed interpretation of evolutionary results.
+We performed a cross-validation test to compare node ages resulting from a DateLife analysis (i.e, phylogenetic dating using secondary calibrations) to those from the original chronograms (i.e, obtained with primary calibrations), and found that DateLife's node age estimates are consistent with the age estimates from the original chronograms, with the largest variation in ages occurring around topologically deeper nodes.
+Results from any software for scientific analysis can only be as good as the data used as input, we highlight the importance of considering the results of a DateLife analysis in the context of the input chronograms.
+We encourage the use of DateLife to help increase awareness of the existing disparities among alternative hypotheses of dates for the same diversification events, and to support exploration of the effect of alternative chronogram hypotheses on downstream analyses, providing a framework for a more informed interpretation of evolutionary results.
 
@@
 
-Word count: 6111
+Word count: 6380
 
@@
 
-Chronograms --phylogenies with branch lengths proportional to time-- provide key data on evolutionary time frame for the study of natural processes in many areas of biological research, such as comparative analysis (Freckleton, Harvey, \& Pagel, 2002; Harvey, Pagel, \& others, 1991), developmental biology (Delsuc et al., 2018; Laubichler \& Maienschein, 2009), conservation biology and ecology (Felsenstein, 1985; Webb, 2000), historical biogeography (Posadas, Crisci, \& Katinas, 2006), and species diversification (Magallon \& Sanderson, 2001; Morlon, 2014).
+Chronograms --phylogenies with branch lengths proportional to time-- provide key data on evolutionary time frame for the study of natural processes in many areas of biological research, such as comparative analysis (Freckleton, Harvey, \& Pagel, 2002; Harvey, Pagel, et al., 1991), developmental biology (Delsuc et al., 2018; Laubichler \& Maienschein, 2009), conservation biology and ecology (Felsenstein, 1985; Webb, 2000), historical biogeography (Posadas, Crisci, \& Katinas, 2006), and species diversification (Magallon \& Sanderson, 2001; Morlon, 2014).
 
@@
 It requires obtaining and curating a homology hypothesis to construct a phylogeny, selecting and placing appropriate calibrations on the phylogeny using independent age data points from the fossil record or other dated events, and inferring a full dated tree.
-All of this entails specialized biological training, taxonomic domain knowledge, and a non-negligible amount of research time, computational resources and funding.
-
-Here we present the DateLife project which has the main goal of extracting and exposing age data from published chronograms, making age data readily accessible to the wider community for reuse and reanalysis, research, teaching, and science communication and policy.
+All of this entails specialized biological training, taxonomic domain knowledge, and a significant amount of research time, computational resources and funding.
+
+Here we present the DateLife project which has the main goal of extracting and exposing age data from published chronograms, making age data readily accessible to the wider community for reuse and reanalysis in research, teaching, science communication and conservation policy.
 DateLife's core software application is available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org.
-It features key elements for scientific reproducibility, such as a curated, versioned, open and fully public chronogram database (McTavish et al., 2015) that stores data in a computer-readable format (Vos et al., 2012); automated and programmatic ways of accessing and downloading the data in a computer-readable format also (Stoltzfus et al., 2013); and methods to summarize and compare the data.
+It features key elements for scientific reproducibility, such as a curated, versioned, open and fully public chronogram database (McTavish et al., 2015) that stores data in a computer-readable format (Vos et al., 2012); automated and programmatic ways of accessing and downloading the data, also in a computer-readable format (Stoltzfus et al., 2013); and methods to summarize and compare the data.
 
@@
 
-DateLife's core software applications are implemented in the R package \texttt{datelife}.
-Its current stable version -- v0.6.6, is available from The Comprehensive R Archive Network (CRAN) repository (Sanchez-Reyes et al., 2022), and relies on functionalities from other biological R packages:
+DateLife's core software applications are implemented in the R package \texttt{datelife},
+and relies on functionalities from other biological R packages:
 ape (Paradis, Claude, \& Strimmer, 2004),
-bold (Chamberlain, 2018),
+bold (Chamberlain et al., 2019),
 geiger (Pennell et al., 2014),
@@
 rotl (Michonneau, Brown, \& Winter, 2016), and
-taxize (Chamberlain, 2018; Chamberlain \& Szöcs, 2013).
+taxize (Chamberlain \& Szöcs, 2013; Chamberlain et al., 2019).
 Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife workflow: creating a search query, searching a database, and summarizing results from the search.
@@
 
-DateLife starts by processing an input consisting of at least one taxon scientific name. Multiple input names can be provided as a comma separated character string or as tip labels on a tree.
-If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a \enquote{phylo} R object (Paradis et al., 2004).
+DateLife starts by processing an input consisting of the scientific name of at least one taxon. Multiple input names can be provided as a comma separated character string or as tip labels on a tree.
+If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis et al., 2004).
 The input tree is not required to have branch lengths, and its topology is used in the summary steps described in the next section.
@@
 DateLife implements TNRS using the Open Tree of Life (OpenTree) unified Taxonomy (OTT, Open Tree Of Life et al., 2016; Rees \& Cranston, 2017) as standard, storing taxonomic identification numbers (OTT ids) for further processing and analysis.
-Other taxonomies currently supported by DateLife are the National Center of Biotechnology Information (NCBI) taxonomic database (Schoch et al., 2020), the Global Biodiversity Information Facility (GBIF) taxonomic backbone (GBIF Secretariat, 2022), and the Interim Register of Marine and Non-marine Genera (IRMNG) database (Rees et al., 2017).
+Other taxonomies currently supported by DateLife are the National Center of Biotechnology Information (NCBI) taxonomic database (Schoch et al., 2020), the Global Biodiversity Information Facility (GBIF) taxonomic backbone (GBIF Secretariat, 2022), and the Interim Register of Marine and Non-marine Genera (IRMNG) database (Rees, Vandepitte, Decock, \& Vanhoorne, 2017).
 
 Besides binomial species names, DateLife accepts scientific names from any inclusive taxonomic group (e.g., genus, family, tribe), as well as subspecific taxonomic variants (e.g., subspecies, variants, strains).
-If a taxon name belongs to an inclusive taxonomic group, DateLife has two alternative behaviors defined by the \enquote{get species from taxon} flag.
+If a taxon name belongs to an inclusive taxonomic group, DateLife has two alternative behaviors defined by the ``get species from taxon'' flag.
 If the flag is active, DateLife retrieves all species names within the taxonomic group from the standard taxonomy of choice, and adds them to the search query.
@@
 At the time of writing of this manuscript
-(Feb 21, 2023),
+(May 04, 2023),
 DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies.
@@
 The updated database is assigned a new version number, followed by a package release on CRAN.
-Users can directly implement \texttt{datelife} functions to trigger an update of the local chronogram database, to incorporate any new chronograms to their DateLife analysis before an official database update is released on CRAN.
+Users can directly run \texttt{datelife} functions to trigger an update of their local chronogram database, to incorporate any new chronograms to their DateLife analysis before an official database update is released on CRAN.
 
@@
 Total distance (in units of millions of years) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Figure \ref{fig:workflow}).
-The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a \enquote{phylo} object or newick string.
+The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a ``phylo'' object or newick string.
 Finally, the patristic matrices are associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.
@@
 \item
-  The matching pruned chronograms as newick strings or \enquote{phylo} objects.
+  The matching pruned chronograms as newick strings or ``phylo'' objects.
 \item
@@
 In rare cases, a group of trees can have multiple groves.
-By default, DateLife chooses the grove with the most taxa, however, the \enquote{criterion = trees} flag allows the user to choose the grove with the most trees instead.
+By default, DateLife chooses the grove with the most taxa, however, the ``criterion = trees'' flag allows the user to choose the grove with the most trees instead.
 The result is a single summary (or supertree) topology, that combines topologies from source chronograms in a grove.
@@
 
-For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method (SDM; Criscuolo et al., 2006), which deforms patristic distance matrices by minimizing variance and then averaging them.
+For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method {[}SDM; Criscuolo et al.
+(2006){]}, which deforms patristic distance matrices by minimizing variance and then averaging them.
 These single summary taxon pair age matrices are stored as summarized calibrations that can be used as secondary calibrations to date a tree topology - with or without initial branch lengths, using phylogenetic dating methods currently supported within DateLife: BLADJ (Webb, Ackerly, \& Kembel, 2008; Webb \& Donoghue, 2005), MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), and treePL (Smith \& O'Meara, 2012).
@@
 \textbf{\emph{With no branch lengths.--}}
-When producing or obtaining a tree with branch lengths for a group of interest is not possible, DateLife can date a topology without branch lengths, obtained from OpenTree or by implementing the supertree approach described above, by implementing the Branch Length Adjuster (BLADJ; Webb et al., 2008; Webb \& Donoghue, 2005) algorithm, which requires no initial branch lengths.
+When producing or obtaining a tree with branch lengths for a group of interest is not possible, DateLife can date a topology without branch lengths, obtained from OpenTree or by implementing the supertree approach described above, by applying the Branch Length Adjuster {[}BLADJ; Webb et al.
+(2008); Webb and Donoghue (2005){]} algorithm, which requires no initial branch lengths.
 The algorithm starts by fixing ages for nodes with calibration data upon the given topology.
-Then, it distributes time for nodes with no data evenly between calibrated nodes, minimizing age variance in the resulting chronogram (Webb et al., 2008), which has proven useful for ecological analyses that require age data (Webb et al., 2008).
+Then, it distributes time for nodes with no data evenly between calibrated nodes, minimizing age variance in the resulting chronogram (Webb et al., 2008).
+This approach has has proven useful for ecological analyses that require age data (Webb et al., 2008).
 When there is conflict between ages of calibrated nodes, BLADJ ignores node ages that are older than the age of a parent node.
@@
 
-In the absence of genetic data, we consider that BLADJ is the most agnostic way to assign ages to nodes with no available data, as it does not require any assumptions on the underlying model of branch length distribution.
+In the absence of genetic data, BLADJ is a very agnostic way to assign ages to nodes with no available data, as it does not require any assumptions on the underlying model of branch length distribution.
 It is however common practice in the literature to use a birth-death model to assign ages to nodes with no genetic data (Jetz, Thomas, Joy, Hartmann, \& Mooers, 2012; Rabosky et al., 2018; Smith \& Brown, 2018).
@@
 \textbf{\emph{Dating a tree with branch lengths.--}}
-Relative branch lengths provide key information for phylogenetic dating, specifically for nodes without any calibration data available.
-While, using initial branch lengths data is the golden standard for phylogenetic dating analyses, producing such requires assembling and curating a homology hypothesis, and choosing and implementing a method for phylogenetic inference.
-DateLife implements a workflow to streamline this process by leveraging on genetic data from the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain markers for input taxa.
-By default, mined genetic sequences are aligned with MUSCLE (Edgar, 2004) using functions from the msa R package (Bodenhofer et al., 2015).
+Relative branch lengths can provide key information for phylogenetic dating, specifically for nodes without any calibration data available.
+While, using initial branch lengths data is the golden standard for phylogenetic dating analyses, producing such trees requires assembling and curating a homology hypothesis, and choosing and implementing a method for phylogenetic inference.
+DateLife implements a workflow to streamline this process by applying open data from the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain genetic markers for input taxa.
+By default, BOLD genetic sequences are aligned with MUSCLE (Edgar, 2004) using functions from the msa R package (Bodenhofer et al., 2015).
 Alternatively, sequences can be aligned with MAFFT (Katoh, Asimenos, \& Toh, 2009), using functions from the ape R package (Paradis et al., 2004).
@@
 Both parsimony and ML optimizations are done with functions from the \texttt{phangorn} package (Schliep, 2011).
-The BOLD workflow is currently only available on DateLife's R package and not on the web application.
+Due to the computing load it requires, the BOLD workflow is currently only supported through DateLife's R package. It is not yet available through the web application.
 
@@
 
-Finally, users can save all source and summary chronograms in formats allowing for reuse and reanalysis, such as newick and the R \enquote{phylo} format.
+Finally, users can save all source and summary chronograms in formats allowing for reuse and reanalysis, such as newick and the R ``phylo'' format.
 Input and summary chronograms can be visualized and compared graphically, and users can construct their own graphs using DateLife's chronogram plot generation functions available from the R package \texttt{datelifeplot} (Sanchez-Reyes \& O'Meara, 2022).
@@
 DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and congruified and mapped age data to nodes in this chosen topology (Table \ref{tbl:table1}).
-The name processing step allowed including five data points for node \enquote{n4} (parent of \emph{Schoeniclus elegans}; Fig.
+The name processing step allowed including five data points for node ``n4'' (parent of \emph{Schoeniclus elegans}; Fig.
 \ref{fig:figure2}A) that would not have had any data otherwise due to name mismatch.
 Age summary statistics per node were calculated (Table \ref{tbl:table2}) and used as calibrations to date the tree topology using the BLADJ algorithm.
-As expected, more inclusive nodes (e.g., node \enquote{n1}) have more variance in age data than less inclusive nodes (e.g., node \enquote{n5}).
-Summary age data for node \enquote{n2} were excluded as final calibration because they are older than age data of the more inclusive node, \enquote{n1} (Fig. \ref{fig:figure2}C4).
+As expected, more inclusive nodes (e.g., node ``n1'') have more variance in age data than less inclusive nodes (e.g., node ``n5'').
+Summary age data for node ``n2'' were excluded as final calibration because they are older than age data of the more inclusive node, ``n1'' (Fig. \ref{fig:figure2}C4).
 
@@
 \textbf{\emph{Creating a query.--}}
-To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the \enquote{get species from taxon} flag,
+To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the ``get species from taxon'' flag,
 which gets all recognized species names within a named group from a taxonomy of choice.
@@
 The final step entailed congruifying and summarizing the age data available for the Fringillidae species into two single summary chronograms, using two different types of summary ages, median and SDM.
-As explained in the \enquote{Description} section, a tree topology to summarize age data upon is required.
+As explained in the ``Description'' section, a tree topology to summarize age data upon is required.
 By default, DateLife uses the topology from OpenTree's synthetic tree that contains all taxa from the search query.
@@
 We congruified node ages extracted from chronograms from all other studies upon the individual topologies, effectively excluding original ages from each topology.
-Average node ages per node were applied as secondary calibrations, smoothed with the BLADJ algorithm.
+Finally, average node ages per node were applied as secondary calibrations and smoothed with the BLADJ algorithm.
 We found that node ages from the original studies, and ages estimated using all other age data available are largely correlated (Fig. \ref{fig:cvXY}).
-For five studies, DateLife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka \enquote{closer to the root}) relative to the original estimate, and overestimate ages for nodes closer to the tips.
+For five studies, DateLife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka `closer to the root') relative to the original estimate, and overestimate ages for nodes closer to the tips.
 Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig.
@@
 
-DateLife makes state-of-the-art data on evolutionary time frame easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
-It is an open service that does not require any expert biological knowledge from users --besides the names of the species or group they want to work with, for any of its functionality.
+DateLife's goal is to improve availability and accessibility of state-of-the-art data on evolutionary time frame of organisms, to allow users from all areas of science and with all levels of expertise to compare, reuse, and reanalyse expert age data for their own applications.
+As such, it is designed as an open service that does not require any expert biological knowledge --besides the scientific names of the species or group users want to work with, to use any of its functionalities.
 
@@
 
-While it is known that incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance in ecological and conservation biology studies (Webb et al., 2008).
-Next, we discuss some particularities of using summarized node age data from a DateLife analysis that users should be aware of.
+While we envision that DateLife will have many interesting applications in research and beyond, we recognise that as
+any other automated tool for biological data analysis, DateLife's results can only be as good as the input data, and as give as good as how we decide to apply them for research.\\
+We recognise that this might be a problem in some research areas, but it might not be the case for all.
+For example, in ecological and conservation biology studies, it has been seen that incorporating some age data provides better results than when not using any age data at all, even if the age data is not accurate (Webb et al., 2008).
+In the following sections we discuss the particularities of divergence times resulting from a DateLife analysis for consideration of the readers and users in different research areas.
 
@@
 See, for example, the robust ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández, 2015; Ramshaw et al., 1972; Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021).
-Source chronograms available for the same organisms have potentially been estimated implementing calibrations very differently.
+Alternative source chronograms available for the same taxa have potentially been estimated implementing calibrations very differently.
 In the Fringillidae example above, the chronograms from Burns et al.
@@
 Yet, if one researcher concludes that a fossil should calibrate the ingroup of a clade, while another researcher concludes that the same fossil should calibrate the outgroup of the clade, the resulting age estimates will differ, as the placement of calibrations as stem or crown group has been proven to significantly affect time of lineage divergence estimates (Sauquet, 2013).
+In addition, nodes that are sandwiched between a calibrated node and a calibrated root have less freedom of movement and hence narrower confidence intervals (Vos \& Mooers, 2004), this inflates precision for nodes without calibrations but does not necessarily improve accuracy of the estimated ages.
 
@@
 (2017).
-There have been attempts to create fossil calibration databases (Ksepka et al., 2015), though these still have room to grow.
+
+The Fossil Calibration Database provides availability of 220 primary calibration points encompassing flowering plants and metazoans, that have been curated by experts and used in peer-reviewd publications (Ksepka et al., 2015).
+The database facilitates the use of expert primary fossil calibrations in new phylogenetic dating analysis.
+Despite the wealth of information displayed in the database, users still need the expertise to locate the equivalent node in their phylogenies to be dated.
+Open databases for reuse of fossil calibration data still have room to grow and should be a great addition to the phylogenetics tool set.
 
@@
 Hence, DateLife's workflow features different ways of assigning node ages in the absence of calibrations and branch length information for certain taxa.
-Completing a phylogeny using a stochastic birth-death politomy resolver and a backbone taxonomy is a common practice in scientific publications: Jetz et al.
+Completing a phylogeny using a stochastic birth-death polytomy resolver and a backbone taxonomy is a common practice in scientific publications: Jetz et al.
 (2012), created a chronogram of all 9, 993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
@@
 
-Taxonomy based politomy resolvers also introduce topological differences.
+Taxonomy based polytomy resolvers also introduce topological differences.
 The study of macroevolutionary processes largely depends on an understanding of the timing of species diversification events, and different phylogenetic and chronogram hypothesis can provide very different overviews of the macroevolutionary history of a biological group.
@@
 It is also crucial for education, science communication and policy, but generating chronograms is difficult, especially for those who want to use phylogenies but who are not systematists, or do not have the time to acquire and develop the necessary knowledge and skills to construct them on their own.
-Importantly, years of primarily publicly funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
-
-The DateLife project allows for easy and fast summary of public and state-of-the-art data on time of lineage divergence.
+Importantly, years of primarily publicly funded research have resulted in vast amounts of chronograms that are already available in scientific publications, but functionally hidden from the public and scientific community for reuse.
+
+The DateLife project allows for easy and fast summarization of public and state-of-the-art data on time of lineage divergence.
 It provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identifying regions that require more research, or that have conflicting information.
-It is available as an R package, and as a web-based R shiny application at www.datelife.org
+It is available as an R package, and as a web-based R shiny application at www.datelife.org.
 Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
-The DateLife project should improve awareness of the existing variation in expert time of divergence data, and foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
+The DateLife project will improve awareness of the existing variation in expert estimations time of divergence data, and foster exploration of the effect of alternative divergence time hypotheses on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
 
@@
 The DateLife software is free and open source.
-It can be used online through its R shiny web application hosted at \url{http://www.datelife.org}, and locally through the \texttt{datelife} R package, downloadable from Zenodo (\url{https://doi.org/10.5281/zenodo.593938}) and the CRAN repository (Sanchez-Reyes et al., 2022).
-DateLife's web application is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker and OpenTree's infrastructure (dates.opentreeoflife.org/datelife).
-\texttt{datelife}'s stable version is available for installation from CRAN repository using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R.
+It can be used online through its R shiny web application hosted at \url{http://www.datelife.org}, and locally through the \texttt{datelife} R package, available from Zenodo (\url{https://doi.org/10.5281/zenodo.593938} and the CRAN repository (Sanchez-Reyes et al., 2022).
+DateLife's web application is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker and OpenTree's infrastructure (datelife.opentreeoflife.org).
+\texttt{datelife}'s stable version can be installed from the CRAN repository using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R.
 Development versions are available from DateLife's GitHub repository (\url{https://github.com/phylotastic/datelife}) and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.
@@
 Supplementary material, including code, biological examples, benchmark results, data files and online-only appendices, can be found in the Dryad data repository (\url{https://doi.org/10.5061/dryad.cnp5hqc6w}), as well as in the Zenodo stable repositories that host the reproducible manuscript (\url{https://doi.org/10.5281/zenodo.7435094}), the biological examples (\url{https://doi.org/10.5281/zenodo.7435101}), and the software benchmark (\url{https://doi.org/10.5281/zenodo.7435106}).
-The corresponding development versions hosted on GitHub can be found at \url{https://github.com/LunaSare/datelifeMS1\#readme}, \url{https://github.com/LunaSare/datelife_examples\#readme}, and \url{https://github.com/LunaSare/datelife_benchmark\#readme}.
+The corresponding development versions hosted on GitHub can be found at \url{https://github.com/LunaSare/datelifeMS1}, \url{https://github.com/LunaSare/datelife_examples}, and \url{https://github.com/LunaSare/datelife_benchmark}.
 
@@
 \hypertarget{refs}{}
-\leavevmode\hypertarget{ref-agnarsson2008acctran}{}%
-Agnarsson, I., \& Miller, J. A. (2008). Is acctran better than deltran? \emph{Cladistics}, \emph{24}(6), 1032--1038.
-
-\leavevmode\hypertarget{ref-alstrom2014discovery}{}%
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-agnarsson2008acctran}{}}%
+Agnarsson, I., \& Miller, J. A. (2008). {Is ACCTRAN better than DELTRAN?} \emph{Cladistics}, \emph{24}(6), 1032--1038.
+
+\leavevmode\vadjust pre{\hypertarget{ref-alstrom2014discovery}{}}%
 Alström, P., Hooper, D.
@@
 
-\leavevmode\hypertarget{ref-ane2009groves}{}%
-Ané, C., Eulenstein, O., Piaggio-Talice, R., \& Sanderson, M. J. (2009). Groves of phylogenetic trees. \emph{Annals of Combinatorics}, \emph{13}(2), 139--167.
-
-\leavevmode\hypertarget{ref-antonelli2017supersmart}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-ane2009groves}{}}%
+Ané, C., Eulenstein, O., Piaggio-Talice, R., \& Sanderson, M. J. (2009). Groves of phylogenetic trees. \emph{{Annals of Combinatorics}}, \emph{13}(2), 139--167.
+
+\leavevmode\vadjust pre{\hypertarget{ref-antonelli2017supersmart}{}}%
 Antonelli, A., Hettling, H., Condamine, F.
@@
 
-\leavevmode\hypertarget{ref-archie1986newick}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-archie1986newick}{}}%
 Archie, J., Day, W.
@@
 (1986).
-The Newick tree format.
-Retrieved from \href{\%7Bhttps://evolution.genetics.washington.edu/phylip/newicktree.html\%7D}{\{https://evolution.genetics.washington.edu/phylip/newicktree.html\}}
-
-\leavevmode\hypertarget{ref-avibase-emberiza}{}%
+{The Newick tree format}.
+Retrieved from \url{\%7Bhttps://evolution.genetics.washington.edu/phylip/newicktree.html\%7D}
+
+\leavevmode\vadjust pre{\hypertarget{ref-avibase-emberiza}{}}%
 Avibase.
 (2022).
-Yellow-throated Bunting.
-\emph{Avibase - the World Bird Database}, (Online Resource).
-Retrieved from \href{\%7Bhttps://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\%7D}{\{https://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\}}
-
-\leavevmode\hypertarget{ref-Bapst2012a}{}%
+{Yellow-throated Bunting}.
+\emph{{Avibase - The World Bird Database}}, (Online Resource).
+Retrieved from \url{\%7Bhttps://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\%7D}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Bapst2012a}{}}%
 Bapst, D.
@@
 (2012).
-Paleotree: An R package for paleontological and phylogenetic analyses of evolution.
-\emph{Methods in Ecology and Evolution}, \emph{3}(5), 803--807.
+{Paleotree: An R package for paleontological and phylogenetic analyses of evolution}.
+\emph{{Methods in Ecology and Evolution}}, \emph{3}(5), 803--807.
 \url{https://doi.org/10.1111/j.2041-210X.2012.00223.x}
 
-\leavevmode\hypertarget{ref-barba2018constraining}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-barba2018constraining}{}}%
 Barba-Montoya, J., Reis, M.
@@
 
-\leavevmode\hypertarget{ref-barker2014mitogenomic}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-barker2014mitogenomic}{}}%
 Barker, F. K. (2014). Mitogenomic data resolve basal relationships among passeriform and passeridan birds. \emph{Molecular Phylogenetics and Evolution}, \emph{79}, 313--324.
 
-\leavevmode\hypertarget{ref-barker2013going}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-barker2013going}{}}%
 Barker, F.
@@
 Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds.
-\emph{Systematic Biology}, \emph{62}(2), 298--320.
-
-\leavevmode\hypertarget{ref-barker2015new}{}%
+\emph{{Systematic Biology}}, \emph{62}(2), 298--320.
+
+\leavevmode\vadjust pre{\hypertarget{ref-barker2015new}{}}%
 Barker, F.
@@
 
-\leavevmode\hypertarget{ref-barker2004phylogeny}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-barker2004phylogeny}{}}%
 Barker, F.
@@
 
-\leavevmode\hypertarget{ref-beresford2005african}{}%
-Beresford, P., Barker, F., Ryan, P., \& Crowe, T.
+\leavevmode\vadjust pre{\hypertarget{ref-beresford2005african}{}}%
+Beresford, P., Barker, F., Ryan, Pg., \& Crowe, T.
 (2005).
-African endemics span the tree of songbirds (passeri): Molecular systematics of several evolutionary ``enigmas''.
-\emph{Proceedings of the Royal Society B: Biological Sciences}, \emph{272}(1565), 849--858.
-
-\leavevmode\hypertarget{ref-bodenhofer2015msa}{}%
+African endemics span the tree of songbirds (passeri): Molecular systematics of several evolutionary {``enigmas.''} \emph{Proceedings of the Royal Society B: Biological Sciences}, \emph{272}(1565), 849--858.
+
+\leavevmode\vadjust pre{\hypertarget{ref-bodenhofer2015msa}{}}%
 Bodenhofer, U., Bonatesta, E., Horejš-Kainrath, C., \& Hochreiter, S. (2015). Msa: An r package for multiple sequence alignment. \emph{Bioinformatics}, \emph{31}(24), 3997--3999.
 
-\leavevmode\hypertarget{ref-Boyle2013}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Boyle2013}{}}%
 Boyle, B., Hopkins, N., Lu, Z., Raygoza Garay, J.
@@
 (2013).
-The taxonomic name resolution service: An online tool for automated standardization of plant names.
-\emph{BMC Bioinformatics}, \emph{14}(1).
+{The taxonomic name resolution service: An online tool for automated standardization of plant names}.
+\emph{{BMC Bioinformatics}}, \emph{14}(1).
 \url{https://doi.org/10.1186/1471-2105-14-16}
 
-\leavevmode\hypertarget{ref-Britton2007}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Britton2007}{}}%
 Britton, T., Anderson, C.
@@
 (2007).
-Estimating Divergence Times in Large Phylogenetic Trees.
-\emph{Systematic Biology}, \emph{56}(788777878), 741--752.
+{Estimating Divergence Times in Large Phylogenetic Trees}.
+\emph{{Systematic Biology}}, \emph{56}(788777878), 741--752.
 \url{https://doi.org/10.1080/10635150701613783}
 
-\leavevmode\hypertarget{ref-bryson2014diversification}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-bryson2014diversification}{}}%
 Bryson Jr, R.
@@
 (2014).
-Diversification across the new world within the `blue'cardinalids (aves: Cardinalidae).
+Diversification across the new world within the `blue'cardinalids (aves: cardinalidae).
 \emph{Journal of Biogeography}, \emph{41}(3), 587--599.
 
-\leavevmode\hypertarget{ref-burleigh2015building}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-burleigh2015building}{}}%
 Burleigh, J. G., Kimball, R. T., \& Braun, E. L. (2015). Building the avian tree of life using a large-scale, sparse supermatrix. \emph{Molecular Phylogenetics and Evolution}, \emph{84}, 53--63.
 
-\leavevmode\hypertarget{ref-burns2014phylogenetics}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-burns2014phylogenetics}{}}%
 Burns, K.
@@
 Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds.
-\emph{Molecular Phylogenetics and Evolution}, \emph{75}, 41--77.
-
-\leavevmode\hypertarget{ref-Chamberlain2018}{}%
-Chamberlain, S. (2018). \emph{bold: Interface to Bold Systems API}. Retrieved from \url{https://CRAN.R-project.org/package=bold}
-
-\leavevmode\hypertarget{ref-Chamberlain2013}{}%
+\emph{{Molecular Phylogenetics and Evolution}}, \emph{75}, 41--77.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Chamberlain2013}{}}%
 Chamberlain, S.
@@
 (2013).
-taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}.
-\emph{F1000Research}, \emph{2}(191), 1--29.
+{taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}}.
+\emph{{F1000Research}}, \emph{2}(191), 1--29.
 \url{https://doi.org/10.12688/f1000research.2-191.v2}
 
-\leavevmode\hypertarget{ref-chang2020estimating}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Chamberlain2018}{}}%
+Chamberlain, S.
+A., Szöcs, E., Foster, Z., Arendsee, Z., Boettiger, C., Ram, K., \ldots{} Li, G.
+(2019).
+\emph{{taxize: Taxonomic information from around the web}}.
+Retrieved from \url{https://github.com/ropensci/taxize}
+
+\leavevmode\vadjust pre{\hypertarget{ref-chang2020estimating}{}}%
 Chang, J., Rabosky, D.
@@
 
-\leavevmode\hypertarget{ref-chaves2013biogeography}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-chaves2013biogeography}{}}%
 Chaves, J.
@@
 
-\leavevmode\hypertarget{ref-claramunt2015new}{}%
-Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{Science Advances}, \emph{1}(11), e1501005.
-
-\leavevmode\hypertarget{ref-Criscuolo2006}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-claramunt2015new}{}}%
+Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{{Science Advances}}, \emph{1}(11), e1501005.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Criscuolo2006}{}}%
 Criscuolo, A., Berry, V., Douzery, E.
-J., \& Gascuel, O.
+J.
+P., \& Gascuel, O.
 (2006).
-SDM: A fast distance-based approach for (super)tree building in phylogenomics.
-\emph{Systematic Biology}, \emph{55}(5), 740--755.
+{SDM: A fast distance-based approach for (super)tree building in phylogenomics}.
+\emph{{Systematic Biology}}, \emph{55}(5), 740--755.
 \url{https://doi.org/10.1080/10635150600969872}
 
-\leavevmode\hypertarget{ref-cusimano2010slowdowns}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-cusimano2010slowdowns}{}}%
 Cusimano, N., \& Renner, S. S. (2010). Slowdowns in diversification rates from real phylogenies may not be real. \emph{Systematic Biology}, \emph{59}(4), 458--464.
 
-\leavevmode\hypertarget{ref-cusimano2012new}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-cusimano2012new}{}}%
 Cusimano, N., Stadler, T., \& Renner, S.
@@
 
-\leavevmode\hypertarget{ref-delsuc2018phylogenomic}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-delsuc2018phylogenomic}{}}%
 Delsuc, F., Philippe, H., Tsagkogeorga, G., Simion, P., Tilak, M.-K., Turon, X., \ldots{} Douzery, E.
@@
 
-\leavevmode\hypertarget{ref-Eastman2013}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Eastman2013}{}}%
 Eastman, J.
@@
 (2013).
-Congruification: Support for time scaling large phylogenetic trees.
-\emph{Methods in Ecology and Evolution}, \emph{4}(7), 688--691.
+{Congruification: Support for time scaling large phylogenetic trees}.
+\emph{{Methods in Ecology and Evolution}}, \emph{4}(7), 688--691.
 \url{https://doi.org/10.1111/2041-210X.12051}
 
-\leavevmode\hypertarget{ref-edgar2004muscle}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-edgar2004muscle}{}}%
 Edgar, R. C. (2004). MUSCLE: Multiple sequence alignment with high accuracy and high throughput. \emph{Nucleic Acids Research}, \emph{32}(5), 1792--1797.
 
-\leavevmode\hypertarget{ref-Felsenstein1985a}{}%
-Felsenstein, J. (1985). Phylogenies and the Comparative Method. \emph{The American Naturalist}, \emph{125}(1), 1--15. Retrieved from \url{http://www.jstor.org/stable/2461605}
-
-\leavevmode\hypertarget{ref-forest2005teasing}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Felsenstein1985a}{}}%
+Felsenstein, J. (1985). {Phylogenies and the Comparative Method}. \emph{The American Naturalist}, \emph{125}(1), 1--15. Retrieved from \url{http://www.jstor.org/stable/2461605}
+
+\leavevmode\vadjust pre{\hypertarget{ref-forest2005teasing}{}}%
 Forest, F., Savolainen, V., Chase, M.
@@
 
-\leavevmode\hypertarget{ref-freckleton2002phylogenetic}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-freckleton2002phylogenetic}{}}%
 Freckleton, R. P., Harvey, P. H., \& Pagel, M. (2002). Phylogenetic analysis and comparative data: A test and review of evidence. \emph{The American Naturalist}.
 
-\leavevmode\hypertarget{ref-gbif2022taxonomy}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-gbif2022taxonomy}{}}%
 GBIF Secretariat.
 (2022).
-GBIF Backbone Taxonomy.
-\emph{Checklist dataset}, (Online Resource accessed via GBIF.org).
+{GBIF Backbone Taxonomy}.
+\emph{{Checklist Dataset}}, (Online Resource accessed via GBIF.org).
 Retrieved from \href{\%7Bhttps://doi.org/10.15468/39omei\%20\%7D}{\{https://doi.org/10.15468/39omei \}}
 
-\leavevmode\hypertarget{ref-gibb2015new}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-gibb2015new}{}}%
 Gibb, G.
@@
 New zealand passerines help clarify the diversification of major songbird lineages during the oligocene.
-\emph{Genome Biology and Evolution}, \emph{7}(11), 2983--2995.
-
-\leavevmode\hypertarget{ref-graur2004reading}{}%
+\emph{{Genome Biology and Evolution}}, \emph{7}(11), 2983--2995.
+
+\leavevmode\vadjust pre{\hypertarget{ref-graur2004reading}{}}%
 Graur, D., \& Martin, W. (2004). Reading the entrails of chickens: Molecular timescales of evolution and the illusion of precision. \emph{TRENDS in Genetics}, \emph{20}(2), 80--86.
 
-\leavevmode\hypertarget{ref-hackett2008phylogenomic}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-hackett2008phylogenomic}{}}%
 Hackett, S.
@@
 L., Braun, M.
-J., \ldots{} others.
+J., et al.others.
 (2008).
@@
 
-\leavevmode\hypertarget{ref-harvey1991comparative}{}%
-Harvey, P. H., Pagel, M. D., \& others. (1991). \emph{The comparative method in evolutionary biology} (Vol. 239). Oxford university press Oxford.
-
-\leavevmode\hypertarget{ref-Hedges2015}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-harvey1991comparative}{}}%
+Harvey, P. H., Pagel, M. D., et al. (1991). \emph{The comparative method in evolutionary biology} (Vol. 239). Oxford university press Oxford.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Hedges2015}{}}%
 Hedges, S.
@@
 (2015).
-Tree of life reveals clock-like speciation and diversification.
-\emph{Molecular Biology and Evolution}, \emph{32}(4), 835--845.
+{Tree of life reveals clock-like speciation and diversification}.
+\emph{{Molecular Biology and Evolution}}, \emph{32}(4), 835--845.
 \url{https://doi.org/10.1093/molbev/msv037}
 
-\leavevmode\hypertarget{ref-Heibl2008}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Heibl2008}{}}%
 Heibl, C. (2008). \emph{PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages.} Retrieved from \url{http://www.christophheibl.de/Rpackages.html}
 
-\leavevmode\hypertarget{ref-hooper2017chromosomal}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-hooper2017chromosomal}{}}%
 Hooper, D. M., \& Price, T. D. (2017). Chromosomal inversion differences correlate with range overlap in passerine birds. \emph{Nature Ecology \& Evolution}, \emph{1}(10), 1526.
 
-\leavevmode\hypertarget{ref-Huelsenbeck2001}{}%
-Huelsenbeck, J. P., \& Ronquist, F. (2001). MRBAYES: Bayesian inference of phylogenetic trees. \emph{Bioinformatics}, \emph{17}(8), 754--755. \url{https://doi.org/10.1093/bioinformatics/17.8.754}
-
-\leavevmode\hypertarget{ref-Jetz2012}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Huelsenbeck2001}{}}%
+Huelsenbeck, J. P., \& Ronquist, F. (2001). {MRBAYES: Bayesian inference of phylogenetic trees}. \emph{Bioinformatics}, \emph{17}(8), 754--755. \url{https://doi.org/10.1093/bioinformatics/17.8.754}
+
+\leavevmode\vadjust pre{\hypertarget{ref-Jetz2012}{}}%
 Jetz, W., Thomas, G., Joy, J.
-J., Hartmann, K., \& Mooers, A.
+J.
+B., Hartmann, K., \& Mooers, A.
 (2012).
-The global diversity of birds in space and time.
+{The global diversity of birds in space and time}.
 \emph{Nature}, \emph{491}(7424), 444--448.
@@
 
-\leavevmode\hypertarget{ref-johansson2008phylogenetic}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-johansson2008phylogenetic}{}}%
 Johansson, U.
@@
 
-\leavevmode\hypertarget{ref-katoh2009multiple}{}%
-Katoh, K., Asimenos, G., \& Toh, H. (2009). Multiple alignment of dna sequences with mafft. In \emph{Bioinformatics for dna sequence analysis} (pp. 39--64). Springer.
-
-\leavevmode\hypertarget{ref-kimball2019phylogenomic}{}%
-Kimball, R. T., Oliveros, C. H., Wang, N., White, N. D., Barker, F. K., Field, D. J., \ldots{} others. (2019). A phylogenomic supertree of birds. \emph{Diversity}, \emph{11}(7), 109.
-
-\leavevmode\hypertarget{ref-klicka2014comprehensive}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-katoh2009multiple}{}}%
+Katoh, K., Asimenos, G., \& Toh, H. (2009). Multiple alignment of DNA sequences with MAFFT. In \emph{Bioinformatics for DNA sequence analysis} (pp. 39--64). Springer.
+
+\leavevmode\vadjust pre{\hypertarget{ref-kimball2019phylogenomic}{}}%
+Kimball, R. T., Oliveros, C. H., Wang, N., White, N. D., Barker, F. K., Field, D. J., et al.others. (2019). A phylogenomic supertree of birds. \emph{Diversity}, \emph{11}(7), 109.
+
+\leavevmode\vadjust pre{\hypertarget{ref-klicka2014comprehensive}{}}%
 Klicka, J., Barker, F.
@@
 
-\leavevmode\hypertarget{ref-ksepka2015fossil}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-ksepka2015fossil}{}}%
 Ksepka, D.
@@
 T., Cranston, K.
-A., \ldots{} others.
+A., et al.others.
 (2015).
@@
 
-\leavevmode\hypertarget{ref-lamichhaney2015evolution}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-lamichhaney2015evolution}{}}%
 Lamichhaney, S., Berglund, J., Almén, M.
-S., Maqbool, K., Grabherr, M., Martinez-Barrio, A., \ldots{} others.
+S., Maqbool, K., Grabherr, M., Martinez-Barrio, A., et al.others.
 (2015).
@@
 
-\leavevmode\hypertarget{ref-laubichler2009form}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-laubichler2009form}{}}%
 Laubichler, M. D., \& Maienschein, J. (2009). \emph{Form and function in developmental evolution}. Cambridge University Press.
 
-\leavevmode\hypertarget{ref-lepage2004avibase}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-lepage2004avibase}{}}%
 Lepage, D. (2004). \emph{Avibase: The world bird database}. Bird Studies Canada.
 
-\leavevmode\hypertarget{ref-lepage2014avibase}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-lepage2014avibase}{}}%
 Lepage, D., Vaidya, G., \& Guralnick, R. (2014). Avibase--a database system for managing and organizing taxonomic concepts. \emph{ZooKeys}, (420), 117.
 
-\leavevmode\hypertarget{ref-lerner2011multilocus}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-lerner2011multilocus}{}}%
 Lerner, H.
@@
 
-\leavevmode\hypertarget{ref-lovette2010comprehensive}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-lovette2010comprehensive}{}}%
 Lovette, I.
@@
 P., Banks, R.
-C., Fiorentino, I., Córdoba-Córdoba, S., \ldots{} others.
+C., Fiorentino, I., Córdoba-Córdoba, S., et al.others.
 (2010).
@@
 
-\leavevmode\hypertarget{ref-magallon2001absolute}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-magallon2001absolute}{}}%
 Magallon, S., \& Sanderson, M. (2001). Absolute diversification rates in angiosperm clades. \emph{Evolution}, \emph{55}(9), 1762--1780.
 
-\leavevmode\hypertarget{ref-magallon2010using}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-magallon2010using}{}}%
 Magallón, S. (2010). Using fossils to break long branches in molecular dating: A comparison of relaxed clocks applied to the origin of angiosperms. \emph{Systematic Biology}, \emph{59}(4), 384--399.
 
-\leavevmode\hypertarget{ref-magallon2015metacalibrated}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-magallon2015metacalibrated}{}}%
 Magallón, S., Gómez-Acevedo, S., Sánchez-Reyes, L.
@@
 (2016).
-rotl: an R package to interact with the Open Tree of Life data.
-\emph{Methods in Ecology and Evolution}, \emph{7}(12), 1476--1481.
+{rotl: an R package to interact with the Open Tree of Life data}.
+\emph{{Methods in Ecology and Evolution}}, \emph{7}(12), 1476--1481.
 \url{https://doi.org/10.1111/2041-210X.12593}
 
-\leavevmode\hypertarget{ref-Morlon2014}{}%
-Morlon, H. (2014). Phylogenetic approaches for studying diversification. \emph{Ecology Letters}, \emph{17}(4), 508--525. \url{https://doi.org/10.1111/ele.12251}
-
-\leavevmode\hypertarget{ref-moyle2016tectonic}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Morlon2014}{}}%
+Morlon, H. (2014). {Phylogenetic approaches for studying diversification.} \emph{{Ecology Letters}}, \emph{17}(4), 508--525. \url{https://doi.org/10.1111/ele.12251}
+
+\leavevmode\vadjust pre{\hypertarget{ref-moyle2016tectonic}{}}%
 Moyle, R.
@@
 
-\leavevmode\hypertarget{ref-oliveros2019earth}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-odeen2011evolution}{}}%
+Ödeen, A., Håstad, O., \& Alström, P. (2011). Evolution of ultraviolet vision in the largest avian radiation-the passerines. \emph{BMC Evolutionary Biology}, \emph{11}(1), 1--8.
+
+\leavevmode\vadjust pre{\hypertarget{ref-oliveros2019earth}{}}%
 Oliveros, C.
@@
 K., Aleixo, A., Andersen, M.
-J., \ldots{} others.
+J., et al.others.
 (2019).
@@
 
-\leavevmode\hypertarget{ref-Ooms2018}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Ooms2018}{}}%
 Ooms, J., \& Chamberlain, S. (2018). \emph{Phylocomr: Interface to 'phylocom'}. Retrieved from \url{https://CRAN.R-project.org/package=phylocomr}
 
-\leavevmode\hypertarget{ref-opentreeAPIs}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-opentreeAPIs}{}}%
 Open Tree Of Life, Redelings, B., Cranston, K.
@@
 (2016).
-Open Tree of Life APIs v3.0.
-\emph{Open Tree of Life Project}, (Online Resources).
-Retrieved from \href{\%7Bhttps://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\%7D}{\{https://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\}}
-
-\leavevmode\hypertarget{ref-opentreeoflife2019synth}{}%
+{Open Tree of Life APIs v3.0}.
+\emph{{Open Tree of Life Project}}, (Online Resources).
+Retrieved from \url{\%7Bhttps://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\%7D}
+
+\leavevmode\vadjust pre{\hypertarget{ref-opentreeoflife2019synth}{}}%
 Open Tree Of Life, Redelings, B., Sánchez Reyes, L.
@@
 
-\leavevmode\hypertarget{ref-odeen2011evolution}{}%
-Ödeen, A., Håstad, O., \& Alström, P. (2011). Evolution of ultraviolet vision in the largest avian radiation-the passerines. \emph{BMC Evolutionary Biology}, \emph{11}(1), 1--8.
-
-\leavevmode\hypertarget{ref-paradis2004}{}%
-Paradis, E., Claude, J., \& Strimmer, K. (2004). APE: analyses of phylogenetics and evolution in R language. \emph{Bioinformatics}, \emph{20}(2), 289--290.
-
-\leavevmode\hypertarget{ref-parchman2007coevolution}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-packert2012horizontal}{}}%
+Päckert, M., Martens, J., Sun, Y.-H., Severinghaus, L.
+L., Nazarenko, A.
+A., Ting, J., \ldots{} Tietze, D.
+T.
+(2012).
+Horizontal and elevational phylogeographic patterns of himalayan and southeast asian forest passerines (aves: passeriformes).
+\emph{Journal of Biogeography}, \emph{39}(3), 556--573.
+
+\leavevmode\vadjust pre{\hypertarget{ref-paradis2004}{}}%
+Paradis, E., Claude, J., \& Strimmer, K. (2004). {APE: analyses of phylogenetics and evolution in R language}. \emph{Bioinformatics}, \emph{20}(2), 289--290.
+
+\leavevmode\vadjust pre{\hypertarget{ref-parchman2007coevolution}{}}%
 Parchman, T.
@@
 
-\leavevmode\hypertarget{ref-packert2012horizontal}{}%
-Päckert, M., Martens, J., Sun, Y.-H., Severinghaus, L.
-L., Nazarenko, A.
-A., Ting, J., \ldots{} Tietze, D.
-T.
-(2012).
-Horizontal and elevational phylogeographic patterns of himalayan and southeast asian forest passerines (aves: Passeriformes).
-\emph{Journal of Biogeography}, \emph{39}(3), 556--573.
-
-\leavevmode\hypertarget{ref-pennell2014geiger}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-pennell2014geiger}{}}%
 Pennell, M.
@@
 
-\leavevmode\hypertarget{ref-posadas2006historical}{}%
-Posadas, P., Crisci, J. V., \& Katinas, L. (2006). Historical biogeography: A review of its basic concepts and critical issues. \emph{Journal of Arid Environments}, \emph{66}(3), 389--403.
-
-\leavevmode\hypertarget{ref-powell2014comprehensive}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-posadas2006historical}{}}%
+Posadas, P., Crisci, J. V., \& Katinas, L. (2006). Historical biogeography: A review of its basic concepts and critical issues. \emph{{Journal of Arid Environments}}, \emph{66}(3), 389--403.
+
+\leavevmode\vadjust pre{\hypertarget{ref-powell2014comprehensive}{}}%
 Powell, A.
@@
 
-\leavevmode\hypertarget{ref-price2014niche}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-price2014niche}{}}%
 Price, T.
@@
 S., Tietze, D.
-T., Alström, P., \ldots{} others.
+T., Alström, P., et al.others.
 (2014).
@@
 
-\leavevmode\hypertarget{ref-pulgarin2013multilocus}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-pulgarin2013multilocus}{}}%
 Pulgarín-R, P.
@@
 (2013).
-Multilocus phylogeny and biogeography of the new world pheucticus grosbeaks (aves: Cardinalidae).
+Multilocus phylogeny and biogeography of the new world pheucticus grosbeaks (aves: cardinalidae).
 \emph{Molecular Phylogenetics and Evolution}, \emph{69}(3), 1222--1227.
 
-\leavevmode\hypertarget{ref-rabosky2015no}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-RCoreTeam2018}{}}%
+R Core Team. (2018). \emph{{R: a language and environment for statistical computing}}. Vienna, Austria: R Foundation for Statistical Computing.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rabosky2015no}{}}%
 Rabosky, D.
@@
 
-\leavevmode\hypertarget{ref-rabosky2018inverse}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-rabosky2018inverse}{}}%
 Rabosky, D.
@@
 O., Cowman, P.
-F., Sallan, L., Friedman, M., \ldots{} others.
+F., Sallan, L., Friedman, M., et al.others.
 (2018).
@@
 
-\leavevmode\hypertarget{ref-ratnasingham2007bold}{}%
-Ratnasingham, S., \& Hebert, P. D. (2007). BOLD: The barcode of life data system (http://www. Barcodinglife. Org). \emph{Molecular Ecology Notes}, \emph{7}(3), 355--364.
-
-\leavevmode\hypertarget{ref-RCoreTeam2018}{}%
-R Core Team. (2018). \emph{R: a language and environment for statistical computing}. Vienna, Austria: R Foundation for Statistical Computing.
-
-\leavevmode\hypertarget{ref-rees2017automated}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-ratnasingham2007bold}{}}%
+Ratnasingham, S., \& Hebert, P. D. (2007). BOLD: The barcode of life data system (http://www. Barcodinglife. org). \emph{Molecular Ecology Notes}, \emph{7}(3), 355--364.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rees2017automated}{}}%
 Rees, \& Cranston, K. (2017). Automated assembly of a reference taxonomy for phylogenetic data synthesis. \emph{Biodiversity Data Journal}, (5).
 
-\leavevmode\hypertarget{ref-rees2017irmng}{}%
-Rees, Vandepitte, L., Decock, W., \& Vanhoorne, B. (2017). IRMNG 2006--2016: 10 Years of a Global Taxonomic Database. \emph{Biodiversity Informatics}, \emph{12}.
-
-\leavevmode\hypertarget{ref-Revell2012}{}%
-Revell, L. J. (2012). Phytools: An r package for phylogenetic comparative biology (and other things). \emph{Methods in Ecology and Evolution}, \emph{3}, 217--223.
-
-\leavevmode\hypertarget{ref-Ronquist2003}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-rees2017irmng}{}}%
+Rees, Vandepitte, L., Decock, W., \& Vanhoorne, B. (2017). {IRMNG 2006--2016: 10 Years of a Global Taxonomic Database}. \emph{Biodiversity Informatics}, \emph{12}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Revell2012}{}}%
+Revell, L. J. (2012). Phytools: An r package for phylogenetic comparative biology (and other things). \emph{{Methods in Ecology and Evolution}}, \emph{3}, 217--223.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Ronquist2003}{}}%
 Ronquist, F., \& Huelsenbeck, J.
@@
 (2003).
-MrBayes 3: Bayesian phylogenetic inference under mixed models.
+{MrBayes 3: Bayesian phylogenetic inference under mixed models}.
 \emph{Bioinformatics}, \emph{19}(12), 1572--1574.
@@
 
-\leavevmode\hypertarget{ref-roquet2014one}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-roquet2014one}{}}%
 Roquet, C., Lavergne, S., \& Thuiller, W. (2014). One tree to link them all: A phylogenetic dataset for the european tetrapoda. \emph{PLoS Currents}, \emph{6}.
 
-\leavevmode\hypertarget{ref-datelifeplot}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-datelifeplot}{}}%
 Sanchez-Reyes, L.
@@
 (2022).
-\texttt{datelifeplot}: Methods to plot chronograms and outputs of the \texttt{datelife} package.
+{\texttt{datelifeplot}: Methods to plot chronograms and outputs of the \texttt{datelife} package}.
 \emph{R Package Release V0.2.2}.
@@
 
-\leavevmode\hypertarget{ref-sanchez2022}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-sanchez2022}{}}%
 Sanchez-Reyes, L.
@@
 (2022).
-\texttt{datelife}: Scientific Data on Time of Lineage Divergence for Your Taxa.
+{\texttt{datelife}: Scientific Data on Time of Lineage Divergence for Your Taxa}.
 In \emph{R package version 0.6.6}.
@@
 
-\leavevmode\hypertarget{ref-sanderson2002}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-sanderson2002}{}}%
 Sanderson, M.
+J.
 (2002).
-Estimating Absolute Rates of Molecular Evolution and Divergence Times: A Penalized Likelihood Approach.
+{Estimating Absolute Rates of Molecular Evolution and Divergence Times: A Penalized Likelihood Approach}.
 \emph{Molecular Biology and Evolution}, \emph{19}(1), 101--109.
@@
 
-\leavevmode\hypertarget{ref-sanderson2003r8s}{}%
-Sanderson, M. (2003). r8s: Inferring Absolute Rates of Molecular Evolution and Divergence Times in the Absence of a Molecular Clock. \emph{Bioinformatics}, \emph{19}(2), 301--302.
-
-\leavevmode\hypertarget{ref-sanderson2001sources}{}%
-Sanderson, M., \& Doyle, J.
+\leavevmode\vadjust pre{\hypertarget{ref-sanderson2003r8s}{}}%
+Sanderson, M. J. (2003). {r8s: Inferring Absolute Rates of Molecular Evolution and Divergence Times in the Absence of a Molecular Clock}. \emph{Bioinformatics}, \emph{19}(2), 301--302.
+
+\leavevmode\vadjust pre{\hypertarget{ref-sanderson2001sources}{}}%
+Sanderson, M.
+J., \& Doyle, J.
+A.
 (2001).
@@
 
-\leavevmode\hypertarget{ref-sauquet2013practical}{}%
-Sauquet, H. (2013). A practical guide to molecular dating. \emph{Comptes Rendus Palevol}, \emph{12}(6), 355--367.
-
-\leavevmode\hypertarget{ref-sauquet2021age}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-sauquet2013practical}{}}%
+Sauquet, H. (2013). A practical guide to molecular dating. \emph{{Comptes Rendus Palevol}}, \emph{12}(6), 355--367.
+
+\leavevmode\vadjust pre{\hypertarget{ref-sauquet2021age}{}}%
 Sauquet, H., Ramírez-Barahona, S., \& Magallón, S. (2021). \emph{The age of flowering plants is unknown}.
 
-\leavevmode\hypertarget{ref-schenk2016sec}{}%
-Schenk, J. J. (2016). Consequences of secondary calibrations on divergence time estimates. \emph{PLoS ONE}, \emph{11}(1). \url{https://doi.org/10.1371/journal.pone.0148228}
-
-\leavevmode\hypertarget{ref-schliep2011phangorn}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-schenk2016sec}{}}%
+Schenk, J. J. (2016). {Consequences of secondary calibrations on divergence time estimates}. \emph{PLoS ONE}, \emph{11}(1). \url{https://doi.org/10.1371/journal.pone.0148228}
+
+\leavevmode\vadjust pre{\hypertarget{ref-schliep2011phangorn}{}}%
 Schliep, K. P. (2011). Phangorn: Phylogenetic analysis in r. \emph{Bioinformatics}, \emph{27}(4), 592--593.
 
-\leavevmode\hypertarget{ref-schoch2020ncbi}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-schoch2020ncbi}{}}%
 Schoch, C.
 L., Ciufo, S., Domrachev, M., Hotton, C.
-L., Kannan, S., Khovanskaya, R., \ldots{} others.
+L., Kannan, S., Khovanskaya, R., et al.others.
 (2020).
-NCBI Taxonomy: a Comprehensive Update on Curation, Resources and Tools.
+{NCBI Taxonomy: a Comprehensive Update on Curation, Resources and Tools}.
 \emph{Database}, \emph{2020}.
 
-\leavevmode\hypertarget{ref-selvatti2015paleogene}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-selvatti2015paleogene}{}}%
 Selvatti, A.
@@
 
-\leavevmode\hypertarget{ref-smith2018constructing}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-smith2018constructing}{}}%
 Smith, S., \& Brown, J. (2018). Constructing a broadly inclusive seed plant phylogeny. \emph{American Journal of Botany}, \emph{105}(3), 302--314.
 
-\leavevmode\hypertarget{ref-Smith2012}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Smith2012}{}}%
 Smith, S., \& O'Meara, B.
 (2012).
-TreePL: Divergence time estimation using penalized likelihood for large phylogenies.
+{TreePL: Divergence time estimation using penalized likelihood for large phylogenies}.
 \emph{Bioinformatics}, \emph{28}(20), 2689--2690.
@@
 
-\leavevmode\hypertarget{ref-Stoltzfus2013}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Stoltzfus2013}{}}%
 Stoltzfus, A., Lapp, H., Matasci, N., Deus, H., Sidlauskas, B., Zmasek, C.
@@
 (2013).
-Phylotastic!
-Making tree-of-life knowledge accessible, reusable and convenient.
-\emph{BMC Bioinformatics}, \emph{14}.
+{Phylotastic!
+Making tree-of-life knowledge accessible, reusable and convenient}.
+\emph{{BMC Bioinformatics}}, \emph{14}.
 \url{https://doi.org/10.1186/1471-2105-14-158}
 
-\leavevmode\hypertarget{ref-sun2020estimating}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-sun2020estimating}{}}%
 Sun, M., Folk, R.
@@
 
-\leavevmode\hypertarget{ref-tietze2013complete}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-tietze2013complete}{}}%
 Tietze, D.
@@
 (2013).
-Complete phylogeny and historical biogeography of true rosefinches (aves: Carpodacus).
+Complete phylogeny and historical biogeography of true rosefinches (aves: carpodacus).
 \emph{Zoological Journal of the Linnean Society}, \emph{169}(1), 215--234.
 
-\leavevmode\hypertarget{ref-title2016macrophylogenies}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-title2016macrophylogenies}{}}%
 Title, P.
@@
 
-\leavevmode\hypertarget{ref-treplin2008molecular}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-treplin2008molecular}{}}%
 Treplin, S., Siegert, R., Bleidorn, C., Thompson, H.
@@
 
-\leavevmode\hypertarget{ref-uyeda2017evolution}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-uyeda2017evolution}{}}%
 Uyeda, J.
@@
 
-\leavevmode\hypertarget{ref-vos2012nexml}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-vos2012nexml}{}}%
 Vos, R.
@@
 T., Lapp, H., Maddison, W.
-P., \ldots{} others.
+P., et al.others.
 (2012).
@@
 \emph{Systematic Biology}, \emph{61}(4), 675--689.
-\url{https://doi.org/10.1093/sysbio/sys025}
-
-\leavevmode\hypertarget{ref-Webb2000}{}%
-Webb, C. (2000). Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees. \emph{The American Naturalist}, \emph{156}(2), 145--155.
-
-\leavevmode\hypertarget{ref-Webb2008}{}%
+
+\leavevmode\vadjust pre{\hypertarget{ref-vos2004reconstructing}{}}%
+Vos, R. A., \& Mooers, A. Ø. (2004). Reconstructing divergence times for supertrees: A molecular approach. \emph{Phylogenetic Supertrees: Combining Information to Reveal the Tree of Life}, 281--299.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Webb2000}{}}%
+Webb, C. (2000). {Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees}. \emph{{The American Naturalist}}, \emph{156}(2), 145--155.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Webb2008}{}}%
 Webb, C., Ackerly, D., \& Kembel, S.
 (2008).
-Phylocom: Software for the analysis of phylogenetic community structure and trait evolution.
+{Phylocom: Software for the analysis of phylogenetic community structure and trait evolution}.
 \emph{Bioinformatics}, \emph{24}(18), 2098--2100.
@@
 
-\leavevmode\hypertarget{ref-webb2005phylomatic}{}%
-Webb, C., \& Donoghue, M. (2005). Phylomatic: Tree assembly for applied phylogenetics. \emph{Molecular Ecology Notes}, \emph{5}(1), 181--183.
-
-\leavevmode\hypertarget{ref-weir2008calibrating}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-webb2005phylomatic}{}}%
+Webb, C., \& Donoghue, M. (2005). Phylomatic: Tree assembly for applied phylogenetics. \emph{{Molecular Ecology Notes}}, \emph{5}(1), 181--183.
+
+\leavevmode\vadjust pre{\hypertarget{ref-weir2008calibrating}{}}%
 Weir, J., \& Schluter, D. (2008). Calibrating the avian molecular clock. \emph{Molecular Ecology}, \emph{17}(10), 2321--2328.
 
-\leavevmode\hypertarget{ref-zuccon2012phylogenetic}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-zuccon2012phylogenetic}{}}%
 Zuccon, D., Prŷs-Jones, R., Rasmussen, P.
@@
 
+\end{CSLReferences}
+
 \endgroup
 
-\newpage
-
-\captionof{table}{Ages of congruified nodes. See Figure 3, step C2.}
-\label{tbl:table1}
-\includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example.pdf}
-
-\vspace{15mm}
-
-\captionof{table}{Summary of congruified nodes ages. See Figure 3, step C3.}
-
-\label{tbl:table2}
-\includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example-summary.png}
-
-\begin{minipage}[t]{17.3cm}
-\includegraphics{../figures/figure-workflow/figure1-new.pdf}
-\captionof{figure}{\textit{Main DateLife workflow.
-Analyses can be performed via DateLife's interactive website at \url{www.datelife.org}, or using the \texttt{datelife} R package.
-Details on the R functions used to perform the analyses are available from \texttt{datelife}'s R package vignettes at \url{https://phylotastic.org/datelife}.}}
-\label{fig:workflow}
-\end{minipage}
-
-\includegraphics{../figures/fig_runtime_main.pdf}
-\captionof{figure}{\textit{DateLife's benchmarking results.
-Computation time used to process a query and a search across \texttt{datelife}'s chronogram database, relative to number of input taxon names.
-For each N = \{10, 100, 200, ..., 1 000, ...
-, 9 000, 10 000\}, we sampled N species names from the class Aves a hundred times, and then performed a datelife search processing the input names with Taxon Names Resolution Service (TNRS; light gray), and without processing names (dark gray).
-For comparison, we performed a search using an input that had been pre-processed with TNRS (light blue).}}
-\label{fig:runtime_main}
-
-\begin{center}
-\includegraphics{../figures/figure-small-example/figure2.pdf}
-\end{center}
-\vspace{10mm}
-\captionof{figure}{\textit{DateLife analysis results for a small sample of A) 6 bird species within the Passeriformes.
-B) Processed species names were found across 9 chronograms within 6 independent studies (Barker et al.
-(\protect\hyperlink{ref-barker2012going}{2012}), Barker et al.
-(\protect\hyperlink{ref-barker2015new}{2015}), Burns et al.
-(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al.
-(\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al.
-(\protect\hyperlink{ref-Jetz2012}{2012}).) C) This revealed 28 source age data points for the queried species names.
-Summarized age data is used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree, resulting in a summary chronogram of source ages.}}
-
-\label{fig:figure2}
-
-\begin{minipage}{18.2cm}
-\includegraphics{../figures/fringillidae-topologies/fringillidae-topology.pdf}
-\captionof{figure}{Tree topologies extracted from Open Tree of Life's (OpenTree) synthetic phylogenetic tree.
-A) Topology of 2,333 tips and 1,305 internal nodes, encompassing bird species within the family Fringillidae following the NCBI taxonomy (black), as well as all other bird species that share the same Most Recent Common Ancestor (MRCA) node in OpenTree's synthetic tree (purple).
-B) Subtree topology of 289 tips and 253 internal nodes, resulting from pruning species that do not belong to the family Fringillidae according to the NCBI taxonomy (purple branches in topology A).
-Bird species within the Fringillidae are paraphyletic
-(Alström et al. \protect\hyperlink{ref-alstrom2014discovery}{2014},
-Barker, Cibois, Schikler, Feinstein, \& Cracraft \protect\hyperlink{ref-barker2004phylogeny}{2004},
-Barker et al. \protect\hyperlink{ref-barker2013going}{2013},
-Barker \protect\hyperlink{ref-barker2014mitogenomic}{2014},
-Barker et al. \protect\hyperlink{ref-barker2015new}{2015},
-Beresford, Barker, Ryan, \& Crowe \protect\hyperlink{ref-beresford2005african}{2005},
-Bryson Jr et al. \protect\hyperlink{ref-bryson2014diversification}{2014},
-Burleigh, Kimball, \& Braun \protect\hyperlink{ref-burleigh2015building}{2015},
-Burns et al. \protect\hyperlink{ref-burns2014phylogenetics}{2014},
-Chaves, Hidalgo, \& Klicka \protect\hyperlink{ref-chaves2013biogeography}{2013},
-Claramunt \& Cracraft \protect\hyperlink{ref-claramunt2015new}{2015},
-Gibb et al. \protect\hyperlink{ref-gibb2015new}{2015},
-Hackett et al. \protect\hyperlink{ref-hackett2008phylogenomic}{2008},
-Jetz et al. \protect\hyperlink{ref-Jetz2012}{2012},
-Johansson, Fjeldså, \& Bowi \protect\hyperlink{ref-johansson2008phylogenetic}{200},
-Kimball et al. \protect\hyperlink{ref-kimball2019phylogenomic}{2019},
-Klicka et al. \protect\hyperlink{ref-klicka2014comprehensive}{2014},
-Lamichhaney et al. \protect\hyperlink{ref-lamichhaney2015evolution}{2015},
-Lerner, Meyer, James, Hofreiter, \& Fleischer \protect\hyperlink{ref-lerner2011multilocus}{2011},
-Lovette et al. \protect\hyperlink{ref-lovette2010comprehensive}{2010},
-Moyle et al. \protect\hyperlink{ref-moyle2016tectonic}{2016},
-Ödeen, Håstad, \& Alström \protect\hyperlink{ref-odeen2011evolution}{2011},
-Oliveros et al. \protect\hyperlink{ref-oliveros2019earth}{2019},
-Päckert et al. \protect\hyperlink{ref-packert2012horizontal}{2012},
-Parchman, Benkman, \& Mezquida \protect\hyperlink{ref-parchman2007coevolution}{2007},
-Powell et al. \protect\hyperlink{ref-powell2014comprehensive}{2014},
-  Price et al. \protect\hyperlink{ref-price2014niche}{2014},
-  Pulgarín-R, Smith, Bryson Jr, Spellman, \& Klicka \protect\hyperlink{ref-pulgarin2013multilocus}{2013},
-  Selvatti, Gonzaga, \& Moraes Russo \protect\hyperlink{ref-selvatti2015paleogene}{2015},
-  Tietze, Päckert, Martens, Lehmann, \& Sun \protect\hyperlink{ref-tietze2013complete}{2013},
-  Treplin et al. \protect\hyperlink{ref-treplin2008molecular}{2008},
-  Zuccon, Prŷs-Jones, Rasmussen, \& Ericson \protect\hyperlink{ref-zuccon2012phylogenetic}{2012}).}
-\label{fig:fringillidae-topologies}
-\end{minipage}
-
-\begin{minipage}{18cm}
-\begin{center}
-\includegraphics{../figures/figure-fringillidae/median_and_calibration_ages-congruified.png}
-\end{center}
-
-\captionof{figure}{Fringillidae median summary chronogram generated with DateLife.
-It has 289 tips and 253 nodes, from which 212 have age data from at least one published chronogram.
-These nodes have age data points from 19 different chronograms from 13 different studies (colored circles), which were summarized into 194 data points that were used as secondary calibrations (purple slashes) with the Branch Length Adjuster (BLADJ ) software from Webb et al., (\protect\hyperlink{ref-Webb2008}{2008}).}
-\label{fig:fringillidages}
-\end{minipage}
-
-\newpage
-
-\begin{minipage}{18cm}
-\begin{center}
-\includegraphics{../figures/fringillidae-cross-validation/fig-cross-validation-xy-plots.pdf}
-\end{center}
-\captionof{figure}{Results from cross validation analysis. Each plot compares the original age estimate (x axis) with the age obtained with a DateLife analysis (y axis), per node.}
-\label{fig:cvXY}
-\end{minipage}
-
 \end{document}

### 2023-05-04 review.Rmd
add some more reviews

--- 2023-02-17 review.Rmd
@@
 ---
-title: "Response to reviews"
-output: pdf_document
-date: '`r format(Sys.time(), "%b %d, %Y")`'
+title: "Response to decision letter"
+output: 
+  pdf_document:
+    keep_md: yes
+date: "2023-05-01"
 ---
@@
 ```
+
+**Dear _Editor-in-Chief_ Isabel Sanmartín,**
+
+&nbsp;
+&nbsp;
+
+**Please find below response to each point raised by the _Associate Editor_ Daniele Silvestro and the two _Reviewers_. We greatly appreciate your time and consideration for our resubmission.**
+
+&nbsp;
+&nbsp;
+
+**Kind regards,**
+
+&nbsp;
+
+**Luna L. Sanchez Reyes, on behalf of all coauthors.**
+
+&nbsp;
+&nbsp;
 
@@
 
+**Thank you for your decision! We made sure to address all comments carefully and hopefully you find the manuscript improved.**
+
 > Please acknowledge (by email to sysbio.editorialoffice@oup.com) receipt of the reviews and give a probable time frame for the return of your revised manuscript.
@@
 
+**Noted and checked.**
+
 > Log into https://mc.manuscriptcentral.com/systbiol and enter your Author Center, where you will find your manuscript title listed under "Manuscripts with Decisions." Under "Actions" click on "Create a Revision." Your manuscript number will be appended to denote a revision.
@@
 
+**Noted and checked.**
+
 > When submitting your revised manuscript, address each point made by the Editor, AE and Reviewers IN THE SPACE INDICATED (under "Response to Decision Letter").
@@
 If you feel a reviewer did not understand a point you made, in your response keep in mind that as an author it is your responsibility to make your points clear to the readers.
+
+**Noted.**
 
@@
 
+**Noted and checked.**
+
 > The cost of printed color figures is $600 each.
@@
 In your cover letter, please make your intentions regarding the use of color clear.
+
+**Noted.**
 
@@
 Then, in a separate section after the main text, include a statement such as:  
-
-> SUPPLEMENTARY MATERIAL
+> SUPPLEMENTARY MATERIAL. 
 > Supplementary material, including data files and/or online-only appendices, can be found in the Dryad data repository (put your Dryad doi in parentheses here).
 
-We added DOIs for Dryad package with Supplementary Material, L551.
+**We added DOIs for Dryad package with Supplementary Material, L551.**
 
@@
 
-Noted and checked.
+**Noted and checked.**
 
@@
 
-Done on L556-558.
+**Done on L556-558.**
 
@@
 
-Checked.
+**Checked.**
 
@@
 
+**Noted.**
+
 > If you are the first author and your manuscript is based on work done while you were a student, you would be eligible for the "Publisher's Award for Excellence in Systematic Research." When submitting your revised version be sure to indicate student work in the checkbox provided.
@@
 
-This work was done while the first author was a Postdoctoral researcher.
+**This work was done while the first author was a Postdoctoral researcher.**
 
@@
 
-Noted and checked.
+**Noted and checked.**
 
----
----
+***
 
@@
 
-Thanks for your reviews and comments! They were most helpful. We carefully revised the manuscript to address all points raised by the reviewers and yourself and we hope you find the revision suitable.
+**Thanks for your reviews and comments!
+They were most helpful and we believe they greatly improved the ms.
+We carefully revised the manuscript to address all points raised by the reviewers and yourself, and we hope you find the revision suitable.**
 
----
+***
 > In addition to that, please also address the following points:
@@
 
-Thanks for this question. We elaborate on this on the Description section, subsection "Dating a tree with branch lengths" L221-228. 
-Briefly, we implement the parsimony algorithm ACCTRAN using functions from th R package phangorn.
-We chose this algorithm as a quick way to obtaining initial branch lengths that can be optimized under ML.
-ACCTRAN resolves ambiguous character optimization by assigning changes along branches of the tree as close to the root as possible (agnarsson2008acctran), so this works well with the type of matrices generated with BOLD data.
-Once estimated, the parsimony branch lengths become parameters that are then optimized using Maximum Likelihood, given the alignment, the topology and a simple JC model.
+**Thanks for this question.
+We elaborate on this on the Description section, subsection "Dating a tree with branch lengths" L221-228.
+Briefly, we implement the parsimony algorithm ACCTRAN using functions from the R package phangorn.
+We chose this algorithm as a quick way to obtain initial branch lengths that can be optimized afterwards using ML.
+The DNA matrices generated from data mined from the Barcode of Life Database (BOLD) do not have too much variation, so ACCTRAN resolves ambiguous character optimization by assigning changes along branches of the tree as close to the root as possible (Agnarsson & Miller, 2008, doi:10.1111/j.1096-0031.2008.00229.x).Once parsimony branch lengths are estimated, they become parameters that are then optimized using Maximum Likelihood (ML), given the alignment, the topology and a simple JC model.
+The branches that are returned are ML branches.**
 <!--ACCTRAN has been implemented in Forest et al.
@@
 Systematic Botany.-->
-We also add a vignette to the R package in which we showcase the whole workflow of branch length optimization, from parsimony to ML.
+**We added a vignette to the R package in which we showcase the whole workflow of branch length optimization, from parsimony to ML.
+This vignette is available at http://phylotastic.org/datelife/articles/making_bold_trees.html.**
 
----
+***
 > 2. Why are the node ages evenly distributed between calibrations? I would expect an exponential distribution of node ages under a standard birth-death process.
 
-Thanks for pointing this out. We elaborate on this on the "Description" section, at the end of subsection "Dating a tree topology with no branch lengths", L 202-210
-(Note that we expanded the subsection "Dating a tree topology" into three sections: "Applying secondary calibrations", "Dating a tree topology" and "Dating a tree with branch lengths").
+**Thanks for this questions.
+We elaborate on it on the "Description" section, at the end of subsection "Dating a tree topology with no branch lengths", L 202-210.Note that we expanded the subsection "Dating a tree topology" into three sections: "Applying secondary calibrations", "Dating a tree topology" and "Dating a tree with branch lengths".**
 
-Briefly, in the absence of genetic data from an alignment, nodes that have no age data from published studies in the chronogram database are assigned an age based solely on other nodes that do have age data from published studies.
-MrBayes has an option in which you can use a birth-death strict-clock model to sample branch lengths in the absence of genetic data. 
-We have implemented a DateLife function called "make_mrbayes_tree", that wraps this MrBayes service.
+**We want to point out that the main goal of the summarizing step in DateLife (when we distribute ages evenly in the chronogram) is to provide a single chronogram that can quickly show at a glance, and in the most agnostic way possible, the distribution of node ages obtained from published studies.
+Using an exponential distribution for missing node ages would make additional assumptions about the underlying evolutionary model of the chronogram, which can bias results in downstream analyses, as we elaborate in the discussion section, "Effects of phylogenetic sampling on downstream analyses" (L445 of manuscript without differences; L527 of manuscript with differences).**
+
+**In the DateLife package we have implemented a function called "make_mrbayes_tree", that runs the MrBayes algorithm using the option in which you can use a birth-death strict-clock model to sample branch lengths in the absence of genetic data.
 Inherited from MrBayes implementation, the function requires setting parameter values for birth and death rates.
-If the amount of missing data is large, the chosen parameters can determine the "shape" of the resulting summary chronogram in terms of branch length distribution.
+If the amount of missing genetic data is large, the birth and death rate values applied will determine the shape of the chronogram branch length distribution and bias the tree shape (Rabosky 2015, doi:10.1111/evo.12817).**
 
-We added a discussion section, "Effects of phylogenetic sampling on downstream analyses" (L445 of manuscript without differences; L527 of manuscript with differences), in which we discuss the effects of chronograms with missing data that have been completed under a certain diversification model, especially when used for analyses that require estimating a diversification rate, which might introduce circularity (Rabosky, 2015.
-Evolution doi.org/10.1111/evo.12817).
+**We discuss the effects of using this type of chronograms with missing data that have been completed at random following a certain diversification model, especially when used for analyses that require estimating a diversification rate, which might introduce circularity (Rabosky, 2015.
+Evolution doi:10.1111/evo.12817).**
 
-MrBayes also allow us to generate a cloud of trees produced with different birth-death rates that could be used for downstream analyses.
-However, the main goal of the summarizing step in DateLife is to provide a single chronogram that can quickly show in one glance the distribution of node ages based on published data, in the most agnostic way possible.
+<!--MrBayes also allow us to generate a cloud of trees produced with different birth-death rates that could be used for downstream analyses. 
+-->
 
-This is why we chose an algorithm that distributes node ages evenly between calibrations and does not require any assumptions on the underlying model of branch length distribution.
+**This is why we chose an algorithm that distributes node ages evenly between calibrations and does not require any assumptions on the underlying model of branch length distribution.
 The Branch Length Adjuster (BLADJ) algorithm allows this and is the one set as default for the summarizing step.
 However, users that wish to use MrBayes instead con do so with the "make_mrbayes_tree" function.
+**
 
----
+***
 > 4.
@@
 
-We agree, we changed the behaviour of the code to fix this, and adjusted the ms accordingly (L196-201).
+**Thanks for raising this point.
+We agree that this practice could be dangerous if a user is not aware that the age of the root was set in an arbitrary way.
+We changed the behaviour of the code to return a very conspicuous warning message when the root is arbitrary to make sure that users are aware of the fact (L196-201).**
 
-It now stops with an informative message on how users can provide an age for the root. The goal is to make sure that users are aware that there is no age data available for that root!
+**In line with our thoughts on the previous point, we want our program to be able to return a chronogram if there is at least one age data point available from the literature.
+Unfortunately, it is impossible to return a chronogram if there is no age for the root, so in the absence of that information, and to be able to automatize the algorithm, the only option available to us is to randomly chose an age that will allow to generate a chronogram using the real data that is available.
+We agree that choosing the 10% is probably too arbitrary, so the function now adds one standard deviation unit of the mean of ages available to the max age value, if there is more than one age data point, and uses that as the root age.**
 
----
+**The warning message that we implemented for this version of datelife now suggests ways on how users can provide an age for the root that is informed in the literature.
+The main goal of the message is to make sure that users are aware that there is no age data available for that root in the datelife database.**
+
+***
 > Please make sure to carefully revise the text to remove typos.
 
-We revised the text throughout to remove all typos we could detect.
+**We revised the text throughout to remove all typos we could detect.**
 
----
+***
 > As one of the reviewers pointed out, it is good to provide links to permanent repositories for your code.
@@
 
-Right! We added Zenodo links for our code and other materials used for this research in the Availability section (L505) and Supplementary Material (L518-520).
+**Right! We added Zenodo links for our code and other materials used for this research in the Availability section (L505) and Supplementary Material (L518-520).**
 
----
+***
 > I hope you will be willing to revise and resubmit your paper and that you’ll find these and the Reviewers’ comments useful.
@@
 
-Yes, thanks!
+**Yes, thanks!**
 
----
+***
 > Reviewer(s)' comments to author:
@@
 
-Thanks for pointing us to these references.
-We elaborated on this subject more thoroughly by adding results from these studies to the corresponding section on the Discussion, "Effects of phylogenetic sampling on downstream analyses" (L481 of manuscript without differences; L527 of manuscript with differences), as well as changing the wording from "making up" to "in the absence of genetic data, simulating following a birth-death process" (L489; L540).
+**Thanks for pointing us to these references.
+We elaborated on this subject more thoroughly by adding results from these studies to the corresponding section on the Discussion, "Effects of phylogenetic sampling on downstream analyses" (L481 of manuscript without differences; L527 of manuscript with differences), as well as changing the wording from "making up" to "in the absence of genetic data, simulating following a birth-death process" (L489; L540).**
 
----
+***
 > There were many typographical errors in the manuscript which should be corrected prior to publication.
 
-We corrected typos across the text.
+**We corrected typos across the text.**
 
----
+***
 > Reviewer: 2
@@
 
-This is a good point. We would not like for this to become a common use for the software. We mention this in the abstract, introduction and discussion.
+**This is a good point.
+We are aware of this issue, and we think that any biological software is subject to this potential problem.
+For example, if someone generates bad DNA sequences, no matter how good the aligner software they use is, they will end up with a really bad alignment and subsequently a very bad phylogeny.**
 
----
+**We agree that users of all biological software should keep in mind that the results they get are only as good as the data they provide, and that they should implement some quality control.
+We mention this in the abstract and 3rd paragraph of discussion.**
+
+***
 > A separate but related point that I would also like to see discussed is that synthesizing services such as DL and OTOL seem capable of ending up in loops where bad trees with bad calibration points provide the skeleton for further bad trees based on the former - with their own seemingly well-supported but in fact dodgy secondary calibrations.
@@
 
-It is a risk that we are aware of, but it is possible to avoid it. We explain how in the text.
+**Thanks for mentioning this.
+We think DateLife actually helps preventing this issue.
+As shown in Figure 5, DateLife allows to compare all age data available for a group, and users can immediately identify chronograms that are outliers, and explore whether it is related to an artifact of the chronogram or something else about the data or methodology used to generate that chronogram.
+Users can then choose to drop those calibrations from the final analysis.**
 
----
+**We agree that it is a responsibility of the users to check the trees and chonograms before using them as secondary calibrations, in the same way researchers curate fossils to use as primary calibration points.**
+
+***
 > Also related: will we gradually start developing a body of literature with trees where the root always just happens to be ±10% older than the oldest nodes? Might that be bad?
 
-Good point.
-It would certainly be a bad thing.
-The reason we implemented this initially is because we wanted DateLife to be able to provide a summary chronogram even if tehre is just one point of age data available.
+**The Associate Editor also pointed this out, and we agree that this feature of DateLife is not ideal, so we corrected it and address it in the code and in the manuscript.
+First, adding the 10% is too arbitrary.
+If there is more than one age data point, the function now adds one standard deviation unit of the mean of published ages to the maximum age value, and uses that as the root age.
+Second, the function now returns a conspicuous warning message when the root is not based on published data to make sure that users are aware of the fact (L196-201).**
 
-As you pointed out, when the root age is absent, it is not possible to return any dated tree.
+**We want to highlight that the main goal of DateLife is to provide a single chronogram that can quickly show at a glance the distribution of node ages based on published data, so we want our program to be able to return a chronogram if there is at least one age data point available from the literature.
+Unfortunately, as you pointed out, when the root age is absent, it is not possible to return any dated tree.**
+**In the absence of that information, and to be able to automatize the algorithm, the only option available to us is to randomly chose an age that will allow to generate a chronogram using the real data that is available.**
 
----
+**The warning message that we implemented now suggests ways on how users can provide an age for the root that is informed in the literature.
+The main goal of the message is to make sure that users are aware that there is no age data available for that root in the datelife database and that the one used for the chronogram is arbitrary.**
+
+***
 > Apart from these general points that might be touched upon a bit more in the Discussion, here now some specifics about the manuscript:
@@
 
----
+**Thanks for your comment, the abstract was indeed "a bit" forsaken on the version of the ms that you first revised.
+We rewrote most of the abstract, checked for spelling, and added more detail about the findings.**
+
+***
 > - In the first paragraph of the Intro you might want to add something like 'comparative analysis' (Harvey & Pagel, yada yada yada).
@@
 
-Thanks for pointing that out. We added that topic and a couple of references to the intro, L48-49.
+**Thanks for pointing that out. We added that topic to the intro along with a couple of references, L48-49.**
 
----
+***
 > - On page 7, second paragraph, you state that subspecies are ignored.
@@
 
-Thanks for pointing this out.
+**Thanks for pointing this out.
 We actually only ignore subspecies when retrieving data from a more inclusive taxonomic group.
-When provided by the user, subspecific taxa are processed and searched as regular species.
-We clarify this in paragraph X, page 7
+When provided by the user, subspecific taxa are processed and searched fully.
+We clarify this in the Description section "Creating a search query".**
 
-It would be possible to drop the subspecific epithet and perform a more general search.
-We added this functionality to our software development plan, so that it can be soon implemented in a future iteration of the software.
+**It would be possible to drop the subspecific epithet and perform a more general search, so, we added this possibility to our software development plan, so that it can be implemented in a future iteration of the software.**
 
----
+***
 > - On page 7, third paragraph: how does the TNRS deal with homonyms?
@@
 
-Yeah!
+**Yeah!
 this is a pretty cool quality of TNRS that we implement in datelife.
-It is possible to provide or identify the "biological context" of a toxonomic name, so that if a list of names has mainly monkeys, then it will pick _Aotus_ the monkey, but if it has mainly plants, it will pick the Australian legume.
+It is possible to provide or identify the "biological context" of a toxonomic name, so that if a list of names has mainly monkeys, then it will pick _Aotus_ the monkey, but if it has mainly plants, it will pick the Australian legume.**
 
-Moreover, homonyms in the Open Tree taxonomy specify the group they belong too, so users can easily identify if the numeric identifier they obtained from TNRS processing belongs to the group they wanted or not.
-For example _Aotus_ the legume is referred to as "Aotus (genus in kingdom Archaeplastida)" and the monkey as "Aotus (genus in Opisthokonta)".
+**Moreover, homonyms in the Open Tree taxonomy specify the group they belong too, so users can easily identify if the numeric identifier they obtained from TNRS processing belongs to the group they wanted or not.
+For example _Aotus_ the legume is referred to as "Aotus (genus in kingdom Archaeplastida)" and the monkey as "Aotus (genus in Opisthokonta)".**
 
-We added examples for the two points above in a new vignette for the package, available at http://phylotastic.org/datelife/articles/make_datelife_query.html
+**We added examples for the two points above in a new vignette for the package, available at http://phylotastic.org/datelife/articles/make_datelife_query.html**
 
----
+***
 
@@
 
-It is indeed currently synced "manually". However we are working to have the syncs done automatically.
-We comment this in the discussion and clarified that it is currently a manual sync on page 8.
+**It is indeed currently synced "manually". We specify this in the Description, L125. We comment in the discussion the benefits of having syncs done automatically, and ways to do so.**
 
----
+***
 > - On page 10, second paragraph: mining BOLD and aligning the sequences automatically is very cool functionality but I did not see it exposed on the website at all.
@@
 
-Ah, yes, thanks!
+**Ah, yes, thanks for the comment!
 Well, the BOLD workflow is faster than a classic workflow, but it still takes considerable time, specifically computational time required for the database search, which the server can't afford yet.
-The functions are currently only available on DateLife's R package and not on the web application, but we plan on making them available there in the future.
-Regarding possible performance issues, the alignments from BOLD data are not very large (around 1000bp), so we expect both MUSCLE and MAFFT to perform approximately the same, and MUSCLE is the default aligner.
+The functions are currently only available on DateLife's R package and not on the web application, and we think it would be good to make them available there in the future.**
+**Regarding possible performance issues, the alignments from BOLD data are not very long, so we expect both MUSCLE and MAFFT to perform approximately the same.
+We chose MUSCLE as the default aligner.**
 
----
+**We comment on this on L231-232. **
+
+***
 > - On page 25 you mention the fossilcalibrations.org initiative.
@@
 
----
+**Indeed!
+The golden standard for obtaining divergence time estimates entails the use of primary calibrations (fossil or from geologic events).
+Fossil preservation bias makes it impossible to get calibrations to constrain all nodes in a phylogenetic tree, hence, we will always have nodes that we need to infer using a model.**
+**Moreover, uncertainty in placement of fossil calibrations and phylogenetic hypotheses makes it so that there are many different hypothesis supported by the golden standard that can be even conflicting.**
+**The goal of DateLife is to provide the user with an insight on the current state of knowledge of time of lineage divergence for any group of taxa, as well as a tree summarizing the data.**
+
+**We addressed that in the discussion.**
+
+***
 > - Page 26, line 416 has some typos.
 
-Fixed :)
+**Fixed :)**
 
----
+***
 > - On page 26 you discuss some criteria for scoring quality of chronograms.
@@
 
----
+**Thanks for pointing us to that paper! It gave us some nice ideas for discussion. We added a paragraph on the Discussion.**
+
+***
 > - Page 27 line 448, chronogram should be plural, I think.
 
-Fixed!
+**Fixed!**
 
----
+***
 > - Page 28 line 473: I think it should be either 'public-funded' or 'publicly funded'
 
-Went for "publicly funded", thanks!
+**Went for "publicly funded", thanks!**
 
----
+***
 > - Page 29, Supplementary Material: it's probably better to sync the repos with Zenodo and cite the DOI, just so that it's guaranteed unchanging.
 
-Thanks for the suggestion! We created stable versions for all three repositories on Zenodo, and refer the doi instead of the GitHub address for the repos.
+**Thanks for the suggestion! We created stable versions for all three repositories on Zenodo, and refer the doi in addition to the GitHub addresses, L522-525.**

### 2023-05-04 tables/README-tables.Rmd
add some more reviews

---
title: "Tables"
output: html_document
date: "2023-05-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#install.packages("xtable")
library(xtable)
```

Use of package `xtable` to create latex tables:
Trick from https://www.youtube.com/watch?v=N-hLp8ADhVE

```{r}
table1 <- read.csv(file = "table-fringillidae-small-example.csv")
head(table1)
print(xtable(table1, type = "latex"))
```

### 2023-05-11 manuscript/manuscript.Rmd
add EJM edits

@@
 authornote: |
-  School of Natural Sciences, University of California, Merced, 258 Science and Engineering Building 1, Merced, CA 95340, USA.
+  School of Natural Sciences, University of California, Merced, Science and Engineering Building 1, Merced, CA 95340, USA.
 
@@
 If a taxon name belongs to an inclusive taxonomic group, DateLife has two alternative behaviors defined by the "get species from taxon" flag.
-If the flag is active, DateLife retrieves all species names within the taxonomic group from the standard taxonomy of choice, and adds them to the search query.
+If the flag is active, DateLife retrieves all species names within a taxonomic group provided, from a standard taxonomy of choice, and adds them to the search query.
 In this case, subspecific variants are excluded.
-If the flag is inactive, DateLife excludes any taxon names above the species level from the search query.
-Species and subspecific variant names are processed and searched as provided by the user.
+If the flag is inactive, DateLife excludes inclusive taxon names from the search query, and species and subspecific variant names are processed as provided by the user.
 The processed taxon names are saved as an R object of a newly defined class, \texttt{datelifeQuery}, that is used in the following steps.
@@
 
-A unique feature of the Phylesystem is that any user can add new published, state-of-the-art chronograms any time, through their curator application (https://tree.opentreeoflife.org/curator).
+A unique feature of the Phylesystem is that any user can add new published, state-of-the-art chronograms any time, through OpenTree's curator application (https://tree.opentreeoflife.org/curator).
 As chronograms are added to Phylesystem, they can be incorporated into the chronogram database of the `datelife` R package, which is currently manually updated as new chronogram data is added to Phylesystem.
@@
 These ages can correspond to the age of the most recent common ancestor (mrca) of the user's group of interest if the source chronograms have all taxa belonging to the group.
-If not, the root corresponds to the mrca of a subgroup withing the group of interest.
+If not, the root corresponds to the mrca of a subgroup within the group of interest.
 3. Study citations where original chronograms were published.
@@
 
-**_With no branch lengths.--_**
+**_Dating a tree with no branch lengths.--_**
 When producing or obtaining a tree with branch lengths for a group of interest is not possible, DateLife can date a topology without branch lengths, obtained from OpenTree or by implementing the supertree approach described above, by applying the Branch Length Adjuster [BLADJ; @Webb2008; @webb2005phylomatic] algorithm, which requires no initial branch lengths.
 The algorithm starts by fixing ages for nodes with calibration data upon the given topology.
-Then, it distributes time for nodes with no data evenly between calibrated nodes, minimizing age variance in the resulting chronogram [@Webb2008].
-This approach has has proven useful for ecological analyses that require age data [@Webb2008].
+Then, it distributes time for nodes with no data evenly between calibrated nodes, minimizing age variance in the resulting chronogram.
+This approach has proven useful for ecological analyses that require a phylogenetic time context [@Webb2008].
 When there is conflict between ages of calibrated nodes, BLADJ ignores node ages that are older than the age of a parent node.
@@
 Relative branch lengths can provide key information for phylogenetic dating, specifically for nodes without any calibration data available.
-While, using initial branch lengths data is the golden standard for phylogenetic dating analyses, producing such trees requires assembling and curating a homology hypothesis, and choosing and implementing a method for phylogenetic inference.
+While using initial branch length data is the golden standard for phylogenetic dating analyses, estimating trees with branch lengths proportional to substitution rates per site requires obtaining primary data, assembling and curating a homology hypothesis, and choosing and implementing a method for phylogenetic inference.
 DateLife implements a workflow to streamline this process by applying open data from the Barcode of Life Data System, BOLD [@ratnasingham2007bold] to obtain genetic markers for input taxa.
@@
 Alternatively, sequences can be aligned with MAFFT [@katoh2009multiple], using functions from the ape R package [@paradis2004].
-The BOLD sequence alignment is then used to obtain initial branch lengths with the accelerated transformation (ACCTRAN) parsimony algorithm, which resolves ambiguous character optimization by assigning changes along branches of the tree as close to the root as possible [@agnarsson2008acctran], resulting in older nodes [@forest2005teasing].
+The BOLD sequence alignment is then used to obtain initial branch lengths with the accelerated transformation (ACCTRAN) parsimony algorithm, which resolves ambiguous character optimization by assigning changes along branches of the tree as close to the root as possible [@agnarsson2008acctran], resulting in older internal nodes as compared to other parsimony algorithms [@forest2005teasing].
 The parsimony branch lengths are then optimized using Maximum Likelihood, given the alignment, the topology and a simple Jukes-Cantor model, producing a BOLD tree with branch lengths proportional to expected number of substitutions per site.
@@
 **_Searching the database.--_**
-DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies (Fig.
+DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms from 6 different studies (Fig.
 \ref{fig:figure2}B).
@@
 
-While we envision that DateLife will have many interesting applications in research and beyond, we recognise that as 
-<!-- sort of a "data mining tool",  -->
-any other automated tool for biological data analysis, DateLife's results can only be as good as the input data, and as give as good as how we decide to apply them for research.  
-<!-- that they might not be appropriate for research. -->
-We recognise that this might be a problem in some research areas, but it might not be the case for all.
-For example, in ecological and conservation biology studies, it has been seen that incorporating some age data provides better results than when not using any age data at all, even if the age data is not accurate [@Webb2008].
-In the following sections we discuss the particularities of divergence times resulting from a DateLife analysis for consideration of the readers and users in different research areas.
+As we envision that DateLife will have many interesting applications in research and beyond, we emphasize that DateLife's results --as well as any insights gleaned from them, largely depend on the quality of the source chronograms: low quality chronograms will produce low quality results.
+This "garbage in, garbage out" problem has long been recognised in supertree methods for summarizing phylogenetic trees [@bininda2004garbage].
+We note that this is a surfacing issue of any automated tool for biological data analysis.
+For example, DNA riddled with sequencing errors will produce generally poor alignments that will return biased evolutionary hypothesis, independently of the quality of the analysis software used.
+Again, we urge readers and DateLife users to explore all input chronograms before using a summary chronogram resulting from a DateLife workflow.
+Finally, uncertainty and variability of chronogram node age estimates might pose larger issues in some research areas than others.
+For example, in ecological and conservation biology studies, it has been shown that incorporating some chronogram data provides better results than when not using any age data at all, even if the node ages are not good quality [@Webb2008].
+In the following sections we discuss the particularities of divergence times from DateLife's summary chronograms and their impact on certain evolutionary analyses, for consideration of the readers and users in different research areas.
 <!--
@@
 
-While most chronograms in DateLife's database are constructed using primary calibrations (molecular substitution rates or ages obtained from the fossil record or geological events), DateLife summarizes chronograms using secondary calibrations (ages coming from other chronograms).
-@graur2004reading cautioned on the increased error and uncertainty in estimated ages when using secondary calibrations in dating analyses.
-@schenk2016sec showed that, in simulations, divergence times inferred using secondary calibrations are significantly younger than those inferred with primary calibrations, when obtained with Bayesian inference methods, and when priors are implemented in similar ways in both analyses.
-Accordingly, the scientific community seems to have more confidence in chronograms obtained from a single analysis, using fossil data as primary sources of calibrations [@schenk2016sec], and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data reflect a coherent evolutionary history [@sauquet2013practical], as for example done by @antonelli2017supersmart.
+While most chronograms in DateLife's database are constructed using primary calibrations (molecular substitution rates or ages obtained from the fossil record or geological events), DateLife also summarizes chronograms using secondary calibrations (ages coming from other chronograms).
+The scientific community has generally more confidence in chronograms generated from a single analysis, where fossil data constitute the primary source of calibrations [@sauquet2013practical; @schenk2016sec; @antonelli2017supersmart].
+@graur2004reading cautioned on the increased error and uncertainty in estimated node ages when using secondary calibrations in dating analyses, and more recently, @schenk2016sec found that divergence times obtained using secondary calibrations with Bayesian inference methods are significantly younger than those obtained using primary calibrations.
+While this trend still needs to be evaluated with alternative methods of divergence time estimation,
+ many efforts and discussions on best practices for implementing primary calibrations have been advanced in recent years,
+for example, using fossils that have been widely discussed and curated as calibrations to date other trees and making sure that all data reflect a coherent evolutionary history [@sauquet2013practical], as for example done by @antonelli2017supersmart.
 <!-- Will this be implemented in future `datelife` versions?
@@
 
+Obtaining vetted primary calibrations from the litterature is another bottleneck for facilitating obtention of chronograms.
 The Fossil Calibration Database provides availability of 220 primary calibration points encompassing flowering plants and metazoans, that have been curated by experts and used in peer-reviewd publications [@ksepka2015fossil].
-The database facilitates the use of expert primary fossil calibrations in new phylogenetic dating analysis.
+This database facilitates the use of expert primary fossil calibrations in new phylogenetic dating analysis.
 <!--though these still have room to grow...
@@
 
-Analysis of species diversification using simulated and empirical phylogenies suggest that using a more completely sampled phylogeny provides estimates that are closer to the true diversification history than when analysing incompletely sampled phylogenies [@cusimano2012new; @chang2020estimating; @sun2020estimating].
-Ideally, phylogenies should be completed using genetic data, but this is a difficult task to achieve for many biological groups.
-Hence, DateLife's workflow features different ways of assigning node ages in the absence of calibrations and branch length information for certain taxa.
+Analysis of species diversification of simulated and empirical phylogenies suggest that using a more completely sampled phylogeny provides estimates that are closer to the true diversification history than when analysing incompletely sampled phylogenies [@cusimano2012new; @chang2020estimating; @sun2020estimating].
+Ideally, phylogenies should be completed using genetic data, but this is a time-consuming and difficult task to achieve for many biological groups.
+Hence, DateLife's workflow features different ways of assigning divergence times to taxa with missing the absence of branch length data and calibrations and branch lengths for certain taxa.
+
 Completing a phylogeny using a stochastic birth-death polytomy resolver and a backbone taxonomy is a common practice in scientific publications: @Jetz2012, created a chronogram of all 9, 993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a chronogram of 31, 536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a chronogram of 353, 185 seed plants where only 23% had molecular data.
@@
 
-Taxonomy based polytomy resolvers also introduce topological differences.
+Taxonomy-based stochastic polytomy resolvers also introduce topological differences in phylogenetic trees.
 The study of macroevolutionary processes largely depends on an understanding of the timing of species diversification events, and different phylogenetic and chronogram hypothesis can provide very different overviews of the macroevolutionary history of a biological group.
@@
 However, it is unknown how using very different age data could affect hypothesis testing in other areas of research and more effort is needed in that direction -->
-Similarly there are worries that patterns of morphological evolution cannot be accurately inferred with these types of phylogenies, as any patterns would be erased by randomization [@rabosky2015no].
-We note that the same applies for geographical and morphological dependent diversification analysis.
-Hence, we suggest that these types of randomly resolved phylogenies can be used as null or neutral models, representing the case of diversification independent of traits and geographical scenario.
+Similarly, there are worries that patterns of morphological evolution cannot be accurately inferred with phylogenies that have been resolved stochastically over a taxonomic backbone, as any patterns would be erased by randomization [@rabosky2015no].
+We note that the same applies for geography- and morphology-dependent diversification analysis.
+Hence, we suggest that phylogenies that have been processed with taxonomy-based stochastic polytomy resolvers, including certain summary chronograms from a DateLife analysis, can be useful as null or neutral models, representing the case of a diversification process that is independent of traits and geographical scenario.
 
-While stochastic polytomy resolvers have been useful to advance research, notably, risks come with this practice.
+Taxonomy-based stochastic polytomy resolvers have been used to advance research in evolution, still, risks come with this practice.
 Taken to the extreme, one could generate a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy, a single calibration, and assigning branch lengths following a birth-death diversification model.
 Clearly, this can lead to a misrepresentation of the true evolutionary history.
-We urge DateLife users to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, potential biases, and assessing the consistency of the results with prior work.
+We urge DateLife users to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, potential biases, and assessing the consistency of DateLife's results with prior work.
 

### 2023-05-12 manuscript/manuscript.Rmd
add EJM edits on discussion

@@
 See, for example, the robust ongoing debate about crown group age of angiosperms [@sauquet2021age; @barba2018constraining; @magallon2015metacalibrated; @sanderson2001sources; @ramshaw1972time].
-Alternative source chronograms available for the same taxa have potentially been estimated implementing calibrations very differently.
-In the Fringillidae example above, the chronograms from @burns2014phylogenetics were inferred using molecular substitution rate estimates across birds [@weir2008calibrating], and have much older age estimates for the same nodes than chronograms that were inferred using fossils as calibrations (Figs.
+Alternative source chronograms available for the same taxa have potentially been estimated implementing different types of calibrations, which affects the resulting node age estimates.
+For example, in the DateLife analysis of the Fringillidae shown above, the chronograms from one study [@burns2014phylogenetics] were inferred using molecular substitution rate estimates across birds [@weir2008calibrating], and have much older age estimates for the same nodes than chronograms that were inferred using fossil calibrations (Figs.
 \ref{fig:fringillidages}, \ref{fig:cvXY}; Supplementary Figs.
@@
 <!--\textcolor{red}{EJM: MORE ON the FINCH examples and the cross validation here}-->
+Another source of conflict in estimated node ages can arise from different placements for the same calibration, which would imply fundamentally distinct evolutionary hypotheses [@antonelli2017supersmart].
+For example, two independent researchers working on the same clade should both carefully select and justify their choices of fossil calibration placement.
+Yet, if one researcher concludes that a fossil should calibrate the ingroup of a clade, while another researcher concludes that the same fossil should calibrate the outgroup of the clade, the resulting age estimates will differ, as the placement of calibrations as stem or crown group is known to significantly affect estimates of time of lineage divergence [@sauquet2013practical].
+Finally, placement of calibrations also affects uncertainty of node age estimates.
+For example, nodes that are sandwiched between a calibrated node and a calibrated root have less freedom of movement and hence narrower confidence intervals [@vos2004reconstructing], which inflates precision for nodes without calibrations but does not necessarily improve accuracy of the estimated ages.
 
-Different calibration implementations might also imply fundamentally distinct evolutionary hypotheses [@antonelli2017supersmart].
-For example, two independent researchers working on the same clade should both carefully select and justify their choices of fossil calibration placement.
-Yet, if one researcher concludes that a fossil should calibrate the ingroup of a clade, while another researcher concludes that the same fossil should calibrate the outgroup of the clade, the resulting age estimates will differ, as the placement of calibrations as stem or crown group has been proven to significantly affect time of lineage divergence estimates [@sauquet2013practical].
-In addition, nodes that are sandwiched between a calibrated node and a calibrated root have less freedom of movement and hence narrower confidence intervals [@vos2004reconstructing], this inflates precision for nodes without calibrations but does not necessarily improve accuracy of the estimated ages.
+DateLife's summary chronograms are intended to represent all variation in estimated node ages from source chronograms.
+Node age distribution ranges allow to visually explore ages from source chronograms individually and contextualize and compare them against other chronograms.
+Researchers that wish to use summary chronograms in downstream evolutionary analysis may select multiple trees sampled from the summary distribution of node ages, to account for variation in source chronograms.
 
@@
 
-While most chronograms in DateLife's database are constructed using primary calibrations (molecular substitution rates or ages obtained from the fossil record or geological events), DateLife also summarizes chronograms using secondary calibrations (ages coming from other chronograms).
-The scientific community has generally more confidence in chronograms generated from a single analysis, where fossil data constitute the primary source of calibrations [@sauquet2013practical; @schenk2016sec; @antonelli2017supersmart].
-@graur2004reading cautioned on the increased error and uncertainty in estimated node ages when using secondary calibrations in dating analyses, and more recently, @schenk2016sec found that divergence times obtained using secondary calibrations with Bayesian inference methods are significantly younger than those obtained using primary calibrations.
-While this trend still needs to be evaluated with alternative methods of divergence time estimation,
- many efforts and discussions on best practices for implementing primary calibrations have been advanced in recent years,
-for example, using fossils that have been widely discussed and curated as calibrations to date other trees and making sure that all data reflect a coherent evolutionary history [@sauquet2013practical], as for example done by @antonelli2017supersmart.
+While most chronograms in DateLife's database are constructed using primary calibrations, i.e., molecular substitution rates or absolute ages in Myr obtained from the fossil record or geological events, DateLife also constructs summary chronograms using secondary calibrations, i.e., node ages in Myr extracted from existing chronograms.
+The scientific community has generally more confidence in chronograms generated from a single analysis, where carefully chosen primary calibrations constitute the main source of absolute time information used in the dating analysis [@shaul2002playing; @graur2004reading; @sauquet2012testing; @sauquet2013practical; @garzon2015incompatible; @schenk2016sec; @antonelli2017supersmart].
+However, implementation of primary calibrations does not come without difficulties: it requires high expertise and training to discover, place and apply  calibrations appropriately [@ksepka2011synthesizing; @hipsley2014beyond].
+One way around it is to use fossils that have been widely discussed and previously curated as calibrations to date other trees [@ksepka2011synthesizing; @sauquet2013practical], and making sure that all data reflect a coherent evolutionary history [@sauquet2013practical], as for example done by @antonelli2017supersmart.
 <!-- Will this be implemented in future `datelife` versions?
@@
 -->
+The Fossil Calibration Database provides availability of 220 primary calibration points encompassing flowering plants and metazoans, that have been curated by experts and used for dating analysis in peer-reviewd publications [@ksepka2015fossil].
+This database facilitates the use of expert primary fossil calibrations in new phylogenetic dating analyses.
+Yet, users still require the expertise to locate appropriate nodes in their phylogenies to corresponding fossils available in the database, potentially delaying increased usage of the database in research.
+Open databases for effective reuse of fossil calibration data still have room to grow and should constitute a key addition to the phylogenetics tool set.
+ <!--though these still have room to grow... BCO: Polite way to say 'no new additions in four years?' LLSR: lol, yes-->
 
-Obtaining vetted primary calibrations from the litterature is another bottleneck for facilitating obtention of chronograms.
-The Fossil Calibration Database provides availability of 220 primary calibration points encompassing flowering plants and metazoans, that have been curated by experts and used in peer-reviewd publications [@ksepka2015fossil].
-This database facilitates the use of expert primary fossil calibrations in new phylogenetic dating analysis.
-<!--though these still have room to grow...
-BCO: Polite way to say 'no new additions in four years?' LLSR: lol, yes-->
-Despite the wealth of information displayed in the database, users still need the expertise to locate the equivalent node in their phylogenies to be dated.
-Open databases for reuse of fossil calibration data still have room to grow and should be a great addition to the phylogenetics tool set.
+In some cases, secondary calibrations might represent the only information available (e.g., groups with a poor fossil record).
+Recently, @powell2020quantifying showed in a simulation analysis that secondary calibrations are as good as primary calibrations.
+It seems that using several secondary calibrations (as opposed to just one) can provide sufficient information to alleviate or even neutralize potential biases described before [@shaul2002playing; @graur2004reading; @sauquet2013practical].
+<!--\textcolor{red}{LLSR: What does our cross validation analysis show in relation to  the effect of using secondary calibrations for a dating analysis that does not make any prior assumptions on the nature of the calibrations themselves?
+the age difference between primary and secondary calibrations?}--> Our cross validation analysis also provides some insight into the application of secondary calibrations.
+In general, node ages summarized with DateLife and those from the original studies appear positively correlated (Supplementary Figures S2-S20).
+We also note that nodes closer to the root tend to be slightly younger than ages from the original studies.
+In contrast, nodes closer to the tips tend to be slightly older when estimated using secondary calibrations than ages from the original studies.
+The only exception to this trend was observed in @burns2014phylogenetics chronogram, which generally displays much younger node ages when estimated using secondary calibrations than the original study (Supplementary Figs.
+S1, S5), supporting previous observations [@sauquet2012testing; @schenk2016sec].
 
-<!--\textcolor{red}{LLSR: What does our cross validation analysis show in relation to  the effect of using secondary calibrations for a dating analysis that does not make any prior assumptions on the nature of the calibrations themselves?
-the age difference between primary and secondary calibrations?}-->
-
-It seems that using several (as opposed to just a few) secondary calibrations can provide sufficient information to alleviate or even neutralize potential biases [@sauquet2013practical].
-Certainly, further studies are required to fully understand the effect of secondary calibrations on outputs from different tree dating methods, and on downstream analyses.
-It is possible that secondary calibrations can be safely used with dating methods that do not require setting priors, such as penalized likelihood [@sanderson2003r8s], with methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ [@Webb2008; @webb2005phylomatic], or methods that summarize age data unto a tree topology.
-
-Our cross validation analysis might provide some insight in this regard.
-When ages are estimated with secondary calibrations, nodes closer to the root do tend to be slightly younger than ages estimated with primary calibrations.
-However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures S2-S20).
-The only exception to this was observed on cross validation results of the @burns2014phylogenetics chronogram, which displays much younger node ages when estimated using secondary calibrations (Supplementary Figs.
-S1, S5).
+We unite to the call of our colleagues: further research is needed to fully understand the effects of using secondary calibrations with alternative tree dating methods, as well as usage of resulting chronograms in downstream analyses [@shaul2002playing; @schenk2016sec; @hipsley2014beyond; @powell2020quantifying].
+It is possible that secondary calibrations can be used successfully with dating methods that do not require setting priors, such as penalized likelihood [@sanderson2003r8s]; with methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ [@Webb2008; @webb2005phylomatic]; and with methods that summarize age data unto a tree topology, such as DateLife.
 
@@
 Completing a phylogeny using a stochastic birth-death polytomy resolver and a backbone taxonomy is a common practice in scientific publications: @Jetz2012, created a chronogram of all 9, 993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a chronogram of 31, 536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a chronogram of 353, 185 seed plants where only 23% had molecular data.
-These stochastically resolved chronograms provide less biased diversification rates estimated with methods that account for sampling fractions [@chang2020estimating; @cusimano2012new], but can also introduce spurious patterns of early bursts of diversification [@cusimano2010slowdowns; @sun2020estimating].
+These stochastically resolved chronograms return diversification rates estimates that appear less biased than those estimated from their incompletely sampled counterparts, even with methods that account for missing lineages by using sampling fractions [@cusimano2012new; @chang2020estimating], but can also introduce spurious patterns of early bursts of diversification [@cusimano2010slowdowns; @sun2020estimating].
 

### 2023-05-12 notes-effects-downstream-analysis-review.Rmd
add EJM edits on discussion

@@
 
-@cusimano2012new: Simulating branch lengths following a birth-death species diversification model for
-missing taxa to chronograms with non random sampling, lowers type I error (false positive,
+@cusimano2012new: Simulating branch lengths following a birth-death species diversification model for missing taxa to chronograms with non random sampling, lowers type I error (false positive,
 incorrectly accepting the null hypothesis of a constant-rate or temporally varying rate
@@
 improves insight from phylogeny-based analyses of diversification rates [@chang2020estimating].
-@chang2020estimating state that stochastic politomy resolution alleviate low statistical power to
+@chang2020estimating state that stochastic polytomy resolution alleviate low statistical power to
 identify shifts in diversification rate with methods that account for sampling fractions.
@@
 corrections based on “sampling fractions” can lead to low statistical power to infer rate variation
-when it is present, depending on the likelihood function used for inference. In the extreme, the sampling
-fraction correction can lead to spurious patterns of diversification that are driven solely by unbalanced
-sampling across the tree in concert with low overall power to infer shifts."
+when it is present, depending on the likelihood function used for inference.
+In the extreme, the sampling fraction correction can lead to spurious patterns of diversification that are driven solely by unbalanced sampling across the tree in concert with low overall power to infer shifts."
 
@@
 tested the effect of using incompletely sampled phylogenies on one hand, and incomplete phylogenies that
-have been made near complete, using a stochastic politomy resolver based on taxonomy, on the other hand
+have been made near complete, using a stochastic polytomy resolver based on taxonomy, on the other hand
 (with two different methods: 1) adding missing species at random following a backbone taxonomy, and 2)

### 2023-05-12 manuscript/manuscript.Rmd
add reviwers to acknowledgements
change figure and table refs for actual numbers

@@
 taxize [@Chamberlain2013; @Chamberlain2018].
-Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife workflow: creating a search query, searching a database, and summarizing results from the search.
+Figure 1 provides a graphical summary of the three main steps of the DateLife workflow: creating a search query, searching a database, and summarizing results from the search.
 
@@
 These matching pruned chronograms are referred to as source chronograms.
-Total distance (in units of millions of years) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Figure \ref{fig:workflow}).
+Total distance in units of million years (Myr) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Fig. 1).
 The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a "phylo" object or newick string.
@@
 Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
-Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig.
-\ref{fig:runtime_main}).
+Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig. 2).
 <!--Speed of other datelife workflows? NO, bc it reflects speed from other software mainly. But I will put a sentence about it...
@@
 We illustrate the DateLife workflow using a family within the passeriform birds encompassing the true finches, Fringillidae, as case study.
-On a small example, we analysed 6 bird species, and results from each step of the workflow are shown in Fig.
-\ref{fig:figure2}.
+On a small example, we analysed 6 bird species, and results from each step of the workflow are shown in Figure 3.
 As a second example, we analysed 289 bird species in the family Fringillidae that are included in the NCBI taxonomy.
-The resulting summary chronogram is shown in Fig.
-\ref{fig:fringillidages}, and results from previous steps of the workflow are available as Supplementary Figures.
+The resulting summary chronogram is shown in Figure 4, and results from previous steps of the workflow are available as Supplementary Figures.
 <!--We used `r cite_r("r-references.bib")` for all our analyses.-->
@@
 For a detailed discussion on the state of the synonym, refer to Avibase [@lepage2004avibase; @lepage2014avibase; @avibase-emberiza].
-Discovering this synonym allowed assigning five age data points for the parent node of *Emberiza elegans*, shown as *Schoeniclus elegans* in figure \ref{fig:figure2}A, which would not have had any data otherwise.
+Discovering this synonym allowed assigning five age data points for the parent node of *Emberiza elegans*, shown as *Schoeniclus elegans* in Figure 3A, which would not have had any data otherwise.
 <!--\textcolor{red}{EJM: I don't understand what this sentence means}
@@
 DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms from 6 different studies (Fig.
-\ref{fig:figure2}B).
+3B).
 Three studies matched five input names [@barker2015new; @Hedges2015; @Jetz2012], one study matched four input names [@hooper2017chromosomal] and two studies matched two input names [@barker2013going; @burns2014phylogenetics].
 No studies matched all input names.
-Together, source chronograms provide 28 unique age data points, covering all nodes on our chosen tree topology to date (Table \ref{tbl:table1}).
+Together, source chronograms provide 28 unique age data points, covering all nodes on our chosen tree topology to date (Table 1).
 
@@
 **_Summarizing search results.--_**
-DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and congruified and mapped age data to nodes in this chosen topology (Table \ref{tbl:table1}).
-The name processing step allowed including five data points for node "n4" (parent of _Schoeniclus elegans_; Fig. \ref{fig:figure2}A) that would not have had any data otherwise due to name mismatch.
-Age summary statistics per node were calculated (Table \ref{tbl:table2}) and used as calibrations to date the tree topology using the BLADJ algorithm.
+DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. 3C), and congruified and mapped age data to nodes in this chosen topology (Table 1).
+The name processing step allowed including five data points for node "n4" (parent of _Schoeniclus elegans_; Fig. 3A) that would not have had any data otherwise due to name mismatch.
+Age summary statistics per node were calculated (Table 2) and used as calibrations to date the tree topology using the BLADJ algorithm.
 As expected, more inclusive nodes (e.g., node "n1") have more variance in age data than less inclusive nodes (e.g., node "n5").
-Summary age data for node "n2" were excluded as final calibration because they are older than age data of the more inclusive node, "n1" (Fig. \ref{fig:figure2}C4).
+Summary age data for node "n2" were excluded as final calibration because they are older than age data of the more inclusive node, "n1" (Fig. 3C4).
 
@@
 According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig.
-\ref{fig:fringillidae-topologies}).
+5).
 Hence, a topology containing only the 289 species from the original query was extracted from Open Tree of Life’s synthetic tree v12.3 [@opentreeoflife2019synth].
@@
 Age data from source chronograms was congruified to OpenTree's topology (Fig.
-\ref{fig:fringillidae-topologies}B), reducing the age data set to 818 different data points (Supplementary Table S2).
+5B), reducing the age data set to 818 different data points (Supplementary Table S2).
 For each congruent node, age summary statistics were calculated and used as fixed secondary calibrations over the chosen tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
-\ref{fig:fringillidages}).
+4).
 
@@
 Finally, average node ages per node were applied as secondary calibrations and smoothed with the BLADJ algorithm.
-We found that node ages from the original studies, and ages estimated using all other age data available are largely correlated (Fig. \ref{fig:cvXY}).
+We found that node ages from the original studies, and ages estimated using all other age data available are largely correlated (Fig. 5).
 For five studies, DateLife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka 'closer to the root') relative to the original estimate, and overestimate ages for nodes closer to the tips.
@@
 For example, in the DateLife analysis of the Fringillidae shown above, the chronograms from one study [@burns2014phylogenetics] were inferred using molecular substitution rate estimates across birds [@weir2008calibrating], and have much older age estimates for the same nodes than chronograms that were inferred using fossil calibrations (Figs.
-\ref{fig:fringillidages}, \ref{fig:cvXY}; Supplementary Figs.
+4, 5; Supplementary Figs.
 S1, S5).
@@
 
+We thank Isabel Sanmartín, Daniele Silvestro, Rutger Vos and an anonymous reviewer, for comments that greatly improved this manuscript.
 The DateLife project was born as a prototype tool aiming to provide the services describe in this paper, and was initially developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA [@Stoltzfus2013].

### 2023-05-12 notes-secondary-calibrations.Rmd
add discussion notes about secondary calibrations

---
title             : "Review: Effects of initial phylogenetic and chronogram hypothesis on evolutionary downstream analyses"
shorttitle        : "Litt review"

author:
  - name          : "Luna L. Sánchez Reyes"
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : ""
    email         : "sanchez.reyes.luna@gmail.com"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Data curation
      - Investigation
      - Software
      - Visualization
      - Validation
      - Writing
      - Original Draft Preparation
      - Writing
      - Review & Editing

affiliation:
  - id            : "1"
    institution   : "University of California, Merced, USA"
  - id            : "2"
    institution   : "University of Tennessee, Knoxville, USA"

authornote: |
  School of Natural Sciences, University of California, Merced, 258 Science and Engineering Building 1, Merced, CA 95340, USA.
  Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 446 Hesler Biology Building, Knoxville, TN 37996, USA.
bibliography      : ["paper_references.bib"]
wordcount         : "`r wordcountaddin::word_count()`"
floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : no
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            :
  papaja::apa6_pdf:
    includes      :
      in_header: "preamble.tex"
---

```{r setup, include = FALSE}
library("papaja")
r_refs("paper_references.bib")
embed_figure <- TRUE
```
An early set of litterature cautions against the use of secondary calibrations in divergence time estimation.

The scientific community has generally more confidence in chronograms generated from a single analysis, where primary calibrations constitute the main source of information in a dating analysis. Early, @shaul2002playing and @graur2004reading warned about the misleading lack of uncertainty in divergence times estimated using secondary calibrations, and more recently, @sauquet2012testing and @schenk2016sec found that node ages obtained using secondary calibrations with Maximum Likelihood and Bayesian inference dating in two different empirical data sets, are significantly younger than those obtained using primary calibrations. However, more recently, @powell2020quantifying reported that secondary calibrations appear to be as good as primary calibrations when dating simulated phylogenies. It seems that using several secondary calibrations (as opposed to just one) can provide sufficient information to alleviate or even neutralize potential biases [@shaul2002playing; @graur2004reading; @sauquet2013practical].

@shaul2002playing: They assess the reliability of divergence time estimates obtained
using secondary calibration points by testing the consistency of the age of the secondary calibration with divergence time estimates calculated assuming no rate heterogeneity, estimating pairwise aminoacid rate substitutions from 21 different protein alignments, following a Poisson model.
They use the same secondary calibrations with all alignments and find that all divergence times are indeed older than the secondary calibration. I believe this is expected, as actual divergence times should be at least as old or older than a calibration point.

However, because this is a secondary calibration, it can't be treated with the same rationale as a fossil calibration, the key difference beig that secondary calibrations already represent the divergence time, so they can't follow the same priors as fossil, such as being necessarily equal or younger than the actual divergence time. An estimated divergence time could in fact be older than the actual divergence time.5tdsx89c v

They list set of methodological steps that can affect divergence time estimates
for the same group or "evolutionary event. High on the list are":
- (1) different molecular datasets;
- (2) different criteria for inclusion or exclusion of [molecular] data;
- (3) different methodologies for the derivation of genetic distances; and
- (4) different calibration points for converting genetic distances into evolutionary rates and subsequently into dates of divergence (for discussion, see Easteal, 1999, Wang et al., 1999, Bromham et al., 2000).
- (5) whether calibrations are primary or secondary
They recommend using multiple primary calibrations, using a model with rate heterogeneity and discussing confidence intervals that could show different hypothesis of divergence time.

@graur2004reading: made a review of divergence times estimated using secondary calibrations (from the Hedges and Kumar group) and found that they were completely ignoring uncertainty of the estimate, giving an illusion of precision to all other node age estimates in the tree.
Their main recommendation is to use multiple calibration points.

@garzon2015incompatible empirical data, bayesian inference, secondary calibrations

@shcenck2016 simulated trees, Bayesian inference with program BEAST, relaxed-clock methods
"Applying secondary calibrations has been said to increase the accuracy of the age estimates across a secondary study as long as the estimate was derived from a robust primary calibration @hedges2004precision. Shaul and Graur's [@shaul2002playing] evidence of inconsistency makes intuitive sense, but their methodology has been criticized [@morrison2010counting; @hedges2004precision]".

@powell2020quantifying find that secondary calibrations are as useful as primary calibrations when used appropriately.
"We quantify the amount of errors in estimates produced by the use of secondary calibrations relative to true times and primary calibrations placed on distant nodes. We find that, overall, the inaccuracies in estimates based on secondary calibrations are predictable and mirror errors associated with primary calibrations and their confidence intervals.

@hipsley2014beyond: Beyond fossil calibrations: realities of molecular clock practices.

## Difficulties of using fossil calibrations

Ho & Phillips 2009: Accounting for calibration uncertainty in phylogenetic estimation of evolutionary
divergence times

Ho, Saarma, Barnett, Haile, & Shapiro, 2008: The effect of inappropriate calibration: three case studies in molecular ecology

Inoue, P.C.J. Donoghue, Z. Yang, 2010: The impact of the representation of fossil calibrations on Bayesian estimation of species divergence times

Ksepka, M.J. Benton, M.T. Carrano, M. Gandolfo, J.J. a Head, E.J. Hermsen, W.G. Joyce, K.S. Lamm, J.S.L. Patané, M.J. Phillips, P.D. Polly, M. Van Tuinen, J.L. Ware, R.C.M. Warnock, J.F. Parham, 2011: Synthesizing and databasing fossil calibrations: divergence dating and beyond

Sebastian S. Groh, Paul Upchurch, Paul M. Barrett, Julia J. Day, 2022: How to date a crocodile: estimation of neosuchian clade ages and a comparison of four time-scaling methods

### Difficulties of dating divergence events

Susana A. Magallón 2004, Dating Lineages: Molecular and Paleontological Approaches to the Temporal Framework of Clades

Smith 2002, Dating the Time of Origin of Major Clades: Molecular Clocks and the Fossil Record

## Open database of fossil calibrations

After Kspeka et al. 2011, a fossil calibration database should have the following components:
- Should provide a tree-based representation of phylogenetic placement of a fossil
- Only include fossils that have been vetted as calibrations
- Clearly provide attribution of original fossil descriptions (to incentivize paleontologists to contribute more) and papers were fossil was vetted as calibrations, for appropriate citation of all original data in research papers that use fossils from the database.

## Review of studeis that cite the Fossil Calibrations Database

## Review of studies from Ksepka in google scholar

### 2023-05-14 data-raw/fringillidae-topologies.R
change subfigure letters to lowercase

@@ -1,0 +1,25 @@
+getwd()
+
+fringillidae_paraphyletic <- ape::read.tree("figures/fringillidae-topologies/fringillidae_paraphyletic_topology.tre")
+
+usethis::use_data(fringillidae_paraphyletic)
+
+fringillidae_monophyletic <- ape::read.tree("figures/fringillidae-topologies/fringillidae_monophyletic_topology.tre")
+
+usethis::use_data(fringillidae_monophyletic)
+
+rotl::tol_about()
+
+# For documentation:
+
+#' OpenTree Synthetic Tree of Life.
+#'
+#' Tree version: opentree13.4
+#' Taxonomy version: 3.3draft1
+#' Constructed on: 2021-06-18 11:13:49
+#' Number of terminal taxa: 2392042
+#' Number of source trees: 1239
+#' Number of source studies: 1184
+#' Source list present: false
+#' Root taxon: cellular organisms
+#' Root ott_id: 93302

### 2023-05-14 manuscript/manuscript.Rmd
add link to suppl figures on zenodo

@@
 At the time of writing of this manuscript
-<!-- (Jun 22, 2022), -->
-(`r format(Sys.time(), "%b %d, %Y")`),
+(May 15, 2023),
+<!-- (`r format(Sys.time(), "%b %d, %Y")`), -->
 DateLife's chronogram database latest version consist of `r format(length(opentree_chronograms$trees), big.mark=",")` chronograms published in `r length(unique(opentree_chronograms$studies))` different studies.
@@
 
-In some cases, secondary calibrations might represent the only information available (e.g., groups with a poor fossil record).
+In some cases, secondary calibrations might represent the only information available for a dating analysis (e.g., groups with a poor fossil record).
 Recently, @powell2020quantifying showed in a simulation analysis that secondary calibrations are as good as primary calibrations.
@@
 the age difference between primary and secondary calibrations?}--> Our cross validation analysis also provides some insight into the application of secondary calibrations.
-In general, node ages summarized with DateLife and those from the original studies appear positively correlated (Supplementary Figures S2-S20).
+In general, node ages summarized with DateLife and those from the original studies appear positively correlated (Supplementary Figs.
+S2-S20).
 We also note that nodes closer to the root tend to be slightly younger than ages from the original studies.
@@
 
-Supplementary material, including code, biological examples, benchmark results, data files and online-only appendices, can be found in the Dryad data repository (<https://doi.org/10.5061/dryad.cnp5hqc6w>), as well as in the Zenodo stable repositories that host the reproducible manuscript (<https://doi.org/10.5281/zenodo.7435094>), the biological examples (<https://doi.org/10.5281/zenodo.7435101>), and the software benchmark (<https://doi.org/10.5281/zenodo.7435106>).
-The corresponding development versions hosted on GitHub can be found at <https://github.com/LunaSare/datelifeMS1>, <https://github.com/LunaSare/datelife_examples>, and <https://github.com/LunaSare/datelife_benchmark>.
+Supplementary Figures can be viewed and dowloaded from their Zenodo repository (<https://doi.org/10.5281/zenodo.6683667>).
+Supplementary material, including code, biological examples, benchmark results, data files and online-only appendices, can be downloaded from the Dryad data repository (<https://doi.org/10.5061/dryad.cnp5hqc6w>), as well as in the Zenodo stable repositories that host the reproducible manuscript (<https://doi.org/10.5281/zenodo.7435094>), the biological examples (<https://doi.org/10.5281/zenodo.7435101>), and the software benchmark (<https://doi.org/10.5281/zenodo.7435106>).
+Development versions corresponding to all of the above are hosted on GitHub, accesible at <https://github.com/LunaSare/datelifeMS1>, <https://github.com/LunaSare/datelife_examples>, and <https://github.com/LunaSare/datelife_benchmark>.
 

### 2023-05-14 manuscript/manuscript.tex
add link to suppl figures on zenodo

--- 2023-02-21 manuscript/manuscript.tex
@@
 
-School of Natural Sciences, University of California, Merced, 258 Science and Engineering Building 1, Merced, CA 95340, USA.
+School of Natural Sciences, University of California, Merced, Science and Engineering Building 1, Merced, CA 95340, USA.
 
@@
 
-Word count: 6111
+Word count: 6715
 
@@
 
-DateLife's core software applications are implemented in the R package \texttt{datelife}.
-Its current stable version -- v0.6.6, is available from The Comprehensive R Archive Network (CRAN) repository (Sanchez-Reyes et al., 2022), and relies on functionalities from other biological R packages:
+DateLife's core software applications are implemented in the R package \texttt{datelife},
+and relies on functionalities from other biological R packages:
 ape (Paradis, Claude, \& Strimmer, 2004),
@@
 taxize (Chamberlain, 2018; Chamberlain \& Szöcs, 2013).
-Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife workflow: creating a search query, searching a database, and summarizing results from the search.
+Figure 1 provides a graphical summary of the three main steps of the DateLife workflow: creating a search query, searching a database, and summarizing results from the search.
 
@@
 
-DateLife starts by processing an input consisting of at least one taxon scientific name. Multiple input names can be provided as a comma separated character string or as tip labels on a tree.
+DateLife starts by processing an input consisting of the scientific name of at least one taxon. Multiple input names can be provided as a comma separated character string or as tip labels on a tree.
 If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a \enquote{phylo} R object (Paradis et al., 2004).
@@
 If a taxon name belongs to an inclusive taxonomic group, DateLife has two alternative behaviors defined by the \enquote{get species from taxon} flag.
-If the flag is active, DateLife retrieves all species names within the taxonomic group from the standard taxonomy of choice, and adds them to the search query.
+If the flag is active, DateLife retrieves all species names within a taxonomic group provided, from a standard taxonomy of choice, and adds them to the search query.
 In this case, subspecific variants are excluded.
-If the flag is inactive, DateLife excludes any taxon names above the species level from the search query.
-Species and subspecific variant names are processed and searched as provided by the user.
+If the flag is inactive, DateLife excludes inclusive taxon names from the search query, and species and subspecific variant names are processed as provided by the user.
 The processed taxon names are saved as an R object of a newly defined class, \texttt{datelifeQuery}, that is used in the following steps.
@@
 At the time of writing of this manuscript
-(Feb 21, 2023),
+(May 15, 2023),
 DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies.
@@
 
-A unique feature of the Phylesystem is that any user can add new published, state-of-the-art chronograms any time, through their curator application (\url{https://tree.opentreeoflife.org/curator}).
+A unique feature of the Phylesystem is that any user can add new published, state-of-the-art chronograms any time, through OpenTree's curator application (\url{https://tree.opentreeoflife.org/curator}).
 As chronograms are added to Phylesystem, they can be incorporated into the chronogram database of the \texttt{datelife} R package, which is currently manually updated as new chronogram data is added to Phylesystem.
 The updated database is assigned a new version number, followed by a package release on CRAN.
-Users can directly implement \texttt{datelife} functions to trigger an update of the local chronogram database, to incorporate any new chronograms to their DateLife analysis before an official database update is released on CRAN.
+Users can directly run \texttt{datelife} functions to trigger an update of their local chronogram database, to incorporate any new chronograms to their DateLife analysis before an official database update is released on CRAN.
 
@@
 These matching pruned chronograms are referred to as source chronograms.
-Total distance (in units of millions of years) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Figure \ref{fig:workflow}).
+Total distance in units of million years (Myr) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Fig. 1).
 The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a \enquote{phylo} object or newick string.
@@
 These ages can correspond to the age of the most recent common ancestor (mrca) of the user's group of interest if the source chronograms have all taxa belonging to the group.
-If not, the root corresponds to the mrca of a subgroup withing the group of interest.
+If not, the root corresponds to the mrca of a subgroup within the group of interest.
 \item
@@
 
-\textbf{\emph{With no branch lengths.--}}
-When producing or obtaining a tree with branch lengths for a group of interest is not possible, DateLife can date a topology without branch lengths, obtained from OpenTree or by implementing the supertree approach described above, by implementing the Branch Length Adjuster (BLADJ; Webb et al., 2008; Webb \& Donoghue, 2005) algorithm, which requires no initial branch lengths.
+\textbf{\emph{Dating a tree with no branch lengths.--}}
+When producing or obtaining a tree with branch lengths for a group of interest is not possible, DateLife can date a topology without branch lengths, obtained from OpenTree or by implementing the supertree approach described above, by applying the Branch Length Adjuster (BLADJ; Webb et al., 2008; Webb \& Donoghue, 2005) algorithm, which requires no initial branch lengths.
 The algorithm starts by fixing ages for nodes with calibration data upon the given topology.
-Then, it distributes time for nodes with no data evenly between calibrated nodes, minimizing age variance in the resulting chronogram (Webb et al., 2008), which has proven useful for ecological analyses that require age data (Webb et al., 2008).
+Then, it distributes time for nodes with no data evenly between calibrated nodes, minimizing age variance in the resulting chronogram.
+This approach has proven useful for ecological analyses that require a phylogenetic time context (Webb et al., 2008).
 When there is conflict between ages of calibrated nodes, BLADJ ignores node ages that are older than the age of a parent node.
@@
 \textbf{\emph{Dating a tree with branch lengths.--}}
-Relative branch lengths provide key information for phylogenetic dating, specifically for nodes without any calibration data available.
-While, using initial branch lengths data is the golden standard for phylogenetic dating analyses, producing such requires assembling and curating a homology hypothesis, and choosing and implementing a method for phylogenetic inference.
-DateLife implements a workflow to streamline this process by leveraging on genetic data from the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain markers for input taxa.
-By default, mined genetic sequences are aligned with MUSCLE (Edgar, 2004) using functions from the msa R package (Bodenhofer et al., 2015).
+Relative branch lengths can provide key information for phylogenetic dating, specifically for nodes without any calibration data available.
+While using initial branch length data is the golden standard for phylogenetic dating analyses, estimating trees with branch lengths proportional to substitution rates per site requires obtaining primary data, assembling and curating a homology hypothesis, and choosing and implementing a method for phylogenetic inference.
+DateLife implements a workflow to streamline this process by applying open data from the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain genetic markers for input taxa.
+By default, BOLD genetic sequences are aligned with MUSCLE (Edgar, 2004) using functions from the msa R package (Bodenhofer et al., 2015).
 Alternatively, sequences can be aligned with MAFFT (Katoh, Asimenos, \& Toh, 2009), using functions from the ape R package (Paradis et al., 2004).
-The BOLD sequence alignment is then used to obtain initial branch lengths with the accelerated transformation (ACCTRAN) parsimony algorithm, which resolves ambiguous character optimization by assigning changes along branches of the tree as close to the root as possible (Agnarsson \& Miller, 2008), resulting in older nodes (Forest et al., 2005).
+The BOLD sequence alignment is then used to obtain initial branch lengths with the accelerated transformation (ACCTRAN) parsimony algorithm, which resolves ambiguous character optimization by assigning changes along branches of the tree as close to the root as possible (Agnarsson \& Miller, 2008), resulting in older internal nodes as compared to other parsimony algorithms (Forest et al., 2005).
 The parsimony branch lengths are then optimized using Maximum Likelihood, given the alignment, the topology and a simple Jukes-Cantor model, producing a BOLD tree with branch lengths proportional to expected number of substitutions per site.
 Both parsimony and ML optimizations are done with functions from the \texttt{phangorn} package (Schliep, 2011).
-The BOLD workflow is currently only available on DateLife's R package and not on the web application.
+Due to the computing load it requires, the BOLD workflow is currently only supported through DateLife's R package. It is not yet available through the web application.
 
@@
 Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
-Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig.
-\ref{fig:runtime_main}).
+Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig. 2).
 
@@
 We illustrate the DateLife workflow using a family within the passeriform birds encompassing the true finches, Fringillidae, as case study.
-On a small example, we analysed 6 bird species, and results from each step of the workflow are shown in Fig.
-\ref{fig:figure2}.
+On a small example, we analysed 6 bird species, and results from each step of the workflow are shown in Figure 3.
 As a second example, we analysed 289 bird species in the family Fringillidae that are included in the NCBI taxonomy.
-The resulting summary chronogram is shown in Fig.
-\ref{fig:fringillidages}, and results from previous steps of the workflow are available as Supplementary Figures.
+The resulting summary chronogram is shown in Figure 4, and results from previous steps of the workflow are available as Supplementary Figures.
 
@@
 For a detailed discussion on the state of the synonym, refer to Avibase (Avibase, 2022; Lepage, 2004; Lepage, Vaidya, \& Guralnick, 2014).
-Discovering this synonym allowed assigning five age data points for the parent node of \emph{Emberiza elegans}, shown as \emph{Schoeniclus elegans} in figure \ref{fig:figure2}A, which would not have had any data otherwise.
+Discovering this synonym allowed assigning five age data points for the parent node of \emph{Emberiza elegans}, shown as \emph{Schoeniclus elegans} in Figure 3A, which would not have had any data otherwise.
 
 \textbf{\emph{Searching the database.--}}
-DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies (Fig.
-\ref{fig:figure2}B).
+DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms from 6 different studies (Fig.
+3B).
 Three studies matched five input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2015; Hedges, Marin, Suleski, Paymer, \& Kumar, 2015; Jetz et al., 2012), one study matched four input names (Hooper \& Price, 2017) and two studies matched two input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2013; Burns et al., 2014).
 No studies matched all input names.
-Together, source chronograms provide 28 unique age data points, covering all nodes on our chosen tree topology to date (Table \ref{tbl:table1}).
+Together, source chronograms provide 28 unique age data points, covering all nodes on our chosen tree topology to date (Table 1).
 
 \textbf{\emph{Summarizing search results.--}}
-DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and congruified and mapped age data to nodes in this chosen topology (Table \ref{tbl:table1}).
-The name processing step allowed including five data points for node \enquote{n4} (parent of \emph{Schoeniclus elegans}; Fig.
-\ref{fig:figure2}A) that would not have had any data otherwise due to name mismatch.
-Age summary statistics per node were calculated (Table \ref{tbl:table2}) and used as calibrations to date the tree topology using the BLADJ algorithm.
+DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. 3C), and congruified and mapped age data to nodes in this chosen topology (Table 1).
+The name processing step allowed including five data points for node \enquote{n4} (parent of \emph{Schoeniclus elegans}; Fig. 3A) that would not have had any data otherwise due to name mismatch.
+Age summary statistics per node were calculated (Table 2) and used as calibrations to date the tree topology using the BLADJ algorithm.
 As expected, more inclusive nodes (e.g., node \enquote{n1}) have more variance in age data than less inclusive nodes (e.g., node \enquote{n5}).
-Summary age data for node \enquote{n2} were excluded as final calibration because they are older than age data of the more inclusive node, \enquote{n1} (Fig. \ref{fig:figure2}C4).
+Summary age data for node \enquote{n2} were excluded as final calibration because they are older than age data of the more inclusive node, \enquote{n1} (Fig. 3C4).
 
@@
 According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig.
-\ref{fig:fringillidae-topologies}).
+5).
 Hence, a topology containing only the 289 species from the original query was extracted from Open Tree of Life's synthetic tree v12.3 (Open Tree Of Life et al., 2019).
@@
 We congruified node ages extracted from chronograms from all other studies upon the individual topologies, effectively excluding original ages from each topology.
-Average node ages per node were applied as secondary calibrations, smoothed with the BLADJ algorithm.
-We found that node ages from the original studies, and ages estimated using all other age data available are largely correlated (Fig. \ref{fig:cvXY}).
+Finally, average node ages per node were applied as secondary calibrations and smoothed with the BLADJ algorithm.
+We found that node ages from the original studies, and ages estimated using all other age data available are largely correlated (Fig. 5).
 For five studies, DateLife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka \enquote{closer to the root}) relative to the original estimate, and overestimate ages for nodes closer to the tips.
@@
 
-While it is known that incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance in ecological and conservation biology studies (Webb et al., 2008).
-Next, we discuss some particularities of using summarized node age data from a DateLife analysis that users should be aware of.
+As we envision that DateLife will have many interesting applications in research and beyond, we emphasize that DateLife's results --as well as any insights gleaned from them, largely depend on the quality of the source chronograms: low quality chronograms will produce low quality results.
+This \enquote{garbage in, garbage out} problem has long been recognised in supertree methods for summarizing phylogenetic trees (Bininda-Emonds et al., 2004).
+We note that this is a surfacing issue of any automated tool for biological data analysis.
+For example, DNA riddled with sequencing errors will produce generally poor alignments that will return biased evolutionary hypothesis, independently of the quality of the analysis software used.
+Again, we urge readers and DateLife users to explore all input chronograms before using a summary chronogram resulting from a DateLife workflow.
+Finally, uncertainty and variability of chronogram node age estimates might pose larger issues in some research areas than others.
+For example, in ecological and conservation biology studies, it has been shown that incorporating some chronogram data provides better results than when not using any age data at all, even if the node ages are not good quality (Webb et al., 2008).
+In the following sections we discuss the particularities of divergence times from DateLife's summary chronograms and their impact on certain evolutionary analyses, for consideration of the readers and users in different research areas.
 
@@
 See, for example, the robust ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández, 2015; Ramshaw et al., 1972; Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021).
-Source chronograms available for the same organisms have potentially been estimated implementing calibrations very differently.
-In the Fringillidae example above, the chronograms from Burns et al.
-(2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older age estimates for the same nodes than chronograms that were inferred using fossils as calibrations (Figs.
-\ref{fig:fringillidages}, \ref{fig:cvXY}; Supplementary Figs.
+Alternative source chronograms available for the same taxa have potentially been estimated implementing different types of calibrations, which affects the resulting node age estimates.
+For example, in the DateLife analysis of the Fringillidae shown above, the chronograms from one study (Burns et al., 2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older age estimates for the same nodes than chronograms that were inferred using fossil calibrations (Figs.
+4, 5; Supplementary Figs.
 S1, S5).
-
-Different calibration implementations might also imply fundamentally distinct evolutionary hypotheses (Antonelli et al., 2017).
+Another source of conflict in estimated node ages can arise from different placements for the same calibration, which would imply fundamentally distinct evolutionary hypotheses (Antonelli et al., 2017).
 For example, two independent researchers working on the same clade should both carefully select and justify their choices of fossil calibration placement.
-Yet, if one researcher concludes that a fossil should calibrate the ingroup of a clade, while another researcher concludes that the same fossil should calibrate the outgroup of the clade, the resulting age estimates will differ, as the placement of calibrations as stem or crown group has been proven to significantly affect time of lineage divergence estimates (Sauquet, 2013).
+Yet, if one researcher concludes that a fossil should calibrate the ingroup of a clade, while another researcher concludes that the same fossil should calibrate the outgroup of the clade, the resulting age estimates will differ, as the placement of calibrations as stem or crown group is known to significantly affect estimates of time of lineage divergence (Sauquet, 2013).
+Finally, placement of calibrations also affects uncertainty of node age estimates.
+For example, nodes that are sandwiched between a calibrated node and a calibrated root have less freedom of movement and hence narrower confidence intervals (Vos \& Mooers, 2004), which inflates precision for nodes without calibrations but does not necessarily improve accuracy of the estimated ages.
+
+DateLife's summary chronograms are intended to represent all variation in estimated node ages from source chronograms.
+Node age distribution ranges allow to visually explore ages from source chronograms individually and contextualize and compare them against other chronograms.
+Researchers that wish to use summary chronograms in downstream evolutionary analysis may select multiple trees sampled from the summary distribution of node ages, to account for variation in source chronograms.
 
@@
 
-While most chronograms in DateLife's database are constructed using primary calibrations (molecular substitution rates or ages obtained from the fossil record or geological events), DateLife summarizes chronograms using secondary calibrations (ages coming from other chronograms).
-Graur and Martin (2004) cautioned on the increased error and uncertainty in estimated ages when using secondary calibrations in dating analyses.
-Schenk (2016) showed that, in simulations, divergence times inferred using secondary calibrations are significantly younger than those inferred with primary calibrations, when obtained with Bayesian inference methods, and when priors are implemented in similar ways in both analyses.
-Accordingly, the scientific community seems to have more confidence in chronograms obtained from a single analysis, using fossil data as primary sources of calibrations (Schenk, 2016), and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data reflect a coherent evolutionary history (Sauquet, 2013), as for example done by Antonelli et al.
+While most chronograms in DateLife's database are constructed using primary calibrations, i.e., molecular substitution rates or absolute ages in Myr obtained from the fossil record or geological events, DateLife also constructs summary chronograms using secondary calibrations, i.e., node ages in Myr extracted from existing chronograms.
+The scientific community has generally more confidence in chronograms generated from a single analysis, where carefully chosen primary calibrations constitute the main source of absolute time information used in the dating analysis (Antonelli et al., 2017; Garzón-Orduña, Silva-Brandão, Willmott, Freitas, \& Brower, 2015; Graur \& Martin, 2004; Sauquet, 2013; Sauquet et al., 2012; Schenk, 2016; Shaul \& Graur, 2002).
+However, implementation of primary calibrations does not come without difficulties: it requires high expertise and training to discover, place and apply calibrations appropriately (Hipsley \& Müller, 2014; Ksepka et al., 2011).
+One way around it is to use fossils that have been widely discussed and previously curated as calibrations to date other trees (Ksepka et al., 2011; Sauquet, 2013), and making sure that all data reflect a coherent evolutionary history (Sauquet, 2013), as for example done by Antonelli et al.
 (2017).
-There have been attempts to create fossil calibration databases (Ksepka et al., 2015), though these still have room to grow.
-
-It seems that using several (as opposed to just a few) secondary calibrations can provide sufficient information to alleviate or even neutralize potential biases (Sauquet, 2013).
-Certainly, further studies are required to fully understand the effect of secondary calibrations on outputs from different tree dating methods, and on downstream analyses.
-It is possible that secondary calibrations can be safely used with dating methods that do not require setting priors, such as penalized likelihood (Sanderson, 2003), with methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ (Webb et al., 2008; Webb \& Donoghue, 2005), or methods that summarize age data unto a tree topology.
-
-Our cross validation analysis might provide some insight in this regard.
-When ages are estimated with secondary calibrations, nodes closer to the root do tend to be slightly younger than ages estimated with primary calibrations.
-However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures S2-S20).
-The only exception to this was observed on cross validation results of the Burns et al.
-(2014) chronogram, which displays much younger node ages when estimated using secondary calibrations (Supplementary Figs.
-S1, S5).
+The Fossil Calibration Database provides availability of 220 primary calibration points encompassing flowering plants and metazoans, that have been curated by experts and used for dating analysis in peer-reviewd publications (Ksepka et al., 2015).
+This database facilitates the use of expert primary fossil calibrations in new phylogenetic dating analyses.
+Yet, users still require the expertise to locate appropriate nodes in their phylogenies to corresponding fossils available in the database, potentially delaying increased usage of the database in research.
+Open databases for effective reuse of fossil calibration data still have room to grow and should constitute a key addition to the phylogenetics tool set.
+
+In some cases, secondary calibrations might represent the only information available for a dating analysis (e.g., groups with a poor fossil record).
+Recently, Powell, Waskin, and Battistuzzi (2020) showed in a simulation analysis that secondary calibrations are as good as primary calibrations.
+It seems that using several secondary calibrations (as opposed to just one) can provide sufficient information to alleviate or even neutralize potential biases described before (Graur \& Martin, 2004; Sauquet, 2013; Shaul \& Graur, 2002).
+Our cross validation analysis also provides some insight into the application of secondary calibrations.
+In general, node ages summarized with DateLife and those from the original studies appear positively correlated (Supplementary Figs.
+S2-S20).
+We also note that nodes closer to the root tend to be slightly younger than ages from the original studies.
+In contrast, nodes closer to the tips tend to be slightly older when estimated using secondary calibrations than ages from the original studies.
+The only exception to this trend was observed in Burns et al.
+(2014) chronogram, which generally displays much younger node ages when estimated using secondary calibrations than the original study (Supplementary Figs.
+S1, S5), supporting previous observations (Sauquet et al., 2012; Schenk, 2016).
+
+We unite to the call of our colleagues: further research is needed to fully understand the effects of using secondary calibrations with alternative tree dating methods, as well as usage of resulting chronograms in downstream analyses (Hipsley \& Müller, 2014; Powell et al., 2020; Schenk, 2016; Shaul \& Graur, 2002).
+It is possible that secondary calibrations can be used successfully with dating methods that do not require setting priors, such as penalized likelihood (Sanderson, 2003); with methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ (Webb et al., 2008; Webb \& Donoghue, 2005); and with methods that summarize age data unto a tree topology, such as DateLife.
 
@@
 \begin{center}
-\emph{Effects of Phylogenetic Sampling on Downstream Analyses}
-\end{center}
-
-Analysis of species diversification using simulated and empirical phylogenies suggest that using a more completely sampled phylogeny provides estimates that are closer to the true diversification history than when analysing incompletely sampled phylogenies (Chang, Rabosky, \& Alfaro, 2020; Cusimano, Stadler, \& Renner, 2012; Sun et al., 2020).
-Ideally, phylogenies should be completed using genetic data, but this is a difficult task to achieve for many biological groups.
-Hence, DateLife's workflow features different ways of assigning node ages in the absence of calibrations and branch length information for certain taxa.
-Completing a phylogeny using a stochastic birth-death politomy resolver and a backbone taxonomy is a common practice in scientific publications: Jetz et al.
+\emph{Effects of Taxon Sampling on Downstream Analyses}
+\end{center}
+
+Analysis of species diversification of simulated and empirical phylogenies suggest that using a more completely sampled phylogeny provides estimates that are closer to the true diversification history than when analysing incompletely sampled phylogenies (Chang, Rabosky, \& Alfaro, 2020; Cusimano, Stadler, \& Renner, 2012; Sun et al., 2020).
+Ideally, phylogenies should be completed using genetic data, but this is a time-consuming and difficult task to achieve for many biological groups.
+Hence, DateLife's workflow features different ways of assigning divergence times to taxa with missing the absence of branch length data and calibrations and branch lengths for certain taxa.
+
+Completing a phylogeny using a stochastic birth-death polytomy resolver and a backbone taxonomy is a common practice in scientific publications: Jetz et al.
 (2012), created a chronogram of all 9, 993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
 (2018) created a chronogram of 31, 536 ray-finned fishes, of which only 37\% had molecular data; Smith and Brown (2018) constructed a chronogram of 353, 185 seed plants where only 23\% had molecular data.
-These stochastically resolved chronograms provide less biased diversification rates estimated with methods that account for sampling fractions (Chang et al., 2020; Cusimano et al., 2012), but can also introduce spurious patterns of early bursts of diversification (Cusimano \& Renner, 2010; Sun et al., 2020).
-
-Taxonomy based politomy resolvers also introduce topological differences.
+These stochastically resolved chronograms return diversification rates estimates that appear less biased than those estimated from their incompletely sampled counterparts, even with methods that account for missing lineages by using sampling fractions (Chang et al., 2020; Cusimano et al., 2012), but can also introduce spurious patterns of early bursts of diversification (Cusimano \& Renner, 2010; Sun et al., 2020).
+
+Taxonomy-based stochastic polytomy resolvers also introduce topological differences in phylogenetic trees.
 The study of macroevolutionary processes largely depends on an understanding of the timing of species diversification events, and different phylogenetic and chronogram hypothesis can provide very different overviews of the macroevolutionary history of a biological group.
 For example, alternative topologies in chronograms from the same biological group can infer very different species diversification patterns (Rabosky, 2015; Title \& Rabosky, 2016).
-Similarly there are worries that patterns of morphological evolution cannot be accurately inferred with these types of phylogenies, as any patterns would be erased by randomization (Rabosky, 2015).
-We note that the same applies for geographical and morphological dependent diversification analysis.
-Hence, we suggest that these types of randomly resolved phylogenies can be used as null or neutral models, representing the case of diversification independent of traits and geographical scenario.
-
-While stochastic politomy resolvers have been useful to advance research, notably, risks come with this practice.
+Similarly, there are worries that patterns of morphological evolution cannot be accurately inferred with phylogenies that have been resolved stochastically over a taxonomic backbone, as any patterns would be erased by randomization (Rabosky, 2015).
+We note that the same applies for geography- and morphology-dependent diversification analysis.
+Hence, we suggest that phylogenies that have been processed with taxonomy-based stochastic polytomy resolvers, including certain summary chronograms from a DateLife analysis, can be useful as null or neutral models, representing the case of a diversification process that is independent of traits and geographical scenario.
+
+Taxonomy-based stochastic polytomy resolvers have been used to advance research in evolution, still, risks come with this practice.
 Taken to the extreme, one could generate a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy, a single calibration, and assigning branch lengths following a birth-death diversification model.
 Clearly, this can lead to a misrepresentation of the true evolutionary history.
-We urge DateLife users to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, potential biases, and assessing the consistency of the results with prior work.
+We urge DateLife users to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, potential biases, and assessing the consistency of DateLife's results with prior work.
 
@@
 
-Supplementary material, including code, biological examples, benchmark results, data files and online-only appendices, can be found in the Dryad data repository (\url{https://doi.org/10.5061/dryad.cnp5hqc6w}), as well as in the Zenodo stable repositories that host the reproducible manuscript (\url{https://doi.org/10.5281/zenodo.7435094}), the biological examples (\url{https://doi.org/10.5281/zenodo.7435101}), and the software benchmark (\url{https://doi.org/10.5281/zenodo.7435106}).
-The corresponding development versions hosted on GitHub can be found at \url{https://github.com/LunaSare/datelifeMS1\#readme}, \url{https://github.com/LunaSare/datelife_examples\#readme}, and \url{https://github.com/LunaSare/datelife_benchmark\#readme}.
+Supplementary Figures can be viewed and dowloaded from their Zenodo repository (\url{https://doi.org/10.5281/zenodo.6683667}).
+Supplementary material, including code, biological examples, benchmark results, data files and online-only appendices, can be downloaded from the Dryad data repository (\url{https://doi.org/10.5061/dryad.cnp5hqc6w}), as well as in the Zenodo stable repositories that host the reproducible manuscript (\url{https://doi.org/10.5281/zenodo.7435094}), the biological examples (\url{https://doi.org/10.5281/zenodo.7435101}), and the software benchmark (\url{https://doi.org/10.5281/zenodo.7435106}).
+Development versions corresponding to all of the above are hosted on GitHub, accesible at \url{https://github.com/LunaSare/datelifeMS1}, \url{https://github.com/LunaSare/datelife_examples}, and \url{https://github.com/LunaSare/datelife_benchmark}.
 
@@
 
+We thank Isabel Sanmartín, Daniele Silvestro, Rutger Vos and an anonymous reviewer, for comments that greatly improved this manuscript.
 The DateLife project was born as a prototype tool aiming to provide the services describe in this paper, and was initially developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA (Stoltzfus et al., 2013).
@@
 \leavevmode\hypertarget{ref-agnarsson2008acctran}{}%
-Agnarsson, I., \& Miller, J. A. (2008). Is acctran better than deltran? \emph{Cladistics}, \emph{24}(6), 1032--1038.
+Agnarsson, I., \& Miller, J. A. (2008). Is ACCTRAN better than DELTRAN? \emph{Cladistics}, \emph{24}(6), 1032--1038.
 
@@
 \emph{Proceedings of the Royal Society B: Biological Sciences}, \emph{272}(1565), 849--858.
+
+\leavevmode\hypertarget{ref-bininda2004garbage}{}%
+Bininda-Emonds, O.
+R., Jones, K.
+E., Price, S.
+A., Cardillo, M., Grenyer, R., \& Purvis, A.
+(2004).
+Garbage in, garbage out: Data issues in supertree construction.
+\emph{Phylogenetic Supertrees: Combining Information to Reveal the Tree of Life}, 267--280.
 
@@
 
+\leavevmode\hypertarget{ref-garzon2015incompatible}{}%
+Garzón-Orduña, I.
+J., Silva-Brandão, K.
+L., Willmott, K.
+R., Freitas, A.
+V., \& Brower, A.
+V.
+(2015).
+Incompatible ages for clearwing butterflies based on alternative secondary calibrations.
+\emph{Systematic Biology}, \emph{64}(5), 752--767.
+
 \leavevmode\hypertarget{ref-gbif2022taxonomy}{}%
@@
 Heibl, C. (2008). \emph{PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages.} Retrieved from \url{http://www.christophheibl.de/Rpackages.html}
+
+\leavevmode\hypertarget{ref-hipsley2014beyond}{}%
+Hipsley, C. A., \& Müller, J. (2014). Beyond fossil calibrations: Realities of molecular clock practices in evolutionary biology. \emph{Frontiers in Genetics}, \emph{5}, 138.
 
@@
 \emph{Molecular Phylogenetics and Evolution}, \emph{77}, 177--182.
+
+\leavevmode\hypertarget{ref-ksepka2011synthesizing}{}%
+Ksepka, D.
+T., Benton, M.
+J., Carrano, M.
+T., Gandolfo, M.
+A., Head, J.
+J., Hermsen, E.
+J., \ldots{} others.
+(2011).
+\emph{Synthesizing and databasing fossil calibrations: Divergence dating and beyond}.
+The Royal Society.
 
@@
 
+\leavevmode\hypertarget{ref-powell2020quantifying}{}%
+Powell, C.
+L.
+E., Waskin, S., \& Battistuzzi, F.
+U.
+(2020).
+Quantifying the error of secondary vs.
+Distant primary calibrations in a simulated environment.
+\emph{Frontiers in Genetics}, \emph{11}, 252.
+
 \leavevmode\hypertarget{ref-price2014niche}{}%
@@
 
+\leavevmode\hypertarget{ref-sauquet2012testing}{}%
+Sauquet, H., Ho, S.
+Y.
+W., Gandolfo, M.
+a, Jordan, G.
+J., Wilf, P., Cantrill, D.
+J., \ldots{} Udovicic, F.
+(2012).
+Testing the impact of calibration on molecular divergence times using a fossil-rich group: the case of Nothofagus (Fagales).
+\emph{Systematic Biology}, \emph{61}(2), 289--313.
+\url{https://doi.org/10.1093/sysbio/syr116}
+
 \leavevmode\hypertarget{ref-sauquet2021age}{}%
@@
 \emph{Molecular Phylogenetics and Evolution}, \emph{88}, 1--15.
+
+\leavevmode\hypertarget{ref-shaul2002playing}{}%
+Shaul, S., \& Graur, D.
+(2002).
+Playing chicken (gallus gallus): Methodological inconsistencies of molecular divergence date estimates due to secondary calibration points.
+\emph{Gene}, \emph{300}(1-2), 59--61.
 
@@
 
+\leavevmode\hypertarget{ref-vos2004reconstructing}{}%
+Vos, R. A., \& Mooers, A. Ø. (2004). Reconstructing divergence times for supertrees: A molecular approach. \emph{Phylogenetic Supertrees: Combining Information to Reveal the Tree of Life}, 281--299.
+
 \leavevmode\hypertarget{ref-Webb2000}{}%
@@
 
-\newpage
-
-\captionof{table}{Ages of congruified nodes. See Figure 3, step C2.}
-\label{tbl:table1}
-\includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example.pdf}
-
-\vspace{15mm}
-
-\captionof{table}{Summary of congruified nodes ages. See Figure 3, step C3.}
-
-\label{tbl:table2}
-\includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example-summary.png}
-
-\begin{minipage}[t]{17.3cm}
-\includegraphics{../figures/figure-workflow/figure1-new.pdf}
-\captionof{figure}{\textit{Main DateLife workflow.
-Analyses can be performed via DateLife's interactive website at \url{www.datelife.org}, or using the \texttt{datelife} R package.
-Details on the R functions used to perform the analyses are available from \texttt{datelife}'s R package vignettes at \url{https://phylotastic.org/datelife}.}}
-\label{fig:workflow}
-\end{minipage}
-
-\includegraphics{../figures/fig_runtime_main.pdf}
-\captionof{figure}{\textit{DateLife's benchmarking results.
-Computation time used to process a query and a search across \texttt{datelife}'s chronogram database, relative to number of input taxon names.
-For each N = \{10, 100, 200, ..., 1 000, ...
-, 9 000, 10 000\}, we sampled N species names from the class Aves a hundred times, and then performed a datelife search processing the input names with Taxon Names Resolution Service (TNRS; light gray), and without processing names (dark gray).
-For comparison, we performed a search using an input that had been pre-processed with TNRS (light blue).}}
-\label{fig:runtime_main}
-
-\begin{center}
-\includegraphics{../figures/figure-small-example/figure2.pdf}
-\end{center}
-\vspace{10mm}
-\captionof{figure}{\textit{DateLife analysis results for a small sample of A) 6 bird species within the Passeriformes.
-B) Processed species names were found across 9 chronograms within 6 independent studies (Barker et al.
-(\protect\hyperlink{ref-barker2012going}{2012}), Barker et al.
-(\protect\hyperlink{ref-barker2015new}{2015}), Burns et al.
-(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al.
-(\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al.
-(\protect\hyperlink{ref-Jetz2012}{2012}).) C) This revealed 28 source age data points for the queried species names.
-Summarized age data is used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree, resulting in a summary chronogram of source ages.}}
-
-\label{fig:figure2}
-
-\begin{minipage}{18.2cm}
-\includegraphics{../figures/fringillidae-topologies/fringillidae-topology.pdf}
-\captionof{figure}{Tree topologies extracted from Open Tree of Life's (OpenTree) synthetic phylogenetic tree.
-A) Topology of 2,333 tips and 1,305 internal nodes, encompassing bird species within the family Fringillidae following the NCBI taxonomy (black), as well as all other bird species that share the same Most Recent Common Ancestor (MRCA) node in OpenTree's synthetic tree (purple).
-B) Subtree topology of 289 tips and 253 internal nodes, resulting from pruning species that do not belong to the family Fringillidae according to the NCBI taxonomy (purple branches in topology A).
-Bird species within the Fringillidae are paraphyletic
-(Alström et al. \protect\hyperlink{ref-alstrom2014discovery}{2014},
-Barker, Cibois, Schikler, Feinstein, \& Cracraft \protect\hyperlink{ref-barker2004phylogeny}{2004},
-Barker et al. \protect\hyperlink{ref-barker2013going}{2013},
-Barker \protect\hyperlink{ref-barker2014mitogenomic}{2014},
-Barker et al. \protect\hyperlink{ref-barker2015new}{2015},
-Beresford, Barker, Ryan, \& Crowe \protect\hyperlink{ref-beresford2005african}{2005},
-Bryson Jr et al. \protect\hyperlink{ref-bryson2014diversification}{2014},
-Burleigh, Kimball, \& Braun \protect\hyperlink{ref-burleigh2015building}{2015},
-Burns et al. \protect\hyperlink{ref-burns2014phylogenetics}{2014},
-Chaves, Hidalgo, \& Klicka \protect\hyperlink{ref-chaves2013biogeography}{2013},
-Claramunt \& Cracraft \protect\hyperlink{ref-claramunt2015new}{2015},
-Gibb et al. \protect\hyperlink{ref-gibb2015new}{2015},
-Hackett et al. \protect\hyperlink{ref-hackett2008phylogenomic}{2008},
-Jetz et al. \protect\hyperlink{ref-Jetz2012}{2012},
-Johansson, Fjeldså, \& Bowi \protect\hyperlink{ref-johansson2008phylogenetic}{200},
-Kimball et al. \protect\hyperlink{ref-kimball2019phylogenomic}{2019},
-Klicka et al. \protect\hyperlink{ref-klicka2014comprehensive}{2014},
-Lamichhaney et al. \protect\hyperlink{ref-lamichhaney2015evolution}{2015},
-Lerner, Meyer, James, Hofreiter, \& Fleischer \protect\hyperlink{ref-lerner2011multilocus}{2011},
-Lovette et al. \protect\hyperlink{ref-lovette2010comprehensive}{2010},
-Moyle et al. \protect\hyperlink{ref-moyle2016tectonic}{2016},
-Ödeen, Håstad, \& Alström \protect\hyperlink{ref-odeen2011evolution}{2011},
-Oliveros et al. \protect\hyperlink{ref-oliveros2019earth}{2019},
-Päckert et al. \protect\hyperlink{ref-packert2012horizontal}{2012},
-Parchman, Benkman, \& Mezquida \protect\hyperlink{ref-parchman2007coevolution}{2007},
-Powell et al. \protect\hyperlink{ref-powell2014comprehensive}{2014},
-  Price et al. \protect\hyperlink{ref-price2014niche}{2014},
-  Pulgarín-R, Smith, Bryson Jr, Spellman, \& Klicka \protect\hyperlink{ref-pulgarin2013multilocus}{2013},
-  Selvatti, Gonzaga, \& Moraes Russo \protect\hyperlink{ref-selvatti2015paleogene}{2015},
-  Tietze, Päckert, Martens, Lehmann, \& Sun \protect\hyperlink{ref-tietze2013complete}{2013},
-  Treplin et al. \protect\hyperlink{ref-treplin2008molecular}{2008},
-  Zuccon, Prŷs-Jones, Rasmussen, \& Ericson \protect\hyperlink{ref-zuccon2012phylogenetic}{2012}).}
-\label{fig:fringillidae-topologies}
-\end{minipage}
-
-\begin{minipage}{18cm}
-\begin{center}
-\includegraphics{../figures/figure-fringillidae/median_and_calibration_ages-congruified.png}
-\end{center}
-
-\captionof{figure}{Fringillidae median summary chronogram generated with DateLife.
-It has 289 tips and 253 nodes, from which 212 have age data from at least one published chronogram.
-These nodes have age data points from 19 different chronograms from 13 different studies (colored circles), which were summarized into 194 data points that were used as secondary calibrations (purple slashes) with the Branch Length Adjuster (BLADJ ) software from Webb et al., (\protect\hyperlink{ref-Webb2008}{2008}).}
-\label{fig:fringillidages}
-\end{minipage}
-
-\newpage
-
-\begin{minipage}{18cm}
-\begin{center}
-\includegraphics{../figures/fringillidae-cross-validation/fig-cross-validation-xy-plots.pdf}
-\end{center}
-\captionof{figure}{Results from cross validation analysis. Each plot compares the original age estimate (x axis) with the age obtained with a DateLife analysis (y axis), per node.}
-\label{fig:cvXY}
-\end{minipage}
-
 \end{document}

### 2023-05-23 submission/final-captions.tex
add submission folder and contents

--- 2023-02-21 manuscript/manuscript-1st-submission.tex
@@
-\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
-\PassOptionsToPackage{hyphens}{url}
-%
-\documentclass[english,man]{apa6}
-\usepackage{lmodern}
-\usepackage{amssymb,amsmath}
-\usepackage{ifxetex,ifluatex}
-\usepackage{fixltx2e} % provides \textsubscript
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provides euro and other symbols
-\else % if luatex or xelatex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
-\fi
-% use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-% use microtype if available
-\IfFileExists{microtype.sty}{%
-\usepackage[]{microtype}
-\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\IfFileExists{parskip.sty}{%
-\usepackage{parskip}
-}{% else
-\setlength{\parindent}{0pt}
-\setlength{\parskip}{6pt plus 2pt minus 1pt}
-}
-\usepackage{hyperref}
-\hypersetup{
-            pdftitle={DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life},
-            pdfauthor={Luna L. Sánchez Reyes1,2, Emily Jane McTavish1, \& Brian O'Meara2},
-            pdfborder={0 0 0},
-            breaklinks=true}
-\urlstyle{same}  % don't use monospace font for urls
-\usepackage{graphicx,grffile}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-\setlength{\emergencystretch}{3em}  % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{0}
+Figure 1
 
-% set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
+Main DateLife workflow.
+Analyses can be performed via DateLife's interactive website at \url{www.datelife.org}, or using the \texttt{datelife} R package.
+Details on the R functions used to perform the analyses are available from \texttt{datelife}'s R package vignettes at \url{https://phylotastic.org/datelife}.
 
-\DeclareUnicodeCharacter{0301}{*************************************}
+Figure 2:
 
-% leave the figure captions outside\ caption{} if you want them
-% to be formatted in the same way as the general text (double spaced and linenumbered)
-% line numbers for captions trick from https://latex.org/forum/viewtopic.php?t=34614
-
-\usepackage{ragged2e}
-
-\usepackage{caption}
-
-\DeclareCaptionFont{linenumbers}{\internallinenumbers}
-
-\usepackage{float}
-\restylefloat{table}
-
-\captionsetup[figure]{font+=linenumbers, labelfont={sc}, labelformat={default}, labelsep=period, name={Figure}}
-
-\captionsetup[table]{font+=linenumbers, labelfont={sc}, labelformat={default}, labelsep=period, name={Table}}
-% Manuscript styling
-\usepackage{upgreek}
-\captionsetup{font=singlespacing,justification=justified}
-
-% Table formatting
-\usepackage{longtable}
-\usepackage{lscape}
-% \usepackage[counterclockwise]{rotating}   % Landscape page setup for large tables
-\usepackage{multirow}		% Table styling
-\usepackage{tabularx}		% Control Column width
-\usepackage[flushleft]{threeparttable}	% Allows for three part tables with a specified notes section
-\usepackage{threeparttablex}            % Lets threeparttable work with longtable
-
-% Create new environments so endfloat can handle them
-% \newenvironment{ltable}
-%   {\begin{landscape}\centering\begin{threeparttable}}
-%   {\end{threeparttable}\end{landscape}}
-\newenvironment{lltable}{\begin{landscape}\centering\begin{ThreePartTable}}{\end{ThreePartTable}\end{landscape}}
-
-% Enables adjusting longtable caption width to table width
-% Solution found at http://golatex.de/longtable-mit-caption-so-breit-wie-die-tabelle-t15767.html
-\makeatletter
-\newcommand\LastLTentrywidth{1em}
-\newlength\longtablewidth
-\setlength{\longtablewidth}{1in}
-\newcommand{\getlongtablewidth}{\begingroup \ifcsname LT@\roman{LT@tables}\endcsname \global\longtablewidth=0pt \renewcommand{\LT@entry}[2]{\global\advance\longtablewidth by ##2\relax\gdef\LastLTentrywidth{##2}}\@nameuse{LT@\roman{LT@tables}} \fi \endgroup}
-
-% \setlength{\parindent}{0.5in}
-% \setlength{\parskip}{0pt plus 0pt minus 0pt}
-
-% Overwrite redefinition of paragraph and subparagraph by the default LaTeX template
-% See https://github.com/crsh/papaja/issues/292
-\makeatletter
-\renewcommand{\paragraph}{\@startsection{paragraph}{4}{\parindent}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-1em}%
-  {\normalfont\normalsize\bfseries\itshape\typesectitle}}
-
-\renewcommand{\subparagraph}[1]{\@startsection{subparagraph}{5}{1em}%
-  {0\baselineskip \@plus 0.2ex \@minus 0.2ex}%
-  {-\z@\relax}%
-  {\normalfont\normalsize\itshape\hspace{\parindent}{#1}\textit{\addperi}}{\relax}}
-\makeatother
-
-% \usepackage{etoolbox}
-\makeatletter
-\patchcmd{\HyOrg@maketitle}
-  {\section{\normalfont\normalsize\abstractname}}
-  {\section*{\normalfont\normalsize\abstractname}}
-  {}{\typeout{Failed to patch abstract.}}
-\patchcmd{\HyOrg@maketitle}
-  {\section{\protect\normalfont{\@title}}}
-  {\section*{\protect\normalfont{\@title}}}
-  {}{\typeout{Failed to patch title.}}
-\makeatother
-
-\usepackage{xpatch}
-\makeatletter
-\xapptocmd\appendix
-  {\xapptocmd\section
-    {\addcontentsline{toc}{section}{\appendixname\ifoneappendix\else~\theappendix\fi\\: #1}}
-    {}{\InnerPatchFailed}%
-  }
-{}{\PatchFailed}
-\DeclareDelayedFloatFlavor{ThreePartTable}{table}
-\DeclareDelayedFloatFlavor{lltable}{table}
-\DeclareDelayedFloatFlavor*{longtable}{table}
-\makeatletter
-\renewcommand{\efloat@iwrite}[1]{\immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
-\makeatother
-\usepackage{lineno}
-
-\linenumbers
-\usepackage{csquotes}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[shorthands=off,main=english]{babel}
-\else
-  % load polyglossia as late as possible as it *could* call bidi if RTL lang (e.g. Hebrew or Arabic)
-  \usepackage{polyglossia}
-  \setmainlanguage[]{english}
-\fi
-
-\title{DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life}
-\author{Luna L. Sánchez Reyes\textsuperscript{1,2}, Emily Jane McTavish\textsuperscript{1}, \& Brian O'Meara\textsuperscript{2}}
-\date{}
-
-\shorttitle{DateLife: revealing the dated Tree of Life}
-
-\authornote{
-
-School of Natural Sciences, University of California, Merced, 258 Science and Engineering Building 1, Merced, CA 95340, USA.
-
-Department of Ecology and Evolutionary Biology, University of Tennessee, Knoxville, 446 Hesler Biology Building, Knoxville, TN 37996, USA.
-
-The authors made the following contributions.
-Luna L.
-Sánchez Reyes: Data curation, Investigation, Software, Visualization, Validation, Writing - Original Draft Preparation, Writing - Review \& Editing; Emily Jane McTavish: Resources, Software, Writing - Review \& Editing; Brian O'Meara: Conceptualization, Funding acquisition, Methodology, Resources, Software, Supervision, Writing - Review \& Editing.
-
-Correspondence concerning this article should be addressed to Luna L. Sánchez Reyes, . E-mail: \href{mailto:sanchez.reyes.luna@gmail.com}{\nolinkurl{sanchez.reyes.luna@gmail.com}}
-
-}
-
-\affiliation{\vspace{0.5cm}\textsuperscript{1} University of California, Merced, USA\\\textsuperscript{2} University of Tennessee, Knoxville, USA}
-
-\begin{document}
-\maketitle
-
-\begin{center}
-Abstract
-\end{center}
-
-Achieving a high-quality reconstruction of a phylogenetic tree with branch lengths proportional to absolute time (chronogram) is a difficult and time-consuming task.
-But the increased availability of fossil and molecular data, and time-efficient analytical techniques has resulted in many recent publications of large chronograms for a large number and wide diversity of organisms.
-Knowledge of the evolutionary time frame of organisms is key for research in the natural sciences. It also represent valuable information for education, science communication, and policy decisions.
-When chronograms are shared in public and open databases, this wealth of expertly-curated and peer-reviewed data on evolutionary timeframe is exposed in a programatic and reusable way, as intensive and localized efforts have improved data sharing practices, as well as incentivizited open science in biology.
-Here we present DateLife, a service implemented as an R package and an R Shiny website application available at www.datelife.org, that provides functionalities for efficient and easy finding, summary, reuse, and reanalysis of expert, peer-reviewed, public data on time frame of evolution.
-The main DateLife workflow constructs a chronogram for any given combination of taxon names by searching a local chronogram database constructed and curated from the Open Tree of Life Phylesystem phylogenetic database, which incorporates phylogenetic data from the TreeBASE database as well.
-We implement and test methods for summarizing time data from multiple source chronograms using supertree and congruification algorithms, and using age data extracted from source chronograms as secondary calibration points to add branch lengths proportional to absolute time to a tree topology.
-DateLife will be useful to increase awareness of the existing variation in alternative hypothesis of evolutionary time for the same organisms, and can foster exploration of the effect of alternative evolutionary timing hypotheses on the results of downstream analyses, providing a framework for a more informed interpretation of evolutionary results.
-
-\emph{Keywords}: Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations
-
-Word count: 5393
-
-\newpage
-
-Chronograms --phylogenies with branch lengths proportional to time-- provide key data on evolutionary time frame for the study of natural processes in many areas of biological research, such as developmental biology (Delsuc et al., 2018; Laubichler \& Maienschein, 2009), conservation biology (Felsenstein, 1985; Webb, 2000), historical biogeography (Posadas, Crisci, \& Katinas, 2006), and species diversification (Magallon \& Sanderson, 2001; Morlon, 2014).
-
-Building a chronogram is not an easy task.
-It requires obtaining and curating data to construct a phylogeny, selecting and placing appropriate calibrations on the phylogeny using independent age data points from the fossil record or other dated events, and inferring the full dated tree;
-it also generally requires specialized biological training, taxonomic domain knowledge, and a non-negligible amount of research time, computational resources and funding.
-
-Here we present the DateLife project which has the main goal of capturing age data from published chronograms, and making these data readily accessible to the community for reuse and reanalysis, for research, teaching, and science communication and policy.
-DateLife's core software application is available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org.
-It features key elements for scientific reproducibility, such as a versioned, open and fully public source database (McTavish et al., 2015), data stored and available in a computer readable format (Vos et al., 2012), automated and programmatic ways of accessing the data (Stoltzfus et al., 2013) and methods to summarize and compare the data.
-
-\begin{center}
-\textsc{Description}
-\end{center}
-
-DateLife's core software application consists of the R package \texttt{datelife}.
-Its current stable version -- v0.6.6, is available from The Comprehensive R Archive Network (CRAN) repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
-ape (Paradis, Claude, \& Strimmer, 2004),
-bold (Chamberlain, 2018),
-geiger (Pennell et al., 2014),
-paleotree (Bapst, 2012),
-phyloch (Heibl, 2008),
-phylocomr (Ooms \& Chamberlain, 2018),
-phytools (Revell, 2012),
-rotl (Michonneau, Brown, \& Winter, 2016), and
-taxize (Chamberlain, 2018; Chamberlain \& Szöcs, 2013).
-Figure \ref{fig:workflow} provides a graphical summary of the three main steps of the DateLife workflow: creating a search query, searching a database, and summarizing results from the search.
-
-\begin{center}
-\emph{Creating a Search Query}
-\end{center}
-
-DateLife starts by processing an input consisting of at least two taxon names, which can be provided as a comma separated character string or as tip labels on a tree.
-If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a \enquote{phylo} R object (Paradis et al., 2004).
-The input tree is not required to have branch lengths, and its topology is used in the summary steps described in the next section.
-
-DateLife accepts scientific names that can belong to any inclusive taxonomic group (e.g., genus, family, tribe, etc.) or a binomial species name.
-Subspecies and variants are ignored.
-If an input taxon name belongs to an inclusive taxonomic group, DateLife has two alternative behaviors defined by the \enquote{get species from taxon} flag.
-If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group following a standard taxonomy of choice, and adds them to the input string.
-Taxonomies currently supported by DateLife are Open Tree of Life (OpenTree) unified Taxonomy (OTT, Rees \& Cranston, 2017), the National Center of Biotechnology Information (NCBI) taxonomic database (Schoch et al., 2020), the Global Biodiversity Information Facility (GBIF) taxonomic backbone (GBIF Secretariat, 2022), and the Interim Register of Marine and Nonmarine Genera (IRMNG) database (Rees et al., 2017).
-If the flag is inactive, DateLife excludes any taxon names above the species level from the search query.
-
-DateLife processes input scientific names using a Taxonomic Name Resolution Service (TNRS), which increases the probability of correctly finding the queried taxon names in the chronogram database.
-TNRS detects, corrects and standardizes name misspellings and typos, variant spellings and authorities, and nomenclatural synonyms to a single taxonomic standard (Boyle et al., 2013).
-DateLife implements TNRS with OTT as standard (Open Tree Of Life et al., 2016; Rees \& Cranston, 2017), storing taxonomic identification numbers for further processing.
-
-The processed input taxon names are saved as an R object of a newly defined class, \texttt{datelifeQuery}, that is used in the following steps.
-This object contains the standardized names, the corresponding OTT identification numbers, and the topology of the input tree if any was provided.
-
-\begin{center}
-\emph{Searching a Chronogram Database}
-\end{center}
-
-At the time of writing of this manuscript (Jun 22, 2022),
-DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies.
-It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata (McTavish et al., 2015), which allows automatic and reproducible assembly of our chronogram database.
-Datelife's chronogram database is navigable as an R data object within the \texttt{datelife} R package.
-
-A unique feature of the Phylesystem is that any user can add new published, state-of-the-art chronograms any time, through their curator application (\url{https://tree.opentreeoflife.org/curator}).
-As chronograms are added to Phylesystem, they are incorporated into the chronogram database of the \texttt{datelife} package.
-The updated database is assigned a new version number, followed by a package release on CRAN.
-\texttt{datelife}'s chronogram database is updated as new chronogram data is added to Phylesystem, at a minimum of once a month and a maximum of every 6 months.
-Users can also implement functions from the \texttt{datelife} R package to trigger an update of the local chronogram database, to incorporate any new chronograms to the user's DateLife analysis before an official database update is released on CRAN.
-
-A DateLife search is implemented by matching processed taxon names provided by the user to tip labels in the chronogram database.
-Chronograms with at least two matching taxon names on their tip labels are identified and pruned down to preserve only the matched taxa.
-These matching pruned chronograms are referred to as source chronograms.
-Total distance (in units of millions of years) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Figure \ref{fig:workflow}).
-The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a \enquote{phylo} object or newick string.
-Finally, the patristic matrices are associated to the study citation where the original chronogram was published, and stored as an R object of the newly defined class \texttt{datelifeResult}.
-
-\begin{center}
-\emph{Summarizing Search Results}
-\end{center}
-
-Summary information is extracted from the \texttt{datelifeResult} object to inform decisions for subsequent steps in the analysis workflow. Basic summary information available to the user is:
-
-\begin{enumerate}
-\def\labelenumi{\arabic{enumi}.}
-\tightlist
-\item
-  The matching pruned chronograms as newick strings or \enquote{phylo} objects.
-\item
-  The ages of the root of all source chronograms.
-These ages can correspond to the age of the most recent common ancestor (mrca) of the user's group of interest if the source chronograms have all taxa belonging to the group.
-If not, the root corresponds to the mrca of a subgroup withing the group of interest.
-\item
-  Study citations where original chronograms were published.
-\item
-  A report of input taxon names matches across source chronograms.
-\item
-  The source chronogram(s) with the most input taxon names.
-\item
-  Various single summary chronograms resulting from summarizing age data, generated using the methodology described next.
-\end{enumerate}
-
-\begin{center}
-\emph{Choosing a Topology}
-\end{center}
-
-DateLife requires a tree topology to summarize age data upon.
-We recommend that users provide as input a tree topology from the literature, or one of their own making.
-If no topology is provided, DateLife automatically extracts one from the OpenTree synthetic tree, a phylogeny encompassing 2.3 million taxa across all life, assembled from 1, 239 published phylogenetic trees and OpenTree's unified Taxonomy, OTT (Open Tree Of Life et al., 2019).
-Alternatively, DateLife can combine topologies from source chronograms using a supertree approach.
-To combine topologies from source chronograms into a single summary (or supertree) topology, the DateLife workflow identifies the source chronograms that form a grove, roughly, a sufficiently overlapping set of taxa between trees, by implementing definition 2.8 for n-overlap from Ané et al.
-(2009).
-In rare cases, a group of trees can have multiple groves.
-By default, DateLife chooses the grove with the most taxa, however, the \enquote{criterion = trees} flag allows the user to choose the grove with the most trees instead.
-If source chronograms do not form a grove, the supertree reconstruction will fail.
-
-\begin{center}
-\emph{Dating the Topology}
-\end{center}
-
-Input topologies from OpenTree or the supertree approach described above do not include branch length estimates of any kind.
-Optionally, to estimate branch lengths proportional to substitution rates on these topologies, DateLife can mine the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain genetic markers for the input taxa.
-These markers are aligned with MUSCLE (Edgar, 2004) (by default) or MAFFT (Katoh, Asimenos, \& Toh, 2009).
-This alignment can be used to estimate branch lengths on input topologies that lack branch lengths.
-Currently, branch length reconstruction in DateLife is performed using parsimony and the likelihood of the phylogenetic tree given a sequence alignment is computed (Schliep, 2011).
-While relative branch length information provides additional data for nodes without secondary date calibrations, topologies without branch lengths can also be dated.
-
-Once a topology is chosen, DateLife applies the congruification method (Eastman, Harmon, \& Tank, 2013) to find nodes belonging to the same clade across source chronograms, and extract the corresponding node ages from the patristic distance matrices stored as \texttt{datelifeResult}.
-By definition, the matrices store total distance (time from tip to tip), hence, node ages correspond to half the values stored in the patristic distance matrices.
-This assumes that the terminal taxa are coeval and occur at the present.
-A table of congruified node ages that can be used as calibrations for a dating analysis is stored as a \texttt{congruifiedCalibrations} object.
-
-For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method (SDM; Criscuolo, Berry, Douzery, \& Gascuel, 2006), which deforms patristic distance matrices by minimizing variance and then averaging them.
-These single summary taxon pair age matrices (Summarized calibrations) can be applied as calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Webb, Ackerly, \& Kembel, 2008; Webb \& Donoghue, 2005), and treePL (Smith \& O'Meara, 2012).
-
-By default, DateLife implements the Branch Length Adjuster (BLADJ) algorithm to obtain a fully dated topology.
-BLADJ fixes node ages that have calibration data, and distributes time between nodes with no data evenly between nodes with calibration data.
-This minimizes age variance in the resulting chronogram (Webb et al., 2008).
-BLADJ does not use branch lengths even when they are present in the input tree or summarizing topology.
-When there is conflict in ages between nodes with calibration data, BLADJ ignores node ages that are older than the age of a parent node.
-BLADJ requires a root age estimate.
-If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature.
-If none is provided, DateLife assigns an arbitrary age to the root as 10\% older than the oldest age available within the group.
-
-Alternative phylogenetic dating options supported in DateLife (MrBayes, PATHD8, TreePL) incorporate branch length information from the input topology in combination with the calibrations.
-PATHd8 is a non-clock, rate-smoothing method (Britton et al., 2007) to date trees.
-treePL (Smith \& O'Meara, 2012), is a semi-parametric, rate-smoothing, penalized likelihood dating method (Sanderson, 2002).
-The MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003) approach in DateLife uses the calibrations as priors on node ages.
-
-\begin{center}
-\emph{Visualizing Results}
-\end{center}
-
-Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R \enquote{phylo} format), as well as visualize and compare results graphically, or construct their own graphs using DateLife's chronogram plot generation functions available from the R package \texttt{datelifeplot} (Sanchez-Reyes \& O'Meara, 2022).
-
-\newpage
-
-\begin{center}
-\textsc{Benchmark}
-\end{center}
-
-\texttt{datelife}'s R package code speed was tested on an Apple iMac
-with one 3.4 GHz Intel Core i5 processor.
-We registered variation in computing time of query processing and search through the database relative to number of queried taxon names.
-Query processing time increases roughly linearly with number of input taxon names, and
-increases considerably if Taxonomic Name Resolution Service (TNRS) is activated.
-Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
-Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig.
-\ref{fig:runtime_main}).
-
-\texttt{datelife}'s code performance was evaluated with a set of unit tests designed and
-implemented with the R package testthat (R Core Team, 2018) that were run both locally
-with the devtools package (R Core Team, 2018), and on a public server using the continuous integration tool of GitHub actions (\url{https://docs.github.com/en/actions}).
-At present, unit tests cover more than 40\% of \texttt{datelife}'s code (\url{https://codecov.io/gh/phylotastic/datelife}).
-Unit testing helps identify potential issues as code is updated or, more critically, as services code relies upon may change.
-
-\begin{center}
-\textsc{Case Studies}
-\end{center}
-
-We illustrate the DateLife workflow using a family within the Passeriform birds encompassing the true finches, Fringillidae, as case study.
-On a small example, we analysed 6 bird species, and results from each step of the workflow are shown in Fig.
-\ref{fig:figure2}.
-As a second example, we analysed 289 bird species in the family Fringillidae that are included in the NCBI taxonomy.
-The resulting summary chronogram is shown in Fig.
-\ref{fig:fringillidages}, and results from previous steps of the workflow are available as Supplementary Figures.
-
-\begin{center}
-\emph{A Small Example}
-\end{center}
-
-\textbf{\emph{Creating a search query.--}}
-We chose 6 bird species within the Passeriformes. The sample includes
-two species of cardinals:
-the black-thighed grosbeak -- \emph{Pheucticus tibialis} and
-the crimson-collared grosbeak -- \emph{Rhodothraupis celaeno};
-three species of buntings:
-the yellowhammer -- \emph{Emberiza citrinella},
-the pine bunting -- \emph{Emberiza leucocephalos} and
-the yellow-throated bunting -- \emph{Emberiza elegans};
-and one species of tanager, the vegetarian finch -- \emph{Platyspiza crassirostris}.
-Processing of input names found that \emph{Emberiza elegans} is synonym for \emph{Schoeniclus elegans} in the default reference taxonomy (OTT v3.3, June 1, 2021).
-For a detailed discussion on the state of the synonym, refer to Avibase (Avibase, 2022; Lepage, 2004; Lepage, Vaidya, \& Guralnick, 2014).
-Discovering this synonym allowed assigning five age data points for the parent node of \emph{Emberiza elegans}, shown as \emph{Schoeniclus elegans} in figure \ref{fig:figure2}A, which would not have had any data otherwise.
-
-\newpage
-
-\textbf{\emph{Searching the database.--}}
-DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms in 6 different studies (Fig.
-\ref{fig:figure2}B).
-Three studies matched five input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2015; Hedges, Marin, Suleski, Paymer, \& Kumar, 2015; Jetz, Thomas, Joy, Hartmann, \& Mooers, 2012), one study matched four input names (Hooper \& Price, 2017) and two studies matched two input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2013; Burns et al., 2014).
-No studies matched all input names.
-Together, source chronograms provide 28 unique age data points, covering all nodes on our chosen tree topology to date (Table \ref{tbl:table1}).
-
-\vspace{3mm}
-
-\textbf{\emph{Summarizing search results.--}}
-DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. \ref{fig:figure2}C), and congruified and mapped age data to nodes in this chosen topology (Table \ref{tbl:table1}).
-The name processing step allowed including five data points for node \enquote{n4} (parent of \emph{Schoeniclus elegans}; Fig.
-\ref{fig:figure2}A) that would not have had any data otherwise due to name mismatch.
-Age summary statistics per node were calculated (Table \ref{tbl:table2}) and used as calibrations to date the tree topology using the BLADJ algorithm.
-As expected, more inclusive nodes (e.g., node \enquote{n1}) have more variance in age data than less inclusive nodes (e.g., node \enquote{n5}).
-Summary age data for node \enquote{n2} were excluded as final calibration because they are older than age data of the more inclusive node, \enquote{n1} (Fig. \ref{fig:figure2}C4).
-
-\newpage
-
-\newpage
-
-\begin{center}
-\emph{An Example with the Family of True Finches}
-\end{center}
-
-\textbf{\emph{Creating a query.--}}
-To obtain ages for all species within the family of true finches, Fringillidae, we ran a DateLife query using the \enquote{get species from taxon} flag,
-which gets all recognized species names within a named group from a taxonomy of choice.
-Following the NCBI taxonomy, our DateLife query has 289 Fringillidae species.
-This taxon-constrained approach implies that the final results of a full DateLife analysis will be done using a tree topology and ages for the species in a named group, which do not necessarily correspond to a monophyletic group.
-Users can change this behaviour by providing a monophyletic tree as input for a DateLife search, or as a tree topology for a DateLife summary.
-
-\textbf{\emph{Searching the database.--}}
-Next, we used the processed species names in our DateLife query to identify chronograms with at least two Fringillidae species.
-The DateLife search identified 13 chronograms containing at least two Fringillidae species, published in 9 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Price et al., 2014).
-Once identified, DateLife pruned matching chronograms to keep Fringillidae species names on tips only, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementray Table S1).
-
-\textbf{\emph{Summarizing search results.--}}
-The final step is to congruify and summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM.
-As explained in the \enquote{Description} section, a tree topology to summarize age data upon is required.
-By default, DateLife uses the topology from OpenTree's synthetic tree that contains the species in the search query to summarize age data upon.
-According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig. \ref{fig:fringillidae-topologies}).
-
-\vspace{50mm}
-
-Age data from source chronograms was congruified to OpenTree's topology (Fig.
-\ref{fig:fringillidae-topologies}B), reducing the age data set to 818 different data points (Supplementray Table S2).
-For each congruent node, age summary statistics were calculated and used as fixed secondary calibrations over the chosen tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
-\ref{fig:fringillidages}).
-
-\begin{center}
-\textsc{Cross-Validation Test}
-\end{center}
-
-We performed a cross validation analysis of the DateLife workflow using the Fringillidae chronograms.
-We used the individual tree topologies from each of the 19 source chronograms from 13 studies as inputs, treating their node ages as unknown.
-We then estimated dates for these topologies using the node ages from the chronograms from the other studies as calibrations and smoothing using BLADJ.
-We found that node ages from original study, and ages estimated using all other age data available are correlated (Fig. \ref{fig:cvXY}).
-For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka \enquote{closer to the root}) relative to the original estimate, and overestimate ages for nodes closer to the tips.
-Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. S1).
-
-\begin{center}
-\textsc{Discussion}
-\end{center}
-
-DateLife makes state-of-the-art data on evolutionary time frame easily accessible for comparison, reuse, and reanalysis, to researchers in all areas of science and with all levels of expertise in the matter.
-It is an open service that does not require any expert biological knowledge from users --besides the names of the species or group they want to work with, for any of its functionality.
-
-A total of 99,474 unique terminal taxa are represented in DateLife's database.
-Incorporation of more chronograms into the database will continue to improve DateLife's services.
-One option to increase the number of chronograms in the DateLife database is the Dryad data repository.
-Methods to automatically mine chronograms from Dryad could be designed and implemented.
-However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined manually by a human curator.
-We would like to emphasize on the importance of sharing chronogram data, including systematically curated metadata, into open repositories, such as OpenTree's Phylesystem (McTavish et al., 2015) for the benefit of the scientific community as a whole.
-
-\begin{center}
-\emph{Age Variation in Source Chronograms}
-\end{center}
-
-Conflict in estimated ages among alternative studies is common in the literature.
-See, for example, the robust ongoing debate about crown group age of angiosperms (Barba-Montoya, Reis, Schneider, Donoghue, \& Yang, 2018; Magallón, Gómez-Acevedo, Sánchez-Reyes, \& Hernández-Hernández, 2015; Ramshaw et al., 1972; Sanderson \& Doyle, 2001; Sauquet, Ramírez-Barahona, \& Magallón, 2021).
-Source chronograms available for the same organisms have potentially been estimated implementing calibrations very differently.
-For example, the chronograms from Burns et al.
-(2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older age estimates for the same nodes than chronograms that were inferred using fossils as calibrations (Figs.
-\ref{fig:fringillidages}, \ref{fig:cvXY}; Supplementary Figs.
-S1, S5).
-
-Different calibration implementations might also imply fundamentally distinct evolutionary hypotheses (Antonelli et al., 2017).
-For example, two independent researchers working on the same clade should both carefully select and justify their choices of fossil calibration placement.
-Yet, if one researcher concludes that a fossil should calibrate the ingroup of a clade, while another researcher concludes that the same fossil should calibrate the outgroup of the clade, the resulting age estimates will differ, as the placement of calibrations as stem or crown group has been proven to significantly affect time of lineage divergence estimates (Sauquet, 2013).
-
-\begin{center}
-\emph{Primary vs Secondary Calibrations}
-\end{center}
-
-While most chronograms in DateLife's database are constructed using primary calibrations (molecular substitution rates or ages obtained from the fossil record or geological events), DateLife summarizes chronograms using secondary calibrations (ages coming from other chronograms).
-Graur and Martin (2004) cautioned on the increased error and uncertainty in estimated ages when using secondary calibrations in dating analyses.
-Schenk (2016) showed that, in simulations, divergence times inferred using secondary calibrations are significantly younger than those inferred with primary calibrations, when obtained with Bayesian inference methods, and when priors are implemented in similar ways in both analyses.
-Accordingly, the scientific community seems to have more confidence in chronograms obtained from a single analysis, using fossil data as primary sources of calibrations (Schenk, 2016), and using fossils that have been widely discussed and curated as calibrations to date other trees, making sure that all data reflect a coherent evolutionary history (Sauquet, 2013), as for example done by Antonelli et al.
-(2017).
-There have been attempts to create fossil calibration databases (Ksepka et al., 2015), though these still have room to grow.
-
-It seems that using several (as opposed to just a few) secondary calibrations can provide sufficient information to alleviate or even neutralize potential biases (Sauquet, 2013).
-Certainly, further studies are required to fully understand the effect of secondary calibrations on outputs from different tree dating methods, and on downstream analyses.
-It is possible that secondary calibrations can be safely used with dating methods that do not require setting priors, such as penalized likelihood (Sanderson, 2003), with methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ (Webb et al., 2008; Webb \& Donoghue, 2005), or methods that summarize age data unto a tree topology.
-
-Our cross validation analysis might provide some insight in this regard.
-When ages are estimated with secondary calibrations, nodes closer to the root do tend to be slightly younger than ages estimated with primary calibrations.
-However, nodes closer to the tip tend to be older when estimated using secondary calibrations with a dating method that does not make any prior assumptions on the nature of the calibrations themselves (Supplementary Figures S2-S20).
-The only excpetion to tjis was observed on results of the cross validation analysis of the Burns et al.
-(2014) chronogram, which results in much younger node ages when estimated using secondary calibrations (Supplementary Figs.
-S1, S5).
-
-\begin{center}
-\emph{Sumarizing Chronograms}
-\end{center}
-
-By default, DateLife currently summarizes all source chronograms that overlap with at least two species names. Users can exclude source chronograms if they have reasons to do so.
-Strictly speaking, a good chronogram should reflect the real time of lineage divergence accurately and precisely.
-To our knowledge, there are no tested measures to determine independently when a chronogram is better than another.
-Yet, several characteristics of the data used for dating analyses, as well as from the output chronogram itself, could be used to score the quality of source chronograms.
-
-Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used (Magallón, 2010; Magallón et al., 2015).
-Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
-
-DateLife provides an opportunity to capture concordance and conflict among date estimates, which can also be used as a metric for chronogram reliability.
-Its open database of chronograms allows other researchers to do such analyses themselves reproducibly, and without needing permission.
-Though, of course, they should follow proper citation practices, especially for the source chronogram studies.
-
-The exercise of summarizing age data from across multiple studies provides the opportunity to work with a more inclusive chronogram,
-that reflects a unified evolutionary history for a lineage, by putting together evidence from different hypotheses.
-The largest, and taxonomically broadest chronogram currently available from OpenTree was constructed summarizing age data from 2,274 published chronograms using NCBI's taxonomic tree as backbone (Hedges et al., 2015).
-A summarizing exercise may also amplify the effect of uncertainty and errors in source data, and blur parts of the evolutionary history of a lineage that might only be reflected in source chronograms and lost on the summary chronogram (Sauquet et al., 2021).
-
-\begin{center}
-\emph{Effects on Downstream Analyses}
-\end{center}
-
-For downstream analyses, using alternative chronogram may deeply affect our inferences (Title \& Rabosky, 2016), particularly when studying phenomena dependent on the timing of species diversification events, such as macroevolutionary processes.
-
-In ecology and conservation biology, incorporating at least some data on lineage divergence times represents a relevant improvement for testing alternative hypothesis using phylogenetic distance (Webb et al., 2008).
-Hence, DateLife's workflow features different ways of estimating node ages in the absence of calibrations and branch length information for certain taxa.
-\enquote{Making up} branch lengths is a common practice in scientific publications: Jetz et al.
-(2012), created a chronogram of all 9, 993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
-(2018) created a chronogram of 31, 536 ray-finned fishes, of which only 37\% had molecular data; Smith and Brown (2018) constructed a chronogram of 353, 185 seed plants where only 23\% had molecular data.
-
-Notably, risks come with this practice.
-Taken to the extreme, one could make a fully resolved, calibrated tree of all modern and extinct taxa using a single taxonomy and a single calibration, using polytomy resolution and branch estimation methods.
-There has yet to be a thorough analysis of what can go wrong when one extends inferences beyond the data in this way, so we urge caution; we also urge readers to follow the example of the large tree papers cited above, by carefully considering the statistical assumptions being made, and assessing the consistency of the results with prior work.
-
-\begin{center}
-\textsc{Conclusions}
-\end{center}
-
-Knowledge of the evolutionary time frame of organisms is key to many research areas: trait evolution,
-species diversification, biogeography, macroecology and more. It is also crucial for education, science communication and policy, but generating chronograms is difficult,
-especially for those who want to use phylogenies but who are not systematists, or
-do not have the time to acquire and develop the necessary knowledge and skills to construct them on their own.
-Importantly, years of primarily public funded research have resulted in vast amounts of chronograms that are already available on scientific publications, but hidden to the public and scientific community for reuse.
-
-The DateLife project allows for easy and fast summary of public and state-of-the-art data on time of lineage divergence.
-It provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identification of regions that require more research, or that have conflicting information.
-It is available as an R package, and as a web-based R shiny application at www.datelife.org
-Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
-The DateLife project helps with awareness of the existing variation in expert time of divergence data, and will foster exploration of the effect of alternative divergence time hypothesis on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
-
-\begin{center}
-\textsc{Availability}
-\end{center}
-
-The DateLife software is free and open source and it can be used through its R shiny web application at \url{http://www.datelife.org}, through the \texttt{datelife} R package, and through Phylotastic's project web portal \url{https://phylo.cs.nmsu.edu/}.
-DateLife's web application is maintained using RStudio's shiny server and the shiny package open infrastructure, as well as Docker and OpenTree's infrastructure (dates.opentreeoflife.org/datelife).
-\texttt{datelife}'s R package stable version is available for installation from the CRAN repository (\url{https://cran.r-project.org/package=datelife}) using the command \texttt{install.packages(pkgs\ =\ "datelife")} from within R.
-Development versions are available from the GitHub repository (\url{https://github.com/phylotastic/datelife}) and can be installed using the command \texttt{devtools::install\_github("phylotastic/datelife")}.
-
-\begin{center}
-\textsc{Supplementary Material}
-\end{center}
-
-Code used to generate all versions of this manuscript, the biological examples, as well as the benchmark of functionalities are available at \href{https://github.com/LunaSare/datelifeMS1}{datelifeMS1}, \href{https://github.com/LunaSare/datelife_examples}{datelife\_examples}, and \href{https://github.com/LunaSare/datelife_benchmark}{datelife\_benchmark} repositories in LLSR's GitHub account.
-
-\begin{center}
-\textsc{Funding}
-\end{center}
-
-Funding was provided by the US National Science Foundation (NSF) grants ABI-1458603 to the Datelife project; DBI-0905606 to the National Evolutionary Synthesis Center (NESCent), ABI-1458572 to the Phylotastic project, and ABI-1759846 to the Open Tree of Life project.
-
-\begin{center}
-\textsc{Acknowledgements}
-\end{center}
-
-The DateLife project was born as a prototype tool aiming to provide these services, and was initially developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA (Stoltzfus et al., 2013).
-We thank colleagues from the O'Meara Lab at the University
-of Tennesse Knoxville for suggestions, discussions and software testing.
-The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
-that led to initial work on this project.
-The team that assembled DateLife's first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
-The Open Tree of Life project that provides the open, metadata rich repository of
-trees used to construct DateLife's chronogram database.
-The many scientists who publish their chronograms in an open, reusable form, and the scientists who curate them for deposition in the Open Tree of Life repository.
-The NSF for funding nearly all the above, in addition to the ABI grant that funded this project itself.
-
-\newpage
-
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\begingroup
-\setlength{\parindent}{-0.5in}
-\setlength{\leftskip}{0.5in}
-
-\hypertarget{refs}{}
-\leavevmode\hypertarget{ref-alstrom2014discovery}{}%
-Alström, P., Hooper, D.
-M., Liu, Y., Olsson, U., Mohan, D., Gelang, M., \ldots{} Price, T.
-D.
-(2014).
-Discovery of a relict lineage and monotypic family of passerine birds.
-\emph{Biology Letters}, \emph{10}(3), 20131067.
-
-\leavevmode\hypertarget{ref-ane2009groves}{}%
-Ané, C., Eulenstein, O., Piaggio-Talice, R., \& Sanderson, M. J. (2009). Groves of phylogenetic trees. \emph{Annals of Combinatorics}, \emph{13}(2), 139--167.
-
-\leavevmode\hypertarget{ref-antonelli2017supersmart}{}%
-Antonelli, A., Hettling, H., Condamine, F.
-L., Vos, K., Nilsson, R.
-H., Sanderson, M.
-J., \ldots{} Vos, R.
-A.
-(2017).
-Toward a self-updating platform for estimating rates of speciation and migration, ages, and relationships of Taxa.
-\emph{Systematic Biology}, \emph{66}(2), 153--166.
-\url{https://doi.org/10.1093/sysbio/syw066}
-
-\leavevmode\hypertarget{ref-archie1986newick}{}%
-Archie, J., Day, W.
-H., Felsenstein, J., Maddison, W., Meacham, C., Rohlf, F.
-J., \& Swofford, D.
-(1986).
-The Newick tree format.
-Retrieved from \href{\%7Bhttps://evolution.genetics.washington.edu/phylip/newicktree.html\%7D}{\{https://evolution.genetics.washington.edu/phylip/newicktree.html\}}
-
-\leavevmode\hypertarget{ref-avibase-emberiza}{}%
-Avibase.
-(2022).
-Yellow-throated Bunting.
-\emph{Avibase - the World Bird Database}, (Online Resource).
-Retrieved from \href{\%7Bhttps://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\%7D}{\{https://avibase.bsc-eoc.org/species.jsp?lang=EN\&avibaseid=82D1EE0049D8D927\}}
-
-\leavevmode\hypertarget{ref-Bapst2012a}{}%
-Bapst, D.
-W.
-(2012).
-Paleotree: An R package for paleontological and phylogenetic analyses of evolution.
-\emph{Methods in Ecology and Evolution}, \emph{3}(5), 803--807.
-\url{https://doi.org/10.1111/j.2041-210X.2012.00223.x}
-
-\leavevmode\hypertarget{ref-barba2018constraining}{}%
-Barba-Montoya, J., Reis, M.
-dos, Schneider, H., Donoghue, P.
-C., \& Yang, Z.
-(2018).
-Constraining uncertainty in the timescale of angiosperm evolution and the veracity of a cretaceous terrestrial revolution.
-\emph{New Phytologist}, \emph{218}(2), 819--834.
-
-\leavevmode\hypertarget{ref-barker2014mitogenomic}{}%
-Barker, F. K. (2014). Mitogenomic data resolve basal relationships among passeriform and passeridan birds. \emph{Molecular Phylogenetics and Evolution}, \emph{79}, 313--324.
-
-\leavevmode\hypertarget{ref-barker2013going}{}%
-Barker, F.
-K., Burns, K.
-J., Klicka, J., Lanyon, S.
-M., \& Lovette, I.
-J.
-(2013).
-Going to extremes: Contrasting rates of diversification in a recent radiation of new world passerine birds.
-\emph{Systematic Biology}, \emph{62}(2), 298--320.
-
-\leavevmode\hypertarget{ref-barker2015new}{}%
-Barker, F.
-K., Burns, K.
-J., Klicka, J., Lanyon, S.
-M., \& Lovette, I.
-J.
-(2015).
-New insights into new world biogeography: An integrated view from the phylogeny of blackbirds, cardinals, sparrows, tanagers, warblers, and allies.
-\emph{The Auk: Ornithological Advances}, \emph{132}(2), 333--348.
-
-\leavevmode\hypertarget{ref-barker2004phylogeny}{}%
-Barker, F.
-K., Cibois, A., Schikler, P., Feinstein, J., \& Cracraft, J.
-(2004).
-Phylogeny and diversification of the largest avian radiation.
-\emph{Proceedings of the National Academy of Sciences}, \emph{101}(30), 11040--11045.
-
-\leavevmode\hypertarget{ref-beresford2005african}{}%
-Beresford, P., Barker, F., Ryan, P., \& Crowe, T.
-(2005).
-African endemics span the tree of songbirds (passeri): Molecular systematics of several evolutionary ``enigmas''.
-\emph{Proceedings of the Royal Society B: Biological Sciences}, \emph{272}(1565), 849--858.
-
-\leavevmode\hypertarget{ref-Boyle2013}{}%
-Boyle, B., Hopkins, N., Lu, Z., Raygoza Garay, J.
-A., Mozzherin, D., Rees, T., \ldots{} Enquist, B.
-J.
-(2013).
-The taxonomic name resolution service: An online tool for automated standardization of plant names.
-\emph{BMC Bioinformatics}, \emph{14}(1).
-\url{https://doi.org/10.1186/1471-2105-14-16}
-
-\leavevmode\hypertarget{ref-Britton2007}{}%
-Britton, T., Anderson, C.
-L., Jacquet, D., Lundqvist, S., \& Bremer, K.
-(2007).
-Estimating Divergence Times in Large Phylogenetic Trees.
-\emph{Systematic Biology}, \emph{56}(788777878), 741--752.
-\url{https://doi.org/10.1080/10635150701613783}
-
-\leavevmode\hypertarget{ref-bryson2014diversification}{}%
-Bryson Jr, R.
-W., Chaves, J., Smith, B.
-T., Miller, M.
-J., Winker, K., Pérez-Emán, J.
-L., \& Klicka, J.
-(2014).
-Diversification across the new world within the `blue'cardinalids (aves: Cardinalidae).
-\emph{Journal of Biogeography}, \emph{41}(3), 587--599.
-
-\leavevmode\hypertarget{ref-burleigh2015building}{}%
-Burleigh, J. G., Kimball, R. T., \& Braun, E. L. (2015). Building the avian tree of life using a large-scale, sparse supermatrix. \emph{Molecular Phylogenetics and Evolution}, \emph{84}, 53--63.
-
-\leavevmode\hypertarget{ref-burns2014phylogenetics}{}%
-Burns, K.
-J., Shultz, A.
-J., Title, P.
-O., Mason, N.
-A., Barker, F.
-K., Klicka, J., \ldots{} Lovette, I.
-J.
-(2014).
-Phylogenetics and diversification of tanagers (passeriformes: Thraupidae), the largest radiation of neotropical songbirds.
-\emph{Molecular Phylogenetics and Evolution}, \emph{75}, 41--77.
-
-\leavevmode\hypertarget{ref-Chamberlain2018}{}%
-Chamberlain, S. (2018). \emph{bold: Interface to Bold Systems API}. Retrieved from \url{https://CRAN.R-project.org/package=bold}
-
-\leavevmode\hypertarget{ref-Chamberlain2013}{}%
-Chamberlain, S.
-A., \& Szöcs, E.
-(2013).
-taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}.
-\emph{F1000Research}, \emph{2}(191), 1--29.
-\url{https://doi.org/10.12688/f1000research.2-191.v2}
-
-\leavevmode\hypertarget{ref-chaves2013biogeography}{}%
-Chaves, J.
-A., Hidalgo, J.
-R., \& Klicka, J.
-(2013).
-Biogeography and evolutionary history of the n eotropical genus s altator (a ves: T hraupini).
-\emph{Journal of Biogeography}, \emph{40}(11), 2180--2190.
-
-\leavevmode\hypertarget{ref-claramunt2015new}{}%
-Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{Science Advances}, \emph{1}(11), e1501005.
-
-\leavevmode\hypertarget{ref-Criscuolo2006}{}%
-Criscuolo, A., Berry, V., Douzery, E.
-J., \& Gascuel, O.
-(2006).
-SDM: A fast distance-based approach for (super)tree building in phylogenomics.
-\emph{Systematic Biology}, \emph{55}(5), 740--755.
-\url{https://doi.org/10.1080/10635150600969872}
-
-\leavevmode\hypertarget{ref-delsuc2018phylogenomic}{}%
-Delsuc, F., Philippe, H., Tsagkogeorga, G., Simion, P., Tilak, M.-K., Turon, X., \ldots{} Douzery, E.
-J.
-(2018).
-A phylogenomic framework and timescale for comparative studies of tunicates.
-\emph{BMC Biology}, \emph{16}(1), 1--14.
-
-\leavevmode\hypertarget{ref-Eastman2013}{}%
-Eastman, J.
-M., Harmon, L.
-J., \& Tank, D.
-C.
-(2013).
-Congruification: Support for time scaling large phylogenetic trees.
-\emph{Methods in Ecology and Evolution}, \emph{4}(7), 688--691.
-\url{https://doi.org/10.1111/2041-210X.12051}
-
-\leavevmode\hypertarget{ref-edgar2004muscle}{}%
-Edgar, R. C. (2004). MUSCLE: Multiple sequence alignment with high accuracy and high throughput. \emph{Nucleic Acids Research}, \emph{32}(5), 1792--1797.
-
-\leavevmode\hypertarget{ref-Felsenstein1985a}{}%
-Felsenstein, J. (1985). Phylogenies and the Comparative Method. \emph{The American Naturalist}, \emph{125}(1), 1--15. Retrieved from \url{http://www.jstor.org/stable/2461605}
-
-\leavevmode\hypertarget{ref-gbif2022taxonomy}{}%
-GBIF Secretariat.
-(2022).
-GBIF Backbone Taxonomy.
-\emph{Checklist dataset}, (Online Resource accessed via GBIF.org).
-Retrieved from \href{\%7Bhttps://doi.org/10.15468/39omei\%20\%7D}{\{https://doi.org/10.15468/39omei \}}
-
-\leavevmode\hypertarget{ref-gibb2015new}{}%
-Gibb, G.
-C., England, R., Hartig, G., McLenachan, P.
-A., Taylor Smith, B.
-L., McComish, B.
-J., \ldots{} Penny, D.
-(2015).
-New zealand passerines help clarify the diversification of major songbird lineages during the oligocene.
-\emph{Genome Biology and Evolution}, \emph{7}(11), 2983--2995.
-
-\leavevmode\hypertarget{ref-graur2004reading}{}%
-Graur, D., \& Martin, W. (2004). Reading the entrails of chickens: Molecular timescales of evolution and the illusion of precision. \emph{TRENDS in Genetics}, \emph{20}(2), 80--86.
-
-\leavevmode\hypertarget{ref-hackett2008phylogenomic}{}%
-Hackett, S.
-J., Kimball, R.
-T., Reddy, S., Bowie, R.
-C., Braun, E.
-L., Braun, M.
-J., \ldots{} others.
-(2008).
-A phylogenomic study of birds reveals their evolutionary history.
-\emph{Science}, \emph{320}(5884), 1763--1768.
-
-\leavevmode\hypertarget{ref-Hedges2015}{}%
-Hedges, S.
-B., Marin, J., Suleski, M., Paymer, M., \& Kumar, S.
-(2015).
-Tree of life reveals clock-like speciation and diversification.
-\emph{Molecular Biology and Evolution}, \emph{32}(4), 835--845.
-\url{https://doi.org/10.1093/molbev/msv037}
-
-\leavevmode\hypertarget{ref-Heibl2008}{}%
-Heibl, C. (2008). \emph{PHYLOCH: R language tree plotting tools and interfaces to diverse phylogenetic software packages.} Retrieved from \url{http://www.christophheibl.de/Rpackages.html}
-
-\leavevmode\hypertarget{ref-hooper2017chromosomal}{}%
-Hooper, D. M., \& Price, T. D. (2017). Chromosomal inversion differences correlate with range overlap in passerine birds. \emph{Nature Ecology \& Evolution}, \emph{1}(10), 1526.
-
-\leavevmode\hypertarget{ref-Huelsenbeck2001}{}%
-Huelsenbeck, J. P., \& Ronquist, F. (2001). MRBAYES: Bayesian inference of phylogenetic trees. \emph{Bioinformatics}, \emph{17}(8), 754--755. \url{https://doi.org/10.1093/bioinformatics/17.8.754}
-
-\leavevmode\hypertarget{ref-Jetz2012}{}%
-Jetz, W., Thomas, G., Joy, J.
-J., Hartmann, K., \& Mooers, A.
-(2012).
-The global diversity of birds in space and time.
-\emph{Nature}, \emph{491}(7424), 444--448.
-\url{https://doi.org/10.1038/nature11631}
-
-\leavevmode\hypertarget{ref-johansson2008phylogenetic}{}%
-Johansson, U.
-S., Fjeldså, J., \& Bowie, R.
-C.
-(2008).
-Phylogenetic relationships within passerida (aves: Passeriformes): A review and a new molecular phylogeny based on three nuclear intron markers.
-\emph{Molecular Phylogenetics and Evolution}, \emph{48}(3), 858--876.
-
-\leavevmode\hypertarget{ref-katoh2009multiple}{}%
-Katoh, K., Asimenos, G., \& Toh, H. (2009). Multiple alignment of dna sequences with mafft. In \emph{Bioinformatics for dna sequence analysis} (pp. 39--64). Springer.
-
-\leavevmode\hypertarget{ref-kimball2019phylogenomic}{}%
-Kimball, R. T., Oliveros, C. H., Wang, N., White, N. D., Barker, F. K., Field, D. J., \ldots{} others. (2019). A phylogenomic supertree of birds. \emph{Diversity}, \emph{11}(7), 109.
-
-\leavevmode\hypertarget{ref-klicka2014comprehensive}{}%
-Klicka, J., Barker, F.
-K., Burns, K.
-J., Lanyon, S.
-M., Lovette, I.
-J., Chaves, J.
-A., \& Bryson Jr, R.
-W.
-(2014).
-A comprehensive multilocus assessment of sparrow (aves: Passerellidae) relationships.
-\emph{Molecular Phylogenetics and Evolution}, \emph{77}, 177--182.
-
-\leavevmode\hypertarget{ref-ksepka2015fossil}{}%
-Ksepka, D.
-T., Parham, J.
-F., Allman, J.
-F., Benton, M.
-J., Carrano, M.
-T., Cranston, K.
-A., \ldots{} others.
-(2015).
-The fossil calibration database---a new resource for divergence dating.
-\emph{Systematic Biology}, \emph{64}(5), 853--859.
-
-\leavevmode\hypertarget{ref-lamichhaney2015evolution}{}%
-Lamichhaney, S., Berglund, J., Almén, M.
-S., Maqbool, K., Grabherr, M., Martinez-Barrio, A., \ldots{} others.
-(2015).
-Evolution of darwin's finches and their beaks revealed by genome sequencing.
-\emph{Nature}, \emph{518}(7539), 371--375.
-
-\leavevmode\hypertarget{ref-laubichler2009form}{}%
-Laubichler, M. D., \& Maienschein, J. (2009). \emph{Form and function in developmental evolution}. Cambridge University Press.
-
-\leavevmode\hypertarget{ref-lepage2004avibase}{}%
-Lepage, D. (2004). \emph{Avibase: The world bird database}. Bird Studies Canada.
-
-\leavevmode\hypertarget{ref-lepage2014avibase}{}%
-Lepage, D., Vaidya, G., \& Guralnick, R. (2014). Avibase--a database system for managing and organizing taxonomic concepts. \emph{ZooKeys}, (420), 117.
-
-\leavevmode\hypertarget{ref-lerner2011multilocus}{}%
-Lerner, H.
-R., Meyer, M., James, H.
-F., Hofreiter, M., \& Fleischer, R.
-C.
-(2011).
-Multilocus resolution of phylogeny and timescale in the extant adaptive radiation of hawaiian honeycreepers.
-\emph{Current Biology}, \emph{21}(21), 1838--1844.
-
-\leavevmode\hypertarget{ref-lovette2010comprehensive}{}%
-Lovette, I.
-J., Pérez-Emán, J.
-L., Sullivan, J.
-P., Banks, R.
-C., Fiorentino, I., Córdoba-Córdoba, S., \ldots{} others.
-(2010).
-A comprehensive multilocus phylogeny for the wood-warblers and a revised classification of the parulidae (aves).
-\emph{Molecular Phylogenetics and Evolution}, \emph{57}(2), 753--770.
-
-\leavevmode\hypertarget{ref-magallon2001absolute}{}%
-Magallon, S., \& Sanderson, M. (2001). Absolute diversification rates in angiosperm clades. \emph{Evolution}, \emph{55}(9), 1762--1780.
-
-\leavevmode\hypertarget{ref-magallon2010using}{}%
-Magallón, S. (2010). Using fossils to break long branches in molecular dating: A comparison of relaxed clocks applied to the origin of angiosperms. \emph{Systematic Biology}, \emph{59}(4), 384--399.
-
-\leavevmode\hypertarget{ref-magallon2015metacalibrated}{}%
-Magallón, S., Gómez-Acevedo, S., Sánchez-Reyes, L.
-L., \& Hernández-Hernández, T.
-(2015).
-A metacalibrated time-tree documents the early rise of flowering plant phylogenetic diversity.
-\emph{New Phytologist}, \emph{207}(2), 437--453.
-
-\leavevmode\hypertarget{ref-mctavish2015phylesystem}{}%
-McTavish, E.
-J., Hinchliff, C.
-E., Allman, J.
-F., Brown, J.
-W., Cranston, K.
-A., Holder, M.
-T., \ldots{} Smith, S.
-(2015).
-Phylesystem: A git-based data store for community-curated phylogenetic estimates.
-\emph{Bioinformatics}, \emph{31}(17), 2794--2800.
-
-\leavevmode\hypertarget{ref-Michonneau2016}{}%
-Michonneau, F., Brown, J.
-W., \& Winter, D.
-J.
-(2016).
-rotl: an R package to interact with the Open Tree of Life data.
-\emph{Methods in Ecology and Evolution}, \emph{7}(12), 1476--1481.
-\url{https://doi.org/10.1111/2041-210X.12593}
-
-\leavevmode\hypertarget{ref-Morlon2014}{}%
-Morlon, H. (2014). Phylogenetic approaches for studying diversification. \emph{Ecology Letters}, \emph{17}(4), 508--525. \url{https://doi.org/10.1111/ele.12251}
-
-\leavevmode\hypertarget{ref-moyle2016tectonic}{}%
-Moyle, R.
-G., Oliveros, C.
-H., Andersen, M.
-J., Hosner, P.
-A., Benz, B.
-W., Manthey, J.
-D., \ldots{} Faircloth, B.
-C.
-(2016).
-Tectonic collision and uplift of wallacea triggered the global songbird radiation.
-\emph{Nature Communications}, \emph{7}(1), 1--7.
-
-\leavevmode\hypertarget{ref-oliveros2019earth}{}%
-Oliveros, C.
-H., Field, D.
-J., Ksepka, D.
-T., Barker, F.
-K., Aleixo, A., Andersen, M.
-J., \ldots{} others.
-(2019).
-Earth history and the passerine superradiation.
-\emph{Proceedings of the National Academy of Sciences}, \emph{116}(16), 7916--7925.
-
-\leavevmode\hypertarget{ref-Ooms2018}{}%
-Ooms, J., \& Chamberlain, S. (2018). \emph{Phylocomr: Interface to 'phylocom'}. Retrieved from \url{https://CRAN.R-project.org/package=phylocomr}
-
-\leavevmode\hypertarget{ref-opentreeAPIs}{}%
-Open Tree Of Life, Redelings, B., Cranston, K.
-A., Allman, J., Holder, M.
-T., \& McTavish, E.
-J.
-(2016).
-Open Tree of Life APIs v3.0.
-\emph{Open Tree of Life Project}, (Online Resources).
-Retrieved from \href{\%7Bhttps://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\%7D}{\{https://github.com/OpenTreeOfLife/germinator/wiki/Open-Tree-of-Life-Web-APIs\}}
-
-\leavevmode\hypertarget{ref-opentreeoflife2019synth}{}%
-Open Tree Of Life, Redelings, B., Sánchez Reyes, L.
-L., Cranston, K.
-A., Allman, J., Holder, M.
-T., \& McTavish, E.
-J.
-(2019).
-Open tree of life synthetic tree v12.3.
-\emph{Zenodo}.
-Retrieved from \url{https://doi.org/10.5281/zenodo.3937742}
-
-\leavevmode\hypertarget{ref-odeen2011evolution}{}%
-Ödeen, A., Håstad, O., \& Alström, P. (2011). Evolution of ultraviolet vision in the largest avian radiation-the passerines. \emph{BMC Evolutionary Biology}, \emph{11}(1), 1--8.
-
-\leavevmode\hypertarget{ref-paradis2004}{}%
-Paradis, E., Claude, J., \& Strimmer, K. (2004). APE: analyses of phylogenetics and evolution in R language. \emph{Bioinformatics}, \emph{20}(2), 289--290.
-
-\leavevmode\hypertarget{ref-parchman2007coevolution}{}%
-Parchman, T.
-L., Benkman, C.
-W., \& Mezquida, E.
-T.
-(2007).
-Coevolution between hispaniolan crossbills and pine: Does more time allow for greater phenotypic escalation at lower latitude?
-\emph{Evolution}, \emph{61}(9), 2142--2153.
-
-\leavevmode\hypertarget{ref-packert2012horizontal}{}%
-Päckert, M., Martens, J., Sun, Y.-H., Severinghaus, L.
-L., Nazarenko, A.
-A., Ting, J., \ldots{} Tietze, D.
-T.
-(2012).
-Horizontal and elevational phylogeographic patterns of himalayan and southeast asian forest passerines (aves: Passeriformes).
-\emph{Journal of Biogeography}, \emph{39}(3), 556--573.
-
-\leavevmode\hypertarget{ref-pennell2014geiger}{}%
-Pennell, M.
-W., Eastman, J.
-M., Slater, G.
-J., Brown, J.
-W., Uyeda, J.
-C., FitzJohn, R.
-G., \ldots{} Harmon, L.
-J.
-(2014).
-Geiger v2.
-0: An expanded suite of methods for fitting macroevolutionary models to phylogenetic trees.
-\emph{Bioinformatics}, \emph{30}(15), 2216--2218.
-
-\leavevmode\hypertarget{ref-posadas2006historical}{}%
-Posadas, P., Crisci, J. V., \& Katinas, L. (2006). Historical biogeography: A review of its basic concepts and critical issues. \emph{Journal of Arid Environments}, \emph{66}(3), 389--403.
-
-\leavevmode\hypertarget{ref-powell2014comprehensive}{}%
-Powell, A.
-F., Barker, F.
-K., Lanyon, S.
-M., Burns, K.
-J., Klicka, J., \& Lovette, I.
-J.
-(2014).
-A comprehensive species-level molecular phylogeny of the new world blackbirds (icteridae).
-\emph{Molecular Phylogenetics and Evolution}, \emph{71}, 94--112.
-
-\leavevmode\hypertarget{ref-price2014niche}{}%
-Price, T.
-D., Hooper, D.
-M., Buchanan, C.
-D., Johansson, U.
-S., Tietze, D.
-T., Alström, P., \ldots{} others.
-(2014).
-Niche filling slows the diversification of himalayan songbirds.
-\emph{Nature}, \emph{509}(7499), 222.
-
-\leavevmode\hypertarget{ref-pulgarin2013multilocus}{}%
-Pulgarín-R, P.
-C., Smith, B.
-T., Bryson Jr, R.
-W., Spellman, G.
-M., \& Klicka, J.
-(2013).
-Multilocus phylogeny and biogeography of the new world pheucticus grosbeaks (aves: Cardinalidae).
-\emph{Molecular Phylogenetics and Evolution}, \emph{69}(3), 1222--1227.
-
-\leavevmode\hypertarget{ref-rabosky2018inverse}{}%
-Rabosky, D.
-L., Chang, J., Title, P.
-O., Cowman, P.
-F., Sallan, L., Friedman, M., \ldots{} others.
-(2018).
-An inverse latitudinal gradient in speciation rate for marine fishes.
-\emph{Nature}, \emph{559}(7714), 392.
-
-\leavevmode\hypertarget{ref-ramshaw1972time}{}%
-Ramshaw, J., Richardson, D., Meatyard, B., Brown, R., Richardson, M., Thompson, E., \& Boulter, D.
-(1972).
-The time of origin of the flowering plants determined by using amino acid sequence data of cytochrome c.
-\emph{New Phytologist}, \emph{71}(5), 773--779.
-
-\leavevmode\hypertarget{ref-ratnasingham2007bold}{}%
-Ratnasingham, S., \& Hebert, P. D. (2007). BOLD: The barcode of life data system (http://www. Barcodinglife. Org). \emph{Molecular Ecology Notes}, \emph{7}(3), 355--364.
-
-\leavevmode\hypertarget{ref-RCoreTeam2018}{}%
-R Core Team. (2018). \emph{R: a language and environment for statistical computing}. Vienna, Austria: R Foundation for Statistical Computing.
-
-\leavevmode\hypertarget{ref-rees2017automated}{}%
-Rees, \& Cranston, K. (2017). Automated assembly of a reference taxonomy for phylogenetic data synthesis. \emph{Biodiversity Data Journal}, (5).
-
-\leavevmode\hypertarget{ref-rees2017irmng}{}%
-Rees, Vandepitte, L., Decock, W., \& Vanhoorne, B. (2017). IRMNG 2006--2016: 10 Years of a Global Taxonomic Database. \emph{Biodiversity Informatics}, \emph{12}.
-
-\leavevmode\hypertarget{ref-Revell2012}{}%
-Revell, L. J. (2012). Phytools: An r package for phylogenetic comparative biology (and other things). \emph{Methods in Ecology and Evolution}, \emph{3}, 217--223.
-
-\leavevmode\hypertarget{ref-Ronquist2003}{}%
-Ronquist, F., \& Huelsenbeck, J.
-P.
-(2003).
-MrBayes 3: Bayesian phylogenetic inference under mixed models.
-\emph{Bioinformatics}, \emph{19}(12), 1572--1574.
-\url{https://doi.org/10.1093/bioinformatics/btg180}
-
-\leavevmode\hypertarget{ref-datelifeplot}{}%
-Sanchez-Reyes, L.
-L., \& O'Meara, B.
-(2022).
-\texttt{datelifeplot}: Methods to plot chronograms and outputs of the \texttt{datelife} package.
-\emph{R Package Release V0.2.2}.
-Retrieved from \url{https://zenodo.org/badge/latestdoi/381501451}
-
-\leavevmode\hypertarget{ref-sanchez2022}{}%
-Sanchez-Reyes, L.
-L., O'Meara, B., Eastman, J., Heath, T., Wright, A., Schliep, K., \ldots{} Alfaro, M.
-(2022).
-\texttt{datelife}: Scientific Data on Time of Lineage Divergence for Your Taxa.
-In \emph{R package version 0.6.6}.
-Retrieved from \href{https://CRAN.R-project.org/package=datelife\%20and\%20https://doi.org/10.5281/zenodo.593938}{https://CRAN.R-project.org/package=datelife and https://doi.org/10.5281/zenodo.593938}
-
-\leavevmode\hypertarget{ref-sanderson2002}{}%
-Sanderson, M.
-(2002).
-Estimating Absolute Rates of Molecular Evolution and Divergence Times: A Penalized Likelihood Approach.
-\emph{Molecular Biology and Evolution}, \emph{19}(1), 101--109.
-\url{https://doi.org/10.1093/oxfordjournals.molbev.a003974}
-
-\leavevmode\hypertarget{ref-sanderson2003r8s}{}%
-Sanderson, M. (2003). r8s: Inferring Absolute Rates of Molecular Evolution and Divergence Times in the Absence of a Molecular Clock. \emph{Bioinformatics}, \emph{19}(2), 301--302.
-
-\leavevmode\hypertarget{ref-sanderson2001sources}{}%
-Sanderson, M., \& Doyle, J.
-(2001).
-Sources of error and confidence intervals in estimating the age of angiosperms from rbcL and 18S rDNA data.
-\emph{American Journal of Botany}, \emph{88}(8), 1499--1516.
-
-\leavevmode\hypertarget{ref-sauquet2013practical}{}%
-Sauquet, H. (2013). A practical guide to molecular dating. \emph{Comptes Rendus Palevol}, \emph{12}(6), 355--367.
-
-\leavevmode\hypertarget{ref-sauquet2021age}{}%
-Sauquet, H., Ramírez-Barahona, S., \& Magallón, S. (2021). \emph{The age of flowering plants is unknown}.
-
-\leavevmode\hypertarget{ref-schenk2016sec}{}%
-Schenk, J. J. (2016). Consequences of secondary calibrations on divergence time estimates. \emph{PLoS ONE}, \emph{11}(1). \url{https://doi.org/10.1371/journal.pone.0148228}
-
-\leavevmode\hypertarget{ref-schliep2011phangorn}{}%
-Schliep, K. P. (2011). Phangorn: Phylogenetic analysis in r. \emph{Bioinformatics}, \emph{27}(4), 592--593.
-
-\leavevmode\hypertarget{ref-schoch2020ncbi}{}%
-Schoch, C.
-L., Ciufo, S., Domrachev, M., Hotton, C.
-L., Kannan, S., Khovanskaya, R., \ldots{} others.
-(2020).
-NCBI Taxonomy: a Comprehensive Update on Curation, Resources and Tools.
-\emph{Database}, \emph{2020}.
-
-\leavevmode\hypertarget{ref-selvatti2015paleogene}{}%
-Selvatti, A.
-P., Gonzaga, L.
-P., \& Moraes Russo, C.
-A.
-de.
-(2015).
-A paleogene origin for crown passerines and the diversification of the oscines in the new world.
-\emph{Molecular Phylogenetics and Evolution}, \emph{88}, 1--15.
-
-\leavevmode\hypertarget{ref-smith2018constructing}{}%
-Smith, S., \& Brown, J. (2018). Constructing a broadly inclusive seed plant phylogeny. \emph{American Journal of Botany}, \emph{105}(3), 302--314.
-
-\leavevmode\hypertarget{ref-Smith2012}{}%
-Smith, S., \& O'Meara, B.
-(2012).
-TreePL: Divergence time estimation using penalized likelihood for large phylogenies.
-\emph{Bioinformatics}, \emph{28}(20), 2689--2690.
-\url{https://doi.org/10.1093/bioinformatics/bts492}
-
-\leavevmode\hypertarget{ref-Stoltzfus2013}{}%
-Stoltzfus, A., Lapp, H., Matasci, N., Deus, H., Sidlauskas, B., Zmasek, C.
-M., \ldots{} Jordan, G.
-(2013).
-Phylotastic!
-Making tree-of-life knowledge accessible, reusable and convenient.
-\emph{BMC Bioinformatics}, \emph{14}.
-\url{https://doi.org/10.1186/1471-2105-14-158}
-
-\leavevmode\hypertarget{ref-tietze2013complete}{}%
-Tietze, D.
-T., Päckert, M., Martens, J., Lehmann, H., \& Sun, Y.-H.
-(2013).
-Complete phylogeny and historical biogeography of true rosefinches (aves: Carpodacus).
-\emph{Zoological Journal of the Linnean Society}, \emph{169}(1), 215--234.
-
-\leavevmode\hypertarget{ref-title2016macrophylogenies}{}%
-Title, P.
-O., \& Rabosky, D.
-L.
-(2016).
-Do Macrophylogenies Yield Stable Macroevolutionary Inferences?
-An Example from Squamate Reptiles.
-\emph{Systematic Biology}, syw102.
-\url{https://doi.org/10.1093/sysbio/syw102}
-
-\leavevmode\hypertarget{ref-treplin2008molecular}{}%
-Treplin, S., Siegert, R., Bleidorn, C., Thompson, H.
-S., Fotso, R., \& Tiedemann, R.
-(2008).
-Molecular phylogeny of songbirds (aves: Passeriformes) and the relative utility of common nuclear marker loci.
-\emph{Cladistics}, \emph{24}(3), 328--349.
-
-\leavevmode\hypertarget{ref-vos2012nexml}{}%
-Vos, R.
-A., Balhoff, J.
-P., Caravas, J.
-A., Holder, M.
-T., Lapp, H., Maddison, W.
-P., \ldots{} others.
-(2012).
-NeXML: Rich, extensible, and verifiable representation of comparative data and metadata.
-\emph{Systematic Biology}, \emph{61}(4), 675--689.
-\url{https://doi.org/10.1093/sysbio/sys025}
-
-\leavevmode\hypertarget{ref-Webb2000}{}%
-Webb, C. (2000). Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees. \emph{The American Naturalist}, \emph{156}(2), 145--155.
-
-\leavevmode\hypertarget{ref-Webb2008}{}%
-Webb, C., Ackerly, D., \& Kembel, S.
-(2008).
-Phylocom: Software for the analysis of phylogenetic community structure and trait evolution.
-\emph{Bioinformatics}, \emph{24}(18), 2098--2100.
-\url{https://doi.org/10.1093/bioinformatics/btn358}
-
-\leavevmode\hypertarget{ref-webb2005phylomatic}{}%
-Webb, C., \& Donoghue, M. (2005). Phylomatic: Tree assembly for applied phylogenetics. \emph{Molecular Ecology Notes}, \emph{5}(1), 181--183.
-
-\leavevmode\hypertarget{ref-weir2008calibrating}{}%
-Weir, J., \& Schluter, D. (2008). Calibrating the avian molecular clock. \emph{Molecular Ecology}, \emph{17}(10), 2321--2328.
-
-\leavevmode\hypertarget{ref-zuccon2012phylogenetic}{}%
-Zuccon, D., Prŷs-Jones, R., Rasmussen, P.
-C., \& Ericson, P.
-G.
-(2012).
-The phylogenetic relationships and generic limits of finches (fringillidae).
-\emph{Molecular Phylogenetics and Evolution}, \emph{62}(2), 581--596.
-
-\endgroup
-
-\newpage
-
-\captionof{table}{Ages of congruified nodes. See Figure 3, step C2.}
-\label{tbl:table1}
-\includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example.pdf}
-
-\vspace{15mm}
-
-\captionof{table}{Summary of congruified nodes ages. See Figure 3, step C3.}
-
-\label{tbl:table2}
-\includegraphics[width=\linewidth]{../tables/table-fringillidae-small-example-summary.png}
-
-\begin{minipage}[t]{17.3cm}
-\includegraphics{../figures/figure-workflow/figure1-new.pdf}
-\captionof{figure}{\textit{Main DateLife workflow.
-Analyses can be performed via DateLife's interactive website at \url{www.datelife.org}, or using the \texttt{datelife} R package.
-Details on the R functions used to perform the analyses are available from \texttt{datelife}'s R package vignettes at \url{https://phylotastic.org/datelife}.}}
-\label{fig:workflow}
-\end{minipage}
-
-\includegraphics{../figures/fig_runtime_main.pdf}
-\captionof{figure}{\textit{DateLife's benchmarking results.
+DateLife's benchmarking results.
 Computation time used to process a query and a search across \texttt{datelife}'s chronogram database, relative to number of input taxon names.
 For each N = \{10, 100, 200, ..., 1 000, ...
-, 9 000, 10 000\}, we sampled N species names from the class Aves a hundred times, and then performed a datelife search processing the input names with Taxon Names Resolution Service (TNRS; light gray), and without processing names (dark gray).
-For comparison, we performed a search using an input that had been pre-processed with TNRS (light blue).}}
-\label{fig:runtime_main}
+, 9 000, 10 000\}, we sampled N species names from the class Aves a hundred times, and then performed a \texttt{datelife} search processing the input names using the Taxon Names Resolution Service (TNRS; light gray), and without processing names (dark gray).
+For comparison, we performed a search using an input that has been pre-processed with TNRS (light blue).
 
-\begin{center}
-\includegraphics{../figures/figure-small-example/figure2.pdf}
-\end{center}
-\vspace{10mm}
-\captionof{figure}{\textit{DateLife analysis results for a small sample of A) 6 bird species within the Passeriformes.
-B) Processed species names were found across 9 chronograms within 6 independent studies (Barker et al.
+Figure 3:
+
+DateLife analysis results for a small sample of 6 bird species within the Passeriformes (a).
+Processed species names were found across 9 chronograms within 6 independent studies (b; Barker et al.
 (\protect\hyperlink{ref-barker2012going}{2012}), Barker et al.
@@
 (\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al.
-(\protect\hyperlink{ref-Jetz2012}{2012}).) C) This revealed 28 source age data points for the queried species names.
-Summarized age data is used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree, resulting in a summary chronogram of source ages.}}
+(\protect\hyperlink{ref-Jetz2012}{2012}).
+This revealed 28 source age data points for the queried species names.
+Summarized age data was used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree v13.4, resulting in a chronogram of summary source ages (c).
 
-\label{fig:figure2}
+Figure 4:
 
-\begin{minipage}{18.2cm}
-\includegraphics{../figures/fringillidae-topologies/fringillidae-topology.pdf}
-\captionof{figure}{Tree topologies extracted from Open Tree of Life's (OpenTree) synthetic phylogenetic tree.
-A) Topology of 2,333 tips and 1,305 internal nodes, encompassing bird species within the family Fringillidae following the NCBI taxonomy (black), as well as all other bird species that share the same Most Recent Common Ancestor (MRCA) node in OpenTree's synthetic tree (purple).
-B) Subtree topology of 289 tips and 253 internal nodes, resulting from pruning species that do not belong to the family Fringillidae according to the NCBI taxonomy (purple branches in topology A).
+Tree topologies extracted from Open Tree of Life's (OpenTree) synthetic phylogenetic tree.
+Topology of 2,333 tips and 1,305 internal nodes (a), encompassing bird species within the family Fringillidae following the NCBI taxonomy (black), as well as all other bird species that share the same Most Recent Common Ancestor (MRCA) node in OpenTree's synthetic tree (purple).
+(Subtree topology of 289 tips and 253 internal nodes (b), resulting from pruning species that do not belong to the family Fringillidae according to the NCBI taxonomy (purple branches in topology a).
 Bird species within the Fringillidae are paraphyletic
@@
 Treplin et al. \protect\hyperlink{ref-treplin2008molecular}{2008},
-  Zuccon, Prŷs-Jones, Rasmussen, \& Ericson \protect\hyperlink{ref-zuccon2012phylogenetic}{2012}).}
-\label{fig:fringillidae-topologies}
-\end{minipage}
+Zuccon, Prŷs-Jones, Rasmussen, \& Ericson \protect\hyperlink{ref-zuccon2012phylogenetic}{2012}).
 
-\begin{minipage}{18cm}
-\begin{center}
-\includegraphics{../figures/figure-fringillidae/median_and_calibration_ages-congruified.png}
-\end{center}
+Figure 5:
 
-\captionof{figure}{Fringillidae median summary chronogram generated with DateLife.
+Fringillidae median summary chronogram generated with DateLife.
 It has 289 tips and 253 nodes, from which 212 have age data from at least one published chronogram.
-These nodes have age data points from 19 different chronograms from 13 different studies (colored circles), which were summarized into 194 data points that were used as secondary calibrations (purple slashes) with the Branch Length Adjuster (BLADJ ) software from Webb et al., (\protect\hyperlink{ref-Webb2008}{2008}).}
-\label{fig:fringillidages}
-\end{minipage}
+These nodes have age data points from 19 different chronograms from 13 different studies (colored circles), which were summarized into 194 data points that were used as secondary calibrations (purple slashes) with the Branch Length Adjuster (BLADJ ) software from Webb et al., (\protect\hyperlink{ref-Webb2008}{2008}).
 
-\newpage
+Figure 6:
 
-\begin{minipage}{18cm}
-\begin{center}
-\includegraphics{../figures/fringillidae-cross-validation/fig-cross-validation-xy-plots.pdf}
-\end{center}
-\captionof{figure}{Results from cross validation analysis. Each plot compares the original age estimate (x axis) with the age obtained with a DateLife analysis (y axis), per node.}
-\label{fig:cvXY}
-\end{minipage}
-
-\end{document}
+Cross validation of results from a DateLife analysis of the family Fringillidae, shown in Fig.
+5.
+Each plot compares node age estimate from an original study chronogram (x axis) with the corresponding node age obtained with a DateLife analysis (y axis).

### 2023-05-30 manuscript/manuscript.Rmd
edits to primary secondary calib

@@
 
-While most chronograms in DateLife's database are constructed using primary calibrations, i.e., molecular substitution rates or absolute ages in Myr obtained from the fossil record or geological events, DateLife also constructs summary chronograms using secondary calibrations, i.e., node ages in Myr extracted from existing chronograms.
-The scientific community has generally more confidence in chronograms generated from a single analysis, where carefully chosen primary calibrations constitute the main source of absolute time information used in the dating analysis [@shaul2002playing; @graur2004reading; @sauquet2012testing; @sauquet2013practical; @garzon2015incompatible; @schenk2016sec; @antonelli2017supersmart].
-However, implementation of primary calibrations does not come without difficulties: it requires high expertise and training to discover, place and apply  calibrations appropriately [@ksepka2011synthesizing; @hipsley2014beyond].
-One way around it is to use fossils that have been widely discussed and previously curated as calibrations to date other trees [@ksepka2011synthesizing; @sauquet2013practical], and making sure that all data reflect a coherent evolutionary history [@sauquet2013practical], as for example done by @antonelli2017supersmart.
+DateLife constructs summary chronograms using node ages extracted from existing chronograms, i.e. secondary calibrations.
+In general, the scientific community has more confidence in chronograms using primary calibrations, where the dated tree is generated from a single analysis where carefully chosen fossil calibrations are the source of absolute time information, than in analyses dated using secondary calibrations [@shaul2002playing; @graur2004reading; @sauquet2012testing; @sauquet2013practical; @garzon2015incompatible; @schenk2016sec; @antonelli2017supersmart].
+However, implementation of primary calibrations is difficult: it requires specialized expertise and training to discover, place and apply calibrations appropriately [@ksepka2011synthesizing; @hipsley2014beyond].
+One approach is to use fossils that have been widely discussed and previously curated as calibrations to date other trees [@ksepka2011synthesizing; @sauquet2013practical], and making sure that all data reflect a coherent evolutionary history [@sauquet2013practical], as for example done by @antonelli2017supersmart.
 <!-- Will this be implemented in future `datelife` versions?
@@
 -->
-The Fossil Calibration Database provides availability of 220 primary calibration points encompassing flowering plants and metazoans, that have been curated by experts and used for dating analysis in peer-reviewd publications [@ksepka2015fossil].
+The Fossil Calibration Database provides data for 220 primary calibration points encompassing flowering plants and metazoans, that have been curated by experts and used for dating analysis in peer-reviewed publications [@ksepka2015fossil].
 This database facilitates the use of expert primary fossil calibrations in new phylogenetic dating analyses.
-Yet, users still require the expertise to locate appropriate nodes in their phylogenies to corresponding fossils available in the database, potentially delaying increased usage of the database in research.
-Open databases for effective reuse of fossil calibration data still have room to grow and should constitute a key addition to the phylogenetics tool set.
+Yet, users still require the expertise to locate and calibrate appropriate nodes in their phylogenies which correspond with fossils available in the database.
  <!--though these still have room to grow... BCO: Polite way to say 'no new additions in four years?' LLSR: lol, yes-->
 
-In some cases, secondary calibrations might represent the only information available for a dating analysis (e.g., groups with a poor fossil record).
-Recently, @powell2020quantifying showed in a simulation analysis that secondary calibrations are as good as primary calibrations.
-It seems that using several secondary calibrations (as opposed to just one) can provide sufficient information to alleviate or even neutralize potential biases described before [@shaul2002playing; @graur2004reading; @sauquet2013practical].
+Recently, @powell2020quantifying showed in a simulation analysis that secondary calibrations using node dates based on previous molecular clock analyses can be as good as primary calibrations. 
+Using several secondary calibrations (as opposed to just one) can provide sufficient information to alleviate or even neutralize potential biases [@shaul2002playing; @graur2004reading; @sauquet2013practical].
 <!--\textcolor{red}{LLSR: What does our cross validation analysis show in relation to  the effect of using secondary calibrations for a dating analysis that does not make any prior assumptions on the nature of the calibrations themselves?
-the age difference between primary and secondary calibrations?}--> Our cross validation analysis also provides some insight into the application of secondary calibrations.
-In general, node ages summarized with DateLife and those from the original studies appear positively correlated (Supplementary Figs.
+the age difference between primary and secondary calibrations?}--> Our cross validation analysis also provides insight into the application of secondary calibrations.
+Node ages summarized with DateLife and those from the original studies are well correlated (Supplementary Figs.
 S2-S20).
-We also note that nodes closer to the root tend to be slightly younger than ages from the original studies.
-In contrast, nodes closer to the tips tend to be slightly older when estimated using secondary calibrations than ages from the original studies.
+We also note that DaetLife estimates for nodes closer to the root tend to be slightly younger than ages from the original studies.
+In contrast, nodes closer to the tips tend to be slightly older when estimated using our secondary calibrations than ages from the original studies.
 The only exception to this trend was observed in @burns2014phylogenetics chronogram, which generally displays much younger node ages when estimated using secondary calibrations than the original study (Supplementary Figs.
 S1, S5), supporting previous observations [@sauquet2012testing; @schenk2016sec].
+However, these younger dates are likely an example of how multiple secondary calibrations can correct erroneous estimates, as dates on the @burns2014phylogenetics tree were estimated using a secondary calibration based on a previously estimated molecular evolution rate across birds, and were major outliers compared to alternate estimates for the same nodes based on primary fossil calibrations (figure ??).
 
-We unite to the call of our colleagues: further research is needed to fully understand the effects of using secondary calibrations with alternative tree dating methods, as well as usage of resulting chronograms in downstream analyses [@shaul2002playing; @schenk2016sec; @hipsley2014beyond; @powell2020quantifying].
-It is possible that secondary calibrations can be used successfully with dating methods that do not require setting priors, such as penalized likelihood [@sanderson2003r8s]; with methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ [@Webb2008; @webb2005phylomatic]; and with methods that summarize age data unto a tree topology, such as DateLife.
+Further research is needed to fully understand the effects of using secondary calibrations and the use of resulting chronograms in downstream analyses [@shaul2002playing; @schenk2016sec; @hipsley2014beyond; @powell2020quantifying].
 

### 2023-05-30 notes-secondary-calibrations.Rmd
add more notes on secondary vs primary calibrations

@@
 ## Review of studies from Ksepka in google scholar
+
+## Review of studies that infer molecular substitution rates in birds
+
+Burns et al. 2014: "[we] used a substitution rate of 0.0105 mean substitutions per million years along each branch (Weir and Schluter, 2008)"
+
+Weir & Schluter 2008: "A mitochondrial DNA clock of approximately 2% has been widely used for birds, mammals, and other vertebrate groups (Brown et al.
+1979; Klicka & Zink 1997; Garcia-Moreno 2004; Lovette 2004).
+This traditional rate was based on relatively few calibrations and disparate methods of estimating genetic distances [i.e.
+distances based on third codon positions, all codon positions or restriction fragment length polymorphism (RFLP) data; see review in Lovette 2004].
+Nevertheless, using consistent methods of calibration (all codon positions for a single gene) and an unbiased method for choosing calibration points, our analysis supports an average molecular rate of 2.1% with little variation in mean rate across most orders (Table 1).
+These results suggest that the 2% rate is highly conserved in birds.""

### 2023-05-30 manuscript/manuscript.Rmd
add Brian's edits

@@
 
-Completing a phylogeny using a stochastic birth-death polytomy resolver and a backbone taxonomy is a common practice in scientific publications: @Jetz2012, created a chronogram of all 9, 993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a chronogram of 31, 536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a chronogram of 353, 185 seed plants where only 23% had molecular data.
+Completing a phylogeny using a stochastic birth-death polytomy resolver and a backbone taxonomy is a common practice in scientific publications: @Jetz2012, created a chronogram of all 9,993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a chronogram of 31,536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a chronogram of 353,185 seed plants where only 23% had molecular data.
 These stochastically resolved chronograms return diversification rates estimates that appear less biased than those estimated from their incompletely sampled counterparts, even with methods that account for missing lineages by using sampling fractions [@cusimano2012new; @chang2020estimating], but can also introduce spurious patterns of early bursts of diversification [@cusimano2010slowdowns; @sun2020estimating].

### 2023-05-30 manuscript/manuscript.Rmd
add edits to EJM's edits

@@
 
-Recently, @powell2020quantifying showed in a simulation analysis that secondary calibrations using node dates based on previous molecular clock analyses can be as good as primary calibrations.
+Recently, @powell2020quantifying showed in a simulation study that secondary calibrations using node ages based on previous molecular clock analyses can be as good as primary calibrations.
 Using several secondary calibrations (as opposed to just one) can provide sufficient information to alleviate or even neutralize potential biases [@shaul2002playing; @graur2004reading; @sauquet2013practical].
@@
 S2-S20).
-We also note that DaetLife estimates for nodes closer to the root tend to be slightly younger than ages from the original studies.
+We also note that DateLife estimates for nodes closer to the root tend to be slightly younger than ages from the original studies.
 In contrast, nodes closer to the tips tend to be slightly older when estimated using our secondary calibrations than ages from the original studies.
@@
 S1, S5), supporting previous observations [@sauquet2012testing; @schenk2016sec].
-However, these younger dates are likely an example of how multiple secondary calibrations can correct erroneous estimates, as dates on the @burns2014phylogenetics tree were estimated using a secondary calibration based on a previously estimated molecular evolution rate across birds, and were major outliers compared to alternate estimates for the same nodes based on primary fossil calibrations (figure ??).
+However, these younger dates are more likely an example of how multiple secondary calibrations can correct erroneous estimates, as dates on the @burns2014phylogenetics tree were obtained using a single secondary calibration based on a previously estimated molecular evolution rate across birds from @weir2008calibrating, and appear as major outliers compared to alternate estimates for the same nodes based on primary fossil calibrations (Fig.
+5).
 

### 2023-05-30 notes-secondary-calibrations.Rmd
add edits to EJM's edits

@@
 
-Burns et al. 2014: "[we] used a substitution rate of 0.0105 mean substitutions per million years along each branch (Weir and Schluter, 2008)"
+Burns et al.
+2014 used a single secondary calibration point with Bayesian analyses using BEAST v1.7.1: "[we] used a substitution rate of 0.0105 mean substitutions per million years along each branch [from] (Weir and Schluter, 2008)".
 

### 2023-06-02 manuscript/manuscript.Rmd
fix some fig numbers that were wrong

@@
 As a second example, we analysed 289 bird species in the family Fringillidae that are included in the NCBI taxonomy.
-The resulting summary chronogram is shown in Figure 4, and results from previous steps of the workflow are available as Supplementary Figures.
+The resulting summary chronogram is shown in Figure 5, and results from previous steps of the workflow are available as Supplementary Figures.
 <!--We used `r cite_r("r-references.bib")` for all our analyses.-->
@@
 For a detailed discussion on the state of the synonym, refer to Avibase [@lepage2004avibase; @lepage2014avibase; @avibase-emberiza].
-Discovering this synonym allowed assigning five age data points for the parent node of *Emberiza elegans*, shown as *Schoeniclus elegans* in Figure 3A, which would not have had any data otherwise.
+Discovering this synonym allowed assigning five age data points for the parent node of *Emberiza elegans*, shown as *Schoeniclus elegans* in Figure 3a, which would not have had any data otherwise.
 <!--\textcolor{red}{EJM: I don't understand what this sentence means}
@@
 DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms from 6 different studies (Fig.
-3B).
+3b).
 Three studies matched five input names [@barker2015new; @Hedges2015; @Jetz2012], one study matched four input names [@hooper2017chromosomal] and two studies matched two input names [@barker2013going; @burns2014phylogenetics].
@@
 **_Summarizing search results.--_**
-DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. 3C), and congruified and mapped age data to nodes in this chosen topology (Table 1).
+DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. 3c), and congruified and mapped age data to nodes in this chosen topology (Table 1).
 The name processing step allowed including five data points for node "n4" (parent of _Schoeniclus elegans_; Fig. 3A) that would not have had any data otherwise due to name mismatch.
@@
 As expected, more inclusive nodes (e.g., node "n1") have more variance in age data than less inclusive nodes (e.g., node "n5").
-Summary age data for node "n2" were excluded as final calibration because they are older than age data of the more inclusive node, "n1" (Fig. 3C4).
+Summary age data for node "n2" were excluded as final calibration because they are older than age data of the more inclusive node, "n1" (Fig. 3c4).
 
@@
 The DateLife search identified `r length(dres)` chronograms matching this criteria, published in `r length(unique(names(dres)))` different studies [@barker2013going; @barker2015new; @burns2014phylogenetics; @claramunt2015new; @gibb2015new; @Hedges2015; @hooper2017chromosomal; @Jetz2012; @kimball2019phylogenomic; @oliveros2019earth; @price2014niche; @roquet2014one; @uyeda2017evolution].
-Once identified, DateLife pruned these matching chronograms to remove tips that do not belong to the queried taxon names, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementary Table S1).
+Once identified, DateLife pruned these matching chronograms to remove tips that do not belong to the queried taxon names, and transformed these pruned chronograms to pairwise distance matrices, revealing 1,206 different age data points available for species within the Fringillidae (Supplementary Table S1).
 
@@
 According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig.
-5).
-Hence, a topology containing only the 289 species from the original query was extracted from Open Tree of Life’s synthetic tree v12.3 [@opentreeoflife2019synth].
+4a).
+Hence, a topology containing only the 289 species from the original query was extracted from Open Tree of Life’s synthetic tree v12.3 [Fig.
+4b; @opentreeoflife2019synth].
 
@@
 
-Age data from source chronograms was congruified to OpenTree's topology (Fig.
-5B), reducing the age data set to 818 different data points (Supplementary Table S2).
-For each congruent node, age summary statistics were calculated and used as fixed secondary calibrations over the chosen tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
-4).
+Source chronograms (Supplementary Figs.
+S2-S20) were congruified to OpenTree's topology shown in Figure 4b, reducing the original 1,206 node age data set to 818 different data points (Supplementary Table S2) that can be used as calibrations for the chosen topology (Fig.
+4b).
+The congruent node age data points were summarized for each node, resulting in 194 summary node ages.
+From these 21 were excluded as secondary calibrations because they were older than the ancestral node and/or younger than a descendant node.
+The remaining 173 summary node ages were used as secondary calibrations to obtain a fully dated (and resolved) phylogeny with the program BLADJ (Fig.
+5).
 
@@
 Finally, average node ages per node were applied as secondary calibrations and smoothed with the BLADJ algorithm.
-We found that node ages from the original studies, and ages estimated using all other age data available are largely correlated (Fig. 5).
+We found that node ages from the original studies, and ages estimated using all other age data available are largely correlated (Fig. 6).
 For five studies, DateLife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka 'closer to the root') relative to the original estimate, and overestimate ages for nodes closer to the tips.
@@
 For example, in the DateLife analysis of the Fringillidae shown above, the chronograms from one study [@burns2014phylogenetics] were inferred using molecular substitution rate estimates across birds [@weir2008calibrating], and have much older age estimates for the same nodes than chronograms that were inferred using fossil calibrations (Figs.
-4, 5; Supplementary Figs.
-S1, S5).
+5, 6; Supplementary Figs.
+S1c, S4).
 <!--\textcolor{red}{EJM: MORE ON the FINCH examples and the cross validation here}-->

### 2023-06-02 submission/final-captions.tex
update figure 5 caption

@@
 Fringillidae median summary chronogram generated with DateLife.
-It has 289 tips and 253 nodes, from which 212 have age data from at least one published chronogram.
-These nodes have age data points from 19 different chronograms from 13 different studies (colored circles), which were summarized into 194 data points that were used as secondary calibrations (purple slashes) with the Branch Length Adjuster (BLADJ ) software from Webb et al., (\protect\hyperlink{ref-Webb2008}{2008}).
+The original topology has 289 tips and 253 nodes, from which 194 have age data from at least one published chronogram.
+In total, 19 different chronograms from 13 different studies contribute 818 age data points, which were summarized to obtain a single value for each one of the 194 nodes with age data.
+From the 194 summary ages available, 21 were not used as calibrations (asterisk), because they were older than a parent node and/or younger than a descendant node; the remianing 173 were used as secondary calibrations (forward slash) with the Branch Length Adjuster (BLADJ ) software from Webb et al., (\protect\hyperlink{ref-Webb2008}{2008}).
 

### 2023-06-03 R/plot_node_ages_v2.1.R
fix confidence interval bars in figure 5

@@ -1,0 +1,450 @@
+# @details The node position on the y axis is given by pp$yy
+# pp$yy also stores the positions of the tips
+# the y coordinate of nodes start at pp$yy[ape::Ntip + 1]
+
+#' Plot a Chronogram and Add Age Data to Corresponding Nodes
+#' @description `plot_node_ages` plots a given chronogram and adds age
+#'   data [points()] given in `node_ages` and `summary_ages` to the corresponding nodes.
+#'
+#' @param chronogram A `phylo` object with branch length proportional to time.
+#' @param matched_ages An output of [datelife:::summary.matchedCalibrations()] with age data matched to nodes in `chronogram`.
+#' @param pch_color A named vector or list of colors. Names must correspond to names
+#'  in `matched_ages$in_phy$references`.
+#'  If vector is not named and length > 1, colors will be recycled.
+#' @param pch_transparency A numeric value ranging from 10-99, indicating
+#' color transparency for color defined by `pch_color`. Default to `NULL`, no transparency.
+#' @param pch_type A numeric vector or named list indicating the symbol for calibration age points.
+#'   See [graphics::par()] for symbol options. Defaults to 20 = "bullet circle".
+#' @param pch_cex A numeric value indicating **c**haracter **ex**pansion (i.e.,
+#'  size scaling factor) of symbols defined by `pch`. Default to 1.
+#' @param bars_color A character vector of one element indicating the color for
+#'  node age distribution bars. Defaults to "#80808050", which is hex for gray
+#' ("#808080") with 50\% transparency.
+#' @param bars_lwd A numeric vector indicating the line width for age distribution bars.
+#'   See [graphics::par()] for options. Default to 7.
+#' @param add_legend Default to `TRUE`, adds a legend to the left of the plot.
+#' @param legend_cex A numeric value indicating **c**haracter **ex**pansion (i.e.,
+#'  size scaling factor) of legend. Default to one half the size of the axis label, `cex_axislabel * 0.5`.
+#' @param legend_pt_cex A numeric value indicating **c**haracter **ex**pansion (i.e.,
+#'  character size scaling factor) of legend points. Default to one half the size of the axis label, `cex_axislabel * 0.5`.
+#' @param legend_x the x co-ordinate to be used to position the legend on the left side of the plot.
+#' @param legend_y the y co-ordinate to be used to position the legend on the left side of the plot.
+#' @param legend_box Default to `TRUE`, adds a box around the legend.
+#' @param legend_pch A numeric vector indicating the symbol to use in legend.
[…]

### 2023-06-03 submission/final-captions.tex
fix typos

@@
 In total, 19 different chronograms from 13 different studies contribute 818 age data points, which were summarized to obtain a single value for each one of the 194 nodes with age data.
-From the 194 summary ages available, 21 were not used as calibrations (asterisk), because they were older than a parent node and/or younger than a descendant node; the remianing 173 were used as secondary calibrations (forward slash) with the Branch Length Adjuster (BLADJ ) software from Webb et al., (\protect\hyperlink{ref-Webb2008}{2008}).
+From the 194 summary ages available, 21 were not used as calibrations (asterisk, *), because they were older than a parent node or younger than a descendant node; the remaining 173 summary ages were used as secondary calibrations (forward slash, /) with the Branch Length Adjuster (BLADJ) software from Webb et al., (\protect\hyperlink{ref-Webb2008}{2008}).
 

### 2023-06-06 R/plot_node_ages_v2.2.R
update image of median chronogram on workflow figure (fig 1)

@@ -38,11 +38,11 @@
 #' @inheritParams plot_phylo
 #' @inheritDotParams ape::plot.phylo
 #' @importFrom ape .PlotPhyloEnv
-#' @details Plot are margin sizes as defined by [graphics::par()$mai] and [graphics::par()$omi]
+#' @details Plot area margin sizes, as defined by [graphics::par()$mai] and [graphics::par()$omi],
 #' are overruled within the function. To modify them you have to use the arguments
 #' `mai1`, `mai2`, `mai3` and `mai4`, and omi1, omi2, omi3 and omi4.
 #' @export
-plot_node_ages_2.1 <- function(chronogram,
+plot_node_ages_2.2 <- function(chronogram,
                            matched_ages,
                            time_depth = NULL,
                            plot_type = "phyloch",
@@ -82,6 +82,7 @@ plot_node_ages_2.1 <- function(chronogram,
   }
   # get calibrations that are in_phy only
   # create a list of data.frames with info per node, one with congruified ages, another one with summary ages:
+  # matched_ages <- small_all_ages
   in_phy <- lapply(matched_ages, "[[", "in_phy")
   # obtain max x lim from calibration ages, chronogram and chronogram + root:
   phylo_length <- max(ape::branching.times(chronogram))
@@ -98,12 +99,12 @@ plot_node_ages_2.1 <- function(chronogram,
   ############################################################################
   # start chunk that can be replaced by plot_chronogram
   datelifeplot::plot_chronogram(chronogram,
-                  title = "",
+                  title = title,
                   time_depth = time_depth,
                   plot_type = "phyloch",
                   time_axis = TRUE,
-                  mai1 = 0, mai2 = 0, mai3 = 0.5, mai4 = 2.22314453125,
-                  omi1 = 0, omi2 = 0, omi3 = 0, omi4 = 0,
+                  mai1 = mai1, mai2 = mai2, mai3 = mai3, mai4 = mai4,
+                  omi1 = omi1, omi2 = omi2, omi3 = omi3, omi4 = omi4,
                   plot_height = 30, plot_width = 20,
                   geologic_timescale = "strat2012",
                   geologic_unit = "period",
@@ -111,10 +112,11 @@ plot_node_ages_2.1 <- function(chronogram,
                   cex_axislabel = cex_axislabel,
                   cex_axis = cex_axis,
                   cex_title = cex_title,
-                  pos_title = 1,
-                  pos_axis = 1,
+                  pos_title = pos_title,
[…]

### 2023-06-07 R/plot_node_ages_v2.2.R
update median chronogram fig of small example (fig 3)

@@ -182,6 +182,9 @@ plot_node_ages_2.2 <- function(chronogram,
                 "nodes with congruified age data."))
   mismatches <- c()
   # i = 7
+  if(is.null(in_phy[[1]]$nodeAge)) {
+    stop("nodeAge column is missing in argument `matched_ages[[1]]$in_phy")
+  }
   for (i in mrca_node_numbers) {
     print(i)
     rowsies1 <- in_phy[[1]]$mrca_node_number == i
@@ -382,6 +385,9 @@ plot_node_ages_2.2 <- function(chronogram,
       }
       # determine legend cex:
       legend_cex_i <- legend_cex[i]
+      if (!inherits(legend_pt_cex, "list")) {
+        legend_pt_cex <- rep(list(legend_pt_cex), 2)
+      }
       legend_pt_cex_i <- legend_pt_cex[[i]]
       # determine legend pch:
       legend_pch_i <- unlist(ifelse(missing(legend_pch),

### 2023-06-09 R/plot_node_ages_v2.2.R
fix bug in plot for figure 3 and add opentree chronogram background

@@ -98,6 +98,7 @@ plot_node_ages_2.2 <- function(chronogram,
   }
   ############################################################################
   # start chunk that can be replaced by plot_chronogram
+  # mai4 <- 1
   datelifeplot::plot_chronogram(chronogram,
                   title = title,
                   time_depth = time_depth,
@@ -234,7 +235,7 @@ plot_node_ages_2.2 <- function(chronogram,
   ############################################################################
   ############################################################################
   legend_color_in <- legend_pch_in <- pch_color_in <- vector(mode = "list")
-  # data_set <- "a2"
+  # data_set <- "a1"
   for (data_set in names(in_phy)) {
     ############################################################################
     # we use lastPP$xx positions to get plot x position for node ages
@@ -263,12 +264,14 @@ plot_node_ages_2.2 <- function(chronogram,
       if (is.null(names(pch_color[[data_set]]))) {
         warning("Argument pch_color$", data_set, "has no names, plotting of points will fail.")
       }
-      study_names <- unique(as.character(in_phy[[data_set]]$reference))
-      mm <- names(pch_color[[data_set]]) %in% study_names
+      study_names <- as.character(in_phy[[data_set]]$reference)
+      mm <- study_names %in% names(pch_color[[data_set]])
       if (sum(mm) == 0) {
         stop("Something is wrong with pch_color study names. Do they match node_ages data sets??")
+      } else if (sum(mm) != length(study_names)) {
+        warning("Some study references in data set are not in pch_color")
       }
-      color_pch_all <- pch_color[[data_set]][mm]
+      color_pch_all <- pch_color[[data_set]][study_names]
     }
     ############################################################################
     # Use study references to get vector point types (pch_type)

### 2023-06-09 tables/README-tables.Rmd
create tex file for table 2

@@
 
+tex file of small examples table 1 developed in vignettes/figure-small-example.Rmd L289-295.
+It is the same as the tex resulting in the code at the end.
+
+Tex file for table 2 generated here:
+
 Use of package `xtable` to create latex tables:
 Trick from https://www.youtube.com/watch?v=N-hLp8ADhVE
+
+```{r}
+table2 <- read.csv(file = "table-fringillidae-small-example-summary.csv")
+head(table2)
+t2 <- knitr::kable(table2, 
+                   format = 'latex',
+                   row.names = FALSE,
+                   align = "c",
+                   caption = "Summary of congruified ages per node.",
+                   label = "table2")
+
+# t2 <- xtable(table2, type = "latex")
+class(t2)
+writeLines(t2, '../tables/table-fringillidae-small-example-summary.tex')
+
+```
 

### 2023-06-10 R/plot_node_ages_v2.3.R
fix tables and phylo median on figure 1

@@ -1,0 +1,443 @@
+# @details The node position on the y axis is given by pp$yy
+# pp$yy also stores the positions of the tips
+# the y coordinate of nodes start at pp$yy[ape::Ntip + 1]
+
+#' Plot a Chronogram and Add Age Data to Corresponding Nodes
+#' @description `plot_node_ages` plots a given chronogram and adds age
+#'   data [points()] given in `node_ages` and `summary_ages` to the corresponding nodes.
+#'
+#' @param chronogram A `phylo` object with branch length proportional to time.
+#' @param matched_ages An output of [datelife:::summary.matchedCalibrations()] with age data matched to nodes in `chronogram`.
+#' @param pch_color A named vector or list of colors. Names must correspond to names
+#'  in `matched_ages$in_phy$references`.
+#'  If vector is not named and length > 1, colors will be recycled.
+#' @param pch_transparency A numeric value ranging from 10-99, indicating
+#' color transparency for color defined by `pch_color`. Default to `NULL`, no transparency.
+#' @param pch_type A numeric vector or named list indicating the symbol for calibration age points.
+#'   See [graphics::par()] for symbol options. Defaults to 20 = "bullet circle".
+#' @param pch_cex A numeric value indicating **c**haracter **ex**pansion (i.e.,
+#'  size scaling factor) of symbols defined by `pch`. Default to 1.
+#' @param bars_color A character vector of one element indicating the color for
+#'  node age distribution bars. Defaults to "#80808050", which is hex for gray
+#' ("#808080") with 50\% transparency.
+#' @param bars_lwd A numeric vector indicating the line width for age distribution bars.
+#'   See [graphics::par()] for options. Default to 7.
+#' @param add_legend Default to `TRUE`, adds a legend to the left of the plot.
+#' @param legend_cex A numeric value indicating **c**haracter **ex**pansion (i.e.,
+#'  size scaling factor) of legend. Default to one half the size of the axis label, `cex_axislabel * 0.5`.
+#' @param legend_pt_cex A numeric value indicating **c**haracter **ex**pansion (i.e.,
+#'  character size scaling factor) of legend points. Default to one half the size of the axis label, `cex_axislabel * 0.5`.
+#' @param legend_x the x co-ordinate to be used to position the legend on the left side of the plot.
+#' @param legend_y the y co-ordinate to be used to position the legend on the left side of the plot.
+#' @param legend_box Default to `TRUE`, adds a box around the legend.
[…]

### 2023-06-28 manuscript/manuscript-1st-submission.Rmd
add updated files for review submission

@@
 Hence, DateLife's workflow features different ways of estimating node ages in the absence of calibrations and branch length information for certain taxa.
-"Making up" branch lengths is a common practice in scientific publications: @Jetz2012, created a chronogram of all 9, 993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a chronogram of 31, 536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a chronogram of 353, 185 seed plants where only 23% had molecular data.
+
+Making up branch lengths is a common practice in scientific publications: @Jetz2012, created a chronogram of all 9, 993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a chronogram of 31, 536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a chronogram of 353, 185 seed plants where only 23% had molecular data.
 

### 2023-06-28 manuscript/manuscript-1st-submission.tex
add updated files for review submission

@@
 \usepackage{csquotes}
-\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
-  \usepackage[shorthands=off,main=english]{babel}
-\else
-  % load polyglossia as late as possible as it *could* call bidi if RTL lang (e.g. Hebrew or Arabic)
-  \usepackage{polyglossia}
-  \setmainlanguage[]{english}
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
 \fi
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same} % disable monospaced font for URLs
+\hypersetup{
+  pdftitle={DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life},
+  pdfauthor={Luna L. Sánchez Reyes1,2, Emily Jane McTavish1, \& Brian O'Meara2},
+  pdflang={en-EN},
+  hidelinks,
+  pdfcreator={LaTeX via pandoc}}
 
@@
 DateLife's core software application consists of the R package \texttt{datelife}.
-Its current stable version -- v0.6.6, is available from The Comprehensive R Archive Network (CRAN) repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
+Its current stable version -- v0.6.8, is available from The Comprehensive R Archive Network (CRAN) repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
 ape (Paradis, Claude, \& Strimmer, 2004),
-bold (Chamberlain, 2018),
+bold (Chamberlain et al., 2019),
 geiger (Pennell et al., 2014),
@@
 DateLife starts by processing an input consisting of at least two taxon names, which can be provided as a comma separated character string or as tip labels on a tree.
-If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a \enquote{phylo} R object (Paradis et al., 2004).
+If the input is a tree, it can be provided as a classic newick character string (Archie et al., 1986), or as a ``phylo'' R object (Paradis et al., 2004).
 The input tree is not required to have branch lengths, and its topology is used in the summary steps described in the next section.
@@
 Subspecies and variants are ignored.
-If an input taxon name belongs to an inclusive taxonomic group, DateLife has two alternative behaviors defined by the \enquote{get species from taxon} flag.
+If an input taxon name belongs to an inclusive taxonomic group, DateLife has two alternative behaviors defined by the ``get species from taxon'' flag.
 If the flag is active, DateLife retrieves all species names within the inclusive taxonomic group following a standard taxonomy of choice, and adds them to the input string.
-Taxonomies currently supported by DateLife are Open Tree of Life (OpenTree) unified Taxonomy (OTT, Rees \& Cranston, 2017), the National Center of Biotechnology Information (NCBI) taxonomic database (Schoch et al., 2020), the Global Biodiversity Information Facility (GBIF) taxonomic backbone (GBIF Secretariat, 2022), and the Interim Register of Marine and Nonmarine Genera (IRMNG) database (Rees et al., 2017).
+Taxonomies currently supported by DateLife are Open Tree of Life (OpenTree) unified Taxonomy (OTT, Rees \& Cranston, 2017), the National Center of Biotechnology Information (NCBI) taxonomic database (Schoch et al., 2020), the Global Biodiversity Information Facility (GBIF) taxonomic backbone (GBIF Secretariat, 2022), and the Interim Register of Marine and Nonmarine Genera (IRMNG) database (Rees, Vandepitte, Decock, \& Vanhoorne, 2017).
 If the flag is inactive, DateLife excludes any taxon names above the species level from the search query.
@@
 In rare cases, a group of trees can have multiple groves.
-By default, DateLife chooses the grove with the most taxa, however, the \enquote{criterion = trees} flag allows the user to choose the grove with the most trees instead.
+By default, DateLife chooses the grove with the most taxa, however, the ``criterion = trees'' flag allows the user to choose the grove with the most trees instead.
 If source chronograms do not form a grove, the supertree reconstruction will fail.
@@
 
-For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method (SDM; Criscuolo, Berry, Douzery, \& Gascuel, 2006), which deforms patristic distance matrices by minimizing variance and then averaging them.
+For each congruent node, the pairwise distances that traverse that node are summarized into a single summary matrix using classic summary statistics (i.e., mean, median, minimum and maximum ages), and the Supermatrix Distance Method {[}SDM; Criscuolo, Berry, Douzery, and Gascuel (2006){]}, which deforms patristic distance matrices by minimizing variance and then averaging them.
 These single summary taxon pair age matrices (Summarized calibrations) can be applied as calibrations to date a tree topology, using different dating methods currently supported within DateLife: MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), PATHd8 (Britton, Anderson, Jacquet, Lundqvist, \& Bremer, 2007), BLADJ (Webb, Ackerly, \& Kembel, 2008; Webb \& Donoghue, 2005), and treePL (Smith \& O'Meara, 2012).
@@
 
-Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R \enquote{phylo} format), as well as visualize and compare results graphically, or construct their own graphs using DateLife's chronogram plot generation functions available from the R package \texttt{datelifeplot} (Sanchez-Reyes \& O'Meara, 2022).
+Finally, users can save all source and summary chronograms in formats that permit reuse and reanalyses (newick and R ``phylo'' format), as well as visualize and compare results graphically, or construct their own graphs using DateLife's chronogram plot generation functions available from the R package \texttt{datelifeplot} (Sanchez-Reyes \& O'Meara, 2022).
 
@@
 The final step is to congruify and summarize the age data available for the Fringillidae species into single summary chronograms, using different types of summary ages, median and SDM.
-As explained in the \enquote{Description} section, a tree topology to summarize age data upon is required.
+As explained in the ``Description'' section, a tree topology to summarize age data upon is required.
 By default, DateLife uses the topology from OpenTree's synthetic tree that contains the species in the search query to summarize age data upon.
@@
 We found that node ages from original study, and ages estimated using all other age data available are correlated (Fig. \ref{fig:cvXY}).
-For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka \enquote{closer to the root}) relative to the original estimate, and overestimate ages for nodes closer to the tips.
+For five studies, Datelife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka `closer to the root') relative to the original estimate, and overestimate ages for nodes closer to the tips.
 Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig. S1).
@@
 Hence, DateLife's workflow features different ways of estimating node ages in the absence of calibrations and branch length information for certain taxa.
-\enquote{Making up} branch lengths is a common practice in scientific publications: Jetz et al.
+
+Making up branch lengths is a common practice in scientific publications: Jetz et al.
 (2012), created a chronogram of all 9, 993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
@@
 \hypertarget{refs}{}
-\leavevmode\hypertarget{ref-alstrom2014discovery}{}%
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-alstrom2014discovery}{}}%
 Alström, P., Hooper, D.
@@
 
-\leavevmode\hypertarget{ref-beresford2005african}{}%
-Beresford, P., Barker, F., Ryan, P., \& Crowe, T.
+\leavevmode\vadjust pre{\hypertarget{ref-beresford2005african}{}}%
+Beresford, P., Barker, F., Ryan, Pg., \& Crowe, T.
 (2005).
-African endemics span the tree of songbirds (passeri): Molecular systematics of several evolutionary ``enigmas''.
-\emph{Proceedings of the Royal Society B: Biological Sciences}, \emph{272}(1565), 849--858.
-
-\leavevmode\hypertarget{ref-Boyle2013}{}%
+African endemics span the tree of songbirds (passeri): Molecular systematics of several evolutionary {``enigmas.''} \emph{Proceedings of the Royal Society B: Biological Sciences}, \emph{272}(1565), 849--858.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Boyle2013}{}}%
 Boyle, B., Hopkins, N., Lu, Z., Raygoza Garay, J.
@@
 (2013).
-taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}.
-\emph{F1000Research}, \emph{2}(191), 1--29.
+{taxize : taxonomic search and retrieval in R {[}version 2; referees: 3 approved{]}}.
+\emph{{F1000Research}}, \emph{2}(191), 1--29.
 \url{https://doi.org/10.12688/f1000research.2-191.v2}
 
-\leavevmode\hypertarget{ref-chaves2013biogeography}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Chamberlain2018}{}}%
+Chamberlain, S.
+A., Szöcs, E., Foster, Z., Arendsee, Z., Boettiger, C., Ram, K., \ldots{} Li, G.
+(2019).
+\emph{{taxize: Taxonomic information from around the web}}.
+Retrieved from \url{https://github.com/ropensci/taxize}
+
+\leavevmode\vadjust pre{\hypertarget{ref-chaves2013biogeography}{}}%
 Chaves, J.
@@
 
-\leavevmode\hypertarget{ref-claramunt2015new}{}%
-Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{Science Advances}, \emph{1}(11), e1501005.
-
-\leavevmode\hypertarget{ref-Criscuolo2006}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-claramunt2015new}{}}%
+Claramunt, S., \& Cracraft, J. (2015). A new time tree reveals earth history's imprint on the evolution of modern birds. \emph{{Science Advances}}, \emph{1}(11), e1501005.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Criscuolo2006}{}}%
 Criscuolo, A., Berry, V., Douzery, E.
-J., \& Gascuel, O.
+J.
+P., \& Gascuel, O.
 (2006).
-SDM: A fast distance-based approach for (super)tree building in phylogenomics.
-\emph{Systematic Biology}, \emph{55}(5), 740--755.
+{SDM: A fast distance-based approach for (super)tree building in phylogenomics}.
+\emph{{Systematic Biology}}, \emph{55}(5), 740--755.
 \url{https://doi.org/10.1080/10635150600969872}
 
-\leavevmode\hypertarget{ref-delsuc2018phylogenomic}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-delsuc2018phylogenomic}{}}%
 Delsuc, F., Philippe, H., Tsagkogeorga, G., Simion, P., Tilak, M.-K., Turon, X., \ldots{} Douzery, E.
@@
 (2013).
-Congruification: Support for time scaling large phylogenetic trees.
-\emph{Methods in Ecology and Evolution}, \emph{4}(7), 688--691.
+{Congruification: Support for time scaling large phylogenetic trees}.
+\emph{{Methods in Ecology and Evolution}}, \emph{4}(7), 688--691.
 \url{https://doi.org/10.1111/2041-210X.12051}
 
-\leavevmode\hypertarget{ref-edgar2004muscle}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-edgar2004muscle}{}}%
 Edgar, R. C. (2004). MUSCLE: Multiple sequence alignment with high accuracy and high throughput. \emph{Nucleic Acids Research}, \emph{32}(5), 1792--1797.
 
-\leavevmode\hypertarget{ref-Felsenstein1985a}{}%
-Felsenstein, J. (1985). Phylogenies and the Comparative Method. \emph{The American Naturalist}, \emph{125}(1), 1--15. Retrieved from \url{http://www.jstor.org/stable/2461605}
-
-\leavevmode\hypertarget{ref-gbif2022taxonomy}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Felsenstein1985a}{}}%
+Felsenstein, J. (1985). {Phylogenies and the Comparative Method}. \emph{The American Naturalist}, \emph{125}(1), 1--15. Retrieved from \url{http://www.jstor.org/stable/2461605}
+
+\leavevmode\vadjust pre{\hypertarget{ref-gbif2022taxonomy}{}}%
 GBIF Secretariat.
 (2022).
-GBIF Backbone Taxonomy.
-\emph{Checklist dataset}, (Online Resource accessed via GBIF.org).
+{GBIF Backbone Taxonomy}.
+\emph{{Checklist Dataset}}, (Online Resource accessed via GBIF.org).
 Retrieved from \href{\%7Bhttps://doi.org/10.15468/39omei\%20\%7D}{\{https://doi.org/10.15468/39omei \}}
 
-\leavevmode\hypertarget{ref-gibb2015new}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-gibb2015new}{}}%
 Gibb, G.
@@
 
-\leavevmode\hypertarget{ref-Hedges2015}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-Hedges2015}{}}%
 Hedges, S.
@@
 (2013).
-Multilocus phylogeny and biogeography of the new world pheucticus grosbeaks (aves: Cardinalidae).
+Multilocus phylogeny and biogeography of the new world pheucticus grosbeaks (aves: cardinalidae).
 \emph{Molecular Phylogenetics and Evolution}, \emph{69}(3), 1222--1227.
 
-\leavevmode\hypertarget{ref-rabosky2018inverse}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-RCoreTeam2018}{}}%
+R Core Team. (2018). \emph{{R: a language and environment for statistical computing}}. Vienna, Austria: R Foundation for Statistical Computing.
+
+\leavevmode\vadjust pre{\hypertarget{ref-rabosky2018inverse}{}}%
 Rabosky, D.
@@
 
-\leavevmode\hypertarget{ref-datelifeplot}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-datelifeplot}{}}%
 Sanchez-Reyes, L.
@@
 (2013).
-Phylotastic!
-Making tree-of-life knowledge accessible, reusable and convenient.
-\emph{BMC Bioinformatics}, \emph{14}.
+{Phylotastic!
+Making tree-of-life knowledge accessible, reusable and convenient}.
+\emph{{BMC Bioinformatics}}, \emph{14}.
 \url{https://doi.org/10.1186/1471-2105-14-158}
 
-\leavevmode\hypertarget{ref-tietze2013complete}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-tietze2013complete}{}}%
 Tietze, D.
@@
 \emph{Systematic Biology}, \emph{61}(4), 675--689.
-\url{https://doi.org/10.1093/sysbio/sys025}
-
-\leavevmode\hypertarget{ref-Webb2000}{}%
-Webb, C. (2000). Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees. \emph{The American Naturalist}, \emph{156}(2), 145--155.
-
-\leavevmode\hypertarget{ref-Webb2008}{}%
+
+\leavevmode\vadjust pre{\hypertarget{ref-Webb2000}{}}%
+Webb, C. (2000). {Exploring the Phylogenetic Structure of Ecological Communities : An Example for Rain Forest Trees}. \emph{{The American Naturalist}}, \emph{156}(2), 145--155.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Webb2008}{}}%
 Webb, C., Ackerly, D., \& Kembel, S.
 (2008).
-Phylocom: Software for the analysis of phylogenetic community structure and trait evolution.
+{Phylocom: Software for the analysis of phylogenetic community structure and trait evolution}.
 \emph{Bioinformatics}, \emph{24}(18), 2098--2100.
@@
 \emph{Molecular Phylogenetics and Evolution}, \emph{62}(2), 581--596.
+
+\end{CSLReferences}
 

### 2023-06-28 manuscript/manuscript-original-submission.tex
add updated files for review submission

--- 2023-01-25 manuscript/manuscript-1st-submission.tex
@@
 {}{\PatchFailed}
-\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4959}
+\keywords{Tree; Phylogeny; Scaling; Dating; Ages; Divergence times; Open Science; Congruification; Supertree; Calibrations; Secondary calibrations\newline\indent Word count: 4960}
 \DeclareDelayedFloatFlavor{ThreePartTable}{table}
@@
 Here we present the DateLife project which has the main goal of capturing age data from published chronograms, and making these data readily accessible to the community for reuse and reanalysis, for research, teaching, and science communication and policy.
-DateLife's core software application is available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org.
+DateLife's core software application is available as an R package ({\textbf{???}}), and as an online Rshiny interactive website at www.datelife.org.
 It features key elements for scientific reproducibility, such as a versioned, open and fully public source database (McTavish et al., 2015), data stored and available in a computer readable format (Vos et al., 2012), automated and programmatic ways of accessing the data (Stoltzfus et al., 2013) and methods to summarize and compare the data.
@@
 DateLife's core software application consists of the R package \texttt{datelife}.
-Its current stable version -- v0.6.6, is available from The Comprehensive R Archive Network (CRAN) repository (Sanchez-Reyes et al., 2022), and relies on functionalities from various biological R packages:
+Its current stable version -- v0.6.8, is available from the The Comprehensive R Archive Network (CRAN) repository ({\textbf{???}}), and relies on functionalities from various biological R packages:
 ape (Paradis, Claude, \& Strimmer, 2004),
@@
 DateLife requires a tree topology to summarize age data upon.
-We recommend that users provide as input a tree topology from the literature, or one of their own making.
+We recommend that users provide a tree topology as input from the literature, or one of their own making.
 If no topology is provided, DateLife automatically extracts one from the OpenTree synthetic tree, a phylogeny encompassing 2.3 million taxa across all life, assembled from 1, 239 published phylogenetic trees and OpenTree's unified Taxonomy, OTT (Open Tree Of Life et al., 2019).
@@
 
-Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used (Magallón, 2010; Magallón et al., 2015).
+Some measures that have been proposed are the proportion of lineage sampling and the number of calibrations used (Magallón, 2010, @magallon2015metacalibrated).
 Some characteristics that are often cited in published studies as a measure of improved age estimates as compared to previously published estimates are: quality of alignment (missing data, GC content), lineage sampling (strategy and proportion), phylogenetic and dating inference method, number of fossils used as calibrations, support for nodes and ages, and magnitude of confidence intervals.
@@
 
-\leavevmode\hypertarget{ref-datelifeR}{}%
-Sanchez-Reyes, L.
-L., O'Meara, B., Eastman, J., Heath, T., Wright, A., Schliep, K., \ldots{} Alfaro, M.
-(2022).
-\texttt{datelife}: Scientific Data on Time of Lineage Divergence for Your Taxa.
-\emph{R Package Release V0.6.2}.
-Retrieved from \url{https://doi.org/10.5281/zenodo.593938}
-
 \leavevmode\hypertarget{ref-sanderson2002}{}%

### 2023-06-28 manuscript/manuscript.tex
add updated files for review submission

@@
 
-School of Natural Sciences, University of California, Merced, Science and Engineering Building 1, Merced, CA 95340, USA.
+Department of Life and Environmental Sciences, University of California, Merced, CA 95343, USA.
 
@@
 
-Word count: 6715
+Word count: 6707
 
@@
 
-Here we present the DateLife project which has the main goal of extracting and exposing age data from published chronograms, making age data readily accessible to the wider community for reuse and reanalysis in research, teaching, science communication and conservation policy.
+Here we present the DateLife project which has the main goal of extracting and exposing age data from published chronograms, making age data readily accessible to a wider community for reuse and reanalysis in research, teaching, science communication and conservation policy.
 DateLife's core software application is available as an R package (Sanchez-Reyes et al., 2022), and as an online Rshiny interactive website at www.datelife.org.
@@
 At the time of writing of this manuscript
-(May 15, 2023),
+(Jun 11, 2023),
 DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies.
@@
 We recommend that users provide as input a tree topology from the literature, or one of their own making.
-If no topology is provided, DateLife automatically extracts one from the OpenTree synthetic tree, a phylogeny currently encompassing 2.3 million taxa across all life, assembled from 1, 239 published phylogenetic trees and OpenTree's unified Taxonomy, OTT (Open Tree Of Life et al., 2019).
+If no topology is provided, DateLife automatically extracts one from the OpenTree synthetic tree, a phylogeny currently encompassing 2.3 million taxa across all life, assembled from 1,239 published phylogenetic trees and OpenTree's unified Taxonomy, OTT (Open Tree Of Life et al., 2019).
 Alternatively, DateLife can combine topologies from source chronograms using a supertree approach (Criscuolo, Berry, Douzery, \& Gascuel, 2006).
@@
 
-\textbf{\emph{Dating a tree with no branch lengths.--}}
+\textbf{\emph{Dating a tree without branch lengths.--}}
 When producing or obtaining a tree with branch lengths for a group of interest is not possible, DateLife can date a topology without branch lengths, obtained from OpenTree or by implementing the supertree approach described above, by applying the Branch Length Adjuster (BLADJ; Webb et al., 2008; Webb \& Donoghue, 2005) algorithm, which requires no initial branch lengths.
@@
 If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature.
-If none is provided, DateLife will not return a dated topology and will provide a warning message along with suggestions on how the user can provide an age for the root so that a fully dated tree can be returned.
+If none is provided, DateLife will choose an age for the root so that it can return a dated topology.
+It will also provide a conspicuous warning message so that users are aware that the root of the chronogram does not have information available in the chronogram database, along with suggestions on how the user can provide an appropriate age for the root.
 
@@
 As a second example, we analysed 289 bird species in the family Fringillidae that are included in the NCBI taxonomy.
-The resulting summary chronogram is shown in Figure 4, and results from previous steps of the workflow are available as Supplementary Figures.
+The summary chronogram resulting from the DateLife analysis is shown in Figure 5, and results from previous steps of the workflow are available as Supplementary Figures.
 
@@
 For a detailed discussion on the state of the synonym, refer to Avibase (Avibase, 2022; Lepage, 2004; Lepage, Vaidya, \& Guralnick, 2014).
-Discovering this synonym allowed assigning five age data points for the parent node of \emph{Emberiza elegans}, shown as \emph{Schoeniclus elegans} in Figure 3A, which would not have had any data otherwise.
+Discovering this synonym allowed assigning five age data points for the parent node of \emph{Emberiza elegans}, shown as \emph{Schoeniclus elegans} in Figure 3a, which would not have had any data otherwise.
 
@@
 DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms from 6 different studies (Fig.
-3B).
+3b).
 Three studies matched five input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2015; Hedges, Marin, Suleski, Paymer, \& Kumar, 2015; Jetz et al., 2012), one study matched four input names (Hooper \& Price, 2017) and two studies matched two input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2013; Burns et al., 2014).
@@
 \textbf{\emph{Summarizing search results.--}}
-DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. 3C), and congruified and mapped age data to nodes in this chosen topology (Table 1).
+DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. 3c), and congruified and mapped age data to nodes in this chosen topology, shown in Table 1.
 The name processing step allowed including five data points for node \enquote{n4} (parent of \emph{Schoeniclus elegans}; Fig. 3A) that would not have had any data otherwise due to name mismatch.
@@
 As expected, more inclusive nodes (e.g., node \enquote{n1}) have more variance in age data than less inclusive nodes (e.g., node \enquote{n5}).
-Summary age data for node \enquote{n2} were excluded as final calibration because they are older than age data of the more inclusive node, \enquote{n1} (Fig. 3C4).
+Median summary age data for node \enquote{n2} was excluded as final calibration because it is older than the median age of a more inclusive node, \enquote{n1} (Fig. 3c4).
 
@@
 The DateLife search identified 19 chronograms matching this criteria, published in 13 different studies (Barker et al., 2013, 2015; Burns et al., 2014; Claramunt \& Cracraft, 2015; Gibb et al., 2015; Hedges et al., 2015; Hooper \& Price, 2017; Jetz et al., 2012; Kimball et al., 2019; Oliveros et al., 2019; Price et al., 2014; Roquet, Lavergne, \& Thuiller, 2014; Uyeda, Pennell, Miller, Maia, \& McClain, 2017).
-Once identified, DateLife pruned these matching chronograms to remove tips that do not belong to the queried taxon names, and transformed these pruned chronograms to pairwise distance matrices, revealing 1, 206 different age data points available for species within the Fringillidae (Supplementary Table S1).
+Once identified, DateLife pruned these matching chronograms to remove tips that do not belong to the queried taxon names, and transformed these pruned chronograms to pairwise distance matrices, revealing 1,206 different age data points available for species within the Fringillidae (Supplementary Table S1).
 
@@
 According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig.
+4a).
+Hence, a topology containing only the 289 species from the original query was extracted from Open Tree of Life's synthetic tree v12.3 (Fig.
+4b; Open Tree Of Life et al., 2019).
+
+Source chronograms (Supplementary Figs.
+S2-S20) were congruified to OpenTree's topology shown in Figure 4b, reducing the original 1,206 node age data set to 818 different data points (Supplementary Table S2) that can be used as calibrations for the chosen topology (Fig.
+4b).
+The congruent node age data points were summarized for each node, resulting in 194 summary node ages.
+From these 21 were excluded as secondary calibrations because they were older than the ancestral node.
+The remaining 173 summary node ages were used as secondary calibrations to obtain a fully dated (and resolved) phylogeny with the program BLADJ (Fig.
 5).
-Hence, a topology containing only the 289 species from the original query was extracted from Open Tree of Life's synthetic tree v12.3 (Open Tree Of Life et al., 2019).
-
-Age data from source chronograms was congruified to OpenTree's topology (Fig.
-5B), reducing the age data set to 818 different data points (Supplementary Table S2).
-For each congruent node, age summary statistics were calculated and used as fixed secondary calibrations over the chosen tree topology, to obtain a fully dated phylogeny with the program BLADJ (Fig.
-4).
 
@@
 Finally, average node ages per node were applied as secondary calibrations and smoothed with the BLADJ algorithm.
-We found that node ages from the original studies, and ages estimated using all other age data available are largely correlated (Fig. 5).
+We found that node ages from the original studies, and ages estimated using all other age data available are largely correlated (Fig. 6).
 For five studies, DateLife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka \enquote{closer to the root}) relative to the original estimate, and overestimate ages for nodes closer to the tips.
@@
 As we envision that DateLife will have many interesting applications in research and beyond, we emphasize that DateLife's results --as well as any insights gleaned from them, largely depend on the quality of the source chronograms: low quality chronograms will produce low quality results.
-This \enquote{garbage in, garbage out} problem has long been recognised in supertree methods for summarizing phylogenetic trees (Bininda-Emonds et al., 2004).
+The \enquote{garbage in, garbage out} problem has long been recognised in supertree methods for summarizing phylogenetic trees (Bininda-Emonds et al., 2004).
 We note that this is a surfacing issue of any automated tool for biological data analysis.
@@
 Again, we urge readers and DateLife users to explore all input chronograms before using a summary chronogram resulting from a DateLife workflow.
+
 Finally, uncertainty and variability of chronogram node age estimates might pose larger issues in some research areas than others.
@@
 For example, in the DateLife analysis of the Fringillidae shown above, the chronograms from one study (Burns et al., 2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older age estimates for the same nodes than chronograms that were inferred using fossil calibrations (Figs.
-4, 5; Supplementary Figs.
-S1, S5).
+5, 6; Supplementary Figs.
+S1c, S4).
 Another source of conflict in estimated node ages can arise from different placements for the same calibration, which would imply fundamentally distinct evolutionary hypotheses (Antonelli et al., 2017).
@@
 
-While most chronograms in DateLife's database are constructed using primary calibrations, i.e., molecular substitution rates or absolute ages in Myr obtained from the fossil record or geological events, DateLife also constructs summary chronograms using secondary calibrations, i.e., node ages in Myr extracted from existing chronograms.
-The scientific community has generally more confidence in chronograms generated from a single analysis, where carefully chosen primary calibrations constitute the main source of absolute time information used in the dating analysis (Antonelli et al., 2017; Garzón-Orduña, Silva-Brandão, Willmott, Freitas, \& Brower, 2015; Graur \& Martin, 2004; Sauquet, 2013; Sauquet et al., 2012; Schenk, 2016; Shaul \& Graur, 2002).
-However, implementation of primary calibrations does not come without difficulties: it requires high expertise and training to discover, place and apply calibrations appropriately (Hipsley \& Müller, 2014; Ksepka et al., 2011).
-One way around it is to use fossils that have been widely discussed and previously curated as calibrations to date other trees (Ksepka et al., 2011; Sauquet, 2013), and making sure that all data reflect a coherent evolutionary history (Sauquet, 2013), as for example done by Antonelli et al.
+DateLife constructs summary chronograms using node ages extracted from existing chronograms, i.e.~secondary calibrations.
+In general, the scientific community has more confidence in chronograms using primary calibrations, where the dated tree is generated from a single analysis where carefully chosen fossil calibrations are the source of absolute time information, than in analyses dated using secondary calibrations (Antonelli et al., 2017; Garzón-Orduña, Silva-Brandão, Willmott, Freitas, \& Brower, 2015; Graur \& Martin, 2004; Sauquet, 2013; Sauquet et al., 2012; Schenk, 2016; Shaul \& Graur, 2002).
+However, implementation of primary calibrations is difficult: it requires specialized expertise and training to discover, place and apply calibrations appropriately (Hipsley \& Müller, 2014; Ksepka et al., 2011).
+One approach is to use fossils that have been widely discussed and previously curated as calibrations to date other trees (Ksepka et al., 2011; Sauquet, 2013), and making sure that all data reflect a coherent evolutionary history (Sauquet, 2013), as for example done by Antonelli et al.
 (2017).
-The Fossil Calibration Database provides availability of 220 primary calibration points encompassing flowering plants and metazoans, that have been curated by experts and used for dating analysis in peer-reviewd publications (Ksepka et al., 2015).
+The Fossil Calibration Database provides data for 220 primary calibration points encompassing flowering plants and metazoans, that have been curated by experts and used for dating analysis in peer-reviewed publications (Ksepka et al., 2015).
 This database facilitates the use of expert primary fossil calibrations in new phylogenetic dating analyses.
-Yet, users still require the expertise to locate appropriate nodes in their phylogenies to corresponding fossils available in the database, potentially delaying increased usage of the database in research.
-Open databases for effective reuse of fossil calibration data still have room to grow and should constitute a key addition to the phylogenetics tool set.
-
-In some cases, secondary calibrations might represent the only information available for a dating analysis (e.g., groups with a poor fossil record).
-Recently, Powell, Waskin, and Battistuzzi (2020) showed in a simulation analysis that secondary calibrations are as good as primary calibrations.
-It seems that using several secondary calibrations (as opposed to just one) can provide sufficient information to alleviate or even neutralize potential biases described before (Graur \& Martin, 2004; Sauquet, 2013; Shaul \& Graur, 2002).
-Our cross validation analysis also provides some insight into the application of secondary calibrations.
-In general, node ages summarized with DateLife and those from the original studies appear positively correlated (Supplementary Figs.
+Yet, users still require the expertise to locate and calibrate appropriate nodes in their phylogenies which correspond with fossils available in the database.
+
+Recently, Powell, Waskin, and Battistuzzi (2020) showed in a simulation study that secondary calibrations using node ages based on previous molecular clock analyses can be as good as primary calibrations.
+Using several secondary calibrations (as opposed to just one) can provide sufficient information to alleviate or even neutralize potential biases (Graur \& Martin, 2004; Sauquet, 2013; Shaul \& Graur, 2002).
+Our cross validation analysis also provides insight into the application of secondary calibrations.
+Node ages summarized with DateLife and those from the original studies are well correlated (Supplementary Figs.
 S2-S20).
-We also note that nodes closer to the root tend to be slightly younger than ages from the original studies.
-In contrast, nodes closer to the tips tend to be slightly older when estimated using secondary calibrations than ages from the original studies.
+We also note that DateLife estimates for nodes closer to the root tend to be slightly younger than ages from the original studies.
+In contrast, nodes closer to the tips tend to be slightly older when estimated using our secondary calibrations than ages from the original studies.
 The only exception to this trend was observed in Burns et al.
@@
 S1, S5), supporting previous observations (Sauquet et al., 2012; Schenk, 2016).
-
-We unite to the call of our colleagues: further research is needed to fully understand the effects of using secondary calibrations with alternative tree dating methods, as well as usage of resulting chronograms in downstream analyses (Hipsley \& Müller, 2014; Powell et al., 2020; Schenk, 2016; Shaul \& Graur, 2002).
-It is possible that secondary calibrations can be used successfully with dating methods that do not require setting priors, such as penalized likelihood (Sanderson, 2003); with methods that do not make any assumptions on the ages and fix them to a node on a tree topology, such as BLADJ (Webb et al., 2008; Webb \& Donoghue, 2005); and with methods that summarize age data unto a tree topology, such as DateLife.
+However, these younger dates are more likely an example of how multiple secondary calibrations can correct erroneous estimates, as dates on the Burns et al.
+(2014) tree were obtained using a single secondary calibration based on a previously estimated molecular evolution rate across birds from Weir and Schluter (2008), and appear as major outliers compared to alternate estimates for the same nodes based on primary fossil calibrations (Fig.
+5).
+
+Further research is needed to fully understand the effects of using secondary calibrations and the use of resulting chronograms in downstream analyses (Hipsley \& Müller, 2014; Powell et al., 2020; Schenk, 2016; Shaul \& Graur, 2002).
 
@@
 Completing a phylogeny using a stochastic birth-death polytomy resolver and a backbone taxonomy is a common practice in scientific publications: Jetz et al.
-(2012), created a chronogram of all 9, 993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
-(2018) created a chronogram of 31, 536 ray-finned fishes, of which only 37\% had molecular data; Smith and Brown (2018) constructed a chronogram of 353, 185 seed plants where only 23\% had molecular data.
+(2012), created a chronogram of all 9,993 bird species, where 67\% had molecular data and the rest was simulated; Rabosky et al.
+(2018) created a chronogram of 31,536 ray-finned fishes, of which only 37\% had molecular data; Smith and Brown (2018) constructed a chronogram of 353,185 seed plants where only 23\% had molecular data.
 These stochastically resolved chronograms return diversification rates estimates that appear less biased than those estimated from their incompletely sampled counterparts, even with methods that account for missing lineages by using sampling fractions (Chang et al., 2020; Cusimano et al., 2012), but can also introduce spurious patterns of early bursts of diversification (Cusimano \& Renner, 2010; Sun et al., 2020).
@@
 The DateLife project allows for easy and fast summarization of public and state-of-the-art data on time of lineage divergence.
-It provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identifying regions that require more research, or that have conflicting information.
 It is available as an R package, and as a web-based R shiny application at www.datelife.org.
-Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
-The DateLife project will improve awareness of the existing variation in expert estimations time of divergence data, and foster exploration of the effect of alternative divergence time hypotheses on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
+DateLife provides a straightforward way to get an informed picture of the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identifying regions that require more research, or that have conflicting information.
+Additionally, both summary and newly generated trees using the DateLife workflow are useful to evaluate evolutionary hypotheses in different areas of research.
+We hope that the DateLife project will increase awareness of the existing variation in expert estimations of time of divergence, and foster exploration of the effect of alternative divergence time hypotheses on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
 
@@
 
-Supplementary Figures can be viewed and dowloaded from their Zenodo repository (\url{https://doi.org/10.5281/zenodo.6683667}).
+Supplementary Figures can be viewed and downloaded from their Zenodo repository (\url{https://doi.org/10.5281/zenodo.6683667}).
 Supplementary material, including code, biological examples, benchmark results, data files and online-only appendices, can be downloaded from the Dryad data repository (\url{https://doi.org/10.5061/dryad.cnp5hqc6w}), as well as in the Zenodo stable repositories that host the reproducible manuscript (\url{https://doi.org/10.5281/zenodo.7435094}), the biological examples (\url{https://doi.org/10.5281/zenodo.7435101}), and the software benchmark (\url{https://doi.org/10.5281/zenodo.7435106}).
@@
 
-\leavevmode\hypertarget{ref-sanderson2003r8s}{}%
-Sanderson, M. (2003). r8s: Inferring Absolute Rates of Molecular Evolution and Divergence Times in the Absence of a Molecular Clock. \emph{Bioinformatics}, \emph{19}(2), 301--302.
-
 \leavevmode\hypertarget{ref-sanderson2001sources}{}%

### 2023-06-28 review.Rmd
add updated files for review submission

@@
     keep_md: yes
-date: "2023-05-01"
+date: "2023-06-09"
 ---
@@
 
-**Please find below response to each point raised by the _Associate Editor_ Daniele Silvestro and the two _Reviewers_. We greatly appreciate your time and consideration for our resubmission.**
+**Please find below response to each point raised by the _Associate Editor_ Daniele Silvestro and the two _Reviewers_.
+We experienced a delay due to a dependency to our software that was orphaned, and kicked out from the official R repository (CRAN), putting at risk the status of our own software.
+DateLife is hosted in other stable resources, but CRAN is the most widely used, so we decided to adopt the orphaned dependency to get it back on CRAN.
+However, the dependency is written in C, a language that we are not very familiar with, so it took us a while to figure it out.
+It was worthy, as our package is now stable in the official CRAN repository.
+We also carefully addressed all points made by the reviewers and AE, and we think they greatly improved the manuscript.
+We greatly appreciate your time and consideration for our resubmission.**
 
@@
 
-**We added DOIs for Dryad package with Supplementary Material, L551.**
+**We added the DOI of our Dryad package containing all Supplementary Material on L690 (manuscript-differences.pdf).**
 
@@
 
-**Done on L556-558.**
+**Done on L700 (manuscript-differences.pdf).**
 
@@
 **Thanks for this question.
-We elaborate on this on the Description section, subsection "Dating a tree with branch lengths" L221-228.
+We elaborate on this on the Description section, subsection "Dating a tree with branch lengths" L278 (manuscript-differences.pdf).
 Briefly, we implement the parsimony algorithm ACCTRAN using functions from the R package phangorn.
@@
 **Thanks for this questions.
-We elaborate on it on the "Description" section, at the end of subsection "Dating a tree topology with no branch lengths", L 202-210.Note that we expanded the subsection "Dating a tree topology" into three sections: "Applying secondary calibrations", "Dating a tree topology" and "Dating a tree with branch lengths".**
+We elaborate on it on the "Description" section, at the end of subsection "Dating a tree topology without branch lengths", L 2268 (manuscript-differences.pdf).
+Note that we expanded the subsection "Dating a tree topology" into three sections: "Applying secondary calibrations", "Dating a tree topology" and "Dating a tree with branch lengths".**
 
 **We want to point out that the main goal of the summarizing step in DateLife (when we distribute ages evenly in the chronogram) is to provide a single chronogram that can quickly show at a glance, and in the most agnostic way possible, the distribution of node ages obtained from published studies.
-Using an exponential distribution for missing node ages would make additional assumptions about the underlying evolutionary model of the chronogram, which can bias results in downstream analyses, as we elaborate in the discussion section, "Effects of phylogenetic sampling on downstream analyses" (L445 of manuscript without differences; L527 of manuscript with differences).**
+Using an exponential distribution for missing node ages would make additional assumptions about the underlying evolutionary model of the chronogram, which can bias results in downstream analyses, as we elaborate in the discussion section, "Effects of taxon sampling on downstream analyses" (L598 of file manuscript-differences.pdf).**
 
@@
 We agree that this practice could be dangerous if a user is not aware that the age of the root was set in an arbitrary way.
-We changed the behaviour of the code to return a very conspicuous warning message when the root is arbitrary to make sure that users are aware of the fact (L196-201).**
+We changed the behaviour of the code to return a very conspicuous warning message when the root is arbitrary to make sure that users are aware of the fact that an arbitrary age is assigned to the root (L262 manuscript-differences.pdf).**
 
@@
 
-**Right! We added Zenodo links for our code and other materials used for this research in the Availability section (L505) and Supplementary Material (L518-520).**
+**Right!
+We added Zenodo links for our code and other materials used for this research in the Availability section (L677 manuscript-differences.pdf) and Supplementary Material section (L5689 manuscript-differences.pdf).**
 
@@
 **Thanks for pointing us to these references.
-We elaborated on this subject more thoroughly by adding results from these studies to the corresponding section on the Discussion, "Effects of phylogenetic sampling on downstream analyses" (L481 of manuscript without differences; L527 of manuscript with differences), as well as changing the wording from "making up" to "in the absence of genetic data, simulating following a birth-death process" (L489; L540).**
+We elaborated on this subject more thoroughly by adding results from these studies to the corresponding section on the Discussion, "Effects of taxon sampling on downstream analyses" (L600 manuscript-differences.pdf), as well as changing the wording from "making up" to "completing a phylogeny using a stochastic birth-death polytomy".**
 
@@
 **We agree that users of all biological software should keep in mind that the results they get are only as good as the data they provide, and that they should implement some quality control.
-We mention this in the abstract and 3rd paragraph of discussion.**
+We mention this in the abstract and 3rd paragraph of discussion L456 (manuscript-differences.pdf).**
 
@@
 As shown in Figure 5, DateLife allows to compare all age data available for a group, and users can immediately identify chronograms that are outliers, and explore whether it is related to an artifact of the chronogram or something else about the data or methodology used to generate that chronogram.
-Users can then choose to drop those calibrations from the final analysis.**
+Users can then choose to drop those calibrations from the final analysis.
+We mention this in several points of the disucssion.**
 
-**We agree that it is a responsibility of the users to check the trees and chonograms before using them as secondary calibrations, in the same way researchers curate fossils to use as primary calibration points.**
+**Generaly, it is a responsibility of the users to check the trees and chonograms before using them as secondary calibrations, in the same way researchers curate fossils to use as primary calibration points.**
 
@@
 If there is more than one age data point, the function now adds one standard deviation unit of the mean of published ages to the maximum age value, and uses that as the root age.
-Second, the function now returns a conspicuous warning message when the root is not based on published data to make sure that users are aware of the fact (L196-201).**
+Second, the function now returns a conspicuous warning message when the root is not based on published data to make sure that users are aware of the fact (L266 manuscript-differences.pdf).**
 
@@
 
-**Thanks for pointing that out. We added that topic to the intro along with a couple of references, L48-49.**
+**Thanks for pointing that out. We added that topic to the intro along with a couple of references, L71 manuscript-differences.pdf.**
 
@@
 When provided by the user, subspecific taxa are processed and searched fully.
-We clarify this in the Description section "Creating a search query".**
+We clarify this in the Description section "Creating a search query" L104 manuscript-differences.pdf.**
 
@@
 
-**It is indeed currently synced "manually". We specify this in the Description, L125. We comment in the discussion the benefits of having syncs done automatically, and ways to do so.**
+**It is indeed currently synced "manually".
+We specify this in the Description, L163 manuscript-differences.pdf.
+On the discussion, we elaborate on the benefits of having syncs done automatically, and ways to do so L446 manuscript-differences.pdf.**
 
@@
 
-**We comment on this on L231-232. **
+**We comment on this on L288 manuscript-differences.pdf. **
 
@@
 
-**Indeed!
-The golden standard for obtaining divergence time estimates entails the use of primary calibrations (fossil or from geologic events).
+**Indeed! We addressed that in the discussion L530 (manuscript-differences.pdf).**
+**The golden standard for obtaining divergence time estimates entails the use of primary calibrations (fossil or from geologic events).
 Fossil preservation bias makes it impossible to get calibrations to constrain all nodes in a phylogenetic tree, hence, we will always have nodes that we need to infer using a model.**
@@
 **The goal of DateLife is to provide the user with an insight on the current state of knowledge of time of lineage divergence for any group of taxa, as well as a tree summarizing the data.**
-
-**We addressed that in the discussion.**
 
@@
 
-**Thanks for pointing us to that paper! It gave us some nice ideas for discussion. We added a paragraph on the Discussion.**
+**Thanks for pointing us to that paper! It gave us some nice ideas for discussion. We added a paragraph on the Discussion L495 of manuscript-differences.pdf.**
 
@@
 
-**Thanks for the suggestion! We created stable versions for all three repositories on Zenodo, and refer the doi in addition to the GitHub addresses, L522-525.**
+**Thanks for the suggestion! We created stable versions for all three repositories on Zenodo, and refer the doi in addition to the GitHub URLs, L677 and L699 of manuscript-differences.pdf.**

### 2023-06-28 submission/final-captions.tex
add updated files for review submission

@@
+Table 1
+
+Ages of nodes extracted from 9 chronograms containing at least 2 of the 6 bird species used in the Fringillidae small example of the DateLife workflow.
+All nodes are congruent with the chosen tree topology.
+See Figure 3, step c2.
+
+Table 2
+
+Summary statistics of node ages from Table 1. Median ages are used by default to date a tree topology using DateLife, see Figure 3, step c3.
+
 Figure 1

### 2023-07-14 manuscript/manuscript-second-submission.Rmd
Second round of reviews

--- 2023-06-02 manuscript/manuscript.Rmd
@@
 authornote: |
-  School of Natural Sciences, University of California, Merced, Science and Engineering Building 1, Merced, CA 95340, USA.
+  Department of Life and Environmental Sciences, University of California, Merced, CA 95343, USA.
 
@@
 
-Here we present the DateLife project which has the main goal of extracting and exposing age data from published chronograms, making age data readily accessible to the wider community for reuse and reanalysis in research, teaching, science communication and conservation policy.
+Here we present the DateLife project which has the main goal of extracting and exposing age data from published chronograms, making age data readily accessible to a wider community for reuse and reanalysis in research, teaching, science communication and conservation policy.
 <!--  R has become mainstream in the biological-data analyst toolkit.
@@
 At the time of writing of this manuscript
-(May 15, 2023),
-<!-- (`r format(Sys.time(), "%b %d, %Y")`), -->
+(`r format(Sys.time(), "%b %d, %Y")`),
 DateLife's chronogram database latest version consist of `r format(length(opentree_chronograms$trees), big.mark=",")` chronograms published in `r length(unique(opentree_chronograms$studies))` different studies.
@@
 We recommend that users provide as input a tree topology from the literature, or one of their own making.
-If no topology is provided, DateLife automatically extracts one from the OpenTree synthetic tree, a phylogeny currently encompassing 2.3 million taxa across all life, assembled from 1, 239 published phylogenetic trees and OpenTree's unified Taxonomy, OTT [@opentreeoflife2019synth].
+If no topology is provided, DateLife automatically extracts one from the OpenTree synthetic tree, a phylogeny currently encompassing 2.3 million taxa across all life, assembled from 1,239 published phylogenetic trees and OpenTree's unified Taxonomy, OTT [@opentreeoflife2019synth].
 Alternatively, DateLife can combine topologies from source chronograms using a supertree approach [@Criscuolo2006].
@@
 
-**_Dating a tree with no branch lengths.--_**
+**_Dating a tree without branch lengths.--_**
 When producing or obtaining a tree with branch lengths for a group of interest is not possible, DateLife can date a topology without branch lengths, obtained from OpenTree or by implementing the supertree approach described above, by applying the Branch Length Adjuster [BLADJ; @Webb2008; @webb2005phylomatic] algorithm, which requires no initial branch lengths.
@@
 If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature.
-If none is provided, DateLife will not return a dated topology and will provide a warning message along with suggestions on how the user can provide an age for the root so that a fully dated tree can be returned.
+If none is provided, DateLife will choose an age for the root so that it can return a dated topology.
+It will also provide a conspicuous warning message so that users are aware that the root of the chronogram does not have information available in the chronogram database, along with suggestions on how the user can provide an appropriate age for the root.
 <!-- assigns an arbitrary age to the root as 10% older than the oldest age available within the group. -->
@@
 As a second example, we analysed 289 bird species in the family Fringillidae that are included in the NCBI taxonomy.
-The resulting summary chronogram is shown in Figure 5, and results from previous steps of the workflow are available as Supplementary Figures.
+The summary chronogram resulting from the DateLife analysis is shown in Figure 5, and results from previous steps of the workflow are available as Supplementary Figures.
 <!--We used `r cite_r("r-references.bib")` for all our analyses.-->
@@
 **_Summarizing search results.--_**
-DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. 3c), and congruified and mapped age data to nodes in this chosen topology (Table 1).
+DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. 3c), and congruified and mapped age data to nodes in this chosen topology, shown in Table 1.
 The name processing step allowed including five data points for node "n4" (parent of _Schoeniclus elegans_; Fig. 3A) that would not have had any data otherwise due to name mismatch.
@@
 As expected, more inclusive nodes (e.g., node "n1") have more variance in age data than less inclusive nodes (e.g., node "n5").
-Summary age data for node "n2" were excluded as final calibration because they are older than age data of the more inclusive node, "n1" (Fig. 3c4).
+Median summary age data for node "n2" was excluded as final calibration because it is older than the median age of a more inclusive node, "n1" (Fig. 3c4).
 
@@
 The congruent node age data points were summarized for each node, resulting in 194 summary node ages.
-From these 21 were excluded as secondary calibrations because they were older than the ancestral node and/or younger than a descendant node.
+From these 21 were excluded as secondary calibrations because they were older than the ancestral node.
 The remaining 173 summary node ages were used as secondary calibrations to obtain a fully dated (and resolved) phylogeny with the program BLADJ (Fig.
@@
 As we envision that DateLife will have many interesting applications in research and beyond, we emphasize that DateLife's results --as well as any insights gleaned from them, largely depend on the quality of the source chronograms: low quality chronograms will produce low quality results.
-This "garbage in, garbage out" problem has long been recognised in supertree methods for summarizing phylogenetic trees [@bininda2004garbage].
+The "garbage in, garbage out" problem has long been recognised in supertree methods for summarizing phylogenetic trees [@bininda2004garbage].
 We note that this is a surfacing issue of any automated tool for biological data analysis.
@@
 The DateLife project allows for easy and fast summarization of public and state-of-the-art data on time of lineage divergence.
-It provides a straightforward way to get an informed idea on the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identifying regions that require more research, or that have conflicting information.
 It is available as an R package, and as a web-based R shiny application at www.datelife.org.
+DateLife provides a straightforward way to get an informed picture of the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identifying regions that require more research, or that have conflicting information.
 <!-- It can also summarize public data, or use it to generate new chronograms based on trees provided by the user. -->
-Both summary and newly generated trees are useful to evaluate evolutionary hypotheses in different areas of research.
-The DateLife project will improve awareness of the existing variation in expert estimations time of divergence data, and foster exploration of the effect of alternative divergence time hypotheses on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
+Additionally, both summary and newly generated trees using the DateLife workflow are useful to evaluate evolutionary hypotheses in different areas of research.
+We hope that the DateLife project will increase awareness of the existing variation in expert estimations of time of divergence, and foster exploration of the effect of alternative divergence time hypotheses on the results of analyses, nurturing a culture of more cautious interpretation of evolutionary results.
 
@@
 
-Supplementary Figures can be viewed and dowloaded from their Zenodo repository (<https://doi.org/10.5281/zenodo.6683667>).
+Supplementary Figures can be viewed and downloaded from their Zenodo repository (<https://doi.org/10.5281/zenodo.6683667>).
 Supplementary material, including code, biological examples, benchmark results, data files and online-only appendices, can be downloaded from the Dryad data repository (<https://doi.org/10.5061/dryad.cnp5hqc6w>), as well as in the Zenodo stable repositories that host the reproducible manuscript (<https://doi.org/10.5281/zenodo.7435094>), the biological examples (<https://doi.org/10.5281/zenodo.7435101>), and the software benchmark (<https://doi.org/10.5281/zenodo.7435106>).

### 2023-07-14 review2.Rmd
Second round of reviews

--- 2023-02-21 review.Rmd
@@
 ---
-title: "Response to reviews"
-output: pdf_document
-date: '`r format(Sys.time(), "%b %d, %Y")`'
+title: "Response to decision letter"
+output: 
+  pdf_document:
+    keep_md: yes
+date: "2023-07-20"
 ---
@@
 
-> 19-Oct-2022
+**Dear _Editor-in-Chief_ Isabel Sanmartín,**
+
+&nbsp;
+&nbsp;
+
+**Please find below response to each point raised by you and the _Associate Editor_ Daniele Silvestro. We greatly appreciate your time and consideration for our resubmission.**
+
+&nbsp;
+&nbsp;
+
+**Kind regards,**
+
+&nbsp;
+
+**Luna L. Sanchez Reyes, on behalf of all coauthors.**
+
+&nbsp;
+&nbsp;
+
+> 30-Jun-2023
 
@@
 
-> Decision on USYB-2022-152, DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life:
+> Decision on USYB-2022-152.R1, DateLife: leveraging databases and analytical tools to reveal the dated Tree of Life:
 
@@
 
-> Thank you for your Systematic Biology submission.
-It has been reviewed by Associate Editor Dr Daniele Silvestro and two reviewers with the relevant expertise.
-Their comments are listed at the end of this letter.
-Both reviewers and the AE provide some excellent constructive suggestions that I am sure you will appreciate.
-The general consensus is that this software is of general interest to and expected to be widely used by the systematics and evolutionary biology community, providing capabilities that will help in the construction of supertrees.
-I agree with this assessment.
+> I thank the authors for their extensive and careful revisions to the manuscript.
+It has been reviewed by Associate Editor Dr Daniele Silvestro and myself.
+We both agree that the manuscript has been greatly improved and meets all the previous requests of the referees and the AE.
+The AE has several other suggestions that you should find useful while preparing the revision, mainly concerning editorial and formatting issues.
+I agree that some of your figures are difficult to visualize, for example Figs.
+1 and 4, where you could use less color and more simple lines (they resemble a poster).
+Figure 5 is too large and detailed: it could never be published as an A4.
+I also wonder if all figures are necessary: for example, Figures 5 and 6 could be moved to the Supplementary; there is no limit in the number of figures in Software papers but we typically publish 4 figures per paper.
+Please consider these recommendations as you revise the manuscript.
 
-> I think that your paper will be a valuable contribution to Systematic Biology once the detailed comments provided below are addressed, especially in relation to data curation and the English style.
-Thank you very much for your submission.
+**Thank you for your decision! We modified Figures 1 and 4.**
 
-> Sincerely,
-
-> Dr Isabel Sanmartín
-> Editor-in-Chief, Systematic Biology
-> isanmartin@rjb.csic.es
-
-> Please acknowledge (by email to sysbio.editorialoffice@oup.com) receipt of the reviews and give a probable time frame for the return of your revised manuscript.
+> Please acknowledge (by email to sysbio.editorialoffice@oup.com, SYSBIO.OUP.EA@straive.com) receipt of the reviews and give a probable time frame for the return of your revised manuscript.
 Your revision should be submitted as soon as possible.
 Any revision not received in a timely manner may have to be considered a new submission.
+
+**Noted and done.**
 
@@
 Delete the old files and replace them with your modified versions.
-Files that did not require revision can simply be retained
+Files that did not require revision can simply be retained.
+
+**Noted and done.**
 
 > You will be unable to make your revisions in ScholarOne Manuscripts. Instead, revise your paper using a word processing program and save it on your computer.
-HIGHLIGHT the changes to your manuscript within the document using the track changes mode in MS Word or by using bold, strikethrough, or colored text; any method you choose which indicates all changes made.
+> HIGHLIGHT the changes to your manuscript within the document using the track changes mode in MS Word or by using bold, strikethrough, or colored text; any method you choose which indicates all changes made.
 If a large number of changes were made and you feel the document is too cluttered to read easily with all of them shown, feel free to submit a "clean" copy as well, in which changes are not indicated.
 If you decide a clean copy would likely be helpful to reviewers and editors, please upload it in the category called Related Files.
+
+**Noted. We highlighted changes using latex and submitted the PDF file.**
 
@@
 
+**Noted and done. We submitted response to reviews as a PDF file**
+
 > Feel free to argue your case, with careful consideration and documentation, if you disagree with any of the suggestions.
@@
 
+**Noted.**
+
 > DO NOT submit your revision in .pdf format. This applies to the main text, tables, and appendices. Individual figures may be in .pdf format.
+
+**Noted.**
 
@@
 Upload the .pdf in the ScholarOne Manuscripts category called Related Files.
-
 > Figures must be uploaded separately, with each file name including the figure number. No figures may be imbedded in or tied to the .pdf or .tex files.
 
+**Noted and done.**
+
 > After you've uploaded your revision, carefully view the ScholarOne Manuscripts version to verify that all figures and other files display correctly, and that you’ve followed all of our author instructions.
+
+**Noted and checked.**
 
@@
 
+**Noted and done.**
+
 > The cost of printed color figures is $600 each.
@@
 
+**Noted.**
+
 > Journal Covers: Please consider submitting a suggested cover image.
@@
 
+**Noted.**
+
 > Data and Online Appendices: All such files should be deposited in the Dryad data repository.
@@
 Check over your online appendices or supplemental material (if any) carefully, because they will not be copyedited or proofread, and cannot be changed later.
-The first time you mention online-only material in the text of your paper, give the doi provided by Dryad as the location where the material can be found.
-Then, in a separate section after the main text, include a statement such as:
-
+> The first time you mention online-only material in the text of your paper, give the doi provided by Dryad as the location where the material can be found.
+> Then, in a separate section after the main text, include a statement such as:
 > SUPPLEMENTARY MATERIAL
@@
 
-We added DOIs for Dryad package with Supplementary Material, L551.
+**We added the DOI of our Dryad package containing all Supplementary Material on L690 (manuscript-differences.pdf).**
 
@@
 
-Noted and checked.
+**Noted, checked and done.**
 
@@
 
-Done on L556-558.
+**Done on L700 (manuscript-differences.pdf).**
 
@@
 
-This work was done while the first author was a Postdoctoral researcher.
+**This work was done while the first author was a Postdoctoral researcher.**
 
-> Please note that the accepted version of your manuscript will be published on Advance Access prior to typesetting and copyediting; it is therefore important that all files, including tables, figures, and supplementary files, are properly labelled and with the editorial office at this stage of the process -  it is the author’s responsibility to ensure that the correct files are submitted.
-Please also keep in mind that files will be published as submitted (ie, please review table formatting and labelling and position of figure images, etc).
-For further information please see the Advance Access Publication section at: https://academic.oup.com/sysbio/pages/General_Instructions.
+> I think that your paper will be a valuable contribution to Systematic Biology once the comments provided below are addressed. Thank you very much for your submission.
 
-Noted and checked.
+> Sincerely,
 
----
----
+> Dr Isabel Sanmartín
+> Editor-in-Chief, Systematic Biology
+> isanmartin@rjb.csic.es
+
+**Thanks so much for your decision! We made sure to address all comments carefully and hopefully you find the manuscript and figures improved.**
+
+***
 
@@
 
-> Recommendation #1: Accept with minor revisions
+> Associate Editor Recommendation: Accept with minor revisions
 
 > Comments to the Author:
-> Dear Dr Luna L Sanchez Reyes and co-authors,
+> Dear Dr Sanchez Reyes and co-authors,
 
-> many thanks for submitting your manuscript to Systematic Biology.
-Your study was reviewed by two highly qualified reviewers who provided an overall positive assessment of the paper while listing a number of things that should be revised and clarified.
-Based on their assessment and my own reading of the manuscript I invite you to resubmit it after carefully revising to address each and all points raised by the reviewers.
+> Many thanks for the thorough revisions of your paper, which I find nearly ready for acceptance and have decided not to send it back to the reviewers.
+I do have a few additional comments, which I would like you to address in a minor revision.
 
-Thanks for your reviews and comments! They were most helpful. We carefully revised the manuscript to address all points raised by the reviewers and yourself and we hope you find the revision suitable.
+Thanks you for your decision! We addressed all the following comments.
 
----
-> In addition to that, please also address the following points:
+> Lines 24-25 “A service […] that provide services” rephrase?
 
-> 1. how and why is a parsimony method used to estimate branch lengths and how is this used in combination with a likelihood method (line 174)?
+Rephrased to "a phylogenetic software that provides services"
 
-Thanks for this question. We elaborate on this on the Description section, subsection "Dating a tree with branch lengths" L221-228. 
-Briefly, we implement the parsimony algorithm ACCTRAN using functions from th R package phangorn.
-We chose this algorithm as a quick way to obtaining initial branch lengths that can be optimized under ML.
-ACCTRAN resolves ambiguous character optimization by assigning changes along branches of the tree as close to the root as possible (agnarsson2008acctran), so this works well with the type of matrices generated with BOLD data.
-Once estimated, the parsimony branch lengths become parameters that are then optimized using Maximum Likelihood, given the alignment, the topology and a simple JC model.
-<!--ACCTRAN has been implemented in Forest et al.
-2005.
-Teasing apart molecular-versus fossil-based error estimates when dating phylogenetic trees: a case study in the birch family Betulaceae.
-Systematic Botany.-->
-We also add a vignette to the R package in which we showcase the whole workflow of branch length optimization, from parsimony to ML.
+> Line 33: how is the topology chosen? User defined?
 
----
-> 2. Why are the node ages evenly distributed between calibrations? I would expect an exponential distribution of node ages under a standard birth-death process.
+By default it uses OpenTree's tree topology, ig the user does not provide a topology that they want.
+We specified this as "to corresponding nodes on a tree topology, either extracted from Open Tree of Life's synthetic phylogeny or one provided by the user"
 
-Thanks for pointing this out. We elaborate on this on the "Description" section, at the end of subsection "Dating a tree topology with no branch lengths", L 202-210
-(Note that we expanded the subsection "Dating a tree topology" into three sections: "Applying secondary calibrations", "Dating a tree topology" and "Dating a tree with branch lengths").
+> Line 40: I would rephrase to something like “Since the results of any […], we highlight that […]”
 
-Briefly, in the absence of genetic data from an alignment, nodes that have no age data from published studies in the chronogram database are assigned an age based solely on other nodes that do have age data from published studies.
-MrBayes has an option in which you can use a birth-death strict-clock model to sample branch lengths in the absence of genetic data. 
-We have implemented a DateLife function called "make_mrbayes_tree", that wraps this MrBayes service.
-Inherited from MrBayes implementation, the function requires setting parameter values for birth and death rates.
-If the amount of missing data is large, the chosen parameters can determine the "shape" of the resulting summary chronogram in terms of branch length distribution.
+Thanks for pointing this out, we rephrased to  "Because the results of any […], we highlight that […]"
 
-We added a discussion section, "Effects of phylogenetic sampling on downstream analyses" (L445 of manuscript without differences; L527 of manuscript with differences), in which we discuss the effects of chronograms with missing data that have been completed under a certain diversification model, especially when used for analyses that require estimating a diversification rate, which might introduce circularity (Rabosky, 2015.
-Evolution doi.org/10.1111/evo.12817).
+> Line 43: “DateLife can help […]”
 
-MrBayes also allow us to generate a cloud of trees produced with different birth-death rates that could be used for downstream analyses.
-However, the main goal of the summarizing step in DateLife is to provide a single chronogram that can quickly show in one glance the distribution of node ages based on published data, in the most agnostic way possible.
+Yes, we changed to "DateLife can help to increase [...]"
 
-This is why we chose an algorithm that distributes node ages evenly between calibrations and does not require any assumptions on the underlying model of branch length distribution.
-The Branch Length Adjuster (BLADJ) algorithm allows this and is the one set as default for the summarizing step.
-However, users that wish to use MrBayes instead con do so with the "make_mrbayes_tree" function.
+> Line 116: can you provide more information about the taxonomic scope of the available phylogenies?
+How many species are included?
+Can a researcher working on e.g.
+plants or invertebrates still use it?
+What happens if none of the user's species are found?
 
----
-> 4.
-I think the use of an arbitrary root age set by default as 10% older than the oldest age is unjustified and dangerous.
-If no root age is provided by the user, I think the function should return an interpretable error message and refuse to run.
+> Can a user choose to include a chronogram of their own? I think this would be useful if you want to plug a clade you are working on into a broader taxonomic scope.
 
-We agree, we changed the behaviour of the code to fix this, and adjusted the ms accordingly (L196-201).
+> Line 206-207: Equally spacing branches does not mean “not requiring any assumptions”.
+I think it’s OK to use it as default as it does not require prior estimates of speciation and extinction rates, but it is still making an assumption.
+The assumption translates into a possibly strange distribution of non-homogenous branching rates.
+I would focus on the fact that it does not require additional input (the rates) rather than on the assumption aspect.
 
-It now stops with an informative message on how users can provide an age for the root. The goal is to make sure that users are aware that there is no age data available for that root!
+Thanks for pointing it out! We changed it to:
 
----
-> Please make sure to carefully revise the text to remove typos.
+"BLADJ is an agnostic way to assign ages to nodes with no available genetic or branch length data, as it does not require initial diversification rates to draw from a branch length distribution"
 
-We revised the text throughout to remove all typos we could detect.
+An alternative to BLADJ to date tree topologies in the absence of initial branch lengths that is common practice in the literature is to use a birth-death model to draw branch lengths [@Jetz2012; @rabosky2018inverse; @smith2018constructing].
+In addition to the initial tree topology and nodes with age data, these methods require initial values of speciation and extinction rate parameters provided by the user.
+DateLife implements this approach with MrBayes [@Huelsenbeck2001; @Ronquist2003], using nodes with published age data as calibration priors on nodes of a tree topology with no branch lengths, a simple birth-death model with speciation and extinction rate parameters that are provided by the user, and no genetic data.
+However, BLADJ is the default option in DateLife, as it does not require any information on diversification rates for the phylogenetic sample to draw from a branch length distribution.
 
----
-> As one of the reviewers pointed out, it is good to provide links to permanent repositories for your code.
-I see that DateLife actually is already hosted in a Zenodo repository, so maybe you can add the link to your Availability section to make it more visible.
+> Line 217: I think you mean branches that are proportional to the expected number of substitutions per site, i.e. substitution rate multiplied by elapsed time.
 
-Right! We added Zenodo links for our code and other materials used for this research in the Availability section (L505) and Supplementary Material (L518-520).
+> Line 218: orthology hypothesis?
 
----
-> I hope you will be willing to revise and resubmit your paper and that you’ll find these and the Reviewers’ comments useful.
+> Lines 460-462: I don’t have strong opinion on this, but the statement seems unnecessary.
 
-> Best regards, Daniele Silvestro
+> I think the figures require some work.
+They look nice but are currently way too packed and I wonder if they need to include so much detail.
+I don’t know if anybody is still reading printed papers, but I doubt the figures would be readable on an A4.
+I recommend reducing the amount of text and substantially simplifying the graphics in general.
 
-Yes, thanks!
+> Fig. 1. I like the trees in the background, but they contribute to making the figure really busy. Also, there is a lot of text and tables, I think this should be strongly streamlined.
 
----
-> Reviewer(s)' comments to author:
+> Fig.
+3.
+Way too much text, tip labels not readable.
+In b2 I would reduce the trees to like 4 examples and make sure all the boxes have the same size and are aligned.
+Again, consider removing the trees in the background, especially in the colored boxes.
+In c) fonts in references are way too small, the numbers on the time axis are not explained.
+In a figure I would not expect to see full sentences like “We used median summary ages of congruent nodes to date […]”.
+Font types and sizes seem inconsistent.
 
-> Reviewer: 1
+> Arguably none of the text in Figs.
+4 and 5 is readable without a 200% magnification and the title of the plot is unnecessary.
+Tip labels should not be shown if they can’t be read, consider summarizing them by e.g.
+higher taxonomy.
 
-> Comments to the Author
-> The manuscript introduces DateLife, an R package and web service that provides time-calibrated phylogenies across a number of organismal groups.
-It integrates with a number of other services including the Open Tree of Life project.
-The authors provide some benchmarks and walk through a worked example of using their service.
+> I hope this helps and look forward to seeing your revised version of the paper.
 
-> I don't agree with the characterization that stochastic polytomy resolutions methods such as PASTIS (used in Jetz et al 2012) is "making up" these branch lengths (line 455).
-My understanding of the phrase "making up" implies that these are solely inventions of the researcher, rather than generated through the use of well-tested statistical models.
-The manuscript also claims that there are no thorough analyses of phylogenies generated in this way (lines 462-465), which is not the case, and I suggest the authors revise this section in light of some of the relevant literature in this area.
-
-> * Cusimano et al Syst Bio 2012
-> * Thomas et al MEE 2013
-> * Rabosky Evol 2015
-> * Chang et al Syst Bio 2019
-> * Title and Rabosky MEE 2019
-> * Sun et al AJB 2020
-
-Thanks for pointing us to these references.
-We elaborated on this subject more thoroughly by adding results from these studies to the corresponding section on the Discussion, "Effects of phylogenetic sampling on downstream analyses" (L481 of manuscript without differences; L527 of manuscript with differences), as well as changing the wording from "making up" to "in the absence of genetic data, simulating following a birth-death process" (L489; L540).
-
----
-> There were many typographical errors in the manuscript which should be corrected prior to publication.
-
-We corrected typos across the text.
-
----
-> Reviewer: 2
-
-> Comments to the Author
-> As I have checked the box that I don't need to remain anonymous, there is also no point in being mysterious about this: I have been aware of DateLife for a good long while because I've seen its earliest prototype develop at a workshop at NESCent ages ago.
-I've loved the idea ever since - combined with some healthy reservations that I am happy to share here.
-
-> DL synthesizes results from previous research.
-On the one hand that's great, but on the other, it invites the 'garbage in - garbage out' problem.
-Although OTOL has its own curation and QC facilities, the fact that users can provide their own garbage trees makes it so that the service might end up decorating nonsensical data, tainting its own reputation in the process.
-It would be good if the authors could emphasize this a bit more.
-
-This is a good point. We would not like for this to become a common use for the software. We mention this in the abstract, introduction and discussion.
-
----
-> A separate but related point that I would also like to see discussed is that synthesizing services such as DL and OTOL seem capable of ending up in loops where bad trees with bad calibration points provide the skeleton for further bad trees based on the former - with their own seemingly well-supported but in fact dodgy secondary calibrations.
-Is that a risk?
-What can be done about it?
-
-It is a risk that we are aware of, but it is possible to avoid it. We explain how in the text.
-
----
-> Also related: will we gradually start developing a body of literature with trees where the root always just happens to be ±10% older than the oldest nodes? Might that be bad?
-
-Good point.
-It would certainly be a bad thing.
-The reason we implemented this initially is because we wanted DateLife to be able to provide a summary chronogram even if tehre is just one point of age data available.
-
-As you pointed out, when the root age is absent, it is not possible to return any dated tree.
-
----
-> Apart from these general points that might be touched upon a bit more in the Discussion, here now some specifics about the manuscript:
-
-> - The Abstract looks like an extreme afterthought.
-I understand how that works, but please have another look.
-I see verb disagreement on line 21 and on line 23.
-Probably needs a comma after databases on line 25.
-On the same line, 'timeframe' is spelt as one word (fine by me), but elsewhere it's two words.
-Line 27: 'incetivizited' is not a thing.
-Line 29, 'finding' scans weird, maybe use 'discovery'?
-Line 36 probably needs 'use' instead of 'using' but the sentence is hard to parse.
-Line 38, 'awereness' is wrong.
-In this way, the Abstract is quite different from the rest of the MS, which is otherwise well written.
-
-Thanks for your comment, the abstract was indeed slightly forsaken on the previous version of the ms.
-We rewrote most of the abstract and checked for spelling.
-We hope you find it a good improvement from the original.
-
----
-> - In the first paragraph of the Intro you might want to add something like 'comparative analysis' (Harvey & Pagel, yada yada yada).
-It's clearly something that's on your mind because in the Conclusions, 'trait evolution' is the first research area you mention as needing chronograms.
-
-Thanks for pointing that out. We added that topic to the intro along with a couple of references, L48-49.
-
----
-> - On page 7, second paragraph, you state that subspecies are ignored.
-What do you mean precisely?
-My guess is that you ignore the subspecific epithet and collapse to species level.
-Maybe state that more clearly.
-
-Thanks for pointing this out.
-We actually only ignore subspecies when retrieving data from a more inclusive taxonomic group.
-When provided by the user, subspecific taxa are processed and searched as regular species.
-We clarify this in paragraph X, page 7
-
-It would be possible to drop the subspecific epithet and perform a more general search.
-We added this functionality to our software development plan, so that it can be soon implemented in a future iteration of the software.
-
----
-> - On page 7, third paragraph: how does the TNRS deal with homonyms?
-Given that we are in the tree realm it should be possible to infer intelligently whether some label is zoological or botanical code.
-Or is Aotus simply always the monkey, which is much cooler than that Australian legume genus?
-
-Yeah!
-this is a pretty cool quality of TNRS that we implement in datelife.
-It is possible to provide or identify the "biological context" of a toxonomic name, so that if a list of names has mainly monkeys, then it will pick _Aotus_ the monkey, but if it has mainly plants, it will pick the Australian legume.
-
-Moreover, homonyms in the Open Tree taxonomy specify the group they belong too, so users can easily identify if the numeric identifier they obtained from TNRS processing belongs to the group they wanted or not.
-For example _Aotus_ the legume is referred to as "Aotus (genus in kingdom Archaeplastida)" and the monkey as "Aotus (genus in Opisthokonta)".
-
-We added examples for the two points above in a new vignette for the package, available at http://phylotastic.org/datelife/articles/make_datelife_query.html
-
----
-
-> - On page 8, fourth paragraph, it's not quite clear whether DL's database syncs automatically with Phylesystem or whether you have volunteered yourself for this task.
-Which would be noble, but hard to sustain.
-
-It is indeed currently synced "manually". However we are working to have the syncs done automatically.
-We comment this in the discussion and clarified that it is currently a manual sync on page 8.
-
----
-> - On page 10, second paragraph: mining BOLD and aligning the sequences automatically is very cool functionality but I did not see it exposed on the website at all.
-How can users get at those alignments?
-Also, might there be performance issues?
-MAFFT can be quite greedy with larger data sets.
-
-Ah, yes, thanks!
-Well, the BOLD workflow is faster than a classic workflow, but it still takes considerable time, specifically computational time required for the database search, which the server can't afford yet.
-The functions are currently only available on DateLife's R package and not on the web application, but we plan on making them available there in the future.
-Regarding possible performance issues, the alignments from BOLD data are not very large (around 1000bp), so we expect both MUSCLE and MAFFT to perform approximately the same, and MUSCLE is the default aligner.
-
----
-> - On page 25 you mention the fossilcalibrations.org initiative.
-Maybe that's a good opportunity to go a bit into what we need as a community.
-I suspect that, in general, most people in this field think that doing it by themself is 'better', i.e.
-do a bunch of sequencing (hybseq right now, I guess?) and then get good primary calibration points.
-Natural history collections must have many more of those, both as fossils but also from geology (i.e.
-vicariant events having to do with tectonics, orogeny, etc.).
-Shouldn't we want *that*?
-
----
-> - Page 26, line 416 has some typos.
-
-Fixed :)
-
----
-> - On page 26 you discuss some criteria for scoring quality of chronograms.
-One additional criterion might be where the calibration points are placed.
-Nodes that have a calibration point between them and the root have less freedom of movement and hence narrower confidence intervals.
-Ages ago, I did a bit of simulation work on that (Vos & Mooers, 2004 - definitely no need to cite).
-Maybe someone else has discussed this a bit better?
-
----
-> - Page 27 line 448, chronogram should be plural, I think.
-
-Fixed!
-
----
-> - Page 28 line 473: I think it should be either 'public-funded' or 'publicly funded'
-
-Went for "publicly funded", thanks!
-
----
-> - Page 29, Supplementary Material: it's probably better to sync the repos with Zenodo and cite the DOI, just so that it's guaranteed unchanging.
-
-Thanks for the suggestion! We created stable versions for all three repositories on Zenodo, and refer the doi instead of the GitHub address for the repos.
+> Best regards,
+> Daniele

### 2023-08-09 review2.Rmd
streamline fig 1

@@
 The AE has several other suggestions that you should find useful while preparing the revision, mainly concerning editorial and formatting issues.
-I agree that some of your figures are difficult to visualize, for example Figs.
-1 and 4, where you could use less color and more simple lines (they resemble a poster).
+
+I agree that some of your figures are difficult to visualize, for example Figs. 1 and 4, where you could use less color and more simple lines (they resemble a poster).
+
+**Thank you for your decision! We modified Figures 1 and 4.**
+
 Figure 5 is too large and detailed: it could never be published as an A4. 
@@
 
-**Thank you for your decision! We modified Figures 1 and 4.**
+**We moved figures 5 and 6 to supplementary materials.**
 
@@
 
+We streamlined the figure by removing text in the figure and adding it to the figure caption.
+We considered removing the tables, but we decided to keep them as they are important for the explanation of the workflow.
+We hope this new version of the figure is considerably clearer now.
+
 > Fig.

### 2023-08-09 submission/final-captions.tex
streamline fig 1

@@
 
-Main DateLife workflow.
-Analyses can be performed via DateLife's interactive website at \url{www.datelife.org}, or using the \texttt{datelife} R package.
+Main DateLife workflow to obtain sceintific data of time of lineage divergence for a set of taxon names.
+a) The workflow starts with at least two taxon names provided by the user as a list or as tip names on a tree.
+Input names can contain synonyms and misspellings.
+b) Taxon names are processed using the Taxonomic Name Resolution Service (TNRS) and are standardized to the taxonomy of the chronogram database.
+In this mock example, 4 names (in bold) are synonyms in the standardized taxonomy.
+c) Processed taxon names are searched in the chronogram database.
+Chronograms that have at least two taxon names are identified (*), pruned down to matching taxon names only, and stored as source chronograms for the set of input taxon names.
+d) A topology to map ages from source chronograms can be provided by the user.
+If no topology is provided, OpenTree's synthetic tree topology is used.
+Alternative topologies that can be used are the largest (with the most input taxon names) source chronogram or a phylogeny constructed using genetic data from the Barcode of Life Database (BOLD).
+e) Nodes from source chronogram are congruified to nodes from the tree topology chosen in step (d).
+In this example all ages from source chronograms are congruent with nodes on the tree topology and provide time of divergenve information for all but one node (n2).
+f) Source node ages per node are summarized.
+g) Summary node ages are used as secondary calibrations to date the chosen tree topology. In this example we show the chronogram of median ages.
+All this analyses can be performed via DateLife's interactive website at \url{www.datelife.org}, or using the \texttt{datelife} R package.
 Details on the R functions used to perform the analyses are available from \texttt{datelife}'s R package vignettes at \url{https://phylotastic.org/datelife}.

### 2023-10-27 manuscript/manuscript.Rmd
some old changes!

@@
   Yet, achieving a high-quality reconstruction of a chronogram is a difficult and resource-consuming task.
-  Here we present DateLife, a service implemented as an R package and an R Shiny web application available at www.datelife.org, that provides services for efficient and easy discovery, summary, reuse, and reanalysis of node age data mined from a curated database of expert, peer-reviewed, and openly available chronograms.
+  Here we present DateLife, a phylogenetic software implemented as an R package and an R Shiny web application available at www.datelife.org, that provides services for efficient and easy discovery, summary, reuse, and reanalysis of node age data mined from a curated database of expert, peer-reviewed, and openly available chronograms.
     The main DateLife workflow starts with one or more scientific taxon names provided by a user.
 Names are processed and standardized to a unified taxonomy, allowing DateLife to run a name match across its local chronogram database that is curated from Open Tree of Life's phylogenetic repository, and extract all chronograms that contain at least two queried taxon names, along with their metadata.
-    Finally, node ages from matching chronograms are mapped to corresponding nodes from a chosen tree topology using the congruification algorithm.
+    Finally, node ages from matching chronograms are mapped using the congruification algorithm to corresponding nodes on a tree topology, either extracted from Open Tree of Life's synthetic phylogeny or one provided by the user.
 Congruified node ages are used as secondary calibrations to date the chosen topology, with or without initial branch lengths, using different phylogenetic dating methods such as BLADJ, treePL, PATHd8 and MrBayes.
     We performed a cross-validation test to compare node ages resulting from a DateLife analysis (i.e, phylogenetic dating using secondary calibrations) to those from the original chronograms (i.e, obtained with primary calibrations), and found that DateLife's node age estimates are consistent with the age estimates from the original chronograms, with the largest variation in ages occurring around topologically deeper nodes.
-  Results from any software for scientific analysis can only be as good as the data used as input, we highlight the importance of considering the results of a DateLife analysis in the context of the input chronograms.
-  We encourage the use of DateLife to help increase awareness of the existing disparities among alternative hypotheses of dates for the same diversification events, and to support exploration of the effect of alternative chronogram hypotheses on downstream analyses, providing a framework for a more informed interpretation of evolutionary results.
+    Because the results from any software for scientific analysis can only be as good as the data used as input, we highlight the importance of considering the results of a DateLife analysis in the context of the input chronograms.
+    DateLife can help to increase awareness of the existing disparities among alternative hypotheses of dates for the same diversification events, and to support exploration of the effect of alternative chronogram hypotheses on downstream analyses, providing a framework for a more informed interpretation of evolutionary results.
 
@@
 
-**_Dating a tree with no branch lengths.--_**
-When producing or obtaining a tree with branch lengths for a group of interest is not possible, DateLife can date a topology without branch lengths, obtained from OpenTree or by implementing the supertree approach described above, by applying the Branch Length Adjuster [BLADJ; @Webb2008; @webb2005phylomatic] algorithm, which requires no initial branch lengths.
-The algorithm starts by fixing ages for nodes with calibration data upon the given topology.
-Then, it distributes time for nodes with no data evenly between calibrated nodes, minimizing age variance in the resulting chronogram.
+**_Dating a tree without branch lengths.--_**
+To date a tree topology when branch lengths are unavailable, DateLife implements the Branch Length Adjuster (BLADJ) algorithm [@Webb2008; @webb2005phylomatic], which only requires a tree topology with no branch lengths and at least two node ages to use as calibrations, one for the tree root and one for any internal node of the topology.
+The BLADJ algorithm fixes ages for nodes with calibration data upon the given tree topology.
+Then, it assigns ages to nodes with no available age information by distributing time evenly between calibrated nodes, minimizing age variance in the resulting chronogram.
 This approach has proven useful for ecological analyses that require a phylogenetic time context [@Webb2008].
 When there is conflict between ages of calibrated nodes, BLADJ ignores node ages that are older than the age of a parent node.
-The BLADJ algorithm requires a root age estimate to run.
-If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature.
-If none is provided, DateLife will not return a dated topology and will provide a warning message along with suggestions on how the user can provide an age for the root so that a fully dated tree can be returned.
+The BLADJ algorithm requires a root age to run.
+Users can provide an appropriate root age estimate of their own or one obtained from the literature.
+If a root age is not provided and there is no information on the age of the root in the chronogram database, DateLife chooses a random age for the root, so that a dated tree topology can be generated with BLADJ.
+In this case, DateLife will provide a conspicuous warning message, so that users are aware that the root of the chronogram was chosen at random because there was no information available for it in the chronogram database, along with suggestions on how the user can find and provide an appropriate age for the root of the initial topology.
 <!-- assigns an arbitrary age to the root as 10% older than the oldest age available within the group. -->
 
-In the absence of genetic data, BLADJ is a very agnostic way to assign ages to nodes with no available data, as it does not require any assumptions on the underlying model of branch length distribution.
-It is however common practice in the literature to use a birth-death model to assign ages to nodes with no genetic data [@Jetz2012; @rabosky2018inverse; @smith2018constructing].
-To do so, DateLife implements MrBayes  [@Huelsenbeck2001; @Ronquist2003], using nodes with published age data as calibration priors on a fixed topology, a simple birth-death model with parameters that can be determined by the user, and no genetic data.
+An alternative to BLADJ to date tree topologies in the absence of initial branch lengths that is common practice in the literature is to use a birth-death model to draw branch lengths [@Jetz2012; @rabosky2018inverse; @smith2018constructing].
+In addition to the initial tree topology and nodes with age data, these methods require initial values of speciation and extinction rate parameters provided by the user.
+DateLife implements this approach with MrBayes [@Huelsenbeck2001; @Ronquist2003], using nodes with published age data as calibration priors on nodes of a tree topology with no branch lengths, a simple birth-death model with speciation and extinction rate parameters that are provided by the user, and no genetic data.
+However, BLADJ is the default option in DateLife, as it does not require any information on diversification rates for the phylogenetic sample to draw from a branch length distribution.
 
@@
 Relative branch lengths can provide key information for phylogenetic dating, specifically for nodes without any calibration data available.
-While using initial branch length data is the golden standard for phylogenetic dating analyses, estimating trees with branch lengths proportional to substitution rates per site requires obtaining primary data, assembling and curating a homology hypothesis, and choosing and implementing a method for phylogenetic inference.
+While using initial branch length data is the golden standard for phylogenetic dating analyses, estimating trees with branch lengths proportional to substitution rates per site requires obtaining primary data, assembling and curating a homology (orthology) hypothesis, and choosing and implementing a method for phylogenetic inference.
 DateLife implements a workflow to streamline this process by applying open data from the Barcode of Life Data System, BOLD [@ratnasingham2007bold] to obtain genetic markers for input taxa.

### 2023-10-27 manuscript/manuscript.tex
Rendered pdfs

@@
 Yet, achieving a high-quality reconstruction of a chronogram is a difficult and resource-consuming task.
-Here we present DateLife, a service implemented as an R package and an R Shiny web application available at www.datelife.org, that provides services for efficient and easy discovery, summary, reuse, and reanalysis of node age data mined from a curated database of expert, peer-reviewed, and openly available chronograms.
+Here we present DateLife, a phylogenetic software implemented as an R package and an R Shiny web application available at www.datelife.org, that provides services for efficient and easy discovery, summary, reuse, and reanalysis of node age data mined from a curated database of expert, peer-reviewed, and openly available chronograms.
 The main DateLife workflow starts with one or more scientific taxon names provided by a user.
 Names are processed and standardized to a unified taxonomy, allowing DateLife to run a name match across its local chronogram database that is curated from Open Tree of Life's phylogenetic repository, and extract all chronograms that contain at least two queried taxon names, along with their metadata.
-Finally, node ages from matching chronograms are mapped to corresponding nodes from a chosen tree topology using the congruification algorithm.
+Finally, node ages from matching chronograms are mapped using the congruification algorithm to corresponding nodes on a tree topology, either extracted from Open Tree of Life's synthetic phylogeny or one provided by the user.
 Congruified node ages are used as secondary calibrations to date the chosen topology, with or without initial branch lengths, using different phylogenetic dating methods such as BLADJ, treePL, PATHd8 and MrBayes.
 We performed a cross-validation test to compare node ages resulting from a DateLife analysis (i.e, phylogenetic dating using secondary calibrations) to those from the original chronograms (i.e, obtained with primary calibrations), and found that DateLife's node age estimates are consistent with the age estimates from the original chronograms, with the largest variation in ages occurring around topologically deeper nodes.
-Results from any software for scientific analysis can only be as good as the data used as input, we highlight the importance of considering the results of a DateLife analysis in the context of the input chronograms.
-We encourage the use of DateLife to help increase awareness of the existing disparities among alternative hypotheses of dates for the same diversification events, and to support exploration of the effect of alternative chronogram hypotheses on downstream analyses, providing a framework for a more informed interpretation of evolutionary results.
+Because the results from any software for scientific analysis can only be as good as the data used as input, we highlight the importance of considering the results of a DateLife analysis in the context of the input chronograms.
+DateLife can help to increase awareness of the existing disparities among alternative hypotheses of dates for the same diversification events, and to support exploration of the effect of alternative chronogram hypotheses on downstream analyses, providing a framework for a more informed interpretation of evolutionary results.
 
@@
 
-Word count: 6707
+Word count: 6797
 
@@
 At the time of writing of this manuscript
-(Jun 11, 2023),
+(Oct 27, 2023),
 DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies.
@@
 \textbf{\emph{Dating a tree without branch lengths.--}}
-When producing or obtaining a tree with branch lengths for a group of interest is not possible, DateLife can date a topology without branch lengths, obtained from OpenTree or by implementing the supertree approach described above, by applying the Branch Length Adjuster (BLADJ; Webb et al., 2008; Webb \& Donoghue, 2005) algorithm, which requires no initial branch lengths.
-The algorithm starts by fixing ages for nodes with calibration data upon the given topology.
-Then, it distributes time for nodes with no data evenly between calibrated nodes, minimizing age variance in the resulting chronogram.
+To date a tree topology when branch lengths are unavailable, DateLife implements the Branch Length Adjuster (BLADJ) algorithm (Webb et al., 2008; Webb \& Donoghue, 2005), which only requires a tree topology with no branch lengths and at least two node ages to use as calibrations, one for the tree root and one for any internal node of the topology.
+The BLADJ algorithm fixes ages for nodes with calibration data upon the given tree topology.
+Then, it assigns ages to nodes with no available age information by distributing time evenly between calibrated nodes, minimizing age variance in the resulting chronogram.
 This approach has proven useful for ecological analyses that require a phylogenetic time context (Webb et al., 2008).
 When there is conflict between ages of calibrated nodes, BLADJ ignores node ages that are older than the age of a parent node.
-The BLADJ algorithm requires a root age estimate to run.
-If there is no information on the age of the root in the chronogram database, users can provide an estimate from the literature.
-If none is provided, DateLife will choose an age for the root so that it can return a dated topology.
-It will also provide a conspicuous warning message so that users are aware that the root of the chronogram does not have information available in the chronogram database, along with suggestions on how the user can provide an appropriate age for the root.
-
-In the absence of genetic data, BLADJ is a very agnostic way to assign ages to nodes with no available data, as it does not require any assumptions on the underlying model of branch length distribution.
-It is however common practice in the literature to use a birth-death model to assign ages to nodes with no genetic data (Jetz, Thomas, Joy, Hartmann, \& Mooers, 2012; Rabosky et al., 2018; Smith \& Brown, 2018).
-To do so, DateLife implements MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), using nodes with published age data as calibration priors on a fixed topology, a simple birth-death model with parameters that can be determined by the user, and no genetic data.
+The BLADJ algorithm requires a root age to run.
+Users can provide an appropriate root age estimate of their own or one obtained from the literature.
+If a root age is not provided and there is no information on the age of the root in the chronogram database, DateLife chooses a random age for the root, so that a dated tree topology can be generated with BLADJ.
+In this case, DateLife will provide a conspicuous warning message, so that users are aware that the root of the chronogram was chosen at random because there was no information available for it in the chronogram database, along with suggestions on how the user can find and provide an appropriate age for the root of the initial topology.
+
+An alternative to BLADJ to date tree topologies in the absence of initial branch lengths that is common practice in the literature is to use a birth-death model to draw branch lengths (Jetz, Thomas, Joy, Hartmann, \& Mooers, 2012; Rabosky et al., 2018; Smith \& Brown, 2018).
+In addition to the initial tree topology and nodes with age data, these methods require initial values of speciation and extinction rate parameters provided by the user.
+DateLife implements this approach with MrBayes (Huelsenbeck \& Ronquist, 2001; Ronquist \& Huelsenbeck, 2003), using nodes with published age data as calibration priors on nodes of a tree topology with no branch lengths, a simple birth-death model with speciation and extinction rate parameters that are provided by the user, and no genetic data.
+However, BLADJ is the default option in DateLife, as it does not require any information on diversification rates for the phylogenetic sample to draw from a branch length distribution.
 
@@
 Relative branch lengths can provide key information for phylogenetic dating, specifically for nodes without any calibration data available.
-While using initial branch length data is the golden standard for phylogenetic dating analyses, estimating trees with branch lengths proportional to substitution rates per site requires obtaining primary data, assembling and curating a homology hypothesis, and choosing and implementing a method for phylogenetic inference.
+While using initial branch length data is the golden standard for phylogenetic dating analyses, estimating trees with branch lengths proportional to substitution rates per site requires obtaining primary data, assembling and curating a homology (orthology) hypothesis, and choosing and implementing a method for phylogenetic inference.
 DateLife implements a workflow to streamline this process by applying open data from the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain genetic markers for input taxa.

### 2023-10-27 manuscript/manuscript.tex
update figure 5 description

@@
 As a second example, we analysed 289 bird species in the family Fringillidae that are included in the NCBI taxonomy.
-The summary chronogram resulting from the DateLife analysis is shown in Figure 5, and results from previous steps of the workflow are available as Supplementary Figures.
+One clade from the full summary chronogram result from the DateLife analysis is shown figure 5.
+The full chronogram for all 289 species and the Fresults from previous steps of the workflow are available as Supplementary Figures.
 

### 2023-10-28 manuscript/manuscript.Rmd
update figures in submission folder

@@
 As a second example, we analysed 289 bird species in the family Fringillidae that are included in the NCBI taxonomy.
-The summary chronogram resulting from the DateLife analysis is shown in Figure 5, and results from previous steps of the workflow are available as Supplementary Figures.
+One clade from the full summary chronogram result from the DateLife analysis is shown figure 5.
+The full chronogram for all 289 species and the Fresults from previous steps of the workflow are available as Supplementary Figures.
 <!--We used `r cite_r("r-references.bib")` for all our analyses.-->

### 2023-10-31 dryad/README.Rmd
updating supplementray material and figure numbers in manuscript

@@
 	institution: University of Tennessee, Knoxville, USA
+
+	Co-researcher 3    
+	name: Emily Jane McTavish    
+	institution: University of California, Merced, USA
 
@@
 These data were generated to investigate and showcase the performance of the datelife R package (<https://github.com/phylotastic/datelife>).
-We showcased the application of the package with one mock example and two different biological examples.
+We showcased the application of the package with two different biological examples.
 
@@
 
-File 7 Name: Sanchez-Reyes_etal_2022_figure_1_chronogram_mock_example.tre    
-File 7 Description: Newick version of chronogram from the mock example shown in Figure 1 of the manuscript (https://www.biorxiv.org/content/10.1101/782094v2).
+File 7 Name: Sanchez-Reyes_etal_2022_figure_3_chronogram_small_example.tre    
+File 7 Description: Newick version of chronogram from the small example shown in Figure 3 of the manuscript (https://www.biorxiv.org/content/10.1101/782094v2).
 
-File 8 Name: Sanchez-Reyes_etal_2022_figure_3_chronogram_small_example.tre    
-File 8 Description: Newick version of chronogram from the small example shown in Figure 3 of the manuscript (https://www.biorxiv.org/content/10.1101/782094v2).
+File 8 Name: Sanchez-Reyes_etal_2022_figure_4A_topology_finches_mrca.tre    
+File 8 Description: Newick version of topology extracted from the Open Tree of Life synthetic tree, shown in Figure 4A of the manuscript (https://www.biorxiv.org/content/10.1101/782094v2).
 
-File 9 Name: Sanchez-Reyes_etal_2022_figure_4A_topology_finches_mrca.tre    
-File 9 Description: Newick version of topology extracted from the Open Tree of Life synthetic tree, shown in Figure 4A of the manuscript (https://www.biorxiv.org/content/10.1101/782094v2).
+File 9 Name: Sanchez-Reyes_etal_2022_figure_4B_topology_finches_ncbi.tre    
+File 9 Description: Newick version of topology extracted from the Open Tree of Life synthetic tree, shown in Figure 4B of the manuscript (https://www.biorxiv.org/content/10.1101/782094v2).
 
-File 10 Name: Sanchez-Reyes_etal_2022_figure_4B_topology_finches_ncbi.tre    
-File 10 Description: Newick version of topology extracted from the Open Tree of Life synthetic tree, shown in Figure 4B of the manuscript (https://www.biorxiv.org/content/10.1101/782094v2).
+File 10 Name: Sanchez-Reyes_etal_2022_figure_5_chronogram_finches_example.tre    
+File 10 Description: Newick version of chronogram from the finch example shown in Figure 5 of the manuscript (https://www.biorxiv.org/content/10.1101/782094v2).
 
-File 11 Name: Sanchez-Reyes_etal_2022_figure_5_chronogram_finches_example.tre    
-File 11 Description: Newick version of chronogram from the finch example shown in Figure 5 of the manuscript (https://www.biorxiv.org/content/10.1101/782094v2).
+File 11 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S1.pdf    
+File 11 Description: PDf file showing a topology extracted from OpenTree of Life's synthetic tree, of 2,333 tips and 1,305 internal nodes encompassing bird species within the family Fringillidae following the NCBI taxonomy (black), as well as all other bird species that share the same Most Recent Common Ancestor (MRCA) node in OpenTree's synthetic tree (purple).
 
-File 12 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S1.pdf    
-File 12 Description: PDf file showing results of cross validation analysis of datelife's chronogram generating functions. 
+File 12 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S2.pdf    
+File 12 Description: PDf file showing a subtree topology of 289 tips and 253 internal nodes resulting from pruning species from supplementray figure S1 (purple branches), that do not belong to the family Fringillidae according to the NCBI taxonomy.
+
+File 13 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S3.pdf    
+File 13 Description: PDf file showing the Fringillidae median summary chronogram resulting from a DateLife analysis.
+
+File 14 Name: Sanchez-Reyes_etal_2022_supplementary_figure_S4.pdf    
+File 14 Description: PDf file showing results of cross validation analysis of datelife's chronogram generating functions.
+
 "
@@
                       "Uyeda et al 2017")
-file_index <- 12
-fig_index <- 1
+file_index <- 15
+fig_index <- 5
 for (name in names_phylo_all) {

### 2023-10-31 manuscript/manuscript.Rmd
updating supplementray material and figure numbers in manuscript

--- 2023-10-27 manuscript/manuscript.Rmd
@@
 taxize [@Chamberlain2013; @Chamberlain2018].
-Figure 1 provides a graphical summary of the three main steps of the DateLife workflow: creating a search query, searching a database, and summarizing results from the search.
+There are three main steps to the DateLife workflow: 1) creating a search query, 2) searching a database, and 3) summarizing results from the search.
 
@@
 These matching pruned chronograms are referred to as source chronograms.
-Total distance in units of million years (Myr) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Fig. 1).
+Total distance in units of million years (Myr) between taxon pairs within each source chronogram are stored as a patristic distance matrix.
 The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a "phylo" object or newick string.
@@
 Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
-Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig. 2).
+Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig. 1).
 <!--Speed of other datelife workflows? NO, bc it reflects speed from other software mainly. But I will put a sentence about it...
@@
 We illustrate the DateLife workflow using a family within the passeriform birds encompassing the true finches, Fringillidae, as case study.
-On a small example, we analysed 6 bird species, and results from each step of the workflow are shown in Figure 3.
+On a small example, we analysed 6 bird species, and results from each step of the workflow are shown in Figure 2.
 As a second example, we analysed 289 bird species in the family Fringillidae that are included in the NCBI taxonomy.
-The summary chronogram resulting from the DateLife analysis is shown in Figure 5, and results from previous steps of the workflow are available as Supplementary Figures.
+One clade from the full summary chronogram result from the DateLife analysis is shown figure 3.
+The full chronogram for all 289 species and the results from previous steps of the workflow are available as Supplementary Figures.
 <!--We used `r cite_r("r-references.bib")` for all our analyses.-->
@@
 For a detailed discussion on the state of the synonym, refer to Avibase [@lepage2004avibase; @lepage2014avibase; @avibase-emberiza].
-Discovering this synonym allowed assigning five age data points for the parent node of *Emberiza elegans*, shown as *Schoeniclus elegans* in Figure 3a, which would not have had any data otherwise.
+Discovering this synonym allowed assigning five age data points for the parent node of *Emberiza elegans*, shown as *Schoeniclus elegans* in Figure 2, which would not have had any data otherwise.
 <!--\textcolor{red}{EJM: I don't understand what this sentence means}
@@
 DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms from 6 different studies (Fig.
-3b).
+2c).
 Three studies matched five input names [@barker2015new; @Hedges2015; @Jetz2012], one study matched four input names [@hooper2017chromosomal] and two studies matched two input names [@barker2013going; @burns2014phylogenetics].
@@
 **_Summarizing search results.--_**
-DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. 3c), and congruified and mapped age data to nodes in this chosen topology, shown in Table 1.
-The name processing step allowed including five data points for node "n4" (parent of _Schoeniclus elegans_; Fig. 3A) that would not have had any data otherwise due to name mismatch.
+DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. 2d), and congruified and mapped age data to nodes in this chosen topology, shown in Table 1.
+The name processing step allowed including five data points for node "n4" (parent of _Schoeniclus elegans_) that would not have had any data otherwise due to name mismatch.
 Age summary statistics per node were calculated (Table 2) and used as calibrations to date the tree topology using the BLADJ algorithm.
 As expected, more inclusive nodes (e.g., node "n1") have more variance in age data than less inclusive nodes (e.g., node "n5").
-Median summary age data for node "n2" was excluded as final calibration because it is older than the median age of a more inclusive node, "n1" (Fig. 3c4).
+Median summary age data for node "n2" was excluded as final calibration because it is older than the median age of a more inclusive node, "n1" (Fig. 2g).
 
@@
 By default, DateLife uses the topology from OpenTree's synthetic tree that contains all taxa from the search query.
-According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig.
-4a).
-Hence, a topology containing only the 289 species from the original query was extracted from Open Tree of Life’s synthetic tree v12.3 [Fig.
-4b; @opentreeoflife2019synth].
+According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Supplementary Fig.
+S1).
+Hence, a topology containing only the 289 species from the original query was extracted from Open Tree of Life’s synthetic tree v12.3 [Supplementary Fig.
+S2; @opentreeoflife2019synth].
 
@@
 
-Source chronograms (Supplementary Figs.
-S2-S20) were congruified to OpenTree's topology shown in Figure 4b, reducing the original 1,206 node age data set to 818 different data points (Supplementary Table S2) that can be used as calibrations for the chosen topology (Fig.
-4b).
+All `r length(dres)` source chronograms (Supplementary Figs.
+S5-S23) were congruified to OpenTree's topology shown in Supplementary Figure S2, reducing the original 1,206 node age data set to 818 different data points (Supplementary Table S2) that could be used as calibrations for the chosen topology (Supplementary Fig.
+S2).
 The congruent node age data points were summarized for each node, resulting in 194 summary node ages.
 From these 21 were excluded as secondary calibrations because they were older than the ancestral node.
-The remaining 173 summary node ages were used as secondary calibrations to obtain a fully dated (and resolved) phylogeny with the program BLADJ (Fig.
-5).
+The remaining 173 summary node ages were used as secondary calibrations to obtain a fully dated (and resolved) phylogeny with the program BLADJ (Supplementary Figure S3).
+Results for a subgroup are shown in Figure 3.
 
@@
 <!--Data from source chronograms can be used to date tree topologies with no branch lengths, as well as trees with branch lengths as relative substitution rates-->
-We performed a cross validation test of a DateLife analysis using the Fringillidae source chronograms obtained above (Supplementary Figs. S2-S20).
-We used as inputs for a DateLife analysis all individual tree topologies from each of the 19 source chronograms from 13 studies, treating their node ages as unknown.
+We performed a cross validation test of a DateLife analysis using the Fringillidae source chronograms obtained above (Supplementary Figs. S5-S23).
+As inputs for a DateLife analysis, we used all individual tree topologies from each of the `r length(dres)` source chronograms from `r length(unique(names(dres)))` studies, treating their node ages as unknown.
 We congruified node ages extracted from chronograms from all other studies upon the individual topologies, effectively excluding original ages from each topology.
 Finally, average node ages per node were applied as secondary calibrations and smoothed with the BLADJ algorithm.
-We found that node ages from the original studies, and ages estimated using all other age data available are largely correlated (Fig. 6).
+We found that node ages from the original studies, and ages estimated using all other age data available are largely correlated (Fig. 4).
 For five studies, DateLife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka 'closer to the root') relative to the original estimate, and overestimate ages for nodes closer to the tips.
 Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig.
-S1).
+S4).
 In general, topologically deeper nodes display the largest age variation between node ages from the original chronograms and ages summarized with DateLife.
@@
 Alternative source chronograms available for the same taxa have potentially been estimated implementing different types of calibrations, which affects the resulting node age estimates.
-For example, in the DateLife analysis of the Fringillidae shown above, the chronograms from one study [@burns2014phylogenetics] were inferred using molecular substitution rate estimates across birds [@weir2008calibrating], and have much older age estimates for the same nodes than chronograms that were inferred using fossil calibrations (Figs.
-5, 6; Supplementary Figs.
-S1c, S4).
+For example, in the DateLife analysis of the Fringillidae shown above, the chronograms from one study [@burns2014phylogenetics] were inferred using molecular substitution rate estimates across birds [@weir2008calibrating], and have much older age estimates for the same nodes than chronograms that were inferred using fossil calibrations (Fig.
+4; Supplementary Figs.
+S4c, S10).
 <!--\textcolor{red}{EJM: MORE ON the FINCH examples and the cross validation here}-->
@@
 Node ages summarized with DateLife and those from the original studies are well correlated (Supplementary Figs.
-S2-S20).
+S5-S23).
 We also note that DateLife estimates for nodes closer to the root tend to be slightly younger than ages from the original studies.
@@
 The only exception to this trend was observed in @burns2014phylogenetics chronogram, which generally displays much younger node ages when estimated using secondary calibrations than the original study (Supplementary Figs.
-S1, S5), supporting previous observations [@sauquet2012testing; @schenk2016sec].
+S4c, S10), supporting previous observations [@sauquet2012testing; @schenk2016sec].
 However, these younger dates are more likely an example of how multiple secondary calibrations can correct erroneous estimates, as dates on the @burns2014phylogenetics tree were obtained using a single secondary calibration based on a previously estimated molecular evolution rate across birds from @weir2008calibrating, and appear as major outliers compared to alternate estimates for the same nodes based on primary fossil calibrations (Fig.
-5).
+3, Supplementary Fig.
+S3).
 

### 2023-10-31 submission/captions-final.tex
updating supplementray material and figure numbers in manuscript

--- 2023-06-28 submission/final-captions.tex
@@
 
-Figure 1
-
-Main DateLife workflow.
-Analyses can be performed via DateLife's interactive website at \url{www.datelife.org}, or using the \texttt{datelife} R package.
-Details on the R functions used to perform the analyses are available from \texttt{datelife}'s R package vignettes at \url{https://phylotastic.org/datelife}.
-
-Figure 2:
+Figure 1:
 
@@
 
-Figure 3:
+Figure 2:
 
@@
 
+Figure 3:
+
+Subgroup of the Fringillidae median summary chronogram generated with DateLife.
+The original topology has 289 tips and 253 nodes, from which 194 have age data from at least one published chronogram.
+In total, 19 different chronograms from 13 different studies contributed 818 age data points, which were summarized to obtain a single value for each one of the 194 nodes with age data.
+From the 194 summary ages available, 21 were not used as calibrations (asterisk, *), because they were older than a parent node or younger than a descendant node; the remaining 173 summary ages were used as secondary calibrations (forward slash, /) with the Branch Length Adjuster (BLADJ) software from Webb et al., (\protect\hyperlink{ref-Webb2008}{2008}).
+The full figure is shown in Supplementary Figure S3.
+
 Figure 4:
 
-Tree topologies extracted from Open Tree of Life's (OpenTree) synthetic phylogenetic tree.
-Topology of 2,333 tips and 1,305 internal nodes (a), encompassing bird species within the family Fringillidae following the NCBI taxonomy (black), as well as all other bird species that share the same Most Recent Common Ancestor (MRCA) node in OpenTree's synthetic tree (purple).
-(Subtree topology of 289 tips and 253 internal nodes (b), resulting from pruning species that do not belong to the family Fringillidae according to the NCBI taxonomy (purple branches in topology a).
-Bird species within the Fringillidae are paraphyletic
-(Alström et al. \protect\hyperlink{ref-alstrom2014discovery}{2014},
-Barker, Cibois, Schikler, Feinstein, \& Cracraft \protect\hyperlink{ref-barker2004phylogeny}{2004},
-Barker et al. \protect\hyperlink{ref-barker2013going}{2013},
-Barker \protect\hyperlink{ref-barker2014mitogenomic}{2014},
-Barker et al. \protect\hyperlink{ref-barker2015new}{2015},
-Beresford, Barker, Ryan, \& Crowe \protect\hyperlink{ref-beresford2005african}{2005},
-Bryson Jr et al. \protect\hyperlink{ref-bryson2014diversification}{2014},
-Burleigh, Kimball, \& Braun \protect\hyperlink{ref-burleigh2015building}{2015},
-Burns et al. \protect\hyperlink{ref-burns2014phylogenetics}{2014},
-Chaves, Hidalgo, \& Klicka \protect\hyperlink{ref-chaves2013biogeography}{2013},
-Claramunt \& Cracraft \protect\hyperlink{ref-claramunt2015new}{2015},
-Gibb et al. \protect\hyperlink{ref-gibb2015new}{2015},
-Hackett et al. \protect\hyperlink{ref-hackett2008phylogenomic}{2008},
-Jetz et al. \protect\hyperlink{ref-Jetz2012}{2012},
-Johansson, Fjeldså, \& Bowi \protect\hyperlink{ref-johansson2008phylogenetic}{200},
-Kimball et al. \protect\hyperlink{ref-kimball2019phylogenomic}{2019},
-Klicka et al. \protect\hyperlink{ref-klicka2014comprehensive}{2014},
-Lamichhaney et al. \protect\hyperlink{ref-lamichhaney2015evolution}{2015},
-Lerner, Meyer, James, Hofreiter, \& Fleischer \protect\hyperlink{ref-lerner2011multilocus}{2011},
-Lovette et al. \protect\hyperlink{ref-lovette2010comprehensive}{2010},
-Moyle et al. \protect\hyperlink{ref-moyle2016tectonic}{2016},
-Ödeen, Håstad, \& Alström \protect\hyperlink{ref-odeen2011evolution}{2011},
-Oliveros et al. \protect\hyperlink{ref-oliveros2019earth}{2019},
-Päckert et al. \protect\hyperlink{ref-packert2012horizontal}{2012},
-Parchman, Benkman, \& Mezquida \protect\hyperlink{ref-parchman2007coevolution}{2007},
-Powell et al. \protect\hyperlink{ref-powell2014comprehensive}{2014},
-Price et al. \protect\hyperlink{ref-price2014niche}{2014},
-Pulgarín-R, Smith, Bryson Jr, Spellman, \& Klicka \protect\hyperlink{ref-pulgarin2013multilocus}{2013},
-Selvatti, Gonzaga, \& Moraes Russo \protect\hyperlink{ref-selvatti2015paleogene}{2015},
-Tietze, Päckert, Martens, Lehmann, \& Sun \protect\hyperlink{ref-tietze2013complete}{2013},
-Treplin et al. \protect\hyperlink{ref-treplin2008molecular}{2008},
-Zuccon, Prŷs-Jones, Rasmussen, \& Ericson \protect\hyperlink{ref-zuccon2012phylogenetic}{2012}).
-
-Figure 5:
-
-Fringillidae median summary chronogram generated with DateLife.
-The original topology has 289 tips and 253 nodes, from which 194 have age data from at least one published chronogram.
-In total, 19 different chronograms from 13 different studies contribute 818 age data points, which were summarized to obtain a single value for each one of the 194 nodes with age data.
-From the 194 summary ages available, 21 were not used as calibrations (asterisk, *), because they were older than a parent node or younger than a descendant node; the remaining 173 summary ages were used as secondary calibrations (forward slash, /) with the Branch Length Adjuster (BLADJ) software from Webb et al., (\protect\hyperlink{ref-Webb2008}{2008}).
-
-Figure 6:
-
 Cross validation of results from a DateLife analysis of the family Fringillidae, shown in Fig.
-5.
-Each plot compares node age estimate from an original study chronogram (x axis) with the corresponding node age obtained with a DateLife analysis (y axis).
+3 and Supplementary Figure S3.
+Each plot compares node age estimates from an original study chronogram (x axis) with the corresponding node age obtained with a DateLife analysis (y axis).

### 2023-11-11 manuscript/manuscript.Rmd
Edits from second review.

@@
 (`r format(Sys.time(), "%b %d, %Y")`),
-DateLife's chronogram database latest version consist of `r format(length(opentree_chronograms$trees), big.mark=",")` chronograms published in `r length(unique(opentree_chronograms$studies))` different studies.
-It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata [@mctavish2015phylesystem], which allows automatic and reproducible assembly of our chronogram database.
-Datelife's chronogram database is navigable as an R data object within the `datelife` R package.
+DateLife's chronogram database latest version consist of `r format(length(opentree_chronograms$trees), big.mark=",")` chronograms published in `r length(unique(opentree_chronograms$studies))` different studies, encompassing close to 100 k (99 474) different species.
+It is curated from OpenTree's phylogenetic database, the Phylesystem, an open database of expert and peer-reviewed phylogenetic knowledge with rich metadata and a wide taxonomic scope [@mctavish2015phylesystem].
+We expect DateLife's database to largely overlap with OpenTree's phylogenetic database taxonomic coverage, where Chordata and Embryophyta are nearly fully sampled.
+In contrast, Bacteria, Fungi, Nematoda, and Insecta, currently present a large gap between the number of named species and what has phylogenetic information in OpenTree’s synthetic tree.
+It is likely that users working with the former groups will get results from a DateLife analysis.
+If none of the user’s species are found, the software will indicate the lack of age data for the queried taxa in the database.
 
 A unique feature of the Phylesystem is that any user can add new published, state-of-the-art chronograms any time, through OpenTree's curator application (https://tree.opentreeoflife.org/curator).
-As chronograms are added to Phylesystem, they can be incorporated into the chronogram database of the `datelife` R package, which is currently manually updated as new chronogram data is added to Phylesystem.
-The updated database is assigned a new version number, followed by a package release on CRAN.
-Users can directly run `datelife` functions to trigger an update of their local chronogram database, to incorporate any new chronograms to their DateLife analysis before an official database update is released on CRAN.
+Relying on an open source database permits an automatic and reproducible assembly of DateLife's chronogram database, which is stored and navigable as an R data object within the `datelife` R package.
+As chronograms are added to Phylesystem, they can be incorporated into the chronogram database within the `datelife` R package, by manually triggering an update.
+The updated `datelife` database is assigned a new version number, followed by a package release on CRAN.
+We encourage users to submit published chronograms to OpenTree's phylogenetic database, so that their taxon of interest can be included in future DateLife searches.
+Users can directly run `datelife` functions to trigger an update of their local chronogram database, to incorporate any new chronograms to their DateLife analysis before a `datelife` database update is released on CRAN.
 <!--BCO: What is immediately?
@@
 
-Summary information is extracted from the \texttt{datelifeResult} object to inform decisions for subsequent steps in the analysis workflow. Basic summary information available to the user is:
+Summary information is extracted from the \texttt{datelifeResult} object to inform decisions for subsequent steps in the analysis workflow. Basic summary information available to the user includes:
 
@@
 By default, DateLife chooses the grove with the most taxa, however, the "criterion = trees" flag allows the user to choose the grove with the most trees instead.
-The result is a single summary (or supertree) topology, that combines topologies from source chronograms in a grove.
+The result is a single summary (i.e., supertree) topology, that combines topologies from source chronograms in a grove.
 
@@
 **_Dating a tree without branch lengths.--_**
-To date a tree topology when branch lengths are unavailable, DateLife implements the Branch Length Adjuster (BLADJ) algorithm [@Webb2008; @webb2005phylomatic], which only requires a tree topology with no branch lengths and at least two node ages to use as calibrations, one for the tree root and one for any internal node of the topology.
+To date a tree topology when initial branch lengths are unavailable, DateLife implements the Branch Length Adjuster (BLADJ) algorithm [@Webb2008; @webb2005phylomatic], which only requires a tree topology with no branch lengths and at least two node ages to use as calibrations, one for the tree root and one for any internal node of the topology.
 The BLADJ algorithm fixes ages for nodes with calibration data upon the given tree topology.
@@
 Relative branch lengths can provide key information for phylogenetic dating, specifically for nodes without any calibration data available.
-While using initial branch length data is the golden standard for phylogenetic dating analyses, estimating trees with branch lengths proportional to substitution rates per site requires obtaining primary data, assembling and curating a homology (orthology) hypothesis, and choosing and implementing a method for phylogenetic inference.
+While using initial branch length data is the golden standard for phylogenetic dating analyses, obtaining such information from scratch is not an easy task: it requires obtaining primary data, assembling and curating a homology (orthology) hypothesis, and choosing and implementing a method for phylogenetic inference.
 DateLife implements a workflow to streamline this process by applying open data from the Barcode of Life Data System, BOLD [@ratnasingham2007bold] to obtain genetic markers for input taxa.
@@
 As a second example, we analysed 289 bird species in the family Fringillidae that are included in the NCBI taxonomy.
-One clade from the full summary chronogram result from the DateLife analysis is shown figure 3.
+One clade from the full summary chronogram result from the DateLife analysis is shown Figure 3.
 The full chronogram for all 289 species and the results from previous steps of the workflow are available as Supplementary Figures.
@@
 Finally, average node ages per node were applied as secondary calibrations and smoothed with the BLADJ algorithm.
-We found that node ages from the original studies, and ages estimated using all other age data available are largely correlated (Fig. 4).
+We found that node ages from the original studies, and ages estimated using all other age data available are generally correlated (Fig. 4).
 For five studies, DateLife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka 'closer to the root') relative to the original estimate, and overestimate ages for nodes closer to the tips.
@@
 
-Further research is needed to fully understand the effects of using secondary calibrations and the use of resulting chronograms in downstream analyses [@shaul2002playing; @schenk2016sec; @hipsley2014beyond; @powell2020quantifying].
+<!-- Further research is needed to fully understand the effects of using secondary calibrations and the use of resulting chronograms in downstream analyses [@shaul2002playing; @schenk2016sec; @hipsley2014beyond; @powell2020quantifying].
+-->
 
@@
 It is available as an R package, and as a web-based R shiny application at www.datelife.org.
-DateLife provides a straightforward way to get an informed picture of the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identifying regions that require more research, or that have conflicting information.
+DateLife provides a straightforward way to get an informed picture of the state of knowledge for the time frame of evolution of different regions of the tree of life, and allows identifying regions that require more research, or that have conflicting information.
 <!-- It can also summarize public data, or use it to generate new chronograms based on trees provided by the user. -->

### 2023-11-11 manuscript/manuscript.tex
Edits from second review.

--- 2023-10-27 manuscript/manuscript.tex
@@
 
-Word count: 6797
+Word count: 6946
 
@@
 taxize (Chamberlain, 2018; Chamberlain \& Szöcs, 2013).
-Figure 1 provides a graphical summary of the three main steps of the DateLife workflow: creating a search query, searching a database, and summarizing results from the search.
+There are three main steps to the DateLife workflow: 1) creating a search query, 2) searching a database, and 3) summarizing results from the search.
 
@@
 At the time of writing of this manuscript
-(Oct 27, 2023),
-DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies.
-It is curated from OpenTree's phylogenetic database, the Phylesystem, which constitutes an open source of expert and peer-reviewed phylogenetic knowledge with rich metadata (McTavish et al., 2015), which allows automatic and reproducible assembly of our chronogram database.
-Datelife's chronogram database is navigable as an R data object within the \texttt{datelife} R package.
+(Nov 07, 2023),
+DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies, encompassing close to 100 k (99 474) different species.
+It is curated from OpenTree's phylogenetic database, the Phylesystem, an open database of expert and peer-reviewed phylogenetic knowledge with rich metadata and a wide taxonomic scope (McTavish et al., 2015).
+We expect DateLife's database to largely overlap with OpenTree's phylogenetic database taxonomic coverage, where Chordata and Embryophyta are nearly fully sampled.
+In contrast, Bacteria, Fungi, Nematoda, and Insecta, currently present a large gap between the number of named species and what has phylogenetic information in OpenTree's synthetic tree.
+It is likely that users working with the former groups will get results from a DateLife analysis.
+If none of the user's species are found, the software will indicate the lack of age data for the queried taxa in the database.
 
 A unique feature of the Phylesystem is that any user can add new published, state-of-the-art chronograms any time, through OpenTree's curator application (\url{https://tree.opentreeoflife.org/curator}).
-As chronograms are added to Phylesystem, they can be incorporated into the chronogram database of the \texttt{datelife} R package, which is currently manually updated as new chronogram data is added to Phylesystem.
-The updated database is assigned a new version number, followed by a package release on CRAN.
-Users can directly run \texttt{datelife} functions to trigger an update of their local chronogram database, to incorporate any new chronograms to their DateLife analysis before an official database update is released on CRAN.
+Relying on an open source database permits an automatic and reproducible assembly of DateLife's chronogram database, which is stored and navigable as an R data object within the \texttt{datelife} R package.
+As chronograms are added to Phylesystem, they can be incorporated into the chronogram database within the \texttt{datelife} R package, by manually triggering an update.
+The updated \texttt{datelife} database is assigned a new version number, followed by a package release on CRAN.
+We encourage users to submit published chronograms to OpenTree's phylogenetic database, so that their taxon of interest can be included in future DateLife searches.
+Users can directly run \texttt{datelife} functions to trigger an update of their local chronogram database, to incorporate any new chronograms to their DateLife analysis before a \texttt{datelife} database update is released on CRAN.
 
@@
 These matching pruned chronograms are referred to as source chronograms.
-Total distance in units of million years (Myr) between taxon pairs within each source chronogram are stored as a patristic distance matrix (Fig. 1).
+Total distance in units of million years (Myr) between taxon pairs within each source chronogram are stored as a patristic distance matrix.
 The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a \enquote{phylo} object or newick string.
@@
 By default, DateLife chooses the grove with the most taxa, however, the \enquote{criterion = trees} flag allows the user to choose the grove with the most trees instead.
-The result is a single summary (or supertree) topology, that combines topologies from source chronograms in a grove.
+The result is a single summary (i.e., supertree) topology, that combines topologies from source chronograms in a grove.
 
@@
 \textbf{\emph{Dating a tree without branch lengths.--}}
-To date a tree topology when branch lengths are unavailable, DateLife implements the Branch Length Adjuster (BLADJ) algorithm (Webb et al., 2008; Webb \& Donoghue, 2005), which only requires a tree topology with no branch lengths and at least two node ages to use as calibrations, one for the tree root and one for any internal node of the topology.
+To date a tree topology when initial branch lengths are unavailable, DateLife implements the Branch Length Adjuster (BLADJ) algorithm (Webb et al., 2008; Webb \& Donoghue, 2005), which only requires a tree topology with no branch lengths and at least two node ages to use as calibrations, one for the tree root and one for any internal node of the topology.
 The BLADJ algorithm fixes ages for nodes with calibration data upon the given tree topology.
@@
 Up to ten thousand names can be processed and searched in less than 30 minutes with the most time consuming settings.
-Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig. 2).
+Once names have been processed as described in methods, a name search through the chronogram database can be performed in less than a minute, even with a very large number of taxon names (Fig. 1).
 
@@
 We illustrate the DateLife workflow using a family within the passeriform birds encompassing the true finches, Fringillidae, as case study.
-On a small example, we analysed 6 bird species, and results from each step of the workflow are shown in Figure 3.
+On a small example, we analysed 6 bird species, and results from each step of the workflow are shown in Figure 2.
 As a second example, we analysed 289 bird species in the family Fringillidae that are included in the NCBI taxonomy.
-The summary chronogram resulting from the DateLife analysis is shown in Figure 5, and results from previous steps of the workflow are available as Supplementary Figures.
+One clade from the full summary chronogram result from the DateLife analysis is shown Figure 3.
+The full chronogram for all 289 species and the results from previous steps of the workflow are available as Supplementary Figures.
 
@@
 For a detailed discussion on the state of the synonym, refer to Avibase (Avibase, 2022; Lepage, 2004; Lepage, Vaidya, \& Guralnick, 2014).
-Discovering this synonym allowed assigning five age data points for the parent node of \emph{Emberiza elegans}, shown as \emph{Schoeniclus elegans} in Figure 3a, which would not have had any data otherwise.
+Discovering this synonym allowed assigning five age data points for the parent node of \emph{Emberiza elegans}, shown as \emph{Schoeniclus elegans} in Figure 2, which would not have had any data otherwise.
 
@@
 DateLife used the processed input names to search the local chronogram database and found 9 matching chronograms from 6 different studies (Fig.
-3b).
+2c).
 Three studies matched five input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2015; Hedges, Marin, Suleski, Paymer, \& Kumar, 2015; Jetz et al., 2012), one study matched four input names (Hooper \& Price, 2017) and two studies matched two input names (Barker, Burns, Klicka, Lanyon, \& Lovette, 2013; Burns et al., 2014).
@@
 \textbf{\emph{Summarizing search results.--}}
-DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. 3c), and congruified and mapped age data to nodes in this chosen topology, shown in Table 1.
-The name processing step allowed including five data points for node \enquote{n4} (parent of \emph{Schoeniclus elegans}; Fig. 3A) that would not have had any data otherwise due to name mismatch.
+DateLife obtained OpenTree's synthetic tree topology for these taxa (Fig. 2d), and congruified and mapped age data to nodes in this chosen topology, shown in Table 1.
+The name processing step allowed including five data points for node \enquote{n4} (parent of \emph{Schoeniclus elegans}) that would not have had any data otherwise due to name mismatch.
 Age summary statistics per node were calculated (Table 2) and used as calibrations to date the tree topology using the BLADJ algorithm.
 As expected, more inclusive nodes (e.g., node \enquote{n1}) have more variance in age data than less inclusive nodes (e.g., node \enquote{n5}).
-Median summary age data for node \enquote{n2} was excluded as final calibration because it is older than the median age of a more inclusive node, \enquote{n1} (Fig. 3c4).
+Median summary age data for node \enquote{n2} was excluded as final calibration because it is older than the median age of a more inclusive node, \enquote{n1} (Fig. 2g).
 
@@
 By default, DateLife uses the topology from OpenTree's synthetic tree that contains all taxa from the search query.
-According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Fig.
-4a).
-Hence, a topology containing only the 289 species from the original query was extracted from Open Tree of Life's synthetic tree v12.3 (Fig.
-4b; Open Tree Of Life et al., 2019).
-
-Source chronograms (Supplementary Figs.
-S2-S20) were congruified to OpenTree's topology shown in Figure 4b, reducing the original 1,206 node age data set to 818 different data points (Supplementary Table S2) that can be used as calibrations for the chosen topology (Fig.
-4b).
+According to OpenTree's synthetic tree, species belonging to the family Fringillidae do not form a monophyletic group (Supplementary Fig.
+S1).
+Hence, a topology containing only the 289 species from the original query was extracted from Open Tree of Life's synthetic tree v12.3 (Supplementary Fig.
+S2; Open Tree Of Life et al., 2019).
+
+All 19 source chronograms (Supplementary Figs.
+S5-S23) were congruified to OpenTree's topology shown in Supplementary Figure S2, reducing the original 1,206 node age data set to 818 different data points (Supplementary Table S2) that could be used as calibrations for the chosen topology (Supplementary Fig.
+S2).
 The congruent node age data points were summarized for each node, resulting in 194 summary node ages.
 From these 21 were excluded as secondary calibrations because they were older than the ancestral node.
-The remaining 173 summary node ages were used as secondary calibrations to obtain a fully dated (and resolved) phylogeny with the program BLADJ (Fig.
-5).
+The remaining 173 summary node ages were used as secondary calibrations to obtain a fully dated (and resolved) phylogeny with the program BLADJ (Supplementary Figure S3).
+Results for a subgroup are shown in Figure 3.
 
@@
 
-We performed a cross validation test of a DateLife analysis using the Fringillidae source chronograms obtained above (Supplementary Figs. S2-S20).
-We used as inputs for a DateLife analysis all individual tree topologies from each of the 19 source chronograms from 13 studies, treating their node ages as unknown.
+We performed a cross validation test of a DateLife analysis using the Fringillidae source chronograms obtained above (Supplementary Figs. S5-S23).
+As inputs for a DateLife analysis, we used all individual tree topologies from each of the 19 source chronograms from 13 studies, treating their node ages as unknown.
 We congruified node ages extracted from chronograms from all other studies upon the individual topologies, effectively excluding original ages from each topology.
 Finally, average node ages per node were applied as secondary calibrations and smoothed with the BLADJ algorithm.
-We found that node ages from the original studies, and ages estimated using all other age data available are largely correlated (Fig. 6).
+We found that node ages from the original studies, and ages estimated using all other age data available are generally correlated (Fig. 4).
 For five studies, DateLife tended to underestimate ages for topologically deeper nodes (those with many descendant taxa, aka \enquote{closer to the root}) relative to the original estimate, and overestimate ages for nodes closer to the tips.
 Accordingly, root ages are generally older in the original study than estimated using cross-validated ages (Supplementary Fig.
-S1).
+S4).
 In general, topologically deeper nodes display the largest age variation between node ages from the original chronograms and ages summarized with DateLife.
@@
 Alternative source chronograms available for the same taxa have potentially been estimated implementing different types of calibrations, which affects the resulting node age estimates.
-For example, in the DateLife analysis of the Fringillidae shown above, the chronograms from one study (Burns et al., 2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older age estimates for the same nodes than chronograms that were inferred using fossil calibrations (Figs.
-5, 6; Supplementary Figs.
-S1c, S4).
+For example, in the DateLife analysis of the Fringillidae shown above, the chronograms from one study (Burns et al., 2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older age estimates for the same nodes than chronograms that were inferred using fossil calibrations (Fig.
+4; Supplementary Figs.
+S4c, S10).
 Another source of conflict in estimated node ages can arise from different placements for the same calibration, which would imply fundamentally distinct evolutionary hypotheses (Antonelli et al., 2017).
@@
 (2014) chronogram, which generally displays much younger node ages when estimated using secondary calibrations than the original study (Supplementary Figs.
-S1, S5), supporting previous observations (Sauquet et al., 2012; Schenk, 2016).
+S4c, S10), supporting previous observations (Sauquet et al., 2012; Schenk, 2016).
 However, these younger dates are more likely an example of how multiple secondary calibrations can correct erroneous estimates, as dates on the Burns et al.
 (2014) tree were obtained using a single secondary calibration based on a previously estimated molecular evolution rate across birds from Weir and Schluter (2008), and appear as major outliers compared to alternate estimates for the same nodes based on primary fossil calibrations (Fig.
-5).
+3, Supplementary Fig.
+S3).
 

### 2023-11-11 submission/captions-final.tex
Edits from second review.

@@
 
-DateLife's benchmarking results.
-Computation time used to process a query and a search across \texttt{datelife}'s chronogram database, relative to number of input taxon names.
+DateLife's benchmarking results showing computation time used for taxon name processing and search across \texttt{datelife}'s chronogram database, as a function of number of input taxon names (N).
 For each N = \{10, 100, 200, ..., 1 000, ...
-, 9 000, 10 000\}, we sampled N species names from the class Aves a hundred times, and then performed a \texttt{datelife} search processing the input names using the Taxon Names Resolution Service (TNRS; light gray), and without processing names (dark gray).
-For comparison, we performed a search using an input that has been pre-processed with TNRS (light blue).
+, 9 000, 10 000\}, we randomly sampled N species names from the class Aves, a hundred times, and then performed a \texttt{datelife} search processing the input names using the Taxon Names Resolution Service (TNRS; light gray), and without processing input names (dark gray).
+For comparison, we performed a chronogram search using names that have been pre-processed with TNRS (light blue).
 
@@
 
-DateLife analysis results for a small sample of 6 bird species within the Passeriformes (a).
-Processed species names were found across 9 chronograms within 6 independent studies (b; Barker et al.
+DateLife results of an analysis of a small sample of 6 bird species within the Passeriformes (a, b).
+Processed species names were found across 9 chronograms within 6 independent studies (c; Barker et al.
 (\protect\hyperlink{ref-barker2012going}{2012}), Barker et al.
@@
 (\protect\hyperlink{ref-Jetz2012}{2012}).
-This revealed 28 source age data points for the queried species names.
-Summarized age data was used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree v13.4, resulting in a chronogram of summary source ages (c).
+This revealed 28 source age data points for the queried species names (e; Table 1).
+Summarized age data (f; Table 2) was used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree v13.4 (d), resulting in the chronogram of summary source ages shown in (g).
 
@@
 
-Subgroup of the Fringillidae median summary chronogram generated with DateLife.
-The original topology has 289 tips and 253 nodes, from which 194 have age data from at least one published chronogram.
+Median summary chronogram resulting from a Datelife analysis of bird species within the family Fringillidae.
+For visualization purposes, we are showing a portion of the final median summary chronogram encompassing 57 species out of the 289 total included in the analysis.
+The complete chronogram is available as Supplementary Figure S3.
+The working tree topology has 289 tip species and 253 nodes, from which 194 have age data from at least one published chronogram.
 In total, 19 different chronograms from 13 different studies contributed 818 age data points, which were summarized to obtain a single value for each one of the 194 nodes with age data.
-From the 194 summary ages available, 21 were not used as calibrations (asterisk, *), because they were older than a parent node or younger than a descendant node; the remaining 173 summary ages were used as secondary calibrations (forward slash, /) with the Branch Length Adjuster (BLADJ) software from Webb et al., (\protect\hyperlink{ref-Webb2008}{2008}).
-The full figure is shown in Supplementary Figure S3.
+From the 194 summary ages available, 21 were discarded and not used as calibrations (asterisk, *), because they were older than a parent node or younger than a descendant node; the remaining 173 summary ages were used as secondary calibrations (forward slash, /) with the Branch Length Adjuster (BLADJ) software from Webb et al., (\protect\hyperlink{ref-Webb2008}{2008}).
 
@@
 3 and Supplementary Figure S3.
-Each plot compares node age estimates from an original study chronogram (x axis) with the corresponding node age obtained with a DateLife analysis (y axis).
+Each plot compares original node age estimates from an input source study chronogram (x axis) with the corresponding node age resulting from a DateLife analysis excluding data from that study (y axis).

### 2023-11-11 submission/review2.Rmd
Edits from second review.

--- 2023-07-14 review2.Rmd
@@
     keep_md: yes
-date: "2023-07-20"
+date: "2023-11-07"
 ---
@@
 The AE has several other suggestions that you should find useful while preparing the revision, mainly concerning editorial and formatting issues.
-I agree that some of your figures are difficult to visualize, for example Figs.
-1 and 4, where you could use less color and more simple lines (they resemble a poster).
+
+I agree that some of your figures are difficult to visualize, for example Figs. 1 and 4, where you could use less color and more simple lines (they resemble a poster).
+
+**Thank you for your decision! We modified Figures 1 and 4, and consolidated into a single one. They are more streamlined now and usage of color was reduced; we followed all suggestions from AE.**
+
 Figure 5 is too large and detailed: it could never be published as an A4.
+
+**We reduced the number of tips shown in Fig 5 and moved the tree with all tips to supplementary data. Fig 5 is now Fig 3**
+
 I also wonder if all figures are necessary: for example, Figures 5 and 6 could be moved to the Supplementary; there is no limit in the number of figures in Software papers but we typically publish 4 figures per paper.
@@
 
-**Thank you for your decision! We modified Figures 1 and 4.**
+**We have now 4 figures in our paper.
+We achieved this by consolidating fig 1 and 3 into one single figure (now fig 2) and we moved figures 4 and 5 to supplementary material.
+We would like to request keeping Figure 6, which is now Figure 4 in the new submission, as we believe it adds good visual information for the cross validation analysis.**
 
@@
 
-**Noted. We highlighted changes using latex and submitted the PDF file.**
+**Noted. We highlighted changes using latex and submitted the PDF file with track changes names manuscript-differences.pdf.**
 
@@
 
-**We added the DOI of our Dryad package containing all Supplementary Material on L690 (manuscript-differences.pdf).**
+**We added the DOI of our Dryad package containing all Supplementary Material at the en dof the manuscript (https://datadryad.org/stash/dataset/doi:10.5061/dryad.cnp5hqc6w).**
 
@@
 
-**Done on L700 (manuscript-differences.pdf).**
+**Added in "Funding" section, after "Supplementary Material".**
 
@@
 
-**Checked.**
+**Checked and implemented.**
 
@@
 
-Thanks you for your decision! We addressed all the following comments.
+**Thanks you for your decision! We addressed all your comments as follows.**
 
@@
 
-Rephrased to "a phylogenetic software that provides services"
+**Rephrased to "a phylogenetic software that provides services"**
 
@@
 
-By default it uses OpenTree's tree topology, ig the user does not provide a topology that they want.
-We specified this as "to corresponding nodes on a tree topology, either extracted from Open Tree of Life's synthetic phylogeny or one provided by the user"
+**By default it uses OpenTree's tree topology, ig the user does not provide a topology that they want.
+We specified this as "to corresponding nodes on a tree topology, either extracted from Open Tree of Life's synthetic phylogeny or one provided by the user"**
 
@@
 
-Thanks for pointing this out, we rephrased to  "Because the results of any […], we highlight that […]"
+**Thanks for pointing this out, we rephrased to  "Because the results of any […], we highlight that […]"**
 
@@
 
-Yes, we changed to "DateLife can help to increase [...]"
+**Yes, we changed to "DateLife can help to increase [...]"**
 
@@
 
+**We address this on the manuscript, section "Searching a chronogram database", second paragraph (L 126-132).
+Briefly, it depends on the taxonomic scope that the researcher is looking for.
+The chronogram database encompasses close to 100 k (99 474) different species.
+The chronograms in datelife’s database encompass almost the same amount of tips with phylogenetic information in OpenTree’s synthetic tree (112 669), in which Chordata and Embryophyta are nearly fully sampled.
+In contrast, Bacteria, Fungi, Nematoda, and Insecta, present a large gap between the estimated number of species and what has phylogenetic information in OpenTree’s synthetic tree.
+We would expect that users working with the former groups will always get results from datelife.
+If none of the user’s species are found, datelife will return a message saying that there are no chronograms containing any of those species in the database.
+It will also encourage users to submit chronograms to a public database, so that their taxon of interest can be included in future searches.**
+
 > Can a user choose to include a chronogram of their own? I think this would be useful if you want to plug a clade you are working on into a broader taxonomic scope.
+
+**Yes! They can add any chronograms they want to the database, or by providing an initial chronogram, node ages and tip names will be used as input for the analysis.**
 
@@
 
-Thanks for pointing it out! We changed it to:
+**Thanks for pointing it out! We changed it to:**
 
-"BLADJ is an agnostic way to assign ages to nodes with no available genetic or branch length data, as it does not require initial diversification rates to draw from a branch length distribution"
+**"BLADJ is an agnostic way to assign ages to nodes with no available genetic or branch length data, as it does not require initial diversification rates to draw from a branch length distribution"**
 
-An alternative to BLADJ to date tree topologies in the absence of initial branch lengths that is common practice in the literature is to use a birth-death model to draw branch lengths [@Jetz2012; @rabosky2018inverse; @smith2018constructing].
+**An alternative to BLADJ to date tree topologies in the absence of initial branch lengths that is common practice in the literature is to use a birth-death model to draw branch lengths [@Jetz2012; @rabosky2018inverse; @smith2018constructing].
 In addition to the initial tree topology and nodes with age data, these methods require initial values of speciation and extinction rate parameters provided by the user.
 DateLife implements this approach with MrBayes [@Huelsenbeck2001; @Ronquist2003], using nodes with published age data as calibration priors on nodes of a tree topology with no branch lengths, a simple birth-death model with speciation and extinction rate parameters that are provided by the user, and no genetic data.
-However, BLADJ is the default option in DateLife, as it does not require any information on diversification rates for the phylogenetic sample to draw from a branch length distribution.
+However, BLADJ is the default option in DateLife, as it does not require any information on diversification rates for the phylogenetic sample to draw from a branch length distribution.**
 
@@
 
+**In this instance we are referring to branch lengths from phylogenetic trees that have not been dated yet. We modified the phrasing to make it more clear.**
+
 > Line 218: orthology hypothesis?
 
+**Yes! we clarified that, too.**
+
 > Lines 460-462: I don’t have strong opinion on this, but the statement seems unnecessary.
+
+**We removed it.**
 
@@
 I recommend reducing the amount of text and substantially simplifying the graphics in general.
+
+**We simplified all figures and reduced the number of final figures in the main manuscript.**
 
@@
 
+**We consolidated Fig 1 and 3 and streamlined the figure by removing text in the figure and adding it to the figure caption.
+We removed the tables ad only reference them.
+We hope this new version of the figure is considerably clearer now.**
+
 > Arguably none of the text in Figs.
@@
 
+**We reduced the number of tips shown in Figure 5 and sent the full figure (as well as Figure 4) to supplementary material. Figure 5 is now Figure 3.**
+
 > I hope this helps and look forward to seeing your revised version of the paper.
+
+**Thanks so much!**
 
@@
 > Daniele
+
+**We greatly appreciate your consideration of our resubmission.**
+
+**Luna Sanchez-Reyes on behalf of all coauthors.**

### 2023-11-15 manuscript/manuscript.Rmd
Updates

@@
 
-DateLife's goal is to improve availability and accessibility of state-of-the-art data on evolutionary time frame of organisms, to allow users from all areas of science and with all levels of expertise to compare, reuse, and reanalyse expert age data for their own applications.
-As such, it is designed as an open service that does not require any expert biological knowledge --besides the scientific names of the species or group users want to work with, to use any of its functionalities.
+DateLife's goal is to improve availability, accessibility, and reusability of state-of-the-art data on evolutionary time frame of organisms, to allow users from all areas of science and with all levels of expertise to compare, use and reanalyse expert age data for their own applications.
+As such, it is designed as an open service that does not require any expert biological knowledge --besides the scientific names of the species or group that users want to work with-- to use any of its functionalities.
 
@@
 However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined manually by a human curator.
-We would like to emphasize on the importance of sharing chronogram data, including systematically curated metadata, into open repositories, such as OpenTree's Phylesystem [@mctavish2015phylesystem] for the benefit of the scientific community as a whole.
+We would like to emphasize on the importance of sharing chronogram data, including systematically curated metadata, into open repositories, such as OpenTree's Phylesystem [@mctavish2015phylesystem] for the benefit of research and the scientific community as a whole.
+Another important source of expert data on time of lineage divergence is TimeTree's database [@Hedges2006], which holds chronograms from more than 4k published studies, and is fully browsable using its graphical user interface (timetree.org).
+TimeTree's chronogram database was not accessible in computer readable format until very recently [@kumar2022timetree], when its terms of use and website application were updated, now allowing some kinds of reuse, but not redistribution.
+<!-- Programmatic access and redistribution is still prohibited by copyright, which prevents adding chronograms to any other databases.--> The inaccessibility of TimeTree's database was an inspiration for the DateLife project, which was born as a prototype tool initially developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA [@Stoltzfus2013], as the need to make scientific information that is funded by the public practically available to the public was acknowledged and prioritized.
+
+<!--
+We believe that separating science from a capitalist view is important.
+
+However the source chronograms used as input data for this tree are not available in computer readable format, preventing reuse and reanalysis.
+As this tree is part of datelife's database, the amount of lineages that can be queried using `datelife` (`r length(unique(unname(unlist(lapply(opentree_chronograms$trees, "[[", "tip.label")))))` unique terminal taxa) is substantial.
+Access to the input chronograms used to generate the Hedges et al.
+summary tree would improve measures of uncertainty in DateLife, but they are available only as image files and not as usable data (timetree.org).
+-->
 
@@
 Alternative source chronograms available for the same taxa have potentially been estimated implementing different types of calibrations, which affects the resulting node age estimates.
-For example, in the DateLife analysis of the Fringillidae shown above, the chronograms from one study [@burns2014phylogenetics] were inferred using molecular substitution rate estimates across birds [@weir2008calibrating], and have much older age estimates for the same nodes than chronograms that were inferred using fossil calibrations (Fig.
-4; Supplementary Figs.
+For example, in the DateLife analysis of the Fringillidae shown above, the chronograms from one study [@burns2014phylogenetics] were inferred using molecular substitution rate estimates across birds [@weir2008calibrating], and have much older age estimates for the same nodes than chronograms that were inferred using fossil calibrations (Figs.
+3, 4c, Supplementary Figs.
 S4c, S10).
@@
 
-The exercise of summarizing age data from across multiple studies provides the opportunity to work with a chronogram that reflects a unified evolutionary history for a lineage, by putting together evidence from different hypotheses.
-The largest, and taxonomically broadest chronogram currently available from OpenTree was constructed summarizing age data from 2,274 published chronograms using NCBI's taxonomic tree as backbone [@Hedges2015].
-<!--BCO: I just checked: TIMETREEE NOW ALLOWS SOURCE TREE DOWNLOADS!
-This is great news.
-They have a copyright (not legal in US on trees, IMHO) notice on individual trees, but they do say "To download all individual study timetrees, please send an email request to info@timetree.org." -- I think this could be great to do-->
-A summarizing exercise may also amplify the effect of uncertainty and errors in source data, and blur parts of the evolutionary history of a lineage that might only be reflected in source chronograms and lost on the summary chronogram [@sauquet2021age].
+The exercise of summarizing age data from across multiple studies is a common resource in research, as it provides the opportunity to work with a chronogram that reflects a unified evolutionary history for a lineage, by putting together evidence from different hypotheses.
+For example, the largest, and taxonomically broadest chronogram currently available from OpenTree was constructed summarizing age data from 2,274 published chronograms using NCBI's taxonomic tree as backbone [@Hedges2015], which has been widely reused for research.
+Finally, we note that summarizing chronograms should be done with caution, as it may amplify the effect of uncertainty and errors in source data, and blur parts of the evolutionary history of a lineage that might only be reflected in source chronograms and lost on the summary chronogram [@sauquet2021age].
 <!--\textcolor{red}{EJM: ??? it's not clear to me what this last part means! What might you be losing?
@@
 <!-- Still, even if all source chronograms have been generated by excellent standards and using similar methods, the evolutionary history they depict might be very different.  -->
+<!--
+
+BCO: I just checked: TIMETREEE NOW ALLOWS SOURCE TREE DOWNLOADS!
+This is great news.
+They have a copyright (not legal in US on trees, IMHO) notice on individual trees, but they do say "To download all individual study timetrees, please send an email request to info@timetree.org." -- I think this could be great to do-->
 
@@
 Completing a phylogeny using a stochastic birth-death polytomy resolver and a backbone taxonomy is a common practice in scientific publications: @Jetz2012, created a chronogram of all 9,993 bird species, where 67% had molecular data and the rest was simulated; @rabosky2018inverse created a chronogram of 31,536 ray-finned fishes, of which only 37% had molecular data; @smith2018constructing constructed a chronogram of 353,185 seed plants where only 23% had molecular data.
-These stochastically resolved chronograms return diversification rates estimates that appear less biased than those estimated from their incompletely sampled counterparts, even with methods that account for missing lineages by using sampling fractions [@cusimano2012new; @chang2020estimating], but can also introduce spurious patterns of early bursts of diversification [@cusimano2010slowdowns; @sun2020estimating].
+Stochastically resolved chronograms can return diversification rates estimates that appear less biased than those estimated from their incompletely sampled counterparts, even with methods that account for missing lineages by using sampling fractions [@cusimano2012new; @chang2020estimating], but can also introduce spurious patterns of early bursts of diversification [@cusimano2010slowdowns; @sun2020estimating].
 
@@
 We thank Isabel Sanmartín, Daniele Silvestro, Rutger Vos and an anonymous reviewer, for comments that greatly improved this manuscript.
-The DateLife project was born as a prototype tool aiming to provide the services describe in this paper, and was initially developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA [@Stoltzfus2013].
 We thank colleagues from the O'Meara Lab at the University
 of Tennessee Knoxville for suggestions, discussions and software testing.
-The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
-that led to initial work on this project.
+The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons that led to initial work on this project.
 The team that assembled DateLife's first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.

### 2023-11-15 manuscript/manuscript.tex
Updates

@@
 
-Word count: 6946
+Word count: 7042
 
@@
 At the time of writing of this manuscript
-(Nov 07, 2023),
+(Nov 14, 2023),
 DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies, encompassing close to 100 k (99 474) different species.
@@
 Relative branch lengths can provide key information for phylogenetic dating, specifically for nodes without any calibration data available.
-While using initial branch length data is the golden standard for phylogenetic dating analyses, estimating trees with branch lengths proportional to substitution rates per site requires obtaining primary data, assembling and curating a homology (orthology) hypothesis, and choosing and implementing a method for phylogenetic inference.
+While using initial branch length data is the golden standard for phylogenetic dating analyses, obtaining such information from scratch is not an easy task: it requires obtaining primary data, assembling and curating a homology (orthology) hypothesis, and choosing and implementing a method for phylogenetic inference.
 DateLife implements a workflow to streamline this process by applying open data from the Barcode of Life Data System, BOLD (Ratnasingham \& Hebert, 2007) to obtain genetic markers for input taxa.
@@
 However, Dryad's metadata system has no information to automatically detect branch length units, and those would still need to be determined manually by a human curator.
-We would like to emphasize on the importance of sharing chronogram data, including systematically curated metadata, into open repositories, such as OpenTree's Phylesystem (McTavish et al., 2015) for the benefit of the scientific community as a whole.
+We would like to emphasize on the importance of sharing chronogram data, including systematically curated metadata, into open repositories, such as OpenTree's Phylesystem (McTavish et al., 2015) for the benefit of research and the scientific community as a whole.
+Another important source of expert data on time of lineage divergence is TimeTree's database (Hedges, Dudley, \& Kumar, 2006), which holds chronograms from more than 4k published studies, and is fully browsable using its graphical user interface (timetree.org).
+TimeTree's chronogram database was not accessible in computer readable format until very recently (Kumar et al., 2022), when its terms of use and website application were updated, now allowing some kinds of reuse, but not redistribution.\\
+ The inaccessibility of TimeTree's database was an inspiration for the DateLife project, which was born as a prototype tool initially developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA (Stoltzfus et al., 2013), as the need to make scientific information that is funded by the public practically available to the public was acknowledged and prioritized.
 
@@
 Alternative source chronograms available for the same taxa have potentially been estimated implementing different types of calibrations, which affects the resulting node age estimates.
-For example, in the DateLife analysis of the Fringillidae shown above, the chronograms from one study (Burns et al., 2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older age estimates for the same nodes than chronograms that were inferred using fossil calibrations (Fig.
-4; Supplementary Figs.
+For example, in the DateLife analysis of the Fringillidae shown above, the chronograms from one study (Burns et al., 2014) were inferred using molecular substitution rate estimates across birds (Weir \& Schluter, 2008), and have much older age estimates for the same nodes than chronograms that were inferred using fossil calibrations (Figs.
+3, 4c, Supplementary Figs.
 S4c, S10).
@@
 
-Further research is needed to fully understand the effects of using secondary calibrations and the use of resulting chronograms in downstream analyses (Hipsley \& Müller, 2014; Powell et al., 2020; Schenk, 2016; Shaul \& Graur, 2002).
-
 \begin{center}
@@
 
-The exercise of summarizing age data from across multiple studies provides the opportunity to work with a chronogram that reflects a unified evolutionary history for a lineage, by putting together evidence from different hypotheses.
-The largest, and taxonomically broadest chronogram currently available from OpenTree was constructed summarizing age data from 2,274 published chronograms using NCBI's taxonomic tree as backbone (Hedges et al., 2015).
-A summarizing exercise may also amplify the effect of uncertainty and errors in source data, and blur parts of the evolutionary history of a lineage that might only be reflected in source chronograms and lost on the summary chronogram (Sauquet et al., 2021).
+The exercise of summarizing age data from across multiple studies is a common resource in research, as it provides the opportunity to work with a chronogram that reflects a unified evolutionary history for a lineage, by putting together evidence from different hypotheses.
+For example, the largest, and taxonomically broadest chronogram currently available from OpenTree was constructed summarizing age data from 2,274 published chronograms using NCBI's taxonomic tree as backbone (Hedges et al., 2015), which has been widely reused for research.
+Finally, we note that summarizing chronograms should be done with caution, as it may amplify the effect of uncertainty and errors in source data, and blur parts of the evolutionary history of a lineage that might only be reflected in source chronograms and lost on the summary chronogram (Sauquet et al., 2021).
 
@@
 (2018) created a chronogram of 31,536 ray-finned fishes, of which only 37\% had molecular data; Smith and Brown (2018) constructed a chronogram of 353,185 seed plants where only 23\% had molecular data.
-These stochastically resolved chronograms return diversification rates estimates that appear less biased than those estimated from their incompletely sampled counterparts, even with methods that account for missing lineages by using sampling fractions (Chang et al., 2020; Cusimano et al., 2012), but can also introduce spurious patterns of early bursts of diversification (Cusimano \& Renner, 2010; Sun et al., 2020).
+Stochastically resolved chronograms can return diversification rates estimates that appear less biased than those estimated from their incompletely sampled counterparts, even with methods that account for missing lineages by using sampling fractions (Chang et al., 2020; Cusimano et al., 2012), but can also introduce spurious patterns of early bursts of diversification (Cusimano \& Renner, 2010; Sun et al., 2020).
 
@@
 It is available as an R package, and as a web-based R shiny application at www.datelife.org.
-DateLife provides a straightforward way to get an informed picture of the state of knowledge of the time frame of evolution of different regions of the tree of life, and allows identifying regions that require more research, or that have conflicting information.
+DateLife provides a straightforward way to get an informed picture of the state of knowledge for the time frame of evolution of different regions of the tree of life, and allows identifying regions that require more research, or that have conflicting information.
 Additionally, both summary and newly generated trees using the DateLife workflow are useful to evaluate evolutionary hypotheses in different areas of research.
@@
 We thank Isabel Sanmartín, Daniele Silvestro, Rutger Vos and an anonymous reviewer, for comments that greatly improved this manuscript.
-The DateLife project was born as a prototype tool aiming to provide the services describe in this paper, and was initially developed over a series of hackathons at the National Evolutionary Synthesis Center, NC, USA (Stoltzfus et al., 2013).
 We thank colleagues from the O'Meara Lab at the University
 of Tennessee Knoxville for suggestions, discussions and software testing.
-The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons
-that led to initial work on this project.
+The late National Evolutionary Synthesis Center (NESCent), which sponsored hackathons that led to initial work on this project.
 The team that assembled DateLife's first proof of concept: Tracy Heath, Jonathan Eastman, Peter Midford, Joseph Brown, Matt Pennell, Mike Alfaro, and Luke Harmon.
@@
 
+\leavevmode\hypertarget{ref-Hedges2006}{}%
+Hedges, S.
+B., Dudley, J., \& Kumar, S.
+(2006).
+TimeTree: A public knowledge-base of divergence times among organisms.
+\emph{Bioinformatics}, \emph{22}(23), 2971--2972.
+\url{https://doi.org/10.1093/bioinformatics/btl505}
+
 \leavevmode\hypertarget{ref-Hedges2015}{}%
@@
 \emph{Systematic Biology}, \emph{64}(5), 853--859.
+
+\leavevmode\hypertarget{ref-kumar2022timetree}{}%
+Kumar, S., Suleski, M., Craig, J.
+M., Kasprowicz, A.
+E., Sanderford, M., Li, M., \ldots{} Hedges, S.
+B.
+(2022).
+TimeTree 5: An expanded resource for species divergence times.
+\emph{Molecular Biology and Evolution}, \emph{39}(8), msac174.
 

### 2023-11-15 manuscript/manuscript.Rmd
resubmission

@@
 All `r length(dres)` source chronograms (Supplementary Figs.
-S5-S23) were congruified to OpenTree's topology shown in Supplementary Figure S2, reducing the original 1,206 node age data set to 818 different data points (Supplementary Table S2) that could be used as calibrations for the chosen topology (Supplementary Fig.
-S2).
+S5-S23) were congruified to OpenTree's topology shown in Supplementary Figure S2, reducing the original 1,206 node age data set to 818 different data points (Supplementary Table S2) that could be used as calibrations for that chosen topology.
 The congruent node age data points were summarized for each node, resulting in 194 summary node ages.
@@
 Finally, uncertainty and variability of chronogram node age estimates might pose larger issues in some research areas than others.
-For example, in ecological and conservation biology studies, it has been shown that incorporating some chronogram data provides better results than when not using any age data at all, even if the node ages are not good quality [@Webb2008].
+For example, in ecological and conservation biology studies, it has been shown that incorporating some chronogram data provides better results than when not using any age data at all, even if the node ages are not the best quality [@Webb2008].
 In the following sections we discuss the particularities of divergence times from DateLife's summary chronograms and their impact on certain evolutionary analyses, for consideration of the readers and users in different research areas.
@@
 
-Supplementary Figures can be viewed and downloaded from their Zenodo repository (<https://doi.org/10.5281/zenodo.6683667>).
-Supplementary material, including code, biological examples, benchmark results, data files and online-only appendices, can be downloaded from the Dryad data repository (<https://doi.org/10.5061/dryad.cnp5hqc6w>), as well as in the Zenodo stable repositories that host the reproducible manuscript (<https://doi.org/10.5281/zenodo.7435094>), the biological examples (<https://doi.org/10.5281/zenodo.7435101>), and the software benchmark (<https://doi.org/10.5281/zenodo.7435106>).
-Development versions corresponding to all of the above are hosted on GitHub, accesible at <https://github.com/LunaSare/datelifeMS1>, <https://github.com/LunaSare/datelife_examples>, and <https://github.com/LunaSare/datelife_benchmark>.
+Supplementary material, including figures, tables, code, biological examples, benchmark results, data files and online-only appendices, can be viewed and downloaded from the Dryad data repository (<https://doi.org/10.5061/dryad.cnp5hqc6w>), as well as from the Zenodo stable repositories that host the reproducible manuscript (<https://doi.org/10.5281/zenodo.7435094>), the biological examples (<https://doi.org/10.5281/zenodo.7435101>), the software benchmark (<https://doi.org/10.5281/zenodo.7435106>), and the figures (<https://doi.org/10.5281/zenodo.6683667>).
+Dryad's data publication fee is covered by the Society of Systematic Biologists.
+Development versions corresponding to all of the above are hosted on GitHub, accessible at <https://github.com/LunaSare/datelifeMS1>, <https://github.com/LunaSare/datelife_examples>, and <https://github.com/LunaSare/datelife_benchmark>.
 
@@
 
+# Figure Captions
+
+Figure 1.
+DateLife's benchmarking results showing computation time used for taxon name processing and search across \texttt{datelife}'s chronogram database, as a function of number of input taxon names (N).
+For each N = \{10, 100, 200, ..., 1 000, ...
+, 9 000, 10 000\}, we randomly sampled N species names from the class Aves, a hundred times, and then performed a \texttt{datelife} search processing the input names using the Taxon Names Resolution Service (TNRS; light gray), and without processing input names (dark gray).
+For comparison, we performed a chronogram search using names that have been pre-processed with TNRS.
+
+Figure 2.
+DateLife results of an analysis of a small sample of 6 bird species within the Passeriformes (a, b).
+Processed species names were found across 9 chronograms within 6 independent studies (c; Barker et al.
+(\protect\hyperlink{ref-barker2012going}{2012}), Barker et al.
+(\protect\hyperlink{ref-barker2015new}{2015}), Burns et al.
+(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al.
+(\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al.
+(\protect\hyperlink{ref-Jetz2012}{2012}).
+This revealed 28 source age data points for the queried species names (e; Table 1).
+Summarized age data (f; Table 2) was used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree v13.4 (d), resulting in the chronogram of summary source ages shown in (g).
+
+Figure 3.
+Median summary chronogram resulting from a Datelife analysis of bird species within the family Fringillidae.
+For visualization purposes, we are showing a portion of the final median summary chronogram encompassing 57 species out of the 289 total included in the analysis.
+The complete final chronogram is available as Supplementary Figure S3.
+The starting tree topology (Supplementary Fig.
+S2) has 289 tip species and 253 nodes; DateLife revealed age data for 194 of these nodes from at least one published chronogram.
+In total, 19 different chronograms from 13 different studies contributed 818 age data points, which were summarized to obtain a single value for each one of the 194 nodes with age data.
+From the 194 summary ages available, 21 were discarded and not used as calibrations (asterisk, *), because they were older than a parent node or younger than a descendant node; the remaining 173 summary ages were used as secondary calibrations (forward slash, /) with the Branch Length Adjuster (BLADJ) software from Webb et al., (\protect\hyperlink{ref-Webb2008}{2008}).
+
+Figure 4.
+Cross validation of results from a DateLife analysis of the family Fringillidae, shown in Figure 3 and Supplementary Figure S3.
+Each plot compares the original node age estimates from an input source study chronogram (x axis) with the corresponding node age resulting from a dating analysis using the DateLife workflow, excluding data from that study (y axis).
+
 ```{r fig1, child= if (figures_at_end) c("table1.Rmd", "table2.Rmd", "fig-workflow.Rmd", "fig-benchmark.Rmd", "fig-small-example.Rmd", "fig-topologies.Rmd", "fig-fringillidae-full-example.Rmd", "fig-cross-validation.Rmd")}

### 2023-11-15 submission/captions-final.tex
resubmission

@@
 
-Ages of nodes extracted from 9 chronograms containing at least 2 of the 6 bird species used in the Fringillidae small example of the DateLife workflow.
-All nodes are congruent with the chosen tree topology.
-See Figure 3, step c2.
+Node ages extracted with DateLife from 9 chronograms (shown in Fig.
+2c) containing at least 2 of the 6 bird species analysed in the Fringillidae small example.
+All nodes are congruent with the chosen tree topology shown in Figure 2d.
 
@@
 
-Summary statistics of node ages from Table 1. Median ages are used by default to date a tree topology using DateLife, see Figure 3, step c3.
+Summary statistics of node ages from Table 1. DateLife uses median ages by default to date a chosen tree topology (Fig. 3g).
 

### 2023-11-15 submission/manuscript.tex
resubmission

--- 2023-11-15 manuscript/manuscript.tex
@@
 
-Word count: 7042
+Word count: 7484
 
@@
 At the time of writing of this manuscript
-(Nov 14, 2023),
+(Nov 21, 2023),
 DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies, encompassing close to 100 k (99 474) different species.
@@
 All 19 source chronograms (Supplementary Figs.
-S5-S23) were congruified to OpenTree's topology shown in Supplementary Figure S2, reducing the original 1,206 node age data set to 818 different data points (Supplementary Table S2) that could be used as calibrations for the chosen topology (Supplementary Fig.
-S2).
+S5-S23) were congruified to OpenTree's topology shown in Supplementary Figure S2, reducing the original 1,206 node age data set to 818 different data points (Supplementary Table S2) that could be used as calibrations for that chosen topology.
 The congruent node age data points were summarized for each node, resulting in 194 summary node ages.
@@
 Finally, uncertainty and variability of chronogram node age estimates might pose larger issues in some research areas than others.
-For example, in ecological and conservation biology studies, it has been shown that incorporating some chronogram data provides better results than when not using any age data at all, even if the node ages are not good quality (Webb et al., 2008).
+For example, in ecological and conservation biology studies, it has been shown that incorporating some chronogram data provides better results than when not using any age data at all, even if the node ages are not the best quality (Webb et al., 2008).
 In the following sections we discuss the particularities of divergence times from DateLife's summary chronograms and their impact on certain evolutionary analyses, for consideration of the readers and users in different research areas.
@@
 
-Supplementary Figures can be viewed and downloaded from their Zenodo repository (\url{https://doi.org/10.5281/zenodo.6683667}).
-Supplementary material, including code, biological examples, benchmark results, data files and online-only appendices, can be downloaded from the Dryad data repository (\url{https://doi.org/10.5061/dryad.cnp5hqc6w}), as well as in the Zenodo stable repositories that host the reproducible manuscript (\url{https://doi.org/10.5281/zenodo.7435094}), the biological examples (\url{https://doi.org/10.5281/zenodo.7435101}), and the software benchmark (\url{https://doi.org/10.5281/zenodo.7435106}).
-Development versions corresponding to all of the above are hosted on GitHub, accesible at \url{https://github.com/LunaSare/datelifeMS1}, \url{https://github.com/LunaSare/datelife_examples}, and \url{https://github.com/LunaSare/datelife_benchmark}.
+Supplementary material, including figures, tables, code, biological examples, benchmark results, data files and online-only appendices, can be viewed and downloaded from the Dryad data repository (\url{https://doi.org/10.5061/dryad.cnp5hqc6w}), as well as from the Zenodo stable repositories that host the reproducible manuscript (\url{https://doi.org/10.5281/zenodo.7435094}), the biological examples (\url{https://doi.org/10.5281/zenodo.7435101}), the software benchmark (\url{https://doi.org/10.5281/zenodo.7435106}), and the figures (\url{https://doi.org/10.5281/zenodo.6683667}).
+Dryad's data publication fee is covered by the Society of Systematic Biologists.
+Development versions corresponding to all of the above are hosted on GitHub, accessible at \url{https://github.com/LunaSare/datelifeMS1}, \url{https://github.com/LunaSare/datelife_examples}, and \url{https://github.com/LunaSare/datelife_benchmark}.
 
@@
 
+\hypertarget{figure-captions}{%
+\section{Figure Captions}\label{figure-captions}}
+
+Figure 1.
+DateLife's benchmarking results showing computation time used for taxon name processing and search across \texttt{datelife}'s chronogram database, as a function of number of input taxon names (N).
+For each N = \{10, 100, 200, \ldots{}, 1 000, \ldots{} , 9 000, 10 000\}, we randomly sampled N species names from the class Aves, a hundred times, and then performed a \texttt{datelife} search processing the input names using the Taxon Names Resolution Service (TNRS; light gray), and without processing input names (dark gray).
+For comparison, we performed a chronogram search using names that have been pre-processed with TNRS.
+
+Figure 2.
+DateLife results of an analysis of a small sample of 6 bird species within the Passeriformes (a, b).
+Processed species names were found across 9 chronograms within 6 independent studies (c; Barker et al.
+(\protect\hyperlink{ref-barker2012going}{2012}), Barker et al.
+(\protect\hyperlink{ref-barker2015new}{2015}), Burns et al.
+(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al.
+(\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al.
+(\protect\hyperlink{ref-Jetz2012}{2012}).
+This revealed 28 source age data points for the queried species names (e; Table 1).
+Summarized age data (f; Table 2) was used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree v13.4 (d), resulting in the chronogram of summary source ages shown in (g).
+
+Figure 3.
+Median summary chronogram resulting from a Datelife analysis of bird species within the family Fringillidae.
+For visualization purposes, we are showing a portion of the final median summary chronogram encompassing 57 species out of the 289 total included in the analysis.
+The complete final chronogram is available as Supplementary Figure S3.
+The starting tree topology (Supplementary Fig.
+S2) has 289 tip species and 253 nodes; DateLife revealed age data for 194 of these nodes from at least one published chronogram.
+In total, 19 different chronograms from 13 different studies contributed 818 age data points, which were summarized to obtain a single value for each one of the 194 nodes with age data.
+From the 194 summary ages available, 21 were discarded and not used as calibrations (asterisk, *), because they were older than a parent node or younger than a descendant node; the remaining 173 summary ages were used as secondary calibrations (forward slash, /) with the Branch Length Adjuster (BLADJ) software from Webb et al., (\protect\hyperlink{ref-Webb2008}{2008}).
+
+Figure 4.
+Cross validation of results from a DateLife analysis of the family Fringillidae, shown in Figure 3 and Supplementary Figure S3.
+Each plot compares the original node age estimates from an input source study chronogram (x axis) with the corresponding node age resulting from a dating analysis using the DateLife workflow, excluding data from that study (y axis).
+
 \end{document}

### 2023-11-15 tables/README-tables.Rmd
resubmission

@@
 
+Doc files developed in files:
+- table-fringillidae-small-example.Rmd
+- table-fringillidae-small-example-summary.Rmd
+
+***
 tex file of small examples table 1 developed in vignettes/figure-small-example.Rmd L289-295.

### 2023-11-15 tables/table-fringillidae-small-example-summary.Rmd
tables as doc

---
title: ""
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
table2 <- read.csv(file = "table-fringillidae-small-example-summary.csv")
head(table2)
colnames_edited <- gsub("\\.", " ", colnames(table2))
colnames_edited
```

```{r, echo =FALSE}
knitr::kable(table2, 
             digits = 3,
             row.names = FALSE,
             col.names = colnames_edited,
             align = "l",
            caption = "Table 2. Summary statistics of node ages from Table 1. DateLife uses median ages by default to date a chosen tree topology (Fig. 3g).")
```

### 2023-11-15 tables/table-fringillidae-small-example.Rmd
tables as doc

---
title: ""
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
table10 <- read.csv(file = "table-fringillidae-small-example.csv")
head(table10)
colnames_edited <- gsub("\\.", " ", colnames(table10))
colnames_edited 
colnames_edited[5] <- "Study reference"
```

```{r, echo =FALSE}
knitr::kable(table10, 
             digits = 3,
             row.names = FALSE,
             col.names = colnames_edited,
             align = "l",
            caption = "Table 1. Node ages extracted with DateLife from 9 chronograms (shown in Fig. 2c) containing at least 2 of the 6 bird species analysed in the Fringillidae small example. All nodes are congruent with the chosen tree topology shown in Figure 2d.")
```

### 2023-12-15 data-raw/felis-chronogram-plot.Rmd
workflow to plot calibrations for felis

---
title: "felis plot"
output: pdf_document
date: "2023-12-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
getwd()
source(file = "R/plot_node_ages_v2.3.R")
library(ape)
library(datelife)
```

```{r}
felis_node_subtree <- rotl::tol_subtree(node_id = 'ott563165', label = "name")
my_datelife_query <- datelife::make_datelife_query(input = felis_node_subtree)
ls(my_datelife_query)
my_datelife_result <-  datelife::get_datelife_result(input = my_datelife_query, cache = opentree_chronograms)
```

```{r}
my_datelife_result$`Roquet, C., Lavergne, S., & Thuiller, W. (2014). One Tree to Link Them All: A Phylogenetic Dataset for the European Tetrapoda. PLoS Currents. doi:10.1371/currents.tol.5102670fff8aa5c918e78f5592790e48
`
```

```{r}
phylo_median <- datelife::datelife_search(input = my_datelife_query,
                                           summary_format = "phylo_median")
#dsumm <- datelife:::summary.datelifeResult(datelife_query = my_datelife_query, object = my_datelife_result)

phylo_all <- datelife::datelife_search(input = my_datelife_query,
                                           summary_format = "phylo_all")
class(phylo_all)
ls(phylo_all)
length(phylo_all)
```

### Getting oldest age among all source chronograms in a `multiPhylo` object:

```{r}
lapply(phylo_all, ape::branching.times)

plot_max_time <- 846
```

```{r}
all_calibrations <- datelife:::extract_calibrations_phylo(
  input = phylo_all,
  each = FALSE)
nrow(all_calibrations)
```
Congruify calibrations:

```{r}
congruified_calibs <- datelife::congruify_and_mrca_multiPhylo(phy = phylo_median,
                              source_chronograms = phylo_all[-9])

```

```{r}

calibrations <-  datelife:::summary.matchedCalibrations(phylo_median)
ls(calibrations)
calibrations$in_phy
congruified_calibs

calibrations <- list(in_phy = congruified_calibs)
plot_node_ages_2.3(phylo_median, calibrations)
```

```{r}
datelifeplot::plot_node_ages2(chronogram = phylo_median, matched_ages = calibrations)
```

### 2024-02-15 data-raw/felis-chronogram-plot.Rmd
Edits from final decision

@@
 getwd()
-source(file = "R/plot_node_ages_v2.3.R")
+source(file = "../R/plot_node_ages_v2.3.R")
 library(ape)
+# detach("package:datelife", unload=TRUE)
+devtools::install(pkg = "../../lunasare/datelife/.")
+
+# devtools::install_github(repo = "phylotastic/datelife", ref = "ott-names")
 library(datelife)
@@
 ```
+### Visualizing all source chronograms
+
+```{r}
+plot_max_time <- 845
+source_chronogram_file_names <- paste0("felis-source-chronogram-", 1:length(phylo_all))
+datelifeplot::plot_phylo_all(chronograms = phylo_all,
+                             folder_name = paste0("felis-source_chronograms"),
+                             file_name = source_chronogram_file_names,
+                             plot_height = 600,
+                             plot_width = 600,
+                             mai1 = 1,
+                             mai2 = 0,
+                             mai3 = 1,
+                             mai4 = 5,
+                             omi1 = 1,
+                             omi2 = 0,
+                             omi3 = 0,
+                             omi4 = 0,
+                             write = "pdf",
+                             plot_type = "phyloch",
+                             max_depth = plot_max_time,
+                             edge.width = 2,
+                             cex_tiplabels = 1,
+                             cex_axislabel = 1.5,
+                             cex_axis = 2,
+                             cex_title = 2.5,
+                             pos_title = -2.8,
+                             pos_axis = 2,
+                             axis_label = "")
+```
 
@@
 lapply(phylo_all, ape::branching.times)
-
-plot_max_time <- 846
+class(phylo_all[[9]]) <- "phylo"
+datelifeplot::plot_chronogram(chronogram = plot(phylo_all[[9]]))
 ```

### 2024-02-15 manuscript/manuscript.Rmd
Edits from final decision

@@
 (`r format(Sys.time(), "%b %d, %Y")`),
-DateLife's chronogram database latest version consist of `r format(length(opentree_chronograms$trees), big.mark=",")` chronograms published in `r length(unique(opentree_chronograms$studies))` different studies, encompassing close to 100 k (99 474) different species.
+DateLife's chronogram database latest version consist of `r format(length(opentree_chronograms$trees), big.mark=",")` chronograms published in `r length(unique(opentree_chronograms$studies))` studies, encompassing 99 474 species.
 It is curated from OpenTree's phylogenetic database, the Phylesystem, an open database of expert and peer-reviewed phylogenetic knowledge with rich metadata and a wide taxonomic scope [@mctavish2015phylesystem].

### 2024-03-09 manuscript/manuscript.Rmd
Last edits

@@
 These matching pruned chronograms are referred to as source chronograms.
-Total distance in units of million years (Myr) between taxon pairs within each source chronogram are stored as a patristic distance matrix.
+Total distance in units of million years (myr) between taxon pairs within each source chronogram are stored as a patristic distance matrix.
 The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a "phylo" object or newick string.
@@
 Figure 3.
-Median summary chronogram resulting from a Datelife analysis of bird species within the family Fringillidae.
+Median summary chronogram resulting from a DateLife analysis of bird species within the family Fringillidae.
 For visualization purposes, we are showing a portion of the final median summary chronogram encompassing 57 species out of the 289 total included in the analysis.

### 2024-03-09 manuscript/manuscript.tex
Last edits

@@
 
-% \usepackage{etoolbox}
 \makeatletter
-\patchcmd{\HyOrg@maketitle}
+\usepackage{etoolbox}
+\patchcmd{\maketitle}
   {\section{\normalfont\normalsize\abstractname}}
@@
   {}{\typeout{Failed to patch abstract.}}
-\patchcmd{\HyOrg@maketitle}
+\patchcmd{\maketitle}
   {\section{\protect\normalfont{\@title}}}
@@
 
-Word count: 7042
+Word count: 7478
 
@@
 At the time of writing of this manuscript
-(Nov 14, 2023),
-DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies, encompassing close to 100 k (99 474) different species.
+(Mar 08, 2024),
+DateLife's chronogram database latest version consist of 253 chronograms published in 187 studies, encompassing 99 474 species.
 It is curated from OpenTree's phylogenetic database, the Phylesystem, an open database of expert and peer-reviewed phylogenetic knowledge with rich metadata and a wide taxonomic scope (McTavish et al., 2015).
@@
 These matching pruned chronograms are referred to as source chronograms.
-Total distance in units of million years (Myr) between taxon pairs within each source chronogram are stored as a patristic distance matrix.
+Total distance in units of million years (myr) between taxon pairs within each source chronogram are stored as a patristic distance matrix.
 The matrix format speeds up extraction of pairwise taxon ages of any queried taxa, as opposed to searching the ancestor node of a pair of taxa in a \enquote{phylo} object or newick string.
@@
 
+\hypertarget{figure-captions}{%
+\section{Figure Captions}\label{figure-captions}}
+
+Figure 1.
+DateLife's benchmarking results showing computation time used for taxon name processing and search across \texttt{datelife}'s chronogram database, as a function of number of input taxon names (N).
+For each N = \{10, 100, 200, \ldots{}, 1 000, \ldots{} , 9 000, 10 000\}, we randomly sampled N species names from the class Aves, a hundred times, and then performed a \texttt{datelife} search processing the input names using the Taxon Names Resolution Service (TNRS; light gray), and without processing input names (dark gray).
+For comparison, we performed a chronogram search using names that have been pre-processed with TNRS.
+
+Figure 2.
+DateLife results of an analysis of a small sample of 6 bird species within the Passeriformes (a, b).
+Processed species names were found across 9 chronograms within 6 independent studies (c; Barker et al.
+(\protect\hyperlink{ref-barker2012going}{2012}), Barker et al.
+(\protect\hyperlink{ref-barker2015new}{2015}), Burns et al.
+(\protect\hyperlink{ref-burns2014phylogenetics}{2014}), Hedges et al.
+(\protect\hyperlink{ref-Hedges2015}{2015}), Hooper and Price (\protect\hyperlink{ref-hooper2017chromosomal}{2017}), Jetz et al.
+(\protect\hyperlink{ref-Jetz2012}{2012}).
+This revealed 28 source age data points for the queried species names (e; Table 1).
+Summarized age data (f; Table 2) was used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree v13.4 (d), resulting in the chronogram of summary source ages shown in (g).
+
+Figure 3.
+Median summary chronogram resulting from a DateLife analysis of bird species within the family Fringillidae.
+For visualization purposes, we are showing a portion of the final median summary chronogram encompassing 57 species out of the 289 total included in the analysis.
+The complete final chronogram is available as Supplementary Figure S3.
+The starting tree topology (Supplementary Fig.
+S2) has 289 tip species and 253 nodes; DateLife revealed age data for 194 of these nodes from at least one published chronogram.
+In total, 19 different chronograms from 13 different studies contributed 818 age data points, which were summarized to obtain a single value for each one of the 194 nodes with age data.
+From the 194 summary ages available, 21 were discarded and not used as calibrations (asterisk, *), because they were older than a parent node or younger than a descendant node; the remaining 173 summary ages were used as secondary calibrations (forward slash, /) with the Branch Length Adjuster (BLADJ) software from Webb et al., (\protect\hyperlink{ref-Webb2008}{2008}).
+
+Figure 4.
+Cross validation of results from a DateLife analysis of the family Fringillidae, shown in Figure 3 and Supplementary Figure S3.
+Each plot compares the original node age estimates from an input source study chronogram (x axis) with the corresponding node age resulting from a dating analysis using the DateLife workflow, excluding data from that study (y axis).
+
 \end{document}

### 2024-03-29 submission/manuscript.tex
3rd review edits

@@
 
-Word count: 7484
+Word count: 7530
 
@@
 At the time of writing of this manuscript
-(Nov 21, 2023),
-DateLife's chronogram database latest version consist of 253 chronograms published in 187 different studies, encompassing close to 100 k (99 474) different species.
+(Mar 08, 2024),
+DateLife's chronogram database latest version consist of 253 chronograms published in 187 studies, encompassing 99 474 species.
 It is curated from OpenTree's phylogenetic database, the Phylesystem, an open database of expert and peer-reviewed phylogenetic knowledge with rich metadata and a wide taxonomic scope (McTavish et al., 2015).
@@
 Summarized age data (f; Table 2) was used as secondary calibrations to date a tree topology obtained from OpenTree's synthetic tree v13.4 (d), resulting in the chronogram of summary source ages shown in (g).
+The Paleogene Period spans from the end of the Cretaceous Period 66 million years ago (Ma) to the beginning of the Neogene Period 23.03 Ma.
 
 Figure 3.
-Median summary chronogram resulting from a Datelife analysis of bird species within the family Fringillidae.
+Median summary chronogram resulting from a DateLife analysis of bird species within the family Fringillidae.
 For visualization purposes, we are showing a portion of the final median summary chronogram encompassing 57 species out of the 289 total included in the analysis.
@@
 From the 194 summary ages available, 21 were discarded and not used as calibrations (asterisk, *), because they were older than a parent node or younger than a descendant node; the remaining 173 summary ages were used as secondary calibrations (forward slash, /) with the Branch Length Adjuster (BLADJ) software from Webb et al., (\protect\hyperlink{ref-Webb2008}{2008}).
+The Paleogene Period spans from the end of the Cretaceous Period 66 million years ago (Ma) to the beginning of the Neogene Period 23.03 Ma.
 
